# Lecture 1: Simple linear regression {#i01}

```{r}
#| label: i01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 1, you can</span>

- explain what a simple regression model describes and what its parts mean
- tell apart the errors of the model and the residuals of a fit
- explain how least squares chooses the best fitting line
- fit a simple regression in R and interpret the slope with units
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Every founder who pitches puts a price on the company. Asking 100,000 dollars for 10 percent of the company claims that the company is worth 1 million dollars. The question is how much of that claimed value is left after the negotiation. A line fitted through the data turns the question into one number a founder can use.</p>

Among pitches that reached a deal, what is a claimed valuation worth: how
much higher is the agreed valuation for each extra dollar of valuation that
the founder claims?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=vCHtY6Me5FI&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=9">3.1
Simple Linear Regression</a>. Focus on the population line, the fitted line,
and least squares. The video writes RSS for the quantity this book calls SSE,
the sum of squared residuals. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
The videos explain key ideas in another way and can use a different
notation. The lecture material and its notation always come first.
</div>

**Warm up: five quantities from QM Basic.** QM Intermediate
builds directly on five quantities from the end of QM Basic: estimate,
standard error, t value, p-value, confidence interval. The sections for
Lectures 9 and 10 of the
[QM Basic formula sheet](../basic/formula-sheet.qmd#formula-l9) list their
formulas. With the sheet open, answer the five items below.

1. A study reports an estimate of 12 with standard error 4. What is the t
   value for testing whether the underlying quantity is zero?
2. With a large sample, what critical value goes into a 95% confidence
   interval?
3. Compute that 95% interval for the estimate in item 1.
4. The p-value is 0.003. Complete: "If the true value were zero, ..."
5. Which of the five quantities has the same units as the estimate itself?

::: {.callout-note collapse="true"}
## Check your preparation

1. $t = 12/4 = 3$.
2. About 1.96. The critical value $t_{0.975,\,df}$ is close to
   $z_{0.975}=1.96$ when the degrees of freedom are large.
3. $12 \pm 1.96 \times 4 = (4.2,\; 19.8)$.
4. "... an estimate at least this far from zero would occur in about 0.3% of
   samples."
5. The standard error and both interval endpoints. The t value and p-value
   are unit-free.
:::

**Define the variables.** The data of this course are 1,441
pitches from the television show Shark Tank, where founders ask investors
(the sharks) for money in exchange for a share of their company. This lecture
uses the 874 pitches that ended in a deal for a share of the company. It
measures all money in **millions of dollars**:

- $X_i$: the valuation the founder claimed, in \$ millions. Asking \$100,000
  for 10% claims \$1 million, so $X_i = 1$.
- $Y_i$: the valuation agreed on air (during the show), in \$ millions. A deal of \$100,000 for
  20% values the company at \$0.5 million, so $Y_i = 0.5$.

**Predict.** Answer this question before you come to class: what are the **sign** and the **unit** of
the slope in a regression of the agreed valuation on the claimed valuation,
and is the slope above or below 1? Choose one
answer before class. The lecture material gives the correct answer.

[Shark Tank data]{.shark-tag} **R warm-up.** Every Intermediate lecture has a short R warm-up:
you run a few lines of R and bring the output to class. This week, load the
data, keep the deals for a share of the company, and plot the agreed
valuation against the claimed valuation:

```{r}
#| label: i01-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
plot(vals$claimed_m, vals$agreed_m)
```

**Code along (bring your laptop).** Bring a charged laptop with RStudio and
the course project open. The code of the lecture is in this chapter, ready to
copy, so you can run each step yourself during the lecture and get the same
numbers and figures.

If you are not sure about lines, slopes, intercepts, or subscripts, review
[Mathematics Refresher: straight lines, slopes, and intercepts](../resources/math-refresher.html#straight-lines-slopes-and-intercepts)
and [subscripts and indices](../resources/math-refresher.html#subscripts-and-indices).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section fits a straight line to the valuations in the Shark Tank data.
It explains what the line is, how it is calculated, and how to read its slope
and its intercept.

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=vCHtY6Me5FI">3.1 Simple Linear Regression</a>.</p>

### Start with a founder's question

A founder who asks \$100,000 for 10% of the company claims that it is worth
\$1 million. How much of this claimed value is left after the negotiation with the
sharks? Your warm-up plot shows the answer as a cloud of points. A fitted
line turns that cloud into one number.

### Begin with the data

[Shark Tank data]{.shark-tag} The data are the Shark Tank teaching file. One row is one pitch on the
television show: 1,441 pitches from Seasons 1 to 16. This lecture uses the
874 pitches that ended in a deal for a share of the company, and two
variables. The valuation claimed by the founder (`requested_valuation_usd`)
is the amount asked divided by the share offered. The valuation agreed on air
(`deal_valuation_usd`) is the amount paid divided by the share the sharks
receive. The
[variable table in Basic Lecture 1](../basic/b01-introduction-data-analysis.qmd#b01-variable-table)
describes every variable of the file.

The code below is the warm-up code. It keeps the deals for a share of the
company and creates both valuations in \$ millions. Of the 882 deals, 8 gave
the sharks no share of the company. These 8 deals have no valuation, so we
leave them out. This leaves 874 deals.

```{r}
#| label: i01-build-deals
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sum(sharks$deal_on_show == 1)   # all pitches that ended in a deal
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
nrow(vals)                      # deals for a share of the company
```

```{r}
#| label: fig-i01-scatter
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Valuation agreed on air against the valuation claimed by the founder, for the 874 deals for a share of the company. Both axes in $ millions. The display zooms in on valuations up to $10 million (819 of 874 deals). The orange line is the fitted regression line, calculated from all 874 deals. On the grey dashed line the agreed valuation equals the claimed one."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
display_vals <- subset(vals, claimed_m <= 10 & agreed_m <= 10)

ggplot(display_vals, aes(x = claimed_m, y = agreed_m)) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  geom_abline(intercept = coef(val_model)[1],
              slope = coef(val_model)[2],
              colour = "#D55E00", linewidth = 0.9) +
  geom_abline(intercept = 0, slope = 1, colour = "#5B6976",
              linetype = "dashed", linewidth = 0.6) +
  annotate("text", x = 6.8, y = 7.5, label = "agreed = claimed",
           hjust = 1, colour = "#5B6976", size = 3.2) +
  annotate("text", x = 0.3, y = 9.85,
           label = "Fitted regression (estimated on all 874 deals)",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  coord_cartesian(xlim = c(0, 10.3), ylim = c(0, 10.3), expand = FALSE) +
  labs(x = "Valuation claimed by the founder ($ millions)",
       y = "Valuation agreed on air ($ millions)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each point in @fig-i01-scatter is one deal. The claimed valuation is on
the horizontal axis and the agreed valuation is on the vertical axis. Both axes use units of
\$1 million. The orange line is the straight line that fits the points best.
It shows the average agreed valuation at each claimed valuation, not the
outcome of every deal. The rest of this lecture explains what "fits best"
means and how to read the line.

The figure shows 819 of the 874 deals, so that the area with many points is
easy to read. The figure leaves out the 55 deals with a valuation above
\$10 million on at least one axis. R still uses these deals to calculate the
orange line. Always say when a graph is zoomed in, that is, when it shows
only part of the data. Otherwise readers may think that the displayed range
is the complete data.

Most points lie below the grey dashed line. In 81% of the deals the agreed
valuation is lower than the valuation the founder claimed. In 13% it is the
same, and in 6% it is higher. The sharks usually pay the amount that the
founder asked for, but they take a larger share of the company for it. In
this way the agreed valuation ends up lower than the claimed valuation.

### Units first: money in millions

Every number in this lecture has the units of these two columns. So we fix
the units before we fit any model. We divide the raw dollar valuations by
1,000,000: a claimed valuation of \$2,000,000 becomes `claimed_m = 2`. This
change of scale does not change the relationship between the two valuations. @exm-i01-units, after the model is
fitted, shows what happens to the coefficients when the units change.

### The population model and the conditional mean

A regression has an **outcome** $Y$, the variable we want to describe, and a
**predictor** $X$, the variable we describe it with. Here the outcome is the
agreed valuation and the predictor is the claimed valuation. The simple
linear regression model is

$$
Y_i=\beta_0+\beta_1X_i+\varepsilon_i.
$$

- $Y_i$ is the outcome for observation $i$ (here: deal $i$).
- $X_i$ is the predictor for observation $i$.
- $\beta_0$ is the population intercept.
- $\beta_1$ is the population slope.
- $\varepsilon_i$ is the error: the part of $Y_i$ that the line does not
  capture.

The model is a statement about averages. The **conditional mean**
$E[Y\mid X=x]$ is the average outcome among observations whose predictor
equals $x$. An example is the average agreed valuation among deals in which
the founder claimed exactly \$2 million. The model says that this average
follows a line:

$$
E[Y\mid X=x]=\beta_0+\beta_1x.
$$

If you compare two values of the predictor that are one unit apart, the two
averages differ by the slope:

$$
\beta_1
=E[Y\mid X=x+1]-E[Y\mid X=x].
$$

::: {.definition-box}
<span class="label">Definition</span>

**Slope** $\beta_1$. The difference in the average outcome $Y$ that is
associated with a one-unit difference in the predictor $X$. Its unit is
units of $Y$ per unit of $X$.

**Intercept** $\beta_0$. The average outcome $Y$ at $X=0$. It has the units
of $Y$, and it has a useful meaning only if $X=0$ lies inside the range of
the data.
:::

Here the slope is the difference in the average agreed valuation between
deals whose claimed valuations differ by \$1 million. The words "associated
with" matter. A regression line compares averages across
groups of observed deals. It does not say what would happen if one founder
changed their claimed valuation. The box "What this fitted line can and cannot say" later
in this section explains this limit.

You can read the slope as an effect only under one condition: nothing
outside the model affects both the claimed and the agreed valuation. Lecture 2 calls this
condition no confounding.

### Errors versus residuals

The population line $\beta_0+\beta_1x$ is part of the model. Nobody can observe
it, because $\beta_0$ and $\beta_1$ are unknown. What we calculate from the
data is the fitted line. Its coefficients $\hat\beta_0$ and $\hat\beta_1$
have a hat symbol on top. The hat marks an estimate: these are the values
that we calculate from the data for the unknown $\beta_0$ and $\beta_1$. The fitted line is

$$
\widehat Y_i=\hat\beta_0+\hat\beta_1X_i.
$$

::: {.definition-box}
<span class="label">Definition</span>

**Fitted value** $\widehat Y_i$. The value on the fitted line at the
predictor value of observation $i$: $\widehat Y_i=\hat\beta_0+\hat\beta_1X_i$.

**Residual** $e_i$. The observed outcome minus the fitted value:
$e_i=Y_i-\widehat Y_i$. It is the vertical distance from the point to the
fitted line, positive above the line and negative below it.
:::

The error $\varepsilon_i$ and the residual $e_i$ have the same role in the
two equations, but they are different quantities, as @fig-i01-error-residual
shows.

```{r}
#| label: fig-i01-error-residual
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 8
#| fig-height: 3.6
#| fig-cap: "The same seven observations twice. Left: the unknown population line with the errors ε (distances to the true line). Right: the fitted line with the residuals e (distances to the fitted line). Only the right panel can be calculated from data."
library(ggplot2)
set.seed(101)
x_er <- c(1, 2.2, 3.4, 4.6, 5.8, 7.0, 8.2)
beta0_true <- 1; beta1_true <- 0.8
y_er <- beta0_true + beta1_true * x_er + c(1.3, -1.5, 0.9, -0.8, 1.6, -1.2, 0.4)
fit_er <- lm(y_er ~ x_er)

panel_names <- c("Population line (unknown)", "Fitted line (computed)")
er_df <- data.frame(
  panel = factor(rep(panel_names, each = 7), levels = panel_names),
  x = rep(x_er, 2),
  y = rep(y_er, 2),
  line_y = c(beta0_true + beta1_true * x_er, fitted(fit_er))
)
line_df <- data.frame(
  panel = factor(panel_names, levels = panel_names),
  intercept = c(beta0_true, coef(fit_er)[1]),
  slope = c(beta1_true, coef(fit_er)[2])
)
label_df <- data.frame(
  panel = factor(panel_names, levels = panel_names),
  x = x_er[5] + 0.45,
  y = c((y_er[5] + beta0_true + beta1_true * x_er[5]) / 2,
        (y_er[5] + fitted(fit_er)[5]) / 2),
  lab = c("epsilon[i]", "e[i]")
)

ggplot(er_df, aes(x = x, y = y)) +
  geom_abline(data = line_df[1, ],
              aes(intercept = intercept, slope = slope),
              colour = "#5B6976", linewidth = 0.9, linetype = "dashed") +
  geom_abline(data = line_df[2, ],
              aes(intercept = intercept, slope = slope),
              colour = "#D55E00", linewidth = 0.9) +
  geom_segment(aes(xend = x, yend = line_y),
               colour = "#0072B2", linewidth = 0.8) +
  geom_point(size = 2.6, colour = "#009E73") +
  geom_text(data = label_df, aes(x = x, y = y, label = lab),
            parse = TRUE, colour = "#0072B2", size = 4) +
  coord_cartesian(xlim = c(0, 9.4), ylim = c(0, 9.4)) +
  facet_wrap(~ panel) +
  labs(x = "Predictor X", y = "Outcome Y") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

- $\varepsilon_i = Y_i - (\beta_0+\beta_1X_i)$ measures the distance to the
  **population** line. Because $\beta_0$ and $\beta_1$ are unknown,
  $\varepsilon_i$ is never observed.
- $e_i = Y_i - (\hat\beta_0+\hat\beta_1X_i)$ measures the distance to the **fitted** line.
  It is computed from data, one number per observation.

We never see the errors. The residuals are what we can calculate in their
place, so they are estimates of the errors and not the errors themselves.
Lecture 2 uses residuals to check the model.

::: {.callout-note collapse="true"}
## Check: which object is observable?

Only the residual $e_i$. The observed data give $Y_i$ and $X_i$. The fitted
line gives $\widehat Y_i$, so $e_i = Y_i - \widehat Y_i$ is a computable
number. The error $\varepsilon_i$ depends on the unknown population
coefficients $\beta_0$ and $\beta_1$, so we can think about it but we can
never calculate it.
:::

### Least squares: why vertical, why squared

Many lines can be drawn through a cloud of points. To choose one, we need a
score that says how far a line misses the points. The residuals are the
misses. Each residual is measured vertically, in the units of the outcome:
millions of dollars of agreed valuation, not distance on the page. We
measure vertically because we use the line to describe $Y$ at a given $X$.
So a miss is a miss in $Y$.

Adding up the residuals is not a useful score, because positive and negative
misses cancel. Adding up their absolute values avoids this, but then there is
no simple formula for the best line. Squared residuals are never negative,
they count a large miss more heavily than a small one, and they lead to a
direct formula for the best line. For these reasons squaring is the standard
choice. But it is not the only possible choice.

::: {.definition-box}
<span class="label">Definition</span>

**SSE.** The sum of the squared residuals of a line. The letters come from
its usual name, "sum of squared errors", although it adds up squared
residuals.

$$
\operatorname{SSE}=\sum_{i=1}^{n}e_i^2
=\sum_{i=1}^{n}(Y_i-\hat\beta_0-\hat\beta_1X_i)^2.
$$

**Ordinary least squares (OLS).** The rule that chooses the intercept
$\hat\beta_0$ and the slope $\hat\beta_1$ so that SSE is as small as
possible. The resulting line is called the least-squares line.
:::

#### Try it yourself: make SSE as small as you can

Drag either square handle to move the line. The vertical dashed lines are the
residuals. The SSE below the figure changes every time you move the line. Find the line with
the smallest SSE that you can. Then click "Show least-squares fit" to see the
line with the smallest possible SSE, and compare it with your best attempt.

<style>
.ls-lab {
  background: #fff;
  border: 1px solid #d5dbe0;
  border-radius: 8px;
  margin: 1.2em 0 1.5em;
  padding: 1em;
}
.ls-lab-heading,
.ls-actions {
  align-items: center;
  display: flex;
  flex-wrap: wrap;
  gap: 0.8em 1.2em;
  justify-content: space-between;
}
.ls-lab svg {
  display: block;
  height: auto;
  margin: 0.5em auto;
  max-width: 700px;
  touch-action: none;
  width: 100%;
}
.ls-controls {
  display: grid;
  gap: 0.8em 1.4em;
  grid-template-columns: repeat(2, minmax(0, 1fr));
  margin: 0.6em 0 1em;
}
.ls-controls label {
  color: #17324d;
  font-size: 0.9em;
  font-weight: 600;
}
.ls-controls input {
  display: block;
  margin-top: 0.3em;
  width: 100%;
}
.ls-actions button {
  background: #17324d;
  border: 1px solid #17324d;
  border-radius: 4px;
  color: #fff;
  cursor: pointer;
  font: inherit;
  padding: 0.45em 0.8em;
}
.ls-actions button + button {
  background: #fff;
  color: #17324d;
}
.ls-caption {
  color: #53606b;
  font-size: 0.9em;
  margin-bottom: 0;
}
.ls-axis { stroke: #65727e; stroke-width: 1.2; }
.ls-tick { fill: #53606b; font-size: 13px; }
.ls-label { fill: #263b4b; font-size: 15px; }
.ls-point { fill: #0072b2; stroke: #17324d; stroke-width: 1; }
.ls-fit-line { stroke: #d55e00; stroke-width: 3; }
.ls-residual { stroke: #755f93; stroke-width: 2; stroke-dasharray: 4 3; }
.ls-handle { fill: #f0e442; stroke: #17324d; stroke-width: 2; cursor: ns-resize; }
.ls-handle:focus { outline: 3px solid #56b4e9; outline-offset: 2px; }
@media (max-width: 767px) {
  .ls-controls {
    grid-template-columns: 1fr;
  }
}
</style>

<div id="least-squares-lab" class="ls-lab" role="group" aria-labelledby="ls-lab-title">
<div class="ls-lab-heading">
<strong id="ls-lab-title">Least-squares line lab</strong>
<span id="ls-equation" aria-live="polite"></span>
</div>
<svg id="ls-svg" viewBox="0 0 700 390" role="img" aria-label="Scatterplot with a draggable line and vertical residual segments"></svg>
<div class="ls-controls">
<label>Line height at x = 1
<input id="ls-left" type="range" min="0" max="10" step="0.05" value="3">
</label>
<label>Line height at x = 9
<input id="ls-right" type="range" min="0" max="10" step="0.05" value="7">
</label>
</div>
<div class="ls-actions">
<button id="ls-best" type="button">Show least-squares fit</button>
<button id="ls-reset" type="button">Try another line</button>
<strong id="ls-sse" aria-live="polite"></strong>
</div>
<p class="ls-caption">Squaring makes every contribution to SSE zero or positive. A residual that is twice as long adds four times as much to SSE.</p>
</div>

<script>
(function () {
  const root = document.getElementById("least-squares-lab");
  if (!root || root.dataset.ready) return;
  root.dataset.ready = "true";
  const svg = root.querySelector("#ls-svg");
  const left = root.querySelector("#ls-left");
  const right = root.querySelector("#ls-right");
  const equation = root.querySelector("#ls-equation");
  const sseText = root.querySelector("#ls-sse");
  const points = [
    {x:1, y:1.8}, {x:2, y:1.4}, {x:3, y:3.2},
    {x:4, y:4.8}, {x:5, y:4.2}, {x:6, y:6.8},
    {x:7, y:6.4}, {x:8, y:8.6}, {x:9, y:8.2}
  ];
  const ns = "http://www.w3.org/2000/svg";
  const box = {left:58, right:680, top:18, bottom:336};
  const sx = x => box.left + x / 10 * (box.right - box.left);
  const sy = y => box.bottom - y / 10 * (box.bottom - box.top);
  const add = (name, attrs, parent=svg) => {
    const el = document.createElementNS(ns, name);
    Object.entries(attrs).forEach(([k,v]) => el.setAttribute(k, v));
    parent.appendChild(el); return el;
  };
  add("line", {x1:box.left, y1:box.bottom, x2:box.right, y2:box.bottom, class:"ls-axis"});
  add("line", {x1:box.left, y1:box.top, x2:box.left, y2:box.bottom, class:"ls-axis"});
  [0,2,4,6,8,10].forEach(v => {
    add("line", {x1:sx(v), y1:box.bottom, x2:sx(v), y2:box.bottom+6, class:"ls-axis"});
    const tx=add("text", {x:sx(v), y:box.bottom+25, class:"ls-tick"}); tx.textContent=v;
    add("line", {x1:box.left-6, y1:sy(v), x2:box.left, y2:sy(v), class:"ls-axis"});
    const ty=add("text", {x:box.left-12, y:sy(v)+4, class:"ls-tick", "text-anchor":"end"}); ty.textContent=v;
  });
  const xl=add("text", {x:(box.left+box.right)/2, y:382, class:"ls-label", "text-anchor":"middle"}); xl.textContent="Predictor x";
  const yl=add("text", {x:16, y:(box.top+box.bottom)/2, class:"ls-label", transform:`rotate(-90 16 ${(box.top+box.bottom)/2})`, "text-anchor":"middle"}); yl.textContent="Outcome y";
  const residualGroup=add("g", {class:"ls-residuals"});
  const line=add("line", {class:"ls-fit-line"});
  const handleLeft=add("rect", {width:16, height:16, rx:2, class:"ls-handle", "aria-hidden":"true"});
  const handleRight=add("rect", {width:16, height:16, rx:2, class:"ls-handle", "aria-hidden":"true"});
  points.forEach(p => add("circle", {cx:sx(p.x), cy:sy(p.y), r:5.5, class:"ls-point"}));
  const meanX=points.reduce((a,p)=>a+p.x,0)/points.length;
  const meanY=points.reduce((a,p)=>a+p.y,0)/points.length;
  const bestSlope=points.reduce((a,p)=>a+(p.x-meanX)*(p.y-meanY),0) /
                  points.reduce((a,p)=>a+(p.x-meanX)**2,0);
  const bestIntercept=meanY-bestSlope*meanX;
  const fitFromHeights = () => {
    const y1=Number(left.value), y9=Number(right.value);
    const slope=(y9-y1)/8, intercept=y1-slope;
    return {intercept, slope};
  };
  function draw() {
    const {intercept, slope}=fitFromHeights();
    residualGroup.replaceChildren();
    let sse=0;
    points.forEach(p => {
      const fitted=intercept+slope*p.x;
      sse+=(p.y-fitted)**2;
      add("line", {x1:sx(p.x), y1:sy(p.y), x2:sx(p.x), y2:sy(fitted), class:"ls-residual"}, residualGroup);
    });
    line.setAttribute("x1", sx(0)); line.setAttribute("y1", sy(intercept));
    line.setAttribute("x2", sx(10)); line.setAttribute("y2", sy(intercept+10*slope));
    [[handleLeft,1,Number(left.value)],[handleRight,9,Number(right.value)]].forEach(([h,x,y])=>{
      h.setAttribute("x",sx(x)-8); h.setAttribute("y",sy(y)-8);
    });
    equation.textContent=`ŷ = ${intercept.toFixed(2)} ${slope<0?"−":"+"} ${Math.abs(slope).toFixed(2)}x`;
    sseText.textContent=`SSE = ${sse.toFixed(2)}`;
  }
  function drag(handle, slider) {
    handle.addEventListener("pointerdown", event => {
      handle.setPointerCapture(event.pointerId);
      const move = e => {
        const rect=svg.getBoundingClientRect();
        const svgY=(e.clientY-rect.top)*390/rect.height;
        const y=Math.max(0,Math.min(10,(box.bottom-svgY)*10/(box.bottom-box.top)));
        slider.value=y.toFixed(2); draw();
      };
      handle.addEventListener("pointermove", move);
      handle.addEventListener("pointerup", () => handle.removeEventListener("pointermove", move), {once:true});
    });
  }
  drag(handleLeft,left); drag(handleRight,right);
  root.querySelector("#ls-best").addEventListener("click",()=>{
    left.value=(bestIntercept+bestSlope).toFixed(2);
    right.value=(bestIntercept+9*bestSlope).toFixed(2); draw();
  });
  root.querySelector("#ls-reset").addEventListener("click",()=>{
    left.value=3; right.value=7; draw();
  });
  draw();
})();
</script>

#### The formulas for the slope and the intercept

You do not have to find the least-squares line by trying many lines. There
is a direct formula for the slope and one for the intercept, so no search is
needed.

In words, the slope compares two amounts: how much $Y$ moves together with
$X$, and how much $X$ moves at all. Two quantities from QM Basic measure
them.

- The **sample variance** of $X$, written $s_X^2$, measures how far the
  values of $X$ spread around their mean $\bar x$. Its square root $s_X$ is
  the standard deviation.
  [Basic Lecture 3](../basic/b03-discrete-distributions.qmd) explains the
  variance and the standard deviation.
- The **sample covariance** of $X$ and $Y$, written $s_{XY}$, measures
  whether $X$ and $Y$ move together. It is positive when high values of $X$
  go with high values of $Y$, and negative when high values of $X$ go with
  low values of $Y$.
  [Basic Lecture 5](../basic/b05-joint-covariance-correlation.qmd) explains
  the covariance and the correlation.

The variance is an average of squared deviations from the mean. The
covariance is an average of products of deviations. Both divide by $n-1$ in
place of $n$:

$$
s_X^2=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar x)^2,
\qquad
s_{XY}=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar x)(y_i-\bar y).
$$

The slope is the covariance of $X$ and $Y$ divided by the variance of $X$:

$$
\hat\beta_1=\frac{s_{XY}}{s_X^2}.
$$

The covariance has the units of $X$ times the units of $Y$, and the variance
has the squared units of $X$. So their ratio has the units of a slope:
units of $Y$ per unit of $X$.

The same slope can be written with the correlation
$r_{XY}=s_{XY}/(s_X s_Y)$. The correlation is the covariance rescaled so that
it has no units and lies between $-1$ and $1$. Then

$$
\hat\beta_1=r_{XY}\,\frac{s_Y}{s_X}.
$$

In words: the slope is the correlation, rescaled into units of $Y$ per unit
of $X$. Two facts follow from this. The slope always has the sign of the correlation,
because standard deviations are positive. And if $X$ and $Y$ are both
standardised, so that each has standard deviation 1, the slope is equal to
the correlation.

The intercept is chosen so that the line passes through the point of means
$(\bar x,\bar y)$:

$$
\hat\beta_0=\bar y-\hat\beta_1\bar x.
$$

You can check this by putting $x=\bar x$ into the fitted line:
$\hat\beta_0+\hat\beta_1\bar x=\bar y$. At the average value of $X$, the
line gives the average value of $Y$.

::: {#exm-i01-toy}
## Worked example: least squares by hand on three points

Three observations: $(1,2)$, $(2,5)$, $(3,5)$. We write out every sum.

**Means.** $\bar x = (1+2+3)/3 = 2$ and $\bar y = (2+5+5)/3 = 4$.

**Deviations and their products.**

| $i$ | $x_i-\bar x$ | $y_i-\bar y$ | $(x_i-\bar x)(y_i-\bar y)$ | $(x_i-\bar x)^2$ | $(y_i-\bar y)^2$ |
|---|---:|---:|---:|---:|---:|
| 1 | $-1$ | $-2$ | $2$ | $1$ | $4$ |
| 2 | $0$ | $1$ | $0$ | $0$ | $1$ |
| 3 | $1$ | $1$ | $1$ | $1$ | $1$ |
| **Sum** | | | $3$ | $2$ | $6$ |

**Covariance and variance.** Divide the sums by $n-1=2$. The covariance is
$s_{XY} = 3/2 = 1.5$ and the variance of $X$ is $s_X^2 = 2/2 = 1$.

**Slope: the covariance divided by the variance of $X$.**

$$
\hat\beta_1 = \frac{s_{XY}}{s_X^2} = \frac{1.5}{1} = 1.5.
$$

**The same slope from the correlation.** The standard deviations are
$s_X=\sqrt{1}=1$ and $s_Y=\sqrt{6/2}\approx 1.732$, so the correlation is
$r_{XY}=1.5/(1\times 1.732)\approx 0.866$. Then

$$
\hat\beta_1 = r_{XY}\,\frac{s_Y}{s_X} = 0.866\times\frac{1.732}{1} = 1.5.
$$

**Intercept: the line passes through the point of means.**

$$
\hat\beta_0 = \bar y - \hat\beta_1\bar x = 4 - 1.5\times 2 = 1.
$$

The fitted line is $\widehat y = 1 + 1.5x$. @fig-i01-toy shows that the line
passes exactly through the point of means $(\bar x,\bar y)=(2,4)$. The
residuals are $-0.5$, $+1$, and $-0.5$. They sum to zero, as least-squares
residuals always do when the model has an intercept.
:::

```{r}
#| label: fig-i01-toy
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 5.4
#| fig-height: 4.2
#| fig-cap: "The three-point toy dataset with its least-squares line. The open diamond marks the point of means (2, 4), which the fitted line passes through exactly."
library(ggplot2)
x_toy <- c(1, 2, 3); y_toy <- c(2, 5, 5)
toy_fit <- lm(y_toy ~ x_toy)
toy_df <- data.frame(x = x_toy, y = y_toy, fitted = fitted(toy_fit))

ggplot(toy_df, aes(x = x, y = y)) +
  geom_abline(intercept = coef(toy_fit)[1], slope = coef(toy_fit)[2],
              colour = "#D55E00", linewidth = 0.9) +
  geom_segment(aes(xend = x, yend = fitted),
               colour = "#0072B2", linewidth = 0.8, linetype = "dashed") +
  geom_point(size = 3, colour = "#009E73") +
  annotate("point", x = 2, y = 4, shape = 5, size = 3.4, stroke = 1.1,
           colour = "#17324D") +
  annotate("text", x = 2.34, y = 3.72,
           label = "'('*bar(x)*', '*bar(y)*')'", parse = TRUE,
           colour = "#17324D", size = 3.4) +
  coord_cartesian(xlim = c(0.5, 3.5), ylim = c(1, 6)) +
  labs(x = "x", y = "y") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

You can verify the hand calculation in R:

```{r}
#| label: i01-toy-check
x_toy <- c(1, 2, 3); y_toy <- c(2, 5, 5)
c(covariance = cov(x_toy, y_toy), variance_x = var(x_toy),
  sd_y = sd(y_toy), correlation = cor(x_toy, y_toy))
coef(lm(y_toy ~ x_toy))
```

### Fit and read the Shark Tank model in R

[Shark Tank data]{.shark-tag} Now the Shark Tank data. The function `lm()` (for "linear model") fits the
least-squares line. Its first argument is a formula of the form
`outcome ~ predictor`. The block below loads the data again, so you can run
it on its own:

```{r}
#| label: i01-fit
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6

val_model <- lm(agreed_m ~ claimed_m, data = vals)
coef(val_model)
```

The fitted equation is approximately

$$
\widehat{\text{agreed valuation}}
=-0.068+0.649\times\text{claimed valuation},
$$

with both valuations measured in \$ millions.

R calculated these two numbers with the formulas of the previous section.
You can check this with the covariance, the variance, the correlation, and
the standard deviations of the two columns:

```{r}
#| label: i01-formula-check
x <- vals$claimed_m
y <- vals$agreed_m
c(correlation = cor(x, y), sd_x = sd(x), sd_y = sd(y))
c(cov_over_var = cov(x, y) / var(x),
  cor_times_sd_ratio = cor(x, y) * sd(y) / sd(x),
  slope_from_lm = unname(coef(val_model)["claimed_m"]))
c(mean_y_minus_slope_times_mean_x = mean(y) - cov(x, y) / var(x) * mean(x),
  intercept_from_lm = unname(coef(val_model)["(Intercept)"]))
```

Both formulas give the slope 0.649 that `lm()` reports, and the intercept
formula gives $-0.068$. The second form of the slope also shows why it is
below 1. The correlation of the two valuations is 0.893, and the agreed
valuations spread less than the claimed ones: their standard deviation is
\$3.70 million against \$5.09 million. So
$\hat\beta_1 = 0.893 \times 3.70/5.09 \approx 0.649$.

**The slope, with units.** A claimed valuation that is \$1 million higher is
associated with an agreed valuation that is about \$0.65 million higher,
among deals for a share of the company. A founder would say: the sharks
accept about 65 cents of every claimed dollar. This answers the prediction
question from before class. The sign is positive, the unit is millions of
dollars agreed per million claimed, and the slope is clearly below 1. The
slope does not mean that claiming more causes the sharks to pay more.

**The intercept, with units.** The intercept gives a fitted agreed valuation
of about $-\$68{,}000$ at a claimed valuation of zero. A company claimed to
be worth nothing is not in the data (the smallest claimed valuation is
\$40,000), and a negative valuation has no meaning. The line needs an
intercept so that it is at the right height. But this intercept has no
useful meaning of its own.

::: {#exm-i01-units}
## Worked example: the same coefficients in three unit systems

[Shark Tank data]{.shark-tag} The fitted intercept is about $-0.068$ and the fitted slope is about $0.649$,
with both valuations in \$ millions. Read the intercept in dollars first:

$$
-0.068 \text{ (in \$ millions)} = -0.068 \times \$1{,}000{,}000 = -\$68{,}000.
$$

Suppose a report measures both valuations in \$1,000s. Every value of $Y$
and of $X$ is then 1,000 times as large. The intercept has the units of $Y$,
so it is multiplied by 1,000. The slope is a ratio, units of $Y$ per unit of
$X$. Both parts of the ratio change by the same factor, so the slope stays
the same:

$$
\text{intercept: } -0.068 \times 1000 = -68 \text{ (in \$1,000s)},
\qquad
\text{slope: } 0.649 \text{ (unchanged)}.
$$

R confirms this result:

```{r}
#| label: i01-units-check
vals$claimed_k <- vals$claimed_m * 1000   # claimed valuation in $1,000s
vals$agreed_k <- vals$agreed_m * 1000     # agreed valuation in $1,000s
coef(lm(agreed_k ~ claimed_k, data = vals))
```

In raw dollars, the intercept is $-68{,}000$ dollars and the slope is again
0.649: dollars of agreed valuation per dollar of claimed valuation.

The general rule has three cases. If you multiply the outcome and the
predictor by the same factor, the intercept is multiplied by that factor and
the slope does not change. If you multiply only the outcome by a factor, the
slope is multiplied by the same factor. If you multiply only the predictor
by a factor, the slope is divided by it. When two sources disagree about a
coefficient, check the units first.
:::

The table below shows the observed value, the fitted value, and the residual
for the first six rows of the data:

```{r}
#| label: i01-fitted-residual
head(data.frame(
  observed = vals$agreed_m,
  fitted = fitted(val_model),
  residual = residuals(val_model)
))
```

For every row, `observed = fitted + residual`. Check it on the first row.
The founders asked \$50,000 for 15%, a claimed valuation of \$0.333 million.
They agreed to \$50,000 for 55%, an agreed valuation of \$0.091 million. The
fitted value is $-0.068 + 0.649 \times 0.333 = 0.148$, and indeed
$0.091 = 0.148 + (-0.057)$.

### Prediction inside the range of the data

[Shark Tank data]{.shark-tag} The function `predict()` calculates the fitted value for a new value of the
predictor:

```{r}
#| label: i01-one-prediction
predict(val_model, newdata = data.frame(claimed_m = 2))
range(vals$claimed_m)   # smallest and largest claimed valuation
```

For a claimed valuation of \$2 million (`claimed_m = 2`), the fitted agreed
valuation is about 1.23, that is, \$1.23 million. This value is the average that
the model gives for such deals. It is not a guarantee for a new pitch. If such a
deal closed at a valuation of \$1 million, its residual would be

$$
e = 1.00 - 1.23 = -0.23,
$$

about \$230,000 below the fitted value. Use predictions only inside the range
of the data. The claimed valuations run from \$40,000 to \$50 million. Outside
this range there are no deals to support the line. A prediction there
depends only on the assumption that the line continues.

::: {.callout-caution collapse="true"}
## Try it: a fitted value and a residual

[Shark Tank data]{.shark-tag} Try this with your neighbour first. A founder claims a valuation of
\$8 million, and the deal closes at a valuation of \$6 million.

*Step 1 (given).* The fitted equation is
$\widehat{\text{agreed}}=-0.068+0.649\times\text{claimed}$, in \$ millions.

*Step 2 (you).* Calculate the fitted agreed valuation.

*Step 3 (you).* Calculate the residual and say what its sign means.

::: {.callout-note collapse="true"}
## Check your answer

Fitted: $-0.068+0.649\times8=5.12$, about \$5.12 million. Residual:
$6-5.12=0.88$, about \$880,000. The sign is positive: this deal closed
above the fitted value. R confirms the residual:

```{r}
#| label: i01-tryit-check
unname(6 - predict(val_model, newdata = data.frame(claimed_m = 8)))
```
:::
:::

::: {.caveat-box}
<span class="label">What this fitted line can and cannot say</span>

- The slope 0.649 describes how deals differed. Founders who claim more also
  differ in sales, stage, and industry.
- Only pitches that reached a deal are in the data.
- The model does not say what happens if you raise your own claim.

**What you can do with it:** expect the sharks to accept about two thirds of
a claimed valuation, and prepare your negotiation for that.
:::

### From an association to a decision

The box above protects you from one mistake: reading a description as a
promise. It does not make the slope useless. Founders and managers decide
with imperfect evidence every day. Three questions help.

1. **How expensive is a wrong decision, and can I undo it?** If the action
   is cheap and easy to undo, act on the association and watch what happens.
   If the action is expensive or hard to undo, ask for more evidence first.
2. **What else differs between the cases I compare?** Founders who claim
   more also differ in sales, stage, and industry. Lecture 4 shows how to
   hold measured differences fixed.
3. **Can I test it myself?** A founder can often change one feature on purpose
   for some customers and not for others. Chance decides which customers get
   the change. Lecture 5 shows
   how to read such a test.

### Answering the driving question

Among pitches that reached a deal, a claimed valuation is worth about two
thirds of the claimed amount. The fitted slope is 0.649, so each extra \$1
million of claimed valuation is associated with about \$0.65 million more
agreed valuation. The limits in the box "What this fitted line can and cannot
say" still apply.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You regress weekly orders on weekly advertising spending in euros and get: fitted orders = 120 + 0.15 × spending. Each extra 100 euros of advertising goes with 15 more orders. You earn 4 euros per order, so 100 euros of advertising goes with 60 euros of extra margin. The decision: at this slope, advertising does not pay for itself, unless new customers come back and order again. The slope turned a feeling that ads help into a number that you can compare with the cost. But the slope only describes your past weeks. Before a large budget change, vary the spending on purpose for a few weeks and see whether the orders follow.
:::

**Next:** R fits a line to any data, even to data that a line describes
badly. R gives no warning. Lecture 2 states the assumptions behind the
fitted line and shows how to check them with residual plots.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions

1. What is the difference between $\varepsilon_i$ and $e_i$?
2. What does least squares minimise, and why are the residuals squared
   before they are added up?
3. Why must a slope interpretation include units?
4. Why can an intercept be calculated correctly and still have no useful
   meaning?
5. Why is the slope not automatically causal?

::: {.callout-note collapse="true"}
## Check your answers

1. $\varepsilon_i$ is the error of the population model: the distance to
   the unknown population line. It is never observed. $e_i$ is the residual:
   the distance to the fitted line, calculated from the data.
2. The sum of squared residuals, $\operatorname{SSE}=\sum_i e_i^2$. In a
   plain sum the positive and the negative residuals would cancel. Squares
   are never negative, and they count a large miss more heavily than a small
   one.
3. The slope is a difference in outcome units per one unit of the predictor.
   Without units the number 0.649 could mean anything.
4. The value $X=0$ may lie outside the observed range, or it may make no
   sense for the variable.
5. The predictor was not randomly assigned, and other factors may affect
   both variables.
:::

### Finish a half-worked example

[Shark Tank data]{.shark-tag} Re-read the prediction calculation in the chapter (fitted value at
`claimed_m = 2`). Then complete the same calculation for a new claim.

*Step 1 (given).* A founder claims a valuation of \$4 million, so
`claimed_m = 4`. The fitted equation is
$\widehat{\text{agreed}} = -0.068 + 0.649 \times \text{claimed}$.

*Step 2 (you).* Compute the fitted agreed valuation, in \$ millions and in
dollars.

*Step 3 (you).* The deal actually closed at a valuation of \$2 million.
Compute the residual and interpret its sign in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

Fitted: $-0.068 + 0.649 \times 4 = 2.53$, about \$2.53 million. Residual:
$2 - 2.53 = -0.53$, about $-\$530{,}000$. The negative sign means this deal
closed below the model's fitted conditional mean for a claimed valuation of
\$4 million.

```{r}
#| label: i01-faded-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
unname(2 - predict(val_model, newdata = data.frame(claimed_m = 4)))
```
:::

Then do five more tasks. These tasks have no given steps:

1. Use the fitted equation to predict the agreed valuation for claimed
   valuations of \$1 million and \$10 million.
2. Calculate the residual for a deal with a claimed valuation of \$2 million
   and an agreed valuation of \$1 million.
3. Explain why `lm(y ~ x, data = dat)` treats the variable left of `~` as
   the outcome.
4. A report keeps the agreed valuation in \$ millions but measures the
   claimed valuation in \$1,000s. Which intercept and which slope does it
   report? Use the rule of @exm-i01-units.
5. Under which condition could the slope 0.649 be read as the effect of a
   higher claim? Name one factor outside the model that breaks this condition
   for the Shark Tank deals.

::: {.callout-note collapse="true"}
## Check your answers

1. Substitute `claimed_m = 1` and `claimed_m = 10`. The fitted values are
   about 0.58 and 6.42 in \$ millions.
2. The fitted valuation is about 1.23 in \$ millions, so the residual is
   $1-1.23\approx-0.23$, or about $-\$230{,}000$.
3. R formulas use `outcome ~ predictors`: the left-hand side is what the
   model predicts. The right-hand side is what the model uses to predict it.
4. Only the predictor is rescaled, by the factor 1,000. The slope is divided
   by 1,000: $0.649/1000 = 0.000649$ million dollars of agreed valuation per
   \$1,000 claimed, which is \$649 per \$1,000. The intercept stays at
   $-0.068$, because the outcome keeps its units.
5. The condition: nothing outside the model affects both the claimed and the
   agreed valuation. Here it fails. A company with high sales can claim a
   higher valuation, and the sharks also agree to a higher valuation for it.
   The slope then mixes the effect of the claim with the effect of the sales.

The R check for task 4:

```{r}
#| label: i01-practice-units
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_k <- vals$requested_valuation_usd / 1e3   # in $1,000s
vals$agreed_m <- vals$deal_valuation_usd / 1e6         # in $ millions
coef(lm(agreed_m ~ claimed_k, data = vals))
```
:::

### Practice: a regression of your own

[Shark Tank data]{.shark-tag} Fit `equity_offered_pct ~ season` on the full dataset (all 1,441 pitches)
and interpret both coefficients with units. State the unit of every quantity
you report.

::: {.callout-note collapse="true"}
## Check your answers

```{r}
#| label: i01-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
coef(lm(equity_offered_pct ~ season, data = sharks))
```

The fitted intercept is about 20.46 percentage points and the slope about
$-0.824$ percentage points per season. Later seasons are associated with
smaller initial equity offers on average. The outcome and the intercept are in
percentage points. The predictor is in seasons. The slope is in percentage
points per season. The intercept is the fitted value at Season 0, which is
outside the data (Seasons 1 to 16), so it has no useful meaning of its own.
:::

### Review: one question from QM Basic

An invented example. A market survey of $n = 400$ customers gives an average
willingness to pay of €24.60 with standard error €1.10. Write down the t
value for testing a benchmark of €22 and the 95% confidence interval, and
say in one sentence what the interval estimates.

::: {.callout-note collapse="true"}
## Check your answer

$t_{obs}=(24.60-22)/1.10\approx2.36$. With $df=399$, the closest listed
value on the formula sheet is $t_{0.975,\,120}=1.980$. The interval is
$24.60\pm1.980\times1.10=24.60\pm2.18=(22.42,\ 26.78)$ euros. It is a
range of plausible values for the population mean willingness to pay, if
the 400 customers are a random sample.
:::

### Worked exam interpretation

**Exam-style question.** An invented example. A model of the
monthly revenue of cafés (€1,000s) on their opening hours per week reports
$\widehat{\text{revenue}} = 4.20 + 0.85\,\text{hours}$. Interpret the slope
with units, and calculate the fitted value and the residual for a café that
is open 40 hours per week and has a monthly revenue of €35,000.

::: {.callout-note collapse="true"}
## Check a model answer

Each additional opening hour per week is associated with €850 more fitted
monthly revenue among the sampled cafés. For 40 hours the fitted value is
$4.20+0.85\times40=38.2$, about €38,200. The residual is $35-38.2=-3.2$,
about $-$€3,200: the observed revenue is below the fitted value.
:::

### Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant: "The slope of my valuation regression is 0.649 and the intercept is -0.068. What do they mean for my pitch?" It answered:

> "If you raise your claimed valuation by 1 million dollars, the sharks will pay you 649,000 dollars more. And a company with a claimed valuation of zero gets -68,000 dollars."

Find what is wrong and write the corrected answer. This task is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

Three errors. The slope describes how deals differed. It does not say what happens if you change your own claim, because stronger companies both claim more and get more. The outcome is the agreed valuation of the company, not money paid to the founder. And a claimed valuation of zero is outside the data, so the intercept has no meaning of its own.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Name one
numerical outcome and one numerical input that you think affects it. Then
write two sentences. First, state the slope you would expect (sign and
units). Second, say what decision would change if the fitted slope turned
out to be near zero. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student venture that sells boxes of prepared meals on campus. "Outcome: weekly orders. Input: euros spent on social-media advertising that week. I expect a positive slope of a few orders per extra €100. If the fitted slope came out near zero, I would move the whole budget to campus flyers."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it what else, besides your input,
could affect your outcome.
::::

[Continue to Lecture 2: What the model assumes and how residuals reveal it](i02-assumptions-diagnostics.qmd).

[Code sheet: Lecture 1 commands and formulas](code-sheet.qmd#lecture-1-simple-linear-regression).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.1 and 3.6.2, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 19.

:::
