# Lecture 1: Simple linear regression {#i01}

```{r}
#| label: i01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 1, you can</span>

- state a simple linear regression model and identify its components;
- read the slope as a difference of conditional means, with units;
- convert a coefficient between measurement units (dollars, $1,000s,
  $100,000s);
- distinguish observations, fitted values, errors, and residuals;
- explain what least squares minimises and why;
- compute a slope and intercept by hand on a tiny dataset;
- fit a regression with `lm()` in R and interpret both coefficients within
  the observed range.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">The first question of deal-making: does asking for more get you more? A line fitted through the data turns that vague hunch into one number a founder can actually use.</p>

Among pitches that reached a deal, does asking for more money go together
with getting more money --- and by how much per dollar asked?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=vCHtY6Me5FI&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=9">3.1
Simple Linear Regression</a>. Focus on the population line, the fitted line,
and least squares. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate the pentad (10 minutes).** QM Intermediate builds directly on
the five quantities you met at the end of QM Basic --- estimate, standard
error, t value, p-value, confidence interval. With the Basic formula sheet
open, answer the five items below.

1. A study reports an estimate of 12 with standard error 4. What is the t
   value for testing whether the underlying quantity is zero?
2. With a large sample, what multiplier goes into a 95% confidence interval?
3. Compute that 95% interval for the estimate in item 1.
4. The p-value is 0.003. Complete: "If the true value were zero, ..."
5. Which of the five quantities has the same units as the estimate itself?

::: {.callout-note collapse="true"}
## Check your answers

1. $t = 12/4 = 3$.
2. Approximately 1.96 (the t multiplier is close to the normal 1.96 when the
   degrees of freedom are large).
3. $12 \pm 1.96 \times 4 = (4.2,\; 19.8)$.
4. "... an estimate at least this far from zero would occur in about 0.3% of
   samples."
5. The standard error and both interval endpoints; the t value and p-value
   are unit-free.
:::

**Define the variables (2 minutes).** For the 882 pitches with an on-air
deal, this lecture works in **thousands of dollars**:

- $Y_i$: agreed on-air deal amount, in $1,000s;
- $X_i$: amount initially requested, in $1,000s.

A pitch that asked for \$300,000 and agreed \$250,000 has $X_i = 300$ and
$Y_i = 250$.

**Predict (3 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: what are the **sign** and the **unit** of
the slope in a regression of agreed amount on requested amount? Commit to an
answer; the lecture resolves it.

**R warm-up (5 minutes).** The standing warm-up for every Intermediate
lecture: run a few lines of R and bring the output to class. This week, load
the data, subset the deals, and plot agreed amount against requested amount:

```{r}
#| label: i01-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
plot(deals$ask_1k, deals$deal_1k)
```

If lines, slopes, intercepts, or subscripts are not secure, review
[Mathematics Refresher: straight lines and indices](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question.

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=vCHtY6Me5FI">3.1 Simple Linear Regression</a>.</p>

### Begin with the data

The warm-up plot, polished. The construction is exactly the warm-up code:

```{r}
#| label: i01-build-deals
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
nrow(deals)
```

```{r}
#| label: fig-i01-scatter
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Agreed on-air deal amount against amount requested, for the 882 pitches with a deal. Both axes in $1,000s. The display zooms in on requests and deals up to $1 million (864 of 882 records); the orange least-squares line is fitted on all 882."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
display_deals <- subset(deals, ask_1k <= 1000 & deal_1k <= 1000)

ggplot(display_deals, aes(x = ask_1k, y = deal_1k)) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  geom_abline(intercept = coef(deal_model)[1],
              slope = coef(deal_model)[2],
              colour = "#D55E00", linewidth = 0.9) +
  annotate("text", x = 40, y = 985,
           label = "Fitted regression (estimated on all 882 deals)",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  coord_cartesian(xlim = c(0, 1030), ylim = c(0, 1030), expand = FALSE) +
  labs(x = "Amount requested ($1,000s)",
       y = "Agreed on-air deal amount ($1,000s)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each point in @fig-i01-scatter is one pitch with an on-air deal. Read
requested amount horizontally and agreed amount vertically; both axes use
units of $1,000. The orange line gives the fitted conditional mean, not the
outcome of every pitch. The display shows 864 of the 882 deal records so
that the dense region is readable. Eighteen records above \$1 million on at
least one axis are omitted from this display but retained when estimating
the orange line. Always say when a graph is zoomed; otherwise readers may
mistake the displayed range for the complete data.

### Units first: money in thousands

Every number this lecture produces carries the units of these two columns,
so settle the units before any model. We divide raw dollar amounts by 1,000:
a request of \$300,000 becomes `ask_1k = 300`. Scaling changes nothing about
the relationship; it only makes axes and coefficients readable
(`ask_amount_usd = 300000` and `ask_1k = 300` are the same fact).

::: {#exm-i01-units}
## Worked example: the same coefficient in three unit systems

Later in this lecture, R will print an intercept of about 27.571 with both
variables in $1,000s. Two lines of arithmetic read it:

$$
27.571 \text{ (in \$1,000s)} = 27.571 \times \$1{,}000 = \$27{,}571.
$$

Suppose a published paper measures both variables in \$100,000s instead.
Dividing both $Y$ and $X$ by a further 100 shrinks the intercept by that
factor but leaves the slope untouched, because the slope is a ratio of the
two:

$$
\text{intercept: } 27.571 / 100 = 0.276 \text{ (in \$100,000s)},
\qquad
\text{slope: } 1.0085 \text{ (unchanged)}.
$$

And in raw dollars? Multiplying both variables by 1,000 gives an intercept
of $27.571 \times 1000 = 27{,}571$ dollars --- and again the same slope,
1.0085 dollars per dollar. The general rule: rescaling the outcome and the
predictor by the **same** factor moves the intercept and leaves the slope
alone; rescaling only one of them rescales the slope by that factor. When
two sources disagree about a coefficient, check the units before suspecting
the analysis.
:::

### The population model and the conditional mean

The simple linear regression model is

$$
Y_i=\beta_0+\beta_1X_i+\varepsilon_i.
$$

Equivalently, the conditional mean is

$$
E[Y_i\mid X_i=x]=\beta_0+\beta_1x.
$$

- $Y_i$ is the outcome for observation $i$.
- $X_i$ is the predictor.
- $\beta_0$ is the population intercept.
- $\beta_1$ is the population slope.
- $\varepsilon_i$ is the part of $Y_i$ not represented by the line.

The slope is a **difference of conditional means**:

$$
\beta_1
=E[Y\mid X=x+1]-E[Y\mid X=x].
$$

It is the expected difference in $Y$ **associated with** a one-unit
difference in $X$ --- here, the expected difference in agreed amount between
pitches whose requests differ by \$1,000. "Associated with" is deliberate,
and this is the one place the phrase is flagged: a regression line alone
compares averages across observed groups; it does not identify what would
happen if a given pitch *changed* its request. The caveat box at the end of
the lecture collects the full scope statement.

### Errors versus residuals

The population line is a model statement; nobody has seen it. What we
compute from data is the fitted sample line

$$
\widehat Y_i=b_0+b_1X_i,
$$

and, for each observation, the **residual** --- the vertical difference

$$
e_i=Y_i-\widehat Y_i.
$$

The error $\varepsilon_i$ and the residual $e_i$ play parallel roles in the
two equations but are different objects, as @fig-i01-error-residual shows.

```{r}
#| label: fig-i01-error-residual
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 8
#| fig-height: 3.6
#| fig-cap: "The same seven observations twice. Left: the unknown population line with the errors ε (distances to the true line). Right: the fitted least-squares line with the residuals e (distances to the estimated line). Only the right panel can be computed from data."
library(ggplot2)
set.seed(101)
x_er <- c(1, 2.2, 3.4, 4.6, 5.8, 7.0, 8.2)
beta0_true <- 1; beta1_true <- 0.8
y_er <- beta0_true + beta1_true * x_er + c(1.3, -1.5, 0.9, -0.8, 1.6, -1.2, 0.4)
fit_er <- lm(y_er ~ x_er)

panel_names <- c("Population line (unknown)", "Fitted line (computed)")
er_df <- data.frame(
  panel = factor(rep(panel_names, each = 7), levels = panel_names),
  x = rep(x_er, 2),
  y = rep(y_er, 2),
  line_y = c(beta0_true + beta1_true * x_er, fitted(fit_er))
)
line_df <- data.frame(
  panel = factor(panel_names, levels = panel_names),
  intercept = c(beta0_true, coef(fit_er)[1]),
  slope = c(beta1_true, coef(fit_er)[2])
)
label_df <- data.frame(
  panel = factor(panel_names, levels = panel_names),
  x = x_er[5] + 0.45,
  y = c((y_er[5] + beta0_true + beta1_true * x_er[5]) / 2,
        (y_er[5] + fitted(fit_er)[5]) / 2),
  lab = c("epsilon[i]", "e[i]")
)

ggplot(er_df, aes(x = x, y = y)) +
  geom_abline(data = line_df[1, ],
              aes(intercept = intercept, slope = slope),
              colour = "#5B6976", linewidth = 0.9, linetype = "dashed") +
  geom_abline(data = line_df[2, ],
              aes(intercept = intercept, slope = slope),
              colour = "#D55E00", linewidth = 0.9) +
  geom_segment(aes(xend = x, yend = line_y),
               colour = "#0072B2", linewidth = 0.8) +
  geom_point(size = 2.6, colour = "#009E73") +
  geom_text(data = label_df, aes(x = x, y = y, label = lab),
            parse = TRUE, colour = "#0072B2", size = 4) +
  coord_cartesian(xlim = c(0, 9.4), ylim = c(0, 9.4)) +
  facet_wrap(~ panel) +
  labs(x = "Predictor X", y = "Outcome Y") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

- $\varepsilon_i = Y_i - (\beta_0+\beta_1X_i)$ measures the distance to the
  **population** line. Because $\beta_0$ and $\beta_1$ are unknown,
  $\varepsilon_i$ is never observed.
- $e_i = Y_i - (b_0+b_1X_i)$ measures the distance to the **fitted** line.
  It is computed from data, one number per observation.

Residuals are our window on the errors, but they are estimates, not the
thing itself --- Lecture 2 lives entirely in that window.

::: {.callout-note collapse="true"}
## Check: which object is observable?

Only the residual $e_i$. The observed data give $Y_i$ and $X_i$; the fitted
line gives $\widehat Y_i$, so $e_i = Y_i - \widehat Y_i$ is a computable
number. The error $\varepsilon_i$ depends on the unknown population
coefficients $\beta_0$ and $\beta_1$, so it can be reasoned about but never
calculated.
:::

### Least squares: why vertical, why squared

For observation $i$, the model predicts $\widehat Y_i$ at its observed
$X_i$. The residual $e_i=Y_i-\widehat Y_i$ is therefore a difference in the
**outcome's units** --- thousands of dollars of agreed amount, not distance
on the page. A positive residual lies above the line; a negative residual
lies below it.

Adding raw residuals is not a useful score: positive and negative misses
can cancel. Adding absolute residuals avoids cancellation, but the resulting
objective has sharp corners and is harder to optimise algebraically. Squared
residuals are non-negative, penalise large misses more strongly, and give a
smooth objective with a tractable solution. This is a modelling choice, not
a claim that squared error is the only possible loss function.

Ordinary least squares chooses $b_0$ and $b_1$ to minimise

$$
\operatorname{SSE}=\sum_{i=1}^{n}e_i^2
=\sum_{i=1}^{n}(Y_i-b_0-b_1X_i)^2.
$$

The initials **OLS** stand for **ordinary least squares**: the rule that
fits the linear model by minimising this squared-error criterion.

#### Try the criterion yourself

Drag either square handle. The vertical segments are residuals; the
displayed SSE updates after every move. In class the widget stays open over
the break with a target SSE to beat; then reveal the line with the smallest
SSE.

<style>
.ls-lab {
  background: #fff;
  border: 1px solid #d5dbe0;
  border-radius: 8px;
  margin: 1.2em 0 1.5em;
  padding: 1em;
}
.ls-lab-heading,
.ls-actions {
  align-items: center;
  display: flex;
  flex-wrap: wrap;
  gap: 0.8em 1.2em;
  justify-content: space-between;
}
.ls-lab svg {
  display: block;
  height: auto;
  margin: 0.5em auto;
  max-width: 700px;
  touch-action: none;
  width: 100%;
}
.ls-controls {
  display: grid;
  gap: 0.8em 1.4em;
  grid-template-columns: repeat(2, minmax(0, 1fr));
  margin: 0.6em 0 1em;
}
.ls-controls label {
  color: #17324d;
  font-size: 0.9em;
  font-weight: 600;
}
.ls-controls input {
  display: block;
  margin-top: 0.3em;
  width: 100%;
}
.ls-actions button {
  background: #17324d;
  border: 1px solid #17324d;
  border-radius: 4px;
  color: #fff;
  cursor: pointer;
  font: inherit;
  padding: 0.45em 0.8em;
}
.ls-actions button + button {
  background: #fff;
  color: #17324d;
}
.ls-caption {
  color: #53606b;
  font-size: 0.9em;
  margin-bottom: 0;
}
.ls-axis { stroke: #65727e; stroke-width: 1.2; }
.ls-tick { fill: #53606b; font-size: 13px; }
.ls-label { fill: #263b4b; font-size: 15px; }
.ls-point { fill: #0072b2; stroke: #17324d; stroke-width: 1; }
.ls-fit-line { stroke: #d55e00; stroke-width: 3; }
.ls-residual { stroke: #755f93; stroke-width: 2; stroke-dasharray: 4 3; }
.ls-handle { fill: #f0e442; stroke: #17324d; stroke-width: 2; cursor: ns-resize; }
.ls-handle:focus { outline: 3px solid #56b4e9; outline-offset: 2px; }
@media (max-width: 767px) {
  .ls-controls {
    grid-template-columns: 1fr;
  }
}
</style>

<div id="least-squares-lab" class="ls-lab" role="group" aria-labelledby="ls-lab-title">
<div class="ls-lab-heading">
<strong id="ls-lab-title">Least-squares line lab</strong>
<span id="ls-equation" aria-live="polite"></span>
</div>
<svg id="ls-svg" viewBox="0 0 700 390" role="img" aria-label="Scatterplot with a draggable line and vertical residual segments"></svg>
<div class="ls-controls">
<label>Line height at x = 1
<input id="ls-left" type="range" min="0" max="10" step="0.05" value="3">
</label>
<label>Line height at x = 9
<input id="ls-right" type="range" min="0" max="10" step="0.05" value="7">
</label>
</div>
<div class="ls-actions">
<button id="ls-best" type="button">Show least-squares fit</button>
<button id="ls-reset" type="button">Try another line</button>
<strong id="ls-sse" aria-live="polite"></strong>
</div>
<p class="ls-caption">Squares make every contribution to SSE non-negative. Longer residuals contribute disproportionately because their lengths are squared.</p>
</div>

<script>
(function () {
  const root = document.getElementById("least-squares-lab");
  if (!root || root.dataset.ready) return;
  root.dataset.ready = "true";
  const svg = root.querySelector("#ls-svg");
  const left = root.querySelector("#ls-left");
  const right = root.querySelector("#ls-right");
  const equation = root.querySelector("#ls-equation");
  const sseText = root.querySelector("#ls-sse");
  const points = [
    {x:1, y:1.8}, {x:2, y:1.4}, {x:3, y:3.2},
    {x:4, y:4.8}, {x:5, y:4.2}, {x:6, y:6.8},
    {x:7, y:6.4}, {x:8, y:8.6}, {x:9, y:8.2}
  ];
  const ns = "http://www.w3.org/2000/svg";
  const box = {left:58, right:680, top:18, bottom:336};
  const sx = x => box.left + x / 10 * (box.right - box.left);
  const sy = y => box.bottom - y / 10 * (box.bottom - box.top);
  const add = (name, attrs, parent=svg) => {
    const el = document.createElementNS(ns, name);
    Object.entries(attrs).forEach(([k,v]) => el.setAttribute(k, v));
    parent.appendChild(el); return el;
  };
  add("line", {x1:box.left, y1:box.bottom, x2:box.right, y2:box.bottom, class:"ls-axis"});
  add("line", {x1:box.left, y1:box.top, x2:box.left, y2:box.bottom, class:"ls-axis"});
  [0,2,4,6,8,10].forEach(v => {
    add("line", {x1:sx(v), y1:box.bottom, x2:sx(v), y2:box.bottom+6, class:"ls-axis"});
    const tx=add("text", {x:sx(v), y:box.bottom+25, class:"ls-tick"}); tx.textContent=v;
    add("line", {x1:box.left-6, y1:sy(v), x2:box.left, y2:sy(v), class:"ls-axis"});
    const ty=add("text", {x:box.left-12, y:sy(v)+4, class:"ls-tick", "text-anchor":"end"}); ty.textContent=v;
  });
  const xl=add("text", {x:(box.left+box.right)/2, y:382, class:"ls-label", "text-anchor":"middle"}); xl.textContent="Predictor x";
  const yl=add("text", {x:16, y:(box.top+box.bottom)/2, class:"ls-label", transform:`rotate(-90 16 ${(box.top+box.bottom)/2})`, "text-anchor":"middle"}); yl.textContent="Outcome y";
  const residualGroup=add("g", {class:"ls-residuals"});
  const line=add("line", {class:"ls-fit-line"});
  const handleLeft=add("rect", {width:16, height:16, rx:2, class:"ls-handle", "aria-hidden":"true"});
  const handleRight=add("rect", {width:16, height:16, rx:2, class:"ls-handle", "aria-hidden":"true"});
  points.forEach(p => add("circle", {cx:sx(p.x), cy:sy(p.y), r:5.5, class:"ls-point"}));
  const meanX=points.reduce((a,p)=>a+p.x,0)/points.length;
  const meanY=points.reduce((a,p)=>a+p.y,0)/points.length;
  const bestSlope=points.reduce((a,p)=>a+(p.x-meanX)*(p.y-meanY),0) /
                  points.reduce((a,p)=>a+(p.x-meanX)**2,0);
  const bestIntercept=meanY-bestSlope*meanX;
  const fitFromHeights = () => {
    const y1=Number(left.value), y9=Number(right.value);
    const slope=(y9-y1)/8, intercept=y1-slope;
    return {intercept, slope};
  };
  function draw() {
    const {intercept, slope}=fitFromHeights();
    residualGroup.replaceChildren();
    let sse=0;
    points.forEach(p => {
      const fitted=intercept+slope*p.x;
      sse+=(p.y-fitted)**2;
      add("line", {x1:sx(p.x), y1:sy(p.y), x2:sx(p.x), y2:sy(fitted), class:"ls-residual"}, residualGroup);
    });
    line.setAttribute("x1", sx(0)); line.setAttribute("y1", sy(intercept));
    line.setAttribute("x2", sx(10)); line.setAttribute("y2", sy(intercept+10*slope));
    [[handleLeft,1,Number(left.value)],[handleRight,9,Number(right.value)]].forEach(([h,x,y])=>{
      h.setAttribute("x",sx(x)-8); h.setAttribute("y",sy(y)-8);
    });
    equation.textContent=`ŷ = ${intercept.toFixed(2)} ${slope<0?"−":"+"} ${Math.abs(slope).toFixed(2)}x`;
    sseText.textContent=`SSE = ${sse.toFixed(2)}`;
  }
  function drag(handle, slider) {
    handle.addEventListener("pointerdown", event => {
      handle.setPointerCapture(event.pointerId);
      const move = e => {
        const rect=svg.getBoundingClientRect();
        const svgY=(e.clientY-rect.top)*390/rect.height;
        const y=Math.max(0,Math.min(10,(box.bottom-svgY)*10/(box.bottom-box.top)));
        slider.value=y.toFixed(2); draw();
      };
      handle.addEventListener("pointermove", move);
      handle.addEventListener("pointerup", () => handle.removeEventListener("pointermove", move), {once:true});
    });
  }
  drag(handleLeft,left); drag(handleRight,right);
  root.querySelector("#ls-best").addEventListener("click",()=>{
    left.value=(bestIntercept+bestSlope).toFixed(2);
    right.value=(bestIntercept+9*bestSlope).toFixed(2); draw();
  });
  root.querySelector("#ls-reset").addEventListener("click",()=>{
    left.value=3; right.value=7; draw();
  });
  draw();
})();
</script>

The line that minimises SSE has closed-form coefficients. For simple
regression,

$$
b_1=\frac{s_{XY}}{s_X^2},
\qquad
b_0=\bar y-b_1\bar x,
$$

where $s_{XY}$ is the sample covariance of $X$ and $Y$ and $s_X^2$ is the
sample variance of $X$. The second formula rearranges to
$\bar y = b_0 + b_1\bar x$: the fitted line always passes through the point
of means $(\bar x,\bar y)$.

::: {#exm-i01-toy}
## Worked example: least squares by hand on three points

Three observations: $(1,2)$, $(2,5)$, $(3,5)$. Every sum written out.

**Means.** $\bar x = (1+2+3)/3 = 2$ and $\bar y = (2+5+5)/3 = 4$.

**Deviations and their products.**

| $i$ | $x_i-\bar x$ | $y_i-\bar y$ | $(x_i-\bar x)(y_i-\bar y)$ | $(x_i-\bar x)^2$ |
|---|---:|---:|---:|---:|
| 1 | $-1$ | $-2$ | $2$ | $1$ |
| 2 | $0$ | $1$ | $0$ | $0$ |
| 3 | $1$ | $1$ | $1$ | $1$ |
| **Sum** | | | $3$ | $2$ |

**Covariance and variance.** Dividing both sums by $n-1=2$:
$s_{XY} = 3/2 = 1.5$ and $s_X^2 = 2/2 = 1$.

**Slope and intercept.**

$$
b_1 = \frac{s_{XY}}{s_X^2} = \frac{1.5}{1} = 1.5,
\qquad
b_0 = \bar y - b_1\bar x = 4 - 1.5\times 2 = 1.
$$

The fitted line is $\widehat y = 1 + 1.5x$. @fig-i01-toy confirms the third
fact: the line passes exactly through $(\bar x,\bar y)=(2,4)$. The residuals
are $-0.5$, $+1$, $-0.5$ --- and they sum to zero, as least-squares
residuals always do when the model has an intercept.
:::

```{r}
#| label: fig-i01-toy
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 5.4
#| fig-height: 4.2
#| fig-cap: "The three-point toy dataset with its least-squares line. The open diamond marks the point of means (2, 4), which the fitted line passes through exactly."
library(ggplot2)
x_toy <- c(1, 2, 3); y_toy <- c(2, 5, 5)
toy_fit <- lm(y_toy ~ x_toy)
toy_df <- data.frame(x = x_toy, y = y_toy, fitted = fitted(toy_fit))

ggplot(toy_df, aes(x = x, y = y)) +
  geom_abline(intercept = coef(toy_fit)[1], slope = coef(toy_fit)[2],
              colour = "#D55E00", linewidth = 0.9) +
  geom_segment(aes(xend = x, yend = fitted),
               colour = "#0072B2", linewidth = 0.8, linetype = "dashed") +
  geom_point(size = 3, colour = "#009E73") +
  annotate("point", x = 2, y = 4, shape = 5, size = 3.4, stroke = 1.1,
           colour = "#17324D") +
  annotate("text", x = 2.34, y = 3.72,
           label = "'('*bar(x)*', '*bar(y)*')'", parse = TRUE,
           colour = "#17324D", size = 3.4) +
  coord_cartesian(xlim = c(0.5, 3.5), ylim = c(1, 6)) +
  labs(x = "x", y = "y") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

You can verify the hand calculation in one line:

```{r}
#| label: i01-toy-check
coef(lm(c(2, 5, 5) ~ c(1, 2, 3)))
```

### Fit and read the Shark Tank model in R

Now the real data. The construction is repeated so this block runs on its
own --- it is the same subset and the same $1,000-unit columns as the
warm-up:

```{r}
#| label: i01-fit
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000

deal_model <- lm(deal_1k ~ ask_1k, data = deals)
coef(deal_model)
```

The fitted equation is approximately

$$
\widehat{\text{deal amount}}
=27.571+1.0085\times\text{amount requested},
$$

with both monetary variables measured in $1,000s.

**The slope, with units.** One additional \$1,000 requested is associated
with about \$1,008.50 more in the fitted agreed amount among pitches that
reached an on-air deal --- close to dollar-for-dollar. It does not mean that
asking for more causes investors to offer more.

**The intercept, with units.** The intercept predicts about \$27,571 at a
request of zero. A zero request lies outside the observed range (the
smallest request in the data is \$10,000) and is not substantively
meaningful; an intercept can be mathematically necessary without having a
useful standalone interpretation.

Observed, fitted, and residual for the first rows:

```{r}
#| label: i01-fitted-residual
head(data.frame(
  observed = deals$deal_1k,
  fitted = fitted(deal_model),
  residual = residuals(deal_model)
))
```

For every row, `observed = fitted + residual`. Check it on the first row:
the pitch asked \$50,000 and agreed \$50,000; the model fits
$27.571 + 1.0085 \times 50 = 77.996$, and indeed
$50 = 77.996 + (-27.996)$.

### Prediction inside the evidence boundary

```{r}
#| label: i01-one-prediction
predict(deal_model, newdata = data.frame(ask_1k = 300))
```

For a \$300,000 request (`ask_1k = 300`), the fitted mean agreed amount is
about 330.12, that is, \$330,124. This is a model-based average, not a
guarantee for a new pitch. If such a pitch actually agreed \$250,000, its
residual would be

$$
e = 250 - 330.12 = -80.12,
$$

about \$80,124 **below** the fitted conditional mean. Predictions are
defensible only inside the range of requests the data contain; extrapolating
far beyond it would rely almost entirely on the assumed line.

::: {.caveat-box}
<span class="label">What this fitted line can and cannot say</span>

The slope 1.0085 describes an **association** among the 882 televised
pitches that reached an on-air deal. Three limits apply.

- **Association, not causation.** Requests were not randomly assigned.
  Pitches asking for more differ in many other ways --- stage, industry,
  revenue --- that also move agreed amounts.
- **Selection into deals.** The analysis conditions on reaching a deal.
  Pitches without a deal are structurally excluded, so the line describes
  deal records, not all pitches, and certainly not all start-ups.
- **No advice about changing an ask.** The model does not say that a
  founder who raises their request by \$1,000 will walk away with \$1,008.50
  more. That is a claim about an intervention, which this observational
  comparison cannot support.
:::

### Answering the driving question

Among pitches that reached a deal: yes, asking for more goes together with
getting more, almost one-for-one. The fitted slope is 1.0085, so each extra
\$1,000 requested is associated with about \$1,008.50 more agreed on
air --- about \$1.01 per dollar asked --- with the caveat box's limits
attached.

**Next:** We now have a fitted line and can read its slope in money terms. But
least squares will draw an equally confident line through terrible data ---
the software never warns you. Lecture 2 lays out the assumptions under which
the line means what we claimed (linearity of the average, equal spread,
independence, and design conditions no plot can certify) and teaches the
diagnostics: reading residual plots, and spotting the single influential pitch
that can bend an entire fit.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 60 minutes.</span>

### Retrieval (10 minutes)

1. What is the difference between $\varepsilon_i$ and $e_i$?
2. What does least squares minimise?
3. Why must a slope interpretation include units?
4. Why can an intercept be mathematically valid but substantively unhelpful?
5. Why is the slope not automatically causal?

::: {.callout-note collapse="true"}
## Check your answers

1. $\varepsilon_i$ is an unobserved population-model error (distance to the
   unknown population line); $e_i$ is the observed residual after fitting
   the sample line.
2. The sum of squared residuals, $\operatorname{SSE}=\sum_i e_i^2$.
3. The slope is change in outcome units per one predictor unit; without
   units the number 1.0085 could mean anything.
4. The value $X=0$ may lie outside the observed range or lack substantive
   meaning.
5. The predictor was not randomly assigned and other factors may affect both
   variables.
:::

### From worked to faded (20 minutes)

Re-read the prediction calculation in the chapter (fitted value at
`ask_1k = 300`). Then complete the same calculation for a new request.

*Step 1 (given).* A pitch requests \$400,000, so `ask_1k = 400`. The fitted
equation is $\widehat{\text{deal}} = 27.571 + 1.0085 \times \text{ask}$.

*Step 2 (you).* Compute the fitted agreed amount, in $1,000s and in dollars.

*Step 3 (you).* The pitch actually agreed \$350,000. Compute the residual
and interpret its sign in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

Fitted: $27.571 + 1.0085 \times 400 = 430.97$, about \$430,975. Residual:
$350 - 430.97 = -80.97$, about $-\$80{,}975$. The negative sign means this
pitch agreed less than the model's fitted conditional mean for a \$400,000
request.

```{r}
#| label: i01-faded-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
unname(350 - predict(deal_model, newdata = data.frame(ask_1k = 400)))
```
:::

Then the unassisted set:

1. Use the fitted equation to predict the agreed amount for requests of
   \$100,000 and \$500,000.
2. Calculate the residual for a pitch that requested \$300,000 and agreed
   \$250,000.
3. Explain why `lm(y ~ x, data = df)` treats the variable left of `~` as
   the outcome.

::: {.callout-note collapse="true"}
## Check your answers

1. Substitute `ask_1k = 100` and `ask_1k = 500`; the fitted values are about
   128.42 and 531.83 in $1,000s --- \$128,422 and \$531,826.
2. The fitted amount is about 330.12 in $1,000s, so the residual is
   $250-330.12\approx-80.12$, or about $-\$80{,}124$.
3. R formulas use `outcome ~ predictors`: the left-hand side is what the
   model predicts, the right-hand side is what it predicts from.
:::

### Practice: a regression of your own (15 minutes)

Fit `equity_offered_pct ~ season` on the full dataset (all 1,441 pitches)
and interpret both coefficients with units. State the unit of every quantity
you report.

::: {.callout-note collapse="true"}
## Check your answers

```{r}
#| label: i01-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
coef(lm(equity_offered_pct ~ season, data = sharks))
```

The fitted intercept is about 20.46 percentage points and the slope about
$-0.824$ percentage points per season. Later seasons are associated with
smaller initial equity offers on average. The outcome and intercept use
percentage points; the predictor uses seasons; the slope uses percentage
points per season.
:::

### Spaced review: one pentad item from QM Basic (5 minutes)

A market survey of $n = 400$ customers gives an average willingness to pay
of €24.60 with standard error €1.10. Write down the t value for testing a
benchmark of €22 and the 95% confidence interval, and say in one sentence
what the interval estimates.

::: {.callout-note collapse="true"}
## Check your answer

$t = (24.60-22)/1.10 \approx 2.36$; the interval is
$24.60 \pm 1.96\times1.10 = (22.44,\; 26.76)$ euros. It is a range of
plausible values for the population mean willingness to pay, under the
sampling model from QM Basic.
:::

### Worked exam interpretation (5 minutes)

**Question** (mock exam Question 1 pattern). A model of venture sales
growth (percentage points) on marketing spending (€10,000s) reports
$\widehat{\text{growth}} = 6.60 + 1.38\,\text{spending}$. Interpret the
slope with units, and calculate the fitted value and residual for a venture
spending €30,000 with observed growth of 15 percentage points.

::: {.callout-note collapse="true"}
## Check a model answer

Each additional €10,000 of marketing spending is associated with 1.38
percentage points higher fitted sales growth among sampled ventures.
€30,000 is `spending = 3`, so the fitted value is
$6.60+1.38\times3=10.74$ percentage points; the residual is
$15-10.74=4.26$ percentage points --- observed growth exceeds the fitted
value.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well. Name one
numerical outcome and one numerical input you suspect drives it, and write
two sentences: first state the slope you would expect (sign and units), then
say what decision would change if the fitted slope turned out to be near
zero. There is no wrong answer; the point is connecting today's tool to
something you care about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Outcome: weekly orders; input: euros of social-media spend that week --- I expect a positive slope of a few orders per extra €100. If the fitted slope came out near zero, I would move the whole budget to campus flyers."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it whether your
outcome-input pair is more likely to suffer from selection (like analysing
only deals) or from reverse causation.
::::

Continue to Lecture 2: What the model assumes and how residuals betray it.

Formula sheet: Lecture 1 formulas.

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.1 and 3.6.2, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 19.

:::
