# Lecture 2: What the model assumes and how residuals reveal it {#i02}

```{r}
#| label: i02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 2, you can</span>

- explain what the main assumptions of a regression are needed for
- check the main assumptions of a regression with residual plots
- say which assumptions need knowledge of the data rather than plots
- find influential observations and handle them openly and carefully
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">R draws a fitted line in the same way through good data and through bad data. Whether you can trust the line depends on assumptions that no software checks for you.</p>

The fitted line from Lecture 1 looks convincing. Under what conditions
should we trust it, and what do its residuals reveal?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">No video this week</span><br>
The preparation for this lecture is your own residual plot. Do the R warm-up
below and bring the plot to class.
</div>

**Predict.** Answer this question before you come to class: the agreed valuations are strongly
right-skewed (most are small and a few are very large), far from a normal
distribution. Does that alone make the
Lecture 1 regression invalid, **yes** or **no**? Choose one answer before
class and write one sentence explaining why. The lecture material gives the
correct answer.

[Shark Tank data]{.shark-tag} **R warm-up.** Rebuild the Lecture 1 model and make its
residual-versus-fitted plot: the fitted values on the horizontal axis and the
residuals on the vertical axis. Bring the plot to class:

```{r}
#| label: i02-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
plot(fitted(val_model), residuals(val_model))
abline(h = 0, lty = 2)
```

**Code along (bring your laptop).** Bring a charged laptop with RStudio and
the course project open, so that you can run the code of this chapter
yourself during the lecture and get the same numbers and figures.

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section states what the regression model assumes. It then shows how to
check these assumptions with the residuals of the Lecture 1 model. It also
shows which assumptions no plot can check.

### Start with a founder's question

[Shark Tank data]{.shark-tag} Lecture 1 gave you a number: the sharks accept about 65 cents of every
claimed dollar. Before you plan a negotiation around this number, ask whether you
can trust the line behind it. R draws a line through any data, even through
data that do not support a line. R never warns you.

### The residual plot from the warm-up

[Shark Tank data]{.shark-tag} The code below is the warm-up code. It rebuilds the data and the model of
Lecture 1: the valuation agreed on air (`agreed_m`) regressed on the
valuation claimed by the founder (`claimed_m`), both in \$ millions, for the
874 Shark Tank deals for a share of the company.

```{r}
#| label: i02-build-model
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
```

```{r}
#| label: fig-i02-warmup-residuals
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Residuals against fitted values for the Lecture 1 valuation model, both in $ millions. On the left the residuals lie close to zero. On the right they spread over several million dollars."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
resid_df <- data.frame(
  fitted = fitted(val_model),
  residual = residuals(val_model)
)

ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  annotate("text", x = 31.5, y = 0.9, label = "residual = 0",
           hjust = 1, colour = "#17324D", size = 3.4) +
  labs(x = "Fitted agreed valuation ($ millions)",
       y = "Residual ($ millions)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

::: {.definition-box}
<span class="label">Definition</span>

**Residual plot** (residual-versus-fitted plot). A scatterplot with the
fitted values $\widehat Y_i$ on the horizontal axis and the residuals $e_i$
on the vertical axis. It shows how the misses of the line (the residuals) change from the
left to the right of the plot.
:::

In @fig-i02-warmup-residuals, both axes are in the units of the outcome,
\$1 million. A point above zero is a deal that closed at a higher valuation
than its fitted value. A point below zero is a deal that closed at a lower valuation than its
fitted value. To
read patterns in this plot, you first need to know what the model assumes.

### What each assumption supports

The model of Lecture 1 is

$$
Y_i=\beta_0+\beta_1X_i+\varepsilon_i.
$$

The model depends on several separate assumptions. Each assumption supports
a different part of the analysis.

::: {.definition-box}
<span class="label">Definition</span>

**Model assumptions.** The conditions under which we can trust the fitted
line and its standard errors:

- **Linearity.** The average of $Y$ at each value of $X$ lies on a straight
  line.
- **Independence.** The observations do not depend on each other. Every row
  brings new information.
- **Constant variance.** The errors spread equally at every value of $X$.
- **Normal errors.** At each value of $X$, the errors follow a normal
  distribution. This assumption matters only in small samples.
:::

The table repeats these four assumptions in symbols and adds two more
conditions. Without variation in $X$, we cannot calculate a slope. And to read
the slope as an effect, we need one more condition: no confounding. A
**confounder** is a variable that is not in the model and that affects both
$X$ and $Y$.

| Condition | Meaning | What it supports |
|---|---|---|
| Linear conditional mean | $E[Y\mid X=x]=\beta_0+\beta_1x$ over the range of the data | One constant slope summarises how the average of $Y$ changes with $X$ |
| No confounding (for a causal reading only) | Nothing that is missing from the data affects both $X$ and $Y$ | Reading the slope as the effect of changing $X$. Describing how average outcomes differ across values of $X$ does not need it |
| Independent observations | The errors of different rows are unrelated, so no row repeats the information of another | The usual standard-error formula |
| Variation in $X$ | Not all observations have the same predictor value | A slope can be calculated |
| Constant conditional variance | $\operatorname{Var}(\varepsilon\mid X)=\sigma^2$, the same at every value of $X$ | The usual standard-error formula |
| Conditional normality | At each value of $X$, $\varepsilon$ follows a normal distribution | Exact t and F distributions in small samples |

No row of the table says that the data are normal. This answers the prediction
question from before class: the outcome does not need to be normally
distributed, and neither does the predictor. The agreed valuations are
strongly right-skewed, and this skew alone does not break any assumption.
The assumptions are about the average of $Y$ at each $X$, about the errors,
and about how the data were produced. They are not about the histogram of $Y$ or the histogram of
$X$.

### Linearity is about the average, not every point

Individual outcomes do not have to lie on a line. Linearity says that the
average outcome at each predictor value follows a line. If the residual
plot shows a curve, one constant slope misses part of that average pattern.

Constant variance is about the vertical spread. It says that the variance of
the errors is the same at each predictor value. It does not say that all
residuals have the same size. The technical name for constant variance is
**homoskedasticity**, and for changing variance **heteroskedasticity**. In a
residual plot, changing variance looks like a fan: the vertical band of points
widens or narrows from left to right. @fig-i02-assumption-panels shows both
problems.

```{r}
#| label: fig-i02-assumption-panels
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 6
#| fig-cap: "Four simulated situations. Top row: a linear mean fits (left, outcome against predictor), a curved mean leaves a U-shaped residual pattern (right). Bottom row: constant vertical spread (left), spread growing with the fitted value, a fan (right). The three residual panels share the gold dashed line at residual zero."
library(ggplot2)
set.seed(802)
n_demo <- 90
x_demo <- runif(n_demo, 0, 10)

y_linear <- 2 + 0.65 * x_demo + rnorm(n_demo, 0, 0.8)
y_curved <- 2 + 0.10 * x_demo + 0.09 * x_demo^2 + rnorm(n_demo, 0, 0.55)
m_curved <- lm(y_curved ~ x_demo)
m_linear <- lm(y_linear ~ x_demo)

fit_demo <- seq(1, 9, length.out = n_demo)
res_constant <- rnorm(n_demo, 0, 0.75)
res_fan <- rnorm(n_demo, 0, 0.18 + 0.13 * fit_demo)

panel_names <- c("Linear mean (outcome)", "Curved mean (residuals)",
                 "Constant spread (residuals)", "Changing spread (residuals)")
demo_df <- data.frame(
  panel = factor(rep(panel_names, each = n_demo), levels = panel_names),
  x = c(x_demo, fitted(m_curved), fit_demo, fit_demo),
  y = c(y_linear, residuals(m_curved), res_constant, res_fan)
)
line_df <- data.frame(
  panel = factor(panel_names[1], levels = panel_names),
  intercept = coef(m_linear)[1],
  slope = coef(m_linear)[2]
)
zero_df <- data.frame(
  panel = factor(panel_names[-1], levels = panel_names),
  yintercept = 0
)
zero_lab <- data.frame(
  panel = factor(panel_names[-1], levels = panel_names),
  x = c(min(fitted(m_curved)), 1, 1),
  y = c(max(residuals(m_curved)), max(res_constant), max(res_fan))
)

ggplot(demo_df, aes(x = x, y = y)) +
  geom_hline(data = zero_df, aes(yintercept = yintercept),
             colour = "#C98B2E", linetype = "dashed", linewidth = 0.8) +
  geom_abline(data = line_df, aes(intercept = intercept, slope = slope),
              colour = "#D55E00", linewidth = 0.9) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  geom_text(data = zero_lab, aes(x = x, y = y, label = "residual = 0"),
            hjust = 0, colour = "#17324D", size = 3) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "Predictor (top left) or fitted value",
       y = "Outcome (top left) or residual") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

::: {.callout-note collapse="true"}
## Check: match each residual pattern to the assumption it challenges

A U-shaped (curved) residual pattern challenges the linear
conditional mean: the average residual is not zero everywhere, so a
straight line misses part of the mean relationship. A fan challenges
constant conditional variance: the vertical spread of residuals changes
with the fitted value. A real plot can show both problems at once.
:::

### Reading the valuation model's vertical spread

[Shark Tank data]{.shark-tag} Now look for a fan in @fig-i02-warmup-residuals. At small fitted values the
residuals stay close to the zero line. At larger fitted values they spread
over several million dollars. A short calculation confirms what the plot
shows. The code sorts the deals by the claimed valuation into five groups.
Each group has about a fifth of the deals. The code then gives the standard
deviation of the residuals in each group:

```{r}
#| label: i02-spread-by-fifth
fifth <- cut(vals$claimed_m,
             breaks = quantile(vals$claimed_m, probs = seq(0, 1, 0.2)),
             include.lowest = TRUE)
round(tapply(residuals(val_model), fifth, sd), 2)
```

The standard deviation of the residuals is about \$0.14 million in the fifth
of deals with the smallest claimed valuations and about \$3.6 million in the
fifth with the largest, about 25 times as much. The spread of agreed
valuations grows with the size of the claim. This makes economic sense: founders
and sharks can disagree by much more money about a \$20 million company than
about a \$200,000 company. This pattern is a fan. The usual standard errors
assume one common variance $\sigma^2$, so here they depend on an assumption
that probably does not hold. Lecture 3 returns to
this point.

::: {.callout-caution collapse="true"}
## Try it: annotate a residual plot

[Shark Tank data]{.shark-tag} Try this with your neighbour first. Use a printed copy or a sketch of
@fig-i02-warmup-residuals.

*Step 1 (given).* The horizontal dashed line marks residual zero. Points
above it are deals with an agreed valuation above the fitted value. Points
below it are deals with an agreed valuation below the fitted value.

*Step 2 (you).* Draw two rough curves around the residuals, one along
the top of the points and one along the bottom, from the left edge to the
right edge of the plot. Does the band between the two curves stay equally
wide?

*Step 3 (you).* Name the assumption that your two curves challenge. Then
write one sentence that says whether the average of the residuals also
moves away from zero in some part of the plot.

::: {.callout-note collapse="true"}
## Check your answer

The band between the two curves widens strongly from left to right. Deals
with small fitted valuations lie within a few hundred thousand dollars of the
line. Deals with large fitted valuations lie within several million. That fan challenges
constant conditional variance. The
average residual stays near zero across the range, so the plot does not
show a clear violation of the linear conditional mean. You read the same
plot twice, once for each assumption.
:::
:::

### What plots cannot show

Two rows of the assumptions table never appear in any residual plot.

**No confounding.** A curve in the residual plot can show that a straight
line misses the average pattern. No plot can show whether a confounder
exists, that is, whether something that is missing from the data affects
both the predictor and the outcome. Least squares always gives residuals with a mean of zero and no
straight-line relation to $X$. So the residual plot looks clean even when a
confounder exists. Ask how the observations were produced.
Which common causes of $X$ and $Y$ are plausible but missing from the data?

**Independence.** Whether rows share information is a fact about how the
data were collected, not about the residuals. Suppose that rows are related
but you treat them as independent. Then the data seem to contain more
information than they really do, and the standard errors are too small.

Two examples from Shark Tank:

- **Shared episodes.** Pitches in the same episode face the same panel of
  sharks on the same day, and the same few sharks appear in hundreds of
  deals. Their rows share negotiators, moods, and editing decisions. This is
  an **independence** problem: the 874 deals contain somewhat less information
  than 874 unrelated deals would.
- **Network ties.** Suppose founders with strong investor networks both
  claim higher valuations (they know the market) and get better deals (the
  sharks know them). Network strength is not recorded in the data. It is related to
  the predictor, and it affects the outcome. This is a **confounding** problem.
  The slope still describes how average agreed valuations differ across
  claims. But you cannot read the slope as what a higher claim would achieve.

Neither problem is visible in @fig-i02-warmup-residuals. You can find both
problems only if you know how the show works.

### Influence: when a few rows move the line a lot

[Shark Tank data]{.shark-tag} A first check needs no plot. The function `summary()` gives the smallest
residual, the largest residual, the mean, the median, and the two quartiles.
[Basic Lecture 1](../basic/b01-introduction-data-analysis.qmd#b01-quartiles)
defines the median and the quartiles: a quarter of the values lie below the
first quartile, half below the median, and three quarters below the third
quartile.

```{r}
#| label: i02-residual-line
summary(residuals(val_model))
```

::: {#exm-i02-residual-line}
## Worked example: reading the summary of the residuals

[Shark Tank data]{.shark-tag} Read the numbers one by one, all in \$ millions:

- The **median** residual is about $-0.01$: the typical deal lies almost
  exactly on the fitted line.
- The middle half of all residuals, from the first to the third quartile,
  runs from $-0.40$ to $0.17$: a band about \$570,000 wide.
- The **minimum** is about $-7.8$ and the **maximum** about $13.4$: one deal
  closed \$7.8 million below its fitted valuation and another \$13.4 million
  above.

The mean is zero, as it always is when the model has an intercept. If no
deal were extreme, the largest misses would be a few times the width of the
middle half. Here they are 14 to 24 times as large. A few deals miss the line
by far more than the typical deal does, and such deals can strongly change
the fitted line. The same numbers, without the mean, appear at the top of
`summary(val_model)`. Lecture 3 explains this output.
:::

Two standard tools make this check precise. A **standardised residual**,
`rstandard()`, is a residual divided by its standard deviation, so it counts
the miss in standard deviations. Values beyond about $\pm 3$ are large by any
common standard. **Cook's distance**, `cooks.distance()`, measures how much the
whole fitted line would move if one observation were removed. A common
threshold is $4/n$: you look more closely at observations above it.

::: {.definition-box}
<span class="label">Definition</span>

**Outlier.** An observation with an unusually large residual: its outcome
lies far from the fitted line. A common rule is a standardised residual
beyond $\pm 3$.

**Influential observation.** An observation that changes the fitted line
noticeably when it is removed. A common rule is a Cook's distance above
$4/n$.
:::

```{r}
#| label: i02-influence
rs <- rstandard(val_model)
cd <- cooks.distance(val_model)

sum(abs(rs) > 3)               # standardised residuals beyond 3
sum(cd > 4 / nobs(val_model))  # Cook's distance above 4/n
vals[order(cd, decreasing = TRUE)[1:3],
      c("claimed_m", "agreed_m")]
```

The command `nobs(val_model)` gives the number of observations $n$ that the
model uses, here 874. Twenty-five deals have standardised residuals beyond
$\pm 3$, and 64 have a Cook's distance above $4/n$. These are 3% and 7% of
the 874 deals. The most
influential deal claimed a valuation of \$50 million and closed at
\$25 million, half of the claim. @fig-i02-influence marks the 64 deals in the
scatterplot.

```{r}
#| label: fig-i02-influence
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The 874 deals with the fitted line, full range shown. Orange points have a Cook's distance above 4/n. The labelled deal (claimed $50 million, agreed $25 million) moves the fitted line most: it sits far to the right, so it pulls on the line like a lever."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
cd <- cooks.distance(val_model)
vals$flagged <- cd > 4 / nrow(vals)
top <- which.max(cd)

ggplot(vals, aes(x = claimed_m, y = agreed_m)) +
  geom_point(data = subset(vals, !flagged),
             colour = "grey75", alpha = 0.45, size = 1.6) +
  geom_abline(intercept = coef(val_model)[1],
              slope = coef(val_model)[2],
              colour = "#D55E00", linewidth = 0.9) +
  geom_point(data = subset(vals, flagged),
             colour = "#D55E00", size = 2.2) +
  annotate("text", x = vals$claimed_m[top] - 1.2, y = vals$agreed_m[top],
           label = "claimed 50, agreed 25",
           hjust = 1, colour = "#17324D", size = 3.4) +
  labs(x = "Valuation claimed by the founder ($ millions)",
       y = "Valuation agreed on air ($ millions)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

What should you do with a flagged observation? Follow three steps, in this
order:

1. **Investigate.** Is it a data error, or a real pitch? Here the valuations
   of the flagged deals match their recorded amounts and shares, so we treat
   these deals as real.
2. **Report.** State that the results depend on a few observations and say
   which ones. Show the fit with and without them if the results change.
3. **Never silently delete.** If you drop rows that do not fit your story
   and do not say so, you give a false picture of the data.

### Q--Q plots: how to read them, when normality matters

[Shark Tank data]{.shark-tag} The last assumption in the table is conditional normality of the errors. A
**normal Q--Q plot** (quantile-quantile plot) puts the standardised residuals
in order and compares them with the values that a normal distribution would
give. If the residuals are close to normal, the points lie along the
reference line.

```{r}
#| label: fig-i02-qq
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Normal Q--Q plot of the standardised residuals of the valuation model. Both ends leave the line: a few deals closed far above and far below their fitted valuation."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
qq_df <- data.frame(rs = rstandard(val_model))

ggplot(qq_df, aes(sample = rs)) +
  geom_qq_line(colour = "#D55E00", linewidth = 0.9) +
  geom_qq(colour = "#009E73", alpha = 0.4, size = 1.6) +
  labs(x = "Normal theoretical quantile",
       y = "Standardised residual") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The horizontal axis shows where a residual of this rank would lie if the
errors were normal. The vertical axis shows where the residual actually lies. In
@fig-i02-qq the middle of the distribution is close to the line. Both ends
bend away from it, the upper end more strongly. The residuals have more
extreme values on both sides than a normal distribution has. We say
that such a distribution has heavy tails.

Two base R commands draw the same plot without the styling of the figure:

```{r}
#| label: i02-qq-base
qqnorm(rstandard(val_model))
qqline(rstandard(val_model))
```

When does this matter? Normal errors are needed for exact t and F
distributions in small samples. With 874 observations the tests and
intervals of Lecture 3 still hold approximately without normal errors,
because averages over many observations are close to normal. The
approximation is weaker when a few extreme deals decide where the line lies.
Very extreme outliers distort both the line and its
standard errors, in samples of any size.

::: {.caveat-box}
<span class="label">What a residual check can and cannot show</span>

- A residual plot without patterns agrees with the assumptions. It does not
  prove them. Confounders and dependent rows do not show up in a plot.
- A fan or an influential deal is a finding to look into and to report. It is
  not a final judgement on the model.
- Keeping or dropping an unusual observation is a decision that you make from
  knowledge of the data and that you state openly.

**What you can do with it:** before you trust a fitted line, look at its
residual plot and at the few points that move it most. If the conclusion changes
without them, say so.
:::

### Answering the driving question

You can trust the Lecture 1 line only under certain conditions. Its
residuals reveal two problems. The vertical spread grows with the size of the claim (the
fan). And a few extreme deals, flagged by standardised residuals beyond
$\pm3$ and Cook's distances above $4/n$, strongly affect both the line and
its standard errors. Nothing in the plots contradicts a linear conditional mean.
Independence and the absence of confounding cannot be seen in any plot. They
depend on how the show works. Lecture 3 calculates standard errors,
intervals, and tests for this model. These two problems limit how much you
can trust those results.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

One week in your data includes a catering order of 200 meal boxes, in a week with high advertising. With that week the advertising slope is 0.15. Without that week the slope is 0.09. The decision: do not base next term's budget on a fit that depends on one single week. Catering is a different business from selling single meal boxes every day. You report both fits, plan the daily business with the slope of 0.09, and look at catering separately.
:::

**Next:** The slope is an estimate from one set of deals. Lecture 3 shows
how precise it is, with its standard error, its confidence interval, and a
t-test. It also shows how much of the variation in the outcome the line
explains ($R^2$) and how to test whether the model is useful at all (the
F-test).

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions

1. Which assumption concerns the conditional mean?
2. Must the outcome variable be normally distributed? Why or why not?
3. What does a fan shape in a residual-versus-fitted plot suggest?
4. Which assumptions can you not check with a residual plot alone?
5. What do a standardised residual and Cook's distance each measure?
6. What are the three steps for handling a flagged observation?

::: {.callout-note collapse="true"}
## Check your answers

1. The linear conditional mean, $E[Y\mid X=x]=\beta_0+\beta_1x$.
2. No. The assumptions are about the errors at each value of the predictor
   and about how the data were produced. They are not about the
   distribution of $Y$ or of $X$ on its own.
3. Non-constant conditional variance (heteroskedasticity).
4. Independence and the absence of confounding. They require knowledge of
   how the data were collected.
5. A standardised residual measures how unusual one observation's vertical
   miss is, in standard-deviation units. Cook's distance measures how much
   the whole fitted line moves if that observation is removed.
6. Investigate, report, never silently delete.
:::

### Finish a half-worked example: diagnose three residual plots

For each panel of @fig-i02-three-plots, write one sentence: which
assumption, if any, does the plot challenge?

Plot A (given). The residuals follow a U shape. The average residual is
negative in the middle and positive at the ends, so the plot challenges the
linear conditional mean.

Plot B and Plot C (you). Write the sentence for each of them in the same
way.

```{r}
#| label: fig-i02-three-plots
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.2
#| fig-cap: "Three simulated residual-versus-fitted plots. The gold dashed line marks residual zero in each panel."
library(ggplot2)
set.seed(1204)
n_p <- 120
f_p <- seq(1, 9, length.out = n_p)
res_a <- 1.6 * (f_p - 5)^2 / 8 - 1 + rnorm(n_p, 0, 0.55)
res_b <- rnorm(n_p, 0, 0.15 + 0.16 * f_p)
res_c <- rnorm(n_p, 0, 0.8)

plots_df <- data.frame(
  panel = rep(c("Plot A", "Plot B", "Plot C"), each = n_p),
  fitted = rep(f_p, 3),
  residual = c(res_a, res_b, res_c)
)
zero_lab <- data.frame(
  panel = c("Plot A", "Plot B", "Plot C"),
  x = 1,
  y = c(max(res_a), max(res_b), max(res_c))
)

ggplot(plots_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  geom_text(data = zero_lab, aes(x = x, y = y, label = "residual = 0"),
            hjust = 0, colour = "#17324D", size = 3) +
  facet_wrap(~ panel, scales = "free_y") +
  labs(x = "Fitted value", y = "Residual") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

::: {.callout-note collapse="true"}
## Check your answers

Plot B: the vertical band widens with the fitted
value. This fan challenges constant conditional variance. Plot C: no
visible pattern. The plot agrees with linearity and constant variance, the
two assumptions it can show. This result is not proof that all assumptions hold.
:::

### Practice: diagnostics for your own model

[Shark Tank data]{.shark-tag} Fit `equity_offered_pct ~ season` on all 1,441 pitches. Produce the
residual-versus-fitted plot and the normal Q--Q plot, and interpret each in
one or two sentences. Then count the standardised residuals beyond $\pm 3$
and the observations with a Cook's distance above $4/n$.

::: {.callout-note collapse="true"}
## Check your answers

```{r}
#| label: i02-practice-equity
#| fig-height: 3.6
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
equity_model <- lm(equity_offered_pct ~ season, data = sharks)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(equity_model), residuals(equity_model),
     pch = 16, col = grDevices::adjustcolor("#0072B2", alpha.f = 0.30),
     bty = "n", xlab = "Fitted equity offered (pp)",
     ylab = "Residual (pp)")
abline(h = 0, lty = 2, lwd = 2, col = "#D55E00")
qqnorm(rstandard(equity_model), pch = 16,
       col = grDevices::adjustcolor("#009E73", alpha.f = 0.35),
       bty = "n", main = "", xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(equity_model), col = "#D55E00", lwd = 2)
par(old_par)
max(residuals(equity_model))
c(big_rstandard = sum(abs(rstandard(equity_model)) > 3),
  cooks_flagged = sum(cooks.distance(equity_model) > 4 / nobs(equity_model)))
```

The residual plot shows sixteen vertical stripes. The season
(`season`) takes only whole-number values, so only sixteen fitted values
exist. That is a feature of the predictor, not a violation. The vertical
spread is broadly similar across the stripes, so there is no strong fan. The
Q--Q plot bends upwards at the right end: some pitches offered far more
equity than the fitted trend. The largest residual is about 83 percentage
points, from founders who offered almost the whole company. So the normality
assumption probably does not hold. With $n = 1441$ this matters little for
tests and intervals, but you should look more closely at the extreme rows. Eighteen
standardised residuals lie beyond $\pm 3$, and 58 pitches have a Cook's
distance above $4/n$. These are about 1% and 4% of the 1,441 pitches.
:::

### Review: one residual-arithmetic question from Lecture 1

[Shark Tank data]{.shark-tag} Using the Lecture 1 fitted equation
$\widehat{\text{agreed}} = -0.068 + 0.649 \times \text{claimed}$ (both in
\$ millions): a founder claimed a valuation of \$3 million and the deal
closed at \$1.5 million. Compute the fitted value and the residual.

::: {.callout-note collapse="true"}
## Check your answer

Fitted: $-0.068 + 0.649\times3 = 1.88$, about \$1.88 million. Residual:
$1.5 - 1.88 = -0.38$, about $-\$380{,}000$: the deal closed below the
model's fitted conditional mean.
:::

### Worked exam interpretation

**Exam-style question.** An invented example. A regression
describes the monthly revenue of cafés by their opening hours per week. Its
residual-versus-fitted plot shows vertical spread increasing with the fitted
value. Name the pattern and the assumption it challenges, and state one
assumption the plot cannot assess and what information would assess it.

::: {.callout-note collapse="true"}
## Check a model answer

The pattern is a fan: the spread of the residuals grows with the fitted
value. This pattern challenges constant conditional variance
(homoskedasticity). The plot cannot assess independence of observations. To
assess it, you need knowledge of the data collection, for example whether some cafés appear in several rows or
belong to the same chain.
:::

### Check the AI

You asked an AI assistant: "My residuals look strange. What should I do?" It answered:

> "The agreed valuations are not normally distributed, so the regression is invalid. I removed the 25 outliers and now the model is fine."

Find what is wrong and write the corrected answer. This task is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

Two errors. The outcome does not have to be normally distributed. The assumptions are about the errors and about how the data were produced. And you never remove unusual deals without saying so. Investigate them, report the fit with and without them, and say which one you use and why.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Think of a
regression you might one day run on its data (sales on advertising, orders
on price, sign-ups on campaign week). Write two sentences. First, name the
assumption from today's table that you would worry about most for these
data. Second, describe one observation that could dominate the fit, and say
what you would do about it. Investigate and report, do not delete.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student venture that sells boxes of prepared meals on campus. "For orders-on-price data I would worry most about confounding. I cut prices exactly in the weeks when demand is already weak. So weak demand affects both my price and my orders. One catering order of 200 meal boxes would dominate any fit. I would report the model with and without it and investigate whether catering is a different business, not delete the row."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: describe your venture's data
and ask it to argue which assumption is least plausible.
::::

[Continue to Lecture 3: Coefficient inference, fit, and the F-test](i03-inference-fit-ftest.qmd).

[Code sheet: Lecture 2 commands and formulas](code-sheet.qmd#lecture-2-assumptions-residual-diagnostics-and-influence).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.1.2 and 3.3.3, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 19 and 22.

:::
