# Lecture 2: What the model assumes and how residuals betray it {#i02}

```{r}
#| label: i02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 2, you can</span>

- state the assumptions behind ordinary least-squares estimation and
  inference, and say what each one supports;
- explain why the outcome itself does not need to be normally distributed;
- inspect residual plots for non-linearity and unequal variance;
- explain which assumptions no plot can establish, and what design
  knowledge assesses them;
- compute standardised residuals and Cook's distance, and identify
  influential observations;
- read a normal Q--Q plot and say when normality matters;
- respond to an influential observation properly: investigate, report,
  never silently delete.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Every fitted line looks authoritative --- the software draws it with equal confidence through good data and bad. Whether the line deserves trust depends on assumptions no software checks for you.</p>

The fitted line from Lecture 1 looks convincing. Under what conditions
should we trust it, and what do its residuals confess?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch the first half of Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=3GiWpRfkSjc&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=10">3.2
Hypothesis Testing and Confidence Intervals</a>, focusing on what can go
wrong in a regression rather than on the test mechanics --- Lecture 3
handles those. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: can a **steep** fitted slope still be an
**imprecise** estimate? Commit to an answer and write one sentence
explaining why; the lecture resolves it.

**R warm-up (10 minutes).** Rebuild the Lecture 1 model and produce its
residual-versus-fitted plot. Bring the plot to class:

```{r}
#| label: i02-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
plot(fitted(deal_model), residuals(deal_model))
abline(h = 0, lty = 2)
```

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question --- with the class's own warm-up plots on
screen.

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=3GiWpRfkSjc">3.2 Hypothesis Testing and Confidence Intervals</a>.</p>

### The plot you brought

The warm-up code, run once more so every object on this page is built in
the open:

```{r}
#| label: i02-build-model
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
```

```{r}
#| label: fig-i02-warmup-residuals
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Residuals against fitted values for the Lecture 1 deal model, both in $1,000s. Reading this plot --- and knowing what it cannot show --- is the whole of today's lecture."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
resid_df <- data.frame(
  fitted = fitted(deal_model),
  residual = residuals(deal_model)
)

ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  annotate("text", x = 5050, y = 170, label = "residual = 0",
           hjust = 1, colour = "#17324D", size = 3.4) +
  labs(x = "Fitted agreed amount ($1,000s)",
       y = "Residual ($1,000s)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

In @fig-i02-warmup-residuals, fitted outcomes run horizontally and
observed-minus-fitted residuals vertically, both in the outcome's units of
$1,000. A point above zero is a pitch that agreed more than its fitted
value. Before reading patterns out of this cloud, we need to know what the
model actually assumes.

### What each assumption supports

For

$$
Y_i=\beta_0+\beta_1X_i+\varepsilon_i,
$$

separate the conditions below rather than memorising one vague statement
that "the data are normal." Each condition supports a different part of the
analysis --- take the table one row at a time.

| Condition | Meaning | What it supports |
|---|---|---|
| Linear conditional mean | $E[Y\mid X=x]=\beta_0+\beta_1x$ over the modelled range | One constant slope summarises the mean relationship |
| Zero conditional mean | $E[\varepsilon\mid X]=0$ | The slope is not systematically absorbing omitted components related to $X$ |
| Independent observations | Errors from different rows do not supply duplicate information | The usual standard-error calculation and effective sample size |
| Positive variation in $X$ | Not all observations have the same predictor value | A slope can be estimated |
| Constant conditional variance | $\operatorname{Var}(\varepsilon\mid X)=\sigma^2$ | Classical OLS standard errors have the stated form |
| Conditional normality | $\varepsilon\mid X$ follows a normal distribution | Exact t and F reference distributions in small samples |

Say it early and loudly: **the outcome does not need to be normally
distributed, and neither does the predictor.** Agreed deal amounts are
strongly right-skewed, and that alone breaks nothing. The assumptions
concern the conditional mean, the errors, and the sampling process --- not
the marginal histograms of $Y$ and $X$.

### Linearity is about the average, not every point

Individual outcomes need not lie on a line. Linearity says that the
*average* outcome at each predictor value follows a line. A systematic
curve in residuals means one constant slope misses part of that average
pattern.

Constant variance is the vertical counterpart: homoskedasticity says the
conditional variance is the same at each predictor value. It does not say
all residuals have the same size. A fan-shaped residual plot --- in which
the vertical band widens or narrows --- suggests changing conditional
variance. @fig-i02-assumption-panels shows both stories.

```{r}
#| label: fig-i02-assumption-panels
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 6
#| fig-cap: "Four simulated situations. Top row: a linear mean fits (left, outcome against predictor); a curved mean leaves a U-shaped residual pattern (right). Bottom row: constant vertical spread (left); spread growing with the fitted value, a fan (right). The three residual panels share the gold dashed line at residual zero."
library(ggplot2)
set.seed(802)
n_demo <- 90
x_demo <- runif(n_demo, 0, 10)

y_linear <- 2 + 0.65 * x_demo + rnorm(n_demo, 0, 0.8)
y_curved <- 2 + 0.10 * x_demo + 0.09 * x_demo^2 + rnorm(n_demo, 0, 0.55)
m_curved <- lm(y_curved ~ x_demo)
m_linear <- lm(y_linear ~ x_demo)

fit_demo <- seq(1, 9, length.out = n_demo)
res_constant <- rnorm(n_demo, 0, 0.75)
res_fan <- rnorm(n_demo, 0, 0.18 + 0.13 * fit_demo)

panel_names <- c("Linear mean (outcome)", "Curved mean (residuals)",
                 "Constant spread (residuals)", "Changing spread (residuals)")
demo_df <- data.frame(
  panel = factor(rep(panel_names, each = n_demo), levels = panel_names),
  x = c(x_demo, fitted(m_curved), fit_demo, fit_demo),
  y = c(y_linear, residuals(m_curved), res_constant, res_fan)
)
line_df <- data.frame(
  panel = factor(panel_names[1], levels = panel_names),
  intercept = coef(m_linear)[1],
  slope = coef(m_linear)[2]
)
zero_df <- data.frame(
  panel = factor(panel_names[-1], levels = panel_names),
  yintercept = 0
)
zero_lab <- data.frame(
  panel = factor(panel_names[-1], levels = panel_names),
  x = c(min(fitted(m_curved)), 1, 1),
  y = c(max(residuals(m_curved)), max(res_constant), max(res_fan))
)

ggplot(demo_df, aes(x = x, y = y)) +
  geom_hline(data = zero_df, aes(yintercept = yintercept),
             colour = "#C98B2E", linetype = "dashed", linewidth = 0.8) +
  geom_abline(data = line_df, aes(intercept = intercept, slope = slope),
              colour = "#D55E00", linewidth = 0.9) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  geom_text(data = zero_lab, aes(x = x, y = y, label = "residual = 0"),
            hjust = 0, colour = "#17324D", size = 3) +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "Predictor (top left) or fitted value",
       y = "Outcome (top left) or residual") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

::: {.callout-note collapse="true"}
## Check: match each residual pattern to the assumption it challenges

A **U-shaped** (curved) residual pattern challenges the **linear
conditional mean**: the average residual is not zero everywhere, so a
straight line misses part of the mean relationship. A **fan** challenges
**constant conditional variance**: the vertical spread of residuals changes
with the fitted value. A real plot can show both problems at once.
:::

### Reading the deal model's vertical spread

Now apply the fan test to @fig-i02-warmup-residuals. At small fitted values
the residuals hug the zero line; at larger fitted values they range over
hundreds of thousands of dollars. The conditional variance of agreed
amounts grows with the size of the deal --- an economically unsurprising
fan. Classical standard errors, which assume one common $\sigma^2$, are
therefore built on a doubtful assumption here. Lecture 3 returns to exactly
this point.

::: {.callout-caution collapse="true"}
## Try it: annotate a residual plot

In class this is done in pairs on a printed copy of
@fig-i02-warmup-residuals. On your own copy (or a sketch):

*Step 1 (given).* The horizontal dashed line marks residual zero: points
above it agreed more than fitted, points below agreed less.

*Step 2 (you).* Draw the rough envelope of the residuals --- one curve
above, one below --- from the left edge to the right edge of the plot. Does
the band stay equally wide?

*Step 3 (you).* Name the assumption your envelope challenges, and write
one sentence saying whether the *average* of the residuals also drifts away
from zero anywhere.

::: {.callout-note collapse="true"}
## Check your answer

The envelope widens sharply from left to right: small fitted deals sit
within tens of thousands of the line, large fitted deals within hundreds of
thousands. That fan challenges **constant conditional variance**. The
average residual stays near zero across the range, so the plot does not
show a clear violation of the **linear conditional mean** --- two different
assumptions, two different readings of the same picture.
:::
:::

### What plots cannot show

Two rows of the assumptions table never appear in any residual plot.

**Zero conditional mean.** Fitted OLS residuals are mechanically centred
around the fitted line --- the least-squares algebra forces it --- so a
residual plot cannot reveal whether $E[\varepsilon\mid X]=0$ holds in the
population. Unobserved confounders are, by definition, not on the plot. Ask
instead how the observations came about: what plausible common causes were
omitted, whether measurement error or reverse direction is credible.

**Independence.** Whether rows share information is a fact about the data
collection, not about the residuals. If rows are related, treating them as
independent makes the effective sample size look larger than it is and
standard errors too small.

Two short Shark Tank stories, classified:

- **Repeat pitchers.** Some ventures appear in more than one season, and
  some founders pitch multiple products. Their rows share a founder, a
  negotiation style, a network of sharks. This is an **independence**
  problem: the 882 deal rows carry less than 882 rows' worth of
  information.
- **Network ties.** Suppose founders with strong investor networks both ask
  for more (they know the market) and get systematically better deals (the
  sharks know them). Network strength is unrecorded, related to the
  predictor, and moves the outcome: this is a **zero conditional mean**
  problem --- the slope partly absorbs the omitted network effect.

Neither story is visible in @fig-i02-warmup-residuals. Both come from
knowing how the show works.

### Influence: when a few rows steer the line

A diagnosis can start before any plot, in one line of printed output. The
five-number summary of the residuals is the first thing `summary()` reports
for any model; here it is on its own:

```{r}
#| label: i02-residual-line
summary(residuals(deal_model))
```

::: {#exm-i02-residual-line}
## Worked example: reading the residual line as a red flag

Same five numbers, read slowly, all in $1,000s:

- The **median** residual is about $-28.8$: the typical deal sits
  \$28,800 *below* the fitted line.
- The middle half of all residuals --- first to third quartile --- spans
  only $-30.1$ to $-28.2$: a band about \$2,000 wide. Most deals sit in a
  tight bundle just under the line.
- The **maximum** residual is about $2821$: one deal came in \$2.82
  million *above* its fitted value.

A symmetric, well-behaved residual distribution would have a median near
zero and a maximum and minimum of comparable size. Here the mean is zero
(it always is, with an intercept), but the distribution is wildly
asymmetric: a dense mass slightly below the line, balanced by a few
enormous positive misses. Before computing anything else, this one line
says the fit is being shaped by a handful of extreme pitches. The same five
numbers sit at the top of `summary(deal_model)`, which Lecture 3 reads in
full.
:::

Two standard tools quantify the suspicion. A **standardised residual**,
`rstandard()`, re-expresses each residual in standard-deviation units, so
values beyond about $\pm 3$ are large by any convention. **Cook's
distance**, `cooks.distance()`, measures how much the entire fitted line
would move if one observation were removed; a common screening threshold is
$4/n$.

```{r}
#| label: i02-influence
rs <- rstandard(deal_model)
cd <- cooks.distance(deal_model)

sum(abs(rs) > 3)          # unusually large standardised residuals
sum(cd > 4 / nrow(deals)) # influential by the 4/n screen
deals[order(cd, decreasing = TRUE)[1:3],
      c("ask_1k", "deal_1k")]
```

Twelve pitches have standardised residuals beyond $\pm 3$, and 21 exceed
the $4/n$ Cook's screen. The single most influential record asked for
\$150,000 and agreed \$3,000,000 --- twenty times its request.
@fig-i02-influence places the flagged pitches on the scatter.

```{r}
#| label: fig-i02-influence
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The 882 deal records with the fitted line, full range shown. Orange points exceed the 4/n Cook's distance screen; the labelled pitch (ask $150,000, deal $3,000,000) moves the fitted line most. The $5 million ask at the far right is also influential --- not because its residual is large, but because it sits so far right that it pulls on the line like a lever."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
cd <- cooks.distance(deal_model)
deals$flagged <- cd > 4 / nrow(deals)
top <- which.max(cd)

ggplot(deals, aes(x = ask_1k, y = deal_1k)) +
  geom_point(data = subset(deals, !flagged),
             colour = "grey75", alpha = 0.45, size = 1.6) +
  geom_abline(intercept = coef(deal_model)[1],
              slope = coef(deal_model)[2],
              colour = "#D55E00", linewidth = 0.9) +
  geom_point(data = subset(deals, flagged),
             colour = "#D55E00", size = 2.2) +
  annotate("text", x = deals$ask_1k[top] + 60, y = deals$deal_1k[top],
           label = "ask 150, deal 3,000",
           hjust = 0, colour = "#17324D", size = 3.4) +
  labs(x = "Amount requested ($1,000s)",
       y = "Agreed on-air deal amount ($1,000s)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

What to do with a flagged observation --- the standing rule, in order:

1. **Investigate.** Is it a data error, or a real pitch? (These are real:
   some sharks buy whole companies.)
2. **Report.** State that results depend on a few identifiable
   observations; show the fit with and without them if it matters.
3. **Never silently delete.** Dropping inconvenient rows without saying so
   is not cleaning, it is misrepresentation.

### Q--Q plots: how to read them, when normality matters

The last assumption in the table is conditional normality of the errors. A
**normal Q--Q plot** compares the ordered standardised residuals with the
values expected under a normal distribution; agreement appears as points
along the reference line.

```{r}
#| label: fig-i02-qq
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Normal Q--Q plot of the standardised residuals of the deal model. The heavy departure in the upper tail is the same handful of enormous positive residuals flagged by Cook's distance."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
qq_df <- data.frame(rs = rstandard(deal_model))

ggplot(qq_df, aes(sample = rs)) +
  geom_qq_line(colour = "#D55E00", linewidth = 0.9) +
  geom_qq(colour = "#009E73", alpha = 0.4, size = 1.6) +
  labs(x = "Normal theoretical quantile",
       y = "Standardised residual") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read the horizontal axis as "where a residual of this rank *should* sit if
errors were normal" and the vertical axis as where it actually sits. In
@fig-i02-qq the middle of the distribution is close to the line while the
upper tail shoots far above it: the residual distribution has a much
heavier right tail than a normal.

When does this matter? Normality is used to derive **exact** t and F
distributions in **small samples**. With 882 reasonably independent
observations, central-limit arguments make the inference of Lecture 3
approximately valid even without normal errors --- but "approximately" is
doing real work when the same few pitches that break normality also
dominate the fit. Severe outliers distort both the line and its
uncertainty, whatever the sample size.

::: {.caveat-box}
<span class="label">What a residual check can and cannot certify</span>

- **No visible pattern is not proof.** A clean residual plot is consistent
  with the assumptions; it does not establish them. Omitted confounders and
  dependent rows leave no trace on any plot.
- **Diagnostics start the conversation, they do not end it.** A fan or an
  influential pitch is a finding to investigate and report, not an
  automatic verdict on the model --- and not a licence to delete rows until
  the pictures look better.
- **Keeping or dropping an influential observation is a substantive
  decision**, made from knowledge of the data, stated openly in the
  report --- never made silently by the software or the analyst.
:::

### Answering the driving question

Trust in the Lecture 1 line is conditional. Its residuals confess two
things: the vertical spread grows with the size of the deal (the fan), and
a handful of extreme pitches --- flagged by standardised residuals beyond
$\pm3$ and Cook's distances above $4/n$ --- steer both the line and its
apparent precision. Nothing in the plots contradicts a linear conditional
mean, but zero conditional mean and independence rest on how the show
works, not on any picture. Lecture 3 computes standard errors, intervals,
and tests for this model --- and must respect exactly these two confessions.

**Next:** We now know when a fitted line deserves provisional trust. What we
still lack is precision and worth: the slope is an estimate from one sample,
and a model can pass every diagnostic while explaining almost nothing. Lecture
3 brings the inference machinery from QM Basic to the coefficient --- standard
error, interval, t-test --- and adds the fit toolkit: the SST = SSR + SSE
decomposition, $R^2$, and the overall F-test for whether the model is useful
at all.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 55 minutes.</span>

### Retrieval (10 minutes)

1. Which assumption concerns the conditional mean?
2. Must the outcome variable be normally distributed? Why or why not?
3. What does a fan shape in a residual-versus-fitted plot suggest?
4. Which assumptions cannot be established from a residual plot alone?
5. What do a standardised residual and Cook's distance each measure?
6. What are the three steps for handling a flagged observation?

::: {.callout-note collapse="true"}
## Check your answers

1. The linear conditional mean, $E[Y\mid X=x]=\beta_0+\beta_1x$ (and zero
   conditional mean of the errors).
2. No. The assumptions concern the errors conditional on the predictor and
   the sampling process, not the marginal distribution of $Y$ or $X$.
3. Non-constant conditional variance (heteroskedasticity).
4. Zero conditional mean and independence; they require knowledge of the
   design and data-generating process.
5. A standardised residual measures how unusual one observation's vertical
   miss is, in standard-deviation units; Cook's distance measures how much
   the whole fitted line moves if that observation is removed.
6. Investigate, report, never silently delete.
:::

### Diagnose three residual plots (15 minutes)

For each panel of @fig-i02-three-plots, write one sentence: which
assumption, if any, does the plot challenge?

```{r}
#| label: fig-i02-three-plots
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.2
#| fig-cap: "Three simulated residual-versus-fitted plots for diagnosis practice. The gold dashed line marks residual zero in each panel."
library(ggplot2)
set.seed(1204)
n_p <- 120
f_p <- seq(1, 9, length.out = n_p)
res_a <- 1.6 * (f_p - 5)^2 / 8 - 1 + rnorm(n_p, 0, 0.55)
res_b <- rnorm(n_p, 0, 0.15 + 0.16 * f_p)
res_c <- rnorm(n_p, 0, 0.8)

plots_df <- data.frame(
  panel = rep(c("Plot A", "Plot B", "Plot C"), each = n_p),
  fitted = rep(f_p, 3),
  residual = c(res_a, res_b, res_c)
)
zero_lab <- data.frame(
  panel = c("Plot A", "Plot B", "Plot C"),
  x = 1,
  y = c(max(res_a), max(res_b), max(res_c))
)

ggplot(plots_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  geom_text(data = zero_lab, aes(x = x, y = y, label = "residual = 0"),
            hjust = 0, colour = "#17324D", size = 3) +
  facet_wrap(~ panel, scales = "free_y") +
  labs(x = "Fitted value", y = "Residual") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

::: {.callout-note collapse="true"}
## Check your answers

**Plot A**: the residuals follow a U shape --- the average residual is
negative in the middle and positive at the ends --- challenging the linear
conditional mean. **Plot B**: the vertical band widens with the fitted
value --- a fan challenging constant conditional variance. **Plot C**: no
visible pattern; consistent with the assumptions the plot *can* speak to,
which is not the same as proof that all assumptions hold.
:::

### Practice: diagnostics for your own model (15 minutes)

Fit `equity_offered_pct ~ season` on all 1,441 pitches. Produce the
residual-versus-fitted plot and the normal Q--Q plot, and interpret each in
one or two sentences.

::: {.callout-note collapse="true"}
## Check your answers

```{r}
#| label: i02-practice-equity
#| fig-height: 3.6
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
equity_model <- lm(equity_offered_pct ~ season, data = sharks)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(equity_model), residuals(equity_model),
     pch = 16, col = grDevices::adjustcolor("#0072B2", alpha.f = 0.30),
     bty = "n", xlab = "Fitted equity offered (pp)",
     ylab = "Residual (pp)")
abline(h = 0, lty = 2, lwd = 2, col = "#D55E00")
qqnorm(rstandard(equity_model), pch = 16,
       col = grDevices::adjustcolor("#009E73", alpha.f = 0.35),
       bty = "n", main = "", xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(equity_model), col = "#D55E00", lwd = 2)
par(old_par)
```

The residual plot shows sixteen vertical stripes --- season takes only
whole-number values, so only sixteen fitted values exist; that is a feature
of the predictor, not a violation. The vertical spread is broadly similar
across stripes, so no strong fan. The Q--Q plot shows a heavy right tail:
a minority of pitches offered far more equity than the fitted trend (the
maximum residual is above 80 percentage points --- founders offering
essentially the whole company), so the normality assumption is doubtful,
which matters little for inference at $n = 1441$ but flags rows worth
investigating.
:::

### Spaced review: one residual-arithmetic item from Lecture 1 (5 minutes)

Using the Lecture 1 fitted equation
$\widehat{\text{deal}} = 27.571 + 1.0085 \times \text{ask}$ (both in
$1,000s): a pitch asked \$200,000 and agreed \$180,000. Compute the fitted
value and the residual.

::: {.callout-note collapse="true"}
## Check your answer

Fitted: $27.571 + 1.0085\times200 = 229.27$, about \$229,273. Residual:
$180 - 229.27 = -49.27$, about $-\$49{,}273$ --- the pitch agreed less than
the model's fitted conditional mean.
:::

### Worked exam interpretation (5 minutes)

**Question** (mock exam Question 2 pattern). A residual-versus-fitted plot
for a growth-on-marketing regression shows vertical spread increasing with
the fitted value. Name the pattern and the assumption it challenges, and
state one assumption the plot cannot assess and what information would
assess it.

::: {.callout-note collapse="true"}
## Check a model answer

The pattern is a fan: residual spread grows with the fitted value,
challenging constant conditional variance (homoskedasticity). The plot
cannot assess independence of observations; that requires knowledge of the
data collection --- for example whether some ventures appear in multiple
rows or were sampled in clusters.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well --- and a
regression you might one day run on its data (sales on advertising, orders
on price, sign-ups on campaign week). Write two sentences: first name the
assumption from today's table you would worry about most for that data, then
describe one observation that could dominate the fit and what you would do
about it --- investigate and report, not delete.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "For orders-on-price data I would worry most about zero conditional mean: I cut prices exactly in the weeks demand is already weak, so the error and the predictor move together. One 200-box catering order would dominate any fit --- I would report the model with and without it and investigate whether catering is a different business, not delete the row."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: describe your venture's data
and ask it to argue which assumption is least plausible.
::::

Continue to Lecture 3: Coefficient inference, fit, and the F-test.

Formula sheet: Lecture 2 formulas.

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.1.2 and 3.3.3, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 19 and 22.

:::
