# Lecture 3: Coefficient inference, fit, and the F-test {#i03}

```{r}
#| label: i03-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 3, you can</span>

- read a coefficient's standard error and build its confidence interval by
  hand, using the correct degrees of freedom;
- test a regression coefficient against zero and say what the p-value does
  and does not establish;
- decompose outcome variation into explained and residual components;
- calculate and interpret $R^2$ --- and correct four common misreadings;
- state and interpret the overall F-test for model usefulness;
- connect the simple-regression F-test to the slope t-test;
- distinguish fit, statistical evidence, prediction, and causality.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">A slope is an estimate, not a fact. Before a number goes into a pitch deck or a report, we need to know how precise it is and whether the model behind it earns its keep.</p>

The ask--deal slope is about 1.01. How precisely do we know that number, how
much of the outcome does the line explain, and is the model useful at all?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=50sv4UTjE90&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=12">3.4
Some Important Questions</a>. Focus on the questions "Is there a
relationship?" and "How accurately can we predict the response?" All links
are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate the recipes (5 minutes).** With the Basic formula sheet open,
write from memory the general 95% confidence interval recipe and the
two-sided test of a benchmark value. Today both recipes return, applied to a
regression slope.

::: {.callout-note collapse="true"}
## Check your preparation

Interval: $\text{estimate} \pm (\text{multiplier}) \times
\text{standard error}$, with the multiplier close to 1.96 for large samples.
Test: $t = (\text{estimate} - \text{benchmark})/\text{standard error}$,
compared with a t distribution; small p-values are evidence against the
benchmark.
:::

**Classify (5 minutes).** For a *useful* model, decide whether each quantity
should be relatively small or relatively large: SSE, SSR, $R^2$. Commit to
your three answers; the lecture defines all three quantities precisely.

::: {.callout-note collapse="true"}
## Check your preparation

For a useful model, SSE should be relatively small, SSR relatively large,
and $R^2$ relatively large. These are comparisons to total outcome
variation, not universal cut-offs.
:::

**R warm-up (10 minutes).** The standing warm-up: rebuild the Lecture 1
model and extract its coefficient table. Bring the table to class ---
today's first hour reads it line by line:

```{r}
#| label: i03-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
summary(deal_model)$coefficients
```

If squared terms, ratios, or the interval recipe feel rusty, review
[Mathematics Refresher: squares, proportions, and standardisation](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question --- with the warm-up coefficient table on
screen.

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=50sv4UTjE90">3.4 Some Important Questions</a>.</p>

### How precisely do we know the slope?

The warm-up construction once more, so every object on this page is built in
the open:

```{r}
#| label: i03-build-model
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
summary(deal_model)$coefficients
```

Lecture 1 read only the first column. The other three columns --- and the
interval R computes from them --- are the **pentad** from QM Basic,
returning unchanged. Nothing in this table is new; only the quantity being
estimated is:

| Pentad ingredient | Where it sits | For the ask--deal slope |
|---|---|---:|
| estimate | `Estimate` column | $b_1 = 1.0085$ |
| standard error | `Std. Error` column | $\operatorname{SE}(b_1) = 0.0182$ |
| t value | `t value` column | $t = 55.4$ |
| p-value | `Pr(>|t|)` column | $< 0.001$ |
| confidence interval | `confint()` | computed below |

The standard error answers the pentad's standing question, now about a
slope: **if we could redraw the sample, how much would $b_1$ wobble?** The
estimated slope 1.0085 came from these 882 deals; a different set of deals
would give a slightly different line. The SE of 0.0182 --- in the slope's
own units, $1,000 of deal per $1,000 of ask --- estimates the size of that
sampling wobble under the Lecture 2 model conditions.

**Degrees of freedom.** The interval multiplier comes from a t distribution
with $n - 2 = 882 - 2 = 880$ degrees of freedom. The subtraction is the same
bookkeeping as in QM Basic: estimating the intercept and the slope uses two
pieces of information before any is left over to estimate the residual
spread. (With $k$ slopes plus an intercept, the residual degrees of freedom
become $n - k - 1$; that general form matters from Lecture 4 onwards.) At
880 degrees of freedom, the multiplier $t_{0.975,\,880} = 1.963$ is
indistinguishable in practice from the familiar 1.96.

::: {#exm-i03-interval}
## Worked example: the slope interval by hand

Everything comes from the coefficient table: $b_1 = 1.0085$,
$\operatorname{SE}(b_1) = 0.0182$, multiplier $\approx 1.96$.

$$
b_1 \pm t_{0.975,\,880}\operatorname{SE}(b_1)
= 1.0085 \pm 1.96 \times 0.0182
= 1.0085 \pm 0.0357
= (0.973,\; 1.044).
$$

In words: under the model conditions, the data are consistent with a
population slope between about \$973 and \$1,044 of agreed amount per
\$1,000 requested. The interval's endpoints carry the slope's units ---
that is the pentad rule that the estimate, SE, and interval share units,
while t and the p-value are unit-free.

R agrees:

```{r}
#| label: i03-confint
confint(deal_model)
```
:::

The interval describes **sampling uncertainty under the model conditions**.
It does not include uncertainty from selection into televised pitches, the
decision to analyse only on-air deals, or measurement choices --- no
formula widens an interval for a flawed design.

### The t-test against zero

The standard significance test asks whether the data are consistent with
**no** linear association:

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1\ne0,
\qquad
t=\frac{b_1-0}{\operatorname{SE}(b_1)}.
$$

By hand, once: $t = 1.0085 / 0.0182 \approx 55.4$. An estimate 55 standard
errors from zero is astronomically far into the tail of a t distribution.
Live verification, computing the p-value exactly as QM Basic did:

```{r}
#| label: i03-manual-t
coef_table <- summary(deal_model)$coefficients
b1 <- coef_table["ask_1k", "Estimate"]
se_b1 <- coef_table["ask_1k", "Std. Error"]
t_value <- b1 / se_b1
p_value <- 2 * pt(-abs(t_value), df = df.residual(deal_model))
c(estimate = b1, standard_error = se_b1,
  t_statistic = t_value, p_value = p_value)
```

What the tiny p-value **does** say: if the population slope were zero and
the model conditions held, a sample slope this far from zero would
essentially never occur --- strong evidence against a zero linear
association in this selected sample.

What it does **not** say, one line each:

- it does not say the model is correctly specified --- Lecture 2's fan and
  influential pitches stand as the permanent caveat under this table;
- it does not say the association is large or important --- with 882
  observations, even tiny slopes can earn tiny p-values;
- it does not say that requesting more causes a larger deal --- no test
  statistic repairs an observational design.

::: {.callout-note collapse="true"}
## Check: the p-value is 4×10^-289^. Is this slope therefore trustworthy?

No. The p-value measures distance from zero in standard-error units, and
the standard error was computed under the classical conditions --- constant
error variance and independent rows among them. Lecture 2 showed the
residual fan and the handful of pitches that steer the fit, so this
particular SE rests on doubtful ground. "Significant" means the data are
inconsistent with a zero slope, not that 0.0182 is the right measure of
wobble.
:::

### Decomposing variation

The second part of the driving question --- how much does the line explain?
--- needs a benchmark. Without a predictor, the best single prediction for
every deal is the mean agreed amount $\bar y$. Regression is useful to the
extent that the fitted values $\widehat y_i$ improve on that flat
prediction.

For each observation, the total deviation from the mean splits exactly into
an explained part and a residual part, as @fig-i03-decomposition shows for
one point:

$$
\underbrace{y_i-\bar y}_{\text{total}}
=
\underbrace{(\widehat y_i-\bar y)}_{\text{explained by the line}}
+
\underbrace{(y_i-\widehat y_i)}_{\text{residual}}.
$$

```{r}
#| label: fig-i03-decomposition
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 6.6
#| fig-height: 4.4
#| fig-cap: "One observation's total deviation from the mean of Y splits into the part the fitted line explains and the residual. Squaring and summing over all observations gives SST = SSR + SSE."
library(ggplot2)
x_d <- c(1, 2.2, 3.4, 4.6, 5.8, 7.0, 8.2)
y_d <- 1 + 0.8 * x_d + c(1.3, -1.5, 0.9, -0.8, 1.6, -1.2, 0.4)
fit_d <- lm(y_d ~ x_d)
decomp_df <- data.frame(x = x_d, y = y_d)
i <- 5
xh <- x_d[i]; yh <- y_d[i]; fh <- fitted(fit_d)[i]; mh <- mean(y_d)

ggplot(decomp_df, aes(x = x, y = y)) +
  geom_hline(yintercept = mh, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_abline(intercept = coef(fit_d)[1], slope = coef(fit_d)[2],
              colour = "#D55E00", linewidth = 0.9) +
  annotate("segment", x = xh + 0.55, xend = xh + 0.55, y = mh, yend = yh,
           colour = "#17324D", linewidth = 0.9) +
  annotate("segment", x = xh, xend = xh, y = mh, yend = fh,
           colour = "#009E73", linewidth = 1.3) +
  annotate("segment", x = xh, xend = xh, y = fh, yend = yh,
           colour = "#0072B2", linewidth = 1.3) +
  geom_point(size = 2.6, colour = "#009E73") +
  annotate("text", x = xh + 1.25, y = (mh + yh) / 2,
           label = "y[i] - bar(y)", parse = TRUE,
           colour = "#17324D", size = 3.4) +
  annotate("text", x = xh - 0.85, y = (mh + fh) / 2,
           label = "hat(y)[i] - bar(y)", parse = TRUE,
           colour = "#009E73", size = 3.4) +
  annotate("text", x = xh - 0.75, y = (fh + yh) / 2 + 0.15,
           label = "y[i] - hat(y)[i]", parse = TRUE,
           colour = "#0072B2", size = 3.4) +
  annotate("text", x = 9.0, y = mh - 0.45, label = "bar(y)",
           parse = TRUE, colour = "#17324D", size = 3.4) +
  annotate("text", x = 9.4, y = coef(fit_d)[1] + coef(fit_d)[2] * 9.4 + 0.5,
           label = "fitted line", hjust = 1,
           colour = "#D55E00", size = 3.4) +
  coord_cartesian(xlim = c(0, 9.6), ylim = c(0, 9.6)) +
  labs(x = "Predictor X", y = "Outcome Y") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Square each piece and sum over all observations. For ordinary least squares
with an intercept, the cross-terms cancel and the identity holds exactly:

$$
\underbrace{\sum_i(y_i-\bar y)^2}_{\text{SST: total}}
=
\underbrace{\sum_i(\widehat y_i-\bar y)^2}_{\text{SSR: explained}}
+
\underbrace{\sum_i(y_i-\widehat y_i)^2}_{\text{SSE: residual}}.
$$

Live, on the deal model, with the identity checked:

```{r}
#| label: i03-sums-squares
y <- deals$deal_1k
yhat <- fitted(deal_model)

SST <- sum((y - mean(y))^2)
SSR <- sum((yhat - mean(y))^2)
SSE <- sum((y - yhat)^2)

c(SST = SST, SSR = SSR, SSE = SSE, check = SSR + SSE)
```

The units are squared thousands of dollars, so the raw magnitudes (around
$10^8$) mean little on their own --- the sums of squares matter only
relative to each other, which is exactly what the next section computes.

::: {#exm-i03-sst-sse}
## Worked example: from SST and SSE to SSR and R²

A regression output reports $\text{SST} = 500$ and $\text{SSE} = 125$. Find
SSR and $R^2$.

By the identity, $\text{SSR} = \text{SST} - \text{SSE} = 500 - 125 = 375$.
Then

$$
R^2 = \frac{\text{SSR}}{\text{SST}} = \frac{375}{500} = 0.75
\qquad\text{or equivalently}\qquad
R^2 = 1 - \frac{\text{SSE}}{\text{SST}} = 1 - \frac{125}{500} = 0.75.
$$

The fitted values account for 75% of the observed sample variation in the
outcome; 25% remains as residual variation around the line. In class, the
one-click Mentimeter check on this calculation is posed right before the
break.
:::

### R²: what 0.777 means --- and four things it is not

The general definition, applied to the live numbers:

$$
R^2=\frac{\text{SSR}}{\text{SST}}
=1-\frac{\text{SSE}}{\text{SST}}.
$$

```{r}
#| label: i03-r-squared
c(
  manual = 1 - SSE / SST,
  from_R = summary(deal_model)$r.squared,
  correlation_squared = cor(deals$ask_1k, deals$deal_1k)^2
)
```

Here $R^2 \approx 0.777$: **about 77.7% of the observed sample variation in
agreed on-air deal amounts, among deal records, is explained by the linear
association with the amount requested.** The third number above shows a
simple-regression bonus fact: with an intercept, $R^2 = r_{XY}^2$, the
squared correlation. Squaring discards the sign, so inspect the slope or
correlation for direction.

$R^2$ is one of the most misquoted numbers in applied work. Four things it
is **not**, each with a wrong sentence corrected. The first two are worked
here; the last two are yours.

**Not the percentage of correct predictions.**
*Wrong:* "The model predicts 77.7% of deal amounts correctly."
*Corrected:* the model explains 77.7% of the *variation* in deal amounts;
individual predictions can still miss by hundreds of thousands of dollars
(the residual standard error is about \$162,000).

**Not proof that the assumptions hold.**
*Wrong:* "With $R^2$ this high, the model is clearly well specified."
*Corrected:* $R^2$ measures fit, not validity --- Lecture 2 found a
residual fan and influential pitches in this very model, and 0.777 is
silent about both.

::: {.callout-caution collapse="true"}
## Try it: correct the last two wrong sentences

In class this is the pair task. Correct each sentence in one line, as
above.

*Sentence 3 (you).* "Asking explains 78% of the deal, so raising your ask
will raise your deal."

*Sentence 4 (you).* "The model will explain 77.7% of deal-amount variation
in next year's pitches too."

::: {.callout-note collapse="true"}
## Check your answers

*Sentence 3* claims causation: $R^2$ describes an association among
observed deals; requests were not randomly assigned, and nothing in a fit
statistic says what would happen if a founder *changed* their ask.
*Sentence 4* claims out-of-sample performance: $R^2$ is computed on this
sample; a new season, a different selection of pitches, or a changed
negotiation culture can all lower it. In-sample fit is evidence about these
882 records, not a guarantee for the next 882.
:::
:::

### The overall F-test

The last part of the driving question: is the model useful **at all**,
compared with predicting $\bar y$ for everyone? For a regression with $k$
predictors, the overall usefulness test is

$$
H_0:\beta_1=\cdots=\beta_k=0
$$

against the alternative that at least one slope is non-zero. The statistic
is

$$
F=\frac{\text{SSR}/k}{\text{SSE}/(n-k-1)}.
$$

Read the ratio the way it is built: the numerator is **explained variation
per predictor**; the denominator is **residual variation per residual
degree of freedom**. F asks how many times more explanation the model buys,
per coefficient spent, than leftover noise would suggest. Values near 1 are
what pure noise delivers; large values are evidence that the predictors
jointly matter.

The whole lecture now sits in one printed block --- read `summary()` from
top to bottom in this order: (1) formula and analysis sample, (2)
coefficient estimates and units, (3) standard errors, t statistics, and
p-values, (4) residual scale, (5) $R^2$ and adjusted $R^2$, (6) the overall
F-test --- and (7) everything the table does *not* print: assumptions and
design limits.

```{r}
#| label: i03-summary
summary(deal_model)
```

The output reports $F \approx 3069$ on 1 and 880 degrees of freedom, with a
p-value too small to print meaningfully. Rejecting the null says the model
with requested amount fits better than an intercept-only model, under the
inferential conditions.

**In simple regression, F and t answer the same question.** With one
predictor and a two-sided alternative, $H_0:\beta_1 = 0$ is the entire
joint null, and

$$
F = t^2.
$$

```{r}
#| label: i03-f-t-link
t_slope <- summary(deal_model)$coefficients["ask_1k", "t value"]
f_model <- unname(summary(deal_model)$fstatistic["value"])
c(t_squared = t_slope^2, F = f_model)
```

Both equal about 3069. This equivalence **dies in multiple regression**,
where the F-test evaluates several slopes jointly while each t-test
evaluates one --- Lecture 5 builds a genuine joint test on exactly that
difference.

### Usefulness has more than one meaning

"Is the model useful?" has at least four distinct readings, and a model can
pass one while failing another:

- **Statistically non-zero slope, low $R^2$.** A loyalty-programme analysis
  on 400,000 customers finds a reliably positive effect of mailings on
  spending --- real, precisely estimated, and explaining 1% of spending
  variation. Useful for a mass decision, useless for predicting any one
  customer.
- **High $R^2$, poor causal interpretation.** Store revenue regressed on
  staff count fits beautifully --- because both track store size. Hiring
  more staff at a small store will not produce a big store's revenue.
- **Good in-sample fit, fragile out-of-sample prediction.** A travel-demand
  model fitted on 2015--2019 data explained bookings superbly and collapsed
  in 2020. In-sample $R^2$ never promised stability.
- **Accurate predictions, no structural story.** A black-box model may
  forecast weekly orders extremely well while its coefficients support no
  explanation of *why* --- fine for stocking decisions, useless for
  designing a price change.

Always align the evaluation with the goal: explanation, prediction, or
decision support.

::: {.optional-box}
<span class="label">Optional: robust standard errors</span><br>
Heteroskedasticity-consistent, or **robust**, standard errors relax the
constant-variance assumption when computing uncertainty. The fitted
coefficients remain the ordinary least-squares values; only the estimated
SEs (and hence t values, p-values, and intervals) change. Given Lecture 2's
residual fan, they are a natural upgrade for this model:

```{r}
#| label: i03-robust-code
#| eval: false
# Run this if these packages are available in the course R environment
library(lmtest)
library(sandwich)

coeftest(deal_model, vcov. = vcovHC(deal_model, type = "HC1"))
```

They fix exactly **one** thing: a misspecified conditional error variance.
They do not repair non-linearity, omitted-variable bias, dependence across
rows, influential data errors, or non-random selection. A robust SE on a
biased slope is a precise measurement of the wrong quantity.
:::

::: {.caveat-box}
<span class="label">What a precise slope and a high R² can and cannot say</span>

Every number this lecture produced --- the SE of 0.0182, the interval
$(0.973, 1.044)$, the p-value, $R^2 = 0.777$, $F \approx 3069$ --- is
computed *under the model conditions of Lecture 2*, and the diagnostics
there flagged two of them: the residual spread grows with deal size, and a
handful of extreme pitches steer the fit. So the classical SE (and
everything built from it) rests on a doubtful variance assumption, and both
the slope and $R^2$ are partly the work of a few identifiable records.
Beyond that, the sample is on-air deals only --- selection travels with
every estimate --- and none of today's statistics moves the association any
closer to a causal claim. Precision and fit are internal report cards, not
external validation.
:::

### Answering the driving question

Precisely, substantially, and yes --- with conditions. The slope 1.0085 has
standard error 0.0182, so the 95% interval $(0.973, 1.044)$ pins the
association down to within about seven cents per dollar asked; the t-test
rejects a zero slope beyond any doubt ($t \approx 55.4$). The line explains
77.7% of the variation in agreed amounts, and the overall F-test
($F \approx 3069$) confirms the model beats a flat mean. Every one of those
statements is conditional on the Lecture 2 diagnostics --- the fan and the
influential pitches --- and none of them is causal.

**Next:** Tutorial 1 rehearses Lectures 1--3 in exam format and in TestVision
style: fit a model in R, diagnose it, and defend its coefficients with
intervals, tests, and $R^2$ --- the complete simple-regression pipeline you
now own. Bring your attempts; the session starts from them.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 70 minutes.</span>

### Retrieval (10 minutes)

1. Why are the residual degrees of freedom $n-2$ in simple regression?
2. Which pentad quantities share the slope's units, and which are
   unit-free?
3. What do SST, SSR, and SSE measure, and how are they related?
4. Why is $R^2$ between 0 and 1 in an OLS model with an intercept?
5. What is the null hypothesis of the overall F-test, and why does $F=t^2$
   in simple regression?
6. What single problem do robust standard errors address?

::: {.callout-note collapse="true"}
## Check your answers

1. Two coefficients --- intercept and slope --- are estimated before
   residual variation can be measured, using up two pieces of information.
2. The estimate, its SE, and the interval endpoints share the slope's
   units; the t statistic and p-value are unit-free.
3. Total, explained (fitted), and residual sample variation in the outcome;
   $\text{SST}=\text{SSR}+\text{SSE}$ for OLS with an intercept.
4. SSE is non-negative and no larger than SST, so $1-\text{SSE}/\text{SST}$
   lies between 0 and 1.
5. All population slope coefficients are zero. With one predictor that is
   the same hypothesis as the two-sided slope t-test, and $F=t^2$.
6. Misspecified (non-constant) conditional error variance --- nothing else.
:::

### From worked to faded: an interval for a new row (15 minutes)

Re-read the worked slope interval (@exm-i03-interval). Then build the
interval for the **intercept row** of the same table.

*Step 1 (given).* From the coefficient table: estimate $27.571$, standard
error $7.241$, degrees of freedom $880$, multiplier $\approx 1.96$.

*Step 2 (you).* Compute the 95% interval.

*Step 3 (you).* Interpret it in one sentence --- and say why this
particular interval has little standalone business value (Lecture 1 told
you).

::: {.callout-note collapse="true"}
## Check your answer

$27.571 \pm 1.96 \times 7.241 = 27.571 \pm 14.19 = (13.4,\; 41.8)$, in
$1,000s --- R's exact answer is $(13.36, 41.78)$. Under the model
conditions, the data are consistent with a population intercept between
about \$13,400 and \$41,800. But the intercept is the fitted mean at a
request of zero, outside the observed range of asks, so the interval is
mathematically sound and substantively unhelpful.
:::

Then the unassisted set:

1. A study reports a coefficient of 2.40 with standard error 0.80 in a
   large sample. Compute the t statistic, decide the 5% two-sided test
   against zero, and give the 95% interval.
2. Interpret $R^2=0.20$ without calling the model "80% wrong."
3. A model has $k=3$, $n=104$, SSR = 300, and SSE = 500. Calculate F.
4. Explain why adding a predictor cannot increase the in-sample SSE in OLS.

::: {.callout-note collapse="true"}
## Check your answers

1. $t = 2.40/0.80 = 3$; reject at 5% (well beyond 1.96). Interval:
   $2.40 \pm 1.96 \times 0.80 = (0.83,\; 3.97)$.
2. The fitted values explain 20% of observed outcome variation; 80% remains
   as residual variation around the fitted line. Nothing is "wrong" ---
   the model's predictions are simply far from individual outcomes.
3. $F=(300/3)/(500/(104-3-1))=100/5=20$.
4. The expanded model can reproduce the old fit by setting the new slope to
   zero, so least squares cannot end at a larger minimum SSE.
:::

### Practice: the full pipeline on equity and season (20 minutes)

Fit `equity_offered_pct ~ season` on all 1,441 pitches. Report the slope's
95% confidence interval, the t-test against zero, $R^2$, and the overall
F-test --- then write a 120-word conclusion that separates estimate,
uncertainty, fit, and scope.

::: {.callout-note collapse="true"}
## Check a model answer

```{r}
#| label: i03-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
equity_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(equity_model)$coefficients
confint(equity_model)
c(r_squared = summary(equity_model)$r.squared,
  F = unname(summary(equity_model)$fstatistic["value"]))
```

A complete conclusion reports: one later season is associated with about
0.824 percentage points less initial equity offered (SE 0.048); the 95%
interval $(-0.917, -0.731)$ excludes zero and the t-test rejects a zero
slope ($t\approx-17.3$, p $<$ 0.001); season explains about 17.3% of
observed equity-offer variation, and the overall F-test
($F\approx300.7 = t^2$) rejects the intercept-only model. Scope: an
observational time trend among recorded pitches --- not a causal effect of
season, and the classical SEs assume the Lecture 2 conditions.
:::

### Spaced review: one diagnostics item from Lecture 2 (5 minutes)

The residual-versus-fitted plot of the deal model fans out to the right.
Which of today's printed numbers does that pattern undermine most directly,
and why?

::: {.callout-note collapse="true"}
## Check your answer

The classical standard error 0.0182 --- and with it the t statistic,
p-value, and confidence interval built from it. The classical SE formula
assumes one common error variance $\sigma^2$; a fan says the conditional
variance grows with the fitted value, so the stated precision is doubtful.
Robust SEs address exactly this one problem; they repair nothing else.
:::

### Worked exam interpretation (5 minutes)

**Question** (mock exam Question 3 pattern). A model of venture sales
growth (percentage points) on marketing spending (€10,000s), fitted on
$n = 102$ ventures, reports a slope of 1.38 with standard error 0.46,
$R^2 = 0.083$, and $F = 9.0$. Build the 95% interval for the slope, decide
the 5% test against zero, verify the F statistic from the t statistic, and
state in one sentence what the low $R^2$ does and does not imply.

::: {.callout-note collapse="true"}
## Check a model answer

$t = 1.38/0.46 = 3.0$; with $df = 100$ the multiplier is about 1.98, so the
interval is $1.38 \pm 1.98\times0.46 \approx (0.47,\; 2.29)$ percentage
points per €10,000. The interval excludes zero, so the two-sided 5% test
rejects $H_0:\beta_1=0$; consistently, $F = t^2 = 9.0$. The low $R^2$ says
marketing spending explains about 8% of growth variation --- predictions
for individual ventures will be poor --- but it does not undo the evidence
that the association is non-zero, and neither number makes the association
causal.
:::

### From the reading ladder (10 minutes)

The [reading ladder](../resources/reading-ladder.qmd) returns to
[Tierney & Farmer (2002)](https://doi.org/10.2307/3069429) --- the paper
whose correlation table you read in Basic Lecture 5. This time find a
**regression table** and pick one coefficient row. Apply today's pentad
reading: the estimate and its sign, the reported SE or significance marker,
what null hypothesis the star implicitly tests, and what the careful
one-sentence interpretation is. Then find the reported $R^2$ (or change in
$R^2$) and say in one sentence what it does --- and does not --- tell you
about the model's usefulness.

::: {.callout-note collapse="true"}
## Check your approach

Your sentence should name the outcome, the predictor's units or scale, and
"holding the other included variables fixed", and the significance star
should translate to "a t-test against zero rejects at the marked level" ---
nothing more. If the $R^2$ sentence avoided both "the model is good" and
"the model is bad" and instead said what share of outcome variation the
predictors account for in this sample, you read the table the way this
course intends.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well --- and one
regression you might run on its data. Write two sentences: first say which
you would rather have for your next decision, a precise slope (narrow
interval) or a high $R^2$, and why; then name the usefulness meaning from
today's four-case list that matters most for that decision. There is no
wrong answer; the point is noticing that precision and fit answer different
questions.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "For setting next month's ad budget I would take a precise slope over a high $R^2$: I need to know what an extra €100 buys, even if week-to-week orders stay noisy. That is the levers meaning of usefulness --- predicting individual weeks matters less to me than the payoff of an input I control."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to invent a business
model with a significant slope and a low $R^2$ and challenge you to say
what the model is still useful for.
::::

Continue to Tutorial 1: Simple regression, diagnostics, and model
usefulness.

Formula sheet: Lecture 3 formulas.

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.1.2--3.1.3, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 19--20.

:::
