# Lecture 3: Coefficient inference, fit, and the F-test {#i03}

```{r}
#| label: i03-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 3, you can</span>

- read a coefficient's standard error and build its confidence interval by
  hand, using the correct degrees of freedom
- test a regression coefficient against zero and say what the p-value does
  and does not establish
- split the variation of the outcome into an explained part and a residual
  part
- calculate and interpret $R^2$, and correct four common misreadings
- state and interpret the overall F-test for model usefulness
- connect the simple-regression F-test to the slope t-test
- distinguish fit, statistical evidence, prediction, and causality.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- read a coefficient table: standard error, t value, p-value
- build a 95% interval for a slope by hand
- calculate $R^2$ from SST and SSE and say what it is not
- state the null hypothesis of the overall F-test.

**Enough to recognise:**

- the formula of adjusted $R^2$
- what robust standard errors fix, and what they do not fix.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">A slope is an estimate, not a fact. Before a number goes into a pitch deck or a report, we need to know how precise it is and whether the model behind it is of any use.</p>

The valuation slope is about 0.65. How precisely do we know that number, how
much of the outcome does the line explain, and is the model useful at all?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Minimum path:** the video and the R task. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=3GiWpRfkSjc&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=10">3.2
Hypothesis Testing and Confidence Intervals</a>. Focus on the standard error of
the slope, its confidence interval, and the test against zero. The video
writes RSS for the quantity this book calls SSE. All links
are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate the recipes (5 minutes).** Write from memory the general recipe
for a 95% confidence interval and the two-sided test of a benchmark value.
Then compare with the blocks for Lectures 9 and 10 of the
[QM Basic formula sheet](../basic/formula-sheet.qmd#formula-l9).
[Basic Lecture 9: The t distribution and confidence intervals](../basic/b09-t-confidence-intervals.qmd)
explains the interval, and
[Basic Lecture 10](../basic/b10-testing-errors-power.qmd) explains the
p-value and the test decision. This lecture applies both recipes to a
regression slope.

::: {.callout-note collapse="true"}
## Check your preparation

Interval: $\text{estimate} \pm (\text{multiplier}) \times
\text{standard error}$, with the multiplier close to 1.96 for large samples.
Test: $t = (\text{estimate} - \text{benchmark})/\text{standard error}$,
compared with a t distribution. Small p-values are evidence against the
benchmark.
:::

**Classify (5 minutes).** Three quantities measure how well a line fits. SSE
is the sum of squared residuals from Lecture 1. SSR is the part of the
variation in the outcome that the line explains. $R^2$ is the share of the
variation that the line explains. For a useful model, decide whether each of
the three should be relatively small or relatively large. Commit to your
three answers. The lecture material defines all three quantities precisely.

::: {.callout-note collapse="true"}
## Check your preparation

For a useful model, SSE should be relatively small, SSR relatively large,
and $R^2$ relatively large. These are comparisons to total outcome
variation, not universal cut-offs.
:::

**R warm-up (10 minutes).** Rebuild the Lecture 1 model and print its
coefficient table. Bring the table to class:

```{r}
#| label: i03-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
summary(val_model)$coefficients
```

**Code along (bring your laptop).** Bring a charged laptop with RStudio and
the course project open, so that you can run the code of this chapter
yourself during the lecture and get the same numbers and figures.

If you are not sure about squared terms, ratios, or the interval recipe,
review
[Mathematics Refresher: squares, proportions, and standardisation](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section reads the coefficient table of the valuation model: the
standard error, the t value, the p-value, and the confidence interval of the
slope. It then measures how well the line fits, with $R^2$, and tests whether
the model is useful at all, with the F-test.

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=3GiWpRfkSjc">3.2 Hypothesis Testing and Confidence Intervals</a>.</p>

### Start with a founder's question

The slope says 65 cents per claimed dollar. Would another set of deals say
50 cents, or 80? A founder who negotiates with this number wants to know how
far it can be off. R and every AI assistant answer with a table of standard
errors, t values, and p-values. This lecture shows how to read that table
and how to check what is said about it.

### How precisely do we know the slope?

The code below is the warm-up code. It rebuilds the data and the model of
Lecture 1: the valuation agreed on air (`agreed_m`) regressed on the
valuation claimed by the founder (`claimed_m`), both in \$ millions, for the
874 Shark Tank deals for a share of the company. It then prints the
coefficient table.

```{r}
#| label: i03-build-model
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
summary(val_model)$coefficients
```

Lecture 1 read only the first column, the estimates. The other three
columns, and the interval that R calculates from them, complete the pentad:
the five quantities from QM Basic that Lecture 1 reviewed. They work here as
in QM Basic. Only the quantity that is estimated is new, a slope:

| Pentad ingredient | Where it sits | For the valuation slope |
|---|---|---:|
| estimate | `Estimate` column | $\hat\beta_1 = 0.649$ |
| standard error | `Std. Error` column | $\widehat{\operatorname{SE}}(\hat\beta_1) = 0.0111$ |
| t value | `t value` column | $t = 58.6$ |
| p-value | `Pr(>|t|)` column | $< 0.001$ |
| confidence interval | `confint()` | computed below |

The 874 deals are all the deals for a share of the company in the file, so
nothing was sampled from it. We treat them as one outcome of a process that
could have produced a different set of deals. A different set of deals would
give a slightly different line. The standard error says how much the
estimated slope $\hat\beta_1$ would vary from one such set to the next.

::: {.definition-box}
<span class="label">Definition</span>

**Standard error of a coefficient**,
$\widehat{\operatorname{SE}}(\hat\beta_1)$. An estimate of how much the
estimated coefficient would vary from sample to sample. It has the units of
the coefficient.
:::

Here the standard error is 0.0111, in the units of the slope: \$ million
agreed per \$ million claimed. It is small compared with the slope, about 2%
of 0.649. This number holds under the model assumptions of Lecture 2.

**Degrees of freedom.** The interval multiplier comes from a t distribution
with $n - 2 = 874 - 2 = 872$ degrees of freedom. Two is subtracted because
two coefficients, the intercept and the slope, are estimated before the
spread of the residuals can be estimated. With $k$ slopes and an intercept,
the residual degrees of freedom are $n - k - 1$. This general form is needed
from Lecture 4 onwards. The multiplier for a 95% interval is written
$t_{0.975,\,872}$. Basic Lecture 9 calls this number the critical value. It
is the value that a t distribution with 872 degrees of
freedom stays below with probability 0.975, so 2.5% lies above it and 2.5%
lies below its negative. In R it is `qt(0.975, df)`:

```{r}
#| label: i03-multiplier
qt(0.975, df = df.residual(val_model))
```

With 872 degrees of freedom the multiplier is 1.963, almost the same as the
1.96 of the normal distribution.

::: {.definition-box}
<span class="label">Definition</span>

**95% confidence interval for a slope.**
$\hat\beta_1 \pm t_{0.975,\,n-2}\,\widehat{\operatorname{SE}}(\hat\beta_1)$.
It is the range of values of the population slope $\beta_1$ that the data
are consistent with, under the model assumptions. In large samples the
multiplier is close to 1.96.
:::

::: {#exm-i03-interval}
## Worked example: the slope interval by hand

Everything comes from the coefficient table. Take the estimate and the
standard error with one digit more than in the table above, so that rounding
does not change the result: $\hat\beta_1 = 0.6487$ and
$\widehat{\operatorname{SE}}(\hat\beta_1) = 0.01107$. The multiplier is about
1.96.

$$
\hat\beta_1 \pm t_{0.975,\,872}\,\widehat{\operatorname{SE}}(\hat\beta_1)
= 0.6487 \pm 1.96 \times 0.01107
= 0.6487 \pm 0.0217
= (0.627,\; 0.670).
$$

In words: under the model assumptions, the data are consistent with a
population slope between about 63 and 67 cents of agreed valuation per dollar
of claimed valuation. A slope of 1, which would mean that the sharks accept
the claimed valuation, is far outside this range. The endpoints of the
interval have the units of the slope. The estimate, its standard error, and
the interval share these units. The t value and the p-value have no units.

R agrees:

```{r}
#| label: i03-confint
confint(val_model)
```
:::

The interval describes only the variation from sample to sample, under the
model assumptions. It does not cover other sources of doubt: that only
televised pitches are recorded, that only pitches with a deal are analysed,
or how the valuations are measured. No formula makes an interval wider for a
weak design.

### The t-test against zero

The standard test asks whether the data are consistent with no linear
association between $X$ and $Y$, that is, with a population slope of zero.
The test is two-sided:

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1\ne0,
\qquad
t=\frac{\hat\beta_1-0}{\widehat{\operatorname{SE}}(\hat\beta_1)}.
$$

By hand: $t = 0.6487 / 0.01107 \approx 58.6$. The estimate lies about 59
standard errors away from zero. For comparison, a t value beyond about 2 in
either direction is already rare when the population slope is zero. The
p-value is the probability, if the population slope is zero, of a t value at
least this far from zero in either direction. The code below calculates it
as in QM Basic:

```{r}
#| label: i03-manual-t
coef_table <- summary(val_model)$coefficients
b1 <- coef_table["claimed_m", "Estimate"]
se_b1 <- coef_table["claimed_m", "Std. Error"]
t_value <- b1 / se_b1
p_value <- 2 * pt(-abs(t_value), df = df.residual(val_model))
c(estimate = b1, standard_error = se_b1,
  t_statistic = t_value, p_value = p_value)
```

What the tiny p-value does say: if the population slope were zero and the
model assumptions held, a sample slope this far from zero would practically
never occur. This is strong evidence against a zero slope among these deals.

What it does not say:

- It does not say that the model is correct. The fan and the influential
  deals of Lecture 2 remain a limit on everything in this table.
- It does not say that the association is large or important. With 874
  observations, even a very small slope can have a tiny p-value.
- It does not say that claiming more causes a higher agreed valuation. No
  test can turn observed deals into an experiment.

::: {.callout-note collapse="true"}
## Check: the p-value is 1×10^-304^. Is this slope therefore trustworthy?

No. The p-value measures the distance from zero in standard errors, and the
standard error was calculated under the model assumptions, among them
constant error variance and independent rows. Lecture 2 showed the fan in
the residuals and the few deals that steer the fit, so this standard error
is doubtful. "Significant" means that the data are not consistent with a
zero slope. It does not mean that 0.0111 is the right standard error.
:::

### Decomposing variation

The second part of the driving question asks how much the line explains.
This needs a point of comparison. Without a predictor, the best single
prediction for every deal is the mean agreed valuation $\bar y$. A
regression is useful to the extent that its fitted values $\widehat y_i$
come closer to the observed values than this one mean does.

For each observation, the distance from the mean $\bar y$ to the observed
value $y_i$ is its total deviation. It splits exactly into two parts. The
part explained by the line runs from the mean to the fitted value
$\widehat y_i$. The residual runs from the fitted value to the observed
value.

$$
\underbrace{y_i-\bar y}_{\text{total (navy)}}
=
\underbrace{(\widehat y_i-\bar y)}_{\text{explained by the line (green)}}
+
\underbrace{(y_i-\widehat y_i)}_{\text{residual (blue)}}.
$$

@fig-i03-decomposition shows the split for one observation of an invented
example. On the left, the green bar (explained by the line) and the blue bar
(residual) are stacked. On the right, the navy bracket (total) has the same
height as the two bars together.

```{r}
#| label: fig-i03-decomposition
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7.5
#| fig-height: 4.8
#| fig-cap: "One observation of an invented example. The dotted lines mark the fitted value and the observed value. Left: the part explained by the line (green, from the mean of Y to the fitted value) and the residual (blue, from the fitted value to the observed value). Right: the total deviation from the mean (navy). The green bar and the blue bar together are exactly as long as the navy bracket."
library(ggplot2)
# Invented data: eight points, the sixth one is highlighted
x_d <- c(1, 2, 3, 4, 5, 7, 8.6, 9.6)
y_d <- c(2.0, 1.7, 3.3, 3.0, 4.0, 8.3, 7.1, 8.0)
fit_d <- lm(y_d ~ x_d)
decomp_df <- data.frame(x = x_d, y = y_d)
i <- 6
xh <- x_d[i]                       # x of the highlighted observation
yh <- y_d[i]                       # its observed value
fh <- unname(fitted(fit_d)[i])     # its fitted value
mh <- mean(y_d)                    # the mean of Y
x_parts <- xh - 2.6                # where the two stacked parts are drawn
x_total <- xh + 0.7                # where the total is drawn
cap <- 0.12                        # half-width of the bracket ends

ggplot(decomp_df, aes(x = x, y = y)) +
  geom_hline(yintercept = mh, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_abline(intercept = coef(fit_d)[1], slope = coef(fit_d)[2],
              colour = "#D55E00", linewidth = 0.9) +
  # dotted guides at the fitted value and at the observed value
  annotate("segment", x = x_parts, xend = x_total, y = fh, yend = fh,
           colour = "#5B6976", linetype = "dotted", linewidth = 0.6) +
  annotate("segment", x = x_parts, xend = x_total, y = yh, yend = yh,
           colour = "#5B6976", linetype = "dotted", linewidth = 0.6) +
  # the two stacked parts, left of the observation
  annotate("segment", x = x_parts, xend = x_parts, y = mh, yend = fh,
           colour = "#009E73", linewidth = 2.2) +
  annotate("segment", x = x_parts, xend = x_parts, y = fh, yend = yh,
           colour = "#0072B2", linewidth = 2.2) +
  # the total, right of the observation, drawn as a bracket
  annotate("segment", x = x_total, xend = x_total, y = mh, yend = yh,
           colour = "#17324D", linewidth = 1.1) +
  annotate("segment", x = x_total - cap, xend = x_total + cap,
           y = c(mh, yh), yend = c(mh, yh),
           colour = "#17324D", linewidth = 1.1) +
  # the other observations, the fitted value, and the observed value
  geom_point(data = decomp_df[-i, ], size = 2.6, colour = "grey75") +
  annotate("point", x = xh, y = fh, shape = 21, size = 3.2, stroke = 1.2,
           colour = "#17324D", fill = "white") +
  annotate("point", x = xh, y = yh, size = 3.6, colour = "#17324D") +
  annotate("text", x = xh, y = yh + 0.55,
           label = "y[i]*' (observed)'", parse = TRUE,
           colour = "#17324D", size = 4.2) +
  annotate("text", x = xh - 0.2, y = fh + 0.38,
           label = "hat(y)[i]*' (on the line)'", parse = TRUE,
           hjust = 1, colour = "#17324D", size = 4.2) +
  annotate("text", x = x_parts - 0.25, y = (mh + fh) / 2,
           label = "'explained by the line: '*hat(y)[i] - bar(y)",
           parse = TRUE, hjust = 1, colour = "#009E73", size = 4.2) +
  annotate("text", x = x_parts - 0.25, y = (fh + yh) / 2,
           label = "'residual: '*y[i] - hat(y)[i]",
           parse = TRUE, hjust = 1, colour = "#0072B2", size = 4.2) +
  annotate("text", x = x_total + 0.3, y = (mh + fh) / 2,
           label = "'total: '*y[i] - bar(y)",
           parse = TRUE, hjust = 0, colour = "#17324D", size = 4.2) +
  annotate("text", x = 11.9, y = mh - 0.38,
           label = "bar(y)*' (mean of Y)'", parse = TRUE,
           hjust = 1, colour = "#17324D", size = 4.2) +
  annotate("text", x = 2.5, y = 1.2, label = "fitted line",
           hjust = 0, colour = "#D55E00", size = 4.2) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  scale_y_continuous(breaks = seq(0, 10, 2)) +
  coord_cartesian(xlim = c(0, 12), ylim = c(0, 10.5), expand = FALSE) +
  labs(x = "Predictor X", y = "Outcome Y") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Now square each of the three pieces and add up over all observations. This
gives three sums of squares. SST is the total sum of squares, SSR is the
sum of squares explained by the regression, and SSE is the sum of squared
residuals from Lecture 1. For a least-squares line with an intercept, the
same split holds for the sums:

$$
\underbrace{\sum_i(y_i-\bar y)^2}_{\text{SST: total}}
=
\underbrace{\sum_i(\widehat y_i-\bar y)^2}_{\text{SSR: explained by the line}}
+
\underbrace{\sum_i(y_i-\widehat y_i)^2}_{\text{SSE: residual}}.
$$

The code below calculates the three sums for the valuation model and checks
that SSR and SSE add up to SST:

```{r}
#| label: i03-sums-squares
y <- vals$agreed_m
yhat <- fitted(val_model)
mean(y)   # mean agreed valuation, in $ millions

SST <- sum((y - mean(y))^2)
SSR <- sum((yhat - mean(y))^2)
SSE <- sum((y - yhat)^2)

c(SST = SST, SSR = SSR, SSE = SSE, check = SSR + SSE)
```

The units are squared millions of dollars, so the raw numbers (SST is about
12,000) mean little on their own. The sums of squares matter only relative
to each other. The next section compares them.

::: {#exm-i03-sst-sse}
## Worked example: from SST and SSE to SSR and R²

A regression output reports $\text{SST} = 500$ and $\text{SSE} = 125$. Find
SSR and $R^2$.

Because $\text{SST}=\text{SSR}+\text{SSE}$, the explained part is
$\text{SSR} = \text{SST} - \text{SSE} = 500 - 125 = 375$. Then

$$
R^2 = \frac{\text{SSR}}{\text{SST}} = \frac{375}{500} = 0.75
\qquad\text{or equivalently}\qquad
R^2 = 1 - \frac{\text{SSE}}{\text{SST}} = 1 - \frac{125}{500} = 0.75.
$$

The fitted values account for 75% of the variation of the outcome in the
sample. The other 25% remains as residual variation around the line.
:::

### R²: what 0.798 means, and four things it is not

::: {.definition-box}
<span class="label">Definition</span>

**$R^2$** (R squared). The share of the variation of the outcome in the
sample that the fitted line explains:

$$
R^2=\frac{\text{SSR}}{\text{SST}}
=1-\frac{\text{SSE}}{\text{SST}}.
$$

It lies between 0 and 1.
:::

For the valuation model:

```{r}
#| label: i03-r-squared
c(
  manual = 1 - SSE / SST,
  from_R = summary(val_model)$r.squared,
  correlation_squared = cor(vals$claimed_m, vals$agreed_m)^2
)
```

Here $R^2 \approx 0.798$. Among these deals, about 80% of the variation in
the agreed valuations is explained by the linear association with the
claimed valuation. The third number in the output shows a fact that holds
in simple regression: $R^2$ is the squared correlation of $X$ and $Y$,
$R^2 = r_{XY}^2$. Squaring removes the sign, so look at the slope or the
correlation for the direction of the association.

$R^2$ is often misread. Below are four things that it is not, each with a
wrong sentence and its correction. The first two are worked out here. The
last two are in the Try it task.

**Not the percentage of correct predictions.**
*Wrong:* "The model predicts 80% of agreed valuations correctly."
*Corrected:* the model explains 80% of the *variation* in agreed valuations.
Single predictions can still miss by millions of dollars. The standard
deviation of the residuals is about \$1.66 million, as the output of
`summary()` shows later in this section.

**Not proof that the assumptions hold.**
*Wrong:* "With $R^2$ this high, the model is clearly correct."
*Corrected:* $R^2$ measures fit. It does not check the assumptions. Lecture 2
found a fan in the residuals and influential deals in this same model, and
the value 0.798 says nothing about either.

::: {.callout-caution collapse="true"}
## Try it: correct the last two wrong sentences

Try this with your neighbour first. Correct each sentence in one line, as
above.

*Sentence 3 (you).* "The claim explains 80% of the valuation, so claiming
more will raise your valuation."

*Sentence 4 (you).* "The model will explain 80% of the differences in
valuations in next year's pitches too."

::: {.callout-note collapse="true"}
## Check your answers

*Sentence 3* claims causation. $R^2$ describes an association among
observed deals. Claims were not randomly assigned, and no measure of fit
says what would happen if a founder *changed* the claim.
*Sentence 4* claims that the fit holds for new data. $R^2$ is calculated on
this sample. A new season, a different selection of pitches, or a change in
how the sharks negotiate can all lower it. The fit is evidence about these
874 deals, not a guarantee for the next 874.
:::
:::

### The overall F-test

The last part of the driving question asks whether the model is useful at
all, compared with predicting the mean $\bar y$ for every deal. The overall
F-test answers it. It is stated here for a regression with $k$ predictors,
because Lecture 4 adds more predictors. In this lecture $k=1$.

::: {.definition-box}
<span class="label">Definition</span>

**Overall F-test.** A test of the null hypothesis that all slopes of the
model are zero, $H_0:\beta_1=\cdots=\beta_k=0$, against the alternative that
at least one slope is not zero. The test statistic is

$$
F=\frac{\text{SSR}/k}{\text{SSE}/(n-k-1)}.
$$

A large F with a small p-value is evidence that the model explains more than
the mean alone.
:::

Read the ratio part by part. The numerator is the explained variation per
predictor. The denominator is the residual variation per residual degree of
freedom. If no slope differs from zero, the two are of similar size and F is
near 1. A large F is evidence that the predictors together matter.

All numbers of this lecture appear together in the output of `summary()`.
Read it from top to bottom in this order: (1) the formula and the data, (2)
the coefficient estimates and their units, (3) the standard errors, t
values, and p-values, (4) the residual standard error, (5) $R^2$ and adjusted
$R^2$, (6) the overall F-test. Then remember (7), which the output does not
print: the assumptions and the limits of the data.

```{r}
#| label: i03-summary
summary(val_model)
```

Two lines of this output are new. The **residual standard error**, 1.664, is
the standard deviation of the residuals in the units of the outcome:
$s=\sqrt{\text{SSE}/(n-2)}$, here about \$1.66 million. A few very large
misses drive this number: Lecture 2 showed that half of the residuals lie
between $-0.40$ and $0.17$. Still, it is large next to the mean agreed
valuation of \$2.22 million, so single deals can lie far from the line even
though $R^2$ is high. **Adjusted $R^2$** is $R^2$ with a small
penalty for every predictor in the model, $1-(1-R^2)(n-1)/(n-k-1)$. It is
always a little below $R^2$. Unlike $R^2$, it can fall when a predictor that
adds almost nothing is included. That makes it the fairer number for
comparing models with different numbers of predictors. With one predictor
and 874 observations the two hardly differ (0.7975 and 0.7973).

The last line reports $F \approx 3435$ on 1 and 872 degrees of freedom. The
p-value is printed as `< 2.2e-16`, which means practically zero. The test
rejects the null hypothesis: under the model assumptions, the model with the
claimed valuation fits better than a model with only an intercept.

**In simple regression, F and t answer the same question.** With one
predictor, the null hypothesis of the F-test is $H_0:\beta_1 = 0$, the same
as in the two-sided t-test of the slope, and

$$
F = t^2.
$$

```{r}
#| label: i03-f-t-link
t_slope <- summary(val_model)$coefficients["claimed_m", "t value"]
f_model <- unname(summary(val_model)$fstatistic["value"])
c(t_squared = t_slope^2, F = f_model)
```

Both equal about 3435. This holds only in simple regression. In multiple
regression the F-test looks at several slopes together, while each t-test
looks at one slope. Lecture 5 uses this difference to test several slopes at
once.

### Usefulness has more than one meaning

The question "Is the model useful?" has at least four different meanings,
and a model can pass one and fail another. The four examples below are
invented:

- **Slope clearly different from zero, low $R^2$.** A loyalty-programme
  analysis on 400,000 customers finds a positive association between
  mailings and spending. It is precisely estimated, and it explains 1% of
  the variation in spending. This is useful for a decision about all
  customers, and useless for predicting the spending of one customer.
- **High $R^2$, no causal reading.** Store revenue regressed on the number
  of staff fits very well, because both follow the size of the store.
  Hiring more staff at a small store will not produce the revenue of a big
  store.
- **Good fit on the data used, poor prediction for new data.** A
  travel-demand model fitted on data for 2015--2019 explained bookings very
  well and failed in 2020. The $R^2$ of the fitted data never promised that
  the pattern would continue.
- **Accurate predictions, no explanation.** A complex prediction model may
  forecast weekly orders very well while its coefficients do not explain
  *why* orders change. That is fine for stocking decisions, and useless for
  planning a price change.

Always judge a model by the goal: explanation, prediction, or support for a
decision.

### Robust standard errors: what they fix {#i03-robust}

Heteroskedasticity-consistent standard errors, called **robust** standard
errors for short, do not need the constant-variance assumption. The fitted
coefficients stay the ordinary least-squares values. Only the standard
errors change, and with them the t values, p-values, and intervals. Lecture 2
found a fan in the residuals of this model, so robust standard errors are a
sensible choice here:

```{r}
#| label: i03-robust-code
library(lmtest)
library(sandwich)

coeftest(val_model, vcov. = vcovHC(val_model, type = "HC1"))
coefci(val_model, vcov. = vcovHC(val_model, type = "HC1"))
```

The function `coeftest()` prints the coefficient table with robust standard
errors, and `coefci()` prints the robust 95% confidence intervals. The
robust standard error of the slope is 0.034, about three times the usual
0.0111. The robust 95% interval runs from about 0.58 to 0.72. It is three
times as wide as the usual interval and still lies far below 1.

Robust standard errors fix exactly one problem: an error variance that is
not constant. They
do not repair a curved relationship, confounding, dependence between rows,
influential data errors, or a selected sample. A robust standard error on a
biased slope measures the precision of the wrong number. You do not have to
calculate robust standard errors by hand. You should know what they are for.

::: {.caveat-box}
<span class="label">What a precise slope and a high R² can and cannot say</span>

- Every number of this lecture is calculated under the assumptions of
  Lecture 2. The fan makes the usual standard error doubtful, and a few
  deals shape the fit.
- The data are deals on a television show, not a random sample of
  companies. This selection affects every estimate.
- Precision and fit are not causation.

**What you can do with it:** use the slope and its interval for the decision,
use the residual standard error to judge how far single cases lie from the
line, and prefer robust standard errors when the residual plot shows a fan.
:::

### Answering the driving question

The slope is known precisely, the line explains much of the outcome, and the
model is useful, all under conditions. The slope 0.649 has standard error
0.0111. Its 95% interval $(0.627, 0.670)$ is only about four cents per
claimed dollar wide. With robust standard errors it is about 13 cents wide,
from 0.58 to 0.72. The t-test clearly rejects a zero slope
($t \approx 58.6$). The line explains about 80% of the variation in agreed
valuations, and the overall F-test ($F \approx 3435$) confirms that the
model fits better than the mean alone. All of these statements depend on the
assumptions checked in Lecture 2, where the residuals showed a fan and a few
influential deals. None of them is causal.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your advertising slope is 0.15 orders per euro, with a standard error of 0.05. The 95% interval runs from 0.05 to 0.25: each 100 euros of advertising goes with between 5 and 25 extra orders. At 4 euros of margin per order, that is between 20 and 100 euros. Even the best case only breaks even. The decision: reduce the advertising budget. The model has an $R^2$ of 0.20, so it predicts single weeks badly. That does not matter here. The decision depends on the slope and its interval, not on predicting next week.
:::

**Next:** Tutorial 1 practises Lectures 1--3 in the format of the exam: fit a
model in R, check its residuals, and report its coefficients with intervals,
tests, and $R^2$. The exam is taken in TestVision with R available.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 80 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Why are the residual degrees of freedom $n-2$ in simple regression?
2. Which pentad quantities share the slope's units, and which are
   unit-free?
3. What do SST, SSR, and SSE measure, and how are they related?
4. Why is $R^2$ between 0 and 1 in an OLS model with an intercept?
5. What is the null hypothesis of the overall F-test, and why does $F=t^2$
   in simple regression?
6. What single problem do robust standard errors address?

::: {.callout-note collapse="true"}
## Check your answers

1. Two coefficients, intercept and slope, are estimated before
   residual variation can be measured, using up two pieces of information.
2. The estimate, its SE, and the interval endpoints share the slope's
   units. The t statistic and p-value are unit-free.
3. Total, explained (fitted), and residual sample variation in the outcome.
   $\text{SST}=\text{SSR}+\text{SSE}$ for OLS with an intercept.
4. SSE is non-negative and no larger than SST, so $1-\text{SSE}/\text{SST}$
   lies between 0 and 1.
5. All population slope coefficients are zero. With one predictor that is
   the same hypothesis as the two-sided slope t-test, and $F=t^2$.
6. An error variance that is not constant, and nothing else.
:::

### Finish a half-worked example: an interval for a new row (15 minutes)

Re-read the worked slope interval (@exm-i03-interval). Then build the
interval for the **intercept row** of the same table.

*Step 1 (given).* From the coefficient table: estimate $-0.068$, standard
error $0.0685$, degrees of freedom $872$, multiplier $\approx 1.96$.

*Step 2 (you).* Compute the 95% interval.

*Step 3 (you).* Interpret it in one sentence, and say why this interval is of
little use for a business decision (see the intercept in Lecture 1).

::: {.callout-note collapse="true"}
## Check your answer

$-0.068 \pm 1.96 \times 0.0685 = -0.068 \pm 0.134 = (-0.202,\; 0.066)$, in
\$ millions, which matches R. Under the model assumptions, the data are
consistent with a population intercept between about $-\$202{,}000$ and
\$66,000. The interval includes zero, so the line may well pass through the
origin. But the intercept is the fitted mean at a claimed valuation of zero,
outside the observed range of claims. The interval is calculated correctly
and still tells a founder little.
:::

Then four more tasks without given steps:

1. A study reports a coefficient of 2.40 with standard error 0.80 in a
   large sample. Compute the t statistic, decide the 5% two-sided test
   against zero, and give the 95% interval.
2. Interpret $R^2=0.20$ without calling the model "80% wrong."
3. A model has $k=3$, $n=104$, SSR = 300, and SSE = 500. Calculate $R^2$
   and F.
4. Explain why adding a predictor to a least-squares model cannot increase
   its SSE.

::: {.callout-note collapse="true"}
## Check your answers

1. $t = 2.40/0.80 = 3$. Reject at 5% (well beyond 1.96). Interval:
   $2.40 \pm 1.96 \times 0.80 = (0.83,\; 3.97)$.
2. The fitted values explain 20% of the variation of the outcome in the
   sample. 80% remains as residual variation around the fitted line. Nothing
   is "wrong". The fitted values are simply far from many single outcomes.
3. $\text{SST}=300+500=800$, so $R^2=300/800=0.375$.
   $F=(300/3)/(500/(104-3-1))=100/5=20$.
4. The larger model can reproduce the old fit by setting the new slope to
   zero. Least squares picks the smallest SSE, so the SSE of the larger
   model cannot be above the old one.
:::

### Practice: all steps on equity and season (20 minutes)

Fit `equity_offered_pct ~ season` on all 1,441 pitches. Report the slope's
95% confidence interval, the t-test against zero, $R^2$, and the overall
F-test, then write a 120-word conclusion that separates estimate,
uncertainty, fit, and scope.

::: {.callout-note collapse="true"}
## Check a model answer

```{r}
#| label: i03-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
equity_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(equity_model)$coefficients
confint(equity_model)
c(r_squared = summary(equity_model)$r.squared,
  F = unname(summary(equity_model)$fstatistic["value"]))
```

A complete conclusion reports four things. Estimate: one later season
(`season`) is associated with about 0.824 percentage points less equity
offered (`equity_offered_pct`), with standard error 0.048. Uncertainty: the
95% interval $(-0.917, -0.731)$ excludes zero, and the t-test rejects a zero
slope ($t\approx-17.3$, p $<$ 0.001). Fit: the season explains about 17.3%
of the variation in the equity offered, and the overall F-test
($F\approx300.7 = t^2$) rejects the model with only an intercept. Scope:
this is a trend over time among recorded pitches, not a causal effect of
the season, and the standard errors rest on the assumptions of Lecture 2.
:::

### Review: one diagnostics question from Lecture 2 (5 minutes)

The residual-versus-fitted plot of the valuation model shows unequal spread,
the fan of Lecture 2. Which numbers of the coefficient table in this lecture
does that pattern put most directly in doubt, and why?

::: {.callout-note collapse="true"}
## Check your answer

The standard error 0.0111, and with it the t value, the p-value, and the
confidence interval that are built from it. The usual formula for the
standard error assumes one common error variance $\sigma^2$. A fan says that
the variance is not constant, so the stated precision is doubtful. Robust
standard errors address exactly this one problem. They repair nothing else.
:::

### Worked exam interpretation (5 minutes)

**Question** (mock exam Question 3 pattern). A model of venture sales
growth (percentage points) on marketing spending (€10,000s), fitted on
$n = 102$ ventures, reports a slope of 1.38 with standard error 0.46,
$R^2 = 0.083$, and $F = 9.0$. Build the 95% interval for the slope, decide
the 5% test against zero, verify the F statistic from the t statistic, and
state in one sentence what the low $R^2$ does and does not imply.

::: {.callout-note collapse="true"}
## Check a model answer

$t = 1.38/0.46 = 3.0$. With $n-2 = 100$ degrees of freedom the multiplier
`qt(0.975, 100)` is about 1.98, so the
interval is $1.38 \pm 1.98\times0.46 \approx (0.47,\; 2.29)$ percentage
points per €10,000. The interval excludes zero, so the two-sided 5% test
rejects $H_0:\beta_1=0$. In agreement with this, $F = t^2 = 9.0$. The low
$R^2$ says that marketing spending explains about 8% of the variation in
growth, so predictions for single ventures will be poor. It does not remove
the evidence that the association is not zero. Neither number makes the
association causal.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "R squared is 0.80 and the p-value is below 0.001. Is my model good?" It answered:

> "Yes. The model predicts 80% of the valuations correctly, and the tiny p-value proves that the model assumptions hold."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Two errors. R squared is the share of the variation that the line explains, not a share of correct predictions. Single predictions can still miss by millions of dollars (the residual standard error is about 1.7 million). And the p-value is calculated under the assumptions. It cannot prove them. Lecture 2 showed a fan in the residuals, so the standard error behind this p-value is doubtful.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture, or any business you know well, and one
regression you might run on its data. Write two sentences: first say which
you would rather have for your next decision, a precise slope (narrow
interval) or a high $R^2$, and why. Then name the meaning of "useful" from
the list of four in this lecture that matters most for that decision. There
is no wrong answer. Precision and fit answer different questions.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "For setting next month's ad budget I would take a precise slope over a high $R^2$: I need to know what an extra €100 buys, even if week-to-week orders stay noisy. That is the first meaning on the list, a reliable slope with a low $R^2$: predicting individual weeks matters less to me than the payoff of an input I control."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to invent a business
model with a significant slope and a low $R^2$ and challenge you to say
what the model is still useful for.
::::

[Continue to Tutorial 1: Simple regression, diagnostics, and model usefulness](t01-exam-prep.qmd).

[Code sheet: Lecture 3 commands and formulas](code-sheet.qmd#lecture-3-coefficient-inference-fit-and-the-f-test).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.1.2--3.1.3, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 19--20.

:::
