# Lecture 4: Multiple regression and what controls do {#i04}

```{r}
#| label: i04-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 4, you can</span>

- interpret a coefficient in a multiple regression, holding other predictors fixed
- explain why a coefficient can change when you add a predictor
- compare several models side by side in one regression table
- say what adding a control variable can and cannot achieve
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Two different explanations can produce the same pattern. Perhaps the equity offered (<code>equity_offered_pct</code>) in the Shark Tank file really fell over the seasons. Or perhaps later pitches simply ask for different amounts of money. Multiple regression separates two explanations like these.</p>

Later seasons show smaller equity offers. Is the season the reason? Or is
the requested amount the real reason, because later pitches also ask for
different amounts?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=o9hoLdylWKo&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=11">3.3
Multiple Linear Regression</a>. Focus on how one coefficient is interpreted
when the other predictors are held fixed, and on why several predictors are
included together.
All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
The videos explain key ideas in another way and can use a different
notation. The lecture material and its notation always come first.
</div>

**Predict.** Answer this question before you come to class: once the requested amount is added, will the
season coefficient move towards zero, away from zero, or barely
move? Think about why the slope might change at all. If the requested
amount is related to both the season and the equity offered, the simple slope
mixes the two relationships. Choose one answer. You will find the answer in
the lecture material.

[Shark Tank data]{.shark-tag} **R warm-up.** Make the prediction above first, then run this
code. In Lecture 3 you fitted the simple model of the equity offered
(`equity_offered_pct`) on season and tested its slope. Fit it again. Then fit
the model with two predictors, which adds the requested amount in thousands
of dollars (`ask_1k`). Write down both season coefficients and bring them to
class:

```{r}
#| label: i04-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
coef(lm(equity_offered_pct ~ season, data = sharks))
coef(lm(equity_offered_pct ~ season + ask_1k, data = sharks))
```

**Code along (bring your laptop).** Bring a charged laptop with RStudio and
the course project open. All the code of the lecture is in this chapter,
ready to copy. You can run each step yourself during the lecture and get the
same numbers and figures.

If you are unsure how subscripts or equations with several variables work,
review
[Mathematics Refresher: subscripts and indices](../resources/math-refresher.html#subscripts-and-indices).

:::

::: {.course-phase .phase-in-class}

## Lecture material

The R task before class produced two season coefficients: about $-0.824$
without the requested amount and about $-0.814$ with it. This section
explains what the second number means, why the two numbers can differ, and
why they hardly differ here.

### Start with a founder's question

Your sales rose after you cut the price. But it was also December. Was it
the lower price or the time of year? Founders often face this problem: two
factors changed at the same time, and a simple comparison cannot separate
them. The Shark
Tank file has the same problem. The equity offered (`equity_offered_pct`)
fell over the seasons, and the requested amounts also changed over the
seasons. Multiple regression solves this problem by comparing pitches that asked for
the same amount.

### More than one predictor

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=o9hoLdylWKo">3.3 Multiple Linear Regression</a>.</p>

For predictors $X_{i1},\ldots,X_{ik}$, the model is

$$
Y_i=\beta_0+\beta_1X_{i1}+\cdots+\beta_kX_{ik}+\varepsilon_i,
$$

with conditional mean

$$
E[Y_i\mid X_{i1}=x_1,\ldots,X_{ik}=x_k]
=\beta_0+\beta_1x_1+\cdots+\beta_kx_k.
$$

::: {.definition-box}
<span class="label">Definition</span>

**Partial regression coefficient** (often called partial slope). The
coefficient $\beta_j$ of one predictor in a model with several predictors. It is the expected difference in $Y$ that goes with a
one-unit difference in $X_j$, holding the other included predictors fixed.

**Holding the other predictors fixed.** The comparison is between cases that
have the same values of all other predictors in the model and differ only in
$X_j$. Econometrics textbooks use the Latin term ceteris paribus, which
means "other things equal".
:::

Write the phrase "holding the other included predictors fixed" in every
interpretation of a slope from a multiple regression.

The model makes this comparison. The data do not need to contain two
real pitches that are the same in every other respect. The model compares
conditional means at fixed values of the included predictors.

### Why coefficients change: an invented example

Start with an invented example. We build the data ourselves, so we know the
true relationships. Imagine 200 small firms. One background variable, the size of the
firm (`size`), influences both advertising spend (`adspend`) and revenue
(`revenue`). Advertising itself adds nothing to revenue: the line of code
that builds `revenue` contains `size` but not `adspend`.

```{r}
#| label: i04-sim-build
set.seed(42)
n <- 200
size    <- rnorm(n, mean = 10, sd = 2)          # Z: firm size (drives both)
adspend <- 2 + 0.8 * size + rnorm(n, sd = 1)    # X: bigger firms spend more
revenue <- 5 + 1.5 * size + rnorm(n, sd = 1.5)  # Y: size drives revenue,
demo <- data.frame(revenue, adspend, size)      #    adspend does not
coef(lm(revenue ~ adspend, data = demo))
coef(lm(revenue ~ adspend + size, data = demo))
```

The simple slope of `adspend` is about $1.39$, clearly positive. Yet we know
that advertising does nothing in these data. Once `size` is in the model, the
slope of `adspend` falls to about $0.10$, close to the zero that we built in.
The simple slope counted the effect of differences in size as an effect of
`adspend`. As a
description the simple slope is correct: firms that advertise more do have
higher revenue on average. As an answer to the question of what advertising
does, it is wrong.

::: {.definition-box}
<span class="label">Definition</span>

**Omitted variable.** A variable that is related to the outcome but is left
out of the model.

**Confounder.** An omitted variable that is related to both a predictor and
the outcome. The slope of that predictor then also picks up the relationship
between the confounder and the outcome. This distortion is called
omitted variable bias. In the invented example, `size` is the confounder of
`adspend`.
:::

@fig-i04-ovb-scatter shows why. Larger firms (orange) are high on both axes.
Smaller firms (blue) are low on both. If you move from left to right through
the whole cloud of points, you move from small firms to large firms. So the
simple slope compares different kinds of firms. Within each colour the points
form a nearly flat cloud: at a given size, more advertising does not go with more revenue.

```{r}
#| label: fig-i04-ovb-scatter
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Invented data: revenue against advertising spend for 200 simulated firms, coloured by firm size. The upward direction of the whole cloud comes from size alone. Within a colour group the relationship is nearly flat."
library(ggplot2)
set.seed(42)
n <- 200
size    <- rnorm(n, mean = 10, sd = 2)          # Z: firm size (drives both)
adspend <- 2 + 0.8 * size + rnorm(n, sd = 1)    # X: bigger firms spend more
revenue <- 5 + 1.5 * size + rnorm(n, sd = 1.5)  # Y: size drives revenue,
demo <- data.frame(revenue, adspend, size)      #    adspend does not
demo$group <- ifelse(demo$size > median(demo$size),
                     "Larger firms (size above median)",
                     "Smaller firms (size below median)")
simple_fit <- lm(revenue ~ adspend, data = demo)

ggplot(demo, aes(x = adspend, y = revenue,
                 colour = group, shape = group)) +
  geom_abline(intercept = coef(simple_fit)[1],
              slope = coef(simple_fit)[2],
              colour = "#17324D", linewidth = 1.1) +
  geom_point(alpha = 0.75, size = 2) +
  annotate("text", x = max(demo$adspend) - 0.2,
           y = coef(simple_fit)[1] + coef(simple_fit)[2] *
             (max(demo$adspend) - 0.2) - 2.4,
           label = "simple regression line",
           hjust = 1, colour = "#17324D", size = 3.4) +
  scale_colour_manual(values = c(
    "Larger firms (size above median)" = "#D55E00",
    "Smaller firms (size below median)" = "#0072B2"
  )) +
  scale_shape_manual(values = c(
    "Larger firms (size above median)" = 17,
    "Smaller firms (size below median)" = 16
  )) +
  labs(x = "Advertising spend (invented units)",
       y = "Revenue (invented units)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

A coefficient changes when the added predictor explains differences that
the simple slope had picked up. This is why the phrase "holding the other
included predictors fixed" matters: it says which comparison the estimate
makes.

### The real-data contrast: the season slope barely moves

[Shark Tank data]{.shark-tag} Now make the same comparison on the Shark Tank data. From here on, money is
measured in thousands of dollars: `ask_1k` is the requested amount divided by
1,000. One unit is \$1,000 and 100 units are \$100,000. Lecture 1 showed how
a coefficient changes when the units of measurement change.

```{r}
#| label: i04-compare-slopes
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
m_simple   <- lm(equity_offered_pct ~ season, data = sharks)
m_adjusted <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
c(
  simple           = coef(m_simple)["season"],
  adjusted_for_ask = coef(m_adjusted)["season"]
)
```

The season slope moves from about $-0.824$ to about $-0.814$. That is a
change of 0.01 percentage points per season, about 1% of the slope. This
answers the prediction
question from before class: the coefficient barely moves.
In the invented example, adding `size` removed almost the whole simple slope.
Here, adding the requested amount leaves the season trend almost unchanged.
In these data, the season trend does not come from differences in the
requested amount.

The stable slope tells you something useful: the requested amount does not
explain the season trend. It does not rule out every other explanation. A
variable that is not in the file, such as the maturity of the venture or the
experience of the founder, could still play the role that `size` played in
the invented data. A stable slope tells you about the variables you added.
It tells you nothing about variables that are not in the file.

### A three-predictor Shark Tank model

[Shark Tank data]{.shark-tag} Now add a third predictor. The outcome is the equity that the founders
offered at the start of the pitch (`equity_offered_pct`), in percent. The
predictors are:

- `season`: the season number, 1 to 16
- `ask_1k`: the requested amount in thousands of dollars
- `description_words`: the number of words in the short description of the
  business in the data source.

```{r}
#| label: i04-fit
equity_model <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
summary(equity_model)
```

The fitted equation is approximately

$$
\widehat{\text{equity}}
=22.025-0.836\,\text{season}-0.0029\,\text{ask\_1k}
-0.128\,\text{description\_words}.
$$

#### The coefficient-interpretation protocol

Use the same five checks every time you interpret a slope:

1. **Estimate and direction:** state the fitted change and its sign.
2. **Predictor step and unit:** say what a one-unit increase means.
3. **Outcome unit:** state the unit in which the outcome changes.
4. **Conditioning:** name the other included predictors that are held fixed.
5. **Scope:** name the analysed records and write "associated with", unless
   the data come from an experiment that justifies a causal claim.

The three predictors have different units:

| Term | A one-unit increase means | Coefficient unit |
|---|---|---|
| `season` | one later season | equity percentage points per season |
| `ask_1k` | \$1,000 more requested | equity percentage points per \$1,000 |
| `description_words` | one additional word | equity percentage points per word |

Apply the five checks to each slope in turn. All three sentences are about
the 1,441 recorded pitches:

- **Season.** At the same requested amount and description length, one later
  season is associated with about $0.836$ percentage points less equity
  offered.
- **Requested amount.** At the same season and description length, an
  additional \$1,000 requested is associated with about $0.0029$ percentage
  points less equity offered. The number is tiny because the unit step is
  tiny. Per additional \$100,000 the association is about $0.29$ percentage
  points. The t statistic (about $-5.0$) does not depend on the unit.
- **Description length.** At the same season and requested amount, one
  additional description word is associated with about $0.128$ percentage
  points less equity offered. This estimate is not statistically
  distinguishable from zero at 5% (p $\approx$ 0.26). Report the estimate
  anyway. "Not significant" does not mean "exactly zero". It means that the
  data are consistent with a range of values that includes zero.

::: {.callout-note collapse="true"}
## Check: the season slope was −0.814 with two predictors and −0.836 with three. Contradiction?

No. The two coefficients answer different questions. $-0.814$ holds only
`ask_1k` fixed. $-0.836$ holds `ask_1k` and `description_words` fixed.
Adding a predictor changes what is held fixed, so the coefficient may change.
Here the change is small, as it was when `ask_1k` was added.
:::

### Three models in one table {#i04-spec-table}

[Shark Tank data]{.shark-tag} This lecture has now fitted three models of the equity offered
(`equity_offered_pct`): season alone (`m_simple`), season and requested
amount (`m_adjusted`), and the three-predictor model (`equity_model`).
Research papers show such models side by side in one table, with one column
per model. The table needs the standard error of every coefficient and the
fit of every model. `summary()` prints these numbers for the first two models
below. The output for the third model is in the previous section.

```{r}
#| label: i04-spec-summaries
summary(m_simple)
summary(m_adjusted)
```

The table below collects the numbers of the three outputs above.

::: {.reg-table}
| | (1) | (2) | (3) |
|:---|---:|---:|---:|
| Season | −0.824\*\*\* | −0.814\*\*\* | −0.836\*\*\* |
| | (0.048) | (0.047) | (0.051) |
| Amount requested (\$1,000s) | | −0.0029\*\*\* | −0.0029\*\*\* |
| | | (0.0006) | (0.0006) |
| Description length (words) | | | −0.128 |
| | | | (0.114) |
| Intercept | 20.457\*\*\* | 21.185\*\*\* | 22.025\*\*\* |
| | (0.464) | (0.483) | (0.888) |
| Observations | 1,441 | 1,441 | 1,441 |
| $R^2$ | 0.173 | 0.187 | 0.188 |
| Adjusted $R^2$ | 0.172 | 0.186 | 0.186 |
:::

Outcome: equity offered (%). Ordinary least squares. Standard errors in
parentheses. All three columns use the same 1,441 pitches.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001.

Each column is one model and each row is one predictor. An empty space means
that the predictor is not in that model. The number in parentheses under a
coefficient is its standard error, and the stars show the p-value of the
two-sided test of that coefficient against zero. Read along a row to see how
a coefficient changes when predictors are added: the season coefficient stays
between $-0.836$ and $-0.814$ in all three columns. Read down a column to see
one model: column (3) is `equity_model`. The bottom rows give the number of
observations and the fit of each model. This is the layout to use for
regression results in your thesis.

The last two rows need one more term, which Lecture 3 introduced.

::: {.definition-box}
<span class="label">Definition</span>

**Adjusted $R^2$.** $R^2$ with a small penalty for every predictor in the
model. $R^2$ never falls when a predictor is added. Adjusted $R^2$ can fall
when the new predictor adds almost nothing. Use adjusted $R^2$ to compare
models with different numbers of predictors.
:::

The formula is on the
[code sheet](code-sheet.qmd#lecture-3-coefficient-inference-fit-and-the-f-test).
In the table, adding the requested amount raises $R^2$ from 0.173 to 0.187.
The model then accounts for 18.7% of the variation in the equity offered, up
from 17.3%. Adjusted $R^2$ rises with it, from 0.172 to 0.186. Adding the
description length leaves adjusted $R^2$ at 0.186: this predictor adds almost
nothing to the fit.

R can build such a table for you with the package `modelsummary`:

```{r}
#| label: i04-modelsummary
#| eval: false
library(modelsummary)
modelsummary(
  list("(1)" = m_simple, "(2)" = m_adjusted, "(3)" = equity_model),
  stars = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)
```

You need to install the package once with
`install.packages("modelsummary")`.
It rounds to three decimals, so in its table the coefficient of the
requested amount appears as −0.003.

### Scaling to comparisons readers care about

A reader rarely cares about one season or one thousand dollars. Scale the
coefficient to a step that matters to the reader: fitted change =
coefficient × number of unit steps, holding the other included predictors
fixed.

::: {#exm-i04-scaling}
## Worked example: five seasons, half a million dollars, ten words

[Shark Tank data]{.shark-tag} Holding the other included predictors fixed in `equity_model`:

- **Five later seasons:** $5\times(-0.836)=-4.18$ fitted equity percentage
  points.
- **$500,000 more requested:** $500{,}000/1{,}000 = 500$ units of `ask_1k`,
  so $500\times(-0.0029)=-1.45$ fitted percentage points.
- **Ten additional description words:** $10\times(-0.128)=-1.28$ fitted
  percentage points. This coefficient is not significantly different from
  zero, so the scaled version is not either.

Changing the units of measurement of a predictor changes the coefficient and
its standard error by the same factor. It does not change the fitted values, the t
statistic, or the relationship itself. A coefficient "per dollar" of
$-0.0000029$ would be correct but hard to read. Per \$1,000 is readable. Per
\$100,000 is often the best unit for a sentence.
:::

::: {.callout-caution collapse="true"}
## Try it: three seasons later and $200,000 more

[Shark Tank data]{.shark-tag} Try this with your neighbour first. Two pitches have the same description
length. One is three seasons later and requests $200,000 more. Complete the
steps before opening the answer.

*Step 1 (given).* Three seasons contribute $3\times(-0.836)$ fitted
percentage points.

*Step 2 (you).* Convert $200,000 into `ask_1k` units and multiply by the
`ask_1k` coefficient.

*Step 3 (you).* Add the two contributions. This is allowed because the model
is additive: each predictor has its own term, and the terms are added up.
State the combined fitted difference in one sentence that passes the five
checks.

::: {.callout-note collapse="true"}
## Check your answer

$200{,}000/1{,}000=200$ units, so the request contributes
$200\times(-0.0029)\approx-0.58$. Combined:
$3(-0.836)+200(-0.0029)\approx-2.51-0.58=-3.09$ percentage points. Sentence:
among the recorded pitches, a pitch three seasons later that requests
\$200,000 more is fitted to offer about 3.1 percentage points less equity,
holding description length fixed.

```{r}
#| label: i04-faded-check
b <- coef(equity_model)
3 * b["season"] + 200 * b["ask_1k"]
```
:::
:::

### What control variables can and cannot do

A control variable (often just called a control) is a predictor that is
included in the model so that the comparison holds it fixed. Including a
control variable can:

- make a comparison fairer, because the compared cases are alike on the
  measured variables
- reduce the residual variation, which makes confidence intervals narrower
- show that a simple association partly came from comparing different kinds
  of cases.

A control variable also has a cost: it makes a coefficient less precise when
the predictors are strongly correlated.

Including a control variable cannot:

- adjust for variables that were not measured
- fix reverse causality, where the outcome influences the predictor
- repair selection, where the cases in the file are not typical of the
  cases you want to talk about
- remove errors in how a variable was measured.

A control variable is simply a predictor like any other. The word
"control" does not make any coefficient causal. Only the way the data
were produced can make a coefficient causal, for example when the data come
from an experiment.

::: {.callout-note collapse="true"}
## Check: which of these four problems does adding `ask_1k` address?

&#40;a) Later pitches request different amounts. (b) The experience of the
founder is not recorded. (c) The equity offered may influence how much is
asked. (d) Only pitches that were broadcast are in the file.

Only (a). The requested amount is measured, and holding `ask_1k` fixed
removes it from the season comparison. Problem (b) is an unmeasured
confounder, (c) is reverse causality, and (d) is selection. Adding a measured
control variable does not solve any of these three.
:::

### Assumptions extend to the full predictor set

For multiple regression, the Lecture 2 assumption list applies to the whole
set of included predictors:

1. The conditional mean is linear in the included terms (linearity).
2. Each predictor varies, and no predictor can be calculated exactly from
   the others, for example as the sum of two of them (no perfect
   multicollinearity).
3. The observations are independent (random sampling).
4. The usual standard errors require homoskedasticity, a constant
   conditional error variance.
5. In small samples, exact t and F inference also requires normally
   distributed errors.
6. Zero conditional mean of the error given all included predictors. For a
   causal reading of a coefficient, this means that no omitted confounder
   drives both an included predictor and the outcome.

A high correlation among the predictors is called **multicollinearity**.
It does not bias the coefficients. It makes the single coefficients imprecise and
unstable, because the data contain little information to separate predictors
that move together. It is worse when one predictor can be calculated exactly
from the others. This is called perfect multicollinearity. R then cannot
estimate all coefficients and leaves one term out. Multicollinearity makes
the estimates less precise. With perfect multicollinearity, one coefficient
cannot be estimated at all.

### A coefficient plot keeps units visible

[Shark Tank data]{.shark-tag} @fig-i04-coefplot shows the three slope estimates of `equity_model` with
their 95% confidence intervals.

```{r}
#| label: fig-i04-coefplot
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Coefficient estimates from the three-predictor model of the equity offered, with 95% confidence intervals. Each row uses a different predictor unit, so the sizes cannot be compared across rows without the table of units."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
equity_model <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
ci <- confint(equity_model)[-1, , drop = FALSE]
term_labels <- c("Season", "Amount requested ($1,000s)",
                 "Description length (words)")
coef_df <- data.frame(
  term = factor(term_labels, levels = term_labels),
  estimate = coef(equity_model)[-1],
  lower = ci[, 1],
  upper = ci[, 2]
)

ggplot(coef_df, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  colour = "#009E73", linewidth = 0.7, fatten = 2.6) +
  annotate("text", x = -0.02, y = 3.35, label = "no change = 0",
           hjust = 1, colour = "#17324D", size = 3.4) +
  labs(x = "Estimated change in equity offered (percentage points)",
       y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-i04-coefplot with the
[five steps for reading a figure](../basic/overview.qmd#how-to-read-every-graph):

1. **Unit and variables:** the rows are the predictors. The horizontal axis
   is the estimated change in the equity offered, in percentage points.
2. **Scales:** each estimate is per one unit of its own predictor: per
   season, per \$1,000, per word. The rows are not on a common scale.
3. **One element:** a dot is a coefficient estimate, and its horizontal line
   is the 95% confidence interval. The dashed line marks zero.
4. **Pattern:** the season interval lies clearly left of zero. The interval
   for the description length crosses zero. The `ask_1k` row lies almost on
   the zero line only because its unit step (\$1,000) is tiny: its t
   statistic is about $-5$. The distance from zero in this plot depends on
   the unit. It does not show how strong the evidence is.
5. **Limits:** the intervals show sampling uncertainty under the model,
   among these recorded pitches.

::: {.caveat-box}
<span class="label">What an adjusted coefficient can and cannot say</span>

- "Holding fixed" covers only the predictors in the model. The maturity of
  the venture, its revenue, and the experience of the founder are not held
  fixed.
- If a coefficient stays stable across models, the measured control
  variables do not explain it. This proves nothing about variables that were not
  measured.
- The equity offered is the opening offer in the pitch, not a completed
  investment.

**What you can do with it:** add the control variables that a critic would name first.
If your coefficient stays about the same after you add them, you can act on it
for cheap decisions. For expensive decisions, test it first.
:::

### Answering the driving question

At the same requested amount and description length, the season trend
remains: about $-0.836$ equity percentage points per season, against $-0.824$
without the two control variables. The requested amount does not explain the season trend.
Later pitches offer less equity even when they are compared at the same
requested amount.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Suppose you run a campus venture that sells meal boxes. In weeks with lower prices you get more orders: the simple slope is 60 more orders per euro of price cut. But you lowered prices mainly in exam weeks, when demand is high anyway. With exam weeks included in the model as a control variable, the slope is 25 orders per euro. You sell about 150 meal boxes a week at a margin of 4 euros per box. A price cut of 1 euro costs 150 euros of margin on the boxes that you would sell anyway. After the cut, each extra order earns a margin of 3 euros. With the simple slope, the cut would bring 60 more orders at 3 euros each, so 180 euros, and the cut would be profitable. With the adjusted slope, the cut brings 25 orders, so 75 euros, which is less than the 150 euros that the cut costs. The decision: keep the price. The control variable changed the decision.
:::

**Next:** You can now hold other predictors fixed, as long as they are
numbers. Many business comparisons are between categories: sales channels,
customer segments, kinds of teams. Lecture 5 shows how categories enter a
regression and how to test a whole set of category terms at once.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main ideas. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly, the course AI
assistant.

### Recall questions

1. Which phrase belongs in every interpretation of a slope from a multiple
   regression?
2. Why can a coefficient change after another predictor is added?
3. What does the zero conditional mean condition (no omitted confounder) say, and when is it needed?
4. Why does adding control variables not make a coefficient causal?
5. What problem can strongly correlated predictors create?
6. Which five elements belong in the interpretation of a slope?

::: {.callout-note collapse="true"}
## Check your answers

1. "Holding the other included predictors fixed."
2. The simple slope also picks up the relationship between the outcome and an
   omitted variable that is related to the predictor. When that variable is
   added, the slope no longer includes this part.
3. The error averages zero at every value of the predictors: nothing outside
   the model drives both an included predictor and the outcome. It is needed only when a coefficient is read as the effect of
   changing that predictor.
4. Unmeasured confounders, selection, reverse causality, and measurement
   error can all remain after a control variable is added.
5. Imprecise and unstable estimates of the single coefficients. With perfect
   multicollinearity R cannot estimate all coefficients and leaves one term out.
6. Estimate and direction, predictor step and unit, outcome unit, the
   predictors that are held fixed, and the scope (which records, association
   or effect).
:::

### Finish a half-worked example: interpret the three slopes

[Shark Tank data]{.shark-tag} Interpret the three slopes of `equity_model` with units and with the
predictors that are held fixed.

Step 1 (given). Season: holding the requested amount and the description
length fixed, among the recorded pitches, one later season is associated
with about 0.836 percentage points less equity offered.

Step 2 (you). Write the same kind of sentence for `ask_1k` and for
`description_words`.

::: {.callout-note collapse="true"}
## Check your answer

Holding the season and the description length fixed, an additional \$1,000
requested is associated with about 0.0029 percentage points less equity
offered (0.29 per \$100,000). Holding the season and the requested amount
fixed, one additional description word is associated with about 0.128
percentage points less.
:::

### Practice

[Shark Tank data]{.shark-tag} Items 5 to 7 use the calculations of @exm-i04-scaling. Item 8 is the Try it
task, now done from start to finish.

1. Calculate 95% confidence intervals with `confint(equity_model)`.
2. Test each coefficient against zero. Explain how these three single tests
   differ from the one overall F-test of the model.
3. Remove `description_words`. Compare the coefficients of season and
   requested amount with those of the three-predictor model.
4. Name one plausible omitted variable and explain which interpretation of
   the coefficients it would make questionable.
5. Interpret a difference of five seasons with the fitted season
   coefficient.
6. Interpret a difference of \$500,000 in the requested amount.
7. Interpret a difference of ten words in description length and use the 95%
   interval to judge $H_0:\beta=0$.
8. Compare two fitted pitches that are the same except that one is three
   seasons later and requests \$200,000 more. Calculate the combined fitted
   difference.

::: {.callout-note collapse="true"}
## Check your answers

1. The season interval is approximately $(-0.936, -0.736)$, the `ask_1k`
   interval $(-0.0040, -0.0018)$, and the description interval
   $(-0.351, 0.095)$.
2. Each t-test evaluates one $\beta_j=0$ at a time. The overall F-test
   (F $\approx$ 110.6 on 3 and 1437 df) evaluates all three slope
   coefficients together.
3. Without `description_words`, the season coefficient is about $-0.814$ and
   the `ask_1k` coefficient about $-0.0029$. These are the numbers in column (2)
   of the table in the lecture material. Compare them with column (3).
   The coefficients change only slightly.
4. Examples are the maturity of the venture, its revenue, the experience of
   the founder, or decisions of the producers about what is broadcast. If
   such a variable is related to both an included predictor and the equity
   offered, it is a confounder. The coefficient is then still a correct
   description, but it is not the effect of the predictor.
5. Holding requested amount and description length fixed, a pitch five
   seasons later is fitted to offer about $5(-0.836)=-4.18$ equity
   percentage points less.
6. \$500,000 equals 500 units of `ask_1k`. Holding season and description
   length fixed, the fitted difference is $500\times(-0.0029)\approx-1.45$
   percentage points.
7. Ten words go with about $-1.28$ fitted percentage points, holding season
   and requested amount fixed. Multiplying the interval by 10 gives
   approximately $(-3.51, 0.95)$. The interval contains zero, so the
   two-sided 5% test does not reject a zero coefficient.
8. The fitted difference is $3(-0.836)+200(-0.0029)\approx-3.09$ percentage
   points, holding description length fixed. The two contributions add
   because the model is additive.
:::

### Practice: all steps together

[Shark Tank data]{.shark-tag} Write one paragraph that connects the diagnostics of Lecture 2 to
`equity_model`. Which residual checks would you run on this three-predictor
model? What would each check look for? What would you do about a highly
influential pitch if you found one?

::: {.callout-note collapse="true"}
## Check a model answer

A plot of residuals against fitted values checks that the conditional mean
is roughly linear in the included terms and that the spread is roughly
constant. A curve would suggest that assumption 1 fails. A fan shape
would suggest heteroskedasticity, and then the usual standard errors are wrong. Standardised residuals (`rstandard`)
flag pitches that the model fits unusually badly. Cook's distance flags
pitches that move the coefficients on their own. Investigate a flagged
pitch, for example one with an extreme requested amount, but do not delete
it. Check the recorded values, fit the model with and without it, and report
both fits if they disagree. Never delete an observation without reporting
it.
:::

### Review: one question from Lecture 3

[Shark Tank data]{.shark-tag} For the simple model `equity_offered_pct ~ season`, the slope's t statistic
is about $-17.34$. What is the overall F statistic of this one-predictor
model, and why?

::: {.callout-note collapse="true"}
## Check your answer

$F = t^2 \approx 300.7$. With a single slope, the overall F-test and the
two-sided t-test of that slope test the same hypothesis, and F is exactly the
square of t. With several predictors this is no longer true. So Lecture 5
introduces a test of several coefficients at once.
:::

### Worked exam interpretation

[Shark Tank data]{.shark-tag} **Exam-style question.** A summary table reports, for the three-predictor model,
`season  -0.836  (SE 0.051)` and `description_words  -0.128  (SE 0.114)`. A
student writes: "Season matters and word count does not, so descriptions have
no effect on equity." Give two corrections.

::: {.callout-note collapse="true"}
## Check a model answer

First, "no effect" is a wrong reading of a result that is not significant. The interval
$(-0.351, 0.095)$ contains zero and also values near $-0.35$. The data are
consistent with no association and also with a small negative one. Not significant does
not mean zero. Second, "effect" claims causation. Both estimates are
associations among the recorded pitches, holding the other included
predictors fixed. The data do not come from an experiment, so they do not
show a causal effect.
:::

### Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant: "I added the requested amount as a control. Is the season coefficient of -0.81 now the effect of time?" It answered:

> "Yes. Because you controlled for the requested amount, one later season now causes 0.81 percentage points less equity."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

A control variable only holds fixed what was measured and included. Venture maturity, founder experience, and the choices of the producers are not in the model. The correct sentence: at the same requested amount, one later season is associated with about 0.81 percentage points less equity offered.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Name one
outcome you care about and two factors that influence it and are correlated
with each other (for example price and quality tier, or marketing spend and
firm age). Then write two sentences. What would a regression of the outcome
on only one of the two factors mix together? What would "holding the other
factor fixed" let you see? There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "Outcome: weekly orders. Factors: price and quality tier, which move together because premium boxes cost more. A regression of orders on price alone would mix cheaper boxes with lower-tier boxes. Holding the tier fixed shows what a price cut does within a tier, and that is the number I can act on."
:::

You can discuss this task, and any other task in this chapter, with
Tilly for QM Intermediate, the course AI assistant. Ask it to propose an omitted variable for
the regression of your venture, and to ask you whether a measured control
variable could fix it.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask Tilly for
QM Intermediate, the course AI assistant. For example: "Give me an extra
exercise for Intermediate Lecture 4. Show the solution only after I send my
answer."

[Continue to Lecture 5: Categories as predictors and joint tests](i05-categories-joint-tests.qmd).

[Code sheet: Lecture 4 commands and formulas](code-sheet.qmd#lecture-4-multiple-linear-regression).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.2 and 3.3.3, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 20.

:::
