# Lecture 4: Multiple regression and what controls do {#i04}

```{r}
#| label: i04-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 4, you can</span>

- state a multiple linear regression model;
- interpret a coefficient while holding the other included predictors fixed;
- explain, with a constructed example, why simple and multiple-regression
  coefficients can differ;
- fit and compare multiple-regression specifications in R;
- scale a coefficient to a comparison a reader cares about;
- distinguish statistical control from causal identification.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Two stories can produce the same pattern: perhaps equity offers genuinely fell over time, or perhaps later pitches simply ask for different amounts. Separating overlapping explanations is the core job of multiple regression --- and of most business analytics.</p>

Later seasons show smaller equity offers. Is that a time story, or is it
really a money story in disguise, since later pitches also ask for different
amounts?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=o9hoLdylWKo&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=11">3.3
Multiple Linear Regression</a>. Focus on the conditional interpretation of
one coefficient and the reason several predictors may be included together.
All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (10 minutes).** In Lecture 3 you fitted the simple season model
and tested its slope. Fit it again, then fit the two-predictor model that adds
the requested amount in thousands of dollars, and bring **both season
coefficients** to class:

```{r}
#| label: i04-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
coef(lm(equity_offered_pct ~ season, data = sharks))
coef(lm(equity_offered_pct ~ season + ask_1k, data = sharks))
```

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: once the requested amount is added, will the
season coefficient move **towards zero**, **away from zero**, or **barely
move**? Think about why the slope might change at all: if the requested
amount is associated with both season and offered equity, the simple slope
mixes their relationships. Commit to an answer; the lecture resolves it.

If subscripts or multivariable equations feel rusty, review
[Mathematics Refresher: subscripts and multivariable equations](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. Your warm-up produced two season
coefficients --- about $-0.824$ without the requested amount and $-0.814$
with it. This lecture explains what the second number means, why the two can
differ, and why here they barely do.

### More than one predictor

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=o9hoLdylWKo">3.3 Multiple Linear Regression</a>.</p>

For predictors $X_{i1},\ldots,X_{ik}$, the model is

$$
Y_i=\beta_0+\beta_1X_{i1}+\cdots+\beta_kX_{ik}+\varepsilon_i,
$$

with conditional mean

$$
E[Y_i\mid X_{i1}=x_1,\ldots,X_{ik}=x_k]
=\beta_0+\beta_1x_1+\cdots+\beta_kx_k.
$$

The coefficient $\beta_j$ is the expected change in $Y$ for a one-unit
increase in $X_j$, **holding the other included predictors fixed**. That
phrase must appear in every multiple-regression slope interpretation you
write in this course.

It describes a **model-based comparison**. It does not require that two real
pitches exist that are identical on every feature; it compares fitted
conditional means at fixed values of the included variables. The model does
the holding fixed --- the data need not contain twins.

### Why coefficients change: a constructed demonstration

Before touching the Shark Tank data, watch the mechanism operate where we
control it completely, because we built the data ourselves. Imagine 200 small
firms. A single background variable --- firm **size** --- drives both
advertising spend and revenue. Advertising itself contributes nothing: the
revenue line below contains `size` but no `adspend` term.

```{r}
#| label: i04-sim-build
set.seed(42)
n <- 200
size    <- rnorm(n, mean = 10, sd = 2)          # Z: firm size (drives both)
adspend <- 2 + 0.8 * size + rnorm(n, sd = 1)    # X: bigger firms spend more
revenue <- 5 + 1.5 * size + rnorm(n, sd = 1.5)  # Y: size drives revenue,
demo <- data.frame(revenue, adspend, size)      #    adspend does not
coef(lm(revenue ~ adspend, data = demo))
coef(lm(revenue ~ adspend + size, data = demo))
```

The simple slope of `adspend` is about $1.39$: strongly positive, and highly
significant if you check its t statistic. Yet we *know* advertising does
nothing here. Once `size` enters the model, the `adspend` slope collapses to
about $0.10$ with a p-value around $0.39$ --- indistinguishable from the zero
we built in. The simple slope was entirely an **omitted-variable** artefact:
`adspend` earned credit for the size differences it was correlated with.

@fig-i04-ovb-scatter makes the mechanism visible. Larger firms (orange) sit
high on *both* axes; smaller firms (blue) sit low on both. Sliding along the
overall cloud from left to right, you move from small firms to large firms
--- so the simple slope compares different kinds of firms. *Within* each
colour, the cloud is nearly flat: at a given size, spending more is not
associated with more revenue.

```{r}
#| label: fig-i04-ovb-scatter
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Constructed data: revenue against advertising spend for 200 simulated firms, coloured by firm size. The overall upward drift comes entirely from size; within a colour group the relationship is nearly flat."
library(ggplot2)
set.seed(42)
n <- 200
size    <- rnorm(n, mean = 10, sd = 2)          # Z: firm size (drives both)
adspend <- 2 + 0.8 * size + rnorm(n, sd = 1)    # X: bigger firms spend more
revenue <- 5 + 1.5 * size + rnorm(n, sd = 1.5)  # Y: size drives revenue,
demo <- data.frame(revenue, adspend, size)      #    adspend does not
demo$group <- ifelse(demo$size > median(demo$size),
                     "Larger firms (size above median)",
                     "Smaller firms (size below median)")
simple_fit <- lm(revenue ~ adspend, data = demo)

ggplot(demo, aes(x = adspend, y = revenue,
                 colour = group, shape = group)) +
  geom_abline(intercept = coef(simple_fit)[1],
              slope = coef(simple_fit)[2],
              colour = "#17324D", linewidth = 1.1) +
  geom_point(alpha = 0.75, size = 2) +
  annotate("text", x = max(demo$adspend) - 0.2,
           y = coef(simple_fit)[1] + coef(simple_fit)[2] *
             (max(demo$adspend) - 0.2) - 2.4,
           label = "simple regression line",
           hjust = 1, colour = "#17324D", size = 3.4) +
  scale_colour_manual(values = c(
    "Larger firms (size above median)" = "#D55E00",
    "Smaller firms (size below median)" = "#0072B2"
  )) +
  scale_shape_manual(values = c(
    "Larger firms (size above median)" = 17,
    "Smaller firms (size below median)" = 16
  )) +
  labs(x = "Advertising spend (constructed units)",
       y = "Revenue (constructed units)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

A coefficient changes when an added predictor accounts for variation that the
simpler association had absorbed. That is the entire mechanism --- and it is
why "holding the other included predictors fixed" is not decoration but the
substance of the estimate.

### The real-data contrast: the season slope barely moves

Now the same comparison on the Shark Tank data. The running money unit from
here on is **thousands of dollars**: `ask_1k` is the requested amount divided
by 1,000, so one unit is $1,000 and 100 units are $100,000 (Lecture 1 showed
how a coefficient rescales when the unit changes).

```{r}
#| label: i04-compare-slopes
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
m_simple   <- lm(equity_offered_pct ~ season, data = sharks)
m_adjusted <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
c(
  simple           = coef(m_simple)["season"],
  adjusted_for_ask = coef(m_adjusted)["season"]
)
```

The season slope moves from about $-0.824$ to about $-0.814$: hardly at all.
Contrast this with the constructed example, where adjustment destroyed the
simple slope. Here, adjusting for the requested amount leaves the season
trend essentially intact, so within these data the time story is not a
disguised money story.

Read the stability correctly. It is **informative**: the measured requested
amount does not account for the season trend. It is **not proof** that
omitted-variable bias is absent: an unmeasured variable --- venture maturity,
founder experience, editorial selection --- could still play exactly the role
`size` played in the constructed data. Stability with respect to the
variables you have says nothing about the variables you do not have.

### A three-predictor Shark Tank model

Add a third predictor and fit the model this lecture works with. The outcome
is initial equity offered, in percentage points. The predictors are:

- `season` --- the season number, 1 to 16;
- `ask_1k` --- requested amount in thousands of dollars;
- `description_words` --- the number of words in the source's short
  description of the business.

```{r}
#| label: i04-fit
equity_model <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
summary(equity_model)
```

The fitted equation is approximately

$$
\widehat{equity}
=22.025-0.836\,season-0.0029\,ask_{1k}
-0.128\,description\_words.
$$

#### The coefficient-interpretation protocol

Use the same five checks every time you interpret a slope:

1. **Estimate and direction:** state the fitted change and its sign.
2. **Predictor step and unit:** say what a one-unit increase means.
3. **Outcome unit:** state the unit in which the outcome changes.
4. **Conditioning:** name the other included predictors being held fixed.
5. **Scope:** identify the analysed records and say "associated with" unless
   a research design justifies a causal claim.

The variables in this model use deliberately different units:

| Term | A one-unit increase means | Coefficient unit |
|---|---|---|
| `season` | one later season | equity percentage points per season |
| `ask_1k` | $1,000 more requested | equity percentage points per $1,000 |
| `description_words` | one additional word | equity percentage points per word |

Applying the protocol to each slope in turn, always among the 1,441 recorded
pitches:

- **Season.** At the same requested amount and description length, one later
  season is associated with about $0.836$ percentage points less initial
  equity offered.
- **Requested amount.** At the same season and description length, an
  additional $1,000 requested is associated with about $0.0029$ percentage
  points less initial equity offered. The number is tiny because the unit
  step is tiny; per additional $100,000 the association is about $0.29$
  percentage points, and its t statistic (about $-5.0$) does not depend on
  the unit chosen.
- **Description length.** At the same season and request, one additional
  description word is associated with about $0.128$ percentage points less
  initial equity offered. This estimate is **not statistically
  distinguishable from zero** at 5% (p $\approx$ 0.26). Report the point
  estimate anyway: "not significant" is not the same as "exactly zero" ---
  the data are simply consistent with a range of values that includes zero.

::: {.callout-note collapse="true"}
## Check: the season slope was −0.814 with two predictors and −0.836 with three. Contradiction?

No. Each coefficient answers a different conditional question: $-0.814$ holds
only `ask_1k` fixed; $-0.836$ holds `ask_1k` *and* `description_words` fixed.
Adding a predictor changes what is being held fixed, so the coefficient may
shift. Here the shift is small --- the same stability, and the same limits to
it, as before.
:::

### Scaling to comparisons readers care about

Nobody plans in single seasons of a TV show or single thousands of dollars.
Scale the coefficient to the step a reader cares about: **fitted change =
coefficient × number of unit steps**, holding the other included predictors
fixed. This resolves the Mentimeter question posed before the break.

::: {#exm-i04-scaling}
## Worked example: five seasons, half a million dollars, ten words

Holding the other included predictors fixed in `equity_model`:

- **Five later seasons:** $5\times(-0.836)=-4.18$ fitted equity percentage
  points.
- **$500,000 more requested:** $500{,}000/1{,}000 = 500$ units of `ask_1k`,
  so $500\times(-0.0029)=-1.45$ fitted percentage points.
- **Ten additional description words:** $10\times(-0.128)=-1.28$ fitted
  percentage points --- with the reminder that this coefficient is not
  significantly different from zero, so the scaled version is not either.

Rescaling a predictor changes the numerical coefficient and its SE by the
same factor, but not the fitted values, t statistic, or substantive
relationship. A "per dollar" coefficient of $-0.0000029$ would be correct and
unreadable; per $1,000 is readable on the data's scale; per $100,000 is often
the best unit for a sentence.
:::

::: {.callout-caution collapse="true"}
## Try it: three seasons later and $200,000 more

In class this is the pair task. Two pitches are identical in description
length; one is three seasons later and requests $200,000 more. Complete the
steps before opening the answer.

*Step 1 (given).* Three seasons contribute $3\times(-0.836)$ fitted
percentage points.

*Step 2 (you).* Convert $200,000 into `ask_1k` units and multiply by the
`ask_1k` coefficient.

*Step 3 (you).* Add the two contributions --- the model is additive --- and
state the combined fitted difference in one protocol-complete sentence.

::: {.callout-note collapse="true"}
## Check your answer

$200{,}000/1{,}000=200$ units, so the request contributes
$200\times(-0.0029)\approx-0.58$. Combined:
$3(-0.836)+200(-0.0029)\approx-2.51-0.58=-3.09$ percentage points. Sentence:
among the recorded pitches, a pitch three seasons later that requests
$200,000 more is fitted to offer about 3.1 percentage points less initial
equity, holding description length fixed --- an association, not an effect.

```{r}
#| label: i04-faded-check
b <- coef(equity_model)
3 * b["season"] + 200 * b["ask_1k"]
```
:::
:::

### What controls can and cannot do

Including a control **can**:

- make a comparison more like-for-like on measured dimensions;
- reduce residual variation, tightening intervals;
- reveal that a simple association was partly compositional;
- increase uncertainty when predictors overlap strongly.

Including a control **cannot**:

- adjust for unmeasured factors;
- fix reverse causality;
- repair selection into the sample;
- remove measurement error.

A "control variable" is just an included predictor. Nothing about the word
*control* grants it a causal role; only a research design can do that.

::: {.callout-note collapse="true"}
## Check: which of these four problems does adding `ask_1k` address?

(a) later pitches request different amounts; (b) founder experience is not
recorded; (c) equity offers may influence how much is asked; (d) only pitches
that reached broadcast are in the file.

Only (a). It is a measured compositional difference, and holding `ask_1k`
fixed removes it from the season comparison. Problems (b), (c), and (d) are
unmeasured confounding, reverse causality, and sample selection --- none of
them touched by adding a measured control.
:::

### Assumptions extend to the full predictor set

For multiple regression, the Lecture 2 assumption list applies to the whole
set of included predictors:

1. the conditional mean is linear in the included terms;
2. $E[\varepsilon_i\mid X_{i1},\ldots,X_{ik}]=0$ --- zero conditional mean at
   every combination of predictor values;
3. observations are independent under the design;
4. no predictor is an exact linear combination of the others;
5. classical SEs require constant conditional error variance;
6. exact small-sample t/F inference uses conditional normality.

High correlation among predictors --- **collinearity** --- does not bias OLS
by itself, but it makes individual coefficients unstable and imprecise: the
data contain little information to separate overlapping predictors. Exact
linear dependence is worse: the coefficients are then **not identified** at
all, and R drops a term. Instability is a precision problem; exact dependence
is a no-answer-exists problem.

### A coefficient plot keeps units visible

```{r}
#| label: fig-i04-coefplot
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Coefficient estimates from the three-predictor model with 95% confidence intervals. Each row uses a different predictor unit, so magnitudes are not comparable across rows without the units table."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
equity_model <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
ci <- confint(equity_model)[-1, , drop = FALSE]
term_labels <- c("Season", "Amount requested ($1,000s)",
                 "Description length (words)")
coef_df <- data.frame(
  term = factor(term_labels, levels = term_labels),
  estimate = coef(equity_model)[-1],
  lower = ci[, 1],
  upper = ci[, 2]
)

ggplot(coef_df, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  colour = "#009E73", linewidth = 0.7, fatten = 2.6) +
  annotate("text", x = -0.02, y = 3.35, label = "no change = 0",
           hjust = 1, colour = "#17324D", size = 3.4) +
  labs(x = "Estimated change in equity offered (percentage points)",
       y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-i04-coefplot with the standing five-step sequence:

1. **Axes:** rows are predictors; the horizontal axis is the estimated change
   in initial equity offered, in percentage points.
2. **Units:** each row's estimate is *per one unit of that predictor* ---
   per season, per $1,000, per word. The rows are not on a common scale.
3. **Marks:** a dot is the coefficient estimate; its horizontal line is the
   95% confidence interval; the dashed line marks zero.
4. **Pattern:** season and requested amount sit clearly left of zero; the
   description-words interval crosses zero. The `ask_1k` row hugs the zero
   line only because its unit step ($1,000) is tiny --- its t statistic is
   about $-5$. Visual size on this plot is a unit artefact, not evidence
   strength.
5. **Limits:** intervals reflect model-based sampling uncertainty among these
   recorded pitches; they say nothing about causation or unmeasured
   variables.

::: {.caveat-box}
<span class="label">What an adjusted coefficient can and cannot say</span>

Every slope in `equity_model` is a model-based comparison among the 1,441
recorded pitches: fitted conditional means at fixed values of the *included*
predictors. "Holding the other included predictors fixed" never extends to
variables outside the model --- maturity, revenue, founder experience, and
editorial selection are all unheld. The stability of the season slope across
specifications is informative about the measured controls only; it proves no
absence of omitted-variable bias. Initial equity offered is an on-air opening
position, not a completed investment, and none of these associations is a
causal effect of time, money, or word count.
:::

### Answering the driving question

At the same requested amount (and description length), the season trend
persists: about $-0.836$ equity percentage points per season, versus $-0.824$
unadjusted. The money story does not absorb the time story here --- later
pitches offer less equity even compared at the same ask. And neither story is
causal: an unmeasured driver could still lie behind both, exactly as firm
size did in the constructed example.

**Next:** We can now hold other predictors fixed --- as long as they are
numerical. But many of the most interesting business comparisons are
categorical: channels, segments, team compositions. Lecture 5 shows how
categories enter a regression correctly --- dummy variables and a reference
group chosen deliberately --- and how to judge a whole set of category terms
at once with the partial F-test, because individually insignificant dummies
can still be jointly worth having.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 75 minutes.</span>

### Retrieval (10 minutes)

1. What phrase must appear in a multiple-regression slope interpretation?
2. Why can a coefficient change after another predictor is added?
3. What is the zero conditional mean assumption?
4. Why do controls not automatically identify causality?
5. What problem can strongly overlapping predictors create?
6. What five elements belong in a slope interpretation?

::: {.callout-note collapse="true"}
## Check your answers

1. "Holding the other included predictors fixed."
2. The simpler slope may absorb relationships involving omitted correlated
   predictors; the added predictor takes back its share.
3. The error has mean zero at every combination of included predictor values.
4. Unmeasured confounding, selection, reverse causality, and measurement
   error can all remain after adjustment.
5. Unstable, imprecise individual coefficient estimates; exact overlap makes
   the model unidentified.
6. Estimate and direction, predictor step and unit, outcome unit, held-fixed
   predictors, and the sample/causal scope.
:::

### Practice (30 minutes)

Items 1--5 are unassisted checks of the core skills. **Items 6--9 form the
faded-to-unassisted ladder**: item 6 repeats a calculation worked in class,
items 7--8 change the numbers, and item 9 is the combined contrast you met
half-finished in the pair task --- now do it start to finish.

1. Interpret all three slopes from `equity_model` with units and
   conditioning.
2. Calculate 95% confidence intervals with `confint(equity_model)`.
3. Test each coefficient against zero and distinguish three individual tests
   from one joint model test.
4. Remove `description_words`. Compare the season and request coefficients
   with the full model's.
5. Name one plausible omitted variable and explain which model assumption it
   could threaten.
6. Interpret a five-season difference using the fitted season coefficient.
7. Interpret a $500,000 difference in requested amount.
8. Interpret a ten-word difference in description length and use its 95%
   interval to assess $H_0:\beta=0$.
9. Compare two otherwise identical fitted pitches when one is three seasons
   later and requests $200,000 more. Calculate the combined fitted
   difference.

::: {.callout-note collapse="true"}
## Check your answers

1. Holding the other two predictors fixed, among the recorded pitches: one
   later season is associated with about 0.836 percentage points less
   initial equity; an additional $1,000 requested with about 0.0029
   percentage points less (0.29 per $100,000); one additional description
   word with about 0.128 percentage points less.
2. The season interval is approximately $(-0.936, -0.736)$, the `ask_1k`
   interval $(-0.0040, -0.0018)$, and the description interval
   $(-0.351, 0.095)$.
3. Individual t-tests evaluate one $\beta_j=0$ at a time; the overall F-test
   (F $\approx$ 110.6 on 3 and 1437 df) evaluates all slope coefficients
   jointly.
4. Without `description_words`, the season coefficient is about $-0.814$ and
   the `ask_1k` coefficient about $-0.0029$. They change only slightly here;
   that empirical stability does not prove absence of omitted-variable bias.
5. Examples include venture maturity, revenue, founder experience, or editing
   decisions; if related to both an included predictor and offered equity,
   the omitted variable violates zero conditional mean.
6. Holding request and description length fixed, a pitch five seasons later
   is fitted to offer about $5(-0.836)=-4.18$ equity percentage points less.
7. $500,000 equals 500 `ask_1k` units. Holding season and description length
   fixed, the fitted difference is $500\times(-0.0029)\approx-1.45$
   percentage points.
8. Ten words correspond to about $-1.28$ fitted percentage points, holding
   season and request fixed. Multiplying the interval by 10 gives
   approximately $(-3.51, 0.95)$; because zero remains plausible, the
   two-sided 5% test does not reject a zero coefficient.
9. The fitted difference is $3(-0.836)+200(-0.0029)\approx-3.09$ percentage
   points, holding description length fixed. Coefficient contributions add
   because the model is additive.
:::

### Cumulative mini-task (10 minutes)

Write **one paragraph** connecting the Lecture 2 diagnostics to
`equity_model`: which residual checks would you run on this three-predictor
fit, what would each look for, and what would you do about a highly
influential pitch if you found one?

::: {.callout-note collapse="true"}
## Check a model answer

A residuals-versus-fitted plot checks that the conditional mean is roughly
linear in the included terms and that spread is roughly constant; curvature
would question assumption 1, a funnel shape would question the classical SEs.
Standardised residuals (`rstandard`) flag pitches the model fits unusually
badly, and Cook's distance flags pitches that individually move the
coefficients. A flagged pitch --- say, an extreme requested amount --- gets
investigated, not deleted: check the recorded values, refit with and without
it, and report both fits if they disagree. Silent deletion is never an
option.
:::

### Spaced review: one item from Lecture 3 (5 minutes)

For the simple model `equity_offered_pct ~ season`, the slope's t statistic
is about $-17.34$. What is the overall F statistic of this one-predictor
model, and why?

::: {.callout-note collapse="true"}
## Check your answer

$F = t^2 \approx 300.7$. With a single slope, the overall F-test and the
two-sided t-test of that slope test the same hypothesis, and F is exactly the
square of t. With several predictors that equivalence disappears --- which is
why Lecture 5 needs a genuine joint test.
:::

### Worked exam interpretation (5 minutes)

**Question.** A summary table reports, for the three-predictor model,
`season  -0.836  (SE 0.051)` and `description_words  -0.128  (SE 0.114)`. A
student writes: "Season matters and word count does not, so descriptions have
no effect on equity." Give two corrections in the mock-exam Q4 style.

::: {.callout-note collapse="true"}
## Check a model answer

First, "no effect" misreads non-significance: the interval
$(-0.351, 0.095)$ contains zero *and* values near $-0.35$, so the data are
consistent with no association and with a modest negative one --- not
significant is not zero. Second, "effect" claims causation; both estimates
are associations among recorded pitches, holding the other included
predictors fixed, with no design that identifies a causal effect.
:::

### From the reading ladder (10 minutes)

Next rung of the [reading ladder](../resources/reading-ladder.qmd): open
[Zhao, Seibert & Hills (2005)](https://doi.org/10.1037/0021-9010.90.6.1265)
via the university library and find the table predicting **entrepreneurial
intentions**. Identify the outcome, the predictors of substantive interest,
and the variables included as controls. Pick one coefficient and write the
full protocol sentence --- including "holding the other included predictors
fixed" and the sample scope. Then name one variable *not* in the model that
could drive both the predictor and the outcome, exactly like the
constructed demonstration from today's lecture.

::: {.callout-note collapse="true"}
## Check your approach

Any defensible answer names the model-based nature of the comparison (no
claim that identical students exist), keeps the units or scale of the
predictor visible, and proposes an omitted variable with a plausible path to
both sides --- family business background, access to capital, and general
self-confidence are all fair candidates. If your sentence could be read as
causal advice ("increase X to raise intentions"), tighten it.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well. Name one
outcome you care about and two of its drivers that are correlated with each
other (say, price and quality tier, or marketing spend and firm age), and
write two sentences: which comparison would a regression of the outcome on
only one driver mix together, and what would "holding the other fixed" let
you see? There is no wrong answer; the point is spotting the confound in a
business you actually know.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Outcome: weekly orders; drivers: price and quality tier, which move together because premium boxes cost more. Orders-on-price alone would mix cheaper with lower-tier; holding tier fixed shows what a price cut does within a tier --- the number I actually control."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to propose an omitted
variable for your venture's regression and challenge you to say whether a
measured control could fix it.
::::

Continue to Lecture 5: Categories as predictors and joint tests.

Formula sheet: Lecture 4 formulas.

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.2 and 3.3.3, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 20.

:::
