# Lecture 5: Categories as predictors and joint tests {#i05}

```{r}
#| label: i05-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 5, you can</span>

- encode a two-level or multi-level categorical predictor with dummy
  variables;
- interpret dummy coefficients relative to a named reference group, in
  percentage points where the outcome is a percentage;
- avoid the dummy-variable trap;
- compute a partial F statistic by hand from an anova table and read the
  R output;
- separate joint usefulness from individual significance;
- separate adjusted associations from causal or essentialist claims.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Markets come in categories: sales channels, customer segments, team compositions. Putting categories into a model correctly --- and judging whether they belong in it at all --- is a craft this lecture teaches.</p>

Adjusted for season and ask, do all-women presenting teams offer different
equity than all-men teams --- and are the category terms worth having in the
model at all?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=13">3.5
Extensions of the Linear Model</a>, focusing on qualitative predictors and
how a reference category determines coefficient meaning. All links are also
listed in the <a href="../resources/video-guide.html">External Video
Guide</a>.
</div>

**Reactivate and predict (5 minutes).** With three categories --- `all_men`,
`all_women`, and `mixed` --- how many dummy variables does a model with an
intercept need? Commit to a number, then answer the one-click Mentimeter
question posted on Canvas: at the same season and requested amount, do you
expect all-women teams to offer **more**, **less**, or **about the same**
equity as all-men teams? The lecture resolves both.

**R warm-up (10 minutes).** Class starts from the subset and factor you
build here --- there are no hidden objects, so this warm-up *is* the setup.
Bring the contrasts table:

```{r}
#| label: i05-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
levels(known_gender$gender)
contrasts(known_gender$gender)
```

If indicators and reference categories are unfamiliar, review
[Mathematics Refresher: indicators and reference categories](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question --- and, before any model, to the
measurement behind it.

### Measurement first: what the category records

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk">3.5 Extensions of the Linear Model</a>.</p>

The source categories describe the presenting team as all men, all women, or
mixed. They do not measure the share of women, gender identity beyond the
source categories, or the composition of the full venture. Seven records with
an `unknown` classification are excluded from every model in this lecture;
the reason comes at the end. Every sentence this chapter produces inherits
this measurement boundary.

### A two-level indicator

For a binary predictor $D_i\in\{0,1\}$,

$$
E[Y_i\mid D_i]=\beta_0+\beta_1D_i.
$$

Then

$$
E[Y_i\mid D_i=0]=\beta_0,
\qquad
E[Y_i\mid D_i=1]=\beta_0+\beta_1.
$$

Thus $\beta_1$ is the expected difference between the 1 group and the 0
**reference group**. A dummy regression with one indicator is exactly a
comparison of two group means, dressed as a regression.

::: {#exm-i05-two-ticks}
## Worked example: two group means from two coefficients

Restrict to the all-men and all-women teams (1,158 records) and let
`women_team` equal 1 for all-women teams:

```{r}
#| label: i05-two-level
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
two_groups <- subset(known_gender,
                     pitcher_gender_group %in% c("all_men", "all_women"))
two_groups$women_team <- as.integer(
  two_groups$pitcher_gender_group == "all_women"
)
coef(lm(equity_offered_pct ~ women_team, data = two_groups))
tapply(two_groups$equity_offered_pct, two_groups$women_team, mean)
```

Read the two coefficients as two ticks on the equity axis
(@fig-i05-two-ticks): the intercept $\beta_0\approx 13.20$ *is* the all-men
mean, and $\beta_0+\beta_1\approx 13.20+0.82=14.02$ *is* the all-women mean.
The slope $\beta_1\approx 0.82$ is the gap between the ticks --- nothing
more.
:::

```{r}
#| label: fig-i05-two-ticks
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-height: 2.2
#| fig-cap: "The two-level dummy model places the two group means as two ticks on the outcome axis: the intercept is the reference-group mean and the dummy coefficient is the distance between the ticks."
old_par <- par(no.readonly = TRUE)
par(mar = c(3.5, 1, 1, 1))
plot(NA, xlim = c(12.6, 14.6), ylim = c(0, 1), axes = FALSE,
     xlab = "", ylab = "")
axis(1, at = seq(12.6, 14.6, by = 0.4))
mtext("Mean initial equity offered (%)", side = 1, line = 2.4, cex = 0.9)
segments(13.20, 0.15, 13.20, 0.55, col = "#0072B2", lwd = 4)
segments(14.02, 0.15, 14.02, 0.55, col = "#D55E00", lwd = 4)
arrows(13.20, 0.38, 14.02, 0.38, length = 0.08, code = 3, col = "#17324D")
text(13.20, 0.72, expression(beta[0] == 13.20), col = "#0072B2")
text(14.02, 0.72, expression(beta[0] + beta[1] == 14.02), col = "#D55E00")
text(13.61, 0.25, expression(beta[1] == 0.82), col = "#17324D", cex = 0.9)
par(old_par)
```

### More than two categories

With $G$ categories and an intercept, use $G-1$ dummies. Including all $G$
dummies plus an intercept creates exact linear dependence, because the
dummies sum to 1 in every row --- the model from Lecture 4 whose coefficients
are not identified. This is the **dummy-variable trap**.

R creates the dummies automatically when a predictor is a factor. Set the
reference category deliberately and verify it --- this is exactly what your
warm-up built:

```{r}
#| label: i05-reference
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
levels(known_gender$gender)
contrasts(known_gender$gender)
```

Each row of the contrasts table is a category; each column is one dummy. The
reference category `all_men` is the all-zero row: it lives in the intercept.

::: {.callout-note collapse="true"}
## Check: four industries and an intercept --- how many dummies?

Three. One industry is the reference group carried by the intercept; the
three dummy coefficients each compare one remaining industry with it. Four
dummies plus an intercept would fall into the trap.
:::

### Percent versus percentage points

One habit must be in place before any coefficient is read. The outcome
`equity_offered_pct` is measured in percent, so its regression coefficients
are in **percentage points** --- and the sentence "the all-women coefficient
is 1.28%" is ambiguous. It could mean:

- a **relative** change: offers 1.28% *of* the all-men level higher (13.0%
  would become about 13.2%); or
- an **absolute** change of 1.28 **percentage points**: 13.0% becomes
  14.28%.

These differ by a factor of about eight here. A coefficient on a
percent-scaled outcome is always the absolute kind. The precise sentence
replacing the ambiguous one is: "At the same season and requested amount,
all-women presenting teams are fitted to offer 1.28 **equity percentage
points** more initially than all-men presenting teams." Every full-credit
dummy interpretation in this course uses *percentage points* for this
outcome.

### The adjusted comparison

Now the model of the driving question: both category dummies, with season and
requested amount (in thousands of dollars) held fixed.

```{r}
#| label: i05-fit
gender_model <- lm(
  equity_offered_pct ~ season + ask_1k + gender,
  data = known_gender
)
summary(gender_model)
```

Both dummy coefficients, each with the full sentence --- categories named,
unit, held-fixed set, sample boundary:

- **All-women versus all-men.** Among pitches with a recorded source
  classification, all-women presenting teams are fitted to offer about
  **1.28 equity percentage points more** initially than all-men presenting
  teams at the same season and requested amount (p $\approx$ 0.008).
- **Mixed versus all-men.** Among the same pitches, mixed teams are fitted
  to differ from all-men teams by about **0.12 percentage points** at the
  same season and requested amount (p $\approx$ 0.823) --- an estimate not
  statistically distinguishable from zero.

Note the adjusted all-women gap (1.28 points) is larger than the raw gap
from the worked example (0.82 points): adjusting for season and ask moved
the comparison, which is precisely why the held-fixed set belongs in the
sentence. These are two reference-group comparisons, not a claim that gender
causes the offers.

Changing the reference group changes coefficient labels, not fitted values or
overall fit --- verify rather than trust:

```{r}
#| label: i05-change-reference
known_gender$gender_women_ref <- relevel(known_gender$gender,
                                         ref = "all_women")
coef(lm(equity_offered_pct ~ season + ask_1k + gender_women_ref,
        data = known_gender))
```

The same three group means are now described from the all-women vantage
point: the all-men coefficient is $-1.28$, the mirror image of before, and
the fitted values are identical.

### Are the category terms worth having at all?

Block 1 read the coefficients one at a time. The driving question also asks
something no single t-test answers: do the two category terms **jointly** add
information beyond season and ask?

The logic needs two **nested** models --- the reduced model is the complete
model with the tested coefficients forced to zero, fitted on the same rows:

```{r}
#| label: fig-i05-nested
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-height: 2.6
#| fig-cap: "The nested pair behind the partial F-test: the reduced model is the complete model with the two category coefficients set to zero, on the same analysis rows."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.3, 0.3, 0.3, 0.3))
plot(NA, xlim = c(0, 10), ylim = c(0, 5), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")
rect(0.4, 1.1, 4.6, 3.9, col = "#DCEAF3", border = "#17324D", lwd = 2)
rect(5.4, 1.1, 9.6, 3.9, col = "#FDE5C1", border = "#17324D", lwd = 2)
text(2.5, 3.35, "Reduced model", font = 2, col = "#17324D")
text(2.5, 2.55, "season + ask_1k", cex = 0.95)
text(2.5, 1.85, expression(beta[aw] == 0 ~ "," ~ beta[mx] == 0),
     cex = 0.95, col = "#5B6976")
text(7.5, 3.35, "Complete model", font = 2, col = "#17324D")
text(7.5, 2.55, "season + ask_1k", cex = 0.95)
text(7.5, 1.85, "+ two category dummies", cex = 0.95)
arrows(4.7, 2.5, 5.3, 2.5, length = 0.1, lwd = 2, col = "#17324D")
text(5.0, 3.0, "free the\ntwo coefficients", cex = 0.8, col = "#5B6976")
par(old_par)
```

The hypotheses:

$$
H_0:\beta_{all\_women}=\beta_{mixed}=0
\quad\text{against at least one non-zero.}
$$

::: {#exm-i05-partial-f}
## Worked example: the partial F statistic by hand

Fit both models and ask R for the anova table:

```{r}
#| label: i05-partial-f
reduced_gender <- lm(equity_offered_pct ~ season + ask_1k,
                     data = known_gender)
anova(reduced_gender, gender_model)
```

Plug the table's numbers into the formula once by hand. Adding the two
category terms reduces the residual sum of squares by $436.25$ using $2$
degrees of freedom; the complete model leaves $82{,}792$ on $1{,}429$
degrees of freedom:

$$
F=\frac{436.25/2}{82{,}792/1{,}429}
=\frac{218.13}{57.94}
\approx 3.76.
$$

Read the ratio: the numerator is explained variation *per tested
coefficient*; the denominator is unexplained variation *per residual degree
of freedom*. The category terms buy about 3.8 times as much explanation per
coefficient as noise alone would --- with p $\approx$ 0.023, more than chance
comfortably explains.
:::

At 5%, the two source-category terms are **jointly useful** in this
specification. Hold that next to the individual results: joint p-value
$0.023$, yet the mixed dummy's own p-value is $0.823$ *in the very same
output*. Joint usefulness is not individual significance --- the F-test asks
whether the pair of terms together improves the model, not whether each
survives on its own. (The overall F in `summary(gender_model)`, about 84.5
with a tiny p-value, answers a third, cruder question: whether *all four*
predictors jointly beat an intercept-only model.)

### Visualise the adjusted comparison

```{r}
#| label: fig-i05-adjusted
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted initial equity by source classification at the average season and requested amount, with 95% confidence intervals. The intervals express model-based uncertainty, not uncertainty about how the categories were recorded."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
gender_model <- lm(
  equity_offered_pct ~ season + ask_1k + gender,
  data = known_gender
)
newdat <- data.frame(
  season = mean(known_gender$season),
  ask_1k = mean(known_gender$ask_1k),
  gender = factor(c("all_men", "all_women", "mixed"),
                  levels = levels(known_gender$gender))
)
pred <- predict(gender_model, newdata = newdat, interval = "confidence")
adj_df <- data.frame(
  team = factor(c("All-men", "All-women", "Mixed"),
                levels = c("All-men", "All-women", "Mixed")),
  fit = pred[, "fit"],
  lwr = pred[, "lwr"],
  upr = pred[, "upr"]
)

ggplot(adj_df, aes(x = team, y = fit)) +
  geom_pointrange(aes(ymin = lwr, ymax = upr),
                  colour = "#009E73", linewidth = 0.7, fatten = 2.6) +
  labs(x = "Source classification of pitching team",
       y = "Fitted initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-i05-adjusted with the standing five-step sequence:

1. **Axes:** source categories across, fitted initial equity (in percent) up.
2. **Units:** each point is a fitted mean at the *same* average season and
   requested amount --- an adjusted comparison, not raw group means.
3. **Marks:** points are fitted values (about 12.9, 14.1, and 13.0); vertical
   lines are 95% confidence intervals, wider for the smaller categories.
4. **Pattern:** the all-women point sits about 1.28 points above the all-men
   point; the mixed interval overlaps both.
5. **Limits:** the intervals express **model-based** sampling uncertainty.
   They say nothing about the **recording** uncertainty in the broad source
   categories themselves, and nothing causal.

### Why `unknown` is excluded, not a fourth group

Seven records carry the classification `unknown`. That label records the
*absence of information*, not a kind of team: treating it as a substantive
fourth category would estimate a "mean equity of teams the source failed to
classify," a quantity with no business meaning. Excluding the records keeps
every coefficient interpretable; the price, stated openly, is that the
analysed sample is the classified records only.

::: {.caveat-box}
<span class="label">What these category coefficients can and cannot say</span>

The categories are broad source codings of the on-screen presenting team ---
not shares, identities, or full founding teams --- so the coefficients
compare codings, not people. The analysed sample is the 1,434 classified
records of broadcast pitches; selection into the show and into
classification travels with every estimate. The two dummy coefficients are
two reference-group contrasts at fixed season and ask, and they change
labels when the reference changes. None of this supports an essentialist or
causal reading: the model reports that offers *differ* by coded composition
in these data, not that composition *makes* anyone offer differently.
:::

### Answering the driving question

At the same season and requested amount, all-women presenting teams are
fitted to offer about 1.28 equity percentage points more than all-men teams
(p $\approx$ 0.008); mixed teams are indistinguishable from all-men teams.
The two category terms are jointly useful (partial F $\approx$ 3.76,
p $\approx$ 0.023), so they earn their place in the model --- as an adjusted
association within these recorded pitches, nothing more.

**Next:** We can now compare categories at fixed values of the other
predictors --- but so far every group shares one slope, and every trend is a
straight line. Both restrictions are modelling choices, not facts. Lecture 6
relaxes them: interactions let the season trend differ by team composition,
and polynomial terms let the trend bend --- each with the discipline for
reading the result honestly.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 70 minutes.</span>

### Retrieval (10 minutes)

1. Why are only $G-1$ dummies used with an intercept?
2. What does a dummy coefficient compare?
3. What changes when the reference group changes?
4. What does a partial F-test evaluate?
5. Why can a joint F-test reject when one individual t-test does not?

::: {.callout-note collapse="true"}
## Check your answers

1. All $G$ dummies sum to 1, the intercept column: exact collinearity, so
   the coefficients would not be identified.
2. A category with the named reference category, holding the other included
   predictors fixed.
3. Coefficient labels and values change to describe the same fits from a new
   vantage point; fitted values and overall fit do not change.
4. Whether a specified group of coefficients is jointly zero, comparing
   nested models on the same rows.
5. The F-test evaluates the terms together; individual tests ask narrower
   questions, each with its own uncertainty, and can miss what the pair
   achieves jointly.
:::

### Releveling and verification practice (15 minutes)

1. Relevel `gender` so `mixed` is the reference group and interpret both
   dummy coefficients. Save the relevelled model as `relevelled_model`.
2. Verify that fitted values are unchanged after releveling by comparing
   with `gender_model`.
3. State the hypotheses for the partial F-test of both gender dummies.
4. Explain why the seven `unknown` records were excluded rather than treated
   as evidence of a substantive fourth group.
5. Explain why "the all-women coefficient is 1.28%" is ambiguous, then
   rewrite it precisely.

::: {.callout-note collapse="true"}
## Check your answers

1. Use:

   ```r
   mixed_reference <- relevel(known_gender$gender, ref = "mixed")
   relevelled_model <- lm(
     equity_offered_pct ~ season + ask_1k + mixed_reference,
     data = known_gender
   )
   coef(relevelled_model)
   ```

   At fixed season and request, the all-men coefficient is about $-0.121$
   percentage points and the all-women coefficient about $1.163$ percentage
   points, each relative to mixed teams.
2. `all.equal(fitted(gender_model), fitted(relevelled_model))` returns
   `TRUE` up to numerical tolerance.
3. $H_0:\beta_{all\_women}=\beta_{mixed}=0$ against at least one non-zero
   coefficient, with all-men as the current reference.
4. `unknown` records lack a known classification; absence of recorded
   information is not a meaningful team-composition category.
5. It confuses percent with percentage points and omits the comparison and
   conditioning. A precise version: "At the same season and requested
   amount, all-women presenting teams are fitted to offer 1.28 equity
   percentage points more initially than all-men presenting teams."
:::

### From worked to faded: a partial F with new numbers (10 minutes)

Test the two category dummies in a *smaller* complete model that contains
only `season`:

```{r}
#| label: i05-faded-anova
reduced_season <- lm(equity_offered_pct ~ season, data = known_gender)
full_season    <- lm(equity_offered_pct ~ season + gender,
                     data = known_gender)
anova(reduced_season, full_season)
```

*Step 1 (given).* Adding the two dummies reduces the residual sum of squares
by $643.43$ using $2$ degrees of freedom; the complete model leaves
$84{,}003$ on $1{,}430$ degrees of freedom.

*Step 2 (you).* Plug these into the F formula from the worked example.

*Step 3 (you).* Compare with the table's F and p-value and state the
conclusion at 5% in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$F=(643.43/2)/(84{,}003/1{,}430)=321.7/58.7\approx 5.48$, matching the
table, with p $\approx$ 0.004. At 5%, the two category terms are jointly
useful even in the model that adjusts only for season.
:::

### The 120-word interpretation (10 minutes)

Write a 120-word interpretation of the full `gender_model` that includes the
source measurement boundary and the causal limit.

::: {.callout-note collapse="true"}
## Check a model answer

A complete answer reports: the outcome and sample (initial equity offered,
1,434 broadcast pitches with a recorded source classification); both
estimates with units and conditioning (all-women teams fitted 1.28
percentage points above all-men teams, mixed teams 0.12 points above, at the
same season and requested amount); uncertainty (p $\approx$ 0.008 and 0.823;
jointly useful, partial F p $\approx$ 0.023); the measurement boundary
(broad source coding of the on-screen team, not shares or identities); and
the scope (adjusted associations among selected, recorded pitches --- not
causal, not essentialist).
:::

### Spaced review: one item from Lecture 4 (5 minutes)

Using the Lecture 4 three-predictor model, compute the combined fitted
difference for a pitch three seasons later requesting $200,000 more, at the
same description length.

::: {.callout-note collapse="true"}
## Check your answer

$3(-0.836)+200(-0.0029)\approx-3.09$ percentage points: contributions add
because the model is additive, and $200,000 is 200 units of `ask_1k`.
:::

### Worked exam interpretation (5 minutes)

**Question.** In the mock-exam Q4 style: a model of equity on season,
`ask_1k`, and industry (four categories, reference `Food and Beverage`)
reports `industryTechnology 1.9 (p = 0.04)`. A student writes: "Technology
pitches offer 1.9% more equity, so entering tech makes founders more
generous." Correct the sentence.

::: {.callout-note collapse="true"}
## Check a model answer

Three repairs. Units: the estimate is 1.9 equity *percentage points*, not a
relative 1.9%. Comparison and conditioning: technology pitches are fitted
1.9 points above *Food and Beverage* pitches at the same season and
requested amount, among the recorded pitches. Scope: nothing here is causal
--- "entering tech makes founders more generous" reads a reference-group
contrast as an effect of industry choice.
:::

### From the reading ladder (10 minutes)

Next rung of the [reading ladder](../resources/reading-ladder.qmd): open
[Brooks et al. (2014)](https://doi.org/10.1073/pnas.1321202111) --- an
open-access study of entrepreneurial pitch success. Find where the analysis
compares pitches across a **category** (the presenter's gender). Identify
the categorical predictor and what serves as the reference comparison, and
restate the paper's headline gap in one careful sentence that respects
today's two rules: percentage points are not percent, and an adjusted gap
between categories is not a causal claim about any attribute. Then note one
way their setting differs from our Shark Tank teaching file.

::: {.callout-note collapse="true"}
## Check your approach

A good sentence names both categories explicitly, states the outcome and
its unit, and attributes the gap to the sample and design ("in these
pitch competitions, pitches presented by ... were funded ... percentage
points more often than ...") without sliding into "because investors
prefer...". Differences from our file worth noting: their data include
experiments with manipulated recordings, whereas ours are selected,
edited television records with a broad source category for the presenting
team.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well. Name one
categorical feature of your customers or products (segment, channel, region,
plan tier), and write two sentences: which category would you set as the
reference group and why, and what would a "jointly useful" F-test verdict on
the whole set of categories change about how you describe your market? There
is no wrong answer; the point is choosing a reference deliberately.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "My channels are the campus stand, the app, and the website; I would set the campus stand as reference because it is the original channel every decision gets compared against. A jointly useful F-verdict would justify describing my market by channel at all --- individually mixed p-values on single dummies would not settle that."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to check whether your
two sentences confuse percent with percentage points, or to quiz you on the
dummy-variable trap with your own categories.
::::

Continue to Lecture 6: Interactions and curvature.

Formula sheet: Lecture 5 formulas.

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.2.2 and 3.3.1, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 20--21.

:::
