# Lecture 5: Categories as predictors and joint tests {#i05}

```{r}
#| label: i05-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 5, you can</span>

- use dummy variables to put a categorical predictor into a regression
- interpret dummy coefficients as differences from a named reference category
- test whether a group of terms is jointly significant, and whether group means
  differ in a one-way ANOVA
- tell an adjusted comparison between groups apart from the effect that an
  A/B test estimates
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Many business variables are categories: sales channels, customer segments, kinds of teams. This lecture shows how to put categories into a model and how to judge whether they belong in it.</p>

At the same season and requested amount, do all-women presenting teams offer
a different equity (`equity_offered_pct`) than all-men teams in the Shark
Tank file? And do the category terms improve the model at all?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=13">3.5
Extensions of the Linear Model</a>. Focus on predictors that are categories
(the video calls them qualitative predictors) and on how the reference
category decides what a coefficient means. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
The videos explain key ideas in another way and can use a different
notation. The lecture material and its notation always come first.
</div>

**Warm up and predict.** The Shark Tank file sorts the
presenting teams into three categories: `all_men`, `all_women`, and `mixed`.
The video explains dummy variables, which take only the values 0 and 1. How
many dummy variables does a model with an intercept need for three
categories? Write down a number. Then answer this question before you come to
class: at the same season and requested
amount, do you expect all-women teams to offer more, less, or
about the same equity as all-men teams? You will find both answers in
the lecture material.

[Shark Tank data]{.shark-tag} **R warm-up.** The code below builds the subset of pitches and
the factor that the lecture material uses. Run the code and bring the
contrasts table that it prints to class:

```{r}
#| label: i05-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
levels(known_gender$gender)
contrasts(known_gender$gender)
```

**Code along (bring your laptop).** Bring a charged laptop with RStudio and
the course project open. All the code of the lecture is in this chapter,
ready to copy. You can run each step yourself during the lecture and get the
same numbers and figures.

If indicators and reference categories are unfamiliar, review
[Mathematics Refresher: indicators and reference categories](../resources/math-refresher.html#indicators-and-reference-categories).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section first compares the groups of teams with plain averages. It then
shows how a regression uses a variable made of categories, how to analyse
an A/B test, and how to test several coefficients at once.

### Start with a founder's question

[Shark Tank data]{.shark-tag} Founders often compare groups: customers from the app against
customers from the website, one price against another, one kind of team
against another. Start with the comparison in the driving question. The
Shark Tank file sorts the presenting teams into all men, all women, and
mixed (`pitcher_gender_group`). What equity (`equity_offered_pct`) did each
group offer on average?

```{r}
#| label: i05-opening
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known <- subset(sharks, pitcher_gender_group != "unknown")
round(tapply(known$equity_offered_pct, known$pitcher_gender_group, mean), 2)
table(known$pitcher_gender_group)
```

All-men teams offered 13.20% on average, all-women teams 14.02%, and mixed
teams 12.17%. The difference between all-women and all-men teams is 0.82 percentage
points, about 6% of the all-men average. Three questions follow. Is this difference
larger than chance alone would produce? Does it remain when we compare teams from the same season
that asked for the same amount? And how can `lm()` use a variable that
contains words, not numbers?

### Measurement first: what the category records

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk">3.5 Extensions of the Linear Model</a>.</p>

The categories describe the people who presented on screen: all men, all
women, or mixed. They do not record the share of women in a team, and they
do not describe the full team behind the venture. In seven records the
source classifies the team as `unknown`. We leave these records out of every
model in this lecture. A later section of this lecture explains why. Keep these
limits in mind for every result in this chapter.

### A two-level indicator

::: {.definition-box}
<span class="label">Definition</span>

**Dummy variable.** A variable that takes only the values 0 and 1. It is 1
when an observation belongs to a category and 0 when it does not. Other
names are indicator variable and binary variable. A dummy stands in for a
category, so that the category can enter a regression as a number.

**Reference category.** The category for which all dummies are 0. Every
dummy coefficient is a comparison with this category. Other names are
reference group and base group.
:::

The question at the start of this lecture was how `lm()` can use a variable
that contains words. The answer is a dummy. For example, a dummy for
all-women teams is 1 for a pitch by an all-women team and 0 for every other
pitch. The words become numbers, and the regression can use them. You met a
dummy in Lecture 1 as a predictor that is 0 or 1. This lecture shows how to
use one or more dummies for a categorical variable.

For a dummy variable $D_i\in\{0,1\}$,

$$
E[Y_i\mid D_i]=\beta_0+\beta_1D_i.
$$

Then

$$
E[Y_i\mid D_i=0]=\beta_0,
\qquad
E[Y_i\mid D_i=1]=\beta_0+\beta_1.
$$

So $\beta_1$ is the expected difference between the group with $D=1$ and
the reference category with $D=0$. A regression on one dummy is the same
as a comparison of two group means.

::: {#exm-i05-two-ticks}
## Worked example: two group means from two coefficients

[Shark Tank data]{.shark-tag} Keep only the all-men and all-women teams (1,158 records) and let
`women_team` equal 1 for all-women teams:

```{r}
#| label: i05-two-level
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
two_groups <- subset(known_gender,
                     pitcher_gender_group %in% c("all_men", "all_women"))
two_groups$women_team <- as.integer(
  two_groups$pitcher_gender_group == "all_women"
)
coef(lm(equity_offered_pct ~ women_team, data = two_groups))
tapply(two_groups$equity_offered_pct, two_groups$women_team, mean)
```

Read the two coefficients as two marks on the equity axis
(@fig-i05-two-ticks). The intercept $\hat\beta_0\approx 13.20$ is the mean
of the all-men teams. The sum $\hat\beta_0+\hat\beta_1\approx 13.20+0.82=14.02$
is the mean of the all-women teams. The coefficient $\hat\beta_1\approx 0.82$
is the distance between the two marks: the difference between the two
group means.
:::

```{r}
#| label: fig-i05-two-ticks
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 2.2
#| fig-cap: "The model with one dummy places the two group means as two marks on the axis of the equity offered. The intercept is the mean of the reference category (all-men teams), and the dummy coefficient is the distance between the marks."
old_par <- par(no.readonly = TRUE)
par(mar = c(3.5, 1, 1, 1))
plot(NA, xlim = c(12.6, 14.6), ylim = c(0, 1), axes = FALSE,
     xlab = "", ylab = "")
axis(1, at = seq(12.6, 14.6, by = 0.4))
mtext("Mean initial equity offered (%)", side = 1, line = 2.4, cex = 0.9)
segments(13.20, 0.15, 13.20, 0.55, col = "#0072B2", lwd = 4)
segments(14.02, 0.15, 14.02, 0.55, col = "#D55E00", lwd = 4)
arrows(13.20, 0.38, 14.02, 0.38, length = 0.08, code = 3, col = "#17324D")
text(13.20, 0.72, expression(hat(beta)[0] == 13.20), col = "#0072B2")
text(14.02, 0.72, expression(hat(beta)[0] + hat(beta)[1] == 14.02), col = "#D55E00")
text(13.61, 0.25, expression(hat(beta)[1] == 0.82), col = "#17324D", cex = 0.9)
par(old_par)
```

### Comparing two versions: the A/B test {#i05-ab-test}

An invented example. A student venture sells meal-prep boxes through its
website. The team has designed a new checkout page that suggests a drink
with every box. Should the venture switch to the new page? The founders
want to know whether the new page raises the order value, and by how much.

One idea is to compare the orders of last month (old page) with the orders
of this month (new page). But other things also change from one month to
the next: exams, the weather, the prices of competitors. The difference
would mix the effect of the page with all of these. An A/B test avoids this
problem.

::: {.definition-box}
<span class="label">Definition</span>

**A/B test.** A randomised experiment with two groups. The **treatment group** gets the new version (B), and the
**control group** gets the current version (A). With **random assignment**,
chance decides which customer gets which version. The **outcome** $Y_i$ is
what you measure for each customer, such as the order value. The
**treatment dummy** $D_i$ is 1 for the treatment group and 0 for the control
group.
:::

Random assignment is the key step. Chance decides who sees the new page.
So the budget, the taste, or the device of a customer cannot decide it. On
average, the two groups are then alike in everything else, in what you
measure and in what you do not measure. The only systematic difference
between them is the page. So the difference in the mean outcomes estimates
the average effect of the new page on the order value. This is called the
**average treatment effect**.

Compare this with the Shark Tank comparisons of this lecture. Nobody
assigned the gender of a team by chance. Teams of different composition can
also differ in industry, experience, or product. The regression later in this
lecture holds season and requested amount fixed, but not these other differences. That is why
the coefficients of this lecture are adjusted comparisons, not effects. You
met the same contrast in
[QM Basic Lecture 7](../basic/b07-samples-estimators.qmd): random sampling
lets a sample represent a population, and random assignment supports claims
about cause. An A/B test uses random assignment.

Analyse an A/B test with the regression on one dummy from the previous
section:

$$
Y_i=\beta_0+\beta_1D_i+\varepsilon_i.
$$

As shown above, $\hat\beta_0$ is the mean outcome of the control group, and
$\hat\beta_0+\hat\beta_1$ is the mean outcome of the treatment group. So
$\hat\beta_1$ is the difference in means: treatment group minus control
group. Two tools from the
[exam procedure of Lecture 3](i03-inference-fit-ftest.qmd#i03-exam-procedure)
answer the two questions of the founders:

- Does the new version have an effect? Test $H_0:\beta_1=0$ against
  $H_1:\beta_1\ne0$ with the t-test in five steps.
- How large is the effect? Give the 95% confidence interval for $\beta_1$.

This is the comparison of two means that
[QM Basic Lecture 10](../basic/b10-testing-errors-power.qmd) only previewed.
The regression gives the same estimate, the difference between the two
group means, and it adds the standard error, the test, and the interval in
one table.

::: {#exm-i05-ab-test}
## Worked example: an A/B test of a checkout page

An invented example. The venture runs the test with 200 new customers.
Chance puts 100 customers in each group. The code below builds the invented
data. It sets a true effect of 2 euros, which a real founder never knows.

```{r}
#| label: i05-ab-data
set.seed(12)
# random assignment: 100 zeros (old page) and 100 ones (new page),
# shuffled by chance
new_page <- sample(rep(c(0, 1), each = 100))
# invented order values in euros, with a true effect of 2 euros
order_value <- round(25 + 2 * new_page + rnorm(200, mean = 0, sd = 6), 2)
ab_test <- data.frame(new_page, order_value)
head(ab_test)
table(ab_test$new_page)
```

Now regress the order value on the treatment dummy `new_page`:

```{r}
#| label: i05-ab-fit
ab_model <- lm(order_value ~ new_page, data = ab_test)
summary(ab_model)$coefficients
confint(ab_model)
tapply(ab_test$order_value, ab_test$new_page, mean)
qt(0.975, df = 198)
```

The intercept $\hat\beta_0\approx 24.83$ euros is the mean order value of
the control group. The treatment group has a mean of 26.87 euros. The
coefficient $\hat\beta_1\approx 2.04$ euros is the difference, with a
standard error of 0.80.

The test in five steps.

1. **Hypotheses.** $H_0:\beta_1=0$, the new page has no effect on the mean
   order value. $H_1:\beta_1\ne0$. The significance level is 5%.
2. **Test statistic.** $T=\hat\beta_1/\operatorname{SE}(\hat\beta_1)$, with

   $$
   df=n-k-1=200-1-1=198.
   $$
3. **Critical value and rejection region.** The critical value is 1.972,
   from `qt(0.975, df = 198)`. Reject $H_0$ if $|t_{obs}|\ge 1.972$.
4. **Observed value.** $t_{obs}=2.042/0.799\approx 2.56$.
5. **Conclusion.** $2.56\ge1.972$, so reject $H_0$ (p $\approx$ 0.011). At
   the 5% level, the new page raises the mean order value.

The 95% interval for $\beta_1$ in five parts.

1. **Formula.**
   $\hat\beta_1\pm t_{0.975,\,df}\,\operatorname{SE}(\hat\beta_1)$.
2. **Given values.** $\hat\beta_1=2.042$,
   $\operatorname{SE}(\hat\beta_1)=0.799$, $df=198$, and
   $t_{0.975,\,198}=1.972$.
3. **Calculations.** The margin of error is $1.972\times0.799\approx1.576$.
4. **Interval.** $2.042\pm1.576$, from about 0.47 to 3.62 euros.
5. **Conclusion.** With 95% confidence, the new page raises the mean order
   value by between about 0.47 and 3.62 euros. The interval lies above 0,
   which agrees with the test.

The interval contains the true effect of 2 euros that the code built in. To
judge the size, compare the estimate with the mean of the control group:

```{r}
#| label: i05-ab-share
ab_coef <- coef(ab_model)
round(100 * ab_coef["new_page"] / ab_coef["(Intercept)"], 1)  # effect, % of control mean
round(100 * confint(ab_model)["new_page", ] / ab_coef["(Intercept)"], 1)
```

The new page raises the order value by about 8.2% of the control mean. The
interval runs from about 1.9% to 14.6%. Even the lower end is a gain. If the
new page costs nothing extra to run, the decision is to switch to it.
:::

An A/B test also has limits.

- The result holds for the customers and the period of the test. Customers
  in another season, or another kind of customer, may react differently.
- Each group needs enough customers. With few customers per group, the
  interval is wide, and a real effect can go unnoticed.
- Each customer must see only one version. A customer who sees both
  versions belongs to neither group.
- Testing many versions or many outcomes needs care. With many tests at the
  5% level, some will reject by chance alone, even if no version has an
  effect.

### When you may use causal words {#i05-causal-words}

Since Lecture 1 you describe a coefficient with "is associated with" and
"on average". You do not use causal words such as "increases", "raises",
"leads to", or "the effect of". Most data in business are observational:
who got what was not randomly assigned. Then a coefficient mixes the role of
the predictor with the role of everything that goes with it, measured or
not. Control variables remove only the part that you measured.

A well-run A/B test is the exception. Here causal words are correct, because
the way the data were produced rules out the other explanations. Check three
conditions before you use them:

1. Random assignment. Chance decided which customer got which version. The
   customers did not choose it, and neither did the team.
2. One difference. During the test, the two groups had the same period, the
   same prices, and the same website, apart from the version you test.
3. Same measurement. You measured the outcome in the same way and over the
   same time in both groups, and every customer kept the version assigned.

If all three hold, the difference in means estimates the average treatment
effect. You may then write "the new page raises the mean order value". If
one condition fails, go back to the default wording. An example: the
customers chose between the old and the new page themselves. Customers who
like extras may pick the new page and also order more. Then the comparison
is observational again.

Compare two sentences about the same question:

- Observational data, towns with and without free delivery: "Orders in
  towns with free delivery are on average 2 euros higher." This is an
  association. The towns may differ in more than delivery.
- A/B test, free delivery assigned by chance to customers: "Free delivery
  raises the mean order value by about 2 euros." This is a causal claim.

The limits above still apply to a causal claim. It holds for the customers
and the period of the test. It is also an average: it does not say that
every customer orders more.

### More than two categories

[Shark Tank data]{.shark-tag} With $G$ categories and an intercept, use $G-1$ dummies. With all $G$
dummies and an intercept, the dummies sum to 1 in every row, which is the
same as the intercept column. This is perfect multicollinearity from
Lecture 4: one column can be calculated exactly from the others, so R cannot
estimate all the coefficients. With dummies, this situation is sometimes
called the dummy variable trap.

R creates the dummies itself when a predictor is a factor. Choose the
reference category yourself and check it. This is the code of the R warm-up:

```{r}
#| label: i05-reference
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
levels(known_gender$gender)
contrasts(known_gender$gender)
```

Each row of the contrasts table is a category. Each column is one dummy. The
reference category `all_men` is the row with only zeros. The intercept
belongs to this category. This answers the first question from before class: three
categories need two dummies.

::: {.callout-note collapse="true"}
## Check: four industries and an intercept, how many dummies?

Three. One industry is the reference category, and the intercept belongs to
this category. Each
of the three dummy coefficients compares one of the other industries with
it. Four dummies and an intercept would give perfect multicollinearity.
:::

::: {.callout-caution collapse="true"}
## Try it: three sales channels

Try this with your neighbour first. In an invented example, a venture sells
through three channels: the campus stand, the app, and the website. A model
of the order value in euros uses the campus stand as the reference and
reports `channelApp 6.0` and `channelWebsite 2.5`, with an intercept of 9.0.

*Step 1 (given).* With three categories and an intercept, the model needs
two dummies. The campus stand has no dummy. Its mean is the intercept.

*Step 2 (you).* Give the fitted mean order value for each of the three
channels.

*Step 3 (you).* By how much does the fitted order value in the app differ
from the fitted order value on the website? Neither is the reference category.

::: {.callout-note collapse="true"}
## Check your answer

Campus stand: 9.0 euros. App: 9.0 + 6.0 = 15.0 euros. Website: 9.0 + 2.5 =
11.5 euros. App minus website: 6.0 - 2.5 = 3.5 euros. You can subtract two
coefficients that share a reference category from each other. The result
compares their two categories directly.
:::
:::

### Percent versus percentage points

[Shark Tank data]{.shark-tag} The outcome `equity_offered_pct` is measured in percent, so its regression
coefficients are in **percentage points**. The model in the next section
estimates a coefficient of 1.28 for all-women teams. The sentence "the all-women coefficient is 1.28%" would be ambiguous.
It could mean:

- a relative change: 1.28% of the all-men level more (13.2% would become
  about 13.4%)
- an absolute change of 1.28 percentage points (13.2% would become 14.48%).

The two readings differ by a factor of about eight here. A coefficient for
an outcome in percent is always an absolute change in percentage points. The precise sentence is:
"At the same season and requested amount, all-women presenting teams are
fitted to offer 1.28 equity percentage points more than all-men presenting
teams." Always write percentage points for this outcome.

### The adjusted comparison

[Shark Tank data]{.shark-tag} Now fit the model of the driving question. It contains both category
dummies, and it holds season and requested amount (in thousands of dollars,
`ask_1k`) fixed.

```{r}
#| label: i05-fit
gender_model <- lm(
  equity_offered_pct ~ season + ask_1k + gender,
  data = known_gender
)
summary(gender_model)
```

Interpret both dummy coefficients with a full sentence. Each sentence names
the two categories, the unit, the predictors that are held fixed, and the
records it is about:

- **All-women versus all-men.** Among the 1,434 pitches with a known team
  classification, all-women presenting teams are fitted to offer about 1.28
  equity percentage points more than all-men presenting teams at the same
  season and requested amount (p $\approx$ 0.008).
- **Mixed versus all-men.** Among the same pitches, mixed teams are fitted
  to offer about 0.12 percentage points more than all-men teams at the same
  season and requested amount (p $\approx$ 0.823). This estimate is not
  statistically distinguishable from zero.

The first sentence answers the prediction
question from before class: all-women
teams offer more. The adjusted difference between all-women and all-men teams (1.28
points) is larger than the unadjusted difference from the worked example (0.82 points). Holding season and
requested amount fixed changed the comparison. This is why the sentence must
name what is held fixed. Both coefficients are comparisons with the
reference category. They do not show that the gender of a team causes the
offer.

Changing the reference category changes the names and values of the
coefficients. It does not change the fitted values or the overall fit. Check
this in R:

```{r}
#| label: i05-change-reference
known_gender$gender_women_ref <- relevel(known_gender$gender,
                                         ref = "all_women")
coef(lm(equity_offered_pct ~ season + ask_1k + gender_women_ref,
        data = known_gender))
```

The same three fitted group means are now described as comparisons with the
all-women teams. The all-men coefficient is $-1.28$, the same difference as before
with the opposite sign, and the fitted values are identical.

### Are the category terms worth having at all?

So far we read the coefficients one at a time. The driving question also
asks something that no single t-test answers: do the two category terms
together add information beyond season and requested amount?

::: {.definition-box}
<span class="label">Definition</span>

**Partial F-test** (F-test for joint significance). A test of whether
several coefficients are all zero at the same time. It compares a complete
model with a reduced model that leaves out the tested terms. Econometrics
textbooks call these two models the unrestricted and the restricted model.
Both models are fitted on the same observations.

$$
F=\dfrac{(\text{SSE}_R-\text{SSE}_C)/q}{\text{SSE}_C/(n-k_C-1)},
$$

with $\text{SSE}_R$ and $\text{SSE}_C$ the sums of squared residuals of the
reduced and the complete model, $q$ the number of tested coefficients, and
$k_C$ the number of slopes of the complete model.
:::

Two such models are called **nested**: the reduced model is the complete
model with the tested coefficients set to zero. @fig-i05-nested shows the
pair for the two category dummies.

```{r}
#| label: fig-i05-nested
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 2.6
#| fig-cap: "The two nested models of the partial F-test. The reduced model is the complete model with the two category coefficients set to zero, fitted on the same observations."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.3, 0.3, 0.3, 0.3))
plot(NA, xlim = c(0, 10), ylim = c(0, 5), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")
rect(0.4, 1.1, 4.6, 3.9, col = "#DCEAF3", border = "#17324D", lwd = 2)
rect(5.4, 1.1, 9.6, 3.9, col = "#FDE5C1", border = "#17324D", lwd = 2)
text(2.5, 3.35, "Reduced model", font = 2, col = "#17324D")
text(2.5, 2.55, "season + ask_1k", cex = 0.95)
text(2.5, 1.85, expression(beta["all_women"] == 0 ~ "," ~ beta["mixed"] == 0),
     cex = 0.95, col = "#5B6976")
text(7.5, 3.35, "Complete model", font = 2, col = "#17324D")
text(7.5, 2.55, "season + ask_1k", cex = 0.95)
text(7.5, 1.85, "+ two category dummies", cex = 0.95)
arrows(4.7, 2.5, 5.3, 2.5, length = 0.1, lwd = 2, col = "#17324D")
text(5.0, 3.0, "free the\ntwo coefficients", cex = 0.8, col = "#5B6976")
par(old_par)
```

The hypotheses:

$$
H_0:\beta_{\text{all\_women}}=\beta_{\text{mixed}}=0
\quad\text{against at least one non-zero.}
$$

::: {#exm-i05-partial-f}
## Worked example: the partial F statistic by hand

[Shark Tank data]{.shark-tag} Fit both models and ask R for the anova table:

```{r}
#| label: i05-partial-f
reduced_gender <- lm(equity_offered_pct ~ season + ask_1k,
                     data = known_gender)
anova(reduced_gender, gender_model)
```

Now calculate F by hand from the numbers in the table. The residual sum of
squares (RSS in the R output, SSE in this course) falls by $436.25$ when you
add the two category terms. This uses $2$ degrees of freedom. The complete
model leaves a residual sum of squares of $82{,}792$ on $1{,}429$ degrees of
freedom:

$$
F=\frac{436.25/2}{82{,}792/1{,}429}
=\frac{218.13}{57.94}
\approx 3.76.
$$

Read the ratio. The numerator is the variation that the two terms explain,
per tested coefficient. The denominator is the variation that is left
unexplained, per residual degree of freedom. If the two coefficients were
really zero, the ratio would be close to 1. Here it is about 3.8, and the
p-value is about 0.023.
:::

At 5%, the two category terms are jointly significant in this model. Compare
this with the single results: the p-value of the joint test is $0.023$, and
the p-value of the mixed dummy alone is $0.823$. Joint significance is not the
same as individual significance. The F-test asks whether the two terms
together improve the model. It does not ask whether each term is significant
on its own. The overall F-test in `summary(gender_model)` is about 84.5, with a
very small p-value. It answers a third question: do all four coefficients
together improve on a model with only an intercept?

### The same model as a one-way ANOVA

Take away season and requested amount for a moment. Then the only
predictors are the dummies of one categorical variable. Such a regression
compares group means, and it has its own name.

::: {.definition-box}
<span class="label">Definition</span>

**One-way analysis of variance (ANOVA).** A regression of the outcome on the
dummies of one categorical variable with $G$ groups, and nothing else. It
compares the $G$ group means. Its F-test has the null hypothesis that all
group means are equal, against the alternative that at least one differs.
:::

In this model each fitted value is the mean of the group of the
observation. Write $\bar y_g$ for the mean of group $g$, $n_g$ for its
number of observations, and $\bar y$ for the overall mean. The intercept is
the mean of the reference category. Each dummy coefficient is the difference
between the mean of its group and the mean of the reference category.

The variation splits as in Lecture 3, $\text{SST}=\text{SSR}+\text{SSE}$.
Put the group mean $\bar y_g$ in place of the fitted value $\widehat y_i$.
Then SSR measures the variation between the groups: how far the group means
lie from the overall mean, weighted by the size of each group. SSE measures
the variation within the groups: how far each observation lies from the
mean of its own group.

$$
\text{SSR}=\sum_{g=1}^{G}n_g(\bar y_g-\bar y)^2,
\qquad
\text{SSE}=\sum_{g=1}^{G}\ \sum_{i\text{ in group }g}(y_i-\bar y_g)^2.
$$

The model has $k=G-1$ slopes, so the residual degrees of freedom are
$n-k-1=n-G$. The F statistic of the ANOVA is

$$
F=\frac{\text{SSR}/(G-1)}{\text{SSE}/(n-G)}.
$$

This is the overall F-test of Lecture 3 for the dummy regression. Its null
hypothesis $H_0:\mu_1=\mu_2=\cdots=\mu_G$ says that the population means
of all groups are equal. This is the same as saying that all dummy
coefficients are zero. A small p-value says that at least one group mean
differs. It does not say which one.

R prints the numbers of this test in an **ANOVA table**. It has one row for
the group variable and one row for the residuals. The columns are:

- `Df`: the degrees of freedom, $G-1$ for the groups and $n-G$ for the
  residuals.
- `Sum Sq`: the sum of squares, SSR for the groups and SSE for the
  residuals.
- `Mean Sq`: the mean square, `Sum Sq` divided by `Df`.
- `F value`: the F statistic, the mean square of the groups divided by the
  mean square of the residuals.
- `Pr(>F)`: the p-value of the F-test.

[Shark Tank data]{.shark-tag} Return to the raw comparison at the start of
this lecture. Are the three group means of the equity offered
(`equity_offered_pct`) further apart than chance alone would produce? Fit
the model with the gender groups only, without season and requested
amount, and ask R for the ANOVA table:

```{r}
#| label: i05-one-way-anova
anova_model <- lm(equity_offered_pct ~ gender, data = known_gender)
anova(anova_model)
summary(anova_model)
```

The coefficients repeat the group means from the start of this lecture. The
intercept 13.20 is the mean of the all-men teams. The all-women mean is
$13.20+0.82=14.02$, and the mixed mean is $13.20-1.03=12.17$.

Now check the table by hand. Each mean square is the sum of squares divided
by its degrees of freedom:

$$
\frac{551}{2}\approx 275.5,
\qquad
\frac{101{,}822}{1{,}431}\approx 71.15,
\qquad
F=\frac{275.5}{71.15}\approx 3.87.
$$

This is the `F-statistic` in the last line of `summary()`: 3.872 on 2 and
1,431 degrees of freedom, with the same p-value 0.021. The mean square of
the residuals is $s^2$, and $\sqrt{71.15}\approx 8.435$ is the residual
standard error. The code below computes SSR from the group means with the
formula above, and the overall mean and standard deviation for comparison:

```{r}
#| label: i05-anova-by-hand
y <- known_gender$equity_offered_pct
group_means <- tapply(y, known_gender$gender, mean)
group_sizes <- table(known_gender$gender)
sum(group_sizes * (group_means - mean(y))^2)   # SSR, between groups
c(mean = mean(y), sd = sd(y))
```

At the 5% level, the three groups of teams do not all offer the same mean
equity (F $\approx$ 3.87, p $\approx$ 0.021). The largest difference, between
all-women teams (14.02%) and mixed teams (12.17%), is 1.85 percentage
points. That is about 14% of the overall mean of 13.22% and about 0.22
standard deviations of the equity offered. The groups explain only about
0.5% of the variation ($R^2\approx 0.005$), so most of the differences
between offers lie within the groups.

Once season and requested amount are added, as in the adjusted comparison
of this lecture, the model is no longer a one-way ANOVA. Its fitted values
are then no longer the raw group means. ANOVA with more than one
categorical variable, and its extensions, come in QM Advanced.

::: {#exm-i05-anova-hand}
## Worked example: a one-way ANOVA by hand

An invented example. A venture sells through three channels: the campus
stand, the app, and the website. It records the order value in euros of 10
orders per channel, so $n=30$ and $G=3$. The mean order values are 10 euros
(campus stand), 14 euros (app), and 12 euros (website). The variation within
the channels is $\text{SSE}=270$. Two values are given: the critical value
3.35, from `qf(0.95, 2, 27)`, and the p-value 0.030.

First the overall mean and SSR. All groups have the same size, so the
overall mean is $(10+14+12)/3=12$ euros. Then

$$
\text{SSR}=10(10-12)^2+10(14-12)^2+10(12-12)^2=40+40+0=80.
$$

Now the test in five steps.

1. **Hypotheses.** $H_0:\mu_{\text{stand}}=\mu_{\text{app}}=\mu_{\text{website}}$,
   the three channels have the same mean order value. $H_1$: at least one
   mean differs. The significance level is 5%.
2. **Test statistic.** $F=\dfrac{\text{SSR}/(G-1)}{\text{SSE}/(n-G)}$,
   with $G-1=2$ and $n-G=27$ degrees of freedom.
3. **Critical value and rejection region.** The critical value is 3.35.
   Reject $H_0$ if $F_{obs}\ge 3.35$.
4. **Observed value.**
   $F_{obs}=\dfrac{80/2}{270/27}=\dfrac{40}{10}=4.0$.
5. **Conclusion.** $F_{obs}=4.0$ lies in the rejection region, so reject
   $H_0$ (p $\approx$ 0.030). At the 5% level, the mean order value differs
   between at least two of the channels. The test does not say which ones.
   The largest difference, 4 euros between the app and the campus stand, is a
   third of the overall mean.
:::

### Visualise the adjusted comparison

[Shark Tank data]{.shark-tag} @fig-i05-adjusted shows the fitted equity offered for the three groups of
teams at the same season and requested amount.

```{r}
#| label: fig-i05-adjusted
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted equity offered for the three groups of presenting teams at the average season and requested amount, with 95% confidence intervals. The all-women point lies above the other two. The intervals show sampling uncertainty under the model, not uncertainty about how the categories were recorded."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
gender_model <- lm(
  equity_offered_pct ~ season + ask_1k + gender,
  data = known_gender
)
newdat <- data.frame(
  season = mean(known_gender$season),
  ask_1k = mean(known_gender$ask_1k),
  gender = factor(c("all_men", "all_women", "mixed"),
                  levels = levels(known_gender$gender))
)
pred <- predict(gender_model, newdata = newdat, interval = "confidence")
adj_df <- data.frame(
  team = factor(c("All-men", "All-women", "Mixed"),
                levels = c("All-men", "All-women", "Mixed")),
  fit = pred[, "fit"],
  lwr = pred[, "lwr"],
  upr = pred[, "upr"]
)

ggplot(adj_df, aes(x = team, y = fit)) +
  geom_pointrange(aes(ymin = lwr, ymax = upr),
                  colour = "#009E73", linewidth = 0.7, fatten = 2.6) +
  labs(x = "Source classification of pitching team",
       y = "Fitted initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-i05-adjusted with the
[five steps for reading a figure](../basic/overview.qmd#how-to-read-every-graph):

1. **Unit and variables:** the horizontal axis shows the three categories of
   teams. The vertical axis shows the fitted equity offered, in percent.
2. **Scales:** each point is a fitted mean at the same average season and
   requested amount. This is an adjusted comparison, not a comparison of raw
   group means.
3. **One element:** the points are fitted values (about 12.9, 14.1, and
   13.0). The vertical lines are 95% confidence intervals. They are wider
   for the smaller groups.
4. **Pattern:** the all-women point lies about 1.28 points above the all-men
   point. The mixed interval overlaps both.
5. **Limits:** the intervals show sampling uncertainty under the model. They
   say nothing about how well the broad categories were recorded, and
   nothing about causes.

### Why `unknown` is excluded, not a fourth group

Seven records have the classification `unknown`. This label says that
information is missing. It is not a kind of team. A fourth category would
estimate the mean equity of teams that the source could not classify. This
number has no business meaning. If we leave these records out, every
coefficient has a clear meaning. The cost is that the results describe the 1,434
classified records only.

::: {.caveat-box}
<span class="label">What these category coefficients can and cannot say</span>

- The categories are a broad description of the people who presented. They
  do not record shares of women, and they do not describe the full founding
  team.
- Each coefficient compares one group with the reference category, at the same
  season and requested amount.
- The model says that offers differ between the coded groups. It does not say
  that the composition of a team makes anyone offer differently.

**What you can do with it:** when you can decide who gets which version,
run an A/B test with random assignment ([Comparing two versions: the A/B
test](#i05-ab-test)). Then the coefficient of the treatment dummy estimates
the average treatment effect.
:::

### Answering the driving question

At the same season and requested amount, all-women presenting teams are
fitted to offer about 1.28 equity percentage points more than all-men teams
(p $\approx$ 0.008). That is about one tenth of the average equity offered by
all-men teams (13.20%). The estimate for mixed teams is not statistically
different from all-men teams. The two category terms are jointly significant (partial F $\approx$ 3.76,
p $\approx$ 0.023), so they belong in the model. The result is an adjusted
association among these recorded pitches, not an effect of gender.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You want to know whether free delivery raises the order value in your web shop. Do not compare towns with free delivery against towns without it. Such towns differ in more than delivery, just as the Shark Tank teams differ in more than gender. Run an A/B test instead, as in [Comparing two versions: the A/B test](#i05-ab-test). Offer free delivery to a treatment group of new customers chosen by chance, and keep a control group without it. Regress the order value on the treatment dummy. Switch to free delivery if the whole 95% interval for $\beta_1$ lies above zero and the gain per order is larger than the delivery cost per order.
:::

**Next:** You can now compare categories at fixed values of the other
predictors. So far every group has the same slope, and every trend is a
straight line. Both are choices of the model, not facts. Lecture 6 removes
these two restrictions. An interaction lets the season trend differ between
groups of teams, and a quadratic term lets the trend curve.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main ideas. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly, the course AI
assistant.

### Recall questions

1. Why are only $G-1$ dummies used with an intercept?
2. What does a dummy coefficient compare?
3. What changes when the reference category changes?
4. What does a partial F-test evaluate?
5. Why can a joint F-test reject when one individual t-test does not?
6. In an A/B test, what do $\hat\beta_0$ and $\hat\beta_1$ of the
   regression on the treatment dummy measure, and why can you read
   $\hat\beta_1$ as an effect?
7. What does the F-test of a one-way ANOVA test, and how is it related to
   the regression on the dummies of one categorical variable?
8. In this course you describe a coefficient without causal words. When
   may you use them, and which three conditions do you check first?

::: {.callout-note collapse="true"}
## Check your answers

1. All $G$ dummies sum to 1 in every row, which is the same as the intercept
   column. This is perfect multicollinearity, so R could not estimate all
   the coefficients.
2. It compares a category with the named reference category, holding the
   other included predictors fixed.
3. The names and values of the coefficients change, because they now compare
   with a different category. The fitted values and the overall fit do not
   change.
4. Whether several coefficients are all zero at the same time. It compares
   two nested models on the same observations.
5. The F-test evaluates the terms together. Each t-test asks a narrower
   question about one comparison with the reference category, and it can miss
   what the terms explain together.
6. $\hat\beta_0$ is the mean outcome of the control group, and
   $\hat\beta_1$ is the difference in means, treatment group minus control
   group. Random assignment makes the two groups alike on average in
   everything else, measured or not. So $\hat\beta_1$ estimates the average
   treatment effect. In the Shark Tank file nobody assigned team gender, so
   the coefficients there are adjusted comparisons only.
7. It tests whether all group means are equal, against the alternative
   that at least one differs. It is the overall F-test of the regression on
   the $G-1$ dummies, with nothing else in the model. A rejection says that
   some group mean differs, not which one.
8. Only for a well-run randomised experiment, such as an A/B test. Check
   random assignment (chance decided who got which version), one difference
   (same period, prices and website apart from the tested version), and the
   same measurement in both groups, with every customer keeping the assigned
   version. If one condition fails, use "is associated with" and "on
   average".
:::

### Finish a half-worked example: a partial F with new numbers

[Shark Tank data]{.shark-tag} Test the two team-category dummies again, now in a smaller complete
model. The outcome is the equity offered, in percent (`equity_offered_pct`).
The data are the 1,434 pitches with a known team classification
(`known_gender`). The reduced model contains only the season (`season`). The
complete model adds the team category (`gender`, reference all-men teams):

```{r}
#| label: i05-faded-anova
reduced_season <- lm(equity_offered_pct ~ season, data = known_gender)
full_season    <- lm(equity_offered_pct ~ season + gender,
                     data = known_gender)
anova(reduced_season, full_season)
```

*Step 1 (given).* Adding the two dummies reduces the residual sum of squares
by $643.43$ using $2$ degrees of freedom. The complete model leaves
$84{,}003$ on $1{,}430$ degrees of freedom.

*Step 2 (you).* Put these numbers into the partial F formula of the
definition box in the section "Are the category terms worth having at all?".

*Step 3 (you).* Compare your F with the F and the p-value in the R output
above. State the conclusion at 5% in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$F=(643.43/2)/(84{,}003/1{,}430)=321.7/58.7\approx 5.48$, matching the
table, with p $\approx$ 0.004. At 5%, the two category terms are jointly
significant even in the model that adjusts only for season.
:::

### Practice: relevelling, hypotheses, an A/B test, and an ANOVA table

Questions 1, 2, 3, and 5 use the model of the section "The adjusted
comparison" in this lecture, saved as `gender_model`. It regresses the
equity offered (`equity_offered_pct`) on the season, the requested amount
in thousands of dollars (`ask_1k`), and the team category `gender`, with
all-men teams as the reference category. The data are the 1,434 pitches
with a known team classification (`known_gender`).

1. [Shark Tank data]{.shark-tag} Fit the same model again, but with mixed
   teams as the reference category. Save it as `relevelled_model`, and
   interpret both dummy coefficients.
2. [Shark Tank data]{.shark-tag} Show that the fitted values of
   `relevelled_model` are the same as those of `gender_model`.
3. State the hypotheses of the partial F-test of both gender dummies in
   `gender_model`.
4. An invented A/B test, analysed as in
   [Comparing two versions: the A/B test](#i05-ab-test). Random assignment
   puts 200 new customers in a control group (old checkout page) and a
   treatment group (new page), 100 each. A regression of the order value
   (euros) on the treatment dummy `new_page` reports an intercept of 18.40
   and a coefficient of 1.20 with standard error 0.50. Give the mean order
   value of each group. Test at 5% whether the new page has an effect, in
   five steps. Give a 95% interval for $\beta_1$, and the effect as a share
   of the control mean. The critical value is
   $t_{0.975,\,198}$ = `qt(0.975, 198)` = 1.972.
5. [Shark Tank data]{.shark-tag} In `gender_model`, the all-women
   coefficient is about 1.28. A student writes "the all-women coefficient is
   1.28%". Explain why this sentence is unclear, then rewrite it precisely.
6. An invented example. A franchise compares the weekly revenue (in
   thousands of euros) of its shops in four regions. Fill in the three
   missing numbers A, B, and C of this ANOVA table:

   ```text
               Df Sum Sq Mean Sq F value  Pr(>F)
   region       3    240       A       B 0.00053
   Residuals   96      C    12.5
   ```

   How many shops are in the data? State the conclusion at 5%. The critical
   value is `qf(0.95, 3, 96)` = 2.70.

::: {.callout-note collapse="true"}
## Check your answers

1. Use:

   ```r
   mixed_reference <- relevel(known_gender$gender, ref = "mixed")
   relevelled_model <- lm(
     equity_offered_pct ~ season + ask_1k + mixed_reference,
     data = known_gender
   )
   coef(relevelled_model)
   ```

   At the same season and requested amount, the all-men coefficient is
   about $-0.121$ percentage points and the all-women coefficient about
   $1.163$ percentage points. Both compare with mixed teams.
2. `all.equal(fitted(gender_model), fitted(relevelled_model))` returns
   `TRUE`.
3. $H_0:\beta_{\text{all\_women}}=\beta_{\text{mixed}}=0$ against the alternative that at
   least one of the two is not zero, with all-men as the reference category.
4. Control group (old page): 18.40 euros. Treatment group (new page):
   $18.40+1.20=19.60$ euros.

   The test in five steps.

   1. **Hypotheses.** $H_0:\beta_1=0$, the new page has no effect on the
      mean order value. $H_1:\beta_1\ne0$. The significance level is 5%.
   2. **Test statistic.** $T=\hat\beta_1/\operatorname{SE}(\hat\beta_1)$,
      with $df=n-k-1=198$.
   3. **Critical value and rejection region.** $t_{0.975,\,198}=1.972$.
      Reject $H_0$ if $|t_{obs}|\ge1.972$.
   4. **Observed value.** $t_{obs}=1.20/0.50=2.40$.
   5. **Conclusion.** $2.40\ge1.972$, so reject $H_0$. At the 5% level, the
      new page raises the mean order value.

   The interval in five parts.

   1. **Formula.**
      $\hat\beta_1\pm t_{0.975,\,df}\,\operatorname{SE}(\hat\beta_1)$.
   2. **Given values.** $\hat\beta_1=1.20$,
      $\operatorname{SE}(\hat\beta_1)=0.50$, $df=198$, and
      $t_{0.975,\,198}=1.972$.
   3. **Calculations.** The margin of error is $1.972\times0.50=0.986$.
   4. **Interval.** $1.20\pm0.986$, from about 0.21 to 2.19 euros.
   5. **Conclusion.** With 95% confidence, the new page raises the mean
      order value by between about 0.21 and 2.19 euros.

   The effect is $1.20/18.40\approx6.5\%$ of the control mean. Random
   assignment makes the two groups alike on average in everything else, so
   the coefficient estimates the average treatment effect of the page.
5. It confuses percent with percentage points, and it does not name the
   comparison group or what is held fixed. A precise version: "At the same
   season and requested amount, all-women presenting teams are fitted to
   offer 1.28 equity percentage points more than all-men presenting teams."
6. A $=240/3=80$. C $=12.5\times96=1{,}200$. B $=80/12.5=6.4$. With
   $G=4$ regions and $n-G=96$, there are $n=100$ shops. Because
   $6.4\ge2.70$ (p $\approx$ 0.0005), reject $H_0$: at the 5% level, the
   mean weekly revenue differs between at least two regions. The regions
   explain $240/(240+1{,}200)\approx17\%$ of the variation in weekly
   revenue.
:::

### The 120-word interpretation

[Shark Tank data]{.shark-tag} Write an interpretation of `gender_model`
(see the Practice above) in about 120 words. Include what the team
categories do and do not record, and say why the result is not causal.

::: {.callout-note collapse="true"}
## Check a model answer

A complete answer reports five things. First, the outcome and the sample:
the equity offered, among the 1,434 broadcast pitches with a known team
classification. Second, both estimates with units and with what is held
fixed: all-women teams are fitted 1.28 percentage points above all-men
teams and mixed teams 0.12 points above, at the same season and requested
amount. Third, the uncertainty: p $\approx$ 0.008 and 0.823 for the two
coefficients, and p $\approx$ 0.023 for the partial F-test, so the terms are
jointly significant. Fourth, the limits of the measurement: the categories
describe the people who presented, not shares of women or the full team.
Fifth, the scope: these are adjusted associations among the recorded
pitches, not causal effects.
:::

### Review: one question from Lecture 4

[Shark Tank data]{.shark-tag} Lecture 4 regressed the equity offered on
season, requested amount (`ask_1k`, in thousands of dollars), and the number
of words in the description. Its slopes are about $-0.836$ per season and
$-0.0029$ per \$1,000 requested. Compute the combined fitted difference for
a pitch three seasons later that requests \$200,000 more, at the same
description length.

::: {.callout-note collapse="true"}
## Check your answer

$3(-0.836)+200(-0.0029)\approx-3.09$ percentage points. The two
contributions add because the model is additive, and \$200,000 is 200 units
of `ask_1k`.
:::

### Worked exam interpretation

**Exam-style question.** An invented example.
A model of equity offered on season, the requested amount in thousands of
dollars (`ask_1k`), and industry (four categories, reference
`Food and Beverage`) reports
`industryTechnology 1.9 (p = 0.04)`. A student
writes: "Technology pitches offer 1.9% more equity, so entering tech makes
founders more generous." Correct the sentence.

::: {.callout-note collapse="true"}
## Check a model answer

Three corrections. Units: the estimate is 1.9 equity percentage points, not
a relative 1.9%. Comparison and what is held fixed: technology pitches are
fitted 1.9 points above Food and Beverage pitches at the same season and
requested amount, among the recorded pitches. Scope: nothing here is causal.
"Entering tech makes founders more generous" treats a comparison with the
reference category as an effect of choosing an industry.
:::

### Check the AI

[Shark Tank data]{.shark-tag} You regressed the equity offered on the season, the requested amount, and the team category, with all-men teams as the reference. You asked an AI assistant: "The coefficient of the all-women dummy is 1.28. What does that mean?" It answered:

> "All-women teams offer 1.28% more equity than the average team."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

Three errors. The unit is percentage points, not percent. The comparison is with the reference category, all-men teams, not with the average team. And the answer leaves out that the comparison is at the same season and requested amount. Correct: at the same season and requested amount, all-women teams are fitted to offer 1.28 percentage points more equity than all-men teams.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Name one
categorical feature of your customers or products (segment, channel, region,
plan tier). Then write two sentences. Which category would you set as the
reference category, and why? Suppose the F-test for the whole set of categories
shows that they are jointly significant. How would this change the way you
describe your market?
There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "My channels are the campus stand, the app, and the website. I would set the campus stand as the reference, because it is the first channel and every decision is compared with it. If the F-test shows that the channel dummies are jointly significant, it makes sense to describe my market by channel. The p-values of the single dummies, some small and some large, could not answer that question."
:::

You can discuss this task, and any other task in this chapter, with
Tilly for QM Intermediate, the course AI assistant. Ask it to check whether your two
sentences confuse percent with percentage points, or to ask you questions
about perfect multicollinearity with your own categories.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask Tilly for
QM Intermediate, the course AI assistant. For example: "Give me an extra
exercise for Intermediate Lecture 5. Show the solution only after I send my
answer."

[Continue to Lecture 6: Interactions and curvature](i06-interactions-curvature.qmd).

[Code sheet: Lecture 5 commands and formulas](code-sheet.qmd#lecture-5-factors-reference-categories-and-partial-f-tests).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.2.2 and 3.3.1, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 20--21.

:::
