# Lecture 5: Categories as predictors and joint tests {#i05}

```{r}
#| label: i05-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 5, you can</span>

- put a categorical predictor with two or more categories into a model with
  dummy variables
- interpret dummy coefficients against a named reference group, in
  percentage points where the outcome is a percentage
- avoid the dummy-variable trap
- compute a partial F statistic by hand from an anova table and read the
  R output
- tell whether a set of terms is jointly useful, and tell this apart from
  the significance of a single term
- tell an adjusted association apart from a causal claim.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- set a reference group and interpret dummy coefficients against it
- use percentage points for a percent outcome
- calculate a partial F statistic from an `anova()` table
- tell joint usefulness from individual significance
- read an A/B test as a dummy regression.

**Enough to recognise:**

- the `contrasts()` table.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Markets come in categories: sales channels, customer segments, kinds of teams. This lecture shows how to put categories into a model and how to judge whether they belong in it.</p>

At the same season and requested amount, do all-women presenting teams offer
a different equity (`equity_offered_pct`) than all-men teams in the Shark
Tank file? And are the category terms worth having in the model at all?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

**Minimum path:** the video and the R task. **Full path:** everything in this section.

Bring a charged laptop with RStudio and the course project open. The code of
the lecture is in this chapter, ready to copy, so you can run each step
yourself during the lecture and get the same numbers and figures.

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=13">3.5
Extensions of the Linear Model</a>. Focus on predictors that are categories
(the video calls them qualitative predictors) and on how the reference
category decides what a coefficient means. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate and predict (5 minutes).** The Shark Tank file sorts the
presenting teams into three categories: `all_men`, `all_women`, and `mixed`.
The video explains dummy variables, which take only the values 0 and 1. How
many dummy variables does a model with an intercept need for three
categories? Write down a number. Then answer the one-click Wooclap question
posted on Canvas before you come to class: at the same season and requested
amount, do you expect all-women teams to offer **more**, **less**, or
**about the same** equity as all-men teams? You find both answers in the
lecture material.

**R warm-up (10 minutes).** The lecture material starts from the subset and
the factor that you build here. Run the code and bring the contrasts table
that it prints:

```{r}
#| label: i05-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
levels(known_gender$gender)
contrasts(known_gender$gender)
```

If indicators and reference categories are unfamiliar, review
[Mathematics Refresher: indicators and reference categories](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section first compares the groups of teams with plain averages. It then
shows how a variable made of categories enters a regression, and how to test
several coefficients at once.

### Start with a founder's question

Founders compare groups all the time: customers from the app against
customers from the website, one price against another, one kind of team
against another. Start with the comparison in the driving question. The
Shark Tank file sorts the presenting teams into all men, all women, and
mixed (`pitcher_gender_group`). What equity (`equity_offered_pct`) did each
group offer on average?

```{r}
#| label: i05-opening
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known <- subset(sharks, pitcher_gender_group != "unknown")
round(tapply(known$equity_offered_pct, known$pitcher_gender_group, mean), 2)
table(known$pitcher_gender_group)
```

All-men teams offered 13.20% on average, all-women teams 14.02%, and mixed
teams 12.17%. The gap between all-women and all-men teams is 0.82 percentage
points, about 6% of the all-men average. Three questions follow. Is this gap
more than chance? Does it remain when we compare teams from the same season
that asked for the same amount? And how does a variable made of words, not
numbers, get into `lm()` at all? The sections below answer them in reverse
order.

### Measurement first: what the category records

<p class="concept-video"><strong>Stanford explanation (review):</strong>
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk">3.5 Extensions of the Linear Model</a>.</p>

The categories describe the people who presented on screen: all men, all
women, or mixed. They do not record the share of women in a team, and they
do not describe the full team behind the venture. Seven records are
classified as `unknown`. They are left out of every model in this lecture.
The section on the `unknown` records further down explains why. Keep these
limits in mind for every result in this chapter.

### A two-level indicator

::: {.definition-box}
<span class="label">Definition</span>

**Dummy variable.** A variable that takes only the values 0 and 1. It is 1
when an observation belongs to a category and 0 when it does not. Another
name is indicator.

**Reference category.** The category for which all dummies are 0. Every
dummy coefficient is a comparison with this category. Another name is
reference group.
:::

For a dummy variable $D_i\in\{0,1\}$,

$$
E[Y_i\mid D_i]=\beta_0+\beta_1D_i.
$$

Then

$$
E[Y_i\mid D_i=0]=\beta_0,
\qquad
E[Y_i\mid D_i=1]=\beta_0+\beta_1.
$$

So $\beta_1$ is the expected difference between the group with $D=1$ and
the reference group with $D=0$. A regression on one dummy is the same thing
as a comparison of two group means.

This is also how an A/B test is analysed. In an A/B test you show two
versions of a page, a price, or an offer to customers, and chance decides who
sees which version. Code the version as a dummy and read its coefficient.
Because chance decided who saw which version, nothing else differs between
the two groups on average. In that case, and only in that case, the
coefficient can be read as the effect of the change.

::: {#exm-i05-two-ticks}
## Worked example: two group means from two coefficients

Restrict to the all-men and all-women teams (1,158 records) and let
`women_team` equal 1 for all-women teams:

```{r}
#| label: i05-two-level
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
two_groups <- subset(known_gender,
                     pitcher_gender_group %in% c("all_men", "all_women"))
two_groups$women_team <- as.integer(
  two_groups$pitcher_gender_group == "all_women"
)
coef(lm(equity_offered_pct ~ women_team, data = two_groups))
tapply(two_groups$equity_offered_pct, two_groups$women_team, mean)
```

Read the two coefficients as two marks on the equity axis
(@fig-i05-two-ticks). The intercept $\hat\beta_0\approx 13.20$ is the mean
of the all-men teams. The sum $\hat\beta_0+\hat\beta_1\approx 13.20+0.82=14.02$
is the mean of the all-women teams. The coefficient $\hat\beta_1\approx 0.82$
is the gap between the two marks.
:::

```{r}
#| label: fig-i05-two-ticks
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-height: 2.2
#| fig-cap: "The model with one dummy places the two group means as two marks on the axis of the equity offered. The intercept is the mean of the reference group (all-men teams), and the dummy coefficient is the distance between the marks."
old_par <- par(no.readonly = TRUE)
par(mar = c(3.5, 1, 1, 1))
plot(NA, xlim = c(12.6, 14.6), ylim = c(0, 1), axes = FALSE,
     xlab = "", ylab = "")
axis(1, at = seq(12.6, 14.6, by = 0.4))
mtext("Mean initial equity offered (%)", side = 1, line = 2.4, cex = 0.9)
segments(13.20, 0.15, 13.20, 0.55, col = "#0072B2", lwd = 4)
segments(14.02, 0.15, 14.02, 0.55, col = "#D55E00", lwd = 4)
arrows(13.20, 0.38, 14.02, 0.38, length = 0.08, code = 3, col = "#17324D")
text(13.20, 0.72, expression(hat(beta)[0] == 13.20), col = "#0072B2")
text(14.02, 0.72, expression(hat(beta)[0] + hat(beta)[1] == 14.02), col = "#D55E00")
text(13.61, 0.25, expression(hat(beta)[1] == 0.82), col = "#17324D", cex = 0.9)
par(old_par)
```

### More than two categories

With $G$ categories and an intercept, use $G-1$ dummies. With all $G$
dummies and an intercept, the dummies sum to 1 in every row. This is the
exact linear dependence from Lecture 4, and the coefficients cannot be
estimated. The mistake is called the **dummy-variable trap**.

R creates the dummies itself when a predictor is a factor. Choose the
reference category yourself and check it. This is the code of the R warm-up:

```{r}
#| label: i05-reference
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
levels(known_gender$gender)
contrasts(known_gender$gender)
```

Each row of the contrasts table is a category. Each column is one dummy. The
reference category `all_men` is the row with only zeros. Its mean is carried
by the intercept. This answers the first question from before class: three
categories need two dummies.

::: {.callout-note collapse="true"}
## Check: four industries and an intercept, how many dummies?

Three. One industry is the reference group, carried by the intercept. Each
of the three dummy coefficients compares one of the other industries with
it. Four dummies and an intercept would be the dummy-variable trap.
:::

::: {.callout-caution collapse="true"}
## Try it: three sales channels

Try this with your neighbour first. In an invented example, a venture sells
through three channels: the campus stand, the app, and the website. A model
of the order value in euros uses the campus stand as the reference and
reports `channelApp 6.0` and `channelWebsite 2.5`, with an intercept of 9.0.

*Step 1 (given).* With three categories and an intercept, the model needs
two dummies. The campus stand has no dummy. Its mean is the intercept.

*Step 2 (you).* Give the fitted mean order value for each of the three
channels.

*Step 3 (you).* By how much does the fitted order value in the app differ
from the website? Neither is the reference group.

::: {.callout-note collapse="true"}
## Check your answer

Campus stand: 9.0 euros. App: 9.0 + 6.0 = 15.0 euros. Website: 9.0 + 2.5 =
11.5 euros. App minus website: 6.0 - 2.5 = 3.5 euros. Two coefficients that
share a reference group can be subtracted from each other.
:::
:::

### Percent versus percentage points

One point about units comes first. The outcome `equity_offered_pct` is
measured in percent, so its regression coefficients are in **percentage
points**. The next section estimates a coefficient of 1.28 for all-women
teams. The sentence "the all-women coefficient is 1.28%" would be ambiguous.
It could mean:

- a relative change: 1.28% of the all-men level more (13.0% would become
  about 13.2%)
- an absolute change of 1.28 percentage points (13.0% would become 14.28%).

The two readings differ by a factor of about eight here. A coefficient for
an outcome in percent is always the absolute kind. The precise sentence is:
"At the same season and requested amount, all-women presenting teams are
fitted to offer 1.28 equity percentage points more than all-men presenting
teams." Always write percentage points for this outcome.

### The adjusted comparison

Now fit the model of the driving question. It contains both category
dummies, and it holds season and requested amount (in thousands of dollars,
`ask_1k`) fixed.

```{r}
#| label: i05-fit
gender_model <- lm(
  equity_offered_pct ~ season + ask_1k + gender,
  data = known_gender
)
summary(gender_model)
```

Interpret both dummy coefficients with a full sentence. Each sentence names
the two categories, the unit, the predictors that are held fixed, and the
records it is about:

- **All-women versus all-men.** Among the 1,434 pitches with a known team
  classification, all-women presenting teams are fitted to offer about 1.28
  equity percentage points more than all-men presenting teams at the same
  season and requested amount (p $\approx$ 0.008).
- **Mixed versus all-men.** Among the same pitches, mixed teams are fitted
  to offer about 0.12 percentage points more than all-men teams at the same
  season and requested amount (p $\approx$ 0.823). This estimate is not
  statistically distinguishable from zero.

The first sentence answers the Wooclap question from before class: all-women
teams offer more. The adjusted gap between all-women and all-men teams (1.28
points) is larger than the raw gap from the worked example (0.82 points). Holding season and
requested amount fixed changed the comparison. This is why the sentence must
name what is held fixed. Both coefficients are comparisons with the
reference group. They do not show that the gender of a team causes the
offer.

Changing the reference group changes the names and values of the
coefficients. It does not change the fitted values or the overall fit. Check
it:

```{r}
#| label: i05-change-reference
known_gender$gender_women_ref <- relevel(known_gender$gender,
                                         ref = "all_women")
coef(lm(equity_offered_pct ~ season + ask_1k + gender_women_ref,
        data = known_gender))
```

The same three fitted group means are now described as comparisons with the
all-women teams. The all-men coefficient is $-1.28$, the same gap as before
with the opposite sign, and the fitted values are identical.

### Are the category terms worth having at all?

So far the coefficients were read one at a time. The driving question also
asks something that no single t-test answers: do the two category terms
together add information beyond season and requested amount?

::: {.definition-box}
<span class="label">Definition</span>

**Partial F-test.** A test of whether several coefficients are all zero at
the same time. It compares a complete model with a reduced model that leaves
out the tested terms. Both models are fitted on the same observations.
:::

Two such models are called **nested**: the reduced model is the complete
model with the tested coefficients set to zero. @fig-i05-nested shows the
pair for the two category dummies.

```{r}
#| label: fig-i05-nested
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-height: 2.6
#| fig-cap: "The two nested models of the partial F-test. The reduced model is the complete model with the two category coefficients set to zero, fitted on the same observations."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.3, 0.3, 0.3, 0.3))
plot(NA, xlim = c(0, 10), ylim = c(0, 5), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")
rect(0.4, 1.1, 4.6, 3.9, col = "#DCEAF3", border = "#17324D", lwd = 2)
rect(5.4, 1.1, 9.6, 3.9, col = "#FDE5C1", border = "#17324D", lwd = 2)
text(2.5, 3.35, "Reduced model", font = 2, col = "#17324D")
text(2.5, 2.55, "season + ask_1k", cex = 0.95)
text(2.5, 1.85, expression(beta[aw] == 0 ~ "," ~ beta[mx] == 0),
     cex = 0.95, col = "#5B6976")
text(7.5, 3.35, "Complete model", font = 2, col = "#17324D")
text(7.5, 2.55, "season + ask_1k", cex = 0.95)
text(7.5, 1.85, "+ two category dummies", cex = 0.95)
arrows(4.7, 2.5, 5.3, 2.5, length = 0.1, lwd = 2, col = "#17324D")
text(5.0, 3.0, "free the\ntwo coefficients", cex = 0.8, col = "#5B6976")
par(old_par)
```

The hypotheses:

$$
H_0:\beta_{all\_women}=\beta_{mixed}=0
\quad\text{against at least one non-zero.}
$$

::: {#exm-i05-partial-f}
## Worked example: the partial F statistic by hand

Fit both models and ask R for the anova table:

```{r}
#| label: i05-partial-f
reduced_gender <- lm(equity_offered_pct ~ season + ask_1k,
                     data = known_gender)
anova(reduced_gender, gender_model)
```

Now calculate F by hand from the numbers in the table. The residual sum of
squares (RSS in the R output, SSE in this course) falls by $436.25$ when the
two category terms are added. This uses $2$ degrees of freedom. The complete
model leaves a residual sum of squares of $82{,}792$ on $1{,}429$ degrees of
freedom:

$$
F=\frac{436.25/2}{82{,}792/1{,}429}
=\frac{218.13}{57.94}
\approx 3.76.
$$

Read the ratio. The numerator is the variation that the two terms explain,
per tested coefficient. The denominator is the variation that is left
unexplained, per residual degree of freedom. If the two coefficients were
really zero, the ratio would be close to 1. Here it is about 3.8, and the
p-value is about 0.023.
:::

At 5%, the two category terms are **jointly useful** in this model. Compare
this with the single results: the p-value of the joint test is $0.023$, and
the p-value of the mixed dummy alone is $0.823$. Joint usefulness is not the
same as individual significance. The F-test asks whether the two terms
together improve the model. It does not ask whether each term is significant
on its own. The overall F-test in `summary(gender_model)`, about 84.5 with a
very small p-value, answers a third question: whether all four coefficients
together improve on a model with only an intercept.

### Visualise the adjusted comparison

@fig-i05-adjusted shows the fitted equity offered for the three groups of
teams at the same season and requested amount.

```{r}
#| label: fig-i05-adjusted
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted equity offered for the three groups of presenting teams at the average season and requested amount, with 95% confidence intervals. The all-women point lies above the other two. The intervals show sampling uncertainty under the model, not uncertainty about how the categories were recorded."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
gender_model <- lm(
  equity_offered_pct ~ season + ask_1k + gender,
  data = known_gender
)
newdat <- data.frame(
  season = mean(known_gender$season),
  ask_1k = mean(known_gender$ask_1k),
  gender = factor(c("all_men", "all_women", "mixed"),
                  levels = levels(known_gender$gender))
)
pred <- predict(gender_model, newdata = newdat, interval = "confidence")
adj_df <- data.frame(
  team = factor(c("All-men", "All-women", "Mixed"),
                levels = c("All-men", "All-women", "Mixed")),
  fit = pred[, "fit"],
  lwr = pred[, "lwr"],
  upr = pred[, "upr"]
)

ggplot(adj_df, aes(x = team, y = fit)) +
  geom_pointrange(aes(ymin = lwr, ymax = upr),
                  colour = "#009E73", linewidth = 0.7, fatten = 2.6) +
  labs(x = "Source classification of pitching team",
       y = "Fitted initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-i05-adjusted with the
[five steps for reading a figure](../basic/overview.qmd#how-to-read-every-graph):

1. **Unit and variables:** the horizontal axis shows the three categories of
   teams. The vertical axis shows the fitted equity offered, in percent.
2. **Scales:** each point is a fitted mean at the same average season and
   requested amount. This is an adjusted comparison, not a comparison of raw
   group means.
3. **One element:** the points are fitted values (about 12.9, 14.1, and
   13.0). The vertical lines are 95% confidence intervals. They are wider
   for the smaller groups.
4. **Pattern:** the all-women point lies about 1.28 points above the all-men
   point. The mixed interval overlaps both.
5. **Limits:** the intervals show sampling uncertainty under the model. They
   say nothing about how well the broad categories were recorded, and
   nothing about causes.

### Why `unknown` is excluded, not a fourth group

Seven records have the classification `unknown`. This label says that
information is missing. It is not a kind of team. A fourth category would
estimate the mean equity of teams that the source could not classify, a
number with no business meaning. Leaving these records out keeps every
coefficient interpretable. The cost is that the results describe the 1,434
classified records only.

::: {.caveat-box}
<span class="label">What these category coefficients can and cannot say</span>

- The categories are a broad description of the people who presented. They
  do not record shares of women, and they do not describe the full founding
  team.
- Each coefficient compares one group with the reference group, at the same
  season and requested amount.
- The model says that offers differ between the coded groups. It does not say
  that the composition of a team makes anyone offer differently.

**What you can do with it:** for groups that you can assign yourself, such as
two versions of a page, run an A/B test. Then the group coefficient is the
effect.
:::

### Answering the driving question

At the same season and requested amount, all-women presenting teams are
fitted to offer about 1.28 equity percentage points more than all-men teams
(p $\approx$ 0.008). That is about one tenth of the average equity offered by
all-men teams (13.20%). Mixed teams cannot be distinguished from all-men
teams. The two category terms are jointly useful (partial F $\approx$ 3.76,
p $\approx$ 0.023), so they belong in the model. The result is an adjusted
association among these recorded pitches, not an effect of gender.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You test two checkout pages. Each new customer sees the old or the new page by chance, 100 customers each. You regress the order value in euros on a dummy for the new page. The coefficient is 1.20 with a standard error of 0.50, so the 95% interval runs from about 0.20 to 2.20 euros. Chance decided who saw which page, so the two groups differ in nothing else on average. Here the coefficient is the effect of the page. The decision: switch to the new page. This is an A/B test, the founder's own way to get from an association to an effect.
:::

**Next:** You can now compare categories at fixed values of the other
predictors. So far every group has the same slope, and every trend is a
straight line. Both are choices of the model, not facts. Lecture 6 relaxes
them. An interaction lets the season trend differ between groups of teams,
and a squared term lets the trend bend.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 80 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Why are only $G-1$ dummies used with an intercept?
2. What does a dummy coefficient compare?
3. What changes when the reference group changes?
4. What does a partial F-test evaluate?
5. Why can a joint F-test reject when one individual t-test does not?
6. How is an A/B test analysed with a regression, and why can its
   coefficient be read as an effect?

::: {.callout-note collapse="true"}
## Check your answers

1. All $G$ dummies sum to 1 in every row, which is the same as the intercept
   column. This is an exact linear dependence, so the coefficients could not
   be estimated.
2. A category with the named reference category, holding the other included
   predictors fixed.
3. The names and values of the coefficients change, because they now compare
   with a different category. The fitted values and the overall fit do not
   change.
4. Whether several coefficients are all zero at the same time. It compares
   two nested models on the same observations.
5. The F-test evaluates the terms together. Each t-test asks a narrower
   question about one comparison with the reference group, and it can miss
   what the terms explain together.
6. Code the version (A or B) as a dummy and regress the outcome on it. The
   coefficient is the difference between the two group means. Chance decided
   who saw which version, so the groups differ in nothing else on average,
   and the difference is the effect of the version.
:::

### Releveling and verification practice (15 minutes)

1. Relevel `gender` so that `mixed` is the reference group, and interpret
   both dummy coefficients. Save the relevelled model as `relevelled_model`.
2. Check that the fitted values are unchanged after releveling by comparing
   them with those of `gender_model`.
3. State the hypotheses for the partial F-test of both gender dummies.
4. Explain why the seven `unknown` records were left out and not treated as
   a fourth group.
5. Explain why "the all-women coefficient is 1.28%" is ambiguous, then
   rewrite it precisely.

::: {.callout-note collapse="true"}
## Check your answers

1. Use:

   ```r
   mixed_reference <- relevel(known_gender$gender, ref = "mixed")
   relevelled_model <- lm(
     equity_offered_pct ~ season + ask_1k + mixed_reference,
     data = known_gender
   )
   coef(relevelled_model)
   ```

   At the same season and requested amount, the all-men coefficient is
   about $-0.121$ percentage points and the all-women coefficient about
   $1.163$ percentage points. Both compare with mixed teams.
2. `all.equal(fitted(gender_model), fitted(relevelled_model))` returns
   `TRUE`.
3. $H_0:\beta_{all\_women}=\beta_{mixed}=0$ against the alternative that at
   least one of the two is not zero, with all-men as the reference group.
4. The `unknown` records have no known classification. Missing information
   is not a kind of team.
5. It confuses percent with percentage points, and it does not name the
   comparison group or what is held fixed. A precise version: "At the same
   season and requested amount, all-women presenting teams are fitted to
   offer 1.28 equity percentage points more than all-men presenting teams."
:::

### Finish a half-worked example: a partial F with new numbers (10 minutes)

Test the two category dummies again, now in a smaller complete model. The
reduced model contains only `season`:

```{r}
#| label: i05-faded-anova
reduced_season <- lm(equity_offered_pct ~ season, data = known_gender)
full_season    <- lm(equity_offered_pct ~ season + gender,
                     data = known_gender)
anova(reduced_season, full_season)
```

*Step 1 (given).* Adding the two dummies reduces the residual sum of squares
by $643.43$ using $2$ degrees of freedom. The complete model leaves
$84{,}003$ on $1{,}430$ degrees of freedom.

*Step 2 (you).* Plug these into the F formula from the worked example.

*Step 3 (you).* Compare with the table's F and p-value and state the
conclusion at 5% in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$F=(643.43/2)/(84{,}003/1{,}430)=321.7/58.7\approx 5.48$, matching the
table, with p $\approx$ 0.004. At 5%, the two category terms are jointly
useful even in the model that adjusts only for season.
:::

### The 120-word interpretation (10 minutes)

Write an interpretation of the full `gender_model` in about 120 words.
Include what the team categories do and do not record, and say why the
result is not causal.

::: {.callout-note collapse="true"}
## Check a model answer

A complete answer reports five things. First, the outcome and the sample:
the equity offered, among the 1,434 broadcast pitches with a known team
classification. Second, both estimates with units and with what is held
fixed: all-women teams are fitted 1.28 percentage points above all-men
teams and mixed teams 0.12 points above, at the same season and requested
amount. Third, the uncertainty: p $\approx$ 0.008 and 0.823 for the two
coefficients, and p $\approx$ 0.023 for the partial F-test, so the terms are
jointly useful. Fourth, the limits of the measurement: the categories
describe the people who presented, not shares of women or the full team.
Fifth, the scope: these are adjusted associations among the recorded
pitches, not causal effects.
:::

### Review: one question from Lecture 4 (5 minutes)

Using the Lecture 4 three-predictor model, compute the combined fitted
difference for a pitch three seasons later requesting $200,000 more, at the
same description length.

::: {.callout-note collapse="true"}
## Check your answer

$3(-0.836)+200(-0.0029)\approx-3.09$ percentage points. The two
contributions add because the model is additive, and \$200,000 is 200 units
of `ask_1k`.
:::

### Worked exam interpretation (5 minutes)

**Question.** This question is in the style of Question 4 of the mock exam.
A model of equity offered on season, `ask_1k`, and industry (four
categories, reference `Food and Beverage`) reports
`industryTechnology 1.9 (p = 0.04)`. The numbers are invented. A student
writes: "Technology pitches offer 1.9% more equity, so entering tech makes
founders more generous." Correct the sentence.

::: {.callout-note collapse="true"}
## Check a model answer

Three corrections. Units: the estimate is 1.9 equity percentage points, not
a relative 1.9%. Comparison and what is held fixed: technology pitches are
fitted 1.9 points above Food and Beverage pitches at the same season and
requested amount, among the recorded pitches. Scope: nothing here is causal.
"Entering tech makes founders more generous" reads a comparison with the
reference group as an effect of choosing an industry.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "The coefficient of the all-women dummy is 1.28. What does that mean?" It answered:

> "All-women teams offer 1.28% more equity than the average team."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Three errors. The unit is percentage points, not percent. The comparison is with the reference group, all-men teams, not with the average team. And it holds at the same season and requested amount. Correct: at the same season and requested amount, all-women teams are fitted to offer 1.28 percentage points more equity than all-men teams.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture, or any business you know well. Name one
categorical feature of your customers or products (segment, channel, region,
plan tier). Then write two sentences. Which category would you set as the
reference group, and why? What would a "jointly useful" result of the F-test
for the whole set of categories change in how you describe your market?
There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus (an invented example). "My channels are the campus stand, the app, and the website. I would set the campus stand as the reference, because it is the first channel and every decision is compared with it. A jointly useful result of the F-test would justify describing my market by channel at all. Mixed p-values of the single dummies would not settle that."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant. Ask it to check whether your two
sentences confuse percent with percentage points, or to ask you questions
about the dummy-variable trap with your own categories.
::::

[Continue to Lecture 6: Interactions and curvature](i06-interactions-curvature.qmd).

[Code sheet: Lecture 5 commands and formulas](code-sheet.qmd#lecture-5-factors-reference-categories-and-partial-f-tests).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Sections 3.2.2 and 3.3.1, and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 20--21.

:::
