# Lecture 6: Interactions and curvature {#i06}

```{r}
#| label: i06-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 6, you can</span>

- read a two-lines picture as four numbers: two intercepts and two slopes;
- explain an interaction coefficient as a difference in slopes, with units;
- interpret lower-order terms in a model containing an interaction, and use
  centering to make them meaningful;
- report a non-significant interaction without overclaiming in either
  direction;
- apply the hierarchy principle and the `x * z` formula shorthand;
- fit a quadratic term, explain why it is still linear regression, and
  compute a fitted change without reading coefficients separately;
- test added terms with a partial F-test and refuse to extrapolate a
  polynomial.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">"On average" can hide the interesting part: an effect may differ across groups, or bend as scale grows. Interactions and curvature are how a model is allowed to say so.</p>

Is the downward season trend in equity offers the same for every team
composition --- and is the trend even a straight line?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=13">3.5
Extensions of the Linear Model</a> and
<a href="https://www.youtube.com/watch?v=F-D3lZzYn50&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=48">7.1
Polynomials and Step Functions</a>. Focus on why an interaction changes a
slope and why a polynomial remains a linear regression model in its
coefficients. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Derive the two lines (5 minutes).** In the model
$Y=\beta_0+\beta_1X+\beta_2D+\beta_3XD+\varepsilon$ with $D\in\{0,1\}$,
write the fitted line for $D=0$ and for $D=1$.

::: {.callout-note collapse="true"}
## Check your preparation

For $D=0$: $E[Y\mid X,D=0]=\beta_0+\beta_1X$. For $D=1$:
$E[Y\mid X,D=1]=(\beta_0+\beta_2)+(\beta_1+\beta_3)X$. Setting $D=1$ shifts
the intercept by $\beta_2$ and the slope by $\beta_3$.
:::

**Predict (2 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: how many separate numbers do you need to
fully describe a picture of **two straight lines with different slopes**?
Commit to an answer; the lecture resolves it.

**R warm-up (8 minutes).** Build the analysis variables --- the same
`known_gender` subset as Lecture 5, plus a `women` indicator and a centred
season --- fit one equity-on-season line per group, and bring the plot:

```{r}
#| label: i06-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

plot(equity_offered_pct ~ season, data = known_gender,
     pch = 16, col = "grey80")
abline(lm(equity_offered_pct ~ season,
          data = subset(known_gender, women == 0)),
       col = "#0072B2", lwd = 3)
abline(lm(equity_offered_pct ~ season,
          data = subset(known_gender, women == 1)),
       col = "#D55E00", lwd = 3)
```

If products, powers, or centering are not secure, review
[Mathematics Refresher: products, powers, and centering](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question --- with the class's own warm-up two-line
plots on screen.

<p class="concept-video"><strong>Stanford explanations (review):</strong>
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk">3.5 Extensions of the Linear Model</a> and
<a href="https://www.youtube.com/watch?v=F-D3lZzYn50">7.1 Polynomials and Step Functions</a>.</p>

### The picture before the formula

Every model so far forced all groups to share one slope. Lecture 5 let the
*heights* differ by group; today the **slopes** get the same freedom. Start
with the picture, not the algebra:

```{r}
#| label: fig-i06-two-lines
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-height: 3.8
#| fig-cap: "Two groups, two lines, different slopes. Before any formula: how many numbers does it take to describe this picture completely?"
old_par <- par(no.readonly = TRUE)
par(mar = c(4.2, 4.2, 1, 1), las = 1)
plot(NA, xlim = c(0, 10), ylim = c(0, 10), bty = "n",
     xlab = "Predictor X", ylab = "Outcome Y")
abline(a = 2, b = 0.75, col = "#009E73", lwd = 3)
abline(a = 4.5, b = 0.25, col = "#D55E00", lwd = 3, lty = 2)
text(8.4, 2 + 0.75 * 8.4 + 0.9, "Group D = 0", col = "#009E73")
text(8.4, 4.5 + 0.25 * 8.4 - 0.8, "Group D = 1", col = "#D55E00")
par(old_par)
```

This resolves the prediction from your prep: **four numbers** --- an
intercept and a slope for each line. The interaction model is nothing more
than a reparameterisation of those four numbers:

$$
E[Y\mid X,D]
=\beta_0+\beta_1X+\beta_2D+\beta_3XD.
$$

Map each coefficient to a feature of the picture:

- $\beta_0$: where the $D=0$ line meets the vertical axis;
- $\beta_1$: the slope of the $D=0$ line;
- $\beta_2$: how much *higher* the $D=1$ line starts;
- $\beta_3$: how much *steeper* the $D=1$ line runs.

Or as the two-row table that your prep derivation produced:

| Group | Intercept | Slope |
|---|---|---|
| $D=0$ | $\beta_0$ | $\beta_1$ |
| $D=1$ | $\beta_0+\beta_2$ | $\beta_1+\beta_3$ |

Four coefficients, four picture features. When $\beta_3=0$ the lines are
parallel and the model collapses to Lecture 5's common-slope dummy model.

### The interaction coefficient is a difference in slopes

In the additive model the slope of $X$ is the same at every value of the
other predictor. Adding the product $XD$ lets it depend: the change in the
conditional mean for a one-unit increase in $X$ is

$$
\beta_1+\beta_3D,
$$

so $\beta_3$ is a **difference in slopes**, not a standalone effect. Its
units combine the variables. If $X$ is measured in seasons and $D$ is a
group indicator, $\beta_3$ is measured in *outcome units per season,
between the groups* --- here, equity percentage points per season, group
minus reference group. (The same algebra works when both variables are
quantitative: the slope of $X$ becomes $\beta_1+\beta_3Z$, a slope that
slides with $Z$.)

**Centering.** Once $XD$ is in the model, $\beta_2$ compares the groups *at
$X=0$* --- for us, the unobserved Season 0. Replacing season with

$$
season\_c = season - \overline{season}
$$

moves that comparison to the average season (about 8.8). Centering changes
exactly one kind of thing and cannot change another:

- it **changes** the meaning (and value) of lower-order terms --- the
  intercept and the group coefficient now refer to the average season;
- it **cannot change** the fitted values, the residuals, $R^2$, or the
  interaction estimate itself. The lines are the same lines; only the
  place where we read off their vertical gap has moved.

::: {.callout-note collapse="true"}
## Check: which printed numbers change when season is centred?

The intercept and the `women` coefficient (lower-order terms) change; the
interaction coefficient, the `ask_1k` coefficient, all fitted values, and
$R^2$ do not. The after-class practice has you verify this in R.
:::

### The Shark Tank interaction model

Now the driving question's first half. For this lecture, `women = 1` means
the source classifies the presenting team as all-women or mixed;
`women = 0` means all-men. It is a broader split than Lecture 5's
three-category factor, it does not measure a share of women or the full
venture team, and the seven `unknown` records are excluded as before. The
whole setup is visible --- it is the warm-up code:

```{r}
#| label: i06-interaction-fit
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

interaction_model <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
summary(interaction_model)
```

Read the two group slopes off the two-row table, holding requested amount
fixed. For all-men teams (`women = 0`), the fitted season slope is about
$-0.799$ percentage points per season. For teams with women represented, it
is

$$
-0.799+(-0.061)=-0.861
\text{ percentage points per season.}
$$

The interaction estimate is $-0.061$ equity percentage points per season
(between groups), with p-value about 0.520. Read that carefully, in both
directions: this model does **not provide evidence that the season trend
differs** between these two broad source-defined groups. It also does
**not prove the slopes are identical** --- the estimate is imprecise (SE
about 0.096), and the data are consistent with anything from a moderately
steeper to a moderately flatter trend for teams with women represented.
No evidence of different trends is not proof of equal trends.

```{r}
#| label: fig-i06-interaction
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted initial equity by season for the two source-defined groups, at the same average requested amount. Each line's slope is that group's fitted change per season; the vertical separation is the fitted group difference at a given season."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)
interaction_model <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
season_grid <- 1:16
ask_bar <- mean(known_gender$ask_1k)
newdat <- expand.grid(season = season_grid, women = c(0, 1))
newdat$season_c <- newdat$season - mean(known_gender$season)
newdat$ask_1k <- ask_bar
newdat$fit <- predict(interaction_model, newdata = newdat)
newdat$group <- ifelse(newdat$women == 1, "Women represented", "All-men")
fit_men_13 <- newdat$fit[newdat$women == 0 & newdat$season == 13]
fit_women_13 <- newdat$fit[newdat$women == 1 & newdat$season == 13]

ggplot(newdat, aes(x = season, y = fit,
                   colour = group, linetype = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(data = subset(newdat, season %in% c(1, 16)), size = 2.2) +
  annotate("text", x = 13, y = fit_women_13 + 1.1,
           label = "Women represented",
           colour = "#D55E00", size = 3.4) +
  annotate("text", x = 13, y = fit_men_13 - 1.1,
           label = "All-men",
           colour = "#009E73", size = 3.4) +
  scale_colour_manual(values = c("All-men" = "#009E73",
                                 "Women represented" = "#D55E00"),
                      guide = "none") +
  scale_linetype_manual(values = c("All-men" = "solid",
                                   "Women represented" = "dashed"),
                        guide = "none") +
  scale_x_continuous(breaks = seq(1, 16, by = 3)) +
  labs(x = "Season", y = "Fitted initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Graph the model whenever an interaction is present: in
@fig-i06-interaction the two nearly-parallel lines *are* the finding, and
neither $\beta_1$ nor $\beta_2$ alone is an overall effect once $XD$ is in
the model.

### The hierarchy principle and the `*` shorthand

If a model contains $XD$, ordinarily retain both $X$ and $D$, even when a
lower-order term is not individually significant. The lower-order terms
define the two lines whose *difference* the interaction describes; dropping
one forces the lines through an arbitrary common feature and turns the
interaction coefficient into something unintended. R's formula shorthand
respects the hierarchy automatically:

```{r}
#| label: i06-interaction-formula
#| eval: false
lm(y ~ x * z, data = df)  # x * z expands to x + z + x:z
```

Before the break, the parallel that organises the whole lecture: block 1
let the slope differ **by group**; block 2 lets the slope differ **along
$x$ itself**. Both are the same move --- feeding least squares a new
constructed column (a product, a square) and reading the fitted
combination.

### Curvature: polynomials are still linear regression

Is the season trend even a straight line? A quadratic conditional mean

$$
E[Y\mid X=x]=\beta_0+\beta_1x+\beta_2x^2
$$

is curved in $x$ but **linear in the unknown coefficients**
$\beta_0,\beta_1,\beta_2$ --- and that is all `lm()` requires. The square
is just another constructed predictor column.

What changes is how to read the output. The slope is no longer one number:
moving from $x$ to $x+1$ changes the fitted mean by

$$
(\beta_1(x+1)+\beta_2(x+1)^2) - (\beta_1x+\beta_2x^2)
=\beta_1+\beta_2(2x+1),
$$

a change that depends on where you start. The standing rule: **never read
$\beta_1$ alone while $x^2$ is in the model.** Compute fitted changes or
graph the curve.

```{r}
#| label: i06-polynomial-fit
linear_season <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
poly_season <- lm(equity_offered_pct ~ season + I(season^2) + ask_1k,
                  data = sharks)
summary(poly_season)$coefficients
```

The `I()` wrapper tells R to square arithmetically rather than read `^` as
formula syntax. The fitted coefficients are $b_1 \approx -1.713$ and
$b_2 \approx 0.051$: a steep early decline that flattens season by season.

::: {#exm-i06-quadratic-change}
## Worked example: the fitted change from Season 5 to Season 6

Holding requested amount fixed, the fitted change from season $x$ to season
$x+1$ is $b_1+b_2(2x+1)$. From Season 5 to Season 6, $x = 5$:

$$
b_1+b_2(2\times5+1)
=-1.713+0.051\times11
=-1.713+0.561
\approx-1.15
\text{ percentage points.}
$$

Reading $b_1=-1.713$ alone would overstate the drop by half a point ---
that number is the fitted change at the meaningless $x=0$, not a constant
slope. Verification in R, as a difference of two fitted values:

```{r}
#| label: i06-quadratic-check
b <- coef(poly_season)
unname(b["season"] + b["I(season^2)"] * (2 * 5 + 1))
```
:::

### Is the curvature worth having? The partial F

The quadratic model and the linear model are nested --- Lecture 5's test
applies directly to the one added term:

```{r}
#| label: i06-partial-f
anova(linear_season, poly_season)
```

The partial F-test for adding `I(season^2)` gives $F \approx 20.1$ with
p-value below 0.001 in the classical model: the curvature is statistically
detectable. Detectable is not dramatic --- adjusted $R^2$ moves from about
0.186 to 0.196 --- so also assess the substantive shape:

```{r}
#| label: fig-i06-polynomial
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Raw season means of initial equity offered (points) and the fitted quadratic at the average requested amount (curve). The curve flattens towards the last observed seasons; beyond Season 16 (shaded) it would turn upward, a region with no data support."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
poly_season <- lm(equity_offered_pct ~ season + I(season^2) + ask_1k,
                  data = sharks)
season_means <- aggregate(equity_offered_pct ~ season, data = sharks, mean)
curve_data <- data.frame(season = seq(1, 16, length.out = 200),
                         ask_1k = mean(sharks$ask_1k))
curve_data$fit <- predict(poly_season, newdata = curve_data)

ggplot(season_means, aes(x = season, y = equity_offered_pct)) +
  annotate("rect", xmin = 16, xmax = 17.6, ymin = -Inf, ymax = Inf,
           fill = "grey75", alpha = 0.25) +
  geom_vline(xintercept = 16, colour = "grey60",
             linetype = "dashed", linewidth = 0.8) +
  geom_line(data = curve_data, aes(x = season, y = fit),
            colour = "#D55E00", linewidth = 1.1) +
  geom_point(size = 2.6, colour = "#009E73") +
  annotate("text", x = 16.8, y = max(season_means$equity_offered_pct),
           label = "no data beyond\nseason 16", colour = "#5B6976",
           size = 3, lineheight = 0.95) +
  scale_x_continuous(breaks = seq(1, 16, by = 3)) +
  coord_cartesian(xlim = c(1, 17.6)) +
  labs(x = "Season", y = "Initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The points in @fig-i06-polynomial are raw season means; the curve holds
requested amount at its sample mean. **Do not extrapolate the quadratic
beyond Seasons 1--16.** The fitted parabola bottoms out just past Season 16
and would climb steeply afterwards --- an artefact of the functional form,
not a forecast. Polynomials can turn sharply the moment they leave the
data.

::: {.caveat-box}
<span class="label">What flexible terms can and cannot repair</span>

Interactions and polynomials make the conditional mean more flexible ---
and that is *all* they do. They repair nothing about the design: dependence
across rows, selection into broadcast, measurement error, unequal error
variance, and unmeasured confounding all pass through untouched. Nor is
flexibility free: every added term spends precision and invites overfitting
if terms are added because they are significant rather than because a
substantive question asks for them. Choose terms from the question ---
"could this trend differ by group?", "could this relationship bend?" ---
inspect the implied predictions, and report combined effects, never single
coefficients ripped out of a product or a polynomial.
:::

### Answering the driving question

Two halves, two verdicts, both with limits. **Same trend for every
composition?** The estimated slope difference is $-0.061$ percentage points
per season (p $\approx$ 0.52): no detectable difference between the
source-defined groups --- and no proof of equality, since the estimate is
imprecise. **A straight line?** Not quite: the quadratic term is
detectable (partial $F \approx 20.1$, p $<$ 0.001), and the fitted decline
flattens from about $-1.56$ points per season early on to about $-0.64$
around Season 10 --- a mild, data-supported curvature valid only inside
Seasons 1--16, in a model that remains observational throughout.

**Next:** Tutorial 2 rehearses Lectures 4--6 in exam format and closes with
the integrated modelling task: one full analysis from research question to a
150-word conclusion --- outcome, predictors, an extension a real question
motivates, and the caveat that headlines it. That is a rehearsal of exactly
what the exam, and later your thesis, will ask of you.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 65 minutes.</span>

### Retrieval (10 minutes)

1. What does an interaction coefficient measure, and what are its units in
   the season-by-group model?
2. Why centre a predictor before interacting it, and what can centering
   never change?
3. State the hierarchy principle.
4. Why is a quadratic model still a linear regression model?
5. Why should a polynomial be graphed, and never extrapolated casually?

::: {.callout-note collapse="true"}
## Check your answers

1. How one predictor's slope differs with the other predictor --- here,
   equity percentage points per season, as a difference between groups.
2. To make lower-order terms refer to a meaningful value such as the sample
   mean; it never changes fitted values, residuals, $R^2$, or the
   interaction estimate.
3. Retain the component lower-order terms whenever their interaction or
   higher-order term is in the model.
4. It is linear in the unknown coefficients; the squared column is just
   another constructed predictor.
5. No single coefficient is the slope, so only fitted values or the curve
   show the model's claim --- and outside the observed range the curve can
   turn sharply with no data support.
:::

### Centering verification (10 minutes)

Fit the interaction model with and without centering season. Verify that
the fitted values and the interaction coefficient are unchanged, and report
which coefficients do change.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: i06-centering-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

uncentered <- lm(equity_offered_pct ~ season * women + ask_1k,
                 data = known_gender)
centered <- lm(equity_offered_pct ~ season_c * women + ask_1k,
               data = known_gender)
all.equal(fitted(uncentered), fitted(centered))
c(uncentered = unname(coef(uncentered)["season:women"]),
  centered = unname(coef(centered)["season_c:women"]))
c(women_uncentered = unname(coef(uncentered)["women"]),
  women_centered = unname(coef(centered)["women"]))
```

The fitted values agree and both interaction estimates are about $-0.061$.
The lower-order `women` coefficient changes from about $1.35$ (a comparison
at the unobserved Season 0) to about $0.81$ (a comparison at the average
season, about 8.8); the intercept changes correspondingly. Centering moved
the reading point, not the lines.
:::

### From worked to faded: a quadratic change (15 minutes)

Re-read the worked Season 5 to 6 calculation
(@exm-i06-quadratic-change). Then compute the fitted change from **Season
10 to Season 11** in the same model.

*Step 1 (given).* The recipe is $b_1+b_2(2x+1)$ with $x=10$, so the
multiplier on $b_2$ is $21$.

*Step 2 (you).* Plug in $b_1\approx-1.713$ and $b_2\approx0.051$.

*Step 3 (you).* Compare the result with the Season 5 to 6 change and state
in one sentence what the comparison shows about the trend.

::: {.callout-note collapse="true"}
## Check your answer

$-1.713+0.051\times21\approx-1.713+1.072\approx-0.64$ percentage points
(exactly: $b_1+21b_2=-0.642$). The drop per season has shrunk from about
$-1.15$ (Seasons 5--6) to about $-0.64$ (Seasons 10--11): the decline
flattens in later seasons, which is precisely what a positive $b_2$ says.
:::

Then unassisted, with new coefficients. A model of monthly revenue
(€1,000s) on shop age $x$ (years) reports
$\widehat Y = 40 - 2.00x + 0.060x^2$.

1. Compute the fitted change from age 3 to age 4 and from age 8 to age 9.
2. A colleague reports "revenue falls €2,000 per year of age." Correct the
   sentence.
3. In $Y=10+2X+3D-0.5XD$, write both group lines and interpret the
   interaction coefficient.

::: {.callout-note collapse="true"}
## Check your answers

1. Age 3 to 4: $-2.00+0.060(2\times3+1)=-2.00+0.42=-1.58$, about
   $-€1{,}580$. Age 8 to 9: $-2.00+0.060\times17=-0.98$, about $-€980$.
2. $-2.00$ is the fitted change at age 0 only. With $x^2$ in the model the
   change depends on age: about $-1.58$ at age 3, about $-0.98$ at age 8
   --- the decline weakens as shops age.
3. For $D=0$: $\widehat Y=10+2X$. For $D=1$: $\widehat Y=13+1.5X$. The
   interaction $-0.5$ says the $D=1$ slope is 0.5 outcome units per unit of
   $X$ smaller than the $D=0$ slope.
:::

### Cumulative mini-task (15 minutes)

In Lecture 4 you wrote a paragraph connecting the Lecture 2 diagnostics to
the three-predictor equity model. Extend that paragraph by one argument:
propose **either** an interaction **or** a polynomial term for that model,
state the substantive question that motivates it, and say how you would
test whether the term earns its place.

::: {.callout-note collapse="true"}
## Check a model answer

A defensible extension: "The season decline need not be linear ---
early-seasons novelty and later-seasons professionalisation suggest a
flattening trend, so I would add `I(season^2)`. The term is motivated by
the shape question, not by hunting significance. I would test it with a
partial F-test against the nested linear model on the same rows (here
$F\approx20.1$, p $<$ 0.001), check that adjusted $R^2$ improves, graph the
fitted curve inside Seasons 1--16, and report fitted season-to-season
changes rather than the raw `season` coefficient." An interaction argument
(for example season by gender group, testing whether the trend differs;
$F\approx0.41$, p $\approx$ 0.52, so the term is not retained) is equally
valid if it names the motivating question and the nested test.
:::

### Spaced review: one partial-F item from Lecture 5 (5 minutes)

Adding the two gender dummies to the season-and-ask model of equity reduced
the residual sum of squares by 436.25 using 2 degrees of freedom, and the
complete model left 82,792 on 1,429 degrees of freedom. Compute the partial
F statistic and state the 5% conclusion.

::: {.callout-note collapse="true"}
## Check your answer

$F=(436.25/2)/(82{,}792/1{,}429)=218.1/57.9\approx3.76$, with
p $\approx$ 0.023. At 5%, the two category terms are jointly useful in that
specification --- even though the mixed dummy's own t-test is far from
significant, which is the standing lesson that joint and individual tests
answer different questions.
:::

### Worked exam interpretation (5 minutes)

**Question** (mock exam Question 5 pattern). A model of venture sales
growth (percentage points) on marketing spending (€10,000s), an
online-only indicator, and their interaction reports
`spending 1.20`, `online 2.00`, `spending:online -0.50`. A student writes:
"Spending raises growth by 1.2 percentage points per €10,000 for all
ventures." Correct the sentence.

::: {.callout-note collapse="true"}
## Check a model answer

Two repairs. First, with the interaction present, 1.20 is the fitted
spending slope for **offline** ventures only (`online = 0`); for
online-only ventures the slope is $1.20-0.50=0.70$ percentage points per
€10,000 --- there is no single all-venture slope. Second, "raises" claims
causation; these are fitted associations among sampled ventures, holding
the other included terms fixed, with no design that identifies an effect of
spending.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well. Name one
relationship in its data whose *slope* plausibly depends on something else
(price sensitivity differing by segment, campaign effect differing by
channel, growth flattening with scale), and write two sentences: first
state the interaction or quadratic term you would add and the substantive
question behind it, then say what a non-significant result would --- and
would not --- tell you. There is no wrong answer; the point is letting the
question, not the software, choose the term.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I would add an ad-spend × exam-week interaction, because the substantive question is whether ads only work when students are actually on campus. A non-significant interaction would mean no evidence the slopes differ --- not proof that exam weeks make no difference."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to read your two
sentences and check whether you treated "no evidence of a difference" as
"proof of no difference."
::::

Continue to Tutorial 2: Multiple regression and extensions.

Formula sheet: Lecture 6 formulas.

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Section 3.3.2 and Sections 3.6.4--3.6.6,
and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 21.

:::
