# Lecture 6: Interactions and curvature {#i06}

```{r}
#| label: i06-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 6, you can</span>

- read a picture of two lines as four numbers: two intercepts and two slopes
- explain an interaction coefficient as a difference in slopes, with units
- interpret the other coefficients in a model with an interaction, and use
  centering to give them a useful meaning
- report an interaction that is not significant without claiming too much in
  either direction
- apply the hierarchy principle and the `x * z` formula shorthand
- fit a quadratic term, explain why the model is still a linear regression,
  and compute a fitted change from both coefficients together
- test added terms with a partial F-test, and explain why a curve must not be
  used outside the range of the data.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- write the two lines of an interaction model
- interpret an interaction as a difference in slopes
- say what centering changes and what it cannot change
- calculate a fitted change in a quadratic model with $\hat\beta_1+\hat\beta_2(2x+1)$
- explain why a curve must not be used beyond the range of the data.

**Enough to recognise:**

- the `I()` syntax, which is on the code sheet
- the name hierarchy principle.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">An average slope can hide the interesting part. A relationship may differ between groups, or it may bend as a predictor grows. Interactions and curves let a model show this.</p>

The equity offered (`equity_offered_pct`) in the Shark Tank file falls over
the seasons. Is this trend the same for every kind of presenting team, and
is it a straight line at all?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Minimum path:** the video and the R task. **Full path:** everything in this section.

Bring a charged laptop with RStudio and the course project open. The code of
the lecture is in this chapter, ready to copy, so you can run each step
yourself during the lecture and get the same numbers and figures.

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=13">3.5
Extensions of the Linear Model</a> and
<a href="https://www.youtube.com/watch?v=F-D3lZzYn50&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=48">7.1
Polynomials and Step Functions</a>. Focus on why an interaction changes a
slope, and why a model with a squared term is still a linear regression. All
links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Derive the two lines (5 minutes).** In the model
$Y=\beta_0+\beta_1X+\beta_2D+\beta_3XD+\varepsilon$ with a dummy variable
$D\in\{0,1\}$, write the line for $D=0$ and the line for $D=1$.

::: {.callout-note collapse="true"}
## Check your preparation

For $D=0$: $E[Y\mid X,D=0]=\beta_0+\beta_1X$. For $D=1$:
$E[Y\mid X,D=1]=(\beta_0+\beta_2)+(\beta_1+\beta_3)X$. Setting $D=1$ shifts
the intercept by $\beta_2$ and the slope by $\beta_3$.
:::

**Predict (2 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: how many separate numbers do you need to
fully describe a picture of **two straight lines with different slopes**?
Choose one answer. You find the answer in the lecture material.

**R warm-up (8 minutes).** Build the variables for the analysis. The code
uses the same `known_gender` subset as Lecture 5. It adds a dummy `women`
and a centred season `season_c`, which is the season minus the mean season.
It then fits one line of equity offered on season for each group and draws
both lines. Bring the plot:

```{r}
#| label: i06-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

plot(equity_offered_pct ~ season, data = known_gender,
     pch = 16, col = "grey80")
abline(lm(equity_offered_pct ~ season,
          data = subset(known_gender, women == 0)),
       col = "#0072B2", lwd = 3)
abline(lm(equity_offered_pct ~ season,
          data = subset(known_gender, women == 1)),
       col = "#D55E00", lwd = 3)
```

If you are not sure about products, powers, or centering, review
[Mathematics Refresher: products, powers, and centering](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section has two parts. The first part adds an interaction, which lets
a slope differ between two groups. The second part adds a squared term,
which lets a slope change along the predictor itself.

<p class="concept-video"><strong>Stanford explanations (review):</strong>
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk">3.5 Extensions of the Linear Model</a> and
<a href="https://www.youtube.com/watch?v=F-D3lZzYn50">7.1 Polynomials and Step Functions</a>.</p>

### Start with a founder's question

Does advertising work equally well in every week? Does a trend go on in a
straight line forever? Often the answer is "it depends". The two tools of
this lecture let you check this in the data. An interaction lets a slope
differ between groups. A curve lets a slope change along the way.

### The picture before the formula

In every model so far, all groups had the same slope. Lecture 5 let the
height of the line differ between groups. Now the slope may differ as well.
Start with the picture in @fig-i06-two-lines:

```{r}
#| label: fig-i06-two-lines
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-height: 3.8
#| fig-cap: "Two groups, two lines, different slopes (an invented picture). How many numbers do you need to describe this picture completely?"
old_par <- par(no.readonly = TRUE)
par(mar = c(4.2, 4.2, 1, 1), las = 1)
plot(NA, xlim = c(0, 10), ylim = c(0, 10), bty = "n",
     xlab = "Predictor X", ylab = "Outcome Y")
abline(a = 2, b = 0.75, col = "#009E73", lwd = 3)
abline(a = 4.5, b = 0.25, col = "#D55E00", lwd = 3, lty = 2)
text(8.4, 2 + 0.75 * 8.4 + 0.9, "Group D = 0", col = "#009E73")
text(8.4, 4.5 + 0.25 * 8.4 - 0.8, "Group D = 1", col = "#D55E00")
par(old_par)
```

This answers the Wooclap question from before class: four numbers, an
intercept and a slope for each line. The interaction model writes the same
four numbers in a different way:

$$
E[Y\mid X,D]
=\beta_0+\beta_1X+\beta_2D+\beta_3XD.
$$

Each coefficient is one feature of the picture:

- $\beta_0$: where the $D=0$ line meets the vertical axis
- $\beta_1$: the slope of the $D=0$ line
- $\beta_2$: how much higher the $D=1$ line starts
- $\beta_3$: how much steeper the $D=1$ line is.

The same in a table with one row per group, as you derived it before class:

| Group | Intercept | Slope |
|---|---|---|
| $D=0$ | $\beta_0$ | $\beta_1$ |
| $D=1$ | $\beta_0+\beta_2$ | $\beta_1+\beta_3$ |

When $\beta_3=0$ the two lines are parallel, and the model is the dummy
model of Lecture 5 with one common slope.

### The interaction coefficient is a difference in slopes

::: {.definition-box}
<span class="label">Definition</span>

**Interaction.** A term that is the product of two predictors, for example
$X\times D$. It lets the slope of one predictor depend on the value of the
other. Its coefficient is a difference in slopes.
:::

In the additive model the slope of $X$ is the same at every value of the
other predictor. With the product $XD$ in the model, a one-unit increase in
$X$ changes the conditional mean by

$$
\beta_1+\beta_3D.
$$

So $\beta_3$ is the difference between the slope of the group with $D=1$ and
the slope of the reference group. It is not an effect on its own. Its unit
combines the two variables. If $X$ is measured in seasons and $D$ is a
dummy, $\beta_3$ is in outcome units per season: here, equity percentage
points per season, one group minus the reference group. The same algebra
works when the second variable is a number $Z$ and not a dummy. The slope of
$X$ is then $\beta_1+\beta_3Z$, a slope that changes with $Z$.

**Centering.** Centering a predictor means subtracting its mean. With $XD$
in the model, $\beta_2$ compares the two groups at $X=0$. For the Shark Tank
data this would be Season 0, which does not exist. Replacing season with

$$
season\_c = season - \overline{season}
$$

moves that comparison to the average season (about 8.8). Centering changes
one kind of thing and cannot change another:

- It changes the meaning and the value of the intercept and of the group
  coefficient. Both now refer to the average season.
- It cannot change the fitted values, the residuals, $R^2$, or the
  interaction coefficient. The lines are the same lines. Only the place
  where we read the vertical gap between them has moved.

::: {.callout-note collapse="true"}
## Check: which printed numbers change when season is centred?

The intercept and the `women` coefficient change. The interaction
coefficient, the `ask_1k` coefficient, all fitted values, and $R^2$ do not.
You check this in R in the after-class practice.
:::

::: {.callout-caution collapse="true"}
## Try it: two lines from four coefficients

Try this with your neighbour first. In an invented example, a venture models
weekly orders with advertising spending in euros, a dummy for exam weeks,
and their interaction:
fitted orders = 100 + 0.10 × ads + 15 × exam + 0.12 × ads × exam.

*Step 1 (given).* In ordinary weeks (exam = 0) the line is 100 + 0.10 × ads.

*Step 2 (you).* Write the line for exam weeks.

*Step 3 (you).* Compare the fitted orders at 200 euros of advertising in the
two kinds of weeks, and say what the coefficient 0.12 means.

::: {.callout-note collapse="true"}
## Check your answer

Exam weeks: (100 + 15) + (0.10 + 0.12) × ads = 115 + 0.22 × ads. At 200
euros: 120 orders in an ordinary week and 159 in an exam week. The
coefficient 0.12 is the difference between the two slopes: each euro of
advertising goes with 0.12 more orders in an exam week than in an ordinary
week.
:::
:::

### The Shark Tank interaction model

Now take the first half of the driving question. In this lecture,
`women = 1` means that the presenting team is classified as all-women or
mixed, and `women = 0` means all-men. This split is broader than the three
categories of Lecture 5. It does not measure the share of women or the full
team behind the venture. The seven `unknown` records are left out as before.
The code first builds the variables, as in the R warm-up:

```{r}
#| label: i06-interaction-fit
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

interaction_model <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
summary(interaction_model)
```

Use the table with one row per group to get the two season slopes, holding
requested amount fixed. For all-men teams (`women = 0`), the fitted season
slope is about $-0.799$ percentage points per season. For teams with women
(`women = 1`), it is

$$
-0.7994+(-0.0615)=-0.861
\text{ percentage points per season.}
$$

The interaction estimate is $-0.061$ equity percentage points per season,
with a p-value of about 0.520. The difference is small: about 8% of the
all-men slope. Read this result in both directions. The model gives no
evidence that the season trend differs between the two groups. It also does
not prove that the two slopes are equal. The estimate is imprecise (standard
error about 0.096), so the data fit a somewhat steeper trend and also a
somewhat flatter trend for teams with women. No evidence of a difference is
not proof of no difference.

```{r}
#| label: fig-i06-interaction
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted equity offered by season for all-men teams and for teams with women, at the average requested amount. The two lines are almost parallel. The slope of a line is the fitted change per season for that group, and the vertical gap is the fitted difference between the groups in a given season."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)
interaction_model <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
season_grid <- 1:16
ask_bar <- mean(known_gender$ask_1k)
newdat <- expand.grid(season = season_grid, women = c(0, 1))
newdat$season_c <- newdat$season - mean(known_gender$season)
newdat$ask_1k <- ask_bar
newdat$fit <- predict(interaction_model, newdata = newdat)
newdat$group <- ifelse(newdat$women == 1, "Women represented", "All-men")
fit_men_13 <- newdat$fit[newdat$women == 0 & newdat$season == 13]
fit_women_13 <- newdat$fit[newdat$women == 1 & newdat$season == 13]

ggplot(newdat, aes(x = season, y = fit,
                   colour = group, linetype = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(data = subset(newdat, season %in% c(1, 16)), size = 2.2) +
  annotate("text", x = 13, y = fit_women_13 + 1.1,
           label = "Women represented",
           colour = "#D55E00", size = 3.4) +
  annotate("text", x = 13, y = fit_men_13 - 1.1,
           label = "All-men",
           colour = "#009E73", size = 3.4) +
  scale_colour_manual(values = c("All-men" = "#009E73",
                                 "Women represented" = "#D55E00"),
                      guide = "none") +
  scale_linetype_manual(values = c("All-men" = "solid",
                                   "Women represented" = "dashed"),
                        guide = "none") +
  scale_x_continuous(breaks = seq(1, 16, by = 3)) +
  labs(x = "Season", y = "Fitted initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Always graph a model that contains an interaction. In @fig-i06-interaction
the two almost parallel lines are the result. Once $XD$ is in the model,
neither $\hat\beta_1$ nor $\hat\beta_2$ alone describes all teams.

### The hierarchy principle and the `*` shorthand

The **hierarchy principle** says: if a model contains $XD$, keep both $X$
and $D$ in the model, even when one of them is not significant on its own.
The terms $X$ and $D$ define the two lines, and the interaction describes
the difference between these lines. Without one of them the two lines are
restricted for no reason, for example to the same intercept, and this
restriction distorts the interaction coefficient. The `*` shorthand in an R
formula follows the principle for you:

```{r}
#| label: i06-interaction-formula
#| eval: false
lm(y ~ x * z, data = df)  # x * z expands to x + z + x:z
```

The first part of this lecture let the slope differ by group. The second
part lets the slope differ along $x$ itself. Both parts do the same thing:
they give least squares a new column that is built from the data (a product
or a square), and they read several coefficients together.

### Curvature: polynomials are still linear regression

Is the season trend a straight line at all? To let it bend, add the square
of the predictor.

::: {.definition-box}
<span class="label">Definition</span>

**Quadratic term.** The square of a predictor, $x^2$, added to a model that
also contains $x$. It lets the fitted line bend, so the slope changes with
$x$. A model with powers of a predictor is called a polynomial.
:::

The quadratic conditional mean

$$
E[Y\mid X=x]=\beta_0+\beta_1x+\beta_2x^2
$$

is curved in $x$, but it is linear in the unknown coefficients
$\beta_0,\beta_1,\beta_2$. That is all `lm()` requires. The square is one
more column of data.

What changes is how to read the output. The slope is no longer one number.
Moving from $x$ to $x+1$ changes the conditional mean by

$$
(\beta_1(x+1)+\beta_2(x+1)^2) - (\beta_1x+\beta_2x^2)
=\beta_1+\beta_2(2x+1).
$$

This change depends on where you start. The rule is: **never read $\beta_1$
alone while $x^2$ is in the model.** Calculate fitted changes or graph the
curve.

```{r}
#| label: i06-polynomial-fit
linear_season <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
poly_season <- lm(equity_offered_pct ~ season + I(season^2) + ask_1k,
                  data = sharks)
summary(poly_season)$coefficients
```

`I()` tells R to calculate the square. Without it, R would read `^` as
formula syntax. The fitted coefficients are $\hat\beta_1 \approx -1.713$ for
`season` and $\hat\beta_2 \approx 0.051$ for `I(season^2)`. Together they
describe a steep decline in the early seasons that becomes flatter season by
season.

::: {#exm-i06-quadratic-change}
## Worked example: the fitted change from Season 5 to Season 6

Holding requested amount fixed, the fitted change from season $x$ to season
$x+1$ is $\hat\beta_1+\hat\beta_2(2x+1)$. From Season 5 to Season 6, $x = 5$:

$$
\hat\beta_1+\hat\beta_2(2\times5+1)
=-1.713+0.051\times11
=-1.713+0.561
\approx-1.15
\text{ percentage points.}
$$

Reading $\hat\beta_1=-1.713$ alone would overstate the drop by about 0.56
points, which is about half of the fitted drop of 1.15 points. The number
$-1.713$ is not a constant slope. It belongs to $x=0$, a season that does
not exist. Check the calculation in R with the stored coefficients:

```{r}
#| label: i06-quadratic-check
b <- coef(poly_season)
unname(b["season"] + b["I(season^2)"] * (2 * 5 + 1))
```
:::

### Is the curvature worth having? The partial F

The quadratic model and the linear model are nested, so the partial F-test
of Lecture 5 applies to the one added term:

```{r}
#| label: i06-partial-f
anova(linear_season, poly_season)
```

The partial F-test for adding `I(season^2)` gives $F \approx 20.1$ with a
p-value below 0.001. The data show a curve. The gain in fit is small:
adjusted $R^2$ rises from about 0.186 to about 0.196, one percentage point
more of the variation in the equity offered. So also look at the shape of
the curve in @fig-i06-polynomial:

```{r}
#| label: fig-i06-polynomial
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Season means of the equity offered (points) and the fitted quadratic curve at the average requested amount. The curve becomes flatter towards the last observed seasons. Beyond Season 16 (shaded) it would turn upward, but there are no data in that region."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
poly_season <- lm(equity_offered_pct ~ season + I(season^2) + ask_1k,
                  data = sharks)
season_means <- aggregate(equity_offered_pct ~ season, data = sharks, mean)
curve_data <- data.frame(season = seq(1, 16, length.out = 200),
                         ask_1k = mean(sharks$ask_1k))
curve_data$fit <- predict(poly_season, newdata = curve_data)

ggplot(season_means, aes(x = season, y = equity_offered_pct)) +
  annotate("rect", xmin = 16, xmax = 17.6, ymin = -Inf, ymax = Inf,
           fill = "grey75", alpha = 0.25) +
  geom_vline(xintercept = 16, colour = "grey60",
             linetype = "dashed", linewidth = 0.8) +
  geom_line(data = curve_data, aes(x = season, y = fit),
            colour = "#D55E00", linewidth = 1.1) +
  geom_point(size = 2.6, colour = "#009E73") +
  annotate("text", x = 16.8, y = max(season_means$equity_offered_pct),
           label = "no data beyond\nseason 16", colour = "#5B6976",
           size = 3, lineheight = 0.95) +
  scale_x_continuous(breaks = seq(1, 16, by = 3)) +
  coord_cartesian(xlim = c(1, 17.6)) +
  labs(x = "Season", y = "Initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The points in @fig-i06-polynomial are the plain means of each season. The
curve holds the requested amount at its sample mean. To extrapolate means to
use a fitted model outside the range of the data. **Do not extrapolate the
quadratic beyond Seasons 1--16.** The fitted curve reaches its lowest point
just after Season 16 and would rise steeply afterwards. This rise comes from
the shape of a quadratic, not from any data, so it is not a forecast.

::: {.caveat-box}
<span class="label">What flexible terms can and cannot repair</span>

- Interactions and curves make the model more flexible. They repair nothing
  about how the data came about.
- Every added term costs precision. Add a term because a question asks for
  it, not because it turns out significant.
- A curve can be trusted only inside the range of the data.

**What you can do with it:** ask whether a relationship could differ between
groups or bend, add that one term, graph the fitted lines, and report
combined results.
:::

### Answering the driving question

**The same trend for every kind of team?** The estimated difference in
slopes is $-0.061$ percentage points per season (p $\approx$ 0.52). The data
show no difference between all-men teams and teams with women. They also do
not prove that the trends are equal, because the estimate is imprecise.

**A straight line?** Not quite. The quadratic term is significant (partial
$F \approx 20.1$, p $<$ 0.001). The fitted decline becomes flatter: from
about $-1.56$ points per season between Seasons 1 and 2 to about $-0.64$
between Seasons 10 and 11. The curve can be used only inside Seasons 1--16.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You suspect that advertising works better in exam weeks. The model with an interaction gives an advertising slope of 0.10 orders per euro in ordinary weeks and 0.10 + 0.12 = 0.22 in exam weeks. After a price increase, each order now brings 6 euros of margin. So 100 euros of advertising goes with 10 more orders and 60 euros of margin in an ordinary week, which is less than the advertising costs. In an exam week it goes with 22 more orders and 132 euros of margin. The decision: advertise in the exam weeks only. One check first: the interaction of 0.12 has its own interval. If that interval includes zero, the difference between the two kinds of weeks is not yet established, and you keep the budget split for one more term before you commit.
:::

**Next:** Tutorial 2 practises Lectures 4--6 in the format of the exam. It
ends with one full analysis, from a research question to a conclusion of 150
words. The exam asks for this kind of analysis, and so will your thesis.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 85 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. What does an interaction coefficient measure, and what is its unit in the
   model with season and the `women` dummy?
2. Why centre a predictor before you use it in an interaction, and what can
   centering never change?
3. State the hierarchy principle.
4. Why is a quadratic model still a linear regression model?
5. Why should you graph a quadratic curve, and why should you not
   extrapolate it?

::: {.callout-note collapse="true"}
## Check your answers

1. How the slope of one predictor differs with the other predictor. Here its
   unit is equity percentage points per season, as a difference between the
   two groups.
2. So that the intercept and the group coefficient refer to a useful value,
   such as the sample mean, and not to zero. Centering never changes the
   fitted values, the residuals, $R^2$, or the interaction coefficient.
3. Keep $X$ and $D$ in the model whenever their interaction $XD$ is in it,
   and keep $x$ whenever $x^2$ is in it.
4. It is linear in the unknown coefficients. The squared column is one more
   predictor built from the data.
5. No single coefficient is the slope, so only fitted values or the curve
   show what the model says. Outside the observed range the curve can turn
   sharply, and no data support that turn.
:::

### Centering verification (10 minutes)

Fit the interaction model with and without centering season. Check that the
fitted values and the interaction coefficient are unchanged, and report
which coefficients do change.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: i06-centering-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

uncentered <- lm(equity_offered_pct ~ season * women + ask_1k,
                 data = known_gender)
centered <- lm(equity_offered_pct ~ season_c * women + ask_1k,
               data = known_gender)
all.equal(fitted(uncentered), fitted(centered))
c(uncentered = unname(coef(uncentered)["season:women"]),
  centered = unname(coef(centered)["season_c:women"]))
c(women_uncentered = unname(coef(uncentered)["women"]),
  women_centered = unname(coef(centered)["women"]))
```

The fitted values agree, and both interaction estimates are about $-0.061$.
The `women` coefficient changes from about $1.35$ (a comparison at Season 0,
which does not exist) to about $0.81$ (a comparison at the average season,
about 8.8). The intercept changes as well. Centering moved the place where
the gap is read. It did not move the lines.
:::

### Finish a half-worked example: a quadratic change (15 minutes)

Read the worked calculation for Season 5 to Season 6 again
(@exm-i06-quadratic-change). Then calculate the fitted change from **Season
10 to Season 11** in the same model.

*Step 1 (given).* The formula is $\hat\beta_1+\hat\beta_2(2x+1)$ with $x=10$,
so $\hat\beta_2$ is multiplied by $21$.

*Step 2 (you).* Plug in $\hat\beta_1\approx-1.713$ and $\hat\beta_2\approx0.051$.

*Step 3 (you).* Compare the result with the Season 5 to 6 change and state
in one sentence what the comparison shows about the trend.

::: {.callout-note collapse="true"}
## Check your answer

$-1.713+0.051\times21\approx-1.713+1.071\approx-0.64$ percentage points
(with the unrounded coefficients: $\hat\beta_1+21\hat\beta_2=-0.642$). The
drop per season has become smaller, from about $-1.15$ (Seasons 5--6) to
about $-0.64$ (Seasons 10--11). The decline becomes flatter in later
seasons. This is what a positive $\hat\beta_2$ means.
:::

Now do the next tasks on your own, with new coefficients from an invented
example. A model of monthly revenue (€1,000s) on the age of a shop $x$
(years) reports $\widehat Y = 40 - 2.00x + 0.060x^2$.

1. Calculate the fitted change from age 3 to age 4 and from age 8 to age 9.
2. A colleague reports "revenue falls €2,000 per year of age." Correct the
   sentence.
3. In $\widehat Y=10+2X+3D-0.5XD$, write the lines of both groups and
   interpret the interaction coefficient.

::: {.callout-note collapse="true"}
## Check your answers

1. Age 3 to 4: $-2.00+0.060(2\times3+1)=-2.00+0.42=-1.58$, about
   $-€1{,}580$. Age 8 to 9: $-2.00+0.060\times17=-0.98$, about $-€980$.
2. $-2.00$ is not the fitted change per year. With $x^2$ in the model the
   change depends on age: about $-1.58$ from age 3 to 4 and about $-0.98$
   from age 8 to 9. The decline becomes weaker as shops get older.
3. For $D=0$: $\widehat Y=10+2X$. For $D=1$: $\widehat Y=13+1.5X$. The
   interaction coefficient $-0.5$ says that the slope of the $D=1$ group is
   0.5 outcome units per unit of $X$ smaller than the slope of the $D=0$
   group.
:::

### Cumulative mini-task (15 minutes)

In Lecture 4 you wrote a paragraph that connects the diagnostics of Lecture
2 to the three-predictor model of the equity offered. Add one argument to
that paragraph. Propose either an interaction or a quadratic term for that
model. State the question about the business or the data that motivates the
term, and say how you would test whether the term belongs in the model.

::: {.callout-note collapse="true"}
## Check a model answer

A good answer: "The decline over the seasons need not be a straight line.
The show was new in the early seasons and pitches became more professional
later, so the trend may become flatter. I would add `I(season^2)`. I add the
term because of this question about the shape, not because I am searching
for a significant result. I would test it with a partial F-test against the
nested linear model on the same observations (here $F\approx20.1$,
p $<$ 0.001), check that adjusted $R^2$ rises, graph the fitted curve inside
Seasons 1--16, and report fitted changes from season to season in place of
the `season` coefficient alone." An answer with an interaction is equally
good if it names the question and the nested test. An example is the
interaction of season with the `women` dummy, which asks whether the trend
differs between the groups. Its test gives $F\approx0.41$ and
p $\approx$ 0.52, so the term is not kept.
:::

### Review: one partial-F question from Lecture 5 (5 minutes)

Adding the two gender dummies to the season-and-ask model of equity reduced
the residual sum of squares by 436.25 using 2 degrees of freedom, and the
complete model left 82,792 on 1,429 degrees of freedom. Compute the partial
F statistic and state the 5% conclusion.

::: {.callout-note collapse="true"}
## Check your answer

$F=(436.25/2)/(82{,}792/1{,}429)=218.1/57.9\approx3.76$, with
p $\approx$ 0.023. At 5%, the two category terms are jointly useful in that
model, even though the t-test of the mixed dummy alone is far from
significant. Joint tests and individual tests answer different questions.
:::

### Worked exam interpretation (5 minutes)

**Question.** This question is in the style of Question 5 of the mock exam.
The numbers are invented. A model of the sales growth of ventures
(percentage points) on marketing spending (€10,000s), a dummy for
online-only ventures, and their interaction reports `spending 1.20`,
`online 2.00`, `spending:online -0.50`. A student writes: "Spending raises
growth by 1.2 percentage points per €10,000 for all ventures." Correct the
sentence.

::: {.callout-note collapse="true"}
## Check a model answer

Two corrections. First, with the interaction in the model, 1.20 is the
fitted spending slope only for ventures that are not online-only
(`online = 0`). For online-only ventures the slope is $1.20-0.50=0.70$
percentage points per €10,000. There is no single slope for all ventures.
Second, "raises" claims causation. These are fitted associations among the
sampled ventures, holding the other included terms fixed. The data do not
come from an experiment, so they do not show an effect of spending.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "The interaction has a coefficient of -0.061 and a p-value of 0.52. What do I conclude?" It answered:

> "The p-value proves that the season trend is identical for both groups. And the season coefficient of -0.80 is the season trend for all teams."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Two errors. A large p-value means the data do not show a difference. It does not prove that the trends are the same, because the estimate is imprecise. And with an interaction in the model, -0.80 is the trend for the group coded 0, the all-men teams. For teams with women it is -0.80 - 0.06 = -0.86.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture, or any business you know well. Name one
relationship in its data whose slope could depend on something else. For
example, the response to price may differ by customer segment, the result of
a campaign may differ by channel, or growth may become flatter as the
business gets larger. Then write two sentences. First, state the interaction
or quadratic term you would add and the question behind it. Second, say what
a result that is not significant would tell you, and what it would not tell
you. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus (an invented example). "I would add an interaction of advertising spend with exam week, because my question is whether ads only work when students are on campus. An interaction that is not significant would mean no evidence that the slopes differ. It would not prove that exam weeks make no difference."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant. Ask it to read your two sentences and to
check whether you treated "no evidence of a difference" as "proof of no
difference".
::::

[Continue to Tutorial 2: Multiple regression and extensions](t02-exam-prep.qmd).

[Code sheet: Lecture 6 commands and formulas](code-sheet.qmd#lecture-6-interactions-and-polynomial-terms).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Section 3.3.2 and Sections 3.6.4--3.6.6,
and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 21.

:::
