# Lecture 6: Interactions and curvature {#i06}

```{r}
#| label: i06-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 6, you can</span>

- interpret an interaction coefficient as a difference in slopes, also when it is not significant
- tell a moderator from a mediator and a confounder
- fit a quadratic term and compute a fitted change from the curve
- test added terms jointly and avoid using a curve beyond the data
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">An average slope can hide important differences. A relationship may differ between groups, or its slope may change as a predictor grows. Interactions and curved terms let a model show these patterns.</p>

The equity offered (`equity_offered_pct`) in the Shark Tank file falls over
the seasons. Is this trend the same for every kind of presenting team, and
is it a straight line at all?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch Stanford Statistical Learning
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=13">3.5
Extensions of the Linear Model</a> and
<a href="https://www.youtube.com/watch?v=F-D3lZzYn50&list=PLoROMvodv4rOzrYsAxzQyHb8n_RWNuS1e&index=48">7.1
Polynomials and Step Functions</a>. Focus on why an interaction changes a
slope, and why a model with a quadratic term (a squared predictor) is still a
linear regression. All
links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
The videos explain key ideas in another way and can use a different
notation. The lecture material and its notation always come first.
</div>

**Derive the two lines.** In the model
$Y=\beta_0+\beta_1X+\beta_2D+\beta_3XD+\varepsilon$ with a dummy variable
$D\in\{0,1\}$, write the line for $D=0$ and the line for $D=1$.

::: {.callout-note collapse="true"}
## Check your preparation

For $D=0$: $E[Y\mid X,D=0]=\beta_0+\beta_1X$. For $D=1$:
$E[Y\mid X,D=1]=(\beta_0+\beta_2)+(\beta_1+\beta_3)X$. Setting $D=1$ shifts
the intercept by $\beta_2$ and the slope by $\beta_3$.
:::

**Predict.** Answer this question before you come to class: how many separate numbers do you need to
fully describe a picture of two straight lines with different slopes?
Choose one answer. You will find the answer in the lecture material.

[Shark Tank data]{.shark-tag} **R warm-up.** Build the variables for the analysis. The code
uses the same `known_gender` subset as Lecture 5. It adds a dummy `women`
and a centred season `season_c`, which is the season minus the mean season.
It then fits one line of equity offered on season for each group and draws
both lines. Bring the plot:

```{r}
#| label: i06-prep-warmup
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

plot(equity_offered_pct ~ season, data = known_gender,
     pch = 16, col = "grey80")
abline(lm(equity_offered_pct ~ season,
          data = subset(known_gender, women == 0)),
       col = "#0072B2", lwd = 3)
abline(lm(equity_offered_pct ~ season,
          data = subset(known_gender, women == 1)),
       col = "#D55E00", lwd = 3)
```

**Code along (bring your laptop).** Bring a charged laptop with RStudio and
the course project open. All the code of the lecture is in this chapter,
ready to copy. You can run each step yourself during the lecture and get the
same numbers and figures.

If you are not sure about products, powers, or centring, review
[Mathematics Refresher: products, interactions, and combined effects](../resources/math-refresher.html#products-interactions-and-combined-effects)
and [polynomial terms](../resources/math-refresher.html#polynomial-terms).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section adds two terms to the regression model. An interaction term
lets a slope differ between two groups, or change with the value of another
predictor. A quadratic term lets a slope change along the predictor itself.

<p class="concept-video"><strong>Stanford explanations (review):</strong>
<a href="https://www.youtube.com/watch?v=dEBQmiXv9fk">3.5 Extensions of the Linear Model</a> and
<a href="https://www.youtube.com/watch?v=F-D3lZzYn50">7.1 Polynomials and Step Functions</a>.</p>

### Start with a founder's question

Does advertising work equally well in every week? Does a trend go on in a
straight line forever? Often the answer is "it depends". The two tools of
this lecture let you check this in the data. An interaction lets a slope
differ between groups. A curve lets a slope change as the predictor grows.

### The picture before the formula

In every model so far, all groups had the same slope. Lecture 5 let the
height of the line differ between groups. Now the slope may differ as well.
Start with the picture in @fig-i06-two-lines:

```{r}
#| label: fig-i06-two-lines
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.8
#| fig-cap: "Two groups, two lines, different slopes (an invented picture). How many numbers do you need to describe this picture completely?"
old_par <- par(no.readonly = TRUE)
par(mar = c(4.2, 4.2, 1, 1), las = 1)
plot(NA, xlim = c(0, 10), ylim = c(0, 10), bty = "n",
     xlab = "Predictor X", ylab = "Outcome Y")
abline(a = 2, b = 0.75, col = "#009E73", lwd = 3)
abline(a = 4.5, b = 0.25, col = "#D55E00", lwd = 3, lty = 2)
text(8.4, 2 + 0.75 * 8.4 + 0.9, "Group D = 0", col = "#009E73")
text(8.4, 4.5 + 0.25 * 8.4 - 0.8, "Group D = 1", col = "#D55E00")
par(old_par)
```

This answers the prediction
question from before class: four numbers, an
intercept and a slope for each line. The interaction model writes the same
four numbers in a different way:

$$
E[Y\mid X,D]
=\beta_0+\beta_1X+\beta_2D+\beta_3XD.
$$

Each coefficient is one feature of the picture:

- $\beta_0$: where the $D=0$ line meets the vertical axis
- $\beta_1$: the slope of the $D=0$ line
- $\beta_2$: how much higher the $D=1$ line starts
- $\beta_3$: how much steeper the $D=1$ line is.

The same in a table with one row per group, as you derived it before class:

| Group | Intercept | Slope |
|---|---|---|
| $D=0$ | $\beta_0$ | $\beta_1$ |
| $D=1$ | $\beta_0+\beta_2$ | $\beta_1+\beta_3$ |

When $\beta_3=0$ the two lines are parallel, and the model is the dummy
model of Lecture 5 with one common slope.

### The interaction coefficient is a difference in slopes

::: {.definition-box}
<span class="label">Definition</span>

**Interaction term.** A term that is the product of two predictors, for
example $X\times D$. It lets the slope of one predictor depend on the value of the
other. Its coefficient is a difference in slopes.
:::

In the additive model the slope of $X$ is the same at every value of the
other predictor. With the product $XD$ in the model, a one-unit increase in
$X$ changes the conditional mean by

$$
\beta_1+\beta_3D.
$$

So $\beta_3$ is the difference between the slope of the group with $D=1$ and
the slope of the reference category ($D=0$). It is not an effect on its own. Its unit
combines the two variables. If $X$ is measured in seasons and $D$ is a
dummy, $\beta_3$ is in outcome units per season: here, equity percentage
points per season, the group with $D=1$ minus the reference category. The same algebra
works when the second variable is a number $Z$ and not a dummy. The slope of
$X$ is then $\beta_1+\beta_3Z$, a slope that changes with $Z$. The section
on a continuous moderator below works through this case.

[**Centring.**]{#i06-centring} Centring a predictor means subtracting its mean. With $XD$
in the model, $\beta_2$ compares the two groups at $X=0$. For the Shark Tank
data this would be Season 0, which does not exist. Replacing season with

$$
\text{season\_c} = \text{season} - \overline{\text{season}}
$$

moves that comparison to the average season (about 8.8). Centring changes
some results and cannot change others:

- It changes the meaning and the value of the intercept and of the dummy
  coefficient. Both now refer to the average season.
- It cannot change the fitted values, the residuals, $R^2$, or the
  interaction coefficient. The lines are the same lines. Only the place
  where we compare the two groups has moved.

::: {.callout-note collapse="true"}
## Check: which printed numbers change when season is centred?

The intercept and the `women` coefficient change. The interaction
coefficient, the `ask_1k` coefficient, all fitted values, and $R^2$ do not.
You will check this in R in the after-class practice.
:::

::: {.callout-caution collapse="true"}
## Try it: two lines from four coefficients

Try this with your neighbour first. In an invented example, a venture models
weekly orders with advertising spend in euros, a dummy for exam weeks,
and their interaction:
fitted orders = 100 + 0.10 × ads + 15 × exam + 0.12 × ads × exam.

*Step 1 (given).* In ordinary weeks (exam = 0) the line is 100 + 0.10 × ads.

*Step 2 (you).* Write the line for exam weeks.

*Step 3 (you).* Compare the fitted orders at 200 euros of advertising in the
two kinds of weeks, and say what the coefficient 0.12 means.

::: {.callout-note collapse="true"}
## Check your answer

Exam weeks: (100 + 15) + (0.10 + 0.12) × ads = 115 + 0.22 × ads. At 200
euros: 120 orders in an ordinary week and 159 in an exam week. The
coefficient 0.12 is the difference between the two slopes: each euro of
advertising goes with 0.12 more orders in an exam week than in an ordinary
week.
:::
:::

### The Shark Tank interaction model

[Shark Tank data]{.shark-tag} Now answer the first part of the driving question. In this lecture,
`women = 1` means that the presenting team is classified as all-women or
mixed, and `women = 0` means all-men. This split is broader than the three
categories of Lecture 5. It does not measure the share of women or the full
team behind the venture. As before, we leave out the seven `unknown` records.
The code first builds the variables, as in the R warm-up:

```{r}
#| label: i06-interaction-fit
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

interaction_model <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
summary(interaction_model)
```

::: {#exm-i06-reading-interaction}
## Worked example: reading each coefficient of the interaction model

[Shark Tank data]{.shark-tag} The outcome is the equity offered in percent, so every coefficient is in
percentage points. The predictor `season_c` is in seasons, measured from the
average season. The requested amount `ask_1k` is in \$1,000s. The reference
category is all-men teams (`women = 0`). The code below stores the
coefficients and computes the two group lines at the average requested
amount:

```{r}
#| label: i06-interaction-read
b_int <- coef(interaction_model)
ask_bar <- mean(known_gender$ask_1k)
intercept_men <- unname(b_int["(Intercept)"] + b_int["ask_1k"] * ask_bar)
intercept_women <- intercept_men + unname(b_int["women"])
c(intercept_men = intercept_men,
  intercept_women = intercept_women,
  slope_men = unname(b_int["season_c"]),
  slope_women = unname(b_int["season_c"] + b_int["season_c:women"]))
c(mean_season = mean(known_gender$season),
  min_ask_1k = min(known_gender$ask_1k),
  gap_share_of_men = unname(b_int["women"]) / intercept_men,
  interaction_share_of_slope = unname(b_int["season_c:women"] /
                                        b_int["season_c"]))
```

Step 1: the two lines. Fill in the table with one row per group. At the
average requested amount, with $\text{season\_c}$ as the predictor:

$$
\begin{aligned}
\text{all-men teams: } & \widehat{\text{equity}} = 12.866 - 0.799\times\text{season\_c},\\
\text{teams with women: } & \widehat{\text{equity}} = (12.866+0.807) + (-0.7994-0.0615)\times\text{season\_c}\\
& \phantom{\widehat{\text{equity}}} = 13.673 - 0.861\times\text{season\_c}.
\end{aligned}
$$

Step 2: one sentence per coefficient.

- `season_c` ($-0.799$). Among all-men teams, one season later is
  associated with an equity offered that is on average 0.80 percentage points
  lower, at the same requested amount. This is the slope of the reference
  category only.
- `season_c:women` ($-0.061$). Per one season more, the equity offered by
  teams with women falls 0.061 percentage points more than that of all-men
  teams. This is the difference between the two slopes. It is about 8% of
  the all-men slope.
- `women` ($0.807$). In the average season (about 8.8), teams with women are
  fitted to offer about 0.81 percentage points more equity than all-men teams,
  at the same requested amount. With the interaction in the model, this
  difference holds only where $\text{season\_c}=0$. In other seasons the gap
  is $0.807-0.061\times\text{season\_c}$.
- `ask_1k` ($-0.0027$). A request that is \$1,000 higher is associated with
  an equity offered that is on average 0.0027 percentage points lower, in
  both groups. Per \$100,000 more, this is about 0.27 percentage points.
- The intercept ($13.62$) is the fitted equity offered for all-men teams in
  the average season with a requested amount of \$0. The smallest request in
  the data is \$10,000, so the intercept has no useful meaning of its own.

Step 3: how large is it? In the average season, all-men teams are fitted to
offer 12.87% and teams with women 13.67%. The gap is 0.81 percentage points.
Measured as a percent of the all-men level, the same gap is about 6%. So
"0.81 percent more" would be wrong. The coefficient is a difference in
percentage points.

Why does the model use `season_c` and not `season`? The dummy coefficient
always compares the groups at $X=0$. Without [centring](#i06-centring), $X=0$
would be Season 0, a season outside the data. Centring moves the comparison
to the average season, where there are many pitches.
:::

The interaction estimate is $-0.061$ equity percentage points per season,
with a p-value of about 0.520. The difference is small, about 8% of the
all-men slope (@exm-i06-reading-interaction). Draw two conclusions from
this result. First, the model gives no
evidence that the season trend differs between the two groups. Second, the model
does not prove that the two slopes are equal. The estimate is imprecise
(standard error about 0.096), so the data are consistent with a somewhat
steeper trend and also with a somewhat flatter trend for teams with women. No evidence of a difference is
not proof of no difference.

```{r}
#| label: fig-i06-interaction
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted equity offered by season for all-men teams and for teams with women, at the average requested amount. The two lines are almost parallel. The slope of a line is the fitted change per season for that group, and the vertical distance between the lines is the fitted difference between the groups in a given season."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)
interaction_model <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
season_grid <- 1:16
ask_bar <- mean(known_gender$ask_1k)
newdat <- expand.grid(season = season_grid, women = c(0, 1))
newdat$season_c <- newdat$season - mean(known_gender$season)
newdat$ask_1k <- ask_bar
newdat$fit <- predict(interaction_model, newdata = newdat)
newdat$group <- ifelse(newdat$women == 1, "Teams with women", "All-men")
fit_men_13 <- newdat$fit[newdat$women == 0 & newdat$season == 13]
fit_women_13 <- newdat$fit[newdat$women == 1 & newdat$season == 13]

ggplot(newdat, aes(x = season, y = fit,
                   colour = group, linetype = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(data = subset(newdat, season %in% c(1, 16)), size = 2.2) +
  annotate("text", x = 13, y = fit_women_13 + 1.1,
           label = "Teams with women",
           colour = "#D55E00", size = 3.4) +
  annotate("text", x = 13, y = fit_men_13 - 1.1,
           label = "All-men",
           colour = "#009E73", size = 3.4) +
  scale_colour_manual(values = c("All-men" = "#009E73",
                                 "Teams with women" = "#D55E00"),
                      guide = "none") +
  scale_linetype_manual(values = c("All-men" = "solid",
                                   "Teams with women" = "dashed"),
                        guide = "none") +
  scale_x_continuous(breaks = seq(1, 16, by = 3)) +
  labs(x = "Season", y = "Fitted initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Always graph a model that contains an interaction. In @fig-i06-interaction,
the main result is that the two lines are almost parallel. Once $XD$ is in the model,
neither $\hat\beta_1$ nor $\hat\beta_2$ alone describes all teams.

### A continuous moderator

The second variable of an interaction need not be a dummy. It can be any
number $Z$, such as a price, a firm size, or a number of stores. The model is

$$
E[Y\mid X,Z]=\beta_0+\beta_1X+\beta_2Z+\beta_3XZ,
$$

and a one-unit increase in $X$ changes the conditional mean by
$\beta_1+\beta_3Z$. There are now not two lines but one line for every value
of $Z$.

::: {.definition-box}
<span class="label">Definition</span>

**Moderator.** A variable $Z$ that changes the slope of another predictor
$X$ through the interaction term $XZ$. The slope of $X$ at one value of $Z$,
$\beta_1+\beta_3Z$, is called the conditional effect of $X$ at that value.
:::

The name "conditional effect" is standard, but it is still a fitted slope.
It is a causal effect only when the design of the study allows a causal
reading. Two coefficients now need care. $\beta_1$ is the slope of $X$ only
where $Z=0$, and $\beta_2$ is the slope of $Z$ only where $X=0$. If zero lies
outside the data, neither number describes any real case. The coefficients
$\beta_1$ and $\beta_2$ are often called the main effects. With $XZ$ in the
model they are conditional too. [Centring](#i06-centring) $Z$ makes
$\beta_1$ the slope of $X$ at the mean of $Z$.

::: {#exm-i06-moderator}
## Worked example: marketing pays more in regions with more stores

An invented example. A franchise chain has 150 regions. For each region it
records the monthly sales and the monthly marketing spending, both in
€1,000s, and the number of stores. The question: does an extra €1,000 of
marketing go with more sales in regions with many stores? The code below
builds the invented data and fits the model with the interaction term:

```{r}
#| label: i06-moderator-fit
set.seed(606)
n_regions <- 150
stores <- sample(1:12, n_regions, replace = TRUE)
marketing <- round(runif(n_regions, min = 2, max = 20), 1)
sales <- round(30 + 1.0 * marketing + 6 * stores +
               0.25 * marketing * stores + rnorm(n_regions, sd = 12), 1)
franchise <- data.frame(sales, marketing, stores)
summary(franchise)

moderator_model <- lm(sales ~ marketing * stores, data = franchise)
summary(moderator_model)$coefficients
```

Step 1: the slope of marketing as a function of stores. With the
coefficients from the output,

$$
\widehat{\text{slope of marketing}} = 1.745 + 0.147\times\text{stores}.
$$

Step 2: the slope at three meaningful values of the moderator. Take the
lower quartile, the median, and the upper quartile of the number of stores:

```{r}
#| label: i06-moderator-slopes
b_mod <- coef(moderator_model)
store_values <- quantile(franchise$stores, c(0.25, 0.5, 0.75))
store_values
marketing_slopes <- b_mod["marketing"] + b_mod["marketing:stores"] * store_values
marketing_slopes
marketing_slopes[3] / marketing_slopes[1]
```

In a region with 3 stores (the lower quartile), €1,000 more marketing is
associated with about €2,190 more monthly sales on average. With 6 stores
(the median) it is about €2,630, and with 10 stores (the upper quartile)
about €3,220. The slope at the upper quartile is about 1.47 times the slope
at the lower quartile.

Step 3: one sentence per coefficient.

- `marketing:stores` ($0.147$). A region with one more store has a
  marketing slope that is 0.147 higher: €147 more monthly sales per €1,000
  of marketing. Read the other way round, a region that spends €1,000 more
  on marketing has a stores slope that is €147 per store higher.
- `marketing` ($1.745$). This is the slope of marketing in a region with 0
  stores. No region has 0 stores (the minimum is 1), so this number does not
  describe the marketing slope of any region.
- `stores` ($6.768$). This is the slope of stores at €0 of marketing. The
  smallest spending in the data is €2,300, so this number also lies outside
  the data.

Step 4: centre the moderator. With $\text{stores\_c}$, the number of stores
minus its mean, the marketing coefficient becomes the slope at the average
number of stores (about 6.3):

```{r}
#| label: i06-moderator-centred
franchise$stores_c <- franchise$stores - mean(franchise$stores)
coef(lm(sales ~ marketing * stores_c, data = franchise))
```

The marketing coefficient is now 2.676: at the average number of stores,
€1,000 more marketing is associated with about €2,680 more monthly sales.
The interaction coefficient stays 0.147. As with the dummy, centring moves
the place of the comparison and not the fitted lines.
:::

@fig-i06-moderator draws the fitted lines at the three values of the
moderator. The lines fan out because their slopes differ.

```{r}
#| label: fig-i06-moderator
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Fitted monthly sales against marketing spending (invented example) in regions with 3, 6, and 10 stores, the quartiles of the number of stores. The lines become steeper as the number of stores grows: the slope of marketing depends on the moderator."
library(ggplot2)
set.seed(606)
n_regions <- 150
stores <- sample(1:12, n_regions, replace = TRUE)
marketing <- round(runif(n_regions, min = 2, max = 20), 1)
sales <- round(30 + 1.0 * marketing + 6 * stores +
               0.25 * marketing * stores + rnorm(n_regions, sd = 12), 1)
franchise <- data.frame(sales, marketing, stores)
moderator_model <- lm(sales ~ marketing * stores, data = franchise)

lines_data <- expand.grid(marketing = seq(2, 20, length.out = 50),
                          stores = c(3, 6, 10))
lines_data$fit <- predict(moderator_model, newdata = lines_data)
lines_data$group <- factor(paste(lines_data$stores, "stores"),
                           levels = c("3 stores", "6 stores", "10 stores"))
label_data <- subset(lines_data, marketing == 20)

ggplot(franchise, aes(x = marketing, y = sales)) +
  geom_point(colour = "grey80", size = 1.6) +
  geom_line(data = lines_data,
            aes(x = marketing, y = fit, colour = group, linetype = group),
            linewidth = 1.1) +
  geom_text(data = label_data,
            aes(x = marketing + 0.4, y = fit, label = group, colour = group),
            hjust = 0, size = 3.4) +
  scale_colour_manual(values = c("3 stores" = "#009E73",
                                 "6 stores" = "#0072B2",
                                 "10 stores" = "#D55E00"),
                      guide = "none") +
  scale_linetype_manual(values = c("3 stores" = "solid",
                                   "6 stores" = "longdash",
                                   "10 stores" = "dashed"),
                        guide = "none") +
  coord_cartesian(xlim = c(2, 23)) +
  labs(x = "Monthly marketing spending (EUR 1,000s)",
       y = "Monthly sales (EUR 1,000s)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

### The hierarchy principle and the `*` shorthand

The **hierarchy principle** says: if a model contains $XD$, keep both $X$
and $D$ in the model, even when one of them is not significant on its own.
The same holds for $XZ$ with a continuous moderator $Z$. The terms $X$ and
$D$ define the two lines, and the interaction describes the difference
between these lines. If you leave out $D$, the model forces both lines to
start at the same height. This distorts the interaction coefficient. The `*` shorthand in an R formula follows the principle for
you:

```{r}
#| label: i06-interaction-formula
#| eval: false
lm(y ~ x * z, data = df)  # x * z expands to x + z + x:z
```

### Moderation and mediation

Two words often appear in business research about a third variable: a
moderator and a mediator. They describe different roles, and a regression
handles them in different ways.

Moderation first. A moderator changes the strength or the direction of the
relationship between $X$ and $Y$. With a moderator, the answer to "how is $X$
related to $Y$?" is "it depends on $Z$". In a regression, moderation shows up
as an interaction term. You have seen two cases. In
@exm-i06-reading-interaction the moderator is a dummy, the `women` group,
and the season slope differs between two groups. In @exm-i06-moderator the
moderator is a number, the number of stores, and the marketing slope grows
with every extra store. A moderator can also reverse a relationship, for
example a price cut that raises orders for students and lowers them for a
luxury segment.

::: {.definition-box}
<span class="label">Definition</span>

**Mediator.** A variable $M$ that lies on the path from $X$ to $Y$: $X$
affects $M$, and $M$ affects $Y$. The part of the effect of $X$ on $Y$ that
runs through $M$ is called the indirect effect.
:::

Take marketing spending ($X$), website visits ($M$), and sales ($Y$).
Marketing brings visitors to the website, and visitors buy. Marketing can
also raise sales in other ways, for example through people who see an
advert and buy in a shop. @fig-i06-mediation shows the two paths. Three
effects describe them:

- the total effect: how much $Y$ changes when $X$ changes, by all paths
  together
- the direct effect: the part that does not run through $M$
- the indirect effect: the part that runs through $M$.

The total effect is the direct effect plus the indirect effect.

```{r}
#| label: fig-i06-mediation
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3
#| fig-cap: "A mediation diagram (invented example). The indirect path runs from marketing spending through website visits to sales. The direct path goes from marketing spending to sales without visits. The total effect is the sum of both paths."
library(ggplot2)
boxes <- data.frame(x = c(1, 3, 5), y = c(1, 2, 1),
                    label = c("Marketing\nspending (X)",
                              "Website\nvisits (M)",
                              "Sales (Y)"))
arrows_data <- data.frame(x = c(1.45, 3.55, 1.6),
                          y = c(1.25, 1.75, 1),
                          xend = c(2.45, 4.55, 4.4),
                          yend = c(1.75, 1.25, 1),
                          path = c("indirect", "indirect", "direct"))

ggplot() +
  geom_segment(data = arrows_data,
               aes(x = x, y = y, xend = xend, yend = yend, colour = path),
               linewidth = 1.1,
               arrow = arrow(length = grid::unit(0.25, "cm"), type = "closed")) +
  geom_label(data = boxes, aes(x = x, y = y, label = label),
             colour = "#17324D", size = 3.6, label.padding = grid::unit(0.4, "lines")) +
  annotate("text", x = 3, y = 1.55, label = "indirect path",
           colour = "#009E73", size = 3.4) +
  annotate("text", x = 3, y = 0.85, label = "direct path",
           colour = "#D55E00", size = 3.4) +
  scale_colour_manual(values = c("indirect" = "#009E73",
                                 "direct" = "#D55E00"),
                      guide = "none") +
  coord_cartesian(xlim = c(0.4, 5.6), ylim = c(0.7, 2.3)) +
  theme_void()
```

::: {#exm-i06-mediation}
## Worked example: how much of the marketing slope runs through visits?

An invented example. For 120 small web shops we know the monthly marketing
spending and the monthly sales, both in €1,000s, and the monthly website
visits in 1,000s. The code below builds the invented data. It sets a
direct path and a path through visits, which a real founder never knows.

```{r}
#| label: i06-mediation-fit
set.seed(616)
n_shops <- 120
marketing <- round(runif(n_shops, min = 2, max = 12), 1)
visits <- round(4 + 1.5 * marketing + rnorm(n_shops, sd = 2), 1)
sales <- round(20 + 2 * marketing + 2 * visits + rnorm(n_shops, sd = 5), 1)
shop <- data.frame(sales, marketing, visits)

total_model <- lm(sales ~ marketing, data = shop)
direct_model <- lm(sales ~ marketing + visits, data = shop)
visits_model <- lm(visits ~ marketing, data = shop)
coef(total_model)
coef(direct_model)
coef(visits_model)

total <- unname(coef(total_model)["marketing"])
direct <- unname(coef(direct_model)["marketing"])
c(total = total, direct = direct, through_visits = total - direct,
  share_through_visits = (total - direct) / total)
unname(coef(visits_model)["marketing"] * coef(direct_model)["visits"])
```

Step 1: the total association. Without visits in the model, €1,000 more
marketing is associated with about €5,040 more monthly sales on average.

Step 2: the direct part. With visits added, the marketing coefficient falls
to 2.62. At the same number of visits, €1,000 more marketing is associated
with about €2,620 more sales.

Step 3: the part through visits. The difference is $5.04-2.62=2.42$, about
€2,420 of sales per €1,000 of marketing. That is about 48% of the total.
The same number comes from the two paths. €1,000 more marketing goes with
about 1,490 more visits ($1.488$ in 1,000s), and 1,000 more visits go with
about €1,630 more sales at the same marketing ($1.625$). The product is
$1.488\times1.625\approx2.42$.
:::

A mediator is not a control variable. In
[Lecture 4](i04-multiple-regression.qmd#what-control-variables-can-and-cannot-do)
you added control variables to hold confounders fixed. A mediator looks like
one more predictor, but adding it changes the question. With visits in the
model, the marketing coefficient no longer contains the part that runs
through visits. If you want the total effect of marketing, do not control
for visits. So ask first what role a variable plays. Control for a
confounder. Do not control for a mediator when you want the total effect.

Mediation is a causal idea. It needs a causal design, such as an
experiment, or strong assumptions. With observational data, the split into
a direct and an indirect part only describes the data.

The table compares the three roles of a third variable $Z$.

| Role | What it is | How it appears in a regression | Example |
|---|---|---|---|
| Moderator | Changes the slope of $X$ | Interaction term $XZ$ | Number of stores changes the marketing slope |
| Mediator | Lies on the path from $X$ to $Y$ | Adding it shrinks the coefficient of $X$ to the direct part | Website visits between marketing and sales |
| Confounder | Affects both $X$ and $Y$ | Control variable, to remove omitted variable bias | Firm size behind both advertising and sales |

### Curvature: polynomials are still linear regression

An interaction term lets the slope of $x$ differ by group or change with a
moderator. A quadratic term, defined below, lets the slope change as $x$
itself changes. Both add one new column to the data, a product or a square.
In both, you read two coefficients together.

[Shark Tank data]{.shark-tag} Is the season trend a straight line at all? To let it bend, add the square
of the predictor.

::: {.definition-box}
<span class="label">Definition</span>

**Quadratic term.** The square of a predictor, $x^2$, added to a model that
also contains $x$. It lets the fitted line bend, so the slope changes with
$x$. A model with powers of a predictor is called a polynomial.
:::

The quadratic conditional mean

$$
E[Y\mid X=x]=\beta_0+\beta_1x+\beta_2x^2
$$

is curved in $x$, but it is linear in the unknown coefficients
$\beta_0,\beta_1,\beta_2$. Linearity in the coefficients is all that `lm()` requires. The square is one
more column of data.

What changes is how you read the output. The slope is no longer one number.
Moving from $x$ to $x+1$ changes the conditional mean by

$$
(\beta_1(x+1)+\beta_2(x+1)^2) - (\beta_1x+\beta_2x^2)
=\beta_1+\beta_2(2x+1).
$$

This change depends on the starting value $x$. The rule is: never read $\beta_1$
alone while $x^2$ is in the model. Calculate fitted changes or graph the
curve.

```{r}
#| label: i06-polynomial-fit
linear_season <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
poly_season <- lm(equity_offered_pct ~ season + I(season^2) + ask_1k,
                  data = sharks)
summary(poly_season)$coefficients
```

`I()` tells R to calculate the square. Without `I()`, R would read `^` as
part of the formula syntax and not as a square. The fitted coefficients are $\hat\beta_1 \approx -1.713$ for
`season` and $\hat\beta_2 \approx 0.051$ for `I(season^2)`. Together they
describe a steep decline in the early seasons that becomes flatter season by
season.

::: {#exm-i06-quadratic-change}
## Worked example: the fitted change from Season 5 to Season 6

[Shark Tank data]{.shark-tag} Holding requested amount fixed, the fitted change from season $x$ to season
$x+1$ is $\hat\beta_1+\hat\beta_2(2x+1)$. From Season 5 to Season 6, $x = 5$:

$$
\hat\beta_1+\hat\beta_2(2\times5+1)
=-1.713+0.051\times11
=-1.713+0.561
\approx-1.15
\text{ percentage points.}
$$

Reading $\hat\beta_1=-1.713$ alone would make the drop about 0.56 points
too large, which is about half of the fitted drop of 1.15 points. The number
$-1.713$ is not a constant slope. It describes the slope at $x=0$, a season
that does not exist. Check the calculation in R with the stored coefficients:

```{r}
#| label: i06-quadratic-check
b <- coef(poly_season)
unname(b["season"] + b["I(season^2)"] * (2 * 5 + 1))
```
:::

### Is the curvature worth having? The partial F

[Shark Tank data]{.shark-tag} The quadratic model and the linear model are nested, so the partial F-test
of Lecture 5 applies to the one added term:

```{r}
#| label: i06-partial-f
anova(linear_season, poly_season)
```

The partial F-test for adding `I(season^2)` gives $F \approx 20.1$ with a
p-value below 0.001. The data show a curve. But the gain in fit is small:
adjusted $R^2$ rises from about 0.186 to about 0.196, one percentage point
more of the variation in the equity offered. So also look at the shape of
the curve in @fig-i06-polynomial:

```{r}
#| label: fig-i06-polynomial
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Season means of the equity offered (points) and the fitted quadratic curve at the average requested amount. The curve becomes flatter towards the last observed seasons. Beyond Season 16 (shaded) it would turn upward, but there are no data in that region."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
poly_season <- lm(equity_offered_pct ~ season + I(season^2) + ask_1k,
                  data = sharks)
season_means <- aggregate(equity_offered_pct ~ season, data = sharks, mean)
curve_data <- data.frame(season = seq(1, 16, length.out = 200),
                         ask_1k = mean(sharks$ask_1k))
curve_data$fit <- predict(poly_season, newdata = curve_data)

ggplot(season_means, aes(x = season, y = equity_offered_pct)) +
  annotate("rect", xmin = 16, xmax = 17.6, ymin = -Inf, ymax = Inf,
           fill = "grey75", alpha = 0.25) +
  geom_vline(xintercept = 16, colour = "grey60",
             linetype = "dashed", linewidth = 0.8) +
  geom_line(data = curve_data, aes(x = season, y = fit),
            colour = "#D55E00", linewidth = 1.1) +
  geom_point(size = 2.6, colour = "#009E73") +
  annotate("text", x = 16.8, y = max(season_means$equity_offered_pct),
           label = "no data beyond\nseason 16", colour = "#5B6976",
           size = 3, lineheight = 0.95) +
  scale_x_continuous(breaks = seq(1, 16, by = 3)) +
  coord_cartesian(xlim = c(1, 17.6)) +
  labs(x = "Season", y = "Initial equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The points in @fig-i06-polynomial are the plain means of each season. The
curve keeps the requested amount fixed at its sample mean. To extrapolate means to
use a fitted model outside the range of the data. Do not extrapolate the
quadratic beyond Seasons 1--16. The fitted curve reaches its lowest point
just after Season 16 and would rise steeply afterwards. This rise comes from
the shape of a quadratic, not from any data, so it is not a forecast.

::: {.caveat-box}
<span class="label">What flexible terms can and cannot repair</span>

- Interactions and curves make the model more flexible. They cannot fix
  problems in how the data were produced.
- Every added term costs precision. Add a term because your question needs
  it, not because it turns out to be significant.
- You can trust a curve only inside the range of the data.

**What you can do with it:** ask whether a relationship could differ between
groups or follow a curve, add that one term, graph the fitted lines, and report
combined results.
:::

### Answering the driving question

**The same trend for every kind of team?** The estimated difference in
slopes is $-0.061$ percentage points per season (p $\approx$ 0.52). The data
show no difference between all-men teams and teams with women. They also do
not prove that the trends are equal, because the estimate is imprecise.

**A straight line?** Not quite. The quadratic term is significant (partial
$F \approx 20.1$, p $<$ 0.001). The fitted decline becomes flatter: from
about $-1.56$ points per season between Seasons 1 and 2 to about $-0.64$
between Seasons 10 and 11. Use the curve only inside Seasons 1--16.

The founder's questions at the start of the lecture get the same two kinds
of answer.
Whether advertising works equally well in every week is a question about an
interaction. In the invented example of the Try it task, advertising does not
work equally well: the slope is 0.10 orders per euro in ordinary weeks and 0.22 in exam weeks.
Whether a trend goes on in a straight line is a question about a curve. The
season trend of the equity offered does not go on in a straight line: the
fitted drop per season
shrinks from about 1.56 to about 0.64 points, and beyond Season 16 the data
say nothing.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You suspect that advertising works better in exam weeks. The model with an interaction gives an advertising slope of 0.10 orders per euro in ordinary weeks and 0.10 + 0.12 = 0.22 in exam weeks. After a price increase, each order now brings 6 euros of margin. So 100 euros of advertising goes with 10 more orders and 60 euros of margin in an ordinary week, which is less than the advertising costs. In an exam week it goes with 22 more orders and 132 euros of margin. The decision: advertise in the exam weeks only. One check first: the interaction of 0.12 has its own interval. If that interval includes zero, the data do not yet show that the two kinds of weeks differ. Then keep spreading the advertising budget over both kinds of weeks for one more study period before you decide.
:::

**Next:** Tutorial 2 practises Lectures 4--6 with exam-style questions. It
ends with one full analysis, from a research question to a conclusion of 150
words. Your thesis asks for this kind of analysis too.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main ideas. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly, the course AI
assistant.

### Recall questions

1. What does an interaction coefficient measure, and what is its unit in the
   model with season and the `women` dummy?
2. Why centre a predictor before you use it in an interaction, and what can
   centring never change?
3. State the hierarchy principle.
4. Why is a quadratic model still a linear regression model?
5. Why should you graph a quadratic curve, and why should you not
   extrapolate it?
6. In the invented franchise model, `marketing` is 1.745 and
   `marketing:stores` is 0.147, with sales and marketing in €1,000s. What
   is the slope of marketing in a region with 6 stores, and why does 1.745
   alone not describe any region?
7. A researcher finds that training is related to productivity, and that
   the relationship is stronger in large firms. A second researcher finds
   that training raises skills and that skills raise productivity. Which
   finding is about a moderator and which about a mediator, and how does
   each appear in a regression?

::: {.callout-note collapse="true"}
## Check your answers

1. How the slope of one predictor changes with the value of the other
   predictor. Here its
   unit is equity percentage points per season, as a difference between the
   two groups.
2. So that the intercept and the dummy coefficient refer to a useful value,
   such as the sample mean, and not to zero. Centring never changes the
   fitted values, the residuals, $R^2$, or the interaction coefficient.
3. Keep $X$ and $D$ in the model whenever their interaction $XD$ is in it,
   and keep $x$ whenever $x^2$ is in it.
4. It is linear in the unknown coefficients. The squared column is one more
   predictor built from the data.
5. No single coefficient is the slope, so only fitted values or the curve
   show what the model says. Outside the observed range the curve can turn
   sharply, and no data support that turn.
6. $1.745+0.147\times6\approx2.63$: in a region with 6 stores, €1,000 more
   marketing is associated with about €2,630 more monthly sales on average.
   The number 1.745 is the slope at 0 stores. No region has 0 stores, so it
   is a conditional effect at a value outside the data.
7. Firm size is a moderator: it changes the slope of training. It appears
   as an interaction term of training and firm size. Skills are a mediator:
   they lie on the path from training to productivity. Adding skills to the
   model shrinks the training coefficient to the direct part, so leave
   skills out if you want the total effect.
:::

### Centring verification

[Shark Tank data]{.shark-tag} Fit the interaction model with and without centring season. Check that the
fitted values and the interaction coefficient are unchanged, and report
which coefficients do change.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: i06-centering-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)

uncentred <- lm(equity_offered_pct ~ season * women + ask_1k,
                 data = known_gender)
centred <- lm(equity_offered_pct ~ season_c * women + ask_1k,
               data = known_gender)
all.equal(fitted(uncentred), fitted(centred))
c(uncentred = unname(coef(uncentred)["season:women"]),
  centred = unname(coef(centred)["season_c:women"]))
c(women_uncentred = unname(coef(uncentred)["women"]),
  women_centred = unname(coef(centred)["women"]))
```

The fitted values agree, and both interaction estimates are about $-0.061$.
The `women` coefficient changes from about $1.35$ (a comparison at Season 0,
which does not exist) to about $0.81$ (a comparison at the average season,
about 8.8). The intercept changes as well. Centring moved the place where
we compare the two groups. It did not move the lines.
:::

### Finish a half-worked example: a quadratic change

[Shark Tank data]{.shark-tag} Read the worked calculation for Season 5 to Season 6 again
(@exm-i06-quadratic-change). Then calculate the fitted change from **Season
10 to Season 11** in the same model.

*Step 1 (given).* The formula is $\hat\beta_1+\hat\beta_2(2x+1)$ with $x=10$,
so you multiply $\hat\beta_2$ by $21$.

*Step 2 (you).* Insert $\hat\beta_1\approx-1.713$ and $\hat\beta_2\approx0.051$.

*Step 3 (you).* Compare the result with the Season 5 to 6 change and state
in one sentence what the comparison shows about the trend.

::: {.callout-note collapse="true"}
## Check your answer

$-1.713+0.051\times21\approx-1.713+1.071\approx-0.64$ percentage points
(with the unrounded coefficients: $\hat\beta_1+21\hat\beta_2=-0.642$). The
drop per season has become smaller, from about $-1.15$ (Seasons 5--6) to
about $-0.64$ (Seasons 10--11). The decline becomes flatter in later
seasons. This is what a positive $\hat\beta_2$ means.
:::

Now do the next tasks on your own, with new coefficients from an invented
example. A model of monthly revenue (€1,000s) on the age of a shop $x$
(years) reports $\widehat Y = 40 - 2.00x + 0.060x^2$.

1. Calculate the fitted change from age 3 to age 4 and from age 8 to age 9.
2. A colleague reports "revenue falls €2,000 per year of age." Correct the
   sentence.
3. In $\widehat Y=10+2X+3D-0.5XD$, write the lines of both groups and
   interpret the interaction coefficient.

::: {.callout-note collapse="true"}
## Check your answers

1. Age 3 to 4: $-2.00+0.060(2\times3+1)=-2.00+0.42=-1.58$, about
   $-\text{€}1{,}580$. Age 8 to 9: $-2.00+0.060\times17=-0.98$, about $-\text{€}980$.
2. $-2.00$ is not the fitted change per year. With $x^2$ in the model the
   change depends on age: about $-1.58$ from age 3 to 4 and about $-0.98$
   from age 8 to 9. The decline becomes weaker as shops get older.
3. For $D=0$: $\widehat Y=10+2X$. For $D=1$: $\widehat Y=13+1.5X$. The
   interaction coefficient $-0.5$ says that the slope of the $D=1$ group is
   0.5 outcome units per unit of $X$ smaller than the slope of the $D=0$
   group.
:::

### Practice: all steps together

[Shark Tank data]{.shark-tag} In Lecture 4 you wrote a paragraph that connects the diagnostics of Lecture
2 to the three-predictor model of the equity offered. Add one argument to
that paragraph. Propose either an interaction or a quadratic term for that
model. State the question about the business or the data that motivates the
term, and say how you would test whether the term belongs in the model.

::: {.callout-note collapse="true"}
## Check a model answer

A good answer: "The decline over the seasons need not be a straight line.
The show was new in the early seasons and pitches became more professional
later, so the trend may become flatter. I would add `I(season^2)`. I add the
term because of this question about the shape, not because I am searching
for a significant result. I would test it with a partial F-test against the
nested linear model on the same observations (here $F\approx20.1$,
p $<$ 0.001). I would also check that adjusted $R^2$ rises and graph the
fitted curve inside Seasons 1--16. Finally, I would report fitted changes
from season to season, not the `season` coefficient alone." An answer with an interaction is equally
good if it names the question and the nested test. An example is the
interaction of season with the `women` dummy, which asks whether the trend
differs between the groups. Its test gives $F\approx0.41$ and
p $\approx$ 0.52, so the term is not kept.
:::

### Review: one partial-F question from Lecture 5

[Shark Tank data]{.shark-tag} Adding the two gender dummies to the model of equity on season and requested amount reduced
the residual sum of squares by 436.25 using 2 degrees of freedom, and the
complete model left 82,792 on 1,429 degrees of freedom. Compute the partial
F statistic and state the 5% conclusion.

::: {.callout-note collapse="true"}
## Check your answer

$F=(436.25/2)/(82{,}792/1{,}429)=218.1/57.9\approx3.76$, with
p $\approx$ 0.023. At 5%, the two category terms are jointly significant in that
model, even though the t-test of the mixed dummy alone is far from
significant. Joint tests and individual tests answer different questions.
:::

### Worked exam interpretation

**Exam-style question.** An invented example. A model of the weekly
revenue of market stalls (in €100s) on opening hours per week, a dummy for
indoor stalls, and their interaction reports `hours 1.20`,
`indoor 2.00`, `hours:indoor -0.50`. A student writes: "One more opening
hour raises weekly revenue by €120 for all stalls." Correct the
sentence.

::: {.callout-note collapse="true"}
## Check a model answer

Two corrections. First, with the interaction in the model, 1.20 is the
fitted slope of opening hours only for stalls that are not indoor
(`indoor = 0`). For indoor stalls the slope is $1.20-0.50=0.70$, that is,
€70 of weekly revenue per opening hour. There is no single slope for all
stalls. Second, "raises" claims causation. These are fitted associations
among the sampled stalls, holding the other included terms fixed. The data
do not come from an experiment, so they do not show an effect of opening
hours.
:::

### Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant: "The interaction has a coefficient of -0.061 and a p-value of 0.52. What do I conclude?" It answered:

> "The p-value proves that the season trend is identical for both groups. And the season coefficient of -0.80 is the season trend for all teams."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

Two errors. A large p-value means the data do not show a difference. It does not prove that the trends are the same, because the estimate is imprecise. And with an interaction in the model, -0.80 is the trend for the group coded 0, the all-men teams. For teams with women it is -0.80 - 0.06 = -0.86.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Name one
relationship in its data whose slope could depend on something else. For
example, the response to price may differ by customer segment, the result of
a campaign may differ by channel, or growth may become flatter as the
business gets larger. Then write two sentences. First, state the interaction
or quadratic term you would add and the question behind it. Second, say what
a result that is not significant would tell you, and what it would not tell
you. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "I would add an interaction of advertising spend with exam week, because my question is whether advertising works only when students are on campus. An interaction that is not significant would mean no evidence that the slopes differ. It would not prove that exam weeks make no difference."
:::

You can discuss this task, and any other task in this chapter, with
Tilly for QM Intermediate, the course AI assistant. Ask it to read your two sentences and to
check whether you treated "no evidence of a difference" as "proof of no
difference".
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask Tilly for
QM Intermediate, the course AI assistant. For example: "Give me an extra
exercise for Intermediate Lecture 6. Show the solution only after I send my
answer."

[Continue to Tutorial 2: Multiple regression and extensions](t02-exam-prep.qmd).

[Code sheet: Lecture 6 commands and formulas](code-sheet.qmd#lecture-6-interactions-and-polynomial-terms).

For a more formal treatment, see James et al., *An Introduction to
Statistical Learning*, 2nd ed., Section 3.3.2 and Sections 3.6.4--3.6.6,
and
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 21.

:::
