# Mock Exam: QM Intermediate {#intermediate-mock-exam}

```{r}
#| label: imock-setup
#| include: false
# Chunk options only.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

This mock exam covers all six lectures of the course. It has **100
points**, so your test score is

$$
\text{test score} = \frac{\text{points earned}}{10},
$$

on a scale from 1.00 to 10.00, with two decimals. A score below 1.00 counts
as 1.00. The lecturer enters this test score in Osiris. Osiris then rounds it to a
whole or half grade for your final course grade. A score from 5.50 to 5.74
becomes 6.0, and a score from 5.25 to 5.49 becomes 5.0. For example, 72
points give a test score of 7.20 and a final grade of 7.0. You pass with a
final grade of 6.0 or higher, that is, with 55 points or more.

You take the Intermediate exam in TestVision. It lasts 120 minutes. You may
use RStudio, Excel, the calculator in TestVision, a simple pocket
calculator, and scrap paper. The practice
exam in TestVision appears in your own TestVision account once the
lecturer publishes it. You log in with your university account. There is no separate
link. General information about TestVision for students is on the
[university page about TestVision](https://www.tilburguniversity.edu/students/studying/exams/e-assessment/testvision).

You receive the [QM Intermediate Code Sheet](code-sheet.qmd) at
the exam. Keep it open while you work on this mock exam.

Use **120 minutes** as the practice time limit, as in the exam.

## How this mock exam works

This mock exam gives an impression of what an exam in this course can look
like. You download a data file, you run R yourself, and you type your answers.

All parts are open questions. You type a number with your calculation, or
one to three sentences. An answer without a calculation or an explanation
gets no points, unless the part asks only to report or to state something.

You do not have to simulate data, and you do not have to reproduce a
simulation. The practice data are invented. The code that made them is in a
collapsed box at the end of this page. It is there only so that anyone can
check how we made the data. It is not part of the exam.

A part that starts with "In R:" needs code that you write and run yourself.
Its answer is not printed in the question. For these parts, type the R
command that you used and the numbers that the part asks for. You can
answer all other parts from your own results, from printed output, or with a
short calculation. You may use R as a calculator everywhere.

Question 3 shows an R output of a colleague, and Question 4 shows a
regression table in the layout of a research paper. You interpret them.
Question 6 shows an answer of an AI assistant. You assess it.

All tests in this course are two-sided. Use a 5% significance level. Some parts that build
on an "In R:" part print stand-in values. Use them if your code does not run.

## Dataset used throughout

The practice data are invented. Treat them as a simple random sample of 320
young ventures from a business registry. One row is one venture. The data
are observational: none of the predictors was assigned by chance.

- `growth_pct`: growth of annual sales compared with the year before, in
  percent. A value of 12 means that sales were 12% higher than in the year
  before
- `marketing_10k`: annual marketing spending, in €10,000s. The values run
  from 0.25 (€2,500) to 7.99 (€79,900)
- `experience_years`: prior industry experience of the lead founder, in
  years, from 0 to 18
- `team_size`: number of current team members, from 1 to 10
- `digital`: `No` for a mainly physical offering and `Yes` for a mainly
  digital offering. `No` is the reference category
- `spend_c`: `marketing_10k - 4`, marketing spending centred at €40,000.
  This is close to the sample mean of about €40,400.

The population is the set of ventures in this registry. The data do not
measure later survival or market conditions.

[Download `qm-intermediate-mock-exam.csv`](../data/qm-intermediate-mock-exam.csv),
save it in the folder `data` of your R project, and load it with the code
below. If you save the file directly in your project folder, write
`"qm-intermediate-mock-exam.csv"` as the file name.

```{r}
#| label: imock-load
ventures <- read.csv("data/qm-intermediate-mock-exam.csv")
ventures$digital <- factor(ventures$digital, levels = c("No", "Yes"))
```

## Question 1: Simple regression (13 points)

A founder asks whether ventures that spend more on marketing grow faster.
The outcome is `growth_pct` and the predictor is `marketing_10k`.

1. In R: load the data and check that the file has 320 rows. Report the
   mean and the largest value of `growth_pct`. **(2 points: 1 + 1)**
2. In R: fit the simple regression of `growth_pct` on `marketing_10k`.
   Report the intercept and the slope. **(2 points: 1 + 1)**

   Stand-in values, if your code does not run: intercept 6.6 and slope 1.4.
   Use them in parts 3 and 4.
3. Interpret the slope in one sentence. **(4 points)**
4. In R: use `predict()` to calculate the fitted growth of a venture that
   spends €50,000 on marketing. The actual sales growth of this venture is 11%.
   Calculate its residual. State what the sign of the residual means.
   **(5 points: 3 + 1 + 1)**

::: {.callout-note collapse="true"}
## Answer indication

```{r}
#| label: imock-q1-fit
dim(ventures)
summary(ventures)
mean(ventures$growth_pct)
simple_model <- lm(growth_pct ~ marketing_10k, data = ventures)
coef(simple_model)
```

1. The mean sales growth is 12.172%. The largest sales growth is 31.940%.
   Both numbers are in the output of `summary(ventures)`.
2. Intercept 6.598 and slope 1.378. The fitted line is
   $\widehat{growth\_pct}=6.598+1.378\,marketing\_10k$.
3. Among the 320 sampled ventures, a venture that spends €10,000 more on
   marketing has on average 1.38 percentage points higher fitted sales
   growth. The sentence names the size and direction (1.38 higher), the
   step of the predictor (€10,000 more marketing spending), the unit of the
   outcome (percentage points of sales growth), and is worded as an
   association, without "causes", "leads to", or "will raise". With the
   stand-in value the size is 1.4.
4. €50,000 is `marketing_10k = 5`. The command
   `predict(simple_model, newdata = data.frame(marketing_10k = 5))` gives a
   fitted growth of 13.490%. The calculation by hand gives the same value:
   $6.598+1.378\times5=13.49$%. The residual is $11-13.49=-2.49$
   percentage points. The sign is negative: this venture grew less than
   the fitted growth for ventures with the same marketing spending. With
   the stand-in values the fitted growth is $6.6+1.4\times5=13.60$% and the
   residual is $-2.60$ percentage points.

```{r}
#| label: imock-q1-case
new_venture <- data.frame(marketing_10k = 5)
fit_50k <- unname(predict(simple_model, newdata = new_venture))
c(fitted = fit_50k, residual = 11 - fit_50k)
c(fitted_stand_in = 6.6 + 1.4 * 5, residual_stand_in = 11 - (6.6 + 1.4 * 5))
```
:::

## Question 2: Assumptions and diagnostics (15 points)

Check the model of Question 1, `growth_pct ~ marketing_10k`. Parts 1 and 4
ask for two plots. If your plots do not work, open the stand-in figure
below part 4 and answer from it.

1. In R: make the residual-versus-fitted plot of the model. Describe in one
   sentence the pattern that you see. State the assumption that this pattern
   challenges. **(5 points: 3 + 2)**
2. In R: count the ventures with a standardised residual beyond $\pm3$.
   Then count the ventures with a Cook's distance above $4/n$. Report both
   counts. **(2 points: 1 + 1)**
3. A colleague wants to delete all ventures that part 2 flags, "to get a
   cleaner model". Respond to her plan in two or three sentences. State the
   rule of Lecture 2 for flagged observations, and say what you would do
   with these ventures. **(4 points: 1 + 3)**
4. In R: make the normal Q--Q plot of the standardised residuals. Describe
   in one sentence how the points lie relative to the line. Then explain in
   one sentence whether a departure from normality is a serious problem for
   this model with 320 ventures. **(4 points: 1 + 3)**

::: {.callout-note collapse="true"}
## Stand-in figure for parts 1 and 4 (open it only if your own plots do not work)

```{r}
#| label: fig-imock-diagnostics
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Residual-versus-fitted plot (left) and normal Q-Q plot (right) for the simple model of sales growth (growth_pct) on marketing spending (marketing_10k) in the practice data. The dashed line in the left panel marks a residual of zero."
library(ggplot2)
library(patchwork)
simple_model <- lm(growth_pct ~ marketing_10k, data = ventures)

resid_df <- data.frame(
  fitted = fitted(simple_model),
  residual = residuals(simple_model)
)
qq_df <- data.frame(rs = rstandard(simple_model))
theme_mock <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )

p_resid <- ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  labs(x = "Fitted growth (%)",
       y = "Residual (percentage points)") +
  theme_mock

p_qq <- ggplot(qq_df, aes(sample = rs)) +
  geom_qq_line(colour = "#D55E00", linewidth = 0.9) +
  geom_qq(colour = "#009E73", alpha = 0.45, size = 1.6) +
  labs(x = "Normal theoretical quantile",
       y = "Standardised residual") +
  theme_mock

p_resid + p_qq
```
:::

::: {.callout-note collapse="true"}
## Answer indication

The two plots with the commands of the code sheet:

```{r}
#| label: imock-q2-plots
#| fig-height: 3.8
simple_model <- lm(growth_pct ~ marketing_10k, data = ventures)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(simple_model), residuals(simple_model),
     xlab = "Fitted growth (%)", ylab = "Residual (percentage points)")
abline(h = 0, lty = 2)
qqnorm(rstandard(simple_model), main = "",
       xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(simple_model))
par(old_par)
```

1. The vertical spread of the residuals grows from left to right, that is,
   with the fitted growth. The points form a fan. This challenges the
   assumption of constant variance of the errors (homoskedasticity). No
   clear curve is visible in this plot, so linearity is not the assumption
   that this pattern challenges. The code below confirms what you see in
   the plot.
   The standard deviation of the residuals is 3.40 percentage points in the
   third of the ventures with the lowest fitted growth and 5.42 in the third
   with the highest.

```{r}
#| label: imock-q2-spread
thirds <- cut(
  fitted(simple_model),
  breaks = quantile(fitted(simple_model), c(0, 1/3, 2/3, 1)),
  include.lowest = TRUE, labels = c("lowest", "middle", "highest")
)
tapply(residuals(simple_model), thirds, sd)
```

2. One venture has a standardised residual beyond $\pm3$. 19 ventures
   have a Cook's distance above $4/n=4/320=0.0125$.

```{r}
#| label: imock-q2-flags
sum(abs(rstandard(simple_model)) > 3)
sum(cooks.distance(simple_model) > 4 / nobs(simple_model))
```

3. The rule is: investigate, report, never silently delete. Investigate
   means to look at the flagged ventures in the data and to ask whether a
   value is an error or whether the venture is of a different kind. Report
   means to fit the model with and without these ventures, to show both
   results, and to say which one you use and why.
4. The points follow the line in the middle. At the upper end they lie
   above the line, and near the lower end slightly below it. This is not a
   serious problem here. Normal errors matter in small samples. With 320
   ventures the tests and intervals still hold approximately. The fan of
   part 1 is the bigger problem.
:::

## Question 3: Inference and fit (10 points)

A colleague drew 200 of the 320 ventures at random and fitted the model
below to these 200 ventures. You do not know which ventures she drew, so you
cannot run this model yourself. Answer from her output. The formulas are on
the code sheet.

```text
Call:
lm(formula = growth_pct ~ marketing_10k + experience_years + 
    digital, data = ventures_200)

Residuals:
     Min       1Q   Median       3Q      Max 
-11.2970  -2.7338  -0.0992   2.6132  12.3786 

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)       2.73159    0.78897   3.462 0.000658 ***
marketing_10k     1.35473    0.12868  10.528  < 2e-16 ***
experience_years  0.28585    0.05277   5.417 1.77e-07 ***
digitalYes        3.17544    0.58246   5.452 1.49e-07 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 4.049 on 196 degrees of freedom
Multiple R-squared:  0.4803,	Adjusted R-squared:  0.4723 
F-statistic: 60.38 on 3 and 196 DF,  p-value: < 2.2e-16
```

For this model, SSE = 3212.85 and SST = 6181.93. The critical value for a
95% confidence interval and a 5% test with 196 degrees of freedom is
$t_{0.975,\,196}=1.972$. In R, `qt(0.975, 196)` gives it.

1. &#40;a) Show with a calculation how the Multiple R-squared follows from SSE
   and SST. (b) State in one sentence what this number says about the model
   for these 200 ventures. **(4 points: 2 + 2)**
2. &#40;a) Calculate the 95% confidence interval for the coefficient of
   `marketing_10k`. Write the five parts of the
   [exam procedure of Lecture 3](i03-inference-fit-ftest.qmd#i03-exam-procedure). (b)
   State whether the 5% test rejects a coefficient of zero, and explain how the
   interval shows this. **(6 points: 4 + 2)**

::: {.callout-note collapse="true"}
## Answer indication

The code below is the code of the colleague. We show it so that you can
check every printed number.

```{r}
#| label: imock-colleague
set.seed(2026)
rows_200 <- sort(sample(nrow(ventures), 200))
ventures_200 <- ventures[rows_200, ]
model_200 <- lm(
  growth_pct ~ marketing_10k + experience_years + digital,
  data = ventures_200
)
summary(model_200)
SSE_200 <- sum(residuals(model_200)^2)
SST_200 <- sum((ventures_200$growth_pct - mean(ventures_200$growth_pct))^2)
c(SSE = SSE_200, SST = SST_200, R2 = 1 - SSE_200 / SST_200)
qt(0.975, df.residual(model_200))
confint(model_200)["marketing_10k", ]
```

1. &#40;a) $R^2=1-\text{SSE}/\text{SST}=1-3212.85/6181.93=0.48$. (b) Among
   these 200 ventures, the model explains about 48% of the variation in
   sales growth. It does not mean that 48% of the predictions are correct.
2. &#40;a) The five parts of the interval:
   1. **Formula.** $\hat\beta_j\pm t_{0.975,\,df}\,\operatorname{SE}(\hat\beta_j)$,
      with $df=n-k-1=200-3-1=196$.
   2. **Given values.** $\hat\beta_j=1.35473$,
      $\operatorname{SE}(\hat\beta_j)=0.12868$, and
      $t_{0.975,\,196}=1.972$.
   3. **Calculations.** The margin of error is $1.972\times0.12868=0.254$.
   4. **Interval.** $(1.355-0.254,\ 1.355+0.254)=(1.10,\ 1.61)$ percentage
      points per €10,000.
   5. **Conclusion.** See (b).

   &#40;b) Zero lies outside the interval. The interval contains the values
   of the coefficient that the two-sided 5% test does not reject, so the
   test rejects a coefficient of zero. At the 5% level we can conclude the
   following for all ventures in the registry: at the same experience and
   offering type, more marketing spending goes with higher sales growth.
:::

## Question 4: Multiple regression and a dummy predictor (23 points)

The table below has the layout of a regression table in a research paper.
It shows three models of sales growth (`growth_pct`), fitted to all 320
ventures of the practice data. You do not need to run these models for
parts 1 to 3.

::: {.reg-table}
| | (1) | (2) | (3) |
|:---|---:|---:|---:|
| Marketing spending (€10,000s) | 1.340\*\*\* | 1.338\*\*\* | 1.341\*\*\* |
| | (0.106) | (0.100) | (0.100) |
| Digital offering (1 = yes) | 2.969\*\*\* | 3.016\*\*\* | 3.084\*\*\* |
| | (0.488) | (0.462) | (0.463) |
| Founder experience (years) | | 0.255\*\*\* | 0.258\*\*\* |
| | | (0.042) | (0.041) |
| Team size (members) | | | 0.168 |
| | | | (0.103) |
| Intercept | 5.418\*\*\* | 3.206\*\*\* | 2.449\*\* |
| | (0.528) | (0.617) | (0.771) |
| Observations | 320 | 320 | 320 |
| $R^2$ | 0.395 | 0.459 | 0.464 |
| Adjusted $R^2$ | 0.391 | 0.454 | 0.457 |
:::

Outcome: sales growth (%). Ordinary least squares. Standard errors in
parentheses. All three columns use the same 320 ventures. The reference
group of the dummy is ventures with a mainly physical offering.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001.

The critical value for a 95% confidence interval and a 5% test is
$t_{0.975,\,df}=1.967$ in all three columns. For column (3), with
$df=320-4-1=315$, R gives it with `qt(0.975, 315)`.

1. Team size has no star in column (3). Calculate its t value. Then state
   what the missing star means and what it does not mean.
   **(4 points: 1 + 2 + 1)**
2. Venture B spends €30,000 more on marketing than venture A. Its lead
   founder has 4 more years of experience. B has a digital offering and A a
   physical one. Both have the same team size. Calculate the fitted growth
   of B minus the fitted growth of A with column (3). **(5 points)**
3. Founder experience has three stars in column (2). Still, adding it moves
   the marketing coefficient only from 1.340 to 1.338. Explain in one or two
   sentences what this says about experience and marketing spending in this
   sample. Then state one problem that the stable coefficient does not
   exclude. **(5 points: 3 + 2)**
4. In R: test whether founder experience and team size together improve the
   model of column (1). Fit the models of columns (1) and (3) and compare
   them with `anova()`. Report the F statistic. Then show how F follows from
   the two residual sums of squares in your output.
   **(4 points: 1 + 3)**

   Stand-in values, if your code does not run: the residual sum of squares
   is about 5950 on 317 degrees of freedom for column (1). It is about
   5270 on 315 degrees of freedom for column (3). Calculate F from them.
   For part 5, use a p-value below 0.001.
5. State the null hypothesis of the test in part 4 and your conclusion at
   the 5% level. Then explain in one sentence how this conclusion and the
   missing star of team size can both be true. **(5 points: 1 + 2 + 2)**

::: {.callout-note collapse="true"}
## Answer indication

We typed the table from the three outputs below.

```{r}
#| label: imock-q4-table
col_1 <- lm(growth_pct ~ marketing_10k + digital, data = ventures)
col_2 <- lm(growth_pct ~ marketing_10k + digital + experience_years,
            data = ventures)
col_3 <- lm(
  growth_pct ~ marketing_10k + digital + experience_years + team_size,
  data = ventures
)
summary(col_1)
summary(col_2)
summary(col_3)
qt(0.975, df.residual(col_3))
```

1. The five steps of a test for one coefficient give the answer.
   1. **Hypotheses.** $H_0:\beta_{team}=0$ versus $H_1:\beta_{team}\ne0$,
      where $\beta_{team}$ is the coefficient of team size in the model of
      column (3) for all ventures of the registry. The significance level is
      $\alpha=0.05$.
   2. **Test statistic.** $T=\hat\beta_{team}/\operatorname{SE}(\hat\beta_{team})$,
      which has the $t_{315}$ distribution if $H_0$ is true.
   3. **Critical value and rejection region.** $t_{0.975,\,315}=1.967$.
      Reject $H_0$ if $t\le-1.967$ or $t\ge1.967$.
   4. **Observed value.** The numbers in parentheses are standard errors, so
      $t=0.168/0.103=1.63$.
   5. **Conclusion.** 1.63 lies between $-1.967$ and 1.967, so do not
      reject $H_0$. The p-value is above 0.05, and so there is no star. At
      the 5% level we cannot conclude that team size goes with growth at
      fixed marketing spending, offering type, and experience.

   So $t=1.63$, which is below 1.967 (the p-value is above 0.05). What the
   missing star does not mean: it does not show that the coefficient is
   zero. The 95% interval, $0.168\pm1.967\times0.103$, runs from $-0.035$
   to 0.371 percentage points per team member. It contains zero and also
   clearly positive values.
2. Marketing: €30,000 is 3 steps, so $3\times1.341=4.023$. Experience:
   $4\times0.258=1.032$. Digital offering: 3.084. Sum:
   $4.023+1.032+3.084=8.14$ percentage points.

```{r}
#| label: imock-q4-compare
b3 <- coef(col_3)
unname(3 * b3["marketing_10k"] + 4 * b3["experience_years"] + b3["digitalYes"])
confint(col_3)
cor(ventures$marketing_10k, ventures$experience_years)
```

3. A coefficient changes when the added predictor is related to the outcome
   and also to the predictor of interest. Experience is related to growth,
   but in this sample it is almost unrelated to marketing spending. So
   holding it fixed changes the marketing comparison very little. The
   correlation of the two variables in the data is 0.001. The stable
   coefficient does not exclude a variable that is not in the data and
   that is related to both marketing spending and growth, for example the
   quality of the product.
4. $F=20.216$. From the output, with the partial F statistic, $q=2$, and
   $df=315$:
   $F=\dfrac{(5949.3-5272.6)/2}{5272.6/315}=\dfrac{338.35}{16.74}=20.21$.
   With the stand-in values, $F=\dfrac{(5950-5270)/2}{5270/315}=20.32$.

```{r}
#| label: imock-q4-anova
anova(col_1, col_3)
c(F_by_hand = ((deviance(col_1) - deviance(col_3)) / 2) /
    (deviance(col_3) / df.residual(col_3)),
  F_stand_in = ((5950 - 5270) / 2) / (5270 / 315))
```

5. $H_0$: the coefficients of founder experience and team size are both
   zero. The p-value in the `anova()` output is 5.491e-09, that is
   0.000000005, far below 0.05. So reject $H_0$. At the 5% level we can
   conclude that the two predictors together improve the model of column
   (1). The F-test asks about the two terms together, and the t-test asks
   about team size alone. The joint result comes from experience. A set of
   terms can be jointly useful while one of its terms is not significant.
:::

## Question 5: Interactions and curvature (19 points)

Parts 1 to 3 ask whether marketing spending goes with more growth for
digital ventures than for physical ventures. Parts 4 to 6 use a model in
which the relation between marketing spending and growth can be curved.

1. In R: fit a model of `growth_pct` with `spend_c`, `digital`, the
   interaction of the two, and `experience_years`. Report the coefficients
   of `spend_c`, `digitalYes`, and the interaction.
   **(3 points: 1 + 1 + 1)**

   Stand-in values, if your code does not run: 1.0 for `spend_c`, 3.0 for
   `digitalYes`, and 0.7 for the interaction. Use them in parts 2 and 3.
2. State the marketing slope for physical ventures. Calculate the
   marketing slope for digital ventures. **(3 points: 1 + 2)**
3. Interpret the interaction coefficient in one sentence. **(4 points)**
4. In R: run the model below, which has no interaction and adds the square
   of marketing spending. Report the coefficients of `marketing_10k` and
   `I(marketing_10k^2)`. **(2 points: 1 + 1)**

   ```r
   curve_model <- lm(
     growth_pct ~ marketing_10k + I(marketing_10k^2) + digital +
       experience_years,
     data = ventures
   )
   ```

   Stand-in values, if your code does not run: 2.2 for `marketing_10k` and
   $-0.10$ for `I(marketing_10k^2)`. Use them in part 5.
5. Use the model of part 4. Calculate the fitted change in growth when
   marketing spending rises from €20,000 to €30,000, and when it rises from
   €60,000 to €70,000. Offering type and experience stay fixed. Then state
   in one sentence what the two results show together.
   **(4 points: 1 + 1 + 2)**
6. A founder plans to spend €150,000 on marketing. She wants to use the
   model of part 4 to calculate her fitted growth. Explain in one or two
   sentences why she should not do this. **(3 points)**

::: {.callout-note collapse="true"}
## Answer indication

```{r}
#| label: imock-q5-interaction
interaction_model <- lm(
  growth_pct ~ spend_c * digital + experience_years,
  data = ventures
)
summary(interaction_model)$coefficients
b5 <- coef(interaction_model)
c(slope_physical = unname(b5["spend_c"]),
  slope_digital = unname(b5["spend_c"] + b5["spend_c:digitalYes"]))
```

1. `spend_c` 1.044, `digitalYes` 2.967, interaction `spend_c:digitalYes`
   0.660.
2. Physical ventures (`digital = No`): 1.044 percentage points per €10,000.
   Digital ventures: $1.044+0.660=1.704$ percentage points per €10,000.
   With the stand-in values: 1.0 and 1.7.
3. Among the 320 sampled ventures and at the same founder experience, each
   additional €10,000 of marketing spending goes with 0.66 percentage points
   more sales growth for digital ventures than for physical ventures. The
   interaction is the difference between the two slopes. Its size has a
   unit (0.66 percentage points of growth) and a step (per €10,000). The
   sentence describes an association, without "causes" or "leads to". With
   the stand-in value the size is 0.7.

4. `marketing_10k` 2.179 and `I(marketing_10k^2)` $-0.105$.

```{r}
#| label: imock-q5-quadratic
curve_model <- lm(
  growth_pct ~ marketing_10k + I(marketing_10k^2) + digital +
    experience_years,
  data = ventures
)
bq <- coef(curve_model)
bq
c(from_2_to_3 = unname(bq["marketing_10k"] +
                         bq["I(marketing_10k^2)"] * (2 * 2 + 1)),
  from_6_to_7 = unname(bq["marketing_10k"] +
                         bq["I(marketing_10k^2)"] * (2 * 6 + 1)))
c(stand_in_2_to_3 = 2.2 - 0.10 * 5, stand_in_6_to_7 = 2.2 - 0.10 * 13)
unname(-bq["marketing_10k"] / (2 * bq["I(marketing_10k^2)"]))
```

5. The fitted change from $x$ to $x+1$ is
   $\hat\beta_1+\hat\beta_2(2x+1)$. From €20,000 to €30,000, $x=2$:
   $2.1794-0.1046\times5=1.66$ percentage points. From €60,000 to €70,000,
   $x=6$: $2.1794-0.1046\times13=0.82$ percentage points. With the
   coefficients rounded to three decimals the results are 1.65 and 0.81.
   These are also correct. With the stand-in values: 1.70 and 0.90.
   Together: the same extra €10,000 goes with about half as much extra
   growth at the higher level of spending. The fitted curve becomes
   flatter. The coefficient 2.179 alone does not give these changes,
   because the squared term also changes with spending.
6. €150,000 is far outside the range of the data. The largest spending in
   the sample is €79,900. No venture in the data supports the curve there.
   A quadratic curve does not stop curving. This one turns downward at
   about €104,000 of spending, and no data show such a turn. With the
   stand-in values the turn is at about €110,000.
:::

## Question 6: An answer of an AI assistant, and a decision (20 points)

Sara runs a young venture with a mainly digital offering. She gave column
(3) of the table in Question 4 to an AI assistant. Here is column (3) again:

```text
Outcome: sales growth (%), 320 ventures     Estimate  Std. Error
Marketing spending (€10,000s)                  1.341       0.100
Digital offering (1 = yes)                     3.084       0.463
Founder experience (years)                     0.258       0.041
Team size (members)                            0.168       0.103
Intercept                                      2.449       0.771
```

She asked: "What does this mean for my marketing budget?" The AI answered:

> "Good news. Your model controls for experience, team size, and offering
> type. So the marketing coefficient is a causal effect: if you spend
> €20,000 more on marketing, your sales growth will rise by 2.68 percentage
> points. Your venture is digital, and digital ventures grow 3.084% faster
> than the average venture. And the p-value of marketing is below 0.001, which proves that the assumptions of the
> model hold."

1. Assess the claim that the marketing coefficient is a causal effect.
   &#40;a) Explain whether the controls in the model justify this claim. (b)
   Write a sentence about €20,000 more marketing spending that the table
   does support. **(5 points: 3 + 2)**
2. Assess the sentence about digital ventures. Explain two errors in it,
   and write a correct sentence about the coefficient 3.084.
   **(4 points: 1 + 1 + 2)**
3. Assess the last sentence. (a) Explain whether a small p-value can show
   that the assumptions of the model hold. (b) State what the small p-value
   of marketing does show. **(4 points: 2 + 2)**

Sara's venture had sales of €1,000,000 last year. So 1 percentage point of
sales growth is €10,000 of extra sales. Of each euro of sales, 50 cents are
left as margin after the direct costs. Sara is considering spending €10,000 more
on marketing next year.

4. The 95% confidence interval of the marketing coefficient in column (3)
   runs from 1.14 to 1.54 percentage points per €10,000. Show with a
   calculation that €10,000 more marketing spending goes with extra sales
   between €11,400 and €15,400, and with an extra margin between €5,700 and
   €7,700. **(2 points: 1 + 1)**
5. &#40;a) Decide whether Sara should spend the extra €10,000 on the basis of
   these numbers, and give your reason in one sentence. (b) Explain one
   reason why her true gain could differ from these numbers.
   **(5 points: 3 + 2)**

::: {.callout-note collapse="true"}
## Answer indication

The calculation $2\times1.341=2.68$ in the AI answer is correct. The
wording around it is not.

1. &#40;a) No. A control holds fixed only what is measured and in the model.
   Variables that are not in the data can influence both marketing
   spending and growth, for example the quality of the product. (b) "At
   the same experience, team size, and offering type, ventures that spend
   €20,000 more on marketing have on average 2.68 percentage points higher
   fitted sales growth." The sentence is a comparison of ventures, without
   "will rise" or "causes".
2. The sentence has three errors, and the question asks for two. The
   outcome is in percent, so the difference is 3.084 percentage points,
   not 3.084%: "3.084% faster" would be a relative change. The comparison
   is with the reference group, ventures with a mainly physical offering,
   and not with the average venture. The AI sentence also does not say
   what is held fixed. A correct sentence: "Ventures with a mainly digital
   offering have on average 3.084 percentage points higher fitted sales
   growth than ventures with a mainly physical offering that have the same
   marketing spending, experience, and team size."
3. &#40;a) No. A p-value is calculated under the assumptions of the model,
   so it cannot prove them. (b) The small p-value is evidence against a
   marketing coefficient of zero, if the assumptions hold.

```{r}
#| label: imock-q6-decision
2 * unname(coef(col_3)["marketing_10k"])
ci_marketing <- confint(col_3)["marketing_10k", ]
ci_marketing
extra_sales <- round(ci_marketing, 2) * 10000
extra_margin <- 0.5 * extra_sales
rbind(extra_sales, extra_margin)
```

4. The interval of part 4 follows from column (3):
   $1.341\pm1.967\times0.100$, or `confint(col_3)`, gives 1.144 to 1.538.
   Extra sales: $1.14\times\text{€}10{,}000=\text{€}11{,}400$ at the lower end and
   $1.54\times\text{€}10{,}000=\text{€}15{,}400$ at the upper end. Extra margin
   at 50 cents per euro: $0.5\times\text{€}11{,}400=\text{€}5{,}700$ and
   $0.5\times\text{€}15{,}400=\text{€}7{,}700$.
5. &#40;a) Decision: do not spend the extra €10,000 on this evidence.
   Reason: even at the upper end of the interval, the extra margin of €7,700
   is below the €10,000 that the marketing costs. "Spend it only as a small
   test" is also a correct decision if it comes with this comparison. A
   decision to spend can also be correct if it compares the margin with
   the €10,000 and names a further gain, for example customers who return
   in later years. "Growth rises" alone, without the comparison with the
   cost, is not a reason to spend.

   &#40;b) Possible reasons, each with its explanation. The coefficient
   is an association among 320 ventures in observational data. It is not
   the effect of a change in her own budget, which can be smaller or larger.
   Customers who come back in later years add margin that these numbers
   leave out. Her venture may differ from the average venture of the
   sample.
:::

The questions of this mock exam have 100 points together. Do the mock exam in R before
you open the answers. Use the answer indications to check your numbers, and
also your units, reference groups, statements about what is held fixed, and
the limits of causal claims.

::: {.callout-tip collapse="true"}
## Optional: how the practice data were made (not part of the exam)

You do not need this box for any question. The practice data were made with
the code below. The fixed seed makes sure that the code always gives the
same 320 rows as the file that you download. Answer every question for the
registry data as the text describes them, not for this code.

```{r}
#| label: imock-data
set.seed(734)
n_mock <- 320
marketing_10k <- round(runif(n_mock, 0.25, 8), 2)
experience_years <- sample(0:18, n_mock, replace = TRUE)
team_size <- sample(
  1:10, n_mock, replace = TRUE,
  prob = c(8, 16, 20, 20, 14, 9, 6, 4, 2, 1)
)
digital <- factor(
  ifelse(runif(n_mock) < 0.45, "Yes", "No"),
  levels = c("No", "Yes")
)
spend_c <- marketing_10k - 4
growth_mean <-
  8 + 1.10 * spend_c + 3 * (digital == "Yes") -
  0.10 * spend_c^2 + 0.30 * experience_years +
  0.70 * spend_c * (digital == "Yes") + 0.20 * team_size
growth_pct <- round(
  growth_mean + rnorm(n_mock, 0, 2.5 + 0.40 * marketing_10k),
  2
)
ventures <- data.frame(
  growth_pct, marketing_10k, experience_years, team_size, digital, spend_c
)
```
:::
