# Mock Exam: QM Intermediate {#intermediate-mock-exam}

```{r}
#| label: imock-setup
#| include: false
# Chunk options only. Every object used on this page is constructed in
# visible code in the "Dataset used throughout" section below.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

This mock exam covers all six lectures of the course. It has **100
points**, so the exam grade is

$$
\text{grade} = \frac{\text{points earned}}{10}.
$$

The Intermediate exam is taken in TestVision, with R available. The practice
exam in TestVision appears in your own TestVision account once it is
published. You log in with your university account. There is no separate
link. General information about TestVision for students is on the
[university page about TestVision](https://www.tilburguniversity.edu/students/studying/exams/e-assessment/testvision).

The [QM Intermediate Code Sheet](#intermediate-code-sheet) is provided at
the exam. Use it to review the R commands and the formulas before you try
this mock exam.

Unless a question says otherwise, use a 5% significance level. Interpret
every coefficient in the units and for the sample defined below. Do not use
causal language unless the way the data were produced supports it.

| Question | Topic | Points |
|---|---|---:|
| 1 | Simple regression and least squares | 20 |
| 2 | Assumptions and diagnostics | 20 |
| 3 | Model fit and the overall F-test | 20 |
| 4 | Multiple regression and a dummy predictor | 20 |
| 5 | Interactions, curvature, and a partial F-test | 20 |
| **Total** |  | **100** |

## Dataset used throughout

The exam dataset is simulated. Treat it as a simple random sample of 320
independent young ventures from a business registry. One row is one venture.
The data are observational: none of the predictors was assigned by chance.

- `growth_pct`: change in annual sales, in percentage points
- `marketing_10k`: annual marketing spending, in €10,000s
- `experience_years`: prior industry experience of the lead founder, in
  years
- `team_size`: number of current team members
- `digital`: `No` for a mainly physical offering and `Yes` for a mainly
  digital offering. `No` is the reference category
- `spend_c = marketing_10k - 4`: marketing spending centred at €40,000.

The population is the set of ventures in this registry. Later survival and
market conditions are not measured.

The dataset is simulated with a fixed seed, so you can reproduce every
printed result on this page exactly. The code below generates the data and
fits the four models used in the questions:

```{r}
#| label: imock-data
set.seed(734)
n_mock <- 320
marketing_10k <- round(runif(n_mock, 0.25, 8), 2)
experience_years <- sample(0:18, n_mock, replace = TRUE)
team_size <- sample(
  1:10, n_mock, replace = TRUE,
  prob = c(8, 16, 20, 20, 14, 9, 6, 4, 2, 1)
)
digital <- factor(
  ifelse(runif(n_mock) < 0.45, "Yes", "No"),
  levels = c("No", "Yes")
)
spend_c <- marketing_10k - 4
growth_mean <-
  8 + 1.10 * spend_c + 3 * (digital == "Yes") -
  0.10 * spend_c^2 + 0.30 * experience_years +
  0.70 * spend_c * (digital == "Yes") + 0.20 * team_size
growth_pct <- round(
  growth_mean + rnorm(n_mock, 0, 2.5 + 0.40 * marketing_10k),
  2
)
ventures <- data.frame(
  growth_pct, marketing_10k, experience_years, team_size, digital, spend_c
)

simple_model <- lm(growth_pct ~ marketing_10k, data = ventures)
multiple_model <- lm(
  growth_pct ~ marketing_10k + experience_years + team_size + digital,
  data = ventures
)
additive_model <- lm(
  growth_pct ~ spend_c + digital + experience_years,
  data = ventures
)
extended_model <- lm(
  growth_pct ~ spend_c * digital + I(spend_c^2) + experience_years,
  data = ventures
)
```

```{r}
#| label: imock-write-csv
#| include: false
# Writes the downloadable copy of the dataset into data/, which the project
# copies into the rendered site as a listed resource (this survives frozen
# re-renders, unlike writing into docs/ directly).
# No analysis objects are created here.
write.csv(
  ventures, "data/qm-intermediate-mock-exam.csv", row.names = FALSE
)
```

[Download `qm-intermediate-mock-exam.csv`](../data/qm-intermediate-mock-exam.csv),
save it in your R project, and load it with:

```r
ventures <- read.csv("qm-intermediate-mock-exam.csv")
ventures$digital <- factor(ventures$digital, levels = c("No", "Yes"))
```

The printed results below let you continue if R is not available for a
moment. When you practise with a time limit, reproduce them from the
downloaded file.

## Question 1: Simple regression and least squares (20 points)

The first model is

```r
simple_model <- lm(growth_pct ~ marketing_10k, data = ventures)
```

and R reports

```text
Coefficients:
                Estimate  Std. Error  t value  Pr(>|t|)
(Intercept)        6.60       0.52      12.72   <2e-16
marketing_10k      1.38       0.11      12.35   <2e-16

Residual standard error: 4.57 on 318 degrees of freedom
Multiple R-squared: 0.324
```

1. Write the fitted regression equation with units. **(3 points)**
2. Interpret the slope and intercept in context. Comment on whether the
   intercept has a useful meaning here. **(6 points)**
3. Calculate the fitted growth for a venture spending €30,000 on marketing.
   If its observed growth is 15 percentage points, calculate its residual and
   interpret the sign. **(5 points)**
4. Explain what least squares minimises and why residuals are squared rather
   than simply summed. **(4 points)**
5. State the observational unit and why the fitted slope is not automatically
   causal. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. $\widehat{growth\_pct}=6.60+1.38\,marketing\_10k$, where the outcome is
   in percentage points and the predictor in €10,000s (3 points).
2. For each additional €10,000 of marketing spending, fitted annual sales
   growth is 1.38 percentage points higher among sampled ventures (3 points).
   At zero marketing spending, fitted growth is 6.60 percentage points
   (2 points). This intercept is useful only if zero lies in or near the
   observed spending range and defines a meaningful venture (1 point).
3. €30,000 gives `marketing_10k = 3`, so the fitted value is
   $6.60+1.38(3)=10.74$ percentage points (3 points). The residual is
   $15-10.74=4.26$ percentage points. Positive means observed growth exceeds
   the model's fitted value (2 points).
4. Least squares chooses the coefficients that minimise the sum of squared
   residuals, $\sum_i(y_i-\hat y_i)^2$ (2 points). Residuals that are not
   squared cancel: with an intercept, the positive and negative residuals
   sum to zero. Squaring also counts large misses more heavily (2 points).
5. The unit is one venture (1 point). An association in observational data
   can come from confounding, reverse causality, selection, or measurement
   error (1 point).
:::

## Question 2: Assumptions and diagnostics (20 points)

R produced the following diagnostics for
`lm(growth_pct ~ marketing_10k, data = ventures)`. In the left panel,
each point has fitted growth on the horizontal axis and its residual on the
vertical axis. In the right panel, observed standardised residuals are plotted
against theoretical standard-normal quantiles. A normal pattern would follow
the straight reference line.

```{r}
#| label: fig-imock-diagnostics
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Residual-versus-fitted plot and normal Q-Q plot for the simple model."
library(ggplot2)
library(patchwork)
# Rebuilds the same ventures dataset as the visible chunk above (same seed).
set.seed(734)
n_mock <- 320
marketing_10k <- round(runif(n_mock, 0.25, 8), 2)
experience_years <- sample(0:18, n_mock, replace = TRUE)
team_size <- sample(
  1:10, n_mock, replace = TRUE,
  prob = c(8, 16, 20, 20, 14, 9, 6, 4, 2, 1)
)
digital <- factor(
  ifelse(runif(n_mock) < 0.45, "Yes", "No"),
  levels = c("No", "Yes")
)
spend_c <- marketing_10k - 4
growth_mean <-
  8 + 1.10 * spend_c + 3 * (digital == "Yes") -
  0.10 * spend_c^2 + 0.30 * experience_years +
  0.70 * spend_c * (digital == "Yes") + 0.20 * team_size
growth_pct <- round(
  growth_mean + rnorm(n_mock, 0, 2.5 + 0.40 * marketing_10k),
  2
)
ventures <- data.frame(
  growth_pct, marketing_10k, experience_years, team_size, digital, spend_c
)
simple_model <- lm(growth_pct ~ marketing_10k, data = ventures)

resid_df <- data.frame(
  fitted = fitted(simple_model),
  residual = residuals(simple_model)
)
qq_df <- data.frame(rs = rstandard(simple_model))
theme_mock <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )

p_resid <- ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_point(alpha = 0.45, size = 1.6, colour = "#009E73") +
  annotate("text", x = max(resid_df$fitted), y = 0,
           label = "residual = 0", hjust = 1, vjust = -0.7,
           colour = "#17324D", size = 3) +
  labs(x = "Fitted growth (percentage points)",
       y = "Residual (percentage points)") +
  theme_mock

p_qq <- ggplot(qq_df, aes(sample = rs)) +
  geom_qq_line(colour = "#D55E00", linewidth = 0.9) +
  geom_qq(colour = "#009E73", alpha = 0.45, size = 1.6) +
  labs(x = "Normal theoretical quantile",
       y = "Observed standardised residual") +
  theme_mock

p_resid + p_qq
```

1. Read both axes of the residual-versus-fitted panel and explain what a point
   above zero represents. **(4 points)**
2. What pattern do you see in that panel? Name the assumption it challenges.
   **(4 points)**
3. Explain how to read the Q--Q plot and assess normality here. **(4 points)**
4. Name two regression assumptions that these plots cannot establish, and
   state what information would help assess each. **(4 points)**
5. Would heteroskedasticity-robust standard errors repair a curved conditional
   mean or make the slope causal? Explain. **(4 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. The horizontal axis is model-fitted growth. The vertical axis is
   observed-minus-fitted growth, both in percentage points (2 points). A point
   above zero is a venture whose observed growth is higher than its fitted
   growth (2 points).
2. Vertical spread increases with fitted growth, creating a fan shape
   (2 points). This challenges constant conditional variance
   (homoskedasticity) (2 points).
3. The Q--Q plot compares ordered residuals with values expected under a
   normal distribution. Agreement appears as points near the line (2 points).
   The middle is reasonably aligned but tail departures suggest imperfect
   normality and possible extreme residuals (2 points).
4. Examples. Independence: you need to know whether ventures are clustered,
   observed more than once, or sampled in a particular order. No
   confounding: you need knowledge of how the data came about and of
   relevant variables that are not in the model. Correct measurement: you
   need information about how the variables were recorded. (2 points for
   each valid assumption with the information that would help, maximum 4.)
5. Robust standard errors give better estimates of the uncertainty in large
   samples when the error variance is not constant (2 points). They do not
   change the fitted conditional mean, they do not repair a curve that the
   model leaves out, they do not remove confounding, and they do not make a
   coefficient from observational data causal (2 points).
:::

## Question 3: Model fit and the overall F-test (20 points)

The researcher next fits

```r
multiple_model <- lm(
  growth_pct ~ marketing_10k + experience_years + team_size + digital,
  data = ventures
)
```

R reports:

```text
Residual standard error: 4.09 on 315 degrees of freedom
Multiple R-squared: 0.464, Adjusted R-squared: 0.457
F-statistic: 68.10 on 4 and 315 DF, p-value: < 2.2e-16
```

For the intercept-only model, SST = 9,831.91. For `multiple_model`,
SSE = 5,272.57.

1. Verify $R^2$ from SST and SSE. Interpret it precisely. **(6 points)**
2. Explain why adjusted $R^2$ is lower than $R^2$, and when adjusted
   $R^2$ is useful. **(4 points)**
3. State the null and alternative hypotheses of the reported overall F-test.
   Give its conclusion. **(6 points)**
4. Explain why this F-test is not equivalent to four separate 5% t-tests.
   **(2 points)**
5. Does $R^2=0.464$ show that the model predicts future ventures well?
   Explain briefly. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. $R^2=1-5272.57/9831.91\approx0.464$ (3 points). The fitted values explain
   46.4% of the observed sample variation in sales-growth percentage points.
   It is not a percentage of correct predictions (3 points).
2. Adjusted $R^2$ contains a penalty for every predictor in the model, and
   it can fall when a term adds little to the fit (2 points). It helps to
   compare models with different numbers of predictors that are fitted to
   the same outcome and the same observations (2 points).
3. $H_0:\beta_{marketing}=\beta_{experience}=\beta_{team}=\beta_{digital}=0$
   against the alternative that at least one of them is not zero
   (3 points). The p-value is far below 0.05, so reject $H_0$: the model
   with four predictors fits better than a model with only an intercept, if
   the assumptions of the test hold (3 points).
4. The F-test is one joint test of all four coefficients together. Four
   separate t-tests answer four single questions. A set of terms can be
   jointly useful even when no single t-test rejects (2 points).
5. No. $R^2$ measures the fit in this sample only. To judge predictions for
   future ventures, the model must be checked on new data that were not
   used to fit it (2 points).
:::

## Question 4: Multiple regression and a dummy predictor (20 points)

The coefficient table below comes from
`lm(growth_pct ~ marketing_10k + experience_years + team_size + digital, data = ventures)`.
The reference category for `digital` is `No`.

```text
                  Estimate  Std. Error  t value  Pr(>|t|)
(Intercept)          2.45       0.77       3.18     0.002
marketing_10k        1.34       0.10      13.40    <0.001
experience_years     0.26       0.04       6.23    <0.001
team_size            0.17       0.10       1.63     0.104
digitalYes           3.08       0.46       6.67    <0.001
```

1. Interpret `marketing_10k`, including the predictor step, outcome unit,
   held-fixed variables, and sample. **(5 points)**
2. Interpret `digitalYes` and name its reference category. **(4 points)**
3. Compare two otherwise identical ventures: A spends €20,000 on marketing,
   has 5 years of founder experience, a team of 4, and `digital = No`. B has
   the same values except it spends €40,000 and has `digital = Yes`. Calculate
   B's fitted growth minus A's. **(5 points)**
4. Test $H_0:\beta_{team\_size}=0$ against a two-sided alternative and give
   a careful conclusion. **(3 points)**
5. Explain why including controls does not by itself give the marketing or
   digital coefficients a causal interpretation. **(3 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. Among sampled registry ventures, an additional €10,000 in annual marketing
   spending is associated with 1.34 percentage points higher fitted sales
   growth, holding founder experience, team size, and offering type fixed
   (5 points).
2. At equal marketing spending, experience, and team size, primarily digital
   ventures have fitted growth 3.08 percentage points higher than primarily
   physical ventures. `No` is the reference (4 points).
3. The spending change is two units, contributing $2(1.34)=2.68$. Changing
   from `No` to `Yes` contributes 3.08. The fitted difference is
   $2.68+3.08=5.76$ percentage points (5 points).
4. $t=1.63$, $p=0.104>0.05$, so do not reject $H_0$. The data do not show
   an association between team size and growth, holding the other
   predictors fixed. This does not prove that the coefficient is zero
   (3 points).
5. Controls adjust only for variables that are measured and included in the
   model. Unmeasured confounders, reverse causality, selection, and
   measurement error can remain in these observational data (3 points).
:::

## Question 5: Interactions, curvature, and a partial F-test (20 points)

The researcher fits the extended model

```r
extended_model <- lm(
  growth_pct ~ spend_c * digital + I(spend_c^2) + experience_years,
  data = ventures
)
```

with estimates:

```text
                         Estimate
(Intercept)                  9.20
spend_c                      1.04
digitalYes                   2.96
I(spend_c^2)                -0.11
experience_years             0.24
spend_c:digitalYes           0.68
```

The partial comparison with the additive linear model reports:

```text
Analysis of Variance Table

Model 1: growth_pct ~ spend_c + digital + experience_years
Model 2: growth_pct ~ spend_c * digital + I(spend_c^2) + experience_years
  Res.Df    RSS  Df  Sum of Sq    F   Pr(>F)
1    316  5317.0
2    314  5053.8   2     263.23  8.18  0.00035
```

1. At €40,000 of marketing spending (`spend_c = 0`), give the fitted spending
   slope for each offering type and the fitted `Yes`--`No` difference.
   **(6 points)**
2. At €60,000 (`spend_c = 2`), calculate the fitted `Yes`--`No` difference.
   **(3 points)**
3. For a physical-offering venture, calculate the fitted change when spending
   rises from €60,000 (`spend_c = 2`) to €70,000 (`spend_c = 3`), holding
   experience fixed. Repeat for a digital venture. **(5 points)**
4. State the hypotheses and conclusion of the partial F-test. **(4 points)**
5. State the hierarchy rule and one reason not to extrapolate this quadratic
   far beyond the observed spending range. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. At `spend_c = 0`, the physical-offering slope is 1.04 percentage points per
   €10,000 and the digital slope is $1.04+0.68=1.72$ (2 points each). The
   fitted digital-minus-physical difference is 2.96 percentage points
   (2 points). The slopes 1.04 and 1.72 are the slopes of the two curves
   exactly at `spend_c = 0`. The fitted change from `spend_c = 0` to
   `spend_c = 1` is $1.04-0.11=0.93$ for physical and $0.93+0.68=1.61$ for
   digital offerings. Either pair earns the points if the reasoning is
   shown.
2. The group difference is $2.96+0.68(2)=4.32$ percentage points
   (3 points).
3. For physical offerings, the change from 2 to 3 is
   $1.04(1)-0.11(3^2-2^2)=1.04-0.55=0.49$ percentage points
   (2 points). For digital offerings, add the interaction change of 0.68, so
   the fitted change is 1.17 percentage points (2 points). Units and the
   held-fixed experience condition earn 1 point.
4. $H_0:\beta_{spend\times digital}=\beta_{spend^2}=0$ against the
   alternative that at least one of the two is not zero (2 points). Since
   $p=0.00035<0.05$, reject $H_0$: the two added terms are jointly useful
   beyond the additive linear model, if the assumptions of the test hold
   (2 points).
5. Keep `spend_c` and `digital` in the model when their interaction or the
   square of `spend_c` is in it (1 point). A quadratic curve can turn
   sharply and give unrealistic values outside the observed range
   (1 point).
:::

The points of the five questions sum to 100. Do the mock exam in R before
you open the answers. Use the marking guide to check your calculations, and
also your units, reference groups, statements about what is held fixed,
uncertainty, and the limits of causal claims.
