# Tutorial 1: Simple regression, diagnostics, and model usefulness {#i-ep01}

```{r}
#| label: iep01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- fit and interpret a simple regression in R, with units
- calculate fitted values and residuals by hand
- find problems with the assumptions from residual plots, standardised
  residuals, and Cook's distance
- interpret the coefficient table, $R^2$, and the F-test together
- write a short interpretation of a model that is ready for the exam format.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam, which is taken in TestVision with R available.
You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the solution is discussed together.

The tasks follow the order of the course. Tasks 1 and 2 use Lecture 1, Task 3
uses Lecture 2, and Tasks 4 and 5 use Lecture 3. Task 6 applies all steps to
a new question that matters to founders.

::: {.course-phase .phase-before}

## Before class

Complete the after-class practice from Lectures 1 and 2. Lecture 3 is on the
day before this tutorial: read its lecture material, and do its practice
after the tutorial. Keep the Lectures
1--3 sections of the
[QM Intermediate Code Sheet](#intermediate-code-sheet) open while working.
If the notation slows you down, review
[Mathematics Refresher: lines, squares, and standardisation](../resources/math-refresher.html).

Write from memory:

1. the simple linear regression model
2. the fitted line
3. the residual
4. the formula for $R^2$
5. the null hypothesis of the overall model F-test.

::: {.callout-note collapse="true"}
## Check your preparation

1. $Y_i=\beta_0+\beta_1X_i+\varepsilon_i$.
2. $\widehat Y_i=\hat\beta_0+\hat\beta_1X_i$.
3. $e_i=Y_i-\widehat Y_i$.
4. $R^2=1-\text{SSE}/\text{SST}=\text{SSR}/\text{SST}$.
5. $H_0:\beta_1=0$ in simple regression (all slopes zero in general).
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Attempt each task in R before opening the answer. A complete written answer
names the sample, shows the calculation, reports values with units,
interprets them in context, and states one important limit of the model or
of the data.

All tasks use the Shark Tank teaching file and the valuations of Lectures 1
to 3, in **millions of dollars**: the valuation agreed on air (`agreed_m`)
and the valuation claimed by the founder (`claimed_m`). Build the data once,
in the same way as in Lectures 1--3:

```{r}
#| label: iep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
nrow(vals)   # deals for a share of the company
```

### Task 1: Fit and interpret (Lecture 1)

Fit `agreed_m ~ claimed_m` among the deals for a share of the company.
Report the sample, the fitted equation, the interpretation of the slope with
units, and the limit of the intercept.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task1
val_model <- lm(agreed_m ~ claimed_m, data = vals)
coef(val_model)
```

Among the 874 deals for a share of the company, the fitted equation is
$\widehat y = -0.068 + 0.649x$, with both valuations in \$ millions. A
claimed valuation that is \$1 million higher is associated with an agreed
valuation about \$0.65 million higher. A claimed valuation of zero is outside
the observed range, so the intercept (about $-\$68{,}000$) has no useful
business meaning of its own.
:::

### Task 2: Fitted value and residual (Lecture 1)

A founder claimed a valuation of \$5 million and the deal closed at a
valuation of \$2.5 million.

1. Calculate its fitted value.
2. Calculate its residual.
3. Interpret the sign of the residual.

::: {.callout-note collapse="true"}
## Check your answer

With $x=5$ (in \$ millions),
$\widehat y = -0.068 + 0.649 \times 5 \approx 3.18$, or about \$3.18 million.
The observed $y = 2.5$, so $e = 2.5 - 3.18 = -0.68$, about $-\$680{,}000$.
The negative sign means this deal closed below the model's fitted conditional
mean for a claimed valuation of \$5 million.

```{r}
#| label: iep01-task2
val_model <- lm(agreed_m ~ claimed_m, data = vals)
unname(2.5 - predict(val_model, newdata = data.frame(claimed_m = 5)))
```
:::

### Task 3: Diagnostics and influence (Lecture 2)

Produce the residual-versus-fitted plot and the normal Q--Q plot for the
Task 1 model.

1. Name two problems with the assumptions that you can see in the plots.
2. Count the standardised residuals beyond $\pm 3$ and the observations
   with a Cook's distance above $4/n$.
3. State the three-step rule for handling a flagged observation.
4. Name one assumption these plots cannot assess and what would assess it.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task3
#| fig-height: 3.6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(val_model), residuals(val_model),
     pch = 16, col = grDevices::adjustcolor("#0072B2", alpha.f = 0.35),
     bty = "n", xlab = "Fitted agreed valuation ($ millions)",
     ylab = "Residual ($ millions)")
abline(h = 0, lty = 2, lwd = 2, col = "#D55E00")
qqnorm(rstandard(val_model), pch = 16,
       col = grDevices::adjustcolor("#009E73", alpha.f = 0.40),
       bty = "n", main = "", xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(val_model), col = "#D55E00", lwd = 2)
par(old_par)
c(big_rstandard = sum(abs(rstandard(val_model)) > 3),
  cooks_flagged = sum(cooks.distance(val_model) > 4 / nrow(vals)))
```

The spread of the residuals widens with the fitted value. This fan
challenges constant conditional variance. The Q--Q plot leaves the line at
both ends, with several extreme residuals on both sides. This challenges
normality and points to possibly influential deals. Twenty-five standardised
residuals lie beyond $\pm3$, and 64 observations have a Cook's distance above
$4/n$. The rule for flagged observations: investigate, report, never
silently delete. The plots cannot assess independence or the absence of
confounding. Those need knowledge of how the data were collected, for
example whether the same sharks or founders appear in several rows.
:::

### Task 4: Coefficient inference (Lecture 3)

Use `summary()` and `confint()` on the Task 1 model.

1. Test $H_0:\beta_1=0$ against the two-sided alternative.
2. Report the 95% confidence interval for $\beta_1$, and verify it by hand
   from the estimate and the standard error.
3. Would robust standard errors ([Lecture 3](i03-inference-fit-ftest.qmd#i03-robust)) fix both
   problems from Task 3?

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task4
val_model <- lm(agreed_m ~ claimed_m, data = vals)
summary(val_model)$coefficients
confint(val_model)
```

The slope is $0.649$ with standard error $0.0111$, so $t \approx 58.6$ on
$872$ degrees of freedom and the p-value is far below 0.001: reject a zero
slope. By hand, with one more digit to avoid rounding error,
$0.6487 \pm 1.96 \times 0.01107 = 0.6487 \pm 0.0217 = (0.627,\; 0.670)$. This
matches `confint()`, in \$ million agreed per \$ million claimed. Robust
standard errors address the unequal variance only. They do not help with
influential observations, with the selection of pitches that ended in a
deal, or with a relationship that is not linear.
:::

### Task 5: R² and F (Lecture 3)

1. Calculate SST, SSE, SSR, and $R^2$ manually.
2. Verify them against `summary()`.
3. Verify $F = t^2$.
4. Write one concluding sentence that does not overclaim.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task5
val_model <- lm(agreed_m ~ claimed_m, data = vals)
y <- vals$agreed_m
yhat <- fitted(val_model)
SST <- sum((y - mean(y))^2)
SSE <- sum((y - yhat)^2)
SSR <- sum((yhat - mean(y))^2)
c(R2_manual = 1 - SSE / SST,
  R2_summary = summary(val_model)$r.squared,
  F_summary = unname(summary(val_model)$fstatistic["value"]),
  t_squared = summary(val_model)$coefficients["claimed_m", "t value"]^2)
```

The sums (in squared \$ millions) satisfy $\text{SST}=\text{SSR}+\text{SSE}$,
and $R^2 \approx 0.798$. The overall F statistic is about 3435, equal to the
squared slope t statistic, as it must be in simple regression. A sentence
that does not overclaim: among these deals, the claimed valuation explains
about 80% of the variation in the agreed valuations, but this is an
association in a selected set of deals, shaped partly by a few influential
ones, and not the effect of changing a claim.
:::

### Task 6: A new question: how much more equity do the sharks take?

Lecture 1 showed that the sharks usually pay the amount asked and take a
larger share of the company. Model that directly. Use all 882 deals and
regress the equity agreed in the deal on the equity the founder first
offered, both in percent of the company:

```{r}
#| label: iep01-equity
all_deals <- subset(sharks, deal_on_show == 1)
equity_model <- lm(deal_equity_pct ~ equity_offered_pct, data = all_deals)
nrow(all_deals)
summary(equity_model)$coefficients
confint(equity_model)
summary(equity_model)$r.squared
```

1. Interpret the slope and the intercept with units.
2. A founder offers 10% of the company. Which agreed share does the model
   fit?
3. Compare the $R^2$ with the valuation model. What does the difference tell
   a founder?
4. Advise the founder in two sentences.

::: {.callout-note collapse="true"}
## Check your answer

The slope is about 1.21 percentage points of agreed equity per percentage
point offered, with a 95% interval of about 1.11 to 1.31. The intercept is
about 8.3 percentage points. A founder who offers 10% is fitted to give up
$8.3 + 1.21\times10\approx20.4$% of the company, about twice the offer. $R^2$
is about 0.39, against 0.80 for the valuations. The offered share says much
less about the agreed share than the claimed valuation says about the agreed
valuation, so single deals scatter widely around the line. Advice: expect to
give up roughly twice the share you offer, and decide before the pitch how
far you are willing to go. These are deals that happened on a television
show. They describe what founders who closed a deal gave up, not what would
happen to you if you offered a different share.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice: 20 minutes

Fit `equity_offered_pct ~ season` using all 1,441 records.

1. Make and read the scatterplot with the fitted line.
2. Report the fitted equation and units.
3. Report the standard error of the slope, its t value, its p-value, and its
   95% interval.
4. Report $R^2$ and the overall F-test.
5. Inspect the residual plots and write a conclusion of about 120 words.

::: {.callout-note collapse="true"}
## Check a model answer

```{r}
#| label: iep01-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(season_model)$coefficients
confint(season_model)
c(r_squared = summary(season_model)$r.squared,
  F = unname(summary(season_model)$fstatistic["value"]))
```

The fitted equation is
$\widehat{equity}=20.457-0.824\,season$: one later season is associated
with about 0.824 percentage points less initial equity offered on average.
The standard error is about 0.048, $t\approx-17.3$, and the p-value is below
0.001. The 95% interval $(-0.917,\,-0.731)$ excludes zero.
$R^2\approx0.173$, so the season explains about 17.3% of the variation in
the equity offered, and the overall F-test ($F\approx300.7=t^2$) reaches the
same conclusion as the slope test. The residual plot shows sixteen vertical
stripes (the season takes whole numbers only) with broadly similar spread.
The Q--Q plot bends upwards at the right end because of some very high
offers. This is a trend over time among recorded pitches, not a causal
effect of a later season.
:::

### Check your understanding

Complete from memory:

1. A residual is ...
2. The slope's standard error estimates ...
3. A fan in the residual plot challenges ...
4. $R^2$ is the proportion of ...
5. In simple regression, the overall F statistic equals ...

::: {.callout-note collapse="true"}
## Check your answers

1. The observed outcome minus the fitted value, $e_i=Y_i-\widehat Y_i$.
2. How much the estimated slope would vary from sample to sample, under the
   model assumptions.
3. Constant conditional variance (homoskedasticity), and with it the
   usual standard errors.
4. The variation of the outcome in the sample that the fitted values
   explain, $\text{SSR}/\text{SST}$.
5. The squared slope t statistic, $F=t^2$.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture, or any business you know well. Imagine the
one simple regression you would show an investor (revenue on customers,
orders on ad spend), and write two sentences. First name the number from
this tutorial you would put in the headline, slope, interval, or $R^2$, and
say why. Then name the limit from Task 3 or Task 5 that you would state
openly next to it. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Headline: the slope with its 95% interval. Each extra €100 of ads is associated with roughly 12 to 19 more orders. An investor asks what money buys, and the interval shows how uncertain the number is. Limit: the fan in my residual plot means that busy weeks vary more, so the interval rests on an assumption that my own residual plot puts in doubt."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant: paste your two sentences and
ask it to play the sceptical investor.
::::

[Continue to Lecture 4: Multiple regression and what controls do](i04-multiple-regression.qmd).

:::
