# Tutorial 1: Simple regression, diagnostics, and model usefulness {#i-ep01 .unnumbered}

```{r}
#| label: iep01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- fit a simple regression in R and interpret its results with units
- check the assumptions of a model with residual plots and influence measures
- judge a model with its coefficient table, $R^2$, and the F-test
:::

Tutorials are exam preparation sessions. The tasks below are exam-style
questions. You take the exam in TestVision, and R is available.
You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the class discusses the solution together.

::: {.course-phase .phase-before}

## Before class

Complete the after-class practice from Lectures 1 and 2. Lecture 3 is on the
day before this tutorial: read its lecture material, and do its practice
after the tutorial. Keep the Lectures
1--3 sections of the
[QM Intermediate Code Sheet](code-sheet.qmd) open while you work.
If the notation slows you down, review
[Mathematics Refresher: straight lines, slopes, and intercepts](../resources/math-refresher.html#straight-lines-slopes-and-intercepts)
and [powers, roots, and squared deviations](../resources/math-refresher.html#powers-roots-and-squared-deviations).

Write from memory:

1. the simple linear regression model
2. the fitted line
3. the residual
4. the formula for $R^2$
5. the null hypothesis of the overall F-test.

::: {.callout-note collapse="true"}
## Check your preparation

1. $Y_i=\beta_0+\beta_1X_i+\varepsilon_i$.
2. $\widehat Y_i=\hat\beta_0+\hat\beta_1X_i$.
3. $e_i=Y_i-\widehat Y_i$.
4. $R^2=1-\text{SSE}/\text{SST}=\text{SSR}/\text{SST}$.
5. $H_0:\beta_1=0$ in simple regression (all slopes zero in general).
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try each task in R before you open the answer. A complete written answer
does five things. It names the sample, shows the calculation, and reports
values with units. It interprets the values in context. And it states one
important limit of the model or of the data.

All tasks use the Shark Tank teaching file and the valuations of Lectures 1
to 3, in **millions of dollars**: the valuation agreed on air (`agreed_m`)
and the valuation claimed by the founder (`claimed_m`). Build the data once,
in the same way as in Lectures 1--3:

```{r}
#| label: iep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
nrow(vals)   # deals for a share of the company
```

### Task 1: Fit and interpret (Lecture 1)

[Shark Tank data]{.shark-tag} Fit `agreed_m ~ claimed_m` among the deals for a share of the company.
Then report:

1. the sample
2. the fitted equation
3. the interpretation of the slope with units
4. the limit of the intercept's meaning.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task1
val_model <- lm(agreed_m ~ claimed_m, data = vals)
coef(val_model)
```

1. The sample is the 874 deals for a share of the company.
2. The fitted equation is $\widehat y = -0.068 + 0.649x$, with both
   valuations in \$ millions.
3. A claimed valuation that is \$1 million higher is associated with an
   agreed valuation about \$0.65 million higher.
4. A claimed valuation of zero is outside the observed range, so the
   intercept (about $-\$68{,}000$) has no useful business meaning of its own.
:::

### Task 2: Fitted value and residual (Lecture 1)

[Shark Tank data]{.shark-tag} Use the Task 1 model of the agreed valuation on the claimed valuation,
both in \$ millions. A founder claimed a valuation of \$5 million and the
deal was agreed on air at a valuation of \$2.5 million.

1. Calculate the fitted value of this deal.
2. Calculate the residual of this deal.
3. Interpret the sign of the residual.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task2
val_model <- lm(agreed_m ~ claimed_m, data = vals)
unname(2.5 - predict(val_model, newdata = data.frame(claimed_m = 5)))
```

1. With $x=5$ (in \$ millions),
   $\widehat y = -0.068 + 0.649 \times 5 \approx 3.18$, or about \$3.18
   million.
2. The observed $y = 2.5$, so $e = 2.5 - 3.18 = -0.68$, about
   $-\$680{,}000$.
3. The negative sign means this deal was agreed below the model's fitted
   conditional mean for a claimed valuation of \$5 million.
:::

### Task 3: Diagnostics and influence (Lecture 2)

[Shark Tank data]{.shark-tag} Produce the residual-versus-fitted plot and the normal Q--Q plot for the
Task 1 model of the agreed valuation on the claimed valuation.

1. Name two problems with the assumptions that you can see in the plots.
2. Count the standardised residuals beyond $\pm 3$ and the observations
   with a Cook's distance above $4/n$.
3. State the three-step rule for handling a flagged observation.
4. Name one assumption that these plots cannot assess, and say what
   information would assess it.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task3
#| fig-height: 3.6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(val_model), residuals(val_model),
     pch = 16, col = grDevices::adjustcolor("#0072B2", alpha.f = 0.35),
     bty = "n", xlab = "Fitted agreed valuation ($ millions)",
     ylab = "Residual ($ millions)")
abline(h = 0, lty = 2, lwd = 2, col = "#D55E00")
qqnorm(rstandard(val_model), pch = 16,
       col = grDevices::adjustcolor("#009E73", alpha.f = 0.40),
       bty = "n", main = "", xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(val_model), col = "#D55E00", lwd = 2)
par(old_par)
c(big_rstandard = sum(abs(rstandard(val_model)) > 3),
  cooks_flagged = sum(cooks.distance(val_model) > 4 / nobs(val_model)))
```

1. The spread of the residuals widens with the fitted value. This fan shape
   shows heteroskedasticity, so it challenges homoskedasticity. The Q--Q
   plot leaves the line at both ends, with several extreme residuals on both
   sides. This pattern challenges normality and suggests that some deals may
   be influential.
2. Twenty-five standardised residuals lie beyond $\pm3$, and 64
   observations have a Cook's distance above $4/n$.
3. The rule for flagged observations: investigate, report, never silently
   delete.
4. The plots cannot assess independence or the zero conditional mean
   assumption (no omitted confounder). To assess them, you need knowledge of
   how the data were collected, for example whether the same sharks or
   founders appear in several rows.
:::

### Task 4: Coefficient inference (Lecture 3)

[Shark Tank data]{.shark-tag} Use `summary()` and `confint()` on the Task 1 model of the agreed
valuation on the claimed valuation.

1. Test $H_0:\beta_1=0$ against the two-sided alternative.
2. Report the 95% confidence interval for $\beta_1$, and verify it by hand
   from the estimate and the standard error.
3. Would robust standard errors ([Lecture 3](i03-inference-fit-ftest.qmd#i03-robust)) fix both
   problems with the assumptions that you named in Task 3?

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task4
val_model <- lm(agreed_m ~ claimed_m, data = vals)
summary(val_model)$coefficients
confint(val_model)
```

1. The test of the slope:
   1. **Hypotheses.** $H_0:\beta_1=0$, $H_1:\beta_1\ne0$. The significance
      level is 5%.
   2. **Test statistic.** $T=\hat\beta_1/\operatorname{SE}(\hat\beta_1)$,
      with $872$ degrees of freedom.
   3. **Critical value and rejection region.** $t_{0.975,\,872}$ =
      `qt(0.975, 872)` = 1.963. Reject $H_0$ if $|t_{obs}|\ge1.963$.
   4. **Observed value.** The slope is $0.649$ with standard error
      $0.0111$, so $t_{obs} \approx 58.6$.
   5. **Conclusion.** Since $58.6\ge1.963$, reject a zero slope at the 5%
      level. The p-value, far below 0.001, gives the same decision.
2. The interval for the slope, by hand with one more digit to avoid
   rounding error:
   1. **Formula.** $\hat\beta_1\pm t_{0.975,\,872}\,\operatorname{SE}(\hat\beta_1)$.
   2. **Given values.** $\hat\beta_1=0.6487$,
      $\operatorname{SE}(\hat\beta_1)=0.01107$, $t_{0.975,\,872}=1.963$.
   3. **Calculations.** $0.6487 \pm 1.963 \times 0.01107 = 0.6487 \pm 0.0217$.
   4. **Interval.** $(0.627,\; 0.670)$, in \$ million agreed per \$ million
      claimed.
   5. **Conclusion.** This matches `confint()`.
3. Robust standard errors address the heteroskedasticity only. They do not
   help with influential observations, with the selection of pitches that
   ended in a deal, or with a relationship that is not linear.
:::

### Task 5: R² and F (Lecture 3)

Use the Task 1 model of the agreed valuation on the claimed valuation.

1. [Shark Tank data]{.shark-tag} Calculate SST, SSE, SSR, and $R^2$ manually.
2. Verify them against `summary()`.
3. Verify $F = t^2$, with $t$ the t statistic of the slope.
4. Write one concluding sentence that does not claim more than the data show.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task5
val_model <- lm(agreed_m ~ claimed_m, data = vals)
y <- vals$agreed_m
yhat <- fitted(val_model)
SST <- sum((y - mean(y))^2)
SSE <- sum((y - yhat)^2)
SSR <- sum((yhat - mean(y))^2)
c(R2_manual = 1 - SSE / SST,
  R2_summary = summary(val_model)$r.squared,
  F_summary = unname(summary(val_model)$fstatistic["value"]),
  t_squared = summary(val_model)$coefficients["claimed_m", "t value"]^2)
```

1. The sums (in squared \$ millions) satisfy
   $\text{SST}=\text{SSR}+\text{SSE}$, and $R^2 \approx 0.798$.
2. `R2_manual` and `R2_summary` give the same value.
3. The overall F statistic is about 3435, equal to the squared slope t
   statistic, as it must be in simple regression.
4. A sentence that does not claim more than the data show: among these
   deals, the claimed valuation explains about 80% of the variation in the
   agreed valuations, but this is an association in a selected set of deals,
   partly driven by a few influential deals, and not the effect of changing
   a claim.
:::

### Task 6: A new question: how much more equity do the sharks take?

[Shark Tank data]{.shark-tag} Lecture 1 showed that the sharks usually pay the amount that the founder
asked for and take a larger share of the company. Model this pattern
directly. Use the 874 deals for a share of the company. Regress the equity
agreed in the deal (`deal_equity_pct`) on the equity that the founder first
offered (`equity_offered_pct`), both in percent of the company. Report the
coefficients, their 95% confidence intervals, and $R^2$. The code below
builds the data.

```{r}
#| label: iep01-equity-data
all_deals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
nrow(all_deals)
```

1. Interpret the slope and the intercept with units.
2. A founder offers 10% of the company. Which agreed share does the model
   give as the fitted value?
3. Compare the $R^2$ with that of the valuation model of Task 1 (about
   0.798). What does the difference tell a founder?
4. Advise the founder in two sentences.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-equity
equity_model <- lm(deal_equity_pct ~ equity_offered_pct, data = all_deals)
summary(equity_model)$coefficients
confint(equity_model)
summary(equity_model)$r.squared
```

1. The slope is about 1.21 percentage points of agreed equity per
   percentage point offered, with a 95% interval of about 1.11 to 1.31. The
   intercept is about 8.5%, the fitted agreed share for an offer of 0%. No
   founder offers 0%, so the intercept has no business meaning of its own.
2. For a founder who offers 10%, the fitted share that the founder gives up
   is $8.52 + 1.21\times10\approx20.6$% of the company, about twice the
   offer.
3. $R^2$ is about 0.40, against 0.80 for the valuations. The offered share
   says much less about the agreed share than the claimed valuation says
   about the agreed valuation. So single deals are spread widely around the
   line.
4. Advice: for an offer of 10%, expect to give up about 20%, and decide
   before the pitch how large a share you are willing to give up. These are
   deals that happened on a television show. They describe what founders who
   agreed a deal on air gave up, not what would happen to you if you offered
   a different share.
:::

:::

::: {.course-phase .phase-after}

## After class

### Individual practice

[Shark Tank data]{.shark-tag} Regress the equity offered by the founder, in percent
(`equity_offered_pct`), on the season (`season`), using all 1,441 pitches.

1. Make and read the scatterplot with the fitted line.
2. Report the fitted equation and units.
3. Report the standard error of the slope, its t statistic, its p-value, and its
   95% interval.
4. Report $R^2$ and the overall F-test.
5. Inspect the residual plots and write a conclusion of about 120 words.

::: {.callout-note collapse="true"}
## Check a model answer

```{r}
#| label: iep01-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(season_model)$coefficients
confint(season_model)
c(r_squared = summary(season_model)$r.squared,
  F = unname(summary(season_model)$fstatistic["value"]))
```

This model answer covers points 2 to 5.

2. The fitted equation is
   $\widehat{\text{equity}}=20.457-0.824\,\text{season}$: one later season
   is associated with about 0.824 percentage points less initial equity
   offered on average.
3. The standard error is about 0.048, $t\approx-17.3$, and the p-value is
   below 0.001. The 95% interval $(-0.917,\,-0.731)$ excludes zero.
4. $R^2\approx0.173$, so the season explains about 17.3% of the variation
   in the equity offered, and the overall F-test ($F\approx300.7=t^2$)
   reaches the same conclusion as the slope test.
5. The residual plot shows sixteen vertical stripes (the season takes whole
   numbers only) with broadly similar spread. The Q--Q plot bends upwards at
   the right end because of some very high offers. This is a trend over time
   among recorded pitches, not a causal effect of a later season.
:::

### Check your understanding

Complete from memory:

1. A residual is ...
2. The slope's standard error estimates ...
3. A fan shape in the residual plot challenges ...
4. $R^2$ is the proportion of ...
5. In simple regression, the overall F statistic equals ...

::: {.callout-note collapse="true"}
## Check your answers

1. The observed outcome minus the fitted value, $e_i=Y_i-\widehat Y_i$.
2. How much the estimated slope would vary from sample to sample, under the
   model assumptions.
3. Homoskedasticity, and with it the usual standard errors.
4. The variation of the outcome in the sample that the fitted values
   explain, $\text{SSR}/\text{SST}$.
5. The squared slope t statistic, $F=t^2$.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Imagine the
one simple regression you would show an investor (revenue on customers,
orders on advertising spending). Write two sentences. First, name the number
from this tutorial that you would put in the headline (slope, interval, or
$R^2$), and say why. Second, name the limit from Task 3 or Task 5 that you
would state openly next to that number. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student venture that sells boxes of prepared meals on campus. "Headline: the slope with its 95% interval. Each extra €100 of advertising is associated with about 12 to 19 more orders. An investor wants to know how many orders the money brings, and the interval shows how uncertain the number is. Limit: the fan shape in my residual plot shows that orders vary more in busy weeks. So the interval depends on an assumption that my own residual plot puts in doubt."
:::

You can discuss this task, and any other task in this tutorial, with
Tilly for QM Intermediate, the course AI assistant: paste your two sentences and
ask it to act as an investor who doubts your numbers.
::::

### More practice

For extra exercises on the lectures of this tutorial, with worked solutions,
ask Tilly for QM Intermediate, the course AI assistant. For example: "Give me
an extra exercise for Intermediate Tutorial 1. Show the solution only after
I send my answer."

[Continue to Lecture 4: Multiple regression and what controls do](i04-multiple-regression.qmd).

:::
