# Tutorial 1: Simple regression, diagnostics, and model usefulness {#i-ep01}

```{r}
#| label: iep01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- fit and interpret a simple regression in R, with units
- calculate fitted values and residuals by hand
- find problems with the assumptions from residual plots, standardised
  residuals, and Cook's distance
- interpret the coefficient table, $R^2$, and the F-test together
- write a short interpretation of a model that is ready for the exam format.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You take the exam in TestVision, and R is available.
You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the class discusses the solution together.

The tasks follow the order of the course. Tasks 1 and 2 use Lecture 1, Task 3
uses Lecture 2, and Tasks 4 and 5 use Lecture 3. Task 6 applies all steps to
a new question.

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 15 minutes.</span>

Complete the after-class practice from Lectures 1 and 2. Lecture 3 is on the
day before this tutorial: read its lecture material, and do its practice
after the tutorial. Keep the Lectures
1--3 sections of the
[QM Intermediate Code Sheet](code-sheet.qmd) open while you work.
If the notation slows you down, review
[Mathematics Refresher: straight lines, slopes, and intercepts](../resources/math-refresher.html#straight-lines-slopes-and-intercepts)
and [powers, roots, and squared deviations](../resources/math-refresher.html#powers-roots-and-squared-deviations).

Write from memory:

1. the simple linear regression model
2. the fitted line
3. the residual
4. the formula for $R^2$
5. the null hypothesis of the overall model F-test.

::: {.callout-note collapse="true"}
## Check your preparation

1. $Y_i=\beta_0+\beta_1X_i+\varepsilon_i$.
2. $\widehat Y_i=\hat\beta_0+\hat\beta_1X_i$.
3. $e_i=Y_i-\widehat Y_i$.
4. $R^2=1-\text{SSE}/\text{SST}=\text{SSR}/\text{SST}$.
5. $H_0:\beta_1=0$ in simple regression (all slopes zero in general).
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try each task in R before you open the answer. A complete written answer
does five things. It names the sample, shows the calculation, and reports
values with units. It interprets the values in context. And it states one
important limit of the model or of the data. Tasks 1 to 6 take about 60 minutes together.

All tasks use the Shark Tank teaching file and the valuations of Lectures 1
to 3, in **millions of dollars**: the valuation agreed on air (`agreed_m`)
and the valuation claimed by the founder (`claimed_m`). Build the data once,
in the same way as in Lectures 1--3:

```{r}
#| label: iep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
vals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
vals$claimed_m <- vals$requested_valuation_usd / 1e6
vals$agreed_m <- vals$deal_valuation_usd / 1e6
nrow(vals)   # deals for a share of the company
```

### Task 1: Fit and interpret (Lecture 1, 8 minutes)

[Shark Tank data]{.shark-tag} Fit `agreed_m ~ claimed_m` among the deals for a share of the company.
Report the sample, the fitted equation, the interpretation of the slope with
units, and the limit of the intercept's meaning.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task1
val_model <- lm(agreed_m ~ claimed_m, data = vals)
coef(val_model)
```

Among the 874 deals for a share of the company, the fitted equation is
$\widehat y = -0.068 + 0.649x$, with both valuations in \$ millions. A
claimed valuation that is \$1 million higher is associated with an agreed
valuation about \$0.65 million higher. A claimed valuation of zero is outside
the observed range, so the intercept (about $-\$68{,}000$) has no useful
business meaning of its own.
:::

### Task 2: Fitted value and residual (Lecture 1, 5 minutes)

[Shark Tank data]{.shark-tag} A founder claimed a valuation of \$5 million and the deal closed at a
valuation of \$2.5 million.

1. Calculate the fitted value of this deal.
2. Calculate the residual of this deal.
3. Interpret the sign of the residual.

::: {.callout-note collapse="true"}
## Check your answer

With $x=5$ (in \$ millions),
$\widehat y = -0.068 + 0.649 \times 5 \approx 3.18$, or about \$3.18 million.
The observed $y = 2.5$, so $e = 2.5 - 3.18 = -0.68$, about $-\$680{,}000$.
The negative sign means this deal closed below the model's fitted conditional
mean for a claimed valuation of \$5 million.

```{r}
#| label: iep01-task2
val_model <- lm(agreed_m ~ claimed_m, data = vals)
unname(2.5 - predict(val_model, newdata = data.frame(claimed_m = 5)))
```
:::

### Task 3: Diagnostics and influence (Lecture 2, 12 minutes)

[Shark Tank data]{.shark-tag} Produce the residual-versus-fitted plot and the normal Q--Q plot for the
Task 1 model.

1. Name two problems with the assumptions that you can see in the plots.
2. Count the standardised residuals beyond $\pm 3$ and the observations
   with a Cook's distance above $4/n$.
3. State the three-step rule for handling a flagged observation.
4. Name one assumption that these plots cannot assess, and say what
   information would assess it.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task3
#| fig-height: 3.6
val_model <- lm(agreed_m ~ claimed_m, data = vals)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(val_model), residuals(val_model),
     pch = 16, col = grDevices::adjustcolor("#0072B2", alpha.f = 0.35),
     bty = "n", xlab = "Fitted agreed valuation ($ millions)",
     ylab = "Residual ($ millions)")
abline(h = 0, lty = 2, lwd = 2, col = "#D55E00")
qqnorm(rstandard(val_model), pch = 16,
       col = grDevices::adjustcolor("#009E73", alpha.f = 0.40),
       bty = "n", main = "", xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(val_model), col = "#D55E00", lwd = 2)
par(old_par)
c(big_rstandard = sum(abs(rstandard(val_model)) > 3),
  cooks_flagged = sum(cooks.distance(val_model) > 4 / nobs(val_model)))
```

The spread of the residuals widens with the fitted value. This fan
challenges constant conditional variance. The Q--Q plot leaves the line at
both ends, with several extreme residuals on both sides. This pattern
challenges normality and suggests that some deals may be influential. Twenty-five standardised
residuals lie beyond $\pm3$, and 64 observations have a Cook's distance above
$4/n$. The rule for flagged observations: investigate, report, never
silently delete. The plots cannot assess independence or the absence of
confounding. To assess them, you need knowledge of how the data were
collected, for example whether the same sharks or founders appear in several rows.
:::

### Task 4: Coefficient inference (Lecture 3, 10 minutes)

[Shark Tank data]{.shark-tag} Use `summary()` and `confint()` on the Task 1 model.

1. Test $H_0:\beta_1=0$ against the two-sided alternative.
2. Report the 95% confidence interval for $\beta_1$, and verify it by hand
   from the estimate and the standard error.
3. Would robust standard errors ([Lecture 3](i03-inference-fit-ftest.qmd#i03-robust)) fix both
   problems from Task 3?

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task4
val_model <- lm(agreed_m ~ claimed_m, data = vals)
summary(val_model)$coefficients
confint(val_model)
```

The slope is $0.649$ with standard error $0.0111$, so $t_{obs} \approx 58.6$
on $872$ degrees of freedom. The critical value is
$t_{0.975,\,872}$ = `qt(0.975, 872)` = 1.963. Since $58.6\ge1.963$, reject
a zero slope at the 5% level. The p-value, far below 0.001, gives the same
decision. By hand, with one more digit to avoid rounding error,
$0.6487 \pm 1.963 \times 0.01107 = 0.6487 \pm 0.0217 = (0.627,\; 0.670)$. This
matches `confint()`, in \$ million agreed per \$ million claimed. Robust
standard errors address the unequal variance only. They do not help with
influential observations, with the selection of pitches that ended in a
deal, or with a relationship that is not linear.
:::

### Task 5: R² and F (Lecture 3, 10 minutes)

1. [Shark Tank data]{.shark-tag} Calculate SST, SSE, SSR, and $R^2$ manually.
2. Verify them against `summary()`.
3. Verify $F = t^2$.
4. Write one concluding sentence that does not claim more than the data show.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task5
val_model <- lm(agreed_m ~ claimed_m, data = vals)
y <- vals$agreed_m
yhat <- fitted(val_model)
SST <- sum((y - mean(y))^2)
SSE <- sum((y - yhat)^2)
SSR <- sum((yhat - mean(y))^2)
c(R2_manual = 1 - SSE / SST,
  R2_summary = summary(val_model)$r.squared,
  F_summary = unname(summary(val_model)$fstatistic["value"]),
  t_squared = summary(val_model)$coefficients["claimed_m", "t value"]^2)
```

The sums (in squared \$ millions) satisfy $\text{SST}=\text{SSR}+\text{SSE}$,
and $R^2 \approx 0.798$. The overall F statistic is about 3435, equal to the
squared slope t statistic, as it must be in simple regression. A sentence
that does not claim more than the data show: among these deals, the claimed
valuation explains about 80% of the variation in the agreed valuations, but
this is an association in a selected set of deals, partly driven by a few
influential deals, and not the effect of changing a claim.
:::

### Task 6: A new question: how much more equity do the sharks take? (15 minutes)

[Shark Tank data]{.shark-tag} Lecture 1 showed that the sharks usually pay the amount that the founder
asked for and take a larger share of the company. Model this pattern
directly. Use the 874 deals for a share of the company. Regress the equity
agreed in the deal on the equity that the founder first offered, both in
percent of the company:

```{r}
#| label: iep01-equity
all_deals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
equity_model <- lm(deal_equity_pct ~ equity_offered_pct, data = all_deals)
nrow(all_deals)
summary(equity_model)$coefficients
confint(equity_model)
summary(equity_model)$r.squared
```

1. Interpret the slope and the intercept with units.
2. A founder offers 10% of the company. Which agreed share does the model
   give as the fitted value?
3. Compare the $R^2$ with the valuation model. What does the difference tell
   a founder?
4. Advise the founder in two sentences.

::: {.callout-note collapse="true"}
## Check your answer

The slope is about 1.21 percentage points of agreed equity per percentage
point offered, with a 95% interval of about 1.11 to 1.31. The intercept is
about 8.5%, the fitted agreed share for an offer of 0%. No founder offers
0%, so the intercept has no business meaning of its own. For a founder who offers 10%, the fitted share
that the founder gives up is $8.52 + 1.21\times10\approx20.6$% of the
company, about twice the offer.
$R^2$ is about 0.40, against 0.80 for the valuations. The offered share says much
less about the agreed share than the claimed valuation says about the agreed
valuation. So single deals are spread widely around the line. Advice: for an
offer of 10%, expect to give up about 20%, and decide before the pitch how
large a share you are willing to give up. These are deals that happened on a television
show. They describe what founders who closed a deal gave up, not what would
happen to you if you offered a different share.
:::

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 30 minutes.</span>

### Timed individual practice: 20 minutes

[Shark Tank data]{.shark-tag} Fit `equity_offered_pct ~ season` using all 1,441 pitches.

1. Make and read the scatterplot with the fitted line.
2. Report the fitted equation and units.
3. Report the standard error of the slope, its t value, its p-value, and its
   95% interval.
4. Report $R^2$ and the overall F-test.
5. Inspect the residual plots and write a conclusion of about 120 words.

::: {.callout-note collapse="true"}
## Check a model answer

```{r}
#| label: iep01-practice-equity
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(season_model)$coefficients
confint(season_model)
c(r_squared = summary(season_model)$r.squared,
  F = unname(summary(season_model)$fstatistic["value"]))
```

The fitted equation is
$\widehat{\text{equity}}=20.457-0.824\,\text{season}$: one later season is associated
with about 0.824 percentage points less initial equity offered on average.
The standard error is about 0.048, $t\approx-17.3$, and the p-value is below
0.001. The 95% interval $(-0.917,\,-0.731)$ excludes zero.
$R^2\approx0.173$, so the season explains about 17.3% of the variation in
the equity offered, and the overall F-test ($F\approx300.7=t^2$) reaches the
same conclusion as the slope test. The residual plot shows sixteen vertical
stripes (the season takes whole numbers only) with broadly similar spread.
The Q--Q plot bends upwards at the right end because of some very high
offers. This is a trend over time among recorded pitches, not a causal
effect of a later season.
:::

### Check your understanding (5 minutes)

Complete from memory:

1. A residual is ...
2. The slope's standard error estimates ...
3. A fan in the residual plot challenges ...
4. $R^2$ is the proportion of ...
5. In simple regression, the overall F statistic equals ...

::: {.callout-note collapse="true"}
## Check your answers

1. The observed outcome minus the fitted value, $e_i=Y_i-\widehat Y_i$.
2. How much the estimated slope would vary from sample to sample, under the
   model assumptions.
3. Constant conditional variance (homoskedasticity), and with it the
   usual standard errors.
4. The variation of the outcome in the sample that the fitted values
   explain, $\text{SSR}/\text{SST}$.
5. The squared slope t statistic, $F=t^2$.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Imagine the
one simple regression you would show an investor (revenue on customers,
orders on advertising spending). Write two sentences. First, name the number
from this tutorial that you would put in the headline (slope, interval, or
$R^2$), and say why. Second, name the limit from Task 3 or Task 5 that you
would state openly next to that number. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student venture that sells boxes of prepared meals on campus. "Headline: the slope with its 95% interval. Each extra €100 of advertising is associated with about 12 to 19 more orders. An investor wants to know how many orders the money brings, and the interval shows how uncertain the number is. Limit: the fan in my residual plot shows that orders vary more in busy weeks. So the interval depends on an assumption that my own residual plot puts in doubt."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant: paste your two sentences and
ask it to act as an investor who doubts your numbers.
::::

[Continue to Lecture 4: Multiple regression and what controls do](i04-multiple-regression.qmd).

:::
