# Tutorial 1: Simple regression, diagnostics, and model usefulness {#i-ep01}

```{r}
#| label: iep01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- fit and interpret a simple regression in R, with units;
- calculate fitted values and residuals by hand;
- diagnose assumption problems from residual plots and influence measures;
- interpret coefficient inference, $R^2$, and the F-test together;
- produce a concise TestVision-ready model interpretation.
:::

Tutorials are **exam preparation sessions**: every task below follows the
format of the final exam. The standing routine for each task is the same ---
the TA states the task, pairs work on it first for a few minutes with AI use
allowed, and the solution is then discussed in plenary, starting from the
pair attempts.

One note on the order. The tasks run in course order: Tasks 1--2 rest on
Lecture 1, Task 3 on Lecture 2, and Tasks 4--5 on yesterday's Lecture 3.
This is deliberate --- the tutorial meets the day after Lecture 3, so you
warm up on settled material before testing what you learned yesterday.

::: {.course-phase .phase-before}

## Before class

Complete the after-class practice from Lectures 1--3. Keep the Lectures
1--3 sections of the
[QM Intermediate Code Sheet](#intermediate-code-sheet) open while working,
and review
[Mathematics Refresher: lines, squares, and standardisation](../resources/math-refresher.html)
if the notation is slowing you down.

Write from memory:

1. the simple linear regression model;
2. the fitted line;
3. the residual;
4. the formula for $R^2$;
5. the null hypothesis of the overall model F-test.

::: {.callout-note collapse="true"}
## Check your preparation

1. $Y_i=\beta_0+\beta_1X_i+\varepsilon_i$.
2. $\widehat Y_i=b_0+b_1X_i$.
3. $e_i=Y_i-\widehat Y_i$.
4. $R^2=1-\text{SSE}/\text{SST}=\text{SSR}/\text{SST}$.
5. $H_0:\beta_1=0$ in simple regression (all slopes zero in general).
:::

:::

::: {.course-phase .phase-in-class}

## In class

Attempt each task in R before opening the answer. A complete written answer
names the sample, shows the calculation, reports values with units,
interprets in context, and states an important model or design boundary.

All tasks use the Shark Tank teaching file and the running money unit of
**thousands of dollars**. Build the analysis data once --- this is the same
construction as Lectures 1--3:

```{r}
#| label: iep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
deals$ask_1k <- deals$ask_amount_usd / 1000
deals$deal_1k <- deals$deal_amount_usd / 1000
```

### Task 1 --- Fit and interpret (Lecture 1)

Fit `deal_1k ~ ask_1k` among the on-air deal records. Report the sample,
the fitted equation, the slope interpretation with units, and the intercept
limitation.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task1
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
coef(deal_model)
```

Among the 882 pitches with an on-air deal, the fitted equation is
$\widehat y = 27.571 + 1.0085x$, with both variables in $1,000s. One
additional \$1,000 requested is associated with about \$1,008.50 more in
fitted agreed amount. A zero request is outside the observed range, so the
intercept (about \$27,600) lacks a useful standalone business
interpretation.
:::

### Task 2 --- Fitted value and residual (Lecture 1)

A pitch requested \$500,000 and agreed \$450,000.

1. Calculate its fitted value.
2. Calculate its residual.
3. Interpret the sign of the residual.

::: {.callout-note collapse="true"}
## Check your answer

With $x=500$ (in $1,000s),
$\widehat y = 27.571 + 1.0085 \times 500 \approx 531.83$, or about
\$531,826. The observed $y = 450$, so
$e = 450 - 531.83 = -81.83$, about $-\$81{,}826$. The negative sign means
this pitch agreed less than the model's fitted conditional mean for a
\$500,000 request.

```{r}
#| label: iep01-task2
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
unname(450 - predict(deal_model, newdata = data.frame(ask_1k = 500)))
```
:::

### Task 3 --- Diagnostics and influence (Lecture 2)

Produce the residual-versus-fitted plot and the normal Q--Q plot for the
Task 1 model.

1. Name two visible assumption concerns.
2. Count the standardised residuals beyond $\pm 3$ and the observations
   above the $4/n$ Cook's distance screen.
3. State the three-step rule for handling a flagged observation.
4. Name one assumption these plots cannot assess and what would assess it.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task3
#| fig-height: 3.6
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.5, 4.4, 1.3, 0.7), las = 1)
plot(fitted(deal_model), residuals(deal_model),
     pch = 16, col = grDevices::adjustcolor("#0072B2", alpha.f = 0.35),
     bty = "n", xlab = "Fitted agreed amount ($1,000s)",
     ylab = "Residual ($1,000s)")
abline(h = 0, lty = 2, lwd = 2, col = "#D55E00")
qqnorm(rstandard(deal_model), pch = 16,
       col = grDevices::adjustcolor("#009E73", alpha.f = 0.40),
       bty = "n", main = "", xlab = "Normal theoretical quantile",
       ylab = "Standardised residual")
qqline(rstandard(deal_model), col = "#D55E00", lwd = 2)
par(old_par)
c(big_rstandard = sum(abs(rstandard(deal_model)) > 3),
  cooks_flagged = sum(cooks.distance(deal_model) > 4 / nrow(deals)))
```

Residual spread widens with the fitted value (a fan, challenging constant
conditional variance), and the Q--Q plot shows a heavy right tail with
several extreme positive residuals (challenging normality and flagging
influence). Twelve standardised residuals exceed $\pm3$ and 21 observations
exceed the $4/n$ Cook's screen --- the most influential pitch asked
\$150,000 and agreed \$3,000,000. The rule: investigate, report, never
silently delete. The plots cannot assess independence (or zero conditional
mean); those need knowledge of the data collection --- for example whether
founders or ventures appear in multiple rows.
:::

### Task 4 --- Coefficient inference (Lecture 3)

Use `summary()` and `confint()` on the Task 1 model.

1. Test $H_0:\beta_1=0$ against the two-sided alternative.
2. Report the 95% confidence interval for $\beta_1$, and verify it by hand
   from the estimate and SE.
3. Would robust standard errors fix both concerns from Task 3?

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task4
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
summary(deal_model)$coefficients
confint(deal_model)
```

The slope is $1.0085$ with classical SE $0.0182$, so
$t = 1.0085/0.0182 \approx 55.4$ on $880$ degrees of freedom and the
p-value is far below 0.001: reject a zero slope. By hand,
$1.0085 \pm 1.96 \times 0.0182 = (0.973,\; 1.044)$, matching `confint()`
--- in $1,000s of deal per $1,000 asked. Robust SEs address the unequal
variance only; they do not remove influential observations, selection into
deals, or a misspecified mean.
:::

### Task 5 --- R² and F (Lecture 3)

1. Calculate SST, SSE, SSR, and $R^2$ manually.
2. Verify them against `summary()`.
3. Verify $F = t^2$.
4. Write one concluding sentence that does not overclaim.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep01-task5
deal_model <- lm(deal_1k ~ ask_1k, data = deals)
y <- deals$deal_1k
yhat <- fitted(deal_model)
SST <- sum((y - mean(y))^2)
SSE <- sum((y - yhat)^2)
SSR <- sum((yhat - mean(y))^2)
c(R2_manual = 1 - SSE / SST,
  R2_summary = summary(deal_model)$r.squared,
  F_summary = unname(summary(deal_model)$fstatistic["value"]),
  t_squared = summary(deal_model)$coefficients["ask_1k", "t value"]^2)
```

The sums (in squared $1,000s) satisfy $\text{SST}=\text{SSR}+\text{SSE}$,
and $R^2 \approx 0.777$. The overall F statistic is about 3069, equal to
the squared slope t statistic, as it must be in simple regression. A
sentence that does not overclaim: requested amount explains about 78% of
the observed variation in agreed amounts among on-air deals, but this
selected observational association --- shaped partly by a few influential
pitches --- is not a causal effect of changing an ask.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice --- 20 minutes

Fit `equity_offered_pct ~ season` using all 1,441 records.

1. Make and read the scatterplot with the fitted line.
2. Report the fitted equation and units.
3. Report the slope SE, t statistic, p-value, and 95% interval.
4. Report $R^2$ and the overall F-test.
5. Inspect the residual plots and write a 120-word model conclusion.

::: {.callout-note collapse="true"}
## Check a model answer

The fitted equation is
$\widehat{equity}=20.457-0.824\,season$: one later season is associated
with about 0.824 percentage points less initial equity offered on average.
The classical SE is about 0.048, $t\approx-17.3$, and the p-value is below
0.001; the 95% interval $(-0.917,\,-0.731)$ excludes zero.
$R^2\approx0.173$, so season explains about 17.3% of observed equity-offer
variation, and the overall F-test ($F\approx300.7=t^2$) reaches the same
conclusion as the slope test. The residual plot shows sixteen vertical
stripes (season is whole-numbered) with broadly similar spread, and the
Q--Q plot a heavy right tail from very high offers. This is an
observational time association among recorded pitches, not a causal effect
of progressing to a later season.
:::

### Check your understanding

Complete from memory:

1. A residual is ...
2. The slope's standard error estimates ...
3. A fan in the residual plot challenges ...
4. $R^2$ is the proportion of ...
5. In simple regression, the overall F statistic equals ...

::: {.callout-note collapse="true"}
## Check your answers

1. The observed outcome minus the fitted value, $e_i=Y_i-\widehat Y_i$.
2. How much the estimated slope would vary across repeated samples under
   the model conditions.
3. Constant conditional variance (homoskedasticity) --- and with it the
   classical standard errors.
4. Observed sample variation in the outcome explained by the fitted values,
   $\text{SSR}/\text{SST}$.
5. The squared slope t statistic, $F=t^2$.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well. Imagine the
one simple regression you would show an investor (revenue on customers,
orders on ad spend), and write two sentences: first name the number from
today's pipeline you would put in the headline --- slope, interval, or
$R^2$ --- and why, then name the caveat from Task 3 or Task 5 you would be
honest about in the footnote. There is no wrong answer; the point is
pairing every headline number with its boundary.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Headline: the slope with its 95% interval --- each extra €100 of ads is associated with roughly 12 to 19 more orders --- because an investor asks what money buys, and the interval carries the honesty. Footnote: the fan shape in my residual plot means busy weeks are noisier, so that interval leans on an assumption my own diagnostics question."
:::

You can discuss this task --- and any other task in this tutorial --- with
**Tilly**, the course AI assistant on Canvas: paste your two sentences and
ask it to play the sceptical investor.
::::

Continue to Lecture 4: Multiple regression and what controls do.

:::
