# Tutorial 2: Multiple regression and extensions {#i-ep02}

```{r}
#| label: iep02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 2, you can</span>

- fit and interpret a multiple regression in R
- test single coefficients and groups of coefficients against zero
- interpret dummy, interaction, and quadratic terms correctly
- compare nested models with a partial F-test
- read a regression table as printed in a research paper
- write a complete analysis that is ready for the exam format and states
  its limits.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You first try each task in pairs. You may use AI for this, following
the four steps Attempt, Ask, Verify, Explain. Then the whole group discusses
the solution with the tutor.

The tasks follow the order of the course. Task 1 uses Lecture 4, Task 2
uses Lecture 5, and Tasks 3 and 4 use Lecture 6. A regression table then
puts the models of Tasks 2 to 4 side by side. Task 5 applies Lectures 3 to 5
to an invented table in the layout of a research paper.

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 15 minutes.</span>

Complete the after-class practice from Lectures 4 and 5. Lecture 6 is on the
day before this tutorial: read its lecture material, and do its practice
after the tutorial. Use the complete
[QM Intermediate Code Sheet](code-sheet.qmd) while working. If needed,
review
[Mathematics Refresher: indicators and reference categories](../resources/math-refresher.html#indicators-and-reference-categories),
[products, interactions, and combined effects](../resources/math-refresher.html#products-interactions-and-combined-effects),
and [polynomial terms](../resources/math-refresher.html#polynomial-terms).

Complete from memory:

1. You interpret a multiple-regression slope while ...
2. A dummy coefficient compares ...
3. An interaction coefficient is a difference in ...
4. A partial F-test compares ...

::: {.callout-note collapse="true"}
## Check your preparation

1. Holding the other included predictors fixed.
2. A category with the named reference category.
3. Slopes.
4. A reduced nested model with a complete model on the same observations.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try each task in R before you open the answer. Tasks 1 to 4 use the Shark
Tank file, and money is measured in thousands of dollars (`ask_1k`). First
build the data for the analysis. The code is the same as in Lectures 4--6.
Tasks 1 to 5 and the regression table take about 65 minutes together.

```{r}
#| label: iep02-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)
```

### Task 1: Multiple-regression coefficients (Lecture 4, 12 minutes)

[Shark Tank data]{.shark-tag} Fit:

```{r}
#| label: iep02-base-model
m1 <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
summary(m1)
```

1. Interpret all three slopes with units and with the predictors that are
   held fixed.
2. Find out which 95% confidence interval of a coefficient includes zero.
3. Interpret adjusted $R^2$ and the overall F-test.
4. Give two reasons why the coefficients cannot be read as causal effects.

::: {.callout-note collapse="true"}
## Check your answer

At fixed values of the other included predictors, among the 1,441 recorded
pitches: one later season is associated with about 0.836 percentage points
less equity offered, an extra \$1,000 requested with about 0.0029
percentage points less (about 0.29 points per \$100,000), and one
additional description word with about 0.128 percentage points less.
`confint(m1)` shows that the interval for the description words, about
$(-0.351, 0.095)$, includes zero. $R^2\approx0.188$: the model explains
about 19% of the variation in the equity offered. Adjusted
$R^2\approx0.186$ is almost the same, so the predictors do not simply add
random noise. The overall
F-test ($F\approx110.6$ on 3 and 1437 df) rejects the null hypothesis that
all three slopes are zero. Possible reasons against a causal reading: only
selected pitches reach the show, the quality of the venture is not
measured, the equity offered may influence the amount requested, and the
variables may be measured with error.
:::

### Task 2: Dummy variables and a partial F-test (Lecture 5, 12 minutes)

```{r}
#| label: iep02-gender-models
reduced <- lm(equity_offered_pct ~ season + ask_1k,
              data = known_gender)
complete <- lm(equity_offered_pct ~ season + ask_1k + gender,
               data = known_gender)
summary(complete)
anova(reduced, complete)
```

1. [Shark Tank data]{.shark-tag} Name the reference category.
2. Interpret both dummy coefficients.
3. State the hypotheses of the partial F-test and decide.
4. Explain why the partial F-test and the individual t-tests answer
   different questions.

::: {.callout-note collapse="true"}
## Check your answer

The reference category is all-men. At the same season and requested amount,
all-women presenting teams are fitted to offer about 1.28 equity percentage
points more than all-men teams, and mixed teams about 0.12 points more. The
unit is percentage points, not percent. The null hypothesis of the partial
F-test is that both dummy coefficients are zero. $F\approx3.76$ and
p $\approx$ 0.023, so we reject the null hypothesis at 5%. The two terms are jointly useful, even
though the t-test of the mixed dummy alone (p $\approx$ 0.82) is far from
significant. The F-test asks whether the two terms together improve the
model. Each t-test asks about one comparison with the reference group. The
categories are broad and the data do not come from an experiment, so the
result does not show a causal effect of gender.
:::

### Task 3: Interaction (Lecture 6, 12 minutes)

```{r}
#| label: iep02-interaction
m_interaction <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
summary(m_interaction)
```

1. [Shark Tank data]{.shark-tag} Write the fitted line for `women = 0` and for `women = 1` at a fixed
   requested amount.
2. Calculate both season slopes.
3. Interpret the interaction estimate and its p-value without claiming too
   much in either direction.
4. Explain what centring does.

::: {.callout-note collapse="true"}
## Check your answer

At a fixed requested amount, the fitted line for all-men teams is
$13.625-0.799\,\text{season\_c}$ plus the term for the requested amount. For teams
with women the intercept is $13.625+0.807=14.432$ and the slope is
$-0.7994-0.0615=-0.861$. So the season slope is about $-0.799$ percentage
points per season for all-men teams and about $-0.861$ for teams with
women. The estimated difference in slopes is $-0.061$ percentage points per
season with p $\approx$ 0.520. The model gives no evidence that the season
trends differ between the two groups. It also does not prove that the trends
are equal, because the estimate is imprecise. Centring makes the `women`
coefficient compare the groups at the average season (about 8.8) and not at
Season 0, which does not exist. It changes neither the fitted values nor the
interaction estimate.
:::

### Task 4: Polynomial term (Lecture 6, 10 minutes)

```{r}
#| label: iep02-polynomial
m_linear <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
m_quadratic <- lm(
  equity_offered_pct ~ season + I(season^2) + ask_1k,
  data = sharks
)
summary(m_quadratic)
anova(m_linear, m_quadratic)
```

1. [Shark Tank data]{.shark-tag} Why must you write `I(season^2)` in the R formula?
2. Decide with the partial F-test whether the quadratic term improves the
   model.
3. Calculate the fitted change from Season 10 to Season 11.
4. State why you should not extrapolate the curve beyond Season 16.

::: {.callout-note collapse="true"}
## Check your answer

`I()` tells R to calculate the square. Without `I()`, R would read `^` as
part of the formula syntax and not as a square. The partial F-test gives $F\approx20.1$ with p $<$ 0.001,
so the quadratic term improves the model. With $\hat\beta_1\approx-1.713$
and $\hat\beta_2\approx0.051$, the change from Season 10 to 11 is
$\hat\beta_1+\hat\beta_2(2\times10+1)=\hat\beta_1+21\hat\beta_2\approx-0.64$
percentage points. Never read $\hat\beta_1$ alone while the squared term is
in the model. Beyond Season 16 the fitted curve turns upward, and no data
support that turn. A quadratic curve can change direction sharply outside
the observed range.
:::

### The models of Tasks 2 to 4 in one table (5 minutes) {#iep02-spec-table}

[Shark Tank data]{.shark-tag} Lecture 4 showed how research papers put several models side by side in one
regression table. The table below does this for the models of Tasks 2 to 4.
All columns of such a table must use the same observations. Tasks 2 and 3
use the 1,434 pitches with a known team classification (`known_gender`).
Task 4 uses all 1,441 pitches. So first fit the quadratic model again on the
1,434 pitches. The code then prints the two model outputs that this
tutorial has not shown yet.

```{r}
#| label: iep02-table-models
m_quadratic_known <- lm(
  equity_offered_pct ~ season + I(season^2) + ask_1k,
  data = known_gender
)
summary(reduced)
summary(m_quadratic_known)
```

Column (1) is the reduced model of Task 2 (`reduced`), column (2) the
complete model of Task 2 (`complete`), column (3) the interaction model of
Task 3 (`m_interaction`), and column (4) the quadratic model
(`m_quadratic_known`).

::: {.reg-table}
| | (1) | (2) | (3) | (4) |
|:---|---:|---:|---:|---:|
| Season | −0.815\*\*\* | −0.824\*\*\* | −0.799\*\*\* | −1.713\*\*\* |
| | (0.047) | (0.048) | (0.066) | (0.206) |
| Season squared | | | | 0.051\*\*\* |
| | | | | (0.011) |
| Amount requested (\$1,000s) | −0.0029\*\*\* | −0.0027\*\*\* | −0.0027\*\*\* | −0.0027\*\*\* |
| | (0.0006) | (0.0006) | (0.0006) | (0.0006) |
| All-women team | | 1.284\*\* | | |
| | | (0.481) | | |
| Mixed team | | 0.121 | | |
| | | (0.541) | | |
| Team with women (all-women or mixed) | | | 0.807 | |
| | | | (0.412) | |
| Season × team with women | | | −0.061 | |
| | | | (0.096) | |
| Intercept | 21.185\*\*\* | 20.839\*\*\* | 13.625\*\*\* | 24.168\*\*\* |
| | (0.484) | (0.506) | (0.335) | (0.822) |
| Observations | 1,434 | 1,434 | 1,434 | 1,434 |
| $R^2$ | 0.187 | 0.191 | 0.189 | 0.198 |
| Adjusted $R^2$ | 0.186 | 0.189 | 0.187 | 0.197 |
:::

Outcome: equity offered (%). Ordinary least squares. Standard errors in
parentheses. All four columns use the same 1,434 pitches with a known team
classification. The reference group is all-men teams. In column (3) season
is centred at its mean (about 8.8), so the intercept and the coefficient of
"Team with women" refer to the average season.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001.

Read along the Season row. The coefficient is close to $-0.8$ in
columns (1) to (3). In column (4) it is $-1.713$, because with the squared term in
the model the season coefficient alone is not the trend. Read down a column
to see one model. Columns (2), (3), and (4) each add terms to column (1), so
a partial F-test can compare each of them with column (1). The bottom rows
show that the four models fit almost equally well: adjusted $R^2$ lies
between 0.186 and 0.197, a difference of about one percentage point of the
variation in the equity offered. Use this layout for regression results in
your thesis.

### Task 5: Read a regression table from a paper (14 minutes)

The table below is invented. It has the layout of a table in a research
paper, the same layout as the table above. The outcome is
entrepreneurial intention on a scale from 1 to 7, measured in a survey of
265 students. Self-efficacy is a score from 1 to 7 for how confident a
student is that he or she can start a business. The table does not show the
intercept. Some journals do this.

::: {.reg-table}
| | (1) | (2) |
|:---|---:|---:|
| Age (years) | 0.02 | 0.01 |
| | (0.02) | (0.02) |
| Female (1 = yes) | −0.21\* | −0.15 |
| | (0.10) | (0.10) |
| Prior entrepreneurial experience (1 = yes) | | 0.42\*\* |
| | | (0.13) |
| Entrepreneurial self-efficacy (1--7) | | 0.38\*\*\* |
| | | (0.07) |
| Observations | 265 | 265 |
| $R^2$ | 0.03 | 0.19 |
| Adjusted $R^2$ | 0.02 | 0.18 |
:::

Outcome: entrepreneurial intention (scale from 1 to 7). Ordinary least
squares. Standard errors in parentheses.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001.

1. What do the numbers in parentheses and the stars mean? Check the stars
   of the self-efficacy coefficient with its t statistic.
2. Interpret the self-efficacy coefficient in column (2) in one complete
   sentence.
3. The coefficient of Female changes from $-0.21$ with a star in column (1)
   to $-0.15$ without a star in column (2). Explain what happened with the
   ideas of Lecture 4, and say what the missing star does and does not
   mean.
4. $R^2$ rises from 0.03 to 0.19. What does that say, and which test would
   show whether the two added predictors are jointly useful?

::: {.callout-note collapse="true"}
## Check your answer

1. The numbers in parentheses are standard errors. The stars show the
   p-value of the two-sided t-test of each coefficient against zero. For
   self-efficacy, $t_{obs}=0.38/0.07\approx5.4$. This is far beyond the
   critical value $t_{0.975,\,260}$ = `qt(0.975, 260)` = 1.969, and p is
   below 0.001, so three stars are correct.
2. Among the 265 surveyed students, a self-efficacy score that is one point
   higher is associated with an entrepreneurial intention that is 0.38
   points higher, holding age, gender, and prior experience fixed.
3. Model (2) holds two more variables fixed than model (1). The Female
   coefficient now compares women and men with the same experience and
   self-efficacy. In column (1), at the same age, the gap was 0.21 points. Part of this
   gap went together with differences in experience and self-efficacy. The missing star says that the
   confidence interval around $-0.15$ includes zero. It does not show that
   the gap is zero.
4. The two added predictors explain a further 16 percentage points of
   the variation in intentions in this sample. A partial F-test of the two
   nested models would show whether the two added predictors are jointly
   useful. It compares model (1) with model (2) on the same 265 students. If
   you have the data, `anova()` of the two fitted models gives this test.
:::

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 40 minutes.</span>

### Timed individual practice: 30 minutes

[Shark Tank data]{.shark-tag} Build one model of the equity offered (`equity_offered_pct`) in the Shark
Tank file. Use season, the requested amount in \$1,000s, the presenting
team, and one extension that you can justify: an interaction of season with
the dummy `women`, or a quadratic term in season next to the three known
team categories (`gender`).

Your answer must include:

1. a precise research question
2. the code that prepares the data, so that someone else can repeat it
3. one graph of the data before you fit the model
4. the R formula and the model summary. If you compare several models, show
   them in one regression table in the layout of this tutorial
5. interpretations of the coefficients with units and reference groups
6. a t-test of one coefficient and an overall or partial F-test
7. $R^2$, adjusted $R^2$, and residual diagnostics
8. one graph of fitted values from the model
9. a conclusion of 150 words that clearly separates three things: the
   association, its uncertainty, and the reasons why it is not causal.

::: {.callout-note collapse="true"}
## Check a model structure

A good solution could fit

```r
m <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
```

or a model with a quadratic term for season and the group dummies. Keep
`season_c` and `women` in the model next to their interaction (the hierarchy
principle), and name the reference group. Interpret slopes that combine
several coefficients, or differences between fitted values. Use `anova()` to
compare the model with a nested reduced model. Inspect
the residuals, including `rstandard()` and Cook's distance, and plot fitted
values inside Seasons 1--16. The conclusion must say that the team
categories are broad, that only broadcast pitches are in the file, and that
the results therefore do not show a causal effect of gender or season.
:::

### Check your understanding (5 minutes)

Complete from memory:

1. A coefficient in a multiple regression compares cases that ...
2. In a model with an intercept, R includes a factor with three categories
   as ...
3. In the model `season_c * women + ask_1k`, the season slope for teams
   with women is ...
4. Centring a predictor changes ..., and it does not change ...
5. In a model with $x$ and $x^2$, the fitted change from $x$ to $x+1$ is ...

::: {.callout-note collapse="true"}
## Check your answers

1. Differ in that predictor and have the same values of the other included
   predictors.
2. Two dummy variables. The third category is the reference category.
3. The coefficient of `season_c` plus the coefficient of `season_c:women`.
4. The meaning and the value of the intercept and of the coefficients that
   are read at $x=0$. It does not change the fitted values or the
   interaction coefficient.
5. $\hat\beta_1+\hat\beta_2(2x+1)$.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Describe in two
sentences the full analysis that you would run on its data. First, name the
outcome, two predictors, and one extension (a set of dummies, an
interaction, or a quadratic term) that answers a real question. Second,
say which limit you would mention first in your conclusion: selection,
measurement, or omitted variables. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "Outcome: weekly orders. Predictors: advertising spend and price, plus a dummy for exam weeks and its interaction with advertising spend, because I believe advertising works mainly when students are on campus. The first limit is omitted variables: the weather affects both how many flyers I hand out and how many people are on campus, and nothing in my data measures it."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant. Ask it to review your two sentences
before you spend a lot of time on the data.
::::

This is the end of the QM Intermediate lectures and tutorials. Review the
[QM Intermediate Code Sheet](code-sheet.qmd), attempt the
[QM Intermediate Mock Exam](mock-exam.qmd), then return to the
[R Refresher](../resources/r-refresher.html),
[Mathematics Refresher](../resources/math-refresher.html), or the
[External Video Guide](../resources/video-guide.html) whenever you want to
review a specific topic.

[Continue to QM Intermediate Code Sheet](code-sheet.qmd).

:::
