# Tutorial 2: Multiple regression and extensions {#i-ep02 .unnumbered}

```{r}
#| label: iep02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 2, you can</span>

- fit and interpret a multiple regression with dummies, interactions, and curves
- compare nested models and test groups of terms with a partial F-test
- read a regression table from a research paper and state its limits
:::

Tutorials are exam preparation sessions. The tasks below are exam-style
questions. You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the class discusses the solution together.

::: {.course-phase .phase-before}

## Before class

Complete the after-class practice from Lectures 4 and 5. Lecture 6 is on the
day before this tutorial: read its lecture material, and do its practice
after the tutorial. Use the complete
[QM Intermediate Code Sheet](code-sheet.qmd) while working. If needed,
review
[Mathematics Refresher: indicators and reference categories](../resources/math-refresher.qmd#indicators-and-reference-categories),
[products, interactions, and combined effects](../resources/math-refresher.qmd#products-interactions-and-combined-effects),
and [polynomial terms](../resources/math-refresher.qmd#polynomial-terms).

Complete from memory:

1. You interpret a multiple-regression slope while ...
2. A dummy coefficient compares ...
3. An interaction coefficient is a difference in ...
4. A partial F-test compares ...

::: {.callout-note collapse="true"}
## Check your preparation

1. Holding the other included predictors fixed.
2. A category with the named reference category.
3. Slopes.
4. A reduced nested model with a complete model on the same observations.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try each task in R before you open the answer. Tasks 1 to 4 use the Shark
Tank file, and money is measured in thousands of dollars (`ask_1k`). First
build the data for the analysis. The code is the same as in Lectures 4--6.

```{r}
#| label: iep02-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)
```

### Task 1: Multiple-regression coefficients (Lecture 4)

[Shark Tank data]{.shark-tag} With all 1,441 pitches (`sharks`), fit a regression
of the equity offered, in percent (`equity_offered_pct`), on the season
(`season`), the requested amount in thousands of dollars (`ask_1k`), and the
number of words in the description (`description_words`).
Then answer these questions.

1. Interpret all three slopes with units and with the predictors that are
   held fixed.
2. Find out which 95% confidence interval of a coefficient includes zero.
3. Interpret adjusted $R^2$ and the overall F-test.
4. Give two reasons why the coefficients cannot be read as causal effects.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep02-base-model
m1 <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
summary(m1)
```

1. Among the 1,441 recorded pitches, at fixed values of the other included
   predictors:
   - one later season is associated with about 0.836 percentage points less
     equity offered, on average
   - an extra \$1,000 requested is associated with about 0.0029 percentage
     points less (about 0.29 points per \$100,000), on average
   - one additional description word is associated with about 0.128
     percentage points less, on average.
2. `confint(m1)` shows that the interval for the description words, about
   $(-0.351, 0.095)$, includes zero.
3. $R^2\approx0.188$: the model explains about 19% of the variation in the
   equity offered. Adjusted $R^2\approx0.186$ is almost the same, so the
   fit does not come from extra predictors that only fit chance variation.
   The overall F-test ($F\approx110.6$ on 3 and 1437 df) rejects the null
   hypothesis that all three slopes are zero.
4. Any two of these: only selected pitches reach the show, the quality of
   the venture is not measured, the equity offered may influence the amount
   requested, and the variables may be measured with error.
:::

### Task 2: Dummy variables and a partial F-test (Lecture 5)

Use the 1,434 pitches with a known team classification (`known_gender`).
Fit a reduced model of the equity offered (`equity_offered_pct`) on the
season and the requested amount (`ask_1k`). Fit a complete model that adds
the team category `gender` (all-men, all-women, or mixed). Compare the two
models with `anova()`.

1. [Shark Tank data]{.shark-tag} Name the reference category.
2. Interpret both dummy coefficients.
3. State the hypotheses of the partial F-test and decide.
4. Explain why the partial F-test and the individual t-tests answer
   different questions.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep02-gender-models
reduced <- lm(equity_offered_pct ~ season + ask_1k,
              data = known_gender)
complete <- lm(equity_offered_pct ~ season + ask_1k + gender,
               data = known_gender)
summary(complete)
anova(reduced, complete)
```

1. The reference category is all-men.
2. At the same season and requested amount, all-women presenting teams are
   fitted to offer about 1.28 equity percentage points more than all-men
   teams, and mixed teams about 0.12 points more. The unit is percentage
   points, not percent. The categories are broad and the data do not come
   from an experiment, so the result does not show a causal effect of
   gender.
3. $H_0$: both dummy coefficients are zero. $H_1$: at least one of them is
   not zero. $F\approx3.76$ and p $\approx$ 0.023, so reject $H_0$ at the
   5% level. The two terms are jointly significant.
4. The t-test of the mixed dummy alone (p $\approx$ 0.82) is far from
   significant, but the F-test rejects. The F-test asks whether the two
   terms together improve the model. Each t-test asks about one comparison
   with the reference category.
:::

### Task 3: Interaction (Lecture 6)

With the 1,434 pitches with a known team classification (`known_gender`),
fit a regression of the equity offered (`equity_offered_pct`) on the
centred season `season_c`, the dummy `women` (1 for all-women and mixed
teams), their interaction, and the requested amount (`ask_1k`).

1. [Shark Tank data]{.shark-tag} Write the fitted line for `women = 0` and for `women = 1` at the
   average requested amount.
2. Calculate both season slopes.
3. Interpret the interaction estimate and its p-value without claiming too
   much in either direction.
4. Explain what centring does.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep02-interaction
m_interaction <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
summary(m_interaction)
```

1. At the average requested amount, the fitted line for all-men teams is
   $12.866-0.799\,\text{season\_c}$, and for teams with women it is
   $13.673-0.861\,\text{season\_c}$, as in Lecture 6. The intercept 12.866
   is the intercept 13.625 plus the `ask_1k` coefficient times the average
   requested amount, `mean(known_gender$ask_1k)`. For teams with women the
   intercept is $12.866+0.807=13.673$.
2. The season slope is about $-0.799$ percentage points per season for
   all-men teams. For teams with women it is $-0.7994-0.0615\approx-0.861$.
3. The estimated difference in slopes is $-0.061$ percentage points per
   season, with p $\approx$ 0.520. The model gives no evidence that the
   season trends differ between the two groups. It also does not prove that
   the trends are equal, because the estimate is imprecise.
4. Centring makes the `women` coefficient compare the groups at the average
   season (about 8.8) and not at Season 0, which does not exist. It changes
   neither the fitted values nor the interaction estimate.
:::

### Task 4: Polynomial term (Lecture 6)

With all 1,441 pitches (`sharks`), fit a linear model of the equity offered
(`equity_offered_pct`) on the season and the requested amount (`ask_1k`), and a quadratic model that adds the square of
`season`, written `I(season^2)`. Compare the two models with `anova()`.

1. [Shark Tank data]{.shark-tag} Why must you write `I(season^2)` in the R formula?
2. Decide with the partial F-test whether the quadratic term improves the
   model.
3. Calculate the fitted change from Season 10 to Season 11.
4. State why you should not extrapolate the curve beyond Season 16.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: iep02-polynomial
m_linear <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
m_quadratic <- lm(
  equity_offered_pct ~ season + I(season^2) + ask_1k,
  data = sharks
)
summary(m_quadratic)
anova(m_linear, m_quadratic)
```

1. `I()` tells R to calculate the square. Without `I()`, R would read `^`
   as part of the formula syntax and not as a square.
2. $H_0$: the coefficient of the quadratic term is zero. The partial F-test
   gives $F\approx20.1$ with p $<$ 0.001, so reject $H_0$ at the 5% level.
   The quadratic term improves the model.
3. With $\hat\beta_1\approx-1.713$ and $\hat\beta_2\approx0.051$, the
   change from Season 10 to 11 is
   $\hat\beta_1+\hat\beta_2(2\times10+1)=\hat\beta_1+21\hat\beta_2\approx-0.64$
   percentage points. Never read $\hat\beta_1$ alone while the quadratic
   term is in the model.
4. Beyond Season 16 the fitted curve turns upward, and no data support that
   turn. A quadratic curve can change direction sharply outside the observed
   range.
:::

### The models of Tasks 2 to 4 in one table {#iep02-spec-table}

[Shark Tank data]{.shark-tag} Lecture 4 showed how research papers put several models side by side in one
regression table. The table below does this for the models of Tasks 2 to 4.
All columns of such a table must use the same observations. Tasks 2 and 3
use the 1,434 pitches with a known team classification (`known_gender`).
Task 4 uses all 1,441 pitches. So first fit the quadratic model again on the
1,434 pitches. The code then prints the two model outputs that the answers
above do not show.

```{r}
#| label: iep02-table-models
m_quadratic_known <- lm(
  equity_offered_pct ~ season + I(season^2) + ask_1k,
  data = known_gender
)
summary(reduced)
summary(m_quadratic_known)
```

Column (1) is the reduced model of Task 2 (`reduced`), column (2) the
complete model of Task 2 (`complete`), column (3) the interaction model of
Task 3 (`m_interaction`), and column (4) the quadratic model
(`m_quadratic_known`).

::: {.reg-table}
| | (1) | (2) | (3) | (4) |
|:---|---:|---:|---:|---:|
| Season | −0.815\*\*\* | −0.824\*\*\* | −0.799\*\*\* | −1.713\*\*\* |
| | (0.047) | (0.048) | (0.066) | (0.206) |
| Season squared | | | | 0.051\*\*\* |
| | | | | (0.011) |
| Amount requested (\$1,000s) | −0.0029\*\*\* | −0.0027\*\*\* | −0.0027\*\*\* | −0.0027\*\*\* |
| | (0.0006) | (0.0006) | (0.0006) | (0.0006) |
| All-women team | | 1.284\*\* | | |
| | | (0.481) | | |
| Mixed team | | 0.121 | | |
| | | (0.541) | | |
| Team with women (all-women or mixed) | | | 0.807 | |
| | | | (0.412) | |
| Season × team with women | | | −0.061 | |
| | | | (0.096) | |
| Intercept | 21.185\*\*\* | 20.839\*\*\* | 13.625\*\*\* | 24.168\*\*\* |
| | (0.484) | (0.506) | (0.335) | (0.822) |
| Observations | 1,434 | 1,434 | 1,434 | 1,434 |
| $R^2$ | 0.187 | 0.191 | 0.189 | 0.198 |
| Adjusted $R^2$ | 0.186 | 0.189 | 0.187 | 0.197 |
:::

Outcome: equity offered (%). Ordinary least squares. Standard errors in
parentheses. All four columns use the same 1,434 pitches with a known team
classification. The reference category is all-men teams. In column (3) season
is centred at its mean (about 8.8), so the intercept and the coefficient of
"Team with women" refer to the average season.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001.

Read along the Season row. The coefficient is close to $-0.8$ in
columns (1) to (3). In column (4) it is $-1.713$, because with the quadratic term in
the model the season coefficient alone is not the trend. Read down a column
to see one model. Columns (2), (3), and (4) each add terms to column (1), so
a partial F-test can compare each of them with column (1). The bottom rows
show that the four models fit almost equally well: adjusted $R^2$ lies
between 0.186 and 0.197, a difference of about one percentage point of the
variation in the equity offered. Use this layout for regression results in
your thesis.

### Task 5: Read a regression table from a paper

The table below is invented. It has the layout of a table in a research
paper, the same layout as the table above. The outcome is
entrepreneurial intention on a scale from 1 to 7, measured in a survey of
265 students. Self-efficacy is a score from 1 to 7 for how confident a
student is that he or she can start a business. The table does not show the
intercept. Some journals do this.

::: {.reg-table}
| | (1) | (2) |
|:---|---:|---:|
| Age (years) | 0.02 | 0.01 |
| | (0.02) | (0.02) |
| Female (1 = yes) | −0.21\* | −0.15 |
| | (0.10) | (0.10) |
| Prior entrepreneurial experience (1 = yes) | | 0.42\*\* |
| | | (0.13) |
| Entrepreneurial self-efficacy (1--7) | | 0.38\*\*\* |
| | | (0.07) |
| Observations | 265 | 265 |
| $R^2$ | 0.03 | 0.19 |
| Adjusted $R^2$ | 0.02 | 0.18 |
:::

Outcome: entrepreneurial intention (scale from 1 to 7). Ordinary least
squares. Standard errors in parentheses.
\* p < 0.05, \*\* p < 0.01, \*\*\* p < 0.001.

1. What do the numbers in parentheses and the stars mean? Check the stars
   of the self-efficacy coefficient with its t statistic. The critical
   value is $t_{0.975,\,260}$ = `qt(0.975, 260)` = 1.969.
2. Interpret the self-efficacy coefficient in column (2) in one complete
   sentence.
3. The coefficient of Female changes from $-0.21$ with a star in column (1)
   to $-0.15$ without a star in column (2). Explain what happened with the
   ideas of Lecture 4, and say what the missing star does and does not
   mean.
4. $R^2$ rises from 0.03 to 0.19. What does that say, and which test would
   show whether the two added predictors are jointly significant?

::: {.callout-note collapse="true"}
## Check your answer

1. The numbers in parentheses are standard errors. The stars show the
   p-value of the two-sided t-test of each coefficient against zero. For
   self-efficacy, $t_{obs}=0.38/0.07\approx5.4$. This is far beyond the
   critical value $t_{0.975,\,260}$ = `qt(0.975, 260)` = 1.969, and p is
   below 0.001, so three stars are correct.
2. Among the 265 surveyed students, a self-efficacy score that is one point
   higher is associated with an entrepreneurial intention that is 0.38
   points higher, holding age, gender, and prior experience fixed.
3. Model (2) holds two more variables fixed than model (1). The Female
   coefficient now compares women and men with the same experience and
   self-efficacy. In column (1), at the same age, the difference between women
   and men was 0.21 points. Part of this difference went together with differences in experience and self-efficacy. The missing star says that the
   confidence interval around $-0.15$ includes zero. It does not show that
   the difference is zero.
4. The two added predictors explain a further 16 percentage points of
   the variation in intentions in this sample. A partial F-test of the two
   nested models would show whether the two added predictors are jointly
   significant. It compares model (1) with model (2) on the same 265 students. If
   you have the data, `anova()` of the two fitted models gives this test.
:::

### Task 6: A/B test, one-way ANOVA, moderator and mediator (Lectures 5 and 6)

&#40;a) An invented A/B test. An online shop for refurbished phones shows 120
new visitors either its current product page (`new_page = 0`) or a new page
with customer videos (`new_page = 1`). Chance decides which page each
visitor sees, 60 visitors per page. The outcome is the value of the first
order in euros.

```{r}
#| label: iep02-ab-test
set.seed(21)
# random assignment: 60 zeros (old page) and 60 ones (new page)
new_page <- sample(rep(c(0, 1), each = 60))
# invented order values in euros, with a true effect of 3 euros
order_value <- round(40 + 3 * new_page + rnorm(120, mean = 0, sd = 8), 2)
ab_shop <- data.frame(new_page, order_value)
```

Regress `order_value` on the treatment dummy `new_page`. Test in five steps
whether the new page changes the mean order value. The critical value is
$t_{0.975,\,118}$ = `qt(0.975, 118)` = 1.980.
Explain why $\hat\beta_1$ estimates the average treatment effect of the new
page, and judge its size.

&#40;b) An invented one-way ANOVA. The same shop records the delivery time in
hours of 20 orders for each of three couriers.

```{r}
#| label: iep02-anova
set.seed(7)
courier <- factor(rep(c("cycle", "post", "van"), each = 20))
delivery_hours <- round(c(rnorm(20, 20, 6), rnorm(20, 26, 6),
                          rnorm(20, 22, 6)), 1)
deliveries <- data.frame(courier, delivery_hours)
```

Fit the regression of `delivery_hours` on `courier` and make its ANOVA table
with `anova()`. Which dummies does this regression use? What do its two
dummy coefficients say? Test at 5% whether the mean delivery time is the
same for the three couriers. Give $H_0$, $F$, both degrees of freedom, and
the conclusion. The critical value is `qf(0.95, 2, 57)` = 3.16.

&#40;c) A meal-kit start-up gives a discount coupon at random to half of its
customers. Customers with the coupon order more meal kits. The coupon raises
orders more for students than for other customers. The coupon also brings
customers to the website more often, and more visits go with more orders.
Which variable is a moderator, and which is a mediator? How does each one
appear in a regression of orders on the coupon dummy?

::: {.callout-note collapse="true"}
## Check your answer

&#40;a)

```{r}
#| label: iep02-ab-fit
ab_fit <- lm(order_value ~ new_page, data = ab_shop)
summary(ab_fit)$coefficients
qt(0.975, df = 118)
```

The mean order value is 39.09 euros with the old page (the intercept)
and $39.09+3.41=42.50$ euros with the new page.

1. **Hypotheses.** $H_0:\beta_1=0$, the new page has no effect on the mean
   order value. $H_1:\beta_1\ne0$. The significance level is 5%.
2. **Test statistic.** $T=\hat\beta_1/\operatorname{SE}(\hat\beta_1)$, with
   $df=n-k-1=118$.
3. **Critical value and rejection region.** $t_{0.975,\,118}=1.980$, from
   `qt(0.975, df = 118)`. Reject $H_0$ if $|t_{obs}|\ge1.980$.
4. **Observed value.** $t_{obs}=3.409/1.341\approx2.54$.
5. **Conclusion.** $2.54\ge1.980$, so reject $H_0$ (p $\approx$ 0.012). At
   the 5% level, the new page raises the mean order value.

Chance decided who saw the new page. So the two groups are alike on average
in everything else, measured or not, and the only systematic difference is
the page. That is why $\hat\beta_1\approx3.41$ euros estimates the average
treatment effect of the new page. It is about 8.7% of the mean order value
with the old page.

&#40;b)

```{r}
#| label: iep02-courier-fit
courier_fit <- lm(delivery_hours ~ courier, data = deliveries)
coef(courier_fit)
anova(courier_fit)
mean(deliveries$delivery_hours)   # overall mean
qf(0.95, 2, 57)
```

The ANOVA table comes from the regression of `delivery_hours` on two
dummies, `post` and `van`, with `cycle` as the reference category. The
intercept 22.65 hours is the mean of the cycle courier. The post courier
takes on average 3.94 hours longer, and the van courier 0.02 hours less.

1. **Hypotheses.**
   $H_0:\mu_{\text{cycle}}=\mu_{\text{post}}=\mu_{\text{van}}$.
   $H_1$: at least one mean differs. The significance level is 5%.
2. **Test statistic.** $F$ = mean square of `courier` divided by the mean
   square of the residuals, with $G-1=2$ and $n-G=57$ degrees of freedom.
3. **Critical value and rejection region.** The critical value is 3.16,
   from `qf(0.95, 2, 57)`. Reject $H_0$ if $F_{obs}\ge3.16$.
4. **Observed value.** $F_{obs}=103.62/35.11\approx2.95$.
5. **Conclusion.** $2.95<3.16$ (p $\approx$ 0.060), so do not reject
   $H_0$. At the 5% level, the data do not show that the mean delivery time
   differs between the couriers. This does not prove that the means are
   equal.

The largest difference, 3.95 hours between the post and the van courier,
is about 16% of the overall mean of 23.96 hours.

&#40;c)

- Student status is the moderator: it changes the slope of the coupon.
  It enters as an interaction, `lm(orders ~ coupon * student)`, and the
  coefficient of `coupon:student` is the difference in the coupon effect
  between students and other customers.
- Website visits are the mediator: they lie on the path from the coupon to
  orders. Adding visits, `lm(orders ~ coupon + visits)`, shrinks the coupon
  coefficient to the direct part. For the total effect of the coupon, leave
  visits out of the model.
:::

:::

::: {.course-phase .phase-after}

## After class

### Individual practice

[Shark Tank data]{.shark-tag} Build one model of the equity offered (`equity_offered_pct`) in the Shark
Tank file. Use season, the requested amount in \$1,000s, the presenting
team, and one extension that you can justify: an interaction of season with
the dummy `women` (1 for all-women and mixed teams), or a quadratic term in season next to the three known
team categories (`gender`).

Your answer must include:

1. a precise research question
2. the code that prepares the data, so that someone else can repeat it
3. one graph of the data before you fit the model
4. the R formula and the model summary. If you compare several models, show
   them in one regression table in the layout of this tutorial
5. interpretations of the coefficients with units and reference categories
6. a t-test of one coefficient and an overall or partial F-test
7. $R^2$, adjusted $R^2$, and residual diagnostics
8. one graph of fitted values from the model
9. a conclusion of 150 words that clearly separates three things: the
   association, its uncertainty, and the reasons why it is not causal.

::: {.callout-note collapse="true"}
## Check a model structure

This model structure covers points 4 to 9.

4. A good solution could fit

   ```r
   m <- lm(
     equity_offered_pct ~ season_c * women + ask_1k,
     data = known_gender
   )
   ```

   or a model with a quadratic term for season and the group dummies. Keep
   `season_c` and `women` in the model next to their interaction (the
   hierarchy principle).
5. Name the reference category. Interpret slopes that combine several
   coefficients, or differences between fitted values.
6. Use `anova()` to compare the model with a nested reduced model.
7. Inspect the residuals, including `rstandard()` and Cook's distance.
8. Plot fitted values inside Seasons 1--16.
9. The conclusion must say that the team categories are broad, that only
   broadcast pitches are in the file, and that the results therefore do not
   show a causal effect of gender or season.
:::

### Check your understanding

Complete from memory:

1. A coefficient in a multiple regression compares cases that ...
2. In a model with an intercept, R includes a factor with three categories
   as ...
3. In the model of the equity offered on the centred season, the dummy
   `women`, their interaction, and the requested amount
   (`season_c * women + ask_1k`), the season slope for teams with women
   is ...
4. Centring a predictor changes ..., and it does not change ...
5. In a model with $x$ and $x^2$, the fitted change from $x$ to $x+1$ is ...

::: {.callout-note collapse="true"}
## Check your answers

1. Differ in that predictor and have the same values of the other included
   predictors.
2. Two dummy variables. The third category is the reference category.
3. The coefficient of `season_c` plus the coefficient of `season_c:women`.
4. The meaning and the value of the intercept and of the coefficients that
   are read at $x=0$. It does not change the fitted values or the
   interaction coefficient.
5. $\hat\beta_1+\hat\beta_2(2x+1)$.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA3 venture, or any business you know well. Describe in two
sentences the full analysis that you would run on its data. First, name the
outcome, two predictors, and one extension (a set of dummies, an
interaction, or a quadratic term) that answers a real question. Second,
say which limit you would mention first in your conclusion: selection,
measurement, or omitted variables. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "Outcome: weekly orders. Predictors: advertising spend and price, plus a dummy for exam weeks and its interaction with advertising spend, because I believe advertising works mainly when students are on campus. The first limit is omitted variables: the weather affects both how many flyers I hand out and how many people are on campus, and nothing in my data measures it."
:::

You can discuss this task, and any other task in this tutorial, with
Tilly for QM Intermediate, the course AI assistant. Ask it to review your two sentences
before you spend a lot of time on the data.
::::

### More practice

For extra exercises on the lectures of this tutorial, with worked solutions,
ask Tilly for QM Intermediate, the course AI assistant. For example: "Give me
an extra exercise for Intermediate Tutorial 2. Show the solution only after
I send my answer."

This is the end of the QM Intermediate lectures and tutorials. Review the
[QM Intermediate Code Sheet](code-sheet.qmd), attempt the
[QM Intermediate Mock Exam](mock-exam.qmd), then return to the
[R Refresher](../resources/r-refresher.qmd),
[Mathematics Refresher](../resources/math-refresher.qmd), or the
[External Video Guide](../resources/video-guide.qmd) whenever you want to
review a specific topic.

[Continue to QM Intermediate Code Sheet](code-sheet.qmd).

:::
