# Tutorial 2: Multiple regression and extensions {#i-ep02}

```{r}
#| label: iep02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 2, you can</span>

- fit and interpret multiple regression in R;
- test individual and grouped coefficients against zero;
- interpret dummy, interaction, and polynomial terms correctly;
- compare nested models with a partial F-test;
- write a complete TestVision-ready analysis with design limits.
:::

Tutorials are **exam preparation sessions**: every task below follows the
format of the final exam. The standing routine for each task is the same ---
the TA states the task, pairs work on it first for a few minutes with AI use
allowed, and the solution is then discussed in plenary, starting from the
pair attempts.

One note on the order. The tasks run in course order: Task 1 rests on
Lecture 4, Task 2 on Lecture 5, and Tasks 3--4 on yesterday's Lecture 6.
This is deliberate --- the tutorial meets the day after Lecture 6, so you
warm up on settled material before testing what you learned yesterday.

::: {.course-phase .phase-before}

## Before class

Complete the after-class practice from Lectures 4--6. Use the complete
[QM Intermediate Code Sheet](#intermediate-code-sheet) while working, and
review
[Mathematics Refresher: multivariable equations, indicators, products, and powers](../resources/math-refresher.html)
if needed.

Complete from memory:

1. A multiple-regression slope is interpreted while ...
2. A dummy coefficient compares ...
3. An interaction coefficient is a difference in ...
4. A partial F-test compares ...

::: {.callout-note collapse="true"}
## Check your preparation

1. Holding the other included predictors fixed.
2. A category with the named reference category.
3. Slopes.
4. A reduced nested model with a complete model on the same observations.
:::

:::

::: {.course-phase .phase-in-class}

## In class

Attempt each task in R before opening the answer. All tasks use the Shark
Tank teaching file and the running money unit of **thousands of dollars**.
Build the analysis data once --- the same constructions as Lectures 4--6:

```{r}
#| label: iep02-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$ask_1k <- sharks$ask_amount_usd / 1000
known_gender <- subset(sharks, pitcher_gender_group != "unknown")
known_gender$gender <- relevel(factor(known_gender$pitcher_gender_group),
                               ref = "all_men")
known_gender$ask_1k <- known_gender$ask_amount_usd / 1000
known_gender$women <- as.integer(
  known_gender$pitcher_gender_group %in% c("all_women", "mixed")
)
known_gender$season_c <- known_gender$season - mean(known_gender$season)
```

### Task 1 --- Multiple-regression coefficients (Lecture 4)

Fit:

```{r}
#| label: iep02-base-model
m1 <- lm(
  equity_offered_pct ~ season + ask_1k + description_words,
  data = sharks
)
summary(m1)
```

1. Interpret all three slopes with units and conditioning.
2. Identify which 95% coefficient interval includes zero.
3. Interpret adjusted $R^2$ and the overall F-test.
4. State two threats to causal interpretation.

::: {.callout-note collapse="true"}
## Check your answer

At fixed values of the other included predictors, among the 1,441 recorded
pitches: one later season is associated with about 0.836 percentage points
less initial equity; an extra \$1,000 requested with about 0.0029
percentage points less equity (about 0.29 points per \$100,000); and one
additional description word with about 0.128 percentage points less equity.
The description-word interval, about $(-0.351,\; 0.095)$, includes zero.
Adjusted $R^2\approx0.186$, and the overall F-test
($F\approx110.6$ on 3 and 1437 df) rejects the joint zero-slope null.
Selection into the show, unmeasured venture quality, reverse relationships,
and measurement choices are all possible causal threats.
:::

### Task 2 --- Dummy variables and a partial F-test (Lecture 5)

```{r}
#| label: iep02-gender-models
reduced <- lm(equity_offered_pct ~ season + ask_1k,
              data = known_gender)
complete <- lm(equity_offered_pct ~ season + ask_1k + gender,
               data = known_gender)
summary(complete)
anova(reduced, complete)
```

1. Name the reference category.
2. Interpret both dummy coefficients.
3. State and decide the partial F-test.
4. Explain why the partial test and the individual t-tests answer different
   questions.

::: {.callout-note collapse="true"}
## Check your answer

The reference is all-men. At the same season and requested amount,
all-women presenting teams are fitted to offer about 1.28 equity percentage
points more, and mixed teams about 0.12 points more, than all-men teams ---
percentage points, not percent. The partial null is that both dummy
coefficients equal zero; $F\approx3.76$, p $\approx$ 0.023, so reject at
5%: the terms are jointly useful even though the mixed dummy's own t-test
(p $\approx$ 0.82) is far from significant. The F-test asks whether the
pair improves the model together; each t-test asks about one
reference-group contrast. The broad source coding and observational design
prohibit a causal gender claim.
:::

### Task 3 --- Interaction (Lecture 6)

```{r}
#| label: iep02-interaction
m_interaction <- lm(
  equity_offered_pct ~ season_c * women + ask_1k,
  data = known_gender
)
summary(m_interaction)
```

1. Write the fitted line for `women = 0` and `women = 1` at a fixed
   request.
2. Calculate both season slopes.
3. Interpret the interaction estimate and p-value without overclaiming in
   either direction.
4. Explain the role of centering.

::: {.callout-note collapse="true"}
## Check your answer

For all-men teams the season slope is about $-0.799$ percentage points per
season; for teams with women represented it is
$-0.799-0.061=-0.861$. The estimated slope difference is $-0.061$
percentage points per season with p $\approx$ 0.520: this model provides no
evidence of different season trends between the two source-defined groups
--- and no proof of equal trends, since the estimate is imprecise.
Centering makes the lower-order group coefficient compare the groups at the
average season (about 8.8) instead of at the unobserved Season 0; it
changes neither the fitted values nor the interaction estimate.
:::

### Task 4 --- Polynomial term (Lecture 6)

```{r}
#| label: iep02-polynomial
m_linear <- lm(equity_offered_pct ~ season + ask_1k, data = sharks)
m_quadratic <- lm(
  equity_offered_pct ~ season + I(season^2) + ask_1k,
  data = sharks
)
summary(m_quadratic)
anova(m_linear, m_quadratic)
```

1. Why must `I(season^2)` be used in the R formula?
2. Decide whether the quadratic term adds fit under the classical test.
3. Calculate the fitted change from Season 10 to Season 11.
4. State why the curve should not be extrapolated beyond Season 16.

::: {.callout-note collapse="true"}
## Check your answer

`I()` tells R to square arithmetically rather than read `^` as formula
syntax. The partial F-test gives $F\approx20.1$ with p $<$ 0.001, so the
quadratic term adds detectable fit. With $b_1\approx-1.713$ and
$b_2\approx0.051$, the change from Season 10 to 11 is
$b_1+b_2(2\times10+1)=b_1+21b_2\approx-0.64$ percentage points --- never
read $b_1$ alone while the squared term is in the model. Beyond Season 16
the fitted parabola turns upward with no data support; polynomials can
behave sharply outside the observed range.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed integrated task --- 30 minutes

Develop one model of `equity_offered_pct` using season, the requested
amount in $1,000s, the three known source gender groups, and one justified
extension.

Your submission must include:

1. a precise research question;
2. reproducible data preparation;
3. one pre-model visual;
4. the R formula and model summary;
5. coefficient interpretations with units and reference groups;
6. an individual t-test and an overall or partial F-test;
7. $R^2$, adjusted $R^2$, and residual diagnostics;
8. one graph of fitted conditional predictions;
9. a 150-word conclusion separating association, uncertainty, and causal
   limits.

::: {.callout-note collapse="true"}
## Check a model structure

A defensible solution could fit

```r
m <- lm(
  equity_offered_pct ~ season * gender + ask_1k,
  data = known_gender
)
```

or a quadratic-season model with group dummies. Keep lower-order terms,
name the reference group, interpret combined slopes or fitted contrasts,
compare a nested reduced model with `anova(reduced, complete)`, inspect
residuals (including `rstandard()` and Cook's distance), and plot
predictions within Seasons 1--16. The conclusion must state that the
source's broad presenter classification and the selected televised records
do not identify a causal effect of gender or season.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA3 venture --- or any business you know well. Sketch, in
two sentences, the integrated analysis you would actually run on its data:
first name the outcome, two predictors, and the one extension (a dummy set,
an interaction, or a quadratic) that a real question motivates, then say
which single caveat --- selection, measurement, or omitted variables ---
would headline your 150-word conclusion. There is no wrong answer; the
point is noticing that you now own the entire pipeline.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Outcome: weekly orders; predictors: ad spend and price, plus an exam-week dummy interacted with ad spend, because I believe ads work mainly when students are on campus. The headlining caveat is omitted variables: weather drives both my flyer effort and campus foot traffic, and nothing in my data measures it."
:::

You can discuss this task --- and any other task in this tutorial --- with
**Tilly**, the course AI assistant on Canvas: ask it to referee your
two-sentence design before you invest a weekend in the data.
::::

QM Intermediate is complete. Review the
[QM Intermediate Code Sheet](#intermediate-code-sheet), attempt the
[QM Intermediate Mock Exam](#intermediate-mock-exam), then return to the
[R Refresher](../resources/r-refresher.html),
[Mathematics Refresher](../resources/math-refresher.html), or the
[External Video Guide](../resources/video-guide.html) whenever you need a
targeted review.

:::
