# Lecture 1: Introduction to data analysis {#b01}

```{r}
#| label: b01-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 1, you can</span>

- organise a reproducible R analysis and check work that an AI helped with
- distinguish categorical, ordinal, discrete, continuous, and identifier
  variables
- identify a sample, a target population, sampling variability, and selection
  bias
- inspect and check a real dataset and state its limits
- describe one variable with frequency tables, bar charts, and histograms
- read a histogram by hand, and use cumulative bar counts to find the
  interval that contains the median.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- say what one row of a dataset is and which group the data describe
- classify a variable as categorical, ordinal, discrete, continuous, or an identifier
- tell sampling variability apart from selection bias in a short example
- read a histogram by hand and find the interval that contains the median or a quartile.

**Enough to recognise:**

- the R commands on this page (`dim()`, `str()`, `table()`): the exam shows R output, you do not write code
- the six steps of the analysis cycle.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Shark Tank turns a business decision (fund this venture or not) into recorded data: 1,441 pitches, each with an amount asked for, an equity offer, and an outcome. Investors, founders, and analysts use records like these to see what works.</p>

The Shark Tank file records 1,441 televised pitches. What can this dataset
tell us about entrepreneurship, and what can it never tell us?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 35 minutes.</span>

**Minimum path:** the setup and the prediction. **Full path:** everything in this section.

**Read (5 minutes).** Read *How to use this book* at the start of this book
and locate the sections called **Before class**, **Lecture material**, and
**After class** in this chapter.

[Shark Tank data]{.shark-tag} **Set up (15 minutes).** Open RStudio by opening an R project. Create a new
script and save it before you write code. Then
[download `shark_tank_teaching.csv`](../data/shark_tank_teaching.csv){download="shark_tank_teaching.csv"}
and place it in a folder called `data` inside your project. Keep the
downloaded file unchanged. Run the following commands and bring the output
to class:

```{r}
#| label: b01-preparation
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv",
                   stringsAsFactors = FALSE)

dim(sharks)
names(sharks)
head(
  sharks[c(
    "pitch_id", "startup_name", "season", "industry",
    "pitcher_gender_group", "deal_on_show"
  )],
  3
)
```

**Predict (5 minutes).** Answer this question before you come to class: what do you think one row in this
dataset represents? You find the answer in the lecture material. It is fine
if you change your answer later.

<div class="r-refresher">
<span class="label">R refresher</span><br>
You used scripts, CSV files, data frames, summaries, and plots in
<a href="https://walshc.github.io/ebi-prog/">Programming for E&amp;BI</a>.
If the commands above feel unfamiliar, revisit its chapters on
<a href="https://walshc.github.io/ebi-prog/r-scripts.html">R scripts</a>,
<a href="https://walshc.github.io/ebi-prog/reading-csv-datasets.html">reading CSV files</a>, and
<a href="https://walshc.github.io/ebi-prog/dataframes-summary-statistics.html">summarising data frames</a>.
</div>

If percentages, units, or graph axes are unfamiliar, review
[Mathematics Refresher: numbers, proportions, percentages, and units](../resources/math-refresher.html#numbers-proportions-percentages-and-units).

### Read before class: working reproducibly (5 minutes)

#### Reproducible scripts

A reproducible script recreates the analysis in a new R session. It does not
need objects that you created earlier at the Console.

Use these habits from the start:

- open the project rather than changing the working directory with `setwd()`
- keep downloaded data unchanged
- use relative paths such as `data/shark_tank_teaching.csv`
- write every step (importing, checking, transforming, plotting, and
  modelling) in scripts
- give objects meaningful names
- keep code and written interpretation together
- restart R and run the complete script before you share the analysis.

#### The analysis cycle

Data analysis moves through six steps, and you often go back to an earlier
step:

1. Ask a focused question.
2. Define the unit (what one row of the data stands for) and decide how to
   measure each idea.
3. Obtain or create the data and write down where they come from.
4. Inspect and check the data before calculating.
5. Summarise and visualise patterns.
6. Interpret the result in context and state its limits.

### Read before class: the variables in the Shark Tank file (5 minutes) {#b01-variable-table}

[Shark Tank data]{.shark-tag} One row of the file is one recorded pitch. The table lists all 19 variables
of the file, in the order of its columns. Have the table open when you run
the code above.

The Type column uses four words that the lecture material defines in the
section [From ideas to variables](#b01-variable-types). For now, read them
like this. A categorical variable records a group. A discrete or continuous
variable records a number. An identifier only labels a row.

| Variable | Meaning | Type |
|---|---|---|
| `pitch_id` | number that labels the row | identifier |
| `startup_name` | name of the venture or product | identifier (a name) |
| `season` | television season, 1--16 | discrete, ordered in time |
| `episode` | episode number within the season | discrete, labels the episode |
| `industry` | industry category given by the source | categorical |
| `business_description` | short written description of the venture, written by the source of the data | text, not analysed in this course |
| `description_words` | number of words in `business_description` | discrete (a count) |
| `pitcher_gender_group` | `all_women`, `all_men`, `mixed`, or `unknown`, as classified by the source | categorical |
| `women_represented` | 1 for `all_women` or `mixed`, 0 for `all_men`, missing for `unknown` | categorical with two groups, stored as 0/1 |
| `multiple_entrepreneurs` | 1 if more than one entrepreneur presented, 0 otherwise | categorical with two groups, stored as 0/1, with missing values |
| `pitcher_state` | US state of the entrepreneurs, empty when the source does not report it | categorical |
| `ask_amount_usd` | amount of money requested | continuous, in USD |
| `equity_offered_pct` | ownership share initially offered | continuous, in percent |
| `requested_valuation_usd` | company value implied by the request | continuous, in USD |
| `deal_on_show` | 1 for an on-air deal, 0 otherwise | categorical with two groups, stored as 0/1 |
| `deal_amount_usd` | amount of money in the agreed deal | continuous, in USD, missing without a deal |
| `deal_equity_pct` | ownership share in the agreed deal | continuous, in percent, missing without a deal |
| `deal_valuation_usd` | company value implied by the agreed deal | continuous, in USD, missing without a deal |
| `number_sharks_in_deal` | number of sharks who joined the deal, 1--5 | discrete (a count), missing without a deal |

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section shows how a question becomes data, how to check a dataset, and
how to describe one variable. It uses two datasets: a short class survey and
the Shark Tank file.

### Start with a founder's question

How many students in your year want to start a business in the next ten years?
And how many of them would put real money at risk today? A short survey can
collect the answers. Founders ask the same kind of
question about their market every week: how many people want this, and how do
I know? This lecture shows how such a question becomes data, and what the data
can and cannot tell you.

### How the course works

Every lecture follows the same cycle: you prepare before class, take part in
the lecture, and practise after class. The lecture covers the most important
parts of each chapter. This text has the full material, with every step, every example, and
all the code. The
[How to succeed in this course](../resources/how-to-succeed.qmd) page shows
the weekly plan and the time each part takes.

### AI as a fallible assistant

AI systems write answers that sound right. They do not guarantee that a
calculation, source, or interpretation is correct. Use:

**Attempt → Ask → Verify → Explain**

1. **Attempt** the question and identify where you are stuck.
2. **Ask** for a hint, a question that helps you find the problem, a
   counterexample, or a review of your code.
3. **Verify** every claim against the data, R output, course notation, and
   reliable sources.
4. **Explain** the result in your own words and ask for feedback on that
   explanation.

Do not share confidential or personal material, or exam material that you
may not share. You stay responsible for every claim that you hand in.

### Collecting our own data

The class survey is short and anonymous. It has three questions. Together, the
answers of all students form a small dataset.

#### A choice with money at risk

The first question is: imagine that you receive EUR 100 and make one choice.

| Choice | Amount placed at risk | Possible final amounts |
|---|---:|---|
| Keep it | EUR 0 | EUR 100 for certain |
| Split it | EUR 50 | EUR 50 or EUR 150, each with 50% chance |
| Go all-in | EUR 100 | EUR 0 or EUR 200, each with 50% chance |

All three choices have the same average payoff if the choice is repeated many
times: EUR 100. The choices differ only in how uncertain the result of a
single choice is.

The second question is:

> How do you see yourself in general? Use 0 for "not at all willing to take
> risks" and 10 for "very willing to take risks."

The third question asks for your current belief:

> What is the probability that you will start or co-found a business at
> least once during the next ten years?

The answers can be shown in one graph. It places the general willingness to
take risks on the horizontal axis and the probability of founding a
business on the vertical axis. One anonymous point represents one student.
Before you see such a graph, predict its shape: will students who are more
willing to take risks also report a higher probability of founding a
business?

#### From ideas to variables {#b01-variable-types}

The survey turns three ideas into data. Five terms describe how this works. A
**characteristic** is the feature we want to know about, here a student's
risk attitude. A **variable** is a well-defined rule for measuring a
characteristic, here the amount of 0, 50, or 100 euros that the student puts
at risk. A **unit**, also called an observational unit, is one member of the
group we study, here one student. A **value** is one possible result of the
measurement. An **observation** is the value recorded for one unit. For
example, the variable `amount_at_risk` has the three values 0, 50, and 100,
and the observation for one student may be 50.

In formulas, from Lecture 2 on, a variable is written with a capital letter,
such as $X$, and its values with the matching lower-case letter, $x$. Here
$X$ stands for the rule of measurement, which is the same for every unit,
and $x$ for a number that the rule produced for one unit.

Variables come in different types. The type determines which tables, graphs,
and calculations make sense.

::: {.definition-box}
<span class="label">Definition</span>

**Types of variables.** The examples are from the Shark Tank file and the
class survey.

- **Categorical.** The values are groups without an order. Example:
  `industry`. With only two groups the variable is called binary and is often
  stored as 0 and 1, as in `deal_on_show`.
- **Ordinal.** The values are groups with an order, but the steps between
  them have no fixed size. Example: the answer from 0 to 10 to the second
  survey question (`risk_willingness`). The Shark Tank file has no ordinal
  variable.
- **Discrete.** The values are separate numbers, usually counts. Example:
  `number_sharks_in_deal`.
- **Continuous.** The values are numbers that can lie anywhere in a range.
  Example: `equity_offered_pct`.
- **Identifier.** The value only labels a unit. Calculating with it has no
  meaning. Example: `pitch_id`.
:::

Discrete and continuous variables are together called **quantitative**,
because their values are numbers that count or measure something. The class
survey has one variable of each of four types:

| Variable | What is recorded | Type |
|---|---|---|
| `respondent_id` | anonymous row label | identifier |
| `amount_at_risk` | 0, 50, or 100 euros | discrete |
| `risk_willingness` | one selected point from 0 to 10 | ordinal, often analysed as a number |
| `founder_probability` | belief from 0% to 100% | continuous, stored as a rounded value |

**Risk attitude** is an idea that cannot be observed directly. The money
choice and the 0--10 answer are two imperfect measurements of it. The two
measurements may disagree for three reasons. Risk taking depends on the
situation. An imagined choice differs from a choice with real money. Also,
people do not judge themselves perfectly. One answer does not show a fixed personality type.

### Sample and target population

Data almost never cover everyone we want to know about. Three terms separate
the units that we observed from the units that the question is about.

::: {.definition-box}
<span class="label">Definition</span>

**Population.** The complete set of units about which a question is asked.

**Sample.** The part of the population that is observed.

**Random sample.** A sample drawn so that every unit of the population has
the same chance of being drawn.
:::

A population is usually too large to measure every unit, so we work with a
sample. Inferential statistics asks what a sample can tell us about the
population. Lectures 7 to 10 cover it.

In the class survey, the sample consists of the students who are present and
respond. The population we want to learn about is called the **target
population**. It could be any of these five groups:

- the responding students
- everyone attending the lecture
- all students enrolled in QM Basic
- all first-year business students in the Netherlands
- all young adults in the Netherlands.

::: {.callout-note collapse="true"}
## Check: which of the five possible populations can our responses describe directly?

Only the first: the responding students. For the other groups (absent
classmates, other QM Basic students, Dutch first-year students, young
adults) we would need assumptions, because attendance and participation were
not random. The data describe the respondents. They do not automatically
represent larger groups.
:::

### Sampling variability versus selection bias

A sample can differ from its population for two reasons. Do not mix them
up:

- **Sampling variability:** even a random sample can contain an unusual
  number of students who are very willing to take risks, purely by chance.
  Larger random samples usually reduce this chance variation.
- **Selection bias:** a sample of people who were easy to reach can
  systematically leave out certain people. Collecting more responses from the
  same selected group does not remove that bias.

::: {#exm-b01-bias-contrast}
## Worked example: two scenarios

*Scenario 1.* A researcher draws 40 students at random from the list of all
students enrolled in the programme. By chance, the sample contains an unusually large
number of students who chose "go all-in."

*Scenario 2.* A student shares a survey about founding a business in the
group chat of the entrepreneurship club, and 400 members respond.

Scenario 1 is sampling variability. The selection was random, so the unusual
result is chance variation, and a larger random sample would usually reduce
it. Scenario 2 is selection bias. Club members are systematically more
interested in founding than the average student. Collecting 4,000 responses
from the same chat would not remove this bias. It would only give a more precise
answer to the wrong question.
:::

A random sample of Dutch students would require a clear definition of the
population, a list of all its members, and random selection from that list.
Even then, some students would not answer, and answers can be imprecise.

### Meet the Shark Tank data

[Shark Tank data]{.shark-tag} The class survey is a small dataset that a class creates itself. The rest of
the course uses one larger dataset: the Shark Tank file.
Examples and tasks that use the Shark Tank file carry this tag: [Shark Tank data]{.shark-tag}.

#### Context and unit of observation

*Shark Tank US* is a television programme in which entrepreneurs present a
venture or product to a panel of investors known as sharks. Entrepreneurs
may ask for money in exchange for an ownership share. The investors question
them, negotiate, and may announce a deal during the episode.

The recorded outcome is an **on-air deal**: the sharks and the entrepreneurs
agree on a deal in front of the camera. It does not show that an investment
was completed or that the venture later succeeded.

For a founder, the most interesting numbers in the file are valuations. A
valuation is the value of the whole company that a request or a deal implies.
Asking \$100,000 for 10% of a company values it at \$1 million. The file
stores the valuation implied by the founder's request in
`requested_valuation_usd`. For pitches that ended in a deal, it stores the
valuation implied by the agreed deal in `deal_valuation_usd`. The section
"Summarise in context" below compares the two.

The teaching file contains 1,441 televised pitches from completed Seasons
1--16 and 19 variables. The Before class section lists all 19. The table
below repeats the five that this lecture uses most.

| Variable | Meaning | Type |
|---|---|---|
| `deal_on_show` | 1 for an on-air deal, 0 otherwise | categorical with two groups, stored as 0/1 |
| `season` | television season, 1--16 | discrete, ordered in time |
| `description_words` | number of words in the written description | discrete (a count) |
| `ask_amount_usd` | amount of money requested | continuous, in USD |
| `equity_offered_pct` | ownership share initially offered | continuous, in percent |

The gender variables describe the people who present the pitch, not
necessarily every founder or employee. `women_represented` is not the share
of women. For mixed teams, the source does not report how many women and how
many men presented, so nobody can calculate that share.

Note carefully what `description_words` measures: the length of the
written description in the source dataset, not anything about the pitch
itself. It counts words typed by the person who created the source dataset, not
words spoken on the show. The file also does not measure the spoken pitch,
whether the investment was completed, or how the venture performed later.

#### Inspect and validate

Run the code below yourself, and read each output before you read its
interpretation. All code on this page uses the data that the first line
imports.

```{r}
#| label: b01-inspect
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
dim(sharks)
str(sharks)
```

There are 1,441 rows and 19 columns. This answers the prediction from your
preparation: one row represents one recorded televised pitch from Seasons
1--16 of *Shark Tank US*. In the words of this lecture, the unit is one
pitch, and each of the 19 columns is one variable.

```{r}
#| label: b01-validate
colSums(is.na(sharks))
range(sharks$season)
```

Seasons run from 1 to 16. The four deal variables are missing for the 559
pitches without a deal, as expected. The source does not report
`multiple_entrepreneurs` for 427 pitches, and seven records have no value for
`women_represented` because their gender classification is unknown. Missing
does not mean zero. These checks show that the recorded values are
realistic. They do not show that the source recorded every pitch or measured
every idea well.

#### Summarise in context

```{r}
#| label: b01-deal-rate
deal_count <- sum(sharks$deal_on_show)
pitch_count <- nrow(sharks)
deal_rate <- mean(sharks$deal_on_show)

c(
  on_air_deals = deal_count,
  recorded_pitches = pitch_count,
  recorded_deal_rate = deal_rate
)
```

A **proportion** is a count divided by the total. Here 882 of the 1,441
pitches reached a deal, and 882/1441 = 0.612, or 61.2%. For a variable stored
as 0 and 1, the mean (the average) equals the proportion of 1s, so
`mean(sharks$deal_on_show)` returns the same 0.612. A careful sentence is:

> In this dataset, 882 of 1,441 recorded pitches (61.2%) reached an on-air
> deal.

This sentence names the dataset and outcome. It does not claim that a
randomly selected entrepreneur has the same chance of a completed investment.

The second summary compares the valuation a founder claimed with the
valuation of the agreed deal. In 8 of the 882 deals the recorded share for
the sharks (`deal_equity_pct`) is 0%, so the deal implies no valuation. The
code keeps the other 874 deals.

```{r}
#| label: b01-valuation
share_deals <- subset(sharks, deal_on_show == 1 & deal_equity_pct > 0)
nrow(share_deals)
median(share_deals$requested_valuation_usd)
median(share_deals$deal_valuation_usd)
mean(share_deals$deal_valuation_usd < share_deals$requested_valuation_usd)
```

The summary uses the median, the middle value of the ordered data: at least
half of the values are at or below it. The section
[Reading a histogram by hand](#b01-quartiles) defines it in full. Among these
874 deals, the median claimed valuation (`requested_valuation_usd`) is \$2.0
million. The median agreed valuation (`deal_valuation_usd`) is \$1.0 million,
half as much. In 81% of these deals the agreed valuation was below the one
the founder claimed. QM Intermediate returns to this gap.

### Describing one variable

[Shark Tank data]{.shark-tag} Before comparing groups or fitting models, learn to describe one
variable well. The right tool depends on the variable's type.

#### Frequency tables and bar charts for discrete categories

An **empirical distribution** records how often values occur in a dataset.
For a categorical or discrete variable, a **frequency table** gives counts
and a **relative-frequency table** gives proportions:

```{r}
#| label: b01-season-table
season_counts <- table(sharks$season)
season_counts
round(prop.table(season_counts), 3)
```

The second table divides each count by the total. Season 5 has 116 of the
1,441 pitches, so its proportion is 116/1441 = 0.080, or 8.0%.

```{r}
#| label: fig-b01-season-bar
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Number of recorded pitches per season. Seasons 1 to 3 contain the fewest records, so an overall percentage gives them less weight."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_counts <- as.data.frame(table(sharks$season))
names(season_counts) <- c("season", "pitches")
season_counts$season <- as.integer(as.character(season_counts$season))

ggplot(season_counts, aes(x = season, y = pitches)) +
  geom_col(fill = "#009E73", alpha = 0.85, colour = "white") +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Season", y = "Number of recorded pitches") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

In @fig-b01-season-bar the horizontal axis lists the discrete season
categories. Each bar's height on the vertical count scale is the number of
recorded pitches in that season. Seasons 1 to 3 contain the fewest records
and Seasons 5 to 7 the most (116 each), so an overall percentage gives the
early seasons less weight.

#### Histograms for quantitative variables

For a quantitative variable, a **histogram** groups nearby values into
intervals and counts how many observations fall in each.

The histogram below shows the equity initially offered
(`equity_offered_pct`) in the Shark Tank file:

```{r}
#| label: fig-b01-equity-hist
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Distribution of the equity initially offered across the 1,441 recorded pitches, in 5-point intervals. Each bar includes its upper limit, and the number above a bar is its count. The dashed vertical line marks the median, 10%."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
equity_median <- median(sharks$equity_offered_pct)   # 10%

ggplot(sharks, aes(x = equity_offered_pct)) +
  geom_histogram(breaks = seq(0, 100, by = 5),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_vline(xintercept = equity_median, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.9) +
  stat_bin(breaks = seq(0, 100, by = 5), geom = "text",
           aes(label = ifelse(after_stat(count) >= 30, after_stat(count), "")),
           vjust = -0.5, colour = "#17324D", size = 3.2) +
  annotate("text", x = equity_median + 1.5, y = 470,
           label = "median = 10%\n(counted in the bar that ends at 10)",
           hjust = 0, vjust = 1, lineheight = 0.95,
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = seq(0, 100, by = 5),
                     labels = function(x) ifelse(x %% 10 == 0, x, "")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Equity initially offered (%)",
       y = "Number of recorded pitches") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    axis.ticks.x = element_line(colour = "#5B6976"),
    axis.ticks.length.x = grid::unit(4, "pt")
  )
```

Each bar counts the pitches within one interval of offered equity. Most initial
offers are below 20%, while a small number are much larger. A shape like
this, with most values small and a long tail to the right, is called
**right-skewed**. Changing the interval width can show or hide patterns, so
the analyst must choose the width with care.

### Reading a histogram by hand {#b01-quartiles}

[Shark Tank data]{.shark-tag} The exam gives you a printed histogram and no computer, so you must be able
to read a histogram by hand. Read every graph with the same
**five-step sequence**:

1. **Unit and variables.** What does one observation represent, and what is
   on each axis, in which units?
2. **Scales.** Read the axis scales, including where they start and how wide
   each interval is.
3. **One element.** Describe one bar (or point) in a full sentence.
4. **Pattern.** Describe the overall shape: where most values lie,
   whether the distribution is symmetric or skewed, whether there are
   outliers (single values far away from the rest).
5. **Limits.** State one thing the graph cannot tell you.

A histogram also shows where the middle of the data lies and how wide the
data are spread. Four terms describe this.

::: {.definition-box}
<span class="label">Definition</span>

**Median.** The middle value of the ordered data: at least half of the
observations are at or below it and at least half are at or above it.

**Lower quartile** $Q_1$ **and upper quartile** $Q_3$. At least one quarter
of the observations are at or below $Q_1$, and at least three quarters are
at or above it. At least three quarters of the observations are at or below
$Q_3$, and at least one quarter are at or above it.

**Interquartile range.** $\operatorname{IQR}=Q_3-Q_1$, the width of the
middle half of the data.
:::

With an odd number of observations the median is the single middle value.
With an even number it is the average of the two middle values. R computes
all four numbers for the equity initially offered (`equity_offered_pct`):

```{r}
#| label: b01-equity-quartiles
quantile(sharks$equity_offered_pct, c(0.25, 0.50, 0.75))
IQR(sharks$equity_offered_pct)
```

The lower quartile is 8%, the median is 10%, and the upper quartile is 20%.
The interquartile range is $20-8=12$ percentage points: the middle half of
the offers lies between 8% and 20% equity.

Many pitches offer exactly 10% or exactly 20%. For this reason 59% of the
offers (853 of 1,441) are at or below the median, and 89% (1,288 of 1,441)
are at or below $Q_3$. The table in the worked example below shows these
counts. This is why the definitions say "at least".

A simpler measure of spread is the **range**: the largest value minus the
smallest value.

The worked example below finds the median from a printed histogram, without
R.

::: {#exm-b01-histogram-read}
## Worked example: the five steps and the median interval

[Shark Tank data]{.shark-tag} Apply the sequence to @fig-b01-equity-hist.

1. One observation is one recorded pitch. The horizontal axis is equity
   initially offered in percent, the vertical axis is the number
   of pitches.
2. The horizontal axis runs from 0 to 100 in intervals of width 5. The
   vertical axis counts pitches.
3. The tallest bar covers 5--10%: 560 pitches offered more than 5% and at
   most 10% equity.
4. The distribution is strongly right-skewed: most offers are at or below
   20%, and a few values spread out towards 100% (a thin tail).
5. The histogram cannot say why so many offers lie in this range, and it
   hides the exact values inside each interval.

**Finding the median interval.** With 1,441 ordered observations, the
median is the 721st value. Add bar counts from the left until the running
total reaches 721:

| Interval (%) | Count | Cumulative count |
|---:|---:|---:|
| 0 to 5 | 293 | 293 |
| above 5, up to 10 | 560 | 853 |
| above 10, up to 15 | 194 | 1,047 |
| above 15, up to 20 | 241 | 1,288 |

After the first bar only 293 observations are covered, not yet 721. After
the second bar the running total is $293+560=853\ge721$, so the 721st
observation lies in the **5--10% interval**: the median interval. (R
confirms this: the exact median is 10%. This histogram counts the value 10%
in the 5--10 bar, because each bar includes its right endpoint.)
:::

::: {.callout-caution collapse="true"}
## Try it: locate the upper-quartile interval

[Shark Tank data]{.shark-tag} Try this first with the student next to you. Use the definition of the upper quartile
$Q_3$ in the box above.

*Step 1 (given).* Three-quarters of 1,441 observations is about 1,081, so
you are looking for the interval containing the 1,081st ordered value.

*Step 2 (you).* Extend the cumulative counts in the table above until the
running total first reaches 1,081.

*Step 3 (you).* Name the upper-quartile interval.

::: {.callout-note collapse="true"}
## Check your answer

Up to and including the 10--15% bar, the cumulative count is
$293+560+194=1{,}047$, still below 1,081. Adding the 15--20% bar gives $1{,}047+241=1{,}288\ge1{,}081$, so the
upper quartile lies in the **15--20% interval**. (R confirms this: the `quantile()`
output above gives 20 for the upper quartile, on the upper edge of that bar.)
:::
:::

::: {.caveat-box}
<span class="label">What these records can and cannot say</span>

- The file describes 1,441 pitches that producers chose to film and air. It
  is not a sample of all start-ups, or even of all applicants.
- An on-air deal is made in front of the camera. It is not a
  completed investment, and it says nothing about later success.
- A variable measures only what it records. `description_words` counts words
  in a written description, not anything about the pitch itself.

A **causal** claim says that changing one thing would change another, for
example that offering less equity would get a founder a deal. Records like
these show what went together. They do not show what would happen after a
change.

**What you can do with it:** describe what happened in these pitches and form
expectations about what happens on the show. For claims about founders in general, look for
data that were collected for that purpose.
:::

### Answering the driving question

The dataset can tell us precisely what happened in these 1,441 recorded
pitches, and anyone can repeat the calculation. Of these pitches, 882 (61.2%)
reached an on-air deal. Most initial offers gave up at most 20% equity, and the
median offer lies in the 5--10% interval. The file cannot answer three
questions: what will happen to a new founder, whether a deal on air was
completed later, and why these patterns exist.

The founder's question at the start asked how many students in your year want
to start a business. A class survey answers it only for the students who
responded, because nobody drew them at random. An answer for the whole year
needs a random sample from the enrolment list, such as the 40 students of
Scenario 1 in the worked example above. The founder in the box below has the
same problem with 40 friends.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You plan a venture that sells boxes of prepared meals on campus, and you ask 40 friends whether they would buy such a box. Thirty say yes, which is 75%. Before you order packaging for a market of that size, ask the two questions of this lecture. Who is in the sample? Friends, not the students you want to sell to. Who chose to answer? People who like you. The share among all students is probably lower than 75%. The decision: spend one lunch break and ask 50 students you do not know, for example every tenth person leaving the canteen, and plan with that number.
:::

**Next:** We can now describe what happened in these 1,441 pitches: counts,
shares, and shapes. Description alone cannot answer the questions about
chance that an investor asks. How likely is a deal if I pick a pitch at
random? How likely is a late-season pitch? How likely are both at once?
Lecture 2 builds the tool for such questions, a probability model. The model
selects one pitch at random, so that every pitch has the same chance. It
defines events such as "reached a deal" and "late season". Its rules say how
likely each event is and when you may add chances.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 80 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the most important parts. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

Answer without reopening the chapter, then check your answers.

1. What does one row represent, and which population can it describe
   directly?
2. Why is `pitch_id` an identifier rather than a quantitative measurement?
3. One founder asks 30 customers drawn by chance from her full customer
   list. Another gets 300 answers to a poll on his own Instagram page. Which
   result has sampling variability only, and which has selection bias?
4. What exactly does `description_words` measure?
5. How do you locate the median interval of a histogram by hand?

::: {.callout-note collapse="true"}
## Check your answers

1. One row is one recorded televised pitch. The file directly describes
   these 1,441 records from completed Seasons 1--16.
2. `pitch_id` labels a record. Differences between its values have no
   meaning.
3. The first has sampling variability only: the draw was random, and a
   larger random sample would reduce the chance variation. The second has
   selection bias: followers chose to answer, and more answers from the same
   group do not remove that bias.
4. The number of words in the written dataset description: the source
   text about the pitch, not the spoken pitch itself.
5. Find the position of the median (about $n/2$), then add bar counts from
   the left. The first interval whose cumulative count reaches that position
   contains the median.
:::

### Reproduce the workflow (15 minutes)

[Shark Tank data]{.shark-tag} Save the script. In RStudio, choose **Session > Restart R** to clear objects
from memory. Run the complete script from top to bottom. It should import
the data, perform the checks, calculate the deal rate, and recreate at least
one graph. It should not need any commands that you ran earlier at the
Console.

Add these five further checks to your script:

```{r}
#| label: b01-validate-more
sort(unique(sharks$deal_on_show))
range(sharks$episode)
range(sharks$description_words)
table(sharks$pitcher_gender_group, useNA = "ifany")
anyDuplicated(sharks$pitch_id)
```

Deal status contains only 0 and 1. Episodes run from 1 to 29. Written
descriptions contain 1 to 13 words. Seven records have an unknown gender
classification in the source, and no pitch identifier appears twice.

::: {.callout-note collapse="true"}
## Check the result

A clean run without errors reports 1,441 rows and 19 variables, a recorded
on-air deal rate of 882/1,441 (61.2%), and the ranges stated above. It also
recreates the graph without an "object not found" error.
:::

### Explain one graph (15 minutes)

[Shark Tank data]{.shark-tag} The two graphs below compare raw observations with group means. Read them
first.

```{r}
#| label: fig-b01-raw-points
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Equity initially offered by season, one point per recorded pitch, jittered horizontally to reduce overlap. Colour and shape mark on-air deal status."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$deal_status <- factor(sharks$deal_on_show, levels = c(0, 1),
                             labels = c("No on-air deal", "On-air deal"))
set.seed(2026)
sharks$season_jittered <- jitter(sharks$season, amount = 0.16)

ggplot(sharks, aes(x = season_jittered, y = equity_offered_pct,
                   colour = deal_status, shape = deal_status)) +
  geom_point(alpha = 0.45, size = 1.6) +
  scale_colour_manual(values = c("No on-air deal" = "#009E73",
                                 "On-air deal" = "#D55E00")) +
  scale_shape_manual(values = c("No on-air deal" = 16,
                                "On-air deal" = 17)) +
  scale_x_continuous(breaks = c(seq(1, 15, by = 2), 16)) +
  labs(x = "Season", y = "Equity initially offered (%)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Season is discrete and ordered in time. The code moves the points in
@fig-b01-raw-points slightly to the left or right only so that they overlap
less. The vertical spread within every season shows that a group mean cannot
describe every pitch.

```{r}
#| label: fig-b01-season-means
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Mean equity initially offered per season, separately for pitches with and without an on-air deal."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_means <- aggregate(
  equity_offered_pct ~ season + deal_on_show,
  data = sharks,
  FUN = mean
)
season_means$deal_status <- factor(season_means$deal_on_show,
                                   levels = c(0, 1),
                                   labels = c("No on-air deal",
                                              "On-air deal"))

ggplot(season_means, aes(x = season, y = equity_offered_pct,
                         colour = deal_status, shape = deal_status,
                         linetype = deal_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.4) +
  scale_colour_manual(values = c("No on-air deal" = "#009E73",
                                 "On-air deal" = "#D55E00")) +
  scale_shape_manual(values = c("No on-air deal" = 16,
                                "On-air deal" = 17)) +
  scale_linetype_manual(values = c("No on-air deal" = "solid",
                                   "On-air deal" = "dashed")) +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Season", y = "Mean equity initially offered (%)",
       colour = NULL, shape = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Mean initially offered equity in @fig-b01-season-means is lower in later
seasons for both deal groups. The graph describes selected televised
pitches. It does not show why the offers changed or what would happen
if an entrepreneur changed only the offered equity.

A clear graph labels its axes and its groups, so that a reader can check the
claim.

Write four sentences about either Shark Tank graph
(@fig-b01-raw-points or @fig-b01-season-means):

1. Name the variables and units.
2. Describe the main visible pattern.
3. State one interpretation that the graph supports.
4. State one important limitation or next question.

::: {.callout-note collapse="true"}
## Check a model answer

Each point in the raw-data graph represents one recorded pitch, with season
on the horizontal axis and initially offered equity on the vertical axis.
Offers vary within every season and are generally lower in later seasons.
This is evidence of an association between season and the recorded initial
offer. It does not show why offers changed or identify the causal effect of
changing an offer.
:::

### Finish a half-worked example: a histogram reading (10 minutes)

[Shark Tank data]{.shark-tag} The task uses the histogram of description length:

```{r}
#| label: fig-b01-word-hist
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Distribution of the number of words in the written dataset description. The dashed vertical line marks the median, 5 words."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
word_median <- median(sharks$description_words)   # 5 words

ggplot(sharks, aes(x = description_words)) +
  geom_histogram(breaks = seq(0.5, 13.5, by = 1),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_vline(xintercept = word_median, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = word_median + 0.4, y = 300,
           label = "median = 5 words", hjust = 0,
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = 1:13) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Words in the written dataset description",
       y = "Number of recorded pitches") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

In @fig-b01-word-hist the horizontal axis measures description length in
words and the vertical axis counts pitches in each one-word interval. The
distribution is slightly right-skewed: most descriptions contain 3--7 words,
and a small number of longer descriptions lie further to the right. The median is 5 words.
Remember what this variable measures: the source text about the pitch, not
the pitch itself.

Re-read the worked histogram reading (@exm-b01-histogram-read). Now finish
the same reading for the description-words histogram
(@fig-b01-word-hist). The bar counts, from 1 word upward, are:

| Words | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 | 13 |
|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| Count | 4 | 113 | 217 | 285 | 292 | 241 | 146 | 85 | 36 | 9 | 4 | 6 | 3 |

*Step 1 (given).* One observation is one recorded pitch. The axes show
description length in words and the pitch count. With 1,441 observations,
the median is the 721st ordered value.

*Step 2 (you).* Build the cumulative counts and locate the bar containing
the 721st observation.

*Step 3 (you).* Describe the shape and state one limit of the graph.

::: {.callout-note collapse="true"}
## Check your answer

Cumulative counts: 4, 117, 334, 619, 911, ... The running total first
reaches 721 at the 5-word bar ($619+292=911\ge721$), so the median is 5
words, which matches `median(sharks$description_words)`. The distribution is
slightly right-skewed, with most values at 3--7 words and a thin right tail. The
graph cannot say anything about the pitch itself: it describes the source's
written text.
:::

### Worked exam interpretation (5 minutes)

**Exam-style question.** An invented example. A meal-delivery
start-up records, for each of 200 orders: `order_value_eur` (the amount paid,
in euros), `items` (the number of items in the order), `district` (one of four city
districts), and `repeat_customer` (1 if the customer ordered before, 0
otherwise). Classify each variable. A histogram of `order_value_eur` has
equal-width intervals of EUR 10, starting at zero, with counts 46, 74, 48,
20, 8, 4. Each bar includes its upper limit. Which interval contains the
median?

::: {.callout-note collapse="true"}
## Check a model answer

`order_value_eur` is quantitative and continuous, `items` is quantitative
and discrete, `district` is categorical, and `repeat_customer` is
categorical with two groups (binary), even though it is stored as 0/1. With
200 observations, the median lies between the 100th and 101st ordered
values. Cumulative counts: 46 at or below EUR 10, then $46+74=120$ at or
below EUR 20. So observations 100 and 101 lie in the interval **above EUR 10,
up to EUR 20**, and the median lies there too.
:::

### Check the AI (5 minutes)

[Shark Tank data]{.shark-tag} An AI assistant writes:

> "The dataset contains 1,441 successful companies and proves that offering
> less equity makes investors fund startups."

Use **Attempt → Ask → Verify → Explain** to correct it. Your final response
must use the observed count or percentage, name the measured outcome, and
state why the causal claim is unsupported.

::: {.callout-note collapse="true"}
## Check your answer

The file contains 1,441 recorded televised pitches, of which 882 reached an
on-air deal. It does not measure completed investment or later company
success. The founders chose their own offers (`equity_offered_pct`). Nobody
assigned the offers by chance. So the records cannot show that a lower offer
would get a deal.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Write two
sentences: first name one number you would really like to know about it
(a rate, a count, an amount). Then state what data you would need to collect
to measure that number, and one way in which the data collection could be
biased. There is
no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells boxes of prepared meals on campus. "The number I most want is the share of first-order customers who order again within 30 days. I would need order records linked over time, and counting only app users would bias the share upward, because the most loyal customers are the ones who install the app."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to find weak points in your
data-collection plan or to suggest a selection bias you have not thought of.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 1. Show the solution only after I send my answer."

[Continue to Lecture 2: Probability foundations and random variables](b02-probability-foundations.qmd).

[Formula sheet: Lecture 1 formulas](formula-sheet.qmd#formula-l1).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 1--5.

:::
