# Lecture 1: Introduction to Data Analysis {#b01}

```{r}
#| label: b01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 1, you can</span>

- organise a reproducible R analysis and verify AI-assisted work;
- distinguish categorical, discrete, continuous, and identifier variables;
- identify a sample, a target population, sampling variability, and selection
  bias;
- inspect, validate, and qualify a real dataset;
- describe one variable with frequency tables, bar charts, and histograms;
- read a histogram by hand, including locating the median interval from
  cumulative bar counts.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Shark Tank turns a business decision --- fund this venture or not --- into recorded data: 1,441 pitches, each with an ask, an equity offer, and an outcome. Learning from records like these is how investors, founders, and analysts figure out what works.</p>

1,441 televised pitches are on our screen. What can this dataset tell us
about entrepreneurship --- and what can it never tell us?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Read (5 minutes).** Read *How to use this book* at the start of this book
and locate the sections called **Before class**, **In class**, and
**After class** in this chapter.

**Set up (15 minutes).** Open RStudio through an R project. Create a new
script and save it before writing code. Then
[download `shark_tank_teaching.csv`](../data/shark_tank_teaching.csv){download="shark_tank_teaching.csv"}
and place it in a folder called `data` inside your project. Keep the
downloaded file unchanged. Run the following commands and bring the output
to class:

```{r}
#| label: b01-preparation
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv",
                   stringsAsFactors = FALSE)

dim(sharks)
names(sharks)
head(
  sharks[c(
    "pitch_id", "startup_name", "season", "industry",
    "pitcher_gender_group", "deal_on_show"
  )],
  3
)
```

**Predict (5 minutes).** Write one sentence answering: **what do you think
one row in this dataset represents?** Commit to an answer; the lecture
resolves it. It is fine if your answer changes during class.

<div class="r-refresher">
<span class="label">R reactivation</span><br>
You used scripts, CSV files, data frames, summaries, and plots in
<a href="https://walshc.github.io/ebi-prog/">Programming for E&amp;BI</a>.
If the commands above feel unfamiliar, revisit its chapters on
<a href="https://walshc.github.io/ebi-prog/r-scripts.html">R scripts</a>,
<a href="https://walshc.github.io/ebi-prog/reading-csv-datasets.html">reading CSV files</a>, and
<a href="https://walshc.github.io/ebi-prog/dataframes-summary-statistics.html">summarising data frames</a>.
</div>

If percentages, units, or graph axes are unfamiliar, review
[Mathematics Refresher: numbers, percentages, and graphs](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz --- in this first
week, on the preparation reading and setup --- then turns to the driving
question: what can 1,441 recorded pitches tell us about entrepreneurship,
and what can they never tell us? The hook for the whole session is a graph
the class builds itself: the anonymous class map from the Mentimeter risk
activity below.

### How the course works

Every lecture uses the same rhythm:

1. **Prepare before class.** Encounter the question, notation, data, or short
   explanation before the lecture.
2. **Participate during class.** Predict results, discuss choices, run code,
   and revise your reasoning with the teaching team.
3. **Practise after class.** Retrieve ideas from memory, reproduce the
   analysis, and explain results without copying a model answer.
4. **Use the exam preparation sessions.** Bring your attempts and questions;
   work in groups with classmates and teaching assistants.

Complete the **Before class** section before each lecture; use
**After class** to check that you can reproduce and explain the analysis
independently.

### Data analysis with R and AI

#### Reproducible scripts

A reproducible script recreates the analysis in a new R session, without
depending on objects created earlier at the Console.

Use these habits from the start:

- open the project rather than changing the working directory with `setwd()`;
- keep downloaded data unchanged;
- use relative paths such as `data/shark_tank_teaching.csv`;
- record importing, checking, transforming, plotting, and modelling in
  scripts;
- give objects meaningful names;
- keep code and written interpretation together; and
- restart R and run the complete script before sharing the analysis.

#### The analysis cycle

Data analysis is iterative:

1. Ask a focused question.
2. Define the unit and decide how each idea will be measured.
3. Obtain or create the data and document their provenance.
4. Inspect and validate before calculating.
5. Summarise and visualise patterns.
6. Interpret the result in context and state its limits.

#### Use AI as a fallible assistant

AI systems generate plausible responses; they do not certify that a
calculation, source, or interpretation is correct. Use:

**Attempt → Ask → Verify → Explain**

1. **Attempt** the question and identify where you are stuck.
2. **Ask** for a hint, diagnostic question, counterexample, or code review.
3. **Verify** every claim against the data, R output, course notation, and
   reliable sources.
4. **Explain** the result in your own words and ask for feedback on that
   explanation.

Do not share confidential, personal, or assessment-restricted material. You
remain responsible for every submitted claim.

### Collecting our own data

The class creates its first dataset through an anonymous Mentimeter
activity.

#### A choice with money at risk

Imagine receiving EUR 100 and making one choice:

| Choice | Amount placed at risk | Possible final amounts |
|---|---:|---|
| Keep it | EUR 0 | EUR 100 for certain |
| Split it | EUR 50 | EUR 50 or EUR 150, each with 50% chance |
| Go all-in | EUR 100 | EUR 0 or EUR 200, each with 50% chance |

All three choices have the same average monetary payoff across many
repetitions: EUR 100. The options differ in how uncertain the one-time
result is.

Next, answer:

> How do you see yourself in general? Use 0 for "not at all willing to take
> risks" and 10 for "very willing to take risks."

Finally, give your current belief:

> What is the probability that you will start or co-found a business at
> least once during the next ten years?

Before the class map appears on screen, **predict its shape**: will students
who place more money at risk also report a higher founder probability? The
map places general risk willingness on the horizontal axis and the
subjective founder probability on the vertical axis. One anonymous point
represents one responding student.

#### From ideas to variables

| Variable | What is recorded | Type |
|---|---|---|
| `respondent_id` | anonymous row label | identifier |
| `amount_at_risk` | 0, 50, or 100 euros | discrete quantitative |
| `risk_willingness` | one selected point from 0 to 10 | ordered discrete rating |
| `founder_probability` | belief from 0% to 100% | continuous concept recorded on a finite-precision slider |

**Risk attitude** is a theoretical construct. The monetary choice and the
0--10 response are two imperfect measurements of it. They may disagree
because risk taking depends on context, hypothetical choices differ from
real stakes, and self-perceptions are imperfect. A single response is not a
diagnosis of a permanent personality type.

The first two responses offer separate allowable points. Founder probability
is conceptualised on the continuum from 0% to 100%, although the software
stores a rounded value.

### Sample and target population

The **sample** consists of students who are present and respond. Plausible
target populations include:

- the responding students;
- everyone attending this lecture;
- all students enrolled in QM Basic;
- all first-year business students in the Netherlands; or
- all young adults in the Netherlands.

::: {.callout-note collapse="true"}
## Check: which of the five candidate populations can our responses describe directly?

Only the first: the responding students. Everyone else --- absent
classmates, other QM Basic students, Dutch first-years, young adults ---
enters the picture only through assumptions, because attendance and
participation were not random. The data describe the respondents; they do
not automatically represent wider groups.
:::

### Sampling variability versus selection bias

Two uncertainties must not be confused:

- **Sampling variability:** even a genuinely random sample can contain an
  unusual number of highly risk-tolerant students by chance. Larger random
  samples usually reduce this chance variation.
- **Selection bias:** a convenience sample can systematically exclude
  certain people. Collecting more responses from the same selected group
  does not necessarily repair that bias.

::: {#exm-b01-bias-contrast}
## Worked example: two scenarios, classified aloud

*Scenario 1.* A researcher draws 40 students **at random** from the full
programme enrolment list. By chance, the sample contains an unusually large
number of students who chose "go all-in."

*Scenario 2.* A survey about founding a business is shared **in the group
chat of the entrepreneurship club**, and 400 members respond.

Scenario 1 is sampling variability: the selection mechanism was random, so
the oddity is chance variation, and a larger random sample would usually
tame it. Scenario 2 is selection bias: club members are systematically more
interested in founding than the average student, and collecting 4,000
responses from the same chat would not repair the distortion --- it would
make a biased estimate more precise.
:::

A probability sample of Dutch students would require a defined population,
a suitable sampling frame, and random selection. Non-response and
measurement error could still remain.

### Meet the Shark Tank data

The first block used data the class created; this block turns to the data
the whole course runs on.

#### Context and unit of observation

*Shark Tank US* is a television programme in which entrepreneurs present a
venture or product to a panel of investors known as sharks. Entrepreneurs
may ask for money in exchange for an ownership share. The investors question
them, negotiate, and may announce a deal during the episode.

The recorded outcome is an **on-air agreement**. It does not establish that
an investment was completed or that the venture later succeeded.

The teaching file contains 1,441 televised pitches from completed Seasons
1--16 and 19 variables. The table below introduces the variables used most
often in this course.

| Variable | Meaning | Type |
|---|---|---|
| `pitch_id` | row identifier | identifier |
| `startup_name` | venture or product name | categorical label |
| `deal_on_show` | 1 for an on-air agreement; 0 otherwise | binary categorical, stored as 0/1 |
| `season` | television season, 1--16 | ordered discrete/time index |
| `episode` | episode number within season | discrete identifier within season |
| `industry` | source industry category | categorical |
| `description_words` | words in the written dataset description | discrete quantitative count |
| `pitcher_gender_group` | `all_women`, `all_men`, `mixed`, or `unknown`, based on the source classification | categorical |
| `women_represented` | 1 for `all_women` or `mixed`; 0 for `all_men`; missing for `unknown` | binary categorical, stored as 0/1 |
| `multiple_entrepreneurs` | source indicator for more than one entrepreneur | binary categorical, with missing values |
| `ask_amount_usd` | amount requested | quantitative, USD |
| `equity_offered_pct` | ownership percentage initially offered | quantitative, percentage points |
| `requested_valuation_usd` | valuation implied by the initial ask | quantitative, USD |
| `deal_amount_usd`, `deal_equity_pct`, `deal_valuation_usd` | recorded deal terms | quantitative; structurally missing without a deal |

Gender describes the people presenting the pitch, not necessarily every
founder or employee. `women_represented` is not the share of women: the
source does not report the presenter counts needed to calculate that share
for mixed teams. Names were not used to infer gender.

Note carefully what `description_words` measures: the length of the
**written description in the source dataset**, not anything about the pitch
itself. It counts words typed by whoever compiled the file, not words spoken
in the tank. The file also does not measure the spoken pitch, completed
investment, or later venture performance.

#### Inspect and validate

This is the live-coding part of the lecture: each output appears on screen,
and you read it **before** the interpretation is given. Run the code
yourself --- everything on this page starts from the visible import below.

```{r}
#| label: b01-inspect
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
dim(sharks)
names(sharks)
str(sharks)
```

There are 1,441 rows and 19 columns. This resolves the prediction from your
prep: **one row represents one recorded televised pitch** from Seasons 1--16
of *Shark Tank US*.

```{r}
#| label: b01-validate
colSums(is.na(sharks))
sort(unique(sharks$deal_on_show))
range(sharks$season)
range(sharks$episode)
range(sharks$description_words)
table(sharks$pitcher_gender_group, useNA = "ifany")
anyDuplicated(sharks$pitch_id)
```

Pitch identifiers are unique, deal status contains only 0 and 1, and seasons
run from 1 to 16. Written descriptions contain from 1 to 13 words. Missing
deal terms are expected for pitches without a deal; seven records have an
unknown source gender classification, and the source does not report
`multiple_entrepreneurs` consistently. Missing does not mean zero. These
checks establish internal plausibility, not that the source captured every
pitch or measured every concept well.

#### Summarise in context

```{r}
#| label: b01-deal-rate
deal_count <- sum(sharks$deal_on_show)
pitch_count <- nrow(sharks)
deal_rate <- mean(sharks$deal_on_show)

c(
  on_air_deals = deal_count,
  recorded_pitches = pitch_count,
  recorded_deal_rate = deal_rate
)
```

Because a 0/1 variable has mean equal to its proportion of 1s,
`mean(deal_on_show)` is 61.2%. A defensible sentence is:

> In this dataset, 882 of 1,441 recorded pitches (61.2%) reached an on-air
> agreement.

This sentence names the dataset and outcome. It does not claim that a
randomly selected entrepreneur has the same chance of completed investment.

### Describing one variable

Before comparing groups or fitting models, learn to describe **one**
variable well. The right tool depends on the variable's type.

#### Frequency tables and bar charts for discrete categories

An **empirical distribution** records how often values occur in a dataset.
For a categorical or discrete variable, a **frequency table** gives counts
and a **relative-frequency table** gives proportions:

```{r}
#| label: b01-season-table
season_counts <- table(sharks$season)
season_counts
round(prop.table(season_counts), 3)
```

```{r}
#| label: fig-b01-season-bar
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Number of recorded pitches per season. Later seasons contain more records, so an overall percentage gives those seasons more weight."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_counts <- as.data.frame(table(sharks$season))
names(season_counts) <- c("season", "pitches")
season_counts$season <- as.integer(as.character(season_counts$season))

ggplot(season_counts, aes(x = season, y = pitches)) +
  geom_col(fill = "#009E73", alpha = 0.85, colour = "white") +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Season", y = "Number of recorded pitches") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

In @fig-b01-season-bar the horizontal axis lists the discrete season
categories; each bar's height on the vertical count scale is the number of
recorded pitches in that season. The later seasons contain more records, so
an overall percentage gives those seasons more weight.

#### Histograms for quantitative variables

For a quantitative variable, a **histogram** groups nearby values into
intervals and counts how many observations fall in each.

```{r}
#| label: fig-b01-word-hist
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Distribution of the number of words in the written dataset description. The dashed vertical line marks the median, 5 words."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
word_median <- median(sharks$description_words)   # 5 words

ggplot(sharks, aes(x = description_words)) +
  geom_histogram(breaks = seq(0.5, 13.5, by = 1),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_vline(xintercept = word_median, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = word_median + 0.4, y = 300,
           label = "median = 5 words", hjust = 0,
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = 1:13) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Words in the written dataset description",
       y = "Number of recorded pitches") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

In @fig-b01-word-hist the horizontal axis measures description length in
words and the vertical axis counts pitches in each one-word interval. The
distribution is mildly right-skewed: most descriptions contain 3--7 words,
with a small number extending farther to the right. The median is 5 words.
Remember what this variable measures: the source text about the pitch, not
the pitch itself.

The histogram the rest of this section works with shows the equity initially
offered:

```{r}
#| label: fig-b01-equity-hist
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Distribution of the equity initially offered across the 1,441 recorded pitches, in 5-point intervals. The dashed vertical line marks the median, 10%."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
equity_median <- median(sharks$equity_offered_pct)   # 10%

ggplot(sharks, aes(x = equity_offered_pct)) +
  geom_histogram(breaks = seq(0, 100, by = 5),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_vline(xintercept = equity_median, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.9) +
  annotate("text", x = equity_median + 2, y = 530,
           label = "median = 10%", hjust = 0,
           colour = "#17324D", size = 3.4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Equity initially offered (%)",
       y = "Number of recorded pitches") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each bar counts pitches within an offered-equity interval. Most initial
offers are below 20%, while a small number are much larger. Changing the
interval width can reveal or conceal patterns, so a histogram is an
analytical choice rather than decoration.

### Reading a histogram by hand

The exam gives you a printed histogram and no computer, so this skill is
practised by hand. Every graph in this course is read with the same
**five-step sequence**:

1. **Unit and variables.** What does one observation represent, and what is
   on each axis, in which units?
2. **Scales.** Read the axis scales, including where they start and how wide
   each interval is.
3. **One element.** Put one bar (or point) into a full sentence.
4. **Pattern.** Describe the overall shape: where values concentrate,
   whether the distribution is symmetric or skewed, whether there are
   outliers.
5. **Limits.** State one thing the graph cannot tell you.

::: {#exm-b01-histogram-read}
## Worked example: the five steps and the median interval

Apply the sequence to @fig-b01-equity-hist.

1. One observation is one recorded pitch; the horizontal axis is equity
   initially offered in percentage points, the vertical axis is the number
   of pitches.
2. The horizontal axis runs from 0 to 100 in intervals of width 5; the
   vertical axis counts pitches.
3. The tallest bar covers 5--10%: 560 pitches offered more than 5% and at
   most 10% equity.
4. The distribution is strongly right-skewed: most offers are at or below
   20%, with a thin tail stretching towards 100%.
5. The histogram cannot say why offers cluster there, and it hides the exact
   values inside each interval.

**Locating the median interval.** With 1,441 ordered observations, the
median is the 721st value. Add bar counts from the left until the running
total reaches 721:

| Interval (%) | Count | Cumulative count |
|---:|---:|---:|
| 0--5   | 293 | 293 |
| 5--10  | 560 | 853 |
| 10--15 | 194 | 1,047 |
| 15--20 | 241 | 1,288 |

After the first bar only 293 observations are covered --- not yet 721. After
the second bar the running total is $293+560=853\ge721$, so the 721st
observation lies in the **5--10% interval**: the median interval. (R
confirms: the exact median is 10%, which this histogram counts in the 5--10
bar because each bar includes its right endpoint.) The same cumulative-count
reading is tested in Mock Exam Question 1.
:::

::: {.callout-caution collapse="true"}
## Try it: locate the upper-quartile interval

In class this is the pair task. The upper quartile is the value with
three-quarters of the observations at or below it.

*Step 1 (given).* Three-quarters of 1,441 observations is about 1,081, so
you are looking for the interval containing the 1,081st ordered value.

*Step 2 (you).* Extend the cumulative counts in the table above until the
running total first reaches 1,081.

*Step 3 (you).* Name the upper-quartile interval.

::: {.callout-note collapse="true"}
## Check your answer

Through 10--15% the cumulative count is $293+560+194=1{,}047$, still below
1,081. Adding the 15--20% bar gives $1{,}047+241=1{,}288\ge1{,}081$, so the
upper quartile lies in the **15--20% interval**. (R confirms:
`quantile(sharks$equity_offered_pct, 0.75)` returns 20, on the upper edge
of that bar.)
:::
:::

### Comparing raw observations and group means

```{r}
#| label: fig-b01-raw-points
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Equity initially offered by season, one point per recorded pitch, jittered horizontally to reduce overlap. Colour and shape mark on-air deal status."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
sharks$deal_status <- factor(sharks$deal_on_show, levels = c(0, 1),
                             labels = c("No on-air deal", "On-air deal"))
set.seed(2026)
sharks$season_jittered <- jitter(sharks$season, amount = 0.16)

ggplot(sharks, aes(x = season_jittered, y = equity_offered_pct,
                   colour = deal_status, shape = deal_status)) +
  geom_point(alpha = 0.45, size = 1.6) +
  scale_colour_manual(values = c("No on-air deal" = "#009E73",
                                 "On-air deal" = "#D55E00")) +
  scale_shape_manual(values = c("No on-air deal" = 16,
                                "On-air deal" = 17)) +
  scale_x_continuous(breaks = c(seq(1, 15, by = 2), 16)) +
  labs(x = "Season", y = "Equity initially offered (%)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Season is a discrete ordered time index, so the points in
@fig-b01-raw-points are moved slightly left or right only to reduce overlap.
The vertical spread within every season warns us that a group mean will not
describe every pitch.

```{r}
#| label: fig-b01-season-means
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Mean equity initially offered per season, separately for pitches with and without an on-air deal."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_means <- aggregate(
  equity_offered_pct ~ season + deal_on_show,
  data = sharks,
  FUN = mean
)
season_means$deal_status <- factor(season_means$deal_on_show,
                                   levels = c(0, 1),
                                   labels = c("No on-air deal",
                                              "On-air deal"))

ggplot(season_means, aes(x = season, y = equity_offered_pct,
                         colour = deal_status, shape = deal_status,
                         linetype = deal_status)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.4) +
  scale_colour_manual(values = c("No on-air deal" = "#009E73",
                                 "On-air deal" = "#D55E00")) +
  scale_shape_manual(values = c("No on-air deal" = 16,
                                "On-air deal" = 17)) +
  scale_linetype_manual(values = c("No on-air deal" = "solid",
                                   "On-air deal" = "dashed")) +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Season", y = "Mean equity initially offered (%)",
       colour = NULL, shape = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Mean initially offered equity in @fig-b01-season-means is lower in later
seasons for both deal groups. The graph describes selected televised
pitches; it does not establish why the offers changed or what would happen
if an entrepreneur changed only the offered equity.

A clear graph matches the variable types, labels its axes and groups, and
makes the claim inspectable. Colour is reinforced here by point shape and
line type, so the groups remain distinguishable without relying on colour
alone.

::: {.caveat-box}
<span class="label">What these records can and cannot say</span>

The Shark Tank file describes **1,441 recorded televised pitches**, nothing
more, for three structural reasons:

- **Selection into television.** Producers choose which pitches are filmed
  and aired. The file is not a random sample of startups, entrepreneurs, or
  even Shark Tank applicants.
- **On-air deal is not investment.** `deal_on_show` records an on-air
  agreement. It does not establish that money changed hands or that the
  venture later succeeded.
- **Measured is not the same as meant.** `description_words` counts words in
  the source's written description, not the spoken pitch;
  `women_represented` is a category, not a share; the file does not measure
  later performance at all.

The same discipline applies to the class dataset: it describes the students
who responded, and a bigger convenience sample would not make it describe
Dutch students in general.
:::

### Answering the driving question

The dataset can tell us, precisely and reproducibly, what happened in these
1,441 recorded pitches: 882 of them (61.2%) reached an on-air agreement,
most initial offers gave up at most 20% equity with a median in the 5--10%
interval, and mean offers drifted downward across seasons. It can never tell
us, by itself, what would happen to a new entrepreneur, whether an on-air
agreement became a completed investment, or why any of these patterns arose
--- those questions need assumptions about selection, measurement, and
causality that the file cannot supply.

**Next:** We can now describe what happened in these 1,441 pitches: counts,
shares, and shapes. But description alone cannot answer the chance questions
an investor actually asks --- how likely is a deal if I pick a pitch at
random, how likely is a late-season pitch, how likely are both at once.
Lecture 2 builds the tool for exactly such questions, a probability model:
select one record at random so that every pitch is equally likely, define
events such as "reached a deal" and "late season", and compute how likely each
is --- with rules that say when chances may simply be added.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 75 minutes.</span>

### Reproduce the workflow (15 minutes)

Save the script. In RStudio, choose **Session > Restart R** to clear objects
from memory. Run the complete script from top to bottom. It should import
the data, perform the checks, calculate the deal rate, and recreate at least
one graph without relying on commands run earlier at the Console.

::: {.callout-note collapse="true"}
## Check the result

A successful clean run reports 1,441 rows and 19 variables, a recorded
on-air deal rate of 882/1,441 (61.2%), the stated ranges, and recreates the
graph without an "object not found" error.
:::

### Retrieval (10 minutes)

Answer without reopening the chapter, then check your answers.

1. What does one row represent, and which population can it describe
   directly?
2. Why is `pitch_id` an identifier rather than a quantitative measurement?
3. What is the difference between sampling variability and selection bias?
4. What exactly does `description_words` measure?
5. How do you locate the median interval of a histogram by hand?

::: {.callout-note collapse="true"}
## Check your answers

1. One row is one recorded televised pitch; the file directly describes
   these 1,441 records from completed Seasons 1--16.
2. `pitch_id` labels a record. Arithmetic differences between its values
   have no substantive meaning.
3. Sampling variability is chance variation across random samples; selection
   bias is systematic distortion from how units enter the sample.
4. The number of words in the written dataset description --- the source
   text about the pitch, not the spoken pitch itself.
5. Find the position of the median (about $n/2$), then add bar counts from
   the left; the first interval whose cumulative count reaches that position
   contains the median.
:::

### Explain one graph (10 minutes)

Write four sentences about either Shark Tank graph
(@fig-b01-raw-points or @fig-b01-season-means):

1. name the variables and units;
2. describe the main visible pattern;
3. state one defensible interpretation; and
4. state one important limitation or next question.

::: {.callout-note collapse="true"}
## Check a model answer

Each point in the raw-data graph represents one recorded pitch, with season
on the horizontal axis and initially offered equity on the vertical axis.
Offers vary within every season and are generally lower in later seasons.
This is evidence of an association between season and the recorded initial
offer. It does not show why offers changed or identify the causal effect of
changing an offer.
:::

### From worked to faded: finish a histogram reading (10 minutes)

Re-read the worked histogram reading (@exm-b01-histogram-read). Now finish
the same reading for the description-words histogram
(@fig-b01-word-hist). The bar counts, from 1 word upward, are:

| Words | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 | 13 |
|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| Count | 4 | 113 | 217 | 285 | 292 | 241 | 146 | 85 | 36 | 9 | 4 | 6 | 3 |

*Step 1 (given).* One observation is one recorded pitch; the axes show
description length in words and the pitch count. With 1,441 observations,
the median is the 721st ordered value.

*Step 2 (you).* Build the cumulative counts and locate the bar containing
the 721st observation.

*Step 3 (you).* Describe the shape and state one limit of the graph.

::: {.callout-note collapse="true"}
## Check your answer

Cumulative counts: 4, 117, 334, 619, 911, ... The running total first
reaches 721 at the 5-word bar ($619+292=911\ge721$), so the median is 5
words --- matching `median(sharks$description_words)`. The distribution is
mildly right-skewed, concentrated at 3--7 words with a thin right tail. The
graph cannot say anything about the pitch itself: it describes the source's
written text.
:::

### Audit an AI answer (10 minutes)

An AI assistant writes:

> "The dataset contains 1,441 successful companies and proves that offering
> less equity makes investors fund startups."

Use **Attempt → Ask → Verify → Explain** to correct it. Your final response
must use the observed count or percentage, name the measured outcome, and
state why the causal claim is unsupported.

::: {.callout-note collapse="true"}
## Check a model correction after writing your own

The file contains 1,441 recorded televised pitches, of which 882 reached an
on-air agreement. It does not measure completed investment or later company
success. `equity_offered_pct` records the initial ownership percentage
offered, not a randomly assigned treatment, and the observational records do
not establish that changing only this offer would cause an investor
decision.
:::

### Worked exam interpretation (5 minutes)

**Question (mirrors Mock Exam Q1).** A startup accelerator records, for each
of 240 applications: `funding_requested_k` (euros, in thousands),
`team_size`, `sector` (four categories), and `accepted` (0/1). Classify each
variable. A histogram of `funding_requested_k` has equal-width intervals of
EUR 20,000 with counts 58, 88, 54, 24, 10, 6. Which interval contains the
median?

::: {.callout-note collapse="true"}
## Check a model answer

`funding_requested_k` is quantitative continuous; `team_size` is
quantitative discrete; `sector` is categorical; `accepted` is categorical
binary, even though it is coded 0/1. With 240 observations, the median lies
between the 120th and 121st ordered values. Cumulative counts: 58 below
EUR 20,000, then $58+88=146$ below EUR 40,000 --- so observations 120 and
121, and hence the median, lie in the **EUR 20,000--40,000 interval**.
:::

### From the reading ladder (10 minutes)

This course trains you to read real research, one bounded piece at a time
--- the [reading ladder](../resources/reading-ladder.qmd) explains the plan
and the six-question protocol. First rung: open
[Ajzen (1991), the Theory of Planned Behavior](https://doi.org/10.1016/0749-5978(91)90020-T)
via the university library and read only the **title and abstract**. List
two theoretical constructs the theory uses (for example attitude toward the
behaviour, subjective norm, perceived behavioural control, intention), and
for each propose one concrete measurable variable you could actually record
from a survey of E&BI students --- then classify each proposed variable as
categorical, discrete quantitative, or continuous quantitative.

::: {.callout-note collapse="true"}
## Check an example answer

The construct *entrepreneurial intention* could be measured as the answer to
"do you plan to start a business within five years?" --- a categorical
(binary) variable; the construct *attitude toward entrepreneurship* could be
measured as agreement with "running my own business appeals to me" on a
1--7 scale --- a discrete quantitative measurement of a construct no single
question fully captures. The gap between the construct and the variable is
exactly the imperfect-measurement point from today's lecture.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Write two
sentences: first name one number you would genuinely want to know about it
(a rate, a count, an amount), then state what data you would need to collect
to measure that number and one way the collection could be biased. There is
no wrong answer; the point is connecting today's tools to something you care
about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "The number I most want is the share of first-order customers who order again within 30 days. I would need order records linked over time, and counting only app users would bias the share upward, because the most committed customers are the ones who install the app."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to poke holes in your
data-collection plan or to suggest a selection bias you have not thought of.
::::

Continue to Lecture 2: Probability foundations and random variables.

Formula sheet: Lecture 1 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 1--5.

:::
