# Lecture 2: Probability foundations and random variables {#b02}

```{r}
#| label: b02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 2, you can</span>

- define a random experiment, outcome, sample space, and event;
- use complements, intersections, unions, and disjointness;
- state the probability rules and compute probabilities from a counts table;
- recognise a partition of the sample space;
- distinguish empirical proportions from model probabilities and from
  forecasts;
- define a random variable informally and state its support.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Investors think in chances: how likely is a deal, how likely is a pitch from the show's later, bigger years, how likely are both at once. Before we can compute any such chance, we need a precise language for "how likely" --- that language is today's lecture.</p>

Pick one recorded pitch at random. What is the chance it reached a deal, came
from a late season, or both --- and what do those numbers mean beyond this
file?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=iQ2edOqEQAs">Sample Space</a>
and <a href="https://www.youtube.com/watch?v=pA83XtLeVig">Probability
Axioms</a>, then watch
<a href="https://www.youtube.com/watch?v=vfqPpai_9jI">Definition of Random
Variables</a>. Be ready to define an outcome, sample space, event, and
random variable. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (5 minutes).** In Lecture 1 you built frequency tables for the
Shark Tank data. Rebuild the season frequency table --- one line of R once the
data are loaded:

```{r}
#| label: b02-prep-table
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
table(sharks$season)
```

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: is $P(\text{deal or late season})$ **larger**
or **smaller** than $P(\text{deal}) + P(\text{late season})$? Commit to an
answer; the lecture resolves it.

If set notation, complements, unions, or intersections are unfamiliar, review
[Mathematics Refresher: sets and events](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. The experiment behind the whole lecture:
**select one of the 1,441 recorded pitches uniformly at random.**

### A probability model begins with possible outcomes

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/sample-space/">01.2 Sample Space</a>.</p>

Probability starts by specifying exactly what may happen. A probability model
needs a process, a list of complete possible results, and events that describe
which results matter for the question.

A **random experiment** is a process whose outcome is not known in advance.
The word *random* does not mean "unstructured." It means the model does not
specify in advance which possible outcome will occur.

- An **outcome**, written $\omega$, is one complete possible result.
- The **sample space**, written $\Omega$, is the set of all possible outcomes.
- An **event** is a specified subset of $\Omega$ to which the model assigns a
  probability. It occurs when the realised outcome belongs to that subset.

The distinction is easiest to read from the whole-to-part structure:

| Object | Question it answers | Shark Tank example |
|---|---|---|
| experiment | What is done? | Select one of the 1,441 records uniformly |
| outcome $\omega$ | Which complete result occurred? | One complete selected pitch record |
| sample space $\Omega$ | Which complete results were possible? | The set of all 1,441 records |
| event $A$ | Which outcomes count as "yes" for this question? | The subset of records with an on-air deal |

For a single roll of a fair six-sided die,

$$\Omega=\{1,2,3,4,5,6\}.$$

The event $E=\{2,4,6\}$ means "an even result." The outcome $\omega=4$
belongs to $E$, so $E$ occurs when 4 is rolled. The number 4 is an outcome;
the singleton set $\{4\}$ is an event.

::: {.callout-note collapse="true"}
## Check: is "the record with pitch_id 7" an outcome or an event?

The record itself is one complete possible result of the selection, so it is
an **outcome**. The description "the selected pitch is the record with
pitch_id 7" picks out the singleton set containing that record, and a set of
outcomes is an **event**. The same distinction as 4 versus $\{4\}$ on the die.
:::

#### An empirical selection experiment

Define the experiment as **select one of the 1,441 recorded pitches uniformly
at random**. The sample space is the set of 1,441 pitch records. Define:

- $D$: the selected pitch reached an on-air deal;
- $L$: the selected pitch appeared in a late season (Seasons 9--16).

Making the selection uniform means every record has probability $1/1441$.
Consequently, event probabilities equal relative frequencies in this fixed
dataset. This does not imply the same probabilities for future entrepreneurs
--- the caveat box below returns to this point.

### Combining events

Events are sets of outcomes. Set notation therefore gives us a precise way to
describe how events are combined.

| Notation | Name | Meaning |
|---|---|---|
| $A^c$ | complement | $A$ does not occur |
| $A\cap B$ | intersection | both $A$ and $B$ occur |
| $A\cup B$ | union | $A$, $B$, or both occur |
| $A\subseteq B$ | subset | every outcome in $A$ is also in $B$ |
| $A\cap B=\varnothing$ | disjoint | $A$ and $B$ cannot occur together |

To make the notation concrete, suppose one number is selected at random from

$$\Omega=\{1,2,\ldots,12\}.$$

Define $A=\{1,2,3,5,6,9\}$ and $B=\{1,2,3,4\}$. Each panel of
@fig-b02-events shows the same sample space. A green circle means that the
outcome belongs to the event named above the panel.

```{r}
#| label: fig-b02-events
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-width: 8
#| fig-height: 3.1
#| fig-cap: "Complement, intersection, and union of two events in a 12-outcome sample space. Green outcomes belong to the event named above each panel."
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3), mar = c(0.6, 0.35, 2.8, 0.35),
    xaxs = "i", yaxs = "i")

omega <- 1:12
A <- c(1, 2, 3, 5, 6, 9)
B <- c(1, 2, 3, 4)

event_sets <- list(
  setdiff(omega, A),
  intersect(A, B),
  union(A, B)
)
panel_titles <- c("Complement", "Intersection", "Union")
panel_meanings <- c("not A", "A and B", "A or B (or both)")

draw_outcome_set <- function(selected, title, meaning) {
  plot(NA, xlim = c(0.45, 4.55), ylim = c(0.35, 4.2), axes = FALSE,
       xlab = "", ylab = "", asp = 1)
  title(main = title, line = 1.2, cex.main = 1.05, col.main = "#17324D")
  mtext(meaning, side = 3, line = 0.1, cex = 0.82, col = "#5B6976")

  rect(0.62, 0.55, 4.38, 3.55, col = "#F4F7F9",
       border = "#17324D", lwd = 1.4)
  text(4.18, 3.76, expression(Omega), cex = 0.85, col = "#17324D")

  x <- rep(1:4, 3)
  y <- rep(3:1, each = 4)
  is_selected <- omega %in% selected
  symbols(x, y, circles = rep(0.25, 12), inches = FALSE,
          bg = ifelse(is_selected, "#009E73", "white"),
          fg = ifelse(is_selected, "#009E73", "#9AA7B2"),
          lwd = 1.4, add = TRUE)
  text(x, y, labels = omega,
       col = ifelse(is_selected, "white", "#17324D"),
       font = 2, cex = 0.82)
}

for (i in seq_along(event_sets)) {
  draw_outcome_set(event_sets[[i]], panel_titles[i], panel_meanings[i])
}
par(old_par)
```

Thus $A^c=\{4,7,8,10,11,12\}$, $A\cap B=\{1,2,3\}$, and
$A\cup B=\{1,2,3,4,5,6,9\}$. Notice that **or** includes outcomes that are
in both events: $1,2,3$ belong to the union as well as the intersection.

The same grammar works for business events. In the selection experiment, let
$F$ be "the selected pitch is from the Food and Beverage industry" and $T$
"from the Technology industry." Each record lists exactly one industry, so
$F$ and $T$ are disjoint: $F\cap T=\varnothing$. The union $F\cup T$ is "a
food or technology pitch," and $F^c$ collects every pitch from any other
industry. Nothing new is needed --- only the sets have changed.

Disjoint does **not** mean independent. If two disjoint events both have
positive probability, observing one makes the other impossible. Independence,
introduced in Lecture 4, means that observing one event does not change the
probability of the other.

#### The 2×2 counts table for deals and seasons

Everything this lecture computes comes from one table. The code constructs
$D$ and $L$ exactly as defined above --- run it yourself:

```{r}
#| label: b02-counts-table
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
table(deal = D, late = L)
```

Written with row and column totals:

|                    | Early ($L^c$, S1--8) | Late ($L$, S9--16) | Total |
|--------------------|---------------------:|-------------------:|------:|
| **No deal** ($D^c$) | 330 | 229 | **559** |
| **Deal** ($D$)      | 377 | 505 | **882** |
| **Total**           | **707** | **734** | **1441** |

::: {#exm-b02-events}
## Worked example: three events in words and symbols

From the table, describe each event in words and give its count.

- $D^c$, "no on-air deal": the top row, $330 + 229 = 559$ records.
- $D\cap L$, "deal **and** late season": the single bottom-right cell,
  505 records.
- $D\cup L$, "deal, late season, **or both**": every record except the
  top-left cell, $1441 - 330 = 1111$ records --- the deal row (882) plus the
  late column (734) minus the 505 records counted in both.
:::

### Probability rules

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-axioms/">01.4 Probability Axioms</a>.</p>

A **probability law** assigns a number $P(A)$ to each event $A$. It must
satisfy three rules:

1. **Non-negativity:** $P(A)\ge 0$ for every event $A$.
2. **Normalisation:** $P(\Omega)=1$.
3. **Additivity:** if $A$ and $B$ are disjoint, then
   $$P(A\cup B)=P(A)+P(B),$$
   and the same holds for any finite collection of pairwise disjoint events.

The first two rules set the probability scale: probability cannot be negative,
and something in the sample space must occur. The third says that mutually
exclusive routes can be added because no outcome is counted twice.

::: {.optional-box}
<span class="label">Optional: countable additivity</span><br>
In the formal theory the third rule is stated for infinite sequences of
pairwise disjoint events, $P\!\left(\bigcup_{i=1}^{\infty} A_i\right)
=\sum_{i=1}^{\infty} P(A_i)$, which matters for sample spaces with infinitely
many outcomes. Every sample space in this course is finite, so the finite
version above is all you need.
:::

For a finite sample space with equally likely outcomes,

$$P(A)=\frac{|A|}{|\Omega|}.$$

That ratio is a special case, not a universal definition. It applies because
equal likelihood has been built into the model --- here, because the selection
was made uniform.

#### Rules derived from the three above

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/simple-properties-of-probabilities/">01.5 Simple Properties of Probabilities</a>.</p>

Because $A$ and $A^c$ are disjoint and together form $\Omega$,

$$P(A^c)=1-P(A).$$

When $A$ and $B$ may overlap,

$$P(A\cup B)=P(A)+P(B)-P(A\cap B).$$

The intersection is subtracted once because it was counted in both marginal
probabilities.

::: {#exm-b02-union}
## Worked example: P(deal or late season), by hand

Every number comes from the counts table above --- point to each one as you
read.

$$
P(D\cup L)
=\frac{882}{1441}+\frac{734}{1441}-\frac{505}{1441}
=\frac{882+734-505}{1441}
=\frac{1111}{1441}\approx 0.771.
$$

The 882 is the deal row total, the 734 is the late column total, and the 505
is the deal-and-late cell that both totals contain --- subtracted once so no
record is counted twice.

This resolves the prediction from your prep: $P(D\cup L)$ is **smaller** than
$P(D)+P(L)$ whenever the two events overlap, because the overlap
$P(D\cap L) > 0$ is subtracted.
:::

Verification in R, with $D$ and $L$ built exactly as before:

```{r}
#| label: b02-probability-rules
c(
  P_D            = mean(D),
  P_L            = mean(L),
  P_D_and_L      = mean(D & L),
  P_not_D        = mean(!D),
  P_D_or_L_direct = mean(D | L),
  P_D_or_L_rule   = mean(D) + mean(L) - mean(D & L)
)
```

::: {.callout-caution collapse="true"}
## Try it: P(no deal and early season)

In class this is the pair task set just before the break. Complete the two
missing steps before opening the answer.

*Step 1 (given).* "No deal **and** early season" is the event $D^c\cap L^c$
--- the top-left cell of the counts table.

*Step 2 (you).* Read off the count for that cell.

*Step 3 (you).* Divide by the number of records and interpret the result in
one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(D^c\cap L^c) = 330/1441 \approx 0.229$: about 23% of the recorded pitches
are early-season pitches that did not reach an on-air deal.

Verification in R (the construction of `D` and `L` is repeated so the block
runs on its own):

```{r}
#| label: b02-faded-check
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
mean(!D & !L)
```
:::
:::

### Partitions

A **partition** divides $\Omega$ into events that do not overlap and together
include every possible outcome. Each outcome belongs to exactly one part. For
example, Seasons 1--8 and Seasons 9--16 partition the recorded pitches into an
early-period event $L^c$ and a late-period event $L$ (@fig-b02-partition).

```{r}
#| label: fig-b02-partition
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-width: 6.4
#| fig-height: 2.4
#| fig-cap: "The early-period event and the late-period event partition the sample space: no overlap, nothing left out."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.3, 0.3, 0.3, 0.3))
plot(NA, xlim = c(0, 10), ylim = c(0, 5), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")
rect(0.4, 0.4, 9.6, 4.6, border = "#17324D", lwd = 2)
rect(0.4, 0.4, 4.9, 4.6, col = "#DCEAF3", border = NA)
rect(4.9, 0.4, 9.6, 4.6, col = "#FDE5C1", border = NA)
segments(4.9, 0.4, 4.9, 4.6, col = "#17324D", lwd = 2)
rect(0.4, 0.4, 9.6, 4.6, border = "#17324D", lwd = 2)
text(2.65, 2.5, expression(L^c), col = "#0072B2", cex = 1.25)
text(7.25, 2.5, "L", col = "#D55E00", cex = 1.25)
text(9.15, 4.15, expression(Omega), col = "#17324D")
par(old_par)
```

The two coloured regions do not overlap, and together they fill the whole
rectangle. Lecture 4 uses this idea to combine conditional probabilities from
separate groups.

::: {.callout-note collapse="true"}
## Check: do the sixteen events "the pitch is from Season s" form a partition?

Yes. Every record has exactly one season, so the sixteen single-season events
are pairwise disjoint and together cover all 1,441 records. $D$ and $L$, by
contrast, do **not** form a partition: they overlap (505 records) and leave
records out (the 330 early no-deal records).
:::

### From complete outcomes to numerical measurements

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/definition-of-random-variables/">05.2 Definition of Random Variables</a>.</p>

An outcome is a complete result, but an analysis usually focuses on one
numerical feature of that result. A **random variable** is a numerical
measurement whose value is determined by chance --- through a fixed rule that
turns the realised outcome into a number.

Three random variables on our selection experiment:

- the **deal indicator** $X$: equal to 1 when the selected pitch reached an
  on-air deal and 0 otherwise;
- the **season** $S$: the season number of the selected pitch;
- the **word count** $W$: the number of words in its written description.

The rule is fixed before an outcome is selected; the value is unknown until
then. That is the only sense in which the variable is "random." After
selection, the number produced --- written in lower case, $x$, $s$, or $w$
--- is its **realised value**.

A random variable is **discrete** when its possible values can be listed. Its
**support** is the set of values it can take with positive probability. The
support of the deal indicator is $\{0,1\}$; the support of season is
$\{1,2,\ldots,16\}$. Lecture 3 attaches a probability to every value in the
support.

::: {.optional-box}
<span class="label">Optional: the formal definition</span><br>
Formally, a random variable is a function from the sample space to the real
numbers, $X:\Omega\rightarrow\mathbb{R}$. Read left to right: $X$ takes an
outcome $\omega$ from $\Omega$ and returns the number $X(\omega)$. For the
deal indicator, $X(\omega)=1$ when $\omega\in D$ and $X(\omega)=0$ otherwise,
so the event $D$ and the variable $X$ are linked by
$D=\{\omega : X(\omega)=1\}$. The bookkeeping table:

| Symbol | Meaning |
|---|---|
| $\omega$ | one complete outcome, such as one selected pitch record |
| $X$ | the fixed measurement rule |
| $X(\omega)$ | the number assigned by $X$ to outcome $\omega$ |
| $x$ | a possible or realised numerical value of $X$ |
:::

::: {.caveat-box}
<span class="label">What these numbers can and cannot say</span>

$P(D)=882/1441\approx 0.612$ is exact for one question only: the chance of a
deal when one record is selected uniformly from this fixed file. It is an
**empirical proportion**. Calling it the probability that a *future*
televised pitch reaches an on-air agreement adds four assumptions:

- the future case belongs to a defined target population;
- selection into the show is comparable;
- the process is stable over time;
- "deal" retains the same definition.

Calling it the probability of a completed *investment* would additionally
change the outcome being predicted: completed investment is not recorded in
these data. An on-air agreement is not money in the bank.
:::

### Answering the driving question

For one uniformly selected record: the chance it reached a deal is
$P(D)\approx 0.612$, the chance it came from a late season is
$P(L)\approx 0.509$, the chance of both is $P(D\cap L)\approx 0.350$, and the
chance of at least one is $P(D\cup L)\approx 0.771$. Beyond this file, those
numbers describe future pitches only under the four assumptions in the caveat
box.

**Next:** We can now attach a probability to a single event --- deal or no
deal, late or early. But most business quantities have many possible values:
which season, how many words, how many units sold. Lecture 3 upgrades from
single events to the whole *distribution* of a random variable: a probability
for every possible value (the PMF), running totals (the CDF), and the two
numbers that summarise it all --- expected value and variance --- so that "a
typical pitch" finally gets a precise meaning.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 55 minutes.</span>

### Retrieval (10 minutes)

1. What is the difference between an outcome and an event?
2. What is the difference between a random variable and a realised value?
3. State the three probability rules.
4. Why is the equally-likely ratio a special case?
5. What extra assumptions turn an empirical proportion into a forecast?

::: {.callout-note collapse="true"}
## Check your answers

1. An outcome is one possible complete result; an event is a specified set of
   outcomes.
2. A random variable is a fixed measurement rule that turns outcomes into
   numbers; a realised value is the number produced for the observed outcome.
3. Probabilities are non-negative, $P(\Omega)=1$, and probabilities add
   across disjoint events.
4. The ratio $|A|/|\Omega|$ requires a finite sample space whose outcomes are
   equally likely.
5. A forecast requires a defined target population and assumptions about
   selection, process stability, and a stable outcome definition.
:::

### From worked to faded (10 minutes)

Re-read the worked union calculation (@exm-b02-union). Then complete the same
calculation with new numbers: let $E=L^c$ be the early-period event
(Seasons 1--8).

*Step 1 (given).* From the counts table: $|D| = 882$, $|E| = 707$,
$|D\cap E| = 377$.

*Step 2 (you).* Apply $P(D\cup E)=P(D)+P(E)-P(D\cap E)$.

*Step 3 (you).* Interpret the result in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(D\cup E)=(882+707-377)/1441=1212/1441\approx 0.841$: about 84% of the
recorded pitches reached an on-air deal, came from Seasons 1--8, or both.
:::

### Practice (15 minutes)

In the uniform selection experiment, use
$P(D)=882/1441$, $P(L)=734/1441$, and
$P(D\cap L)=505/1441$.

1. Calculate and interpret $P(D^c)$.
2. Calculate and interpret $P(D\cup L)$.
3. Are $D$ and $D^c$ disjoint? Explain.
4. Define a random variable $S$ that records the season of the selected
   pitch. State its support.
5. Define a deal indicator $X$. Explain how the event $D$ and the random
   variable $X$ are related but not the same object.

::: {.callout-note collapse="true"}
## Check your answers

1. $P(D^c)=1-882/1441=559/1441\approx0.388$: 38.8% of the records did not
   reach an on-air agreement.
2. $P(D\cup L)=(882+734-505)/1441=1111/1441\approx0.771$: 77.1% of the
   records are from Seasons 9--16, reached an on-air deal, or satisfy both
   conditions.
3. They are disjoint because a record cannot simultaneously have and not have
   an on-air deal.
4. $S$ records the season number of the selected pitch record. Its support is
   $\{1,2,\ldots,16\}$.
5. $X$ equals 1 when the selected pitch is in $D$ and 0 otherwise. $D$ is a
   set of outcomes; $X$ is a measurement rule that assigns a number to every
   outcome. They are related by $D=\{X=1\}$.
:::

### Spaced review: one item from Lecture 1 (5 minutes)

A student concludes: "Our class survey had 180 responses, so its average risk
score describes Dutch first-year students well --- with that many responses,
chance variation is small." Which problem does the size of the sample *not*
fix?

::: {.callout-note collapse="true"}
## Check your answer

Selection bias. A larger sample shrinks sampling variability, but the class
is a convenience sample of one programme, not a random sample of Dutch
first-year students. More responses from the same selected group do not make
it more representative.
:::

### Worked exam interpretation (10 minutes)

**Question.** R reports `mean(sharks$deal_on_show) = 0.612`. Is this an
empirical proportion or a universal probability?

::: {.callout-note collapse="true"}
## Check a model answer

It is first an empirical proportion: 882 of the 1,441 recorded
pitches reached an on-air agreement. It is also the event probability in a
model that selects one of those 1,441 records uniformly. Applying it to a
future entrepreneur requires additional population, selection, stability, and
outcome assumptions.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Name one
uncertain event that matters for it, and write two sentences: first define
the event precisely (which outcomes count as "yes"?), then say what knowing
its probability would change about a real decision. There is no wrong
answer; the point is connecting today's tools to something you care about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Let $R$ be the event that a first-time customer places a second order within 30 days --- each order record either shows a repeat order in that window (yes) or does not. If $P(R)$ were known to be above 0.4, I would spend my small budget on loyalty rewards rather than new-customer ads."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to challenge your event
definition or to suggest what else you would need to know.
::::

Continue to Lecture 3: Discrete distributions, expectation, and variance.

Formula sheet: Lecture 2 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 6--8 for probability models and random variables.

:::
