# Lecture 2: Probability foundations and random variables {#b02}

```{r}
#| label: b02-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 2, you can</span>

- define a random experiment, outcome, sample space, and event
- use complements, intersections, unions, and disjointness
- state the probability rules and compute probabilities from a counts table
- recognise a partition of the sample space
- distinguish empirical proportions from model probabilities and from
  forecasts
- say what a random variable is and state its support.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- read a probability off a counts table
- work out "not", "and", and "or" probabilities, with the addition rule for "or"
- say whether two events are disjoint
- say what a random variable is and list its possible values.

**Enough to recognise:**

- the symbols $\Omega$, $\omega$, $\varnothing$, $A^c$, $A\cap B$, $A\cup B$, $A\subseteq B$
- the three probability rules by name.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Investors think in chances: how likely is a deal, how likely is a pitch from the show's later years, how likely are both at once. Before we can compute any such chance, we need a precise language for "how likely". This lecture introduces that language.</p>

Pick one recorded pitch at random. What is the chance that it reached a deal,
came from a late season, or both, and what do those numbers mean beyond this
file?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=iQ2edOqEQAs">Sample Space</a>
and <a href="https://www.youtube.com/watch?v=pA83XtLeVig">Probability
Axioms</a>, then watch
<a href="https://www.youtube.com/watch?v=vfqPpai_9jI">Definition of Random
Variables</a>. Be ready to define an outcome, sample space, event, and
random variable. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

[Shark Tank data]{.shark-tag} **Warm up (5 minutes).** In Lecture 1 you built frequency tables for the
Shark Tank data. Build the season frequency table again. Once the data are
loaded, you need one line of R:

```{r}
#| label: b02-prep-table
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
table(sharks$season)
```

[Shark Tank data]{.shark-tag} **Predict (5 minutes).** Answer this question before you come to class. Pick one recorded pitch at random.
Is the probability that the pitch reached a deal or comes from a late season **larger** or
**smaller** than the probability of a deal plus the probability of a late
season? Late means Seasons 9 to 16. Choose one answer. You find the answer in
the lecture material.

If set notation, complements, unions, or intersections are unfamiliar, review
[Mathematics Refresher: sets and event notation](../resources/math-refresher.html#sets-and-event-notation).

:::

::: {.course-phase .phase-in-class}

## Lecture material

### Start with a founder's question

[Shark Tank data]{.shark-tag} You are preparing a pitch. Before you walk in, you would like answers to three
questions. How likely is a deal at all? Was a deal easier to get in the later
seasons, when the show was better known? How many pitches had both a late
season and a deal? All three answers are in one small table. The code builds it from the
Shark Tank file. "Late" means Seasons 9 to 16.

The code first stores two yes-or-no answers for every pitch. `D` is `TRUE`
when the pitch reached an on-air deal (`deal_on_show` equals 1). `L` is `TRUE`
when the pitch comes from a late season (`season` is 9 or higher). Later in
this lecture the same two letters name the two events.

```{r}
#| label: b02-counts-table
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # TRUE if the pitch reached an on-air deal
L <- sharks$season >= 9         # TRUE if the pitch is from a late season (9-16)
table(deal = D, late = L)
```

The same counts with row and column totals:

|  | Early seasons (1--8) | Late seasons (9--16) | Total |
|---|---:|---:|---:|
| **No deal** | 330 | 229 | **559** |
| **Deal** | 377 | 505 | **882** |
| **Total** | **707** | **734** | **1441** |

#### Answer by counting

- **A deal at all.** 882 of the 1,441 pitches, 61.2%.
- **A late season.** 734 of the 1,441 pitches, 50.9%.
- **Both.** The bottom-right cell: 505 of 1,441, 35.0%.
- **A deal, or a late season, or both.** Do not add 882 and 734. That would
  count the 505 pitches in the bottom-right cell twice. Count them once:
  882 + 734 - 505 = 1,111 of 1,441, 77.1%. This answers the prediction from
  your preparation: the share is smaller than the sum of the two shares.
- **No deal.** 1,441 - 882 = 559 pitches, 38.8%.

One more comparison that a founder would want to see: among the late-season pitches, 505
of 734 got a deal (68.8%). Among the early ones, 377 of 707 did (53.3%).
Lecture 4 is about comparisons of this kind.

You have just used every idea of this lecture. The rest of the lecture gives
these ideas names and symbols: outcomes and events, "and", "or", and "not",
and the rules that say when you may add shares. With these names and symbols,
you can use the ideas on any table and recognise them in any textbook.

### Why data analysis needs probability

The shares above describe one fixed file. Most business questions are about
a larger group, a population, but we see only a sample of it. A sample is one
draw from the population, and another draw would give other numbers.
Probability is the language for this uncertainty.

[Lecture 7](b07-samples-estimators.qmd) shows this with the equity initially
offered (`equity_offered_pct`) in the Shark Tank file. There the 1,441
records play the role of the population. One random sample of 100 records
gives a mean that differs from the mean of all 1,441 records. Lectures 2 to 6 build the language of
probability. Lectures 7 to 10 use it to go from a sample to a population.

### A probability model begins with possible outcomes

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/sample-space/">01.2 Sample Space</a>.</p>

[Shark Tank data]{.shark-tag} Probability starts by describing exactly what may happen. A probability model
needs a process, a list of complete possible results, and events that describe
which results matter for the question.

A **random experiment** is a process whose outcome is not known in advance.
The word *random* does not mean "without any order or rule." It means that
the model does not say in advance which possible outcome will occur.

::: {.definition-box}
<span class="label">Definition</span>

**Outcome, sample space, and event.** An **outcome**, written $\omega$, is
one complete possible result of a random experiment. The **sample space**,
written $\Omega$, is the set of all possible outcomes. An **event** is a set
of outcomes, so it is a part of the sample space. An event occurs when the
outcome that happens belongs to it.
:::

The table shows how these terms fit together:

| Object | Question it answers | Shark Tank example |
|---|---|---|
| experiment | What is done? | Select one of the 1,441 records at random, each with the same chance |
| outcome $\omega$ | Which complete result occurred? | One complete selected pitch record |
| sample space $\Omega$ | Which complete results were possible? | The set of all 1,441 records |
| event $A$ | Which outcomes count as "yes" for this question? | The subset of records with an on-air deal |

For a single roll of a fair six-sided die,

$$\Omega=\{1,2,3,4,5,6\}.$$

The event $E=\{2,4,6\}$ means "an even result." The outcome $\omega=4$
belongs to $E$, so $E$ occurs when you roll a 4. The number 4 is an outcome.
The set $\{4\}$, which contains only this one outcome, is an event.

::: {.callout-note collapse="true"}
## Check: is "the record with pitch_id 7" an outcome or an event?

The record itself is one complete possible result of the selection, so it is
an **outcome**. The description "the selected pitch is the record with
pitch_id 7" describes the set that contains only that record, and a set of
outcomes is an **event**. This is the same difference as the one between 4
and $\{4\}$ for the die.
:::

#### An empirical selection experiment

Define the experiment as **select one of the 1,441 recorded pitches uniformly
at random**. The sample space is the set of 1,441 pitch records. Define two
events:

- $D$: the selected pitch reached an on-air deal
- $L$: the selected pitch appeared in a late season (Seasons 9--16).

These are the events behind the R objects `D` and `L` from the start of this
lecture. Making the selection uniform means that every record has the same
probability, $1/1441$. The probability of an event then equals its share of
the records in this fixed dataset, so $D$ has probability $882/1441$. This
does not mean that future entrepreneurs have the same probabilities. The box "What
these numbers can and cannot say" below returns to this point.

### Combining events

[Shark Tank data]{.shark-tag} Events are sets of outcomes. So set notation gives us a precise way to
describe how we combine events.

| Notation | Name | Meaning |
|---|---|---|
| $A^c$ | complement | $A$ does not occur |
| $A\cap B$ | intersection | both $A$ and $B$ occur |
| $A\cup B$ | union | $A$, $B$, or both occur |
| $A\subseteq B$ | subset | every outcome in $A$ is also in $B$ |
| $\varnothing$ | empty event | the event that contains no outcome, so it never occurs |
| $A\cap B=\varnothing$ | disjoint | $A$ and $B$ cannot occur together |

As an example, suppose that we select one number at random from

$$\Omega=\{1,2,\ldots,12\}.$$

Define $A=\{1,2,3,5,6,9\}$ and $B=\{1,2,3,4\}$. Each panel of
@fig-b02-events shows the same sample space. A green circle means that the
outcome belongs to the event named above the panel.

```{r}
#| label: fig-b02-events
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-width: 8
#| fig-height: 3.1
#| fig-cap: "Complement, intersection, and union of two events in a 12-outcome sample space. Green outcomes belong to the event named above each panel."
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 3), mar = c(0.6, 0.35, 2.8, 0.35),
    xaxs = "i", yaxs = "i")

omega <- 1:12
A <- c(1, 2, 3, 5, 6, 9)
B <- c(1, 2, 3, 4)

event_sets <- list(
  setdiff(omega, A),
  intersect(A, B),
  union(A, B)
)
panel_titles <- c("Complement", "Intersection", "Union")
panel_meanings <- c("not A", "A and B", "A or B (or both)")

draw_outcome_set <- function(selected, title, meaning) {
  plot(NA, xlim = c(0.45, 4.55), ylim = c(0.35, 4.2), axes = FALSE,
       xlab = "", ylab = "", asp = 1)
  title(main = title, line = 1.2, cex.main = 1.05, col.main = "#17324D")
  mtext(meaning, side = 3, line = 0.1, cex = 0.82, col = "#5B6976")

  rect(0.62, 0.55, 4.38, 3.55, col = "#F4F7F9",
       border = "#17324D", lwd = 1.4)
  text(4.18, 3.76, expression(Omega), cex = 0.85, col = "#17324D")

  x <- rep(1:4, 3)
  y <- rep(3:1, each = 4)
  is_selected <- omega %in% selected
  symbols(x, y, circles = rep(0.25, 12), inches = FALSE,
          bg = ifelse(is_selected, "#009E73", "white"),
          fg = ifelse(is_selected, "#009E73", "#9AA7B2"),
          lwd = 1.4, add = TRUE)
  text(x, y, labels = omega,
       col = ifelse(is_selected, "white", "#17324D"),
       font = 2, cex = 0.82)
}

for (i in seq_along(event_sets)) {
  draw_outcome_set(event_sets[[i]], panel_titles[i], panel_meanings[i])
}
par(old_par)
```

So $A^c=\{4,7,8,10,11,12\}$, $A\cap B=\{1,2,3\}$, and
$A\cup B=\{1,2,3,4,5,6,9\}$. Notice that "or" includes outcomes that are
in both events: $1,2,3$ belong to the union as well as the intersection.

The same notation works for business events. In the selection experiment, let
$F$ be "the selected pitch is from the Food and Beverage industry" and $T$
"from the Technology/Software industry." Each record lists exactly one industry, so
$F$ and $T$ are disjoint: $F\cap T=\varnothing$. The union $F\cup T$ is "a
food or technology pitch," and $F^c$ contains every pitch from any other
industry. Nothing new is needed. Only the sets have changed.

Disjoint does **not** mean independent. If two disjoint events both have
positive probability, observing one makes the other impossible. Independence,
introduced in Lecture 4, means that observing one event does not change the
probability of the other.

#### The counts table again, with event names

This is the table from the start of the lecture. The row and column labels now
show the event names $D$ and $L$:

|                    | Early ($L^c$, S1--8) | Late ($L$, S9--16) | Total |
|--------------------|---------------------:|-------------------:|------:|
| **No deal** ($D^c$) | 330 | 229 | **559** |
| **Deal** ($D$)      | 377 | 505 | **882** |
| **Total**           | **707** | **734** | **1441** |

::: {#exm-b02-events}
## Worked example: three events in words and symbols

[Shark Tank data]{.shark-tag} From the table, describe each event in words and give its count.

- $D^c$, "no on-air deal": the top row, $330 + 229 = 559$ records.
- $D\cap L$, "deal **and** late season": the single bottom-right cell,
  505 records.
- $D\cup L$, "deal, late season, **or both**": every record except the
  top-left cell, $1441 - 330 = 1111$ records. This is the deal row (882) plus
  the late column (734) minus the 505 records counted in both.
:::

::: {.callout-tip collapse="true"}
## Optional: more rules for combining events

Four rules help to rewrite a combination of events. $A$, $B$, and $C$ are any
three events.

- First distributive law: $A\cap(B\cup C)=(A\cap B)\cup(A\cap C)$.
- Second distributive law: $A\cup(B\cap C)=(A\cup B)\cap(A\cup C)$.
- First De Morgan rule: $(A\cup B)^c=A^c\cap B^c$. "Not ($A$ or $B$)" is the
  same as "not $A$ and not $B$".
- Second De Morgan rule: $(A\cap B)^c=A^c\cup B^c$. "Not ($A$ and $B$)" is
  the same as "not $A$ or not $B$".

In the counts table, $(D\cup L)^c=D^c\cap L^c$ is the top-left cell: the 330
early-season pitches without a deal.
:::

### Probability rules

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-axioms/">01.4 Probability Axioms</a>.</p>

::: {.definition-box}
<span class="label">Definition</span>

**Probability.** The probability of an event $A$, written $P(A)$, is a number
between 0 and 1 that says how likely the event is. The closer the number is
to 1, the more likely the event. An impossible event has probability 0, and a
certain event has probability 1.
:::

[Shark Tank data]{.shark-tag} A model gives a probability $P(A)$ to each event $A$. These numbers must
follow three rules:

1. **Non-negativity:** $P(A)\ge 0$ for every event $A$.
2. **Normalisation:** $P(\Omega)=1$.
3. **Additivity:** if $A$ and $B$ are disjoint, then
   $$P(A\cup B)=P(A)+P(B),$$
   The same holds for any finite number of events, as long as no two of them
   can occur together.

The first two rules set the probability scale: probability cannot be negative,
and something in the sample space must occur. The third says that you
can add the probabilities of disjoint events, because no outcome is counted
twice.

::: {.callout-tip collapse="true"}
## Optional: countable additivity

In the formal theory the third rule also holds for infinite sequences of
events, as long as no two of them can occur together, $P\!\left(\bigcup_{i=1}^{\infty} A_i\right)
=\sum_{i=1}^{\infty} P(A_i)$. This matters for sample spaces with infinitely
many outcomes. Every sample space in Lectures 2 to 5 is finite, so the finite
version above is all you need there.
:::

Write $|A|$ for the number of outcomes in the event $A$. For a finite sample
space with equally likely outcomes,

$$P(A)=\frac{|A|}{|\Omega|}.$$

For example, $|D|=882$ and $|\Omega|=1441$, so $P(D)=882/1441\approx 0.612$.
This ratio works only when all outcomes are equally likely. Here they are,
because every record has the same chance of being drawn.

#### Rules derived from the three above

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/simple-properties-of-probabilities/">01.5 Simple Properties of Probabilities</a>.</p>

Because $A$ and $A^c$ are disjoint and together form $\Omega$,

$$P(A^c)=1-P(A).$$

When $A$ and $B$ may overlap,

$$P(A\cup B)=P(A)+P(B)-P(A\cap B).$$

This is the addition rule. We subtract the intersection once, because it is
counted in both $P(A)$ and $P(B)$.

::: {#exm-b02-union}
## Worked example: P(deal or late season), by hand

[Shark Tank data]{.shark-tag} Every number comes from the counts table above. Find each one in the table as
you read.

$$
P(D\cup L)
=\frac{882}{1441}+\frac{734}{1441}-\frac{505}{1441}
=\frac{882+734-505}{1441}
=\frac{1111}{1441}\approx 0.771.
$$

The 882 is the deal row total, the 734 is the late column total, and the 505
is the deal-and-late cell that both totals contain. We subtract it once, so
that no record counts twice.

This is the count from the start of the lecture, written as a rule:
$P(D\cup L)$ is smaller than
$P(D)+P(L)$ whenever the two events overlap, because the rule subtracts the
overlap $P(D\cap L) > 0$.
:::

Verification in R, with $D$ and $L$ built exactly as before:

```{r}
#| label: b02-probability-rules
c(
  P_D            = mean(D),
  P_L            = mean(L),
  P_D_and_L      = mean(D & L),
  P_not_D        = mean(!D),
  P_D_or_L_direct = mean(D | L),
  P_D_or_L_rule   = mean(D) + mean(L) - mean(D & L)
)
```

Three more properties follow from the three rules. You can see each one in
the counts table.

- The empty event has probability zero: $P(\varnothing)=0$. No record is
  both a deal and not a deal, so $D\cap D^c=\varnothing$ contains 0 of the
  1,441 records.
- A subset is never more likely: if $A\subseteq B$, then $P(A)\le P(B)$.
  The 505 late-season deals $D\cap L$ are part of the 882 deals $D$, and
  $505/1441\approx 0.350$ is below $882/1441\approx 0.612$.
- An event can be split into two parts: $P(A)=P(A\cap B)+P(A\cap B^c)$. The
  882 deals are the 505 late-season deals plus the 377 early-season deals, so
  $P(D)=P(D\cap L)+P(D\cap L^c)=505/1441+377/1441=882/1441$. Lecture 4 uses
  this property again.

::: {.callout-caution collapse="true"}
## Try it: P(no deal and early season)

[Shark Tank data]{.shark-tag} Try this first with the student next to you. Complete the two missing steps
before you open the answer.

*Step 1 (given).* "No deal **and** early season" is the event $D^c\cap L^c$:
the top-left cell of the counts table.

*Step 2 (you).* Read the count for that cell from the table.

*Step 3 (you).* Divide by the number of records and interpret the result in
one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(D^c\cap L^c) = 330/1441 \approx 0.229$: about 23% of the recorded pitches
are early-season pitches that did not reach an on-air deal.

Verification in R. The first two lines build `D` and `L` again:

```{r}
#| label: b02-faded-check
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
mean(!D & !L)
```
:::
:::

### Partitions

[Shark Tank data]{.shark-tag} A **partition** divides $\Omega$ into events that do not overlap and together
include every possible outcome. Each outcome belongs to exactly one part. For
example, Seasons 1--8 and Seasons 9--16 partition the recorded pitches into an
early-period event $L^c$ and a late-period event $L$ (@fig-b02-partition).

```{r}
#| label: fig-b02-partition
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-width: 6.4
#| fig-height: 2.4
#| fig-cap: "The early-period event and the late-period event partition the sample space: no overlap, nothing left out."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.3, 0.3, 0.3, 0.3))
plot(NA, xlim = c(0, 10), ylim = c(0, 5), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")
rect(0.4, 0.4, 9.6, 4.6, border = "#17324D", lwd = 2)
rect(0.4, 0.4, 4.9, 4.6, col = "#DCEAF3", border = NA)
rect(4.9, 0.4, 9.6, 4.6, col = "#FDE5C1", border = NA)
segments(4.9, 0.4, 4.9, 4.6, col = "#17324D", lwd = 2)
rect(0.4, 0.4, 9.6, 4.6, border = "#17324D", lwd = 2)
text(2.65, 2.5, expression(L^c), col = "#0072B2", cex = 1.25)
text(7.25, 2.5, "L", col = "#D55E00", cex = 1.25)
text(9.15, 4.15, expression(Omega), col = "#17324D")
par(old_par)
```

The two coloured regions do not overlap, and together they fill the whole
rectangle. Lecture 4 uses this idea to combine probabilities from separate
groups.

::: {.callout-note collapse="true"}
## Check: do the sixteen events "the pitch is from Season s" form a partition?

Yes. Every record has exactly one season, so no two of the sixteen
single-season events overlap, and together they cover all 1,441 records. $D$ and $L$, by
contrast, do **not** form a partition: they overlap (505 records) and leave
records out (the 330 early no-deal records).
:::

### From complete outcomes to numerical measurements

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/definition-of-random-variables/">05.2 Definition of Random Variables</a>.</p>

[Shark Tank data]{.shark-tag} An outcome is a complete result, but an analysis usually focuses on one
number that describes that result.

::: {.definition-box}
<span class="label">Definition</span>

**Random variable.** A number whose value depends on the outcome of a random
experiment. A fixed rule turns each outcome into a number.
:::

Random variables are written with capital letters, such as $X$. Their values
are written with lower-case letters, such as $x$.

Here are three random variables for our selection experiment:

- the **deal indicator** $X$: equal to 1 when the selected pitch reached an
  on-air deal and 0 otherwise
- the **season** $S$: the season number of the selected pitch
- the **word count** $W$: the number of words in its written description.

We fix the rule before we select an outcome. The value is unknown until
then. That is the only sense in which the variable is "random." After the
selection, the rule produces a number. This number, written in lower case as
$x$, $s$, or $w$, is the **realised value** of the variable.

A random variable is **discrete** when its possible values can be listed. A
random variable is **continuous** when it can take every value in an
interval, for example a waiting time. Lecture 6 treats continuous
random variables.

The **support** of a discrete random variable is the set of values it can
take with positive probability. The support of the deal indicator is
$\{0,1\}$. The support of season is $\{1,2,\ldots,16\}$. Lecture 3 gives a
probability to every value in the support.

::: {.callout-tip collapse="true"}
## Optional: the formal definition of a random variable

Formally, a random variable is a function from the sample space to the real
numbers, $X:\Omega\rightarrow\mathbb{R}$. Read left to right: $X$ takes an
outcome $\omega$ from $\Omega$ and returns the number $X(\omega)$. For the
deal indicator, $X(\omega)=1$ when $\omega\in D$ and $X(\omega)=0$ otherwise,
so the event $D$ and the variable $X$ are linked by
$D=\{\omega : X(\omega)=1\}$. The table lists the symbols:

| Symbol | Meaning |
|---|---|
| $\omega$ | one complete outcome, such as one selected pitch record |
| $X$ | the fixed measurement rule |
| $X(\omega)$ | the number assigned by $X$ to outcome $\omega$ |
| $x$ | a possible or realised numerical value of $X$ |
:::

::: {.caveat-box}
<span class="label">What these numbers can and cannot say</span>

- $P(D)=882/1441\approx 0.612$ is exact for one question only: the chance of
  a deal when one record is drawn at random from this file. It is an
  **empirical proportion**.
- It becomes a forecast for a future pitch only under four assumptions. The
  future pitch is like these pitches, producers select in the same way, the
  process does not change over time, and "deal" keeps its meaning.
- It says nothing about completed investments, which the file does not
  record.

**What you can do with it:** use 61% as a first guess for a pitch on the
show, and adjust it when you know more about the pitch. Lecture 4 shows how.
:::

### Answering the driving question

Select one record uniformly at random. The chance that it reached a deal is
$P(D)\approx 0.612$. The chance that it came from a late season is
$P(L)\approx 0.509$. The chance of both is $P(D\cap L)\approx 0.350$, and the
chance of at least one is $P(D\cup L)\approx 0.771$. Beyond this file, those
numbers describe future pitches only under the four assumptions in the box
"What these numbers can and cannot say".

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your order records show three shares among your customers: 12% had a late delivery, 8% gave a low rating, and 3% had both. You want to send a personal message to every customer with at least one of these problems. The addition rule gives the share: 12% + 8% - 3% = 17%. With 400 customers that is 68 messages, not the 80 that you would plan for if you simply added the two shares. The decision: 68 messages fit into one evening, so you send them this week.
:::

**Next:** We can now give a probability to a single event, such as deal or
no deal, late or early. Most business quantities have many possible values:
which season, how many words, how many units sold. Lecture 3 moves from
single events to the whole distribution of a random variable, which is a
probability for every possible value. It also introduces two numbers that
summarise a distribution: the expected value and the variance.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 75 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the most important parts. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. What is the difference between an outcome and an event?
2. What is the difference between a random variable and a realised value?
3. State the three probability rules.
4. Why is the ratio $|A|/|\Omega|$ for equally likely outcomes a special case?
5. What extra assumptions turn an empirical proportion into a forecast?

::: {.callout-note collapse="true"}
## Check your answers

1. An outcome is one possible complete result. An event is a set of outcomes
   that we define.
2. A random variable is a fixed measurement rule that turns outcomes into
   numbers. A realised value is the number produced for the observed outcome.
3. Probabilities are non-negative, $P(\Omega)=1$, and probabilities add
   across disjoint events.
4. The ratio $|A|/|\Omega|$ requires a finite sample space whose outcomes are
   equally likely.
5. A forecast needs a clearly defined target population. It also needs
   assumptions: the selection works in the same way, the process stays
   stable, and the definition of the outcome does not change.
:::

### Finish a half-worked example (10 minutes)

[Shark Tank data]{.shark-tag} Re-read the worked union calculation (@exm-b02-union). Then complete the same
calculation with new numbers: let $E=L^c$ be the early-period event
(Seasons 1--8).

*Step 1 (given).* From the counts table: $|D| = 882$, $|E| = 707$,
$|D\cap E| = 377$.

*Step 2 (you).* Apply $P(D\cup E)=P(D)+P(E)-P(D\cap E)$.

*Step 3 (you).* Interpret the result in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(D\cup E)=(882+707-377)/1441=1212/1441\approx 0.841$: about 84% of the
recorded pitches reached an on-air deal, came from Seasons 1--8, or both.
:::

### Practice (15 minutes)

[Shark Tank data]{.shark-tag} In the uniform selection experiment, use
$P(D)=882/1441$, $P(L)=734/1441$, and
$P(D\cap L)=505/1441$.

1. Calculate and interpret $P(D^c)$.
2. Calculate and interpret $P(D^c\cup L^c)$. Hint: which records does this
   event leave out?
3. Are $D$ and $D^c$ disjoint? Explain.
4. Define a random variable $S$ that records the season of the selected
   pitch. State its support.
5. Define a deal indicator $X$. Explain how the event $D$ and the random
   variable $X$ are related but not the same object.

::: {.callout-note collapse="true"}
## Check your answers

1. $P(D^c)=1-882/1441=559/1441\approx0.388$: 38.8% of the records did not
   reach an on-air deal.
2. The event leaves out only the records that are both deals and from a late season,
   so $P(D^c\cup L^c)=1-P(D\cap L)=1-505/1441=936/1441\approx0.650$: 65.0% of
   the records did not reach an on-air deal, came from Seasons 1--8, or both.
   The addition rule gives the same result:
   $(559+707-330)/1441=936/1441$.
3. They are disjoint because a record cannot have and not have an on-air deal
   at the same time.
4. $S$ records the season number of the selected pitch record. Its support is
   $\{1,2,\ldots,16\}$.
5. $X$ equals 1 when the selected pitch is in $D$ and 0 otherwise. $D$ is a
   set of outcomes. $X$ is a measurement rule that assigns a number to every
   outcome. $D$ is the set of outcomes for which $X$ equals 1.
:::

### Review: one question from Lecture 1 (5 minutes)

A student concludes: "Our class survey had 180 responses, so its average risk
score describes Dutch first-year students well. With that many responses,
chance variation is small." Which problem does the size of the sample *not*
solve?

::: {.callout-note collapse="true"}
## Check your answer

Selection bias. A larger sample reduces sampling variability. But the class
is a group from one programme that was easy to reach, not a random sample of
Dutch first-year students. More responses from the same selected group do
not make it more representative.
:::

### Worked exam interpretation (10 minutes)

[Shark Tank data]{.shark-tag} **Question.** R reports `mean(sharks$deal_on_show) = 0.612`. Is this an
empirical proportion or a universal probability?

::: {.callout-note collapse="true"}
## Check a model answer

It is first an empirical proportion: 882 of the 1,441 recorded
pitches reached an on-air deal. It is also the probability of the event in a
model that selects one of those 1,441 records uniformly. To use it for a
future founder, you must assume that the future pitch is like these pitches,
that it is selected in the same way, and that a deal means the same thing.
:::

### Check the AI (5 minutes)

[Shark Tank data]{.shark-tag} You asked an AI assistant: "How likely is it that a pitch got a deal or came from a late season?" It answered:

> "P(deal or late) = 882/1441 + 734/1441 = 1616/1441 = 1.12. So the chance is about 112%."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

A probability can never be above 1, so the answer must be wrong. The AI added the two shares and counted the 505 late-season deals twice. Subtract them once: (882 + 734 - 505)/1441 = 1111/1441, about 77%. A quick check for any probability: it must lie between 0 and 1.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Name one
uncertain event that matters for it. Then write two sentences. First, define
the event precisely (which outcomes count as "yes"?). Second, say how knowing
its probability would change a real decision. There is no wrong
answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells boxes of prepared meals on campus. "Let $R$ be the event that a first-time customer places a second order within 30 days: each order record either shows a repeat order in that window (yes) or does not. If I knew that $P(R)$ is above 0.4, I would spend my small budget on loyalty rewards rather than new-customer ads."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to challenge your event
definition or to suggest what else you would need to know.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 2. Show the solution only after I send my answer."

[Continue to Lecture 3: Discrete distributions, expectation, and variance](b03-discrete-distributions.qmd).

[Formula sheet: Lecture 2 formulas](formula-sheet.qmd#formula-l2).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 6--8 for probability models and random variables.

:::
