# Lecture 3: Discrete distributions, expectation, and variance {#b03}

```{r}
#| label: b03-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 3, you can</span>

- describe a discrete variable with its probability and cumulative distributions
- calculate and interpret the expected value of a discrete variable
- measure spread with variance and standard deviation, also after rescaling
- keep model expectations and variances apart from sample means and variances
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Anyone preparing a pitch wants to know what to expect: the typical outcome, and how far real outcomes usually differ from it. Every summary of business risk starts from two numbers: the centre and the spread.</p>

If your pitch ends in a deal, how many sharks will you work with, and how
much does that number differ from deal to deal?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-mass-functions/">Probability Mass Functions</a> and
<a href="https://www.youtube.com/watch?v=_yJsO5955ZE">Expectation</a>.
Be ready to explain what $p_X(x)$ measures and why an expected
value does not have to be a possible realised value. All links are also listed in
the <a href="../resources/video-guide.html">External Video Guide</a>.
The videos explain key ideas in another way and can use a different notation. The lecture material and its notation always come first.
</div>

[Shark Tank data]{.shark-tag} **Warm up.** Lecture 2 ended with random variables for the
experiment that selects one record uniformly at random. Define $X=1$ if the
selected pitch reached an on-air deal and $X=0$ otherwise. Define $S$ as the
season number of the selected pitch.

1. What are the possible values of $X$ and of $S$?
2. Do these possible values tell us how likely each value is?

::: {.callout-note collapse="true"}
## Check your preparation

1. $X$ can take the values 0 and 1. $S$ can take the values
   $1,2,\ldots,16$.
2. No. The support lists the possible values. A probability distribution
   must also give a probability to each value. This lecture adds that
   step.
:::

[Shark Tank data]{.shark-tag} **Predict.** Answer this question before you come to class: can the **expected season** be a
number that is not a whole season, yes or no? Choose one answer. You find the
answer in the lecture material.

If functions, support, or summation notation are unfamiliar, review
[Mathematics Refresher: summation notation](../resources/math-refresher.html#summation-notation).

:::

::: {.course-phase .phase-in-class}

## Lecture material

### Start with a founder's question

[Shark Tank data]{.shark-tag} Suppose your pitch ends in a deal. Will you work with one investor who gives
you full attention, or with a group of sharks who share the deal? The Shark
Tank file records the number of sharks in each deal in the variable
`number_sharks_in_deal`. The code below counts the deals by this number:

```{r}
#| label: b03-sharks-table
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
deals <- subset(sharks, deal_on_show == 1)
shark_counts <- table(deals$number_sharks_in_deal)
shark_counts
round(prop.table(shark_counts), 3)
```

Of the 882 deals, 648 have one shark, 203 have two, 22 have three, 3 have
four, and 6 have five. As shares: 0.735, 0.230, 0.025, 0.003, and 0.007.

#### Four questions, answered from the shares

1. **How likely is a deal with exactly one shark?** Read it from the shares: 73.5%.
2. **How likely is a deal with at most two sharks?** Add the first two
   shares: 73.5% + 23.0% = 96.5%.
3. **How many sharks should you expect?** Multiply each number by its share
   and add the results: $1\times0.735+2\times0.230+3\times0.025+4\times0.003+5\times0.007
   \approx1.32$. No deal has 1.32 sharks. The section on expectation
   explains how to read such a number.
4. **How much do deals differ from that number?** Not much: most deals have
   one shark and nearly all others have two. The section on variance gives
   one number for this spread.

Each of the four answers uses a tool that has a name. The list of shares is a probability mass
function. The running total in question 2 is a cumulative distribution
function. The weighted average in question 3 is an expected value. The spread
in question 4 is measured by the variance and the standard deviation.

### A PMF assigns probability to each possible value

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-mass-functions/">05.3 Probability Mass Functions</a>.</p>

[Shark Tank data]{.shark-tag} A random variable tells us which numerical values we may observe. Its
distribution adds the missing information: how the probability is divided
between those values.

For a discrete random variable $X$, the **probability mass function (PMF)**
is

$$p_X(x)=P(X=x).$$

Read this notation in parts:

| Symbol | Meaning |
|---|---|
| $X$ | the random variable |
| $x$ | one possible numerical value |
| $p_X$ | the PMF belonging to $X$ |
| $p_X(x)$ | the probability that $X$ equals exactly $x$ |

The probability of one value is also called its probability mass, or mass
for short. A valid PMF must meet two rules:

$$
p_X(x)\ge 0\quad\text{for every }x,
\qquad
\sum_{x\in\mathcal X}p_X(x)=1,
$$

where $\mathcal X$ is the support of $X$. The first rule prevents negative
probability. The second says that one of the possible values must occur.

::: {#exm-b03-deal-pmf}
## Worked example: the deal-indicator PMF, checked

In the uniform selection experiment from Lecture 2, the deal indicator
$X$ has support $\{0,1\}$. The Shark Tank file has 882 deals among 1,441
records, so 559 records have no deal. The PMF is

$$
p_X(0)=\frac{559}{1441}\approx 0.388,
\qquad
p_X(1)=\frac{882}{1441}\approx 0.612.
$$

Check both rules: each mass is non-negative, and

$$
\frac{559}{1441}+\frac{882}{1441}=\frac{1441}{1441}=1.
$$

This is a valid PMF.
:::

The code below builds the same objects in R: the data, the deal proportion
`p_hat`, and the PMF.

```{r}
#| label: b03-deal-pmf
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
p_hat <- mean(sharks$deal_on_show)   # 882/1441

deal_pmf <- prop.table(
  table(factor(sharks$deal_on_show, levels = c(0, 1)))
)
names(deal_pmf) <- c("0: no on-air deal", "1: on-air deal")
deal_pmf
sum(deal_pmf)
```

For the season $S$ (`season` in the Shark Tank file), the PMF in the
uniform selection experiment gives each season its share of the 1,441
records:

```{r}
#| label: b03-season-pmf
season_pmf <- prop.table(table(sharks$season))
round(season_pmf, 4)
sum(season_pmf)
```

```{r}
#| label: fig-b03-season-pmf
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The PMF of season under uniform record selection: sixteen separated probability masses, one per possible value."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_pmf <- prop.table(table(sharks$season))
pmf_df <- data.frame(
  season = as.numeric(names(season_pmf)),
  probability = as.numeric(season_pmf)
)

ggplot(pmf_df, aes(x = season, y = probability)) +
  geom_segment(aes(xend = season, yend = 0),
               linewidth = 1.1, colour = "#009E73") +
  geom_point(size = 2.6, colour = "#009E73") +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Season s", y = expression(p[S](s))) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The probability mass above season 5 in @fig-b03-season-pmf has height
$p_S(5)=P(S=5)=116/1441\approx0.080$. The graph shows the PMF as separate
probability masses, because $S$ can take only the listed discrete values.
The term probability density function (PDF) is used only for continuous
random variables. Lecture 6 introduces it.

### A CDF accumulates probability up to a threshold

[Shark Tank data]{.shark-tag} The PMF asks about one exact value. The **cumulative distribution function
(CDF)** asks how much probability lies at or below a chosen threshold (a
cut-off value). Imagine moving from left to right across the PMF and keeping a
running total.

For any random variable $X$,

$$F_X(a)=P(X\le a).$$

Here $a$ is a threshold, not necessarily a value in the support. For a
discrete random variable,

$$F_X(a)=\sum_{x\le a}p_X(x).$$

::: {.definition-box}
<span class="label">Definition</span>

**Probability mass function (PMF)** and **cumulative distribution function
(CDF).** The PMF, $p_X(x)=P(X=x)$, gives the probability that $X$ equals
exactly the value $x$. The CDF, $F_X(a)=P(X\le a)$, gives the probability
that $X$ is at or below the threshold $a$.
:::

::: {#exm-b03-cdf-hand}
## Worked example: F(3) for season, by hand

"$S\le 3$" contains the first three seasons, so add the first three
probability masses of the PMF. The season counts are 64, 36, and 60:

$$
F_S(3)=p_S(1)+p_S(2)+p_S(3)
=\frac{64}{1441}+\frac{36}{1441}+\frac{60}{1441}
=\frac{160}{1441}\approx0.111.
$$

About 11% of the probability lies at or below season 3.
:::

The code below builds the full running total in R. Each CDF value is the
current PMF value plus all earlier PMF values:

```{r}
#| label: b03-season-distribution
s_values <- as.numeric(names(season_pmf))
p_values <- as.numeric(season_pmf)
season_distribution <- data.frame(
  season = s_values,
  pmf = round(p_values, 4),
  cdf = round(cumsum(p_values), 4)
)
season_distribution
```

```{r}
#| label: fig-b03-season-cdf
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The CDF of season: a step graph that never falls. It is 0 to the left of Season 1, jumps at each possible value by that value's probability mass, and stays at 1 from Season 16 on."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_pmf <- prop.table(table(sharks$season))
cdf_df <- data.frame(
  threshold = as.numeric(names(season_pmf)),
  cdf = cumsum(as.numeric(season_pmf))
)
# For the line only: a flat part at 0 before Season 1 and at 1 after Season 16
step_df <- rbind(
  data.frame(threshold = 0, cdf = 0),
  cdf_df,
  data.frame(threshold = 17, cdf = 1)
)

ggplot(cdf_df, aes(x = threshold, y = cdf)) +
  geom_step(data = step_df, linewidth = 0.9, colour = "#009E73") +
  geom_point(size = 1.9, colour = "#17324D") +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Threshold a", y = expression(F[S](a))) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

A CDF adds up probability as the threshold moves from left to right. It
can stay flat or rise, but it cannot fall. It equals 0 below the smallest
possible value and reaches 1 once it includes all possible values.
In @fig-b03-season-cdf the line is flat at 0 to the left of Season 1 and
flat at 1 to the right of Season 16.

For a discrete variable, the CDF is a step graph. It stays flat between
possible values and jumps at each value that has positive probability. A
larger jump means that this value has more probability.

::: {.callout-note collapse="true"}
## Check: read F(8) from the step graph

$F_S(8)=P(S\le 8)$ is the height of the step graph at threshold 8: the
first eight seasons contain $64+36+60+103+116+116+116+96=707$ records, so
$F_S(8)=707/1441\approx0.491$. Just under half of the probability lies in
Seasons 1--8. This agrees with Lecture 2, where the early-period event
$L^c$ had probability $707/1441$.
:::

The probability of an interval is a difference of two CDF values:

$$P(a<X\le b)=F_X(b)-F_X(a).$$

For example, $P(3<S\le 8)=F_S(8)-F_S(3)=(707-160)/1441=547/1441\approx0.380$:
38.0% of the probability lies in Seasons 4 to 8.

For the deal indicator:

$$
F_X(a)=
\begin{cases}
0, & a<0,\\
1-p, & 0\le a<1,\\
1, & a\ge1,
\end{cases}
$$

where $p=P(X=1)$. So $p_X(1)=P(X=1)=p$, while $F_X(1)=P(X\le1)=1$. The
PMF measures the mass at exactly 1. The CDF includes all mass up to and
including 1.

### Expectation is the distribution's weighted centre

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/expectation/">05.8 Expectation</a>.</p>

[Shark Tank data]{.shark-tag} A distribution can contain many possible values and probabilities. The
**expected value**, or expectation, summarises its centre in one number.
Values with larger probabilities receive more weight:

$$E[X]=\sum_{x\in\mathcal X}x\,p_X(x).$$

The notation $E[X]$ is read "the expected value of $X$". The expected value
is also called the mean of $X$ and written $\mu$ (the Greek letter mu) or
$\mu_X$. The calculation multiplies each possible value by its probability
and adds the products. You can also see the result as the balance point of
the probability distribution: if the masses were weights on a ruler, the
ruler would balance at this point.

::: {.definition-box}
<span class="label">Definition</span>

**Expected value.** The probability-weighted average of the possible values
of $X$, written $E[X]$, $\mu$, or $\mu_X$. If you repeat the experiment
many times, the average of the observed values comes close to $E[X]$.
:::

This long-run meaning does not make the expected value a guarantee for a
single observation. The weighted average in question 3 at the start of the
lecture was this calculation: the expected number of sharks in a deal is about 1.32. No
single deal has 1.32 sharks, but over many deals the average number of
sharks per deal comes close to 1.32.

Before we use the Shark Tank data again, try the calculation once on a very
small invented distribution.

::: {#exm-b03-tiny-expectation}
## Worked example: expectation of a tiny PMF

Let $Y$ have support $\{0,1,2\}$ with masses

$$p_Y(0)=0.2,\qquad p_Y(1)=0.5,\qquad p_Y(2)=0.3.$$

This is a valid PMF: all masses are non-negative and $0.2+0.5+0.3=1$. Write
every product:

$$
E[Y]
=0\times0.2 \;+\; 1\times0.5 \;+\; 2\times0.3
=0 + 0.5 + 0.6
=1.1.
$$

Each possible value contributes its value times its probability. The sum,
1.1, is the balance point. Note that 1.1 is not in the support: $Y$
never equals 1.1.
:::

Now apply the same steps to the Shark Tank distributions. For the deal
indicator,

$$E[X]=0\times(1-p)+1\times p=p.$$

This result explains why the mean of a 0/1 variable equals its proportion
of 1s. Here $E[X]=882/1441\approx0.612$.

For the season, the sum has sixteen products instead of two:

```{r}
#| label: b03-expectation
s <- as.numeric(names(season_pmf))
p_s <- as.numeric(season_pmf)

c(
  E_X = 0 * (1 - p_hat) + 1 * p_hat,
  E_S_from_pmf = sum(s * p_s),
  observed_mean_season = mean(sharks$season)
)
```

The expected season is about 8.8, and there is no Season 8.8. This
answers the prediction question from before class: an expectation is a
probability-weighted average, so it need not be a possible value of the
random variable.

### Expectation after transforming a variable

[Shark Tank data]{.shark-tag} Many quantities come from a random variable through a fixed rule. Examples
for the variable $Y$ of the worked example are the profit $10Y-3$ (10 euros
per unit of $Y$ minus a fixed cost of 3 euros) and the square $Y^2$. Such a
function $g(X)$ of a random variable $X$ is itself a random variable, with
its own possible values and its own distribution.

If $g$ changes each possible value of $X$, the expected value of the
transformed variable is

$$E[g(X)]=\sum_{x\in\mathcal X}g(x)\,p_X(x).$$

The probabilities do not change. Apply $g$ to each possible value, weight
the transformed values by the original PMF, and add.

::: {.callout-caution collapse="true"}
## Try it: E[Y²] for the tiny PMF

Try this first with the student next to you. Use the same distribution as
in the worked example: $Y\in\{0,1,2\}$ with masses $0.2,0.5,0.3$, and $g(y)=y^2$.

*Step 1 (given).* Square each possible value: $0^2=0$, $1^2=1$, $2^2=4$.

*Step 2 (you).* Multiply each squared value by its original probability
and write the three products.

*Step 3 (you).* Add them to get $E[Y^2]$. Is it equal to
$\left(E[Y]\right)^2=1.1^2$?

::: {.callout-note collapse="true"}
## Check your answer

$$
E[Y^2]=0\times0.2+1\times0.5+4\times0.3=0+0.5+1.2=1.7.
$$

It is not $\left(E[Y]\right)^2=1.21$. Squaring and then averaging
differs from averaging and then squaring. The gap between the two is
$1.7-1.21=0.49$. The section on variance gives this gap a name.
:::
:::

The same rule for season in R:

```{r}
#| label: b03-transformed-expectation
c(
  E_S = sum(s * p_s),
  E_S_squared = sum(s^2 * p_s)
)
```

### Variance and standard deviation

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/variance/">06.2 Variance</a>.</p>

[Shark Tank data]{.shark-tag} Expectation gives the centre of a distribution, but it does not describe
how close the values lie to that centre. Two distributions can have
the same expectation and very different spreads. Both distributions in
@fig-b03-same-mean-spread have expectation 1. Which one is riskier, and
why?

```{r}
#| label: fig-b03-same-mean-spread
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.15
#| fig-cap: "Two distributions with the same expectation and very different spreads. The dashed line marks the common mean, 1."
library(ggplot2)
spread_df <- data.frame(
  x = rep(0:2, times = 2),
  probability = c(0.10, 0.80, 0.10,    # Distribution A
                  0.45, 0.10, 0.45),   # Distribution B
  distribution = rep(c("Distribution A (variance = 0.20)",
                       "Distribution B (variance = 0.90)"), each = 3)
)

ggplot(spread_df, aes(x = x, y = probability)) +
  geom_vline(xintercept = 1, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_segment(aes(xend = x, yend = 0),
               linewidth = 1.1, colour = "#009E73") +
  geom_point(size = 2.6, colour = "#009E73") +
  annotate("text", x = 1.1, y = 0.87, label = "mean = 1",
           hjust = 0, colour = "#17324D", size = 3.4) +
  facet_wrap(~ distribution) +
  scale_x_continuous(breaks = 0:2) +
  scale_y_continuous(limits = c(0, 0.9),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Possible value of X", y = "Probability") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

The horizontal axis lists the possible values and the vertical axis gives
their probabilities. The centre is the same, but Distribution B has greater
spread because values 0 and 2 receive much more probability. If these were
payoffs, B would be the riskier choice. Variance is the number that measures
this difference.

To measure spread, start with the deviation $X-\mu$, where $\mu=E[X]$.
If we averaged the deviations directly, positive and negative deviations
would cancel each other. So the **variance** squares each deviation before it
takes the probability-weighted average:

$$
\operatorname{Var}(X)
=E[(X-\mu)^2]
=\sum_x (x-\mu)^2\,p_X(x).
$$

The variance is also written $\sigma^2$, read "sigma squared" ($\sigma$ is
the Greek letter sigma). Read the calculation as four steps:

1. Find the centre, $\mu=E[X]$.
2. Measure each possible deviation, $x-\mu$.
3. Square the deviation, $(x-\mu)^2$.
4. Weight it by $p_X(x)$ and add across the support.

So large deviations contribute more than small deviations. Variance is zero
only when all probability lies at one single value.

The equivalent **computational form** is

$$\operatorname{Var}(X)=E[X^2]-\left(E[X]\right)^2.$$

It gives the same result, but the definition $E[(X-\mu)^2]$ is usually
easier to interpret.

The **standard deviation** is the square root of the variance. It is also
written $\sigma$:

$$\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.$$

Variance has squared units. Standard deviation returns to the units of $X$.
If $X$ is measured in seasons, $\operatorname{Var}(X)$ is in seasons
squared, a unit that nobody can imagine, while $\operatorname{SD}(X)$ is
back in seasons.

::: {.definition-box}
<span class="label">Definition</span>

**Variance and standard deviation.** The variance, $\operatorname{Var}(X)$
or $\sigma^2$, is the probability-weighted average of the squared deviations
of $X$ from its mean $\mu$. The standard deviation, $\operatorname{SD}(X)$ or
$\sigma$, is the square root of the variance and has the same units as $X$.
:::

::: {#exm-b03-tiny-variance}
## Worked example: variance of the tiny PMF, both ways

Same $Y$ as before: values $0,1,2$ with masses $0.2,0.5,0.3$, and
$\mu=E[Y]=1.1$.

**By the definition**, in four steps:

$$
\operatorname{Var}(Y)
=(0-1.1)^2(0.2)+(1-1.1)^2(0.5)+(2-1.1)^2(0.3)
$$
$$
=1.21\times0.2+0.01\times0.5+0.81\times0.3
=0.242+0.005+0.243
=0.49.
$$

**By the computational form**, using $E[Y^2]=1.7$ from the Try it task:

$$
\operatorname{Var}(Y)=E[Y^2]-\left(E[Y]\right)^2=1.7-1.21=0.49.
$$

Both ways give 0.49, so $\operatorname{SD}(Y)=\sqrt{0.49}=0.7$, in the
same units as $Y$.
:::

A variable with only the values 0 and 1, such as the deal indicator, is so
common that its distribution has a name.

::: {.definition-box}
<span class="label">Definition</span>

**Bernoulli distribution.** A random variable $X$ that takes only the values
0 and 1, with $P(X=1)=p$ and $P(X=0)=1-p$, has a Bernoulli distribution,
written $X\sim\text{Bernoulli}(p)$. Its expected value is $p$ and its
variance is $p(1-p)$.
:::

The symbol $\sim$ is read "is distributed as". We found the expected value $p$
in the section on expectation. The variance follows from the
computational form. Since $X^2=X$ when $X$ is 0 or 1, $E[X^2]=E[X]=p$, so

$$
\operatorname{Var}(X)=p-p^2=p(1-p).
$$

The deal indicator is a Bernoulli variable with $p=882/1441\approx0.612$.
Its variance is $0.612\times0.388\approx0.237$ and its standard deviation is
about 0.49.

```{r}
#| label: b03-variance
mu_s <- sum(s * p_s)
var_s <- sum((s - mu_s)^2 * p_s)

c(
  Var_X = p_hat * (1 - p_hat),
  SD_X = sqrt(p_hat * (1 - p_hat)),
  Var_S = var_s,
  SD_S = sqrt(var_s),
  average_distance_S = sum(abs(s - mu_s) * p_s)
)
```

The season has a variance of about 18.1 seasons squared and a standard
deviation of about 4.3 seasons. The season of a randomly selected pitch
typically lies about four seasons away from the expected season 8.8. Four
seasons are a quarter of the 16 seasons. The standard deviation is somewhat
larger than the plain average distance (about 3.7 seasons here), because
squaring gives large deviations more weight.

The opening question asked how much deals differ in their number of sharks.
The same two steps answer it:

```{r}
#| label: b03-sharks-spread
k <- as.numeric(names(shark_counts))
p_k <- as.numeric(prop.table(shark_counts))
mu_k <- sum(k * p_k)
c(expected_sharks = mu_k,
  variance = sum((k - mu_k)^2 * p_k),
  sd = sqrt(sum((k - mu_k)^2 * p_k)))
```

The expected number of sharks is 1.32 and the standard deviation is about
0.61 sharks, which is less than one shark. So the number of sharks differs
little from deal to deal.

### Shifting and rescaling a variable {#b03-shift-rescale}

Business quantities are often built from other quantities: revenue is price
times units, profit is revenue minus a fixed cost. For constants $a$ and $b$,

$$
E[aX+b]=aE[X]+b,
\qquad
\operatorname{Var}(aX+b)=a^2\operatorname{Var}(X),
$$
$$
\operatorname{SD}(aX+b)=|a|\operatorname{SD}(X).
$$

Adding a constant $b$ moves the centre and leaves the spread unchanged.
Multiplying by $a$ multiplies the centre by $a$ and the standard deviation
by $|a|$, the size of $a$ without its sign. So the variance is multiplied
by $a^2$.

An example with the tiny PMF: $Y$ has $E[Y]=1.1$,
$\operatorname{Var}(Y)=0.49$, and $\operatorname{SD}(Y)=0.7$. If each unit of
$Y$ earns 10 euros and there is a fixed cost of 3 euros, the profit is
$10Y-3$. Its expectation is $10\times1.1-3=8$ euros. Its variance is
$10^2\times0.49=49$ euros squared, and its standard deviation is
$10\times0.7=7$ euros.

Standardising in [Lecture 6](b06-continuous-normal.qmd) uses this rule. Subtract the mean $\mu$ from $X$ and divide by the standard deviation
$\sigma$. The result $(X-\mu)/\sigma$ then has mean 0 and standard
deviation 1. Lecture 8 uses the same rules to find the spread of a sample
mean.

### Model expectation and the sample mean

[Shark Tank data]{.shark-tag} For observed values $x_1,\ldots,x_n$, the sample mean is

$$\bar{x}=\frac{1}{n}\sum_{i=1}^n x_i.$$

Now give every observed row the same probability, $1/n$. This distribution
is called the **empirical distribution** of the data. Its expectation is

$$\sum_{i=1}^n x_i\,\frac{1}{n}=\bar{x}.$$

The two numbers are equal, but the two terms have different meanings. Expectation
describes a probability distribution. The sample mean summarises observed
data.

The same difference exists for spread. The variance of the empirical
distribution weights each of the $n$ observations by $1/n$, so it divides
by $n$. The usual **sample variance** divides by $n-1$:

$$s^2=\frac{1}{n-1}\sum_{i=1}^n\left(x_i-\bar{x}\right)^2.$$

Its square root $s$ is the sample standard deviation. R's `var()` and `sd()`
divide by $n-1$. [Lecture 7](b07-samples-estimators.qmd) explains why we divide
by $n-1$ when we estimate a population variance.

An invented example shows the calculation by hand. A founder counts the
orders on five days: 3, 5, 6, 8, and 8. The mean is $\bar x=30/5=6$ orders.
The deviations from the mean are $-3$, $-1$, $0$, $2$, and $2$. Their
squares add up to $9+1+0+4+4=18$. Divide by $n-1=4$:
$s^2=18/4=4.5$ orders squared, so $s=\sqrt{4.5}\approx2.12$ orders. A
typical day lies about two orders, or about a third of the mean, away from
the mean. Dividing by $n=5$ would give $3.6$ instead of $4.5$. R gives the
same numbers:

```{r}
#| label: b03-small-sd
orders <- c(3, 5, 6, 8, 8)
c(mean = mean(orders), var = var(orders), sd = sd(orders),
  var_divide_n = mean((orders - mean(orders))^2))
```

The code below computes both versions for the number of words in the written description (`description_words`) in the
Shark Tank file:

```{r}
#| label: b03-word-mean-variance
w <- sharks$description_words
c(
  mean_words = mean(w),
  empirical_expectation = sum(w * rep(1 / length(w), length(w))),
  empirical_variance_divide_n = mean((w - mean(w))^2),
  sample_variance_divide_n_minus_1 = var(w),
  sample_sd = sd(w)
)
```

Both centre calculations give about 5.0 words for these 1,441 records. The
two variances, 3.647 and 3.649, differ only in the third decimal because $n$
is large. All these numbers summarise the same observations. To use one of
them as an estimate for a larger population, you also need to know how the
sample was drawn. Lecture 7 covers this. There, the model quantities $\mu$
and $\sigma^2$ of this lecture become the parameters of a population: fixed
numbers that describe the whole population. The data quantities $\bar x$ and
$s^2$ become statistics: numbers calculated from a sample. The first figure
of Lecture 7 shows how the model quantities and the data quantities are
connected.

::: {.caveat-box}
<span class="label">What an expected value can and cannot promise</span>

- An expected value is not a promise for one case. No deal has 1.32 sharks,
  and no pitch gets "0.612 of a deal".
- Single outcomes can lie far from the expected value. The standard deviation
  says how far, typically.
- All numbers here describe the recorded pitches. For future pitches they are
  only forecasts, and these forecasts depend on the assumptions of Lecture 2.

**What you can do with it:** plan with the expected value when a decision
repeats many times, and look at the spread when one bad outcome would cause
serious damage.
:::

### Answering the driving question

If your pitch ends in a deal, expect to work with one shark. 73.5% of deals
have exactly one shark and 96.5% have at most two. The expected number is
1.32, with a standard deviation of about 0.61. The same two numbers, centre
and spread, describe the other variables of this lecture. The deal indicator
is a Bernoulli variable: a recorded pitch reaches an on-air deal with
probability 0.612, and the variance is about 0.24. The season has expected
value $E[S]\approx8.8$ and a standard deviation of about 4.3 seasons. Centre
and spread together describe a distribution. One of them alone can give a
wrong picture.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your lunch startup sells boxes with prepared meals. Tomorrow you sell 20 meal boxes with probability 0.2, 30 with probability 0.5, and 40 with probability 0.3. The expected value is 31 meal boxes and the standard deviation is 7. A meal box that you prepare but do not sell costs you 3 euros of ingredients. A customer you cannot serve costs you 5 euros of margin. Compare the expected cost of three plans: preparing 20 meal boxes costs 55 euros a day on average, preparing 30 costs 21 euros, and preparing 40 costs 27 euros. The decision: prepare 30. The expected value told you where to aim. The spread and the two costs told you how much a wrong guess costs.
:::

**Next:** We can now say what a typical pitch looks like and how much pitches
vary around it. But those are answers about a pitch we know nothing about.
In practice we always know something: its season, its industry, or the
rating of an AI tool that screens pitches. Lecture 4 asks how probabilities
change when new information arrives. It covers conditional probability and
the total-probability rule, which rebuilds an overall rate from group rates.
It also covers Bayes' rule, which reverses the direction of a question, and
independence, the case in which new information changes nothing.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the most important parts. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions

1. What information does a PMF add to the support of a random variable?
2. Which two rules must every PMF follow?
3. What is the difference between $p_X(x)$ and $F_X(x)$?
4. Why can an expectation be a value the random variable never takes?
5. What does the variance average, and why do we square the deviations first?
6. Why do variance and standard deviation have different units, and which
   one is easier to interpret?

::: {.callout-note collapse="true"}
## Check your answers

1. The support lists possible values. The PMF gives a probability to each
   value.
2. Every probability mass is non-negative, and the masses sum to 1.
3. $p_X(x)=P(X=x)$ is the mass at exactly $x$. $F_X(x)=P(X\le x)$ is all
   probability accumulated up to and including $x$.
4. Expectation is a probability-weighted average and need not belong to the
   support.
5. Variance averages squared deviations from the distribution mean. The
   probability-weighted raw deviations sum to zero. Squaring prevents
   positive and negative deviations from cancelling each other.
6. Variance is in squared units of $X$. The standard deviation is its
   square root and returns to the units of $X$, which makes it the easier
   summary to interpret.
:::

### Finish a half-worked example

Re-read the two tiny-PMF worked examples (@exm-b03-tiny-expectation and
@exm-b03-tiny-variance). Then do all the steps on a new invented
distribution. Let $W$ have support $\{0,1,2\}$ with masses $0.3$, $0.4$,
and $0.3$.

*Step 1 (given).* Validity: all masses are non-negative and
$0.3+0.4+0.3=1$.

*Step 2 (you).* Calculate $F_W(1)$.

*Step 3 (you).* Calculate $E[W]$, writing each product.

*Step 4 (you).* Calculate $E[W^2]$, then $\operatorname{Var}(W)$ with the
computational form, then $\operatorname{SD}(W)$.

::: {.callout-note collapse="true"}
## Check your answer

$F_W(1)=P(W\le1)=0.3+0.4=0.7$.
$E[W]=0(0.3)+1(0.4)+2(0.3)=1.0$.
$E[W^2]=0(0.3)+1(0.4)+4(0.3)=1.6$, so
$\operatorname{Var}(W)=1.6-1.0^2=0.6$ and
$\operatorname{SD}(W)=\sqrt{0.6}\approx0.77$.
Compare $W$ with the variable $Y$ of the worked examples. Both take the
values 0, 1, and 2, but $W$ has the larger variance ($0.6>0.49$), because
more of its probability lies at the two ends, 0 and 2.
:::

### Practice

[Shark Tank data]{.shark-tag} Tasks 1 to 5 use the Shark Tank distributions of the lecture material.
Task 6 uses an invented example.

1. For the deal indicator with $p=882/1441$, calculate $F_X(0.4)$.
2. Use the season CDF table to calculate $P(3<S\le6)$.
3. A Bernoulli variable has $p=0.25$. Calculate its variance and standard
   deviation.
4. The standard deviation of season is about 4.3. Write one sentence that
   interprets this number and names its units.
5. Why is $\operatorname{Var}(S)\approx18.1$ harder to interpret than
   $\operatorname{SD}(S)\approx4.3$?
6. An invented example. A founder sells phone cases at a weekend market.
   The number of cases sold on one day, $X$, has $E[X]=12$ and
   $\operatorname{SD}(X)=4$. Each case brings in 15 euros and the stall
   costs 60 euros per day, so the profit of one day is $15X-60$ euros.
   Calculate the expected value, the variance, and the standard deviation
   of the profit.

::: {.callout-note collapse="true"}
## Check your answers

1. Because $0\le0.4<1$, $F_X(0.4)=P(X=0)=559/1441\approx0.388$.
2. $P(3<S\le6)=F_S(6)-F_S(3)=(495-160)/1441=335/1441\approx0.232$.
3. $\operatorname{Var}=0.25\times0.75=0.1875$.
   $\operatorname{SD}=\sqrt{0.1875}\approx0.43$.
4. The season of a randomly selected pitch typically differs from the
   expected season (about 8.8) by roughly 4.3 seasons.
5. Variance is measured in seasons squared, a unit that is hard to
   imagine. The standard deviation is back in seasons.
6. Use the rules for $aX+b$ with $a=15$ and $b=-60$. The expected profit is
   $15\times12-60=120$ euros. $\operatorname{Var}(X)=4^2=16$, so the
   variance of the profit is $15^2\times16=3600$ euros squared. The standard
   deviation of the profit is $15\times4=60$ euros, half of the expected
   profit. The stall cost of 60 euros lowers the expected profit and leaves
   the spread unchanged.
:::

### Review: one question from Lecture 2

[Shark Tank data]{.shark-tag} Using the Lecture 2 counts table ($|D^c|=559$, $|L|=734$,
$|D^c\cap L|=229$, $n=1441$), calculate $P(D^c\cup L)$ and interpret it in
one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(D^c\cup L)=(559+734-229)/1441=1064/1441\approx0.738$: about 74% of the
recorded pitches failed to reach an on-air deal, came from Seasons 9--16,
or both. We subtract the overlap (229 late-season no-deal records) once, so
that no record counts twice.
:::

### Worked exam interpretation

[Shark Tank data]{.shark-tag} **Question.** For the number of sharks in a deal, R reports, after
rounding, `expected_sharks` 1.32 and `sd` 0.61. Interpret both numbers in
context, with their units.

::: {.callout-note collapse="true"}
## Check a model answer

The 1.32 is the expected number of sharks in a deal: the average of the
shark counts of the recorded deals, each count weighted by its share. No
deal has 1.32 sharks, because an expected value need not be a possible
value. The 0.61 is the standard deviation, in sharks. The number of sharks
in a deal typically differs from 1.32 by about 0.6 sharks, less than one
shark, so the number of sharks differs little between deals. Both numbers describe the
recorded deals. They do not guarantee the number of sharks for a future
pitch.
:::

### Worked exam interpretation: from a PMF to the standard deviation

**Exam-style question.** An invented example. A venture sells one product per customer,
at one of three prices. Let $V$ be the price paid by a randomly selected customer, in
euros, with PMF $p_V(10)=0.5$, $p_V(20)=0.3$, $p_V(40)=0.2$. (a) Verify
that this is a valid PMF. (b) Calculate $F_V(20)$. (c) Calculate $E[V]$
and $E[V^2]$. (d) Calculate $\operatorname{Var}(V)$ and
$\operatorname{SD}(V)$, stating the units of each.

::: {.callout-note collapse="true"}
## Check a model answer

&#40;a) All masses are non-negative and $0.5+0.3+0.2=1$.
&#40;b) $F_V(20)=P(V\le20)=0.5+0.3=0.8$.
&#40;c) $E[V]=10(0.5)+20(0.3)+40(0.2)=5+6+8=19$ euros.
$E[V^2]=100(0.5)+400(0.3)+1600(0.2)=50+120+320=490$.
&#40;d) $\operatorname{Var}(V)=490-19^2=490-361=129$ euros squared.
$\operatorname{SD}(V)=\sqrt{129}\approx11.4$ euros. The expected revenue
per customer is 19 euros. The typical deviation from it is about 11 euros,
more than half of the expected value.
:::

### Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant: "How many sharks will I work with if I get a deal?" It answered:

> "The expected number of sharks is 1.32. The expected value is the most likely value, so in most deals you negotiate with 1.32 sharks."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

No deal has 1.32 sharks. The expected value is a weighted average, not the most likely value. The most likely number is one shark: 73.5% of deals have exactly one. The 1.32 is still useful, for example to plan for many deals, but it describes the centre of the distribution, not a single deal.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Pick one
uncertain quantity with a few possible values, for example the units
sold next week, the price a customer picks, or the orders per day. Sentence
1: describe its approximate PMF and give its expected value. Sentence 2: does
the centre or the spread matter more for a real decision you would make, and
why? There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells boxes of prepared meals on campus. "Daily orders next week are roughly 20 with probability 0.2, 30 with probability 0.5, and 40 with probability 0.3, so the expected value is about 31. The spread matters more than the centre for me, because I prepare ingredients in advance: a day with 40 orders when I have prepared for only 20 costs me more than an average day earns."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to check whether your
PMF is valid and to compute its standard deviation with you.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 3. Show the solution only after I send my answer."

[Continue to Tutorial 1: Evidence and probability foundations](t01-exam-prep.qmd).

[Formula sheet: Lecture 3 formulas](formula-sheet.qmd#formula-l3).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 8.

:::
