# Lecture 3: Discrete distributions, expectation, and variance {#b03}

```{r}
#| label: b03-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 3, you can</span>

- construct, check, and interpret a probability mass function;
- construct and interpret a discrete cumulative distribution function;
- calculate and interpret expectation and the expectation of a function;
- calculate and interpret variance and standard deviation, including the
  Bernoulli case $p(1-p)$;
- connect model expectation to an empirical mean, and distribution variance
  to sample variance, without confusing them.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Anyone preparing a pitch wants to know what to expect: the typical outcome, and how far reality tends to stray from it. Centre and spread are the two numbers every summary of business risk starts from.</p>

What does a typical recorded pitch look like, and how much do pitches differ
from that typical value?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-mass-functions/">Probability Mass Functions</a>,
<a href="https://www.youtube.com/watch?v=4QeL1ma_XJ0">Cumulative Distribution
Functions</a>, and
<a href="https://www.youtube.com/watch?v=_yJsO5955ZE">Expectation</a>.
Be ready to explain what $p_X(x)$ and $F_X(a)$ measure and why an expected
value need not be a possible realised value. All links are also listed in
the <a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (5 minutes).** Lecture 2 ended with random variables on the
uniform record-selection experiment. Define $X=1$ if the selected pitch
reached an on-air deal and $X=0$ otherwise. Define $S$ as its season number.

1. What are the possible values of $X$ and of $S$?
2. Do these possible values tell us how likely each value is?

::: {.callout-note collapse="true"}
## Check your answers

1. $X$ can take the values 0 and 1. $S$ can take the values
   $1,2,\ldots,16$.
2. No. The support lists the possible values. A probability distribution
   must also assign probability to them --- that is what this lecture adds.
:::

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: can the **expected season** be a number
that is not a whole season --- yes or no? Commit to an answer; the lecture
resolves it.

If functions, support, or summation notation are unfamiliar, review
[Mathematics Refresher: functions and sums](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. On screen is a figure with sixteen
probability masses, one per season, heights unlabelled --- by the end of the first block
you can name every height, and by the end of the session you can compress
all sixteen into two numbers: a centre and a spread.

### A PMF assigns probability to each possible value

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-mass-functions/">05.3 Probability Mass Functions</a>.</p>

A random variable identifies the numerical values that may be observed. Its
distribution adds the missing information: how probability is divided among
those values.

For a discrete random variable $X$, the **probability mass function (PMF)**
is

$$p_X(x)=P(X=x).$$

Read this notation in parts:

| Symbol | Meaning |
|---|---|
| $X$ | the random variable |
| $x$ | one possible numerical value |
| $p_X$ | the PMF belonging to $X$ |
| $p_X(x)$ | the probability that $X$ equals exactly $x$ |

A valid PMF must satisfy

$$
p_X(x)\ge 0\quad\text{for every }x,
\qquad
\sum_{x\in\mathcal X}p_X(x)=1,
$$

where $\mathcal X$ is the support of $X$. The first rule prevents negative
probability. The second says that one of the possible values must occur.

::: {#exm-b03-deal-pmf}
## Worked example: the deal-indicator PMF, checked

Under the uniform selection experiment from Lecture 2, the deal indicator
$X$ has support $\{0,1\}$. From the counts --- 882 deals among 1,441
records ---

$$
p_X(0)=\frac{559}{1441}\approx 0.388,
\qquad
p_X(1)=\frac{882}{1441}\approx 0.612.
$$

Check both rules: each mass is non-negative, and

$$
\frac{559}{1441}+\frac{882}{1441}=\frac{1441}{1441}=1.
$$

This is a valid PMF.
:::

The same objects in R --- the data, the deal proportion `p_hat`, and the PMF
are all built in visible code:

```{r}
#| label: b03-deal-pmf
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
p_hat <- mean(sharks$deal_on_show)   # 882/1441

deal_pmf <- prop.table(
  table(factor(sharks$deal_on_show, levels = c(0, 1)))
)
names(deal_pmf) <- c("0: no on-air deal", "1: on-air deal")
deal_pmf
sum(deal_pmf)
```

For season $S$, the empirical PMF assigns each season its relative
frequency under the uniform record-selection experiment:

```{r}
#| label: b03-season-pmf
season_pmf <- prop.table(table(sharks$season))
round(season_pmf, 4)
sum(season_pmf)
```

```{r}
#| label: fig-b03-season-pmf
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The PMF of season under uniform record selection: sixteen separated probability masses, one per possible value."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_pmf <- prop.table(table(sharks$season))
pmf_df <- data.frame(
  season = as.numeric(names(season_pmf)),
  probability = as.numeric(season_pmf)
)

ggplot(pmf_df, aes(x = season, y = probability)) +
  geom_segment(aes(xend = season, yend = 0),
               linewidth = 1.1, colour = "#009E73") +
  geom_point(size = 2.6, colour = "#009E73") +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Season s", y = expression(p[S](s))) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The probability mass above season 5 in @fig-b03-season-pmf has height
$p_S(5)=P(S=5)=116/1441\approx0.080$. A PMF is represented by separated
probability masses because $S$ can take only the listed discrete values.
The term **probability density function (PDF)** is reserved for continuous
random variables and is introduced later in the course.

### A CDF accumulates probability up to a threshold

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/cumulative-distribution-functions/">08.7 Cumulative Distribution Functions</a>.</p>

The PMF asks about one exact value. The **cumulative distribution function
(CDF)** asks how much probability has accumulated at or below a chosen
threshold. Imagine moving from left to right across the PMF and keeping a
running total.

For any random variable $X$,

$$F_X(a)=P(X\le a).$$

Here $a$ is a threshold, not necessarily a value in the support. For a
discrete random variable,

$$F_X(a)=\sum_{x\le a}p_X(x).$$

::: {#exm-b03-cdf-hand}
## Worked example: F(3) for season, by hand

"$S\le 3$" collects the first three seasons, so add the first three
probability masses of the PMF. The season counts are 64, 36, and 60:

$$
F_S(3)=p_S(1)+p_S(2)+p_S(3)
=\frac{64}{1441}+\frac{36}{1441}+\frac{60}{1441}
=\frac{160}{1441}\approx0.111.
$$

About 11% of the probability sits at or below season 3.
:::

The full running total in R --- each CDF value is the current PMF value
plus all earlier PMF values:

```{r}
#| label: b03-season-distribution
s_values <- as.numeric(names(season_pmf))
p_values <- as.numeric(season_pmf)
season_distribution <- data.frame(
  season = s_values,
  pmf = round(p_values, 4),
  cdf = round(cumsum(p_values), 4)
)
season_distribution
```

```{r}
#| label: fig-b03-season-cdf
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The CDF of season: a step graph that never falls, jumping at each possible value by that value's probability mass."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
season_pmf <- prop.table(table(sharks$season))
cdf_df <- data.frame(
  threshold = as.numeric(names(season_pmf)),
  cdf = cumsum(as.numeric(season_pmf))
)

ggplot(cdf_df, aes(x = threshold, y = cdf)) +
  geom_step(linewidth = 0.9, colour = "#009E73") +
  geom_point(size = 1.9, colour = "#17324D") +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Threshold a", y = expression(F[S](a))) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

A CDF accumulates probability as the threshold moves from left to right. It
can stay flat or rise, but it cannot fall. It begins near 0 before any
possible values and reaches 1 after all possible values have been included.

For a discrete variable, the CDF is a step graph. It stays flat between
possible values and jumps at each value that has positive probability. A
larger jump means that more probability is attached to that value.

::: {.callout-note collapse="true"}
## Check: read F(8) off the step graph

$F_S(8)=P(S\le 8)$ is the height of the step graph at threshold 8: the
first eight seasons contain $64+36+60+103+116+116+116+96=707$ records, so
$F_S(8)=707/1441\approx0.491$. Just under half of the probability lies in
Seasons 1--8 --- consistent with Lecture 2, where the early-period event
$L^c$ had probability $707/1441$.
:::

For the deal indicator:

$$
F_X(a)=
\begin{cases}
0, & a<0,\\
1-p, & 0\le a<1,\\
1, & a\ge1,
\end{cases}
$$

where $p=P(X=1)$. Thus $p_X(1)=P(X=1)=p$, while $F_X(1)=P(X\le1)=1$. The
PMF measures mass at exactly 1; the CDF includes all mass through 1.

### Expectation is the distribution's weighted centre

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/expectation/">05.8 Expectation</a>.</p>

A distribution can contain many possible values and probabilities.
**Expectation** summarises its centre in one number. Values with larger
probabilities receive more weight:

$$E[X]=\sum_{x\in\mathcal X}x\,p_X(x).$$

This definition applies when the probability-weighted sum is well defined
and finite. The notation $E[X]$ is read "the expected value of $X$." The
calculation multiplies each possible value by its probability and adds the
contributions. It can also be viewed as the balance point of the
probability distribution.

Before touching the Shark Tank data, run the calculation once on a
distribution small enough to hold in your head.

::: {#exm-b03-tiny-expectation}
## Worked example: expectation of a tiny PMF

Let $Y$ have support $\{0,1,2\}$ with masses

$$p_Y(0)=0.2,\qquad p_Y(1)=0.5,\qquad p_Y(2)=0.3.$$

(Valid: all non-negative, and $0.2+0.5+0.3=1$.) Write every product:

$$
E[Y]
=0\times0.2 \;+\; 1\times0.5 \;+\; 2\times0.3
=0 + 0.5 + 0.6
=1.1.
$$

Each possible value contributes its value times its probability; the sum,
1.1, is the balance point. Note that 1.1 is not in the support --- $Y$
never equals 1.1.
:::

Now the same recipe on the Shark Tank distributions. For the deal
indicator,

$$E[X]=0\times(1-p)+1\times p=p.$$

This identity explains why the mean of a 0/1 variable equals its proportion
of 1s: here $E[X]=882/1441\approx0.612$.

For season, sixteen products instead of two:

```{r}
#| label: b03-expectation
s <- as.numeric(names(season_pmf))
p_s <- as.numeric(season_pmf)

c(
  E_X = 0 * (1 - p_hat) + 1 * p_hat,
  E_S_from_pmf = sum(s * p_s),
  observed_mean_season = mean(sharks$season)
)
```

The expected season is about **8.8** --- and there is no Season 8.8. This
resolves the prediction from your prep: an expectation is a
probability-weighted average, so it need not be a possible value of the
random variable.

Under independent repetitions of the same distribution, provided
$E[|X|]<\infty$, the sample average tends to become close to $E[X]$ as the
number of observations grows. This long-run interpretation does not make
the expected value a guarantee for a single observation.

### Expectation after transforming a variable

If $g$ transforms each possible value of $X$, the expected transformed
value is

$$E[g(X)]=\sum_{x\in\mathcal X}g(x)\,p_X(x).$$

The probabilities do not change: apply $g$ to each possible value, weight
the transformed values by the original PMF, and add.

::: {.callout-caution collapse="true"}
## Try it: E[Y²] for the tiny PMF

In class this is the pair task before the break. Same distribution as the
worked example: $Y\in\{0,1,2\}$ with masses $0.2,0.5,0.3$, and $g(y)=y^2$.

*Step 1 (given).* Square each possible value: $0^2=0$, $1^2=1$, $2^2=4$.

*Step 2 (you).* Weight each squared value by its **original** probability
and write the three products.

*Step 3 (you).* Add them to obtain $E[Y^2]$. Is it equal to
$\left(E[Y]\right)^2=1.1^2$?

::: {.callout-note collapse="true"}
## Check your answer

$$
E[Y^2]=0\times0.2+1\times0.5+4\times0.3=0+0.5+1.2=1.7.
$$

It is **not** $\left(E[Y]\right)^2=1.21$: squaring then averaging differs
from averaging then squaring. The gap between the two, $1.7-1.21=0.49$, is
about to get a name.
:::
:::

The same rule for season in R:

```{r}
#| label: b03-transformed-expectation
c(
  E_S = sum(s * p_s),
  E_S_squared = sum(s^2 * p_s)
)
```

### Variance and standard deviation

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/variance/">06.2 Variance</a>.</p>

Expectation locates the centre of a distribution, but it does not describe
how tightly values cluster around that centre. Two distributions can have
the same expectation and very different spreads --- this is the Mentimeter
question posed before the break: both distributions in
@fig-b03-same-mean-spread have expectation 1, so which is riskier, and why?

```{r}
#| label: fig-b03-same-mean-spread
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.15
#| fig-cap: "Two distributions with the same expectation and very different spreads. The dashed line marks the common mean, 1."
library(ggplot2)
spread_df <- data.frame(
  x = rep(0:2, times = 2),
  probability = c(0.10, 0.80, 0.10,    # Distribution A
                  0.45, 0.10, 0.45),   # Distribution B
  distribution = rep(c("Distribution A (variance = 0.20)",
                       "Distribution B (variance = 0.90)"), each = 3)
)

ggplot(spread_df, aes(x = x, y = probability)) +
  geom_vline(xintercept = 1, colour = "#C98B2E",
             linetype = "dashed", linewidth = 0.8) +
  geom_segment(aes(xend = x, yend = 0),
               linewidth = 1.1, colour = "#009E73") +
  geom_point(size = 2.6, colour = "#009E73") +
  annotate("text", x = 1.1, y = 0.87, label = "mean = 1",
           hjust = 0, colour = "#17324D", size = 3.4) +
  facet_wrap(~ distribution) +
  scale_x_continuous(breaks = 0:2) +
  scale_y_continuous(limits = c(0, 0.9),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Possible value of X", y = "Probability") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

The horizontal axis lists the possible values and the vertical axis gives
their probabilities. The centre is the same, but Distribution B has greater
spread because values 0 and 2 receive much more probability. If these were
payoffs, most students call B riskier --- variance is the number that backs
that intuition.

To measure spread, start with the deviation $X-\mu$, where $\mu=E[X]$.
Positive and negative deviations would cancel if they were averaged
directly, so variance squares each deviation before taking its
probability-weighted average:

$$
\operatorname{Var}(X)
=E[(X-\mu)^2]
=\sum_x (x-\mu)^2\,p_X(x).
$$

Read the calculation as four steps:

1. locate the centre, $\mu=E[X]$;
2. measure each possible deviation, $x-\mu$;
3. square the deviation, $(x-\mu)^2$;
4. weight it by $p_X(x)$ and add across the support.

Large deviations therefore contribute more than small deviations. Variance
is zero only when all probability is concentrated at one value.

The equivalent **computational form** is

$$\operatorname{Var}(X)=E[X^2]-\left(E[X]\right)^2.$$

It gives the same result, but the definition $E[(X-\mu)^2]$ is usually
easier to interpret.

The **standard deviation** is

$$\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.$$

Variance has squared units; standard deviation returns to the units of $X$.
If $X$ is measured in seasons, $\operatorname{Var}(X)$ is in seasons
squared --- a unit nobody can picture --- while $\operatorname{SD}(X)$ is
back in seasons.

::: {#exm-b03-tiny-variance}
## Worked example: variance of the tiny PMF, both ways

Same $Y$ as before: values $0,1,2$ with masses $0.2,0.5,0.3$, and
$\mu=E[Y]=1.1$.

**By the definition** --- the four steps:

$$
\operatorname{Var}(Y)
=(0-1.1)^2(0.2)+(1-1.1)^2(0.5)+(2-1.1)^2(0.3)
$$
$$
=1.21\times0.2+0.01\times0.5+0.81\times0.3
=0.242+0.005+0.243
=0.49.
$$

**By the computational form** --- using $E[Y^2]=1.7$ from the faded task:

$$
\operatorname{Var}(Y)=E[Y^2]-\left(E[Y]\right)^2=1.7-1.21=0.49.
$$

Both routes give 0.49, so $\operatorname{SD}(Y)=\sqrt{0.49}=0.7$ --- in the
same units as $Y$.
:::

For a Bernoulli variable (the deal indicator), the computational form gives
a formula worth memorising. Since $X^2=X$ when $X$ is 0 or 1,
$E[X^2]=E[X]=p$, so

$$
\operatorname{Var}(X)=p-p^2=p(1-p).
$$

With $p=882/1441\approx0.612$, the variance is
$0.612\times0.388\approx0.237$ and the standard deviation is about 0.49.

```{r}
#| label: b03-variance
mu_s <- sum(s * p_s)
var_s <- sum((s - mu_s)^2 * p_s)

c(
  Var_X = p_hat * (1 - p_hat),
  SD_X = sqrt(p_hat * (1 - p_hat)),
  Var_S = var_s,
  SD_S = sqrt(var_s)
)
```

Season has variance of about 18.1 seasons squared and standard deviation of
about **4.3 seasons**: the typical distance between a randomly selected
pitch's season and the expected season 8.8 is roughly four seasons.

### Model expectation and the sample mean

For observed values $x_1,\ldots,x_n$, the sample mean is

$$\bar{x}=\frac{1}{n}\sum_{i=1}^n x_i.$$

If the empirical distribution assigns probability $1/n$ to every observed
row, its expectation is

$$\sum_{i=1}^n x_i\,\frac{1}{n}=\bar{x}.$$

The numerical answers agree, but the terms identify different objects.
Expectation describes a probability distribution; the sample mean
summarises observed data.

The same split exists for spread. The **empirical-distribution variance**
weights each of the $n$ observations by $1/n$ --- it divides by $n$. The
usual **sample variance** instead divides by $n-1$:

$$s^2=\frac{1}{n-1}\sum_{i=1}^n\left(x_i-\bar{x}\right)^2.$$

R's `var()` and `sd()` use the $n-1$ denominator. Why $n-1$ is the right
choice when estimating a population variance is a Lecture 7 question; for
now, only recognise which convention a number uses. Live in R, on the
description word counts:

```{r}
#| label: b03-word-mean-variance
w <- sharks$description_words
c(
  mean_words = mean(w),
  empirical_expectation = sum(w * rep(1 / length(w), length(w))),
  empirical_variance_divide_n = mean((w - mean(w))^2),
  sample_variance_divide_n_minus_1 = var(w),
  sample_sd = sd(w)
)
```

Both centre calculations give about 5.0 words for these 1,441 records, and
the two variance conventions differ only in the third decimal because $n$
is large. These are different summaries of the same observations; calling
any of them an estimate for a wider population additionally requires a
sampling model.

::: {.caveat-box}
<span class="label">What an expected value can and cannot promise</span>

- **Expectation is not a guarantee.** $E[X]=0.612$ does not mean any single
  pitch gets "0.612 of a deal," and an expected season of 8.8 belongs to no
  actual season. Expectation describes the distribution's balance point;
  single realisations can sit far from it --- how far, on typical, is
  exactly what the standard deviation measures.
- **File values are not forecasts.** Every number in this lecture is
  computed under the model "select one of the 1,441 recorded pitches
  uniformly at random." Treating 0.612, 8.8, or 4.3 as statements about
  future pitches or new entrepreneurs requires the population, selection,
  stability, and outcome-definition assumptions from Lecture 2 --- the file
  cannot supply them.
:::

### Answering the driving question

A typical recorded pitch, in two numbers per variable: it reaches an on-air
deal with probability $0.612$ (a Bernoulli distribution with variance
$0.612\times0.388\approx0.24$), and it comes from around season
$E[S]\approx8.8$ --- with a standard deviation of about $4.3$ seasons, so
"typical" spans a wide stretch of the show's history. Centre and spread
together describe the distribution; either alone misleads.

**Next:** We can now say what a typical pitch looks like and how much pitches
vary around it. But those are answers about a pitch we know *nothing* about
--- and in practice we always know something: its season, its industry, a flag
from a screening model. Lecture 4 asks how probabilities change when
information arrives: conditional probability, the total-probability rule that
rebuilds an overall rate from group rates, Bayes' rule for reversing the
direction of a question, and independence --- the exact case in which
information changes nothing.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 60 minutes.</span>

### Retrieval (10 minutes)

1. What information does a PMF add to the support of a random variable?
2. What two requirements must every PMF satisfy?
3. What is the difference between $p_X(x)$ and $F_X(x)$?
4. Why can an expectation be a value the random variable never takes?
5. What does variance average, and why are deviations squared first?
6. Why do variance and standard deviation have different units, and which
   one is easier to interpret?

::: {.callout-note collapse="true"}
## Check your answers

1. The support lists possible values; the PMF assigns probability to each
   value.
2. Every probability mass is non-negative, and the masses sum to 1.
3. $p_X(x)=P(X=x)$ is the mass at exactly $x$; $F_X(x)=P(X\le x)$ is all
   probability accumulated through $x$.
4. Expectation is a probability-weighted average and need not belong to the
   support.
5. Variance averages squared deviations from the distribution mean; the
   probability-weighted raw deviations sum to zero, so squaring stops
   positive and negative deviations from cancelling.
6. Variance is in squared units of $X$; the standard deviation is its
   square root and returns to the units of $X$, which makes it the easier
   summary to interpret.
:::

### From worked to faded (10 minutes)

Re-read the two tiny-PMF worked examples (@exm-b03-tiny-expectation and
@exm-b03-tiny-variance). Then run the full ladder on a new distribution:
let $W$ have support $\{0,1,2\}$ with masses $0.3$, $0.4$, and $0.3$.

*Step 1 (given).* Validity: all masses are non-negative and
$0.3+0.4+0.3=1$.

*Step 2 (you).* Calculate $F_W(1)$.

*Step 3 (you).* Calculate $E[W]$, writing each product.

*Step 4 (you).* Calculate $E[W^2]$, then $\operatorname{Var}(W)$ with the
computational form, then $\operatorname{SD}(W)$.

::: {.callout-note collapse="true"}
## Check your answer

$F_W(1)=P(W\le1)=0.3+0.4=0.7$.
$E[W]=0(0.3)+1(0.4)+2(0.3)=1.0$.
$E[W^2]=0(0.3)+1(0.4)+4(0.3)=1.6$, so
$\operatorname{Var}(W)=1.6-1.0^2=0.6$ and
$\operatorname{SD}(W)=\sqrt{0.6}\approx0.77$.
Note $W$ has the same expectation as a coin-like centre of 1 but a larger
spread than $Y$ from class ($0.6>0.49$): more mass sits at the ends.
:::

### Practice (10 minutes)

Now unassisted, using the Shark Tank distributions built in class.

1. For the deal indicator with $p=882/1441$, calculate $F_X(0.4)$.
2. Use the season CDF table to calculate $P(3<S\le6)$.
3. A Bernoulli variable has $p=0.25$. Calculate its variance and standard
   deviation.
4. The standard deviation of season is about 4.3. Write one sentence
   interpreting this number, naming its units.
5. Why is $\operatorname{Var}(S)\approx18.1$ harder to interpret than
   $\operatorname{SD}(S)\approx4.3$?

::: {.callout-note collapse="true"}
## Check your answers

1. Because $0\le0.4<1$, $F_X(0.4)=P(X=0)=559/1441\approx0.388$.
2. $P(3<S\le6)=F_S(6)-F_S(3)=(495-160)/1441=335/1441\approx0.232$.
3. $\operatorname{Var}=0.25\times0.75=0.1875$;
   $\operatorname{SD}=\sqrt{0.1875}\approx0.43$.
4. The season of a randomly selected pitch typically differs from the
   expected season (about 8.8) by roughly 4.3 **seasons**.
5. Variance is measured in seasons squared, a unit without a direct
   picture; the standard deviation is back in seasons.
:::

### Spaced review: one item from Lecture 2 (5 minutes)

Using the Lecture 2 counts table ($|D^c|=559$, $|L|=734$,
$|D^c\cap L|=229$, $n=1441$), calculate $P(D^c\cup L)$ and interpret it in
one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(D^c\cup L)=(559+734-229)/1441=1064/1441\approx0.738$: about 74% of the
recorded pitches failed to reach an on-air deal, came from Seasons 9--16,
or both. The overlap (229 late-season no-deal records) is subtracted once
so no record is counted twice.
:::

### Worked exam interpretation (10 minutes)

**Question.** R reports `mean(deal_on_show) = 0.612`. Explain this number
as both a sample mean and an empirical expectation.

::: {.callout-note collapse="true"}
## Check a model answer

Because the indicator equals 1 for an on-air agreement, its sample mean is
the observed deal proportion: 61.2% of the 1,441 records. If the empirical
distribution assigns mass $1/1441$ to each record, its expectation is the
same weighted average, 0.612. This does not show later venture success or
guarantee the probability for a future pitch.
:::

### Exam practice: a full moment set (10 minutes)

**Question.** A venture sells one product per customer at three price
points. Let $V$ be the price paid by a randomly selected customer, in
euros, with PMF $p_V(10)=0.5$, $p_V(20)=0.3$, $p_V(40)=0.2$. (a) Verify
that this is a valid PMF. (b) Calculate $F_V(20)$. (c) Calculate $E[V]$
and $E[V^2]$. (d) Calculate $\operatorname{Var}(V)$ and
$\operatorname{SD}(V)$, stating the units of each.

::: {.callout-note collapse="true"}
## Check a model answer

(a) All masses are non-negative and $0.5+0.3+0.2=1$.
(b) $F_V(20)=P(V\le20)=0.5+0.3=0.8$.
(c) $E[V]=10(0.5)+20(0.3)+40(0.2)=5+6+8=19$ euros;
$E[V^2]=100(0.5)+400(0.3)+1600(0.2)=50+120+320=490$.
(d) $\operatorname{Var}(V)=490-19^2=490-361=129$ euros squared;
$\operatorname{SD}(V)=\sqrt{129}\approx11.4$ euros. The expected revenue
per customer is 19 euros, with a typical deviation of about 11 euros.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Pick one
uncertain quantity with a handful of possible values (units sold next
week, price a customer picks, orders per day) and write two sentences:
first sketch its rough PMF and expected value, then say whether the centre
or the spread matters more for a real decision you would take, and why.
There is no wrong answer; the point is connecting today's tools to
something you care about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Daily orders next week are roughly 20 with probability 0.2, 30 with probability 0.5, and 40 with probability 0.3, so the expected value is about 31. The spread matters more than the centre for me, because I prep ingredients in advance --- a 40-order day with 20-order stock hurts more than an average day helps."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to check whether your
sketched PMF is valid and to compute its standard deviation with you.
::::

Continue to Lecture 4: Conditioning, total probability, Bayes, and
independence.

Formula sheet: Lecture 3 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 8.

:::
