# Lecture 4: Conditioning, Bayes' rule, and independence {#b04}

```{r}
#| label: b04-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 4, you can</span>

- calculate conditional probabilities from a counts table and identify the
  reference class behind each one;
- use the multiplication rule to move between joint and conditional
  probabilities;
- rebuild an overall probability from group-specific rates with the law of
  total probability;
- update a probability after new evidence with Bayes' rule, first by counting
  on a natural-frequency tree and then with the formula;
- check whether two events are independent and explain why disjoint is not
  the same as independent.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">An investor who learns a pitch is late-season updates their expectations. A bank seeing a positive fraud flag does the same. Updating chances on new information is the most used probability skill in business --- and the easiest to get backwards.</p>

Late-season pitches reach deals more often. What exactly do the two possible
"68.8% versus 57.3%" sentences claim, and what does a positive flag from an
AI deal-predictor really tell us?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=MPRKc4UPoJk">Conditional
Probabilities</a> and
<a href="https://www.youtube.com/watch?v=kz2tvO_ZAKI">Bayes' Rule</a>. Be
ready to identify the reference class in a conditional probability and to
explain why reversing a condition changes the denominator. All links are
also listed in the <a href="../resources/video-guide.html">External Video
Guide</a>.
</div>

**Reactivate (10 minutes).** Lecture 2 built the 2×2 counts table for deals
and seasons. Rebuild it --- the construction of $D$ and $L$ is the same as
there:

```{r}
#| label: b04-prep-table
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
table(deal = D, late = L)
```

Then compare, in words only:

- "Among pitches in Seasons 9--16, what proportion reached an on-air deal?"
- "Among pitches that reached an on-air deal, what proportion came from
  Seasons 9--16?"

Both questions will use the same cell of your table. Decide before class
which part of the table each question restricts attention to. The lecture
opens with exactly this exercise.

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: an AI classifier flags pitches likely to
get a deal and catches 80% of actual deals. Of the *flagged* pitches, what
share are deals --- near 80%, near 50%, or near 30%? Commit to an answer; the
lecture resolves it.

If fractions or the conditioning bar are unfamiliar, review
[Mathematics Refresher: fractions, powers, and event
notation](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. Two sentences circulate about the same
dataset:

- **Sentence A.** "68.8% of late-season pitches reached an on-air deal."
- **Sentence B.** "57.3% of the pitches that reached a deal came from the
  late seasons."

Both sentences are built on the *same* 505 records --- the late-season deals.
Yet they answer different questions and produce different numbers. By the end
of this session you can say precisely what each sentence claims, and you can
answer a harder question of the same kind: when an AI classifier flags a
pitch as a likely deal, how much should that flag change your mind?

### One table, five denominators

Everything starts from the Lecture 2 counts table. The code constructs $D$
and $L$ exactly as before --- run it yourself:

```{r}
#| label: b04-counts-table
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
table(deal = D, late = L)
```

Written with row and column totals:

|                    | Early ($L^c$, S1--8) | Late ($L$, S9--16) | Total |
|--------------------|---------------------:|-------------------:|------:|
| **No deal** ($D^c$) | 330 | 229 | **559** |
| **Deal** ($D$)      | 377 | 505 | **882** |
| **Total**           | **707** | **734** | **1441** |

Before any formula, do the class opening exercise: for each probability
below, point --- physically, on the table --- to the count that goes on top
and the count that goes underneath.

1. $P(D)$;
2. $P(L)$;
3. $P(D\cap L)$;
4. $P(D\mid L)$;
5. $P(L\mid D)$.

::: {.callout-note collapse="true"}
## Check your denominators

1. $P(D)$: 882 deal records out of all 1,441 records.
2. $P(L)$: 734 late-season records out of all 1,441 records.
3. $P(D\cap L)$: 505 late-season deal records out of all 1,441 records.
4. $P(D\mid L)$: 505 late-season deal records out of the **734 late-season
   records**.
5. $P(L\mid D)$: 505 late-season deal records out of the **882 deal
   records**.

The first three all divide by 1,441. The last two divide by a *part* of the
table --- and that is the whole subject of this lecture.
:::

### Conditional probability

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/conditional-probabilities/">02.2 Conditional Probabilities</a>.</p>

For events $A$ and $B$ with $P(B)>0$, the **conditional probability of $A$
given $B$** is

$$P(A\mid B)=\frac{P(A\cap B)}{P(B)}.$$

Read the definition as an instruction in three steps: keep only the outcomes
inside $B$, rescale their total probability to 1, and then ask what fraction
of that reduced world is also in $A$. The event after the vertical bar is the
**reference class** --- the group you restrict attention to --- and it
supplies the denominator. For a fixed $B$, $P(\cdot\mid B)$ is itself a
probability law: it satisfies all three rules from Lecture 2.

::: {#exm-b04-conditional}
## Worked example: the two rival sentences, resolved

**Sentence A asks about $P(D\mid L)$.** Conditioning on $L$ means covering
everything except the Late column:

|                    | Early | **Late ($L$)** |
|--------------------|------:|---------------:|
| No deal            | · | **229** |
| Deal               | · | **505** |
| Total              | · | **734** |

Within that one column, 505 of 734 records are deals:

$$P(D\mid L)=\frac{505}{734}\approx 0.688.$$

**Sentence B asks about $P(L\mid D)$.** Conditioning on $D$ means covering
everything except the Deal row:

|                    | Early | Late | Total |
|--------------------|------:|-----:|------:|
| No deal            | · | · | · |
| **Deal ($D$)**     | **377** | **505** | **882** |

Within that one row, 505 of 882 records are late-season:

$$P(L\mid D)=\frac{505}{882}\approx 0.573.$$

The numerator is 505 in both calculations --- the same intersection cell.
The denominators, and therefore the questions, differ: Sentence A restricts
to the 734 late-season pitches, Sentence B to the 882 deal pitches.
:::

::: {.callout-note collapse="true"}
## Check: which sentence matches which number?

Sentence A ("of late-season pitches, ... reached a deal") conditions on the
season, so it is $P(D\mid L)=505/734\approx 68.8\%$. Sentence B ("of deal
pitches, ... came from late seasons") conditions on the deal, so it is
$P(L\mid D)=505/882\approx 57.3\%$. A report that quotes "68.8% of deal
pitches came from late seasons" has attached the right number to the wrong
sentence.
:::

Verification in R --- subsetting with `[L]` or `[D]` is exactly the
"restrict the reference class" step:

```{r}
#| label: b04-conditionals
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  P_deal_given_late = mean(D[L]),
  P_late_given_deal = mean(L[D])
)
```

### The multiplication rule

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-multiplication-rule/">02.6 The Multiplication Rule</a>.</p>

Multiply both sides of the definition by $P(B)$ and the same content appears
in a new role --- a way to *build* a joint probability from a conditional one:

$$P(A\cap B)=P(A\mid B)\,P(B).$$

Reversing the roles of $A$ and $B$ must give the same joint probability,
because $A\cap B$ and $B\cap A$ are the same event:

$$P(A\cap B)=P(A\mid B)\,P(B)=P(B\mid A)\,P(A).$$

This one line is the algebraic reason the two rival sentences can share a
numerator while having different denominators: both products equal
$505/1441$, but they factor it through different reference classes.

::: {.callout-caution collapse="true"}
## Try it: verify the joint probability two ways

In class this is the pair task. Use the counts table to check the identity
above for $D$ and $L$.

*Step 1 (given).* The direct route: $P(D\cap L)=505/1441\approx 0.350$.

*Step 2 (given).* First factorisation:
$P(D\mid L)\,P(L)=\dfrac{505}{734}\cdot\dfrac{734}{1441}$. The 734 cancels
--- carry out the cancellation and confirm you land on $505/1441$.

*Step 3 (you).* Write down the second factorisation,
$P(L\mid D)\,P(D)$, as a product of two fractions from the table, and show
that it also collapses to $505/1441$.

*Step 4 (you).* Confirm all three routes in R.

::: {.callout-note collapse="true"}
## Check your answer

$P(L\mid D)\,P(D)=\dfrac{505}{882}\cdot\dfrac{882}{1441}=\dfrac{505}{1441}
\approx 0.350$: the 882 cancels, just as the 734 did.

```{r}
#| label: b04-multiplication
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  direct_joint            = mean(D & L),
  cond_on_late_times_base = mean(D[L]) * mean(L),
  cond_on_deal_times_base = mean(L[D]) * mean(D)
)
```
:::
:::

### The law of total probability

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/total-probability-theorem/">02.7 Total Probability Theorem</a>.</p>

Lecture 2 introduced a partition: events that do not overlap and together
cover every outcome. Conditioning gives the probability of $A$ *within* each
part; the multiplication rule turns each conditional probability into a joint
piece; and adding the pieces gives the overall probability of $A$. That chain
of three known moves is the **law of total probability**: for a partition
$B_1,\ldots,B_k$ with every $P(B_i)>0$,

$$P(A)=\sum_{i=1}^{k}P(A\mid B_i)\,P(B_i).$$

::: {#exm-b04-total}
## Worked example: rebuilding P(D) from the period-specific rates

The early/late events $\{L^c, L\}$ form a partition, so

$$P(D)=P(D\mid L^c)\,P(L^c)+P(D\mid L)\,P(L)
=\frac{377}{707}\cdot\frac{707}{1441}
+\frac{505}{734}\cdot\frac{734}{1441}
=\frac{377+505}{1441}=\frac{882}{1441}\approx 0.612.$$

Point to the two kinds of number: $377/707\approx 0.533$ and
$505/734\approx 0.688$ are the *period-specific deal rates*, and
$707/1441\approx 0.491$ and $734/1441\approx 0.509$ are the **weights** ---
how much of the sample space each period occupies. The overall deal
probability is a weighted average of the group rates, so it must lie between
0.533 and 0.688, closer to the rate of the slightly larger group.
:::

```{r}
#| label: b04-total-probability
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  direct       = mean(D),
  from_periods = mean(D[!L]) * mean(!L) + mean(D[L]) * mean(L)
)
```

@fig-b04-partition-areas shows the same computation as areas.

```{r}
#| label: fig-b04-partition-areas
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.6
#| fig-cap: "Total probability as areas. The columns are the early/late partition from Lecture 2; column width is the period's probability (the weight), and the height of each coloured region is the period-specific deal rate. The two coloured areas do not overlap, and their sum is P(D)."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
p_early <- mean(!L)
p_deal_early <- mean(D[!L])
p_deal_late <- mean(D[L])

ggplot() +
  # light backgrounds: the two parts of the partition
  annotate("rect", xmin = 0, xmax = p_early, ymin = 0, ymax = 1,
           fill = "#009E73", alpha = 0.25) +
  annotate("rect", xmin = p_early, xmax = 1, ymin = 0, ymax = 1,
           fill = "#D55E00", alpha = 0.25) +
  # solid areas: deal rate within each part, times the part's width
  annotate("rect", xmin = 0, xmax = p_early, ymin = 0,
           ymax = p_deal_early, fill = "#009E73", alpha = 0.85) +
  annotate("rect", xmin = p_early, xmax = 1, ymin = 0,
           ymax = p_deal_late, fill = "#D55E00", alpha = 0.85) +
  annotate("segment", x = p_early, xend = p_early, y = 0, yend = 1,
           colour = "#17324D", linewidth = 0.8) +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1,
           fill = NA, colour = "#17324D", linewidth = 0.8) +
  annotate("text", x = p_early / 2, y = 0.28, label = "D ∩ Lᶜ",
           colour = "white", size = 4.2) +
  annotate("text", x = p_early + (1 - p_early) / 2, y = 0.36,
           label = "D ∩ L", colour = "white", size = 4.2) +
  annotate("text", x = p_early / 2, y = 0.88, label = "L^c",
           parse = TRUE, colour = "#17324D", size = 4.4) +
  annotate("text", x = p_early + (1 - p_early) / 2, y = 0.88,
           label = "L", colour = "#17324D", size = 4.4) +
  annotate("text", x = 0.985, y = 0.955, label = "Omega",
           parse = TRUE, colour = "#17324D", size = 3.8) +
  coord_cartesian(expand = FALSE) +
  theme_void()
```

### Bayes' rule: count first, formula second

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/bayes-rule/">02.8 Bayes' Rule</a>.</p>

Return to your prep prediction. An AI classifier reads pitch transcripts and
flags the pitches it considers likely to reach a deal. Three numbers describe
a (hypothetical) classifier of this kind:

- the **base rate**: 10% of pitches in its target population reach a deal;
- the **sensitivity**: among actual deals, 80% get flagged,
  $P(\text{flag}\mid\text{deal})=0.80$;
- the **false-positive rate**: among non-deals, 20% still get flagged,
  $P(\text{flag}\mid\text{no deal})=0.20$.

A pitch has just been flagged. How likely is a deal now --- near 80%, 50%,
or 30%? Before any formula, count.

::: {#exm-b04-bayes-tree}
## Worked example: 1,000 imaginary pitches

Imagine 1,000 pitches from this population and push them through the three
numbers, as in @fig-b04-bayes-tree.

1. **Split by the base rate.** Of 1,000 pitches, 100 are deals and 900 are
   not.
2. **Flag the deals.** Sensitivity 80%: of the 100 deals, 80 are flagged.
3. **Flag some non-deals too.** False-positive rate 20%: of the 900
   non-deals, 180 are flagged.
4. **Collect the flagged.** In total $80+180=260$ pitches carry a flag ---
   and only 80 of them are deals:

$$P(\text{deal}\mid\text{flag})=\frac{80}{260}\approx 0.31.$$

The classifier catches 80% of deals, yet fewer than a third of its flags are
deals. The reason is visible in the tree: the 900 non-deals are such a large
group that even a modest 20% false-positive rate produces 180 false flags ---
more than double the 80 true ones. A flag moves the deal probability from
10% to 31%: real evidence, but nowhere near certainty.
:::

```{r}
#| label: fig-b04-bayes-tree
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-width: 8
#| fig-height: 4.6
#| fig-cap: "Natural-frequency tree for 1,000 imaginary pitches. Orange boxes are the flagged pitches: 80 + 180 = 260 flags in total, of which only 80 are deals, so P(deal | flag) = 80/260, about 31%."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.2, 0.2, 0.2, 0.2))
plot(NA, xlim = c(0, 10), ylim = c(0.3, 9.7), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")

draw_node <- function(x, y, top, bottom, fill = "white",
                      border = "#17324D", lwd = 1.4, w = 1.05, h = 0.78) {
  rect(x - w, y - h, x + w, y + h, col = fill, border = border, lwd = lwd)
  text(x, y + 0.26, top, font = 2, cex = 1.0, col = "#17324D")
  text(x, y - 0.30, bottom, cex = 0.8, col = "#5B6976")
}
draw_branch <- function(x1, y1, x2, y2, label, above = TRUE) {
  segments(x1, y1, x2, y2, col = "#9AA7B2", lwd = 1.6)
  text((x1 + x2) / 2, (y1 + y2) / 2 + if (above) 0.32 else -0.32,
       label, cex = 0.76, col = "#5B6976")
}

# branches first, then boxes on top
draw_branch(2.45, 5.35, 3.55, 7.15, "10% deal")
draw_branch(2.45, 4.65, 3.55, 2.85, "90% no deal", above = FALSE)
draw_branch(5.65, 7.75, 6.95, 8.45, "80% flagged")
draw_branch(5.65, 7.05, 6.95, 6.35, "20% not", above = FALSE)
draw_branch(5.65, 2.95, 6.95, 3.65, "20% flagged")
draw_branch(5.65, 2.25, 6.95, 1.55, "80% not", above = FALSE)

draw_node(1.35, 5.0, "1,000", "pitches", w = 1.15)
draw_node(4.6, 7.4, "100", "deal")
draw_node(4.6, 2.6, "900", "no deal")
draw_node(8.0, 8.6, "80", "flagged", fill = "#FDE5C1",
          border = "#D55E00", lwd = 2.2)
draw_node(8.0, 6.2, "20", "not flagged", fill = "#F4F7F9",
          border = "#9AA7B2")
draw_node(8.0, 3.8, "180", "flagged", fill = "#FDE5C1",
          border = "#D55E00", lwd = 2.2)
draw_node(8.0, 1.4, "720", "not flagged", fill = "#F4F7F9",
          border = "#9AA7B2")

text(9.55, 6.2, "260\nflagged\nin total", cex = 0.82, col = "#D55E00")
par(old_par)
```

Now the formula. The question reverses a condition: the classifier's quality
is stated as $P(\text{flag}\mid\text{deal})$, but the decision-maker needs
$P(\text{deal}\mid\text{flag})$. The multiplication rule gives the numerator
and total probability gives the denominator:

$$P(B_i\mid A)=\frac{P(A\mid B_i)\,P(B_i)}
{\sum_j P(A\mid B_j)\,P(B_j)}.$$

This is **Bayes' rule**. For the classifier, with the deal/no-deal partition:

$$P(\text{deal}\mid\text{flag})
=\frac{0.80\times 0.10}{0.80\times 0.10+0.20\times 0.90}
=\frac{0.08}{0.08+0.18}=\frac{0.08}{0.26}\approx 0.308.$$

Every term is a branch of the tree, divided by 1,000:

| Formula term | Value | Tree count |
|---|---:|---|
| $P(\text{flag}\mid\text{deal})\,P(\text{deal})$ | $0.08$ | the 80 flagged deals |
| $P(\text{flag}\mid\text{no deal})\,P(\text{no deal})$ | $0.18$ | the 180 flagged non-deals |
| the denominator (total probability of a flag) | $0.26$ | all 260 flagged pitches |

The formula is the tree, written compactly. When a Bayes question appears on
the exam, you may draw the tree with a convenient imaginary total and count
--- the two routes always agree.

Bayes' rule also works on the Shark Tank events themselves. Reversing
$P(D\mid L)$ recovers $P(L\mid D)$, and the R check makes the point that
nothing beyond the multiplication rule and total probability is involved:

```{r}
#| label: b04-bayes
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
bayes_late_given_deal <- mean(D[L]) * mean(L) / mean(D)
c(
  from_bayes = bayes_late_given_deal,
  from_table = mean(L[D])
)
```

### Independence of events

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/independence-of-two-events/">03.3 Independence of Two Events</a>.</p>

Conditioning finally makes precise an idea Lecture 2 could only promise:
what it means for one event to carry *no information* about another. Events
$A$ and $B$ (with $P(B)>0$) are **independent** when learning that $B$
occurred does not change the probability of $A$:

$$P(A\mid B)=P(A).$$

Multiplying both sides by $P(B)$ gives the symmetric **product form**, which
also covers zero-probability events and is the usual textbook definition:

$$P(A\cap B)=P(A)\,P(B).$$

::: {#exm-b04-independence}
## Worked example: are D and L independent?

Check either form with the real numbers.

Via conditioning: $P(D\mid L)=505/734\approx 0.688$, while
$P(D)=882/1441\approx 0.612$. Learning that a pitch is late-season *raises*
the deal probability, so $D$ and $L$ are **not independent**.

Via the product: $P(D\cap L)=505/1441\approx 0.350$, while
$P(D)\,P(L)=\dfrac{882}{1441}\cdot\dfrac{734}{1441}\approx 0.612\times
0.509\approx 0.312$. The joint probability exceeds the product, which says
the same thing: deals and late seasons occur together more often than
independence would predict.
:::

```{r}
#| label: b04-independence
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  P_D_given_L  = mean(D[L]),
  P_D          = mean(D),
  P_D_and_L    = mean(D & L),
  P_D_times_P_L = mean(D) * mean(L)
)
```

**Disjoint is not independent --- one minute, one example.** Take the
early/late pair $L^c$ and $L$. They are disjoint: no pitch is in both
periods, so $P(L^c\cap L)=0$. But the product
$P(L^c)\,P(L)\approx 0.491\times 0.509\approx 0.25$ is far from zero, so they
fail the independence check spectacularly. Disjoint events with positive
probability are always *dependent*: observing one tells you the other is
impossible --- maximal information, not none.

### Second application: deal rates by team composition

The dataset classifies pitching teams as `all_men`, `all_women`, `mixed`, or
`unknown`. Each within-category deal rate is a conditional probability:
$P(D\mid \text{category})$, with the category as reference class. The
construction is shown in full --- this is the live-R segment of the session:

```{r}
#| label: b04-team-deal-rates
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
known_gender <- sharks$pitcher_gender_group != "unknown"
gender_labels <- c("all_men", "all_women", "mixed")

gender_deal_rates <- tapply(
  sharks$deal_on_show[known_gender],
  factor(sharks$pitcher_gender_group[known_gender], levels = gender_labels),
  mean
)
gender_deal_rates
```

```{r}
#| label: fig-b04-team-deal-rates
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Conditional deal proportions by the source's team-composition category, among the 1,434 pitches with a known classification. Each bar is a within-category percentage: P(deal | category)."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- sharks$pitcher_gender_group != "unknown"
gender_labels <- c("all_men", "all_women", "mixed")
gender_deal_rates <- tapply(
  sharks$deal_on_show[known_gender],
  factor(sharks$pitcher_gender_group[known_gender], levels = gender_labels),
  mean
)
gender_counts <- table(
  factor(sharks$pitcher_gender_group[known_gender], levels = gender_labels)
)
rates_df <- data.frame(
  team = paste0(c("All-men", "All-women", "Mixed"),
                "\n(n = ", as.integer(gender_counts), ")"),
  deal_rate = as.numeric(gender_deal_rates)
)
rates_df$team <- factor(rates_df$team, levels = rates_df$team)

ggplot(rates_df, aes(x = team, y = deal_rate)) +
  geom_col(fill = "#009E73", alpha = 0.85, colour = "white",
           width = 0.62) +
  geom_text(aes(label = paste0(round(100 * deal_rate, 1), "%")),
            vjust = -0.6, colour = "#17324D", size = 3.6) +
  scale_y_continuous(limits = c(0, 0.75),
                     breaks = seq(0, 0.75, by = 0.15),
                     labels = paste0(seq(0, 75, by = 15), "%"),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Source classification of pitching team",
       y = "Pitches with an on-air deal") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-b04-team-deal-rates as three conditional proportions, each with its
own denominator: the 63.6% bar means that 63.6% of the 385 all-women pitch
records reached an on-air deal. The bars use three different reference
classes, which is exactly why they can be compared --- each is rescaled
within its own group.

::: {.caveat-box}
<span class="label">What a conditional probability can and cannot say</span>

**Reversing the condition does not reverse cause.** Bayes' rule converts
$P(A\mid B)$ into $P(B\mid A)$ --- an accounting identity about reference
classes. It says nothing about which event influences which:
$P(D\mid L) > P(D)$ does not mean late broadcasting causes deals, and a
flagged pitch is not made more deal-worthy by the flag.

**These are selected televised records.** Every conditional proportion in
this lecture describes pitches that producers selected for broadcast. The
team-composition comparison uses the source's presenter classification (not
every company member, and seven records are unclassified) and does not
isolate a causal effect of team composition. The classifier example uses
stated hypothetical rates, not rates estimated from these data. And, as
always, an on-air deal is not a completed investment.
:::

### Answering the driving question

The two sentences claim different things about the same 505 records:
"68.8% of late-season pitches reached a deal" is $P(D\mid L)=505/734$, with
the late seasons as reference class; "57.3% of deal pitches came from late
seasons" is $P(L\mid D)=505/882$, with the deals as reference class. Both are
correct; swapping their numbers is the classic conditioning error. And a
positive flag from the AI deal-predictor tells us less than its 80%
sensitivity suggests: with a 10% base rate and a 20% false-positive rate,
only $80/260\approx 31\%$ of flagged pitches are deals --- the flag lifts the
probability from 10% to 31%, no further.

**Next:** We can now update the chance of one event when another is known. The
natural next step is to watch two numerical measurements at the same time ---
deals and seasons, seasons and equity asks --- because business questions are
usually about relationships, not single variables. Lecture 5 introduces the
joint distribution of a pair of random variables and the first summaries of
co-movement, covariance and correlation, together with what a correlation
number does and does not license us to say.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 60 minutes.</span>

### Retrieval (10 minutes)

1. What is the denominator of $P(A\mid B)$?
2. Why is $P(A\mid B)$ generally different from $P(B\mid A)$?
3. What makes a collection of events a partition?
4. What does the denominator in Bayes' rule calculate?
5. State the two equivalent definitions of independence of events.
6. Can two disjoint events, each with positive probability, be independent?

::: {.callout-note collapse="true"}
## Check your answers

1. $P(B)$, the probability of the conditioning event.
2. They condition on different reference classes and therefore have
   different denominators.
3. The parts are mutually disjoint and their union is $\Omega$.
4. The total probability of the observed evidence, added across all parts of
   the partition.
5. $P(A\mid B)=P(A)$ (learning $B$ does not change the probability of $A$),
   and the product form $P(A\cap B)=P(A)\,P(B)$.
6. No. Their intersection has probability 0 while the product of their
   probabilities is positive, so the product check fails. Observing one makes
   the other impossible.
:::

### From worked to faded: the Bayes tree (15 minutes)

Re-read the worked count tree (@exm-b04-bayes-tree). Then rebuild it with new
numbers: base rate 20%, sensitivity 90%, false-positive rate 30%.

*Step 1 (given).* Start with 1,000 imaginary pitches: 200 deals, 800
non-deals.

*Step 2 (you).* Fill in the four leaves: flagged and unflagged counts in
each branch.

*Step 3 (you).* Collect the flagged pitches and compute
$P(\text{deal}\mid\text{flag})$.

::: {.callout-note collapse="true"}
## Check your answer

Of the 200 deals, 90% are flagged: 180. Of the 800 non-deals, 30% are
flagged: 240. Flagged total $180+240=420$, of which 180 are deals:
$P(\text{deal}\mid\text{flag})=180/420\approx 0.429$. The formula agrees:
$\dfrac{0.90\times 0.20}{0.90\times 0.20+0.30\times 0.80}
=\dfrac{0.18}{0.42}\approx 0.429$.
:::

**Unassisted.** A stricter classifier operates on a population where 5% of
pitches reach a deal; its sensitivity is 90% and its false-positive rate is
10%. What share of flagged pitches are deals? Use either the tree or the
formula.

::: {.callout-note collapse="true"}
## Check your answer

With 1,000 pitches: 50 deals of which 45 flagged; 950 non-deals of which 95
flagged. $P(\text{deal}\mid\text{flag})=45/(45+95)=45/140\approx 0.321$.
Even with 90% sensitivity and a low false-positive rate, the 5% base rate
keeps the share of true flags below one third.
:::

### Practice (15 minutes)

Use the counts table: 330, 229, 377, 505, with totals 559, 882, 707, 734,
and 1,441.

1. Use the early/late partition to verify
   $P(D)=P(D\mid L^c)\,P(L^c)+P(D\mid L)\,P(L)$.
2. Calculate and interpret $P(L\mid D)$. Explain why it differs from
   $P(D\mid L)$.
3. Decide by calculation whether $D$ and $L$ are independent, using the
   product form.
4. The events $L^c$ and $L$ are disjoint. Are they independent? Explain in
   one sentence.
5. Among the 1,434 pitches with a known team-composition classification, 428
   of 661 pitches with women represented (all-women or mixed) and 449 of 773
   all-men pitches reached an on-air deal. Calculate both conditional deal
   proportions and interpret the difference.

::: {.callout-note collapse="true"}
## Check your answers

1. $(377/707)(707/1441)+(505/734)(734/1441)=(377+505)/1441=882/1441
   \approx 0.612$.
2. $P(L\mid D)=505/882\approx 0.573$: among deal records, 57.3% came from
   Seasons 9--16. It differs from $P(D\mid L)=505/734\approx 0.688$ because
   the conditioning event --- and therefore the denominator --- is different.
3. $P(D\cap L)=505/1441\approx 0.350$, while
   $P(D)\,P(L)=(882/1441)(734/1441)\approx 0.312$. The two are not equal, so
   $D$ and $L$ are not independent.
4. No: $P(L^c\cap L)=0$ but $P(L^c)\,P(L)\approx 0.25>0$, so the product
   check fails --- disjoint events with positive probability are always
   dependent.
5. $428/661\approx 0.648$ and $449/773\approx 0.581$, a descriptive
   difference of about 6.7 percentage points. The comparison uses the
   source's presenter classification and selected televised pitches; it does
   not establish a causal effect of team composition.
:::

### Spaced review: one item from Lecture 3 (5 minutes)

A variable $Y$ has PMF $p_Y(0)=0.2$, $p_Y(1)=0.5$, and $p_Y(2)=0.3$.
Calculate $E[Y]$ and say in one sentence what the number means, given that
$Y$ can never equal it exactly... or can it?

::: {.callout-note collapse="true"}
## Check your answer

$E[Y]=0(0.2)+1(0.5)+2(0.3)=1.1$. It is the probability-weighted average of
the possible values --- the long-run average over many repetitions --- and it
need not be a value the variable can take: here $Y$ is never 1.1.
:::

### Worked exam interpretation (10 minutes)

**Question.** A student sees $P(D\mid L)=0.688$ and writes, "68.8% of deal
pitches came from late seasons."

::: {.callout-note collapse="true"}
## Check a model answer

The calculation conditions on $L$, so the denominator is the 734 pitches in
Seasons 9--16. The correct interpretation is that 68.8% of late-season
records reached an on-air deal. The sentence the student wrote describes the
reverse conditional, $P(L\mid D)=505/882\approx 57.3\%$. Neither quantity
measures completed investment or a causal season effect.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Name one signal
it receives about customers or partners (a positive review, a signed letter
of intent, a click on an ad) and write two sentences: first state the
conditional probability the signal makes you want to know, in
$P(\cdot\mid\cdot)$ form with the reference class named; then say what base
rate you would need before trusting the signal, Bayes-style. There is no
wrong answer; the point is connecting today's tools to something you care
about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I want $P(\text{becomes a paying customer} \mid \text{clicked the ad})$, where the reference class is all students shown the ad this month. Before trusting a high click count I would need the base rate of buyers among all viewers --- with a tiny base rate, most clickers still never buy, exactly the count-tree logic."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to build the
natural-frequency tree for your signal with numbers you consider realistic.
::::

Continue to Lecture 5: Joint random variables, covariance, and correlation.

Formula sheet: Lecture 4 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 7--8.

:::
