# Lecture 4: Conditioning, Bayes' rule, and independence {#b04}

```{r}
#| label: b04-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 4, you can</span>

- calculate conditional probabilities and say which group each one refers to
- rebuild an overall probability from the rates of separate groups
- update a probability after new evidence with Bayes' rule
- check whether two events are independent and tell this apart from disjoint
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">An investor who learns that a pitch is from a late season updates their expectations. A bank that sees a fraud warning on a payment does the same. Updating chances after new information is a core probability skill in business. It is also easy to turn the question around by mistake.</p>

Late-season pitches reach deals more often. Two sentences about these pitches
give the numbers 68.8% and 57.3%. What exactly does each sentence claim, and
what does a positive flag from an AI tool that predicts deals really tell us?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch <a href="https://www.youtube.com/watch?v=MPRKc4UPoJk">Conditional
Probabilities</a> and
<a href="https://www.youtube.com/watch?v=kz2tvO_ZAKI">Bayes' Rule</a>. Be
ready to identify the reference class in a conditional probability and to
explain why turning a condition around changes the denominator. All links are
also listed in the <a href="../resources/video-guide.html">External Video
Guide</a>.
The videos explain key ideas in another way and can use a different notation. The lecture material and its notation always come first.
</div>

[Shark Tank data]{.shark-tag} **Warm up.** Lecture 2 built the counts table for deals and
seasons from the Shark Tank file. Build it again. The events $D$ (on-air
deal) and $L$ (late season) are defined as in Lecture 2:

```{r}
#| label: b04-prep-table
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
table(deal = D, late = L)
```

Then compare, in words only:

- "Among pitches in Seasons 9--16, what proportion reached an on-air deal?"
- "Among pitches that reached an on-air deal, what proportion came from
  Seasons 9--16?"

Both questions use the same cell of your table. Decide before class which
part of the table each question looks at.

**Predict.** Answer this question before you come to class. An AI classifier flags (marks) the
pitches that are likely to get a deal. Of the pitches that it checks, only
10% reach a deal. The classifier flags 80% of the actual deals and also 20%
of the non-deals. Of the flagged pitches, what share are deals: near 80%,
near 50%, or near 30%? Choose one answer. You find the answer in the lecture material.

If fractions or event notation are unfamiliar, review
[Mathematics Refresher: sets and event
notation](../resources/math-refresher.html#sets-and-event-notation) and
[fractions and rearranging
equations](../resources/math-refresher.html#fractions-and-rearranging-equations).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section starts from one counts table. On that table it builds four
tools: conditional probability, the law of total probability, Bayes' rule,
and independence of events.

### Start with a founder's question

[Shark Tank data]{.shark-tag} A founder hears that most deals on the show come from the later seasons. The
founder concludes that a pitch in a later season is likely to end in a deal.
Is that the same statement? Here are two sentences about the 1,441 pitches in
the Shark Tank file:

- **Sentence A.** "68.8% of late-season pitches reached an on-air deal."
- **Sentence B.** "57.3% of the pitches that reached a deal came from the
  late seasons."

Both sentences use the same 505 records: the late-season pitches that
reached a deal. But they answer different questions and give different
numbers. This lecture shows what each sentence claims. It also answers a
harder question of the same kind: when an AI classifier flags a pitch as a
likely deal, how much should that flag change what you believe?

### One table, five probabilities

[Shark Tank data]{.shark-tag} The starting point is the counts table from Lecture 2. The code builds the
events $D$ (on-air deal, `deal_on_show`) and $L$ (late season, `season` 9 or
higher) as before. Run it yourself:

```{r}
#| label: b04-counts-table
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
table(deal = D, late = L)
```

Written with row and column totals:

|                    | Early ($L^c$, S1--8) | Late ($L$, S9--16) | Total |
|--------------------|---------------------:|-------------------:|------:|
| **No deal** ($D^c$) | 330 | 229 | **559** |
| **Deal** ($D$)      | 377 | 505 | **882** |
| **Total**           | **707** | **734** | **1441** |

Each of the five quantities below is a fraction of two counts. Before you
read on, find in the table the count that goes on top and the count that
goes underneath.

1. $P(D)$, the probability that a pitch reached a deal.
2. $P(L)$, the probability that a pitch is from the late seasons.
3. $P(D\cap L)$, the probability that a pitch is from the late seasons and
   reached a deal.
4. Among late-season pitches, the share that reached a deal.
5. Among pitches that reached a deal, the share that came from the late
   seasons.

::: {.callout-note collapse="true"}
## Check your denominators

1. $P(D)$: 882 deal records out of all 1,441 records.
2. $P(L)$: 734 late-season records out of all 1,441 records.
3. $P(D\cap L)$: 505 late-season deal records out of all 1,441 records.
4. Among late-season pitches, the share with a deal: 505 late-season deal
   records out of the 734 late-season records.
5. Among deal pitches, the share from the late seasons: 505 late-season deal
   records out of the 882 deal records.

The first three all divide by 1,441, the number of all pitches. The last two
divide by a part of the table. A share within a part of the table is called
a conditional probability. The next section defines it.
:::

### Conditional probability

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/conditional-probabilities/">02.2 Conditional Probabilities</a>.</p>

[Shark Tank data]{.shark-tag} Items 4 and 5 above ask the same kind of question: what is the probability
of one event if you already know that another event happens? You answer
both in two steps. First, keep only the pitches in one group. This group
is called the **reference class**. Second, take the share of them in which
the event happens. For item 4 the reference class is the 734 late-season
pitches, and 505 of them reached a deal. Looking only at a
reference class is called **conditioning** on it.

The symbol for this is a vertical bar, read as "given". Item 4 is written
$P(D\mid L)$, "the probability of $D$ given $L$". Item 5 is written
$P(L\mid D)$. The event after the bar is always the reference class.

::: {.definition-box}
<span class="label">Definition</span>

**Conditional probability.** The probability that event $A$ happens, given
that event $B$ happens. For events $A$ and $B$ with $P(B)>0$,

$$P(A\mid B)=\frac{P(A\cap B)}{P(B)}.$$

In words: keep only the outcomes in $B$, then take the share of them that
are also in $A$.
:::

The formula does the same two steps as the count. For item 4,

$$P(D\mid L)=\frac{P(D\cap L)}{P(L)}
=\frac{505/1441}{734/1441}=\frac{505}{734}.$$

The reference class gives the denominator. Dividing by $P(B)$ rescales
the probabilities inside $B$ so that they add up to 1 again.

For a fixed event $B$, conditional probabilities follow all three rules from
Lecture 2. So every rule that Lecture 2 derived from them also holds inside
the reference class. The condition is that the event after the bar stays the
same in the whole calculation. Two examples follow. The first is
the complement rule, $P(A^c\mid B)=1-P(A\mid B)$. Among late-season pitches,
the share without a deal is

$$P(D^c\mid L)=1-P(D\mid L)=1-\frac{505}{734}=\frac{229}{734}\approx 0.312.$$

The second is the addition rule. For two events $A$ and $C$ and the same
reference class $B$,

$$P(A\cup C\mid B)=P(A\mid B)+P(C\mid B)-P(A\cap C\mid B).$$

Some calculations mix different reference classes, for example $P(A\mid B)$
and $P(A\mid B^c)$. For these you need the law of total probability,
which a later section explains.

::: {#exm-b04-conditional}
## Worked example: what each of the two sentences means

[Shark Tank data]{.shark-tag} **Sentence A asks about $P(D\mid L)$.** Conditioning on $L$ means that you
hide everything except the Late column:

|                    | Early | **Late ($L$)** |
|--------------------|------:|---------------:|
| No deal            | · | **229** |
| Deal               | · | **505** |
| Total              | · | **734** |

Within that one column, 505 of 734 records are deals:

$$P(D\mid L)=\frac{505}{734}\approx 0.688.$$

**Sentence B asks about $P(L\mid D)$.** Conditioning on $D$ means that you
hide everything except the Deal row:

|                    | Early | Late | Total |
|--------------------|------:|-----:|------:|
| No deal            | · | · | · |
| **Deal ($D$)**     | **377** | **505** | **882** |

Within that one row, 505 of 882 records are late-season:

$$P(L\mid D)=\frac{505}{882}\approx 0.573.$$

The numerator is 505 in both calculations: the same cell of the table. The
denominators differ, and so do the questions. Sentence A looks only at the
734 late-season pitches. Sentence B looks only at the 882 deal pitches.
:::

::: {.callout-note collapse="true"}
## Check: which sentence matches which number?

Sentence A ("of late-season pitches, ... reached a deal") conditions on the
season, so it is $P(D\mid L)=505/734\approx 68.8\%$. Sentence B ("of deal
pitches, ... came from late seasons") conditions on the deal, so it is
$P(L\mid D)=505/882\approx 57.3\%$. A report that quotes "68.8% of deal
pitches came from late seasons" has put the right number into the wrong
sentence.
:::

The same two numbers in R. `D[L]` keeps only the late-season pitches, and
`L[D]` keeps only the deal pitches. This is the step that keeps only the
reference class:

```{r}
#| label: b04-conditionals
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  P_deal_given_late = mean(D[L]),
  P_late_given_deal = mean(L[D])
)
```

### The multiplication rule

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-multiplication-rule/">02.6 The Multiplication Rule</a>.</p>

[Shark Tank data]{.shark-tag} The probability of an intersection, $P(A\cap B)$, is also called a **joint
probability**: both events happen together. Multiply both sides of the
definition of conditional probability by $P(B)$. The result is the
**multiplication rule**. It builds a joint probability from a conditional
one:

$$P(A\cap B)=P(A\mid B)\,P(B).$$

The same holds with the roles of $A$ and $B$ exchanged, because $A\cap B$
and $B\cap A$ are the same event:

$$P(A\cap B)=P(A\mid B)\,P(B)=P(B\mid A)\,P(A).$$

This line explains why the two sentences share a numerator and have
different denominators. Both products equal $505/1441$, but each one uses a
different reference class.

::: {.callout-caution collapse="true"}
## Try it: verify the joint probability two ways

[Shark Tank data]{.shark-tag} Try this first with the student next to you. Use the counts table to check
the equation above for $D$ and $L$.

*Step 1 (given).* The direct way: $P(D\cap L)=505/1441\approx 0.350$.

*Step 2 (given).* The first product:
$P(D\mid L)\,P(L)=\dfrac{505}{734}\cdot\dfrac{734}{1441}$. The 734
cancels. Cancel it and confirm that $505/1441$ is left.

*Step 3 (you).* Write the second product, $P(L\mid D)\,P(D)$, as two
fractions from the table. Show that it also equals $505/1441$.

*Step 4 (you).* Confirm all three ways in R.

::: {.callout-note collapse="true"}
## Check your answer

$P(L\mid D)\,P(D)=\dfrac{505}{882}\cdot\dfrac{882}{1441}=\dfrac{505}{1441}
\approx 0.350$. The 882 cancels, just as the 734 did.

```{r}
#| label: b04-multiplication
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  direct_joint            = mean(D & L),
  cond_on_late_times_base = mean(D[L]) * mean(L),
  cond_on_deal_times_base = mean(L[D]) * mean(D)
)
```
:::
:::

### The law of total probability

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/total-probability-theorem/">02.7 Total Probability Theorem</a>.</p>

[Shark Tank data]{.shark-tag} Lecture 2 introduced a partition: events that do not overlap and together
cover every outcome. The early seasons and the late seasons are a partition
of the pitches. You can rebuild an overall probability from the parts of a
partition in three steps. Take the conditional probability of $A$ within
each part. Multiply it by the probability of that part, which gives a joint
probability. Then add the joint probabilities.

::: {.definition-box}
<span class="label">Definition</span>

**Law of total probability.** For a partition $B_1,\ldots,B_k$ with every
$P(B_i)>0$,

$$P(A)=\sum_{i=1}^{k}P(A\mid B_i)\,P(B_i).$$

In words: the overall probability of $A$ is a weighted average of its
probabilities within the groups. The weight of each group is the probability
of that group.
:::

With only two groups, $B$ and its complement $B^c$, the sum has two terms:

$$P(A)=P(A\mid B)\,P(B)+P(A\mid B^c)\,P(B^c).$$

::: {#exm-b04-total}
## Worked example: rebuilding P(D) from the deal rates of the two periods

[Shark Tank data]{.shark-tag} The early and late events $L^c$ and $L$ form a partition, so

$$P(D)=P(D\mid L^c)\,P(L^c)+P(D\mid L)\,P(L)
=\frac{377}{707}\cdot\frac{707}{1441}
+\frac{505}{734}\cdot\frac{734}{1441}
=\frac{377+505}{1441}=\frac{882}{1441}\approx 0.612.$$

Two kinds of number appear here. $377/707\approx 0.533$ and
$505/734\approx 0.688$ are the deal rates within each period.
$707/1441\approx 0.491$ and $734/1441\approx 0.509$ are the weights: the
share of all pitches that falls in each period. The overall deal probability
is a weighted average of the two rates. So it must lie between 0.533 and
0.688, and it is a little closer to the rate of the larger group.
:::

```{r}
#| label: b04-total-probability
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  direct       = mean(D),
  from_periods = mean(D[!L]) * mean(!L) + mean(D[L]) * mean(L)
)
```

@fig-b04-partition-areas shows the same calculation as areas.

```{r}
#| label: fig-b04-partition-areas
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.6
#| fig-cap: "Total probability as areas. The two columns are the early and late seasons, a partition of all pitches. The width of a column is the probability of that period (the weight). The height of the solid region is the deal rate within that period. The two solid areas do not overlap, and their sum is P(D)."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
p_early <- mean(!L)
p_deal_early <- mean(D[!L])
p_deal_late <- mean(D[L])

ggplot() +
  # light backgrounds: the two parts of the partition
  annotate("rect", xmin = 0, xmax = p_early, ymin = 0, ymax = 1,
           fill = "#009E73", alpha = 0.25) +
  annotate("rect", xmin = p_early, xmax = 1, ymin = 0, ymax = 1,
           fill = "#D55E00", alpha = 0.25) +
  # solid areas: deal rate within each part, times the part's width
  annotate("rect", xmin = 0, xmax = p_early, ymin = 0,
           ymax = p_deal_early, fill = "#009E73", alpha = 0.85) +
  annotate("rect", xmin = p_early, xmax = 1, ymin = 0,
           ymax = p_deal_late, fill = "#D55E00", alpha = 0.85) +
  annotate("segment", x = p_early, xend = p_early, y = 0, yend = 1,
           colour = "#17324D", linewidth = 0.8) +
  annotate("rect", xmin = 0, xmax = 1, ymin = 0, ymax = 1,
           fill = NA, colour = "#17324D", linewidth = 0.8) +
  annotate("text", x = p_early / 2, y = 0.28,
           label = "D ~ symbol('\\307') ~ L^c", parse = TRUE,
           colour = "white", size = 4.2) +
  annotate("text", x = p_early + (1 - p_early) / 2, y = 0.36,
           label = "D ~ symbol('\\307') ~ L", parse = TRUE,
           colour = "white", size = 4.2) +
  annotate("text", x = p_early / 2, y = 0.88, label = "L^c",
           parse = TRUE, colour = "#17324D", size = 4.4) +
  annotate("text", x = p_early + (1 - p_early) / 2, y = 0.88,
           label = "L", colour = "#17324D", size = 4.4) +
  annotate("text", x = 0.985, y = 0.955, label = "Omega",
           parse = TRUE, colour = "#17324D", size = 3.8) +
  coord_cartesian(expand = FALSE) +
  theme_void()
```

### Bayes' rule: count first, formula second

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/bayes-rule/">02.8 Bayes' Rule</a>.</p>

[Shark Tank data]{.shark-tag} Return to the prediction question from before class. An AI classifier reads
the transcripts of pitches (the written text of what was said) and flags
(marks) the pitches that it considers likely to reach a deal.
Three numbers describe an invented classifier of this kind:

- the **base rate**: 10% of the pitches that it checks reach a deal.
- the **sensitivity**: among actual deals, 80% get flagged,
  $P(\text{flag}\mid\text{deal})=0.80$.
- the **false-positive rate**: among non-deals, 20% still get flagged,
  $P(\text{flag}\mid\text{no deal})=0.20$.

A pitch has just been flagged. How likely is a deal now: near 80%, 50%,
or 30%? Count first, and use the formula afterwards.

::: {#exm-b04-bayes-tree}
## Worked example: 1,000 imaginary pitches

Imagine 1,000 pitches and apply the three numbers to them one after the
other, as in @fig-b04-bayes-tree.

1. **Split by the base rate.** Of 1,000 pitches, 100 are deals and 900 are
   not.
2. **Flag the deals.** Sensitivity 80%: of the 100 deals, 80 are flagged.
3. **Flag some non-deals too.** False-positive rate 20%: of the 900
   non-deals, 180 are flagged.
4. **Collect the flagged pitches.** In total $80+180=260$ pitches have a flag,
   and only 80 of them are deals:

$$P(\text{deal}\mid\text{flag})=\frac{80}{260}\approx 0.31.$$

The classifier finds 80% of deals, but fewer than a third of the flagged
pitches are deals. The tree shows the reason. The 900 non-deals are a large group, so
even a 20% false-positive rate produces 180 false flags. That is more than
double the 80 true flags. A flag moves the deal probability from 10% to 31%.
This is real evidence, but it is far from certainty.
:::

```{r}
#| label: fig-b04-bayes-tree
#| code-fold: true
#| code-summary: "See the drawing code (optional, not exam material)"
#| fig-width: 8
#| fig-height: 4.6
#| fig-cap: "Tree of counts for 1,000 imaginary pitches. Orange boxes are the flagged pitches: 80 + 180 = 260 flags in total, of which only 80 are deals, so P(deal | flag) = 80/260, about 31%."
old_par <- par(no.readonly = TRUE)
par(mar = c(0.2, 0.2, 0.2, 0.2))
plot(NA, xlim = c(0, 10), ylim = c(0.3, 9.7), axes = FALSE,
     xlab = "", ylab = "", xaxs = "i", yaxs = "i")

draw_node <- function(x, y, top, bottom, fill = "white",
                      border = "#17324D", lwd = 1.4, w = 1.05, h = 0.78) {
  rect(x - w, y - h, x + w, y + h, col = fill, border = border, lwd = lwd)
  text(x, y + 0.26, top, font = 2, cex = 1.0, col = "#17324D")
  text(x, y - 0.30, bottom, cex = 0.8, col = "#5B6976")
}
draw_branch <- function(x1, y1, x2, y2, label, above = TRUE) {
  segments(x1, y1, x2, y2, col = "#9AA7B2", lwd = 1.6)
  text((x1 + x2) / 2, (y1 + y2) / 2 + if (above) 0.32 else -0.32,
       label, cex = 0.76, col = "#5B6976")
}

# branches first, then boxes on top
draw_branch(2.45, 5.35, 3.55, 7.15, "10% deal")
draw_branch(2.45, 4.65, 3.55, 2.85, "90% no deal", above = FALSE)
draw_branch(5.65, 7.75, 6.95, 8.45, "80% flagged")
draw_branch(5.65, 7.05, 6.95, 6.35, "20% not", above = FALSE)
draw_branch(5.65, 2.95, 6.95, 3.65, "20% flagged")
draw_branch(5.65, 2.25, 6.95, 1.55, "80% not", above = FALSE)

draw_node(1.35, 5.0, "1,000", "pitches", w = 1.15)
draw_node(4.6, 7.4, "100", "deal")
draw_node(4.6, 2.6, "900", "no deal")
draw_node(8.0, 8.6, "80", "flagged", fill = "#FDE5C1",
          border = "#D55E00", lwd = 2.2)
draw_node(8.0, 6.2, "20", "not flagged", fill = "#F4F7F9",
          border = "#9AA7B2")
draw_node(8.0, 3.8, "180", "flagged", fill = "#FDE5C1",
          border = "#D55E00", lwd = 2.2)
draw_node(8.0, 1.4, "720", "not flagged", fill = "#F4F7F9",
          border = "#9AA7B2")

text(9.55, 6.2, "260\nflagged\nin total", cex = 0.82, col = "#D55E00")
par(old_par)
```

Now the formula. The question turns a condition around. We describe the
quality of the classifier with $P(\text{flag}\mid\text{deal})$, but the
person who makes the decision needs $P(\text{deal}\mid\text{flag})$. The multiplication rule gives
the numerator, and the law of total probability gives the denominator.

::: {.definition-box}
<span class="label">Definition</span>

**Bayes' rule.** A formula that turns a conditional probability around: from
$P(A\mid B)$ to $P(B\mid A)$. For events $A$ and $B$ with positive
probability,

$$P(B\mid A)=\frac{P(A\mid B)\,P(B)}{P(A)}.$$

When $B_1,\ldots,B_k$ form a partition, you can write out the denominator
$P(A)$ with the law of total probability:

$$P(B_i\mid A)=\frac{P(A\mid B_i)\,P(B_i)}
{\sum_{j=1}^{k} P(A\mid B_j)\,P(B_j)}.$$
:::

For the classifier, the partition is deal and no deal:

$$P(\text{deal}\mid\text{flag})
=\frac{0.80\times 0.10}{0.80\times 0.10+0.20\times 0.90}
=\frac{0.08}{0.08+0.18}=\frac{0.08}{0.26}\approx 0.308.$$

Every term is a branch of the tree, divided by 1,000:

| Formula term | Value | Tree count |
|---|---:|---|
| $P(\text{flag}\mid\text{deal})\,P(\text{deal})$ | $0.08$ | the 80 flagged deals |
| $P(\text{flag}\mid\text{no deal})\,P(\text{no deal})$ | $0.18$ | the 180 flagged non-deals |
| the denominator (total probability of a flag) | $0.26$ | all 260 flagged pitches |

The formula is a short way to write the tree. Drawing a tree and counting is
a good way to solve a Bayes question. Use an imaginary total that is
easy to calculate with, such as 1,000.
The tree and the formula always give the same answer.

Bayes' rule also works on the Shark Tank events. Starting from
$P(D\mid L)$, it gives $P(L\mid D)$:

$$P(L\mid D)=\frac{P(D\mid L)\,P(L)}{P(D)}
=\frac{(505/734)(734/1441)}{882/1441}=\frac{505}{882}\approx 0.573.$$

The R check computes the same formula and compares it with the direct count
from the table:

```{r}
#| label: b04-bayes
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
bayes_late_given_deal <- mean(D[L]) * mean(L) / mean(D)
c(
  from_bayes = bayes_late_given_deal,
  from_table = mean(L[D])
)
```

### Independence of events

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/independence-of-two-events/">03.3 Independence of Two Events</a>.</p>

[Shark Tank data]{.shark-tag} Sometimes knowing that one event happened tells you nothing about another
event. Lecture 2 mentioned this idea. Conditional probability makes it
precise.

::: {.definition-box}
<span class="label">Definition</span>

**Independent events.** Events $A$ and $B$ are independent when learning that
$B$ happened does not change the probability of $A$. For $P(B)>0$,

$$P(A\mid B)=P(A).$$

The same condition in **product form**:

$$P(A\cap B)=P(A)\,P(B).$$
:::

The product form follows when you multiply both sides of the first equation
by $P(B)$. It treats $A$ and $B$ in the same way, and it also works when an
event has probability zero. Events that are not independent are called
dependent.

::: {#exm-b04-independence}
## Worked example: are D and L independent?

Check both forms with the numbers from the counts table.

With conditioning: $P(D\mid L)=505/734\approx 0.688$, while
$P(D)=882/1441\approx 0.612$. Learning that a pitch is late-season raises
the deal probability, so $D$ and $L$ are not independent.

With the product form: $P(D\cap L)=505/1441\approx 0.350$, while
$P(D)\,P(L)=\dfrac{882}{1441}\cdot\dfrac{734}{1441}\approx 0.612\times
0.509\approx 0.312$. The joint probability is larger than the product. This
says the same thing: deals and late seasons occur together more often than
they would under independence.
:::

```{r}
#| label: b04-independence
D <- sharks$deal_on_show == 1
L <- sharks$season >= 9
c(
  P_D_given_L  = mean(D[L]),
  P_D          = mean(D),
  P_D_and_L    = mean(D & L),
  P_D_times_P_L = mean(D) * mean(L)
)
```

**Disjoint is not independent.** Take the early and late events $L^c$ and
$L$. They are disjoint: no pitch is in both periods, so $P(L^c\cap L)=0$.
But the product $P(L^c)\,P(L)\approx 0.491\times 0.509\approx 0.25$ is not
zero, so the product check fails. Disjoint events with positive probability
are always dependent. If you observe one of them, you know that the other
did not happen. So one event tells you a lot about the other.

### Second application: deal rates by team composition

[Shark Tank data]{.shark-tag} The variable `pitcher_gender_group` in the Shark Tank file puts each team
that pitches into one of four groups: `all_men`, `all_women`, `mixed`, or
`unknown`. The deal rate
within a category is a conditional probability, $P(D\mid \text{category})$,
with the category as reference class. The code computes the three rates for
the pitches with a known category:

```{r}
#| label: b04-team-deal-rates
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1
known_gender <- sharks$pitcher_gender_group != "unknown"
gender_labels <- c("all_men", "all_women", "mixed")

gender_deal_rates <- tapply(
  sharks$deal_on_show[known_gender],
  factor(sharks$pitcher_gender_group[known_gender], levels = gender_labels),
  mean
)
gender_deal_rates
```

```{r}
#| label: fig-b04-team-deal-rates
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Share of pitches with an on-air deal by team category (`pitcher_gender_group`), among the 1,434 Shark Tank pitches with a known category. Each bar is a percentage within its own category: P(deal | category). The three rates lie between 58% and 67%."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
known_gender <- sharks$pitcher_gender_group != "unknown"
gender_labels <- c("all_men", "all_women", "mixed")
gender_deal_rates <- tapply(
  sharks$deal_on_show[known_gender],
  factor(sharks$pitcher_gender_group[known_gender], levels = gender_labels),
  mean
)
gender_counts <- table(
  factor(sharks$pitcher_gender_group[known_gender], levels = gender_labels)
)
rates_df <- data.frame(
  team = paste0(c("All-men", "All-women", "Mixed"),
                "\n(n = ", as.integer(gender_counts), ")"),
  deal_rate = as.numeric(gender_deal_rates)
)
rates_df$team <- factor(rates_df$team, levels = rates_df$team)

ggplot(rates_df, aes(x = team, y = deal_rate)) +
  geom_col(fill = "#009E73", alpha = 0.85, colour = "white",
           width = 0.62) +
  geom_text(aes(label = paste0(round(100 * deal_rate, 1), "%")),
            vjust = -0.6, colour = "#17324D", size = 3.6) +
  scale_y_continuous(limits = c(0, 0.75),
                     breaks = seq(0, 0.75, by = 0.15),
                     labels = paste0(seq(0, 75, by = 15), "%"),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Source classification of pitching team",
       y = "Pitches with an on-air deal") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

@fig-b04-team-deal-rates shows three conditional proportions, each with its
own denominator. The 63.6% bar means that 63.6% of the 385 pitches by
all-women teams reached an on-air deal. The three bars use three different
reference classes. They can still be compared, because each rate is a share
within its own group.

::: {.caveat-box}
<span class="label">What a conditional probability can and cannot say</span>

- Turning a condition around changes the number. Neither number shows a
  cause. 68.8% of late-season pitches got a deal. That does not mean that late
  seasons cause deals.
- The team comparison uses the broad groups of presenters that the data
  source defined. It also uses only pitches that the producers selected.
- The classifier numbers are invented for the example.

**What you can do with it:** before you act on a signal, ask how often the
outcome that you look for happens. With a low base rate, most signals are
false alarms, so a cheap second check is worth its cost.
:::

### Answering the driving question

The two sentences make different claims about the same 505 records.
"68.8% of late-season pitches reached a deal" is $P(D\mid L)=505/734$, with
the late seasons as reference class. "57.3% of deal pitches came from late
seasons" is $P(L\mid D)=505/882$, with the deals as reference class. Both are
correct. The typical error is to use the number of one sentence in the
other. A positive flag from the AI classifier tells us less than its 80%
sensitivity suggests. With a 10% base rate and a 20% false-positive rate,
only $80/260\approx 31\%$ of flagged pitches are deals. The flag raises the
probability of a deal from 10% to 31%, and not more.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your venture sells boxes of prepared meals by subscription, and you give away free trial boxes. One in ten trial customers becomes a subscriber. You notice a signal: 80% of later subscribers open your follow-up email, and 20% of the others open it too. Count with 1,000 trial customers. 100 become subscribers, and 80 of them open the email. Of the other 900, 180 open it. So 80 of 260 openers become subscribers, about 31%. A phone call costs you 2 euros of time, and a subscriber is worth 30 euros. A call to an opener is worth 0.31 × 30 = 9.30 euros on average, so you call the openers. Among the 740 who did not open, only 20 become subscribers, under 3%. A call to one of them is worth about 0.80 euros on average. The decision: call the openers and nobody else.
:::

**Next:** You can now update the probability of one event when another
event is known. Lecture 5 looks at two random variables at the same time,
for example the season of a pitch and the equity offered. It introduces two
numbers that measure how two variables change together, the covariance and the
correlation. It also shows what a correlation can and cannot tell you.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the most important parts. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions

1. What is the denominator of $P(A\mid B)$?
2. Why is $P(A\mid B)$ generally different from $P(B\mid A)$?
3. What makes a collection of events a partition?
4. What does the denominator in Bayes' rule calculate?
5. State the two equivalent definitions of independence of events.
6. Can two disjoint events, each with positive probability, be independent?

::: {.callout-note collapse="true"}
## Check your answers

1. $P(B)$, the probability of the conditioning event.
2. They condition on different reference classes, so they have different
   denominators.
3. No two parts overlap, and together they make up $\Omega$ (their union is
   $\Omega$).
4. The total probability of the observed evidence, added across all parts of
   the partition.
5. $P(A\mid B)=P(A)$ (learning $B$ does not change the probability of $A$),
   and the product form $P(A\cap B)=P(A)\,P(B)$.
6. No. Their intersection has probability 0 while the product of their
   probabilities is positive, so the product check fails. Observing one makes
   the other impossible.
:::

### Finish a half-worked example: the Bayes tree

Re-read the worked example with the tree of 1,000 pitches
(@exm-b04-bayes-tree). Then build the tree again with new numbers: base rate
20%, sensitivity 90%, false-positive rate 30%.

*Step 1 (given).* Start with 1,000 imaginary pitches: 200 deals, 800
non-deals.

*Step 2 (you).* Fill in the four boxes at the end of the tree: the flagged
and the unflagged counts in each branch.

*Step 3 (you).* Collect the flagged pitches and compute
$P(\text{deal}\mid\text{flag})$.

::: {.callout-note collapse="true"}
## Check your answer

Of the 200 deals, 90% are flagged: 180. Of the 800 non-deals, 30% are
flagged: 240. Flagged total $180+240=420$, of which 180 are deals:
$P(\text{deal}\mid\text{flag})=180/420\approx 0.429$. The formula agrees:
$\dfrac{0.90\times 0.20}{0.90\times 0.20+0.30\times 0.80}
=\dfrac{0.18}{0.42}\approx 0.429$.
:::

**On your own.** A stricter classifier checks a group of pitches in which 5%
reach a deal. Its sensitivity is 90% and its false-positive rate is 10%.
What share of flagged pitches are deals? Use either the tree or the formula.

::: {.callout-note collapse="true"}
## Check your answer

With 1,000 pitches: 50 deals, of which 45 are flagged, and 950 non-deals, of
which 95 are flagged. $P(\text{deal}\mid\text{flag})=45/(45+95)=45/140\approx 0.321$.
Even with 90% sensitivity and a low false-positive rate, the 5% base rate
keeps the share of true flags below one third.
:::

### Practice

[Shark Tank data]{.shark-tag} Use the counts table: 330, 229, 377, 505, with totals 559, 882, 707, 734,
and 1,441.

1. Use the early/late partition to verify
   $P(D)=P(D\mid L^c)\,P(L^c)+P(D\mid L)\,P(L)$.
2. Calculate and interpret $P(L\mid D)$. Explain why it differs from
   $P(D\mid L)$.
3. Decide by calculation whether $D$ and $L$ are independent, using the
   product form.
4. The events $L^c$ and $L$ are disjoint. Are they independent? Explain in
   one sentence.
5. Consider the 1,434 pitches with a known classification of the team. Of
   the 661 pitches with women represented (all-women or mixed), 428 reached
   an on-air deal. Of the 773 all-men pitches, 449 reached an on-air deal.
   Calculate both conditional deal proportions and interpret the difference.

::: {.callout-note collapse="true"}
## Check your answers

1. $(377/707)(707/1441)+(505/734)(734/1441)=(377+505)/1441=882/1441
   \approx 0.612$.
2. $P(L\mid D)=505/882\approx 0.573$: among deal records, 57.3% came from
   Seasons 9--16. It differs from $P(D\mid L)=505/734\approx 0.688$ because
   the conditioning event, and so the denominator, is different.
3. $P(D\cap L)=505/1441\approx 0.350$, while
   $P(D)\,P(L)=(882/1441)(734/1441)\approx 0.312$. The two are not equal, so
   $D$ and $L$ are not independent.
4. No: $P(L^c\cap L)=0$ but $P(L^c)\,P(L)\approx 0.25>0$, so the product
   check fails. Disjoint events with positive probability are always
   dependent.
5. $428/661\approx 0.648$ and $449/773\approx 0.581$, a descriptive
   difference of about 6.7 percentage points. The data source defined the
   groups, and the producers chose the pitches for television. The difference does
   not show that the kind of team causes the deal.
:::

### Review: one question from Lecture 3

A variable $Y$ has PMF $p_Y(0)=0.2$, $p_Y(1)=0.5$, and $p_Y(2)=0.3$.
Calculate $E[Y]$ and say in one sentence what the number means. Can
$Y$ ever equal it exactly?

::: {.callout-note collapse="true"}
## Check your answer

$E[Y]=0(0.2)+1(0.5)+2(0.3)=1.1$. It is the probability-weighted average of
the possible values, and the long-run average over many repetitions. It does
not have to be a value that the variable can take: here $Y$ is never 1.1.
:::

### Worked exam interpretation

[Shark Tank data]{.shark-tag} **Question.** A student sees $P(D\mid L)=0.688$ and writes, "68.8% of deal
pitches came from late seasons." Explain what is wrong and give the correct
interpretation.

::: {.callout-note collapse="true"}
## Check a model answer

The calculation conditions on $L$, so the denominator is the 734 pitches in
Seasons 9--16. The correct interpretation is that 68.8% of late-season
records reached an on-air deal. The sentence the student wrote describes the
reversed conditional probability, $P(L\mid D)=505/882\approx 57.3\%$. Neither quantity
measures completed investment or a causal season effect.
:::

### Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant: "If a pitch got a deal, how likely is it that it came from a late season?" It answered:

> "68.8% of late-season pitches got a deal. So if a pitch got a deal, there is a 68.8% chance that it came from a late season."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

The AI turned the condition around. 68.8% is the share of deals among late-season pitches, 505 of 734. The question asks for the share of late-season pitches among deals: 505 of 882, about 57.3%. Always ask which group is in the denominator.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Name one signal
it receives about customers or partners, for example a positive review, a
signed letter of intent, or a click on an ad. Sentence 1: which conditional
probability does the signal make you want to know? Write it with the
vertical bar and name the reference class. Sentence 2: which base rate would
you need to know before you trust the signal? There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells boxes of prepared meals on campus. "I want $P(\text{becomes a paying customer} \mid \text{clicked the ad})$, where the reference class is the students who clicked the ad this month. Before I trust a high number of clicks, I would need the base rate of buyers among all students who saw the ad. With a very low base rate, most students who click still never buy, as the tree of 1,000 cases shows."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to build the tree of 1,000 cases
for your signal with numbers you consider realistic.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 4. Show the solution only after I send my answer."

[Continue to Lecture 5: Joint random variables, covariance, and correlation](b05-joint-covariance-correlation.qmd).

[Formula sheet: Lecture 4 formulas](formula-sheet.qmd#formula-l4).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 7--8.

:::
