# Lecture 5: Joint random variables, covariance, and correlation {#b05}

```{r}
#| label: b05-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 5, you can</span>

- define a pair of random variables on one sample space;
- read a joint PMF and recover marginal and conditional distributions;
- check independence of two random variables with the product rule in every
  cell of the joint table;
- derive covariance from signed products of deviations and compute it two
  ways;
- explain how correlation standardises covariance;
- separate linear association from independence, causality, and importance.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Business questions are rarely about one variable alone: marketing and sales, price and demand, timing and success. The first analytical step is measuring whether two quantities move together at all, and how strongly.</p>

Deals and late seasons, equity offers and deals, seasons and equity offers.
Which pairs move together, how strongly, and what does a correlation number
license us to say?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch
<a href="https://www.youtube.com/watch?v=7nu97OYx4X4">Joint PMFs and the
Expected Value Rule</a>,
<a href="https://www.youtube.com/watch?v=K2Tlj27nkjs">Covariance</a>, and
<a href="https://www.youtube.com/watch?v=HTs6Zhc2S1M">The Correlation
Coefficient</a>. Be ready to state what a joint PMF records, what covariance
averages, and how correlation standardises covariance. All links are also
listed in the <a href="../resources/video-guide.html">External Video
Guide</a>.
</div>

**Reactivate (5 minutes).** From earlier lectures you know the deal
indicator $X$ (1 for an on-air deal, 0 otherwise) and the late-period
indicator $Y$ (1 for Seasons 9--16, 0 for Seasons 1--8). One selected pitch
record now produces *two* numbers at once: the ordered pair $(X,Y)$. Write
down all possible values of the pair.

::: {.callout-note collapse="true"}
## Check your answer

Four ordered pairs: $(0,0)$, $(0,1)$, $(1,0)$, $(1,1)$ --- early no-deal,
late no-deal, early deal, late deal. Order matters: $(1,0)$ and $(0,1)$
describe different pitches.
:::

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: is the correlation between season and the
equity percentage entrepreneurs initially offer **positive**, **near
zero**, or **negative**? Commit to an answer; the lecture resolves it.

If ordered pairs, products, or standardisation are unfamiliar, review
[Mathematics Refresher: products, indices, and
z-scores](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question --- three pairs of variables on one slide:
deal and late period, equity offered and deal, season and equity offered.
Every pair raises the same two questions: do the variables move together,
and how strongly? Until now each random variable had its own distribution,
studied alone. Two separate marginal distributions, however, cannot say how
values are *paired*. For that we need the joint distribution.

### The joint probability mass function

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/joint-pmfs-and-the-expected-value-rule/">06.7 Joint PMFs and the Expected Value Rule</a>.</p>

For discrete random variables $X$ and $Y$ defined on the same experiment,
the **joint PMF** records the probability of every possible pair of values:

$$p_{X,Y}(x,y)=P(X=x,\;Y=y).$$

Like any PMF it must be non-negative, and its entries must sum to 1 over
every possible pair --- each outcome lands in exactly one cell.

The live construction, with $X$ and $Y$ built in full view:

```{r}
#| label: b05-joint-pmf
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
X <- sharks$deal_on_show                # deal indicator: 1 = on-air deal
Y <- as.integer(sharks$season >= 9)     # late-period indicator: 1 = S9-16
joint_counts <- table(Deal = X, Late_period = Y)
joint_pmf <- prop.table(joint_counts)
joint_counts
round(joint_pmf, 3)
sum(joint_pmf)
```

As a table with the marginal sums written in:

| $p_{X,Y}(x,y)$ | $Y=0$ (early) | $Y=1$ (late) | row sum: $p_X(x)$ |
|---|---:|---:|---:|
| $X=0$ (no deal) | 0.229 | 0.159 | **0.388** |
| $X=1$ (deal)    | 0.262 | 0.350 | **0.612** |
| column sum: $p_Y(y)$ | **0.491** | **0.509** | 1 |

::: {.callout-note collapse="true"}
## Check: which entry is P(X = 1, Y = 1)?

The bottom-right cell of the joint table, 0.350: the probability that the
selected pitch reached a deal *and* came from a late season,
$505/1441\approx 0.350$. It is one number for one pair of values --- not a
row, not a column.
:::

#### Recovering marginal PMFs

Each single-variable distribution is recovered by summing the joint PMF over
the values of the other variable:

$$p_X(x)=\sum_y p_{X,Y}(x,y),
\qquad
p_Y(y)=\sum_x p_{X,Y}(x,y).$$

Work it by hand on the table first --- the arrows drawn in class point along
a row and down a column: $p_X(1)=0.262+0.350=0.612$ (sum the $X=1$ row) and
$p_Y(1)=0.159+0.350=0.509$ (sum the $Y=1$ column). These are exactly $P(D)$
and $P(L)$ from earlier lectures, now living in the margins of the joint
table --- which is why single-variable distributions are called *marginal*.

```{r}
#| label: b05-marginals
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
joint_pmf <- prop.table(table(Deal = X, Late_period = Y))
rowSums(joint_pmf)   # marginal PMF of X
colSums(joint_pmf)   # marginal PMF of Y
```

The joint table determines the marginals, but not the other way round: many
different joint tables share the same margins. What the joint table adds is
the *pairing*.

#### Conditional PMFs are rescaled columns

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/conditional-pmfs/">07.2 Conditional PMFs</a>.</p>

Lecture 4's conditioning idea transfers cell by cell. For $p_Y(y)>0$,

$$p_{X\mid Y}(x\mid y)=\frac{p_{X,Y}(x,y)}{p_Y(y)}.$$

::: {#exm-b05-conditional}
## Worked example: one conditional entry, by hand

What is $p_{X\mid Y}(1\mid 1)$ --- the deal probability given a late-season
pitch? Cover everything except the $Y=1$ column and rescale it by its own
total:

| $p_{X,Y}(x,y)$ | $Y=0$ | **$Y=1$** |
|---|---:|---:|
| $X=0$ | · | **0.159** |
| $X=1$ | · | **0.350** |
| column sum | · | **0.509** |

$$p_{X\mid Y}(1\mid 1)=\frac{0.350}{0.509}
=\frac{505/1441}{734/1441}=\frac{505}{734}\approx 0.688.$$

This is precisely Lecture 4's $P(D\mid L)$: the 505-over-734 calculation,
now read off a joint PMF. Dividing the whole column by 0.509 makes its
entries sum to 1 --- a rescaled column is a complete conditional
distribution of $X$.
:::

```{r}
#| label: b05-conditional-pmf
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
joint_counts <- table(Deal = X, Late_period = Y)
round(prop.table(joint_counts, margin = 2), 3)   # each column sums to 1
```

### Independence of random variables

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/independence-of-random-variables/">07.4 Independence of Random Variables</a>.</p>

Lecture 4 defined independence for a *pair of events*: $P(A\cap
B)=P(A)\,P(B)$. Random variables generate many events at once --- one for
every value they can take --- so independence of random variables demands
more. Discrete $X$ and $Y$ are **independent** when the product rule holds
in *every* cell of the joint table:

$$p_{X,Y}(x,y)=p_X(x)\,p_Y(y)
\quad\text{for every pair } (x,y).$$

One mismatch anywhere proves dependence; establishing independence requires
checking every cell.

::: {#exm-b05-independence}
## Worked example: the (1,1) cell check

If deal and late period were independent, the bottom-right cell would equal
the product of its margins:

$$p_X(1)\,p_Y(1)=0.612\times 0.509\approx 0.312.$$

The actual entry is $p_{X,Y}(1,1)=0.350$. The mismatch --- 0.350 observed
against 0.312 required --- proves that $X$ and $Y$ are **not independent**:
deals and late seasons are paired more often than independent variables
would allow. This is the joint-PMF version of Lecture 4's finding that
$P(D\mid L)\ne P(D)$.
:::

```{r}
#| label: b05-joint-independence
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
observed_joint_11 <- mean(X == 1 & Y == 1)
independent_joint_11 <- mean(X == 1) * mean(Y == 1)
c(observed_joint_11 = observed_joint_11,
  independent_joint_11 = independent_joint_11)
```

Had the $(1,1)$ cell matched, the conclusion would still be open: one
matching cell is not enough for independence, but one mismatching cell is
enough for dependence.

### Covariance comes from expectation

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/covariance/">12.5 Covariance</a>.</p>

Independence versus dependence is a yes/no verdict. The next tool grades the
dependence: in which *direction* do the variables move together, and how
strongly? Lecture 3 measured spread as an expected squared deviation,
$\operatorname{Var}(X)=E[(X-\mu_X)^2]$. **Covariance** applies the same idea
to a pair. With $\mu_X=E[X]$ and $\mu_Y=E[Y]$,

$$\operatorname{Cov}(X,Y)=E[(X-\mu_X)(Y-\mu_Y)].$$

The object inside the expectation is a *signed product* of deviations, and
the sign logic comes before any formula manipulation:

- both values above their means, or both below: the product is **positive**;
- one above and one below: the product is **negative**.

Covariance averages those signed products over the joint distribution. A
positive covariance says above-average values of one variable tend to occur
with above-average values of the other; a negative covariance says they tend
to occur with below-average values.

::: {#exm-b05-covariance}
## Worked example: four signed products, one at a time

Take this small joint PMF (it returns in the practice sets):

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.30 | 0.20 |
| $X=1$ | 0.10 | 0.40 |

The marginals are $p_X(1)=0.10+0.40=0.50$ and $p_Y(1)=0.20+0.40=0.60$, so
$\mu_X=E[X]=0.5$ and $\mu_Y=E[Y]=0.6$.

Write out the signed product of deviations for each cell, one by one:

| Cell $(x,y)$ | $(x-0.5)(y-0.6)$ | Sign | Probability | Contribution |
|---|---:|:---:|---:|---:|
| $(0,0)$ | $(-0.5)(-0.6)=+0.30$ | both below | 0.30 | $+0.090$ |
| $(0,1)$ | $(-0.5)(+0.4)=-0.20$ | mixed | 0.20 | $-0.040$ |
| $(1,0)$ | $(+0.5)(-0.6)=-0.30$ | mixed | 0.10 | $-0.030$ |
| $(1,1)$ | $(+0.5)(+0.4)=+0.20$ | both above | 0.40 | $+0.080$ |

Average with the joint probabilities as weights:

$$\operatorname{Cov}(X,Y)=0.090-0.040-0.030+0.080=0.10.$$

The heavy cells are the two "agreeing" corners $(0,0)$ and $(1,1)$, so the
positive contributions dominate: this pair moves together.

**Verify with the computational form.** Expanding the product gives the
identity

$$\operatorname{Cov}(X,Y)=E[XY]-E[X]\,E[Y].$$

Here $XY$ is nonzero only in the $(1,1)$ cell, so $E[XY]=1\times 1\times
0.40=0.40$, and

$$\operatorname{Cov}(X,Y)=0.40-(0.5)(0.6)=0.40-0.30=0.10.$$

The same answer by both routes. The definition explains the sign; the
computational form is quicker on an exam.
:::

The same two routes on the Shark Tank pair, in visible code:

```{r}
#| label: b05-covariance-from-expectation
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
E_X <- mean(X)
E_Y <- mean(Y)
E_XY <- mean(X * Y)

c(
  E_X = E_X,
  E_Y = E_Y,
  E_XY = E_XY,
  covariance_from_moments = E_XY - E_X * E_Y,
  covariance_from_deviations = mean((X - E_X) * (Y - E_Y))
)
```

Covariance has **product units**. If $X$ is measured in dollars and $Y$ in
years, $\operatorname{Cov}(X,Y)$ is in dollar-years. Its sign is
interpretable; its magnitude changes whenever either measurement unit
changes, which makes raw covariances hard to compare.

#### Independence implies zero covariance

If $X$ and $Y$ are independent, then $E[XY]=E[X]\,E[Y]$, so
$\operatorname{Cov}(X,Y)=0$. The reverse is not generally true: zero
covariance rules out linear co-movement, not every form of dependence.

### Correlation standardises covariance

<p class="concept-video"><strong>MIT explanations (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-correlation-coefficient/">12.8 The Correlation Coefficient</a> and
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/interpreting-the-correlation-coefficient/">12.10 Interpreting the Correlation Coefficient</a>.</p>

The unit problem has a standard fix: divide by the two standard deviations.
For variables whose standard deviations are not zero,

$$\rho_{XY}
=\frac{\operatorname{Cov}(X,Y)}
{\operatorname{SD}(X)\operatorname{SD}(Y)},
\qquad -1\le\rho_{XY}\le1.$$

Track the units: the numerator is in dollar-years, and the denominator is
dollars times years --- also dollar-years. They cancel, so **correlation has
no units** and does not change when either variable is rescaled. Its sign
gives the direction of linear association; its magnitude, between 0 and 1,
gives the strength of linear association.

```{r}
#| label: b05-correlation-from-covariance
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
cov_xy <- mean((X - mean(X)) * (Y - mean(Y)))
rho_xy <- cov_xy / (sqrt(mean((X - mean(X))^2)) *
                    sqrt(mean((Y - mean(Y))^2)))
c(from_definition = rho_xy,
  from_R = cor(X, Y))
```

The word *linear* carries real weight. @fig-b05-linear-curved shows two
patterns of the same size.

```{r}
#| label: fig-b05-linear-curved
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.4
#| fig-cap: "Correlation measures linear association only. Pattern A rises almost linearly and has r close to 1. Pattern B is strongly related but U-shaped: its fitted line is flat and its correlation is close to 0."
library(ggplot2)
x_association <- seq(-2, 2, length.out = 61)
y_linear <- as.numeric(scale(0.8 * x_association +
                             0.25 * sin(7 * x_association)))
y_curved <- as.numeric(scale(x_association^2))

assoc_df <- data.frame(
  x = rep(x_association, times = 2),
  y = c(y_linear, y_curved),
  pattern = rep(c("Pattern A", "Pattern B"), each = 61)
)
r_labels <- data.frame(
  pattern = c("Pattern A", "Pattern B"),
  label = paste0("r = ", sprintf("%.2f", c(
    cor(x_association, y_linear),
    cor(x_association, y_curved)
  ))),
  x = -2,
  y = max(assoc_df$y) + 0.45
)

ggplot(assoc_df, aes(x = x, y = y)) +
  geom_point(colour = "#009E73", alpha = 0.72, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "#D55E00", linewidth = 0.9) +
  geom_text(data = r_labels, aes(x = x, y = y, label = label),
            hjust = 0, colour = "#17324D", size = 3.4) +
  facet_wrap(~ pattern) +
  labs(x = "X", y = "Y (standardised units)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Each point is one $(x,y)$ pair; the orange line is the fitted linear
summary. Pattern A follows an almost straight rising path, so its
correlation is close to 1. Pattern B has a clear U-shaped relationship, but
its fitted line is flat and its correlation is close to 0.

::: {.callout-note collapse="true"}
## Check: which panel has correlation near zero --- and why is that not "no relationship"?

Pattern B. Its correlation is near zero because rises on the right cancel
falls on the left in the signed products --- there is no *linear* trend. The
relationship itself is strong: knowing $x$ pins down $y$ almost exactly. A
near-zero correlation licenses only "no linear association," never "no
relationship."
:::

#### Sample covariance and sample correlation

For observed pairs $(x_i,y_i)$, the usual sample covariance is

$$s_{XY}=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y}),$$

and the sample correlation is

$$r_{XY}=\frac{s_{XY}}{s_X s_Y}.$$

These are sample statistics, the two-variable analogues of $s^2$ and $s$
from Lecture 3. Under a model and sampling design, they may be used to
estimate the population quantities $\operatorname{Cov}(X,Y)$ and
$\rho_{XY}$.

::: {.optional-box}
<span class="label">Optional: point-biserial correlation</span><br>
When one of the two variables is a 0/1 indicator --- offered equity paired
with the deal indicator, say --- the ordinary Pearson correlation computed on
the pair also goes by the name *point-biserial correlation*. It is not a new
formula, only a special-case name you may meet in other courses; `cor()`
computes it the same way.
:::

### Inspect first, summarise second

A correlation compresses a whole scatter of pairs into one number, so look
at the scatter before trusting the number. For the season--equity pair
(treating season as an equally spaced time index), @fig-b05-season-equity
shows every pitch.

```{r}
#| label: fig-b05-season-equity
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Season and initially offered equity for all 1,441 pitches. Horizontal jitter only reveals overlapping points. The orange line is the fitted linear summary; its downward direction is the negative linear association that the correlation summarises."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)

ggplot(sharks, aes(x = season, y = equity_offered_pct)) +
  geom_point(position = position_jitter(width = 0.14, height = 0,
                                        seed = 2026),
             colour = "#009E73", alpha = 0.45, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "#D55E00", linewidth = 0.9) +
  scale_x_continuous(breaks = c(seq(1, 15, by = 2), 16)) +
  labs(x = "Season", y = "Equity initially offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each point represents one pitch: season is read horizontally and initially
offered equity in percentage points vertically. The downward drift is
visible but noisy --- the large vertical spread shows why the line does not
predict every pitch closely. Now the live correlations, resolving the prep
prediction:

```{r}
#| label: b05-observed-correlations
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
c(
  season_with_equity_offered =
    cor(sharks$season, sharks$equity_offered_pct),
  equity_offered_with_deal =
    cor(sharks$equity_offered_pct, sharks$deal_on_show)
)
```

Season and initially offered equity have a moderate **negative** linear
association, $r\approx -0.42$: entrepreneurs in later seasons tend to offer
smaller ownership stakes. If you predicted "negative" in the prep, the
scatter and the number now back you up --- and if not, the plot shows what
the single number was hiding. Offered equity and the 0/1 deal indicator
have a weak negative association, $r\approx -0.11$.

::: {.caveat-box}
<span class="label">What a correlation can and cannot say</span>

The weak negative equity--deal correlation invites a tempting reading:
"offer less equity and you are more likely to get a deal." Three patterns
coexist in this file: later seasons have higher deal proportions, later
seasons have lower initial equity offers, and deal records have slightly
lower offers on average. The marginal equity--deal association may therefore
partly reflect differences *between seasons* rather than an effect of the
offer itself --- season is entangled with both variables. Offered equity was
not randomly assigned, and the file records selected televised pitches, not
a causal experiment. Separating the equity--deal association *within* a
season from the season differences requires holding season fixed --- exactly
what multiple regression does in Intermediate Lecture 4. Until then:
correlation alone does not establish causation, independence (only its
absence), or practical importance.
:::

### Answering the driving question

Deal and late period move together: the joint entry 0.350 exceeds the
independence benchmark 0.312, and their correlation is a mild positive
$r\approx 0.16$. Season and offered equity move oppositely, and moderately:
$r\approx -0.42$. Offered equity and deals barely move together at all,
$r\approx -0.11$, and the caveat box explains why even that number should
not be read as advice. A correlation licenses a statement about the
direction and strength of *linear* co-movement in these records --- nothing
about cause, and nothing about curved relationships.

**Next:** Tutorial 2 turns Lectures 4 and 5 into exam practice: conditional
probabilities read off tables, a Bayes tree with new numbers, and covariance
and correlation computed and interpreted. Bring your attempted after-class
tasks --- especially the joint table and the signed-product routine, which the
tutorial reuses directly. After the tutorial, Lecture 6 leaves whole numbers
behind: quantities that can land anywhere in a range.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 70 minutes.</span>

### Retrieval (10 minutes)

1. What information does a joint PMF contain that two marginal PMFs do not?
2. How do you recover a marginal PMF from a joint PMF?
3. Why can covariance change when units change?
4. Why does independence imply zero covariance?
5. Why does zero correlation not imply independence?

::: {.callout-note collapse="true"}
## Check your answers

1. It records how values of the two variables occur together --- the
   pairing, not just the two separate distributions.
2. Sum the joint PMF over every value of the other variable.
3. Covariance has product units, so rescaling either variable rescales the
   covariance.
4. Independence gives $E[XY]=E[X]\,E[Y]$, hence zero covariance for the
   variables considered in this course.
5. Correlation measures linear association; dependent variables can have
   zero linear association (the U-shape in @fig-b05-linear-curved).
:::

### From worked to faded: covariance on a new table (20 minutes)

Re-read the signed-products worked example (@exm-b05-covariance). Then
repeat the routine on a new joint PMF:

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.10 | 0.30 |
| $X=1$ | 0.40 | 0.20 |

*Step 1 (given).* The marginals: $p_X(1)=0.60$, $p_Y(1)=0.50$, so
$\mu_X=0.6$ and $\mu_Y=0.5$.

*Step 2 (you).* Write the four signed products $(x-0.6)(y-0.5)$, one per
cell, with their signs.

*Step 3 (you).* Weight each product by its cell probability and add.

*Step 4 (you).* Verify with $E[XY]-E[X]E[Y]$ and interpret the sign.

::: {.callout-note collapse="true"}
## Check your answer

Contributions: $(0,0)$: $(+0.30)(0.10)=+0.030$; $(0,1)$:
$(-0.30)(0.30)=-0.090$; $(1,0)$: $(-0.20)(0.40)=-0.080$; $(1,1)$:
$(+0.20)(0.20)=+0.040$. Sum:
$\operatorname{Cov}(X,Y)=0.030-0.090-0.080+0.040=-0.10$. Check:
$E[XY]=0.20$ and $E[X]E[Y]=0.6\times 0.5=0.30$, so $0.20-0.30=-0.10$. The
negative sign says the disagreeing corners are heavy: high $X$ tends to come
with low $Y$.
:::

**Unassisted practice table.** For the joint PMF

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.25 | 0.25 |
| $X=1$ | 0.15 | 0.35 |

1. Find the marginal PMFs of $X$ and $Y$.
2. Are $X$ and $Y$ independent?
3. Calculate $\operatorname{Cov}(X,Y)$ and interpret its sign.
4. Calculate $P(X=1\mid Y=1)$ and compare it with $P(X=1)$.
5. Define $Z=100X$. Calculate $\operatorname{Cov}(Z,Y)$ and state what
   happens to $\operatorname{Corr}(Z,Y)$.

::: {.callout-note collapse="true"}
## Check your answers

1. $P(X=0)=0.50$, $P(X=1)=0.50$; $P(Y=0)=0.40$, $P(Y=1)=0.60$.
2. No. $P(X=1,Y=1)=0.35$, whereas $P(X=1)P(Y=1)=0.50\times 0.60=0.30$. One
   mismatching cell proves dependence.
3. $E[X]=0.50$, $E[Y]=0.60$, $E[XY]=0.35$, so
   $\operatorname{Cov}(X,Y)=0.35-0.30=0.05$. Positive: the pair $(1,1)$
   occurs with greater probability than independence would give.
4. $P(X=1\mid Y=1)=0.35/0.60=7/12\approx 0.583$, larger than the marginal
   $P(X=1)=0.50$: conditioning on $Y=1$ shifts the distribution of $X$
   upward.
5. $\operatorname{Cov}(Z,Y)=100\operatorname{Cov}(X,Y)=5$. Multiplication by
   a positive constant rescales covariance but leaves correlation unchanged.
:::

### Spaced review: two items from earlier lectures (10 minutes)

**From Lecture 2.** In the record-selection experiment, is "the selected
pitch is the record with pitch_id 7" an outcome or an event? And what kind
of object is the record itself?

::: {.callout-note collapse="true"}
## Check your answer

The record itself is an **outcome** --- one complete possible result. The
description picks out the singleton set containing that record, and a set of
outcomes is an **event**. Same distinction as 4 versus $\{4\}$ for a die.
:::

**From Lecture 4.** A classifier flags likely deals in a population with a
10% deal base rate; sensitivity is 70% and the false-positive rate is 20%.
Using a 1,000-pitch tree, what share of flagged pitches are deals?

::: {.callout-note collapse="true"}
## Check your answer

Of 100 deals, 70 are flagged; of 900 non-deals, 180 are flagged. So
$P(\text{deal}\mid\text{flag})=70/(70+180)=70/250=0.28$.
:::

### Worked exam interpretation (10 minutes)

The dataset contains 1,441 televised pitches from Seasons 1--16. The
variable `season` records the season number, and `equity_offered_pct`
records the percentage of the business initially offered to the sharks.

**Question.** R reports
`cor(sharks$season, sharks$equity_offered_pct) = -0.416`. Interpret this
result in context. State:

1. the direction and approximate strength of the linear association;
2. what the association means for pitches in earlier and later seasons;
3. whether the correlation has units; and
4. one conclusion that cannot be justified from this result.

::: {.callout-note collapse="true"}
## Check a model answer

Among the 1,441 records, season and initially offered equity have a moderate
negative linear association: later-season pitches tend to offer a smaller
ownership percentage initially, although this is not true of every
individual pitch. The correlation is dimensionless: it has no units. It
summarises only linear co-movement and does not show that the passage of
time caused offers to change.
:::

### Exam-style: independence and covariance sign (5 minutes)

For the joint PMF

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.18 | 0.42 |
| $X=1$ | 0.12 | 0.28 |

decide whether $X$ and $Y$ are independent, and state
$\operatorname{Cov}(X,Y)$ without long calculation.

::: {.callout-note collapse="true"}
## Check your answer

Marginals: $P(X=1)=0.40$, $P(Y=1)=0.70$. Check every cell:
$0.60\times 0.30=0.18$, $0.60\times 0.70=0.42$, $0.40\times 0.30=0.12$,
$0.40\times 0.70=0.28$ --- all four match, so $X$ and $Y$ are
**independent**. Independence implies $\operatorname{Cov}(X,Y)=0$ (directly:
$E[XY]=0.28=E[X]E[Y]$). Note the contrast with the earlier tables: there one
mismatch settled dependence; here independence needed all four cells.
:::

### From the reading ladder (10 minutes)

Next rung of the [reading ladder](../resources/reading-ladder.qmd): open
[Tierney & Farmer (2002)](https://doi.org/10.2307/3069429) via the
university library and find the table that reports means, standard
deviations, and correlations among the study variables (in management
journals this is usually the first numbered table). Pick any two variables:
state their reported correlation, interpret its sign and rough strength in
one sentence, and --- as in today's caveat --- name one third variable that
could produce that correlation without either variable driving the other.
Note the sample size the table reports and why it matters for how seriously
to take a small correlation.

::: {.callout-note collapse="true"}
## Check your approach

Whatever pair you chose, the sentence should have the form "employees higher
in A tended also to be higher/lower in B in this sample (r = ...)" --- a
statement about co-movement in one sample, not about causation and not about
any individual. If your third-variable story was plausible (tenure, role,
ability), you have applied the Lecture 5 box to a published table --- which
is precisely how you will read your thesis sources.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Name two
quantities it could measure per customer, per week, or per order (marketing
spend and sign-ups, price and basket size, delivery time and repeat
purchases) and write two sentences: first predict the sign of their
correlation and why; then name one third variable that could produce that
correlation without either quantity driving the other. There is no wrong
answer; the point is connecting today's tools to something you care about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Weekly flyer spend and weekly sign-ups should correlate positively, since flyers reach students who have never heard of us. But exam weeks could produce that correlation on their own: I hand out more flyers in exam weeks, and students order more prepared food in exactly those weeks."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to propose lurking
third variables for your pair and pick the most plausible one.
::::

Continue to Tutorial 2: Conditioning and relationships.

Formula sheet: Lecture 5 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 5 and 11.

:::
