# Lecture 5: Joint random variables, covariance, and correlation {#b05}

```{r}
#| label: b05-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 5, you can</span>

- read a joint table, find marginal and conditional distributions, and check independence
- measure how two variables move together with covariance and correlation
- find the mean and the variance of a sum of two variables
- explain why correlation differs from independence, causality, and importance
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Business questions are rarely about one variable alone: marketing and sales, price and demand, timing and success. The first step of an analysis is to measure whether two quantities move together at all, and how strongly.</p>

Take three pairs of Shark Tank variables: deals and late seasons, equity
offers and deals, seasons and equity offers. Which pairs move together, how strongly, and what does a correlation allow
us to say?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch
<a href="https://www.youtube.com/watch?v=7nu97OYx4X4">Joint PMFs and the
Expected Value Rule</a>,
<a href="https://www.youtube.com/watch?v=K2Tlj27nkjs">Covariance</a>, and
<a href="https://www.youtube.com/watch?v=HTs6Zhc2S1M">The Correlation
Coefficient</a>. Be ready to state what a joint PMF records, what covariance
averages, and how correlation standardises covariance. All links are also
listed in the <a href="../resources/video-guide.html">External Video
Guide</a>.
The videos explain key ideas in another way and can use a different notation. The lecture material and its notation always come first.
</div>

[Shark Tank data]{.shark-tag} **Warm up.** From earlier lectures you know the deal
indicator $X$ (1 for an on-air deal, 0 otherwise) and the late-period
indicator $Y$ (1 for Seasons 9 to 16, 0 for Seasons 1 to 8). When one pitch
record is selected, it now gives two numbers at once: the ordered pair $(X,Y)$. Write
down all possible values of the pair.

::: {.callout-note collapse="true"}
## Check your preparation

Four ordered pairs: $(0,0)$, $(0,1)$, $(1,0)$, $(1,1)$. In words: early
no-deal, late no-deal, early deal, late deal. Order matters: $(1,0)$ and
$(0,1)$ describe different pitches.
:::

[Shark Tank data]{.shark-tag} **Predict.** Answer this question before you come to class: over the 16 seasons, did the equity
percentage that entrepreneurs initially offer go up, stay about the same, or
go down? In the words of the videos: is the correlation between season and
equity offered positive, near zero, or negative? Choose one answer and write
it down. The lecture material gives the correct answer.

If ordered pairs, products, or standardisation are unfamiliar, review
[Mathematics Refresher: subscripts and indices](../resources/math-refresher.html#subscripts-and-indices).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section studies three pairs of variables from the Shark Tank file: deal
and late period, equity offered and deal, season and equity offered. Every
pair raises the same two questions: do the variables move together, and how
strongly? Until now each random variable was studied alone, with its own
distribution. Now two variables are studied together.

### Start with a founder's question

[Shark Tank data]{.shark-tag} A founder who decides how much equity to offer would like to know two
things. Have offers changed over the years? And do pitches that offer less
equity get a deal more often? A single number, the correlation, is often
used to answer such questions. It lies between $-1$ and $1$. The correlation
summarises many points in one number, so look at the plot before
you trust the number.
@fig-b05-season-equity shows the season (`season`) and the equity initially
offered (`equity_offered_pct`) for every pitch in the Shark Tank file.

```{r}
#| label: fig-b05-season-equity
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Season and equity initially offered (`equity_offered_pct`) for all 1,441 Shark Tank pitches. The points are moved slightly sideways so that overlapping points stay visible. The orange line is the straight line that fits the points best. It slopes downward: in later seasons, entrepreneurs tend to offer less equity."
library(ggplot2)
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)

ggplot(sharks, aes(x = season, y = equity_offered_pct)) +
  geom_point(position = position_jitter(width = 0.14, height = 0,
                                        seed = 2026),
             colour = "#009E73", alpha = 0.45, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "#D55E00", linewidth = 0.9) +
  scale_x_continuous(breaks = c(seq(1, 15, by = 2), 16)) +
  labs(x = "Season", y = "Equity initially offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each point is one pitch. The season is on the horizontal axis. The equity
initially offered, in percent of the company, is on the vertical axis. The
downward trend is visible, but the points spread widely around the line, so
the line does not predict a single pitch closely. The code below computes
the correlation for this pair and for a second pair: equity offered and the
0/1 deal indicator (`deal_on_show`). The first number answers the
prediction question from before class.

```{r}
#| label: b05-observed-correlations
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
c(
  season_with_equity_offered =
    cor(sharks$season, sharks$equity_offered_pct),
  equity_offered_with_deal =
    cor(sharks$equity_offered_pct, sharks$deal_on_show)
)
```

Season and equity offered have a correlation of about $-0.42$. The sign is
negative: entrepreneurs in later seasons tend to offer a smaller share of
their company. The size, 0.42 on a scale from 0 to 1, is moderate. Equity offered
and the deal indicator have a correlation of about $-0.11$. This is also
negative, but weak. As a rough guide, a correlation below about 0.3 in size
is weak, between about 0.3 and 0.7 moderate, and above about 0.7 strong.
Whether a correlation of a given size is important depends on the context.

What does such a number measure, and what can it not tell a founder? The
sections below start with the simplest case, two yes-or-no variables. The
section "Back to the opening numbers" returns to these two numbers.

### The joint probability mass function

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/joint-pmfs-and-the-expected-value-rule/">06.7 Joint PMFs and the Expected Value Rule</a>.</p>

[Shark Tank data]{.shark-tag} Lecture 3 described one random variable by its PMF. For two variables, the
two separate PMFs are not enough, because they cannot say how the values are
paired. They give the share of deals and the share of late-season pitches,
but not the share of late-season pitches with a deal. The joint PMF adds
this information.

::: {.definition-box}
<span class="label">Definition</span>

**Joint PMF.** For discrete random variables $X$ and $Y$ defined on the same
experiment, the joint probability mass function gives the probability of
every possible pair of values:

$$p_{X,Y}(x,y)=P(X=x,\;Y=y).$$

The comma means "and": $X$ takes the value $x$ and $Y$ takes the value $y$.
:::

Like any PMF, a joint PMF satisfies two conditions. Every entry is at least
0, and all entries together sum to 1, because each outcome falls in exactly
one cell. In symbols, as for the PMF of one variable in Lecture 3:

$$p_{X,Y}(x,y)\ge 0\quad\text{for every pair }(x,y),
\qquad
\sum_x\sum_y p_{X,Y}(x,y)=1.$$

The double sum adds the entries over all values of $x$ and all values of
$y$, that is, over all cells of the table.

The code builds the joint PMF of two variables from the Shark Tank file:
the deal indicator $X$ (`deal_on_show`) and the late-period indicator $Y$
(1 for Seasons 9 to 16, 0 for Seasons 1 to 8).

```{r}
#| label: b05-joint-pmf
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
X <- sharks$deal_on_show                # deal indicator: 1 = on-air deal
Y <- as.integer(sharks$season >= 9)     # late-period indicator: 1 = S9-16
joint_counts <- table(Deal = X, Late_period = Y)
joint_pmf <- prop.table(joint_counts)
joint_counts
round(joint_pmf, 3)
sum(joint_pmf)
```

The same joint PMF as a table, with the row sums and the column sums added:

| $p_{X,Y}(x,y)$ | $Y=0$ (early) | $Y=1$ (late) | row sum: $p_X(x)$ |
|---|---:|---:|---:|
| $X=0$ (no deal) | 0.229 | 0.159 | **0.388** |
| $X=1$ (deal)    | 0.262 | 0.350 | **0.612** |
| column sum: $p_Y(y)$ | **0.491** | **0.509** | 1 |

::: {.callout-note collapse="true"}
## Check: which entry is P(X = 1, Y = 1)?

The bottom-right cell of the joint table, 0.350. It is the probability that
the selected pitch reached a deal and came from a late season,
$505/1441\approx 0.350$. It is one number for one pair of values, not a row
and not a column.
:::

#### Recovering marginal PMFs

The PMF of one variable alone can be recovered from the joint PMF. Sum the
joint PMF over all values of the other variable:

$$p_X(x)=\sum_y p_{X,Y}(x,y),
\qquad
p_Y(y)=\sum_x p_{X,Y}(x,y).$$

On the table this means: sum along a row to get $p_X$, and sum down a column
to get $p_Y$. For example, $p_X(1)=0.262+0.350=0.612$ (the $X=1$ row) and
$p_Y(1)=0.159+0.350=0.509$ (the $Y=1$ column). These are $P(D)$ and $P(L)$
from earlier lectures. They now stand in the margins of the joint table. For
this reason the PMF of a single variable is called a **marginal PMF**, or
marginal distribution.

```{r}
#| label: b05-marginals
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
joint_pmf <- prop.table(table(Deal = X, Late_period = Y))
rowSums(joint_pmf)   # marginal PMF of X
colSums(joint_pmf)   # marginal PMF of Y
```

The joint table gives the marginal PMFs, but the marginal PMFs do not give
the joint table: many different joint tables have the same margins. What the joint table adds
is the pairing of the values.

#### Conditional PMFs are rescaled columns

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/conditional-pmfs/">07.2 Conditional PMFs</a>.</p>

Conditioning from Lecture 4 also works on a joint table. The **conditional PMF** of
$X$ given $Y=y$ gives the probabilities of the values of $X$ among the
outcomes with $Y=y$. For $p_Y(y)>0$,

$$p_{X\mid Y}(x\mid y)=\frac{p_{X,Y}(x,y)}{p_Y(y)}.$$

::: {#exm-b05-conditional}
## Worked example: one conditional entry, by hand

[Shark Tank data]{.shark-tag} What is $p_{X\mid Y}(1\mid 1)$, the deal probability given a late-season
pitch? Cover everything except the $Y=1$ column and divide its entries by
the column sum:

| $p_{X,Y}(x,y)$ | $Y=0$ | **$Y=1$** |
|---|---:|---:|
| $X=0$ | · | **0.159** |
| $X=1$ | · | **0.350** |
| column sum | · | **0.509** |

$$p_{X\mid Y}(1\mid 1)=\frac{0.350}{0.509}
=\frac{505/1441}{734/1441}=\frac{505}{734}\approx 0.688.$$

This is $P(D\mid L)$ from Lecture 4, the calculation 505 divided by 734.
Here you read it from a joint PMF. Dividing the whole column by 0.509 makes its entries sum
to 1. The rescaled column is the complete conditional PMF of $X$ given
$Y=1$.
:::

```{r}
#| label: b05-conditional-pmf
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
joint_counts <- table(Deal = X, Late_period = Y)
round(prop.table(joint_counts, margin = 2), 3)   # each column sums to 1
```

#### The conditional mean

A conditional PMF has a mean like any other PMF. The **conditional mean**
$E[X\mid Y=y]$ is the expectation of $X$ computed with the conditional PMF
for $Y=y$, that is, with the rescaled column:

$$E[X\mid Y=y]=\sum_x x\,p_{X\mid Y}(x\mid y).$$

The R output above shows both rescaled columns. For the deal indicator,

$$E[X\mid Y=1]=0\times0.312+1\times0.688=0.688,
\qquad
E[X\mid Y=0]=0\times0.467+1\times0.533=0.533.$$

These are the averages of $X$ among late-season pitches and among
early-season pitches. Because $X$ is a 0/1 variable, they are the deal rates
of the two periods. Regression in QM Intermediate also compares the average
of one variable across the values of another variable.

### Independence of random variables

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/independence-of-random-variables/">07.4 Independence of Random Variables</a>.</p>

[Shark Tank data]{.shark-tag} Lecture 4 defined independence for two events: $P(A\cap B)=P(A)\,P(B)$. A
random variable defines many events, one for every value it can take.
So independence of two random variables requires more: the product
form must hold for every pair of values.

::: {.definition-box}
<span class="label">Definition</span>

**Independent random variables.** Discrete random variables $X$ and $Y$ are
independent when the joint PMF equals the product of the two marginal PMFs
in every cell of the joint table:

$$p_{X,Y}(x,y)=p_X(x)\,p_Y(y)
\quad\text{for every pair } (x,y).$$

In words: knowing the value of one variable does not change the
probabilities of the other.
:::

Under independence, the conditional PMF equals the marginal PMF. Dividing
the product $p_X(x)\,p_Y(y)$ by $p_Y(y)$ leaves $p_X(x)$:

$$p_{X\mid Y}(x\mid y)=p_X(x)
\quad\text{for every } x \text{ and every } y \text{ with } p_Y(y)>0.$$

One cell that does not match is enough to prove dependence. To show
independence, every cell must match.

::: {#exm-b05-independence}
## Worked example: the (1,1) cell check

If deal and late period were independent, the bottom-right cell would equal
the product of its margins:

$$p_X(1)\,p_Y(1)=0.612\times 0.509\approx 0.312.$$

The actual entry is $p_{X,Y}(1,1)=0.350$. The observed 0.350 differs from
the 0.312 that independence requires, so $X$ and $Y$ are not independent.
Deals and late seasons occur together more often than they would under
independence. Lecture 4 found the same result for the events:
$P(D\mid L)\ne P(D)$.
:::

```{r}
#| label: b05-joint-independence
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
observed_joint_11 <- mean(X == 1 & Y == 1)
independent_joint_11 <- mean(X == 1) * mean(Y == 1)
c(observed_joint_11 = observed_joint_11,
  independent_joint_11 = independent_joint_11)
```

In a 2 by 2 table one matching cell is also enough to show independence,
because the margins then fix the other three cells. In larger tables every
cell must be checked.

### Covariance comes from expectation

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/covariance/">12.5 Covariance</a>.</p>

[Shark Tank data]{.shark-tag} Independent or dependent is a yes-or-no answer. The next tool says more: in
which direction do the two variables move together? Lecture 3 measured the
spread of one variable as an expected squared deviation,
$\operatorname{Var}(X)=E[(X-\mu_X)^2]$. The covariance applies the same idea
to a pair of variables. It multiplies the deviation of $X$ from its mean by
the deviation of $Y$ from its mean, and takes the expectation of this
product.

Such a product is a function of two random variables. A function $h(X,Y)$
of two discrete random variables is itself a random variable. Its expected
value is a sum over all pairs of values:

$$E[h(X,Y)]=\sum_x\sum_y h(x,y)\,p_{X,Y}(x,y).$$

This is the two-variable version of the rule for $E[g(X)]$ from Lecture 3.
The covariance uses this rule.

::: {.definition-box}
<span class="label">Definition</span>

**Covariance.** For random variables $X$ and $Y$ with means $\mu_X=E[X]$ and
$\mu_Y=E[Y]$,

$$\operatorname{Cov}(X,Y)=E[(X-\mu_X)(Y-\mu_Y)].$$

A positive covariance says that above-average values of $X$ tend to occur
with above-average values of $Y$. A negative covariance says that
above-average values of $X$ tend to occur with below-average values of $Y$.
:::

The product of the two deviations has a sign:

- both values above their means, or both below: the product is positive.
- one value above its mean and the other below: the product is negative.

For discrete variables, the expectation is a sum over the cells of the joint
table. For every cell, multiply the two deviations, weight the product by
the probability of the cell, and add up:

$$\operatorname{Cov}(X,Y)=\sum_x\sum_y (x-\mu_X)(y-\mu_Y)\,p_{X,Y}(x,y).$$

::: {#exm-b05-covariance}
## Worked example: four signed products, one at a time

Take this small joint PMF, an invented example:

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.30 | 0.20 |
| $X=1$ | 0.10 | 0.40 |

The marginal PMFs give $p_X(1)=0.10+0.40=0.50$ and $p_Y(1)=0.20+0.40=0.60$, so
$\mu_X=E[X]=0.5$ and $\mu_Y=E[Y]=0.6$.

Write the product of the two deviations for each cell, with its sign:

| Cell $(x,y)$ | $(x-0.5)(y-0.6)$ | Sign | Probability | Contribution |
|---|---:|:---:|---:|---:|
| $(0,0)$ | $(-0.5)(-0.6)=+0.30$ | both below | 0.30 | $+0.090$ |
| $(0,1)$ | $(-0.5)(+0.4)=-0.20$ | mixed | 0.20 | $-0.040$ |
| $(1,0)$ | $(+0.5)(-0.6)=-0.30$ | mixed | 0.10 | $-0.030$ |
| $(1,1)$ | $(+0.5)(+0.4)=+0.20$ | both above | 0.40 | $+0.080$ |

The last column is the product times the probability of the cell. Add the
four contributions:

$$\operatorname{Cov}(X,Y)=0.090-0.040-0.030+0.080=0.10.$$

Most of the probability lies in the two corners where the variables agree,
$(0,0)$ and $(1,1)$. The positive contributions are therefore larger, and
the covariance is positive: the two variables move together.

**Verify with the computational form.** Multiplying out the product in the
definition gives a second formula for the same number:

$$\operatorname{Cov}(X,Y)=E[XY]-E[X]\,E[Y].$$

Here $XY$ is different from zero only in the $(1,1)$ cell, so
$E[XY]=1\times 1\times 0.40=0.40$, and

$$\operatorname{Cov}(X,Y)=0.40-(0.5)(0.6)=0.40-0.30=0.10.$$

Both methods give the same answer. The definition explains the sign. The
computational form is quicker by hand.
:::

Here are the same two methods for the pair from the Shark Tank file, the deal
indicator $X$ and the late-period indicator $Y$:

```{r}
#| label: b05-covariance-from-expectation
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
E_X <- mean(X)
E_Y <- mean(Y)
E_XY <- mean(X * Y)

c(
  E_X = E_X,
  E_Y = E_Y,
  E_XY = E_XY,
  covariance_from_moments = E_XY - E_X * E_Y,
  covariance_from_deviations = mean((X - E_X) * (Y - E_Y))
)
```

Both methods give a covariance of about 0.039. The sign is positive: deals
and late seasons go together. The size of this number is hard to judge on
its own. The correlation, introduced below, solves this problem.

::: {.callout-caution collapse="true"}
## Try it: do welcome discounts and reorders go together?

Try this with your neighbour first. An invented example: for a customer
chosen at random, let $X=1$ if the customer reordered and $Y=1$ if the
customer received a welcome discount.

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.40 | 0.20 |
| $X=1$ | 0.10 | 0.30 |

*Step 1 (given).* The marginal PMFs give $p_X(1)=0.40$ and $p_Y(1)=0.50$.

*Step 2 (you).* Before you calculate: which two corners of the table hold
most of the probability, and what sign of the covariance does that suggest?

*Step 3 (you).* Calculate the covariance with $E[XY]-E[X]E[Y]$.

::: {.callout-note collapse="true"}
## Check your answer

Most of the probability is in $(0,0)$ and $(1,1)$, the two corners where the
variables agree, so the covariance should be positive.
$E[XY]=0.30$ and $E[X]E[Y]=0.40\times0.50=0.20$, so
$\operatorname{Cov}(X,Y)=0.10$. Customers with a discount reorder more often.
Whether the discount is the reason for the reorders is a different question.
:::
:::

The units of a covariance are the units of $X$ times the units of $Y$. If
$X$ is measured in dollars and $Y$ in years, $\operatorname{Cov}(X,Y)$ is in
dollar-years. You can interpret the sign of a covariance. Its size changes
whenever the unit of $X$ or $Y$ changes, so it is hard to compare two
covariances.

#### Independence implies zero covariance

If $X$ and $Y$ are independent, then $E[XY]=E[X]\,E[Y]$, so
$\operatorname{Cov}(X,Y)=0$. The opposite is not always true: zero
covariance does not imply independence. A covariance of zero says only that there is no straight-line pattern. Other
forms of dependence are still possible, and @fig-b05-linear-curved below
shows one.

### Sums of two variables {#b05-sums}

A founder often cares about a total, such as the revenue of two products
together. Two rules describe the sum $X+Y$ of two random variables. The
first rule always holds, for independent and for dependent variables:

$$E[X+Y]=E[X]+E[Y].$$

The rule for the variance has an extra term with the covariance:

$$\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)
+2\operatorname{Cov}(X,Y).$$

For independent variables the covariance is zero, and the variances add:
$\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)$.

An invented example: a campus café sells coffee and croissants. Weekly
coffee revenue $X$ has mean 400 euros and variance 900 euros squared, a standard deviation
of 30 euros. Weekly croissant revenue $Y$ has mean 200 euros and variance
400 euros squared, a standard deviation of 20 euros. The two products tend to sell
together, with $\operatorname{Cov}(X,Y)=300$ euros squared. Total revenue then has mean
$400+200=600$ euros and variance $900+400+2\times300=1900$ euros squared, a standard
deviation of about 43.6 euros. Without the covariance term the variance
would be 1300 euros squared, a standard deviation of about 36.1 euros. Because the
products sell together, the standard deviation of the total is about 21%
larger than the 36.1 euros from the two variances alone.

Lecture 8 uses the independent case to find the spread of a sample mean.

::: {.callout-tip collapse="true"}
## Optional: a general linear combination

A linear combination of $X$ and $Y$ is a new variable $W=a+bX+cY$ with fixed
numbers $a$, $b$, and $c$. Its mean and variance are

$$E[W]=a+b\,E[X]+c\,E[Y],$$

$$\operatorname{Var}(W)=b^2\operatorname{Var}(X)+c^2\operatorname{Var}(Y)
+2bc\operatorname{Cov}(X,Y).$$

The sum $X+Y$ is the case $a=0$, $b=1$, $c=1$. The difference $X-Y$ is the
case $b=1$, $c=-1$. The covariance term then has a minus sign.
:::

### Correlation standardises covariance

<p class="concept-video"><strong>MIT explanations (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-correlation-coefficient/">12.8 The Correlation Coefficient</a> and
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/interpreting-the-correlation-coefficient/">12.10 Interpreting the Correlation Coefficient</a>.</p>

[Shark Tank data]{.shark-tag} The size of a covariance depends on the units of measurement. The solution
is to divide the covariance by the two standard deviations.

::: {.definition-box}
<span class="label">Definition</span>

**Correlation.** The covariance divided by both standard deviations. For
variables whose standard deviations are not zero,

$$\rho_{XY}
=\frac{\operatorname{Cov}(X,Y)}
{\operatorname{SD}(X)\operatorname{SD}(Y)},
\qquad -1\le\rho_{XY}\le1.$$

The sign gives the direction of the linear association. The size of the
number without its sign, between 0 and 1, gives the strength. A correlation
has no units.
:::

The symbol $\rho$ is the Greek letter rho. It stands for the correlation of
two random variables, also written $\operatorname{Corr}(X,Y)$. A correlation
computed from data is written $r$. In the record-selection experiment, each
record has probability $1/n$. Then $\rho$ and $r$ give the same number,
because the factors $n$ and $n-1$ cancel in the ratio. The numbers
$-0.42$, $-0.11$, and $0.16$ in this lecture are values of $r$.

Check the units. The numerator is in dollar-years. The denominator is
dollars times years, also dollar-years. The units cancel, so the correlation
has no units. It also stays the same when a variable is multiplied by a
positive number, for example when dollars are converted to euros.

```{r}
#| label: b05-correlation-from-covariance
X <- sharks$deal_on_show
Y <- as.integer(sharks$season >= 9)
cov_xy <- mean((X - mean(X)) * (Y - mean(Y)))
rho_xy <- cov_xy / (sqrt(mean((X - mean(X))^2)) *
                    sqrt(mean((Y - mean(Y))^2)))
c(from_definition = rho_xy,
  from_R = cor(X, Y))
```

Both methods give about 0.16 for the deal indicator and the late-period
indicator: a weak positive linear association.

The word "linear" is important. A correlation measures only how closely the points
follow a straight line. @fig-b05-linear-curved shows two invented patterns
with 61 points each.

```{r}
#| label: fig-b05-linear-curved
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.4
#| fig-cap: "Correlation measures linear association only (two invented patterns). Pattern A rises almost along a straight line and has r close to 1. In Pattern B the two variables are strongly related, but the pattern is U-shaped: the fitted straight line is flat and r is close to 0."
library(ggplot2)
x_association <- seq(-2, 2, length.out = 61)
y_linear <- as.numeric(scale(0.8 * x_association +
                             0.25 * sin(7 * x_association)))
y_curved <- as.numeric(scale(x_association^2))

assoc_df <- data.frame(
  x = rep(x_association, times = 2),
  y = c(y_linear, y_curved),
  pattern = rep(c("Pattern A", "Pattern B"), each = 61)
)
r_labels <- data.frame(
  pattern = c("Pattern A", "Pattern B"),
  label = sub("-0.00", "0.00", fixed = TRUE,
    paste0("r = ", sprintf("%.2f", c(
    cor(x_association, y_linear),
    cor(x_association, y_curved)
  )))),
  x = -2,
  y = max(assoc_df$y) + 0.45
)

ggplot(assoc_df, aes(x = x, y = y)) +
  geom_point(colour = "#009E73", alpha = 0.72, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "#D55E00", linewidth = 0.9) +
  geom_text(data = r_labels, aes(x = x, y = y, label = label),
            hjust = 0, colour = "#17324D", size = 3.4) +
  facet_wrap(~ pattern) +
  labs(x = "X", y = "Y (standardised units)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Each point is one $(x,y)$ pair. The orange line is the straight line that
fits the points best. Pattern A rises almost along a straight line, so its
correlation is close to 1. Pattern B has a clear U-shaped relationship, but
its fitted line is flat and its correlation is close to 0. The two variables
in Pattern B are clearly dependent. A correlation of zero therefore does not
imply independence.

::: {.callout-note collapse="true"}
## Check: which panel has correlation near zero, and why is that not "no relationship"?

Pattern B. Its correlation is near zero because, in the products of deviations, the
rise on the right cancels the fall on the left. There is no
linear trend. The relationship itself is strong: if you know $x$, you know
$y$ almost exactly. A correlation near zero means only "no linear
association". It never means "no relationship".
:::

#### Sample covariance and sample correlation

The formulas above are for random variables. For data, that is, for $n$
observed pairs $(x_i,y_i)$, the sample covariance is

$$s_{XY}=\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})(y_i-\bar{y}),$$

and the sample correlation is

$$r_{XY}=\frac{s_{XY}}{s_X s_Y}.$$

Here $s_X$ and $s_Y$ are the sample standard deviations from Lecture 3. The
sample correlation is written $r$ for short, and the R function `cor()`
calculates it. $s_{XY}$ is the two-variable version of the sample variance
$s^2$. Both $s_{XY}$ and $r_{XY}$ describe the data, and they are used to
estimate $\operatorname{Cov}(X,Y)$ and $\rho_{XY}$.

::: {.callout-tip collapse="true"}
## Optional: point-biserial correlation

Sometimes one of the two variables is a 0/1 indicator, for example the
equity offered paired with the deal indicator. The ordinary correlation of
such a pair is also called the point-biserial correlation. It is not a new
formula, only a special name that you may meet in other courses. `cor()`
computes it in the same way.
:::

### Back to the opening numbers

[Shark Tank data]{.shark-tag} You can now read the two numbers from the start of the lecture. Season and
equity offered have a correlation of $r\approx-0.42$: a moderate negative
linear association, without units. Equity offered and the deal indicator
have a correlation of $r\approx-0.11$: a weak negative one. Both numbers
describe how the variables move together in the 1,441 records of the Shark
Tank file. The box below says what a founder may and may not conclude from
them.

::: {.caveat-box}
<span class="label">What a correlation can and cannot say</span>

- A correlation shows that two variables went together. It does not show
  that one makes the other happen. A third variable can produce a correlation
  between two others. Here the season is such a third variable: later seasons
  have both lower equity offers and more deals. Holding season fixed is what
  multiple regression does in QM Intermediate.
- A correlation measures straight-line patterns only. A strong curved pattern
  can have a correlation near zero.
- A correlation has no units. It says nothing about how important the
  relationship is for a business decision.

**What you can do with it:** treat a correlation as a first hint, not as
proof. If acting on the correlation is cheap and easy to undo, act and watch what happens. If it is expensive,
test first.
:::

### Answering the driving question

Deal and late period move together. The joint entry 0.350 is larger than
the 0.312 expected under independence, and the correlation is weak and
positive, $r\approx 0.16$. Season and equity offered move in opposite
directions, with moderate strength: $r\approx -0.42$. Equity offered and
the deal indicator also move in opposite directions, but only weakly,
$r\approx -0.11$. The box above explains why a founder should not read this
number as advice. A correlation describes the direction and the strength of the linear association in these
records. It says nothing about cause, and nothing about curved
relationships.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Over 30 weeks, your weekly flyer spending and your weekly new sign-ups have a correlation of 0.6. Exam weeks could explain the correlation: you hand out more flyers then, and students order more prepared food then. Should you double the flyer budget? Use a simple rule. If acting is cheap and easy to undo, act on the correlation and watch what happens. If acting is expensive or hard to undo, test first. Flyers are cheap, so you act. You also test: for the next four ordinary weeks you alternate weeks with and without flyers and compare the sign-ups.
:::

**Next:** Lecture 6 turns to continuous variables: quantities that can take
any value in a range. Tutorial 2 follows after Lecture 6. It practises
Lectures 4 and 5 only, with exam-style questions: conditional probabilities from a table,
Bayes' rule, a joint PMF, and covariance and correlation. Bring your
after-class work to that tutorial, especially the tasks on the joint table and
on covariance.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions

1. What information does a joint PMF contain that two marginal PMFs do not?
2. How do you recover a marginal PMF from a joint PMF?
3. Why can covariance change when units change?
4. Why does independence imply zero covariance?
5. Why does zero correlation not imply independence?

::: {.callout-note collapse="true"}
## Check your answers

1. It records how the values of the two variables occur together: the
   pairing, not only the two separate distributions.
2. Sum the joint PMF over every value of the other variable.
3. The units of a covariance are the units of $X$ times the units of $Y$,
   so a change of units in either variable changes the covariance.
4. Independence gives $E[XY]=E[X]\,E[Y]$, so
   $\operatorname{Cov}(X,Y)=E[XY]-E[X]\,E[Y]=0$.
5. Correlation measures linear association. Dependent variables can have
   zero linear association (the U-shape in @fig-b05-linear-curved).
:::

### Finish a half-worked example: covariance on a new table

Re-read the worked example on covariance (@exm-b05-covariance). Then repeat
the same steps on a new joint PMF. It is an invented example: $X$ and $Y$
each take the value 0 or 1.

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.10 | 0.30 |
| $X=1$ | 0.40 | 0.20 |

*Step 1 (given).* The marginal PMFs give $p_X(1)=0.60$ and $p_Y(1)=0.50$, so
$\mu_X=0.6$ and $\mu_Y=0.5$.

*Step 2 (you).* Write the four products of deviations $(x-0.6)(y-0.5)$, one
per cell, with their signs.

*Step 3 (you).* Weight each product by its cell probability and add.

*Step 4 (you).* Verify with $E[XY]-E[X]E[Y]$ and interpret the sign.

::: {.callout-note collapse="true"}
## Check your answer

Contributions: $(0,0)$: $(+0.30)(0.10)=+0.030$, $(0,1)$:
$(-0.30)(0.30)=-0.090$, $(1,0)$: $(-0.20)(0.40)=-0.080$, $(1,1)$:
$(+0.20)(0.20)=+0.040$. Sum:
$\operatorname{Cov}(X,Y)=0.030-0.090-0.080+0.040=-0.10$. Check:
$E[XY]=0.20$ and $E[X]E[Y]=0.6\times 0.5=0.30$, so $0.20-0.30=-0.10$. The
negative sign says that most of the probability is in the two corners where
the variables disagree: $X=1$ tends to come with $Y=0$.
:::

**Practice on your own.** An invented example: $X$ and $Y$ each take the
value 0 or 1, with the joint PMF

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.25 | 0.25 |
| $X=1$ | 0.15 | 0.35 |

1. Find the marginal PMFs of $X$ and $Y$.
2. Are $X$ and $Y$ independent?
3. Calculate $\operatorname{Cov}(X,Y)$ and interpret its sign.
4. Calculate $P(X=1\mid Y=1)$ and compare it with $P(X=1)$.
5. Define $Z=100X$. Calculate $\operatorname{Cov}(Z,Y)$ and state what
   happens to $\operatorname{Corr}(Z,Y)$.
6. Calculate the conditional means $E[X\mid Y=1]$ and $E[X\mid Y=0]$.
7. Calculate $E[X+Y]$ and $\operatorname{Var}(X+Y)$. Use
   $\operatorname{Var}(X)=0.25$ and $\operatorname{Var}(Y)=0.24$.

::: {.callout-note collapse="true"}
## Check your answers

1. $P(X=0)=0.50$, $P(X=1)=0.50$. $P(Y=0)=0.40$, $P(Y=1)=0.60$.
2. No. $P(X=1,Y=1)=0.35$, whereas $P(X=1)P(Y=1)=0.50\times 0.60=0.30$. One
   cell that does not match proves dependence.
3. $E[X]=0.50$, $E[Y]=0.60$, $E[XY]=0.35$, so
   $\operatorname{Cov}(X,Y)=0.35-0.30=0.05$. Positive: the pair $(1,1)$
   occurs with greater probability than independence would give.
4. $P(X=1\mid Y=1)=0.35/0.60=7/12\approx 0.583$, larger than the marginal
   $P(X=1)=0.50$: conditioning on $Y=1$ shifts the distribution of $X$
   upward.
5. $\operatorname{Cov}(Z,Y)=100\operatorname{Cov}(X,Y)=5$. Multiplication by
   a positive constant rescales covariance but leaves correlation unchanged.
6. Rescale each column. For $Y=1$ the column is $0.25$ and $0.35$ with sum
   $0.60$, so $E[X\mid Y=1]=0\times(0.25/0.60)+1\times(0.35/0.60)\approx
   0.583$. For $Y=0$ the column is $0.25$ and $0.15$ with sum $0.40$, so
   $E[X\mid Y=0]=0.15/0.40=0.375$. The average of $X$ is higher when $Y=1$,
   which agrees with the positive covariance.
7. $E[X+Y]=0.50+0.60=1.10$.
   $\operatorname{Var}(X+Y)=0.25+0.24+2\times0.05=0.59$. Without the
   covariance term the answer would be 0.49, which is too small because $X$
   and $Y$ move together.
:::

### Review: two questions from earlier lectures

**From Lecture 2.** One of the 1,441 Shark Tank pitch records is selected
uniformly at random. Is "the selected
pitch is the record with pitch_id 7" an outcome or an event? And what kind
of object is the record itself?

::: {.callout-note collapse="true"}
## Check your answer

The record itself is an outcome: one complete possible result. The
description defines the set that contains only that record, and a set of
outcomes is an event. It is the same difference as between 4 and $\{4\}$
for a die roll.
:::

**From Lecture 4.** An invented classifier marks ("flags") pitches that are
likely to get a deal. In the population, the deal base rate is 10%. The
sensitivity is 70% and the false-positive rate is 20%. Use a tree for 1,000
pitches. What share of flagged pitches are deals?

::: {.callout-note collapse="true"}
## Check your answer

Of 100 deals, 70 are flagged. Of 900 non-deals, 180 are flagged. So
$P(\text{deal}\mid\text{flag})=70/(70+180)=70/250=0.28$.
:::

### Worked exam interpretation

[Shark Tank data]{.shark-tag} The Shark Tank file contains 1,441 televised pitches from Seasons 1 to 16. The
variable `season` records the season number, and `equity_offered_pct`
records the percentage of the business initially offered to the sharks.

**Question.** R reports
`cor(sharks$season, sharks$equity_offered_pct) = -0.416`. Interpret this
result in context. State:

1. the direction and approximate strength of the linear association.
2. what the association means for pitches in earlier and later seasons.
3. whether the correlation has units.
4. one conclusion that cannot be justified from this result.

::: {.callout-note collapse="true"}
## Check a model answer

Among the 1,441 records, season and equity initially offered have a moderate
negative linear association. Later-season pitches tend to offer a smaller
ownership percentage. This is not true of every individual pitch.
The correlation has no units. It summarises only the linear association and
does not show that the passage of time caused offers to change.
:::

**A second short question.** An invented example. A delivery start-up
compares 50 city districts. The number of flyers it handed out in a district
and the number of orders from that district have a correlation of 0.5. Name
one third variable that could produce this correlation, and say how.

::: {.callout-note collapse="true"}
## Check a model answer

The number of residents of the district. A large district gets more flyers,
and it also places more orders, because more people live there. Flyers and
orders then go together even if no flyer leads to an order. A fair
comparison would look at districts of similar size.
:::

### Worked exam interpretation: independence and covariance sign

An invented example: $X$ and $Y$ each take the value 0 or 1, with the
joint PMF

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.18 | 0.42 |
| $X=1$ | 0.12 | 0.28 |

Decide whether $X$ and $Y$ are independent. Then state
$\operatorname{Cov}(X,Y)$ without long calculation.

::: {.callout-note collapse="true"}
## Check your answer

The marginal probabilities are $P(X=1)=0.40$ and $P(Y=1)=0.70$. Check every
cell:
$0.60\times 0.30=0.18$, $0.60\times 0.70=0.42$, $0.40\times 0.30=0.12$,
$0.40\times 0.70=0.28$. All four cells match, so $X$ and $Y$ are
independent. Independence implies $\operatorname{Cov}(X,Y)=0$ (directly:
$E[XY]=0.28=E[X]E[Y]$). In the earlier tables one cell that did not match
was enough to show dependence. Here all four cells match. In a 2 by 2 table,
checking one cell would have been enough, because the margins fix the other
three.
:::

### Check the AI

[Shark Tank data]{.shark-tag} In the Shark Tank file, the equity initially offered (`equity_offered_pct`)
and the deal indicator (`deal_on_show`) have a correlation of about $-0.11$. You
asked an AI assistant: "What does the correlation of -0.11 between offered equity and getting a deal tell me?" It answered:

> "It proves that offering less equity makes the sharks say yes. And because -0.11 is so close to zero, the two variables are independent."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

There are two errors. First, a correlation shows that two variables went together in these records, not that one makes the other happen. Later seasons have both lower offers and more deals, and this alone can produce the correlation. Second, a correlation that is not zero rules out independence. The correlation is weak, but it is not zero.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Name two
quantities it could measure per customer, per week, or per order. Examples
are marketing spend and sign-ups, price and basket size, or delivery time
and repeat purchases. Sentence 1: predict the sign of their correlation and
say why. Sentence 2: name one third variable that could produce that
correlation without one quantity causing the other. There is no wrong
answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus venture that sells boxes of prepared meals. "Weekly flyer spend and weekly sign-ups should correlate positively, since flyers reach students who have never heard of us. But exam weeks could produce that correlation on their own: I hand out more flyers in exam weeks, and students order more prepared food in exactly those weeks."
:::

You can discuss this task, and any other task in this chapter, with
Tilly for QM Basic, the course AI assistant: ask it to propose third variables for
your pair, and choose the most likely one.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask Tilly for
QM Basic, the course AI assistant. For example: "Give me an extra exercise for
Lecture 5. Show the solution only after I send my answer."

[Continue to Lecture 6: Continuous distributions and the normal model](b06-continuous-normal.qmd).

[Formula sheet: Lecture 5 formulas](formula-sheet.qmd#formula-l5).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 5 and 11.

:::
