# Lecture 6: Continuous distributions and the normal model {#b06}

```{r}
#| label: b06-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 6, you can</span>

- distinguish a continuous random variable from a discrete one
- explain why $P(X=x)=0$ for a continuous random variable and interpret a
  probability density function as area rather than point probability
- use a CDF to calculate interval probabilities
- calculate uniform probabilities from interval lengths
- standardise a normal value and compute normal probabilities by hand from
  supplied $Z$ values and with `pnorm()`
- convert a percentile back to a raw value with $x=\mu+z\sigma$ and verify
  with `qnorm()`
- find quantiles of model distributions (the uniform median, normal
  quartiles)
- check a proposed normal model against a histogram of data and say when it
  fails.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- calculate uniform probabilities from interval lengths
- standardise a value and use the supplied $Z$ values, with the complement
  rule and the symmetry rule where needed
- go back from a percentile to a value with $x=\mu+z\sigma$
- say from a histogram whether a normal model fits.

**Enough to recognise:**

- the density notation $f_X(x)$
- `pnorm()` and `qnorm()`
- the central 95% of the standard normal distribution, between $-1.96$ and
  $1.96$.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Not everything comes in whole numbers: equity shares, waiting times, and revenues can take any value in a range. Insurers, lenders, and operations planners price risk on such quantities every day, and that needs a new kind of probability model.</p>

You plan to offer the sharks 30% of your company. How unusual is that, and
can a model with two numbers answer the question without counting every
pitch?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=8QFpZ3FndBc">Probability
Density Functions</a>,
<a href="https://www.youtube.com/watch?v=jPB9zI8F7rE">Uniform and
Piecewise Constant PDFs</a>, and
<a href="https://www.youtube.com/watch?v=6UMv4vb4y7c">Normal Random
Variables</a>. Focus on the difference between density height and
probability, and on the roles of the two normal parameters. All links are
also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (5 minutes).** A discrete random variable has separated
possible values (Lecture 3 attached a PMF and a CDF to such variables). A
continuous random variable can take every real value in an interval. Match
each variable to the better description:

1. number of sharks in an on-air deal
2. equity offered, as an exact percentage of the company
3. season number
4. exact amount invested, before rounding to dollars.

::: {.callout-note collapse="true"}
## Check your answers

Variables 1 and 3 are discrete. Variables 2 and 4 are continuous in a
mathematical model, even if a measuring device or database records rounded
values.
:::

**Predict (5 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: for a *continuous* random variable $X$, is
$P(X = \text{exactly } 10)$ **very small** or **exactly zero**? Commit to an
answer. You find the answer in the lecture material.

If fractions or rearranging an equation feel unfamiliar, review
[Mathematics Refresher: fractions and rearranging equations](../resources/math-refresher.html#fractions-and-rearranging-equations).
The same refresher has short parts on square roots and on inequalities.

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section builds probability models for quantities that can take any
value in a range. It uses them to answer one question about the equity
offered in the Shark Tank file.

### Start with a founder's question

You plan to offer the sharks 30% of your company. Is that unusual? The Shark
Tank file records the equity offered in each pitch (`equity_offered_pct`).
The code below reads the file and counts the pitches that offered 30% or
more.

```{r}
#| label: b06-equity-count
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
nrow(sharks)
sum(sharks$equity_offered_pct >= 30)
mean(sharks$equity_offered_pct >= 30)
```

In the file, 76 of the 1,441 pitches offered 30% or more, about 5%. Now
suppose you did not have the file. You only knew two numbers about the
offers: the mean and the standard deviation.

```{r}
#| label: b06-equity-model
c(mean_equity = mean(sharks$equity_offered_pct),
  sd_equity   = sd(sharks$equity_offered_pct))
```

The mean is 13.2% and the standard deviation is 8.4 percentage points. Could
you still say how unusual 30% is? An AI assistant that gets this question
will often answer with a model called the normal distribution, and it
reports about 2%. That is less than half of the share in the data.

This lecture shows how such models work for quantities that can take any
value in a range, how to calculate with them by hand, and how to check
whether a model fits the data at hand. The check comes near the end, and it
explains the gap between 2% and 5%.

### From counting values to measuring intervals

Equity offered is a continuous quantity: an offer $X$ can be any percentage
in a range, for example 29.87%. What is the probability that an offer is
exactly 30%?

Take a window around 30. The offer lies between 29.5% and 30.5% with some
probability $q$. The window from 29.95% to 30.05% is ten times narrower.
Over such a short range the probability is spread almost evenly, so this
window holds about $q/10$. Every further decimal divides the probability by
about ten. The value "exactly 30%" lies inside every one of these windows,
so its probability cannot be larger than any of theirs. Only zero is that
small. For a continuous random variable,

$$P(X=x)=0 \quad\text{for every single value } x.$$

This answers the Wooclap question from the preparation: exactly zero, not
very small. Probability has not disappeared. It belongs to intervals, not to
single values.

### From probability mass to probability density

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-density-functions/">08.2 Probability Density Functions</a>.</p>

For a discrete random variable, the PMF gives a probability to each single
value. For a continuous variable that no longer works. Look at the two
windows again. The probability fell from $q$ to about $q/10$, and the width
fell from 1 to 0.1 percentage points. Their ratio hardly changed:

$$\frac{q}{1}\approx\frac{q/10}{0.1}.$$

For narrow windows, the probability divided by the width is approximately
constant. This ratio is the probability per unit of width near 30%. It is
called the density at 30%, written $f_X(30)$. The density takes the place of
the PMF.

::: {.definition-box}
<span class="label">Definition</span>

**Probability density function (PDF).** A curve $f_X(x)$ that is never
negative and has a total area of 1 under it. The probability that $X$ lies
between $a$ and $b$ is the area under the curve between $a$ and $b$.
:::

Three reading rules follow. The height of the curve is a rate, not a
probability. The area over an interval is a probability. A rate can be high
where the probability is packed into a narrow range, so a density can exceed
1, as long as the total area is 1. A metal rod gives the same picture: it has
a weight per metre at each point, but no single point has weight.

The Shark Tank file records the equity offered in rounded percentages, so
values repeat.

```{r}
#| label: b06-equity-exactly-30
sum(sharks$equity_offered_pct == 30)
```

In the file, 33 of the 1,441 pitches show exactly 30. That comes from
rounding. The model is continuous.

You do not need calculus in this course. You read probability as an area,
and you calculate it with a CDF, with R, or with supplied values.

::: {.callout-tip collapse="true"}
## Optional: the formal definition of a density

A function $f_X$ is a density of the continuous random variable $X$ if it is
never negative and has total area 1,

$$f_X(x)\ge 0 \ \text{ for every } x,
\qquad
\int_{-\infty}^{\infty} f_X(x)\,dx=1,$$

and if, for all values $a\le b$,

$$P(a\le X\le b)=\int_a^b f_X(x)\,dx.$$

The integral is the area under the curve between $a$ and $b$. For $a=b$ the
area has zero width, which gives $P(X=a)=0$ again.

The normal distribution, which comes later in this lecture, has the density

$$f_X(x)=\frac{1}{\sigma\sqrt{2\pi}}\,
\exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right).$$

Nobody needs to memorise this formula. R evaluates it with `dnorm()`.
:::

::: {.callout-note collapse="true"}
## Check: is a rectangle of width 0.5 and height 2 a valid density?

Yes. The curve is never negative and its total area is
$0.5 \times 2 = 1$, so it is a valid PDF. The height 2 is not a probability.
No probability here exceeds 1, because every probability is an area inside
that rectangle. The height 2 means that the probability is packed into a
narrow range.
:::

#### The CDF works for every random variable

The cumulative distribution function (CDF) is still

$$F_X(x)=P(X\le x),$$

exactly as in Lecture 3. In a picture, $F_X(x)$ is the area under the density
to the left of $x$. For a continuous random variable the CDF rises smoothly,
without jumps, from 0 to 1. It never falls, because a move to the right can
only add area. @fig-b06-uniform-cdf in the next section shows an example.

The probability of an interval is the difference of two CDF values:

$$P(a<X\le b)=F_X(b)-F_X(a).$$

Because a single point has zero probability, changing $<$ to $\le$ does not
change a continuous probability: $P(a<X<b)$, $P(a\le X\le b)$, and the two
mixed versions are all equal. This differs from a discrete variable, where a
point may carry positive probability and the inequality type matters.

### The uniform distribution

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/uniform-piecewise-constant-pdfs">08.3 Uniform and Piecewise Constant PDFs</a>.</p>

The uniform distribution is the simplest continuous model. All possible
values lie in an interval from $a$ to $b$, and all parts of this interval
with the same length are equally likely. We write $X\sim U(a,b)$. The
numbers $a$ and $b$ are the parameters of the distribution. The interval
$[a,b]$ is the set of possible values, also called the support. The density
is constant on the support,

$$f_X(x)=\frac{1}{b-a},\qquad a\le x\le b,$$

and zero outside it. The height is not a free choice. The whole rectangle
must have area 1:

$$\text{width}\times\text{height}=(b-a)\times\frac{1}{b-a}=1.$$

Because the height is constant, the probability of an interval depends only
on its length. For $a\le c<d\le b$,

$$P(c<X<d)=\frac{d-c}{b-a}.$$

The mean, the variance, and the standard deviation are stated here without
derivation:

$$E[X]=\frac{a+b}{2},\qquad \operatorname{Var}(X)=\frac{(b-a)^2}{12},
\qquad \operatorname{SD}(X)=\frac{b-a}{\sqrt{12}}.$$

::: {#exm-b06-uniform-session}
## Worked example: when does the pitch appear?

This is an invented example. A recording session lasts 90 minutes, and a
pitch is equally likely to start at any moment of it. Model the start time
as $X\sim U(0,90)$, measured in minutes.

*Step 1: the height.* The density is $1/(90-0)=1/90\approx 0.011$ per
minute. The total area of 1 fixes this height.

*Step 2: the event.* "The pitch starts during the 18-minute window from
minute 30 to minute 48" is the event $30<X<48$.

*Step 3: the area.* The probability is the area of the shaded rectangle in
@fig-b06-uniform-session:

$$P(30<X<48)=\frac{48-30}{90-0}=\frac{18}{90}=0.20.$$

For a uniform model, only the length of the window matters: any 18-minute
window inside the session has probability 0.20.
:::

```{r}
#| label: fig-b06-uniform-session
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The uniform density for a pitch start time over a 90-minute session. The navy line is the density. The shaded green area is the probability of the 18-minute window, 18/90 = 0.20."
library(ggplot2)

a <- 0
b <- 90
c_cut <- 30
d_cut <- 48
density_height <- 1 / (b - a)

support_rect <- data.frame(x = c(a, b), y = density_height)
window_rect  <- data.frame(x = c(c_cut, d_cut), y = density_height)
density_line <- data.frame(x = c(-9, a, a, b, b, 99),
                           y = c(0, 0, density_height, density_height, 0, 0))

ggplot() +
  geom_area(data = support_rect, aes(x, y), fill = "grey75", alpha = 0.25) +
  geom_area(data = window_rect, aes(x, y), fill = "#009E73", alpha = 0.25) +
  geom_line(data = density_line, aes(x, y), colour = "#17324D",
            linewidth = 1.2) +
  annotate("text", x = 66, y = 0.0127,
           label = "density = 1/(b - a) = 1/90",
           colour = "#17324D", size = 3.4) +
  annotate("segment", x = 39, xend = 39, y = 0.0042, yend = 0.0010,
           colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = 39, y = 0.0052,
           label = "area = P(30 < X < 48) = 0.20",
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = c(a, c_cut, d_cut, b),
                     labels = c("a = 0", "c = 30", "d = 48", "b = 90"),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = c(0, density_height),
                     labels = c("0", "1/90"),
                     limits = c(0, 0.0145),
                     expand = c(0, 0)) +
  labs(x = "Minutes into the session", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read the horizontal axis as possible start times and the vertical axis as
density, not probability. The navy density line is at 0 before minute 0,
jumps to $1/90$ on the support, and drops back to 0 after minute 90. The
green probability is an area: $18\times(1/90)=0.20$. Compare this with the
check above: here the height is far below 1, there it was 2. Both densities
are valid, because only the area is a probability.

#### The CDF of the session model

For the session model, the area to the left of $x$ is a rectangle of width
$x$ and height $1/90$. So the CDF is $F_X(x)=x/90$ for $0\le x\le 90$. It is
0 before minute 0 and 1 after minute 90. @fig-b06-uniform-cdf shows it. The
probability of the window is the difference of two CDF values:

$$P(30<X<48)=F_X(48)-F_X(30)=\frac{48}{90}-\frac{30}{90}=0.20.$$

This is the same 0.20 as the green area in @fig-b06-uniform-session.

```{r}
#| label: fig-b06-uniform-cdf
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The CDF of the pitch start time in the 90-minute session model. The navy line rises from 0 to 1 and never falls. The gold dashed lines read off F(30) and F(48). Their difference, the green bar, is the probability of the 18-minute window, 0.20."
library(ggplot2)

a <- 0
b <- 90
c_cut <- 30
d_cut <- 48
F_c <- (c_cut - a) / (b - a)
F_d <- (d_cut - a) / (b - a)

cdf_line <- data.frame(x = c(-9, a, b, 99), y = c(0, 0, 1, 1))

ggplot() +
  annotate("segment", x = c(c_cut, d_cut), xend = c(c_cut, d_cut),
           y = 0, yend = c(F_c, F_d),
           linetype = "dashed", colour = "#C98B2E", linewidth = 0.6) +
  annotate("segment", x = -9, xend = c(c_cut, d_cut),
           y = c(F_c, F_d), yend = c(F_c, F_d),
           linetype = "dashed", colour = "#C98B2E", linewidth = 0.6) +
  geom_line(data = cdf_line, aes(x, y), colour = "#17324D",
            linewidth = 1.2) +
  annotate("segment", x = 6, xend = 6, y = F_c, yend = F_d,
           colour = "#009E73", linewidth = 1.6) +
  annotate("text", x = 8, y = (F_c + F_d) / 2,
           label = "difference = 0.20", colour = "#17324D",
           size = 3.4, hjust = 0) +
  annotate("text", x = 66, y = 0.58, label = "F(x) = x/90",
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_x_continuous(breaks = c(a, c_cut, d_cut, b), expand = c(0, 0)) +
  scale_y_continuous(breaks = c(0, F_c, F_d, 1),
                     labels = c("0", "F(30) = 0.33", "F(48) = 0.53", "1"),
                     limits = c(0, 1.05), expand = c(0, 0)) +
  labs(x = "Minutes into the session (x)",
       y = "Cumulative probability F(x)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

::: {.callout-caution collapse="true"}
## Try it: the final stretch of the session

Try this with your neighbour first. Same model, $X\sim U(0,90)$. What is the
probability that the pitch starts in the final 27 minutes of the session?

*Step 1 (given).* The event is $63<X<90$, an interval of length
$90-63=27$ minutes inside the support.

*Step 2 (you).* Divide the interval length by the support length.

*Step 3 (you).* Interpret the result in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(63<X<90)=27/90=0.30$: under the uniform model there is a 30% chance the
pitch starts in the final 27 minutes, larger than the 0.20 above only
because the window is longer.
:::
:::

### The normal distribution

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/normal-random-variables/">08.8 Normal Random Variables</a>.</p>

Many quantities are the sum of many small influences: body height,
measurement errors, the weight of a filled package. Such quantities often
have a bell-shaped distribution, with most values near the middle and few
values far away on either side. The normal distribution is the model for
this shape. It is a model, not a law. Many business variables, such as
revenues, are strongly skewed and do not follow it.

A normal model needs two numbers. The first is its mean $\mu$. The second is
its standard deviation $\sigma$. Lecture 3 introduced both symbols.

::: {.definition-box}
<span class="label">Definition</span>

**Normal distribution.** A continuous distribution with a symmetric,
bell-shaped density. Its centre is the mean $\mu$, and its spread is the
standard deviation $\sigma>0$. We write $X\sim N(\mu,\sigma^2)$.
:::

Read the notation with care: the second number in the brackets is the
variance $\sigma^2$, not the standard deviation. For example, $N(50,100)$
has $\sigma=\sqrt{100}=10$. In symbols,

$$E[X]=\mu,\qquad \operatorname{Var}(X)=\sigma^2,
\qquad \operatorname{SD}(X)=\sigma.$$

The two parameters do different jobs, as @fig-b06-normal-parameters shows.
A change in $\mu$ shifts the whole curve to the left or to the right, and
the shape stays the same. A change in $\sigma$ makes the curve wider or
narrower. A wider curve is also lower, because the total area stays 1.

```{r}
#| label: fig-b06-normal-parameters
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The two parameters of the normal model. Left: a larger mean (orange, μ = 3 in place of 0) shifts the curve to the right and leaves its shape unchanged. Right: a larger standard deviation (orange, σ = 2 in place of 1) makes the curve wider and lower."
library(ggplot2)

x_grid <- seq(-6, 8, length.out = 600)
panel_shift <- "Changing the mean shifts the curve"
panel_width <- "Changing the SD widens the curve"

curves <- rbind(
  data.frame(panel = panel_shift, curve = "first", x = x_grid,
             density = dnorm(x_grid, mean = 0, sd = 1)),
  data.frame(panel = panel_shift, curve = "second", x = x_grid,
             density = dnorm(x_grid, mean = 3, sd = 1)),
  data.frame(panel = panel_width, curve = "first", x = x_grid,
             density = dnorm(x_grid, mean = 0, sd = 1)),
  data.frame(panel = panel_width, curve = "second", x = x_grid,
             density = dnorm(x_grid, mean = 0, sd = 2))
)
curves$panel <- factor(curves$panel, levels = c(panel_shift, panel_width))

curve_labels <- data.frame(
  panel = factor(c(panel_shift, panel_shift, panel_width, panel_width),
                 levels = c(panel_shift, panel_width)),
  curve = c("first", "second", "first", "second"),
  x = c(-1.2, 4.2, 1.1, 3.3),
  y = c(0.36, 0.36, 0.36, 0.12),
  label = c("mu == 0*','~sigma == 1", "mu == 3*','~sigma == 1",
            "mu == 0*','~sigma == 1", "mu == 0*','~sigma == 2"),
  hjust = c(1, 0, 0, 0)
)

ggplot(curves, aes(x, density, colour = curve)) +
  geom_line(linewidth = 1.1) +
  geom_text(data = curve_labels, aes(x, y, label = label, hjust = hjust),
            parse = TRUE, size = 3.4) +
  facet_wrap(~panel) +
  scale_colour_manual(values = c(first = "#009E73", second = "#D55E00"),
                      guide = "none") +
  scale_x_continuous(breaks = seq(-6, 8, by = 2)) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of X", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", hjust = 0)
  )
```

@fig-b06-normal-shape shows where the probability of a normal model lies.

```{r}
#| label: fig-b06-normal-shape
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The normal density. The shaded central area between the two grey dashed lines runs from one standard deviation below the mean to one above and contains about 68% of the distribution. The gold dashed line marks the mean μ."
library(ggplot2)

normal_curve <- data.frame(z = seq(-3.6, 3.6, length.out = 800))
normal_curve$density <- dnorm(normal_curve$z)

ggplot(normal_curve, aes(z, density)) +
  geom_area(data = subset(normal_curve, z >= -1 & z <= 1),
            fill = "#009E73", alpha = 0.25) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.16, label = "about 68%",
           colour = "#17324D", size = 3.4) +
  annotate("text", x = 0.15, y = 0.425, label = "mean~mu", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_x_continuous(breaks = c(-2, -1, 0, 1, 2),
                     labels = expression(mu - 2 * sigma, mu - sigma, mu,
                                         mu + sigma, mu + 2 * sigma)) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of X", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The horizontal axis locates possible values relative to $\mu$ and $\sigma$.
The vertical axis is density, and probability is again area. In the normal
model, about 68% of the distribution lies within one standard deviation of
the mean and about 95% within 1.96 standard deviations. The value 1.96 is
explained later in this lecture. These are facts about the mathematical
model. The caveat box below explains why they are not automatic facts about
every dataset.

#### The opening question as an area

Return to the offer of 30%. Take the normal model that has the mean and the
standard deviation of the equity offered in the Shark Tank file,
$X\sim N(13.2,\,8.4^2)$. In this model, the probability of an offer of 30%
or more is the area under the curve to the right of 30.
@fig-b06-equity-normal-area shows that this area is small.

```{r}
#| label: fig-b06-equity-normal-area
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The normal model with the mean and the standard deviation of the equity offered (`equity_offered_pct`) in the Shark Tank file. The gold dashed line marks the mean. The orange area to the right of 30% is the probability that this model gives to an offer of 30% or more, about 2%."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mean_equity <- mean(sharks$equity_offered_pct)
sd_equity <- sd(sharks$equity_offered_pct)

model_curve <- data.frame(x = seq(-15, 45, length.out = 601))
model_curve$density <- dnorm(model_curve$x, mean = mean_equity,
                             sd = sd_equity)

ggplot(model_curve, aes(x, density)) +
  geom_area(data = subset(model_curve, x >= 30),
            fill = "#D55E00", alpha = 0.55) +
  geom_vline(xintercept = mean_equity, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = mean_equity + 0.8, y = 0.0495, label = "mean 13.2%",
           colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("segment", x = 35, xend = 32.2, y = 0.012, yend = 0.003,
           colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = 30.5, y = 0.0168, label = "area == P(X >= 30)",
           parse = TRUE, colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("text", x = 30.5, y = 0.0138, label = "about 2% in this model",
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_x_continuous(breaks = seq(-10, 40, by = 10)) +
  scale_y_continuous(limits = c(0, 0.052),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The figure shows the answer of the model, not the answer of the data. The
curve also gives some area to values below 0%, which no offer can have. The
check of the model later in this lecture returns to both points. To turn the
orange area into a number, you need one more tool: standardising.

#### Standardising: every normal question becomes a Z question

::: {.definition-box}
<span class="label">Definition</span>

**z-score.** The z-score of a value $x$ is $z=\dfrac{x-\mu}{\sigma}$. It
says how many standard deviations $x$ lies above the mean (positive $z$) or
below the mean (negative $z$). Calculating a z-score is called
**standardising**.
:::

If $X\sim N(\mu,\sigma^2)$, then the standardised variable is again normal,
with mean 0 and standard deviation 1:

$$Z=\frac{X-\mu}{\sigma}\sim N(0,1).$$

The distribution $N(0,1)$ is called the **standard normal distribution**.
The mean 0 and the standard deviation 1 follow from the rules for shifting
and rescaling in
[Lecture 3](b03-discrete-distributions.qmd#b03-shift-rescale).

Standardising helps because one set of standard-normal values then serves
every normal model. Whatever $\mu$ and $\sigma$ are, the question "what is
$P(X\le x)$?" becomes the question "what is $P(Z\le z)$?", with
$z=(x-\mu)/\sigma$.

Two general rules turn one value $P(Z\le z)$ into further probabilities.
Both hold for every $z$.

- Complement rule. The total area under the curve is 1, so
  $P(Z>z)=1-P(Z\le z)$.
- Symmetry rule. The curve is symmetric around 0, so
  $P(Z\le -z)=P(Z>z)$.

::: {#exm-b06-supplied-z}
## Worked example: normal probabilities from supplied Z values

This is an invented example. Processing time $Y$ (in minutes) is modelled as
$Y\sim N(50,100)$. The variance is 100, so $\sigma=\sqrt{100}=10$. In the
exam, the standard-normal values that you need are printed in the question,
as here:

```
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
P(Z <= -1.0) = 0.1587
```

**(a) $P(Y\le 65)$.** Standardise the boundary:

$$z=\frac{65-50}{10}=1.5,$$

so

$$P(Y\le 65)=P(Z\le 1.5)=0.9332.$$

**(b) $P(40<Y<60)$.** Standardise both boundaries:

$$z_{\text{low}}=\frac{40-50}{10}=-1.0,\qquad
z_{\text{high}}=\frac{60-50}{10}=1.0,$$

then take the difference of the two cumulative probabilities, exactly as
with any CDF:

$$P(40<Y<60)=P(Z\le 1.0)-P(Z\le -1.0)=0.8413-0.1587=0.6826.$$

That is the 68% rule seen in @fig-b06-normal-shape: the interval runs from
one standard deviation below the mean to one above.

**(c) $P(Y>60)$ and $P(Y\le 35)$.** These two need the rules. The z-score of
60 is 1.0, and the complement rule gives

$$P(Y>60)=P(Z>1.0)=1-P(Z\le 1.0)=1-0.8413=0.1587.$$

The z-score of 35 is $(35-50)/10=-1.5$. No value for $-1.5$ is supplied, so
use the symmetry rule first and then the complement rule:

$$P(Y\le 35)=P(Z\le -1.5)=P(Z>1.5)=1-0.9332=0.0668.$$

The supplied values also show the symmetry rule at work:
$P(Z\le -1.0)=0.1587$ is the same number as $P(Z>1.0)$.
:::

`pnorm()` is the normal CDF in R. Give it the boundary, the mean, and the
standard deviation (not the variance). The four lines below repeat the four
results of the worked example:

```{r}
#| label: b06-pnorm-check
pnorm(65, mean = 50, sd = 10)
pnorm(60, mean = 50, sd = 10) - pnorm(40, mean = 50, sd = 10)
1 - pnorm(60, mean = 50, sd = 10)
pnorm(35, mean = 50, sd = 10)
```

The hand calculation and `pnorm()` agree. Small differences in the fourth
decimal come from the rounding of the supplied values.

#### The answer of the model to the opening question

Now the orange area in @fig-b06-equity-normal-area can be calculated. The
model is $X\sim N(13.2,\,8.4^2)$ and the boundary is 30:

$$z=\frac{30-13.2}{8.4}=2.0.$$

An offer of 30% lies two standard deviations above the mean. The supplied
value is $P(Z\le 2.0)=0.9772$, and the complement rule gives

$$P(X\ge 30)=P(Z\ge 2.0)=1-P(Z\le 2.0)=1-0.9772=0.0228.$$

For a continuous variable, $\ge$ and $>$ give the same probability. R
confirms the result:

```{r}
#| label: b06-opening-pnorm
pnorm(2)
1 - pnorm(2)
1 - pnorm(30, mean = 13.2, sd = 8.4)
```

The normal model says that about 2% of pitches offer 30% or more. The count
in the file gave about 5%. The section on checking the model explains the
difference.

#### The reverse direction: from a probability to a value

`pnorm()` starts from a value and returns the probability at or below it.
The reverse question starts from the probability: which value has 90% of
the probability at or below it? Such a value is called a quantile.

::: {.definition-box}
<span class="label">Definition</span>

**Quantile and percentile.** The 0.90 quantile of a distribution is the
value with 90% of the probability at or below it. It is also called the
90th percentile. In the same way, the 0.25 quantile is the 25th percentile.
For the standard normal we write $z_{0.90}$, and R gives it with
`qnorm(0.90)`.
:::

For the standard normal, $z_{0.90}=1.2816$: 90% of the area lies at or below
1.2816. For any other normal model, solve $z=(x-\mu)/\sigma$ for $x$:

$$x=\mu+z\sigma,$$

where $z$ is the standard-normal quantile for the chosen proportion.

::: {#exm-b06-reverse}
## Worked example: the 90th percentile of processing time

Take the invented processing-time model $Y\sim N(50,100)$ again. Which
processing time is exceeded by only 10% of cases? This is the 90th
percentile of $Y$. The supplied standard-normal quantile is
$z_{0.90}=1.2816$.

*Step 1.* The 90th percentile of $Z$ is 1.2816. The value you look for
therefore lies 1.2816 standard deviations above the mean.

*Step 2.* Translate back to the scale of $Y$:

$$x=\mu+z\sigma=50+1.2816\times 10=62.8 \text{ minutes.}$$

*Step 3.* Verify with `qnorm()`, the reverse of `pnorm()`:

```{r}
#| label: b06-qnorm-check
qnorm(0.90, mean = 50, sd = 10)
```

About 62.8 minutes: 90% of processing times fall at or below it, 10%
above.
:::

#### The central 95% of the standard normal

One pair of quantiles is used again and again in later lectures. Which two
values enclose the central 95% of the standard normal? If 95% of the
probability lies in the middle, 5% is left for the two tails. By symmetry,
each tail holds 2.5%. The upper boundary therefore has 97.5% of the
probability at or below it. It is the 0.975 quantile. The lower boundary is
the 0.025 quantile.

```{r}
#| label: b06-qnorm-central
qnorm(c(0.025, 0.975))
```

So $z_{0.025}=-1.96$ and $z_{0.975}=1.96$. In the standard normal, 95% of the
probability lies between $-1.96$ and $1.96$, with 2.5% in each tail, as
@fig-b06-central-95 shows.

```{r}
#| label: fig-b06-central-95
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The standard normal density. The green area between -1.96 and 1.96 holds 95% of the probability. Each orange tail holds 2.5%."
library(ggplot2)

z_cut <- qnorm(0.975)
z_curve <- data.frame(z = seq(-3.6, 3.6, length.out = 800))
z_curve$density <- dnorm(z_curve$z)
centre <- data.frame(z = seq(-z_cut, z_cut, length.out = 400))
centre$density <- dnorm(centre$z)
left_tail <- data.frame(z = seq(-3.6, -z_cut, length.out = 200))
left_tail$density <- dnorm(left_tail$z)
right_tail <- data.frame(z = seq(z_cut, 3.6, length.out = 200))
right_tail$density <- dnorm(right_tail$z)

ggplot(z_curve, aes(z, density)) +
  geom_area(data = centre, fill = "#009E73", alpha = 0.25) +
  geom_area(data = left_tail, fill = "#D55E00", alpha = 0.55) +
  geom_area(data = right_tail, fill = "#D55E00", alpha = 0.55) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.16, label = "95% of the area",
           colour = "#17324D", size = 3.4) +
  annotate("segment", x = c(-2.9, 2.9), xend = c(-2.3, 2.3),
           y = 0.075, yend = 0.02, colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = c(-2.9, 2.9), y = 0.09, label = "2.5%",
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = c(-z_cut, 0, z_cut),
                     labels = c("-1.96", "0", "1.96")) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of Z", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

With $x=\mu+z\sigma$ the same holds for every normal model: 95% of the
probability lies between $\mu-1.96\sigma$ and $\mu+1.96\sigma$. Lectures 9
and 10 build on this. Lecture 10 uses the value 1.96 in confidence intervals
for proportions. Lecture 9 uses slightly larger values from the t
distribution in its place.

#### Sums of normal variables {#b06-normal-sums}

A sum of independent normal variables is again normal. Its mean and its
variance follow from the sum rules of
[Lecture 5](b05-joint-covariance-correlation.qmd#b05-sums). Lecture 8 uses
this fact for the sample mean.

### Checking the normal model against data

The normal distribution is a model, not a label that every quantitative
variable automatically deserves. Skewness, bounds, extreme outliers, or a
few heavily repeated values can make it a poor description. The check is
visual: draw the histogram of the data, overlay the normal curve with the
same mean and standard deviation, and see whether the curve reproduces the
bars. @fig-b06-equity-normal-check does this for the equity offered
(`equity_offered_pct`) in the Shark Tank file, with the mean of 13.2% and
the standard deviation of 8.4 percentage points from the start of this
lecture.

```{r}
#| label: fig-b06-equity-normal-check
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Initial equity offered across all 1,441 recorded pitches (green bars), with the normal density that has the same mean and standard deviation (orange curve). The curve misses the skew, the bounds, and the spikes at round percentages."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mean_equity <- mean(sharks$equity_offered_pct)
sd_equity <- sd(sharks$equity_offered_pct)

ggplot(sharks, aes(equity_offered_pct)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = seq(0, 100, by = 5),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  stat_function(fun = dnorm,
                args = list(mean = mean_equity, sd = sd_equity),
                colour = "#D55E00", linewidth = 0.9) +
  annotate("text", x = 42, y = 0.062, label = "observed pitches",
           colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("text", x = 42, y = 0.054,
           label = "normal curve with the same mean and SD",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Equity initially offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-b06-equity-normal-check with the five-step sequence from
Lecture 1. *Unit and variables:* one observation is one pitch. The horizontal
axis is the equity offered, in percentage points. *Scales:* the vertical axis
is density, so the areas of the bars, not their heights alone, represent
proportions of pitches. *One element:* the green bars are the 1,441 observed
offers. The orange curve is $N(13.2,\,8.4^2)$, the normal model with the same
mean and SD. *Pattern:* the bars pile up between 5% and 20% and stretch out
a long right tail. The data are bounded between 0% and 100%, strongly
right-skewed, and spiked at round values such as 10% and 20%. The symmetric
curve misses all three features. *Limits:* the figure rejects the normal
model for single equity offers. It says nothing yet about averages of
offers. A model is a choice, and this check rejects the choice.

The opening question shows what happens without the check. The normal model
gave $P(X\ge 30)=0.0228$, about 2%. The data give 76 of 1,441, about 5%. The
model misses the long right tail, and the offers of 30% or more lie in that
tail.

The normal model stays useful all the same. From Lecture 8 onwards it
describes the sampling distribution of an average, that is, how an average
varies from sample to sample. Lectures 7 and 8 explain this idea. The normal
model fits those distributions well, even when the underlying variable is as
skewed as this one. That result is the central limit theorem. It is the main
reason why the normal model is used so much in statistics.

### Quantiles of a model distribution

Lecture 1 located quantiles of data: the median, the quartiles $Q_1$ and
$Q_3$, and the interquartile range of the equity offered (see
[Lecture 1](b01-introduction-data-analysis.qmd#b01-quartiles)). A model
distribution has the same quantiles. The median is the 0.50 quantile, and
the quartiles are the 0.25 and the 0.75 quantiles.

- **Uniform:** by symmetry, the median of $U(a,b)$ is the midpoint
  $(a+b)/2$. For the session model $U(0,90)$, the median start time is 45
  minutes: half of the probability lies on each side.
- **Normal:** by symmetry, the median equals the mean $\mu$. The quartiles
  lie $0.6745$ standard deviations on either side of the mean,
  $\mu\pm 0.6745\sigma$. For the processing-time model $Y\sim N(50,100)$:

```{r}
#| label: b06-model-quantiles
qnorm(c(0.25, 0.75), mean = 50, sd = 10)
```

The middle half of this model distribution runs from about 43.3 to 56.7
minutes. The width of this range is the interquartile range of the model.

::: {.caveat-box}
<span class="label">What a normal label can and cannot say</span>

- "Normally distributed" is a model, not a fact. It has to be checked against
  a histogram.
- The 68% and 95% rules hold for the model. For skewed or bounded data they
  give wrong numbers, as with the 30% offers.
- A curve with the right mean and standard deviation can still have the wrong
  shape.

**What you can do with it:** use the normal model for quick answers when the
histogram looks roughly bell-shaped, and count in the data when it does not.
From Lecture 8 on, the normal model describes averages, where it works well.
:::

### Answering the driving question

An offer of 30% or more is unusual: about 1 pitch in 20 makes one. In the
Shark Tank file, 76 of the 1,441 pitches did, about 5%. The normal model
with the same mean and standard deviation says about 2%, because it ignores
the long right tail of the offers. So a model with two numbers can answer such a question
without counting every pitch, but only when the model fits. The histogram
check tells you whether it does.

The tools behind this answer are general. For a continuous variable, no
single value has positive probability, $P(X=x)=0$. Probabilities belong to
intervals, and they are areas under a density. For a uniform model, only the
length of the interval matters: the 18-minute window in the session model
has probability $18/90=0.20$. For a normal model, you standardise and use
the supplied $Z$ values. In the processing-time model $N(50,100)$ this gave
$P(Y\le 65)=0.9332$ and $P(40<Y<60)=0.6826$, and `pnorm()` confirmed both. A
density is a modelling choice, so check the model before you trust its
areas.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your delivery times look roughly normal, with a mean of 28 minutes and a standard deviation of 6 minutes. You want to promise a delivery time that 90% of orders meet. The 90th percentile is 28 + 1.2816 × 6 = 35.7 minutes. The decision: promise 36 minutes on the website. First check the histogram of your delivery times. If a few deliveries go very wrong, the normal model is too optimistic, and you promise the 90th percentile of the actual times instead.
:::

**Next:** So far the 1,441 pitches were treated as the whole world. In
practice a firm rarely sees the whole world. It sees a sample: 100 audited
transactions, 500 surveyed customers. Lecture 7 moves from one pitch to a
sample of pitches. It covers populations, random sampling, and the
difference between a parameter, an estimator, and an estimate.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 75 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Why is $P(X=x)=0$ for a continuous random variable?
2. What quantity under a PDF represents probability?
3. How do you obtain $P(a<X\le b)$ from a CDF?
4. What do $\mu$ and $\sigma$ control in a normal model?
5. A density curve reaches a height of 2. Why is that not a contradiction?
6. What does a z-score of $-1.5$ mean?

::: {.callout-note collapse="true"}
## Check your answers

1. Probability belongs to intervals. An individual point has zero width and
   therefore zero area.
2. Area under the density curve between the interval boundaries.
3. $F_X(b)-F_X(a)$.
4. $\mu$ controls the centre and $\sigma$ the spread.
5. Density is probability per unit, not probability. A height above 1 is
   fine as long as the total area under the curve is 1.
6. The value lies 1.5 standard deviations below the mean.
:::

### Finish a half-worked example (5 minutes)

Re-read the session-timing calculation (@exm-b06-uniform-session). Then
complete the same calculation for a new model: a waiting time
$X\sim U(10,30)$ minutes.

*Step 1 (given).* The density height is $1/(30-10)=0.05$ on the support
$[10,30]$ and zero outside it.

*Step 2 (you).* Find $P(14<X<20)$.

*Step 3 (you).* State $E[X]$ and interpret both numbers in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(14<X<20)=(20-14)/(30-10)=6/20=0.30$ and $E[X]=(10+30)/2=20$: the wait
averages 20 minutes, and under the uniform model there is a 30% chance it
lands between 14 and 20 minutes.
:::

### Practice (15 minutes)

For items 3 and 4, use only these supplied values, then verify in R:

```
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
P(Z <= -1.0) = 0.1587
z(0.90)      = 1.2816
```

1. For $X\sim U(10,30)$: find $\operatorname{SD}(X)$ and $P(X>25)$.
2. A supplier delivers at a moment uniformly distributed between 09:00 and
   13:00. Your goods lift is reserved for other tenants except during
   12:00--13:00. What is the probability the delivery arrives while the
   lift is available, and what is the density height per minute?
3. Monthly demand for a product is modelled as $Y\sim N(200,400)$.
   Calculate $P(Y\le 230)$, $P(180<Y<220)$, $P(Y>230)$, and $P(Y\le 170)$
   by hand.
4. Weekly revenue is modelled as $N(100,15^2)$ (in hundreds of euros).
   Which revenue level is exceeded in only 10% of weeks? Compute by hand
   with $x=\mu+z\sigma$, then verify with `qnorm()`.
5. Explain in two sentences why the normal curve in
   @fig-b06-equity-normal-check is not a convincing model for offered
   equity.

::: {.callout-note collapse="true"}
## Check your answers

1. $\operatorname{SD}(X)=\sqrt{(30-10)^2/12}=20/\sqrt{12}\approx 5.77$
   minutes, and $P(X>25)=(30-25)/20=0.25$.
2. The window is 60 of the 240 possible minutes, so $60/240=0.25$. The
   density height is $1/240\approx 0.0042$ per minute, and the
   probability is the area $60\times(1/240)=0.25$, not the height.
3. $\sigma=\sqrt{400}=20$. For 230: $z=(230-200)/20=1.5$, so
   $P(Y\le 230)=0.9332$. For the interval:
   $z_{\text{low}}=(180-200)/20=-1.0$ and
   $z_{\text{high}}=(220-200)/20=1.0$, so
   $P(180<Y<220)=0.8413-0.1587=0.6826$. The complement rule gives
   $P(Y>230)=1-0.9332=0.0668$. For 170: $z=(170-200)/20=-1.5$, and the
   symmetry rule gives $P(Y\le 170)=P(Z\le -1.5)=P(Z>1.5)=0.0668$.
4. $x=100+1.2816\times 15=119.2$, so about 119 hundred euros, that is about
   11,900 euros. Only 10% of weeks exceed it.
5. The observed distribution is bounded between 0% and 100%, strongly
   right-skewed, and concentrated at a few round percentages. The symmetric
   normal curve reproduces none of those features.

Verification in R for items 3 and 4:

```{r}
#| label: b06-practice-check
pnorm(230, mean = 200, sd = 20)
pnorm(220, mean = 200, sd = 20) - pnorm(180, mean = 200, sd = 20)
1 - pnorm(230, mean = 200, sd = 20)
pnorm(170, mean = 200, sd = 20)
qnorm(0.90, mean = 100, sd = 15)
```
:::

### Review: one question from Lecture 3 (5 minutes)

The number of sharks who join an on-air deal, $X$, has PMF $p(1)=0.6$,
$p(2)=0.3$, $p(3)=0.1$. Find $F(2)=P(X\le 2)$ and $P(X<2)$. Why do the two
answers differ here, when for a continuous variable they would not?

::: {.callout-note collapse="true"}
## Check your answer

$F(2)=0.6+0.3=0.9$, while $P(X<2)=P(X=1)=0.6$. They differ because the
single point $X=2$ carries positive probability, $p(2)=0.3$. For a
continuous variable every single point has probability zero, so $<$ and
$\le$ give the same answer.
:::

### Changing units (5 minutes)

The session-timing variable $X\sim U(0,90)$ is recorded in minutes and then
converted to seconds. Explain what changes about the PDF and what remains
unchanged.

::: {.callout-note collapse="true"}
## Check your answer

The support stretches from $[0,90]$ minutes to $[0,5400]$ seconds, and the
density height drops from $1/90$ per minute to $1/5400$ per second: the
numerical size of a density depends on the units. Probabilities of
corresponding time intervals are unchanged, because they are areas: the
18-minute window becomes a 1,080-second window, and
$1080/5400=18/90=0.20$.
:::

### Exam-style item (10 minutes)

This item follows the pattern of Mock Exam Question 3. Let
$X\sim U(20,60)$ be a waiting time in minutes and, separately,
$Y\sim N(80,25)$ a processing time in minutes. Supplied:

```
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
P(Z <= -1.0) = 0.1587
```

1. State the PDF height of $X$, including its value outside the support.
2. Calculate $P(30<X<40)$, $E[X]$, and $\operatorname{Var}(X)$.
3. Standardise 87.5 minutes under the model for $Y$, then calculate
   $P(Y\le 87.5)$.
4. Calculate $P(75<Y<85)$.
5. Explain why a PDF height is not itself a probability.

::: {.callout-note collapse="true"}
## Check a model answer

1. $f_X(x)=1/(60-20)=0.025$ for $20\le x\le 60$, and $f_X(x)=0$ outside
   that interval.
2. $P(30<X<40)=(40-30)/40=0.25$, $E[X]=(20+60)/2=40$, and
   $\operatorname{Var}(X)=(60-20)^2/12\approx 133.33$ square minutes.
3. The variance is 25, so $\sigma=5$ and $z=(87.5-80)/5=1.5$. Hence
   $P(Y\le 87.5)=0.9332$.
4. $P(75<Y<85)=P(-1<Z<1)=0.8413-0.1587=0.6826$.
5. Probability is area over an interval: approximately density times a
   small interval width. A single point has zero width and hence zero
   probability. A density value only becomes a probability once multiplied
   by a width.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "Equity offers have a mean of 13.2% and a standard deviation of 8.4%. How many pitches offer 30% or more?" It answered:

> "With the normal distribution, 30% is 2.0 standard deviations above the mean, so about 2.3% of pitches offer 30% or more."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

The calculation is correct for a normal model, but the model does not fit. Equity offers are strongly right-skewed and pile up at round numbers. In the data, 76 of 1,441 pitches offer 30% or more, about 5.3%, more than twice the answer of the AI. Check the histogram before you accept an answer that rests on a normal model.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Pick one
continuous quantity that matters for it (a delivery time, weekly revenue,
time on site) and write two sentences: first name the interval of values
that would trigger a real decision, then say whether a uniform or a normal
model seems defensible for that quantity, and which feature of its
histogram you would check first. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Delivery time is my quantity: anything over 40 minutes triggers a refund, so $P(T > 40)$ is the number that matters. A normal model is doubtful because times cannot be negative and a few deliveries go very wrong: right skew is the first thing I would check on the histogram."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to propose a
distribution for your quantity and then challenge whether the normal
model's assumptions actually hold for it.
::::

[Continue to Lecture 7: Samples, parameters, and estimators](b07-samples-estimators.qmd).

[Formula sheet: Lecture 6 formulas](formula-sheet.qmd#formula-l6).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 10.

:::
