# Lecture 6: Continuous distributions and the normal model {#b06}

```{r}
#| label: b06-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 6, you can</span>

- tell continuous variables apart from discrete ones and read probabilities as areas
- calculate probabilities for a uniform model from interval lengths
- compute normal probabilities and quantiles by hand and in R
- check a normal model against data and say when it fails
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Not everything comes in whole numbers: equity shares, waiting times, and revenues can take any value in a range. Insurers, banks, and operations planners calculate risks for such quantities every day. For this they need a new kind of probability model.</p>

You plan to offer the sharks 30% of your company. How unusual is that, and
can a model with two numbers answer the question without counting every
pitch?
:::

::: {.course-phase .phase-before}

## Before class

Do all the preparation below before class.

<div class="video-prep">
<span class="label">Video preparation</span><br>
Watch <a href="https://www.youtube.com/watch?v=8QFpZ3FndBc">Probability
Density Functions</a>,
<a href="https://www.youtube.com/watch?v=jPB9zI8F7rE">Uniform and
Piecewise Constant PDFs</a>, and
<a href="https://www.youtube.com/watch?v=6UMv4vb4y7c">Normal Random
Variables</a>. Focus on the difference between density height and
probability, and on the roles of the two normal parameters. All links are
also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
The videos explain key ideas in another way and can use a different
notation. The lecture material and its notation always come first.
</div>

**Warm up.** A discrete random variable has separate
possible values with gaps between them (Lecture 3 attached a PMF and a CDF to such variables). A
continuous random variable can take every real value in an interval. Match
each variable to the better description:

1. number of sharks in an on-air deal
2. equity offered, as an exact percentage of the company
3. season number
4. exact amount invested, before rounding to dollars.

::: {.callout-note collapse="true"}
## Check your preparation

Variables 1 and 3 are discrete. Variables 2 and 4 are continuous in a
mathematical model, even if a measuring device or database records rounded
values.
:::

**Predict.** Answer this question before you come to class: for a continuous random variable $X$, is
$P(X = \text{exactly } 10)$ very small or exactly zero? Choose one answer
and write it down. The lecture material gives the correct answer.

If fractions or rearranging an equation feel unfamiliar, review
[Mathematics Refresher: fractions and rearranging equations](../resources/math-refresher.html#fractions-and-rearranging-equations).
The same refresher has short parts on square roots and on inequalities.

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section builds probability models for quantities that can take any
value in a range. It uses them to answer one question about the equity
offered in the Shark Tank file.

### Start with a founder's question

[Shark Tank data]{.shark-tag} You plan to offer the sharks 30% of your company. Is that unusual? The Shark
Tank file records the equity offered in each pitch (`equity_offered_pct`).
The code below reads the file and counts the pitches that offered 30% or
more.

```{r}
#| label: b06-equity-count
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
nrow(sharks)
sum(sharks$equity_offered_pct >= 30)
mean(sharks$equity_offered_pct >= 30)
```

In the file, 76 of the 1,441 pitches offered 30% or more, about 5%. Now
suppose you did not have the file. You only knew two numbers about the
offers: the mean and the standard deviation.

```{r}
#| label: b06-equity-model
# sd() divides by n - 1. With 1,441 records the difference from dividing
# by N is in the third decimal.
c(mean_equity = mean(sharks$equity_offered_pct),
  sd_equity   = sd(sharks$equity_offered_pct))
```

The mean is 13.2% and the standard deviation is 8.4 percentage points. Could
you still say how unusual 30% is? If you ask an AI assistant this question,
it will often use a model called the normal distribution and report about
2%. That is less than half of the share in the data. The section on checking
the normal model against data explains the difference.

### From counting values to measuring intervals

Equity offered is a continuous quantity: an offer $X$ can be any percentage
in a range, for example 29.87%. What is the probability that an offer is
exactly 30%?

Take a narrow interval, a window, around 30. The offer lies between 29.5% and 30.5% with some
probability $c$. The window from 29.95% to 30.05% is ten times narrower.
Over such a short range the probability is spread almost evenly, so this
window holds about $c/10$. Every further decimal divides the probability by
about ten. The value "exactly 30%" lies inside every one of these windows.
So its probability is at most $c$, at most $c/10$, at most $c/100$, and so
on without end. No positive number is below all of these. The probability is
zero. For a continuous random variable,

$$P(X=x)=0 \quad\text{for every single value } x.$$

This answers the prediction question from before class: exactly zero, not
very small. Probability has not disappeared. It belongs to intervals, not to
single values.

Zero probability does not mean impossible. The next founder certainly offers
some exact share of the company. Whatever that share turns out to be, it had
probability zero in advance, and it occurred anyway. A continuous random
variable always takes a value. The result only says that you cannot name a
single value in advance that has a positive probability.

### From probability mass to probability density

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/probability-density-functions/">08.2 Probability Density Functions</a>.</p>

[Shark Tank data]{.shark-tag} For a discrete random variable, the PMF gives a probability to each single
value. For a continuous variable that no longer helps, because every single
value has probability zero. We need something else, and the histogram
already contains the idea.

Imagine a histogram of the equity offered, like the one in
[Lecture 1](b01-introduction-data-analysis.qmd), but for offers that are
recorded exactly, without rounding. The intervals of a histogram are also
called bins. The height of each bar is the share of offers in its bin. With
bins 5 percentage points wide, the bars have a reasonable size. Now make the
bins 1 point wide. Each new bar holds about a fifth of what the old bar
held, so all bars shrink. With bins 0.1 wide they
shrink again, by another factor of ten. The whole histogram sinks towards
the axis. This is the shrinking window of the previous section, now for all
bins at once. A histogram with the share in the bin as its height is not
useful once the bins are narrow.

There is a second way to draw the heights. Divide the share in each bin by
the width of the bin. The height is then the share per percentage point. Now
the picture does not sink when the bins get narrower. A bar 1 point wide
that holds 2% of the offers has the height $0.02/1=0.02$ per point. A bar
0.1 wide that holds 0.2% of the offers has the height $0.002/0.1=0.02$ per
point as well. With narrower bins, the bars stop shrinking and form a
smooth curve, as @fig-b06-density-bins shows. The figure uses bins 5, 1,
and 0.25 wide. This curve is the probability
density, written $f_X(x)$. Its height at 30 is the share per percentage
point near 30. The share in a bin is the height of the bar times its width,
which is the area of the bar. Probability has become area.

The Shark Tank file records rounded percentages, so it cannot show what
happens in very narrow bins. The figure therefore uses 100,000 invented
offers. A computer drew them, without rounding, from a smooth model with the
same mean and standard deviation as the equity offered in the file.

```{r}
#| label: fig-b06-density-bins
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.8
#| fig-cap: "Narrowing the bins, for 100,000 invented offers that are not rounded. The height of a bar is the share of offers in the bin divided by the bin width. With this height the histogram does not sink. It comes closer and closer to the same navy curve, the density."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mean_equity <- mean(sharks$equity_offered_pct)
# sd() divides by n - 1. With 1,441 records the difference from dividing
# by N is in the third decimal.
sd_equity <- sd(sharks$equity_offered_pct)

# 100,000 invented offers that are not rounded. They come from a smooth,
# right-skewed model (a gamma distribution) with the mean and the standard
# deviation of the equity offered in the Shark Tank file.
model_shape <- (mean_equity / sd_equity)^2
model_rate <- mean_equity / sd_equity^2
set.seed(206)
exact_offers <- rgamma(100000, shape = model_shape, rate = model_rate)

panel_names <- c("Bins 5 points wide", "Bins 1 point wide",
                 "Bins 0.25 points wide")
offers_wide <- data.frame(
  panel = factor(panel_names[1], levels = panel_names), x = exact_offers)
offers_mid <- data.frame(
  panel = factor(panel_names[2], levels = panel_names), x = exact_offers)
offers_narrow <- data.frame(
  panel = factor(panel_names[3], levels = panel_names), x = exact_offers)

# The same smooth curve for all three panels: the density of the model.
x_grid <- seq(0, 50, length.out = 400)
smooth_curve <- data.frame(
  panel = factor(rep(panel_names, each = length(x_grid)),
                 levels = panel_names),
  x = rep(x_grid, times = 3),
  density = rep(dgamma(x_grid, shape = model_shape, rate = model_rate),
                times = 3)
)

ggplot() +
  geom_histogram(data = offers_wide, aes(x, after_stat(density)),
                 binwidth = 5, boundary = 0,
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_histogram(data = offers_mid, aes(x, after_stat(density)),
                 binwidth = 1, boundary = 0,
                 fill = "#009E73", alpha = 0.85, colour = "white",
                 linewidth = 0.2) +
  geom_histogram(data = offers_narrow, aes(x, after_stat(density)),
                 binwidth = 0.25, boundary = 0,
                 fill = "#26AD88", colour = "#26AD88", linewidth = 0.1) +
  geom_line(data = smooth_curve, aes(x, density), colour = "#17324D",
            linewidth = 0.9) +
  facet_wrap(~panel) +
  coord_cartesian(xlim = c(0, 50)) +
  scale_x_continuous(breaks = seq(0, 50, by = 10)) +
  scale_y_continuous(limits = c(0, 0.07),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Equity offered (%), invented offers without rounding",
       y = "Density (share per point)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", hjust = 0)
  )
```

A metal rod gives the same picture without data. The rod has a weight per
metre at each point, which is its density, and no single point of the rod
has any weight. The weight of a piece is its density times its length. A
probability density is read in the same way: no single value has
probability, and the probability of an interval is the area under the curve
over that interval.

::: {.definition-box}
<span class="label">Definition</span>

**Probability density function (PDF).** A curve $f_X(x)$ that is never
negative and has a total area of 1 under it. The probability that $X$ lies
between $a$ and $b$ is the area under the curve between $a$ and $b$. The
height $f_X(x)$ is the probability per unit of $x$ near the value $x$. It is
a rate, not a probability.
:::

For example, in @fig-b06-density-bins the probability of an offer between
10% and 20% is the area under the navy curve between 10 and 20.

Three rules for reading a density follow. The histogram helps you to
remember them.

1. The height of the curve is a rate, the share per percentage point. It is
   not a probability, just as the height of a bar in the figure is not the
   share in the bin.
2. The area over an interval is a probability, just as the area of a bar is
   the share in its bin.
3. A rate can be high where the probability is packed into a narrow range.
   So a density can exceed 1, as long as the total area stays 1. A bar 0.2
   points wide that holds 30% of the data has the height $0.30/0.2=1.5$.

The Shark Tank file records the equity offered in rounded percentages, so
values repeat.

```{r}
#| label: b06-equity-exactly-30
sum(sharks$equity_offered_pct == 30)
```

In the file, 33 of the 1,441 pitches show exactly 30. These repeated values
come from rounding the percentages in the file, not from the quantity
itself. The model is
continuous, and in the model an offer of exactly 30% has probability zero.

You do not need calculus in this course. You read probability as an area,
and you calculate it with a CDF, with R, or with supplied values.

::: {.callout-tip collapse="true"}
## Optional: the formal definition of a density

A function $f_X$ is a density of the continuous random variable $X$ if it is
never negative and has total area 1,

$$f_X(x)\ge 0 \ \text{ for every } x,
\qquad
\int_{-\infty}^{\infty} f_X(x)\,dx=1,$$

and if, for all values $a\le b$,

$$P(a\le X\le b)=\int_a^b f_X(x)\,dx.$$

The integral is the area under the curve between $a$ and $b$. For $a=b$ the
area has zero width, which gives $P(X=a)=0$ again.
:::

::: {.callout-note collapse="true"}
## Check: is a rectangle of width 0.5 and height 2 a valid density?

Yes. The curve is never negative and its total area is
$0.5 \times 2 = 1$, so it is a valid PDF. The height 2 is not a probability.
No probability here exceeds 1, because every probability is an area inside
that rectangle. The height 2 means that the probability is packed into a
narrow range.
:::

#### The CDF works for every random variable

The cumulative distribution function (CDF) is still

$$F_X(x)=P(X\le x),$$

exactly as in Lecture 3. In a picture, $F_X(x)$ is the area under the density
to the left of $x$. For a continuous random variable the CDF rises smoothly,
without jumps (it is continuous), from 0 to 1. It never falls (it is
non-decreasing), because a move to the right can only add area. @fig-b06-uniform-cdf in the next section shows an example.

The probability of an interval is the difference of two CDF values:

$$P(a<X\le b)=F_X(b)-F_X(a).$$

Because a single point has zero probability, changing $<$ to $\le$ does not
change a continuous probability: $P(a<X<b)$, $P(a\le X\le b)$, and the two
mixed versions are all equal. For a discrete variable this is different: a
single value may have positive probability, so it matters whether you write
$<$ or $\le$.

### The uniform distribution

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/uniform-piecewise-constant-pdfs">08.3 Uniform and Piecewise Constant PDFs</a>.</p>

The uniform distribution is the simplest continuous model. All possible
values lie in an interval from $a$ to $b$, and all parts of this interval
with the same length are equally likely. We write $X\sim U(a,b)$. The
numbers $a$ and $b$ are the parameters of the distribution. The interval
$[a,b]$ is the set of possible values, also called the support. The density
is constant on the support,

$$f_X(x)=\frac{1}{b-a},\qquad a\le x\le b,$$

and zero outside it. You cannot choose the height freely. The whole rectangle
must have area 1:

$$\text{width}\times\text{height}=(b-a)\times\frac{1}{b-a}=1.$$

Because the height is constant, the probability of an interval depends only
on its length. For $a\le c<d\le b$,

$$P(c<X<d)=\frac{d-c}{b-a}.$$

Here are the mean, the variance, and the standard deviation, without a
derivation:

$$E[X]=\frac{a+b}{2},\qquad \operatorname{Var}(X)=\frac{(b-a)^2}{12},
\qquad \operatorname{SD}(X)=\frac{b-a}{\sqrt{12}}.$$

::: {#exm-b06-uniform-session}
## Worked example: when does the pitch appear?

An invented example. A recording session of the TV show lasts 90 minutes,
and a pitch is equally likely to start at any moment of the session. Model the start time
as $X\sim U(0,90)$, measured in minutes.

*Step 1: the height.* The density is $1/(90-0)=1/90\approx 0.011$ per
minute. The total area must be 1, and this sets the height.

*Step 2: the event.* "The pitch starts during the 18-minute window from
minute 30 to minute 48" is the event $30<X<48$.

*Step 3: the area.* The probability is the area of the shaded rectangle in
@fig-b06-uniform-session:

$$P(30<X<48)=\frac{48-30}{90-0}=\frac{18}{90}=0.20.$$

For a uniform model, only the length of the window matters: any 18-minute
window inside the session has probability 0.20.
:::

```{r}
#| label: fig-b06-uniform-session
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The uniform density for a pitch start time over a 90-minute session. The navy line is the density. The shaded green area is the probability of the 18-minute window, 18/90 = 0.20."
library(ggplot2)

a <- 0
b <- 90
c_cut <- 30
d_cut <- 48
density_height <- 1 / (b - a)

support_rect <- data.frame(x = c(a, b), y = density_height)
window_rect  <- data.frame(x = c(c_cut, d_cut), y = density_height)
density_line <- data.frame(x = c(-9, a, a, b, b, 99),
                           y = c(0, 0, density_height, density_height, 0, 0))

ggplot() +
  geom_area(data = support_rect, aes(x, y), fill = "grey75", alpha = 0.25) +
  geom_area(data = window_rect, aes(x, y), fill = "#009E73", alpha = 0.25) +
  geom_line(data = density_line, aes(x, y), colour = "#17324D",
            linewidth = 1.2) +
  annotate("text", x = 66, y = 0.0127,
           label = "density = 1/(b - a) = 1/90",
           colour = "#17324D", size = 3.4) +
  annotate("segment", x = 39, xend = 39, y = 0.0042, yend = 0.0010,
           colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = 39, y = 0.0052,
           label = "area = P(30 < X < 48) = 0.20",
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = c(a, c_cut, d_cut, b),
                     labels = c("a = 0", "c = 30", "d = 48", "b = 90"),
                     expand = c(0, 0)) +
  scale_y_continuous(breaks = c(0, density_height),
                     labels = c("0", "1/90"),
                     limits = c(0, 0.0145),
                     expand = c(0, 0)) +
  labs(x = "Minutes into the session", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read the horizontal axis as possible start times and the vertical axis as
density, not probability. The navy density line is at 0 before minute 0,
jumps to $1/90$ on the support, and drops back to 0 after minute 90. The
green probability is an area: $18\times(1/90)=0.20$. In the check above,
the height was 2. Here the height is far below 1. Both densities are valid, because only the area is a probability.

#### The CDF of the session model

For the session model, the area to the left of $x$ is a rectangle of width
$x$ and height $1/90$. So the CDF is $F_X(x)=x/90$ for $0\le x\le 90$. It is
0 before minute 0 and 1 after minute 90. @fig-b06-uniform-cdf shows it. The
probability of the window is the difference of two CDF values:

$$P(30<X<48)=F_X(48)-F_X(30)=\frac{48}{90}-\frac{30}{90}=0.20.$$

This is the same 0.20 as the green area in @fig-b06-uniform-session.

```{r}
#| label: fig-b06-uniform-cdf
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The CDF of the pitch start time in the 90-minute session model. The navy line rises from 0 to 1 and never falls. The gold dashed lines read off F(30) and F(48). Their difference, the green bar, is the probability of the 18-minute window, 0.20."
library(ggplot2)

a <- 0
b <- 90
c_cut <- 30
d_cut <- 48
F_c <- (c_cut - a) / (b - a)
F_d <- (d_cut - a) / (b - a)

cdf_line <- data.frame(x = c(-9, a, b, 99), y = c(0, 0, 1, 1))

ggplot() +
  annotate("segment", x = c(c_cut, d_cut), xend = c(c_cut, d_cut),
           y = 0, yend = c(F_c, F_d),
           linetype = "dashed", colour = "#C98B2E", linewidth = 0.6) +
  annotate("segment", x = -9, xend = c(c_cut, d_cut),
           y = c(F_c, F_d), yend = c(F_c, F_d),
           linetype = "dashed", colour = "#C98B2E", linewidth = 0.6) +
  geom_line(data = cdf_line, aes(x, y), colour = "#17324D",
            linewidth = 1.2) +
  annotate("segment", x = 6, xend = 6, y = F_c, yend = F_d,
           colour = "#009E73", linewidth = 1.6) +
  annotate("text", x = 8, y = (F_c + F_d) / 2,
           label = "difference = 0.20", colour = "#17324D",
           size = 3.4, hjust = 0) +
  annotate("text", x = 66, y = 0.58, label = "F(x) = x/90",
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_x_continuous(breaks = c(a, c_cut, d_cut, b), expand = c(0, 0)) +
  scale_y_continuous(breaks = c(0, F_c, F_d, 1),
                     labels = c("0", "F(30) = 0.33", "F(48) = 0.53", "1"),
                     limits = c(0, 1.05), expand = c(0, 0)) +
  labs(x = "Minutes into the session (x)",
       y = "Cumulative probability F(x)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

::: {.callout-caution collapse="true"}
## Try it: the final stretch of the session

Try this with your neighbour first. Same model, $X\sim U(0,90)$. What is the
probability that the pitch starts in the final 27 minutes of the session?

*Step 1 (given).* The event is $63<X<90$, an interval of length
$90-63=27$ minutes inside the support.

*Step 2 (you).* Divide the interval length by the support length.

*Step 3 (you).* Interpret the result in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(63<X<90)=27/90=0.30$: under the uniform model there is a 30% chance that
the pitch starts in the final 27 minutes. This is larger than the 0.20
above only because the window is longer.
:::
:::

### The normal distribution

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/normal-random-variables/">08.8 Normal Random Variables</a>.</p>

[Shark Tank data]{.shark-tag} Many quantities have a bell-shaped distribution, with most values near the
middle and few values far away on either side. Why is this shape so common?
Take the adult height of a person. Think of it as a starting value plus a
long list of small pushes. One gene pushes the height up a little, another
pushes it down. Good food in childhood adds a bit, and an illness takes a
bit away. There are hundreds of such influences. Each one is small, and each
is roughly independent of the others. The final height is the sum of all the
pushes.

How does a person become very tall? Nearly all the pushes must go up, and
that almost never happens. In the same way, a coin almost never lands heads
ten times in a row. A height near the average can arise in very many
ways: in every mixture where the ups and downs roughly cancel each other. With ten coin
tosses there is one way to get ten heads, but there are 252 ways to get five
heads. So most people end up near the middle, and few end up far away. A
very short height is as rare as a very tall one, so the two sides of the
bell fall off at the same rate. That is the bell shape.

@fig-b06-pushes shows this with a computer simulation. One repetition adds up 10
pushes. Each push is $+1$ or $-1$ with equal probability, like heads and
tails of a coin. The figure counts the totals of 10,000 repetitions.

```{r}
#| label: fig-b06-pushes
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Ten small pushes, added up, in 10,000 repetitions. Totals near zero can arise in many ways, so they are common. A total of 10 or of -10 can arise in only one way, so it is rare. The bars form a bell around 0."
library(ggplot2)

# One repetition: 10 pushes, each +1 or -1 with equal probability, added up.
set.seed(206)
pushes <- matrix(sample(c(-1, 1), size = 10 * 10000, replace = TRUE),
                 ncol = 10)
total <- rowSums(pushes)
total_counts <- as.data.frame(
  table(total = factor(total, levels = seq(-10, 10, by = 2))))
total_counts$total <- as.numeric(as.character(total_counts$total))

# Number of ways to reach a total. Five ups and five downs give 0: 252 ways.
# Ten ups give 10: 1 way.
ways_zero <- choose(10, 5)
ways_ten <- choose(10, 10)
top <- max(total_counts$Freq)

ggplot(total_counts, aes(total, Freq)) +
  geom_col(fill = "#009E73", alpha = 0.85, width = 1.6) +
  annotate("text", x = 0, y = top + 200,
           label = paste(ways_zero, "ways to reach 0"),
           colour = "#17324D", size = 3.4) +
  annotate("text", x = c(-10, 10), y = 520,
           label = paste0(ways_ten, " way to\nreach ", c(-10, 10)),
           colour = "#17324D", size = 3.4, lineheight = 0.95) +
  annotate("segment", x = c(-10, 10), xend = c(-10, 10), y = 330, yend = 60,
           colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  scale_x_continuous(breaks = seq(-10, 10, by = 2)) +
  scale_y_continuous(limits = c(0, top + 350),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Total of the 10 pushes",
       y = "Number of repetitions (of 10,000)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The same explanation fits many quantities. The weight of a filled package
is the sum of the weights of everything in it. An exam score adds up many
small successes and small mistakes. Suppose a quantity is the sum of many
small, independent influences, and no single influence is much larger than
the others. Then its distribution tends to be bell-shaped. The normal distribution is the mathematical model for this
shape.

The normal distribution is a model, not a law. The explanation above also
tells you when to doubt it. If one influence is much larger than the others,
for example an upper limit, a cut-off, or one large cause, the bell shape is
lost. If the influences multiply and do not add, as with growth
rates, the distribution is skewed and not symmetric. Many business variables
are like this. Revenues, firm sizes, and the equity offered in the Shark
Tank file are strongly skewed, and the normal model does not fit them. The
section on checking the normal model against data comes back to this.

A normal model needs two numbers. The first is its mean $\mu$. The second is
its standard deviation $\sigma$. Lecture 3 introduced both symbols.

::: {.definition-box}
<span class="label">Definition</span>

**Normal distribution.** A continuous distribution with a symmetric,
bell-shaped density. Its centre is the mean $\mu$, and its spread is the
standard deviation $\sigma>0$. We write $X\sim N(\mu,\sigma^2)$.
:::

Read the notation with care: the second number in the brackets is the
variance $\sigma^2$, not the standard deviation. For example, $N(50,100)$
has $\sigma=\sqrt{100}=10$. In symbols,

$$E[X]=\mu,\qquad \operatorname{Var}(X)=\sigma^2,
\qquad \operatorname{SD}(X)=\sigma.$$

::: {.callout-tip collapse="true"}
## Optional: the formula of the normal density

The density of $N(\mu,\sigma^2)$ is

$$f_X(x)=\frac{1}{\sigma\sqrt{2\pi}}\,
\exp\!\left(-\frac{(x-\mu)^2}{2\sigma^2}\right).$$

You do not have to know this formula by heart. R evaluates it with
`dnorm()`.
:::

The two parameters have different roles, as @fig-b06-normal-parameters shows.
A change in $\mu$ shifts the whole curve to the left or to the right, and
the shape stays the same. A change in $\sigma$ makes the curve wider or
narrower. A wider curve is also lower, because the total area stays 1.

```{r}
#| label: fig-b06-normal-parameters
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The two parameters of the normal model. Left: a larger mean (orange, μ = 3 in place of 0) shifts the curve to the right and leaves its shape unchanged. Right: a larger standard deviation (orange, σ = 2 in place of 1) makes the curve wider and lower."
library(ggplot2)

x_grid <- seq(-6, 8, length.out = 600)
panel_shift <- "Changing the mean shifts the curve"
panel_width <- "Changing the SD widens the curve"

curves <- rbind(
  data.frame(panel = panel_shift, curve = "first", x = x_grid,
             density = dnorm(x_grid, mean = 0, sd = 1)),
  data.frame(panel = panel_shift, curve = "second", x = x_grid,
             density = dnorm(x_grid, mean = 3, sd = 1)),
  data.frame(panel = panel_width, curve = "first", x = x_grid,
             density = dnorm(x_grid, mean = 0, sd = 1)),
  data.frame(panel = panel_width, curve = "second", x = x_grid,
             density = dnorm(x_grid, mean = 0, sd = 2))
)
curves$panel <- factor(curves$panel, levels = c(panel_shift, panel_width))

curve_labels <- data.frame(
  panel = factor(c(panel_shift, panel_shift, panel_width, panel_width),
                 levels = c(panel_shift, panel_width)),
  curve = c("first", "second", "first", "second"),
  x = c(-1.2, 4.2, 1.1, 3.3),
  y = c(0.36, 0.36, 0.36, 0.12),
  label = c("mu == 0*','~sigma == 1", "mu == 3*','~sigma == 1",
            "mu == 0*','~sigma == 1", "mu == 0*','~sigma == 2"),
  hjust = c(1, 0, 0, 0)
)

ggplot(curves, aes(x, density, colour = curve)) +
  geom_line(linewidth = 1.1) +
  geom_text(data = curve_labels, aes(x, y, label = label, hjust = hjust),
            parse = TRUE, size = 3.4) +
  facet_wrap(~panel) +
  scale_colour_manual(values = c(first = "#009E73", second = "#D55E00"),
                      guide = "none") +
  scale_x_continuous(breaks = seq(-6, 8, by = 2)) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of X", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", hjust = 0)
  )
```

@fig-b06-normal-shape shows where the probability of a normal model lies.

```{r}
#| label: fig-b06-normal-shape
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The normal density. The shaded central area between the two grey dashed lines runs from one standard deviation below the mean to one above and contains about 68% of the distribution. The gold dashed line marks the mean μ."
library(ggplot2)

normal_curve <- data.frame(z = seq(-3.6, 3.6, length.out = 800))
normal_curve$density <- dnorm(normal_curve$z)

ggplot(normal_curve, aes(z, density)) +
  geom_area(data = subset(normal_curve, z >= -1 & z <= 1),
            fill = "#009E73", alpha = 0.25) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.16, label = "about 68%",
           colour = "#17324D", size = 3.4) +
  annotate("text", x = 0.15, y = 0.425, label = "mean~mu", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_x_continuous(breaks = c(-2, -1, 0, 1, 2),
                     labels = expression(mu - 2 * sigma, mu - sigma, mu,
                                         mu + sigma, mu + 2 * sigma)) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of X", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The horizontal axis shows possible values in terms of $\mu$ and $\sigma$.
The vertical axis is density, and probability is again area. In the normal
model, about 68% of the distribution lies within one standard deviation of
the mean and about 95% within 1.96 standard deviations. The section on the
central part of the standard normal explains the value 1.96. These are facts
about the mathematical model. The box "What a normal label can and cannot
say" below explains why they are not automatically true for every dataset.

#### The opening question as an area

Return to the offer of 30%. Take the normal model that has the mean and the
standard deviation of the equity offered in the Shark Tank file,
$X\sim N(13.2,\,8.4^2)$. In this model, the probability of an offer of 30%
or more is the area under the curve to the right of 30.
@fig-b06-equity-normal-area shows that this area is small.

```{r}
#| label: fig-b06-equity-normal-area
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The normal model with the mean and the standard deviation of the equity offered (`equity_offered_pct`) in the Shark Tank file. The gold dashed line marks the mean. The orange area to the right of 30% is the probability that this model gives to an offer of 30% or more, about 2%."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mean_equity <- mean(sharks$equity_offered_pct)
# sd() divides by n - 1. With 1,441 records the difference from dividing
# by N is in the third decimal.
sd_equity <- sd(sharks$equity_offered_pct)

model_curve <- data.frame(x = seq(-15, 45, length.out = 601))
model_curve$density <- dnorm(model_curve$x, mean = mean_equity,
                             sd = sd_equity)

ggplot(model_curve, aes(x, density)) +
  geom_area(data = subset(model_curve, x >= 30),
            fill = "#D55E00", alpha = 0.55) +
  geom_vline(xintercept = mean_equity, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = mean_equity + 0.8, y = 0.0495, label = "mean 13.2%",
           colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("segment", x = 35, xend = 32.2, y = 0.012, yend = 0.003,
           colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = 30.5, y = 0.0168, label = "area == P(X >= 30)",
           parse = TRUE, colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("text", x = 30.5, y = 0.0138, label = "about 2% in this model",
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_x_continuous(breaks = seq(-10, 40, by = 10)) +
  scale_y_continuous(limits = c(0, 0.052),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The figure shows the answer of the model, not the answer of the data. The
curve also gives some area to values below 0%, which no offer can have. The
section on checking the normal model against data returns to both points. To
turn the orange area into a number, you need one more tool: standardising.

#### Standardising: every normal question becomes a Z question

::: {.definition-box}
<span class="label">Definition</span>

**z-score.** The z-score of a value $x$ is $z=\dfrac{x-\mu}{\sigma}$. It
says how many standard deviations $x$ lies above the mean (positive $z$) or
below the mean (negative $z$). Calculating a z-score is called
**standardising**.
:::

If $X\sim N(\mu,\sigma^2)$, then the standardised variable is again normal,
with mean 0 and standard deviation 1:

$$Z=\frac{X-\mu}{\sigma}\sim N(0,1).$$

The mean 0 and the standard deviation 1 follow from the rules for shifting
and rescaling in
[Lecture 3](b03-discrete-distributions.qmd#b03-shift-rescale). Write
$Z=\frac{1}{\sigma}X-\frac{\mu}{\sigma}$, so that $a=1/\sigma$ and
$b=-\mu/\sigma$:

$$E[Z]=\frac{1}{\sigma}\,E[X]-\frac{\mu}{\sigma}
=\frac{\mu}{\sigma}-\frac{\mu}{\sigma}=0,
\qquad
\operatorname{SD}(Z)=\frac{1}{\sigma}\,\operatorname{SD}(X)
=\frac{\sigma}{\sigma}=1.$$

The shape does not change. @fig-b06-normal-parameters showed that $\mu$ and
$\sigma$ do only two things to a normal curve: $\mu$ moves it along the
axis, and $\sigma$ stretches it. Every normal curve is the same bell, moved
and stretched. Standardising undoes the move and the stretch. What is left
is the one bell with mean 0 and standard deviation 1. The distribution
$N(0,1)$ is called the **standard normal distribution**.

This is why a z-score is useful. It gives the position of a value in its own
distribution, in standard deviations from the mean and without units. For a
probability, only this position matters. A value 1.5 standard deviations
above the mean is at the same place on every bell. The probability of a
value above it is therefore the same for a height in centimetres, a weight
in grams, or a processing time in minutes. So one set of standard-normal
values works for every normal model.

The questions "what is $P(X\le x)$?" and "what is $P(Z\le z)$?", with
$z=(x-\mu)/\sigma$, are the same question. Subtract $\mu$ and divide by
$\sigma$ on both sides of $X\le x$. This gives $Z\le z$, and exactly the
same outcomes satisfy both inequalities. The two probabilities are equal:

$$P(X\le x)=P\!\left(Z\le\frac{x-\mu}{\sigma}\right).$$

With two general rules you can find more probabilities from one value
$P(Z\le z)$. Both rules hold for every $z$.

- Complement rule. The total area under the curve is 1, so
  $P(Z>z)=1-P(Z\le z)$.
- Symmetry rule. The curve is symmetric around 0, so
  $P(Z\le -z)=P(Z>z)$.

::: {#exm-b06-supplied-z}
## Worked example: normal probabilities from supplied Z values

An invented example. The processing time $Y$ of an order (in minutes) is
modelled as
$Y\sim N(50,100)$. The variance is 100, so $\sigma=\sqrt{100}=10$. In this
book, the standard-normal values that you need are given in the task or on
the formula sheet, as here:

```
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
P(Z <= -1.0) = 0.1587
```

**(a) $P(Y\le 65)$.** Standardise the boundary:

$$z=\frac{65-50}{10}=1.5,$$

so

$$P(Y\le 65)=P(Z\le 1.5)=0.9332.$$

**(b) $P(40<Y<60)$.** Standardise both boundaries:

$$z_{\text{low}}=\frac{40-50}{10}=-1.0,\qquad
z_{\text{high}}=\frac{60-50}{10}=1.0,$$

then take the difference of the two cumulative probabilities, exactly as
with any CDF:

$$P(40<Y<60)=P(Z\le 1.0)-P(Z\le -1.0)=0.8413-0.1587=0.6826.$$

That is the 68% rule seen in @fig-b06-normal-shape: the interval runs from
one standard deviation below the mean to one above.

**(c) $P(Y>60)$ and $P(Y\le 35)$.** These two need the complement rule and
the symmetry rule. The z-score of
60 is 1.0, and the complement rule gives

$$P(Y>60)=P(Z>1.0)=1-P(Z\le 1.0)=1-0.8413=0.1587.$$

The z-score of 35 is $(35-50)/10=-1.5$. No value for $-1.5$ is supplied, so
use the symmetry rule first and then the complement rule:

$$P(Y\le 35)=P(Z\le -1.5)=P(Z>1.5)=1-0.9332=0.0668.$$

The supplied values also show the symmetry rule at work:
$P(Z\le -1.0)=0.1587$ is the same number as $P(Z>1.0)$.
:::

`pnorm()` is the normal CDF in R. Give it the boundary, the mean, and the
standard deviation (not the variance). The four lines below repeat the four
results of the worked example:

```{r}
#| label: b06-pnorm-check
pnorm(65, mean = 50, sd = 10)
pnorm(60, mean = 50, sd = 10) - pnorm(40, mean = 50, sd = 10)
1 - pnorm(60, mean = 50, sd = 10)
pnorm(35, mean = 50, sd = 10)
```

The hand calculation and `pnorm()` agree. Small differences in the fourth
decimal come from the rounding of the supplied values.

#### The answer of the model to the opening question

Now the orange area in @fig-b06-equity-normal-area can be calculated. The
model is $X\sim N(13.2,\,8.4^2)$ and the boundary is 30:

$$z=\frac{30-13.2}{8.4}=2.0.$$

An offer of 30% lies two standard deviations above the mean. The supplied
value is $P(Z\le 2.0)=0.9772$, and the complement rule gives

$$P(X\ge 30)=P(Z\ge 2.0)=1-P(Z\le 2.0)=1-0.9772=0.0228.$$

For a continuous variable, $\ge$ and $>$ give the same probability. R
confirms the result:

```{r}
#| label: b06-opening-pnorm
pnorm(2)
1 - pnorm(2)
1 - pnorm(30, mean = 13.2, sd = 8.4)
```

The normal model says that about 2% of pitches offer 30% or more. The count
in the file gave about 5%. The section on checking the model explains the
difference.

#### The reverse direction: from a probability to a value

`pnorm()` starts from a value and returns the probability at or below it.
The reverse question starts from a probability $q$: which value has the
probability $q$ at or below it? This value is called the $q$ quantile.

::: {.definition-box}
<span class="label">Definition</span>

**Quantile.** Take a probability $q$ between 0 and 1. The $q$ quantile of a
distribution is the value $x_q$ that has the probability $q$ at or below it:
$P(X\le x_q)=q$. A **percentile** is the same value with $q$ written in
percent: the 0.90 quantile is the 90th percentile.
:::

The definition is cumulative: $q$ is always the area to the left of $x_q$,
as @fig-b06-quantile shows. The area to the right of $x_q$ is $1-q$. In
terms of the CDF, $F_X(x_q)=q$, so finding a quantile is the reverse of
reading the CDF.

```{r}
#| label: fig-b06-quantile
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.4
#| fig-cap: "The q quantile of a distribution, marked on the axis as x with the subscript q. The green area to the left of this value is the probability q. The area to the right of it is 1 - q."
library(ggplot2)

# The picture is drawn for q = 0.80. The labels hold for every q.
q_shown <- 0.80
x_q <- qnorm(q_shown)
q_curve <- data.frame(x = seq(-3.6, 3.6, length.out = 800))
q_curve$density <- dnorm(q_curve$x)

ggplot(q_curve, aes(x, density)) +
  geom_area(data = subset(q_curve, x <= x_q),
            fill = "#009E73", alpha = 0.25) +
  annotate("segment", x = x_q, xend = x_q, y = 0, yend = dnorm(x_q),
           linetype = "dashed", colour = "#C98B2E", linewidth = 0.7) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = -0.2, y = 0.15,
           label = "area == P(X <= x[q])", parse = TRUE,
           colour = "#17324D", size = 3.4) +
  annotate("text", x = -0.2, y = 0.105, label = "area == q", parse = TRUE,
           colour = "#17324D", size = 3.4) +
  annotate("text", x = 2.3, y = 0.15, label = "area == 1 - q", parse = TRUE,
           colour = "#17324D", size = 3.4) +
  annotate("segment", x = 2.2, xend = 1.5, y = 0.125, yend = 0.05,
           colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  scale_x_continuous(breaks = x_q, labels = expression(x[q])) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of X", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    axis.text.x = element_text(colour = "#17324D", size = 12)
  )
```

For the standard normal, the $q$ quantile is written $z_q$, and R gives it
with `qnorm(q)`. Now take a number for $q$. The 90th percentile of the
standard normal is the 0.90 quantile:

```{r}
#| label: b06-qnorm-090
qnorm(0.90)
```

So $z_{0.90}=1.2816$: 90% of the area lies at or below 1.2816. For any other
normal model, solve $z=(x-\mu)/\sigma$ for $x$:

$$x=\mu+z\sigma,$$

where $z$ is the standard-normal quantile for the chosen probability. With
$z=z_q$, this $x$ is the $q$ quantile $x_q$ of the model.

::: {#exm-b06-reverse}
## Worked example: the 90th percentile of processing time

Take the invented processing-time model $Y\sim N(50,100)$ again. Which
processing time do only 10% of orders exceed? This is the 90th
percentile of $Y$. The supplied standard-normal quantile is
$z_{0.90}=1.2816$.

*Step 1.* The 90th percentile of $Z$ is 1.2816. The value you look for
therefore lies 1.2816 standard deviations above the mean.

*Step 2.* Translate back to the scale of $Y$:

$$x=\mu+z\sigma=50+1.2816\times 10=62.8 \text{ minutes.}$$

*Step 3.* Verify with `qnorm()`, the reverse of `pnorm()`:

```{r}
#| label: b06-qnorm-check
qnorm(0.90, mean = 50, sd = 10)
```

About 62.8 minutes: 90% of processing times fall at or below it, 10%
above.
:::

#### The central part of the standard normal {#b06-central}

One pair of quantiles is used again and again in later lectures: the two
values that enclose the central part of the standard normal. Start with the
probability that is left outside the central part, and call it $\alpha$
(the Greek letter alpha). The central part then holds the probability
$1-\alpha$. For example, $\alpha=0.05$ gives a central part of 0.95.

The probability $\alpha$ lies in two tails, one on the left and one on the
right. The standard normal is symmetric around 0, so the two tails are
equally large. Each tail holds $\alpha/2$, as @fig-b06-central-general
shows.

Now find the two boundaries. To the left of the upper boundary lie the
central part and the lower tail. The probability at or below it is therefore
$(1-\alpha)+\alpha/2=1-\alpha/2$. The upper boundary is the $1-\alpha/2$
quantile, $z_{1-\alpha/2}$. The lower boundary needs no new number. By
symmetry it lies as far below 0 as the upper boundary lies above 0, so it is
$-z_{1-\alpha/2}$. Together:

$$P\!\left(-z_{1-\alpha/2}\le Z\le z_{1-\alpha/2}\right)=1-\alpha.$$

```{r}
#| label: fig-b06-central-general
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The central part of the standard normal. The green central area holds the probability 1 - α. The rest, α, is split equally over the two orange tails, α/2 in each. The two boundaries lie at the same distance from 0."
library(ggplot2)

# The picture is drawn for alpha = 0.10. The labels hold for every alpha.
z_upper <- qnorm(1 - 0.10 / 2)
g_curve <- data.frame(z = seq(-3.6, 3.6, length.out = 800))
g_curve$density <- dnorm(g_curve$z)

ggplot(g_curve, aes(z, density)) +
  geom_area(data = subset(g_curve, z >= -z_upper & z <= z_upper),
            fill = "#009E73", alpha = 0.25) +
  geom_area(data = subset(g_curve, z <= -z_upper),
            fill = "#D55E00", alpha = 0.55) +
  geom_area(data = subset(g_curve, z >= z_upper),
            fill = "#D55E00", alpha = 0.55) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.17, label = "central area",
           colour = "#17324D", size = 3.4) +
  annotate("text", x = 0, y = 0.13, label = "1 - alpha", parse = TRUE,
           colour = "#17324D", size = 4) +
  annotate("segment", x = c(-2.9, 2.9), xend = c(-2.1, 2.1),
           y = 0.105, yend = 0.03, colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = c(-2.9, 2.9), y = 0.125, label = "alpha / 2",
           parse = TRUE, colour = "#17324D", size = 4) +
  annotate("text", x = c(-2.9, 2.9), y = 0.16,
           label = c("lower tail", "upper tail"),
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = c(-z_upper, 0, z_upper),
                     labels = expression(-z[1 - alpha / 2], 0,
                                         z[1 - alpha / 2])) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of Z", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    axis.text.x = element_text(colour = "#17324D", size = 12)
  )
```

Three central parts are used most often: 90%, 95%, and 99%. They belong to
$\alpha=0.10$, $\alpha=0.05$, and $\alpha=0.01$. The quantiles
$z_{1-\alpha/2}$ are then $z_{0.95}$, $z_{0.975}$, and $z_{0.995}$:

```{r}
#| label: b06-qnorm-three
qnorm(c(0.95, 0.975, 0.995))
```

| Central part $1-\alpha$ | $\alpha$ | In each tail, $\alpha/2$ | Quantile $z_{1-\alpha/2}$ | The central part lies between |
|---|---|---|---|---|
| 90% | 0.10 | 0.05 | $z_{0.95}=1.645$ | $-1.645$ and $1.645$ |
| 95% | 0.05 | 0.025 | $z_{0.975}=1.960$ | $-1.960$ and $1.960$ |
| 99% | 0.01 | 0.005 | $z_{0.995}=2.576$ | $-2.576$ and $2.576$ |

A larger central part needs boundaries that lie further out. The 95% case is
the one used most. R confirms the symmetry for it:

```{r}
#| label: b06-qnorm-central
qnorm(c(0.025, 0.975))
```

The 0.025 quantile is $-1.96$ and the 0.975 quantile is $1.96$. In the
standard normal, 95% of the probability lies between $-1.96$ and $1.96$,
with 2.5% in each tail, as @fig-b06-central-95 shows.

```{r}
#| label: fig-b06-central-95
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The standard normal density. The green area between -1.96 and 1.96 holds 95% of the probability. Each orange tail holds 2.5%."
library(ggplot2)

z_cut <- qnorm(0.975)
z_curve <- data.frame(z = seq(-3.6, 3.6, length.out = 800))
z_curve$density <- dnorm(z_curve$z)
centre <- data.frame(z = seq(-z_cut, z_cut, length.out = 400))
centre$density <- dnorm(centre$z)
left_tail <- data.frame(z = seq(-3.6, -z_cut, length.out = 200))
left_tail$density <- dnorm(left_tail$z)
right_tail <- data.frame(z = seq(z_cut, 3.6, length.out = 200))
right_tail$density <- dnorm(right_tail$z)

ggplot(z_curve, aes(z, density)) +
  geom_area(data = centre, fill = "#009E73", alpha = 0.25) +
  geom_area(data = left_tail, fill = "#D55E00", alpha = 0.55) +
  geom_area(data = right_tail, fill = "#D55E00", alpha = 0.55) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.16, label = "95% of the area",
           colour = "#17324D", size = 3.4) +
  annotate("segment", x = c(-2.9, 2.9), xend = c(-2.3, 2.3),
           y = 0.075, yend = 0.02, colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = c(-2.9, 2.9), y = 0.09, label = "2.5%",
           colour = "#17324D", size = 3.4) +
  scale_x_continuous(breaks = c(-z_cut, 0, z_cut),
                     labels = c("-1.96", "0", "1.96")) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of Z", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

With $x=\mu+z\sigma$ the same holds for every normal model. The central
$1-\alpha$ of the probability lies between $\mu-z_{1-\alpha/2}\,\sigma$ and
$\mu+z_{1-\alpha/2}\,\sigma$. For 95%, these are $\mu-1.96\sigma$ and
$\mu+1.96\sigma$. Lectures 9 and 10 build on the three values 1.645, 1.960,
and 2.576.

#### Sums of normal variables {#b06-normal-sums}

A sum of independent normal variables is again normal. Its mean and its
variance follow from the sum rules of
[Lecture 5](b05-joint-covariance-correlation.qmd#b05-sums). Lecture 8 uses
this fact for the sample mean.

### Checking the normal model against data

[Shark Tank data]{.shark-tag} The normal distribution is a model. It does not automatically fit every
quantitative variable. Skewness, bounds, extreme outliers, or a few values
that repeat very often can make it a poor fit. The check is visual. Draw the
histogram of the data and the normal curve with the same mean and standard
deviation. Then see whether the curve follows the bars.
@fig-b06-equity-normal-check does this for the equity offered
(`equity_offered_pct`) in the Shark Tank file, with the mean of 13.2% and
the standard deviation of 8.4 percentage points from the start of this
lecture.

```{r}
#| label: fig-b06-equity-normal-check
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Initial equity offered across all 1,441 recorded pitches (green bars), with the normal density that has the same mean and standard deviation (orange curve). The curve misses the skew, the bounds, and the spikes at round percentages."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mean_equity <- mean(sharks$equity_offered_pct)
# sd() divides by n - 1. With 1,441 records the difference from dividing
# by N is in the third decimal.
sd_equity <- sd(sharks$equity_offered_pct)

ggplot(sharks, aes(equity_offered_pct)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = seq(0, 100, by = 5),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  stat_function(fun = dnorm,
                args = list(mean = mean_equity, sd = sd_equity),
                colour = "#D55E00", linewidth = 0.9) +
  annotate("text", x = 42, y = 0.062, label = "observed pitches",
           colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("text", x = 42, y = 0.054,
           label = "normal curve with the same mean and SD",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Equity initially offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-b06-equity-normal-check with the reading sequence from
Lecture 1. *Unit and variables:* one observation is one pitch. The horizontal
axis is the equity offered, in percent. *Scales:* the vertical axis
is density, so the areas of the bars, not their heights alone, represent
proportions of pitches. *One element:* the green bars are the 1,441 observed
offers. The orange curve is $N(13.2,\,8.4^2)$, the normal model with the same
mean and SD. *Pattern:* most bars are between 0% and 20%, and a long tail
stretches to the right. The data are bounded between 0% and 100%, strongly
right-skewed, and have spikes at round values such as 10% and 20%. The symmetric
curve misses all three features. *Limits:* the figure rejects the normal
model for single equity offers. It says nothing yet about averages of
offers. A model is a choice, and this check shows that the normal model is
the wrong choice here.

The opening question shows what happens without the check. The normal model
gave $P(X\ge 30)=0.0228$, about 2%. The data give 76 of 1,441, about 5%. The
model misses the long right tail, and the offers of 30% or more lie in that
tail.

Still, the normal model remains useful. From Lecture 8 onwards it
describes the sampling distribution of an average, that is, how an average
varies from sample to sample. Lectures 7 and 8 explain this idea. The normal
model fits those distributions well, even when the variable itself is as
skewed as this one. That result is the central limit theorem. It is the main
reason why the normal model is used so much in statistics.

### Quantiles of a model distribution

Lecture 1 found quantiles of data: the median, the quartiles $Q_1$ and
$Q_3$, and the interquartile range of the equity offered (see
[Lecture 1](b01-introduction-data-analysis.qmd#b01-quartiles)). A model
distribution has the same quantiles. The median is the 0.50 quantile, and
the quartiles are the 0.25 and the 0.75 quantiles.

- **Uniform:** by symmetry, the median of $U(a,b)$ is the midpoint
  $(a+b)/2$. For the session model $U(0,90)$, the median start time is 45
  minutes: half of the probability lies on each side.
- **Normal:** by symmetry, the median equals the mean $\mu$. The quartiles
  lie $0.6745$ standard deviations on either side of the mean,
  $\mu\pm 0.6745\sigma$. For the processing-time model $Y\sim N(50,100)$:

```{r}
#| label: b06-model-quantiles
qnorm(c(0.25, 0.75), mean = 50, sd = 10)
```

The middle half of this model distribution runs from about 43.3 to 56.7
minutes. The width of this range is the interquartile range of the model.

::: {.caveat-box}
<span class="label">What a normal label can and cannot say</span>

- "Normally distributed" is a model, not a fact. You have to check it against
  a histogram.
- The 68% and 95% rules hold for the model. For skewed or bounded data they
  give wrong numbers, as with the 30% offers.
- A curve with the right mean and standard deviation can still have the wrong
  shape.

**What you can do with it:** use the normal model for quick answers when the
histogram looks roughly bell-shaped. When it does not, count the values in
the data.
From Lecture 8 on, the normal model describes averages, where it works well.
:::

### Answering the driving question

An offer of 30% or more is unusual: about 1 pitch in 20 makes such an offer. In the
Shark Tank file, 76 of the 1,441 pitches did, about 5%. The normal model
with the same mean and standard deviation says about 2%, because it ignores
the long right tail of the offers. So a model with two numbers can answer such a question
without counting every pitch, but only when the model fits. The histogram
check tells you whether it does.

The tools behind this answer are general. For a continuous variable, no
single value has positive probability, $P(X=x)=0$. Probabilities belong to
intervals, and they are areas under a density. The density is the share per
unit of width: the curve that a histogram comes close to when its bins get
narrow. For a uniform model, only the length of the interval matters: the
18-minute window in the session model has probability $18/90=0.20$. For a
normal model, you standardise and use the supplied $Z$ values. In the
processing-time model $N(50,100)$ this gave $P(Y\le 65)=0.9332$ and
$P(40<Y<60)=0.6826$, and `pnorm()` confirmed both. A quantile goes the other
way, from a probability to a value: the 90th percentile of the processing
time is $50+1.2816\times 10=62.8$ minutes, and the central 95% of every
normal model lies between $\mu-1.96\sigma$ and $\mu+1.96\sigma$. A density
is a modelling choice, so check the model before you trust its areas.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your delivery times look roughly normal, with a mean of 28 minutes and a standard deviation of 6 minutes. You want to promise a delivery time that 90% of orders meet. The 90th percentile is 28 + 1.2816 × 6 = 35.7 minutes. The decision: promise 36 minutes on the website. First check the histogram of your delivery times. If a few deliveries take much too long, the normal model is too optimistic. Then you promise the 90th percentile of the actual delivery times instead.
:::

**Next:** Tutorial 2, later this week, practises Lectures 4 and 5. So far the lectures treated the 1,441 pitches as all the data
that exist. In practice a firm rarely sees all the data. It sees a sample: 100 audited
transactions, 500 surveyed customers. Lecture 7 moves from one pitch to a
sample of pitches. It covers populations, random sampling, and the
difference between a parameter, an estimator, and an estimate.

:::

::: {.course-phase .phase-after}

## After class

After the lecture, work through all the practice below.

### Read the lecture material

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions

1. Why is $P(X=x)=0$ for a continuous random variable?
2. What quantity under a PDF represents probability?
3. How do you obtain $P(a<X\le b)$ from a CDF?
4. What do $\mu$ and $\sigma$ control in a normal model?
5. A density curve reaches a height of 2. Why is that not a contradiction?
6. What does a z-score of $-1.5$ mean?
7. In a continuous model, an offer of exactly 30% has probability zero. Does
   that mean that such an offer cannot happen?
8. A histogram has the share in each bin as the height of its bars. What
   happens to it when the bins get narrower, and which height avoids this?
9. Why does a sum of many small independent pushes give a bell shape?
10. Which quantile of the standard normal is the upper boundary of the
    central 90%, and why?
11. What is the median of the session model $U(0,90)$, and what is the
    median of the processing-time model $N(50,100)$?

::: {.callout-note collapse="true"}
## Check your answers

1. Probability belongs to intervals. An individual point has zero width and
   therefore zero area.
2. Area under the density curve between the interval boundaries.
3. $F_X(b)-F_X(a)$.
4. $\mu$ controls the centre and $\sigma$ the spread.
5. Density is probability per unit, not probability. A height above 1 is
   fine as long as the total area under the curve is 1.
6. The value lies 1.5 standard deviations below the mean.
7. No. Every offer has some exact value, and that value had probability zero
   in advance. Zero probability only says that no single value can be named
   in advance with a positive probability.
8. All bars shrink, and the histogram sinks towards the axis. Divide the
   share in each bin by the bin width. With this height, the share per unit
   of width, the histogram comes close to a curve, the density.
9. A total near the middle can arise in very many ways, because ups and
   downs cancel. A total far from the middle needs nearly all pushes in the
   same direction, and that is rare. Both sides are equally hard to reach,
   so the shape is symmetric.
10. The 0.95 quantile, $z_{0.95}=1.645$. The central 90% leaves
    $\alpha=0.10$ for the two tails, 0.05 in each. To the left of the upper
    boundary lie $0.90+0.05=0.95$ of the probability.
11. 45 minutes and 50 minutes. Both distributions are symmetric, so the
    median is the midpoint, which is also the mean.
:::

### Finish a half-worked example

Re-read the session-timing calculation (@exm-b06-uniform-session). Then
complete the same calculation for a new model: a waiting time
$X\sim U(10,30)$ minutes.

*Step 1 (given).* The density height is $1/(30-10)=0.05$ on the support
$[10,30]$ and zero outside it.

*Step 2 (you).* Find $P(14<X<20)$.

*Step 3 (you).* State $E[X]$ and interpret both numbers in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$P(14<X<20)=(20-14)/(30-10)=6/20=0.30$ and $E[X]=(10+30)/2=20$: the average
wait is 20 minutes. Under the uniform model there is a 30% chance that the
wait is between 14 and 20 minutes.
:::

### Practice

Items 2 to 6 are invented examples. For items 3, 4, and 5, use only these
supplied values, then verify in R:

```
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
P(Z <= -1.0) = 0.1587
0.90 quantile of Z: z(0.90) = 1.2816
0.95 quantile of Z: z(0.95) = 1.645
```

1. For $X\sim U(10,30)$: find $\operatorname{SD}(X)$ and $P(X>25)$.
2. A supplier delivers at a moment uniformly distributed between 09:00 and
   13:00. Other companies in your building use the goods lift (the lift for
   deliveries) all morning. You can use it only from 12:00 to 13:00. What
   is the probability that the delivery arrives while you can use the lift,
   and what is the density height per minute?
3. Monthly demand for a product is modelled as $Y\sim N(200,400)$.
   Calculate $P(Y\le 230)$, $P(180<Y<220)$, $P(Y>230)$, and $P(Y\le 170)$
   by hand.
4. Weekly revenue is modelled as $N(100,15^2)$ (in hundreds of euros).
   Which revenue level is exceeded in only 10% of weeks? Say which quantile
   of the model this is. Compute it by hand with $x=\mu+z\sigma$, then
   verify with `qnorm()`.
5. Take the weekly revenue model of item 4 again. Between which two revenue
   levels do the central 90% of weeks lie? State $\alpha$, the probability
   in each tail, and the quantile $z_{1-\alpha/2}$ that you use.
6. A histogram of 200 delivery times has bars of 10 minutes, from 10 to 90
   minutes, with the counts 20, 70, 60, 25, 12, 8, 3, 2. Sketch it. Does a
   normal model fit these times? Give two reasons.

::: {.callout-note collapse="true"}
## Check your answers

1. $\operatorname{SD}(X)=\sqrt{(30-10)^2/12}=20/\sqrt{12}\approx 5.77$
   minutes, and $P(X>25)=(30-25)/20=0.25$.
2. The window is 60 of the 240 possible minutes, so $60/240=0.25$. The
   density height is $1/240\approx 0.0042$ per minute, and the
   probability is the area $60\times(1/240)=0.25$, not the height.
3. $\sigma=\sqrt{400}=20$. For 230: $z=(230-200)/20=1.5$, so
   $P(Y\le 230)=0.9332$. For the interval:
   $z_{\text{low}}=(180-200)/20=-1.0$ and
   $z_{\text{high}}=(220-200)/20=1.0$, so
   $P(180<Y<220)=0.8413-0.1587=0.6826$. The complement rule gives
   $P(Y>230)=1-0.9332=0.0668$. For 170: $z=(170-200)/20=-1.5$, and the
   symmetry rule gives $P(Y\le 170)=P(Z\le -1.5)=P(Z>1.5)=0.0668$.
4. A level that 10% of weeks exceed has 90% of the weeks at or below it. It
   is the 0.90 quantile, the 90th percentile.
   $x=100+1.2816\times 15=119.2$, so about 119 hundred euros, that is about
   11,900 euros. Only 10% of weeks exceed it.
5. The central 90% leaves $\alpha=0.10$ for the two tails, 0.05 in each.
   The quantile is $z_{1-\alpha/2}=z_{0.95}=1.645$. The boundaries are
   $100-1.645\times 15=75.3$ and $100+1.645\times 15=124.7$. In 90% of
   weeks, the revenue lies between about 7,500 and 12,500 euros. In 5% of
   weeks it is lower, and in 5% it is higher.
6. No. First, the counts rise fast and fall slowly, so the histogram has a
   long tail to the right. A normal curve is symmetric. Second, the highest
   bars are at the left end and not in the middle: 150 of the 200 times
   (75%) lie in the first three of the eight bars. A normal curve has its
   peak in the middle and falls off equally on both sides.

Verification in R for items 3, 4, and 5:

```{r}
#| label: b06-practice-check
pnorm(230, mean = 200, sd = 20)
pnorm(220, mean = 200, sd = 20) - pnorm(180, mean = 200, sd = 20)
1 - pnorm(230, mean = 200, sd = 20)
pnorm(170, mean = 200, sd = 20)
qnorm(0.90, mean = 100, sd = 15)
qnorm(c(0.05, 0.95), mean = 100, sd = 15)
```
:::

### Review: one question from Lecture 3

An invented example. A student venture sells boxes of prepared meals. The
number of meal boxes in one customer order, $X$, has PMF $p(1)=0.6$, $p(2)=0.3$, $p(3)=0.1$. Find
$F(2)=P(X\le 2)$ and $P(X<2)$. Why do the two answers differ here, when for
a continuous variable they would not?

::: {.callout-note collapse="true"}
## Check your answer

$F(2)=0.6+0.3=0.9$, while $P(X<2)=P(X=1)=0.6$. They differ because the
single point $X=2$ has positive probability, $p(2)=0.3$. For a
continuous variable every single point has probability zero, so $<$ and
$\le$ give the same answer.
:::

### Changing units

The pitch start time $X\sim U(0,90)$ from the session model is recorded in
minutes and then converted to seconds. Explain what changes about the PDF and what remains
unchanged.

::: {.callout-note collapse="true"}
## Check your answer

The support stretches from $[0,90]$ minutes to $[0,5400]$ seconds, and the
density height drops from $1/90$ per minute to $1/5400$ per second: the
numerical size of a density depends on the units. Probabilities of the
same time intervals do not change, because they are areas: the
18-minute window becomes a 1,080-second window, and
$1080/5400=18/90=0.20$.
:::

### Worked exam interpretation: uniform and normal probabilities

**Exam-style question.** An invented example. A web shop models the weight
of a parcel as $X\sim U(2,10)$, in kilograms. Separately, it models the
value of one order as $Y\sim N(80,400)$, in euros. Supplied:

```
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
0.975 quantile of Z: z(0.975) = 1.960
```

1. State the PDF height of $X$, including its value outside the support.
2. Calculate $P(4<X<7)$, $E[X]$, and $\operatorname{Var}(X)$.
3. Standardise 100 euros under the model for $Y$, then calculate
   $P(Y\le 100)$.
4. Calculate $P(50<Y<100)$.
5. Between which two values do the central 95% of the order values lie?
6. Explain why a PDF height is not itself a probability.

::: {.callout-note collapse="true"}
## Check a model answer

1. $f_X(x)=1/(10-2)=0.125$ for $2\le x\le 10$, and $f_X(x)=0$ outside that
   interval.
2. $P(4<X<7)=(7-4)/8=0.375$, $E[X]=(2+10)/2=6$ kilograms, and
   $\operatorname{Var}(X)=(10-2)^2/12\approx 5.33$ square kilograms.
3. The variance is 400, so $\sigma=20$ and $z=(100-80)/20=1.0$. So
   $P(Y\le 100)=0.8413$.
4. The z-score of 50 is $(50-80)/20=-1.5$. No value for $-1.5$ is supplied,
   so use the symmetry rule and the complement rule:
   $P(Z\le -1.5)=P(Z>1.5)=1-0.9332=0.0668$. Then
   $P(50<Y<100)=0.8413-0.0668=0.7745$.
5. The central 95% leaves $\alpha=0.05$ for the two tails, 0.025 in each,
   so the quantile is $z_{0.975}=1.960$. The boundaries are
   $80-1.960\times 20=40.8$ and $80+1.960\times 20=119.2$. In this model,
   95% of the orders have a value between 40.80 and 119.20 euros.
6. Probability is area over an interval: approximately density times a
   small interval width. A single point has zero width and so zero
   probability. A density value becomes a probability only after you
   multiply it by a width.

Verification in R for part 5:

```{r}
#| label: b06-exam-check
qnorm(c(0.025, 0.975), mean = 80, sd = 20)
```
:::

### Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant: "Equity offers have a mean of 13.2% and a standard deviation of 8.4 percentage points. How many pitches offer 30% or more?" It answered:

> "With the normal distribution, 30% is 2.0 standard deviations above the mean, so about 2.3% of pitches offer 30% or more."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain.

::: {.callout-note collapse="true"}
## Check your answer

The calculation is correct for a normal model, but the model does not fit. Equity offers are strongly right-skewed and have many values at round numbers. In the data, 76 of 1,441 pitches offer 30% or more, about 5.3%, more than twice the answer of the AI. Check the histogram before you accept an answer that depends on a normal model.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Pick one
continuous quantity that matters for it, for example a delivery time, the
weekly revenue, or the time a visitor spends on your website. Sentence 1:
which interval of values would make you take a real decision? Sentence 2:
does a uniform or a normal model seem reasonable for that quantity, and which feature of its histogram would
you check first? There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus venture that sells boxes of prepared meals. "Delivery time is my quantity: for any delivery over 40 minutes the customer gets a refund, so $P(T > 40)$ is the number that matters. A normal model is doubtful because times cannot be negative and a few deliveries take very long. Right skew is the first thing I would check on the histogram."
:::

You can discuss this task, and any other task in this chapter, with
Tilly for QM Basic, the course AI assistant: ask it to suggest a
distribution for your quantity. Then question whether the assumptions of
the normal model really hold for it.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask Tilly for QM Basic,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 6. Show the solution only after I send my answer."

[Continue to Tutorial 2: Conditioning and relationships](t02-exam-prep.qmd).

[Formula sheet: Lecture 6 formulas](formula-sheet.qmd#formula-l6).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 10.

:::
