# Lecture 7: Samples, parameters, and estimators {#b07}

```{r}
#| label: b07-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 7, you can</span>

- distinguish a population, sampling frame, sample, and target population
- explain simple random sampling with and without replacement, and apply
  the 10% working rule
- distinguish parameters, estimators, and estimates, and write estimators
  with capital letters and estimates with lower-case letters
- explain why a sample proportion is a sample mean
- say what it means that an estimator is unbiased
- explain the $n-1$ degrees of freedom in a sample variance
- describe how an estimate varies from sample to sample while the parameter
  stays fixed
- classify sampling methods and name the main threat of each.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- name the population, the frame, and the sample in a short story
- tell a parameter, an estimator, and an estimate apart
- classify a sampling method and name its main weakness
- say what it means that an estimator is unbiased
- explain why the sample variance divides by $n-1$.

**Enough to recognise:**

- capital letters for random quantities ($\bar X$) and lower-case letters for observed values ($\bar x$)
- the term independent and identically distributed (iid).
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Checking every record, customer, or transaction is usually too expensive, so firms audit a sample and act on what it shows. That only works if a small sample can be trusted to speak for the whole.</p>

We can only afford to inspect 100 of the 1,441 pitches in the Shark Tank
file. What can 100 randomly chosen records tell us about all of them, and
what can they never tell us about entrepreneurs in general?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=nuXDb9B3y0M">The Sample
Mean and Some Terminology</a>. Be ready to distinguish a population
quantity from a statistic calculated after a sample is observed. The link
is also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (10 minutes).** Suppose we want the mean equity offered
(`equity_offered_pct`) across the 1,441 recorded pitches in completed
Seasons 1--16, but can inspect only 100 randomly selected records. Identify:

1. the finite population
2. the population variable
3. the sample
4. the parameter of interest
5. a statistic that could estimate it.

::: {.callout-note collapse="true"}
## Check your answers

The finite population is the 1,441 recorded pitches, the variable is
initial equity offered, the sample is the 100 selected records, the
parameter is the population mean equity offer, and the sample mean is a
suitable estimator.
:::

**Predict (5 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: two students each draw their own random
sample of 100 pitches from the file and each compute the mean equity
offered. Will their two sample means be **equal**? Commit to an answer. You
find the answer in the lecture material.

If subscripts or summation notation are not secure, review
[Mathematics Refresher: indices and sums](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section draws several random samples from the Shark Tank file and
compares their means. It then gives names to what you see: population,
sample, parameter, estimator, and estimate.

### Start with a founder's question

Imagine a venture with 1,441 customers. You want to know what they pay on
average, and asking all of them costs too much. Can 100 of them, chosen by
chance, tell you? Our file lets us try this out, because here we know the
answer for all 1,441 records. Take the equity offered
(`equity_offered_pct`) as the number of interest:

```{r}
#| label: b07-opening
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mean(sharks$equity_offered_pct)                 # all 1,441 records
set.seed(123)
mean(sample(sharks$equity_offered_pct, 100))    # 100 records chosen by chance
```

All 1,441 records give a mean equity offered of 13.21%. The 100 records
chosen by chance give 12.76%. That is close, but it is not the same. Two
questions follow. Would another 100 records give the same number? And how
far off can 100 records be? This lecture answers the first question.
Lecture 8 answers the second.

### Five samples, five means

The code below draws five different random samples of 100 pitches, one after
another. For each sample it keeps one number: the mean equity offered. The
number we try to recover is the mean of all 1,441 records:

```{r}
#| label: b07-parameter
mu_equity <- mean(sharks$equity_offered_pct)
mu_equity
```

It is about 13.21%. We know it only because we hold the full file. In a real
study you hold only the sample. That is why the behaviour of the sample mean
matters.

Each sample of 100 has its own seed. A seed fixes the random draw, so you get
the same five numbers when you run the code yourself. The option
`replace = FALSE` means that each record can be drawn only once:

```{r}
#| label: b07-five-samples
seeds <- 101:105
sample_means <- numeric(length(seeds))
for (i in seq_along(seeds)) {
  set.seed(seeds[i])
  # replace = FALSE: each record can be drawn only once
  rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
  sample_means[i] <- mean(sharks$equity_offered_pct[rows])
}
round(sample_means, 2)
```

Five samples, five different means, from about 12.0 to about 13.8.
@fig-b07-five-means puts them on one axis, with the population mean drawn as
a fixed vertical line.

```{r}
#| label: fig-b07-five-means
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.2
#| fig-cap: "Five random samples of 100, five sample means (green dots), one fixed population mean (gold dashed line at 13.21). The sample mean varies from draw to draw. The population mean stays fixed."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)

seeds <- 101:105
sample_means <- numeric(length(seeds))
for (i in seq_along(seeds)) {
  set.seed(seeds[i])
  rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
  sample_means[i] <- mean(sharks$equity_offered_pct[rows])
}
five_means <- data.frame(draw = seq_along(seeds),
                         seed = paste("seed", seeds),
                         sample_mean = sample_means)

ggplot(five_means, aes(sample_mean, draw)) +
  geom_vline(xintercept = mu_equity, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_point(colour = "#009E73", size = 3) +
  annotate("text", x = mu_equity + 0.12, y = length(seeds) + 0.45,
           label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_y_continuous(breaks = five_means$draw, labels = five_means$seed) +
  coord_cartesian(xlim = c(10, 16),
                  ylim = c(0.5, length(seeds) + 0.5)) +
  labs(x = "Sample mean equity offered (%)", y = "Draw") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each green dot is the mean of one sample. A number computed from a sample
to approximate a number of the population is called an estimate. The gold
dashed line is the mean of all 1,441 records. Such a fixed number of the
population is called a parameter. The dots scatter around the line. None
hits it exactly and none is far from it. No dot is wrong, and the scatter is
not a mistake in the procedure. Random samples give estimates that differ by
chance.

This answers the prediction question from before class: two students who
draw their own random samples will almost never get equal means. Any two of
the five dots differ, and your own sample (in the after-class task below)
will differ again.

One question remains. If we drew a sixth sample, where would its mean land?
You cannot say exactly, but the five dots suggest a range in which it would
probably fall. The lab below shows that range for many more samples.
Lecture 8 gives a formula for its width.

### Explore it: the sampling-distribution lab {#sampling-lab}

Five dots are a start. The lab below draws many more samples. On the left it
shows the histogram of the equity offered in all 1,441 records, which is
strongly right-skewed. Choose a sample size $n$ with the **slider** and press
**Draw 100 samples**. The mean of each sample is added to the histogram on
the right. **Reset** clears that histogram.

<iframe src="../assets/widgets/sampling-lab.html" width="100%" height="520" style="border:1px solid #C9D4DC; border-radius:6px;" title="Sampling distribution lab"></iframe>

Work through three settings. Each time, look at the centre, the spread, and
the shape of the histogram of means.

1. Set $n = 2$ and redraw several times. The means scatter widely and the
   histogram of means is still skewed.
2. Set $n = 25$. The histogram of means narrows and starts to look
   symmetric, centred near 13.2.
3. Slide up to $n = 100$. The histogram of means is a tight, bell-shaped
   pile around the population mean, even though the population histogram
   beside it has not changed at all.

The pile of sample means on the right has a name.

::: {.definition-box}
<span class="label">Definition</span>

**Sampling distribution.** A statistic is a number computed from a sample,
such as the sample mean. The sampling distribution of a statistic is its
distribution across all possible random samples of the same size. It shows
how the statistic varies from sample to sample.
:::

The histogram in the lab approximates the sampling distribution of the mean.
Lecture 8 explains where it is centred, how wide it is, and why it becomes
bell-shaped.

The rest of this lecture gives names to what you just did. It also says when
100 records can speak for all records, and when they cannot.

### Define the population before calculating

::: {.definition-box}
<span class="label">Definition</span>

**Population and sample.** The population is the complete set of units about
which a question is asked. The sample is the part of the population that is
observed. The number of units in the population is written $N$, and the
number of units in the sample is written $n$.
:::

Two more sets matter. A **sampling frame** is the list from which the sample
is selected. The **target population** is the broader group to which we hope
to generalise.

These sets need not coincide. The teaching file lists the recorded US
*Shark Tank* pitches from completed Seasons 1--16. It is not a random sample
of all entrepreneurs, all funding pitches, or even all applicants to the
show. A random sample of rows from this file supports conclusions about the
population of 1,441 records ($N=1{,}441$). It does not remove the show's
selection process.

::: {.callout-note collapse="true"}
## Check: classify two scenarios

**Scenario A.** A food-delivery startup wants to know how satisfied "its
customers" are. It emails a survey to the app users who opted in to
marketing emails. 800 reply. Identify the target population, the frame,
and the sample.

**Scenario B.** A consultant uses our 100 sampled pitches to advise "US
entrepreneurs seeking investors." What is the frame, and what is the
target population being claimed?

**Answers.** In A, the target population is all customers, the frame is
the opted-in email list (which already excludes many customers), and the
sample is the 800 who chose to reply. These 800 are volunteers from an
incomplete frame. In B, the frame is the 1,441 recorded *Shark Tank*
pitches. The claimed target population is all US entrepreneurs seeking
investors. Random sampling of rows supports conclusions about the frame, not
about that far larger target group.
:::

### Simple random sampling

In a simple random sample of size $n$, every set of $n$ population units
has the same chance of selection.

Before the sample is drawn, we do not know which unit will come first. The
value of the first drawn unit is therefore a random variable, written $X_1$.
The same holds for the second draw, $X_2$, and so on up to $X_n$. The
sequence $X_1,\ldots,X_n$ is also called a random sample. After the draw we
hold numbers, the observed values $x_1,\ldots,x_n$. They are also called
realised values. As in Lecture 2, capital letters stand for random variables
and lower-case letters for their values.

There are two ways to draw:

- **With replacement:** a selected unit is returned before the next draw.
  The draws are independent and each draw has the same distribution.
- **Without replacement:** a unit cannot be selected twice. Each draw taken
  alone still has the same distribution. But the draws are dependent, because
  one draw changes which units are left.

Draws that are independent and all have the same distribution are called
**independent and identically distributed (iid)**.

The samples analysed in this book are drawn without replacement, so their
draws are not exactly independent. When the sample is a small part of a large
population, the dependence is weak. This course uses the 10% condition as a
working rule: if the sample size $n$ is at most 10% of the population size
$N$, written $n\le 0.10N$, we treat the draws as independent, and so as iid.

The sample used in the rest of this lecture and in Lecture 8 is drawn with
seed 123, so every reader gets the same 100 rows:

```{r}
#| label: b07-draw-sample
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

nrow(pitch_sample)
```

Here $100/1441\approx 0.069<0.10$, so the working rule is satisfied.

How does this one observed sample compare with the population it came from?

```{r}
#| label: fig-b07-population-sample
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Initial equity offered: all 1,441 records (grey bars) and the seed-123 random sample of 100 (green connected points). The sample resembles the population but does not reproduce it exactly."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

breaks <- seq(0, 100, by = 5)
pop_hist <- hist(sharks$equity_offered_pct, breaks = breaks, plot = FALSE)
samp_hist <- hist(pitch_sample$equity_offered_pct, breaks = breaks,
                  plot = FALSE)
pop_density <- data.frame(mid = pop_hist$mids, density = pop_hist$density)
samp_density <- data.frame(mid = samp_hist$mids, density = samp_hist$density)

ggplot() +
  geom_col(data = pop_density, aes(mid, density), width = 5,
           fill = "grey75", alpha = 0.85, colour = "white") +
  geom_line(data = samp_density, aes(mid, density),
            colour = "#009E73", linewidth = 0.9) +
  geom_point(data = samp_density, aes(mid, density),
             colour = "#009E73", size = 2.2) +
  annotate("text", x = 40, y = 0.070, label = "all 1,441 records",
           colour = "#5B6976", size = 3.4, hjust = 0) +
  annotate("text", x = 40, y = 0.062, label = "random sample of 100",
           colour = "#009E73", size = 3.4, hjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Equity initially offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The horizontal axis shows the equity offered in intervals of 5 percentage
points. The vertical axis is density, so the two datasets of different size
can be compared by shape. The sample resembles the population but does not
reproduce it exactly. A different random sample would resemble it in a
different way, as the five samples above showed.

### Parameters, estimators, and estimates {#b07-parameters}

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-sample-mean-and-some-terminology/">20.3 The Sample Mean and Some Terminology</a>.</p>

A sample is drawn to learn a number that describes the population. Three
objects must be kept apart.

::: {.definition-box}
<span class="label">Definition</span>

**Parameter, estimator, estimate.** A parameter is a fixed number that
describes the population. It is usually unknown. An estimator is a rule that
calculates a number from a random sample $X_1,\ldots,X_n$. It is a random
variable, because the sample is random. An estimate is the number that the
rule gives for one observed sample.
:::

The table shows the three objects for the mean, the variance, and the
proportion of a 0/1 variable.

| | Parameter (population) | Estimator (rule, capital letters) | Estimate (number from one sample, lower case) |
|---|---|---|---|
| Mean | $\mu=\dfrac{1}{N}\sum_{i=1}^{N}x_i$ | $\bar X=\dfrac{1}{n}\sum_{i=1}^{n}X_i$ | $\bar x=\dfrac{1}{n}\sum_{i=1}^{n}x_i$ |
| Variance | $\sigma^2=\dfrac{1}{N}\sum_{i=1}^{N}(x_i-\mu)^2$ | $S^2=\dfrac{1}{n-1}\sum_{i=1}^{n}(X_i-\bar X)^2$ | $s^2=\dfrac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar x)^2$ |
| Proportion (0/1 variable) | $p=\dfrac{1}{N}\sum_{i=1}^{N}x_i$ | $\widehat P=\dfrac{1}{n}\sum_{i=1}^{n}X_i$ | $\hat p=\dfrac{1}{n}\sum_{i=1}^{n}x_i$ |

In the first column the sums run over all $N$ units of the population. In
the other two columns they run over the $n$ units of the sample. Capital
letters describe random quantities before the draw. Lower-case letters
describe observed values. The standard deviations are the square roots
$\sigma$, $S$, and $s$.

The parameters $\mu$ and $\sigma^2$ are the expectation and the variance of
[Lecture 3](b03-discrete-distributions.qmd#b03), for the value of one unit
drawn at random from the population. The variance squares the deviations
from the mean because the plain deviations always add up to zero: the
positive and the negative ones cancel. The standard deviation is the square
root of the variance, so it is in the units of the variable.

A proportion is the mean of a 0/1 variable. The sum of the zeros and ones
counts the ones, and dividing by the number of units gives their share. For a
0/1 variable the population variance is $\sigma^2=p(1-p)$ (Lecture 3).

The estimator $S^2$ divides by $n-1$ and not by $n$. The sections on
unbiased estimators and on degrees of freedom below give the reason.

::: {.callout-tip collapse="true"}
## Optional: sampling without replacement, formally

Without replacement, the draws $X_1,\ldots,X_n$ all have the same
distribution, with mean $\mu$ and variance $\sigma^2$. They are not
independent.

The population variance has a shortcut formula:
$\sigma^2=\frac{1}{N}\sum_{i=1}^{N}x_i^2-\mu^2$.

Lecture 8 shows that $\operatorname{Var}(\bar X)=\sigma^2/n$ for independent
draws. Without replacement the variance is slightly smaller:

$$\operatorname{Var}(\bar X)=\frac{\sigma^2}{n}\cdot\frac{N-n}{N-1}.$$

The second factor is called the finite population correction. For $n=100$
and $N=1{,}441$ it equals $1341/1440\approx 0.93$. It is close to 1 whenever
the 10% condition holds, which is why the course leaves it out.
:::

The estimates for our seed-123 sample:

```{r}
#| label: b07-sample-estimates
c(
  sample_mean_equity = mean(pitch_sample$equity_offered_pct),
  sample_sd_equity = sd(pitch_sample$equity_offered_pct),
  sample_deal_proportion = mean(pitch_sample$deal_on_show)
)
```

For this sample, the mean equity offered is $\bar x=12.76$%, its standard
deviation is $s=7.71$ percentage points, and the deal proportion is
$\hat p=0.64$.

::: {#exm-b07-three-objects}
## Worked example: three objects for the deal proportion

Identify the parameter, the estimator, and the estimate for the proportion
of deals (`deal_on_show`), using the seed-123 sample.

- **Parameter:** $p$, the proportion of the 1,441 records with an on-air
  deal. One fixed number. Because the file is the whole frame, we can even
  look it up: $882/1441\approx 0.612$. In a real study it would be unknown.
- **Estimator:** $\widehat P=\frac{1}{100}\sum_{i=1}^{100}X_i$, where
  $X_i$ is the deal indicator of the $i$-th sampled record. It is a rule,
  and its value is random because the sample is random.
- **Estimate:** $\hat p=0.64$, the number the rule produced for the
  seed-123 sample. A different seed would produce a different $\hat p$,
  while $p$ would stay exactly where it is.
:::

### Unbiased estimators {#b07-unbiased}

An estimator is a random variable, so it has a distribution: its sampling
distribution. The five dots in @fig-b07-five-means are five draws from the
sampling distribution of $\bar X$. The first question about an estimator is
where this distribution is centred.

::: {.definition-box}
<span class="label">Definition</span>

**Unbiased estimator.** Write $\theta$ (the Greek letter theta) for any
parameter and $\hat\theta$ for an estimator of it. The estimator is unbiased
if its sampling distribution is centred on the parameter:
$E[\hat\theta]=\theta$.
:::

In words: the average of the estimator across all possible random samples
equals the parameter. An unbiased estimator does not overestimate or
underestimate systematically. A biased estimator is centred beside the
parameter, as @fig-b07-unbiased shows.

```{r}
#| label: fig-b07-unbiased
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.6
#| fig-cap: "Sampling distributions of two estimators of the same parameter (an illustration, not data). The green curve is centred on the parameter, marked by the gold dashed line, so this estimator is unbiased. The orange curve is centred to the right of the parameter, so this estimator overestimates on average."
library(ggplot2)

grid <- seq(-3.5, 5.5, length.out = 400)
curves <- rbind(
  data.frame(x = grid, estimator = "unbiased",
             density = dnorm(grid, mean = 0, sd = 1)),
  data.frame(x = grid, estimator = "biased",
             density = dnorm(grid, mean = 2, sd = 1))
)

ggplot(curves, aes(x, density, colour = estimator)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_line(linewidth = 0.9) +
  annotate("text", x = -0.15, y = 0.455, label = "parameter",
           colour = "#17324D", size = 3.4, hjust = 1) +
  annotate("text", x = -1.25, y = 0.30,
           label = "unbiased estimator:\ncentred on the parameter",
           colour = "#009E73", size = 3.4, hjust = 1, lineheight = 0.95) +
  annotate("text", x = 3.25, y = 0.30,
           label = "biased estimator:\ncentred beside the parameter",
           colour = "#D55E00", size = 3.4, hjust = 0, lineheight = 0.95) +
  scale_colour_manual(values = c(unbiased = "#009E73", biased = "#D55E00"),
                      guide = "none") +
  scale_x_continuous(breaks = 0, labels = expression(theta),
                     limits = c(-5.2, 7.6)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     limits = c(0, 0.47)) +
  labs(x = "Value of the estimator", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Unbiased says nothing about one sample. Each of the five dots misses 13.21.
The first one, 12.03, misses by 1.18 percentage points, which is about 9% of
the population mean.

The sample mean $\bar X$ is unbiased for $\mu$. That is why the five dots
scatter around the gold line and not to one side of it. Lecture 8 derives
this result. Dividing by $n-1$ makes $S^2$ an unbiased estimator of $\sigma^2$ under
independent random sampling. $S$ itself slightly underestimates $\sigma$ on
average, which matters little for large samples.

### Why the sample variance has $n-1$ degrees of freedom

Once $\bar X$ has been calculated from the same sample, the deviations must
satisfy

$$\sum_{i=1}^{n}(X_i-\bar X)=0.$$

If $n-1$ deviations are known, the final deviation is forced to make the
sum zero. Only $n-1$ deviations can vary freely: the sample variance has
$n-1$ **degrees of freedom**.

::: {#exm-b07-df}
## Worked example: three deviations, two degrees of freedom

Suppose a sample has $n=3$ observations and the first two deviations from
the sample mean are $2$ and $-5$.

*Step 1.* The deviations must sum to zero:
$2+(-5)+d_3=0$.

*Step 2.* Solve: $d_3=3$. The third deviation cannot be chosen freely:
it is forced.

*Step 3.* Count the freedom: three deviations, but only two carried any
information once the sample mean was fixed. The sample variance for this
sample has $3-1=2$ degrees of freedom.
:::

The deviations are measured from $\bar X$, the mean of the same sample, and
not from $\mu$. This makes the squared deviations slightly too small on
average as a measure of the spread in the population. Dividing by $n-1$ in
place of $n$ corrects this downward bias. Degrees of freedom count the
pieces of information that can still vary freely. They are not a percentage
or a probability.

### How the sample was drawn decides what it can show

The number of rows does not tell you how the sample was drawn. Ask how the
units entered the data:

| Method | How units enter | Main implication |
|---|---|---|
| Simple random sample | selected by a known random mechanism | supports conclusions about the whole frame |
| Stratified random sample | randomly selected within defined groups | can guarantee representation of important groups |
| Convenience sample | selected because they are easy to reach | vulnerable to selection bias |
| Voluntary-response sample | units choose whether to respond | people with stronger views may be over-represented |
| Census of a frame | every unit in the available frame is recorded | no sampling error for that frame, but coverage error can remain |

Sampling error is the chance difference between an estimate and the
parameter. Coverage error arises when the frame leaves out part of the
target population.

The Shark Tank file is close to a census of the recorded pitches. When we
draw 100 rows from it, the chance comes from our own draw. It does not make
the television programme's selection of pitches random.

::: {.callout-caution collapse="true"}
## Try it: classify four recruitment stories

Try this with your neighbour first. For each story, name the sampling method
from the table and state its main threat.

*Story 1 (given, as a model).* "A ministry lists all registered firms,
groups them by province, and randomly selects firms within each province."
Method: stratified random sample. Main implication: representation of
every province is guaranteed by design.

*Story 2 (you).* "A startup posts a poll on its own Instagram account
asking followers to rate a prototype. 412 followers respond."

*Story 3 (you).* "A student emails the survey to friends, family, and two
student associations she belongs to."

*Story 4 (you).* "A researcher numbers all 8,214 firms in the Chamber of
Commerce register and selects 300 using a random number generator."

::: {.callout-note collapse="true"}
## Check your answer

Story 2: voluntary-response sample. Followers already like the brand, and
those with the strongest opinions are most likely to reply. Story 3:
convenience sample. Contacts that are easy to reach differ systematically
from the wider population, so selection bias is the main threat. Story 4:
simple random sample from the register. It supports conclusions about the
registered firms. Firms missing from the register remain out of reach
(coverage error).
:::
:::

Random selection reduces systematic over-representation within the
sampling frame. It does not guarantee that one observed sample is
perfectly balanced: by chance, a sample can contain unusually many large
equity offers or unusually many deals. Larger samples reduce this chance
variation. The dots in @fig-b07-five-means would lie closer together if
each dot summarised 1,000 records in place of 100.

::: {.caveat-box}
<span class="label">What random selection can and cannot repair</span>

- Random sampling lets 100 records speak for the list they were drawn from.
  It does not repair a list that leaves people out.
- Random assignment, which gives a treatment to units chosen by chance, is a
  different tool. It supports claims about cause.
- "Unbiased" describes the method across many samples. Your one sample can
  still be off: 12.76 is not 13.21.

**What you can do with it:** when you survey customers, draw them by chance
from the fullest list you have, and say which customers the list leaves out.
:::

### Answering the driving question

One hundred randomly chosen records give estimates of the parameters of the
Shark Tank file. The seed-123 sample estimates the mean equity offered
(`equity_offered_pct`) at $\bar x=12.76$%, while the population mean is
$\mu=13.21$%. It estimates the deal proportion at $\hat p=0.64$, while
$p\approx 0.612$. Five more samples put the sample mean between about 12.0%
and 13.8%, and the parameter did not move. So random samples tell us about
the frame, with a random error. Lecture 8 measures the size of that error.

What can they never tell us? Anything reliable about entrepreneurs in
general. The frame is the set of pitches that a television programme chose
to show, and a random draw of rows cannot undo that selection.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You have 400 customers and want to know how satisfied they are. An Instagram poll would be easy, and perhaps 200 followers would answer. A simple random sample of 40 customers from your full customer list is more work: you draw the names with `sample()` and write to each of them. The decision: take the random 40. The poll answers come from your biggest fans, and 200 of them tell you less about all customers than 40 chosen by chance.
:::

**Next:** The parameter stays fixed while the estimator varies from sample to
sample. Five sample means landed in five different places. Lecture 8 studies
the pattern behind those dots: the sampling distribution of the mean, the
standard error that measures its spread, and the central limit theorem. This
theorem explains why averages become bell-shaped even when the data are
strongly skewed. With that pattern we can say how far an estimate is likely
to be from the parameter.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 70 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. What is the difference between a target population and a sampling frame?
2. Why are draws without replacement dependent?
3. What is the difference between an estimator and an estimate?
4. Why is a sample proportion a sample mean?
5. Why does the sample variance have $n-1$ degrees of freedom?
6. What does random sampling support that random assignment does not?
7. An estimator is unbiased. What does that say about many samples, and what
   about one sample?

::: {.callout-note collapse="true"}
## Check your answers

1. The target population is the group you want to learn about. The
   sampling frame is the list from which units can actually be selected.
2. Selecting one unit changes which units remain available.
3. An estimator is a random rule before data are observed. An estimate is
   its realised numerical value.
4. A 0/1 indicator sums to the number of successes, so its mean is the
   success proportion.
5. Estimating the sample mean imposes one constraint: all deviations from
   it must sum to zero. Only $n-1$ deviations can vary freely.
6. Random sampling supports generalisation to the sampling frame. Random
   assignment supports causal comparison of treatments.
7. Across all possible random samples, the average of the estimator equals
   the parameter: $E[\hat\theta]=\theta$. It says nothing about one sample,
   which can still be off.
:::

### Practice (15 minutes)

1. For a population of $N=20{,}000$ customers and a simple random sample
   of $n=500$, check the 10% condition.
2. Write the estimator and realised estimate for the population deal
   proportion.
3. Explain why $\mu$ is fixed while $\bar X$ is random.
4. Using `pitch_sample`, calculate the sample mean and sample SD of
   `description_words`.
5. Explain why this random sample cannot justify conclusions about all US
   entrepreneurs.
6. For a sample of five observations, state the degrees of freedom used to
   estimate variance and explain the number.

::: {.callout-note collapse="true"}
## Check your answers

1. $500/20{,}000=0.025<0.10$, so the condition is satisfied.
2. The estimator is $\widehat P=\frac{1}{n}\sum_iX_i$. The realised estimate
   is $\hat p=\frac{1}{n}\sum_ix_i$, here $0.64$.
3. The population and its mean are treated as fixed. The selected
   observations, and therefore their mean, vary across possible samples.
4. Rebuild the sample and compute. The mean is about 4.82 words and the SD
   is about 1.76 words:

```{r}
#| label: b07-practice-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
c(mean = mean(pitch_sample$description_words),
  sd   = sd(pitch_sample$description_words))
```

5. The sample was selected from televised records, not from all US
   entrepreneurs. Random row selection cannot repair that coverage gap.
6. The degrees of freedom are $5-1=4$. Once the sample mean is estimated,
   four deviations may vary freely and the fifth is fixed by their zero
   sum.
:::

### Draw your own sample (10 minutes)

Add a dot of your own to the class version of @fig-b07-five-means. Use
your student number as the seed, draw your own sample of 100, and compute
the mean equity offered in your sample:

```{r}
#| label: b07-own-sample
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
my_seed <- 2412345   # replace with your own student number
set.seed(my_seed)
rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
mean(sharks$equity_offered_pct[rows])
```

Submit your mean to the shared class plot on Canvas (the Wooclap link
is posted under Lecture 7). Every student uses a different seed, so every
student contributes a different dot.

::: {.callout-note collapse="true"}
## Check: what will the class plot show?

The dots of the class will not scatter arbitrarily. They will pile up around
13.2 and thin out towards the edges. That shape is the sampling distribution
of the mean. Lecture 8 describes its centre, its spread, and its shape.
:::

### Review: one question from Lecture 6 (5 minutes)

Processing time is modelled as $Y\sim N(50,100)$, and
$P(Z\le 1.5)=0.9332$ is supplied. Find $P(Y>65)$ by hand.

::: {.callout-note collapse="true"}
## Check your answer

$\sigma=\sqrt{100}=10$ and $z=(65-50)/10=1.5$, so
$P(Y>65)=1-P(Z\le 1.5)=1-0.9332=0.0668$: under this model, about 6.7% of
processing times exceed 65 minutes.
:::

### Worked exam interpretation (5 minutes)

**Question.** A retailer's loyalty programme has 52,000 members. An
analyst draws 400 members at random. Their mean January spend is
&euro;87.40. A slide then claims: "the average member spends &euro;87.40
per month." Identify the parameter, the estimator, and the estimate, and
give two distinct objections to the claim.

::: {.callout-note collapse="true"}
## Check a model answer

The parameter is the mean January spend across all 52,000 members (fixed,
unknown). The estimator is the sample mean $\bar X$ of 400 randomly drawn
members' spends. The estimate is $\bar x=87.40$ euros. Objections: first,
the estimate carries sampling error. Another random 400 members would
give a different mean, so &euro;87.40 should be reported with its
uncertainty (Lectures 8--9). Second, the claim generalises a January
figure to "per month". The data cannot support that, because they cover
the spending of one month only.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "I asked my 200 Instagram followers and 150 would buy my product. I want to sell to all students on campus. Can I plan with 75%?" It answered:

> "Yes. A sample of 200 is large, so 75% is a reliable estimate for all students on campus."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

A large sample does not repair how the people got into it. Followers who answer a poll like the brand already. The 75% describes the 200 followers who answered, not the students on campus. A random sample of 40 students from a full list would tell you more.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Write two
sentences. First name one group you want to learn about (your target
population) and the best list of its members you could actually get (your
frame). Then say where the gap between the two would matter if you sampled
from that list. There is no wrong answer.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I want to learn about all students on campus who buy lunch at least twice a week, but the best list I can actually get is my own mailing list. Sampling from that list overrepresents existing fans, so any satisfaction estimate would be too rosy for the market I want to enter next."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it which groups your frame leaves
out and who would therefore be missing from your sample.
::::

[Continue to Lecture 8: Sampling distributions and the central limit theorem](b08-sampling-distributions.qmd).

[Formula sheet: Lecture 7 formulas](formula-sheet.qmd#formula-l7).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 12.

:::
