# Lecture 7: Samples, parameters, and estimators {#b07}

```{r}
#| label: b07-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 7, you can</span>

- distinguish a population, sampling frame, sample, and target population;
- explain simple random sampling with and without replacement, and apply
  the 10% working rule;
- distinguish parameters, estimators, and estimates, with the capital and
  lower-case notation discipline;
- explain why a sample proportion is a sample mean;
- explain the $n-1$ degrees of freedom in a sample variance;
- describe how an estimate varies from sample to sample while the parameter
  sits still;
- classify sampling methods and name the main threat of each.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Checking every record, customer, or transaction is usually too expensive, so firms audit a sample and act on what it shows. That only works if a small sample can be trusted to speak for the whole.</p>

We can only afford to inspect 100 of the 1,441 pitches. What can 100
randomly chosen records tell us about all of them --- and what can they
never tell us about entrepreneurs in general?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=nuXDb9B3y0M">The Sample
Mean and Some Terminology</a>. Be ready to distinguish a population
quantity from a statistic calculated after a sample is observed. The link
is also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (10 minutes).** Suppose we want the mean initial equity
offered across the 1,441 recorded pitches in completed Seasons 1--16, but
can inspect only 100 randomly selected records. Identify:

1. the finite population;
2. the population variable;
3. the sample;
4. the parameter of interest;
5. a statistic that could estimate it.

::: {.callout-note collapse="true"}
## Check your answers

The finite population is the 1,441 recorded pitches; the variable is
initial equity offered; the sample is the 100 selected records; the
parameter is the population mean equity offer; and the sample mean is a
suitable estimator.
:::

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: two students each draw their own random
sample of 100 pitches from the file and each compute the mean equity
offered. Will their two sample means be **equal**? Commit to an answer; the
lecture resolves it.

If subscripts or summation notation are not secure, review
[Mathematics Refresher: indices and sums](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. Your prep prediction --- do two random
samples give equal means? --- stays on screen, unresolved, until the
sampling activity below settles it with evidence.

### Define the population before calculating

A **population** is the complete set of units about which a question is
asked. A **sample** is the subset observed for analysis. A **sampling
frame** is the list or mechanism from which the sample is selected. The
**target population** is the broader group to which we hope to generalise.

These sets need not coincide. The teaching file is a near-census of
recorded US *Shark Tank* pitches from completed Seasons 1--16, but it is
not a random sample of all entrepreneurs, all funding pitches, or even all
applicants to the show. Randomly sampling rows from this file supports
inference to the 1,441-record population. It does not remove the show's
selection process.

::: {.callout-note collapse="true"}
## Check: classify two scenarios

**Scenario A.** A food-delivery startup wants to know how satisfied "its
customers" are. It emails a survey to the app users who opted in to
marketing emails; 800 reply. Identify the target population, the frame,
and the sample.

**Scenario B.** A consultant uses our 100 sampled pitches to advise "US
entrepreneurs seeking investors." What is the frame, and what is the
target population being claimed?

**Answers.** In A, the target population is all customers; the frame is
the opted-in email list (which already excludes many customers); the
sample is the 800 who chose to reply --- a voluntary-response subset of an
incomplete frame. In B, the frame is the 1,441 recorded *Shark Tank*
pitches; the claimed target population is all US entrepreneurs seeking
investors. Random row sampling licenses conclusions about the frame, not
about that far larger target group.
:::

### Simple random sampling

In a simple random sample of size $n$, every set of $n$ population units
has the same chance of selection.

- **With replacement:** a selected unit is returned before the next draw.
  Draws are independent and each draw has the same distribution.
- **Without replacement:** a unit cannot be selected twice. Draws are
  dependent, although every draw has the same marginal distribution.

When a without-replacement sample is a small fraction of a large
population, the dependence is weak. This course uses the 10% condition as
a practical working rule: if $n\le 0.10N$, the usual independent-sample
standard-error formulas are a good approximation.

The running sample for this lecture and the next is drawn with a visible
seed, so every reader gets the same 100 rows:

```{r}
#| label: b07-draw-sample
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

nrow(pitch_sample)
```

Here $100/1441\approx 0.069<0.10$, so the working rule is satisfied.

How does this one realised sample compare to the population it came from?

```{r}
#| label: fig-b07-population-sample
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Initial equity offered: all 1,441 records (grey bars) and the seed-123 random sample of 100 (green connected points). The sample resembles the population but does not reproduce it exactly."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

breaks <- seq(0, 100, by = 5)
pop_hist <- hist(sharks$equity_offered_pct, breaks = breaks, plot = FALSE)
samp_hist <- hist(pitch_sample$equity_offered_pct, breaks = breaks,
                  plot = FALSE)
pop_density <- data.frame(mid = pop_hist$mids, density = pop_hist$density)
samp_density <- data.frame(mid = samp_hist$mids, density = samp_hist$density)

ggplot() +
  geom_col(data = pop_density, aes(mid, density), width = 5,
           fill = "grey75", alpha = 0.85, colour = "white") +
  geom_line(data = samp_density, aes(mid, density),
            colour = "#009E73", linewidth = 0.9) +
  geom_point(data = samp_density, aes(mid, density),
             colour = "#009E73", size = 2.2) +
  annotate("text", x = 40, y = 0.070, label = "all 1,441 records",
           colour = "#5B6976", size = 3.4, hjust = 0) +
  annotate("text", x = 40, y = 0.062, label = "random sample of 100",
           colour = "#009E73", size = 3.4, hjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Equity initially offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The horizontal axis gives equity intervals in percentage points; the
vertical axis is density, so the differently sized datasets can be compared
by shape. The sample resembles the population but does not reproduce it
exactly --- and a *different* random sample would resemble it differently.
That observation is worth turning into an experiment.

### The live sampling activity: watching an estimator vary

In class this segment runs live. The whole class watches five different
random samples being drawn, one after another, and tracks one number per
sample: the mean equity offered. The target of the exercise is the
population mean of the file,

```{r}
#| label: b07-parameter
mu_equity <- mean(sharks$equity_offered_pct)
mu_equity
```

about 13.21 percentage points --- knowable here only because we happen to
hold the full file. In a real study you would hold *only* the sample, which
is exactly why the behaviour of the sample mean matters.

Five samples of 100, each from its own seed, each summarised by its mean.
The code is fully reproducible --- run it yourself and you will get the
same five numbers:

```{r}
#| label: b07-five-samples
seeds <- 101:105
sample_means <- numeric(length(seeds))
for (i in seq_along(seeds)) {
  set.seed(seeds[i])
  rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
  sample_means[i] <- mean(sharks$equity_offered_pct[rows])
}
round(sample_means, 2)
```

Five samples, five different means --- from about 12.0 to about 13.8.
@fig-b07-five-means puts them on one fixed axis, with the population mean
drawn as a vertical line that never moves.

```{r}
#| label: fig-b07-five-means
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-height: 3.2
#| fig-cap: "Five random samples of 100, five sample means (green dots), one fixed population mean (gold dashed line at 13.21). The estimate varies from draw to draw; the parameter sits still."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)

seeds <- 101:105
sample_means <- numeric(length(seeds))
for (i in seq_along(seeds)) {
  set.seed(seeds[i])
  rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
  sample_means[i] <- mean(sharks$equity_offered_pct[rows])
}
five_means <- data.frame(draw = seq_along(seeds),
                         seed = paste("seed", seeds),
                         sample_mean = sample_means)

ggplot(five_means, aes(sample_mean, draw)) +
  geom_vline(xintercept = mu_equity, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_point(colour = "#009E73", size = 3) +
  annotate("text", x = mu_equity + 0.12, y = length(seeds) + 0.45,
           label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_y_continuous(breaks = five_means$draw, labels = five_means$seed) +
  coord_cartesian(xlim = c(10, 16),
                  ylim = c(0.5, length(seeds) + 0.5)) +
  labs(x = "Sample mean equity offered (%)", y = "Draw") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read the figure slowly, because Lecture 8 is built on it. Each green dot
is one *estimate*: the realised mean of one realised sample. The gold
dashed line is the *parameter*: one fixed, unmoving number. The dots scatter
around the line --- none hits it exactly, none is far from it, and no dot
is "wrong." The scatter is not a mistake in the procedure; it *is* the
procedure. Random samples produce random estimates.

This resolves the prep prediction: two students drawing their own random
samples will almost never get equal means. Any two of the five dots differ,
and yours (in the after-class task below) will differ again.

In class the five-dot plot stays on screen through the break, with one
question to chew on: if we drew a sixth sample, where would its mean land?
You cannot say exactly --- but by now you can say something about the
*range* in which it would probably fall. Making that "something" precise is
the entire subject of Lecture 8.

### Parameters, estimators, and estimates

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-sample-mean-and-some-terminology/">20.3 The Sample Mean and Some Terminology</a>.</p>

A **parameter** is a fixed numerical property of a population, usually
unknown. An **estimator** is a rule based on random sample variables. Its
value changes from sample to sample. An **estimate** is the number obtained
from one realised sample.

For a random sample $X_1,\ldots,X_n$:

$$\bar X=\frac{1}{n}\sum_{i=1}^{n}X_i$$

is an estimator of the population mean $\mu$. After observing values
$x_1,\ldots,x_n$,

$$\bar x=\frac{1}{n}\sum_{i=1}^{n}x_i$$

is the realised estimate. Capital letters describe random quantities before
sampling; lower-case letters describe observed values.

For a 0/1 variable, the sample proportion is a sample mean:

$$\widehat P=\frac{1}{n}\sum_{i=1}^{n}X_i.$$

Its realised value is written $\hat p$.

The usual sample variance and standard deviation are

$$S^2=\frac{1}{n-1}\sum_{i=1}^{n}(X_i-\bar X)^2,
\qquad S=\sqrt{S^2}.$$

The denominator $n-1$ makes $S^2$ an unbiased estimator of population
variance under independent random sampling.

The realised estimates for our seed-123 sample:

```{r}
#| label: b07-sample-estimates
c(
  sample_mean_equity = mean(pitch_sample$equity_offered_pct),
  sample_sd_equity = sd(pitch_sample$equity_offered_pct),
  sample_deal_proportion = mean(pitch_sample$deal_on_show)
)
```

For this realised sample, $\bar x=12.76$ percentage points, $s=7.71$
percentage points, and $\hat p=0.64$.

::: {#exm-b07-three-objects}
## Worked example: three objects for the deal proportion

Identify the parameter, the estimator, and the estimate for the proportion
of deals, using our realised sample.

- **Parameter:** $p$, the proportion of the 1,441 records with an on-air
  deal --- one fixed number. Because the file is a near-census of its
  frame, we can even look it up: $882/1441\approx 0.612$. In a real study
  it would be unknown.
- **Estimator:** $\widehat P=\frac{1}{100}\sum_{i=1}^{100}X_i$, where
  $X_i$ is the deal indicator of the $i$-th sampled record --- a rule
  whose value is random because the sample is.
- **Estimate:** $\hat p=0.64$, the number the rule produced for the
  seed-123 sample. A different seed would produce a different $\hat p$
  while $p$ stayed exactly where it was.
:::

### Why the sample variance has $n-1$ degrees of freedom

Once $\bar X$ has been calculated from the same sample, the deviations must
satisfy

$$\sum_{i=1}^{n}(X_i-\bar X)=0.$$

If $n-1$ deviations are known, the final deviation is forced to make the
sum zero. Only $n-1$ deviations can vary freely: the sample variance has
$n-1$ **degrees of freedom**.

::: {#exm-b07-df}
## Worked example: three deviations, two degrees of freedom

Suppose a sample has $n=3$ observations and the first two deviations from
the sample mean are $2$ and $-5$.

*Step 1.* The deviations must sum to zero:
$2+(-5)+d_3=0$.

*Step 2.* Solve: $d_3=3$. The third deviation cannot be chosen freely ---
it is forced.

*Step 3.* Count the freedom: three deviations, but only two carried any
information once the sample mean was fixed. The sample variance for this
sample has $3-1=2$ degrees of freedom.
:::

Using the fitted sample mean makes squared deviations around $\bar X$
slightly too small on average as a measure of population spread. Dividing
by $n-1$, rather than $n$, corrects that downward bias under the
random-sample model. Degrees of freedom count remaining independent pieces
of information; they are not a percentage or a probability.

### Sampling method determines the evidence boundary

The number of rows does not identify the sampling design. Ask how units
entered the data:

| Method | How units enter | Main implication |
|---|---|---|
| Simple random sample | selected by a known random mechanism | supports design-based generalisation to the frame |
| Stratified random sample | randomly selected within defined groups | can guarantee representation of important groups |
| Convenience sample | selected because they are easy to reach | vulnerable to selection bias |
| Voluntary-response sample | units choose whether to respond | people with stronger views may be over-represented |
| Census of a frame | every unit in the available frame is recorded | no sampling error for that frame, but coverage error can remain |

The Shark Tank teaching file is close to a census of its recorded-pitch
frame. When we draw 100 rows for an inference exercise, the randomisation
is created by our row-sampling procedure. It does not make the television
programme's selection of pitches random.

::: {.callout-caution collapse="true"}
## Try it: classify four recruitment stories

In class this is the pair task. For each story, name the sampling method
from the table and state its main threat.

*Story 1 (given, as a model).* "A ministry lists all registered firms,
groups them by province, and randomly selects firms within each province."
Method: stratified random sample. Main implication: representation of
every province is guaranteed by design.

*Story 2 (you).* "A startup posts a poll on its own Instagram account
asking followers to rate a prototype; 412 followers respond."

*Story 3 (you).* "A student emails the survey to friends, family, and two
student associations she belongs to."

*Story 4 (you).* "A researcher numbers all 8,214 firms in the Chamber of
Commerce register and selects 300 using a random number generator."

::: {.callout-note collapse="true"}
## Check your answer

Story 2: voluntary-response sample --- followers already like the brand,
and those with the strongest opinions are most likely to reply. Story 3:
convenience sample --- easy-to-reach contacts are systematically unlike
the wider population, so selection bias is the main threat. Story 4:
simple random sample from the register --- it supports generalisation to
the registered-firm frame, though firms missing from the register remain
out of reach (coverage error).
:::
:::

Random selection reduces systematic over-representation within the
sampling frame. It does not guarantee that one realised sample is
perfectly balanced: by chance, a sample can contain unusually many large
equity offers or unusually many deals. Larger samples reduce this random
instability --- the five-dot plot would tighten if each dot summarised
1,000 draws instead of 100.

::: {.caveat-box}
<span class="label">What random selection can and cannot repair</span>

Random **sampling** supports generalisation from the sample to the frame
it was drawn from: our 100 rows speak, with quantifiable uncertainty, for
the 1,441 recorded pitches. Random **assignment** --- allocating units to
treatments by chance --- supports causal comparison. They answer different
design questions and neither substitutes for the other. And neither
repairs the show's selection: the frame itself was assembled by producers
choosing telegenic pitches, so no amount of careful random row-sampling
turns conclusions about this file into conclusions about entrepreneurs in
general. Finally, "unbiased" is a promise about the procedure across many
samples, not about your one realised estimate: $\bar x = 12.76$ is not
$\mu = 13.21$, and no single sample can tell you how far off it is.
:::

### Answering the driving question

One hundred randomly chosen records give estimates of the file's
parameters: the seed-123 sample estimates the mean equity offer at
$\bar x=12.76$ against the true $\mu=13.21$, and the deal proportion at
$\hat p=0.64$ against $p\approx 0.612$. Five repeated draws put the sample
mean between about 12.0 and 13.8 while the parameter never moved ---
random samples inform us about the frame, with a random error whose size
Lecture 8 will quantify. What they can never tell us: anything reliable
about entrepreneurs in general, because the frame is a television
programme's curated selection, and our row-sampling randomness cannot
undo that.

**Next:** We now speak carefully about samples: the parameter sits still while
the estimator varies from sample to sample --- you watched five sample means
land in five different places. Lecture 8 studies the pattern behind those
dots: the sampling distribution of the mean, the standard error that measures
its spread, and the central limit theorem, which explains why averages turn
bell-shaped even when the underlying data are wildly skewed. That pattern is
what will let us attach a margin of error to any estimate.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 50 minutes.</span>

### Retrieval (10 minutes)

1. What is the difference between a target population and a sampling frame?
2. Why are draws without replacement dependent?
3. What is the difference between an estimator and an estimate?
4. Why is a sample proportion a sample mean?
5. Why does the sample variance have $n-1$ degrees of freedom?
6. What does random sampling support that random assignment does not?

::: {.callout-note collapse="true"}
## Check your answers

1. The target population is the group of substantive interest; the
   sampling frame is the set from which units can actually be selected.
2. Selecting one unit changes which units remain available.
3. An estimator is a random rule before data are observed; an estimate is
   its realised numerical value.
4. A 0/1 indicator sums to the number of successes, so its mean is the
   success proportion.
5. Estimating the sample mean imposes one constraint: all deviations from
   it must sum to zero. Only $n-1$ deviations can vary freely.
6. Random sampling supports generalisation to the sampling frame; random
   assignment supports causal comparison of treatments.
:::

### Practice (15 minutes)

1. For a population of $N=20{,}000$ customers and a simple random sample
   of $n=500$, check the 10% condition.
2. Write the estimator and realised estimate for the population deal
   proportion.
3. Explain why $\mu$ is fixed while $\bar X$ is random.
4. Using `pitch_sample`, calculate the sample mean and sample SD of
   `description_words`.
5. Explain why this random sample cannot justify conclusions about all US
   entrepreneurs.
6. For a sample of five observations, state the degrees of freedom used to
   estimate variance and explain the number.

::: {.callout-note collapse="true"}
## Check your answers

1. $500/20{,}000=0.025<0.10$, so the condition is satisfied.
2. The estimator is $\widehat P=n^{-1}\sum_iX_i$; the realised estimate is
   $\hat p=n^{-1}\sum_ix_i$ --- here $0.64$.
3. The population and its mean are treated as fixed; the selected
   observations, and therefore their mean, vary across possible samples.
4. Rebuild the sample and compute --- about 4.82 words with an SD of about
   1.76:

```{r}
#| label: b07-practice-check
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
c(mean = mean(pitch_sample$description_words),
  sd   = sd(pitch_sample$description_words))
```

5. The sample was selected from televised records, not from all US
   entrepreneurs; random row selection cannot repair that coverage gap.
6. The degrees of freedom are $5-1=4$. Once the sample mean is estimated,
   four deviations may vary freely and the fifth is fixed by their zero
   sum.
:::

### Draw your own sample (10 minutes)

Add a dot of your own to the class version of @fig-b07-five-means. Use
your student number as the seed, draw your own sample of 100, and compute
your mean equity offer:

```{r}
#| label: b07-own-sample
#| eval: false
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
my_seed <- 2412345   # replace with your own student number
set.seed(my_seed)
rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
mean(sharks$equity_offered_pct[rows])
```

Submit your mean to the shared class plot on Canvas (the Mentimeter link
is posted under Lecture 7). Every student uses a different seed, so every
student contributes a different dot.

::: {.callout-note collapse="true"}
## Check: what will the class plot show?

Two hundred dots will not scatter arbitrarily: they will pile up around
13.2, thin out towards the edges, and form a recognisable shape. That
shape --- the sampling distribution of the mean --- is exactly what
Lecture 8 names, describes, and puts to work.
:::

### Spaced review: one item from Lecture 6 (5 minutes)

Processing time is modelled as $Y\sim N(50,100)$, and
$P(Z\le 1.5)=0.9332$ is supplied. Find $P(Y>65)$ by hand.

::: {.callout-note collapse="true"}
## Check your answer

$\sigma=\sqrt{100}=10$ and $z=(65-50)/10=1.5$, so
$P(Y>65)=1-P(Z\le 1.5)=1-0.9332=0.0668$: under this model, about 6.7% of
processing times exceed 65 minutes.
:::

### Worked exam interpretation (5 minutes)

**Question.** A retailer's loyalty programme has 52,000 members. An
analyst draws 400 members at random; their mean January spend is
&euro;87.40. A slide then claims: "the average member spends &euro;87.40
per month." Identify the parameter, the estimator, and the estimate, and
give two distinct objections to the claim.

::: {.callout-note collapse="true"}
## Check a model answer

The parameter is the mean January spend across all 52,000 members (fixed,
unknown). The estimator is the sample mean $\bar X$ of 400 randomly drawn
members' spends. The estimate is $\bar x=87.40$ euros. Objections: first,
the estimate carries sampling error --- another random 400 members would
give a different mean, so &euro;87.40 should be reported with its
uncertainty (Lectures 8--9); second, the claim generalises a January
figure to "per month," which the data cannot support --- the frame covers
one month's spending only.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Write two
sentences: first name one group you would genuinely want to learn about
(your target population) and the best list of its members you could
actually get your hands on (your frame), then say where the gap between
the two would bite if you sampled from that list. There is no wrong
answer; the point is connecting today's tools to something you care about.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I want to learn about all students on campus who buy lunch at least twice a week, but the best list I can actually get is my own mailing list. Sampling from that list overrepresents existing fans, so any satisfaction estimate would be too rosy for the market I want to enter next."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to hunt for coverage
gaps in your frame and to suggest who your sample would silently leave
out.
::::

Continue to Lecture 8: Sampling distributions and the central limit
theorem.

Formula sheet: Lecture 7 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapter 12.

:::
