# Lecture 8: Sampling distributions and the central limit theorem {#b08}

```{r}
#| label: b08-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 8, you can</span>

- distinguish a population distribution, a sampling distribution, and a
  realised estimate
- calculate the mean, the variance, and the standard deviation of a sample
  mean for given $\mu$, $\sigma$, and $n$, and follow their derivation from
  the rules for rescaling and for sums
- say in plain words what unbiased means for the sample mean, and recognise
  the word consistent
- tell the difference between the standard deviation of the sample mean and
  the standard error, and calculate and interpret both
- use the square-root rule to relate sample size to precision
- state the central limit theorem for this course's conditions and use the
  normal model to calculate a probability for a sample mean
- apply the same results to a sample proportion and check the
  success-failure condition
- explain what consistency and the CLT each say, and which problems neither
  of them solves.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- calculate $E[\bar X]$, $\operatorname{Var}(\bar X)$, and
  $\operatorname{SD}(\bar X)$ for given $\mu$, $\sigma$, and $n$, and name
  the rule used at each step (rescaling, sum of independent variables)
- calculate the standard error of a mean, $s/\sqrt n$
- calculate the standard deviation of a sample proportion for a given $p$,
  and its standard error from $\hat p$
- use the square-root rule: four times the data halves the standard
  deviation of the sample mean, and the standard error
- check the success-failure condition
- calculate a probability for a sample mean with the normal model
  (population normal, or $n$ large) and supplied $Z$ values
- tell apart the standard deviation of the observations, the standard
  deviation of the sample mean, and the standard error.

**Enough to recognise:**

- the derivation of $E[\bar X]=\mu$ and
  $\operatorname{Var}(\bar X)=\sigma^2/n$
- the formal statement of the CLT
- the word consistent: with a larger and larger sample, the estimate comes
  closer and closer to the population value
- the binomial distribution $\text{Bin}(n,p)$ for a number of successes, with
  mean $np$ and variance $np(1-p)$.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Any number computed from a sample would have come out differently with a different sample. Before acting on an estimate, a careful analyst asks how much such numbers vary from sample to sample, and what reduces that variation.</p>

A random sample of 100 pitches from the Shark Tank file had a mean equity
offered of 12.76%. How wrong could that number be, and why can we trust an
average even though the equity offered in single pitches is strongly skewed?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=IrKUM3nNXJE">The Central Limit
Theorem</a>. Focus on which distribution becomes approximately normal as the
sample size grows. The video uses $M_n$ for the sample mean, where this
book uses $\bar X$, and $\hat\Theta$ for an estimator. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Warm up (10 minutes).** Classify each object as a *population
distribution*, a *sampling distribution*, or a *realised estimate*:

1. the distribution of equity offered across individual pitches
2. the distribution of $\bar X$ across all possible samples of 40 pitches
3. the single mean calculated from one observed sample.

::: {.callout-note collapse="true"}
## Check your preparation

The first is a population distribution, the second is a sampling
distribution, and the third is a realised estimate, the number observed in
one sample. Keep the three apart. Lecture 7 named the sampling distribution.
This lecture describes it.
:::

**Predict (5 minutes).** Answer this question before you come to class: to cut the typical error of a sample mean
in half, do you need 2, 4, or 10 times as many
observations? Choose one answer. You find the correct answer in the lecture
material.

If you are not sure about square roots, sums, degrees of freedom, or
standardisation, review [Mathematics Refresher: powers, roots, and
standardisation](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section describes the sampling distribution of the sample mean: where
it is centred, how wide it is, and what shape it has. It then applies the
same results to proportions.

### Start with a founder's question

An invented example. You sell meal boxes. You asked 100 customers, chosen by
chance, to rate the box from 1 to 10, and you calculated the average rating.
How far can that average be from the average rating of all your customers?
And how many customers would you need to ask to halve this distance?

These ratings are not in the Shark Tank file. This lecture answers the two
questions with the Shark Tank file, because for that file the whole
population is known. Lecture 7 showed that averages differ from sample to
sample. This lecture measures how much they differ.

Lecture 7 drew five random samples of 100 pitches from the Shark Tank file
(@fig-b07-five-means). The five sample means of the equity offered
(`equity_offered_pct`) differed, while the population mean stayed fixed at
13.21%. If we drew a sixth sample, where would its mean lie? You cannot say
exactly. But you can say something about the range in which it would
probably fall. This lecture makes that statement precise.

### The distribution behind the five dots

[Shark Tank data]{.shark-tag} The sample mean $\bar X$ is a sample statistic: a number calculated from the
sample. We use it as an estimator of the parameter $\mu$ (Lecture 7). Its
value depends on which random sample is drawn. So $\bar X$ is random, and it
has a distribution. The estimate $\bar x$ is different. It is the one number
that an observed sample gives, such as 12.76.

Each of the five dots is one value of $\bar X$. Imagine not five samples but
every possible random sample of the same size, each with its own sample
mean. The distribution of those possible values of $\bar X$ is the sampling
distribution of the mean, which Lecture 7 named. It describes the
uncertainty of an estimator, not the variation among single pitches. Keep
three objects apart:

- The **population distribution** describes how the equity offered varies
  across the 1,441 pitches.
- The **sampling distribution** describes how $\bar X$ would vary across
  repeated samples.
- A **realised estimate**, such as 12.76, is one draw from that sampling
  distribution.

This lecture uses three population parameters of the Shark Tank file: the
mean and the standard deviation (SD) of the equity offered
(`equity_offered_pct`), and the proportion of pitches that ended with a deal
(`deal_on_show`). The 1,441 records are the whole population, so the SD
divides by $N$ and not by $n-1$:

```{r}
#| label: b08-population-parameters
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
p_deal <- mean(sharks$deal_on_show)
c(mu_equity = mu_equity, sigma_equity = sigma_equity, p_deal = p_deal)
```

So $\mu = 13.21$%, $\sigma = 8.44$ percentage points, and $p = 0.612$. These
are fixed numbers. What varies is any number computed from a random sample
of the 1,441 records.

This lecture uses the random sample of 100 pitches from Lecture 7 (seed 123)
again:

```{r}
#| label: b08-sample
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
c(x_bar = mean(pitch_sample$equity_offered_pct),
  s = sd(pitch_sample$equity_offered_pct),
  p_hat = mean(pitch_sample$deal_on_show))
```

This sample gives the estimates $\bar x=12.76$%, $s=7.71$ percentage points,
and $\hat p=0.64$. Another random sample would give other estimates.

We can make the sampling distribution of $\bar X$ visible. The code below
draws 5,000 random samples of 100 pitches and keeps the mean of each sample.
It draws with replacement, so that the draws are exactly independent.
Lecture 7 drew without replacement. With 100 out of 1,441 pitches the
difference is small (the 10% condition of Lecture 7).

```{r}
#| label: b08-many-means
set.seed(708)
# replace = TRUE: each sample of 100 is drawn with replacement
many_means <- replicate(5000, mean(sample(sharks$equity_offered_pct, 100,
                                          replace = TRUE)))
c(average = mean(many_means), SD = sd(many_means))
quantile(many_means, c(0.025, 0.975))   # the middle 95% of the sample means
```

@fig-b08-many-means shows the 5,000 sample means as a histogram.

```{r}
#| label: fig-b08-many-means
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Histogram of 5,000 sample means of the equity offered, each from a random sample of 100 pitches (green). The histogram is centred on the population mean (gold dashed line at 13.21). The five sample means of Lecture 7 (orange dots) are five draws from this distribution."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)

# 5,000 random samples of 100 pitches, drawn with replacement,
# one mean per sample
set.seed(708)
many_means <- replicate(5000, mean(sample(sharks$equity_offered_pct, 100,
                                          replace = TRUE)))

# the five sample means of Lecture 7 (seeds 101 to 105),
# drawn without replacement as in Lecture 7
seeds <- 101:105
five_means <- numeric(length(seeds))
for (i in seq_along(seeds)) {
  set.seed(seeds[i])
  rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
  five_means[i] <- mean(sharks$equity_offered_pct[rows])
}
five_dots <- data.frame(sample_mean = five_means, density = -0.035)

ggplot(data.frame(sample_mean = many_means), aes(sample_mean)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = seq(10, 17, by = 0.2),
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_vline(xintercept = mu_equity, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_point(data = five_dots, aes(sample_mean, density),
             colour = "#D55E00", size = 3) +
  annotate("text", x = mu_equity + 0.12, y = 0.52,
           label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 0) +
  annotate("text", x = 14.05, y = -0.035,
           label = "the five sample means of Lecture 7",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  annotate("text", x = 14.6, y = 0.33,
           label = "5,000 sample means,\neach from 100 pitches",
           colour = "#009E73", size = 3.4, hjust = 0, lineheight = 0.95) +
  scale_x_continuous(breaks = 10:17) +
  scale_y_continuous(breaks = seq(0, 0.5, by = 0.1),
                     expand = expansion(mult = c(0.02, 0.05))) +
  coord_cartesian(xlim = c(10, 17), ylim = c(-0.05, 0.53)) +
  labs(x = "Sample mean equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Each bar counts sample means, not pitches. The five orange dots below the
histogram are the five sample means of Lecture 7. They are five draws from
this distribution, and the mean of a sixth sample would be one more draw.
About 95% of the 5,000 sample means lie between 11.6% and 14.9%. A sixth
sample mean would probably fall in this range. It would rarely fall far
outside it.

### Back to the lab

[Shark Tank data]{.shark-tag} In Lecture 7 you worked with the
[sampling-distribution lab](b07-samples-estimators.qmd#sampling-lab): the
histogram of sample means was centred near 13.2, became narrower as $n$ grew,
and became bell-shaped, although the distribution of the population is strongly
right-skewed. That histogram, like @fig-b08-many-means, shows the sampling
distribution of the mean. The next sections explain its centre, its spread,
and its shape.

### The mean and variance of a sample mean

[Shark Tank data]{.shark-tag} Let $X_1,\ldots,X_n$ be a random sample of independent and identically
distributed (iid) variables, as in Lecture 7. The sample mean is the
estimator

$$
\bar X=\frac{X_1+\cdots+X_n}{n}=\frac{1}{n}\sum_{i=1}^{n}X_i.
$$

Write $\mu$ for the population mean and $\sigma^2$ for the population
variance. The $X_i$ are identically distributed (Lecture 7), so all of them
have the same mean and the same variance: $E[X_i]=\mu$ and
$\operatorname{Var}(X_i)=\sigma^2$ for every $i$. Then

$$
E[\bar X]=\mu,
\qquad
\operatorname{Var}(\bar X)=\frac{\sigma^2}{n},
\qquad
\operatorname{SD}(\bar X)=\frac{\sigma}{\sqrt n}.
$$

The first equality says that $\bar X$ is an unbiased estimator of $\mu$
([Lecture 7](b07-samples-estimators.qmd#b07-unbiased)). In words: if we draw
many random samples, the average of their sample means equals the population
mean. @fig-b08-many-means shows this result. The 5,000 sample means have an average
of 13.21%, which is the population mean $\mu=13.21$%. A single sample mean
still misses $\mu$, sometimes from above and sometimes from below.

The third quantity, $\operatorname{SD}(\bar X)=\sigma/\sqrt n$, is the
standard deviation of the sample mean. It measures how far a sample mean
typically lies from $\mu$. For samples of 100 pitches it is
$8.44/\sqrt{100}=0.844$ percentage points, about 6% of the population mean.
The 5,000 simulated sample means have a standard deviation of 0.84, as the
formula says.

The three formulas are not new facts. They follow from the rules for
rescaling a variable
in [Lecture 3](b03-discrete-distributions.qmd#b03-shift-rescale) and from the
rules for sums of variables in
[Lecture 5](b05-joint-covariance-correlation.qmd#b05-sums):

- $E[X+Y]=E[X]+E[Y]$ and $E[aX]=aE[X]$, so
  $E[\bar X]=\frac{1}{n}\sum_i E[X_i]=\frac{1}{n}\,n\mu=\mu$.
- $\operatorname{Var}(aX)=a^2\operatorname{Var}(X)$ and, for
  independent variables,
  $\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)$, so

$$
\operatorname{Var}(\bar X)
=\operatorname{Var}\!\left(\frac{X_1+\cdots+X_n}{n}\right)
=\frac{1}{n^2}\sum_{i=1}^{n}\operatorname{Var}(X_i)
=\frac{n\sigma^2}{n^2}
=\frac{\sigma^2}{n}.
$$

Both lines use the fact that every $X_i$ has the same mean $\mu$ and the same
variance $\sigma^2$. Independent positive and negative deviations partly
cancel in an average. That is why the right-hand histogram in the lab is so
much narrower than the population histogram.

### Standard deviation or standard error? {#b08-standard-error}

[Shark Tank data]{.shark-tag} The formula $\operatorname{SD}(\bar X)=\sigma/\sqrt n$ needs the population
standard deviation $\sigma$. In practice $\sigma$ is usually unknown,
because you have only the sample. Then you replace $\sigma$ by the sample
standard deviation $S$ (Lecture 7). The result has its own name.

::: {.definition-box}
<span class="label">Definition</span>

**Standard error (SE).** The estimated standard deviation of an estimator,
calculated from the sample. For the sample mean, $\sigma$ is replaced by
$S$: $\operatorname{SE}(\bar X)=S/\sqrt n$. In one observed sample its value
is $s/\sqrt n$.
:::

Keep three quantities apart:

- The standard deviation of the observations, $\sigma$ in the population and
  $s$ in the sample, describes the variation among single observations.
- The standard deviation of the sample mean, $\sigma/\sqrt n$, describes the
  variation of $\bar X$ across repeated random samples.
- The standard error, $s/\sqrt n$, is the estimate of $\sigma/\sqrt n$ that
  one sample gives.

For the seed-123 sample of 100 pitches:

```{r}
#| label: b08-se-seed123
c(SD_of_sample_mean = sigma_equity / sqrt(100),
  standard_error = sd(pitch_sample$equity_offered_pct) / sqrt(100))
```

The standard error is $7.71/\sqrt{100}=0.771$ percentage points. Here the
population is known, so the standard error can be compared with the standard
deviation of the sample mean, 0.844. The two are close. Replacing a
parameter by an estimate adds a little extra uncertainty. Lecture 9 shows
how to take this extra uncertainty into account.

::: {#exm-b08-se}
## Worked example: E, Var, and SD of the sample mean for σ = 12, n = 36

A population has mean $\mu$ and SD $\sigma=12$. A random sample of $n=36$
observations is drawn. Find the mean, the variance, and the standard
deviation of $\bar X$, naming the rule used at each step.

- **Mean.** By the rules for expectations ($E[X+Y]=E[X]+E[Y]$,
  $E[aX]=aE[X]$), $E[\bar X]=\mu$: whatever $\mu$ is, the sample mean is
  unbiased for it.
- **Variance.** By $\operatorname{Var}(aX)=a^2\operatorname{Var}(X)$ and the
  sum rule for independent variables,
  $\operatorname{Var}(\bar X)=\sigma^2/n=12^2/36=144/36=4$.
- **Standard deviation.** Take the square root:
  $\operatorname{SD}(\bar X)=\sqrt 4=2$, which is $\sigma/\sqrt n=12/6=2$.

A single observation typically lies about $\sigma=12$ units from $\mu$. A
sample mean of 36 observations typically lies about 2 units from $\mu$.
Averaging 36 observations divides the typical distance by $\sqrt{36}=6$, and
$12/6=2$. All three rules are on the formula sheet.

Here $\sigma$ is given, so 2 is the standard deviation of the sample mean.
If $\sigma$ were unknown and the sample gave $s=11.4$, the standard error
would be $11.4/6=1.9$.
:::

### The square-root rule

Precision grows with $\sqrt n$, not with $n$. This answers the prediction
question from before class: to cut the standard deviation of the sample mean
in half you need four times the observations, because

$$
\frac{\sigma}{\sqrt{4n}}=\frac{1}{2}\cdot\frac{\sigma}{\sqrt n}.
$$

Doubling the sample size does not halve the standard deviation of the sample
mean. It only divides it by $\sqrt 2\approx 1.41$. The same rule holds for
the standard error if $s$ stays about the same, because it has the same
$\sqrt n$ in its denominator. So each extra step in precision costs more and
more data. Keep this in mind when you plan the budget for data collection.

The rule also answers the founder's second question. To halve the typical
distance between the average of 100 ratings and the average rating of all
customers, 200 customers are not enough. It takes 400.
[Lecture 9](b09-t-confidence-intervals.qmd#b09-sample-size) turns the rule
around and gives the sample size for a target precision.

::: {.callout-caution collapse="true"}
## Try it: cut the standard deviation of a sample mean from 4 to 1

Try this with your neighbour first. Complete the missing steps
before opening the answer.

*Step 1 (given).* The standard deviation of the sample mean must fall by a
factor of $4/1 = 4$.

*Step 2 (you).* The standard deviation of the sample mean is proportional to
$1/\sqrt n$. By what factor must $n$ change to divide it by 4?

*Step 3 (you).* If the original sample had $n=50$, how many observations
does the new design need?

::: {.callout-note collapse="true"}
## Check your answer

Dividing the standard deviation of the sample mean by 4 requires multiplying
$n$ by $4^2=16$. This is the square-root rule used the other way round. From $n=50$ the design
needs $16\times 50=800$ observations. Four times the precision costs sixteen
times the data.
:::
:::

### A larger sample brings the sample mean closer to μ {#b08-consistent}

[Shark Tank data]{.shark-tag} The variance $\sigma^2/n$ shrinks towards zero as $n$ grows. The sampling
distribution of $\bar X$ becomes narrower and narrower around $\mu$. An
estimator with this property has a name.

::: {.definition-box}
<span class="label">Definition</span>

**Consistent estimator.** An estimator is consistent if, with a larger and
larger sample, the estimate comes closer and closer to the population value.
The sample mean $\bar X$ is a consistent estimator of $\mu$.
:::

The code below shows this for the equity offered (`equity_offered_pct`) in
the Shark Tank file. It draws 1,000 pitches at random, one after another,
and calculates the sample mean after each new pitch.

```{r}
#| label: b08-running-mean
set.seed(196)
draws <- sample(sharks$equity_offered_pct, 1000, replace = TRUE)
running_mean <- cumsum(draws) / seq_along(draws)   # mean of the first n draws
round(running_mean[c(10, 100, 1000)], 2)
```

After 10 pitches the sample mean is 16.15%. After 100 pitches it is 14.22%,
and after 1,000 pitches it is 13.27%. The distance from $\mu=13.21$% falls
from 2.94 points to 1.01 points and then to 0.06 points.
@fig-b08-running-mean shows the whole path.

```{r}
#| label: fig-b08-running-mean
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The sample mean of the equity offered in a growing random sample from the Shark Tank file (green line), for n from 1 to 1,000. For small n the sample mean jumps up and down. As n grows it settles on the population mean (gold dashed line at 13.21)."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)

set.seed(196)
draws <- sample(sharks$equity_offered_pct, 1000, replace = TRUE)
running <- data.frame(n = seq_along(draws),
                      running_mean = cumsum(draws) / seq_along(draws))
marked <- running[running$n %in% c(10, 100, 1000), ]
marked$label <- c("n = 10", "n = 100", "n = 1,000")

ggplot(running, aes(n, running_mean)) +
  geom_hline(yintercept = mu_equity, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  geom_line(colour = "#009E73", linewidth = 0.8) +
  geom_point(data = marked, colour = "#17324D", size = 2.4) +
  annotate("text", x = c(28, 112, 1000), y = c(16.15, 13.72, 13.72),
           label = marked$label, colour = "#17324D", size = 3.4,
           hjust = c(0, 0, 1)) +
  annotate("text", x = 1000, y = mu_equity - 0.45, label = "mu == 13.21",
           parse = TRUE, colour = "#17324D", size = 3.4, hjust = 1) +
  scale_x_continuous(breaks = seq(0, 1000, by = 200),
                     labels = c("0", "200", "400", "600", "800", "1,000")) +
  scale_y_continuous(breaks = seq(10, 18, by = 2)) +
  coord_cartesian(ylim = c(9.5, 18.5)) +
  labs(x = "Number of pitches in the sample (n)",
       y = "Sample mean equity offered (%)") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The green line is the sample mean of the first $n$ pitches. For small $n$ it
jumps up and down, because one pitch with a very high equity offer moves the
mean a lot. As $n$ grows, the jumps become smaller and the line stays closer
and closer to the gold dashed line at 13.21. The line does not come closer at
every step. At $n=350$ it lies further from $\mu$ than at $n=250$.
Consistency describes what happens when $n$ becomes very large.

### The central limit theorem

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-central-limit-theorem/">19.2 The Central Limit Theorem</a>.</p>

[Shark Tank data]{.shark-tag} The formulas for $E[\bar X]$ and $\operatorname{SD}(\bar X)$ give the centre
and the spread of the sampling distribution, not its shape. For some
populations the shape is known exactly. The case that matters in this course
is the normal population. If the population itself is normal, the random
sample is drawn from a normal distribution $N(\mu,\sigma^2)$. Then the
sample mean is exactly normal for every $n$:
$\bar X\sim N(\mu,\sigma^2/n)$. The reason is that a sum of independent
normal variables is again normal
([Lecture 6](b06-continuous-normal.qmd#b06-normal-sums)).

For most populations the exact shape is not known. The equity offered is
an example: the distribution of the population is strongly skewed. Still,
the histogram of sample means is bell-shaped, in the lab and in
@fig-b08-many-means. The central limit theorem describes this pattern.

::: {.definition-box}
<span class="label">Definition</span>

**Central limit theorem (CLT).** For a large random sample (large $n$) of
iid observations with mean $\mu$ and variance $\sigma^2$, the sample mean is
approximately normal, whatever the shape of the population distribution:
$\bar X\ \dot\sim\ N\!\left(\mu,\frac{\sigma^2}{n}\right)$.
:::

The dot above $\sim$ means "approximately". The approximation improves as
$n$ grows.
The mean and the variance of this normal curve are the ones found above:
$E[\bar X]=\mu$ and $\operatorname{Var}(\bar X)=\sigma^2/n$. The CLT adds
the shape.

@fig-b08-clt shows a simulation: 5,000 sample means of $n=5$ draws and 5,000
sample means of $n=40$ draws from the equity offered (`equity_offered_pct`).
As before, the simulation draws with replacement, so its draws are exactly
independent.

```{r}
#| label: fig-b08-clt
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Simulated sampling distributions of the mean equity offered for n = 5 (left) and n = 40 (right), each based on 5,000 samples, with the corresponding normal approximations (orange curves). Both panels use the same axes."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
set.seed(208)
B <- 5000
means_5 <- replicate(B, mean(sample(sharks$equity_offered_pct, 5,
                                    replace = TRUE)))
means_40 <- replicate(B, mean(sample(sharks$equity_offered_pct, 40,
                                     replace = TRUE)))

panel_levels <- c("n = 5", "n = 40")
sim_means <- data.frame(
  sample_mean = c(means_5, means_40),
  panel = factor(rep(panel_levels, each = B), levels = panel_levels)
)

all_means <- c(means_5, means_40)
shared_limits <- c(floor(min(all_means)), ceiling(max(all_means)))
shared_breaks <- seq(shared_limits[1], shared_limits[2], length.out = 36)

normal_grid <- seq(shared_limits[1], shared_limits[2], length.out = 400)
normal_curves <- rbind(
  data.frame(x = normal_grid, panel = panel_levels[1],
             density = dnorm(normal_grid, mu_equity,
                             sigma_equity / sqrt(5))),
  data.frame(x = normal_grid, panel = panel_levels[2],
             density = dnorm(normal_grid, mu_equity,
                             sigma_equity / sqrt(40)))
)
normal_curves$panel <- factor(normal_curves$panel, levels = panel_levels)

ggplot(sim_means, aes(sample_mean)) +
  geom_histogram(aes(y = after_stat(density)), breaks = shared_breaks,
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_line(data = normal_curves, aes(x, density),
            colour = "#D55E00", linewidth = 0.9) +
  facet_wrap(~ panel, scales = "fixed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Sample mean equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Read @fig-b08-clt with the reading sequence of Lecture 1. (1) *Unit and
variables:* each bar counts simulated sample means, not pitches. The
horizontal axis is the sample mean of the equity offered, in percent. (2)
*Scales:* both panels use the same range and the same density scale, so
widths can be compared directly. (3) *One element:* the orange curve in each
panel is the normal approximation for that sample size. (4) *Pattern:* the
histogram for $n=40$ is far narrower and much closer to its normal curve
than the histogram for $n=5$. The histogram for $n=5$ is still skewed to the
right, like the population.
(5) *Limits:* the figure shows random samples from this file. It says nothing
about samples that were not drawn by chance.

How large must $n$ be? Many people use a sample size of at least 30 as a
rough guide (a rule of thumb). But no general rule says that $n=30$ is always
enough.
Thirty is a common starting point for roughly symmetric data. Strongly
skewed data, such as revenues or deal amounts, need more. Extreme outliers
and dependence between observations can also require much more data.

The normal model of Lecture 6 returns here. It is not a model for the data.
For the equity offered in single pitches it did not fit (Lecture 6). It is a
model for the sample mean. You can calculate probabilities for $\bar X$ with
the normal model in two situations:

1. the population is normal, so that $\bar X$ is exactly normal for every
   $n$
2. the sample size $n$ is large, so that $\bar X$ is approximately normal by
   the CLT.

One of the two is enough. In both situations, standardise the sample mean as
in Lecture 6:

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n}.
$$

$Z$ is standard normal in the first situation and approximately standard
normal in the second.

Take the worked example above (@exm-b08-se), with $\sigma=12$ and $n=36$, so
that the standard deviation of the sample mean is 2. Suppose that $\mu=50$.
The sample is large, so by the CLT $\bar X$ is approximately $N(50,2^2)$.
The probability of a sample mean above 53 is

$$
P(\bar X>53)\approx P\!\left(Z>\frac{53-50}{2}\right)=P(Z>1.5)
=1-0.9332=0.0668,
$$

with the supplied value $P(Z\le1.5)=0.9332$. A sample mean above 53 occurs
in about 7% of random samples.

### Sample proportions are sample means

[Shark Tank data]{.shark-tag} Many business questions are about a share and not about an average. An
example from the Shark Tank file: what share of all pitches ends with a
deal? The variable `deal_on_show` is a 0/1 variable. It is 1 if the pitch
ended with a deal and 0 if not. The population proportion $p$ is the share
of ones among the 1,441 pitches, here $p=0.612$.

A 0/1 variable is a Bernoulli variable $X$ with $P(X=1)=p$ (Lecture 3). Its
mean and variance are

$$
E[X]=p,
\qquad
\operatorname{Var}(X)=p(1-p).
$$

The sample proportion $\widehat P$ is the estimator of $p$: the share of
ones in a random sample. It is the sample mean of a 0/1 variable
(Lecture 7), $\widehat P=\bar X$. So every result above applies, with
$\mu=p$ and $\sigma^2=p(1-p)$:

$$
E[\widehat P]=p,
\qquad
\operatorname{Var}(\widehat P)=\frac{p(1-p)}{n},
\qquad
\operatorname{SD}(\widehat P)=\sqrt{\frac{p(1-p)}{n}}.
$$

The formulas show two properties. $\widehat P$ is unbiased, because
$E[\widehat P]=p$. And $\widehat P$ is consistent, because its variance
$p(1-p)/n$ shrinks towards zero as $n$ grows.

The standard deviation of the sample proportion needs the parameter $p$,
which is usually unknown. Replacing $p$ by the estimate $\hat p$ gives the
standard error of the sample proportion:

$$
\operatorname{SE}(\widehat P)=\sqrt{\frac{\hat p(1-\hat p)}{n}}.
$$

Use the standard deviation, with $p$, when the population proportion is
given or assumed. Use the standard error, with $\hat p$, when you only have
the sample, as in Tutorial 3. The seed-123 sample has 64 deals in 100
pitches, so $\hat p=0.64$:

```{r}
#| label: b08-proportion-standard-error
p_hat <- mean(pitch_sample$deal_on_show)
c(p_hat = p_hat, standard_error = sqrt(p_hat * (1 - p_hat) / 100))
```

The standard error is $\sqrt{0.64\times0.36/100}=0.048$, or 4.8 percentage
points.

By the CLT, the sampling distribution of $\widehat P$ is approximately normal
for a large sample. How large is large enough for a proportion? The answer
comes from the distribution of a count.

### Counting successes: the binomial distribution {#b08-binomial}

Take $n$ independent 0/1 trials, each with the same success probability $p$.
The number of successes is the sum $Y=X_1+\cdots+X_n$. Its distribution is
called the **binomial distribution**, written $Y\sim\text{Bin}(n,p)$. The sum
rules of [Lecture 5](b05-joint-covariance-correlation.qmd#b05-sums) give

$$
E[Y]=np,
\qquad
\operatorname{Var}(Y)=np(1-p).
$$

The sample proportion is this count divided by the number of trials:
$\widehat P=Y/n$.

An invented example: 200 people visit your web shop, and each of them buys
with probability 0.05, independently of the others. The number of buyers is
$Y\sim\text{Bin}(200,\,0.05)$. You expect $np=200\times 0.05=10$ buyers, with
a standard deviation of $\sqrt{200\times 0.05\times 0.95}=\sqrt{9.5}\approx 3.1$
buyers.

A count is a sum, so the CLT applies to it: for large $n$ the binomial
distribution is close to a normal distribution. @fig-b08-binomial shows when
$n$ is large enough. With 80 visitors you expect only $np=4$ buyers. The
distribution is skewed, with a longer tail to the right. The normal curve
puts about 2% of its area below zero, where no count can lie. The fit is
worse than in the right panel. With 400 visitors you expect $np=20$ buyers,
and the normal curve fits well.

```{r}
#| label: fig-b08-binomial
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The binomial distribution of the number of buyers when each visitor buys with probability 0.05 (green), with the normal curve of the same mean and variance (orange). With 80 visitors (left) the distribution is skewed and the normal curve reaches below zero. With 400 visitors (right) the normal curve fits well. The two panels have different axes."
library(ggplot2)

p_buy <- 0.05
n_small <- 80
n_large <- 400
panel_levels <- c("n = 80, np = 4: rule violated",
                  "n = 400, np = 20: rule satisfied")

pmf <- rbind(
  data.frame(y = 0:12, panel = panel_levels[1],
             prob = dbinom(0:12, size = n_small, prob = p_buy)),
  data.frame(y = 5:37, panel = panel_levels[2],
             prob = dbinom(5:37, size = n_large, prob = p_buy))
)
pmf$panel <- factor(pmf$panel, levels = panel_levels)

grid_small <- seq(-7, 12, length.out = 300)
grid_large <- seq(5, 37, length.out = 300)
normal_curves <- rbind(
  data.frame(y = grid_small, panel = panel_levels[1],
             density = dnorm(grid_small, n_small * p_buy,
                             sqrt(n_small * p_buy * (1 - p_buy)))),
  data.frame(y = grid_large, panel = panel_levels[2],
             density = dnorm(grid_large, n_large * p_buy,
                             sqrt(n_large * p_buy * (1 - p_buy))))
)
normal_curves$panel <- factor(normal_curves$panel, levels = panel_levels)
below_zero <- normal_curves[normal_curves$panel == panel_levels[1] &
                              normal_curves$y <= 0, ]

notes <- data.frame(
  y = c(-7, 26, 26),
  prob = c(0.055, 0.084, 0.068),
  label = c("normal curve\nreaches\nbelow 0", "normal curve",
            "binomial\nprobabilities"),
  colour = c("#D55E00", "#D55E00", "#009E73"),
  panel = factor(panel_levels[c(1, 2, 2)], levels = panel_levels)
)

ggplot(pmf, aes(y, prob)) +
  geom_area(data = below_zero, aes(y, density),
            fill = "#D55E00", alpha = 0.25) +
  geom_segment(aes(xend = y, yend = 0), colour = "#009E73", linewidth = 0.8) +
  geom_point(colour = "#009E73", size = 1.8) +
  geom_line(data = normal_curves, aes(y, density),
            colour = "#D55E00", linewidth = 0.9) +
  geom_text(data = notes, aes(label = label), colour = notes$colour,
            size = 3.2, hjust = 0, lineheight = 0.95) +
  facet_wrap(~ panel, scales = "free") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Number of buyers", y = "Probability") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold", size = 10)
  )
```

So use the normal approximation for a count or a proportion only if
$np\ge10$ and $n(1-p)\ge10$. This rule is called the
**success-failure condition**: the expected numbers of successes and of
failures must both be at least 10. When $p$ is unknown, check the rule with
the counts in the sample: at least 10 successes and at least 10 failures. In
a test of a proposed value $p_0$, use $p_0$ (Lecture 10). Some textbooks use 5
instead of 10. You do not need to calculate binomial probabilities in this
course.

::: {#exm-b08-proportion}
## Worked example: standard deviation of the deal proportion at n = 100

[Shark Tank data]{.shark-tag} Suppose we draw a random sample of $n=100$ pitches from the Shark Tank file, where the
deal proportion is $p=0.612$. How far is the sample deal proportion
typically from 0.612?

- **Standard deviation of the sample proportion.** Here $p$ is given, so
  $\operatorname{SD}(\widehat P)=\sqrt{0.612\,(1-0.612)/100}
  =\sqrt{0.2375/100}\approx 0.0487$, about 4.9 percentage points.
- **Success-failure check.** $np=100(0.612)=61.2\ge 10$ and
  $n(1-p)=100(0.388)=38.8\ge 10$, so the normal approximation is reasonable.

In R, with `p_deal` as calculated above:

```{r}
#| label: b08-proportion-se
p_deal <- mean(sharks$deal_on_show)
c(SD_deal_proportion_n_100 = sqrt(p_deal * (1 - p_deal) / 100),
  successes = 100 * p_deal, failures = 100 * (1 - p_deal))
```
:::

### Consistency and the CLT answer different questions

Consistency and the CLT are both statements about large random samples. They
answer different questions:

- Consistency says where the sample mean goes as $n$ grows: towards $\mu$.
- The CLT says what the distribution of the sample means looks like for a
  large $n$: its approximate shape and its width.

Consistency is about a value. The CLT is about a distribution. Both assume
that the sampling is random. Neither of them can correct a sample that was
selected badly from the start.

::: {.callout-note collapse="true"}
## Check: match each statement to consistency, the CLT, or neither

1. "With enough random draws, the sample mean equity will be very close to
   13.21."
2. "For $n=100$ random draws, the sample mean equity is approximately
   $N(13.21, 0.844^2)$."
3. "With enough respondents, our LinkedIn poll will represent Dutch
   founders."

Statement 1 is consistency, statement 2 is the CLT, and statement 3 is
neither. The people who answer a LinkedIn poll chose themselves (a
self-selected sample). No theorem turns a self-selected sample into a
representative one (see the box below).
:::

::: {.caveat-box}
<span class="label">What a bigger sample can and cannot fix</span>

- More data reduce the random error, at the square-root rate.
- More data do nothing for selection bias. A million televised pitches still
  describe televised pitches.
- The standard error measures precision inside the list you sampled from
  (the sampling frame). It says nothing about who is on the list.

**What you can do with it:** first make sure you ask the right people, then
decide how many people to ask.
Stop collecting when more data would no longer change your decision.
:::

### Answering the driving question

A random sample of 100 pitches from the Shark Tank file had a mean equity
offered of $\bar x=12.76$%. How wrong could that number be?

Imagine you are an analyst who has only this sample. The population
mean $\mu$ is unknown. That is why a sample was drawn. The sample gives its
own standard deviation, $s=7.71$ percentage points (Lecture 7), and from it
the standard error:

$$
\operatorname{SE}(\bar X)=\frac{s}{\sqrt n}=\frac{7.71}{\sqrt{100}}\approx 0.77.
$$

```{r}
#| label: b08-driving-question
x_bar <- mean(pitch_sample$equity_offered_pct)
s <- sd(pitch_sample$equity_offered_pct)
c(x_bar = x_bar, s = s, standard_error = s / sqrt(100))
```

So the mean of a random sample of this size typically lies within about 0.8
percentage points of $\mu$. By the CLT it rarely lies more than about two
standard deviations of the sample mean away from $\mu$. So, without knowing
$\mu$, the analyst can say this: 12.76 is probably within about one
percentage point of the population mean. It is very unlikely to differ from
it by more than two percentage points. A typical error of 0.8 percentage
points is about 6% of the sample mean.

Why can we trust an average this much, although the equity offered in
single pitches is strongly skewed? Two results explain it. First, positive
and negative deviations partly cancel in an average. The sampling distribution of $\bar X$ is therefore far
narrower than the population distribution: a standard error of about 0.8
points for the sample mean, against a standard deviation of about 8 points
for single pitches. Second, the central limit theorem gives this narrow
distribution a known shape. It is approximately normal, although the
population is not. Together, the two results make the size of the error of
an average of 100 skewed values predictable.

This file is the whole population, so the answer can be checked:

```{r}
#| label: b08-driving-check
c(miss = mu_equity - x_bar,
  miss_in_standard_errors = (mu_equity - x_bar) / (s / sqrt(100)),
  SD_of_sample_mean = sigma_equity / sqrt(100))
```

The population mean is $\mu=13.21$%, so this sample missed by 0.45 points.
That is a little over half a standard error, which is a normal size for such
a difference. The standard
deviation of the sample mean, calculated with $\sigma=8.44$, is 0.84. The
standard error of 0.77, calculated from the sample alone, is close to it.

One detail: our sample is drawn without replacement from 1,441 records. That
makes the exact standard deviation of the sample mean slightly smaller than
$\sigma/\sqrt n$:

```{r}
#| label: b08-without-replacement
sigma_equity / sqrt(100) * sqrt((1441 - 100) / (1441 - 1))
```

It is 0.81 and not 0.84. The 10% condition says that we can ignore the
difference. The optional box in
[Lecture 7](b07-samples-estimators.qmd#b07-srs) gives the exact
formula.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You ask customers to rate your meal boxes from 1 to 10. From earlier surveys you expect the ratings to have a standard deviation of about 2. With 25 customers the standard deviation of the sample mean is then 2/5 = 0.4. With 100 customers it is 0.2. Is the larger survey worth four times the effort? Your rule is to change the menu only if the mean rating is below 7. The first 25 customers give a mean of 8.4 and a sample standard deviation of 2. The standard error is 2/5 = 0.4, so 8.4 is 3.5 standard errors above 7. The decision: stop surveying and keep the menu. More data would not change what you do.
:::

**Next:** We can now say how much a sample mean typically varies: by about
one standard deviation of the sample mean, which the standard error
estimates. A report needs a sentence such as "the population mean plausibly
lies between $a$ and $b$". This needs one more step. In a real sample
$\sigma$ is unknown, so we must estimate it too, and this adds uncertainty.
Lecture 9 introduces the t distribution, which takes this extra uncertainty
into account, and builds the confidence interval from it. Lecture 9
also shows how many customers you must ask to reach a target precision.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 80 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. What varies in a sampling distribution?
2. What do "unbiased" and "consistent" say about the sample mean?
3. What is the difference between the standard deviation of the
   observations, the standard deviation of the sample mean, and the standard
   error?
4. To cut the standard deviation of a sample mean in half, by what factor
   must the sample size grow, and why?
5. What does the CLT say becomes approximately normal?
6. Why does a larger sample not remove selection bias?

::: {.callout-note collapse="true"}
## Check your answers

1. The statistic varies across possible repeated samples. The parameter and
   the population stay fixed.
2. Unbiased: $E[\bar X]=\mu$. Over many random samples, the average of the
   sample means equals the population mean. Consistent: with a larger and
   larger sample, the sample mean comes closer and closer to the population
   mean.
3. The standard deviation of the observations ($\sigma$, estimated by $s$)
   describes variation among single observations. The standard deviation of
   the sample mean, $\sigma/\sqrt n$, describes variation of $\bar X$ across
   samples. The standard error, $s/\sqrt n$, is its estimate from one
   sample.
4. By a factor of 4: the standard deviation of the sample mean is
   proportional to $1/\sqrt n$, so halving it requires $2^2=4$ times the
   observations.
5. The sample mean. For a large sample its sampling distribution is
   approximately normal. The single observations do not become normal.
6. Sample size shrinks random error only. It does not change how units
   entered the sampling frame or the sample.
:::

### Finish a half-worked example (5 minutes)

Re-read the worked calculation for a sample mean (@exm-b08-se). Then
complete the same calculation for a proportion. For a planned survey you
expect $p=0.30$, and you plan to ask $n=100$ people.

*Step 1 (given).* The standard deviation of the sample proportion is
$\sqrt{p(1-p)/n}$.

*Step 2 (you).* Calculate it.

*Step 3 (you).* Check the success-failure condition.

*Step 4 (you).* The survey is done: 36 of the 100 respondents say yes.
Calculate the standard error.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SD}(\widehat P)=\sqrt{0.30(0.70)/100}\approx 0.0458$, about
4.6 percentage points. The expected counts are $np=30$ successes and
$n(1-p)=70$ failures, both at least 10, so the normal approximation is
acceptable. After the survey, $\hat p=0.36$ and the standard error is
$\operatorname{SE}(\widehat P)=\sqrt{0.36(0.64)/100}=0.048$, or 4.8
percentage points. The standard error uses $\hat p$, because $p$ itself is
unknown.
:::

### Practice (15 minutes)

1. If $\sigma=15$ and $n=25$, find $E[\bar X]$,
   $\operatorname{Var}(\bar X)$, and $\operatorname{SD}(\bar X)$, naming
   the rule used at each step.
2. By what factor must $n$ grow to reduce the standard deviation of a
   sample mean from 6 to 1.5?
3. For a sample proportion with $p$ assumed to be 0.50 and $n=64$,
   calculate the standard deviation of the sample proportion and check the
   success-failure condition.
4. Explain why the normal approximation is more convincing in the
   right-hand panel of @fig-b08-clt.
5. State one threat to the CLT-based approximation that a large $n$ alone
   does not solve.
6. An invented example. The preparation times of a founder's orders have
   mean $\mu=30$ minutes and SD $\sigma=5$ minutes. She times a random sample
   of 100 orders. Find the probability that the sample mean is above 31
   minutes. Supplied value: $P(Z\le 2)=0.9772$. Her sample then gives
   $s=4.6$ minutes. What is the standard error of the mean?

::: {.callout-note collapse="true"}
## Check your answers

1. $E[\bar X]=\mu$ (rules for expectations),
   $\operatorname{Var}(\bar X)=\sigma^2/n=225/25=9$ (variance of a rescaled
   sum of independent variables), $\operatorname{SD}(\bar X)=\sqrt 9=3$.
2. The standard deviation of the sample mean must fall by a factor of 4, so
   the sample size must grow by $4^2=16$ times.
3. $\operatorname{SD}(\widehat P)=\sqrt{0.50(0.50)/64}=0.0625$. Expected
   counts are 32 and 32, both at least 10.
4. Averaging 40 observations cancels more of the skewness and shrinks the
   sampling variation more than averaging 5 observations. The histogram
   follows its normal curve much more closely.
5. Dependence between observations. A non-representative sample is a
   different problem: the approximation still works, but the mean then
   estimates the wrong quantity. Extreme outliers are a problem too: with
   them, the approximation needs a much larger sample to work.
6. The sample is large, so by the CLT $\bar X$ is approximately normal with
   mean 30 and $\operatorname{SD}(\bar X)=5/\sqrt{100}=0.5$. Standardise:
   $z=(31-30)/0.5=2$, so $P(\bar X>31)\approx P(Z>2)=1-0.9772=0.0228$, about
   2%. The standard error is $s/\sqrt n=4.6/10=0.46$ minutes. It estimates
   the standard deviation of the sample mean, 0.5.
:::

### Widget homework (10 minutes)

[Shark Tank data]{.shark-tag} Open the
[sampling-distribution lab](b07-samples-estimators.qmd#sampling-lab) in
Lecture 7 again and run these three settings. For each setting, draw new
samples several times and write **one sentence** on what the histogram of sample means shows:

1. $n = 2$
2. $n = 25$
3. $n = 100$.

::: {.callout-note collapse="true"}
## Check a model answer

1. At $n=2$ the sample means scatter widely and their histogram is still
   right-skewed, like the population.
2. At $n=25$ the histogram of means is clearly narrower and roughly
   symmetric, centred near 13.2.
3. At $n=100$ the means form a narrow bell shape around the population mean, with
   a typical deviation of less than one percentage point. This is the
   standard deviation of the sample mean, about 0.84 percentage points.
:::

### Review: one question each from Lectures 7 and 6 (10 minutes)

1. *(Lecture 7.)* The sample variance divides by $n-1$, not $n$. In one or
   two sentences, why?
2. *(Lecture 6.)* An invented example. Daily revenue is modelled as
   $X\sim N(10, 2^2)$ (in thousands of euros). Find $P(X>13)$. Supplied
   value: $P(Z\le 1.5)=0.9332$.

::: {.callout-note collapse="true"}
## Check your answers

1. The deviations are measured from the sample mean, which was itself
   estimated from the data. Once it is fixed, only $n-1$ deviations can
   vary freely, and dividing by $n-1$ removes the resulting downward bias.
2. Standardise: $z=(13-10)/2=1.5$, so
   $P(X>13)=P(Z>1.5)=1-0.9332\approx 0.067$.
:::

### Worked exam interpretation: a probability for a sample mean (5 minutes)

[Shark Tank data]{.shark-tag} **Exam-style question.** The mean equity offered (`equity_offered_pct`) in the Shark
Tank file is $\mu=13.21$% with $\sigma=8.44$ percentage points. A random
sample of $n=100$ pitches is drawn. Using the CLT, find the probability that
the sample mean exceeds 14.5%. Supplied value: $P(Z\le 1.53)=0.937$.

::: {.callout-note collapse="true"}
## Check a model answer

By the CLT, $\bar X\ \dot\sim\ N(13.21, 0.844^2)$ since
$\operatorname{SD}(\bar X)=8.44/\sqrt{100}=0.844$. Standardise:
$z=(14.5-13.21)/0.844\approx 1.53$, so
$P(\bar X>14.5)\approx P(Z>1.53)=1-0.937=0.063$. That is about a 6% chance.
Note that the question is about the mean, not about a single pitch. Lecture
6 rejected the normal model for a single pitch.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "The ratings in my sample have a standard deviation of 12 and I asked 36 customers. How precise is the mean, and how do I halve the error?" It answered:

> "The standard error is 12/36 = 0.33. To halve it you need twice as many customers, 72."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Both steps are wrong. The standard error divides by the square root of the sample size: 12/6 = 2. And to halve it you need four times the data, 144 customers, because the standard error falls with the square root of the sample size.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Name one number
you would estimate from a sample (customer satisfaction, conversion rate,
average basket size) and write two sentences. First say how you would draw
the sample and roughly what standard error you would accept. Then say
whether collecting four times the data would be worth halving that standard
error for the decision you face.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "I would estimate the conversion rate of visitors to our tasting stand from a random sample of 100 visitors. A standard error of about 5 percentage points is fine for deciding whether to open a second stand. Collecting four times the data would halve the standard error to about 2.5 percentage points, which is not worth the extra tasting days for this decision."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to help you work out the
trade-off between sample size and precision for your own numbers.

::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 8. Show the solution only after I send my answer."

[Continue to Lecture 9: The t distribution and confidence intervals](b09-t-confidence-intervals.qmd).

[Formula sheet: Lecture 8 formulas](formula-sheet.qmd#formula-l8).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 13 and 14.

:::
