# Lecture 8: Sampling distributions and the central limit theorem {#b08}

```{r}
#| label: b08-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 8, you can</span>

- distinguish a population distribution, a sampling distribution, and a
  realised estimate
- derive the mean and variance of a sample mean using the rules for
  rescaling and for sums
- calculate and interpret a standard error
- use the square-root rule to relate sample size to precision, and find the
  sample size for a target margin of error
- state the central limit theorem for this course's conditions and apply it
  to means and proportions
- explain what the law of large numbers and the CLT each claim, and what
  neither of them repairs.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- calculate the standard error of a mean and of a proportion
- use the square-root rule: four times the data halves the standard error
- find the sample size needed for a target margin of error
- check the success-failure condition
- calculate a probability for a sample mean with the CLT and supplied $Z$ values
- tell a standard deviation from a standard error.

**Enough to recognise:**

- the derivation of $\operatorname{Var}(\bar X)=\sigma^2/n$
- the formal statements of the law of large numbers and the CLT
- the binomial distribution $\text{Bin}(n,p)$ for a number of successes, with
  mean $np$ and variance $np(1-p)$.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Any number computed from a sample would have come out differently with a different sample. Before acting on an estimate, a careful analyst asks how much such numbers vary from sample to sample, and what reduces that variation.</p>

A random sample of 100 pitches from the Shark Tank file had a mean equity
offered of 12.76%. How wrong could that number be, and why can we trust an
average even though the equity offered in single pitches is strongly skewed?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (10 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=IrKUM3nNXJE">The Central Limit
Theorem</a>. Focus on which distribution becomes approximately normal as the
sample size grows. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (10 minutes).** Classify each object as a *population
distribution*, a *sampling distribution*, or a *realised estimate*:

1. the distribution of equity offered across individual pitches
2. the distribution of $\bar X$ across all possible samples of 40 pitches
3. the single mean calculated from one observed sample.

::: {.callout-note collapse="true"}
## Check your answers

The first is a population distribution, the second is a sampling
distribution, and the third is a realised estimate, the number observed in
one sample. Keep the three apart. Lecture 7 named the sampling distribution.
This lecture describes it.
:::

**Predict (5 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: to cut the typical error of a sample mean
**in half**, do you need $2\times$, $4\times$, or $10\times$ the
observations? Commit to an answer. You find the answer in the lecture
material.

If square roots, sums, degrees of freedom, or standardisation are not secure,
review [Mathematics Refresher: powers, roots, and
standardisation](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

This section describes the sampling distribution of the sample mean: where
it is centred, how wide it is, and what shape it has. It then applies the
same results to proportions.

### Start with a founder's question

You asked 100 customers and calculated an average. How far can that average
be from the truth, and how many customers would it take to halve the
distance? Lecture 7 showed that averages differ from sample to sample. This
lecture measures by how much. Start from the five dots of Lecture 7. Five
random samples of 100 pitches gave five different sample means, while the
population mean stayed fixed. Where would the mean of a sixth sample land?
The answer lies in the pattern behind those dots.

### The pattern behind the five dots

Imagine not five samples but every possible random sample of the same size,
each with its own sample mean. The distribution of those possible values of
$\bar X$ is the sampling distribution of the mean, which Lecture 7 named. It
describes the uncertainty of an estimator, not the variation among single
pitches. Three objects must be kept apart:

- The **population distribution** describes how the equity offered varies
  across the 1,441 pitches.
- The **sampling distribution** describes how $\bar X$ would vary across
  repeated samples.
- A **realised estimate**, such as 12.76, is one draw from that sampling
  distribution.

This lecture uses three parameters of the Shark Tank file: the mean and the
SD of the equity offered (`equity_offered_pct`), and the proportion of
pitches that ended with a deal (`deal_on_show`). The 1,441 records are the
whole population, so the SD divides by $N$ and not by $n-1$:

```{r}
#| label: b08-population-parameters
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
p_deal <- mean(sharks$deal_on_show)
c(mu_equity = mu_equity, sigma_equity = sigma_equity, p_deal = p_deal)
```

So $\mu = 13.21$%, $\sigma = 8.44$ percentage points, and $p = 0.612$. These
are fixed numbers. What varies is any number computed from a random sample
of the 1,441 records.

### Back to the lab

In Lecture 7 you worked with the
[sampling-distribution lab](b07-samples-estimators.qmd#sampling-lab): the
histogram of sample means sat near 13.2, narrowed as $n$ grew, and turned
bell-shaped although the population is strongly right-skewed. That pile of
sample means is the sampling distribution of the mean. The rest of this
lecture explains what you saw there: where the pile sits, how wide it is,
and why it becomes bell-shaped.

### The mean and variance of a sample mean

Let $X_1,\ldots,X_n$ be a random sample of independent and identically
distributed (iid) variables, as in Lecture 7, with population mean $\mu$ and
variance $\sigma^2$. Then

$$
E[\bar X]=\mu,
\qquad
\operatorname{Var}(\bar X)=\frac{\sigma^2}{n},
\qquad
\operatorname{SD}(\bar X)=\frac{\sigma}{\sqrt n}.
$$

The first equality says that $\bar X$ is unbiased for $\mu$ (Lecture 7): the
pile of sample means is centred on the parameter, as in the lab. The third
quantity has its own name.

::: {.definition-box}
<span class="label">Definition</span>

**Standard error (SE).** The standard deviation of an estimator across
repeated random samples. For the sample mean,
$\operatorname{SE}(\bar X)=\sigma/\sqrt n$.
:::

An SD describes the variation among observations. An SE describes the
variation of an estimator.

The three formulas above are not new facts. They follow from the rules for
rescaling a variable
in [Lecture 3](b03-discrete-distributions.qmd#b03-shift-rescale) and from the
rules for sums of variables in
[Lecture 5](b05-joint-covariance-correlation.qmd#b05-sums):

- $E[X+Y]=E[X]+E[Y]$ and $E[aX]=aE[X]$, so
  $E[\bar X]=\frac{1}{n}\sum_i E[X_i]=\frac{1}{n}\,n\mu=\mu$.
- $\operatorname{Var}(aX)=a^2\operatorname{Var}(X)$ and, for
  **independent** variables,
  $\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)$, so

$$
\operatorname{Var}(\bar X)
=\operatorname{Var}\!\left(\frac{X_1+\cdots+X_n}{n}\right)
=\frac{1}{n^2}\sum_{i=1}^{n}\operatorname{Var}(X_i)
=\frac{n\sigma^2}{n^2}
=\frac{\sigma^2}{n}.
$$

Independent positive and negative deviations partly cancel in an average.
That is why the right-hand histogram in the lab is so much narrower than the
population histogram.

The variance $\sigma^2/n$ shrinks to zero as $n$ grows, so $\bar X$ gets
close to $\mu$. Estimators with this property are called **consistent**.

The formulas for $E[\bar X]$ and $\operatorname{SD}(\bar X)$ give the centre
and the spread of the sampling distribution, not its shape. In one case the
shape is known exactly. If the population
itself is normal, the sample mean is exactly normal for every $n$:
$\bar X\sim N(\mu,\sigma^2/n)$. The reason is that a sum of independent
normal variables is again normal
([Lecture 6](b06-continuous-normal.qmd#b06-normal-sums)). The central limit
theorem below deals with populations that are not normal.

In practice $\sigma$ is usually unknown and is estimated by the sample SD
$S$. This gives the estimated standard error
$\widehat{\operatorname{SE}}(\bar X)=S/\sqrt n$. The hat marks an estimate:
$\widehat{\operatorname{SE}}$ is a standard error calculated from the
sample, just as $\hat p$ is a proportion calculated from the sample.
Replacing a parameter by an estimate adds a little extra uncertainty.
Lecture 9 shows how to allow for it.

::: {#exm-b08-se}
## Worked example: E, Var, and SE for σ = 12, n = 36

A population has mean $\mu$ and SD $\sigma=12$. A random sample of $n=36$
observations is drawn. Find the mean, variance, and SE of $\bar X$, naming
the rule used at each step.

- **Mean.** By the rules for expectations ($E[X+Y]=E[X]+E[Y]$,
  $E[aX]=aE[X]$), $E[\bar X]=\mu$: whatever $\mu$ is, the sample mean is
  unbiased for it.
- **Variance.** By $\operatorname{Var}(aX)=a^2\operatorname{Var}(X)$ and the
  sum rule for independent variables,
  $\operatorname{Var}(\bar X)=\sigma^2/n=12^2/36=144/36=4$.
- **SE.** Take the square root:
  $\operatorname{SD}(\bar X)=\sqrt 4=2$, which is $\sigma/\sqrt n=12/6=2$.

The typical distance between a sample mean and $\mu$ is about 2 units. That
is one sixth of the typical distance of a single observation from $\mu$.
All three rules are on the formula sheet.
:::

### The square-root rule

Precision grows with $\sqrt n$, not with $n$. This answers the prediction
question from before class: to cut the SE in half you need four times the
observations, because

$$
\frac{\sigma}{\sqrt{4n}}=\frac{1}{2}\cdot\frac{\sigma}{\sqrt n}.
$$

Doubling the sample size does not halve the SE. It only divides the SE by
$\sqrt 2\approx 1.41$. More precision becomes more and more expensive, which
is worth knowing before you budget for data collection.

::: {.callout-caution collapse="true"}
## Try it: how large must n become to cut an SE from 4 to 1?

Try this with your neighbour first. Complete the missing steps
before opening the answer.

*Step 1 (given).* The SE must fall by a factor of $4/1 = 4$.

*Step 2 (you).* The SE is proportional to $1/\sqrt n$. By what factor must
$n$ change to divide the SE by 4?

*Step 3 (you).* If the original sample had $n=50$, how many observations
does the new design need?

::: {.callout-note collapse="true"}
## Check your answer

Dividing the SE by 4 requires multiplying $n$ by $4^2=16$: the square-root
rule in reverse. From $n=50$ the design needs $16\times 50=800$
observations. Four times the precision costs sixteen times the data.
:::
:::

### The central limit theorem

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-central-limit-theorem/">19.2 The Central Limit Theorem</a>.</p>

The lab showed one more pattern that the rules for the mean and the variance
do not explain. The pile of sample means became bell-shaped, although the
population is strongly skewed. The central limit theorem describes this
pattern.

::: {.definition-box}
<span class="label">Definition</span>

**Central limit theorem (CLT).** For a large random sample of iid
observations with mean $\mu$ and variance $\sigma^2$, the sample mean is
approximately normal, whatever the shape of the population:
$\bar X\ \dot\sim\ N\!\left(\mu,\frac{\sigma^2}{n}\right)$.
:::

The dot above $\sim$ marks an approximation, which improves as $n$ grows. The
same statement can be written with the standardised sample mean

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n},
$$

which is approximately standard normal for large $n$. The normal model of
Lecture 6 returns here. It is not a model for the data, where it failed for
the equity offered. It is a model for the sample mean.

With the CLT you can calculate a probability for a sample mean. Take the
worked example above (@exm-b08-se), with $\sigma=12$ and $n=36$, so that
the SE is 2. Suppose that $\mu=50$. By the CLT, $\bar X$ is approximately
$N(50,2^2)$. The probability of a sample mean above 53 is

$$
P(\bar X>53)\approx P\!\left(Z>\frac{53-50}{2}\right)=P(Z>1.5)
=1-0.9332=0.0668,
$$

with the supplied value $P(Z\le1.5)=0.9332$. A sample mean above 53 occurs
in about 7% of random samples.

There is no universal rule that $n=30$ is always enough. Thirty is a common
starting point for roughly symmetric data. Strongly skewed data, such as
revenues or deal amounts, need more. Extreme outliers and dependence between
observations can also require much more data.

@fig-b08-clt shows a simulation: 5,000 sample means of $n=5$ draws and 5,000
sample means of $n=40$ draws from the equity offered (`equity_offered_pct`).
The simulation draws with replacement, so its draws are exactly independent.

```{r}
#| label: fig-b08-clt
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Simulated sampling distributions of the mean equity offered for n = 5 (left) and n = 40 (right), each based on 5,000 samples, with the corresponding normal approximations (orange curves). Both panels use the same axes."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
set.seed(208)
B <- 5000
means_5 <- replicate(B, mean(sample(sharks$equity_offered_pct, 5,
                                    replace = TRUE)))
means_40 <- replicate(B, mean(sample(sharks$equity_offered_pct, 40,
                                     replace = TRUE)))

panel_levels <- c("n = 5", "n = 40")
sim_means <- data.frame(
  sample_mean = c(means_5, means_40),
  panel = factor(rep(panel_levels, each = B), levels = panel_levels)
)

all_means <- c(means_5, means_40)
shared_limits <- c(floor(min(all_means)), ceiling(max(all_means)))
shared_breaks <- seq(shared_limits[1], shared_limits[2], length.out = 36)

normal_grid <- seq(shared_limits[1], shared_limits[2], length.out = 400)
normal_curves <- rbind(
  data.frame(x = normal_grid, panel = panel_levels[1],
             density = dnorm(normal_grid, mu_equity,
                             sigma_equity / sqrt(5))),
  data.frame(x = normal_grid, panel = panel_levels[2],
             density = dnorm(normal_grid, mu_equity,
                             sigma_equity / sqrt(40)))
)
normal_curves$panel <- factor(normal_curves$panel, levels = panel_levels)

ggplot(sim_means, aes(sample_mean)) +
  geom_histogram(aes(y = after_stat(density)), breaks = shared_breaks,
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_line(data = normal_curves, aes(x, density),
            colour = "#D55E00", linewidth = 0.9) +
  facet_wrap(~ panel, scales = "fixed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Sample mean equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Read @fig-b08-clt with the five-step sequence of Lecture 1. (1) *Unit and
variables:* each bar counts simulated sample means, not pitches. The
horizontal axis is the sample mean of the equity offered, in percent. (2)
*Scales:* both panels use the same range and the same density scale, so
widths can be compared directly. (3) *One element:* the orange curve in each
panel is the normal approximation for that sample size. (4) *Pattern:* the
pile for $n=40$ is far narrower and much closer to its normal curve than the
pile for $n=5$, which still leans right like the population. (5) *Limits:*
the figure shows random samples from this file. It says nothing about samples
that were not drawn by chance.

### How many customers must I ask? {#b08-sample-size}

The square-root rule compares two sample sizes. A founder often asks a more
direct question: how large must the sample be to reach a given precision?

The CLT gives the answer. The central 95% of a normal distribution lies
within 1.96 standard deviations of its mean, because $z_{0.975}=1.96$
(Lecture 6). So in about 95% of
random samples, $\bar X$ lies within $1.96\times\sigma/\sqrt n$ of $\mu$. This
distance is called the margin of error at the 95% level. Lecture 9 uses it to
build confidence intervals. If the margin of error may be at most $m$, solve
for $n$:

$$
1.96\times\frac{\sigma}{\sqrt n}\le m
\qquad\text{gives}\qquad
n\ge\left(\frac{1.96\times\sigma}{m}\right)^2.
$$

::: {#exm-b08-sample-size}
## Worked example: a sample size for a margin of 2 euros

An invented example. You want to know the average order value of your
customers to within 2 euros, at the 95% level. How many customers must you
survey?

- **Planning value.** The formula needs $\sigma$, which you do not know
  before the survey. A small pilot survey gave an SD of about 12 euros, so
  use $\sigma=12$ as a planning value. The target margin of 2 euros is one
  sixth of that SD.
- **Sample size.** $n\ge(1.96\times 12/2)^2=11.76^2\approx 138.3$.
- **Round up.** Survey $n=139$ customers. Always round up: with 138
  customers the margin would be slightly above 2 euros.

Halving the margin to 1 euro needs $(1.96\times 12/1)^2=23.52^2\approx 553.2$,
so 554 customers. Half the margin costs about four times the sample, as the
square-root rule says.

```{r}
#| label: b08-sample-size-calc
sigma_plan <- 12
n_needed <- c(margin_2_euros = (1.96 * sigma_plan / 2)^2,
              margin_1_euro  = (1.96 * sigma_plan / 1)^2)
n_needed
ceiling(n_needed)   # ceiling() rounds up to the next whole number
```
:::

### Sample proportions are sample means

For a Bernoulli variable $X$ with $P(X=1)=p$ (Lecture 3),

$$
E[X]=p,
\qquad
\operatorname{Var}(X)=p(1-p).
$$

A sample proportion is the sample mean of a 0/1 variable (Lecture 7), so
every result above applies: $\widehat P=\bar X$ satisfies

$$
E[\widehat P]=p,
\qquad
\operatorname{SE}(\widehat P)=\sqrt{\frac{p(1-p)}{n}}.
$$

This formula needs the parameter $p$, which is usually unknown. Replacing $p$
by the estimate $\hat p$ gives the estimated standard error

$$
\widehat{\operatorname{SE}}(\widehat P)=\sqrt{\frac{\hat p(1-\hat p)}{n}}.
$$

Use the version with $p$ when the population proportion is given or assumed.
Use the estimated version when you only have the sample, as in Tutorial 3.

By the CLT, the sampling distribution of $\widehat P$ is approximately normal
for a large sample. How large is large enough for a proportion? The next
section answers this with the distribution of a count.

### Counting successes: the binomial distribution {#b08-binomial}

Take $n$ independent 0/1 trials, each with the same success probability $p$.
The number of successes is the sum $Y=X_1+\cdots+X_n$. Its distribution is
called the **binomial distribution**, written $Y\sim\text{Bin}(n,p)$. The sum
rules of [Lecture 5](b05-joint-covariance-correlation.qmd#b05-sums) give

$$
E[Y]=np,
\qquad
\operatorname{Var}(Y)=np(1-p).
$$

The sample proportion is this count divided by the number of trials:
$\widehat P=Y/n$.

An invented example: 200 people visit your web shop, and each of them buys
with probability 0.05, independently of the others. The number of buyers is
$Y\sim\text{Bin}(200,\,0.05)$. You expect $np=200\times 0.05=10$ buyers, with
a standard deviation of $\sqrt{200\times 0.05\times 0.95}=\sqrt{9.5}\approx 3.1$
buyers.

A count is a sum, so the CLT applies to it: for large $n$ the binomial
distribution is close to a normal distribution. @fig-b08-binomial shows when
$n$ is large enough. With 80 visitors you expect only $np=4$ buyers. The
distribution is skewed, with a longer tail to the right. The normal curve
puts about 2% of its area below zero, where no count can lie. The fit is
worse than in the right panel. With 400 visitors you expect $np=20$ buyers,
and the normal curve fits well.

```{r}
#| label: fig-b08-binomial
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The binomial distribution of the number of buyers when each visitor buys with probability 0.05 (green), with the normal curve of the same mean and variance (orange). With 80 visitors (left) the distribution is skewed and the normal curve reaches below zero. With 400 visitors (right) the normal curve fits well. The two panels have different axes."
library(ggplot2)

p_buy <- 0.05
n_small <- 80
n_large <- 400
panel_levels <- c("n = 80, np = 4: rule violated",
                  "n = 400, np = 20: rule satisfied")

pmf <- rbind(
  data.frame(y = 0:12, panel = panel_levels[1],
             prob = dbinom(0:12, size = n_small, prob = p_buy)),
  data.frame(y = 5:37, panel = panel_levels[2],
             prob = dbinom(5:37, size = n_large, prob = p_buy))
)
pmf$panel <- factor(pmf$panel, levels = panel_levels)

grid_small <- seq(-7, 12, length.out = 300)
grid_large <- seq(5, 37, length.out = 300)
normal_curves <- rbind(
  data.frame(y = grid_small, panel = panel_levels[1],
             density = dnorm(grid_small, n_small * p_buy,
                             sqrt(n_small * p_buy * (1 - p_buy)))),
  data.frame(y = grid_large, panel = panel_levels[2],
             density = dnorm(grid_large, n_large * p_buy,
                             sqrt(n_large * p_buy * (1 - p_buy))))
)
normal_curves$panel <- factor(normal_curves$panel, levels = panel_levels)
below_zero <- normal_curves[normal_curves$panel == panel_levels[1] &
                              normal_curves$y <= 0, ]

notes <- data.frame(
  y = c(-7, 26, 26),
  prob = c(0.055, 0.084, 0.068),
  label = c("normal curve\nreaches\nbelow 0", "normal curve",
            "binomial\nprobabilities"),
  colour = c("#D55E00", "#D55E00", "#009E73"),
  panel = factor(panel_levels[c(1, 2, 2)], levels = panel_levels)
)

ggplot(pmf, aes(y, prob)) +
  geom_area(data = below_zero, aes(y, density),
            fill = "#D55E00", alpha = 0.25) +
  geom_segment(aes(xend = y, yend = 0), colour = "#009E73", linewidth = 0.8) +
  geom_point(colour = "#009E73", size = 1.8) +
  geom_line(data = normal_curves, aes(y, density),
            colour = "#D55E00", linewidth = 0.9) +
  geom_text(data = notes, aes(label = label), colour = notes$colour,
            size = 3.2, hjust = 0, lineheight = 0.95) +
  facet_wrap(~ panel, scales = "free") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Number of buyers", y = "Probability") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold", size = 10)
  )
```

This course therefore uses the normal approximation for a count or a
proportion only if $np\ge10$ and $n(1-p)\ge10$. This rule is called the
**success-failure condition**: the expected numbers of successes and of
failures must both be at least 10. When $p$ is unknown, check the rule with
the counts in the sample: at least 10 successes and at least 10 failures. In
a test, use the null value $p_0$ (Lecture 10). Some textbooks use 5 in place
of 10. Ten is the safer choice. You do not need to calculate binomial
probabilities in this course.

::: {#exm-b08-proportion}
## Worked example: SE of the deal proportion at n = 100

Suppose we draw a random sample of $n=100$ pitches from the frame, where the
deal proportion is $p=0.612$. How far is the sample deal proportion
typically from 0.612?

- **SE.**
  $\operatorname{SE}(\widehat P)=\sqrt{0.612\,(1-0.612)/100}
  =\sqrt{0.2375/100}\approx 0.0487$, about 4.9 percentage points.
- **Success-failure check.** $np=100(0.612)=61.2\ge 10$ and
  $n(1-p)=100(0.388)=38.8\ge 10$, so the normal approximation is reasonable.

In R, using `p_deal` as constructed above:

```{r}
#| label: b08-proportion-se
p_deal <- mean(sharks$deal_on_show)
c(SE_deal_proportion_n_100 = sqrt(p_deal * (1 - p_deal) / 100),
  successes = 100 * p_deal, failures = 100 * (1 - p_deal))
```
:::

### The law of large numbers and the CLT answer different questions

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-weak-law-of-large-numbers/">18.4 The Weak Law of Large Numbers</a>.</p>

::: {.definition-box}
<span class="label">Definition</span>

**Law of large numbers (LLN).** As the sample size $n$ grows, the sample
mean $\bar X$ of a random sample gets close to the population mean $\mu$.
:::

This is the consistency of $\bar X$ noted above. The LLN and the CLT answer
different questions:

- The LLN says where the pile of sample means goes: towards $\mu$.
- The CLT says what the pile looks like: its approximate shape and width.

The LLN is about a value. The CLT is about a distribution. Both assume that
the sampling is random. Neither repairs a sample that was selected badly in
the first place.

::: {.callout-note collapse="true"}
## Check: match each statement to LLN, CLT, or neither

1. "With enough random draws, the sample mean equity will be very close to
   13.21."
2. "For $n=100$ random draws, the sample mean equity is approximately
   $N(13.21, 0.844^2)$."
3. "With enough respondents, our LinkedIn poll will represent Dutch
   founders."

Statement 1 is the LLN, statement 2 is the CLT, and statement 3 is
**neither**. No theorem turns a self-selected sample into a representative
one (see the box below).
:::

::: {.caveat-box}
<span class="label">What a bigger sample can and cannot fix</span>

- More data shrink the random error, at the square-root rate.
- More data do nothing for selection bias. A million televised pitches still
  describe televised pitches.
- The standard error measures precision inside the list you sampled from. It
  is silent about who is on the list.

**What you can do with it:** first fix who you ask, then decide how many.
Stop collecting when more data would no longer change your decision.
:::

### Answering the driving question

A random sample of 100 pitches from the Shark Tank file had a mean equity
offered of 12.76%, while the mean of all 1,441 records is 13.21%. How wrong
could 12.76 have been? Its standard error:

```{r}
#| label: b08-driving-question
sigma_equity <- sqrt(mean((sharks$equity_offered_pct -
                             mean(sharks$equity_offered_pct))^2))
c(SE_mean_n_25 = sigma_equity / sqrt(25),
  SE_mean_n_100 = sigma_equity / sqrt(100))
```

At $n=100$ the SE is about 0.84 percentage points, which is about 6% of the
population mean. The mean of a random sample typically lands within about
0.84 points of 13.21, and rarely more than about two SEs away. The sample
with 12.76 missed by 0.45 points, about half an SE. At $n=25$ the SE would be
twice as large, by the square-root rule.

Why can we trust the average although the equity offered in single pitches
is strongly skewed? Deviations cancel in an average (the variance rule), and
the CLT gives the mean a predictable, nearly normal distribution.

One detail: our samples are drawn without replacement from 1,441 records.
That makes the exact standard error slightly smaller than the formula says,
0.81 instead of 0.84. The 10% rule says that the difference can be ignored.
The optional box in
[Lecture 7](b07-samples-estimators.qmd#b07-parameters) gives the exact
formula.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

You ask customers to rate your boxes from 1 to 10. Ratings have a standard deviation of about 2. With 25 customers the standard error of the mean is 2/5 = 0.4. With 100 customers it is 0.2. Is the larger survey worth four times the effort? Your rule is to change the menu only if the mean rating is below 7. The first 25 customers give a mean of 8.4, which is 3.5 standard errors above 7. The decision: stop surveying and keep the menu. More data would not change what you do.
:::

**Next:** We can now say how much a sample mean typically varies: by about
one standard error. A report needs a sentence such as "the population mean
plausibly lies between $a$ and $b$". That takes one more step, because in a
real sample $\sigma$ is unknown and must be estimated too, which adds
uncertainty. Lecture 9 introduces the t distribution, which allows for this
extra uncertainty, and builds the confidence interval from it.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 80 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. What varies in a sampling distribution?
2. Why is $E[\bar X]=\mu$ called unbiasedness?
3. What is the difference between an SD and an SE?
4. To cut an SE in half, by what factor must the sample size grow, and why?
5. What does the CLT say becomes approximately normal?
6. Why does a larger sample not remove selection bias?

::: {.callout-note collapse="true"}
## Check your answers

1. The statistic varies across possible repeated samples. The parameter and
   the population stay fixed.
2. Its long-run average across repeated samples equals the population
   parameter.
3. An SD describes variation among observations. An SE describes variation
   of an estimator across samples.
4. By a factor of 4: the SE is proportional to $1/\sqrt n$, so halving it
   requires $2^2=4$ times the observations.
5. The sample mean. For a large sample its sampling distribution is
   approximately normal. The single observations do not become normal.
6. Sample size shrinks random error only. It does not change how units
   entered the sampling frame or the sample.
:::

### Finish a half-worked example (5 minutes)

Re-read the worked SE calculation (@exm-b08-se). Then complete the same
calculation for a proportion: a planned survey expects $p=0.30$ and
$n=100$.

*Step 1 (given).* The SE formula for a proportion is
$\sqrt{p(1-p)/n}$.

*Step 2 (you).* Calculate the SE.

*Step 3 (you).* Check the success-failure condition.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SE}(\widehat P)=\sqrt{0.30(0.70)/100}\approx 0.0458$, about
4.6 percentage points. The expected counts are $np=30$ successes and
$n(1-p)=70$ failures, both at least 10, so the normal approximation is
acceptable.
:::

### Practice (15 minutes)

1. If $\sigma=15$ and $n=25$, find $E[\bar X]$,
   $\operatorname{Var}(\bar X)$, and $\operatorname{SE}(\bar X)$, naming
   the rule used at each step.
2. How large must $n$ become to reduce an SE from 6 to 1.5?
3. For a sample proportion with $p$ assumed to be 0.50 and $n=64$,
   calculate the SE and check the success-failure condition.
4. Explain why the normal approximation is more convincing in the
   right-hand panel of @fig-b08-clt.
5. State one threat to the CLT-based approximation that a large $n$ alone
   does not solve.
6. An invented example. A founder wants to know the average delivery time of
   her orders to within 1 minute, at the 95% level. A pilot gave an SD of 5
   minutes. How many deliveries must she time? How does the answer change
   for a margin of half a minute?

::: {.callout-note collapse="true"}
## Check your answers

1. $E[\bar X]=\mu$ (rules for expectations),
   $\operatorname{Var}(\bar X)=\sigma^2/n=225/25=9$ (variance of a rescaled
   sum of independent variables), $\operatorname{SE}(\bar X)=\sqrt 9=3$.
2. The SE must fall by a factor of 4, so the sample size must grow by
   $4^2=16$ times.
3. $\sqrt{0.50(0.50)/64}=0.0625$. Expected counts are 32 and 32, both at
   least 10.
4. Averaging 40 observations cancels more of the skewness and shrinks the
   sampling variation more than averaging 5 observations. The histogram
   follows its normal curve much more closely.
5. Dependence between observations. A non-representative sample is a
   different problem: the approximation still works, but the mean then
   estimates the wrong quantity. (Extreme outliers slow the approximation too.)
6. $n\ge(1.96\times 5/1)^2=9.8^2=96.04$, so she must time 97 deliveries. For
   half a minute, $n\ge(1.96\times 5/0.5)^2=19.6^2=384.16$, so 385
   deliveries. Half the margin costs about four times the sample.
:::

### Widget homework (10 minutes)

Open the
[sampling-distribution lab](b07-samples-estimators.qmd#sampling-lab) in
Lecture 7 again and run these three settings. For each, redraw several times
and write **one sentence** on what the histogram of sample means shows:

1. $n = 2$
2. $n = 25$
3. $n = 100$.

::: {.callout-note collapse="true"}
## Check a model answer

1. At $n=2$ the sample means scatter widely and their histogram is still
   right-skewed, like the population.
2. At $n=25$ the histogram of means is clearly narrower and roughly
   symmetric, centred near 13.2.
3. At $n=100$ the means form a tight bell around the population mean, with
   a typical deviation of less than one percentage point. This is the SE of
   about 0.84 percentage points.
:::

### Review: one question each from Lectures 7 and 6 (10 minutes)

1. *(Lecture 7.)* The sample variance divides by $n-1$, not $n$. In one or
   two sentences, why?
2. *(Lecture 6.)* Daily revenue is modelled as $X\sim N(10, 2^2)$ (in
   thousands of euros). Find $P(X>13)$. Supplied value:
   $P(Z\le 1.5)=0.9332$.

::: {.callout-note collapse="true"}
## Check your answers

1. The deviations are measured from the sample mean, which was itself
   estimated from the data. Once it is fixed, only $n-1$ deviations can
   vary freely, and dividing by $n-1$ removes the resulting downward bias.
2. Standardise: $z=(13-10)/2=1.5$, so
   $P(X>13)=P(Z>1.5)=1-0.9332\approx 0.067$.
:::

### Worked exam interpretation (5 minutes)

**Question.** The mean equity offered (`equity_offered_pct`) in the Shark
Tank file is $\mu=13.21$% with $\sigma=8.44$ percentage points. A random
sample of $n=100$ pitches is drawn. Using the CLT, find the probability that
the sample mean exceeds 14.5%. Supplied value: $P(Z\le 1.53)=0.937$.

::: {.callout-note collapse="true"}
## Check a model answer

By the CLT, $\bar X\ \dot\sim\ N(13.21, 0.844^2)$ since
$\operatorname{SE}=8.44/\sqrt{100}=0.844$. Standardise:
$z=(14.5-13.21)/0.844\approx 1.53$, so
$P(\bar X>14.5)\approx P(Z>1.53)=1-0.937=0.063$. That is about a 6% chance.
Note that the question is about the mean, not about a single pitch. For a
single pitch the normal model was rejected in Lecture 6.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "My ratings have a standard deviation of 12 and I asked 36 customers. How precise is the mean, and how do I halve the error?" It answered:

> "The standard error is 12/36 = 0.33. To halve it you need twice as many customers, 72."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Both steps are wrong. The standard error divides by the square root of the sample size: 12/6 = 2. And to halve it you need four times the data, 144 customers, because the standard error falls with the square root of the sample size.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Name one number
you would estimate from a sample (customer satisfaction, conversion rate,
average basket size) and write two sentences. First say how you would draw
the sample and roughly what standard error you would accept. Then say
whether collecting four times the data would be worth halving that standard
error for the decision you face.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I would estimate the conversion rate of tasting-stand visitors from a random 100 of them, and a standard error of about 5 percentage points is fine for deciding on a second stand. Collecting four times the data would halve that to about 2.5 points, which is not worth the extra tasting days for this decision."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to help you sketch the
sample-size trade-off for your own numbers.
::::

[Continue to Lecture 9: The t distribution and confidence intervals](b09-t-confidence-intervals.qmd).

[Formula sheet: Lecture 8 formulas](formula-sheet.qmd#formula-l8).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 13 and 14.

:::
