# Lecture 8: Sampling distributions and the central limit theorem {#b08}

```{r}
#| label: b08-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 8, you can</span>

- distinguish a population distribution, a sampling distribution, and a
  realised estimate;
- derive the mean and variance of a sample mean using the linearity rules;
- calculate and interpret a standard error;
- use the square-root rule to relate sample size to precision;
- state the central limit theorem for this course's conditions and apply it
  to means and proportions;
- explain what the law of large numbers and the CLT each claim --- and what
  neither of them repairs.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Any number computed from a sample would have come out differently with a different sample. Before acting on an estimate, a careful analyst asks how much such numbers wobble --- and what tames the wobble.</p>

Our sample of 100 gave mean equity 12.76. How wrong could that number be, and
why does averaging make it trustworthy even though single pitches are wildly
skewed?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=Yh5bR7X3ch8">The Weak Law
of Large Numbers</a> and
<a href="https://www.youtube.com/watch?v=IrKUM3nNXJE">The Central Limit
Theorem</a>. Focus on what converges as sample size grows and on which
distribution becomes approximately normal. All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (10 minutes).** Classify each object as a *population
distribution*, a *sampling distribution*, or a *realised estimate*:

1. the distribution of equity offered across individual pitches;
2. the distribution of $\bar X$ across all possible samples of 40 pitches;
3. the single mean calculated from one observed sample.

::: {.callout-note collapse="true"}
## Check your answers

The first is a population distribution, the second is a sampling
distribution, and the third is a realised estimate. Keep the three apart all
lecture: the middle one is the new object.
:::

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: to cut the typical error of a sample mean
**in half**, do you need $2\times$, $4\times$, or $10\times$ the
observations? Commit to an answer; the lecture resolves it.

If square roots, sums, degrees of freedom, or standardisation are not secure,
review [Mathematics Refresher: powers, roots, and
standardisation](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. Start from where Lecture 7 ended: **the
five dots.** Five random samples of 100 pitches produced five different
sample means, plotted on one axis, while the population mean sat still
underneath them. During the break you guessed where a sixth mean would land.
This lecture is about the *pattern* behind those dots --- the hardest and most
useful idea in the course.

### The pattern behind the five dots

Imagine not five samples but *every possible* random sample of the same size,
each producing its own sample mean. The distribution of those possible values
of $\bar X$ is its **sampling distribution**. It describes uncertainty in an
*estimator*, not variation among individual pitches:

- the **population distribution** describes how equity offers vary across
  the 1,441 pitches;
- the **sampling distribution** describes how $\bar X$ would vary across
  repeated samples;
- a **realised estimate**, such as 12.76, is one draw from that sampling
  distribution.

The two population parameters this chapter keeps returning to are the frame
mean and SD of equity offered, and the frame deal proportion. Construct them
yourself --- because the 1,441 records are the entire frame, the SD divides by
$N$, not $n-1$:

```{r}
#| label: b08-population-parameters
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
p_deal <- mean(sharks$deal_on_show)
c(mu_equity = mu_equity, sigma_equity = sigma_equity, p_deal = p_deal)
```

So $\mu = 13.21$ percentage points, $\sigma = 8.44$ percentage points, and
$p = 0.612$. These are fixed numbers. What varies is any statistic computed
from a random sample of the frame.

### Explore it: the sampling-distribution lab

Before any formula, see the sampling distribution *form*. The lab below shows
the strongly right-skewed population histogram of the 1,441 equity offers on
the left. Use the **slider** to choose a sample size $n$ and press
**Redraw** to draw many random samples of that size: each sample's mean drops
into the histogram forming on the right, next to the population.

<iframe src="../assets/widgets/sampling-lab.html" width="100%" height="520" style="border:1px solid #C9D4DC; border-radius:6px;" title="Sampling distribution lab"></iframe>

Work through three settings and watch three things --- centre, spread, shape:

1. Set $n = 2$ and redraw several times. The means scatter widely and the
   histogram of means is still skewed.
2. Set $n = 25$. The histogram of means narrows and starts to look
   symmetric, centred near 13.2.
3. Slide up to $n = 100$. The histogram of means is a tight, bell-shaped
   pile around the population mean, even though the population histogram
   beside it has not changed at all.

Everything in the rest of this lecture is the mathematics of what you just
manipulated: *where* the pile sits, *how wide* it is, and *why* it becomes
bell-shaped.

### The mean and variance of a sample mean

For independent, identically distributed variables $X_1,\ldots,X_n$ with
population mean $\mu$ and variance $\sigma^2$,

$$
E[\bar X]=\mu,
\qquad
\operatorname{Var}(\bar X)=\frac{\sigma^2}{n},
\qquad
\operatorname{SD}(\bar X)=\frac{\sigma}{\sqrt n}.
$$

The first equality says that $\bar X$ is **unbiased** for $\mu$: the pile of
sample means is centred on the parameter, exactly as in the lab. The third
quantity --- the SD of the estimator across repeated samples --- is called the
**standard error** (SE). An SD describes variation among *observations*; an
SE describes variation of an *estimator*.

These are not new facts. They follow from the linearity rules you already
have on the formula sheet (the Lecture 8 block, and the Refresher's
linearity rules):

- $E[X+Y]=E[X]+E[Y]$ and $E[aX]=aE[X]$, so
  $E[\bar X]=\frac{1}{n}\sum_i E[X_i]=\frac{1}{n}\,n\mu=\mu$;
- $\operatorname{Var}(aX)=a^2\operatorname{Var}(X)$ and, for
  **independent** variables,
  $\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)$, so

$$
\operatorname{Var}(\bar X)
=\operatorname{Var}\!\left(\frac{X_1+\cdots+X_n}{n}\right)
=\frac{1}{n^2}\sum_{i=1}^{n}\operatorname{Var}(X_i)
=\frac{n\sigma^2}{n^2}
=\frac{\sigma^2}{n}.
$$

Independent positive and negative deviations partly cancel in an average.
That cancellation is why the right-hand histogram in the lab is so much
narrower than the population histogram.

In practice $\sigma$ is usually unknown and is estimated by the sample SD
$S$, giving the estimated standard error
$\operatorname{SE}(\bar X)=S/\sqrt n$. Substituting an estimate for a
parameter adds a little extra uncertainty; Lecture 9 opens with the price of
that substitution.

::: {#exm-b08-se}
## Worked example: E, Var, and SE for σ = 12, n = 36

A population has mean $\mu$ and SD $\sigma=12$. A random sample of $n=36$
observations is drawn. Find the mean, variance, and SE of $\bar X$, naming
the rule used at each step.

- **Mean.** By linearity of expectation ($E[X+Y]=E[X]+E[Y]$, $E[aX]=aE[X]$),
  $E[\bar X]=\mu$ --- whatever $\mu$ is, the sample mean is unbiased for it.
- **Variance.** By $\operatorname{Var}(aX)=a^2\operatorname{Var}(X)$ and
  additivity of variance under independence,
  $\operatorname{Var}(\bar X)=\sigma^2/n=12^2/36=144/36=4$.
- **SE.** Take the square root:
  $\operatorname{SD}(\bar X)=\sqrt 4=2$, i.e. $\sigma/\sqrt n=12/6=2$.

The typical distance between a sample mean and $\mu$ is about 2 units ---
one sixth of the typical distance of a *single* observation from $\mu$.
All three rules are on the formula sheet; the derivations are in the
[Mathematics Refresher](../resources/math-refresher.html).
:::

### The square-root rule

Precision grows with $\sqrt n$, not with $n$. This resolves the prep
prediction: to cut the SE **in half** you need $\mathbf{4\times}$ the
observations, because

$$
\frac{\sigma}{\sqrt{4n}}=\frac{1}{2}\cdot\frac{\sigma}{\sqrt n}.
$$

Doubling the sample buys less than you might hope; each extra digit of
precision gets more expensive. This is worth knowing before you budget for
any data collection.

::: {.callout-caution collapse="true"}
## Try it: how large must n become to cut an SE from 4 to 1?

In class this is the pair task before the break. Complete the missing steps
before opening the answer.

*Step 1 (given).* The SE must fall by a factor of $4/1 = 4$.

*Step 2 (you).* The SE is proportional to $1/\sqrt n$. By what factor must
$n$ change to divide the SE by 4?

*Step 3 (you).* If the original sample had $n=50$, how many observations
does the new design need?

::: {.callout-note collapse="true"}
## Check your answer

Dividing the SE by 4 requires multiplying $n$ by $4^2=16$: the square-root
rule in reverse. From $n=50$ the design needs $16\times 50=800$
observations. Precision is bought at a quadratic price.
:::
:::

### The central limit theorem

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-central-limit-theorem/">19.2 The Central Limit Theorem</a>.</p>

The lab showed one more pattern that the mean-and-variance rules do not
explain: the pile of sample means became *bell-shaped*, even though the
population is strongly skewed. That is the **central limit theorem** (CLT).
Under the independent-sampling conditions used in this course, the
standardised sample mean

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n}
$$

approaches a standard normal distribution as $n$ grows. Equivalently,

$$
\bar X\ \dot\sim\ N\!\left(\mu,\frac{\sigma^2}{n}\right)
$$

for a sufficiently large sample. The dot above $\sim$ signals an
approximation. The normal model from Lecture 6 returns here --- not as a model
for the *data*, where it failed for equity offers, but as a model for the
*sample mean*.

There is **no universal rule that $n=30$ is always enough.** Strong
skewness, extreme outliers, or dependence between observations may require
much more data. If the population itself is normal, the sample mean is
exactly normal for every $n$.

@fig-b08-clt shows the simulation behind the claim: 5,000 sample means of
$n=5$ draws and 5,000 sample means of $n=40$ draws from the equity variable.

```{r}
#| label: fig-b08-clt
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Simulated sampling distributions of the mean equity offered for n = 5 (left) and n = 40 (right), each based on 5,000 samples, with the corresponding normal approximations (orange curves). Both panels use the same axes."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))
set.seed(208)
B <- 5000
means_5 <- replicate(B, mean(sample(sharks$equity_offered_pct, 5,
                                    replace = TRUE)))
means_40 <- replicate(B, mean(sample(sharks$equity_offered_pct, 40,
                                     replace = TRUE)))

panel_levels <- c("n = 5", "n = 40")
sim_means <- data.frame(
  sample_mean = c(means_5, means_40),
  panel = factor(rep(panel_levels, each = B), levels = panel_levels)
)

all_means <- c(means_5, means_40)
shared_limits <- c(floor(min(all_means)), ceiling(max(all_means)))
shared_breaks <- seq(shared_limits[1], shared_limits[2], length.out = 36)

normal_grid <- seq(shared_limits[1], shared_limits[2], length.out = 400)
normal_curves <- rbind(
  data.frame(x = normal_grid, panel = panel_levels[1],
             density = dnorm(normal_grid, mu_equity,
                             sigma_equity / sqrt(5))),
  data.frame(x = normal_grid, panel = panel_levels[2],
             density = dnorm(normal_grid, mu_equity,
                             sigma_equity / sqrt(40)))
)
normal_curves$panel <- factor(normal_curves$panel, levels = panel_levels)

ggplot(sim_means, aes(sample_mean)) +
  geom_histogram(aes(y = after_stat(density)), breaks = shared_breaks,
                 fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_line(data = normal_curves, aes(x, density),
            colour = "#D55E00", linewidth = 0.9) +
  facet_wrap(~ panel, scales = "fixed") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Sample mean equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Read @fig-b08-clt with the five-step sequence. (1) Each bar counts simulated
sample means, not pitches. (2) The horizontal axis is the possible sample
mean of equity offered, in percentage points --- **the same range in both
panels**, so widths can be compared directly. (3) The vertical axis is
density, also on a shared scale. (4) The pattern: the $n=40$ pile is far
narrower and visibly closer to its orange normal curve than the $n=5$ pile,
which still leans right like the population. (5) Takeaway: averaging 40
independent draws from a wildly skewed population already produces a
concentrated, nearly normal sampling distribution.

### Sample proportions are sample means

For a Bernoulli variable $X$ with $P(X=1)=p$ (Lecture 3),

$$
E[X]=p,
\qquad
\operatorname{Var}(X)=p(1-p).
$$

A sample proportion is just the sample mean of an indicator, so every result
above applies immediately: $\widehat P=\bar X$ satisfies

$$
E[\widehat P]=p,
\qquad
\operatorname{SE}(\widehat P)=\sqrt{\frac{p(1-p)}{n}}.
$$

Its sampling distribution is approximately normal when the expected numbers
of successes and failures are both sufficiently large. This course uses
$np\ge10$ **and** $n(1-p)\ge10$ as the working rule (the *success--failure
condition*).

::: {#exm-b08-proportion}
## Worked example: SE of the deal proportion at n = 100

Suppose we draw a random sample of $n=100$ pitches from the frame, where the
deal proportion is $p=0.612$. How far is the sample deal proportion
typically from 0.612?

- **SE.**
  $\operatorname{SE}(\widehat P)=\sqrt{0.612\,(1-0.612)/100}
  =\sqrt{0.2375/100}\approx 0.0487$, about 4.9 percentage points.
- **Success--failure check.** $np=100(0.612)=61.2\ge 10$ and
  $n(1-p)=100(0.388)=38.8\ge 10$, so the normal approximation is reasonable.

In R, using `p_deal` as constructed above:

```{r}
#| label: b08-proportion-se
p_deal <- mean(sharks$deal_on_show)
c(SE_deal_proportion_n_100 = sqrt(p_deal * (1 - p_deal) / 100),
  successes = 100 * p_deal, failures = 100 * (1 - p_deal))
```
:::

### The law of large numbers and the CLT answer different questions

<p class="concept-video"><strong>MIT explanation (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/the-weak-law-of-large-numbers/">18.4 The Weak Law of Large Numbers</a>.</p>

- The **law of large numbers** (LLN) says *where the pile goes*: $\bar X$
  becomes close to $\mu$ as $n$ grows.
- The **CLT** says *what the pile looks like*: the approximate shape and
  scale of the sampling distribution after standardisation.

One is about convergence of a value, the other about convergence of a
distribution. Both assume the sampling is random; neither repairs a sample
that was selected badly in the first place.

::: {.callout-note collapse="true"}
## Check: match each statement to LLN, CLT, or neither

1. "With enough random draws, the sample mean equity will be very close to
   13.21."
2. "For $n=100$ random draws, the sample mean equity is approximately
   $N(13.21, 0.844^2)$."
3. "With enough respondents, our LinkedIn poll will represent Dutch
   founders."

Statement 1 is the LLN, statement 2 is the CLT, and statement 3 is
**neither**: no theorem turns a self-selected sample into a representative
one --- see the caveat box.
:::

::: {.caveat-box}
<span class="label">What a bigger sample can and cannot fix</span>

Everything in this lecture concerns **random sampling error**: the
unavoidable wobble of a statistic across random samples. More observations
shrink that wobble at the square-root rate. They do **nothing** for
**selection bias**: if pitches enter the data because producers chose good
television, a sample of a million such pitches still describes televised
pitches, not entrepreneurs in general. An SE of 0.84 quantifies precision
*within the recorded-pitch frame*; it is silent about who got into the frame,
and an on-air deal is still not a completed investment.
:::

### Answering the driving question

Our sample of 100 gave mean equity 12.76, while the frame mean is 13.21. How
wrong could 12.76 have been? Its standard error:

```{r}
#| label: b08-driving-question
sigma_equity <- sqrt(mean((sharks$equity_offered_pct -
                             mean(sharks$equity_offered_pct))^2))
c(SE_mean_n_25 = sigma_equity / sqrt(25),
  SE_mean_n_100 = sigma_equity / sqrt(100))
```

At $n=100$ the SE is about **0.84 percentage points**: a random-sample mean
typically lands within about 0.84 points of 13.21, and rarely more than
about two SEs away. (At $n=25$ the SE would be twice as large --- the
square-root rule again.) Averaging is trustworthy despite the wild skew of
single pitches because deviations cancel (the variance rule) and the CLT
gives the mean a predictable, nearly normal distribution.

**Next:** We can now say how much a sample mean typically wobbles: one
standard error. Turning that into a sentence a report can carry --- "the frame
mean plausibly lies between $a$ and $b$" --- needs one more ingredient,
because in a real sample the population $\sigma$ is unknown and must be
estimated too, adding uncertainty of its own. Lecture 9 introduces the t
distribution that prices in this extra uncertainty, builds the confidence
interval from it, and runs the course's first formal test of a benchmark
claim.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 55 minutes.</span>

### Retrieval (10 minutes)

1. What varies in a sampling distribution?
2. Why is $E[\bar X]=\mu$ called unbiasedness?
3. What is the difference between an SD and an SE?
4. To cut an SE in half, by what factor must the sample size grow, and why?
5. What does the CLT say becomes approximately normal?
6. Why does a larger sample not remove selection bias?

::: {.callout-note collapse="true"}
## Check your answers

1. The statistic varies across possible repeated samples; the parameter and
   the population stay fixed.
2. Its long-run average across repeated samples equals the population
   parameter.
3. An SD describes variation among observations; an SE describes variation
   of an estimator across samples.
4. By a factor of 4: the SE is proportional to $1/\sqrt n$, so halving it
   requires $2^2=4$ times the observations.
5. The standardised sample mean --- and therefore, approximately, the
   sampling distribution of the mean. Not the original observations.
6. Sample size shrinks random error only. It does not change how units
   entered the sampling frame or the sample.
:::

### From worked to faded (5 minutes)

Re-read the worked SE calculation (@exm-b08-se). Then complete the same
calculation for a proportion: a planned survey expects $p=0.30$ and
$n=100$.

*Step 1 (given).* The SE formula for a proportion is
$\sqrt{p(1-p)/n}$.

*Step 2 (you).* Calculate the SE.

*Step 3 (you).* Check the success--failure condition.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SE}(\widehat P)=\sqrt{0.30(0.70)/100}\approx 0.0458$, about
4.6 percentage points. The expected counts are $np=30$ successes and
$n(1-p)=70$ failures, both at least 10, so the normal approximation is
acceptable.
:::

### Practice (10 minutes)

1. If $\sigma=15$ and $n=25$, find $E[\bar X]$,
   $\operatorname{Var}(\bar X)$, and $\operatorname{SE}(\bar X)$, naming
   the rule used at each step.
2. How large must $n$ become to reduce an SE from 6 to 1.5?
3. For a sample proportion with $\hat p$ expected near 0.50 and $n=64$,
   calculate the SE and check the success--failure condition.
4. Explain why the normal approximation is more convincing in the
   right-hand panel of @fig-b08-clt.
5. State one threat to the CLT-based approximation that a large $n$ alone
   does not solve.

::: {.callout-note collapse="true"}
## Check your answers

1. $E[\bar X]=\mu$ (linearity of expectation);
   $\operatorname{Var}(\bar X)=\sigma^2/n=225/25=9$ (variance of a scaled
   sum of independent variables); $\operatorname{SE}(\bar X)=\sqrt 9=3$.
2. The SE must fall by a factor of 4, so the sample size must grow by
   $4^2=16$ times.
3. $\sqrt{0.50(0.50)/64}=0.0625$. Expected counts are 32 and 32, both at
   least 10.
4. Averaging 40 observations cancels more of the skewness and shrinks the
   sampling variation more than averaging 5 observations; the histogram
   hugs its normal curve far more closely.
5. Dependence between observations, or a non-representative sampling
   process. (Extreme outliers slow the approximation too.)
:::

### Widget homework (10 minutes)

Open the sampling-distribution lab again and run these three settings. For
each, redraw several times and write **one sentence** on what the histogram
of sample means shows:

1. $n = 2$;
2. $n = 25$;
3. $n = 100$.

::: {.callout-note collapse="true"}
## Check a model answer

1. At $n=2$ the sample means scatter widely and their histogram is still
   right-skewed, echoing the population shape.
2. At $n=25$ the histogram of means is clearly narrower and roughly
   symmetric, centred near 13.2.
3. At $n=100$ the means form a tight bell around the population mean, with
   a typical deviation of less than one percentage point --- the SE of about
   0.84 made visible.
:::

### Spaced review: one item each from Lectures 7 and 6 (10 minutes)

1. *(Lecture 7.)* The sample variance divides by $n-1$, not $n$. In one or
   two sentences, why?
2. *(Lecture 6.)* Daily revenue is modelled as $X\sim N(10, 2^2)$ (in
   thousands of euros). Find $P(X>13)$.

::: {.callout-note collapse="true"}
## Check your answers

1. The deviations are measured from the sample mean, which was itself
   estimated from the data; once it is fixed, only $n-1$ deviations can
   vary freely, and dividing by $n-1$ removes the resulting downward bias.
2. Standardise: $z=(13-10)/2=1.5$, so
   $P(X>13)=P(Z>1.5)\approx 0.067$.
:::

### Worked exam interpretation (5 minutes)

**Question.** Mean equity offered in the recorded-pitch frame is
$\mu=13.21$ with $\sigma=8.44$. A random sample of $n=100$ pitches is
drawn. Using the CLT, find the probability that the sample mean exceeds
14.5 percentage points.

::: {.callout-note collapse="true"}
## Check a model answer

By the CLT, $\bar X\ \dot\sim\ N(13.21, 0.844^2)$ since
$\operatorname{SE}=8.44/\sqrt{100}=0.844$. Standardise:
$z=(14.5-13.21)/0.844\approx 1.53$, so
$P(\bar X>14.5)\approx P(Z>1.53)\approx 0.063$. About a 6% chance --- and
note the question is about the *mean*, not a single pitch: for one pitch
the normal model was already rejected in Lecture 6.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Name one number
you would estimate from a sample (customer satisfaction, conversion rate,
average basket size) and write two sentences: first say how you would draw
the sample and roughly what SE-driven wobble you would tolerate, then say
whether collecting four times the data would be worth halving that wobble
for the decision at hand.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I would estimate the conversion rate of tasting-stand visitors from a random 100 of them, and a wobble of about $\pm 5$ percentage points is fine for deciding on a second stand. Collecting four times the data would halve that to about $\pm 2.5$ points, which is not worth four more tasting days for this decision."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to help you sketch the
sample-size trade-off for your own numbers.
::::

Continue to Lecture 9: The t distribution, confidence intervals, and the
two-sided test.

Formula sheet: Lecture 8 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 13 and 14.

:::
