# Lecture 9: The t distribution, confidence intervals, and the two-sided test {#b09}

```{r}
#| label: b09-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 9, you can</span>

- explain why estimating the population SD leads to a t distribution with
  $n-1$ degrees of freedom, and use its critical values;
- construct a t confidence interval for a population mean;
- express comparison with a benchmark as a parameter equal to zero;
- state a two-sided null and alternative hypothesis;
- calculate and interpret a t statistic and a two-sided p-value;
- connect a 95% confidence interval to a 5% two-sided test;
- state correctly what "95% confidence" is a property of.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">A common pitch-deck rule of thumb says founders offer investors about 15 percent equity. Our data can put such a benchmark to the test --- if we account honestly for sampling noise.</p>

Is the typical initial equity ask in the recorded-pitch frame really
different from 15 percent?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=NInNhFm046w">Confidence
Intervals</a>,
<a href="https://www.youtube.com/watch?v=MzvRQFYUEFU">Confidence Intervals
When the Variance Is Unknown</a>, and
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to
Hypothesis Testing</a>. Focus on repeated-sampling coverage, the t reference
distribution, and the role of the null value. All links are also listed in
the <a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (5 minutes).** Let $X$ be initial equity offered and define the
**equity gap**

$$
D = X - 15.
$$

Write two-sided hypotheses for the question: "Does population mean initial
equity differ from 15 percentage points?"

::: {.callout-note collapse="true"}
## Check your preparation

Let $\delta=E[D]=\mu_X-15$. Then

$$
H_0:\delta=0
\qquad\text{versus}\qquad
H_1:\delta\ne0.
$$

This is equivalent to testing $H_0:\mu_X=15$ against $H_1:\mu_X\ne15$.
Writing the contrast as $\delta$ makes the central question "is the
difference zero?" --- the same structure later used for regression
coefficients.
:::

**Predict (5 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: in a real sample, the unknown population SD
$\sigma$ must be replaced by the sample SD $S$. Should that substitution
make a 95% interval **wider**, **narrower**, or leave it **unchanged**?
Commit to an answer; the lecture resolves it.

If intervals, standardisation, or degrees of freedom feel rusty, review
[Mathematics Refresher: intervals, standardisation, and degrees of
freedom](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. Lecture 8 left us with a sample mean and
its standard error. This lecture turns those two numbers into a defensible
claim: a range of plausible values for the parameter, and a yes-or-no
decision about a benchmark.

<p class="concept-video"><strong>Video explanations (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/confidence-intervals/">20.5 Confidence Intervals</a>,
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/confidence-intervals-for-the-mean-when-the-variance-is-unknown/">20.7 Confidence Intervals with Unknown Variance</a>, and
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to Hypothesis Testing</a>.</p>

### From a benchmark to a zero contrast

The driving question compares a population mean to the benchmark 15. The
prep exercise already made the key move: subtract the benchmark and study
the **gap**

$$
D = X - 15,
\qquad
\delta = E[D] = \mu_X - 15.
$$

"Is mean equity different from 15?" becomes "is $\delta$ different from
**zero**?" This looks like a cosmetic rewrite, but it is the single most
reusable idea of the next four weeks: every benchmark question can be turned
into a question about a parameter being zero. Lecture 10 shows that group
differences, proportion gaps, and regression slopes all wear the same
zero-contrast uniform. Learn the pattern once here, and every later test is
a rerun.

### The price of estimating σ: the t distribution

Lecture 8 gave the standardised sample mean

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n},
$$

which is approximately standard normal by the CLT. But this $Z$ contains the
*population* SD $\sigma$ --- a number nobody knows in a real study. In
practice we substitute the sample SD $S$, and $S$ is itself a random
quantity: it changes from sample to sample, and in a small sample it can
miss $\sigma$ badly. Standardising with an *estimated* spread therefore
produces a statistic that is more variable than $Z$. For independent
observations from a normal population,

$$
T=\frac{\bar X-\mu}{S/\sqrt n}
\sim t_{n-1}.
$$

The **t distribution** is symmetric around zero like the standard normal,
but it has **heavier tails**: extreme standardised values are more likely,
precisely because $S$ sometimes underestimates the true spread. Its
**degrees of freedom** control how much heavier:

- few degrees of freedom mean $S$ is estimated imprecisely, so the tails
  are heavier;
- more degrees of freedom mean more information about spread;
- as the degrees of freedom increase, the t distribution approaches
  $N(0,1)$.

For a one-sample mean, $df=n-1$ --- the same $n-1$ you met in Lecture 7:
once the sample mean is estimated, the $n$ deviations must sum to zero, so
only $n-1$ of them vary freely when spread is estimated.

```{r}
#| label: fig-b09-t-curves
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The t distribution for 4 and 10 degrees of freedom against the standard normal. All three are symmetric around zero; the t curves put more probability in the tails, most visibly at low degrees of freedom."
library(ggplot2)

t_grid <- seq(-4, 4, length.out = 800)
curve_levels <- c("t, df = 4", "t, df = 10", "standard normal")
reference_curves <- rbind(
  data.frame(x = t_grid, density = dt(t_grid, df = 4),
             curve = curve_levels[1]),
  data.frame(x = t_grid, density = dt(t_grid, df = 10),
             curve = curve_levels[2]),
  data.frame(x = t_grid, density = dnorm(t_grid),
             curve = curve_levels[3])
)
reference_curves$curve <- factor(reference_curves$curve,
                                 levels = curve_levels)
curve_colours <- c("t, df = 4" = "#D55E00",
                   "t, df = 10" = "#0072B2",
                   "standard normal" = "#009E73")

ggplot(reference_curves, aes(x, density,
                             colour = curve, linetype = curve)) +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 2.3, y = 0.375, label = "t, df = 4",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  annotate("text", x = 2.3, y = 0.335, label = "t, df = 10",
           colour = "#0072B2", size = 3.4, hjust = 0) +
  annotate("text", x = 2.3, y = 0.295, label = "standard normal",
           colour = "#009E73", size = 3.4, hjust = 0) +
  scale_colour_manual(values = curve_colours, guide = "none") +
  scale_linetype_manual(values = c("solid", "dashed", "dotted"),
                        guide = "none") +
  scale_y_continuous(limits = c(0, 0.41),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Standardised value", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The difference in @fig-b09-t-curves is most visible in the tails. This is
why a 95% t critical value is larger than the normal 1.96 when $df$ is
small --- the reference curve must reach further out to enclose 95% of its
probability.

This resolves the prep prediction: replacing $\sigma$ with $S$ makes the
interval **wider**, because the heavier-tailed t distribution demands a
larger critical value. The extra width is the honest price of estimating
one more quantity from the same data.

::: {.callout-note collapse="true"}
## Check: which is larger, $t_{0.975,11}$ or $t_{0.975,99}$?

$t_{0.975,11}\approx 2.20$ is larger than $t_{0.975,99}\approx 1.98$. With
only 11 degrees of freedom, $S$ is an imprecise estimate of $\sigma$, so
the t distribution has heavier tails and its 97.5% quantile sits further
from zero. At $df=99$ the critical value is already close to the normal
1.96 --- with a large sample, the t and normal answers nearly coincide.
Verify in R: `qt(0.975, df = 11)` and `qt(0.975, df = 99)`.
:::

### The interval recipe

A confidence interval combines the estimate with its uncertainty. The
general form is

$$
\text{estimate}
\ \pm\
\text{critical value}\times\text{standard error}.
$$

The product after the $\pm$ is the **margin of error**. A confidence
interval becomes:

- wider when the confidence level increases (a larger critical value);
- wider when observations are more variable (a larger SE);
- narrower when the sample size increases, all else equal.

For an independent sample with unknown population SD,

$$
\bar X\pm t_{1-\alpha/2,\,n-1}\frac{S}{\sqrt n}
$$

is a $100(1-\alpha)\%$ confidence interval for $\mu$. For a large
independent sample, the interval is also a reasonable approximation when
the population is not normal --- the CLT again. The same recipe applied to
the gap variable $D$ gives an interval for $\delta$.

::: {#exm-b09-hand-interval}
## Worked example: a 95% interval by hand

A random sample of $n=64$ ventures reports a mean score of $\bar x = 80$
with sample SD $s = 16$. The industry benchmark is 75. Construct a 95%
confidence interval for $\delta=\mu-75$, voicing every arithmetic step.

- **Estimate.** The estimated gap is the sample mean minus the benchmark:
  $\widehat\delta = 80 - 75 = 5$.
- **Standard error.** $\operatorname{SE}=s/\sqrt n = 16/\sqrt{64}$. The
  square root of 64 is 8, so $\operatorname{SE}=16/8=2$.
- **Degrees of freedom and critical value.** $df=n-1=63$, and
  $t_{0.975,63}\approx 2.00$ (the exam supplies this value; in R it is
  `qt(0.975, df = 63)`, about 1.998).
- **Margin of error.** Critical value times SE: $2.00\times 2 = 4$.
- **Interval.** Estimate plus and minus the margin:
  $5 - 4 = 1$ and $5 + 4 = 9$, so the 95% interval for $\delta$ is
  $(1,\,9)$.

Every plausible value of the gap is positive: the data are consistent with
a population mean between 1 and 9 points *above* the benchmark. Adding 75
back to both endpoints gives the equivalent interval $(76,\,84)$ for $\mu$
itself --- the gap form and the mean form are the same statement.
:::

### Check conditions before calculating

The formula is justified by the sampling design and distributional
conditions, not by the presence of an R command. Before any interval or
test, walk this checklist:

1. **Random selection.** Were the rows selected randomly from a defined
   frame? (Ours will be: the seed-123 sample of 100 from the 1,441-record
   frame.)
2. **The 10% condition.** Sampling without replacement creates weak
   dependence, but $100/1441\approx 0.07 < 0.10$, so the usual
   independent-sample SE is a reasonable approximation.
3. **Shape and sample size.** Pitch-level equity is right-skewed, but
   $n=100$ makes the t approximation for the *sample mean* far more
   credible than it would be for a very small sample. Still check for data
   errors or observations that dominate the mean.

These conditions support inference to the recorded-pitch frame. They do not
make televised pitches representative of all entrepreneurs --- the caveat
box returns to this.

### The Shark Tank case, with nothing hidden

Now the real question. The recipe runs on the reproducible sample of 100
from Lecture 7, and every object is constructed on screen:

```{r}
#| label: b09-gap-objects
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

benchmark <- 15
pitch_sample$equity_gap_pp <-
  pitch_sample$equity_offered_pct - benchmark

gap_bar <- mean(pitch_sample$equity_gap_pp)   # estimated gap
gap_sd  <- sd(pitch_sample$equity_gap_pp)     # sample SD of the gap
n       <- nrow(pitch_sample)
df      <- n - 1
gap_se  <- gap_sd / sqrt(n)                   # standard error

c(
  sample_mean_equity = mean(pitch_sample$equity_offered_pct),
  sample_mean_gap = gap_bar,
  sample_sd_gap = gap_sd,
  standard_error_gap = gap_se,
  degrees_of_freedom = df
)
```

The estimated population mean gap is $-2.24$ percentage points: the sample
mean equity offer of 12.76 sits 2.24 points below the 15-point benchmark.
Note that subtracting the constant 15 changes the mean but not the spread:
`gap_sd` equals the sample SD of equity itself, 7.71.

```{r}
#| label: b09-gap-ci
critical_t <- qt(0.975, df = df)
gap_ci <- gap_bar + c(-1, 1) * critical_t * gap_se
mean_ci <- benchmark + gap_ci

c(critical_t = critical_t)
c(gap_lower = gap_ci[1], gap_upper = gap_ci[2])
c(mean_lower = mean_ci[1], mean_upper = mean_ci[2])
```

The 95% interval for the mean gap is approximately $(-3.77,\,-0.71)$
percentage points. Adding 15 to both endpoints gives the equivalent
interval $(11.23,\,14.29)$ for mean initial equity. Same recipe as the
hand-worked example --- only the numbers changed.

### A hypothesis test starts with a null model

The interval displays every plausible value. A **test** answers one
question about one value. For a two-sided test,

$$
H_0:\delta=0
\qquad\text{versus}\qquad
H_1:\delta\ne0.
$$

The null says there is no population mean difference from the benchmark.
The alternative allows a difference in either direction. Choose the
two-sided alternative from the research question *before* inspecting the
sample estimate. (Directional, one-sided alternatives exist; Lecture 10
treats them and explains when they are --- and are not --- legitimate.)

Standardise the estimated gap using its SE:

$$
T=
\frac{\widehat\delta-0}
{\operatorname{SE}(\widehat\delta)}
=
\frac{\bar D}{S_D/\sqrt n}.
$$

By hand first, from the numbers already on screen: the estimate is
$-2.24$, its SE is $0.77$, so

$$
t=\frac{-2.24}{0.77}\approx -2.91.
$$

The sign records direction --- below the benchmark; the absolute value
records distance from zero in standard-error units --- almost three SEs.
Now the same calculation in R, with the p-value:

```{r}
#| label: b09-gap-test
t_value <- gap_bar / gap_se
p_value <- 2 * pt(-abs(t_value), df = df)

c(
  estimate = gap_bar,
  standard_error = gap_se,
  t_statistic = t_value,
  degrees_of_freedom = df,
  two_sided_p_value = p_value
)
```

### A two-sided p-value uses both tails

The **p-value** is the probability, assuming $H_0$, of obtaining a test
statistic at least as far from zero as the observed statistic in either
direction:

$$
p=P\!\left(|T_{99}|\ge|t_{obs}|\mid H_0\right).
$$

```{r}
#| label: fig-b09-two-sided-p
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The t distribution with 99 degrees of freedom under the null hypothesis. Dashed lines mark statistics as far from zero as the observed t = −2.91; the two orange tail areas together are the two-sided p-value, about 0.005."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
pitch_sample$equity_gap_pp <- pitch_sample$equity_offered_pct - 15
gap_bar <- mean(pitch_sample$equity_gap_pp)
gap_se <- sd(pitch_sample$equity_gap_pp) / sqrt(nrow(pitch_sample))
df <- nrow(pitch_sample) - 1
t_value <- gap_bar / gap_se

null_curve <- data.frame(x = seq(-4.5, 4.5, length.out = 1200))
null_curve$density <- dt(null_curve$x, df = df)
left_tail <- subset(null_curve, x <= -abs(t_value))
right_tail <- subset(null_curve, x >= abs(t_value))

ggplot(null_curve, aes(x, density)) +
  geom_area(data = left_tail, fill = "#D55E00", alpha = 0.85) +
  geom_area(data = right_tail, fill = "#D55E00", alpha = 0.85) +
  geom_vline(xintercept = c(-abs(t_value), abs(t_value)),
             linetype = "dashed", colour = "#D55E00", linewidth = 0.6) +
  geom_line(colour = "#17324D", linewidth = 0.9) +
  annotate("text", x = -abs(t_value), y = 0.33,
           label = "t == -2.91", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 1.15) +
  annotate("text", x = abs(t_value), y = 0.33,
           label = "t == +2.91", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = -0.15) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "t statistic under the null", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-b09-two-sided-p with the five-step sequence. (1) The curve is the
distribution of the t statistic **if the null were true** --- not data, a
null model. (2) The horizontal axis is the possible t statistic; the
vertical axis is density, so probability is area. (3) The dashed lines mark
$-2.91$ and $+2.91$: statistics as far from zero as the one observed.
(4) The pattern: the two orange tail areas beyond the dashed lines are tiny
--- together about 0.0045. (5) Takeaway: if the true mean gap were zero, a
sample as extreme as ours would occur in roughly 5 of every 1,000 samples.
That is **not** the probability that $H_0$ is true, and not the size of
the gap.

At $\alpha=0.05$, reject $H_0$ because $p<0.05$. The random sample
provides evidence that mean initial equity in the recorded-pitch frame
differs from 15 percentage points; the negative estimate indicates that it
is lower.

### The confidence interval and test tell the same story

For the same two-sided t procedure:

- reject $H_0:\delta=0$ at level $\alpha$ exactly when the
  $100(1-\alpha)\%$ confidence interval excludes zero;
- do not reject when the interval includes zero.

Here the gap interval $(-3.77,\,-0.71)$ excludes zero, matching the
rejection. The interval adds information the p-value does not: it displays
the direction, plausible magnitude, and precision of the difference. Course
policy, from here to the end of Intermediate: **report the estimate and
interval alongside any test decision.**

::: {.callout-caution collapse="true"}
## Try it: read the decision straight from an interval

A colleague reports a 95% confidence interval for a mean gap:
$(-1.4,\,3.8)$ percentage points. No t statistic, no p-value.

*Step 1 (given).* The matched two-sided test rejects $H_0:\delta=0$ at 5%
exactly when the 95% interval excludes zero.

*Step 2 (you).* Does this interval exclude zero? State the test decision
without computing anything.

*Step 3 (you).* What may you conclude about the population mean gap ---
and what may you *not* conclude?

::: {.callout-note collapse="true"}
## Check your answer

Zero lies inside $(-1.4,\,3.8)$, so the 5% two-sided test does **not**
reject $H_0:\delta=0$. The data are compatible with a zero gap --- and
also with gaps as large as 3.8 points in one direction and 1.4 in the
other. You may *not* conclude that the gap **is** zero: "do not reject" is
not "accept," a distinction Lecture 10 builds on.
:::
:::

### Interpret 95% confidence correctly

The parameter is fixed; interval endpoints vary from sample to sample.
"95%" describes the **procedure**: if the same valid recipe were repeated
on many random samples, about 95% of the resulting intervals would contain
the parameter.

The simulation in @fig-b09-coverage repeats the random sample and interval
calculation 100 times. The fixed target is the mean initial equity among
all 1,441 recorded pitches.

```{r}
#| label: fig-b09-coverage
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 5.2
#| fig-cap: "One hundred repetitions of the same procedure: draw 100 random pitches, compute a 95% t interval for mean initial equity. Green intervals contain the fixed target mean (gold dashed line); orange intervals miss it. In this run, 96 of 100 intervals cover the target."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(209)
B_coverage <- 100
n_coverage <- 100
target_mean <- mean(sharks$equity_offered_pct)
ci_lower <- ci_upper <- ci_centre <- numeric(B_coverage)

for (i in seq_len(B_coverage)) {
  repeated_sample <- sample(
    sharks$equity_offered_pct,
    size = n_coverage,
    replace = FALSE
  )
  repeated_mean <- mean(repeated_sample)
  repeated_se <- sd(repeated_sample) / sqrt(n_coverage)
  repeated_critical <- qt(0.975, df = n_coverage - 1)
  ci_centre[i] <- repeated_mean
  ci_lower[i] <- repeated_mean - repeated_critical * repeated_se
  ci_upper[i] <- repeated_mean + repeated_critical * repeated_se
}

intervals <- data.frame(
  lower = ci_lower, centre = ci_centre, upper = ci_upper,
  covers = ifelse(ci_lower <= target_mean & ci_upper >= target_mean,
                  "contains the target mean", "misses the target mean")
)
intervals <- intervals[order(intervals$centre), ]
intervals$position <- seq_len(B_coverage)

ggplot(intervals, aes(y = position, colour = covers)) +
  geom_segment(aes(x = lower, xend = upper, yend = position),
               linewidth = 0.7) +
  geom_point(aes(x = centre), size = 0.9) +
  geom_vline(xintercept = target_mean, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  annotate("text", x = target_mean, y = B_coverage + 3.5,
           label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = -0.1) +
  scale_colour_manual(values = c("contains the target mean" = "#009E73",
                                 "misses the target mean" = "#D55E00")) +
  scale_y_continuous(breaks = c(1, 20, 40, 60, 80, 100)) +
  labs(x = "Mean initial equity offered (%)",
       y = "Repeated sample (ordered by estimate)",
       colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Each horizontal segment in @fig-b09-coverage is one 95% confidence
interval; its dot is that sample's mean. The gold dashed vertical line is
the fixed target mean, 13.21. In this run (seed 209), **96 of 100** intervals
cross the line. A new run need not give exactly 95, but across many
repetitions the proportion approaches 95% when the procedure's conditions
hold.

One sentence is now forbidden for the rest of your studies:

> "There is a 95% probability that the population mean is in my interval."

The realised interval $(11.23, 14.29)$ either contains the fixed parameter
or it does not --- there is no probability left once the sample is drawn.
The correct replacement:

> "This interval was produced by a procedure that captures the fixed
> population mean in about 95% of repeated valid samples."

::: {.caveat-box}
<span class="label">What significance can and cannot tell you</span>

Three boundaries around everything in this lecture. First, **significance
is not importance**: $p\approx 0.005$ says the gap is measured precisely
enough to distinguish it from zero, not that a 2.24-point gap matters for
any decision --- magnitude and context decide that. Second, **significance
is not causation**: the evidence concerns a mean in an observational frame;
nothing here says *why* equity asks sit below 15 points. Third, **a narrow
interval repairs nothing about selection**: the interval quantifies random
sampling error within the recorded-pitch frame. If the frame itself is a
producer-curated selection of telegenic pitches --- and it is --- then no
interval width, however impressive, extends the conclusion to entrepreneurs
in general. An on-air agreement is still not a completed investment.
:::

### Answering the driving question

Yes --- within the frame. The seed-123 sample gives an estimated gap of
$-2.24$ percentage points with SE $0.77$; the 95% interval for mean equity
is $(11.23, 14.29)$, which excludes 15, and the two-sided test rejects
$H_0:\delta=0$ at 5% ($t=-2.91$, $p\approx 0.005$). The evidence says the
frame's mean initial equity ask sits *below* 15 percent --- plausibly by
between 0.7 and 3.8 points. Whether that gap matters commercially, and what
caused it, are separate questions the test cannot answer.

**Next:** We can now test a benchmark and report the answer as an interval and
a p-value. But every test decision can be wrong in one of two ways ---
rejecting a true null, or missing a real effect --- and the two mistakes
rarely cost the same. Lecture 10 names the two error types, shows the four
levers that make wrong calls rarer (power), settles when a one-sided
alternative is legitimate, and then opens the door this whole course has been
walking toward: the same estimate--SE--t--p--interval row, sitting in a
regression table.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 60 minutes.</span>

### Retrieval (10 minutes)

1. Why does replacing $\sigma$ with the sample SD $S$ require a t rather
   than a normal reference distribution?
2. Why does a one-sample t procedure use $n-1$ degrees of freedom?
3. What are the three pieces of a confidence interval?
4. What do $H_0:\delta=0$ and $H_1:\delta\ne0$ say?
5. Why does a two-sided p-value use two tails?
6. What is the exact decision link between a 95% t interval and a 5%
   two-sided t-test?

::: {.callout-note collapse="true"}
## Check your answers

1. $S$ is itself estimated and varies from sample to sample; the t
   distribution's heavier tails represent that extra uncertainty.
2. Estimating the sample mean imposes one constraint (deviations sum to
   zero) before spread is estimated, leaving $n-1$ freely varying
   deviations.
3. Estimate, critical value, and standard error.
4. The null states that the population contrast is zero; the alternative
   allows a non-zero contrast in either direction.
5. Values equally far below and above zero are both evidence against a
   two-sided null.
6. The test rejects exactly when the interval excludes zero, provided both
   use the same model and SE.
:::

### From worked to faded (10 minutes)

Re-read the hand-worked interval (@exm-b09-hand-interval). Then complete
the same calculation with new numbers: $n=25$ customers, mean satisfaction
$\bar x = 42$, sample SD $s = 10$, benchmark 40, so $\delta=\mu-40$.

*Step 1 (given).* The estimated gap is $42-40=2$, and
$t_{0.975,24}\approx 2.06$.

*Step 2 (you).* Calculate the SE and the margin of error.

*Step 3 (you).* Write the 95% interval for $\delta$ and state, without any
further computation, the 5% two-sided test decision.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SE}=10/\sqrt{25}=2$; margin of error
$2.06\times 2=4.13$. The interval is $2\pm 4.13=(-2.13,\,6.13)$. Zero lies
inside, so the 5% two-sided test does not reject $H_0:\delta=0$: the data
are compatible with a zero gap --- and with gaps up to about 6 points. (For
completeness, $t=2/2=1.00$, $p\approx 0.33$.)
:::

### Practice (10 minutes)

1. A sample has $n=49$, $\bar x=110$, and $s=14$. Define $\delta=\mu-100$.
   Construct a 95% interval for $\delta$ (use $t_{0.975,48}\approx 2.01$).
2. Test $H_0:\delta=0$ against $H_1:\delta\ne0$ at 5%. Report the
   statistic, degrees of freedom, approximate p-value, decision, and a
   contextual conclusion.
3. A coefficient-like estimate is 0.42 percentage points per €100,000,
   with SE 0.18 and $df=120$. Test a two-sided zero null and construct a
   95% interval (use $t_{0.975,120}\approx 1.98$).
4. Explain why a statistically significant mean gap need not be
   practically important or causal.

::: {.callout-note collapse="true"}
## Check your answers

1. The estimated gap is $110-100=10$, its SE is $14/\sqrt{49}=2$, and the
   interval is $10\pm 2.01(2)=(5.98,\,14.02)$.
2. $t=10/2=5.00$, $df=48$, two-sided p-value below 0.001. Reject $H_0$:
   the data provide strong evidence that the population mean exceeds 100
   --- plausibly by 6 to 14 points.
3. $t=0.42/0.18\approx 2.33$, two-sided p-value about 0.021. The 95%
   interval is $0.42\pm1.98(0.18)=(0.064,\,0.776)$ percentage points per
   €100,000. Reject the zero null at 5%.
4. Statistical significance concerns precision relative to zero. Practical
   importance depends on the magnitude and the decision context, and
   causality requires an appropriate design and assumptions --- neither is
   delivered by a small p-value.
:::

### Rewrite the forbidden sentence (5 minutes)

Rewrite this claim correctly: "There is a 95% chance that the population
mean is in my interval $(11.23, 14.29)$."

::: {.callout-note collapse="true"}
## Check a model answer

"The interval $(11.23, 14.29)$ was produced by a procedure that captures
the fixed population mean in about 95% of repeated valid samples." The
parameter is fixed; it is the interval endpoints that vary across samples,
as in @fig-b09-coverage.
:::

### Spaced review: one item each from Lectures 8 and 4 (10 minutes)

1. *(Lecture 8.)* A planned survey expects a sample SD around 24 with
   $n=64$. Estimate the SE of the sample mean, and state what happens to
   it if $n$ grows to 256.
2. *(Lecture 4.)* From the 2×2 counts table,
   $P(D\mid L)=505/734\approx 0.688$ while
   $P(L\mid D)=505/882\approx 0.573$. Why are these two numbers different
   even though both involve the same 505 records?

::: {.callout-note collapse="true"}
## Check your answers

1. $\operatorname{SE}=24/\sqrt{64}=3$. Quadrupling the sample size halves
   the SE to 1.5 --- the square-root rule.
2. They condition on different reference classes: the first divides the
   505 joint records by the 734 late-season records, the second divides
   the same 505 by the 882 deal records. Changing the event after the bar
   changes the denominator.
:::

### Worked exam interpretation (10 minutes)

**Question (Mock Exam Q4 pattern).** From a frame of 1,000 accelerator
applicants, a simple random sample of 64 was selected. `prep_hours` is
hours spent preparing the application; $\mu$ is the frame mean. The sample
has $\bar x=18.4$ and $s=8.0$. R produced:

```text
One Sample t-test

data: prep_hours
t = 2.400, df = 63, p-value = 0.0193
alternative hypothesis: true mean is not equal to 16
95 percent confidence interval:
 16.40 20.40
sample estimate:
mean of x = 18.40
```

Verify the SE and the t statistic, explain the degrees of freedom,
interpret the p-value and the interval, and give the 5% decision.

::: {.callout-note collapse="true"}
## Check a model answer

$\operatorname{SE}=8/\sqrt{64}=1$ hour, so
$t=(18.4-16)/1=2.4$, matching the output. $df=64-1=63$ because the
deviations from the estimated sample mean must sum to zero, leaving 63
free. If the frame mean were 16 hours, a sample mean at least 2.4 SEs from
16 in either direction would occur in about 1.9% of random samples; since
$0.0193<0.05$, reject $H_0:\mu=16$. The interval $(16.40, 20.40)$ gives
the plausible range for $\mu$ --- it excludes 16, agreeing with the test
--- and was produced by a procedure with about 95% coverage across
repeated samples. The claim concerns the 1,000-applicant frame, not
applicants in general.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Pick one key
metric with a published industry benchmark (conversion rate, average order
value, churn), frame the comparison as a gap $\delta = \mu - \text{benchmark}$,
and write two sentences: first say what sample you could realistically
collect and what a 95% interval for $\delta$ excluding zero would let you
claim, then say what decision would change if the interval turned out to
include zero.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "The sector churn benchmark is 25% per term; from my roughly 120 subscribers I can estimate $\delta = \mu - 25$, and a 95% interval sitting entirely below zero would let me claim genuinely stickier customers in my pitch. If the interval included zero, I would stop pitching retention as my edge and lead with margin instead."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to check whether your
gap is defined on a population you can actually sample.
::::

Continue to Lecture 10: Testing decisions, errors, power, and the road to
regression.

Formula sheet: Lecture 9 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 15 and 16.

:::
