# Lecture 9: The t distribution and confidence intervals {#b09}

```{r}
#| label: b09-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 9, you can</span>

- explain why estimating the population SD leads to a t distribution with
  $n-1$ degrees of freedom, and use its critical values
- construct a t confidence interval for a population mean
- express comparison with a benchmark as a parameter equal to zero
- state a two-sided null and alternative hypothesis
- calculate a t statistic and say what it measures
- state correctly what "95% confidence" is a property of.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- build a 95% interval by hand from the mean, the standard deviation, the sample size, and a supplied t value
- write $H_0$ and $H_1$ for a gap from a benchmark
- calculate a t statistic and say what it measures
- say what "95%" refers to.

**Enough to recognise:**

- the t curves for different degrees of freedom
- `qt()`.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Suppose an advisor tells founders that pitches typically offer investors about 15 percent equity. Our data can check such a benchmark, if we take the variation between samples into account.</p>

Is the mean equity offered (`equity_offered_pct`) across all recorded Shark
Tank pitches really different from 15 percent?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=NInNhFm046w">Confidence
Intervals</a>,
<a href="https://www.youtube.com/watch?v=MzvRQFYUEFU">Confidence Intervals
When the Variance Is Unknown</a>, and
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to
Hypothesis Testing</a>. Focus on repeated-sampling coverage, the t reference
distribution, and the role of the null value. All links are also listed in
the <a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (5 minutes).** Let $X$ be the equity offered in a pitch
(`equity_offered_pct`, in percentage points) and define the **equity gap**

$$
G = X - 15.
$$

Write two-sided hypotheses for the question: "Does the population mean of
the equity offered differ from 15 percentage points?"

::: {.callout-note collapse="true"}
## Check your preparation

Let $\delta=E[G]=\mu_X-15$. Then

$$
H_0:\delta=0
\qquad\text{versus}\qquad
H_1:\delta\ne0.
$$

This is equivalent to testing $H_0:\mu_X=15$ against $H_1:\mu_X\ne15$.
Writing the gap as $\delta$ turns the question into "is the difference
zero?". Regression coefficients are tested in the same way later.
:::

**Predict (5 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: in a real sample, the unknown population SD
$\sigma$ must be replaced by the sample SD $S$. Should that substitution
make a 95% interval **wider**, **narrower**, or leave it **unchanged**?
Commit to an answer. You find the answer in the lecture material.

If intervals, standardisation, or degrees of freedom feel rusty, review
[Mathematics Refresher: intervals, standardisation, and degrees of
freedom](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

Lecture 8 gave a sample mean and its standard error. This section turns
those two numbers into a range of plausible values for the parameter, the
confidence interval. It ends with the t statistic, which measures how far an
estimate lies from a benchmark. Lecture 10 uses it to reach a decision.

<p class="concept-video"><strong>Video explanations (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/confidence-intervals/">20.5 Confidence Intervals</a>,
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/confidence-intervals-for-the-mean-when-the-variance-is-unknown/">20.7 Confidence Intervals with Unknown Variance</a>, and
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to Hypothesis Testing</a>.</p>

### Start with a founder's question

Suppose an advisor tells you that founders on the show offer about 15% of
their company. You have looked at 100 pitches, chosen by chance, and their
average equity offered (`equity_offered_pct`) is 12.76%. Is the advisor
wrong? Or is 12.76 simply what 100 pitches can give by chance when the true
average is 15? R answers with one command:

```{r}
#| label: b09-opening
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
t.test(pitch_sample$equity_offered_pct, mu = 15)
```

R prints the mean of the sample, 12.76, a t value of $-2.91$ with 99 degrees
of freedom, a p-value of 0.0045, and an interval from 11.23 to 14.29. An AI
assistant would hand you the same output with a confident sentence about it.
This lecture and the next give you what you need to read the output yourself
and to check that sentence: where the interval comes from, why it uses a t
value and 99 degrees of freedom, and what the p-value does and does not say.
This lecture builds the interval and the t value. Lecture 10 adds the
p-value.

### From a benchmark to a zero contrast

The driving question compares a population mean to the benchmark 15. As in
the preparation exercise, subtract the benchmark and study the **gap**

$$
G = X - 15,
\qquad
\delta = E[G] = \mu_X - 15.
$$

"Is the mean equity offered different from 15?" becomes "is $\delta$
different from zero?" A difference of this kind is called a **contrast**.
Every benchmark question can be rewritten as a question about a contrast
being zero. Lecture 10 tests a gap between proportions and a regression
slope in this way, and names the difference between two groups as a further
case.

### The price of estimating σ: the t distribution

Lecture 8 gave the standardised sample mean

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n},
$$

which is approximately standard normal by the central limit theorem (CLT).
But this $Z$ contains the population SD $\sigma$, and in a real study
nobody knows $\sigma$. In practice we replace it by the sample SD $S$. But
$S$ is itself random. It changes from sample to sample, and in a small
sample it can be far from $\sigma$. A mean that is standardised with an
estimated spread therefore varies more than $Z$. For independent
observations from a normal population,

$$
T=\frac{\bar X-\mu}{S/\sqrt n}
\sim t_{n-1}.
$$

::: {.definition-box}
<span class="label">Definition</span>

**t distribution and degrees of freedom.** The t distribution describes a
sample mean that is standardised with the sample SD $S$ in place of
$\sigma$. It is symmetric around zero and has heavier tails than the
standard normal. Its shape depends on one number, the **degrees of freedom**
($df$). For one sample mean, $df=n-1$.
:::

Heavier tails mean that values far from zero are more likely than under the
standard normal. The reason is that $S$ sometimes underestimates the true
spread. The degrees of freedom say how much heavier the tails are:

- few degrees of freedom mean that $S$ is an imprecise estimate of
  $\sigma$, so the tails are heavier
- more degrees of freedom mean more information about spread
- as the degrees of freedom increase, the t distribution approaches the
  standard normal $N(0,1)$.

For one sample mean, $df=n-1$. This is the same $n-1$ as in Lecture 7. Once
the sample mean is estimated, the $n$ deviations must sum to zero, so only
$n-1$ of them vary freely when spread is estimated.

```{r}
#| label: fig-b09-t-curves
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The t distribution for 4 and 10 degrees of freedom against the standard normal. All three are symmetric around zero. The t curves put more probability in the tails, most visibly at low degrees of freedom."
library(ggplot2)

t_grid <- seq(-4, 4, length.out = 800)
curve_levels <- c("t, df = 4", "t, df = 10", "standard normal")
reference_curves <- rbind(
  data.frame(x = t_grid, density = dt(t_grid, df = 4),
             curve = curve_levels[1]),
  data.frame(x = t_grid, density = dt(t_grid, df = 10),
             curve = curve_levels[2]),
  data.frame(x = t_grid, density = dnorm(t_grid),
             curve = curve_levels[3])
)
reference_curves$curve <- factor(reference_curves$curve,
                                 levels = curve_levels)
curve_colours <- c("t, df = 4" = "#D55E00",
                   "t, df = 10" = "#0072B2",
                   "standard normal" = "#009E73")

ggplot(reference_curves, aes(x, density,
                             colour = curve, linetype = curve)) +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 2.3, y = 0.375, label = "t, df = 4",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  annotate("text", x = 2.3, y = 0.335, label = "t, df = 10",
           colour = "#0072B2", size = 3.4, hjust = 0) +
  annotate("text", x = 2.3, y = 0.295, label = "standard normal",
           colour = "#009E73", size = 3.4, hjust = 0) +
  scale_colour_manual(values = curve_colours, guide = "none") +
  scale_linetype_manual(values = c("solid", "dashed", "dotted"),
                        guide = "none") +
  scale_y_continuous(limits = c(0, 0.41),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Standardised value", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The difference in @fig-b09-t-curves is most visible in the tails. The
central 95% of the standard normal lies between $-1.96$ and $1.96$
(Lecture 6). A t curve must reach further out to enclose its central 95%.
The point where the central 95% ends is the 0.975 quantile of the t
distribution. It is written $t_{0.975,\,df}$ and computed in R with
`qt(0.975, df)`. This number is called the 95% **critical value**, or the
multiplier. For small $df$ it is clearly larger than 1.96.

::: {.callout-note collapse="true"}
## Check: which is larger, $t_{0.975,11}$ or $t_{0.975,99}$?

$t_{0.975,11}\approx 2.20$ is larger than $t_{0.975,99}\approx 1.98$. With
only 11 degrees of freedom, $S$ is an imprecise estimate of $\sigma$, so
the t distribution has heavier tails and its 97.5% quantile sits further
from zero. At $df=99$ the critical value is already close to the normal
1.96. With a large sample, the t and normal answers are almost the same.
Verify in R: `qt(0.975, df = 11)` and `qt(0.975, df = 99)`.
:::

### The interval recipe

A sample can be used in three ways. A point estimate gives one number for
the parameter, such as the sample mean 12.76. An interval estimate gives a
range of plausible values. A test answers a yes-or-no question about one
value of the parameter. Lectures 7 and 8 gave point estimates. This lecture
builds the interval estimate and prepares the test.

A confidence interval combines the estimate with its uncertainty. The
general form is

$$
\text{estimate}
\ \pm\
\text{critical value}\times\text{standard error}.
$$

::: {.definition-box}
<span class="label">Definition</span>

**Confidence interval.** A range of plausible values for a parameter,
calculated from the sample as estimate $\pm$ critical value $\times$
standard error. A 95% confidence interval is built by a method that captures
the parameter in about 95% of random samples.
:::

::: {.definition-box}
<span class="label">Definition</span>

**Margin of error.** The part after the $\pm$ sign: critical value $\times$
standard error. It is half the width of the confidence interval.
:::

A confidence interval becomes:

- wider when the confidence level increases (a larger critical value)
- wider when observations are more variable (a larger SE)
- narrower when the sample size increases, all else equal.

For an independent sample from a normal population with unknown SD, or
approximately for large $n$,

$$
\bar X\pm t_{1-\alpha/2,\,n-1}\frac{S}{\sqrt n}
$$

is a $100(1-\alpha)\%$ confidence interval for $\mu$. Here $\alpha$ is one
minus the confidence level. A 95% interval has $\alpha=0.05$, and the
multiplier $t_{0.975,\,n-1}$ leaves $\alpha/2=0.025$ in each tail. The words
"approximately for large $n$" refer to the CLT: for a large independent
sample the interval is a reasonable approximation even when the population
is not normal. The same recipe applied to the gap variable $G$ gives an
interval for $\delta$.

The margin of error is the precision of the estimate at the stated
confidence level. A smaller margin means a more precise estimate.

This also answers the prediction question from the preparation. Replacing
$\sigma$ with $S$ makes the interval wider on average, because the t
multiplier is larger than 1.96. The extra width is the price of estimating
one more quantity from the same data.

In the exam, the multiplier you need is printed in the question or on the
formula sheet. For a value of the degrees of freedom that is not on the
sheet, use the closest one listed.

::: {#exm-b09-hand-interval}
## Worked example: a 95% interval by hand

This is an invented example. A random sample of $n=64$ ventures reports a
mean customer satisfaction score of $\bar x = 80$ points, with sample SD
$s = 16$ points. The industry benchmark is 75 points. Construct a 95%
confidence interval for $\delta=\mu-75$, step by step.

- **Estimate.** The estimated gap is the sample mean minus the benchmark:
  $\widehat\delta = 80 - 75 = 5$.
- **Standard error.** $\widehat{\operatorname{SE}}=s/\sqrt n = 16/\sqrt{64}$. The
  square root of 64 is 8, so $\widehat{\operatorname{SE}}=16/8=2$.
- **Degrees of freedom and critical value.** $df=n-1=63$, and
  $t_{0.975,63}\approx 2.00$. In R it is `qt(0.975, df = 63)`, about 1.998.
  The closest value on the formula sheet is the one for $df=60$, which is
  2.000.
- **Margin of error.** Critical value times SE: $2.00\times 2 = 4$.
- **Interval.** Estimate plus and minus the margin:
  $5 - 4 = 1$ and $5 + 4 = 9$, so the 95% interval for $\delta$ is
  $(1,\,9)$.

Every plausible value of the gap is positive. The data are consistent with
a population mean between 1 and 9 points above the benchmark. Adding 75
back to both endpoints gives the equivalent interval $(76,\,84)$ for $\mu$
itself. The gap form and the mean form are the same statement.
:::

### Check conditions before calculating

The formula is valid only when conditions on the sampling design and on the
shape of the data hold. R does not check them for you. Before any interval
or test, go through this checklist:

1. **Random selection.** Were the rows selected randomly from a defined
   frame? In the Shark Tank case below they were: the sample of 100 is drawn
   at random (seed 123) from the 1,441 recorded pitches.
2. **The 10% condition.** Sampling without replacement creates weak
   dependence, but $100/1441\approx 0.07 < 0.10$, so the usual
   independent-sample SE is a reasonable approximation.
3. **Shape and sample size.** The equity offered (`equity_offered_pct`) is
   right-skewed. With $n=100$ the t approximation for the sample mean is
   still reasonable. For a very small sample it would not be. Also check for
   data errors and for single observations that dominate the mean.

These conditions support conclusions about the recorded-pitch frame. They do
not make televised pitches representative of all entrepreneurs. The caveat
box at the end of this section returns to this point.

::: {.callout-caution collapse="true"}
## Try it: is the average order above break-even?

Try this with your neighbour first. This is an invented example. A venture
needs an average order of 20 euros to break even. A random sample of $n=36$
orders has a mean of 21.50 euros and a standard deviation of 6 euros.
Supplied value: $t_{0.975,\,35}=2.030$.

*Step 1 (given).* The estimated standard error is
$\widehat{\operatorname{SE}}=6/\sqrt{36}=1$ euro.

*Step 2 (you).* Calculate the margin of error and the 95% interval.

*Step 3 (you).* Is 20 euros a plausible value for the mean order? What would
you tell the founder?

::: {.callout-note collapse="true"}
## Check your answer

The margin is $2.030\times1=2.03$ euros, so the interval runs from 19.47 to
23.53 euros. The break-even value of 20 euros lies inside the interval. The
sample mean is above break-even, but the data cannot rule out that the true
mean is below it. Advice: collect more orders before any decision that
depends on being above break-even. With four times as many orders the margin
would be about half as wide.
:::
:::

### The Shark Tank case, step by step

Now apply the recipe to the Shark Tank data. The code draws the same random
sample of 100 pitches as in Lecture 7 and builds every object step by step:

```{r}
#| label: b09-gap-objects
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

benchmark <- 15
pitch_sample$equity_gap_pp <-
  pitch_sample$equity_offered_pct - benchmark

gap_bar <- mean(pitch_sample$equity_gap_pp)   # estimated gap
gap_sd  <- sd(pitch_sample$equity_gap_pp)     # sample SD of the gap
n       <- nrow(pitch_sample)
df      <- n - 1
gap_se  <- gap_sd / sqrt(n)                   # standard error

c(
  sample_mean_equity = mean(pitch_sample$equity_offered_pct),
  sample_mean_gap = gap_bar,
  sample_sd_gap = gap_sd,
  standard_error_gap = gap_se,
  degrees_of_freedom = df
)
```

The estimated population mean gap is $-2.24$ percentage points. The sample
mean of the equity offered, 12.76, lies 2.24 points below the benchmark of
15 points. That is about one seventh of the benchmark. Subtracting the
constant 15 changes the mean but not the spread, as the shift rule of
[Lecture 3](b03-discrete-distributions.qmd#b03-shift-rescale) says. So
`gap_sd` equals the sample SD of the equity offered itself, 7.71 points. The
standard error is $7.71/\sqrt{100}=0.77$ points.

```{r}
#| label: b09-gap-ci
critical_t <- qt(0.975, df = df)
gap_ci <- gap_bar + c(-1, 1) * critical_t * gap_se
mean_ci <- benchmark + gap_ci

c(critical_t = critical_t)
c(gap_lower = gap_ci[1], gap_upper = gap_ci[2])
c(mean_lower = mean_ci[1], mean_upper = mean_ci[2])
```

The critical value is $t_{0.975,\,99}=1.984$, so the margin of error is
$1.984\times0.77=1.53$ percentage points. The 95% interval for the mean gap
is approximately $(-3.77,\,-0.71)$ percentage points. Adding 15 to both
endpoints gives the equivalent interval $(11.23,\,14.29)$ for the mean
equity offered. The recipe is the same as in the worked example. Only the
numbers differ.

### Interpret 95% confidence correctly

The parameter is fixed. Interval endpoints vary from sample to sample.
"95%" describes the **procedure**: if the same valid recipe were repeated
on many random samples, about 95% of the resulting intervals would contain
the parameter.

The simulation in @fig-b09-coverage repeats the random sample and the
interval calculation 100 times. The fixed target is the mean equity offered
(`equity_offered_pct`) among all 1,441 recorded pitches, 13.21 points.

```{r}
#| label: fig-b09-coverage
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 5.2
#| fig-cap: "One hundred repetitions of the same procedure: draw 100 random pitches, compute a 95% t interval for the mean equity offered. Green intervals contain the fixed target mean (gold dashed line). Orange intervals miss it. In this run, 96 of 100 intervals cover the target."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(209)
B_coverage <- 100
n_coverage <- 100
target_mean <- mean(sharks$equity_offered_pct)
ci_lower <- ci_upper <- ci_centre <- numeric(B_coverage)

for (i in seq_len(B_coverage)) {
  repeated_sample <- sample(
    sharks$equity_offered_pct,
    size = n_coverage,
    replace = FALSE
  )
  repeated_mean <- mean(repeated_sample)
  repeated_se <- sd(repeated_sample) / sqrt(n_coverage)
  repeated_critical <- qt(0.975, df = n_coverage - 1)
  ci_centre[i] <- repeated_mean
  ci_lower[i] <- repeated_mean - repeated_critical * repeated_se
  ci_upper[i] <- repeated_mean + repeated_critical * repeated_se
}

intervals <- data.frame(
  lower = ci_lower, centre = ci_centre, upper = ci_upper,
  covers = ifelse(ci_lower <= target_mean & ci_upper >= target_mean,
                  "contains the target mean", "misses the target mean")
)
intervals <- intervals[order(intervals$centre), ]
intervals$position <- seq_len(B_coverage)

ggplot(intervals, aes(y = position, colour = covers)) +
  geom_segment(aes(x = lower, xend = upper, yend = position),
               linewidth = 0.7) +
  geom_point(aes(x = centre), size = 0.9) +
  geom_vline(xintercept = target_mean, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  annotate("text", x = target_mean, y = B_coverage + 3.5,
           label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = -0.1) +
  scale_colour_manual(values = c("contains the target mean" = "#009E73",
                                 "misses the target mean" = "#D55E00")) +
  scale_y_continuous(breaks = c(1, 20, 40, 60, 80, 100)) +
  labs(x = "Mean equity offered (%)",
       y = "Repeated sample (ordered by estimate)",
       colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Each horizontal segment in @fig-b09-coverage is one 95% confidence
interval. Its dot is that sample's mean. The gold dashed vertical line is
the fixed target mean, 13.21. In this run (seed 209), **96 of 100** intervals
cross the line. A new run need not give exactly 95, but across many
repetitions the proportion approaches 95% when the procedure's conditions
hold.

The following sentence may look natural, but it is wrong:

> "There is a 95% probability that the population mean is in my interval."

Before the sample is drawn, the endpoints of the interval are random, and
the 95% applies to them. After the sample is drawn, the interval
$(11.23, 14.29)$ consists of two fixed numbers. It either contains the fixed
parameter or it does not, and no probability is left. The correct sentence
is:

> "This interval was produced by a procedure that captures the fixed
> population mean in about 95% of repeated valid samples."

### From interval to test: the hypotheses and the t statistic

The interval shows every plausible value. A **test** answers one question
about one value: is the gap zero? A test starts from two hypotheses:

$$
H_0:\delta=0
\qquad\text{versus}\qquad
H_1:\delta\ne0.
$$

::: {.definition-box}
<span class="label">Definition</span>

**Null hypothesis and alternative hypothesis.** The null hypothesis $H_0$
states that the parameter equals the benchmark value, here $\delta=0$. The
alternative hypothesis $H_1$ states that it does not. A two-sided
alternative, $\delta\ne0$, allows a difference in either direction.
:::

Here the null hypothesis says that the population mean of the equity offered
does not differ from the benchmark. State both hypotheses from the research
question before you look at the sample estimate. Alternatives that allow
only one direction (one-sided alternatives) also exist. They are optional
reading in Lecture 10 and are not examined.

The **t statistic** standardises the estimated gap with its SE. Its general
form is the estimate minus the null value, divided by the standard error:

$$
T=
\frac{\widehat\delta-0}
{\widehat{\operatorname{SE}}(\widehat\delta)}
=
\frac{\bar G}{S_G/\sqrt n}.
$$

First by hand, with the numbers calculated above. The estimate is $-2.24$
and its SE is $0.77$, so

$$
t=\frac{-2.24}{0.77}\approx -2.91.
$$

The sign gives the direction: the estimate is below the benchmark. The
absolute value gives the distance from zero in standard-error units, here
almost three SEs. Now the same calculation in R:

```{r}
#| label: b09-gap-test
t_value <- gap_bar / gap_se

c(
  estimate = gap_bar,
  standard_error = gap_se,
  t_statistic = t_value,
  degrees_of_freedom = df
)
```

A t statistic of $-2.91$ says that the estimate lies almost three standard
errors below zero. Lecture 10 measures how surprising such a value would be
if the null hypothesis were true. That measure is called the p-value.

::: {.caveat-box}
<span class="label">What an interval can and cannot tell you</span>

- Precision is not importance. An interval that excludes 15 says the gap is
  measured well, not that 2.24 points matter for a decision.
- An interval is not a claim about cause. Nothing here says why offers sit
  below 15 points.
- A narrow interval repairs nothing about selection. It describes the
  recorded pitches, not founders in general.

**What you can do with it:** read both ends of the interval as possible
truths, and ask whether your decision would be the same at each end.
:::

### Answering the driving question

For the recorded-pitch frame, the evidence points to yes. The random sample
of 100 pitches gives an estimated gap of $-2.24$ percentage points with SE
$0.77$. The 95% interval for the mean equity offered is $(11.23, 14.29)$,
which excludes 15. The t statistic of $-2.91$ puts the estimate almost three
standard errors below the benchmark. Lecture 10 turns that distance into a
p-value and a formal decision. Whether the gap matters commercially, and
what caused it, are separate questions that no interval can answer.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

From 25 orders, the average order value is 9.60 euros with a standard deviation of 2 euros. You need 9 euros per order to break even on a kitchen lease. The 95% interval is 9.60 ± 2.064 × 0.40, from 8.77 to 10.43 euros. The break-even value of 9 euros lies inside the interval. The decision: do not sign yet. With 100 orders the standard error falls to 0.20. If the mean stays at 9.60, the interval then runs from 9.20 to 10.00 and excludes 9. You collect orders for three more weeks and decide then.
:::

**Next:** We can now report an estimate with an interval, and we have
measured how far the estimate lies from a benchmark in standard-error units.
Lecture 10 turns that distance into a p-value and a decision. It names the two
ways a decision can be wrong and shows what makes wrong decisions rarer. It
ends with a first look at a regression table, which reports the same five
items: estimate, SE, t, p-value, and interval.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 70 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Why does replacing $\sigma$ with the sample SD $S$ require a t rather
   than a normal reference distribution?
2. Why does a one-sample t procedure use $n-1$ degrees of freedom?
3. What are the three pieces of a confidence interval?
4. What do $H_0:\delta=0$ and $H_1:\delta\ne0$ say?
5. What does a t statistic of $-2.91$ tell you?
6. What is "95%" a property of: the interval you computed or the procedure?

::: {.callout-note collapse="true"}
## Check your answers

1. $S$ is itself estimated and varies from sample to sample. The t
   distribution's heavier tails represent that extra uncertainty.
2. Estimating the sample mean imposes one constraint (deviations sum to
   zero) before spread is estimated, leaving $n-1$ freely varying
   deviations.
3. Estimate, critical value, and standard error.
4. The null hypothesis states that the population gap $\delta$ is zero. The
   alternative allows a non-zero gap in either direction.
5. The estimate lies 2.91 standard errors below the null value of zero.
6. The procedure: about 95% of intervals built this way from repeated valid
   samples contain the fixed parameter.
:::

### Finish a half-worked example (10 minutes)

Re-read the hand-worked interval (@exm-b09-hand-interval). Then complete
the same calculation with new invented numbers: a random sample of $n=25$
customers, mean satisfaction $\bar x = 42$ points, sample SD $s = 10$
points, benchmark 40 points, so $\delta=\mu-40$.

*Step 1 (given).* The estimated gap is $42-40=2$, and
$t_{0.975,24}=2.064$.

*Step 2 (you).* Calculate the SE and the margin of error.

*Step 3 (you).* Write the 95% interval for $\delta$ and say whether a gap of zero
is among the plausible values.

::: {.callout-note collapse="true"}
## Check your answer

$\widehat{\operatorname{SE}}=10/\sqrt{25}=2$. The margin of error is
$2.064\times 2=4.13$. The interval is $2\pm 4.13=(-2.13,\,6.13)$. Zero lies
inside, so a zero gap is among the plausible values, and so are gaps up to
about 6 points. The t statistic is $2/2=1.00$. The estimate lies one
standard error from zero.
:::

### Practice (10 minutes)

The numbers in tasks 1 to 3 are invented.

1. A random sample has $n=49$, $\bar x=110$, and $s=14$. Define
   $\delta=\mu-100$. Construct a 95% interval for $\delta$ (use
   $t_{0.975,48}\approx 2.01$).
2. For the sample of task 1, state $H_0$ and $H_1$ for a two-sided test of
   a zero gap. Calculate the t statistic and its degrees of freedom.
3. An estimate is 0.42 percentage points per €100,000, with SE 0.18 and
   $df=120$. Calculate the t statistic and construct a 95% interval (use
   $t_{0.975,120}\approx 1.98$).
4. Explain why an interval that excludes zero does not show that a gap is
   important in practice, or that it has a particular cause.

::: {.callout-note collapse="true"}
## Check your answers

1. The estimated gap is $110-100=10$, its SE is $14/\sqrt{49}=2$, and the
   interval is $10\pm 2.01(2)=(5.98,\,14.02)$.
2. $H_0:\delta=0$ against $H_1:\delta\ne0$. The t statistic is
   $t=10/2=5.00$ with $df=48$, so the estimate lies five standard errors
   above zero. The interval from task 1 lies entirely above zero. Plausible
   gaps run from about 6 to 14 points.
3. $t=0.42/0.18\approx 2.33$. The 95% interval is
   $0.42\pm1.98(0.18)=(0.064,\,0.776)$ percentage points per €100,000,
   which excludes zero.
4. An interval that excludes zero says that the estimate is precise enough
   to tell the gap from zero. Whether the gap is important depends on its
   size and on the decision. A causal claim needs a suitable study design.
   A narrow interval gives neither.
:::

### Rewrite the wrong sentence (5 minutes)

Rewrite this claim correctly: "There is a 95% chance that the population
mean is in my interval $(11.23, 14.29)$."

::: {.callout-note collapse="true"}
## Check a model answer

"The interval $(11.23, 14.29)$ was produced by a procedure that captures
the fixed population mean in about 95% of repeated valid samples." The
parameter is fixed. It is the interval endpoints that vary across samples,
as in @fig-b09-coverage.
:::

### Review: one question each from Lectures 8 and 4 (10 minutes)

1. *(Lecture 8.)* A planned survey expects a sample SD around 24 with
   $n=64$. Estimate the SE of the sample mean, and state what happens to
   it if $n$ grows to 256.
2. *(Lecture 4.)* From the 2×2 counts table,
   $P(D\mid L)=505/734\approx 0.688$ while
   $P(L\mid D)=505/882\approx 0.573$. Why are these two numbers different
   even though both involve the same 505 records?

::: {.callout-note collapse="true"}
## Check your answers

1. $\widehat{\operatorname{SE}}=24/\sqrt{64}=3$. Four times the sample size
   halves the SE to 1.5. This is the square-root rule.
2. They condition on different events. The first divides the 505 joint
   records by the 734 late-season records. The second divides the same 505
   by the 882 deal records. Changing the event after the bar changes the
   denominator.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "What does the 95% interval from 11.23 to 14.29 for the mean equity offered mean?" It answered:

> "It means that 95% of all pitches offer between 11.23% and 14.29% of the company."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

The interval is about the mean, not about single pitches. In the Shark Tank file, the equity offered (`equity_offered_pct`) in single pitches ranges from 1% to 100%. The correct reading: the interval was produced by a procedure that captures the mean equity offered in all recorded pitches in about 95% of samples.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Pick one key
metric with a published industry benchmark (conversion rate, average order
value, churn), frame the comparison as a gap $\delta = \mu - \text{benchmark}$,
and write two sentences: first say what sample you could realistically
collect and what a 95% interval for $\delta$ excluding zero would let you
claim, then say what decision would change if the interval turned out to
include zero.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "The sector churn benchmark is 25% per term. From my roughly 120 subscribers I can estimate $\delta = \mu - 25$, and a 95% interval sitting entirely below zero would let me claim genuinely stickier customers in my pitch. If the interval included zero, I would stop pitching retention as my edge and lead with margin instead."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to check whether your
gap is defined on a population you can actually sample.
::::

[Continue to Lecture 10: Testing decisions, errors, power, and the road to regression](b10-testing-errors-power.qmd).

[Formula sheet: Lecture 9 formulas](formula-sheet.qmd#formula-l9).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 15 and 16.

:::
