# Lecture 9: The t distribution and confidence intervals {#b09}

```{r}
#| label: b09-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 9, you can</span>

- explain why estimating the population SD leads to a t distribution with
  $n-1$ degrees of freedom, and use its critical values
- explain where a confidence interval comes from, and tell the difference
  between the interval estimator and the confidence interval
- construct a confidence interval for a population mean at a given
  confidence level, such as 90%, 95%, or 99%, and write it in five parts
- construct a confidence interval for a population proportion
- calculate the sample size needed for a target margin of error
- state the null and the alternative hypothesis of a two-sided test, and
  calculate the test statistic
- state correctly what the confidence level is a property of.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- construct a confidence interval for a mean in the five parts, from the mean, the standard deviation, the sample size, and a supplied t value, at a given confidence level
- construct a confidence interval for a proportion in the five parts
- calculate the sample size for a target margin of error from a planning value for $\sigma$
- write $H_0$ and $H_1$ for a two-sided test of a mean or a proportion
- calculate the observed value of the test statistic for a mean and say what it measures
- say what the confidence level refers to.

**Enough to recognise:**

- the t curves for different degrees of freedom
- the interval with $\sigma$ known and a z value, as the first step towards the t interval
- `qt()`.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Suppose an advisor tells founders that pitches typically offer investors about 15% equity. Our data can check such a benchmark, if we take the variation between samples into account.</p>

Is the mean equity offered (`equity_offered_pct`) across all recorded Shark
Tank pitches really different from 15%?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 30 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (20 minutes)</span><br>
Watch <a href="https://www.youtube.com/watch?v=NInNhFm046w">Confidence
Intervals</a>,
<a href="https://www.youtube.com/watch?v=MzvRQFYUEFU">Confidence Intervals
When the Variance Is Unknown</a>, and
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to
Hypothesis Testing</a>. Focus on three points: what happens to an interval
when the sample is drawn again and again, why a t distribution takes the
place of the normal distribution, and which value a test compares the
estimate with. The videos use $M_n$ for the sample mean, where this book
uses $\bar X$, and $\hat\Theta$ for an estimator. All links are also listed in
the <a href="../resources/video-guide.html">External Video Guide</a>.
</div>

[Shark Tank data]{.shark-tag} **Warm up (5 minutes).** A random sample of $n=100$ pitches has a mean
equity offered (`equity_offered_pct`) of $\bar x=12.76$% and a sample
SD of $s=7.71$ percentage points.

1. Calculate the standard error of the sample mean.
2. Between which two values does the central 95% of the standard normal
   distribution lie?

::: {.callout-note collapse="true"}
## Check your preparation

1. The standard error is $s/\sqrt n=7.71/\sqrt{100}=0.771$ percentage points
   (Lecture 8).
2. Between $-1.96$ and $1.96$, because $z_{0.975}=1.96$ (Lecture 6).

A confidence interval is built from these pieces: an estimate, its standard
error, and a quantile of a known distribution.
:::

**Predict (5 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: in a real sample, the unknown population SD
$\sigma$ must be replaced by the sample SD $S$. Does this replacement make a
95% interval wider or narrower, or does it leave the width unchanged?
Choose one answer. You find the correct answer in the lecture material.

If you are not sure about intervals, standardisation, or degrees of freedom,
review
[Mathematics Refresher: intervals, standardisation, and degrees of
freedom](../resources/math-refresher.html).

:::

::: {.course-phase .phase-in-class}

## Lecture material

Lecture 8 gave a sample mean, which estimates the population mean, and its
standard error. This lecture turns those two numbers into a confidence
interval. That is an interval built by a method that captures the population
parameter in a chosen share of all samples, for example 95%. The lecture ends with a test
statistic, which measures how far an estimate lies from a proposed value of
the parameter. Lecture 10 turns the test statistic into a decision.

<p class="concept-video"><strong>Video explanations (review):</strong>
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/confidence-intervals/">20.5 Confidence Intervals</a>,
<a href="https://ocw.mit.edu/courses/res-6-012-introduction-to-probability-spring-2018/resources/confidence-intervals-for-the-mean-when-the-variance-is-unknown/">20.7 Confidence Intervals with Unknown Variance</a>, and
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to Hypothesis Testing</a>.</p>

### Start with a founder's question

[Shark Tank data]{.shark-tag} Suppose an advisor tells you that founders on the show offer about 15% of
their company. You have looked at 100 pitches, chosen by chance, and their
average equity offered (`equity_offered_pct`) is 12.76%. Is the advisor
wrong? Or is 12.76 simply what 100 pitches can give by chance when the true
average is 15? R gives an answer with one command:

```{r}
#| label: b09-opening
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
t.test(pitch_sample$equity_offered_pct, mu = 15)
```

R prints the mean of the sample, 12.76, a t value of $-2.91$ with 99 degrees
of freedom, a p-value of 0.0045, and an interval from 11.23 to 14.29. An AI
assistant would give you the same output, together with a sentence that
sounds sure about what it means.
To check that sentence, you need to know where the interval comes from, why
it uses a t distribution with 99 degrees of freedom, and what the t value
and the p-value say. This lecture explains the interval and sets up the
test. Lecture 10 explains the decision and the p-value.

### The price of estimating σ: the t distribution

Lecture 8 gave the standardised sample mean

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n}.
$$

Its distribution is known in two situations. If the population is normal,
$Z$ is exactly standard normal for every sample size (Lecture 8). If the
population is not normal but $n$ is large, $Z$ is approximately standard
normal by the central limit theorem (CLT). In both situations the
distribution of $Z$ is known, although $\mu$ is not. So we can use $Z$ to
draw conclusions about $\mu$. This is the step from sample statistics to
population parameters (Lecture 7).

But $Z$ also contains the population SD $\sigma$, and in a real study nobody
knows $\sigma$. The only replacement we have is the sample SD $S$. This
gives

$$
T=\frac{\bar X-\mu}{S/\sqrt n}.
$$

#### Why $T$ has heavier tails than $Z$

$S$ is itself a random variable. It changes from sample to sample, and in a
small sample it can be far from $\sigma$. In a sample where $S$ comes out
larger than $\sigma$, the denominator is too large and $T$ is pulled towards
zero. In a sample where $S$ comes out smaller than $\sigma$, the denominator
is too small and $T$ is pushed away from zero. When $S$ is far below
$\sigma$, the push is large. The two effects do not cancel. A moderate value
of $Z$ becomes an extreme value of $T$ whenever $S$ is much smaller than
$\sigma$. So $T$ has two sources of variation, $\bar X$ and $S$. Its
distribution is still symmetric around zero, but it puts more probability
far from zero. This is what "heavier tails" means.

#### The exact result for a normal population

If the population is normal, the distribution of $T$ is known exactly, for
every $n$:

$$
T=\frac{\bar X-\mu}{S/\sqrt n}
\sim t_{n-1}.
$$

::: {.definition-box}
<span class="label">Definition</span>

**t distributions and degrees of freedom.** The t distributions are a family
of distributions. There is one member for each value of a single number, the
**degrees of freedom** ($df=1,2,3,\ldots$). Every member is symmetric around
zero and has heavier tails than the standard normal. The smaller the degrees
of freedom, the heavier the tails. As the degrees of freedom grow, the
member approaches $N(0,1)$. For a random sample of size $n$ from a normal
population, $T$ follows the member with $df=n-1$.
:::

```{r}
#| label: fig-b09-t-curves
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The t distribution for 4 and 10 degrees of freedom against the standard normal. All three are symmetric around zero. The t curves put more probability in the tails, most visibly at low degrees of freedom."
library(ggplot2)

t_grid <- seq(-4, 4, length.out = 800)
curve_levels <- c("t, df = 4", "t, df = 10", "standard normal")
reference_curves <- rbind(
  data.frame(x = t_grid, density = dt(t_grid, df = 4),
             curve = curve_levels[1]),
  data.frame(x = t_grid, density = dt(t_grid, df = 10),
             curve = curve_levels[2]),
  data.frame(x = t_grid, density = dnorm(t_grid),
             curve = curve_levels[3])
)
reference_curves$curve <- factor(reference_curves$curve,
                                 levels = curve_levels)
curve_colours <- c("t, df = 4" = "#D55E00",
                   "t, df = 10" = "#0072B2",
                   "standard normal" = "#009E73")

ggplot(reference_curves, aes(x, density,
                             colour = curve, linetype = curve)) +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 2.3, y = 0.375, label = "t, df = 4",
           colour = "#D55E00", size = 3.4, hjust = 0) +
  annotate("text", x = 2.3, y = 0.335, label = "t, df = 10",
           colour = "#0072B2", size = 3.4, hjust = 0) +
  annotate("text", x = 2.3, y = 0.295, label = "standard normal",
           colour = "#009E73", size = 3.4, hjust = 0) +
  scale_colour_manual(values = curve_colours, guide = "none") +
  scale_linetype_manual(values = c("solid", "dashed", "dotted"),
                        guide = "none") +
  scale_y_continuous(limits = c(0, 0.41),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Standardised value", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

@fig-b09-t-curves shows two members of the family and the standard normal.
With $df=4$ the tails are clearly heavier. With $df=10$ the difference is
much smaller. A probability or a quantile for $T$ always comes from the
member with $df=n-1$. A different sample size means a different curve.

The difference is most visible in the tails. The central 95% of the standard
normal lies between $-1.96$ and $1.96$ (Lecture 6). A t curve must reach
further out to include its central 95%. The point where the central 95% ends
is the 0.975 quantile of the t distribution. We write it $t_{0.975,\,df}$
and compute it in R with `qt(0.975, df)`. This number is called the 95%
**critical value**. For small $df$ it is clearly larger than 1.96. Other
levels work in the same way: the central 90% ends at the 0.95 quantile
$t_{0.95,\,df}$, and the central 99% ends at the 0.995 quantile
$t_{0.995,\,df}$.

#### Why $n-1$

The degrees of freedom measure how precisely $S$ estimates $\sigma$. With
few degrees of freedom, $S$ is imprecise and the tails are heavy. With many
degrees of freedom, $S$ is close to $\sigma$ in almost every sample, and $T$
behaves like $Z$. The number is $n-1$ and not $n$ for the reason given in
Lecture 7. Once $\bar X$ is fixed, the $n$ deviations $X_i-\bar X$ must sum
to zero. So only $n-1$ of them can vary freely, and $S$ is built from those
$n-1$ free deviations.

#### The rule of this course {#b09-t-rule}

What if the population is not normal? For a large sample, $T$ is then
approximately standard normal, so a z value would be enough. The optional
box at the end of this section shows why. The t value is still a safe
choice, because $t_{n-1}$ comes closer and closer to $N(0,1)$ as $n$ grows.
At $n\approx 30$ the two curves are already almost the same. At $n=100$
the 95% critical values are $t_{0.975,\,99}=1.984$ and $z_{0.975}=1.96$. So
using the t distribution costs almost nothing when $n$ is large, and it is
exactly right when the population is normal. So this course uses one
rule: $\sigma$ unknown means t with $n-1$ degrees of freedom, whatever the
sample size.

| | Population normal | Population not normal, $n$ large |
|---|---|---|
| $\sigma$ known | $Z\sim N(0,1)$ exactly, for every $n$ | $Z\ \dot\sim\ N(0,1)$ by the CLT |
| $\sigma$ unknown | $T\sim t_{n-1}$ exactly, for every $n$ | $T\ \dot\sim\ N(0,1)$, and $t_{n-1}$ is almost the same, so we use $t_{n-1}$ |

The table shows the distribution of the standardised sample mean in each
situation. One case is missing: a small sample from a population that is not
normal. No result covers it. This is one reason to check the conditions
before calculating, as a later section of this lecture does.

::: {.callout-note collapse="true"}
## Check: which is larger, $t_{0.975,11}$ or $t_{0.975,99}$?

$t_{0.975,\,11}=2.201$ is larger than $t_{0.975,\,99}=1.984$. With
only 11 degrees of freedom, $S$ is an imprecise estimate of $\sigma$, so
the t distribution has heavier tails and its 0.975 quantile lies further
from zero. At $df=99$ the critical value is already close to the normal
1.96. With a large sample, the t and normal answers are almost the same.
Verify in R: `qt(0.975, df = 11)` and `qt(0.975, df = 99)`.
:::

::: {.callout-tip collapse="true"}
## Optional: $T$ is $Z$ times a random factor

The two standardised means differ by one factor:

$$
T=\frac{\bar X-\mu}{\sigma/\sqrt n}\times\frac{\sigma}{S}
=Z\times\frac{\sigma}{S}.
$$

If $S$ were equal to $\sigma$, the factor would be 1 and $T$ would be $Z$.
But $S$ is random, so the factor $\sigma/S$ is random too. When $S$ is
larger than $\sigma$, the factor is below 1 and $T$ is closer to zero than
$Z$. When $S$ is smaller than $\sigma$, the factor is above 1 and $T$ is
further from zero than $Z$. Everything about $T$ follows from how far this
factor moves away from 1.

The factor also explains the large-sample case. If the population is not
normal, $Z$ is approximately normal only for large $n$. The same large $n$
makes $S$ a precise estimate of $\sigma$, because $S$ is a consistent
estimator: with a larger and larger sample it comes closer and closer to
$\sigma$ (Lecture 8). The factor $\sigma/S$ is then close to 1 in almost
every sample, and

$$
T=Z\times\frac{\sigma}{S}\approx Z,\qquad\text{so}\qquad T\ \dot\sim\ N(0,1).
$$

Replacing $\sigma$ by a consistent estimate does not change the distribution
in large samples.
:::

### Confidence intervals

[Shark Tank data]{.shark-tag} Lectures 7 and 8 were about estimators: how a statistic such as $\bar X$ or
$\widehat P$ behaves across samples. From here on, we use a sample to
answer three kinds of question about an unknown parameter.

- A point estimate is a single value, the best guess. For the mean equity
  offered it is 12.76.
- An interval estimate is a range of values. It is built so that the method
  captures the parameter with a known probability. This lecture constructs
  it.
- A hypothesis test decides whether one proposed value of the parameter is
  compatible with the sample, with a known risk of a wrong decision. This
  lecture sets it up and Lecture 10 completes it.

The sample mean $\bar x=12.76$ is a point estimate of $\mu$. Two statements are
true of any such number. First, it is almost certainly not equal to $\mu$.
We treat the equity offered as continuous, so the probability that a
sample mean is exactly equal to $\mu$ is zero (Lecture 6). Second, the number
alone does not say how far from $\mu$ it might be. Lecture 8 measured that
distance with the standard error. A confidence interval combines the
estimate and its standard error into one statement about $\mu$.

There is a practical reason to report an interval. A founder who is told
"the mean equity offered is 12.76%" cannot tell whether 15% is still
possible. A founder who is told "between 11.2% and 14.3%, with 95%
confidence" can. The width of the interval shows the sampling uncertainty.

#### Where the interval comes from: σ known

First take the case in which the population SD $\sigma$ is known. This is
rare in practice, because a study that does not know $\mu$ will rarely know
$\sigma$. But it is the simplest case, and the realistic case follows from
it in one step.

The setting is a population with unknown mean $\mu$ and known SD $\sigma$,
and a random sample $X_1,\ldots,X_n$. As in Lecture 8, the population is
normal or $n$ is large. Then the standardised sample mean

$$
Z=\frac{\bar X-\mu}{\sigma/\sqrt n}
$$

is standard normal, exactly in the first case and approximately in the
second.

Now choose a small probability $\alpha$, for example $\alpha=0.05$. The
central $1-\alpha$ of the standard normal lies between $-z_{1-\alpha/2}$ and
$z_{1-\alpha/2}$, with $\alpha/2$ in each tail (Lecture 6):

$$
P\left(-z_{1-\alpha/2}\le Z\le z_{1-\alpha/2}\right)=1-\alpha .
$$

For $\alpha=0.05$ the two cut-offs are $-1.96$ and $1.96$. For $\alpha=0.10$
they are $\pm1.645$, and for $\alpha=0.01$ they are $\pm2.576$.
@fig-b09-normal-central shows the central area and the two tails for
$\alpha=0.05$.

```{r}
#| label: fig-b09-normal-central
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The standard normal density. The green central area has probability 1 − α and lies between the two cut-offs. Each orange tail has probability α/2. The curve is drawn for α = 0.05, where the cut-offs are −1.96 and 1.96."
library(ggplot2)

z_cut <- qnorm(0.975)
z_curve <- data.frame(z = seq(-3.6, 3.6, length.out = 800))
z_curve$density <- dnorm(z_curve$z)
centre <- data.frame(z = seq(-z_cut, z_cut, length.out = 400))
centre$density <- dnorm(centre$z)
left_tail <- data.frame(z = seq(-3.6, -z_cut, length.out = 200))
left_tail$density <- dnorm(left_tail$z)
right_tail <- data.frame(z = seq(z_cut, 3.6, length.out = 200))
right_tail$density <- dnorm(right_tail$z)

ggplot(z_curve, aes(z, density)) +
  geom_area(data = centre, fill = "#009E73", alpha = 0.25) +
  geom_area(data = left_tail, fill = "#D55E00", alpha = 0.55) +
  geom_area(data = right_tail, fill = "#D55E00", alpha = 0.55) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  annotate("text", x = 0, y = 0.18, label = "central area",
           colour = "#17324D", size = 3.6) +
  annotate("text", x = 0, y = 0.15, label = "1 - alpha", parse = TRUE,
           colour = "#17324D", size = 4.2) +
  annotate("segment", x = c(-2.9, 2.9), xend = c(-2.3, 2.3),
           y = 0.075, yend = 0.02, colour = "#17324D",
           arrow = arrow(length = unit(0.08, "inches"))) +
  annotate("text", x = c(-2.9, 2.9), y = 0.095, label = "alpha/2",
           parse = TRUE, colour = "#17324D", size = 3.8) +
  scale_x_continuous(breaks = c(-z_cut, 0, z_cut),
                     labels = expression(-z[1 - alpha/2], 0,
                                         z[1 - alpha/2])) +
  scale_y_continuous(limits = c(0, 0.43),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of Z", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    axis.text.x = element_text(size = 12, colour = "#17324D")
  )
```

Now write $Z$ out in full. The event in the probability above is

$$
-z_{1-\alpha/2}\ \le\ \frac{\bar X-\mu}{\sigma/\sqrt n}\ \le\ z_{1-\alpha/2}.
$$

It says that $\bar X$ and $\mu$ are at most
$z_{1-\alpha/2}\,\sigma/\sqrt n$ apart. Distance is symmetric. "$\bar X$
lies within this distance of $\mu$" and "$\mu$ lies within this distance of
$\bar X$" are the same event, with the same probability. Solving the inequality for $\mu$
gives

$$
P\left(\bar X-z_{1-\alpha/2}\frac{\sigma}{\sqrt n}
\ \le\ \mu\ \le\
\bar X+z_{1-\alpha/2}\frac{\sigma}{\sqrt n}\right)=1-\alpha .
$$

This is a statement about $\mu$. Write the two bounds as

$$
L=\bar X-z_{1-\alpha/2}\frac{\sigma}{\sqrt n},
\qquad
U=\bar X+z_{1-\alpha/2}\frac{\sigma}{\sqrt n}.
$$

$L$ and $U$ are calculated from the random sample, so they are random
variables, and $P(L\le\mu\le U)=1-\alpha$. The centre of $(L,U)$ is the
estimator $\bar X$. When the data are available, $\bar x$ takes the place
of $\bar X$, and the bounds become two numbers $l$ and $u$:

$$
l=\bar x-z_{1-\alpha/2}\frac{\sigma}{\sqrt n},
\qquad
u=\bar x+z_{1-\alpha/2}\frac{\sigma}{\sqrt n}.
$$

::: {.definition-box}
<span class="label">Definition</span>

**Interval estimator and confidence interval.** The random interval $(L,U)$
is an interval estimator: before the data are drawn, it captures the
parameter with a chosen probability $1-\alpha$. The interval $(l,u)$
calculated from the data is a $100(1-\alpha)\%$ confidence interval. The
probability $1-\alpha$ is the **confidence level**.
:::

::: {.callout-tip collapse="true"}
## Optional: solving the inequality for $\mu$ step by step

Start from

$$
-z_{1-\alpha/2}\ \le\ \frac{\bar X-\mu}{\sigma/\sqrt n}\ \le\ z_{1-\alpha/2}.
$$

Multiply all three parts by $\sigma/\sqrt n$, which is positive:

$$
-z_{1-\alpha/2}\frac{\sigma}{\sqrt n}\ \le\ \bar X-\mu\ \le\
z_{1-\alpha/2}\frac{\sigma}{\sqrt n}.
$$

Subtract $\bar X$ from all three parts:

$$
-\bar X-z_{1-\alpha/2}\frac{\sigma}{\sqrt n}\ \le\ -\mu\ \le\
-\bar X+z_{1-\alpha/2}\frac{\sigma}{\sqrt n}.
$$

Multiply by $-1$. This turns the inequality signs around:

$$
\bar X+z_{1-\alpha/2}\frac{\sigma}{\sqrt n}\ \ge\ \mu\ \ge\
\bar X-z_{1-\alpha/2}\frac{\sigma}{\sqrt n}.
$$

Read from right to left, this is $L\le\mu\le U$. Each step keeps the same
event, so the probability stays $1-\alpha$.
:::

#### The interval with real numbers

With the Shark Tank file we can do this calculation with real numbers. The file is the
whole population of recorded pitches, so its SD of the equity offered is
known: $\sigma=8.44$ percentage points.

```{r}
#| label: b09-sigma-known
x_bar <- mean(pitch_sample$equity_offered_pct)       # sample mean
n     <- nrow(pitch_sample)

# The file is the whole population, so sigma divides by N, not by N - 1.
mu_equity    <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))

sd_xbar  <- sigma_equity / sqrt(n)     # standard deviation of the sample mean
margin_z <- qnorm(0.975) * sd_xbar

c(sample_mean = x_bar, sigma = sigma_equity, sd_of_sample_mean = sd_xbar,
  z_0.975 = qnorm(0.975), margin = margin_z)
c(lower = x_bar - margin_z, upper = x_bar + margin_z)
```

The standard deviation of the sample mean is $8.44/\sqrt{100}=0.844$. With
$z_{0.975}=1.96$, the bounds are $12.76\pm1.96\times0.844=12.76\pm1.65$. The
95% confidence interval with $\sigma$ known is $(11.11,\ 14.41)$.

The difference between $(L,U)$ and $(l,u)$ explains what the word
confidence means. The method captures $\mu$ in a fraction $1-\alpha$ of all
samples. After the sample is drawn, $l$, $u$, and $\mu$ are fixed numbers.
Then $\mu$ either lies between $l$ and $u$ or it does not, and nobody knows
which. The 95% belongs to the method, not to the one interval it produced.
"$\mu$ lies in $(11.11,\ 14.41)$ with 95% confidence" is short for "this
interval was produced by a method that captures $\mu$ in 95% of samples".

#### The realistic case: σ unknown

In practice $\sigma$ is not known, and we estimate it by the sample SD $S$. The
construction stays the same, with one change. The standardised mean is now
$T=(\bar X-\mu)/(S/\sqrt n)$, and its distribution is $t_{n-1}$ in place of
$N(0,1)$. So $T$ replaces $Z$, and the critical value
$t_{1-\alpha/2,\,n-1}$ replaces $z_{1-\alpha/2}$. The same steps give

$$
P\left(\bar X-t_{1-\alpha/2,\,n-1}\frac{S}{\sqrt n}
\ \le\ \mu\ \le\
\bar X+t_{1-\alpha/2,\,n-1}\frac{S}{\sqrt n}\right)=1-\alpha .
$$

The interval estimator is therefore

$$
\bar X\pm t_{1-\alpha/2,\,n-1}\frac{S}{\sqrt n},
$$

and with the data put in, $\bar x\pm t_{1-\alpha/2,\,n-1}\,s/\sqrt n$ is a
$100(1-\alpha)\%$ confidence interval for $\mu$. It is exact for a normal
population and approximate for a large sample from a population that is not
normal. These are the same two situations as before. A 95% interval has
$\alpha=0.05$ and uses $t_{0.975,\,n-1}$. A 90% interval uses
$t_{0.95,\,n-1}$, and a 99% interval uses $t_{0.995,\,n-1}$.

From here on, we use this interval. The version with $\sigma$ known was
only the first step towards it. In the exam, an interval or a test for a
mean always uses $s$ and $t$.

The interval has a form that every confidence interval in this course
shares:

$$
\text{estimate}
\ \pm\
\text{critical value}\times\text{standard error}.
$$

### The margin of error

::: {.definition-box}
<span class="label">Definition</span>

**Margin of error.** The part after the $\pm$ sign of a confidence interval,
written $m$. It is half the width of the interval. For a mean it is the
critical value times the standard error: $m=t_{1-\alpha/2,\,n-1}\,S/\sqrt n$,
with the value $t_{1-\alpha/2,\,n-1}\,s/\sqrt n$ in one sample.
:::

The margin of error is the precision of the estimate at the stated
confidence level. A $100(1-\alpha)\%$ confidence interval estimates $\mu$ to
within $m$, with confidence $1-\alpha$. When $\sigma$ is known,
$m=z_{1-\alpha/2}\,\sigma/\sqrt n$ is a fixed number. When $\sigma$ is
unknown, $m$ contains $S$ and so it varies from sample to sample.

Four facts about the margin of error:

- The interval estimator captures $\mu$ with probability $1-\alpha$. A
  calculated interval either contains $\mu$ or it does not.
- A higher confidence level means a smaller $\alpha$ and a larger critical
  value. The interval becomes wider. More confidence costs width.
- Observations that vary more (a larger $\sigma$ or a larger $s$) give a
  wider interval.
- A larger sample gives a narrower interval, because $\sqrt n$ is in the
  denominator. Four times the observations halve the margin (Lecture 8).

You can also use the formula for $m$ the other way round. If you fix the
margin in advance, you can solve the formula for the sample size that gives
this margin. The
section [How many customers must I ask?](#b09-sample-size) does this.

The formula also answers the prediction question from the preparation.
Replacing $\sigma$ with $S$ makes the interval wider on average, because the
t critical value is larger than the z value. The extra width is the price of
estimating one more quantity from the same data.

In the exam, the critical value you need is printed in the question or on
the formula sheet. For a value of the degrees of freedom that is not on the
sheet, use the closest one listed.

### The interval for a mean in five parts

::: {.exam-procedure}
<span class="label">Exam procedure: confidence interval for a mean</span>

Write these five parts, in this order.

1. **Formula.** The interval estimator:
   $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. Say why it may be used: the
   population is normal, or $n$ is large.
2. **Given values.** $\bar x$, $s$, $n$, $df=n-1$, the confidence level,
   and the critical value $t_{1-\alpha/2,\,n-1}$ from the question or the
   formula sheet.
3. **Calculations.** The standard error $s/\sqrt n$. Then the margin of
   error: critical value $\times$ standard error.
4. **Interval.** $(\bar x-\text{margin},\ \bar x+\text{margin})$, with the
   unit.
5. **Conclusion.** One sentence in the words of the question. If the
   question is about a value $\mu_0$: say whether $\mu_0$ lies inside the
   interval. If it lies outside, you can conclude at level $\alpha$ that
   $\mu$ differs from $\mu_0$, and the interval shows on which side. If it
   lies inside, you cannot conclude that $\mu$ differs from $\mu_0$.

<p class="exam-note">In the exam, the question asks for the formula of the
interval, the interval, and the conclusion with an explanation. Give the
five parts in this order.</p>
:::

In part 5, $\alpha$ is one minus the confidence level, so $\alpha=0.05$ for
a 95% interval. Lecture 10 explains why the interval allows this conclusion.

::: {#exm-b09-hand-interval}
## Worked example: a 95% interval by hand

An invented example. A random sample of $n=64$ ventures reports a
mean customer satisfaction score of $\bar x = 80$ points, with sample SD
$s = 16$ points. The industry benchmark is 75 points. Construct a 95%
confidence interval for the population mean $\mu$ of the score. Is the
benchmark a plausible value for $\mu$?

1. **Formula.** $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used
   because the sample is random and $n=64$ is large.
2. **Given values.** $\bar x=80$, $s=16$, $n=64$, so $df=63$. The confidence
   level is 95%, so $\alpha=0.05$ and the critical value is $t_{0.975,\,63}$.
   The closest value on the formula sheet is the one for $df=60$, which is
   2.000. In R, `qt(0.975, df = 63)` gives 1.998.
3. **Calculations.** The standard error is $s/\sqrt n=16/\sqrt{64}=16/8=2$
   points. The margin of error is $2.000\times2=4$ points.
4. **Interval.** $(80-4,\ 80+4)=(76,\ 84)$ points.
5. **Conclusion.** With 95% confidence, the mean satisfaction score of all
   ventures lies between 76 and 84 points. The benchmark of 75 points lies
   outside the interval. At the 5% level we can conclude that the mean
   differs from the benchmark, and the interval shows that it is higher.
:::

### Check conditions before calculating

[Shark Tank data]{.shark-tag} The interval formula is valid only when certain conditions hold. Some
are about how the sample was drawn, and some are about the shape of the
data. R does not check them for you. Before any
interval or test, go through this checklist:

1. **Random selection.** Were the rows selected at random from a clearly
   defined sampling frame? In the Shark Tank case below they were: the sample of 100 is drawn
   at random (seed 123) from the 1,441 recorded pitches.
2. **The 10% condition.** Sampling without replacement makes the
   observations slightly dependent, but $100/1441\approx 0.07 < 0.10$. So we
   treat the sample as independent and identically distributed (Lecture 7), and the standard
   error formula is a reasonable approximation.
3. **Shape and sample size.** The equity offered (`equity_offered_pct`) is
   right-skewed, so the population is not normal. With $n=100$ the sample is
   large and the t interval for the mean is still reasonable. For a very
   small sample it would not be: that is the missing case of the table
   above. Also check for data errors and for single observations that have
   a very large influence on the mean.

These conditions support conclusions about the recorded pitches, the
sampling frame. They do
not make televised pitches representative of all entrepreneurs.

::: {.callout-caution collapse="true"}
## Try it: is the average order above break-even?

Try this with your neighbour first. An invented example. A venture
needs an average order of 20 euros to break even. A random sample of $n=36$
orders has a mean of 21.50 euros and a standard deviation of 6 euros.
Construct a 95% confidence interval for the mean order value. Supplied
value: $t_{0.975,\,35}=2.030$.

*Parts 1 and 2 (given).* The formula is
$\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used because the
sample is random and $n=36$ is large. The given values are $\bar x=21.50$,
$s=6$, $n=36$, $df=35$, the 95% level, and $t_{0.975,\,35}=2.030$.

*Part 3 (started).* The standard error is $6/\sqrt{36}=1$ euro. Calculate
the margin of error.

*Part 4 (you).* Write the interval.

*Part 5 (you).* Is 20 euros a plausible value for the mean order? What would
you tell the founder?

::: {.callout-note collapse="true"}
## Check your answer

Part 3: the margin of error is $2.030\times1=2.03$ euros. Part 4: the
interval is $(19.47,\ 23.53)$ euros. Part 5: the break-even value of 20
euros lies inside the interval, so we cannot conclude that the mean order
differs from 20 euros. The sample mean is above break-even, but the data
do not exclude that the true mean is below it. Advice: collect more orders
before any decision that depends on being above break-even. With four times
as many orders the margin would be about half as wide.
:::
:::

### The Shark Tank case, step by step

[Shark Tank data]{.shark-tag} Now apply the five parts to the Shark Tank data. The code draws the same
random sample of 100 pitches as in Lecture 7 and builds every object step by
step:

```{r}
#| label: b09-mean-objects
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

x_bar <- mean(pitch_sample$equity_offered_pct)   # sample mean
s     <- sd(pitch_sample$equity_offered_pct)     # sample SD
n     <- nrow(pitch_sample)
df    <- n - 1
se    <- s / sqrt(n)                             # standard error

c(
  sample_mean = x_bar,
  sample_sd = s,
  sample_size = n,
  degrees_of_freedom = df,
  standard_error = se
)
```

```{r}
#| label: b09-mean-ci
critical_t <- qt(0.975, df = df)
margin     <- critical_t * se
mean_ci    <- x_bar + c(-1, 1) * margin

c(critical_t = critical_t, margin_of_error = margin)
c(lower = mean_ci[1], upper = mean_ci[2])
```

The same calculation, written in the five parts:

1. **Formula.** $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used
   because the sample is random and $n=100$ is large.
2. **Given values.** $\bar x=12.76$, $s=7.71$, $n=100$, so $df=99$. The
   confidence level is 95%, so the critical value is
   $t_{0.975,\,99}=1.984$. The formula sheet gives the same value for
   $df=100$, the closest one listed.
3. **Calculations.** The standard error is $7.71/\sqrt{100}=0.771$
   percentage points. The margin of error is $1.984\times0.771=1.53$
   percentage points.
4. **Interval.** $(12.76-1.53,\ 12.76+1.53)=(11.23,\ 14.29)$%.
5. **Conclusion.** With 95% confidence, the mean equity offered in all
   recorded pitches lies between 11.23 and 14.29%. The advisor's
   benchmark of 15% lies outside the interval. At the 5% level we can
   conclude that the mean equity offered differs from 15%, and the
   interval shows that it is lower.

The sample mean lies 2.24 percentage points below the benchmark, about one
seventh of the benchmark.

The interval with $\sigma$ known was $(11.11,\ 14.41)$. In this sample
$s=7.71$ happens to be smaller than $\sigma=8.44$, so the t interval is a
little narrower here. On average over many samples, the t interval is the
wider one.

### Interpret the confidence level correctly

[Shark Tank data]{.shark-tag} The parameter is fixed. The endpoints of the interval vary from sample to
sample. The confidence level $1-\alpha$ describes the procedure. If we repeat
the same valid procedure on many random samples, a fraction $1-\alpha$ of
the resulting intervals contains the parameter. For a 95% level, that is
about 95 intervals in 100. For a 90% level, the intervals are narrower and
about 90 in 100 contain the parameter.

The simulation in @fig-b09-coverage repeats the random sample and the
calculation of a 95% interval 100 times. The fixed target is the mean equity
offered (`equity_offered_pct`) among all 1,441 recorded pitches, 13.21%.
It is known here because the file is the whole population.

```{r}
#| label: fig-b09-coverage
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 5.2
#| fig-cap: "One hundred repetitions of the same procedure: draw 100 random pitches, compute a 95% t interval for the mean equity offered. Green intervals contain the fixed target mean (gold dashed line). Orange intervals miss it. In this run, 96 of 100 intervals cover the target."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(209)
B_coverage <- 100
n_coverage <- 100
target_mean <- mean(sharks$equity_offered_pct)
ci_lower <- ci_upper <- ci_centre <- numeric(B_coverage)

for (i in seq_len(B_coverage)) {
  repeated_sample <- sample(
    sharks$equity_offered_pct,
    size = n_coverage,
    replace = FALSE
  )
  repeated_mean <- mean(repeated_sample)
  repeated_se <- sd(repeated_sample) / sqrt(n_coverage)
  repeated_critical <- qt(0.975, df = n_coverage - 1)
  ci_centre[i] <- repeated_mean
  ci_lower[i] <- repeated_mean - repeated_critical * repeated_se
  ci_upper[i] <- repeated_mean + repeated_critical * repeated_se
}

intervals <- data.frame(
  lower = ci_lower, centre = ci_centre, upper = ci_upper,
  covers = ifelse(ci_lower <= target_mean & ci_upper >= target_mean,
                  "contains the target mean", "misses the target mean")
)
intervals <- intervals[order(intervals$centre), ]
intervals$position <- seq_len(B_coverage)

ggplot(intervals, aes(y = position, colour = covers)) +
  geom_segment(aes(x = lower, xend = upper, yend = position),
               linewidth = 0.7) +
  geom_point(aes(x = centre), size = 0.9) +
  geom_vline(xintercept = target_mean, linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.7) +
  annotate("text", x = target_mean, y = B_coverage + 3.5,
           label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = -0.1) +
  scale_colour_manual(values = c("contains the target mean" = "#009E73",
                                 "misses the target mean" = "#D55E00")) +
  scale_y_continuous(breaks = c(1, 20, 40, 60, 80, 100)) +
  labs(x = "Mean equity offered (%)",
       y = "Repeated sample (ordered by estimate)",
       colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    legend.position = "top"
  )
```

Each horizontal segment in @fig-b09-coverage is one 95% confidence
interval. Its dot is that sample's mean. The gold dashed vertical line is
the fixed target mean, 13.21. In this run (seed 209), 96 of 100 intervals
cross the line. A new run does not have to give exactly 95, but across many
repetitions the proportion approaches 95% when the conditions of the
procedure hold.

The following sentence may look natural, but it is wrong:

> "There is a 95% probability that the population mean is in my interval."

Before the sample is drawn, the endpoints of the interval are random, and
the 95% applies to them. After the sample is drawn, the interval
$(11.23, 14.29)$ consists of two fixed numbers. It either contains the fixed
parameter or it does not. Nothing random is left, so the probability of 95%
no longer applies. The correct sentence is:

> "This interval was produced by a procedure that captures the fixed
> population mean in about 95% of repeated valid samples."

The same holds for every other confidence level, with 90% or 99% in place of
95%.

### How many customers must I ask? {#b09-sample-size}

A larger sample gives a smaller margin of error: four times the
observations halve it. A founder often asks a more direct question before a
survey: how large must the sample be to reach a given precision?

The answer comes from the formula for the margin of error, used the other
way round. Before the survey there is no sample, so there is no $s$. So for
planning you use a value for $\sigma$ from a pilot survey (a small test
survey) or from an earlier study. This value is called a planning value, and
you treat it as known. That is why the formula
uses a z value: $m=z_{1-\alpha/2}\,\sigma/\sqrt n$. Choose the largest
margin of error you accept and call it $m$. Then solve for $n$:

$$
z_{1-\alpha/2}\times\frac{\sigma}{\sqrt n}\le m
\qquad\text{gives}\qquad
n\ge\left(\frac{z_{1-\alpha/2}\times\sigma}{m}\right)^2.
$$

For a 95% confidence level, $z_{0.975}=1.96$ and the formula is
$n\ge(1.96\times\sigma/m)^2$. For 90% use $z_{0.95}=1.645$, and for 99% use
$z_{0.995}=2.576$. Always round the result up to the next whole number.

::: {#exm-b08-sample-size}
## Worked example: a sample size for a margin of 2 euros

An invented example. You want to know the average order value of your
customers to within 2 euros, at the 95% confidence level. How many customers
must you survey?

- **Planning value.** The formula needs $\sigma$, which you do not know
  before the survey. A small pilot survey gave an SD of about 12 euros, so
  use $\sigma=12$ as a planning value. The target margin of 2 euros is one
  sixth of that SD.
- **Sample size.** $n\ge(1.96\times 12/2)^2=11.76^2\approx 138.3$.
- **Round up.** Survey $n=139$ customers. Always round up: with 138
  customers the margin would be slightly above 2 euros.

Halving the margin to 1 euro needs $(1.96\times 12/1)^2=23.52^2\approx 553.2$,
so 554 customers. Half the margin costs about four times the sample, as the
square-root rule of Lecture 8 says. A higher confidence level also needs
more observations. A margin of 2 euros at the 99% level needs
$(2.576\times 12/2)^2=15.456^2\approx 238.9$, so 239 customers.

```{r}
#| label: b08-sample-size-calc
sigma_plan <- 12
n_needed <- c(margin_2_euros = (1.96 * sigma_plan / 2)^2,
              margin_1_euro  = (1.96 * sigma_plan / 1)^2,
              margin_2_euros_99_percent = (2.576 * sigma_plan / 2)^2)
n_needed
ceiling(n_needed)   # ceiling() rounds up to the next whole number
```
:::

### The same recipe for a proportion

[Shark Tank data]{.shark-tag} A proportion is the mean of a 0/1 variable (Lecture 8). Take the deal
indicator (`deal_on_show`), which is 1 if a pitch ends in a deal and 0 if
not. Let $p$ be the proportion of all pitches that end in a deal. The sample
proportion $\widehat P$ is the sample mean of the indicators, so the results
of Lecture 8 apply directly:

$$
E[\widehat P]=p,
\qquad
\operatorname{SD}(\widehat P)=\sqrt{\frac{p(1-p)}{n}}.
$$

There is no separate $\sigma$ to know or to estimate. The spread of a 0/1
variable is determined by $p$ itself, because $\sigma^2=p(1-p)$ (Lecture 7). By
the central limit theorem, if $np\ge10$ and $n(1-p)\ge10$ (Lecture 8),

$$
Z=\frac{\widehat P-p}{\sqrt{p(1-p)/n}}
$$

is approximately standard normal. The steps of the mean case then give two
bounds around $\widehat P$. But these bounds contain $p$ inside the square
root, and $p$ is the unknown parameter. The solution is the standard error of Lecture
8: replace $p$ inside the square root by its estimator $\widehat P$. For
large $n$ this changes little. The interval estimator is

$$
\widehat P\pm z_{1-\alpha/2}\sqrt{\frac{\widehat P(1-\widehat P)}{n}},
$$

and it captures $p$ with a probability of approximately $1-\alpha$. With the
data put in, the $100(1-\alpha)\%$ confidence interval for $p$ is

$$
\hat p\pm z_{1-\alpha/2}\sqrt{\frac{\hat p(1-\hat p)}{n}}.
$$

It has the general form of every interval in this course. The estimate is
$\hat p$, the critical value is a z value, and the standard error is
$\operatorname{SE}(\widehat P)=\sqrt{\hat p(1-\hat p)/n}$.

::: {.callout-tip collapse="true"}
## Optional: the bounds that still contain $p$

The same steps as for the mean, applied to
$Z=(\widehat P-p)/\sqrt{p(1-p)/n}$, give with a probability of approximately
$1-\alpha$

$$
\widehat P-z_{1-\alpha/2}\sqrt{\frac{p(1-p)}{n}}
\ \le\ p\ \le\
\widehat P+z_{1-\alpha/2}\sqrt{\frac{p(1-p)}{n}} .
$$

Here the mean and the proportion differ. These bounds contain $p$, the
unknown that they should capture, so they cannot be calculated. Replacing
$p$ inside the square root by the consistent estimator $\widehat P$ turns
the bounds into sample statistics:

$$
L=\widehat P-z_{1-\alpha/2}\sqrt{\frac{\widehat P(1-\widehat P)}{n}},
\qquad
U=\widehat P+z_{1-\alpha/2}\sqrt{\frac{\widehat P(1-\widehat P)}{n}} .
$$
:::

Three remarks.

- The critical value is a z value, not a t value. We used the t
  distribution for the mean because, in a normal population, there is an
  exact result for replacing $\sigma$ by $S$. No such result exists for a proportion. The interval is
  a large-sample approximation from the start, so $z$ is the right critical
  value.
- Check the condition with $\hat p$. The success-failure rule contains
  the unknown $p$. In practice, you check it with the sample proportion:
  $n\hat p\ge10$ and $n(1-\hat p)\ge10$.
- The margin of error depends on $\hat p$, which is not known before the
  sample is drawn. So this formula does not give a required sample size
  directly, unlike the mean case with a planning value for $\sigma$.

::: {.exam-procedure}
<span class="label">Exam procedure: confidence interval for a proportion</span>

Write these five parts, in this order.

1. **Formula.** The interval estimator:
   $\widehat P\pm z_{1-\alpha/2}\sqrt{\widehat P(1-\widehat P)/n}$. Say why
   it may be used: $n\hat p\ge10$ and $n(1-\hat p)\ge10$.
2. **Given values.** The number of successes, $n$, $\hat p$, the confidence
   level, and the critical value $z_{1-\alpha/2}$ from the question or the
   formula sheet.
3. **Calculations.** The standard error $\sqrt{\hat p(1-\hat p)/n}$. Then
   the margin of error: critical value $\times$ standard error.
4. **Interval.** $(\hat p-\text{margin},\ \hat p+\text{margin})$.
5. **Conclusion.** One sentence in the words of the question. If the
   question is about a value $p_0$: say whether $p_0$ lies inside the
   interval. If it lies outside, you can conclude at level $\alpha$ that
   $p$ differs from $p_0$, and the interval shows on which side. If it lies
   inside, you cannot conclude that $p$ differs from $p_0$.

<p class="exam-note">The interval uses $\hat p$ in the standard error. The
test of Lecture 10 uses $p_0$.</p>
:::

::: {#exm-b09-proportion-interval}
## Worked example: a 95% interval for the deal proportion

[Shark Tank data]{.shark-tag} In the random sample of 100 pitches from Lecture 7, 64 ended in a deal.
Construct a 95% confidence interval for the proportion $p$ of all recorded
pitches that end in a deal. Is $p=0.50$ a plausible value?

```{r}
#| label: b09-proportion-ci
deals <- sum(pitch_sample$deal_on_show)    # number of deals in the sample
p_hat <- deals / n                         # sample proportion
se_p  <- sqrt(p_hat * (1 - p_hat) / n)     # standard error
margin_p <- qnorm(0.975) * se_p

c(deals = deals, p_hat = p_hat, standard_error = se_p,
  z_0.975 = qnorm(0.975), margin_of_error = margin_p)
c(lower = p_hat - margin_p, upper = p_hat + margin_p)
```

1. **Formula.**
   $\widehat P\pm z_{1-\alpha/2}\sqrt{\widehat P(1-\widehat P)/n}$. It may be
   used because $n\hat p=64$ and $n(1-\hat p)=36$ are both at least 10.
2. **Given values.** 64 deals in $n=100$ pitches, so $\hat p=0.64$. The
   confidence level is 95%, so the critical value is $z_{0.975}=1.96$.
3. **Calculations.** The standard error is
   $\sqrt{0.64\times0.36/100}=\sqrt{0.002304}=0.048$. The margin of error is
   $1.96\times0.048=0.094$.
4. **Interval.** $(0.64-0.094,\ 0.64+0.094)=(0.546,\ 0.734)$.
5. **Conclusion.** With 95% confidence, between about 55% and 73% of all
   recorded pitches end in a deal. The value 0.50 lies outside the interval.
   At the 5% level we can conclude that the deal proportion differs from one
   half, and the interval shows that it is higher.
:::

Because the file is the whole population, we can check both intervals.
The mean equity offered among all 1,441 recorded pitches is $\mu=13.21$%,
inside $(11.23,\ 14.29)$. The deal proportion is $p=0.612$, inside
$(0.546,\ 0.734)$.

```{r}
#| label: b09-population-values
# Known only because the file is the whole population.
c(population_mean_equity = mean(sharks$equity_offered_pct),
  population_deal_proportion = mean(sharks$deal_on_show))
range(sharks$equity_offered_pct)   # smallest and largest equity offered
```

### From interval to test

[Shark Tank data]{.shark-tag} An interval answers one kind of question: which values of the parameter are
compatible with the sample? A second kind of question works the other way
round. Someone proposes one specific value, and the question is whether the
sample is compatible with it. The advisor in the driving question says that
pitches typically offer 15% equity. The sample of 100 pitches has a
mean of 12.76. Is the advisor's figure compatible with this sample, or does
the sample show that the figure is wrong? Answering questions of this kind
is called hypothesis testing. This section sets up the question. Lecture 10
answers it.

#### A test compares two hypotheses

A hypothesis test is a decision between two statements about an unknown
population parameter, such as a mean $\mu$ or a proportion $p$. The decision
is based on a random sample. The two statements cannot both be true.

::: {.definition-box}
<span class="label">Definition</span>

**Null hypothesis and alternative hypothesis.** The null hypothesis $H_0$
and the alternative hypothesis $H_1$ are two statements about a parameter
that cannot both be true. A hypothesis test uses a random sample to decide
between them. The decision is either "reject $H_0$" or "do not reject
$H_0$".
:::

You cannot choose freely which statement is the null and which is the
alternative. The alternative is the claim that the researcher wants to establish (to
show with data): the research question, the new finding. The null is its
opposite: the default position, "there is no effect or difference". Take a medical trial that hopes to show
that a new drug changes blood pressure. It puts "the drug has an effect" in
$H_1$ and "the drug has no effect" in $H_0$. The reason follows below.

#### The founder's hypotheses

A founder believes that the advisor's figure is wrong. Her claim is about
the population mean $\mu$ of the equity offered: $\mu\ne15$. The
hypotheses are $H_0:\mu=15$ versus $H_1:\mu\ne15$.

#### Two conclusions that are not symmetric

A test ends with one of two conclusions. They do not have the same strength.

- Reject $H_0$. The evidence in the data is strong enough to support $H_1$.
  In the founder's example: "we found enough evidence that the mean equity
  offered differs from 15%." This conclusion can be wrong, but only
  with a small probability that is fixed in advance. Lecture 10 makes this
  precise. It is the strong conclusion.
- Do not reject $H_0$. The data did not give enough evidence against $H_0$.
  This is not the same as concluding that $H_0$ is true. In the example: "we
  do not have enough evidence that the mean equity offered differs from
  15%, but we cannot claim that it equals 15%." The sample may have been too
  small or too variable to answer the question. It is the weak conclusion.

The difference comes from what a test controls. A test is built so that the
probability of rejecting $H_0$ wrongly stays below a small number fixed in
advance. The construction sets no limit on the other mistake: not rejecting
$H_0$ when $H_0$ is false. So "reject" is a conclusion with a guarantee, and
"do not reject" is a conclusion without one. This is also why we put the
claim to be established in $H_1$. If the founder's claim were in $H_0$, the
best result for her would be "do not reject", which establishes nothing.

A court works in the same way. The court starts from the assumption that the
accused person is innocent ($H_0$). The prosecutor must prove guilt ($H_1$).
The verdict is "guilty" or "not guilty", never "innocent". The system keeps
the probability of finding an innocent person guilty small. In return, it
accepts that a guilty person is sometimes found not guilty.

#### How a test differs from a confidence interval

The interval and the test use the same pieces: an estimate, a standard
error, and a known distribution. But they ask different questions. The
interval starts from the sample and returns a set of values for the
parameter. Its guarantee is how often the method captures the true value.
The test starts from one value proposed in advance and returns a decision
about it. Its guarantee is how often the method rejects a value that is
true. Lecture 10 shows that for a mean the interval and the two-sided test
lead to the same decision.

#### The two-sided test used in this course

The founder's claim has no direction: the advisor's figure can be wrong in
either direction. A test of such a claim is called a two-sided test. It is
the standard choice, and the safer one. This course works with the two-sided test

$$
H_0:\mu=\mu_0
\qquad\text{versus}\qquad
H_1:\mu\ne\mu_0 ,
$$

where $\mu_0$ is a fixed, known number: the value that is tested. An
optional box in Lecture 10 describes tests with a direction, called
one-sided tests. For the advisor's claim, $\mu_0=15$:

$$
H_0:\mu=15
\qquad\text{versus}\qquad
H_1:\mu\ne15 .
$$

State both hypotheses from the question, before you look at the sample.

#### The test statistic

As for the interval, the first step is the case with $\sigma$ known. The
setting is the same: a population with unknown mean $\mu$ and known SD
$\sigma$, a random sample $X_1,\ldots,X_n$, and a population that is normal
or an $n$ that is large. For the Shark Tank file we can calculate this case
with real numbers, because $\sigma=8.44$ is known.

The obvious evidence against $H_0$ is the distance between the sample mean
and the tested value, $\bar x-\mu_0$. For the advisor's claim it is
$12.76-15=-2.24$ percentage points. A distance alone is hard to judge. A
difference of 2.24 points is large if sample means rarely lie more than one
point from $\mu$. It is small if they often lie five points away. The right
scale is the standard deviation of the sample mean, $\sigma/\sqrt n$.
Dividing by it gives the **test statistic**

$$
Z=\frac{\bar X-\mu_0}{\sigma/\sqrt n}.
$$

This is the standardised sample mean of the interval section, with $\mu_0$
in place of $\mu$. If $H_0$ is true, then $\mu=\mu_0$ and the distribution
of $Z$ is known. It is standard normal, exactly for a normal population and
approximately for a large sample. If $H_0$ is false, $\bar X$ is centred at
a value other than $\mu_0$, and $Z$ tends to lie far from zero. The value of the test
statistic calculated from the data is called the observed value, $z_{obs}$.

::: {#exm-b09-test-statistic}
## Worked example: the hypotheses and the observed value

[Shark Tank data]{.shark-tag} A complete test has five steps, which Lecture 10 lists. This example carries
out steps 1, 2, and 4. The advisor claims $\mu_0=15$. The sample of Lecture
7 has $n=100$ and $\bar x=12.76$. The population SD, known from the file, is
$\sigma=8.44$.

- **Step 1. Hypotheses.** $H_0:\mu=15$ versus $H_1:\mu\ne15$, where $\mu$ is
  the mean equity offered in all recorded pitches.
- **Step 2. Test statistic.** $Z=(\bar X-\mu_0)/(\sigma/\sqrt n)$. If $H_0$
  is true, it is approximately standard normal, because $n=100$ is large.
- **Step 4. Observed value.** First the denominator:
  $\sigma/\sqrt n=8.44/\sqrt{100}=0.844$. Then
  $z_{obs}=(12.76-15)/0.844=-2.24/0.844=-2.65$.

The sample mean lies 2.65 standard deviations of the sample mean below the
advisor's figure. Steps 3 and 5, the critical value and the conclusion,
follow in Lecture 10.

```{r}
#| label: b09-test-statistic
mu_0    <- 15
sd_xbar <- sigma_equity / sqrt(n)    # standard deviation of the sample mean
z_obs   <- (x_bar - mu_0) / sd_xbar
t_obs   <- (x_bar - mu_0) / se       # the same with s in place of sigma

c(distance = x_bar - mu_0, sd_of_sample_mean = sd_xbar, z_obs = z_obs)
c(standard_error = se, t_obs = t_obs)
```
:::

In a real study $\sigma$ is unknown. Then $S$ replaces $\sigma$, as it did
in the interval, and the test statistic is

$$
T=\frac{\bar X-\mu_0}{S/\sqrt n},
$$

which has the $t_{n-1}$ distribution if $H_0$ is true. Its observed value
here is $t_{obs}=(12.76-15)/0.771=-2.91$. This is the value `t = -2.9061`
in the R output at the start of this lecture. Lecture 10 first completes
the test with $z_{obs}=-2.65$ and then moves to this version. In the exam, a
test for a mean always uses $s$ and $t$. The figure below shows
$z_{obs}=-2.65$.

```{r}
#| label: fig-b09-observed-z
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The distribution of the test statistic Z if the null hypothesis is true: the standard normal curve. The orange line marks the observed value −2.65, far out in the left tail. The open question is how far from zero the observed value must lie before the null hypothesis is rejected."
library(ggplot2)

z_observed <- -2.65
z_ref <- data.frame(z = seq(-4.5, 4.5, length.out = 900))
z_ref$density <- dnorm(z_ref$z)

ggplot(z_ref, aes(z, density)) +
  geom_line(colour = "#17324D", linewidth = 1.2) +
  geom_vline(xintercept = z_observed, colour = "#D55E00", linewidth = 1) +
  annotate("text", x = z_observed - 0.12, y = 0.31,
           label = "observed value", colour = "#D55E00", size = 3.6,
           hjust = 1) +
  annotate("text", x = z_observed - 0.12, y = 0.28,
           label = "z[obs] == -2.65", parse = TRUE, colour = "#D55E00",
           size = 3.8, hjust = 1) +
  annotate("text", x = 0, y = 0.425,
           label = 'paste("if ", H[0], " is true, Z follows N(0, 1)")',
           parse = TRUE, colour = "#17324D", size = 3.8) +
  annotate("text", x = 2.9, y = 0.30,
           label = "how far from zero\nis far enough?",
           colour = "#17324D", size = 3.6, lineheight = 0.95) +
  scale_x_continuous(breaks = -4:4) +
  scale_y_continuous(limits = c(0, 0.44),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Value of the test statistic Z", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

@fig-b09-observed-z shows the situation. If the advisor is right, $Z$
follows the standard normal curve, and $-2.65$ lies far out in its left
tail. But how far is far enough? A value of $-0.5$ would clearly not be
evidence against $H_0$. A value of $-10$ clearly would be. Somewhere in
between, we must draw a line. We cannot decide where to draw it by looking
at the sample. We decide it in advance, based on the probability of a wrong
conclusion. That is the subject of Lecture 10.

Here is a first hint of what follows. The 95% interval for $\mu$ from the
Shark Tank case is $(11.23,\ 14.29)$, and it does not contain 15. So, in
practice, the interval has already answered the question.

::: {.callout-caution collapse="true"}
## Try it: state the hypotheses

Try this with your neighbour first. For each question, write $H_0$ and
$H_1$ for a two-sided test, and say which hypothesis is the claim to be
established.

1. Does the proportion of pitches that end in a deal differ from one half?
2. A commentator claims that sharks close deals in 70% of pitches.
3. An advisor's handbook says the typical equity offered is 15%. A
   founder wants to check the handbook.

::: {.callout-note collapse="true"}
## Check your answer

1. $H_0:p=0.5$ versus $H_1:p\ne0.5$. The claim to be established is $H_1$.
2. $H_0:p=0.7$ versus $H_1:p\ne0.7$. The commentator's figure is the tested
   value $p_0=0.7$. The claim to be established is $H_1$: the deal
   proportion differs from 0.7.
3. $H_0:\mu=15$ versus $H_1:\mu\ne15$. The claim to be established is
   $H_1$. The handbook could be wrong in either direction.
:::
:::

::: {.caveat-box}
<span class="label">What an interval can and cannot tell you</span>

- Precision is not importance. An interval that excludes 15 says that the
  mean is estimated precisely enough to tell it apart from 15. It does not say
  that a difference of 2.24 points matters for a decision.
- An interval is not a claim about cause. Nothing here says why offers lie
  below 15%.
- A narrow interval does not correct a selection problem. It describes the
  recorded pitches, not founders in general.

**What you can do with it:** treat both ends of the interval as possible
true values, and ask whether your decision would be the same at each end.
:::

### Answering the driving question

For the recorded pitches, the sampling frame, the evidence says yes. The random sample
of 100 pitches gives a mean equity offered of 12.76% with a standard
error of 0.771 percentage points. The 95% confidence interval is
$(11.23,\ 14.29)$%. It does not contain 15, so at the 5% level we can
conclude that the mean differs from 15% and is lower. The test
statistic looks at the same distance from the other side. With
$\sigma=8.44$ known, $z_{obs}=-2.65$: the sample mean lies 2.65 standard
deviations of the sample mean below the advisor's figure. Lecture 10 sets
how far from zero is far enough and turns the test statistic into a formal
decision. Here the answer can also be checked, because the file is the whole
population: $\mu=13.21$%, which lies inside the interval and below
15. Whether a difference of about 2 points matters commercially, and what
caused it, are separate questions that no interval can answer.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

From 25 orders, the average order value is 9.60 euros with a standard deviation of 2 euros. Order values are roughly normal. You need 9 euros per order to break even on a kitchen lease. The standard error is 2/5 = 0.40 euros. With the supplied value $t_{0.975,\,24}=2.064$, the 95% confidence interval is 9.60 ± 2.064 × 0.40, from 8.77 to 10.43 euros. The break-even value of 9 euros lies inside the interval, so you cannot conclude that the mean order value differs from 9 euros. The decision: do not sign the lease yet. With 100 orders the standard error falls to 0.20 and the critical value to 1.984. If the mean stays at 9.60 and the standard deviation at 2, the interval then runs from 9.20 to 10.00 and excludes 9. You collect orders for three more weeks and decide then.
:::

**Next:** We can now report an estimate with a confidence interval, for a
mean and for a proportion. We have also stated two hypotheses and measured
with a test statistic how far the estimate lies from the tested value.
Lecture 10 names the two ways a decision can be wrong and uses them to set
the line between "reject" and "do not reject" with a critical value. It
writes every test in five steps, reads the same decision from a p-value, and
shows what makes wrong decisions rarer. It ends with a first look at a
regression table.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 80 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Why does replacing $\sigma$ with the sample SD $S$ require a t
   distribution in place of the normal distribution?
2. Why does the t distribution for one sample mean have $n-1$ degrees of
   freedom?
3. What are the three pieces of a confidence interval?
4. What is the difference between the interval estimator $(L,U)$ and the
   confidence interval $(l,u)$?
5. Which critical value and which standard error does the interval for a
   proportion use?
6. What do $H_0:\mu=15$ and $H_1:\mu\ne15$ say? Which of the two conclusions
   of a test is the strong one?
7. What is the confidence level a property of: the interval you computed or
   the procedure?

::: {.callout-note collapse="true"}
## Check your answers

1. $S$ is itself estimated and varies from sample to sample. The heavier
   tails of the t distribution represent that extra uncertainty.
2. Before we estimate the spread, we estimate the sample mean. This sets
   one condition: the deviations must sum to zero. So only $n-1$ deviations
   can vary freely.
3. Estimate, critical value, and standard error.
4. $(L,U)$ has random bounds. It captures the parameter with probability
   $1-\alpha$ before the data are drawn. $(l,u)$ is the pair of numbers
   calculated from one sample. It either contains the parameter or it does
   not.
5. The critical value $z_{1-\alpha/2}$, for example 1.96 for 95%, and the
   standard error $\sqrt{\hat p(1-\hat p)/n}$.
6. The null hypothesis states that the population mean equals 15. The
   alternative states that it differs from 15, in either direction. "Reject
   $H_0$" is the strong conclusion. "Do not reject $H_0$" does not show that
   $H_0$ is true.
7. The procedure. With a 95% level, about 95% of intervals built this way
   from repeated valid samples contain the fixed parameter.
:::

### Finish a half-worked example (10 minutes)

Re-read the hand-worked interval (@exm-b09-hand-interval). Then complete
the same calculation with new invented numbers. A random sample of $n=25$
customers gives a mean satisfaction of $\bar x = 42$ points with sample SD
$s = 10$ points. Assume that the scores in the population are roughly
normal. The benchmark is 40 points. Construct a 95% confidence interval for
the population mean $\mu$. Supplied value: $t_{0.975,\,24}=2.064$.

*Parts 1 and 2 (given).* The formula is
$\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used because the
population is roughly normal. The given values are $\bar x=42$, $s=10$,
$n=25$, $df=24$, the 95% level, and $t_{0.975,\,24}=2.064$.

*Part 3 (you).* Calculate the standard error and the margin of error.

*Part 4 (you).* Write the interval.

*Part 5 (you).* Say whether the benchmark of 40 points lies inside the
interval, and what you can conclude.

::: {.callout-note collapse="true"}
## Check your answer

Part 3: the standard error is $10/\sqrt{25}=2$ points. The margin of error
is $2.064\times 2=4.13$ points. Part 4: the interval is
$(42-4.13,\ 42+4.13)=(37.87,\ 46.13)$ points. Part 5: the benchmark of 40
points lies inside the interval. At the 5% level we cannot conclude that the
mean satisfaction differs from 40 points. This does not show that the mean
equals 40: values up to about 46 points are also compatible with the sample.
:::

### Practice (20 minutes)

The numbers in tasks 1 to 4 are invented. Write every interval in the five
parts.

1. An app founder measures the length of a session for a random sample of
   $n=121$ users. The mean is $\bar x=18.4$ minutes and the sample SD is
   $s=5.5$ minutes. The pitch deck claims a mean of 20 minutes. Construct a
   90% confidence interval for the mean session length, and say what it
   shows about the claim. Supplied value: $t_{0.95,\,120}=1.658$.
2. A meal-box start-up asks a random sample of 200 customers whether they
   would recommend the meal box to a friend. 58 say yes. Construct a 95%
   confidence interval for the proportion of all customers who would
   recommend the meal box. The sector benchmark is 25%. What do you conclude? Use
   $z_{0.975}=1.96$.
3. A bike-repair start-up wants to estimate its mean repair time to within
   1.5 minutes, at the 90% confidence level. An earlier study gave an SD of
   about 8 minutes. How many repairs must be sampled? How many at the 95%
   level? Use $z_{0.95}=1.645$ and $z_{0.975}=1.96$.
4. A co-working space advertises that members use a desk for 6 hours a day
   on average. A random sample of $n=64$ members gives a mean of
   $\bar x=5.4$ hours with sample SD $s=2.0$ hours. Part a: state $H_0$ and
   $H_1$ for a two-sided test of the advertised value. Part b: write the
   test statistic and calculate its observed value. Say what the value
   measures. Part c: for task 2, state $H_0$ and $H_1$ for a two-sided test
   of the sector benchmark.
5. Explain why an interval that excludes a benchmark does not show that the
   difference is important in practice, or that it has a particular cause.

::: {.callout-note collapse="true"}
## Check your answers

1. Part 1: $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$, allowed because the
   sample is random and $n=121$ is large. Part 2: $\bar x=18.4$, $s=5.5$,
   $n=121$, $df=120$, level 90%, so $\alpha=0.10$ and
   $t_{0.95,\,120}=1.658$. Part 3: the standard error is
   $5.5/\sqrt{121}=5.5/11=0.5$ minutes, and the margin of error is
   $1.658\times0.5=0.83$ minutes. Part 4: $(17.57,\ 19.23)$ minutes. Part 5:
   the claimed 20 minutes lies outside the interval. At the 10% level we can
   conclude that the mean session length differs from 20 minutes, and the
   interval shows that it is shorter.
2. Part 1: $\widehat P\pm z_{1-\alpha/2}\sqrt{\widehat P(1-\widehat P)/n}$,
   allowed because $n\hat p=58$ and $n(1-\hat p)=142$ are both at least 10.
   Part 2: 58 customers who say yes, $n=200$, $\hat p=58/200=0.29$, level
   95%, $z_{0.975}=1.96$. Part 3: the standard error is
   $\sqrt{0.29\times0.71/200}=0.032$, and the margin of error is
   $1.96\times0.032=0.063$. Part 4: $(0.227,\ 0.353)$. Part 5: the benchmark
   0.25 lies inside the interval. At the 5% level we cannot conclude that
   the proportion who would recommend the box differs from the sector
   benchmark.
3. At the 90% level: $n\ge(1.645\times8/1.5)^2=8.773^2\approx76.97$, so 77
   repairs. At the 95% level:
   $n\ge(1.96\times8/1.5)^2=10.453^2\approx109.3$, so 110 repairs. More
   confidence at the same margin needs more observations. The value 8 is a
   planning value for $\sigma$.
4. Part a. Step 1: $H_0:\mu=6$ versus $H_1:\mu\ne6$, where $\mu$ is the
   mean daily desk use of all members, in hours. Part b. Step 2:
   $T=(\bar X-\mu_0)/(S/\sqrt n)$, which has the $t_{63}$ distribution if
   $H_0$ is true ($n=64$ is large). Step 4: the standard error is
   $2.0/\sqrt{64}=0.25$ hours, so $t_{obs}=(5.4-6)/0.25=-2.40$. The sample
   mean lies 2.4 standard errors below the advertised value. Whether that is
   far enough to reject $H_0$ is decided in Lecture 10. Part c.
   $H_0:p=0.25$ versus $H_1:p\ne0.25$, where $p$ is the proportion of all
   customers who would recommend the box.
5. An interval that excludes the benchmark says that the estimate is precise
   enough to tell the parameter apart from the benchmark. Whether the difference
   is important depends on its size and on the decision. A causal claim
   needs a suitable study design. A narrow interval gives neither.
:::

### Worked exam interpretation: what 95% confidence means (5 minutes)

[Shark Tank data]{.shark-tag} **Exam-style question.** Rewrite this claim correctly: "There is a 95% chance that the population
mean is in my interval $(11.23, 14.29)$."

::: {.callout-note collapse="true"}
## Check a model answer

"The interval $(11.23, 14.29)$ was produced by a procedure that captures
the fixed population mean in about 95% of repeated valid samples." The
parameter is fixed. It is the interval endpoints that vary across samples,
as in @fig-b09-coverage.
:::

### Review: one question each from Lectures 8 and 4 (10 minutes)

1. *(Lecture 8.)* A random sample of $n=64$ customers has a sample SD of
   $s=24$ euros. Calculate the standard error of the sample mean. What
   happens to it if $n$ grows to 256 and $s$ stays the same?
2. [Shark Tank data]{.shark-tag} *(Lecture 4.)* In the Shark Tank file, $D$ is the event that a pitch
   ended with a deal on the show and $L$ the event that it is from the late
   seasons. From the 2×2 table of counts,
   $P(D\mid L)=505/734\approx 0.688$ while
   $P(L\mid D)=505/882\approx 0.573$. Why are these two numbers different
   even though both involve the same 505 records?

::: {.callout-note collapse="true"}
## Check your answers

1. The standard error is $24/\sqrt{64}=3$ euros. Four times the sample size
   halves it to 1.5 euros. This is the square-root rule.
2. They condition on different events. The first divides the 505 joint
   records by the 734 late-season records. The second divides the same 505
   by the 882 deal records. Changing the event after the bar changes the
   denominator.
:::

### Check the AI (5 minutes)

[Shark Tank data]{.shark-tag} You asked an AI assistant: "What does the 95% interval from 11.23 to 14.29 for the mean equity offered mean?" It answered:

> "It means that 95% of all pitches offer between 11.23% and 14.29% of the company."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

The interval is about the mean, not about single pitches. In the Shark Tank file, the equity offered (`equity_offered_pct`) in single pitches ranges from 1% to 100%. The correct reading: the interval was produced by a procedure that captures the mean equity offered in all recorded pitches in about 95% of samples.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Pick one number
of the venture that has an industry benchmark, for example the conversion
rate. Decide whether it is a mean $\mu$ or a proportion $p$. Sentence 1:
which sample could you collect, and what could you claim if the 95%
confidence interval does not contain the benchmark? Sentence 2: which
decision would change if the interval contains the benchmark?

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. Churn is the share of subscribers who cancel. "The sector benchmark for churn is 25% per term. From my roughly 120 subscribers I can estimate my churn proportion. A 95% interval that lies entirely below 0.25 would let me claim in my pitch that my customers stay longer than average. If the interval contained 0.25, I would stop presenting customer retention as my advantage and start my pitch with my margin instead."

:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to check whether your
parameter is defined on a population you can actually sample.
::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 9. Show the solution only after I send my answer."

[Continue to Tutorial 3: Normal models, sampling, and confidence intervals](t03-exam-prep.qmd).

[Formula sheet: Lecture 9 formulas](formula-sheet.qmd#formula-l9).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 15 and 16.

:::
