# Lecture 10: Testing decisions, errors, power, and a first look at regression {#b10}

```{r}
#| label: b10-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 10, you can</span>

- carry out a two-sided test for a mean and for a proportion in five steps,
  and write the conclusion in the words of the question
- explain the significance level, the critical value, and the rejection
  region of a test
- explain a p-value and use it for the test decision
- connect a confidence interval at level $1-\alpha$ to the two-sided test at
  level $\alpha$
- describe Type I error, Type II error, and power in context, and name what
  raises power
- separate statistical significance from size, practical importance, and
  causality
- read the test that is printed for a regression coefficient.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- carry out a two-sided test for a mean (with $s$ and $t$) and for a
  proportion in the five steps, and write the conclusion in the words of
  the question
- use the critical-value rule: reject $H_0$ when
  $|t_{obs}|\ge t_{1-\alpha/2,\,n-1}$, or when $|z_{obs}|\ge z_{1-\alpha/2}$
  for a proportion
- explain a p-value in one sentence and use it for the decision
- link the two-sided test at level $\alpha$ to the confidence interval at
  level $1-\alpha$
- describe a Type I and a Type II error in context and name what raises
  power
- read a regression row: estimate, standard error, t, p-value, interval.

**Enough to recognise:**

- the test for a mean with $\sigma$ known, which uses $z$
- the calculation of a power and the two power figures.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Launch the product or wait. Approve the loan or decline. Decisions made from data can be wrong in two directions, and the two mistakes usually cost very different amounts.</p>

How far from a claimed value must a sample lie before you reject the claim?
Every such decision can be wrong in two ways. Which error is which, and what
makes each of them rarer?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Review <a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to
Hypothesis Testing</a> and watch
<a href="https://www.youtube.com/watch?v=JNOWWJeZBTc">Examples Identifying
Type I and Type II Errors</a>. Focus on two-sided tests, decision errors,
and why "do not reject" is not "accept." All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

[Shark Tank data]{.shark-tag} **Warm up (8 minutes).** Lecture 9 ended with the test statistic for the
advisor's figure of 15%, $z_{obs}=-2.65$.

1. Write the two hypotheses of that test.
2. Say in one sentence what the number $-2.65$ measures.
3. Guess: if the advisor's figure were right, would a value this far from
   zero be common or rare?

::: {.callout-note collapse="true"}
## Check your preparation

$H_0:\mu=15$ against $H_1:\mu\ne15$, where $\mu$ is the mean of the equity
offered among all 1,441 pitches in the Shark Tank file. The statistic says
that the sample mean, 12.76, lies 2.65 standard deviations of the sample
mean below 15. Such a value would be rare: the central 95% of the standard
normal distribution lies between $-1.96$ and $1.96$. The lecture material
turns this into a decision rule.
:::

**Predict (2 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: a test ends with the decision "do not
reject $H_0$". Has the study shown that $H_0$ is true: yes, no, or
cannot tell? Choose one answer. You find the correct answer in the lecture
material.

Review [Mathematics Refresher: inequalities and
standardisation](../resources/math-refresher.html) if needed.

:::

::: {.course-phase .phase-in-class}

## Lecture material

Lecture 9 ended with a test statistic and no decision. This section sets
the rule for the decision, writes the test in five steps, and names the two
ways a decision can be wrong. It ends with the same test in a regression
table.

### Start with a founder's question

[Shark Tank data]{.shark-tag} An advisor said that founders offer 15% of their company on average. Your
sample of 100 pitches says 12.76%. Should you tell the advisor that the
number is wrong? And what does it cost you if you are the one who is wrong?

<p class="concept-video"><strong>Video explanations (review):</strong>
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to Hypothesis Testing</a> and
<a href="https://www.youtube.com/watch?v=JNOWWJeZBTc">Type I and Type II Errors</a>.</p>

### Where Lecture 9 stopped

[Shark Tank data]{.shark-tag} Lecture 9 set up the two-sided test of the advisor's figure. The parameter
$\mu$ is the mean of the equity offered (`equity_offered_pct`) among all
1,441 pitches in the Shark Tank file. The hypotheses are

$$
H_0:\mu=15
\qquad\text{versus}\qquad
H_1:\mu\ne15.
$$

The file is the whole population, so its standard deviation is known:
$\sigma=8.44$ percentage points. The code draws the same random sample of
100 pitches as in Lectures 7 to 9 (seed 123) and repeats the calculation of
Lecture 9:

```{r}
#| label: b10-sample-objects
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)

# The file is the whole population, so sigma divides by N (Lecture 8)
mu_equity <- mean(sharks$equity_offered_pct)
sigma_equity <- sqrt(mean((sharks$equity_offered_pct - mu_equity)^2))

# The random sample of 100 pitches
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

n <- nrow(pitch_sample)
x_bar <- mean(pitch_sample$equity_offered_pct)   # sample mean
mu_0 <- 15                                       # the advisor's figure

sd_x_bar <- sigma_equity / sqrt(n)   # standard deviation of the sample mean
z_obs <- (x_bar - mu_0) / sd_x_bar

c(sigma = sigma_equity, n = n, x_bar = x_bar,
  sd_of_sample_mean = sd_x_bar, z_obs = z_obs)
```

The observed value of the test statistic is

$$
z_{obs}=\frac{\bar x-\mu_0}{\sigma/\sqrt n}=\frac{12.76-15}{0.844}=-2.65.
$$

The sample mean lies 2.65 standard deviations of the sample mean below the
advisor's figure. If $H_0$ is true, the statistic $Z$ follows the standard
normal distribution, and $-2.65$ lies in the left tail of that curve. The
open question was where to draw the line: how far from zero must $z_{obs}$
be before we reject $H_0$?

This lecture answers the question in three parts. The first part gives the
form of the decision rule, with the place of the line still open. The second
part gives the reason that decides where the line lies. The third part
gives the resulting procedure, first with $\sigma$ known and then with
$\sigma$ replaced by $S$, as in every real study.

### A decision rule with an unknown constant

The alternative $H_1:\mu\ne15$ says that $\mu$ is either below 15 or
above 15. Evidence for $H_1$ is a sample mean far below 15 or far above 15.
The test statistic is then far from zero, on the negative or on the
positive side. So the test should reject $H_0$ when $|z_{obs}|$ is large. The bars
stand for the absolute value: the distance of $z_{obs}$ from zero, without
the sign. The decision rule has the form

$$
\text{reject }H_0\quad\text{if}\quad |z_{obs}|\ge k
$$

for some positive number $k$. The set of values of the test statistic for
which we reject $H_0$ is called the **rejection region**. Here it has two
parts: all values at or below $-k$, and all values at or above $k$.
@fig-b10-rejection-k shows them as the two tails of the curve.

```{r}
#| label: fig-b10-rejection-k
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The form of the two-sided decision rule, on the standard normal curve that the statistic follows if the null hypothesis is true. The null hypothesis is rejected when the observed value falls in one of the two orange tails, at or below −k or at or above k. The number k is not yet fixed."
library(ggplot2)

k_shown <- 2.3   # only for the drawing: k is not fixed yet
z_curve <- data.frame(z = seq(-4.5, 4.5, length.out = 1200))
z_curve$density <- dnorm(z_curve$z)

ggplot(z_curve, aes(z, density)) +
  geom_area(data = subset(z_curve, z <= -k_shown),
            fill = "#D55E00", alpha = 0.6) +
  geom_area(data = subset(z_curve, z >= k_shown),
            fill = "#D55E00", alpha = 0.6) +
  geom_vline(xintercept = c(-k_shown, k_shown), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_line(colour = "#17324D", linewidth = 0.9) +
  annotate("text", x = c(-3.4, 3.4), y = 0.07, label = "reject~H[0]",
           parse = TRUE, colour = "#D55E00", size = 3.8) +
  annotate("text", x = 0, y = 0.12, label = "do~not~reject~H[0]",
           parse = TRUE, colour = "#17324D", size = 3.8) +
  scale_x_continuous(breaks = c(-k_shown, 0, k_shown),
                     labels = c("-k", "0", "k")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "z statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Any positive $k$ gives a rule. A small $k$ rejects $H_0$ for almost every
sample, even when $\mu$ really is 15. A large $k$ almost never rejects,
even when $\mu$ is far from 15. Neither rule is useful. The right value of
$k$ follows from the ways in which the decision can be wrong.

### How large must $k$ be? Type I error and the significance level

A test is a decision rule, and a decision can be wrong in two ways:

| Reality | Reject $H_0$ | Do not reject $H_0$ |
|---|---|---|
| $H_0$ is true | Type I error | Correct decision |
| $H_1$ is true | Correct decision | Type II error |

After the decision "do not reject $H_0$", only a Type II error is possible.
After the decision "reject $H_0$", only a Type I error is possible.

::: {.definition-box}
<span class="label">Definition</span>

**Type I error and Type II error.** A Type I error is rejecting $H_0$ when
$H_0$ is true. It is a false alarm: you claim a difference that is not
there. A Type II error is not rejecting $H_0$ when $H_1$ is true. It is a
missed difference.
:::

In the advisor's example, a Type I error would be to conclude that the mean
equity offered differs from 15% when it is exactly 15%. A Type II error
would be to miss that the figure is wrong, for example because the sample
is small or because the true mean is close to 15.

Both errors are unwanted. With a fixed sample size you cannot make both of
them small at the same time. A smaller rejection region gives fewer false
alarms but more missed differences. A larger rejection region does the
opposite. The standard solution is to control the Type I error. We build
the test so that the probability of a false alarm is at most a small number
that we choose in advance.

::: {.definition-box}
<span class="label">Definition</span>

**Significance level $\alpha$.** The largest probability of a Type I error
that the test allows:
$P(\text{reject }H_0\mid H_0\text{ true})\le\alpha$. The standard choice is
$\alpha=0.05$. Other common choices are $\alpha=0.10$ and $\alpha=0.01$.
You fix the level before you look at the data.
:::

Why control the Type I error and not the Type II error? A Type I error
claims a discovery that is not there. A method that produces false
discoveries at an unknown rate is of little use. With $\alpha$ fixed, every
rejection comes with a known risk. Over many tests, a test at
$\alpha=0.05$ rejects at most 5% of the null hypotheses that are true. This
makes "reject $H_0$" the strong conclusion. The Type II error comes back
later in this lecture, in the section on power. It is the reason why "do not reject $H_0$" is the
weak conclusion.

Now we can choose $k$. If $H_0$ is true, $Z$ follows the standard normal
distribution. This holds exactly for a normal population and approximately
for a large sample, by the central limit theorem of Lecture 8. With the
rule "reject $H_0$ if $|Z|\ge k$", the probability of a Type I error is the
area of the two tails beyond $-k$ and $k$. The normal curve is symmetric,
so the two tails have the same area. The total must be $\alpha$, so each
tail gets $\alpha/2$. The value with an area of $\alpha/2$ to its right has
an area of $1-\alpha/2$ to its left. That is the quantile $z_{1-\alpha/2}$
of [Lecture 6](b06-continuous-normal.qmd):

$$
k=z_{1-\alpha/2}.
$$

::: {.callout-tip collapse="true"}
## Optional: the calculation of the constant in symbols

If $H_0$ is true, the probability of a Type I error with the rule "reject
$H_0$ if $|Z|\ge k$" is

$$
P(|Z|\ge k)=P(Z\le-k)+P(Z\ge k)=2\,P(Z\ge k).
$$

The last step uses the symmetry of the normal curve. Setting this
probability equal to $\alpha$ gives $P(Z\ge k)=\alpha/2$, so
$P(Z\le k)=1-\alpha/2$. By the definition of a quantile, $k=z_{1-\alpha/2}$.
:::

This number is called the **critical value** of the test. It is the same
value that the confidence interval with $\sigma$ known used in Lecture 9.
Each of the two cut-offs $-z_{1-\alpha/2}$ and $z_{1-\alpha/2}$ leaves a
probability of $\alpha/2$ in its tail. Together the rejection region has
probability exactly $\alpha$ when $H_0$ is true.

```{r}
#| label: b10-z-critical
# Critical values for alpha = 0.10, 0.05, and 0.01
qnorm(c(0.95, 0.975, 0.995))
```

For $\alpha=0.05$ the critical value is $z_{0.975}=1.960$. For
$\alpha=0.01$ it is $z_{0.995}=2.576$, and for $\alpha=0.10$ it is
$z_{0.95}=1.645$. @fig-b10-critical-value shows the rule for
$\alpha=0.05$.

```{r}
#| label: fig-b10-critical-value
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The constant is fixed. Each orange tail of the standard normal curve has probability α/2 = 0.025, so the rejection region beyond −1.96 and 1.96 has probability α = 0.05 when the null hypothesis is true. The observed value −2.65 (blue line) lies in the left part of the rejection region."
library(ggplot2)

z_critical <- 1.96
z_curve <- data.frame(z = seq(-4.5, 4.5, length.out = 1200))
z_curve$density <- dnorm(z_curve$z)

ggplot(z_curve, aes(z, density)) +
  geom_area(data = subset(z_curve, z <= -z_critical),
            fill = "#D55E00", alpha = 0.6) +
  geom_area(data = subset(z_curve, z >= z_critical),
            fill = "#D55E00", alpha = 0.6) +
  geom_vline(xintercept = c(-z_critical, z_critical), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_line(colour = "#17324D", linewidth = 0.9) +
  geom_vline(xintercept = z_obs, colour = "#0072B2", linewidth = 0.9) +
  annotate("text", x = z_obs - 0.1, y = 0.28, label = "z[obs] == -2.65",
           parse = TRUE, colour = "#0072B2", size = 3.8, hjust = 1) +
  annotate("text", x = c(-3.3, 3.3), y = 0.05, label = "alpha/2 == 0.025",
           parse = TRUE, colour = "#D55E00", size = 3.6) +
  annotate("text", x = 0, y = 0.12, label = "1 - alpha == 0.95",
           parse = TRUE, colour = "#17324D", size = 3.8) +
  scale_x_continuous(breaks = c(-4, -1.96, 0, 1.96, 4),
                     labels = c("-4", "-1.96", "0", "1.96", "4")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "z statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

### The five steps of a test, first with σ known

Everything that a test needs is now in place. Every test in this course is
written in the same five steps, and most tests in QM Intermediate follow
them too. The version below uses $\sigma$ and the standard normal
distribution. It is the first step only: it shows the logic in the simplest
case. The exam does not ask for this version. The version of the next
section, with $s$ and $t$, is the one that you use from then on.

For a two-sided test of a mean with $\sigma$ known:

1. Hypotheses. $H_0:\mu=\mu_0$ versus $H_1:\mu\ne\mu_0$. State the
   significance level $\alpha$.
2. Test statistic. $Z=\dfrac{\bar X-\mu_0}{\sigma/\sqrt n}$, which follows
   the standard normal distribution if $H_0$ is true (normal population, or
   large $n$).
3. Critical value and rejection region. Reject $H_0$ if
   $|z_{obs}|\ge z_{1-\alpha/2}$.
4. Observed value. Calculate $z_{obs}$ from the sample.
5. Conclusion. "Reject $H_0$" or "do not reject $H_0$", and what this means
   in the words of the question.

You fix steps 1 to 3 before you look at the data. You use the data only
from step 4.

::: {#exm-b10-advisor-z}
## Worked example: the advisor's figure with σ known, at the 5% level

[Shark Tank data]{.shark-tag} This version with $\sigma$ is only the first step. The exam does not ask for it.

1. **Hypotheses.** $H_0:\mu=15$ versus $H_1:\mu\ne15$, where $\mu$ is the
   mean equity offered among all pitches in the Shark Tank file. The
   significance level is $\alpha=0.05$.
2. **Test statistic.**
   $Z=(\bar X-15)/(8.44/\sqrt{100})=(\bar X-15)/0.844$. The equity offered
   is skewed, so the population is not normal. But $n=100$ is large, so by
   the central limit theorem $Z$ is approximately standard normal if $H_0$
   is true.
3. **Critical value and rejection region.** $z_{0.975}=1.96$. Reject $H_0$
   if $z_{obs}\le-1.96$ or $z_{obs}\ge1.96$.
4. **Observed value.** The standard deviation of the sample mean is
   $8.44/\sqrt{100}=0.844$. So $z_{obs}=(12.76-15)/0.844=-2.65$.
5. **Conclusion.** $-2.65\le-1.96$, so the observed value lies in the
   rejection region. Reject $H_0$. At the 5% level we can conclude that the
   mean equity offered differs from 15%. The negative sign of $z_{obs}$
   shows that it is below 15%.

At $\alpha=0.01$ the critical value is $z_{0.995}=2.576$. Since
$2.65\ge2.576$, we reject $H_0$ at the 1% level as well, but only by a
small margin.
:::

### In practice σ is unknown

[Shark Tank data]{.shark-tag} The test above used the population standard deviation $\sigma$. It is known
here only because the Shark Tank file is the whole population. A real study
has a sample and nothing else. It does not know $\sigma$, just as it does
not know $\mu$. The solution is the same as for the interval in Lecture 9:
replace $\sigma$ by the sample standard deviation $S$. The denominator is
then the standard error $S/\sqrt n$, and the test statistic becomes

$$
T=\frac{\bar X-\mu_0}{S/\sqrt n}.
$$

Lecture 9 showed what this change does to the distribution. There are two
situations.

- The population is normal. Then $T$ follows exactly the t distribution
  with $n-1$ degrees of freedom, written $t_{n-1}$, if $H_0$ is true. Its
  tails are heavier than those of the standard normal curve, because $S$
  varies from sample to sample. We find the critical value as before, now
  on the t curve: $k=t_{1-\alpha/2,\,n-1}$. This is the critical value of
  the t interval.
- The population is not normal, but $n$ is large. Then $T$ is approximately
  standard normal if $H_0$ is true. For a large $n$ the $t_{n-1}$ curve is
  almost the same as the standard normal curve. So we can use the t
  critical value here too, and one rule covers both situations.

```{r}
#| label: b10-t-versus-z
# Critical values for alpha = 0.05 and alpha = 0.01, with n = 100
c(t_0.975 = qt(0.975, df = 99), z_0.975 = qnorm(0.975),
  t_0.995 = qt(0.995, df = 99), z_0.995 = qnorm(0.995))
```

At $n=100$ the two are close: $t_{0.975,\,99}=1.984$ against
$z_{0.975}=1.960$, and $t_{0.995,\,99}=2.626$ against $z_{0.995}=2.576$.

So the rule for tests is the rule for intervals: when $\sigma$ is unknown,
use $S$ and the $t_{n-1}$ distribution, whatever the sample size. The rule
is exact for a normal population and almost exact for a large sample. It
does not cover a small sample from a population that is not normal. In that
case there is no critical value that we can justify. This is why you check the
conditions before you run a test.

In the exam, a test for a mean always uses $s$ and $t$.

The optional box of
[Lecture 9](b09-t-confidence-intervals.qmd#b09-t-rule) shows why the t
value also works for a large sample from a population that is not normal.

The five steps stay the same. Only the statistic and the critical value
change. The box below gives the steps in the form that the exam asks for.
It also names the p-value and the test for a proportion. Both follow later
in this lecture.

::: {.exam-procedure}
<span class="label">Exam procedure: the five steps of a test</span>

Write these five steps, in this order.

1. **Hypotheses.** $H_0:\mu=\mu_0$ versus $H_1:\mu\ne\mu_0$ (or $p$ and
   $p_0$). Say in words what the parameter is. State the significance
   level $\alpha$.
2. **Test statistic.** The formula and its distribution if $H_0$ is true,
   with the condition. Mean: $T=(\bar X-\mu_0)/(S/\sqrt n)$, which has the
   $t_{n-1}$ distribution (normal population, or large $n$). Proportion:
   $Z=(\widehat P-p_0)/\sqrt{p_0(1-p_0)/n}$, approximately $N(0,1)$ if
   $np_0\ge10$ and $n(1-p_0)\ge10$.
3. **Critical value and rejection region.** $t_{1-\alpha/2,\,n-1}$ or
   $z_{1-\alpha/2}$, from the question or the formula sheet. Reject $H_0$
   if the observed value is at or below minus the critical value, or at or
   above the critical value.
4. **Observed value.** Calculate the denominator first, then $t_{obs}$ or
   $z_{obs}$.
5. **Conclusion.** Say whether the observed value lies in the rejection
   region. Write "reject $H_0$" or "do not reject $H_0$", never "accept
   $H_0$". Then one sentence in the words of the question: "At the
   $\alpha$ level we can (cannot) conclude that ...". If you reject, add
   the direction from the sign of the observed value.

The p-value, when the question asks for it, comes after step 5:
$2\,P(T\ge|t_{obs}|)$ or $2\,P(Z\ge|z_{obs}|)$. It gives the same decision:
$p\le\alpha$ exactly when the observed value lies in the rejection region.

<p class="exam-note">In the exam, the question asks you to give all five
steps of the testing procedure. Number them 1 to 5 and write the formula of
the statistic and the rejection region before the numbers.</p>
:::

Now test the advisor's figure again, using only what the sample gives. The code
calculates the sample standard deviation, the standard error, the observed
value, and the critical values:

```{r}
#| label: b10-critical-value
s <- sd(pitch_sample$equity_offered_pct)   # sample standard deviation
se <- s / sqrt(n)                          # standard error of the sample mean
df <- n - 1
t_obs <- (x_bar - mu_0) / se

c(s = s, standard_error = se, df = df, t_obs = t_obs,
  critical_5_percent = qt(0.975, df = df),
  critical_1_percent = qt(0.995, df = df))
```

::: {#exm-b10-advisor-t}
## Worked example: the advisor's figure with s and t, at the 5% level

[Shark Tank data]{.shark-tag} The sample gives $n=100$, $\bar x=12.76$, and $s=7.71$.

1. **Hypotheses.** $H_0:\mu=15$ versus $H_1:\mu\ne15$, where $\mu$ is the
   mean equity offered among all pitches in the Shark Tank file. The
   significance level is $\alpha=0.05$.
2. **Test statistic.** $T=(\bar X-15)/(S/\sqrt{100})$. The equity offered
   is strongly skewed, so the population is not normal. But $n=100$ is
   large, so $T$ has approximately the $t_{99}$ distribution if $H_0$ is
   true.
3. **Critical value and rejection region.** $t_{0.975,\,99}=1.984$. The
   formula sheet lists $df=100$ as the closest value, which gives the same
   number. Reject $H_0$ if $t_{obs}\le-1.984$ or $t_{obs}\ge1.984$.
4. **Observed value.** The standard error is $s/\sqrt n=7.71/10=0.771$. So
   $t_{obs}=(12.76-15)/0.771=-2.91$.
5. **Conclusion.** $-2.91\le-1.984$, so the observed value lies in the
   rejection region. Reject $H_0$. At the 5% level we can conclude that the
   mean equity offered differs from 15%. The negative sign of $t_{obs}$
   shows that it is below 15%.

At $\alpha=0.01$ the critical value is $t_{0.995,\,99}=2.626$. Since
$2.91\ge2.626$, we reject $H_0$ at the 1% level too.
:::

Two numbers changed between the two versions. The critical value moved from
1.96 to 1.984. This is the small price of using $S$. The statistic moved
from $-2.65$ to $-2.91$, because the standard deviation of this sample,
7.71, happens to be smaller than the 8.44 of the population. Another sample
would have another $s$ and another $t_{obs}$. The heavier tails of the t
curve take this variation in $S$ into account.

### The confidence interval and the test give the same decision

[Shark Tank data]{.shark-tag} The critical value of the test is the value that the interval uses, and the
standard error is the same. So the two procedures always agree for a mean.
The test rejects $H_0$ when the distance between $\bar x$ and $\mu_0$ is at
least the critical value times the standard error. This product is the
margin of error of the interval. And $\bar x$ lies at least one margin of
error away from $\mu_0$ exactly when $\mu_0$ is not inside the interval.

For the two-sided test of a mean:

- reject $H_0:\mu=\mu_0$ at level $\alpha$ exactly when the
  $100(1-\alpha)\%$ confidence interval does not contain $\mu_0$
- do not reject $H_0$ exactly when the interval contains $\mu_0$.

```{r}
#| label: b10-interval-check
# The 95% confidence interval of Lecture 9
x_bar + c(-1, 1) * qt(0.975, df = df) * se
```

The 95% interval $(11.23,\,14.29)$ does not contain 15, and the test at the
5% level rejects. The interval says more than the decision. It shows the
direction of the difference, its plausible size, and the precision of the
estimate. The decision only says that 15 is excluded. In this course and
in QM Intermediate, always report the estimate and its interval together
with the test decision.

::: {.callout-tip collapse="true"}
## Optional: the link between test and interval in symbols

The test rejects $H_0$ exactly when

$$
\left|\frac{\bar x-\mu_0}{s/\sqrt n}\right|\ge t_{1-\alpha/2,\,n-1}.
$$

Multiply both sides by $s/\sqrt n$:

$$
|\bar x-\mu_0|\ge t_{1-\alpha/2,\,n-1}\,\frac{s}{\sqrt n}.
$$

The right-hand side is the margin of error. The distance between $\bar x$
and $\mu_0$ is at least the margin exactly when $\mu_0$ lies at an end of
the interval or outside it:

$$
\mu_0\le\bar x-t_{1-\alpha/2,\,n-1}\,\frac{s}{\sqrt n}
\qquad\text{or}\qquad
\mu_0\ge\bar x+t_{1-\alpha/2,\,n-1}\,\frac{s}{\sqrt n}.
$$
:::

::: {.exam-procedure}
<span class="label">Exam procedure: a test question answered with an interval</span>

Some questions ask whether a value $\mu_0$ is plausible and tell you to
use a confidence interval. Then construct the interval in its five parts
and conclude from where $\mu_0$ lies. For a mean, this gives the same
decision as the two-sided test at level $\alpha$. For a proportion the
two can differ slightly, because the interval uses $\hat p$ and the test
uses $p_0$. Use the procedure that the question names.
:::

::: {.callout-caution collapse="true"}
## Try it: read the decision directly from an interval

Try this with your neighbour first. An invented example. A colleague reports
a 95% confidence interval for a population mean: $(13.6,\,18.8)$%.
The value under test is $\mu_0=15$. The report gives no t statistic and no
p-value.

*Given.* The two-sided t test at the 5% level rejects
$H_0:\mu=15$ exactly when the 95% interval does not contain 15.

*Question 1 (you).* Does this interval contain 15? State the test decision
without a calculation.

*Question 2 (you).* What may you conclude about $\mu$, and what may you *not*
conclude?

::: {.callout-note collapse="true"}
## Check your answer

The value 15 lies inside $(13.6,\,18.8)$, so the test at the 5% level does
not reject $H_0:\mu=15$. The data are compatible with $\mu=15$, and also
with values as low as 13.6 and as high as 18.8. You may not conclude that
$\mu$ is 15. "Do not reject" is not "accept". The section on power below
gives the reason.
:::
:::

### The p-value: a second reading of the same decision

[Shark Tank data]{.shark-tag} The advisor's test rejected $H_0$ at $\alpha=0.05$ and again at
$\alpha=0.01$. There is a pattern here. If a test rejects $H_0$ at one
level $\alpha$, it also rejects $H_0$ at every larger level, because a
larger $\alpha$ gives a larger rejection region. So for a given sample there
is a smallest significance level at which the test still rejects $H_0$. This number
measures the evidence against $H_0$ on a continuous scale.

::: {.definition-box}
<span class="label">Definition</span>

**p-value.** The smallest significance level at which $H_0$ is rejected
with the observed data. For a two-sided test it is the probability,
calculated as if $H_0$ were true, of a test statistic at least as far from
zero as the observed one, in either direction:
$p=P(|T|\ge|t_{obs}|)=2\,P(T\ge|t_{obs}|)$, where $T$ follows the $t_{n-1}$
distribution.
:::

The two descriptions give the same number. At level $\alpha$, the test
rejects when $|t_{obs}|$ is at least the critical value. Suppose that
$|t_{obs}|$ lies exactly on the critical value. Then the two tails beyond
$-|t_{obs}|$ and $|t_{obs}|$ have a total area of exactly $\alpha$. With
any smaller $\alpha$, the critical value moves beyond $|t_{obs}|$ and the
test no longer rejects. So the smallest $\alpha$ that rejects is the area of
the two tails beyond the observed value.

In R, `pt()` gives the area under a t curve to the left of a value, and
`pnorm()` does the same for the standard normal curve. The area of the two
tails is twice the area to the left of $-|t_{obs}|$:

```{r}
#| label: b10-p-value
p_value <- 2 * pt(-abs(t_obs), df = df)           # test with s and t
p_value_sigma <- 2 * pnorm(-abs(z_obs))           # first step, sigma known

c(t_obs = t_obs, p_value = p_value)
c(z_obs = z_obs, p_value_sigma_known = p_value_sigma)
```

For the advisor's test with $s$, the p-value is 0.0045.
@fig-b10-two-sided-p shows it as an area.

```{r}
#| label: fig-b10-two-sided-p
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The p-value of the advisor's test as an area under the t curve with 99 degrees of freedom, the distribution of the statistic if the null hypothesis is true. Dashed lines mark −2.91 and 2.91, the values as far from zero as the observed one. Each blue tail holds half of the p-value. Together the two tails have area 0.0045."
library(ggplot2)

t_curve <- data.frame(x = seq(-4.5, 4.5, length.out = 1200))
t_curve$density <- dt(t_curve$x, df = df)

ggplot(t_curve, aes(x, density)) +
  geom_area(data = subset(t_curve, x <= -abs(t_obs)),
            fill = "#0072B2", alpha = 0.75) +
  geom_area(data = subset(t_curve, x >= abs(t_obs)),
            fill = "#0072B2", alpha = 0.75) +
  geom_vline(xintercept = c(-abs(t_obs), abs(t_obs)),
             linetype = "dashed", colour = "#5B6976", linewidth = 0.5) +
  geom_line(colour = "#17324D", linewidth = 0.9) +
  annotate("text", x = c(-3.6, 3.6), y = 0.035, label = "p/2",
           colour = "#0072B2", size = 3.8) +
  annotate("text", x = 0, y = 0.12,
           label = "the two tails together:\np-value = 0.0045",
           colour = "#17324D", size = 3.8, lineheight = 1) +
  scale_x_continuous(breaks = c(-4, -abs(t_obs), 0, abs(t_obs), 4),
                     labels = c("-4", "-2.91", "0", "2.91", "4")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "t statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

The curve in @fig-b10-two-sided-p is not data. It is the distribution of
the t statistic if $H_0$ were true. The vertical axis is density, so a
probability is an area. The two blue tails beyond the dashed lines are very
small. If the mean equity offered were 15%, a sample mean this far from 15
would occur in fewer than 1 of 200 random samples.

The p-value is a second way to state the decision. It is not a different
decision:

$$
|t_{obs}|\ge t_{1-\alpha/2,\,n-1}
\qquad\text{exactly when}\qquad
p\le\alpha.
$$

For the advisor's test, $p=0.0045$. Since $0.0045\le0.05$, reject $H_0$ at
the 5% level. Since $0.0045\le0.01$, reject at the 1% level too. Since
$0.0045>0.001$, do not reject at the 0.1% level. With $\sigma$ known, the
same definition gives $p=2\,P(Z\ge2.65)=0.008$, with the same three
decisions.

The critical value gives the decision at one level $\alpha$. The p-value
gives the decision at every level at once, and it says how surprising the
sample would be if $H_0$ were true. The significance level and the p-value
are different concepts. The level $\alpha$ is a rule, fixed in advance. The
p-value is a result, calculated from the observed data.

### Reading the test from software

[Shark Tank data]{.shark-tag} Software prints the p-value and not the critical value. So the p-value is
the number that you will see most often. In R, the command `t.test()` runs
the test of this lecture on the sample of 100 pitches:

```{r}
#| label: b10-t-test-output
t.test(pitch_sample$equity_offered_pct, mu = 15)
```

Every line is something that you have calculated by hand. `t` is the
observed value of step 4, and `df` is $n-1$. `p-value` is the area of the
two tails. The line `alternative hypothesis: true mean is not equal to 15`
says that the test is two-sided with $\mu_0=15$. The interval is the 95%
confidence interval of Lecture 9, and `mean of x` is $\bar x$. You make the
decision yourself, by comparing the p-value with $\alpha$. Software does
not choose $\alpha$ for you.

### What a p-value is not

The p-value is the probability of data at least this far from $H_0$,
calculated as if $H_0$ were true. It is nothing else. The task below gives
four sentences that claim more than the p-value says. All four are wrong.

::: {.callout-caution collapse="true"}
## Try it: correct the four wrong p-value sentences

[Shark Tank data]{.shark-tag} Try this with your neighbour first. The four sentences are about the
advisor's test ($t_{obs}=-2.91$, $p=0.0045$). Each one is wrong. Write a
correct version of each sentence.

*Sentence 1 (you).* "There is only a 0.45% probability that the mean equity
offered really is 15%."

*Sentence 2 (you).* "The result has a 0.45% probability of being due to
chance."

*Sentence 3 (you).* "Such a tiny p-value shows that the mean equity offered
is far below 15%."

*Sentence 4 (you).* "The tiny p-value proves that founders have become more
careful in what they offer."

::: {.callout-note collapse="true"}
## Check your answers

1. A p-value is not the probability that $H_0$ is true. Correct: "If the
   mean equity offered were exactly 15%, a sample mean at least this far
   from 15 would occur in about 0.45% of random samples of 100 pitches."
2. The p-value is calculated under the assumption that $H_0$ is true and
   chance alone produced the difference. It is the probability of such data
   under that assumption. It is not the probability that the assumption
   holds. Correct: "If chance alone were at work, a difference this large
   would occur in about 0.45% of random samples."
3. A p-value measures how surprising the data are if $H_0$ is true. It
   does not measure the size of the difference. The size is the estimate, 12.76%,
   with the interval $(11.23,\,14.29)$ around it. A tiny p-value can go
   together with a tiny difference when $n$ is very large. Correct: "The
   mean equity offered is estimated at 12.76%, plausibly between 11.2% and
   14.3%."
4. A significant result says nothing about the cause. The test does not
   know how the data came about, and it compares one mean with one number.
   Correct: "The mean equity offered among the pitches in the file differs
   from 15%."
:::
:::

### Type II error and power

[Shark Tank data]{.shark-tag} The rule "reject $H_0$ if $|t_{obs}|\ge1.984$" keeps the probability of a
false alarm at 5%. It says nothing about the other mistake: not rejecting
$H_0$ when $H_1$ is true. How often does that happen? The answer depends on
how far the true value lies from $\mu_0$. So we calculate it for one given
alternative value.

::: {.definition-box}
<span class="label">Definition</span>

**Power.** The probability that the test rejects $H_0$ when a given
alternative value of the parameter is true:
$\text{power}=P(\text{reject }H_0\mid\text{that value is true})$. The
probability of a Type II error at that value is written $\beta$, and
$\text{power}=1-\beta$.
:::

The calculation below is an illustration. Read it to see where a power
comes from. You are not asked to calculate a power yourself.

The Shark Tank file makes the idea concrete, because the population is
known: $\mu=13.21$ and $\sigma=8.44$. A power is usually calculated with
$\sigma$ known, and here $\sigma$ really is known. The true mean is 13.21,
so the advisor's figure of 15 is wrong. How often would a random sample of
100 pitches lead the test at the 5% level to reject it? The test rejects
when $|z_{obs}|\ge1.96$, that is, when the sample mean is far enough from
15:

$$
\bar x\le15-1.96\times0.844=13.35
\qquad\text{or}\qquad
\bar x\ge15+1.96\times0.844=16.65.
$$

Since $\mu=13.21$, the sample mean $\bar X$ is approximately normal with
mean 13.21 and standard deviation 0.844 (Lecture 8). The code calculates
the probability that $\bar X$ falls at or below 13.35 or at or above 16.65:

```{r}
#| label: b10-power-calc
# Cut-offs of the 5% test for the sample mean, n = 100
lower_cut <- mu_0 - 1.96 * sd_x_bar
upper_cut <- mu_0 + 1.96 * sd_x_bar

# Probability of a sample mean beyond a cut-off when mu = 13.21
power_100 <- pnorm(lower_cut, mean = mu_equity, sd = sd_x_bar) +
  (1 - pnorm(upper_cut, mean = mu_equity, sd = sd_x_bar))

c(mu = mu_equity, lower_cut = lower_cut, upper_cut = upper_cut,
  z_at_lower_cut = (lower_cut - mu_equity) / sd_x_bar, power = power_100)
```

Standardising the lower cut-off with the unrounded numbers gives

$$
P(\bar X\le13.35)=P(Z\le0.16)\approx0.56.
$$

The probability of a sample mean at or above 16.65 is practically zero. So
the power is about 0.56. When the true mean is 13.21, a sample of 100
pitches rejects the advisor's figure in 56% of all samples and misses the
difference in 44%. The true difference is 1.79 percentage points, about
12% of the advisor's figure. Almost half of all samples miss it.
@fig-b10-errors-power shows the two situations.

```{r}
#| label: fig-b10-errors-power
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The two errors of the 5% test of the advisor's figure with σ = 8.44 known, on the scale of the sample mean. Left: if the null hypothesis is true, the sample mean follows the dark curve centred at 15, and the orange tails beyond the cut-offs 13.35 and 16.65 have total probability α = 0.05. Right: the true mean is 13.21, so the sample mean follows the blue curve. The green area is the power, about 0.56. The rest of the blue curve, between the cut-offs, is the Type II error probability β, about 0.44."
library(ggplot2)

mean_grid <- seq(10.5, 18.5, length.out = 1400)
panel_levels <- c("H[0]~is~true*':'~~mu==15", "The~truth~is~~mu==13.21")

mean_curves <- rbind(
  data.frame(x = mean_grid, density = dnorm(mean_grid, mu_0, sd_x_bar),
             panel = panel_levels[1]),
  data.frame(x = mean_grid, density = dnorm(mean_grid, mu_equity, sd_x_bar),
             panel = panel_levels[2])
)
mean_curves$panel <- factor(mean_curves$panel, levels = panel_levels)

rejection <- subset(mean_curves, x <= lower_cut | x >= upper_cut)
rejection$side <- ifelse(rejection$x < mu_0, "left", "right")

panel_labels <- data.frame(
  x = c(12.3, 17.7, 15, 11.45, 15.6),
  density = c(0.06, 0.06, 0.52, 0.42, 0.2),
  label = c("alpha/2", "alpha/2", "do~not~reject~H[0]",
            "power %~~% 0.56", "beta %~~% 0.44"),
  panel = factor(panel_levels[c(1, 1, 1, 2, 2)], levels = panel_levels)
)

ggplot(mean_curves, aes(x, density)) +
  geom_area(data = rejection, aes(group = side, fill = panel), alpha = 0.5) +
  geom_vline(xintercept = c(lower_cut, upper_cut), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_line(aes(colour = panel), linewidth = 0.9) +
  geom_text(data = panel_labels, aes(label = label), parse = TRUE,
            colour = "#17324D", size = 3.4) +
  facet_wrap(~ panel, scales = "fixed", labeller = label_parsed) +
  scale_fill_manual(values = setNames(c("#D55E00", "#009E73"), panel_levels),
                    guide = "none") +
  scale_colour_manual(values = setNames(c("#17324D", "#0072B2"),
                                        panel_levels),
                      guide = "none") +
  scale_x_continuous(breaks = c(11, lower_cut, 15, upper_cut, 18),
                     labels = c("11", "13.35", "15", "16.65", "18")) +
  scale_y_continuous(limits = c(0, 0.56),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "Sample mean of the equity offered (%)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Four factors raise the power of a test:

- A larger true difference. The further $\mu$ lies from $\mu_0$, the more
  of the curve in the right panel falls into the rejection region.
- A larger sample. The standard deviation of the sample mean falls with
  $\sqrt n$. Both curves become narrower and overlap less.
- Less variation in the population. A smaller $\sigma$ has the same effect
  as a larger $n$.
- A larger $\alpha$. A wider rejection region finds more true differences,
  but it also gives more false alarms. Power and the probability of a Type
  I error rise together. This trade-off cannot be avoided.

The code repeats the calculation for a sample of 400 pitches:

```{r}
#| label: b10-power-n400
sd_x_bar_400 <- sigma_equity / sqrt(400)
lower_cut_400 <- mu_0 - 1.96 * sd_x_bar_400
upper_cut_400 <- mu_0 + 1.96 * sd_x_bar_400
power_400 <- pnorm(lower_cut_400, mean = mu_equity, sd = sd_x_bar_400) +
  (1 - pnorm(upper_cut_400, mean = mu_equity, sd = sd_x_bar_400))

c(sd_of_sample_mean = sd_x_bar_400, lower_cut = lower_cut_400,
  power = power_400)
```

With $n=400$ the standard deviation of the sample mean halves to 0.422. The
lower cut-off moves to 14.17, and the power when $\mu=13.21$ rises to about
0.99. @fig-b10-power-curve shows the power for every true mean between 11
and 19.

```{r}
#| label: fig-b10-power-curve
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Power of the 5% two-sided test of the advisor's figure of 15 with σ = 8.44 known, for every true mean between 11 and 19 and for two sample sizes. At a true mean of 15 both curves are at α = 0.05: there the probability of rejecting is the probability of a Type I error. Away from 15 the curves rise, and faster for the larger sample. At the true mean 13.21 the power is 0.56 for n = 100 and 0.99 for n = 400."
library(ggplot2)

power_of_test <- function(mu, n) {
  sd_mean <- sigma_equity / sqrt(n)
  pnorm(mu_0 - 1.96 * sd_mean, mean = mu, sd = sd_mean) +
    1 - pnorm(mu_0 + 1.96 * sd_mean, mean = mu, sd = sd_mean)
}

power_grid <- data.frame(mu = rep(seq(11, 19, length.out = 400), times = 2),
                         n = rep(c(100, 400), each = 400))
power_grid$power <- power_of_test(power_grid$mu, power_grid$n)
power_grid$sample <- factor(paste("n =", power_grid$n))

power_points <- data.frame(mu = mu_equity, n = c(100, 400))
power_points$power <- power_of_test(power_points$mu, power_points$n)

ggplot(power_grid, aes(mu, power)) +
  geom_hline(yintercept = 0.05, linetype = "dotted", colour = "#5B6976") +
  geom_vline(xintercept = c(mu_equity, mu_0), linetype = "dashed",
             colour = "#C98B2E", linewidth = 0.6) +
  geom_line(aes(colour = sample), linewidth = 1) +
  geom_point(data = power_points, colour = "#17324D", size = 2.2) +
  annotate("text", x = 13.1, y = c(0.56, 0.955), label = c("0.56", "0.99"),
           colour = "#17324D", size = 3.6, hjust = 1) +
  annotate("text", x = 17.4, y = 0.62, label = "n = 100",
           colour = "#0072B2", size = 3.8, hjust = 0) +
  annotate("text", x = 16.8, y = 0.93, label = "n = 400",
           colour = "#009E73", size = 3.8, hjust = 0) +
  annotate("text", x = 19, y = 0.085, label = "alpha == 0.05", parse = TRUE,
           colour = "#5B6976", size = 3.4, hjust = 1) +
  annotate("text", x = 13.1, y = 1.06, label = "mu == 13.21", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 1) +
  annotate("text", x = 15.1, y = 1.06, label = "mu[0] == 15", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 0) +
  scale_colour_manual(values = c("n = 100" = "#0072B2",
                                 "n = 400" = "#009E73"),
                      guide = "none") +
  scale_x_continuous(breaks = 11:19) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1.09),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "True population mean of the equity offered (%)",
       y = expression("Probability of rejecting " * H[0])) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

This is why "do not reject $H_0$" is the weak conclusion. With a power of
0.56, almost half of the samples do not reject $H_0$, although the
advisor's figure is wrong by almost two percentage points. So a sample that does not
reject is weak evidence that the advisor is right. The test controls the
probability of a false alarm. It does not promise to detect a difference
that exists. This answers the prediction question from the preparation: a
study that ends with "do not reject $H_0$" has not shown that $H_0$ is
true. It is one more reason to report the estimate with its interval. The
interval shows how large a difference the sample could have missed.

### A test for a proportion

[Shark Tank data]{.shark-tag} The deal indicator (`deal_on_show`) is a 0/1 variable. The question "do
half of all pitches end with a deal?" is a test about the proportion $p$ of
pitches with a deal among all pitches in the file:

$$
H_0:p=p_0
\qquad\text{versus}\qquad
H_1:p\ne p_0,
$$

with $p_0=0.50$. The test statistic is built as for a mean. It is the
distance between the estimate and the value under $H_0$, divided by a
standard deviation. One feature makes the test for a proportion simpler.

The standard deviation of $\widehat P$ under $H_0$ is a known number. For a
0/1 variable the spread is not a separate parameter:
$\operatorname{SD}(\widehat P)=\sqrt{p(1-p)/n}$ depends on $p$ alone
(Lecture 8). If $H_0$ is true, then $p=p_0$, so

$$
\operatorname{SD}_0(\widehat P)=\sqrt{\frac{p_0(1-p_0)}{n}}.
$$

This number is calculated from $p_0$ and $n$. We do not estimate anything. There
is no $\sigma$ to replace by $S$, so there is no t distribution. If
$np_0\ge10$ and $n(1-p_0)\ge10$, the statistic

$$
Z=\frac{\widehat P-p_0}{\sqrt{p_0(1-p_0)/n}}
$$

is approximately standard normal when $H_0$ is true, by the central limit
theorem. The critical value is $z_{1-\alpha/2}$, which is 1.96 at the 5%
level for every sample size. Check the success-failure rule of
[Lecture 8](b08-sampling-distributions.qmd#b08-binomial) with $p_0$,
because we use the approximation under the assumption that $H_0$ is true.

Here the test differs from the interval for a proportion of Lecture 9. The
interval had to use $\hat p$ under the square root, because no value of $p$
was given. That is the standard error. The test has a given value, $p_0$,
and uses it. The test asks how surprising the sample would be if $H_0$ were
true, so the spread comes from $H_0$. The two numbers are close when
$\hat p$ is close to $p_0$. Because they are not equal, the interval and
the test for a proportion can sometimes disagree when the observed value
lies close to the critical value.

The five steps are those of the exam procedure above, with $Z$ and
$z_{1-\alpha/2}$. The code applies them to the sample of 100 pitches:

```{r}
#| label: b10-proportion-objects
phat <- mean(pitch_sample$deal_on_show)   # sample deal proportion
p0 <- 0.50                                # value under H0

c(n = n, deals = sum(pitch_sample$deal_on_show), phat = phat, p0 = p0)
```

```{r}
#| label: b10-proportion-test
sd_0 <- sqrt(p0 * (1 - p0) / n)   # standard deviation of P-hat under H0
z_obs_p <- (phat - p0) / sd_0
p_value_p <- 2 * pnorm(-abs(z_obs_p))

c(sd_under_H0 = sd_0, z_obs = z_obs_p, p_value = p_value_p)
```

```{r}
#| label: b10-proportion-interval
se_phat <- sqrt(phat * (1 - phat) / n)   # standard error of P-hat
proportion_ci <- phat + c(-1, 1) * 1.96 * se_phat

c(standard_error = se_phat,
  lower = proportion_ci[1], upper = proportion_ci[2])
```

::: {#exm-b10-deal-test}
## Worked example: the deal proportion in five steps

[Shark Tank data]{.shark-tag} In the sample, 64 of the 100 pitches end with a deal, so $\hat p=0.64$.
Does the deal proportion among all pitches in the file differ from one
half? Test at the 5% level.

1. **Hypotheses.** $H_0:p=0.50$ versus $H_1:p\ne0.50$, where $p$ is the
   proportion of pitches that end with a deal among all pitches in the
   Shark Tank file. The significance level is $\alpha=0.05$.
2. **Test statistic.** $Z=(\widehat P-0.50)/\sqrt{0.50\times0.50/100}$.
   Since $np_0=50$ and $n(1-p_0)=50$ are both at least 10, $Z$ is
   approximately $N(0,1)$ if $H_0$ is true.
3. **Critical value and rejection region.** $z_{0.975}=1.96$. Reject $H_0$
   if $z_{obs}\le-1.96$ or $z_{obs}\ge1.96$.
4. **Observed value.** The denominator is the standard deviation of
   $\widehat P$ under $H_0$: $\sqrt{0.50\times0.50/100}=0.05$. So
   $z_{obs}=(0.64-0.50)/0.05=2.80$.
5. **Conclusion.** $2.80\ge1.96$, so the observed value lies in the
   rejection region. Reject $H_0$. At the 5% level we can conclude that the
   deal proportion differs from 50%. The positive sign of $z_{obs}$ shows
   that it is above 50%.

**The p-value.** $p=2\,P(Z\ge2.80)=0.005$. If $H_0$ were true, a sample
proportion at least 14 percentage points away from one half, in either
direction, would occur in about 5 of 1,000 samples. Since $0.005\le0.05$,
the p-value gives the same decision.

**Report the interval too.** The 95% interval of Lecture 9 uses the
standard error $\sqrt{0.64\times0.36/100}=0.048$. It is
$0.64\pm1.96\times0.048=(0.546,\,0.734)$. Between about 55% and 73% of the
pitches end with a deal. The interval does not contain 0.50, in line with
the decision.
:::

In R, the command `prop.test()` runs this test. The argument
`correct = FALSE` switches off a small correction that is not part of this
course:

```{r}
#| label: b10-prop-test-output
prop.test(x = 64, n = 100, p = 0.50, correct = FALSE)
```

R reports the statistic as `X-squared = 7.84`. This is the square of
$z_{obs}$: $2.80^2=7.84$. The p-value is the same, 0.005. R builds its
interval with a slightly different method, so its ends, 0.542 and 0.727,
differ a little from those of Lecture 9. Both intervals are acceptable. By
hand, use the method of Lecture 9.

::: {.callout-tip collapse="true"}
## Optional: why the interval and the test for a proportion divide by different numbers

For a mean, the test and the interval use the same standard error,
$s/\sqrt n$, because $\sigma$ must be estimated in both. For a proportion
they do not use the same number. The interval uses the standard error
$\sqrt{\hat p(1-\hat p)/n}$, the only choice when no value of $p$ is given.
The test uses the standard deviation of $\widehat P$ under $H_0$,
$\sqrt{p_0(1-p_0)/n}$, because it judges the sample as if $p=p_0$ were
true. In the deal example these are 0.048 and 0.050. One consequence: for a
proportion, the rule "reject $H_0$ at level $\alpha$ exactly when the
interval does not contain $p_0$" holds only approximately. It can fail when
the observed value lies close to the critical value.
:::

The second example uses other numbers and a business claim.

::: {#exm-b10-launch-test}
## Worked example: a proportion test by hand

An invented example. A language-learning app claims that half of its
new users are still active after 30 days. In a random sample of 400 users
who signed up last year, 232 were still active after 30 days. Test the
claim at the 5% level.

1. **Hypotheses.** $H_0:p=0.50$ versus $H_1:p\ne0.50$, where $p$ is the
   proportion of users who are still active after 30 days among all new
   users of last year. The significance level is $\alpha=0.05$.
2. **Test statistic.** $Z=(\widehat P-0.50)/\sqrt{0.50\times0.50/400}$.
   Since $np_0=400\times0.50=200$ and $n(1-p_0)=200$ are both at least 10,
   $Z$ is approximately $N(0,1)$ if $H_0$ is true.
3. **Critical value and rejection region.** $z_{0.975}=1.96$. Reject $H_0$
   if $z_{obs}\le-1.96$ or $z_{obs}\ge1.96$.
4. **Observed value.** The denominator is
   $\sqrt{0.50\times0.50/400}=\sqrt{0.000625}=0.025$. The sample proportion
   is $\hat p=232/400=0.58$. So $z_{obs}=(0.58-0.50)/0.025=3.20$.
5. **Conclusion.** $3.20\ge1.96$, so the observed value lies in the
   rejection region. Reject $H_0$. At the 5% level we can conclude that the
   proportion of active users among all new users differs from 50%. The
   positive sign of $z_{obs}$ shows that it is above 50%.

**The p-value.** It comes after step 5 and is the area of the two tails
beyond 3.20:

```{r}
#| label: b10-launch-p-value
2 * pnorm(-3.20)
```

The p-value is about 0.0014. Since $0.0014\le0.05$, it gives the same
decision.

**Report the interval too.** A decision alone is not a report. The 95%
interval uses the standard error:
$0.58\pm1.96\sqrt{0.58\times0.42/400}=0.58\pm0.048=(0.532,\,0.628)$. It does
not contain 0.50, in line with the test.

Look again at step 4. The denominator was calculated with $p_0=0.50$, not
with $\hat p$. The test measures how surprising the sample would be if
$H_0$ were true, so the spread comes from the value under $H_0$.
:::

### Other tests with the same pattern

[Shark Tank data]{.shark-tag} The two tests of this course, a mean against $\mu_0$ and a proportion
against $p_0$, follow one pattern:

$$
\text{test statistic}
=\frac{\text{estimate}-\text{value under }H_0}{\text{standard error}}.
$$

For a proportion, the denominator is the standard deviation of
$\widehat P$ under $H_0$. The observed value is compared with a critical
value from the distribution that the statistic has if $H_0$ is true. Two
more tests follow this pattern. You will meet them in QM Intermediate. In
both, the value under $H_0$ is zero.

- A difference between two group means. Do pitches that end with a deal
  offer a different share of equity on average than pitches without a deal? With
  $\mu_1$ and $\mu_2$ the two population means, the hypotheses are
  $H_0:\mu_1-\mu_2=0$ versus $H_1:\mu_1-\mu_2\ne0$. The estimate is
  $\bar x_1-\bar x_2$. Its standard error combines the two sample
  variances. The statistic is the estimate divided by its standard error,
  and it is compared with a t critical value.
- A regression slope. Does the equity offered change with the season? With
  $\beta_1$ the slope of the line in the population, the hypotheses are
  $H_0:\beta_1=0$ versus $H_1:\beta_1\ne0$. The estimate is $\hat\beta_1$,
  and its standard error comes from the regression output. The statistic is
  $\hat\beta_1/\operatorname{SE}(\hat\beta_1)$, and it is compared with a t
  critical value. The preview below shows how to read such a row.

In both tests, "no difference" or "no association" is the null hypothesis
that a parameter is zero. Every row of a regression table has this form.
Once you can read one test, you can read the whole table.

::: {.callout-tip collapse="true"}
## Optional: one-sided alternatives

This box is optional reading and is not examined. All tests in this course
are two-sided. You will meet one-sided tests in reports and in software
output.

Some claims have a direction. A founder may believe that pitches offer less
than 15% on average. A test for such a claim is called a one-sided test. It
differs from the two-sided test in one place only: the rejection region.

There are three testing problems for a mean and a value $\mu_0$:

| | Hypotheses | Reject $H_0$ at level $\alpha$ if |
|---|---|---|
| (a) one-sided, right tail | $H_0:\mu\le\mu_0$ versus $H_1:\mu>\mu_0$ | $t_{obs}\ge t_{1-\alpha,\,n-1}$ |
| (b) one-sided, left tail | $H_0:\mu\ge\mu_0$ versus $H_1:\mu<\mu_0$ | $t_{obs}\le-t_{1-\alpha,\,n-1}$ |
| (c) two-sided | $H_0:\mu=\mu_0$ versus $H_1:\mu\ne\mu_0$ | $\lvert t_{obs}\rvert\ge t_{1-\alpha/2,\,n-1}$ |

In (a) and (b) the alternative lies on one side of $\mu_0$ only. Evidence
for it is a statistic that is far from zero in one direction only. So the
rejection region is one tail, and the whole of $\alpha$ goes into that
tail. At $\alpha=0.05$ with $n=100$ the one-sided critical value is
$t_{0.95,\,99}=1.660$, against 1.984 for the two-sided test. A one-sided
test rejects more easily in its own direction, and it never rejects in the
other direction. @fig-b10-one-sided shows the three rejection regions.

```{r}
#| label: fig-b10-one-sided
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.25
#| fig-cap: "Rejection regions at α = 0.05 for the three testing problems, on the same t curve with 99 degrees of freedom. In (a) and (b) the whole 5% lies in one tail, beyond 1.66 or beyond −1.66. In (c) it is split, with 2.5% in each tail beyond −1.98 and 1.98."
library(ggplot2)

alpha_tail <- 0.05
tail_df <- 99
tail_grid <- seq(-4, 4, length.out = 1400)
critical_two <- qt(c(alpha_tail / 2, 1 - alpha_tail / 2), df = tail_df)
critical_left <- qt(alpha_tail, df = tail_df)
critical_right <- qt(1 - alpha_tail, df = tail_df)

panel_levels <- c("(a) Right tail", "(b) Left tail", "(c) Two-sided")
null_curve <- do.call(rbind, lapply(panel_levels, function(p) {
  data.frame(x = tail_grid, density = dt(tail_grid, df = tail_df),
             panel = p)
}))
null_curve$panel <- factor(null_curve$panel, levels = panel_levels)

rejection <- rbind(
  cbind(subset(null_curve, panel == "(a) Right tail" & x >= critical_right),
        side = "right"),
  cbind(subset(null_curve, panel == "(b) Left tail" & x <= critical_left),
        side = "left"),
  cbind(subset(null_curve, panel == "(c) Two-sided" & x <= critical_two[1]),
        side = "left"),
  cbind(subset(null_curve, panel == "(c) Two-sided" & x >= critical_two[2]),
        side = "right")
)

critical_lines <- data.frame(
  x = c(critical_right, critical_left, critical_two),
  label = c("1.66", "-1.66", "-1.98", "1.98"),
  panel = factor(panel_levels[c(1, 2, 3, 3)], levels = panel_levels)
)

panel_labels <- data.frame(
  x = c(2.9, -2.9, -3, 3, 0, 0, 0),
  density = c(0.045, 0.045, 0.04, 0.04, 0.42, 0.42, 0.42),
  label = c("5%", "5%", "2.5%", "2.5%",
            "H[1] : mu > mu[0]", "H[1] : mu < mu[0]", "H[1] : mu != mu[0]"),
  parse_me = c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE),
  panel = factor(panel_levels[c(1, 2, 3, 3, 1, 2, 3)], levels = panel_levels)
)

ggplot(null_curve, aes(x, density)) +
  geom_area(data = rejection, aes(group = side),
            fill = "#D55E00", alpha = 0.6) +
  geom_segment(data = critical_lines,
               aes(x = x, xend = x, y = 0, yend = 0.3),
               linetype = "dashed", colour = "#5B6976", linewidth = 0.5) +
  geom_text(data = critical_lines, aes(x = x, y = 0.325, label = label),
            colour = "#17324D", size = 2.9) +
  geom_line(colour = "#17324D", linewidth = 0.8) +
  geom_text(data = subset(panel_labels, !parse_me), aes(label = label),
            colour = "#17324D", size = 2.9) +
  geom_text(data = subset(panel_labels, parse_me), aes(label = label),
            parse = TRUE, colour = "#5B6976", size = 3.1) +
  facet_wrap(~ panel, nrow = 1, scales = "fixed") +
  scale_x_continuous(breaks = c(-3, 0, 3)) +
  scale_y_continuous(breaks = c(0, 0.2, 0.4), limits = c(0, 0.44),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "t statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

One point of logic is new. In (a), the null hypothesis $H_0:\mu\le\mu_0$
contains many values, and each has its own probability of a Type I error.
This probability is largest at $\mu=\mu_0$, the value of $H_0$ that is
closest to the alternative. So we build the test to control the Type I
error at $\mu=\mu_0$. The probability is then smaller for every other value
in $H_0$. This is why the statistic still uses $\mu_0$ in the numerator.

The p-value changes with the rejection region. In (a) it is
$P(T\ge t_{obs})$, in (b) it is $P(T\le t_{obs})$, and in (c) it is
$2\,P(T\ge|t_{obs}|)$. For the advisor's figure, with $t_{obs}=-2.91$ and
$df=99$:

```{r}
#| label: b10-one-sided-p
c(one_sided_critical_value = qt(0.95, df = df),
  two_sided = 2 * pt(-abs(t_obs), df = df),
  left_tail = pt(t_obs, df = df),
  right_tail = 1 - pt(t_obs, df = df))
```

Take the founder's claim $H_1:\mu<15$, which is problem (b). The rejection
region at $\alpha=0.05$ is $t_{obs}\le-1.660$. Since $-2.91\le-1.660$,
reject $H_0:\mu\ge15$. The data support the claim that pitches offer less
than 15% on average. The p-value is $P(T\le-2.91)=0.0023$, half of the
two-sided value, because only one tail counts.

Suppose the founder had claimed that pitches offer more than 15%, which is
problem (a). The rejection region would be $t_{obs}\ge1.660$, and
$t_{obs}=-2.91$ would not reject. The p-value would be 0.9977. The sample
mean lies far from 15, but on the wrong side.

You must fix the direction before you see the data. If you choose the
side after you have looked at $\bar x$, you always test in the direction
that the sample points to. The true probability of a Type I error is then
about twice the stated one, 10% in place of 5%. This is the main reason why
this course uses two-sided tests only. Use a one-sided test only when the
direction follows from the question and could have been written down in
advance.
:::

### Preview: the test that comes with a regression slope

[Shark Tank data]{.shark-tag} A regression fits a straight line through a cloud of points. Here the line
describes how the equity offered (`equity_offered_pct`) changes from season
to season (`season`). Its slope says by how much per season on average. R
tests every regression coefficient against zero in the way described above.
The code fits the line to all 1,441 records of the Shark Tank file:

```{r}
#| label: b10-regression-preview
preview_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(preview_model)$coefficients
confint(preview_model)
```

One point before you read the output. In Lectures 7 to 10 we sampled 100
records from the file, so the uncertainty came from the random draw. Here
the model uses all 1,441 records. The standard error now answers a different
question: how much would the slope vary if the same process produced another
set of pitches? We treat the recorded pitches as one possible outcome of
that process.

Now read the `season` row. The slope of the line in the population is
written $\beta_1$, and its estimate is $\hat\beta_1$.

The estimate is $\hat\beta_1=-0.824$. Each additional season goes together
with about 0.82 percentage points less equity offered. For comparison, the
mean equity offered in the file is 13.21%, so this is about 6% of the mean
per season. The standard error, 0.048, is the sampling uncertainty of the
estimate, in the same units. The t value, $-17.3$, is the estimate divided
by its standard error. This is step 4 of the five steps, with zero as the
value under $H_0$. The p-value in the last column belongs to the two-sided
test

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1\ne0.
$$

The p-value is far below 0.05, so reject $H_0$. The critical-value rule
gives the same decision. With 1,439 degrees of freedom the t curve is
almost the standard normal curve, so the critical value is about 1.96, and
$|-17.3|$ lies far beyond it. The 95% interval $(-0.92,\,-0.73)$ does not
contain zero, in line with the rejection.

Every column is something you know from Lectures 9 and 10. The row does not
say that a later season causes lower offers. Nobody assigned pitches to
seasons, so this is an association over time, not a causal effect.

QM Intermediate starts from this row (estimate, standard error, t, p-value,
interval). It adds the interpretation of coefficients, models with several
predictors, and checks of the model.

::: {.caveat-box}
<span class="label">What a significant result does and does not establish</span>

- Significant is not large. A tiny difference can be significant with a lot
  of data, and a real difference can be missed with little.
- Significant is not causal. The test does not know how the data came about.
- No test can correct problems in the data. Problems with selection and
  measurement stay as they were.

**What to report, every time:** the estimate with its units, the standard
error, the observed value of the statistic, the decision at a stated
$\alpha$ or the p-value, and the confidence interval.

**What you can do with it:** compare the costs of the two errors before you
decide how much evidence you need.
:::

### Answering the driving question

A decision made from data can be wrong in two directions. Rejecting a true
null hypothesis is a Type I error. You fix its probability in advance at
$\alpha$ when you choose the critical value. This also answers how far is
far enough: the critical value is the distance at which the probability of
a false alarm is exactly $\alpha$. Missing a true difference is a Type II
error. Its probability is not fixed by the test. It is smaller when the
true difference is larger, the sample is larger, the population varies
less, or $\alpha$ is larger. A smaller $\alpha$ gives fewer false alarms at
the price of more missed differences. No critical value avoids this
trade-off.

The founder's question at the start can now be answered. Yes, you can tell
the advisor that 15% does not fit the pitches in the file. The sample mean
is 12.76%, and $t_{obs}=-2.91$ lies in the rejection region at the 5% level
and at the 1% level. The p-value is 0.0045, and the 95% interval
$(11.23,\,14.29)$ for the mean equity offered does not contain 15. You can
still be the one who is wrong. That would be a Type I error. A test at the
5% level makes this error in about 5 of 100 samples when $H_0$ is true.

Which error costs more depends on the decision, not on the statistics. A
remark to an advisor is easy to take back, so a false alarm costs little
here. The founder in the box below uses the same test for two decisions
with very different costs.

This completes QM Basic. You can now define events and variables on a
frame, describe their distributions, work with conditional probabilities,
and model sampling variation. You can turn "our sample mean is 12.76" into a
claim about a population, with its uncertainty, its decision, and its
limits.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

On your current sign-up page, 10% of visitors sign up. You try a new page on 400 visitors and 52 sign up, which is 13%. Test $H_0:p=0.10$ at the 5% level. The standard deviation of the sample proportion under $H_0$ is $\sqrt{0.10\times0.90/400}=0.015$, so $z_{obs}=0.03/0.015=2.0$. This is just above 1.96, so you reject $H_0$. The p-value is about 0.046. Now compare the costs of the two errors. Switching to the new page when it is not really better costs you almost nothing, because you can switch back. Missing a page that is really better costs you sign-ups every week. The decision: switch. For a decision about a kitchen lease, the logic is the other way round. A false alarm there is expensive and hard to undo. So before you sign, you would choose a smaller $\alpha$ and collect more data, so that the power stays high.
:::

**Next:** Tutorial 4 practises the tests of this lecture in exam format. The
Mock Exam then covers the whole course under exam conditions. QM
Intermediate, in the second year, continues from the regression preview
above.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 95 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the main points. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Name the five steps of a test, in order.
2. What is the rejection region of a two-sided test for a mean at level
   $\alpha$?
3. What is the difference between $\alpha$ and a p-value?
4. What is a Type I error?
5. What is a Type II error?
6. Name three factors that increase power.
7. Why is "do not reject" different from "accept"?
8. What does $H_0:\beta_1=0$ mean in a regression model?

::: {.callout-note collapse="true"}
## Check your answers

1. Hypotheses with the significance level, test statistic with its
   distribution if $H_0$ is true, critical value and rejection region,
   observed value, conclusion in the words of the question.
2. All values of the statistic at or below $-t_{1-\alpha/2,\,n-1}$ and all
   values at or above $t_{1-\alpha/2,\,n-1}$. For these values $H_0$ is
   rejected.
3. $\alpha$ is the largest probability of a Type I error that the test
   allows, fixed in advance. The p-value is calculated from the observed
   statistic as if $H_0$ were true. It is the smallest level at which $H_0$
   is rejected.
4. Rejecting $H_0$ when it is true.
5. Not rejecting $H_0$ when $H_1$ is true.
6. Any three of: a larger sample, a larger true difference, less variation
   in the population, or a larger $\alpha$ (the last at the cost of more
   Type I errors).
7. The test controls only the probability of a Type I error. With low
   power, a test often does not reject although $H_0$ is false. So a result
   that does not reject is weak evidence for $H_0$.
8. The slope of the line in the population is zero. The outcome has no
   linear association with the predictor.
:::

### Finish a half-worked example, then do one on your own (30 minutes)

Re-read the hand-worked app test (@exm-b10-launch-test). Then complete
the same five steps with new invented numbers: in 200 independent trials
there are 120 successes. Test whether the success proportion $p$ differs
from 0.50, at the 5% level.

*Steps 1 to 3 (given).* The hypotheses are $H_0:p=0.50$ versus
$H_1:p\ne0.50$, where $p$ is the success proportion in the population, and
$\alpha=0.05$. The test statistic is
$Z=(\widehat P-0.50)/\sqrt{0.50\times0.50/200}$. Since $np_0=100$ and
$n(1-p_0)=100$ are both at least 10, $Z$ is approximately $N(0,1)$ if $H_0$
is true. The critical value is $z_{0.975}=1.96$, so reject $H_0$ if
$z_{obs}\le-1.96$ or $z_{obs}\ge1.96$.

*Step 4 (you).* Calculate the denominator and then $z_{obs}$.

*Step 5 (you).* Write the conclusion.

*After step 5 (you).* The p-value is about 0.005. Check that it gives the
same decision.

::: {.callout-note collapse="true"}
## Check your answer

Step 4: the denominator is $\sqrt{0.50\times0.50/200}\approx0.0354$. The
sample proportion is $\hat p=120/200=0.60$, so
$z_{obs}=(0.60-0.50)/0.0354\approx2.83$. Step 5: $2.83\ge1.96$, so the
observed value lies in the rejection region. Reject $H_0$. At the 5% level
we can conclude that the success proportion differs from 50%. The positive
sign of $z_{obs}$ shows that it is above 50%. The p-value, about 0.005, is
at most 0.05, so it gives the same decision.
:::

Now solve these tasks on your own:

1. An invented example. A co-working space states that its members
   book 50 hours per month on average. A random sample of $n=49$ members
   has a mean of $\bar x=53$ hours and a sample standard deviation of
   $s=14$ hours. Test $H_0:\mu=50$ against $H_1:\mu\ne50$ at the 5% level
   in the five steps. Supplied value: $t_{0.975,\,48}=2.011$. Then answer
   two more questions. The p-value is about 0.14: does it give the same
   decision? The 95% confidence interval is $(48.98,\,57.02)$ hours: which
   decision does it give without a new calculation?
2. Describe the Type I and Type II errors for the 120-of-200 test above,
   in context.
3. An invented example. A regression of the equity offered (in percentage
   points) on the amount requested has the coefficient $-0.360$ percentage
   points per \$100,000, with standard error 0.090 and a large $df$. Calculate
   its t statistic and an approximate 95% interval, and interpret the row.
4. A coefficient estimate is too small to matter for any business decision,
   but it has $p<0.001$. Explain how both facts can be true.
5. State two reasons why the regression preview cannot support a causal
   conclusion.

::: {.callout-note collapse="true"}
## Check your answers

1. The five steps:
   1. Hypotheses. $H_0:\mu=50$ versus $H_1:\mu\ne50$, where $\mu$ is the
      mean number of hours booked per month among all members. The
      significance level is $\alpha=0.05$.
   2. Test statistic. $T=(\bar X-50)/(S/\sqrt{49})$. The sample is fairly
      large, so $T$ has approximately the $t_{48}$ distribution if $H_0$ is
      true.
   3. Critical value and rejection region. $t_{0.975,\,48}=2.011$. Reject
      $H_0$ if $t_{obs}\le-2.011$ or $t_{obs}\ge2.011$.
   4. Observed value. The standard error is $14/\sqrt{49}=2$ hours. So
      $t_{obs}=(53-50)/2=1.50$.
   5. Conclusion. $1.50$ lies between $-2.011$ and $2.011$, so it is not in
      the rejection region. Do not reject $H_0$. At the 5% level we cannot
      conclude that the mean number of hours booked differs from 50. This
      is not evidence that the mean is exactly 50 hours.

   The p-value, about 0.14, is above 0.05, so it gives the same decision.
   The interval $(48.98,\,57.02)$ contains 50, so it also says: do not
   reject $H_0$.
2. Type I: concluding that the population success proportion differs from
   0.50 when it equals 0.50. Type II: not detecting a success proportion
   that really differs from 0.50.
3. $t_{obs}=-0.360/0.090=-4.00$. With a large $df$, the 95% interval is
   approximately $-0.360\pm1.96\times0.090=(-0.536,\,-0.184)$ percentage
   points per \$100,000. An additional \$100,000 requested goes together
   with between about 0.18 and 0.54 percentage points less equity offered.
   The interval does not contain zero and $|t_{obs}|=4.00$ is far above
   1.96, so reject the null hypothesis that the coefficient is zero.
4. A very large sample or very little variation makes the standard error
   tiny. Then even a very small coefficient lies many standard errors from
   zero. A precise estimate is not automatically an important one.
5. Nobody assigned pitches to seasons at random. Other factors that
   changed over the seasons, the selection of pitches for the show, and
   measurement error can all play a role.
:::

::: {.callout-tip collapse="true"}
## Optional: one-sided reasoning

This task belongs to the optional box on one-sided alternatives. It is not
examined.

The test of the advisor's figure gave $t_{obs}=-2.91$ with $df=99$ and a
two-sided p-value of 0.0045.

1. State the p-value for the left-tail alternative $H_1:\mu<15$ and for the
   right-tail alternative $H_1:\mu>15$.
2. Which of the three alternatives would have been a correct choice, and
   what decides that?
3. A classmate sees that the sample mean is below 15 and then switches to
   the left-tail alternative "because it fits the data". What is wrong?

::: {.callout-note collapse="true"}
## Check your answers

1. Left tail: about 0.0023, half of the two-sided p-value. Right tail:
   about 0.9977, one minus that half.
2. Any of the three, if it follows from the research question and was
   fixed before the sample was inspected.
3. If you choose the tail after you see the estimate, the 5% always lies on
   the side where the estimate happens to fall. The true probability of a
   Type I error is then about 10%, not 5%. All tests in this course are
   two-sided.
:::
:::

### Review: one question each from Lectures 9 and 5 (10 minutes)

1. *(Lecture 9.)* An invented example. A 95% confidence interval for a
   population mean is $(20.4,\,25.2)$ euros. Give the estimate and the
   margin of error, and say what "95%" refers to.
2. [Shark Tank data]{.shark-tag} *(Lecture 5.)* In the full Shark Tank file, the correlation between
   season (`season`) and equity offered (`equity_offered_pct`) is about
   $-0.42$. Interpret this number in one sentence, and state one claim it
   cannot support.

::: {.callout-note collapse="true"}
## Check your answers

1. The estimate is the midpoint, $(20.4+25.2)/2=22.8$ euros. The margin of
   error is half the width, $(25.2-20.4)/2=2.4$ euros. "95%" refers to the
   procedure: about 95% of intervals built in this way from random samples
   contain the parameter.
2. Season and equity offered have a moderate negative linear association.
   Pitches in later seasons tend to offer less equity. The correlation
   cannot support a causal claim. Nobody assigned pitches to seasons, and
   other factors changed over time.

```{r}
#| label: b10-review-correlation
cor(sharks$season, sharks$equity_offered_pct)
```
:::

### Worked exam interpretation: a proportion test and a regression row (10 minutes)

**Exam-style question.** An invented example. A crowdfunding
platform states that 40% of its campaigns reach
their funding goal. In a simple random sample of 600 campaigns from last
year, 270 reached the goal. Let $p$ be the proportion of all campaigns of
last year that reached the goal. The same study fits a regression of the
amount raised (thousands of euros) on the number of updates that the
founder posted during the campaign. One row of the output is:

```text
                 Estimate  Std. Error  t value  Pr(>|t|)  95% CI
updates_posted      0.60       0.25       2.40     0.0167  [0.11, 1.09]
```

Supplied values: the two-sided p-value of the proportion test is 0.0124,
and a 95% confidence interval for $p$ is $(0.410,\,0.490)$.

Test $H_0:p=0.40$ against $H_1:p\ne0.40$ at the 5% level and give all five
steps of the testing procedure. Then define a Type I error for this test,
name what raises power while $\alpha$ stays fixed, and say which of these
the researcher can change. Last, interpret the regression row without
causal language.

::: {.callout-note collapse="true"}
## Check a model answer

1. Hypotheses. $H_0:p=0.40$ versus $H_1:p\ne0.40$, where $p$ is the
   proportion of all campaigns of last year that reached their goal. The
   significance level is $\alpha=0.05$.
2. Test statistic. $Z=(\widehat P-0.40)/\sqrt{0.40\times0.60/600}$. Since
   $np_0=600\times0.40=240$ and $n(1-p_0)=600\times0.60=360$ are both at
   least 10, $Z$ is approximately $N(0,1)$ if $H_0$ is true.
3. Critical value and rejection region. $z_{0.975}=1.96$. Reject $H_0$ if
   $z_{obs}\le-1.96$ or $z_{obs}\ge1.96$.
4. Observed value. The denominator is
   $\sqrt{0.40\times0.60/600}=\sqrt{0.0004}=0.02$. The sample proportion is
   $\hat p=270/600=0.45$. So $z_{obs}=(0.45-0.40)/0.02=2.50$.
5. Conclusion. $2.50\ge1.96$, so the observed value lies in the rejection
   region. Reject $H_0$. At the 5% level we can conclude that the
   proportion of campaigns that reach their goal differs from 40%. The
   positive sign of $z_{obs}$ shows that it is above 40%.

The p-value, 0.0124, is at most 0.05, so it gives the same decision. The
estimate is 45%, and plausible values run from 41.0% to 49.0%. The interval
does not contain 0.40, in line with the test.

A Type I error would be to conclude that the proportion differs from 40%
when it equals 40%. With $\alpha$ fixed, power rises with a larger sample
and with a larger true difference from 0.40. The researcher can change only
the sample size. The true difference is not a choice, and for a 0/1 outcome
the spread is determined by $p$.

The regression row: each additional update goes together with 0.60 thousand
euros (€600) more raised. The observed value is $t_{obs}=0.60/0.25=2.40$.
The p-value, 0.0167, is below 0.05, and the interval $(0.11,\,1.09)$ does
not contain zero, so reject $H_0:\beta_1=0$. The founders decided
themselves how many updates to post, so this is an association, not a
causal effect.
:::

### Worked exam interpretation: a mean test (10 minutes)

**Exam-style question.** An invented example. A webshop promises
that an order arrives after 3 days on average.
From the 1,200 orders that it shipped last month, a simple random sample of
100 orders was selected without replacement. `arrival_days` is the number
of days between the order and its arrival. $\mu$ is the mean of
`arrival_days` among all 1,200 orders. The sample has $\bar x=3.3$ days and
$s=2.0$ days. R produced:

```text
	One Sample t-test

data:  arrival_days
t = 1.5, df = 99, p-value = 0.1368
alternative hypothesis: true mean is not equal to 3
95 percent confidence interval:
 2.903157 3.696843
sample estimates:
mean of x 
      3.3 
```

Test $H_0:\mu=3$ against $H_1:\mu\ne3$ at the 5% level and give all five
steps of the testing procedure. Check your observed value against the
output. Then interpret the p-value and the interval, and check the 10%
condition.

::: {.callout-note collapse="true"}
## Check a model answer

1. Hypotheses. $H_0:\mu=3$ versus $H_1:\mu\ne3$, where $\mu$ is the mean
   number of days until arrival among all 1,200 orders of last month. The
   significance level is $\alpha=0.05$.
2. Test statistic. $T=(\bar X-3)/(S/\sqrt{100})$. The sample is large, so
   $T$ has approximately the $t_{99}$ distribution if $H_0$ is true.
3. Critical value and rejection region. The degrees of freedom are
   $df=100-1=99$. The formula sheet gives $t_{0.975,\,100}=1.984$,
   the closest listed value. Reject $H_0$ if $t_{obs}\le-1.984$ or
   $t_{obs}\ge1.984$.
4. Observed value. The standard error is $2.0/\sqrt{100}=0.2$ days. So
   $t_{obs}=(3.3-3)/0.2=1.50$, as in the output.
5. Conclusion. $1.50$ lies between $-1.984$ and $1.984$, so it is not in
   the rejection region. Do not reject $H_0$. At the 5% level we cannot
   conclude that the mean arrival time of the 1,200 orders differs from 3
   days. This is not evidence that the mean is exactly 3 days.

The p-value: if the mean of all 1,200 orders were 3 days, a sample mean at
least 1.5 standard errors from 3 in either direction would occur in about
14% of random samples. Since $0.1368>0.05$, the p-value gives the same
decision. The interval $(2.90,\,3.70)$ contains 3, in line with the test,
but it also contains values up to 3.7 days. It was produced by a procedure
that captures $\mu$ in about 95% of random samples. The 10% condition
holds, because $100\le0.10\times1{,}200=120$. The conclusion concerns the
1,200 orders of last month, not the orders of other months.
:::

### Check the AI (5 minutes)

[Shark Tank data]{.shark-tag} You asked an AI assistant: "I tested the advisor's figure of 15% and the p-value is 0.0045. What does that tell me?" It answered:

> "There is only a 0.45% probability that the null hypothesis is true, so the difference must be large."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Two errors. The p-value is not the probability that the null hypothesis is true. It says: if the mean equity offered were 15%, a sample mean this far from 15 would occur in about 0.45% of samples. And it says nothing about size. The size is the estimate: the sample mean of 12.76% lies 2.24 percentage points below 15, and the interval runs from 11.23 to 14.29.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Name one
yes-or-no claim that you will have to defend with data, for example "our
conversion rate differs from the sector's 3%" or "our average order differs
from the break-even value". Write two sentences. First state the claim as a
two-sided test with $H_0$ and $H_1$, and say which of the two errors would
cost you more, and why. Then name the one way to raise power that you could
really use before your next pitch.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run venture that sells meal-prep boxes on campus. "My claim: our conversion rate differs from the sector's 3%, $H_0\!: p = 0.03$ against $H_1\!: p \ne 0.03$. A false alarm is the costlier error, because I would sign a bigger kitchen lease on a difference that is not real. The one way to raise power that I can use before the next pitch is a larger sample, by running the tasting stand for two more weeks."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to check the five steps of the
test for your claim and to question your choice of the more costly error.

::::

### More practice

For extra exercises on this lecture, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Lecture 10. Show the solution only after I send my answer."

[Continue to Tutorial 4: Hypothesis tests, errors, and power](t04-exam-prep.qmd).

[Formula sheet: Lecture 10 formulas](formula-sheet.qmd#formula-l10).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 15 and 16.

:::
