# Lecture 10: Testing decisions, errors, power, and the road to regression {#b10}

```{r}
#| label: b10-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 10, you can</span>

- rewrite benchmark and comparison questions as tests of a zero contrast;
- complete and communicate a two-sided test using the seven-step workflow;
- distinguish a p-value, a significance level, and a test decision;
- describe Type I error, Type II error, and power in context;
- explain when a one-sided alternative is legitimate and why this course's
  exams are two-sided;
- separate statistical significance from magnitude, practical importance,
  and causality;
- interpret the zero-null test printed for a regression coefficient.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Launch the product or wait. Approve the loan or decline. Decisions made from data can err in two directions, and the two mistakes usually cost very different amounts.</p>

Every test can be wrong in two ways. Which wrong call is which, what makes
wrong calls rarer, and how does all of this reappear in a regression table?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Review <a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to
Hypothesis Testing</a> and watch
<a href="https://www.youtube.com/watch?v=JNOWWJeZBTc">Examples Identifying
Type I and Type II Errors</a>. Focus on two-sided tests, decision errors,
and why "do not reject" is not "accept." All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (8 minutes).** Suppose R reports the following coefficient row:

| Estimate | SE | t | p-value |
|---:|---:|---:|---:|
| -0.290 | 0.058 | -5.00 | < 0.001 |

State the hypotheses being tested and give a careful one-sentence
statistical conclusion.

::: {.callout-note collapse="true"}
## Check your preparation

The default two-sided coefficient test is $H_0:\beta=0$ against
$H_1:\beta\ne0$. Because the p-value is below 0.05, reject $H_0$: the data
provide evidence that the population-model coefficient differs from zero.
A full interpretation still needs the outcome, predictor, units, held-fixed
variables, population, and causal limit.
:::

**Predict (2 minutes).** Answer the one-click Mentimeter question posted on
Canvas before you come to class: a test at $\alpha=0.05$ returns
$p=0.20$. Has the study **shown that the null is true** --- yes, no, or
cannot tell? Commit to an answer; the lecture resolves it.

Review [Mathematics Refresher: zero contrasts, inequalities, and
standardisation](../resources/math-refresher.html) if needed.

:::

::: {.course-phase .phase-in-class}

## In class

The session opens with the usual four-minute entry quiz on the prep videos,
then turns to the driving question. Lecture 9 delivered one complete test.
This final lecture makes testing a *routine*: one recipe, an honest account
of the two ways any decision can be wrong, and the payoff --- the same
machinery reappearing, unchanged, in every regression table you will read
from Intermediate onwards.

<p class="concept-video"><strong>Video explanations (review):</strong>
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to Hypothesis Testing</a> and
<a href="https://www.youtube.com/watch?v=JNOWWJeZBTc">Type I and Type II Errors</a>.</p>

### Why zero is the universal benchmark

Many empirical questions ask whether a population contrast is absent:

- difference from a target: $\delta=\mu-15$;
- difference in probabilities: $\delta_p=p-0.50$;
- difference between two group means: $\mu_1-\mu_0$;
- a regression slope: $\beta_j$.

In every case, "no difference" or "no linear association" becomes

$$
H_0:\theta=0
\qquad\text{versus}\qquad
H_1:\theta\ne0.
$$

Testing a non-zero benchmark can always be rewritten as a test of a
zero-valued contrast --- Lecture 9's gap device. This common structure is
the bridge from Basic inference to regression inference: learn to test one
zero, and you can test them all.

### The seven-step testing workflow

Use the same sequence every time --- on the exam, in your EiA venture, in
any report you ever write:

1. Define the population parameter and its units.
2. State $H_0$ and the two-sided $H_1$ **before** inspecting the result.
3. Check the sampling and model conditions.
4. Calculate the estimate and its SE under the null model.
5. Standardise: $(\text{estimate}-0)/\text{SE}$.
6. Calculate the two-sided p-value and compare it with the pre-specified
   significance level $\alpha$.
7. Report the estimate, uncertainty, decision, contextual meaning, and
   limits.

The significance level $\alpha$ is a long-run decision threshold, commonly
0.05, fixed in advance. The p-value is calculated from the observed data
under $H_0$. They are not the same object: one is a rule, the other a
result.

If $p\le\alpha$, reject $H_0$. If $p>\alpha$, **do not reject** $H_0$ ---
and do not say "accept $H_0$." This resolves the prep prediction: a study
with $p=0.20$ has *not* shown the null is true; it may simply lack the
precision to distinguish a meaningful non-zero parameter from zero. The
power section below makes that precise.

::: {#exm-b10-launch-test}
## Worked example: a proportion test by hand

An accelerator claims that half of its alumni launch a trading venture
within a year. In a random sample of 400 former participants, 232 had
launched. Let $p$ be the launch proportion in the alumni frame and define
$\delta_p=p-0.50$. Test $H_0:\delta_p=0$ against $H_1:\delta_p\ne0$ at 5%,
voicing every arithmetic step.

- **Estimate.** $\hat p=232/400=0.58$, so the estimated gap is
  $\widehat\delta_p=0.58-0.50=0.08$, eight percentage points.
- **Conditions under the null.** $np_0=400(0.50)=200$ and
  $n(1-p_0)=200$, both far above 10 --- the normal approximation is safe.
- **Null SE.** Under $H_0$ the true proportion is 0.50, so
  $\operatorname{SE}_0=\sqrt{0.50(0.50)/400}=\sqrt{0.25/400}
  =\sqrt{0.000625}=0.025$.
- **Standardise.** $z=0.08/0.025=3.20$: the observed gap sits 3.2 null
  standard errors above zero.
- **p-value and decision.** The two-sided p-value is
  $2\,P(Z\le-3.20)\approx 0.0014 < 0.05$: reject $H_0$. The sample
  provides strong evidence that the frame launch proportion differs from
  50%; the positive estimate places it above.

Note step 4's subtlety: the SE was computed with $p_0=0.50$, not with
$\hat p$ --- the test measures how surprising the sample is *in the null's
world*, so the null supplies the spread.
:::

### The Shark Tank deal proportion, in R

The same recipe on our reproducible sample of 100: is the frame's deal
proportion different from 50%? Every object is constructed on screen:

```{r}
#| label: b10-proportion-objects
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

n <- nrow(pitch_sample)
phat <- mean(pitch_sample$deal_on_show)   # sample deal proportion
p0 <- 0.50                                # null value
deal_gap <- phat - p0                     # estimated gap

c(n = n, phat = phat, p0 = p0, deal_gap = deal_gap)
```

```{r}
#| label: b10-proportion-test
null_se <- sqrt(p0 * (1 - p0) / n)
z_value <- deal_gap / null_se
two_sided_p <- 2 * pnorm(-abs(z_value))

c(
  null_standard_error = null_se,
  z_statistic = z_value,
  two_sided_p_value = two_sided_p
)
```

The estimate is $\widehat\delta_p=0.14$, fourteen percentage points. The
null SE is $\sqrt{0.50(0.50)/100}=0.05$, the statistic is $z=2.80$, and
the two-sided p-value is about 0.005. The success--failure condition holds
under the null: $np_0=50$ and $n(1-p_0)=50$. Reject the zero-gap null at
5%: the random sample provides evidence that the recorded-pitch-frame deal
proportion differs from 50%, and the positive estimate places it above.

### Interval and test: report both

Step 7 of the workflow says a decision alone is not a report. A
large-sample 95% interval for the proportion uses the estimated SE:

$$
\hat p\pm1.96
\sqrt{\frac{\hat p(1-\hat p)}{n}}.
$$

```{r}
#| label: b10-proportion-interval
estimated_se <- sqrt(phat * (1 - phat) / n)
proportion_ci <- phat + c(-1, 1) * 1.96 * estimated_se
gap_ci <- proportion_ci - p0

c(proportion_lower = proportion_ci[1],
  proportion_upper = proportion_ci[2])
c(gap_lower = gap_ci[1], gap_upper = gap_ci[2])
```

The proportion interval is approximately $(0.546,\,0.734)$; the equivalent
gap interval is $(0.046,\,0.234)$, which excludes zero and agrees with the
test decision here.

::: {.optional-box}
<span class="label">Optional: estimated versus null standard errors</span><br>
Look closely: the test divided by the **null** SE
$\sqrt{p_0(1-p_0)/n}=0.050$, while the interval used the **estimated** SE
$\sqrt{\hat p(1-\hat p)/n}=0.048$. The test asks how surprising the sample
is if $H_0$ were true, so the null supplies the spread; the interval makes
no null assumption, so the data supply it. Because these are not exactly
the same procedure, interval and test decisions need not match perfectly
when a result sits very near the boundary. Exact equivalence requires an
interval obtained by inverting the same test --- beyond this course. For
the mean-gap t procedures of Lecture 9 the same SE is used in both, so
there the match is exact.
:::

### What a p-value is not

The two-sided p-value is

$$
P\!\left(|\text{test statistic}|
\ge|\text{observed statistic}|\mid H_0\right)
$$

--- and nothing else. Before the break you receive four sentences about the
launch test of @exm-b10-launch-test ($z=3.20$, $p\approx 0.0014$). Each
one is wrong. In pairs, correct all four; the plenary compares repairs.

::: {.callout-caution collapse="true"}
## Try it: correct the four wrong p-value sentences

*Sentence 1 (you).* "There is only a 0.14% probability that the launch
proportion really is 50%."

*Sentence 2 (you).* "The result has a 0.14% probability of being due to
chance."

*Sentence 3 (you).* "Such a tiny p-value shows that the launch proportion
is far above 50%."

*Sentence 4 (you).* "The tiny p-value proves that the accelerator causes
its alumni to launch ventures."

::: {.callout-note collapse="true"}
## Check your answers

1. A p-value is not $P(H_0\mid\text{data})$. Correct: "If the launch
   proportion were exactly 50%, a sample gap at least this far from zero
   in either direction would occur in about 0.14% of random samples."
2. The p-value is computed *assuming* chance alone (the null model)
   generated the gap; it is the probability of data this extreme under
   that assumption, not the probability that the assumption holds.
3. A p-value measures surprise, not size. The magnitude is the estimate:
   8 percentage points, with an interval around it. A tiny p-value can
   accompany a tiny gap when $n$ is huge.
4. Significance says nothing about why. Alumni self-selected into the
   accelerator; without a design that isolates its effect, "causes" is
   unsupported. Correct: the frame's launch proportion differs from 50%.
:::
:::

### Type I error, Type II error, and power

A test is a decision rule, and any decision rule can be wrong in exactly
two ways:

| Reality | Reject $H_0$ | Do not reject $H_0$ |
|---|---|---|
| $H_0$ true | Type I error | Correct decision |
| $H_1$ true | Correct decision | Type II error |

The significance level controls the first:

$$
P(\text{reject }H_0\mid H_0\text{ true})=\alpha.
$$

**Power** is the ability to avoid the second:

$$
\text{power}=P(\text{reject }H_0\mid H_1\text{ true}).
$$

For the launch test, in context: a **Type I error** would be concluding
that the alumni launch proportion differs from 50% when it is exactly 50%
--- crying wolf about a fictitious difference, embarrassing the analyst who
advises repricing the programme. A **Type II error** would be failing to
detect a launch proportion that genuinely differs from 50% --- missing a
real signal because the sample was too small or noisy.

```{r}
#| label: fig-b10-errors-power
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Two-sided testing at α = 0.05. Left: when the null is true, the z statistic follows the dark curve and the orange rejection regions have total probability 0.05 — the Type I error rate. Right: when a positive alternative is true, the statistic follows the shifted blue curve; the green area is the power, and the unshaded middle is the Type II error probability. Both panels use the same axes."
library(ggplot2)

z_grid <- seq(-4.5, 6.5, length.out = 1600)
critical <- qnorm(0.975)
panel_levels <- c("Null is true", "An alternative value is true")

test_curves <- rbind(
  data.frame(z = z_grid, density = dnorm(z_grid, mean = 0),
             panel = panel_levels[1]),
  data.frame(z = z_grid, density = dnorm(z_grid, mean = 2.5),
             panel = panel_levels[2])
)
test_curves$panel <- factor(test_curves$panel, levels = panel_levels)

rejection <- subset(test_curves, z <= -critical | z >= critical)
rejection$side <- ifelse(rejection$z < 0, "left", "right")

panel_labels <- data.frame(
  z = c(0, 3.6),
  density = c(0.13, 0.14),
  label = c("do not reject", "power"),
  panel = factor(panel_levels[c(1, 2)], levels = panel_levels)
)
alpha_labels <- data.frame(
  z = c(-2.5, 2.5),
  density = c(0.028, 0.028),
  label = c("alpha/2", "alpha/2"),
  panel = factor(panel_levels[c(1, 1)], levels = panel_levels)
)

ggplot(test_curves, aes(z, density)) +
  geom_area(data = rejection, aes(group = side, fill = panel),
            alpha = 0.4) +
  geom_vline(xintercept = c(-critical, critical), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_line(aes(colour = panel), linewidth = 0.9) +
  geom_text(data = panel_labels, aes(label = label),
            colour = "#17324D", size = 3.4) +
  geom_text(data = alpha_labels, aes(label = label), parse = TRUE,
            colour = "#17324D", size = 3.1) +
  facet_wrap(~ panel, scales = "fixed") +
  scale_fill_manual(values = c("Null is true" = "#D55E00",
                               "An alternative value is true" = "#009E73"),
                    guide = "none") +
  scale_colour_manual(values = c("Null is true" = "#17324D",
                                 "An alternative value is true" = "#0072B2"),
                      guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(xlim = c(-4, 6)) +
  labs(x = "z statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

In both panels of @fig-b10-errors-power, the horizontal axis is the
possible z statistic, the vertical axis is density, and probability is
area. The dashed vertical lines mark the same two-sided critical values,
$\pm 1.96$. In the left panel the null is true: the two orange rejection
tails have total probability $\alpha=0.05$, which is exactly the Type I
error rate --- the test is *built* to cry wolf 5% of the time. In the right
panel a particular positive alternative is true, so the curve has shifted:
the green area is the power, and whatever probability remains between the
dashed lines is the Type II error rate.

Four levers raise power:

- a larger true distance from zero (an effect that is easier to see);
- a larger sample (a smaller SE);
- lower unexplained variation (also a smaller SE);
- a larger $\alpha$ --- but this lever is a trade-off, not a free gain:
  widening the rejection region catches more true effects *and* raises the
  Type I error rate in lockstep.

### One-sided alternatives

A two-sided test asks whether a parameter differs from zero in **either**
direction. A one-sided test asks a directional question chosen before the
sample is examined:

| Research question | Alternative | Relevant tail(s) |
|---|---|---|
| Does the gap differ from zero? | $H_1:\delta\ne0$ | both tails |
| Is the gap below zero? | $H_1:\delta<0$ | left tail |
| Is the gap above zero? | $H_1:\delta>0$ | right tail |

In all three cases the test statistic is compared with its distribution
under the zero boundary $H_0:\delta=0$. What changes is which values count
as evidence for the stated alternative --- that is, where the rejection
region sits.

```{r}
#| label: fig-b10-one-sided
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.25
#| fig-cap: "Rejection regions at α = 0.05 for the two-sided and both one-sided alternatives, on the same null t distribution (df = 99). The two-sided test places 2.5% in each tail; a one-sided test moves the full 5% into the tail named by its alternative, so its critical value is closer to zero in that direction."
library(ggplot2)

alpha_tail <- 0.05
tail_df <- 99
tail_grid <- seq(-4, 4, length.out = 1400)
critical_two <- qt(c(alpha_tail / 2, 1 - alpha_tail / 2), df = tail_df)
critical_left <- qt(alpha_tail, df = tail_df)
critical_right <- qt(1 - alpha_tail, df = tail_df)

panel_levels <- c("Two-sided", "Left-sided", "Right-sided")
null_curve <- do.call(rbind, lapply(panel_levels, function(p) {
  data.frame(x = tail_grid, density = dt(tail_grid, df = tail_df),
             panel = p)
}))
null_curve$panel <- factor(null_curve$panel, levels = panel_levels)

rejection <- rbind(
  cbind(subset(null_curve, panel == "Two-sided" & x <= critical_two[1]),
        side = "left"),
  cbind(subset(null_curve, panel == "Two-sided" & x >= critical_two[2]),
        side = "right"),
  cbind(subset(null_curve, panel == "Left-sided" & x <= critical_left),
        side = "left"),
  cbind(subset(null_curve, panel == "Right-sided" & x >= critical_right),
        side = "right")
)

critical_lines <- data.frame(
  x = c(critical_two, critical_left, critical_right),
  panel = factor(c("Two-sided", "Two-sided", "Left-sided", "Right-sided"),
                 levels = panel_levels)
)

panel_labels <- data.frame(
  x = c(-2.8, 2.8, -2.8, 2.8, 0, 0, 0),
  density = c(0.035, 0.035, 0.035, 0.035, 0.42, 0.42, 0.42),
  label = c("2.5%", "2.5%", "5%", "5%",
            "H[1] : delta != 0", "H[1] : delta < 0", "H[1] : delta > 0"),
  parse_me = c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE),
  panel = factor(c("Two-sided", "Two-sided", "Left-sided", "Right-sided",
                   "Two-sided", "Left-sided", "Right-sided"),
                 levels = panel_levels)
)

ggplot(null_curve, aes(x, density)) +
  geom_area(data = rejection, aes(group = side),
            fill = "#D55E00", alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = "dotted",
             colour = "grey75", linewidth = 0.4) +
  geom_vline(data = critical_lines, aes(xintercept = x),
             linetype = "dashed", colour = "#5B6976", linewidth = 0.5) +
  geom_line(colour = "#17324D", linewidth = 0.8) +
  geom_text(data = subset(panel_labels, !parse_me), aes(label = label),
            colour = "#17324D", size = 2.9) +
  geom_text(data = subset(panel_labels, parse_me), aes(label = label),
            parse = TRUE, colour = "#5B6976", size = 3.1) +
  facet_wrap(~ panel, nrow = 1, scales = "fixed") +
  scale_x_continuous(breaks = c(-3, 0, 3)) +
  scale_y_continuous(breaks = c(0, 0.2, 0.4), limits = c(0, 0.44),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "t statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

Compare @fig-b10-one-sided with the left panel of
@fig-b10-errors-power: same logic, different placement of the 5%. Moving
the full 5% into one tail pulls that critical value closer to zero, so a
one-sided test is *more* able to detect an effect in its chosen direction
--- a power gain. The price: results in the opposite direction, however
extreme, do not support the stated alternative at all.

That power gain is exactly why the direction must follow from the research
question and be **fixed before the data are inspected**. Choosing the tail
because the observed estimate happens to point that way --- or simply to
obtain a smaller p-value --- silently doubles the true Type I error rate:
you would take the 5% tail on whichever side the noise landed. For
Lecture 9's equity-gap statistic, $t=-2.91$ with $df=99$:

- the two-sided p-value is about 0.0045;
- a **pre-specified** left-sided alternative $H_1:\delta<0$ gives about
  0.0023;
- a **pre-specified** right-sided alternative $H_1:\delta>0$ gives about
  0.9977.

The negative estimate agrees with the left-sided alternative and points
away from the right-sided one. Course convention, on the exam and in every
chapter: **tests are two-sided**, because in business research a
difference in either direction is usually relevant --- and because the
convention removes any temptation to pick the convenient tail.

### Preview: the test attached to a regression slope

R applies the same zero-null logic to every regression coefficient.
Consider the simple model of initial equity offered on season, fitted to
all 1,441 records:

```{r}
#| label: b10-regression-preview
preview_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(preview_model)$coefficients
confint(preview_model)
```

Read the `season` row with the vocabulary you now own. The **estimate** is
$-0.824$: about 0.82 percentage points less equity offered per additional
season. The **SE**, 0.048, is its sampling uncertainty in the same units.
The **t value**, $-17.3$, is estimate divided by SE --- step 5 of the
workflow. The **p-value** belongs to the printed two-sided test

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1\ne0,
$$

and the 95% **interval** $(-0.92,\,-0.73)$ excludes zero, agreeing with
the rejection. Every column is a Lecture 9--10 object wearing a regression
name. What the row does *not* say: this is an observational time
association, not a causal effect of moving a pitch to a later season.

Intermediate begins from this familiar estimate--SE--t--p-value structure
and adds careful coefficient interpretation, multiple predictors, and
model diagnostics.

::: {.caveat-box}
<span class="label">What a significant result does and does not establish</span>

Three separations to carry out of this course. **Significance is not
magnitude:** a tiny effect can be significant in a large sample, and a
meaningful effect can be non-significant in a small or noisy one --- the
p-value measures surprise under the null, the estimate measures size.
**Significance is not causation:** none of this lecture's machinery knows
whether the predictor was randomly assigned. **And no test repairs the
frame:** selection into the recorded-pitch file, measurement of "deal" as
an on-air agreement, and dependence between observations are untouched by
any p-value. The standing reporting list, for every analysis from now on:
the estimate in substantive units; a confidence interval; the p-value or
decision; whether the magnitude matters for the business question; and the
sampling, measurement, and causal limits.
:::

### Answering the driving question

The two wrong calls: rejecting a true null (Type I, rate set by $\alpha$)
and missing a true effect (Type II, rate cut by power). Wrong calls become
rarer with more data, less noise, bigger true effects --- and, for Type I
only, a stricter $\alpha$, at the price of more Type II. And the
regression table? Each coefficient row *is* a Lecture 9 test: estimate,
SE, t, two-sided p against zero, interval.

That closes the arc of QM Basic. Ten lectures ago a Shark Tank pitch was
an anecdote. Now you can define events and variables on a frame, describe
their distributions, condition without confusing directions, model
sampling variation, and turn "our sample mean is 12.76" into a defensible,
properly hedged claim about a population --- with its uncertainty, its
decision, and its limits stated. That is the difference between an opinion
with a spreadsheet and evidence.

**Next:** Tutorial 4 rehearses Lectures 9--10 in exam format, and the Mock
Exam then integrates the whole course under exam conditions. The story itself
continues next year: QM Intermediate picks up exactly where today's regression
preview left off, and the coefficient row you can already read becomes the
central object of six lectures on linear regression.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 65 minutes.</span>

### Retrieval (10 minutes)

1. Why can most comparison hypotheses be written as a parameter equal to
   zero?
2. What is the difference between $\alpha$ and a p-value?
3. What is a Type I error?
4. What is a Type II error?
5. Name three factors that increase power.
6. Why is "do not reject" different from "accept"?
7. What does $H_0:\beta_j=0$ mean in a regression model?

::: {.callout-note collapse="true"}
## Check your answers

1. Subtracting the benchmark or reference value creates a contrast; "no
   difference" corresponds to a zero contrast.
2. $\alpha$ is a pre-specified long-run decision threshold; the p-value is
   calculated from the observed statistic under the null.
3. Rejecting $H_0$ when it is true.
4. Not rejecting $H_0$ when a specified alternative is true.
5. Any three of: larger sample size, larger true effect, lower noise, or
   larger $\alpha$ (the last at the cost of more Type I errors).
6. A non-significant result may simply be too imprecise to distinguish a
   meaningful parameter from zero.
7. Holding the other included predictors fixed, the population-model
   coefficient for predictor $j$ is zero.
:::

### From worked to faded, then unassisted (20 minutes)

Re-read the hand-worked launch test (@exm-b10-launch-test). Then complete
the same recipe with new numbers: in 200 independent trials there are 120
successes, and the benchmark is $p_0=0.50$.

*Step 1 (given).* $\hat p=120/200=0.60$, so $\widehat\delta_p=0.10$; the
null counts are $200(0.50)=100$ successes and 100 failures, both above 10.

*Step 2 (you).* Calculate the null SE and the z statistic.

*Step 3 (you).* The two-sided p-value and the 5% decision, in context.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SE}_0=\sqrt{0.50(0.50)/200}\approx 0.0354$, so
$z=0.10/0.0354\approx 2.83$ and the two-sided p-value is about 0.005.
Reject the zero-gap null at 5%: the data provide evidence that the success
proportion differs from 50%, with the positive estimate placing it above.
:::

Now unassisted:

1. A sample of $n=49$ difference scores has $\bar d=3$ and $s_d=14$. Test
   $H_0:\delta=0$ against $H_1:\delta\ne0$ at 5%.
2. Describe the Type I and Type II errors for the 120-of-200 test above,
   in context.
3. A multiple-regression coefficient is $-0.290$ percentage points per
   €100,000, with SE 0.058 and a large $df$. Calculate its t statistic and
   approximate 95% interval, and interpret the row.
4. A coefficient estimate is economically negligible but has $p<0.001$.
   Explain how both facts can be true.
5. State two reasons the regression preview cannot support a causal
   conclusion.

::: {.callout-note collapse="true"}
## Check your answers

1. The SE is $14/\sqrt{49}=2$, so $t=3/2=1.50$ with $df=48$; the two-sided
   p-value is about 0.14. Do not reject the zero mean-difference null at
   5% --- which is not evidence that the difference is zero.
2. Type I: concluding the population success proportion differs from 0.50
   when it equals 0.50. Type II: failing to detect a genuine specified
   difference from 0.50.
3. $t=-0.290/0.058=-5.00$. With a large $df$, the 95% interval is
   approximately $-0.290\pm1.96(0.058)=(-0.404,\,-0.176)$ percentage
   points per €100,000. Holding the model's other predictors fixed, an
   additional €100,000 requested is associated with between about 0.18 and
   0.40 percentage points less fitted equity; reject the zero-coefficient
   null.
4. A very large sample or low noise makes the SE tiny, so even a
   negligible coefficient sits many SEs from zero. Precision does not
   confer importance.
5. Season was not randomly assigned, and omitted factors, selection,
   reverse direction, or measurement error can all remain.
:::

### One-sided reasoning (10 minutes)

Lecture 9's equity-gap test gave $t=-2.91$ with $df=99$.

1. State all three p-values: two-sided, left-sided ($H_1:\delta<0$), and
   right-sided ($H_1:\delta>0$).
2. Which of the three alternatives would have been legitimate to
   pre-specify, and what determines that?
3. A classmate, seeing the negative estimate, switches to the left-sided
   alternative "because it fits the data." What exactly is wrong?

::: {.callout-note collapse="true"}
## Check your answers

1. Two-sided: about 0.0045. Left-sided: about 0.0023 (the single left tail
   beyond $-2.91$). Right-sided: about 0.9977 (everything to the right of
   $-2.91$).
2. Any of the three --- **provided** it was fixed before inspecting the
   sample and follows from the research question. Legitimacy is about
   timing and motivation, not about which alternative is "correct."
3. Choosing the tail after seeing the estimate means the 5% rejection mass
   would follow the data to whichever side the noise landed, so the true
   Type I error rate is doubled to about 10%. The smaller p-value is
   bought by an invalid procedure. Course convention: two-sided, always,
   on the exam.
:::

### Spaced review: one item each from Lectures 9 and 5 (10 minutes)

1. *(Lecture 9.)* A 95% confidence interval for a mean gap is
   $(0.4,\,5.2)$ units. State the 5% two-sided test decision without
   computing, and name one thing the interval reports that the decision
   does not.
2. *(Lecture 5.)* In the full file, the correlation between season and
   initial equity offered is about $-0.42$. Interpret this number in one
   sentence, and state one claim it cannot support.

::: {.callout-note collapse="true"}
## Check your answers

1. The interval excludes zero, so the matched two-sided test rejects
   $H_0:\delta=0$ at 5%. The interval additionally shows direction,
   plausible magnitude (0.4 to 5.2 units), and precision.
2. Season and offered equity have a moderate negative linear association:
   later-season pitches tend to offer less equity. It cannot support a
   causal claim --- nothing assigned pitches to seasons, and other factors
   changed over time.
:::

### Worked exam interpretation (10 minutes)

**Question (Mock Exam Q5 pattern).** In a simple random sample of 400
former accelerator participants, 232 had launched a trading venture within
one year; a two-sided test of $H_0:p=0.50$ is planned. The same study
fits a regression of monthly revenue (thousands of euros) on weekly
mentoring hours; one row of the output is:

```text
                 Estimate  Std. Error  t value  Pr(>|t|)  95% CI
mentoring_hours     0.80       0.25       3.20     0.0016  [0.31, 1.29]
```

Complete the proportion test at 5% (a large-sample 95% interval for $p$ is
$(0.532,\,0.628)$), define a Type I error for it, name two power levers
that hold $\alpha$ fixed, and interpret the regression row without causal
language.

::: {.callout-note collapse="true"}
## Check a model answer

Under $H_0$, $np_0=n(1-p_0)=200\ge 10$. $\hat p=232/400=0.58$,
$\operatorname{SE}_0=\sqrt{0.50(0.50)/400}=0.025$, and $z=0.08/0.025=3.20$
with two-sided p-value about 0.0014: reject $H_0$; the frame launch
proportion differs from 50%, plausibly between 53.2% and 62.8% --- the
interval excludes 0.50, agreeing with the test. A Type I error would be
concluding the launch proportion differs from 50% when it equals 50%.
With $\alpha$ fixed, power rises with a larger sample or lower outcome
variability (or a larger true gap, though that is not chosen). The
regression row: each additional weekly mentoring hour is associated with
0.80 thousand euros more fitted monthly revenue; $t=0.80/0.25=3.20$,
$p=0.0016$, and the interval $(0.31,\,1.29)$ excludes zero, so reject
$H_0:\beta=0$. Mentoring hours were not randomly assigned, so this is an
association, not a causal effect.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Name one
yes-or-no claim you will realistically have to defend with data (our
conversion rate beats the sector's 3%; our average order exceeds the
break-even basket), and write two sentences: first state it as a two-sided
zero-contrast test and say which of the two errors would be more costly
for you and why, then name the one power lever you could actually pull
before your next pitch.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "My claim: our conversion rate differs from the sector's 3% --- $H_0\!: p = 0.03$ against a two-sided alternative. A false alarm is the costlier error, because I would sign a bigger kitchen lease on an effect that is not real; the one power lever I can pull before the next pitch is sample size, by running the tasting stand two more weeks."
:::

You can discuss this task --- and any other task in this chapter --- with
**Tilly**, the course AI assistant on Canvas: ask it to draft the
seven-step workflow for your claim and challenge your choice of the more
costly error.
::::

Continue to Tutorial 4: Inference and tests against zero.

Formula sheet: Lecture 10 formulas.

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 15 and 16.

:::
