# Lecture 10: Testing decisions, errors, power, and the road to regression {#b10}

```{r}
#| label: b10-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

::: {.outcomes}
<span class="label">By the end of Lecture 10, you can</span>

- rewrite benchmark and comparison questions as tests of a zero contrast
- complete and communicate a two-sided test using the seven-step workflow
- distinguish a p-value, a significance level, and a test decision
- describe Type I error, Type II error, and power in context
- interpret a two-sided p-value and connect a 95% confidence interval to a 5%
  two-sided test
- separate statistical significance from magnitude, practical importance,
  and causality
- interpret the zero-null test printed for a regression coefficient.
:::

::: {.must-know}
<span class="label">Must know</span>

**Be able to do yourself:**

- explain a p-value in one sentence
- make the decision at 5% and link it to the 95% interval
- test a proportion by hand with the null standard error
- describe a Type I and a Type II error in context and name two ways to raise power
- read a regression row: estimate, standard error, t, p-value, interval.

**Enough to recognise:**

- the critical-value rule: reject $H_0$ at 5% when $|t|\ge t_{0.975,\,df}$
- the power figure.
:::

::: {.driving-question}
<span class="label">Driving question</span>

<p class="dq-context">Launch the product or wait. Approve the loan or decline. Decisions made from data can be wrong in two directions, and the two mistakes usually cost very different amounts.</p>

Every test can be wrong in two ways. Which error is which, what makes
errors rarer, and how does all of this reappear in a regression table?
:::

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 25 minutes.</span>

**Minimum path:** the video preparation and the prediction. **Full path:** everything in this section.

<div class="video-prep">
<span class="label">Video preparation (15 minutes)</span><br>
Review <a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to
Hypothesis Testing</a> and watch
<a href="https://www.youtube.com/watch?v=JNOWWJeZBTc">Examples Identifying
Type I and Type II Errors</a>. Focus on two-sided tests, decision errors,
and why "do not reject" is not "accept." All links are also listed in the
<a href="../resources/video-guide.html">External Video Guide</a>.
</div>

**Reactivate (8 minutes).** Lecture 9 ended with the t statistic for the
equity gap, $t=-2.91$ with 99 degrees of freedom.

1. Write the two hypotheses of that test.
2. Say in one sentence what the number $-2.91$ measures.
3. Guess: if the true gap were zero, would a t statistic this far from zero
   be common or rare?

::: {.callout-note collapse="true"}
## Check your preparation

$H_0:\delta=0$ against $H_1:\delta\ne0$, where $\delta$ is the population
mean of the equity offered minus 15. The statistic says that the estimated
gap lies 2.91 standard errors below zero. Such a value would be rare: about
95% of a t distribution with 99 degrees of freedom lies within 1.98 of zero.
The p-value says exactly how rare. It is the first topic of the lecture
material.
:::

**Predict (2 minutes).** Answer the one-click Wooclap question posted on
Canvas before you come to class: a test at $\alpha=0.05$ returns
$p=0.20$. Has the study **shown that the null is true**: yes, no, or
cannot tell? Commit to an answer. You find the answer in the lecture
material.

Review [Mathematics Refresher: zero contrasts, inequalities, and
standardisation](../resources/math-refresher.html) if needed.

:::

::: {.course-phase .phase-in-class}

## Lecture material

### Start with a founder's question

An advisor said that founders offer 15% of their company on average. Your
sample of 100 pitches says 12.76%. Should you tell the advisor that the
number is wrong? And what does it cost you if you are the one who is wrong?
Lecture 9 ended with a t statistic and no decision. This section adds the
p-value and the decision rule. It then gives one workflow for every test
and names the two ways a decision can be wrong. It ends with the same test
in a regression table, the main object of QM Intermediate.

<p class="concept-video"><strong>Video explanations (review):</strong>
<a href="https://www.youtube.com/watch?v=ti9NFdjf3sM">Introduction to Hypothesis Testing</a> and
<a href="https://www.youtube.com/watch?v=JNOWWJeZBTc">Type I and Type II Errors</a>.</p>

### Where Lecture 9 stopped

Lecture 9 compared the mean equity offered (`equity_offered_pct`) with the
benchmark of 15 percentage points, using the random sample of 100 pitches
(seed 123). The estimated gap was $-2.24$ points with standard error $0.77$,
so $t=-2.91$ with 99 degrees of freedom. The estimate lies almost three
standard errors below zero. The open question is how surprising such a value
would be if the true gap were zero. The p-value answers it.

### A two-sided p-value uses both tails

::: {.definition-box}
<span class="label">Definition</span>

**p-value.** The probability, calculated as if $H_0$ were true, of a test
statistic at least as far from zero as the observed one, in either
direction. A small p-value means that the data would be surprising if $H_0$
were true.
:::

For the equity gap, with $t_{obs}$ the observed t statistic:

$$
p=P\!\left(|T_{99}|\ge|t_{obs}|\mid H_0\right).
$$

Here $T_{99}$ is a variable that follows the t distribution with 99 degrees
of freedom.

In R, `pt()` gives the area under a t curve to the left of a value. The
two-sided p-value is twice the area to the left of $-|t_{obs}|$:

```{r}
#| label: b10-gap-p-value
t_value <- -2.906   # the t statistic from Lecture 9
df <- 99
p_value <- 2 * pt(-abs(t_value), df = df)
c(t_statistic = t_value, degrees_of_freedom = df, two_sided_p_value = p_value)
```

The p-value is 0.0045. @fig-b09-two-sided-p shows it as an area.

```{r}
#| label: fig-b09-two-sided-p
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "The t distribution with 99 degrees of freedom under the null hypothesis. Dashed lines mark statistics as far from zero as the observed t = −2.91. The two orange tail areas together are the two-sided p-value, about 0.005."
library(ggplot2)

sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
pitch_sample$equity_gap_pp <- pitch_sample$equity_offered_pct - 15
gap_bar <- mean(pitch_sample$equity_gap_pp)
gap_se <- sd(pitch_sample$equity_gap_pp) / sqrt(nrow(pitch_sample))
df <- nrow(pitch_sample) - 1
t_value <- gap_bar / gap_se

null_curve <- data.frame(x = seq(-4.5, 4.5, length.out = 1200))
null_curve$density <- dt(null_curve$x, df = df)
left_tail <- subset(null_curve, x <= -abs(t_value))
right_tail <- subset(null_curve, x >= abs(t_value))

ggplot(null_curve, aes(x, density)) +
  geom_area(data = left_tail, fill = "#D55E00", alpha = 0.85) +
  geom_area(data = right_tail, fill = "#D55E00", alpha = 0.85) +
  geom_vline(xintercept = c(-abs(t_value), abs(t_value)),
             linetype = "dashed", colour = "#D55E00", linewidth = 0.6) +
  geom_line(colour = "#17324D", linewidth = 0.9) +
  annotate("text", x = -abs(t_value), y = 0.33,
           label = "t == -2.91", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = 1.15) +
  annotate("text", x = abs(t_value), y = 0.33,
           label = "t == +2.91", parse = TRUE,
           colour = "#17324D", size = 3.4, hjust = -0.15) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(x = "t statistic under the null", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

Read @fig-b09-two-sided-p with the five-step sequence. (1) *Unit and
variables:* the curve is not data. It is the distribution of the t statistic
**if the null were true**. The horizontal axis is the possible t statistic.
(2) *Scales:* the vertical axis is density, so probability is area. (3) *One
element:* the dashed lines mark $-2.91$ and $+2.91$, statistics as far from
zero as the one observed. (4) *Pattern:* the two orange tail areas beyond the
dashed lines are tiny, together about 0.0045. If the true mean gap were zero,
a sample this far from zero would occur in roughly 5 of every 1,000 samples.
(5) *Limits:* that number is **not** the probability that $H_0$ is true, and
it is not the size of the gap.

The convention is to reject $H_0$ when the p-value is at most 0.05. This
threshold is called the significance level.

::: {.definition-box}
<span class="label">Definition</span>

**Significance level $\alpha$.** The threshold for the test decision, fixed
before the data are inspected, usually $\alpha=0.05$. Reject $H_0$ when
$p\le\alpha$. Do not reject $H_0$ when $p>\alpha$.
:::

Here $p=0.0045<0.05$, so reject $H_0$. The random sample provides evidence
that the mean equity offered in the recorded-pitch frame differs from 15
percentage points. The negative estimate indicates that it is lower.

The same decision can be read from the t statistic. Reject $H_0$ at the 5%
level when $|t|$ is at least $t_{0.975,\,df}$, the critical value of
Lecture 9. In symbols, $|t|\ge t_{0.975,\,df}$. That is the same decision as
$p\le0.05$.

```{r}
#| label: b10-critical-value
c(abs_t = abs(t_value), critical_value = qt(0.975, df = df))
```

Here $|t|=2.91$ is larger than $t_{0.975,\,99}=1.984$, so the critical-value
rule also rejects $H_0$.

### The confidence interval and test tell the same story

For the same two-sided t procedure:

- reject $H_0:\delta=0$ at level $\alpha$ exactly when the
  $100(1-\alpha)\%$ confidence interval excludes zero
- do not reject when the interval includes zero.

Here the gap interval $(-3.77,\,-0.71)$ from Lecture 9 excludes zero, in
line with the rejection. The interval adds information that the p-value does
not give. It shows the direction, the plausible size, and the precision of
the difference. In this course and in QM Intermediate, always report the
estimate and the interval together with the test decision.

::: {.callout-caution collapse="true"}
## Try it: read the decision straight from an interval

Try this with your neighbour first. A colleague reports a 95% confidence
interval for a mean gap: $(-1.4,\,3.8)$ percentage points (invented
numbers). The report gives no t statistic and no p-value.

*Step 1 (given).* The two-sided t test rejects $H_0:\delta=0$ at 5% exactly
when the 95% interval excludes zero.

*Step 2 (you).* Does this interval exclude zero? State the test decision
without computing anything.

*Step 3 (you).* What may you conclude about the population mean gap,
and what may you *not* conclude?

::: {.callout-note collapse="true"}
## Check your answer

Zero lies inside $(-1.4,\,3.8)$, so the 5% two-sided test does not reject
$H_0:\delta=0$. The data are compatible with a zero gap, and also with gaps
as large as 3.8 points in one direction and 1.4 in the other. You may not
conclude that the gap is zero. "Do not reject" is not "accept". The workflow
below returns to this difference.
:::
:::

### Why zero is the universal benchmark

Many questions about data ask whether a population contrast is zero:

- difference from a target: $\delta=\mu-15$
- difference in probabilities: $\delta_p=p-0.50$
- difference between two group means: $\mu_1-\mu_0$
- a regression slope: $\beta_j$.

In every case, "no difference" or "no linear association" becomes

$$
H_0:\theta=0
\qquad\text{versus}\qquad
H_1:\theta\ne0.
$$

Here $\theta$ stands for any of the four parameters in the list. A test
against a non-zero benchmark can always be rewritten as a test of a contrast
equal to zero, as with the gap in Lecture 9. The tests in a regression table
have the same structure.

### The seven-step testing workflow

Use the same sequence for every test:

1. Define the population parameter and its units.
2. State $H_0$ and the two-sided $H_1$ before you inspect the result.
3. Check the conditions on the sample and on the shape of the data.
4. Calculate the estimate and its SE. For a proportion test, calculate the
   SE with the null value $p_0$.
5. Standardise: $(\text{estimate}-0)/\text{SE}$.
6. Calculate the two-sided p-value and compare it with the significance
   level $\alpha$ that was fixed in advance.
7. Report the estimate, its uncertainty, the decision, the meaning in
   context, and the limits.

The significance level $\alpha$ and the p-value are different things.
$\alpha$ is a rule, fixed in advance. The p-value is a result, calculated
from the observed data as if $H_0$ were true.

If $p\le\alpha$, reject $H_0$. If $p>\alpha$, do not reject $H_0$, and do
not say "accept $H_0$". This answers the prediction question from the
preparation: a study with $p=0.20$ has not shown that the null hypothesis is
true. The study may only be too imprecise to tell a real non-zero parameter
from zero. The section on power below explains this.

::: {#exm-b10-launch-test}
## Worked example: a proportion test by hand

This is an invented example. An accelerator claims that half of its alumni
launch a trading venture within a year. In a random sample of 400 former
participants, 232 had launched. Let $p$ be the launch proportion among all
alumni and define $\delta_p=p-0.50$. Test $H_0:\delta_p=0$ against
$H_1:\delta_p\ne0$ at 5%, step by step.

- **Estimate.** $\hat p=232/400=0.58$, so the estimated gap is
  $\widehat\delta_p=0.58-0.50=0.08$, eight percentage points.
- **Conditions under the null.** The test uses a normal approximation, so
  check the success-failure rule of
  [Lecture 8](b08-sampling-distributions.qmd#b08-binomial): $np\ge10$ and
  $n(1-p)\ge10$. Lecture 8 gives the reason for the rule. In a test, use the
  null value $p_0$: $np_0=400(0.50)=200$ and $n(1-p_0)=200$. Both are far
  above 10, so the normal approximation can be used.
- **Null SE.** Under $H_0$ the true proportion is 0.50, so
  $\operatorname{SE}_0=\sqrt{0.50(0.50)/400}=\sqrt{0.25/400}
  =\sqrt{0.000625}=0.025$.
- **Standardise.** $z=0.08/0.025=3.20$. The observed gap lies 3.2 null
  standard errors above zero. The statistic is called $z$ here because it is
  compared with the standard normal distribution.
- **p-value and decision.** The two-sided p-value is
  $2\,P(Z\le-3.20)\approx 0.0014 < 0.05$, so reject $H_0$. The sample
  provides strong evidence that the launch proportion among all alumni
  differs from 50%. The positive estimate places it above 50%.

**Report the interval too.** A decision alone is not a report. The
large-sample 95% interval uses the estimated SE:
$0.58\pm1.96\sqrt{0.58(0.42)/400}=0.58\pm0.048=(0.532,\,0.628)$. It excludes
0.50, in line with the test.

Look again at step 4 of the workflow. The SE was calculated with $p_0=0.50$,
not with $\hat p$. The test measures how surprising the sample would be if
$H_0$ were true, so the null value supplies the spread.
:::

### In R: the Shark Tank deal proportion

Tutorial 4, Task 2 uses the code of this section and the next. Read both
sections before the tutorial.

Now apply the same workflow to the random sample of 100 pitches from
Lecture 9. Is the proportion of pitches that end with a deal
(`deal_on_show`) in the recorded-pitch frame different from 50%? The code
builds every object step by step:

```{r}
#| label: b10-proportion-objects
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

n <- nrow(pitch_sample)
phat <- mean(pitch_sample$deal_on_show)   # sample deal proportion
p0 <- 0.50                                # null value
deal_gap <- phat - p0                     # estimated gap

c(n = n, phat = phat, p0 = p0, deal_gap = deal_gap)
```

```{r}
#| label: b10-proportion-test
null_se <- sqrt(p0 * (1 - p0) / n)
z_value <- deal_gap / null_se
two_sided_p <- 2 * pnorm(-abs(z_value))

c(
  null_standard_error = null_se,
  z_statistic = z_value,
  two_sided_p_value = two_sided_p
)
```

In the sample, 64 of the 100 pitches end with a deal, so $\hat p=0.64$. The
estimate is $\widehat\delta_p=0.14$, fourteen percentage points. The null SE
is $\sqrt{0.50(0.50)/100}=0.05$, the statistic is $z=2.80$, and the
two-sided p-value is about 0.005. The success-failure rule holds under the
null: $np_0=50$ and $n(1-p_0)=50$. Reject the zero-gap null at 5%. The
random sample provides evidence that the deal proportion in the
recorded-pitch frame differs from 50%, and the positive estimate places it
above 50%.

### Interval and test: report both

Step 7 of the workflow says a decision alone is not a report. A
large-sample 95% interval for the proportion uses the estimated SE:

$$
\hat p\pm1.96
\sqrt{\frac{\hat p(1-\hat p)}{n}}.
$$

```{r}
#| label: b10-proportion-interval
estimated_se <- sqrt(phat * (1 - phat) / n)
proportion_ci <- phat + c(-1, 1) * 1.96 * estimated_se
gap_ci <- proportion_ci - p0

c(proportion_lower = proportion_ci[1],
  proportion_upper = proportion_ci[2])
c(gap_lower = gap_ci[1], gap_upper = gap_ci[2])
```

The proportion interval is approximately $(0.546,\,0.734)$. The equivalent
gap interval is $(0.046,\,0.234)$, which excludes zero and agrees with the
test decision here.

::: {.callout-tip collapse="true"}
## Optional: estimated versus null standard errors

The test divided by the null SE $\sqrt{p_0(1-p_0)/n}=0.050$. The interval
used the estimated SE $\sqrt{\hat p(1-\hat p)/n}=0.048$. The test asks how
surprising the sample would be if $H_0$ were true, so the null value
supplies the spread. The interval does not assume $H_0$, so the data supply
the spread. The two procedures are therefore not exactly the same. When a
result lies very close to the 5% boundary, the interval and the test can
disagree. An interval that always agrees with the test exists, but it is
beyond this course. For the mean gap, the interval of Lecture 9 and the t
test of this lecture use the same SE, so there they always agree.
:::

### What a p-value is not

The two-sided p-value is the probability

$$
P\!\left(|\text{test statistic}|
\ge|\text{observed statistic}|\mid H_0\right)
$$

and nothing else. The task below gives four sentences that read more into
it. All four are wrong.

::: {.callout-caution collapse="true"}
## Try it: correct the four wrong p-value sentences

Try this with your neighbour first. The four sentences are about the launch
test of @exm-b10-launch-test ($z=3.20$, $p\approx 0.0014$). Each one is
wrong. Write a correct version of each sentence.

*Sentence 1 (you).* "There is only a 0.14% probability that the launch
proportion really is 50%."

*Sentence 2 (you).* "The result has a 0.14% probability of being due to
chance."

*Sentence 3 (you).* "Such a tiny p-value shows that the launch proportion
is far above 50%."

*Sentence 4 (you).* "The tiny p-value proves that the accelerator causes
its alumni to launch ventures."

::: {.callout-note collapse="true"}
## Check your answers

1. A p-value is not $P(H_0\mid\text{data})$. Correct: "If the launch
   proportion were exactly 50%, a sample gap at least this far from zero
   in either direction would occur in about 0.14% of random samples."
2. The p-value is calculated under the assumption that $H_0$ is true and
   chance alone produced the gap. It is the probability of data this
   extreme under that assumption. It is not the probability that the
   assumption holds. Correct: "If chance alone were at work, a gap this
   large would occur in about 0.14% of random samples."
3. A p-value measures surprise, not size. The size is the estimate, 8
   percentage points, with an interval around it. A tiny p-value can go
   together with a tiny gap when $n$ is very large. Correct: "The launch
   proportion is estimated at 58%, plausibly between 53.2% and 62.8%."
4. A significant result says nothing about the cause. The alumni chose to
   join the accelerator themselves, so they may differ from other founders
   in many ways. Correct: "The launch proportion among all alumni differs
   from 50%."
:::
:::

### Type I error, Type II error, and power

A test is a decision rule, and a decision can be wrong in two ways:

| Reality | Reject $H_0$ | Do not reject $H_0$ |
|---|---|---|
| $H_0$ true | Type I error | Correct decision |
| $H_1$ true | Correct decision | Type II error |

::: {.definition-box}
<span class="label">Definition</span>

**Type I error and Type II error.** A Type I error is rejecting $H_0$ when
$H_0$ is true (a false alarm). A Type II error is not rejecting $H_0$ when
$H_1$ is true (a missed difference).
:::

The significance level is the probability of a Type I error:

$$
P(\text{reject }H_0\mid H_0\text{ true})=\alpha.
$$

::: {.definition-box}
<span class="label">Definition</span>

**Power.** The probability that the test rejects $H_0$ when $H_1$ is true:
$\text{power}=P(\text{reject }H_0\mid H_1\text{ true})$. Power is one minus
the probability of a Type II error.
:::

Power depends on how far the true parameter lies from zero, so it is always
calculated for one given alternative value.

In the launch test of @exm-b10-launch-test, a Type I error would be to
conclude that the launch proportion among all alumni differs from 50% when
it is exactly 50%. A Type II error would be to miss a launch proportion
that really differs from 50%, for example because the sample was too small.

```{r}
#| label: fig-b10-errors-power
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Two-sided testing at α = 0.05. Left: when the null is true, the z statistic follows the dark curve and the orange rejection regions have total probability 0.05, the Type I error rate. Right: when a positive alternative is true, the statistic follows the shifted blue curve. The green area is the power, and the unshaded middle is the Type II error probability. Both panels use the same axes."
library(ggplot2)

z_grid <- seq(-4.5, 6.5, length.out = 1600)
critical <- qnorm(0.975)
panel_levels <- c("Null is true", "An alternative value is true")

test_curves <- rbind(
  data.frame(z = z_grid, density = dnorm(z_grid, mean = 0),
             panel = panel_levels[1]),
  data.frame(z = z_grid, density = dnorm(z_grid, mean = 2.5),
             panel = panel_levels[2])
)
test_curves$panel <- factor(test_curves$panel, levels = panel_levels)

rejection <- subset(test_curves, z <= -critical | z >= critical)
rejection$side <- ifelse(rejection$z < 0, "left", "right")

panel_labels <- data.frame(
  z = c(0, 3.6),
  density = c(0.13, 0.14),
  label = c("do not reject", "power"),
  panel = factor(panel_levels[c(1, 2)], levels = panel_levels)
)
alpha_labels <- data.frame(
  z = c(-2.5, 2.5),
  density = c(0.028, 0.028),
  label = c("alpha/2", "alpha/2"),
  panel = factor(panel_levels[c(1, 1)], levels = panel_levels)
)

ggplot(test_curves, aes(z, density)) +
  geom_area(data = rejection, aes(group = side, fill = panel),
            alpha = 0.4) +
  geom_vline(xintercept = c(-critical, critical), linetype = "dashed",
             colour = "#5B6976", linewidth = 0.5) +
  geom_line(aes(colour = panel), linewidth = 0.9) +
  geom_text(data = panel_labels, aes(label = label),
            colour = "#17324D", size = 3.4) +
  geom_text(data = alpha_labels, aes(label = label), parse = TRUE,
            colour = "#17324D", size = 3.1) +
  facet_wrap(~ panel, scales = "fixed") +
  scale_fill_manual(values = c("Null is true" = "#D55E00",
                               "An alternative value is true" = "#009E73"),
                    guide = "none") +
  scale_colour_manual(values = c("Null is true" = "#17324D",
                                 "An alternative value is true" = "#0072B2"),
                      guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  coord_cartesian(xlim = c(-4, 6)) +
  labs(x = "z statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

In both panels of @fig-b10-errors-power, the horizontal axis is the
possible z statistic, the vertical axis is density, and probability is
area. The dashed vertical lines mark the same two-sided critical values,
$\pm 1.96$. The test rejects $H_0$ when the statistic falls beyond these
lines. This part of the axis is called the rejection region. In the left
panel the null is true. The two orange tails have total probability
$\alpha=0.05$, which is the Type I error rate. The test is built to give a
false alarm in 5% of samples when $H_0$ is true. In the right panel one
positive alternative value is true, so the curve has shifted to the right.
The green area is the power. The probability that remains between the
dashed lines is the Type II error rate.

Four changes raise power:

- a larger true distance from zero (a difference that is easier to see)
- a larger sample (a smaller SE)
- less variation in the data (also a smaller SE)
- a larger $\alpha$. This last one has a price. A wider rejection region
  finds more true differences, but it also raises the Type I error rate.

::: {.callout-tip collapse="true"}
## Optional: one-sided alternatives

This box is not covered in the lecture and is not examined. All tests in
this course are two-sided.

A two-sided test asks whether a parameter differs from zero in either
direction. A one-sided test asks about one direction only:

| Research question | Alternative | Tail that counts |
|---|---|---|
| Does the gap differ from zero? | $H_1:\delta\ne0$ | both tails |
| Is the gap below zero? | $H_1:\delta<0$ | left tail |
| Is the gap above zero? | $H_1:\delta>0$ | right tail |

The t statistic is the same in all three cases. Only the rejection region
changes. A one-sided test at $\alpha=0.05$ puts the whole 5% in one tail, as
@fig-b10-one-sided shows.

```{r}
#| label: fig-b10-one-sided
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-width: 7
#| fig-height: 3.25
#| fig-cap: "Rejection regions at α = 0.05 for the two-sided and both one-sided alternatives, on the same null t distribution (df = 99). The two-sided test places 2.5% in each tail. A one-sided test moves the full 5% into the tail named by its alternative, so its critical value is closer to zero in that direction."
library(ggplot2)

alpha_tail <- 0.05
tail_df <- 99
tail_grid <- seq(-4, 4, length.out = 1400)
critical_two <- qt(c(alpha_tail / 2, 1 - alpha_tail / 2), df = tail_df)
critical_left <- qt(alpha_tail, df = tail_df)
critical_right <- qt(1 - alpha_tail, df = tail_df)

panel_levels <- c("Two-sided", "Left-sided", "Right-sided")
null_curve <- do.call(rbind, lapply(panel_levels, function(p) {
  data.frame(x = tail_grid, density = dt(tail_grid, df = tail_df),
             panel = p)
}))
null_curve$panel <- factor(null_curve$panel, levels = panel_levels)

rejection <- rbind(
  cbind(subset(null_curve, panel == "Two-sided" & x <= critical_two[1]),
        side = "left"),
  cbind(subset(null_curve, panel == "Two-sided" & x >= critical_two[2]),
        side = "right"),
  cbind(subset(null_curve, panel == "Left-sided" & x <= critical_left),
        side = "left"),
  cbind(subset(null_curve, panel == "Right-sided" & x >= critical_right),
        side = "right")
)

critical_lines <- data.frame(
  x = c(critical_two, critical_left, critical_right),
  panel = factor(c("Two-sided", "Two-sided", "Left-sided", "Right-sided"),
                 levels = panel_levels)
)

panel_labels <- data.frame(
  x = c(-2.8, 2.8, -2.8, 2.8, 0, 0, 0),
  density = c(0.035, 0.035, 0.035, 0.035, 0.42, 0.42, 0.42),
  label = c("2.5%", "2.5%", "5%", "5%",
            "H[1] : delta != 0", "H[1] : delta < 0", "H[1] : delta > 0"),
  parse_me = c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE),
  panel = factor(c("Two-sided", "Two-sided", "Left-sided", "Right-sided",
                   "Two-sided", "Left-sided", "Right-sided"),
                 levels = panel_levels)
)

ggplot(null_curve, aes(x, density)) +
  geom_area(data = rejection, aes(group = side),
            fill = "#D55E00", alpha = 0.6) +
  geom_vline(xintercept = 0, linetype = "dotted",
             colour = "grey75", linewidth = 0.4) +
  geom_vline(data = critical_lines, aes(xintercept = x),
             linetype = "dashed", colour = "#5B6976", linewidth = 0.5) +
  geom_line(colour = "#17324D", linewidth = 0.8) +
  geom_text(data = subset(panel_labels, !parse_me), aes(label = label),
            colour = "#17324D", size = 2.9) +
  geom_text(data = subset(panel_labels, parse_me), aes(label = label),
            parse = TRUE, colour = "#5B6976", size = 3.1) +
  facet_wrap(~ panel, nrow = 1, scales = "fixed") +
  scale_x_continuous(breaks = c(-3, 0, 3)) +
  scale_y_continuous(breaks = c(0, 0.2, 0.4), limits = c(0, 0.44),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(x = "t statistic", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D"),
    strip.text = element_text(colour = "#17324D", face = "bold")
  )
```

A one-sided alternative is legitimate only when the direction follows from
the research question and is fixed before the data are inspected. Choosing
the tail after seeing the estimate doubles the true Type I error rate, from
5% to about 10%.

The one-sided p-value follows from the two-sided one. When the estimate
points in the direction of the alternative, it is half the two-sided
p-value. When the estimate points in the other direction, it is one minus
that half. For the equity gap, $t=-2.91$ with $df=99$:

```{r}
#| label: b10-one-sided-p
t_gap <- -2.906   # the t statistic of the equity gap
df_gap <- 99
c(two_sided = 2 * pt(-abs(t_gap), df = df_gap),
  left_sided = pt(t_gap, df = df_gap),
  right_sided = 1 - pt(t_gap, df = df_gap))
```

The two-sided p-value is 0.0045. The left-sided alternative $H_1:\delta<0$
gives 0.0023, and the right-sided alternative $H_1:\delta>0$ gives 0.9977.
:::

### Preview: the test attached to a regression slope

A regression fits a straight line through a cloud of points. Here the line
describes how the equity offered (`equity_offered_pct`) changes from season
to season (`season`). Its slope says by how much per season on average. R
tests every regression coefficient against zero in the way described above.
The code fits the line to all 1,441 records of the Shark Tank file:

```{r}
#| label: b10-regression-preview
preview_model <- lm(equity_offered_pct ~ season, data = sharks)
summary(preview_model)$coefficients
confint(preview_model)
```

One point before you read the output. In Lectures 7 to 9 we sampled 100
records from the file, so the uncertainty came from the random draw. Here
the model uses all 1,441 records. The standard error now answers a different
question: how much would the slope vary if the same process produced another
set of pitches? The recorded pitches are treated as one possible outcome of
that process.

Now read the `season` row. The slope of the line in the population is
written $\beta_1$, and its estimate is $\hat\beta_1$.

The estimate is $\hat\beta_1=-0.824$. Each additional season goes together
with about 0.82 percentage points less equity offered. For comparison, the
mean equity offered in the file is 13.21 points, so this is about 6% of the
mean per season. The SE, 0.048, is the sampling uncertainty of the estimate,
in the same units. The t value, $-17.3$, is the estimate divided by the SE.
This is step 5 of the workflow. The p-value in the last column belongs to
the two-sided test

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1\ne0.
$$

The p-value is far below 0.05, so reject $H_0$. The 95% interval
$(-0.92,\,-0.73)$ excludes zero, in line with the rejection.

Every column is something you know from Lectures 9 and 10. The row does not
say that a later season causes lower offers. Nobody assigned pitches to
seasons, so this is an association over time, not a causal effect.

QM Intermediate starts from this row (estimate, SE, t, p-value, interval).
It adds the interpretation of coefficients, models with several predictors,
and checks of the model.

::: {.caveat-box}
<span class="label">What a significant result does and does not establish</span>

- Significant is not large. A tiny gap can be significant with a lot of data,
  and a real gap can be missed with little.
- Significant is not causal. The test does not know how the data came about.
- No test repairs the data. Selection and measurement stay as they were.

**What to report, every time:** the estimate with its units, the interval,
the decision, whether the size matters for the business question, and the
limits of the data.

**What you can do with it:** weigh the two errors before you decide how much
evidence you need.
:::

### Answering the driving question

The two wrong decisions are rejecting a true null hypothesis (Type I error,
with probability $\alpha$) and missing a true difference (Type II error,
less likely when power is high). Type II errors become rarer with more data,
less variation, and larger true differences. Type I errors become rarer with
a smaller $\alpha$, at the price of more Type II errors. In a regression
table, each coefficient row is the test of Lectures 9 and 10: estimate, SE,
t, two-sided p-value against zero, and interval.

This completes QM Basic. You can now define events and variables on a
frame, describe their distributions, work with conditional probabilities,
and model sampling variation. You can turn "our sample mean is 12.76" into a
claim about a population, with its uncertainty, its decision, and its
limits.

::: {.founder-box}
<span class="label">A founder's decision (invented example)</span>

Your sign-up page converts 10% of visitors. You try a new page on 400 visitors and 52 sign up, which is 13%. The null standard error is 0.015, so z = 0.03/0.015 = 2.0 and the two-sided p-value is about 0.046. Now weigh the two errors. Switching pages for nothing costs you almost nothing, because you can switch back. Missing a page that is really better costs you sign-ups every week. The decision: switch. For a kitchen lease the logic turns around. A false alarm there is expensive and hard to undo, so you would ask for more data and stronger evidence before you sign.
:::

**Next:** Tutorial 4 practises Lectures 9--10 in exam format. The Mock Exam
then covers the whole course under exam conditions. QM Intermediate, in the
second year, continues from the regression preview above. The coefficient
row that you can already read is the main object of its six lectures on
linear regression.

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 85 minutes.</span>

**Minimum path (about 35 minutes):** read the lecture material, answer the recall questions, then work for 10 minutes on the tasks that follow. **Full path:** all tasks below.

### Read the lecture material (15 minutes)

Read the Lecture material section above from start to end. The lecture
covered the core of it. The text has every step, every example, and all the
code, including parts that were not shown in class. Note what is still
unclear and bring it to the next tutorial, or ask Tilly.

### Recall questions (10 minutes)

1. Why can most comparison hypotheses be written as a parameter equal to
   zero?
2. What is the difference between $\alpha$ and a p-value?
3. What is a Type I error?
4. What is a Type II error?
5. Name three factors that increase power.
6. Why is "do not reject" different from "accept"?
7. What does $H_0:\beta_j=0$ mean in a regression model?

::: {.callout-note collapse="true"}
## Check your answers

1. Subtracting the benchmark or reference value creates a contrast. "No
   difference" then means that the contrast is zero.
2. $\alpha$ is the decision threshold, fixed in advance. The p-value is
   calculated from the observed statistic as if $H_0$ were true.
3. Rejecting $H_0$ when it is true.
4. Not rejecting $H_0$ when $H_1$ is true.
5. Any three of: a larger sample, a larger true difference, less variation
   in the data, or a larger $\alpha$ (the last at the cost of more Type I
   errors).
6. A result that is not significant may only be too imprecise to tell a
   real non-zero parameter from zero.
7. The population slope for that predictor is zero. The outcome has no
   linear association with it.
:::

### Finish a half-worked example, then do one on your own (20 minutes)

Re-read the hand-worked launch test (@exm-b10-launch-test). Then complete
the same steps with new invented numbers: in 200 independent trials there
are 120 successes, and the benchmark is $p_0=0.50$.

*Step 1 (given).* $\hat p=120/200=0.60$, so $\widehat\delta_p=0.10$. The
null counts are $200(0.50)=100$ successes and 100 failures, both above 10.

*Step 2 (you).* Calculate the null SE and the z statistic.

*Step 3 (you).* The two-sided p-value is about 0.005. Give the 5% decision
in context.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SE}_0=\sqrt{0.50(0.50)/200}\approx 0.0354$, so
$z=0.10/0.0354\approx 2.83$. The two-sided p-value, about 0.005, is below
0.05. Reject the zero-gap null at 5%. The data provide evidence that the
success proportion differs from 50%, and the positive estimate places it
above 50%.
:::

Now solve these tasks on your own:

1. A random sample of $n=49$ differences has mean $\bar d=3$ and sample SD
   $s_d=14$ (invented numbers). Test $H_0:\delta=0$ against
   $H_1:\delta\ne0$ at 5%. The two-sided p-value for your t statistic is
   about 0.14.
2. Describe the Type I and Type II errors for the 120-of-200 test above,
   in context.
3. A regression of the equity offered (in percentage points) on the amount
   requested has the coefficient $-0.360$ percentage points per \$100,000,
   with SE 0.090 and a large $df$ (invented numbers). Calculate its t
   statistic and approximate 95% interval, and interpret the row.
4. A coefficient estimate is too small to matter for any business decision,
   but it has $p<0.001$. Explain how both facts can be true.
5. State two reasons why the regression preview cannot support a causal
   conclusion.

::: {.callout-note collapse="true"}
## Check your answers

1. The SE is $14/\sqrt{49}=2$, so $t=3/2=1.50$ with $df=48$. The two-sided
   p-value, about 0.14, is above 0.05. Do not reject the null hypothesis of
   a zero mean difference at 5%. This is not evidence that the difference
   is zero.
2. Type I: concluding that the population success proportion differs from
   0.50 when it equals 0.50. Type II: not detecting a success proportion
   that really differs from 0.50.
3. $t=-0.360/0.090=-4.00$. With a large $df$, the 95% interval is
   approximately $-0.360\pm1.96(0.090)=(-0.536,\,-0.184)$ percentage
   points per \$100,000. An additional \$100,000 requested goes together
   with between about 0.18 and 0.54 percentage points less equity offered.
   The interval excludes zero and $|t|=4.00$ is far above 1.96, so reject
   the null hypothesis of a zero coefficient.
4. A very large sample or very little variation makes the SE tiny. Then
   even a very small coefficient lies many SEs from zero. A precise
   estimate is not automatically an important one.
5. Nobody assigned pitches to seasons at random. Other factors that
   changed over the seasons, the selection of pitches for the show, and
   measurement error can all play a role.
:::

### Optional: one-sided reasoning

This task belongs to the optional box on one-sided alternatives. It is not
examined.

The equity gap gave $t=-2.91$ with $df=99$ and a two-sided p-value of
0.0045.

1. State the left-sided ($H_1:\delta<0$) and the right-sided
   ($H_1:\delta>0$) p-value.
2. Which of the three alternatives could have been chosen legitimately, and
   what decides that?
3. A classmate sees the negative estimate and then switches to the
   left-sided alternative "because it fits the data". What is wrong?

::: {.callout-note collapse="true"}
## Check your answers

1. Left-sided: about 0.0023, half of the two-sided p-value. Right-sided:
   about 0.9977, one minus that half.
2. Any of the three, if it follows from the research question and was
   fixed before the sample was inspected.
3. If the tail is chosen after seeing the estimate, the 5% always lies on
   the side where the estimate happens to fall. The true Type I error rate
   is then about 10%, not 5%. All tests in this course are two-sided.
:::

### Review: one question each from Lectures 9 and 5 (10 minutes)

1. *(Lecture 9.)* A 95% confidence interval for a mean gap is
   $(0.4,\,5.2)$ units (invented numbers). Give the estimate and the margin
   of error, and say what "95%" refers to.
2. *(Lecture 5.)* In the full Shark Tank file, the correlation between
   season (`season`) and equity offered (`equity_offered_pct`) is about
   $-0.42$. Interpret this number in one sentence, and state one claim it
   cannot support.

::: {.callout-note collapse="true"}
## Check your answers

1. The estimate is the midpoint, $(0.4+5.2)/2=2.8$ units. The margin of
   error is half the width, $(5.2-0.4)/2=2.4$ units. "95%" refers to the
   procedure: about 95% of intervals built in this way from random samples
   contain the parameter.
2. Season and equity offered have a moderate negative linear association.
   Pitches in later seasons tend to offer less equity. The correlation
   cannot support a causal claim. Nothing assigned pitches to seasons, and
   other factors changed over time.
:::

### Worked exam interpretation (10 minutes)

**Question (same format as Mock Exam Question 5).** This is an invented
example. A crowdfunding platform states that 40% of its campaigns reach
their funding goal. In a simple random sample of 600 campaigns from last
year, 270 reached the goal. Let $p$ be the proportion of all campaigns of
last year that reached the goal. A two-sided test of $H_0:p=0.40$ is
planned. The same study fits a regression of the amount raised (thousands
of euros) on the number of updates that the founder posted during the
campaign. One row of the output is:

```text
                 Estimate  Std. Error  t value  Pr(>|t|)  95% CI
updates_posted      0.60       0.25       2.40     0.0167  [0.11, 1.09]
```

Supplied values: the two-sided p-value of the proportion test is 0.0124,
and a large-sample 95% interval for $p$ is $(0.410,\,0.490)$.

Complete the proportion test at 5%, define a Type I error for it, name two
changes that raise power while $\alpha$ stays fixed, and interpret the
regression row without causal language.

::: {.callout-note collapse="true"}
## Check a model answer

Under $H_0$, $np_0=600(0.40)=240$ and $n(1-p_0)=600(0.60)=360$. Both are at
least 10. The estimate is $\hat p=270/600=0.45$. The null SE is
$\operatorname{SE}_0=\sqrt{0.40(0.60)/600}=\sqrt{0.0004}=0.02$, so
$z=(0.45-0.40)/0.02=2.50$. The two-sided p-value, 0.0124, is below 0.05, so
reject $H_0$. The sample provides evidence that the proportion of campaigns
that reach their goal differs from 40%. The estimate is 45%, and plausible
values run from 41.0% to 49.0%. The interval excludes 0.40, in line with
the test.

A Type I error would be to conclude that the proportion differs from 40%
when it equals 40%. With $\alpha$ fixed, power rises with a larger sample
and with a larger true difference from 0.40, although the researcher cannot
choose the true difference. For a 0/1 outcome the spread is fixed by $p$,
so less variation is not an option here.

The regression row: each additional update goes together with 0.60 thousand
euros (€600) more raised. $t=0.60/0.25=2.40$, the p-value, 0.0167, is
below 0.05, and the interval $(0.11,\,1.09)$ excludes zero, so reject
$H_0:\beta=0$. The
founders decided themselves how many updates to post, so this is an
association, not a causal effect.
:::

### Worked exam interpretation: a mean test (10 minutes)

**Question (same format as Mock Exam Question 4).** This is an invented
example. A webshop promises that an order arrives after 3 days on average.
From the 1,200 orders that it shipped last month, a simple random sample of
100 orders was selected without replacement. `arrival_days` is the number
of days between the order and its arrival. $\mu$ is the mean of
`arrival_days` among all 1,200 orders. The sample has $\bar x=3.3$ days and
$s=2.0$ days. R produced:

```text
One Sample t-test

data: arrival_days
t = 1.500, df = 99, p-value = 0.1368
alternative hypothesis: true mean is not equal to 3
95 percent confidence interval:
 2.90 3.70
sample estimate:
mean of x = 3.30
```

Verify the SE and the t statistic, explain the degrees of freedom,
interpret the p-value and the interval, give the 5% decision, and check the
10% condition.

::: {.callout-note collapse="true"}
## Check a model answer

$\widehat{\operatorname{SE}}=2.0/\sqrt{100}=0.2$ days, so
$t=(3.3-3)/0.2=1.5$, as in the output. $df=100-1=99$, because the
deviations from the sample mean must sum to zero, which leaves 99 of them
free. If the mean of all 1,200 orders were 3 days, a sample mean at least
1.5 SEs from 3 in either direction would occur in about 14% of random
samples. Since $0.1368>0.05$, do not reject $H_0:\mu=3$. This is not
evidence that the mean is exactly 3 days. The interval $(2.90,\,3.70)$
contains 3, in line with the test, but it also contains values up to 3.7
days. It was produced by a procedure that captures $\mu$ in about 95% of
random samples. The 10% condition holds, because $100\le0.10(1200)=120$.
The conclusion concerns the 1,200 orders of last month, not the orders of
other months.
:::

### Check the AI (5 minutes)

You asked an AI assistant: "The p-value of my test of the equity gap is 0.0045. What does that tell me?" It answered:

> "There is only a 0.45% probability that the null hypothesis is true, so the difference must be large."

Find what is wrong and write the corrected answer. This is the **Verify**
step of Attempt → Ask → Verify → Explain, and the exam asks for it.

::: {.callout-note collapse="true"}
## Check your answer

Two errors. The p-value is not the probability that the null hypothesis is true. It says: if the true gap were zero, a sample this far from zero would occur in about 0.45% of samples. And it says nothing about size. The size is the estimate, 2.24 points below the benchmark, with its interval.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Name one
yes-or-no claim that you will have to defend with data, for example "our
conversion rate differs from the sector's 3%" or "our average order differs
from the break-even value". Write two sentences. First state the claim as a
two-sided test of a zero contrast and say which of the two errors would cost
you more, and why. Then name the one way to raise power that you could
really use before your next pitch.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "My claim: our conversion rate differs from the sector's 3%, $H_0\!: p = 0.03$ against a two-sided alternative. A false alarm is the costlier error, because I would sign a bigger kitchen lease on a difference that is not real. The one way to raise power that I can use before the next pitch is a larger sample, by running the tasting stand two more weeks."
:::

You can discuss this task, and any other task in this chapter, with
**Tilly**, the course AI assistant: ask it to draft the
seven-step workflow for your claim and challenge your choice of the more
costly error.
::::

[Continue to Tutorial 4: Inference and tests against zero](t04-exam-prep.qmd).

[Formula sheet: Lecture 10 formulas](formula-sheet.qmd#formula-l10).

For a more formal explanation, consult
[Nieuwenhuis, *Statistical Methods for Business and Economics*](https://tilburguniversity.on.worldcat.org/oclc/317545871),
Chapters 15 and 16.

:::
