# Mock Exam: QM Basic {#basic-mock-exam}

```{r}
#| label: bmock-setup
#| include: false
# Chunk options only. Every object used on this page is constructed in
# visible or self-contained code below.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

This mock exam represents the full ten-lecture course. It contains **100
points**, so the exam grade is

$$
\text{grade} = \frac{\text{points earned}}{10}.
$$

The real exam is closed book. The complete [QM Basic Formula
Sheet](#formula-sheet) will be attached. Unless a question states otherwise,
use a 5% significance level, report units, and show enough work for your method
to be checked. Use **180 minutes** as the practice time limit.

| Question | Topic | Points | Suggested minutes |
|---|---|---:|---:|
| 1 | Data and a histogram | 15 | 27 |
| 2 | Probability and a discrete random variable | 20 | 36 |
| 3 | Continuous distributions | 15 | 27 |
| 4 | Sampling, confidence intervals, and a mean test | 25 | 45 |
| 5 | A proportion test, errors, power, and R output | 25 | 45 |
| **Total** |  | **100** | **180** |

## Question 1: Data and a histogram (15 points)

An entrepreneurship centre received 240 applications for its accelerator. One
row of its dataset represents one application. The variables are:

- `funding_requested_k`: funding requested, in thousands of euros
- `team_size`: number of people named as current venture-team members
- `sector`: `Consumer`, `Manufacturing`, `Services`, or `Technology`
- `accepted`: 1 if the application was accepted and 0 otherwise.

The histogram below groups the 240 values of `funding_requested_k` into
equal-width intervals. Each bar includes its upper limit. The centre wants to describe these 240 applications.
It does not claim that they are a random sample of all new ventures.

```{r}
#| label: fig-bmock-histogram
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Funding requested by the 240 accelerator applications."
# Self-contained: constructs everything it plots.
library(ggplot2)
mock_breaks <- seq(0, 120, by = 20)
mock_counts <- c(58, 88, 54, 24, 10, 6)
mock_df <- data.frame(
  xmin = mock_breaks[-length(mock_breaks)],
  xmax = mock_breaks[-1],
  count = mock_counts
)

ggplot(mock_df) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = count),
            fill = "#009E73", alpha = 0.85, colour = "white") +
  scale_x_continuous(breaks = mock_breaks) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Funding requested (EUR 1,000s)",
       y = "Number of applications") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

1. State the observational unit, the sample, and one population to which these
   data alone should **not** automatically be generalised. **(4 points)**
2. Classify each of the four variables as categorical or quantitative. For
   each quantitative variable, also state whether it is discrete or
   continuous as modelled here. **(4 points)**
3. Explain both axes and interpret the bar covering more than €20,000
   up to and including €40,000. **(3 points)**
4. Describe the distribution's shape and identify the interval containing the
   median. Briefly justify both answers. **(4 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. The observational unit is one accelerator application (1 point). The
   sample is the 240 received applications (1 point). A defensible example of
   a population not automatically represented is all new ventures in the
   Netherlands, because applications were not randomly sampled from that
   population (2 points).
2. `funding_requested_k` is quantitative continuous, `team_size` is
   quantitative discrete, `sector` is categorical, `accepted` is categorical
   binary, even though it is coded 0/1 (1 point each).
3. The horizontal axis gives requested funding in thousands of euros. The
   vertical axis gives application counts (1 point each). The bar says 88
   applications requested more than €20,000 and at most €40,000 (1 point).
4. The distribution is right-skewed because most observations are in lower
   intervals and the counts taper toward larger requests (2 points). The
   median lies in €20,000--€40,000: 58 observations lie at or below €20,000, while
   $58+88=146$ lie at or below €40,000, so observations 120 and 121 are in that
   interval (2 points).
:::

## Question 2: Probability and a discrete random variable (20 points)

A separate archive contains 500 televised business pitches. Select one record
uniformly at random. Let $D$ mean that the record shows an on-air deal and
let $W$ mean that at least one woman is represented among the presenters.

| Presenters | On-air deal | No on-air deal | Total |
|---|---:|---:|---:|
| Women represented | 90 | 110 | 200 |
| No women represented | 120 | 180 | 300 |
| **Total** | **210** | **290** | **500** |

1. Calculate $P(D)$, $P(W)$, and $P(D\cap W)$. **(6 points)**
2. Calculate $P(D\mid W)$ and $P(W\mid D)$, and explain why they answer
   different questions. **(6 points)**
3. Are $D$ and $W$ independent for this archive of 500 records? Show
   a valid numerical check. **(4 points)**
4. Define $X=1$ for an on-air deal and $X=0$ otherwise. Give the PMF of
   $X$, $E[X]$, and $\operatorname{Var}(X)$. **(4 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. $P(D)=210/500=0.42$, $P(W)=200/500=0.40$, and
   $P(D\cap W)=90/500=0.18$ (2 points each).
2. $P(D\mid W)=90/200=0.45$ and
   $P(W\mid D)=90/210\approx0.429$ (2 points each). The first restricts the
   denominator to pitches with women represented. The second restricts it to
   pitches with a deal (2 points).
3. They are not independent: for example,
   $P(D\mid W)=0.45\ne P(D)=0.42$ (4 points). Equivalently,
   $P(D\cap W)=0.18\ne0.42\times0.40=0.168$.
4. $p_X(1)=0.42$, $p_X(0)=0.58$ (2 points),
   $E[X]=0.42$, and $\operatorname{Var}(X)=0.42(0.58)=0.2436$
   (1 point each).
:::

## Question 3: Continuous distributions (15 points)

Let $X\sim U(15,40)$ represent a waiting time in minutes. Separately, let
$Y\sim N(120,8^2)$ represent a processing time in minutes. The following
standard-normal probabilities are supplied:

```text
P(Z <=  1.5) = 0.9332
P(Z <=  1.0) = 0.8413
P(Z <= -1.0) = 0.1587
```

1. State the PDF height of $X$, including its value outside the support.
   **(3 points)**
2. Calculate $P(22<X<32)$, $E[X]$, and
   $\operatorname{Var}(X)$. **(5 points)**
3. Standardise 132 minutes under the model for $Y$, then calculate
   $P(Y\le132)$. **(3 points)**
4. Calculate $P(112<Y<128)$. **(2 points)**
5. Explain why a PDF height is not itself a probability. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. $f_X(x)=1/(40-15)=0.04$ for $15\le x\le40$, and
   $f_X(x)=0$ outside that interval (3 points).
2. $P(22<X<32)=(32-22)/(40-15)=0.40$ (2 points),
   $E[X]=(15+40)/2=27.5$ (1 point), and
   $\operatorname{Var}(X)=(40-15)^2/12\approx52.08$ square minutes
   (2 points).
3. $z=(132-120)/8=1.5$, so $P(Y\le132)=0.9332$ (3 points).
4. $P(112<Y<128)=P(-1<Z<1)=0.8413-0.1587=0.6826$
   (2 points).
5. Probability is area over an interval: approximately density times a small
   interval width. A point has zero width and hence zero probability for a
   continuous variable (2 points).
:::

## Question 4: Sampling, confidence intervals, and a mean test (25 points)

From a frame of 1,000 accelerator applicants, a simple random sample of 64 was
selected without replacement. For each sampled applicant, `prep_hours` is the
number of hours spent preparing the application. Let $\mu$ be the mean
preparation time among all 1,000 applicants. The sample has
$\bar x=18.4$ hours and $s=8.0$ hours. R produced:

```text
One Sample t-test

data: prep_hours
t = 2.400, df = 63, p-value = 0.0194
alternative hypothesis: true mean is not equal to 16
95 percent confidence interval:
 16.40 20.40
sample estimate:
mean of x = 18.40
```

1. Identify the parameter, estimator, and realised estimate. **(3 points)**
2. Calculate the estimated standard error of the sample mean. **(3 points)**
3. State $H_0$ and $H_1$, and verify the reported t statistic. **(5 points)**
4. Explain why $df=63$. **(3 points)**
5. Interpret the p-value and give the test decision in context. **(5 points)**
6. Interpret the 95% confidence interval. **(3 points)**
7. Check the 10% condition and state what the sampling design supports. Do
   not make a causal claim. **(3 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. The parameter is $\mu$, the frame's population mean preparation time, the
   estimator is $\bar X$, the realised estimate is 18.4 hours (1 point each).
2. $\widehat{\operatorname{SE}}(\bar X)=S/\sqrt n=8/\sqrt{64}=1$
   hour (3 points).
3. $H_0:\mu=16$ against $H_1:\mu\ne16$ (2 points).
   $t=(18.4-16)/1=2.4$ (3 points).
4. After the sample mean is estimated, 63 of the 64 deviations can vary
   freely. The last is fixed by their sum of zero, so $df=n-1=63$
   (3 points).
5. If the frame mean were 16 hours and the test conditions held, the
   probability of a t statistic at least as far from zero as 2.4 in either
   direction would be 0.0194 (3 points). Because $0.0194<0.05$, reject
   $H_0$. The sample provides evidence that the frame mean differs from 16
   hours (2 points).
6. Under repeated valid sampling, 95% of intervals constructed by this method
   would contain the frame mean. For this sample, plausible values run from
   16.40 to 20.40 hours (3 points).
7. $64\le0.10(1000)=100$, so the course's independence approximation is
   reasonable (1 point). The random sample supports generalisation to the
   1,000-applicant frame, as long as all 64 sampled applicants answered and
   the hours were recorded correctly. It does not identify a causal effect
   because nothing was randomly assigned (2 points).
:::

## Question 5: A proportion test, errors, power, and R output (25 points)

In a simple random sample of 625 former accelerator participants, 350 had
launched a trading venture within one year. Let $p$ be the launch proportion
among all former participants in the sampling frame. A two-sided test of
$H_0:p=0.50$ is planned.

The same study also fits a regression with monthly revenue measured in
thousands of euros as the outcome and weekly mentoring hours as a predictor.
One row of the supplied R output is:

```text
                 Estimate  Std. Error  t value  Pr(>|t|)  95% CI
mentoring_hours     0.80       0.25       3.20     0.0014  [0.31, 1.29]
```

1. Check the success-failure condition under $H_0$. **(3 points)**
2. Calculate $\hat p$, its null standard error, and the z statistic.
   **(5 points)**
3. The two-sided p-value is 0.0027. Interpret it and state the test conclusion.
   **(4 points)**
4. A large-sample 95% confidence interval for $p$ is $(0.521,0.599)$.
   Interpret it and explain why it agrees with the test decision. **(4 points)**
5. Define a Type I error in this test. State two changes that would increase
   power while holding the significance level fixed. **(4 points)**
6. Interpret the mentoring coefficient, its test against zero, and the limit
   on causal language. **(5 points)**

::: {.callout-note collapse="true"}
## Model answer and marking guide

1. Under $H_0$, $np_0=625(0.50)=312.5$ and
   $n(1-p_0)=312.5$, both at least 10 (3 points).
2. $\hat p=350/625=0.56$ (1 point), the null SE is
   $\sqrt{0.50(0.50)/625}=0.02$ (2 points), and
   $z=(0.56-0.50)/0.02=3.00$ (2 points).
3. If $p=0.50$, the probability of a z statistic at least as far from zero
   as 3.00 in either direction is about 0.0027 (2 points). Reject $H_0$.
   There is evidence that the frame launch proportion differs from 0.50
   (2 points).
4. Plausible values for the frame launch proportion range from 0.521 to 0.599
   under the method's conditions (2 points). Zero is excluded from the
   equivalent interval for $p-0.50$, so the interval and two-sided 5% test
   both reject the zero difference (2 points).
5. A Type I error is concluding that the frame launch proportion differs from
   0.50 when it actually equals 0.50 (2 points). A larger sample and a
   larger true difference from 0.50 earn 1 point each. Lower variability is
   not accepted, because the spread of a 0/1 outcome is fixed by $p$. Increasing $\alpha$ is not accepted because the question holds it
   fixed.
6. Among the observed participants, one additional weekly mentoring hour is
   associated with 0.80 thousand euros (€800) higher fitted monthly revenue
   (2 points). The row tests $H_0:\beta=0$. The p-value, 0.0014, is below
   0.05 and the interval excludes zero, so reject at 5% (2 points). Without random assignment or a
   defensible causal design, this is an association rather than the causal
   effect of adding an hour of mentoring (1 point).
:::

Your marks across the five questions should sum to 100. When practising,
complete the exam before expanding the model answers, then use the marking
guide to identify whether lost points came from calculation, interpretation,
units, conditions, or scope.
