# QM Basic Mock Exam {#basic-mock-exam .unnumbered}

```{r}
#| label: bmock-setup
#| include: false
# Chunk options only. Every object used on this page is constructed in
# visible or self-contained code below.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5)
```

This mock exam covers all ten lectures of the course. It contains 100
points. Your test score is

$$
\text{test score} = \frac{\text{points earned}}{10},
$$

on a scale from 1.00 to 10.00, with two decimals. A score below 1.00 counts
as 1.00. The test score goes into Osiris, the student information system.
Osiris then rounds the test score to a whole or half grade for your final
course grade. A score from 5.50 to 5.74
becomes 6.0, and a score from 5.25 to 5.49 becomes 5.0. For example, 72
points give a test score of 7.20 and a final grade of 7.0. You pass with a
final grade of 6.0 or higher, that is, with 55 points or more.

This mock exam gives an impression of what an exam in this course can look
like. Keep the following points in mind:

- You write the exam on paper. It is a closed-book exam: you may not use
  books or notes. When you practise, give yourself 180 minutes.
- Every question of this mock exam has several parts, and every part
  shows its points.
- All exam questions are open questions. You write your calculation or your
  explanation. An answer without a calculation or a reason gets no points,
  unless the part asks only to state or to name something.
- Some questions of this mock exam show output from R or an answer of an
  AI assistant. You interpret the output or you assess the answer. You do
  not write R code in this mock exam.
- You may use a calculator of the Casio FX-82 series or of the Texas
  Instruments TI-30 series, but not the TI-30X Pro. Other calculators are
  not allowed.
- The complete [QM Basic Formula Sheet](#formula-sheet) is attached to the
  exam.
- In this mock exam, the values you need are given in the question or on
  the formula sheet.
- All tests in this course are two-sided. In this mock exam, use a 5%
  significance level.
- At the exam inspection, you can see your marked exam. The date of the
  inspection is announced on Canvas.

## Question 1: Data and a histogram (11 points)

An invented example. An entrepreneurship centre runs an accelerator, a support programme for new
ventures. This year the accelerator received 240 applications. The dataset
of the centre has these variables:

- `application_id`: a number that identifies the application
- `funding_requested_k`: funding requested, in thousands of euros
- `team_size`: number of people in the venture team
- `stage`: `idea`, `prototype`, or `first sales`
- `accepted`: 1 if the application was accepted and 0 otherwise.

The histogram shows the 240 values of `funding_requested_k` in intervals of
20 thousand euros. Each interval includes its upper limit: for example, a
value of exactly 20 is in the first bar. The number above a bar is the
number of applications in that interval.

```{r}
#| label: fig-bmock-histogram
#| code-fold: true
#| code-summary: "See the R code for this figure"
#| fig-cap: "Funding requested in the 240 accelerator applications. The number above each bar is the number of applications in that interval."
library(ggplot2)
mock_breaks <- seq(0, 120, by = 20)
mock_counts <- c(58, 88, 54, 24, 10, 6)
mock_df <- data.frame(
  xmin = mock_breaks[-length(mock_breaks)],
  xmax = mock_breaks[-1],
  count = mock_counts
)

ggplot(mock_df) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = count),
            fill = "#009E73", alpha = 0.85, colour = "white") +
  geom_text(aes(x = (xmin + xmax) / 2, y = count, label = count),
            vjust = -0.5, colour = "#17324D", size = 4.2) +
  scale_x_continuous(breaks = mock_breaks) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Funding requested (thousands of euros)",
       y = "Number of applications") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(colour = "#5B6976"),
    axis.title = element_text(colour = "#17324D")
  )
```

1. State what one row of the dataset represents and which group the 240
   rows describe. **(1 point)**
2. State the type of each of these three variables: `application_id`,
   `stage`, `accepted`. Use the five terms of Lecture 1: categorical,
   ordinal, discrete, continuous, identifier. An example:
   `funding_requested_k` is a continuous variable. **(3 points)**
3. &#40;a) Calculate the cumulative count and the cumulative share of each
   of the six intervals. The cumulative share is the cumulative count
   divided by 240. Use your results to state the interval that contains (b)
   the median and (c) the upper quartile $Q_3$ of `funding_requested_k`. **(3 points)**
4. A colleague wants to describe `funding_requested_k` with a normal model.
   Decide whether a normal model fits, and give one reason that you can see
   in the histogram. **(2 points)**
5. The director of the centre writes: "Most new ventures in the Netherlands
   need at most 40 thousand euros." (a) Explain in one sentence why these
   data cannot support this claim. (b) Rewrite the sentence so that the
   data support it. Use a count or a percentage. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer

1. One row is one application to the accelerator, that is, one venture
   that applied. The rows describe the 240 applications that the centre
   received this year, not all new ventures or all start-ups.
2. `application_id` is an identifier. `stage` is ordinal, because its three
   groups have an order. `accepted` is categorical with two groups (binary),
   although it is stored as 0 and 1.
3. &#40;a) The cumulative counts are 58, 146, 200, 224, 234, and 240. The
   cumulative shares are 0.242, 0.608, 0.833, 0.933, 0.975, and 1.000. (b) With 240 values the median lies
   between the 120th and the 121st ordered value. The cumulative share
   first reaches 0.50 in the second interval (0.608, or 146 of 240), so the
   median is above 20 and at most 40 thousand euros. (c) The upper quartile
   is the 180th value, because $0.75\times240=180$. The cumulative share
   first reaches 0.75 in the third interval (0.833, or 200 of 240), so
   $Q_3$ is above 40 and at most 60 thousand euros.
4. A normal model does not fit. Reason: the histogram is skewed to the
   right. Most applications lie in the first two intervals, and the counts
   fall slowly from 54 to 6 on the right side. A normal curve is symmetric
   around its centre.
5. &#40;a) The 240 ventures chose to apply to this accelerator. They are not a
   random sample of all new ventures in the Netherlands, so they can differ
   in a systematic way from the other new ventures. (b) "Of the 240
   applications to this accelerator, 146 (about 61%) requested at most 40
   thousand euros."
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

```{r}
#| label: bmock-q1-check
cumsum(mock_counts)
round(cumsum(mock_counts) / 240, 3)
c(median_position = 240 / 2, q3_position = 0.75 * 240)
```
:::

## Question 2: Probability from a table, and a decision about checks (14 points)

An invented example. An archive contains 500 televised business pitches. You select one record at
random, so every record has the same chance to be selected. $D$ is the event that the pitch
ended with a deal on air. $W$ is the event that at least one woman is among
the presenters.

| Presenters | Deal on air | No deal on air | Total |
|---|---:|---:|---:|
| Women among the presenters | 90 | 110 | 200 |
| No women among the presenters | 120 | 180 | 300 |
| **Total** | **210** | **290** | **500** |

1. Calculate (a) $P(D)$ and (b) $P(D\cap W)$. **(2 points)**
2. Calculate $P(D\cup W)$. Show your calculation. **(1 point)**
3. Calculate (a) $P(D\mid W)$ and (b) $P(W\mid D)$. **(2 points)**
4. Decide whether $D$ and $W$ are independent. Show the comparison that
   you used. **(2 points)**

Parts 5 and 6 use a new invented example. A founder rents out e-scooters.
Of all rentals, 5% end with damage to the scooter. A sensor in the scooter
puts a flag on some rentals. The sensor flags 80% of the rentals with
damage. It also flags 10% of the rentals without damage.

5. Use a tree of 1,000 rentals. (a) State the four counts at the ends of
   the four branches of the tree. (b) Calculate the probability that a flagged rental ended with
   damage. (c) Calculate the probability that a rental without a flag ended
   with damage. **(4 points)**
6. After a rental, the founder can check the scooter by hand. One check
   costs her 4 euros of working time. When the check finds damage, the
   customer pays 60 euros for the damage. For each group, the expected gain of one
   check is the probability of damage times 60 euros. A check is worth its
   cost when the expected gain is above 4 euros. Decide which rentals the
   founder should check: the flagged rentals, the rentals without a flag,
   both groups, or no rental. Show the expected gain for each group. If you
   could not solve part 5, use 0.25 as the probability of damage for
   flagged rentals and 0.02 for rentals without a flag. **(3 points)**

::: {.callout-note collapse="true"}
## Model answer

1. &#40;a) $P(D)=210/500=0.42$. (b) $P(D\cap W)=90/500=0.18$.
2. $P(W)=200/500=0.40$. The addition rule gives
   $P(D\cup W)=0.42+0.40-0.18=0.64$.
3. &#40;a) $P(D\mid W)=90/200=0.45$. (b) $P(W\mid D)=90/210\approx0.429$.
4. $P(D)\,P(W)=0.42\times0.40=0.168$. This is not equal to
   $P(D\cap W)=0.18$, so $D$ and $W$ are not independent.
5. &#40;a) Of 1,000 rentals, 50 end with damage and 950 do not. Of the 50
   rentals with damage, 40 are flagged and 10 are not. Of the 950 others,
   95 are flagged and 855 are not. (b) $40+95=135$ rentals are flagged, so
   $P(\text{damage}\mid\text{flag})=40/135\approx0.296$. (c) $10+855=865$
   rentals have no flag, so
   $P(\text{damage}\mid\text{no flag})=10/865\approx0.012$.
6. Flagged rentals: $0.2963\times60\approx17.78$ euros, which is above the
   cost of 4 euros. Rentals without a flag: $(10/865)\times60\approx0.69$
   euros, below 4 euros. Decision: check the flagged rentals only.
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

```{r}
#| label: bmock-q2-check
pitch_counts <- matrix(c(90, 110, 120, 180), nrow = 2, byrow = TRUE)
n_pitches <- sum(pitch_counts)
c(P_D = sum(pitch_counts[, 1]) / n_pitches,
  P_W = sum(pitch_counts[1, ]) / n_pitches,
  P_D_and_W = pitch_counts[1, 1] / n_pitches,
  P_D_or_W = (210 + 200 - 90) / n_pitches,
  P_D_given_W = 90 / 200, P_W_given_D = 90 / 210,
  product = (210 / 500) * (200 / 500))

damaged <- 1000 * 0.05
others <- 1000 - damaged
tree <- c(damage_flag = damaged * 0.80, damage_no_flag = damaged * 0.20,
          other_flag = others * 0.10, other_no_flag = others * 0.90)
tree
p_flag <- unname(tree["damage_flag"] / (tree["damage_flag"] + tree["other_flag"]))
p_no_flag <- unname(tree["damage_no_flag"] /
                      (tree["damage_no_flag"] + tree["other_no_flag"]))
c(p_flag = p_flag, p_no_flag = p_no_flag,
  gain_flag = 60 * p_flag, gain_no_flag = 60 * p_no_flag,
  net_flag = 60 * p_flag - 4, net_no_flag = 60 * p_no_flag - 4,
  printed_flag = 60 * 0.25, printed_no_flag = 60 * 0.02)
```
:::

## Question 3: Expected value, spread, and two variables (17 points)

An invented example. A founder repairs phones at a stand on campus. Let $R$
be the number of repairs on one morning. Its PMF is:

| $r$ | 2 | 4 | 7 |
|---|---:|---:|---:|
| $p_R(r)$ | 0.3 | 0.5 | 0.2 |

1. Calculate $E[R]$. **(2 points)**
2. Calculate (a) $\operatorname{Var}(R)$ and (b) $\operatorname{SD}(R)$.
   **(3 points)**
3. One repair brings in 25 euros. The stand costs 40 euros per morning. The
   profit of a morning is $W=25R-40$ euros. Calculate (a) $E[W]$ and (b)
   $\operatorname{SD}(W)$. If you could not solve parts 1 and 2, use
   $E[R]=5$ and $\operatorname{SD}(R)=2$. **(2 points)**

The founder also sells phone cases and screen protectors. For one customer,
$X=1$ if the customer buys a case and $X=0$ otherwise. $Y=1$ if the customer
buys a screen protector and $Y=0$ otherwise. The joint probabilities are:

|  | $Y=0$ | $Y=1$ |
|---|---:|---:|
| $X=0$ | 0.50 | 0.10 |
| $X=1$ | 0.20 | 0.20 |

4. Calculate the conditional mean $E[Y\mid X=1]$ and interpret it in one
   sentence. **(1 point)**
5. Calculate $\operatorname{Cov}(X,Y)$ with $E[XY]-E[X]\,E[Y]$.
   **(2 points)**
6. The sum $X+Y$ is the number of these two products (case and screen
   protector) that one customer buys: 0, 1, or 2.
   Calculate (a) $E[X+Y]$ and (b) $\operatorname{Var}(X+Y)$. Use
   $\operatorname{Var}(X)=0.24$ and $\operatorname{Var}(Y)=0.21$. If you
   could not solve part 5, use $\operatorname{Cov}(X,Y)=0.10$.
   **(3 points)**
7. &#40;a) Calculate the correlation between $X$ and $Y$. If you could not
   solve part 5, use $\operatorname{Cov}(X,Y)=0.10$. (b) Interpret its
   direction and its strength in context. **(2 points)**
8. The founder says: "A case makes customers buy a screen protector. I will
   give cases away for free to sell more protectors." Name one third
   variable that could cause the correlation between $X$ and $Y$, and
   explain how it could do this.
   **(2 points)**

::: {.callout-note collapse="true"}
## Model answer

1. $E[R]=2(0.3)+4(0.5)+7(0.2)=0.6+2.0+1.4=4$ repairs.
2. &#40;a) $E[R^2]=4(0.3)+16(0.5)+49(0.2)=1.2+8+9.8=19$, so
   $\operatorname{Var}(R)=19-4^2=3$. (b) $\operatorname{SD}(R)=\sqrt3\approx1.73$ repairs.
3. &#40;a) $E[W]=25\times4-40=60$ euros. (b)
   $\operatorname{SD}(W)=25\times1.7321\approx43.30$ euros. The cost of 40
   euros does not change the spread.
4. $P(X=1)=0.20+0.20=0.40$, so
   $E[Y\mid X=1]=0\times(0.20/0.40)+1\times(0.20/0.40)=0.50$. Half of the
   customers who buy a case also buy a screen protector.
5. $E[X]=0.40$, $E[Y]=0.10+0.20=0.30$, and $E[XY]=0.20$, because $XY=1$
   only when both $X$ and $Y$ are 1. Filled in:
   $\operatorname{Cov}(X,Y)=0.20-0.40\times0.30=0.08$.
6. &#40;a) $E[X+Y]=0.40+0.30=0.70$ products. (b)
   $\operatorname{Var}(X+Y)=0.24+0.21+2\times0.08=0.61$.
7. &#40;a) $\rho_{XY}=0.08/\sqrt{0.24\times0.21}=0.08/0.2245\approx0.36$.
   &#40;b) The correlation is positive and moderate: customers who buy a
   case tend to buy a screen protector too.
8. A third variable: the customer has a new phone. A customer with a new
   phone wants to protect it and buys both products. Then a free case for
   other customers does not sell more protectors.
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

```{r}
#| label: bmock-q3-check
r_values <- c(2, 4, 7)
r_probs <- c(0.3, 0.5, 0.2)
mean_r <- sum(r_values * r_probs)
var_r <- sum(r_values^2 * r_probs) - mean_r^2
c(sum_of_probs = sum(r_probs), mean = mean_r,
  mean_of_squares = sum(r_values^2 * r_probs),
  var = var_r, sd = sqrt(var_r),
  mean_profit = 25 * mean_r - 40, sd_profit = 25 * sqrt(var_r))

joint <- matrix(c(0.50, 0.10, 0.20, 0.20), nrow = 2, byrow = TRUE,
                dimnames = list(c("X=0", "X=1"), c("Y=0", "Y=1")))
p_x1 <- sum(joint["X=1", ])
p_y1 <- sum(joint[, "Y=1"])
cov_xy <- joint["X=1", "Y=1"] - p_x1 * p_y1
var_x <- p_x1 * (1 - p_x1)
var_y <- p_y1 * (1 - p_y1)
c(sum = sum(joint), E_X = p_x1, E_Y = p_y1,
  E_Y_given_X1 = joint["X=1", "Y=1"] / p_x1,
  cov = cov_xy, var_x = var_x, var_y = var_y,
  mean_sum = p_x1 + p_y1, var_sum = var_x + var_y + 2 * cov_xy,
  var_sum_printed = var_x + var_y + 2 * 0.10,
  correlation = cov_xy / sqrt(var_x * var_y),
  correlation_printed = 0.10 / sqrt(var_x * var_y))
```
:::

## Question 4: A normal model and the precision of a sample mean (13 points)

An invented example. A founder repairs laptops. A courier collects the
repaired laptops. Two models describe the times in her business:

- $X\sim U(15,40)$ is the number of minutes between the founder's call and
  the arrival of the courier.
- $Y\sim N(120,8^2)$ is the number of minutes that one repair takes. The
  standard deviation is 8 minutes.

These values of the standard normal distribution are given for the whole
question:

$$
P(Z\le1.0)=0.8413,\qquad P(Z\le1.5)=0.9332,\qquad
P(Z\le2.0)=0.9772,\qquad z_{0.90}=1.2816.
$$

1. Calculate $P(22<X<32)$. **(1 point)**
2. Calculate $P(Y\le132)$. **(2 points)**
3. Calculate $P(Y\le112)$. **(2 points)**
4. The founder wants to promise her customers a maximum repair time. 90% of
   the repairs should take at most this time. Calculate this time. **(2 points)**

The founder also looks at her prices. Let $V$ be the price of one repair,
with mean $\mu=85$ euros and standard deviation $\sigma=20$ euros. She
plans to draw a random sample of $n=100$ repairs from her records of
3,000 repairs.

5. &#40;a) Calculate the standard deviation of the sample mean. (b) Calculate the
   probability that the sample mean is above 89 euros.
   **(3 points)**
6. State how many repairs the sample must contain so that the standard
   deviation of the sample mean is half as large as in part 5. **(1 point)**
7. Next year the prices change, and she will estimate the new mean price
   from a sample. She wants a margin of error of at most 5 euros at the 95%
   level. Use $\sigma=20$ euros as a planning value. Calculate the sample
   size that she needs. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer

1. $P(22<X<32)=(32-22)/(40-15)=10/25=0.40$.
2. $z=(132-120)/8=1.5$, so $P(Y\le132)=P(Z\le1.5)=0.9332$.
3. $z=(112-120)/8=-1.0$. The symmetry rule and the complement rule give
   $P(Z\le-1.0)=P(Z>1.0)=1-0.8413=0.1587$.
4. $x=\mu+z\sigma=120+1.2816\times8=130.25$ minutes. She can promise about
   130 minutes.
5. &#40;a) $\operatorname{SD}(\bar V)=\sigma/\sqrt n=20/\sqrt{100}=2$ euros.
   &#40;b) By the CLT the sample mean is approximately normal with mean 85 and
   standard deviation 2, because $n=100$ is large. The draws can be treated
   as independent, because $100\le0.10\times3{,}000=300$. $z=(89-85)/2=2.0$,
   so $P(\bar V>89)\approx1-0.9772=0.0228$.
6. The sample must contain 400 repairs. The standard deviation of the
   sample mean falls with the square root of the sample size. To halve it,
   she needs four times as many repairs: $20/\sqrt{400}=1$ euro.
7. $n\ge(1.96\times20/5)^2=7.84^2\approx61.47$, so she needs 62 repairs.
   Round up, because 61 repairs give a margin of error above 5 euros.
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

```{r}
#| label: bmock-q4-check
c(uniform = (32 - 22) / (40 - 15),
  p_132 = pnorm(132, mean = 120, sd = 8),
  p_112 = pnorm(112, mean = 120, sd = 8),
  percentile_90 = qnorm(0.90, mean = 120, sd = 8),
  by_hand_90 = 120 + 1.2816 * 8)
sd_mean <- 20 / sqrt(100)
c(sd_of_sample_mean = sd_mean, z = (89 - 85) / sd_mean,
  p_mean_above_89 = 1 - pnorm(89, mean = 85, sd = sd_mean),
  by_hand = 1 - 0.9772,
  sd_of_sample_mean_n_400 = 20 / sqrt(400),
  n_needed = (1.96 * 20 / 5)^2, n_rounded_up = ceiling((1.96 * 20 / 5)^2))
```
:::

## Question 5: A mean test in five steps, and R output (14 points)

An invented example. A founder runs an online shop. The shop covers its
costs when the mean order value is 60 euros. Last year the shop had 2,400 orders.
She drew a simple random sample of 121 of these orders, without replacement.
The variable `order_value` is the value of an order in euros. The sample has
a mean of 64.60 euros and a standard deviation of $s=22.00$ euros.

1. State (a) in words what the parameter $\mu$ stands for in this question
   and (b) the estimate of $\mu$ with its unit. **(2 points)**
2. Test at the 5% level whether the mean order value differs from 60 euros.
   Give all five steps of the testing procedure. The critical value is on
   the formula sheet. **(7 points)**

This year she drew a new simple random sample of 100 of the 2,600 orders
that the shop has had so far this year. Its standard deviation is 24.00 euros. She ran the same test in R
on these orders. This is the output:

```text
	One Sample t-test

data:  order_value
t = 2.6667, df = 99, p-value = 0.008949
alternative hypothesis: true mean is not equal to 60
95 percent confidence interval:
 61.63788 71.16212
sample estimates:
mean of x 
     66.4 
```

3. Interpret the p-value of 0.008949 in one or two sentences.
   **(2 points)**
4. The output also shows a 95% confidence interval. Show with a
   calculation how R found the two ends of this interval. **(1 point)**
5. The founder will rent a larger storage room only if the mean order value
   of this year is above 60 euros. Write your advice to her in two or three
   sentences. (a) Give a decision and support it with both ends of the
   interval. (b) Name one limit of these data, that is, one thing that they
   cannot tell her. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer

1. &#40;a) $\mu$ is the mean order value of all 2,400 orders of last year,
   not the mean of the sample. (b) The estimate is $\bar x=64.60$ euros.
2. The five steps:

   - Step 1, hypotheses. $H_0:\mu=60$ against $H_1:\mu\ne60$, with
     $\alpha=0.05$, where $\mu$ is the mean order value of all 2,400
     orders (part 1).
   - Step 2, test statistic. $T=(\bar X-60)/(S/\sqrt n)$. If $H_0$ is true,
     $T$ has the $t_{120}$ distribution. This may be used because $n=121$ is
     large. Also $121\le0.10\times2{,}400=240$, so the draws can be treated
     as independent.
   - Step 3, critical value and rejection region. $t_{0.975,\,120}=1.980$.
     Reject $H_0$ if $t_{obs}\le-1.980$ or $t_{obs}\ge1.980$.
   - Step 4, observed value. The standard error is
     $s/\sqrt n=22/\sqrt{121}=2.00$ euros.
     $t_{obs}=(64.60-60)/2.00=2.30$.
   - Step 5, conclusion. $2.30\ge1.980$, so the observed value lies in the
     rejection region: reject $H_0$. At the 5% level we can conclude that
     the mean order value of all 2,400 orders differs from 60 euros. The
     positive sign shows that it is above 60 euros.
3. If the mean of all 2,600 orders of this year were 60 euros, a sample
   mean at least as far from 60 euros as this one, in either direction,
   would occur in about 0.9% of random samples.
4. The standard error is $24.00/\sqrt{100}=2.40$ euros. With $df=99$ the
   closest listed value is $t_{0.975,\,100}=1.984$, so
   $66.40\pm1.984\times2.40=66.40\pm4.76$. The interval runs from 61.64 to
   71.16 euros.
5. &#40;a) "Rent the room, because both ends of the interval, 61.64 and
   71.16 euros, are above 60 euros. At the lower end the mean is 1.64
   euros above 60, about 3% of that value." (b) "The data describe the
   orders of this year so far and do not show that later orders will have
   the same values."
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

The R code below builds two invented samples with exactly the means and
the standard deviations of the question. It runs the test of part 2 on the
first sample. For the second sample, it prints the output of parts 3 to 5:

```{r}
#| label: bmock-q5-check
set.seed(2026)
z_values <- rnorm(121)
z_values <- (z_values - mean(z_values)) / sd(z_values)
order_value <- 64.6 + 22 * z_values
c(n = length(order_value), mean = mean(order_value), sd = sd(order_value),
  standard_error = sd(order_value) / sqrt(length(order_value)))
t.test(order_value, mu = 60)
64.6 + c(-1, 1) * 1.980 * 2   # the interval by hand with the sheet value

set.seed(2026)
z_values <- rnorm(100)
z_values <- (z_values - mean(z_values)) / sd(z_values)
order_value <- 66.4 + 24 * z_values   # the new sample of this year
c(n = length(order_value), mean = mean(order_value), sd = sd(order_value),
  standard_error = sd(order_value) / sqrt(length(order_value)))
t.test(order_value, mu = 60)
66.4 + c(-1, 1) * 1.984 * 2.4   # the interval by hand with the sheet value
```
:::

## Question 6: An interval and a test for a proportion, two errors, and power (22 points)

An invented example. A founder sells a coffee subscription. New customers
start with a trial month. In her industry, 20% of trial customers switch to
a paid subscription. She wants to know whether her own share differs from
20%. She drew 100 trial customers at random from the list of all her 1,500
trial customers of last year. Of these 100, 26 switched to a paid
subscription. Let $p$ be the share of customers who switched to a paid
subscription among all 1,500 trial customers of last year.

1. A friend suggests a faster way: print a QR code for a short survey on
   the next 100 coffee bags that she ships, and use the answers that come
   in. (a) Name this sampling method. (b) Explain why the founder's random
   draw from the list of trial customers is the better choice for her
   question. Name one weakness of the survey and one strength of the random
   draw. **(3 points)**
2. Construct a 95% confidence interval for $p$, and use it to answer the
   question whether her share differs from 20%. Write down the formula of
   the interval, the interval, and your conclusion with an explanation.
   Give the five parts. **(7 points)**
3. Test at the 5% level whether her share differs from 20%. Give all five
   steps of the testing procedure. The critical value is on the formula
   sheet. **(7 points)**
4. Describe (a) a Type I error and (b) a Type II error for the test of
   part 3, in the context of the founder. **(2 points)**
5. &#40;a) State which of the two errors is possible after the decision of
   your test in part 3. (b) Name one change to the study that raises the
   power while the significance level stays at 5%. **(2 points)**
6. Your interval uses $\hat p$ in the standard error, and your test uses
   $p_0$ in the standard deviation of $\widehat P$ under $H_0$. Explain in
   one sentence why. **(1 point)**

::: {.callout-note collapse="true"}
## Model answer

1. &#40;a) A voluntary-response sample.
   &#40;b) Weakness of the survey: the customers who scan the code decide
   themselves to answer. They can differ from the other customers in a
   systematic way, for example because they like the coffee more. Strength of the random draw: every trial customer on the
   list has the same chance to be selected, so the result describes all
   1,500.
2. The five parts:

   - Part 1, formula. $\widehat P\pm z_{0.975}\sqrt{\widehat
     P(1-\widehat P)/n}$. It may be used because $n\hat p=26$ and
     $n(1-\hat p)=74$ are both at least 10.
   - Part 2, given values. 26 of $n=100$, so $\hat p=0.26$. The confidence
     level is 95%, so $z_{0.975}=1.96$.
   - Part 3, calculations. Standard error
     $\sqrt{0.26\times0.74/100}=\sqrt{0.001924}\approx0.0439$. Margin of
     error $1.96\times0.0439\approx0.086$.
   - Part 4, interval. $(0.26-0.086,\ 0.26+0.086)=(0.174,\ 0.346)$.
   - Part 5, conclusion. The value 0.20 lies inside the interval. At the 5%
     level she cannot conclude that her share differs from the 20% of the
     industry. This is not proof that her share is 20%, because the
     interval also contains shares up to 34.6%.
3. The five steps:

   - Step 1, hypotheses. $H_0:p=0.20$ against $H_1:p\ne0.20$, with
     $\alpha=0.05$, where $p$ is the share of all 1,500 trial customers of
     last year who switched to a paid subscription.
   - Step 2, test statistic.
     $Z=(\widehat P-0.20)/\sqrt{0.20\times0.80/100}$. If $H_0$ is true,
     $Z$ is approximately $N(0,1)$, because $np_0=20$ and $n(1-p_0)=80$ are
     both at least 10.
   - Step 3, critical value and rejection region. $z_{0.975}=1.96$.
     Reject $H_0$ if $z_{obs}\le-1.96$ or $z_{obs}\ge1.96$.
   - Step 4, observed value. The standard deviation under $H_0$ is
     $\operatorname{SD}_0(\widehat P)=\sqrt{0.20\times0.80/100}
     =\sqrt{0.0016}=0.04$. $z_{obs}=(0.26-0.20)/0.04=1.50$.
   - Step 5, conclusion. $-1.96<1.50<1.96$, so the observed value does
     not lie in the rejection region: do not reject $H_0$. At the 5% level
     she cannot conclude that her share differs from the 20% of the
     industry. This does not show that her share is 20%.
4. &#40;a) Type I error: the test concludes that her share differs from 20%,
   but in truth it is 20%. (b) Type II error: the test does not conclude
   that her share differs from 20%, but in truth it differs.
5. &#40;a) The test of part 3 did not reject $H_0$, so only a Type II error
   is possible. (b) A larger sample.
6. The test calculates as if $H_0$ were true, so it uses $p_0=0.20$. The
   interval has no claimed value for $p$, so it uses the estimate
   $\hat p=0.26$.
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

```{r}
#| label: bmock-q6-check
n_trial <- 100
p_hat <- 26 / n_trial
standard_error <- sqrt(p_hat * (1 - p_hat) / n_trial)
c(successes = n_trial * p_hat, failures = n_trial * (1 - p_hat),
  p_hat = p_hat, standard_error = standard_error,
  margin = 1.96 * standard_error,
  lower = p_hat - 1.96 * standard_error,
  upper = p_hat + 1.96 * standard_error)
p_null <- 0.20
sd_under_H0 <- sqrt(p_null * (1 - p_null) / n_trial)
z_obs <- (p_hat - p_null) / sd_under_H0
c(sd_under_H0 = sd_under_H0, z_obs = z_obs,
  critical_value = qnorm(0.975),
  p_value = 2 * pnorm(-abs(z_obs)))   # the test of part 3
```
:::

## Question 7: An answer of an AI assistant about a regression row (9 points)

An invented example. An accelerator programme has data on 1,200 former
participants. A regression has the monthly revenue of the venture, in
thousands of euros, as the outcome. The predictor is the number of mentoring
hours per week. The row below is a shortened version of the R output. The
last column shows the 95% confidence interval for the coefficient. This
interval comes from a separate R command.

```text
                 Estimate  Std. Error  t value  Pr(>|t|)  95% CI
mentoring_hours     0.75       0.30       2.50     0.0126  [0.16, 1.34]
```

A founder gave this row to an AI assistant and asked: "What does this row
tell me? Should I book more mentoring hours?" The AI answered:

> "The p-value of 0.0126 means that there is only a 1.26% probability that
> the true coefficient is zero. The p-value is below 0.05, so the link
> between mentoring and revenue is strong and important for your business.
> Mentoring causes higher revenue: if you book two more hours per week, your
> revenue will rise by 1,500 euros per month."

1. Interpret the estimate of 0.75 in one sentence. Name the group, the
   difference in mentoring hours that the estimate refers to, and the
   revenue with its unit. **(2 points)**
2. Assess the first sentence of the AI answer. (a) Explain why this is not
   what a p-value means. (b) Write a correct interpretation of the p-value
   of 0.0126. **(3 points)**
3. Assess the second sentence of the AI answer. (a) Explain whether a small
   p-value shows that a link is strong and important. (b) State which
   numbers of the row describe the size of the link, and what these numbers
   say.
   **(2 points)**
4. Assess the third sentence of the AI answer. (a) Explain whether these
   data support the claim that mentoring causes higher revenue, and give
   one other explanation for the link. (b) Write a sentence about two more
   mentoring hours that the data do support. **(2 points)**

::: {.callout-note collapse="true"}
## Model answer

1. Among the 1,200 former participants, a participant with one more
   mentoring hour per week had on average 0.75 thousand euros (750 euros)
   more monthly revenue.
2. &#40;a) A p-value is calculated under the assumption that the true
   coefficient is zero. It is not the probability that this assumption is
   true. (b) "If the true coefficient were zero, an estimate at least 2.5
   standard errors from zero, in either direction, would occur in about
   1.3% of samples."
3. &#40;a) No. A small p-value shows that the estimate is clearly different
   from zero. It does not show that the link is strong or important.
   &#40;b) The size is the estimate with its interval: 750 euros more monthly
   revenue per mentoring hour, with plausible values from 160 to 1,340
   euros. Whether this is important depends on the revenue and the costs
   of the venture.
4. &#40;a) No. The row shows an association. Nobody assigned the mentoring
   hours at random. The participants chose their hours themselves. One
   other explanation: founders with more drive both book more mentoring
   and earn more revenue. (b) "Participants with two more mentoring hours per week had
   on average 1,500 euros more monthly revenue."
:::

::: {.callout-tip collapse="true"}
## Optional: check the numbers in R

You do not need R in this exam. The code below only shows how the numbers
of the model answer were checked. You can run it if you want to practise R,
but you can skip it.

```{r}
#| label: bmock-q7-check
c(t_value = 0.75 / 0.30,
  p_value = 2 * pt(-2.50, df = 1200 - 2),
  lower = 0.75 - qt(0.975, df = 1200 - 2) * 0.30,
  upper = 0.75 + qt(0.975, df = 1200 - 2) * 0.30)
```
:::

The questions of this mock exam have 100 points together. When you practise,
complete the exam before you open the model answers. Then compare
your answers with them: check your calculations, your interpretations,
your units, your conditions, and your decisions.

[Continue to QM Intermediate: Course overview](../intermediate/overview.qmd).
