# Tutorial 1: Evidence and probability foundations {#b-ep01}

```{r}
#| label: bep01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- distinguish data, events, random variables, and distributions;
- apply probability rules to the Shark Tank record-selection experiment;
- calculate and interpret a Bernoulli PMF, CDF, and expectation;
- calculate a variance and standard deviation from a small PMF;
- write a concise interpretation of supplied R output.
:::

Tutorials are **exam preparation sessions**: every task below follows the
format of the final exam. The standing routine for each task is the same ---
the TA states the task, pairs work on it first for a few minutes with AI use
allowed, and the solution is then discussed in plenary, starting from the
pair attempts.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 1--3. Without reopening the
chapters, complete these definitions:

1. An observational unit is ...
2. An event is ...
3. A random variable is ...
4. A PMF is ...
5. The variance of a random variable averages ...

Use the cumulative formula sheet through Lecture 3.

Review [Mathematics Refresher: sets, fractions, and
sums](../resources/math-refresher.html) if the notation is slowing you down.

::: {.callout-note collapse="true"}
## Check your preparation

1. An observational unit is the entity represented by one row of the data.
2. An event is a specified subset of the sample space.
3. A random variable is a real-valued function on the sample space.
4. A PMF assigns probability $P(X=x)$ to each value in the support of a
   discrete random variable; its masses are non-negative and sum to 1.
5. Variance averages squared deviations from the variable's mean:
   $\operatorname{Var}(X)=E\!\left[(X-E[X])^2\right]$.
:::

:::

::: {.course-phase .phase-in-class}

## In class

Attempt each task before opening its answer. A complete written answer should
name the quantity, show the calculation, report the value and units, interpret
it in context, and state an important model or measurement boundary.

All tasks use the Shark Tank teaching file. Load it once:

```{r}
#| label: bep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
```

### Task 1 --- Data and measurement

R reports:

```{r}
#| label: bep01-task1
#| echo: false
c(
  rows = nrow(sharks),
  columns = ncol(sharks),
  deals = sum(sharks$deal_on_show),
  deal_proportion = mean(sharks$deal_on_show),
  median_description_words = median(sharks$description_words)
)
```

1. What is the observational unit?
2. Verify the deal proportion from 882 deals and 559 non-deals.
3. Explain why the mean of `deal_on_show` is a proportion.
4. Evaluate: "The data contain 1,441 successful companies."
5. What exactly does `description_words` measure?

::: {.callout-note collapse="true"}
## Check your answer

One row is one recorded televised pitch. There are $882+559=1441$ records and
$882/1441\approx0.612$, so 61.2% reached an on-air agreement. The mean of a
0/1 indicator equals the share coded 1 because its sum counts the 1s.
"Successful companies" is incorrect: the outcome is an agreement made on the
show, not completed funding or later venture performance. Word count
describes the written text in the source dataset, not the spoken pitch.
:::

### Task 2 --- Events and probability rules

The experiment selects one of the 1,441 records uniformly. The two events are
built exactly as in Lecture 2:

```{r}
#| label: bep01-events
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
```

R reports:

```{r}
#| label: bep01-task2
#| echo: false
c(P_D = mean(D), P_L = mean(L), P_D_and_L = mean(D & L))
```

1. Define $\Omega$, and distinguish an outcome from event $D$.
2. Translate $D^c$, $D\cap L$, and $D\cup L$ into words.
3. Calculate $P(D^c)$.
4. Calculate $P(D\cup L)$ using the addition rule.
5. Explain why these probabilities do not automatically apply to a future US
   entrepreneur.

::: {.callout-note collapse="true"}
## Check your answer

The sample space is the set of 1,441 complete pitch records; one selected
record is an outcome, while $D$ is the subset of 882 deal records. The
complement is no on-air deal, the intersection is a late-season deal, and the
union is late-season, deal, or both. $P(D^c)=1-882/1441=559/1441\approx0.388$.
The union is $P(D)+P(L)-P(D\cap L)\approx0.771$. Future use requires a
defined target population and comparable selection, time period, and outcome.
:::

### Task 3 --- PMF, CDF, and expectation

Let $X=1$ for an on-air deal and $X=0$ otherwise. Use $p=882/1441$.

1. Write the PMF and verify that its masses sum to 1.
2. Write the full piecewise CDF $F_X(a)$.
3. Calculate $E[X]$ from the PMF.
4. Explain why $E[X]$ equals the observed proportion of 1s.
5. Explain the difference between the random variable $X$ and a realised
   value $x=1$.

::: {.callout-note collapse="true"}
## Check your answer

$$
p_X(0)=1-p,\qquad p_X(1)=p,
$$

and the masses total $1-p+p=1$.

$$
F_X(a)=
\begin{cases}
0,&a<0,\\
1-p,&0\le a<1,\\
1,&a\ge1.
\end{cases}
$$

The expectation is $E[X]=0(1-p)+1p=p\approx0.612$. Because the indicator
contains 882 ones among 1,441 observations, its mean is also $882/1441$.
$X$ is the mapping defined before selection; $x=1$ is one possible
realised value.
:::

### Task 4 --- Variance from a small PMF

A venture studio models $N$, the number of investors who make an offer to one
of its startups, with the PMF

| $n$ | 0 | 1 | 2 |
|---|---:|---:|---:|
| $p_N(n)$ | 0.5 | 0.3 | 0.2 |

1. Calculate $E[N]$.
2. Calculate $\operatorname{Var}(N)$ from the definition
   $E\!\left[(N-E[N])^2\right]$.
3. Verify the result with the computational form
   $E[N^2]-\left(E[N]\right)^2$.
4. Report the standard deviation with its unit, and state in one sentence
   what it measures here.

::: {.callout-note collapse="true"}
## Check a model answer

$E[N]=0(0.5)+1(0.3)+2(0.2)=0.7$ offers.

Definition form:
$$
\operatorname{Var}(N)=(0-0.7)^2(0.5)+(1-0.7)^2(0.3)+(2-0.7)^2(0.2)
=0.245+0.027+0.338=0.61.
$$

Computational form: $E[N^2]=0^2(0.5)+1^2(0.3)+2^2(0.2)=1.1$, so
$\operatorname{Var}(N)=1.1-0.7^2=1.1-0.49=0.61$. The two routes agree.

The standard deviation is $\sqrt{0.61}\approx0.78$ offers. It measures the
typical distance of the realised number of offers from the mean of 0.7,
in the original unit (offers), whereas the variance is in squared offers.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice --- 10 minutes

An output block describes word count:

```{r}
#| label: bep01-timed
#| echo: false
w <- sharks$description_words
c(
  mean = mean(w),
  median = median(w),
  minimum = min(w),
  maximum = max(w)
)
```

Write no more than 100 words. Compare mean and median, calculate the range,
and state the measurement boundary. Then explain which value equals the
expectation if the empirical distribution gives mass $1/1441$ to each record.

::: {.callout-note collapse="true"}
## Check a model answer

The written descriptions have a mean length of about 5.0 words and a median
of 5 words. Their range is $13-1=12$ words. The thin upper tail is consistent
with the mildly right-skewed histogram seen in Lecture 1. If each record
receives probability $1/1441$, the empirical expectation is the mean, about
5.0 words. These values describe written dataset text, not spoken-pitch
length or quality.
:::

### Check your understanding

Complete from memory:

1. An event is ...
2. A PMF must ...
3. A CDF at $a$ reports ...
4. Expectation is ...
5. Variance is ... and its square root is ...

::: {.callout-note collapse="true"}
## Check your answers

1. An event is a specified subset of the sample space.
2. A PMF is non-negative and sums to 1 over the support.
3. $F_X(a)=P(X\le a)$.
4. $E[X]=\sum_x x\, p_X(x)$, the probability-weighted average of the possible
   values.
5. Variance is the probability-weighted average of squared deviations from
   the mean, $\operatorname{Var}(X)=E[X^2]-\left(E[X]\right)^2$; its square
   root is the standard deviation, in the variable's original unit.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Pick one
uncertain count that matters for it (customers who sign up in a week, offers
received, orders returned), and write two sentences: first sketch a small PMF
for it (three or four values with rough probabilities), then say what its
expectation and standard deviation would tell you about a real decision.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Weekly catering requests: 0 with probability 0.4, 1 with 0.4, 2 with 0.15, and 3 with 0.05, so I expect about 0.85 requests per week. The standard deviation tells me whether a weekend helper is justified --- with high spread, some weeks genuinely need one even though the average says no."
:::

You can discuss this task --- and any other task in this tutorial --- with
**Tilly**, the course AI assistant on Canvas.
::::

Continue to Lecture 4: Conditioning, total probability, Bayes, and
independence.

:::
