# Tutorial 1: Evidence and probability foundations {#b-ep01 .unnumbered}

```{r}
#| label: bep01-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- classify variables and apply probability rules to a real dataset
- calculate expectations, variances, and quartiles from small distributions and tables
- interpret R output and correct the errors in an AI answer
:::

Tutorials are exam preparation sessions. The tasks below are exam-style
questions. You first try each task in pairs. You may use AI
for this, following Attempt, Ask, Verify, Explain. Then the class discusses
the solution together.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 1 to 3. Without looking at the
chapters again, complete these definitions:

1. An observational unit is ...
2. An event is ...
3. A random variable is ...
4. A PMF is ...
5. The variance of a random variable averages ...
6. The lower quartile $Q_1$ and the upper quartile $Q_3$ are ...

Use the [formula sheet](formula-sheet.qmd) up to and including Lecture 3.

Review [Mathematics Refresher: sets and event
notation](../resources/math-refresher.html#sets-and-event-notation) and
[summation notation](../resources/math-refresher.html#summation-notation) if the notation is difficult for you.

::: {.callout-note collapse="true"}
## Check your preparation

1. An observational unit is the unit that one row of the data describes.
2. An event is a specified subset of the sample space.
3. A random variable is a numerical measurement of the outcome: a fixed rule
   that turns each outcome into a number.
4. A PMF assigns probability $P(X=x)$ to each value in the support of a
   discrete random variable. Its masses are non-negative and sum to 1.
5. Variance averages squared deviations from the variable's mean:
   $\operatorname{Var}(X)=E\!\left[(X-E[X])^2\right]$.
6. At least one quarter of the observations are at or below $Q_1$, and at
   least three quarters are at or above it. At least three quarters of the
   observations are at or below $Q_3$, and at least one quarter are at or
   above it. The difference $Q_3-Q_1$ is the interquartile range.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Attempt each task before opening its answer. A complete written answer names
the quantity and shows the calculation. It reports the value and units and
interprets the value in context. It also states one important limit of the
model or of the measurement.

Tasks 1 to 8 and Task 11 belong to the tutorial. Tasks 9 and 10 are extra
practice for home.

Tasks 1, 3, and 4 and the individual practice after class use the Shark Tank
teaching file. Tasks 2 and 5 to 7 use invented examples. Load the file
once:

```{r}
#| label: bep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
```

### Task 1: Data and measurement

[Shark Tank data]{.shark-tag} R reports:

```{r}
#| label: bep01-task1
#| echo: false
data.frame(
  rows = nrow(sharks),
  columns = ncol(sharks),
  deals = sum(sharks$deal_on_show),
  deal_proportion = round(mean(sharks$deal_on_show), 3),
  median_description_words = median(sharks$description_words)
)
```

1. What is the observational unit of the Shark Tank file?
2. Verify the deal proportion from 882 deals and 559 non-deals.
3. Explain why the mean of `deal_on_show` is a proportion.
4. Evaluate: "The data contain 1,441 successful companies."
5. What exactly does the variable `description_words` measure?

::: {.callout-note collapse="true"}
## Check your answer

1. One row is one recorded televised pitch.
2. There are $882+559=1441$ records and $882/1441\approx0.612$, so 61.2%
   reached an on-air deal.
3. The mean of a 0/1 indicator equals the share coded 1 because its sum
   counts the 1s.
4. "Successful companies" is incorrect: the outcome is a deal made on the
   show, not completed funding or later venture performance.
5. Word count describes the written text in the source dataset, not the
   spoken pitch.
:::

### Task 2: Variables and samples

An invented example. A coffee-bar start-up records, for each of 150 orders:
`order_id`, `drink` (espresso, latte, or tea), `cup_size` (small, medium, or
large), `items` (the number of items), and `amount_eur` (the amount paid, in
euros).

1. Classify each variable as categorical, ordinal, discrete, continuous, or
   an identifier.
2. The founder asks 150 loyalty-card holders whether they would pay more for
   oat milk. Is the main problem sampling variability or selection bias?
   Explain in two sentences.

::: {.callout-note collapse="true"}
## Check your answer

1. `order_id` is an identifier, `drink` is categorical, `cup_size` is
   ordinal, `items` is discrete, and `amount_eur` is continuous.
2. The main problem is selection bias. Loyalty-card holders are regular
   customers who like the bar, so they may be more willing to pay extra than
   other customers. Asking more card holders would not fix this, because
   the sample was not drawn at random from all customers.
:::

### Task 3: Events and probability rules

[Shark Tank data]{.shark-tag} The experiment selects one of the 1,441 records at random, each with the
same probability. The two events are defined exactly as in Lecture 2:

```{r}
#| label: bep01-events
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
```

R reports:

```{r}
#| label: bep01-task3
#| echo: false
c(P_D = mean(D), P_L = mean(L), P_D_and_L = mean(D & L))
```

1. Define the sample space $\Omega$. Then explain the difference between
   an outcome and the event $D$.
2. Translate $D^c$, $D\cap L$, and $D\cup L$ into words.
3. Calculate $P(D^c)$.
4. Calculate $P(D\cup L)$ using the addition rule.
5. Explain why these probabilities do not automatically apply to a future US
   entrepreneur.

::: {.callout-note collapse="true"}
## Check your answer

1. The sample space is the set of 1,441 complete pitch records. One
   selected record is an outcome, while $D$ is the subset of 882 deal
   records.
2. The complement is no on-air deal, the intersection is a late-season deal,
   and the union is late-season, deal, or both.
3. $P(D^c)=1-882/1441=559/1441\approx0.388$.
4. The union is $P(D)+P(L)-P(D\cap L)\approx0.771$.
5. To use these probabilities for a future entrepreneur, you need a defined
   target population and a comparable selection, time period, and outcome.
:::

### Task 4: PMF, CDF, and expectation

[Shark Tank data]{.shark-tag} Select one of the 1,441 records at random, as in Task 3. Let
$X=1$ if the selected pitch reached an on-air deal and $X=0$ otherwise. Use
$p=882/1441$.

1. Write the PMF and verify that its masses sum to 1.
2. Write the full piecewise CDF $F_X(a)$.
3. Calculate $E[X]$ from the PMF.
4. Explain why $E[X]$ equals the observed proportion of 1s.
5. Explain the difference between the random variable $X$ and a realised
   value $x=1$.

::: {.callout-note collapse="true"}
## Check your answer

1. The PMF is
   $$
   p_X(0)=1-p,\qquad p_X(1)=p,
   $$
   and the masses sum to $1-p+p=1$.
2. The CDF is
   $$
   F_X(a)=
   \begin{cases}
   0,&a<0,\\
   1-p,&0\le a<1,\\
   1,&a\ge1.
   \end{cases}
   $$
3. The expectation is $E[X]=0(1-p)+1p=p\approx0.612$.
4. Because the indicator contains 882 ones among 1,441 observations, its
   mean is also $882/1441$.
5. $X$ is the random variable: a numerical measurement of the outcome, a
   fixed rule that turns each outcome into a number (1 for a deal, 0 for no
   deal). $x=1$ is one realised value, the number obtained when the selected
   pitch has a deal.
:::

### Task 5: Variance from a small PMF

An invented example. A venture studio models $N$, the number of investors who
make an offer to one of its start-ups, with the PMF

| $n$ | 0 | 1 | 2 |
|---|---:|---:|---:|
| $p_N(n)$ | 0.5 | 0.3 | 0.2 |

1. Calculate $E[N]$.
2. Calculate $\operatorname{Var}(N)$ from the definition
   $E\!\left[(N-E[N])^2\right]$.
3. Verify the result with the computational form
   $E[N^2]-\left(E[N]\right)^2$.
4. Report the standard deviation with its unit, and state in one sentence
   what it measures here.
5. Each offer costs the studio 3 hours of meetings, and every start-up needs
   2 hours of preparation. The time spent on one start-up is therefore
   $T=3N+2$ hours. Calculate $E[T]$ and $\operatorname{SD}(T)$.

::: {.callout-note collapse="true"}
## Check a model answer

1. $E[N]=0(0.5)+1(0.3)+2(0.2)=0.7$ offers.
2. Definition form:
   $$
   \operatorname{Var}(N)=(0-0.7)^2(0.5)+(1-0.7)^2(0.3)+(2-0.7)^2(0.2)
   =0.245+0.027+0.338=0.61.
   $$
3. Computational form: $E[N^2]=0^2(0.5)+1^2(0.3)+2^2(0.2)=1.1$, so
   $\operatorname{Var}(N)=1.1-0.7^2=1.1-0.49=0.61$. The two methods agree.
4. The standard deviation is $\sqrt{0.61}\approx0.78$ offers. It measures
   the typical distance of the realised number of offers from the mean of
   0.7, in the original unit (offers), whereas the variance is in squared
   offers.
5. Time per start-up: use the rules for $aX+b$ with $a=3$ and $b=2$.
   $E[T]=3E[N]+2=3\times0.7+2=4.1$ hours.
   $\operatorname{SD}(T)=|3|\operatorname{SD}(N)\approx3\times0.78=2.34$
   hours, more than half of the expected time. The 2 hours of preparation
   raise the expected time and leave the spread unchanged.
:::

### Task 6: Check the AI

You asked an AI assistant about the number of offers $N$ in Task 5. It
answered:

> "E[N] = 0.7, so most start-ups get 0.7 offers. The variance is 0.61, so the
> number of offers typically differs from 0.7 by 0.61 offers."

Check the answer sentence by sentence. Find what is wrong and write the
corrected answer.

::: {.callout-note collapse="true"}
## Check your answer

The first sentence is wrong. No start-up gets 0.7 offers, because $N$ is a
count with the values 0, 1, and 2. The expected value is a
probability-weighted average, not the most likely value. The most likely
number is 0 offers, with probability 0.5. The second sentence is also wrong.
The variance 0.61 is in squared offers. The typical distance from 0.7 is the
standard deviation, $\sqrt{0.61}\approx0.78$ offers. Corrected answer: on
average a start-up gets 0.7 offers, half of the start-ups get no offer, and
the number of offers typically differs from 0.7 by about 0.78 offers.
:::

### Task 7: Read a histogram by hand

An invented example. A venture records the delivery times of 120 orders. A histogram groups them
into intervals of 10 minutes, and each bar includes its upper limit. The bar
counts, from the first interval (0 to 10 minutes) to the fifth (above 40, up
to 50 minutes), are 18, 46, 32, 16, and 8.

1. What does one observation represent, and what is on each axis?
2. Describe the tallest bar in one full sentence.
3. Which interval contains the median? Show the cumulative counts.
4. Describe the shape, and state one thing the histogram cannot tell you.

::: {.callout-note collapse="true"}
## Check your answer

1. One observation is one order. The horizontal axis is the delivery time
   in minutes and the vertical axis is the number of orders.
2. The tallest bar: 46 orders took more than 10 and at most 20 minutes.
3. With 120 orders the median lies between the 60th and the 61st ordered
   value. The cumulative counts are 18 and then 18 + 46 = 64, so both values
   lie in the second interval: the median delivery time is above 10 and at
   most 20 minutes.
4. The distribution is right-skewed: most orders arrive within 30 minutes
   and a few take much longer. The histogram cannot say which orders were
   late or why, and it hides the exact times inside each interval.
:::

### More calculation practice {#b-ep01-calculation}

Tasks 8 to 11 practise the calculations of Lectures 1 to 3 by hand. Tasks 8
and 11 belong to the tutorial. Tasks 9 and 10 are extra practice for home.
All four tasks use invented examples. You need
a simple calculator and the formula sheet.

### Task 8: Mean, spread, and quartiles from a frequency table

An invented example. A founder runs an escape room. Groups of 2 to 6 persons
book the room for one hour. The table shows the group size of the 40
bookings of last week.

| Group size (persons) | 2 | 3 | 4 | 5 | 6 |
|---|---:|---:|---:|---:|---:|
| Number of bookings | 4 | 12 | 10 | 8 | 6 |

Give every booking the same weight, $1/40$. This is the empirical
distribution of Lecture 3, so the relative frequencies work like a PMF.

1. State what one observation is. Calculate the relative frequency of each
   group size and check that the five values sum to 1.
2. Calculate the mean group size.
3. Calculate the variance and the standard deviation of this empirical
   distribution. Use the variance formula of Lecture 3 with the relative
   frequencies as weights. This version divides by $n$. State the unit. Then
   calculate the sample standard deviation $s$ of Lecture 3, which divides
   by $n-1$.
4. Write down the cumulative counts and find the median.
5. Find the lower quartile $Q_1$ and the upper quartile $Q_3$ with
   cumulative counts, as for the median in Lecture 1. Add the counts from the
   left until the running total first reaches one quarter of the
   observations. The group size at which this
   happens is $Q_1$. Do the same with three quarters for $Q_3$. Then
   calculate the interquartile range and interpret it in one sentence.
6. Compare the mean with the median. What does the comparison say about the
   shape of the distribution?

For these 40 values, R's `quantile()` reports the same median and the same
quartiles as the rules in parts 4 and 5.

::: {.callout-note collapse="true"}
## Check your answer

1. One observation is one booking. The relative frequencies are
   $4/40=0.10$, $12/40=0.30$, $10/40=0.25$, $8/40=0.20$, and $6/40=0.15$.
   Their sum is 1.
2. The mean is
   $2(0.10)+3(0.30)+4(0.25)+5(0.20)+6(0.15)=0.20+0.90+1.00+1.00+0.90=4.0$
   persons. The counts give the same result:
   $(2\times4+3\times12+4\times10+5\times8+6\times6)/40=160/40=4.0$.
3. The variance is
   $$
   (2-4)^2(0.10)+(3-4)^2(0.30)+(4-4)^2(0.25)+(5-4)^2(0.20)+(6-4)^2(0.15)
   $$
   $$
   =0.40+0.30+0+0.20+0.60=1.50.
   $$
   The standard deviation is $\sqrt{1.50}\approx1.22$ persons. A booking
   typically differs from the mean of 4 persons by about 1.2 persons, which
   is about 31% of the mean. This variance divides by $n=40$. The sum of
   the squared deviations is $40\times1.50=60$. The sample variance divides
   it by $n-1=39$: $s^2=60/39\approx1.54$ and $s\approx1.24$ persons, the
   value of R's `sd()`.
4. The cumulative counts are 4, 16, 26, 34, and 40. With 40 observations the
   median is the average of the 20th and the 21st ordered value. The ordered
   values 17 to 26 all have group size 4, so the median is 4 persons.
5. One quarter of 40 is 10. The running total first reaches 10 at group size
   3, because $16\ge10$, so $Q_1=3$. Three quarters of 40 is 30. The running
   total first reaches 30 at group size 5, because $34\ge30$, so $Q_3=5$.
   The interquartile range is $5-3=2$ persons: the middle half of the
   bookings lies between 3 and 5 persons. Of the 40 bookings, 16 are at or
   below 3 persons. That is 40%, more than one quarter, which is why the
   definition of $Q_1$ says "at least one quarter".
6. The mean and the median are both 4 persons, so the two agree. The
   distribution has no long tail on one side that pulls the mean away from
   the median.

Verification in R:

```{r}
#| label: bep01-frequency-check
group_size <- rep(2:6, times = c(4, 12, 10, 8, 6))
c(mean = mean(group_size),
  var_divide_n = mean((group_size - mean(group_size))^2),
  sum_of_squares = sum((group_size - mean(group_size))^2),
  var_divide_n_minus_1 = var(group_size),
  sd_divide_n_minus_1 = sd(group_size))
cumsum(table(group_size))
quantile(group_size, c(0.25, 0.50, 0.75))
```
:::

### Task 9: A random experiment, its sample space, and events (extra practice for home)

An invented example. A streetwear start-up sells one hoodie in five sizes:
XS, S, M, L, and XL. The random experiment is this: the next order arrives
and its size is recorded. The founders use the shares of last year's orders
as probabilities:

| Size | XS | S | M | L | XL |
|---|---:|---:|---:|---:|---:|
| Probability | 0.05 | 0.20 | 0.35 | 0.30 | 0.10 |

Two events matter for the founders. $A$ is the event that the order is for a
size up to M, that is XS, S, or M. $B$ is the event that the order is for a
standard size, that is S, M, or L. The start-up keeps the standard sizes
in stock and makes XS and XL only after a customer orders them.

1. Write down the sample space $\Omega$. Is L an outcome or an event? What
   is $\{\text{L}\}$?
2. Check that the table is a valid probability model.
3. Write $A$ and $B$ in set notation, and calculate $P(A)$ and $P(B)$.
4. Describe $A^c$ and $B^c$ in words and write them in set notation.
   Calculate their probabilities with the complement rule.
5. Write $A\cap B$ and $A\cup B$ in set notation and in words. Calculate
   $P(A\cup B)$ with the addition rule. Explain why $P(A)+P(B)$ cannot be
   the answer.
6. Which size is neither in $A$ nor in $B$? Calculate its probability from
   part 5 with the complement rule, and check the result in the table.
7. State one assumption that the founders make when they use these
   probabilities for the next order.

::: {.callout-note collapse="true"}
## Check your answer

1. $\Omega=\{\text{XS},\text{S},\text{M},\text{L},\text{XL}\}$. L is an
   outcome: one complete possible result of the experiment. $\{\text{L}\}$
   is an event: the set that contains only this outcome.
2. Every probability is at least 0, and
   $0.05+0.20+0.35+0.30+0.10=1$, so $P(\Omega)=1$. The five sizes are
   disjoint, so their probabilities may be added.
3. $A=\{\text{XS},\text{S},\text{M}\}$ and $P(A)=0.05+0.20+0.35=0.60$.
   $B=\{\text{S},\text{M},\text{L}\}$ and $P(B)=0.20+0.35+0.30=0.85$.
4. $A^c=\{\text{L},\text{XL}\}$: the order is for a size above M.
   $P(A^c)=1-0.60=0.40$. $B^c=\{\text{XS},\text{XL}\}$: the order is for a
   size that is made only after an order. $P(B^c)=1-0.85=0.15$. The table gives the
   same values: $0.30+0.10=0.40$ and $0.05+0.10=0.15$.
5. $A\cap B=\{\text{S},\text{M}\}$: the size is up to M and it is a standard
   size. Its probability is $0.20+0.35=0.55$.
   $A\cup B=\{\text{XS},\text{S},\text{M},\text{L}\}$: the size is up to M,
   or it is a standard size, or both. The addition rule gives
   $P(A\cup B)=0.60+0.85-0.55=0.90$. The sum $P(A)+P(B)=1.45$ is above 1,
   so it cannot be a probability. It counts the sizes S and M twice.
6. Only XL. In symbols, $(A\cup B)^c=\{\text{XL}\}$, and the complement rule
   gives $1-0.90=0.10$. The table shows the same value for XL.
7. The probabilities are shares of last year's orders. They describe the
   next order only if customers choose sizes in the same way as last year.
   A new group of customers or a new design of the hoodie could change the
   shares.

Verification in R:

```{r}
#| label: bep01-sizes-check
size_prob <- c(XS = 0.05, S = 0.20, M = 0.35, L = 0.30, XL = 0.10)
size_A <- c("XS", "S", "M")
size_B <- c("S", "M", "L")
p_A <- sum(size_prob[size_A])
p_B <- sum(size_prob[size_B])
p_A_and_B <- sum(size_prob[intersect(size_A, size_B)])
c(total = sum(size_prob), P_A = p_A, P_B = p_B,
  P_not_A = 1 - p_A, P_not_B = 1 - p_B,
  P_A_and_B = p_A_and_B,
  P_A_or_B = p_A + p_B - p_A_and_B,
  P_neither = 1 - (p_A + p_B - p_A_and_B))
```
:::

### Task 10: From a CDF to probabilities and to the PMF (extra practice for home)

An invented example. A founder rents out four cargo bikes to students who
move house. Let $X$ be the number of bikes that are rented out on a
Saturday. Her model for $X$ is given as a CDF:

| Threshold $a$ | $F_X(a)=P(X\le a)$ |
|---|---:|
| $a<0$ | 0 |
| $0\le a<1$ | 0.05 |
| $1\le a<2$ | 0.20 |
| $2\le a<3$ | 0.50 |
| $3\le a<4$ | 0.80 |
| $a\ge4$ | 1 |

1. State the possible values of $X$. How does the table show them?
2. Read $F_X(2)$, $F_X(2.5)$, and $F_X(6)$ from the table. Say in words what $F_X(2)$
   means.
3. Calculate $P(X>2)$, $P(1<X\le3)$, and $P(1\le X\le3)$.
4. Recover the PMF of $X$ from the CDF, and check that it is a valid PMF.
5. Check your PMF in two ways. Calculate $F_X(3)$ again as a running total, and
   calculate $P(1\le X\le3)$ again by adding probability masses.
6. Interpret $p_X(4)$ in one sentence. State one thing that this number
   cannot tell the founder.

::: {.callout-note collapse="true"}
## Check your answer

1. $X$ can take the values 0, 1, 2, 3, and 4. These are the thresholds at
   which the CDF jumps. Between two of these values the CDF stays flat,
   because no probability lies there.
2. $F_X(2)=0.50$: with probability 0.50, at most two bikes are rented out on
   a Saturday. $F_X(2.5)=0.50$ as well, because $X$ takes no value between 2
   and 2.5. $F_X(6)=1$, because $X$ is never above 4.
3. $P(X>2)=1-F_X(2)=1-0.50=0.50$.
   $P(1<X\le3)=F_X(3)-F_X(1)=0.80-0.20=0.60$. The event $1\le X\le3$
   includes the value 1. $X$ takes whole numbers only, so this event is the
   same as $0<X\le3$, and
   $P(1\le X\le3)=F_X(3)-F_X(0)=0.80-0.05=0.75$.
4. Each probability mass is the size of the jump of the CDF at that value:
   $p_X(0)=0.05$, $p_X(1)=0.20-0.05=0.15$, $p_X(2)=0.50-0.20=0.30$,
   $p_X(3)=0.80-0.50=0.30$, and $p_X(4)=1-0.80=0.20$.

   | $x$ | 0 | 1 | 2 | 3 | 4 |
   |---|---:|---:|---:|---:|---:|
   | $p_X(x)$ | 0.05 | 0.15 | 0.30 | 0.30 | 0.20 |

   No mass is negative, and $0.05+0.15+0.30+0.30+0.20=1$.
5. $F_X(3)=0.05+0.15+0.30+0.30=0.80$, as in the table.
   $P(1\le X\le3)=0.15+0.30+0.30=0.75$, as in part 3.
6. $p_X(4)=0.20$: with probability 0.20, on about one Saturday in five, all
   four bikes are rented out. The number counts rented bikes. It cannot tell
   how many students asked for a bike on such a day, so it does not show how
   often a fifth bike would be rented.

Verification in R:

```{r}
#| label: bep01-cdf-check
bike_cdf <- c(0.05, 0.20, 0.50, 0.80, 1)
bike_pmf <- diff(c(0, bike_cdf))
names(bike_pmf) <- 0:4
bike_pmf
c(total = sum(bike_pmf),
  P_more_than_2 = 1 - bike_cdf[3],
  P_above_1_to_3 = bike_cdf[4] - bike_cdf[2],
  P_1_to_3 = sum(bike_pmf[c("1", "2", "3")]),
  F_3 = sum(bike_pmf[c("0", "1", "2", "3")]))
```
:::

### Task 11: Two options, compared by expected value and spread

An invented example. Two founders sell second-hand designer clothes online.
For December they want to rent a pop-up store, and they can choose between
two locations. Option A is a shop in a side street of the city centre. Option B
is a booth at the Christmas market, where the rent is high and the number of
visitors depends on the weather. $X$ is the profit of the month with option
A and $Y$ is the profit of the month with option B, both in thousands of
euros. A negative value is a loss. The founders use these two PMFs:

| $x$ | 0 | 1 | 2 | 3 | 4 |
|---|---:|---:|---:|---:|---:|
| $p_X(x)$ | 0.05 | 0.25 | 0.45 | 0.15 | 0.10 |

| $y$ | $-2$ | 0 | 2 | 4 | 8 |
|---|---:|---:|---:|---:|---:|
| $p_Y(y)$ | 0.10 | 0.10 | 0.20 | 0.50 | 0.10 |

1. Check that both tables are valid PMFs.
2. Calculate $E[X]$ and $E[Y]$.
3. Calculate $\operatorname{Var}(X)$ and $\operatorname{Var}(Y)$ from the
   definition $E\!\left[(X-E[X])^2\right]$.
4. Verify both variances with the computational form
   $E[X^2]-\left(E[X]\right)^2$.
5. Calculate both standard deviations and state them in euros. Compare each
   standard deviation with the expected profit of its option.
6. Calculate the probability of a loss for each option.
7. Decide which option you would choose. Give the numbers that support your
   choice.

::: {.callout-note collapse="true"}
## Check your answer

1. No mass is negative. The masses sum to 1 in both tables:
   $0.05+0.25+0.45+0.15+0.10=1$ and $0.10+0.10+0.20+0.50+0.10=1$.
2. $E[X]=0(0.05)+1(0.25)+2(0.45)+3(0.15)+4(0.10)=0+0.25+0.90+0.45+0.40=2.0$,
   an expected profit of 2,000 euros.
   $E[Y]=(-2)(0.10)+0(0.10)+2(0.20)+4(0.50)+8(0.10)=-0.20+0+0.40+2.00+0.80=3.0$,
   an expected profit of 3,000 euros.
3. Definition form:
   $$
   \operatorname{Var}(X)=(0-2)^2(0.05)+(1-2)^2(0.25)+(2-2)^2(0.45)+
   (3-2)^2(0.15)+(4-2)^2(0.10)
   $$
   $$
   =0.20+0.25+0+0.15+0.40=1.00,
   $$
   $$
   \operatorname{Var}(Y)=(-2-3)^2(0.10)+(0-3)^2(0.10)+(2-3)^2(0.20)+
   (4-3)^2(0.50)+(8-3)^2(0.10)
   $$
   $$
   =2.50+0.90+0.20+0.50+2.50=6.60.
   $$
4. Computational form:
   $E[X^2]=0(0.05)+1(0.25)+4(0.45)+9(0.15)+16(0.10)=0+0.25+1.80+1.35+1.60=5.00$,
   so $\operatorname{Var}(X)=5.00-2^2=1.00$.
   $E[Y^2]=4(0.10)+0(0.10)+4(0.20)+16(0.50)+64(0.10)=0.40+0+0.80+8.00+6.40=15.60$,
   so $\operatorname{Var}(Y)=15.60-3^2=6.60$. Both forms agree.
5. $\operatorname{SD}(X)=\sqrt{1.00}=1.00$ thousand euros, that is 1,000
   euros. This is half of the expected profit of option A.
   $\operatorname{SD}(Y)=\sqrt{6.60}\approx2.57$ thousand euros, that is
   about 2,570 euros. This is about 86% of the expected profit of option B.
6. Option A has no negative value, so $P(X<0)=0$. For option B,
   $P(Y<0)=p_Y(-2)=0.10$.
7. Both choices can be correct if the numbers support the reason. Option B has
   the higher expected profit: 3,000 euros against 2,000 euros, which is
   50% more. Option A has the smaller spread: its standard deviation of
   1,000 euros is less than half of the 2,570 euros of option B. Option A
   never makes a loss, and option B loses 2,000 euros with probability
   0.10. The founders rent the store for one December only, so they cannot
   count on an average over many repetitions. If a loss of 2,000 euros
   would put the venture in danger, they choose option A. If they can afford
   such a loss, option B is the better choice because of its higher
   expected profit.

Verification in R:

```{r}
#| label: bep01-options-check
x_val <- 0:4
x_prob <- c(0.05, 0.25, 0.45, 0.15, 0.10)
y_val <- c(-2, 0, 2, 4, 8)
y_prob <- c(0.10, 0.10, 0.20, 0.50, 0.10)
mean_x <- sum(x_val * x_prob)
mean_y <- sum(y_val * y_prob)
c(E_X = mean_x, E_Y = mean_y,
  Var_X_definition = sum((x_val - mean_x)^2 * x_prob),
  Var_X_shortcut = sum(x_val^2 * x_prob) - mean_x^2,
  Var_Y_definition = sum((y_val - mean_y)^2 * y_prob),
  Var_Y_shortcut = sum(y_val^2 * y_prob) - mean_y^2,
  SD_X = sqrt(sum((x_val - mean_x)^2 * x_prob)),
  SD_Y = sqrt(sum((y_val - mean_y)^2 * y_prob)),
  P_loss_X = sum(x_prob[x_val < 0]),
  P_loss_Y = sum(y_prob[y_val < 0]))
```
:::

:::

::: {.course-phase .phase-after}

## After class

### Individual practice

[Shark Tank data]{.shark-tag} The R output below describes the number of words in the written description
(`description_words`) in the Shark Tank file:

```{r}
#| label: bep01-timed
#| echo: false
w <- sharks$description_words
c(
  mean = mean(w),
  median = median(w),
  minimum = min(w),
  maximum = max(w)
)
```

Write no more than 100 words. Compare mean and median, calculate the range,
and say what the variable does and does not measure. Then explain which value
equals the expectation if every record gets probability $1/1441$.

::: {.callout-note collapse="true"}
## Check a model answer

The written descriptions have a mean length of about 5.0 words and a median
of 5 words, so the two agree. Their range is $13-1=12$ words. The histogram
in Lecture 1 shows a thin tail to the right (mildly right-skewed), although
mean and median almost agree. If each record gets probability $1/1441$, the expectation of this
distribution is the mean, about 5.0 words. These values describe the written
text in the dataset, not the length or quality of the spoken pitch.
:::

### Check your understanding

Complete from memory:

1. An event is ...
2. A PMF must ...
3. A CDF at $a$ reports ...
4. Expectation is ...
5. Variance is ... and its square root is ...

::: {.callout-note collapse="true"}
## Check your answers

1. An event is a specified subset of the sample space.
2. A PMF is non-negative and sums to 1 over the support.
3. $F_X(a)=P(X\le a)$.
4. $E[X]=\sum_x x\, p_X(x)$, the probability-weighted average of the possible
   values.
5. Variance is the probability-weighted average of squared deviations from
   the mean, $\operatorname{Var}(X)=E[X^2]-\left(E[X]\right)^2$. Its square
   root is the standard deviation, in the variable's original unit.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Pick one
uncertain count that matters for it, for example customers who sign up in a
week, offers received, or orders returned. Write two sentences. First,
sketch a small PMF for the count: three or four values with rough
probabilities. Then say what its expectation and standard deviation would
tell you about a real decision.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus venture that sells boxes of prepared meals. "Weekly catering requests: 0 with probability 0.4, 1 with 0.4, 2 with 0.15, and 3 with 0.05, so I expect about 0.85 requests per week. The standard deviation tells me whether a weekend helper is justified. With a high spread, some weeks need a helper even though the average says no."
:::

You can discuss this task, and any other task in this tutorial, with
Tilly for QM Basic, the course AI assistant.
::::

### More practice

For extra exercises on the lectures of this tutorial, with worked solutions, ask Tilly for
QM Basic, the course AI assistant. For example: "Give me an extra exercise for
Tutorial 1. Show the solution only after I send my answer."

[Continue to Lecture 4: Conditioning, Bayes' rule, and independence](b04-conditioning-bayes-independence.qmd).

:::
