# Tutorial 1: Evidence and probability foundations {#b-ep01}

```{r}
#| label: bep01-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 1, you can</span>

- distinguish data, events, random variables, and distributions
- apply probability rules to the Shark Tank record-selection experiment
- calculate and interpret a Bernoulli PMF, CDF, and expectation
- calculate a variance and standard deviation from a small PMF
- calculate the expected value and standard deviation of a shifted and
  rescaled variable $aX+b$
- write a concise interpretation of supplied R output
- read a histogram by hand and find the interval that contains the median.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the solution is discussed together.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 1--3. Without reopening the
chapters, complete these definitions:

1. An observational unit is ...
2. An event is ...
3. A random variable is ...
4. A PMF is ...
5. The variance of a random variable averages ...

Use the [formula sheet](formula-sheet.qmd) up to and including Lecture 3.

Review [Mathematics Refresher: sets, fractions, and
sums](../resources/math-refresher.html) if the notation is slowing you down.

::: {.callout-note collapse="true"}
## Check your preparation

1. An observational unit is the entity represented by one row of the data.
2. An event is a specified subset of the sample space.
3. A random variable is a numerical measurement of the outcome: a fixed rule
   that turns each outcome into a number.
4. A PMF assigns probability $P(X=x)$ to each value in the support of a
   discrete random variable. Its masses are non-negative and sum to 1.
5. Variance averages squared deviations from the variable's mean:
   $\operatorname{Var}(X)=E\!\left[(X-E[X])^2\right]$.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Attempt each task before opening its answer. A complete written answer should
name the quantity, show the calculation, report the value and units, interpret
it in context, and state one important limit of the model or of the
measurement.

Tasks 1 to 3 and the timed practice after class use the Shark Tank teaching
file. Tasks 4 and 5 use invented examples. Load the file once:

```{r}
#| label: bep01-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
```

### Task 1: Data and measurement

R reports:

```{r}
#| label: bep01-task1
#| echo: false
c(
  rows = nrow(sharks),
  columns = ncol(sharks),
  deals = sum(sharks$deal_on_show),
  deal_proportion = mean(sharks$deal_on_show),
  median_description_words = median(sharks$description_words)
)
```

1. What is the observational unit?
2. Verify the deal proportion from 882 deals and 559 non-deals.
3. Explain why the mean of `deal_on_show` is a proportion.
4. Evaluate: "The data contain 1,441 successful companies."
5. What exactly does `description_words` measure?

::: {.callout-note collapse="true"}
## Check your answer

One row is one recorded televised pitch. There are $882+559=1441$ records and
$882/1441\approx0.612$, so 61.2% reached an on-air agreement. The mean of a
0/1 indicator equals the share coded 1 because its sum counts the 1s.
"Successful companies" is incorrect: the outcome is an agreement made on the
show, not completed funding or later venture performance. Word count
describes the written text in the source dataset, not the spoken pitch.
:::

### Task 2: Events and probability rules

The experiment selects one of the 1,441 records uniformly. The two events are
built exactly as in Lecture 2:

```{r}
#| label: bep01-events
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
```

R reports:

```{r}
#| label: bep01-task2
#| echo: false
c(P_D = mean(D), P_L = mean(L), P_D_and_L = mean(D & L))
```

1. Define $\Omega$, and distinguish an outcome from event $D$.
2. Translate $D^c$, $D\cap L$, and $D\cup L$ into words.
3. Calculate $P(D^c)$.
4. Calculate $P(D\cup L)$ using the addition rule.
5. Explain why these probabilities do not automatically apply to a future US
   entrepreneur.

::: {.callout-note collapse="true"}
## Check your answer

The sample space is the set of 1,441 complete pitch records. One selected
record is an outcome, while $D$ is the subset of 882 deal records. The
complement is no on-air deal, the intersection is a late-season deal, and the
union is late-season, deal, or both. $P(D^c)=1-882/1441=559/1441\approx0.388$.
The union is $P(D)+P(L)-P(D\cap L)\approx0.771$. To use these probabilities
for a future entrepreneur, you need a defined target population and a
comparable selection, time period, and outcome.
:::

### Task 3: PMF, CDF, and expectation

Let $X=1$ for an on-air deal and $X=0$ otherwise. Use $p=882/1441$.

1. Write the PMF and verify that its masses sum to 1.
2. Write the full piecewise CDF $F_X(a)$.
3. Calculate $E[X]$ from the PMF.
4. Explain why $E[X]$ equals the observed proportion of 1s.
5. Explain the difference between the random variable $X$ and a realised
   value $x=1$.

::: {.callout-note collapse="true"}
## Check your answer

$$
p_X(0)=1-p,\qquad p_X(1)=p,
$$

and the masses total $1-p+p=1$.

$$
F_X(a)=
\begin{cases}
0,&a<0,\\
1-p,&0\le a<1,\\
1,&a\ge1.
\end{cases}
$$

The expectation is $E[X]=0(1-p)+1p=p\approx0.612$. Because the indicator
contains 882 ones among 1,441 observations, its mean is also $882/1441$.
$X$ is the random variable: a numerical measurement of the outcome, a fixed
rule that turns each outcome into a number (1 for a deal, 0 for no deal).
$x=1$ is one realised value, the number obtained when the selected pitch
has a deal.
:::

### Task 4: Variance from a small PMF

An invented example. A venture studio models $N$, the number of investors who
make an offer to one of its startups, with the PMF

| $n$ | 0 | 1 | 2 |
|---|---:|---:|---:|
| $p_N(n)$ | 0.5 | 0.3 | 0.2 |

1. Calculate $E[N]$.
2. Calculate $\operatorname{Var}(N)$ from the definition
   $E\!\left[(N-E[N])^2\right]$.
3. Verify the result with the computational form
   $E[N^2]-\left(E[N]\right)^2$.
4. Report the standard deviation with its unit, and state in one sentence
   what it measures here.
5. Each offer costs the studio 3 hours of meetings, and every startup needs
   2 hours of preparation. The time spent on one startup is therefore
   $T=3N+2$ hours. Calculate $E[T]$ and $\operatorname{SD}(T)$.

::: {.callout-note collapse="true"}
## Check a model answer

$E[N]=0(0.5)+1(0.3)+2(0.2)=0.7$ offers.

Definition form:
$$
\operatorname{Var}(N)=(0-0.7)^2(0.5)+(1-0.7)^2(0.3)+(2-0.7)^2(0.2)
=0.245+0.027+0.338=0.61.
$$

Computational form: $E[N^2]=0^2(0.5)+1^2(0.3)+2^2(0.2)=1.1$, so
$\operatorname{Var}(N)=1.1-0.7^2=1.1-0.49=0.61$. The two routes agree.

The standard deviation is $\sqrt{0.61}\approx0.78$ offers. It measures the
typical distance of the realised number of offers from the mean of 0.7,
in the original unit (offers), whereas the variance is in squared offers.

Time per startup: use the rules for $aX+b$ with $a=3$ and $b=2$.
$E[T]=3E[N]+2=3\times0.7+2=4.1$ hours.
$\operatorname{SD}(T)=|3|\operatorname{SD}(N)\approx3\times0.78=2.34$ hours,
more than half of the expected time. The 2 hours of preparation raise the
expected time and leave the spread unchanged.
:::

### Task 5: Read a histogram by hand

An invented example. A venture records the delivery times of 120 orders. A histogram groups them
into intervals of 10 minutes, and each bar includes its upper limit. The bar
counts, from the first interval (0 to 10 minutes) to the fifth (above 40, up
to 50 minutes), are 18, 46, 32, 16, and 8.

1. What does one observation represent, and what is on each axis?
2. Put the tallest bar into one full sentence.
3. Which interval contains the median? Show the cumulative counts.
4. Describe the shape, and state one thing the histogram cannot tell you.

::: {.callout-note collapse="true"}
## Check your answer

One observation is one order. The horizontal axis is the delivery time in
minutes and the vertical axis is the number of orders. The tallest bar: 46
orders took more than 10 and at most 20 minutes. With 120 orders the median
lies between the 60th and the 61st ordered value. The cumulative counts are
18 and then 18 + 46 = 64, so both values lie in the second interval: the
median delivery time is above 10 and at most 20 minutes. The distribution is
right-skewed: most orders arrive within 30 minutes and a few take much
longer. The histogram cannot say which orders were late or why, and it hides
the exact times inside each interval.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice: 10 minutes

The R output below describes the number of words in the written description
(`description_words`) in the Shark Tank file:

```{r}
#| label: bep01-timed
#| echo: false
w <- sharks$description_words
c(
  mean = mean(w),
  median = median(w),
  minimum = min(w),
  maximum = max(w)
)
```

Write no more than 100 words. Compare mean and median, calculate the range,
and say what the variable does and does not measure. Then explain which value
equals the expectation if every record receives probability $1/1441$.

::: {.callout-note collapse="true"}
## Check a model answer

The written descriptions have a mean length of about 5.0 words and a median
of 5 words, so the two agree. Their range is $13-1=12$ words. The maximum
lies 8 words above the median and the minimum only 4 words below it, so the
distribution has a longer tail to the right (it is mildly right-skewed). If
each record receives probability $1/1441$, the expectation of this
distribution is the mean, about 5.0 words. These values describe the written
text in the dataset, not the length or quality of the spoken pitch.
:::

### Check your understanding

Complete from memory:

1. An event is ...
2. A PMF must ...
3. A CDF at $a$ reports ...
4. Expectation is ...
5. Variance is ... and its square root is ...

::: {.callout-note collapse="true"}
## Check your answers

1. An event is a specified subset of the sample space.
2. A PMF is non-negative and sums to 1 over the support.
3. $F_X(a)=P(X\le a)$.
4. $E[X]=\sum_x x\, p_X(x)$, the probability-weighted average of the possible
   values.
5. Variance is the probability-weighted average of squared deviations from
   the mean, $\operatorname{Var}(X)=E[X^2]-\left(E[X]\right)^2$. Its square
   root is the standard deviation, in the variable's original unit.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Pick one
uncertain count that matters for it (customers who sign up in a week, offers
received, orders returned), and write two sentences: first sketch a small PMF
for it (three or four values with rough probabilities), then say what its
expectation and standard deviation would tell you about a real decision.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "Weekly catering requests: 0 with probability 0.4, 1 with 0.4, 2 with 0.15, and 3 with 0.05, so I expect about 0.85 requests per week. The standard deviation tells me whether a weekend helper is justified. With a high spread, some weeks need a helper even though the average says no."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant.
::::

[Continue to Lecture 4: Conditioning, Bayes' rule, and independence](b04-conditioning-bayes-independence.qmd).

:::
