# Tutorial 2: Conditioning and relationships {#b-ep02}

```{r}
#| label: bep02-setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 2, you can</span>

- calculate conditional probabilities and apply Bayes' rule and independence
- find covariance and correlation from a joint table and interpret them without causal claims
- read a correlation table in a research paper and check an AI answer
:::

Tutorials are exam preparation sessions. The tasks below are exam-style
questions. You first try each task in pairs. You may use AI
for this, following Attempt, Ask, Verify, Explain. Then the class discusses
the solution together.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 4 and 5. Complete these statements
without looking at the chapters again:

1. In $P(A\mid B)$, the reference class is ...
2. Variance averages ...
3. Events $A$ and $B$ are independent when ...
4. A joint PMF records ...
5. Covariance averages ...
6. Random variables $X$ and $Y$ are independent when ...

Use the [formula sheet](formula-sheet.qmd) up to and including Lecture 5.

Review [Mathematics Refresher: fractions and rearranging
equations](../resources/math-refresher.html#fractions-and-rearranging-equations)
and [summation notation](../resources/math-refresher.html#summation-notation)
if needed.

::: {.callout-note collapse="true"}
## Check your preparation

1. The reference class is $B$, the event after the vertical bar.
2. Variance averages squared deviations from a variable's mean.
3. $P(A\cap B)=P(A)P(B)$.
4. A joint PMF records the probability of every possible pair of values.
5. Covariance averages products of deviations from the two means.
6. $p_{X,Y}(x,y)=p_X(x)\,p_Y(y)$ in every cell of the joint table.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try each task before you open its answer. In every answer, name the formula
and the reference class, check the value, interpret it in context, and say
what the result does not show.

Tasks 1 and 3 to 8 belong to the tutorial. Tasks 2, 9, and 10 are extra
practice for home.

Tasks 1 to 3 and Task 5 use the Shark Tank file and the two events from
Lectures 4 and 5, defined in the same way:

```{r}
#| label: bep02-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
```

### Task 1: Conditional probability and Bayes

[Shark Tank data]{.shark-tag} R reports the counts of the 1,441 pitches by deal and by period:

```{r}
#| label: bep02-table
#| echo: false
tab <- table(
  Deal = factor(D, levels = c(FALSE, TRUE),
                labels = c("No on-air deal", "On-air deal")),
  Period = factor(L, levels = c(FALSE, TRUE),
                  labels = c("Seasons 1-8", "Seasons 9-16"))
)
tab
```

1. Calculate and interpret $P(D\mid L)$.
2. Calculate and interpret $P(L\mid D)$.
3. Verify $P(D\cap L)=P(D\mid L)P(L)$.
4. Use the law of total probability with the partition $L^c$ and $L$ to
   recover $P(D)$.
5. Show how Bayes' rule gives $P(L\mid D)$ from $P(D\mid L)$.

::: {.callout-note collapse="true"}
## Check your answer

$P(D\mid L)=505/734\approx 0.688$: among late-season records, 68.8% reached
an on-air deal. $P(L\mid D)=505/882\approx 0.573$: among deal records, 57.3%
came from late seasons. The joint probability is
$P(D\mid L)P(L)=(505/734)(734/1441)=505/1441\approx 0.350$.

The law of total probability gives

$$
P(D)=\frac{377}{707}\frac{707}{1441}
     +\frac{505}{734}\frac{734}{1441}
     =\frac{882}{1441}\approx 0.612.
$$

Bayes' rule gives

$$
P(L\mid D)=\frac{P(D\mid L)P(L)}{P(D)}
=\frac{(505/734)(734/1441)}{882/1441}
=\frac{505}{882}\approx 0.573.
$$
:::

### Task 2: Independence of events (extra practice for home)

[Shark Tank data]{.shark-tag} Lecture 4 defined independence of events in two equivalent ways: through
conditioning and through the product form. R reports:

```{r}
#| label: bep02-independence
#| echo: false
c(
  P_D = mean(D),
  P_D_given_L = mean(D[L]),
  P_D_and_L = mean(D & L),
  P_D_times_P_L = mean(D) * mean(L)
)
```

1. Use the definition with conditioning to decide whether $D$ and $L$ are
   independent in these records.
2. Use the product form to reach the same conclusion.
3. Explain why "not independent" does not mean "late season causes a deal."
4. Explain why the early and late events are disjoint but not independent.

::: {.callout-note collapse="true"}
## Check your answer

The deal probability is about 0.612 overall and 0.688 within late seasons, so
conditioning on $L$ changes it. In the same way, the observed joint
probability is about 0.350, while the product $P(D)P(L)$ is about 0.312. So
$D$ and $L$ are not independent in these records. This is an association in
televised pitches that the producers selected. It does not show that the
season causes deals, because other factors also changed over the seasons.
Early and late are disjoint because they cannot occur together. Observing
one makes the other impossible, so disjoint events with positive probability
are dependent.
:::

### Task 3: Joint PMF, covariance, and correlation

[Shark Tank data]{.shark-tag} Let $X$ be the 0/1 deal indicator and $Y$ the 0/1 late-period indicator,
defined from the same events:

```{r}
#| label: bep02-indicators
x <- as.integer(D)
y <- as.integer(L)
```

R reports:

```{r}
#| label: bep02-joint
#| echo: false
joint <- prop.table(table(X = x, Y = y))
joint

c(
  E_X = mean(x),
  E_Y = mean(y),
  Var_X = mean((x - mean(x))^2),
  Var_Y = mean((y - mean(y))^2),
  E_XY = mean(x * y),
  covariance = mean(x * y) - mean(x) * mean(y),
  correlation = cor(x, y)
)
```

1. Verify that the joint PMF sums to 1.
2. Recover the marginal PMF of $X$ by summing rows.
3. Verify $\operatorname{Var}(X)=E[X^2]-\left(E[X]\right)^2$.
4. Verify $\operatorname{Cov}(X,Y)=E[XY]-E[X]E[Y]$.
5. Interpret the sign of the covariance.
6. Explain what the correlation adds to the covariance, and what it does not
   prove.
7. Calculate the conditional means $E[X\mid Y=1]$ and $E[X\mid Y=0]$ from the
   joint PMF, and say what they mean.
8. The sum $X+Y$ counts how many of the two events, deal and late season,
   happened for a pitch. Calculate $E[X+Y]$ and
   $\operatorname{Var}(X+Y)$ from the values in the output.

::: {.callout-note collapse="true"}
## Check your answer

1. The four joint probabilities sum to 1:
   $0.229+0.159+0.262+0.350=1$.
2. Summing each row across the two $Y$ columns gives
   $P(X=0)=0.229+0.159=0.388$ and $P(X=1)=0.262+0.350=0.612$. As counts,
   these are $559/1441$ and $882/1441$.
3. Because $X$ takes only the values 0 and 1, $X^2=X$ and $E[X^2]=E[X]$. So
   $\operatorname{Var}(X)=E[X]-\left(E[X]\right)^2=0.612-0.612^2\approx 0.237$, as in the
   output.
4. $E[XY]=P(X=1,Y=1)\approx 0.3505$, and
   $E[X]E[Y]\approx 0.6121\times0.5094\approx 0.3118$. The difference is
   about 0.0387, as in the output.
5. The covariance is positive. The value 1 occurs for both variables
   together a little more often than it would under independence: deals and
   late seasons go together.
6. The correlation divides the covariance by both standard deviations. The
   result, about 0.16, has no units and lies between $-1$ and $1$: a weak
   positive linear association. It does not show a cause, and it does not
   say how important the association is in practice. A correlation of zero
   would not always mean independence, as the U-shaped pattern in Lecture 5
   shows.
7. Divide each column of the joint PMF by its sum. The column sums are
   $P(Y=1)\approx 0.5094$ and $P(Y=0)\approx 0.4906$. Because $X$ is 0 or 1,
   only the $X=1$ entry counts:
   $E[X\mid Y=1]\approx 0.3505/0.5094\approx 0.688$ and
   $E[X\mid Y=0]\approx 0.2616/0.4906\approx 0.533$. Among late-season
   pitches the average of the deal indicator is 0.688, so 68.8% reached a
   deal. Among early-season pitches it is 53.3%.
8. $E[X+Y]=E[X]+E[Y]=0.612+0.509=1.121$.
   $\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)
   +2\operatorname{Cov}(X,Y)=0.237+0.250+2\times0.039\approx 0.565$.
   Without the covariance term the result would be 0.487. The sum varies
   more because deals and late seasons go together.
:::

### Task 4: Read a correlation table from a paper

Research papers in management usually start their results with one table of
means, standard deviations (SD), and correlations. The table below is an
invented example in the layout of such a table. It does not come from a real
study. It describes a survey of 240 employees with four variables:

- creative self-efficacy: how much an employee believes in his or her own
  ability to be creative, rated from 1 to 7.
- job tenure: the number of years in the current job.
- supervisor support: how much support the employee gets from the
  supervisor, rated from 1 to 7.
- creative performance: how creative the employee's work is, rated from 1
  to 7.

The columns "Mean" and "SD" describe each variable alone. The columns 1, 2,
and 3 contain correlations. Each number is the correlation between the
variable in that row and the variable with that number. For example, the
0.31 in row 3 and column 1 is the correlation between supervisor support and
creative self-efficacy.

| Variable | Mean | SD | 1 | 2 | 3 |
|---|---:|---:|---:|---:|---:|
| 1. Creative self-efficacy (1 to 7) | 5.10 | 0.90 | | | |
| 2. Job tenure (years) | 6.40 | 4.20 | 0.08 | | |
| 3. Supervisor support (1 to 7) | 4.80 | 1.10 | 0.31 | 0.05 | |
| 4. Creative performance (1 to 7) | 4.20 | 1.00 | 0.36 | 0.12 | 0.28 |

1. Find the correlation between creative self-efficacy and creative
   performance, and interpret its sign and strength in one sentence.
2. Which pair of variables has the weakest linear association?
3. A manager concludes: "Raising supervisor support will raise creative
   performance." Name one third variable, or one other reason, that could
   produce the correlation of 0.28 without support causing performance.
4. State one thing this table cannot tell you about an individual employee.

::: {.callout-note collapse="true"}
## Check your answer

1. The correlation is 0.36, in row 4 and column 1: a moderate positive
   linear association. Employees with higher creative self-efficacy tended
   to be rated higher on creative performance in this sample.
2. The weakest association is between job tenure and supervisor support,
   with a correlation of 0.05.
3. Several other explanations fit the same number. Employees in creative
   roles may get more support and also perform more creatively. Or
   supervisors may give more support to employees who already perform well.
   A correlation table cannot separate these explanations.
4. The table describes how the variables move together across 240
   employees. It says nothing certain about any single employee. Many employees
   with high self-efficacy have average performance ratings.
:::

### Task 5: Check the AI

[Shark Tank data]{.shark-tag} You asked an AI assistant about the deal indicator $X$ and the late-season
indicator $Y$ of Task 3. It answered:

> "The covariance of deal and late season is only 0.039, so the two are
> almost independent. The correlation of 0.16 shows that a late season
> causes a deal for 16% of the pitches."

Check the answer sentence by sentence. Find what is wrong and write the
corrected answer.

::: {.callout-note collapse="true"}
## Check your answer

The first sentence is wrong. The size of a covariance depends on the units
of the two variables, so a small number does not mean a weak association.
Independence needs $P(X=1,Y=1)=P(X=1)P(Y=1)$, and here 0.350 differs from
0.312, so $X$ and $Y$ are not independent. The second sentence is also
wrong. A correlation is not a share of pitches, and it does not show a
cause. Corrected answer: deals and late seasons go together, with a weak
positive linear association ($r\approx0.16$). The records do not show that
the season causes deals.
:::

### More calculation practice {#b-ep02-calculation}

Tasks 6 to 10 practise the calculations of Lectures 4 and 5 by hand. Tasks
6, 7, and 8 belong to the tutorial. Do Tasks 9 and 10 in the tutorial if
time is left, and otherwise at home. All five tasks use invented examples.
You need a simple calculator and the formula sheet.

### Task 6: Total probability and Bayes' rule from given rates

An invented example. A start-up prints spare parts for household appliances
on two 3D printers. The old printer makes 60% of all parts and the new
printer makes 40%. Of the parts from the old printer, 5% fail the quality
check. Of the parts from the new printer, 10% fail, because the team is
still learning how to set up the new printer. One part is selected at random. Let $N$ be the
event that the part comes from the new printer, and $F$ the event that the
part fails the check.

1. Write the four given numbers as probabilities, with the symbols $N$,
   $N^c$, and $F$. Name the reference class of $P(F\mid N)$.
2. Calculate $P(F\cap N)$ and $P(F\cap N^c)$ with the multiplication rule.
3. Calculate $P(F)$ with the law of total probability.
4. Draw the tree for 1,000 parts and state the four counts at its end. Use
   the counts to calculate $P(N\mid F)$.
5. Calculate $P(N\mid F)$ again with the formula of Bayes' rule.
6. Decide whether $N$ and $F$ are independent. Show the comparison that you
   use.
7. A colleague reads your result and says: "So 57% of the parts from the new
   printer fail." Explain what is wrong, and write the correct sentence.

::: {.callout-note collapse="true"}
## Check your answer

1. $P(N)=0.40$, $P(N^c)=0.60$, $P(F\mid N)=0.10$, and
   $P(F\mid N^c)=0.05$. The reference class of $P(F\mid N)$ is the parts
   from the new printer.
2. $P(F\cap N)=P(F\mid N)\,P(N)=0.10\times0.40=0.04$ and
   $P(F\cap N^c)=P(F\mid N^c)\,P(N^c)=0.05\times0.60=0.03$.
3. $P(F)=0.04+0.03=0.07$: 7% of all parts fail. This overall rate lies
   between the two rates of 5% and 10%. It is closer to 5%, because the old
   printer makes more parts.
4. Of 1,000 parts, 400 come from the new printer and 600 from the old one.
   New printer: 40 fail and 360 pass. Old printer: 30 fail and 570 pass.
   In total $40+30=70$ parts fail, and 40 of them come from the new
   printer: $P(N\mid F)=40/70\approx0.571$.
5. The formula gives the same value:
   $$
   P(N\mid F)=\frac{P(F\mid N)\,P(N)}
   {P(F\mid N)\,P(N)+P(F\mid N^c)\,P(N^c)}
   =\frac{0.04}{0.04+0.03}=\frac{0.04}{0.07}\approx0.571.
   $$
6. $P(F\mid N)=0.10$, while $P(F)=0.07$. Learning that a part comes from the
   new printer raises the probability that it fails, so $N$ and $F$ are not
   independent. The product form says the same:
   $P(F)\,P(N)=0.07\times0.40=0.028$, which is not equal to
   $P(F\cap N)=0.04$.
7. The colleague reversed the condition. The 57% is $P(N\mid F)$: of
   the failed parts, 57% come from the new printer. The share of the new
   printer's parts that fail is $P(F\mid N)=10\%$. A correct sentence is:
   the new printer makes 40% of all parts and 57% of the failed parts.

Verification in R:

```{r}
#| label: bep02-printers-check
p_new <- 0.40
fail_new <- 0.10
fail_old <- 0.05
printer_tree <- 1000 * c(new_fail = p_new * fail_new,
                         new_pass = p_new * (1 - fail_new),
                         old_fail = (1 - p_new) * fail_old,
                         old_pass = (1 - p_new) * (1 - fail_old))
printer_tree
p_fail <- fail_new * p_new + fail_old * (1 - p_new)
c(P_F_and_N = fail_new * p_new, P_F_and_not_N = fail_old * (1 - p_new),
  P_F = p_fail, P_N_given_F = fail_new * p_new / p_fail,
  P_F_times_P_N = p_fail * p_new)
```
:::

### Task 7: Read a joint PMF with three values per variable

An invented example. A founder sells houseplants online. One order is
selected at random. $X$ is the number of plants in the order (1, 2, or 3)
and $Y$ is the number of decorative pots in the order (0, 1, or 2). The
joint PMF is:

| $p_{X,Y}(x,y)$ | $Y=0$ | $Y=1$ | $Y=2$ |
|---|---:|---:|---:|
| $X=1$ | 0.25 | 0.10 | 0.05 |
| $X=2$ | 0.10 | 0.15 | 0.05 |
| $X=3$ | 0.05 | 0.15 | 0.10 |

1. Check that the table is a valid joint PMF.
2. Find the marginal PMF of $X$ and the marginal PMF of $Y$.
3. Calculate $P(X\le2,\ Y=2)$, the probability that an order has at most
   two plants and two pots. Then calculate $P(X\le2\text{ or }Y=2)$ with the
   addition rule.
4. Find the conditional PMF of $X$ given $Y=0$, and check that it sums
   to 1.
5. Calculate the conditional mean $E[X\mid Y=0]$ and the mean $E[X]$.
   Interpret the difference in one sentence.
6. Check the product form of independence in the cell $(x,y)=(1,0)$. What do
   you conclude, and why is one cell enough here?

::: {.callout-note collapse="true"}
## Check your answer

1. All nine entries are at least 0. The three rows sum to 0.40, 0.30, and
   0.30, so all entries together sum to 1.
2. Sum along each row for $X$: $p_X(1)=0.25+0.10+0.05=0.40$,
   $p_X(2)=0.10+0.15+0.05=0.30$, and $p_X(3)=0.05+0.15+0.10=0.30$. Sum down
   each column for $Y$: $p_Y(0)=0.25+0.10+0.05=0.40$,
   $p_Y(1)=0.10+0.15+0.15=0.40$, and $p_Y(2)=0.05+0.05+0.10=0.20$.
3. The event "at most two plants and two pots" covers the cells $(1,2)$ and
   $(2,2)$, so $P(X\le2,\ Y=2)=0.05+0.05=0.10$. With
   $P(X\le2)=0.40+0.30=0.70$ and $P(Y=2)=0.20$, the addition rule gives
   $P(X\le2\text{ or }Y=2)=0.70+0.20-0.10=0.80$. Check: only the cells
   $(3,0)$ and $(3,1)$ are outside this event, and $1-0.05-0.15=0.80$.
4. Divide the $Y=0$ column by its sum, 0.40:
   $p_{X\mid Y}(1\mid0)=0.25/0.40=0.625$,
   $p_{X\mid Y}(2\mid0)=0.10/0.40=0.25$, and
   $p_{X\mid Y}(3\mid0)=0.05/0.40=0.125$. The three values sum to 1.
5. $E[X\mid Y=0]=1(0.625)+2(0.25)+3(0.125)=0.625+0.50+0.375=1.5$ plants.
   $E[X]=1(0.40)+2(0.30)+3(0.30)=0.40+0.60+0.90=1.9$ plants. Orders without
   a pot contain 1.5 plants on average. That is 0.4 plants fewer than the
   average of all orders, about 21% fewer.
6. The entry is $p_{X,Y}(1,0)=0.25$. The product of its margins is
   $p_X(1)\,p_Y(0)=0.40\times0.40=0.16$. The two numbers differ, so $X$ and
   $Y$ are not independent. Independence needs the product form in every
   cell, so one cell that does not match is enough to show dependence.

Verification in R:

```{r}
#| label: bep02-plants-check
plant_pot <- matrix(c(0.25, 0.10, 0.05,
                      0.10, 0.15, 0.05,
                      0.05, 0.15, 0.10), nrow = 3, byrow = TRUE,
                    dimnames = list(plants = 1:3, pots = 0:2))
rowSums(plant_pot)
colSums(plant_pot)
cond_plants_no_pot <- plant_pot[, "0"] / sum(plant_pot[, "0"])
cond_plants_no_pot
c(total = sum(plant_pot),
  P_X_le2_and_Y2 = sum(plant_pot[c("1", "2"), "2"]),
  P_X_le2_or_Y2 = sum(plant_pot[c("1", "2"), ]) + sum(plant_pot[, "2"]) -
    sum(plant_pot[c("1", "2"), "2"]),
  E_X_given_Y0 = sum(1:3 * cond_plants_no_pot),
  E_X = sum(1:3 * rowSums(plant_pot)),
  cell_1_0 = plant_pot["1", "0"],
  product_1_0 = sum(plant_pot["1", ]) * sum(plant_pot[, "0"]))
```
:::

### Task 8: Covariance and correlation from the same table

Use the joint PMF of Task 7 for the number of plants $X$ and the number of
pots $Y$. Its marginal PMFs are $p_X=0.40,\ 0.30,\ 0.30$ for the values
1, 2, 3 and $p_Y=0.40,\ 0.40,\ 0.20$ for the values 0, 1, 2.

1. Calculate $E[X]$ and $E[Y]$.
2. Calculate $E[XY]$ as a double sum: in every cell multiply $x$, $y$, and
   the probability of the cell, and add the nine products.
3. Calculate $\operatorname{Cov}(X,Y)$ with $E[XY]-E[X]\,E[Y]$ and interpret
   its sign.
4. Calculate $\operatorname{Var}(X)$ and $\operatorname{Var}(Y)$ with the
   computational form.
5. Calculate the correlation of $X$ and $Y$. Interpret its direction and its
   strength.
6. The founder says: "An order with three plants always has a pot in it."
   Use the joint PMF of Task 7 to show that this is wrong. Then say what the
   correlation of part 5 can and cannot tell you about one order.

::: {.callout-note collapse="true"}
## Check your answer

1. $E[X]=1(0.40)+2(0.30)+3(0.30)=1.9$ plants and
   $E[Y]=0(0.40)+1(0.40)+2(0.20)=0.8$ pots.
2. The three cells with $y=0$ give the product 0. The other six cells give
   $1\times1\times0.10=0.10$, $1\times2\times0.05=0.10$,
   $2\times1\times0.15=0.30$, $2\times2\times0.05=0.20$,
   $3\times1\times0.15=0.45$, and $3\times2\times0.10=0.60$. Their sum is
   $E[XY]=0.10+0.10+0.30+0.20+0.45+0.60=1.75$.
3. $\operatorname{Cov}(X,Y)=1.75-1.9\times0.8=1.75-1.52=0.23$. The sign is
   positive: orders with more plants than average tend to have more pots
   than average.
4. $E[X^2]=1(0.40)+4(0.30)+9(0.30)=0.40+1.20+2.70=4.30$, so
   $\operatorname{Var}(X)=4.30-1.9^2=4.30-3.61=0.69$.
   $E[Y^2]=0(0.40)+1(0.40)+4(0.20)=1.20$, so
   $\operatorname{Var}(Y)=1.20-0.8^2=1.20-0.64=0.56$.
5. The standard deviations are $\operatorname{SD}(X)=\sqrt{0.69}\approx0.831$
   and $\operatorname{SD}(Y)=\sqrt{0.56}\approx0.748$. The correlation is
   $$
   \rho_{XY}=\frac{0.23}{0.831\times0.748}=\frac{0.23}{0.6216}\approx0.37.
   $$
   It is a moderate positive linear association, and it has no units.
6. The cell $(3,0)$ has probability 0.05. Among the orders with three plants
   this is $0.05/0.30\approx0.17$, about one order in six, so an order with
   three plants does not always have a pot. A correlation of 0.37 describes
   a tendency across all orders. It cannot say what one order contains.

Verification in R, with `plant_pot` from the check of Task 7:

```{r}
#| label: bep02-plants-cov-check
plants <- 1:3
pots <- 0:2
p_plants <- rowSums(plant_pot)
p_pots <- colSums(plant_pot)
mean_plants <- sum(plants * p_plants)
mean_pots <- sum(pots * p_pots)
var_plants <- sum(plants^2 * p_plants) - mean_plants^2
var_pots <- sum(pots^2 * p_pots) - mean_pots^2
cov_plants_pots <- sum(outer(plants, pots) * plant_pot) - mean_plants * mean_pots
c(E_X = mean_plants, E_Y = mean_pots,
  E_XY = sum(outer(plants, pots) * plant_pot),
  Cov = cov_plants_pots, Var_X = var_plants, Var_Y = var_pots,
  SD_X = sqrt(var_plants), SD_Y = sqrt(var_pots),
  correlation = cov_plants_pots / sqrt(var_plants * var_pots),
  no_pot_given_3_plants = plant_pot["3", "0"] / p_plants[["3"]])
```
:::

### Task 9: Two independent variables (extra practice for home)

An invented example. A founder makes soap and sells it in gift sets through
two shops: a concept store in Tilburg and a museum shop in Groningen. $X$ is
the number of gift sets sold on a Saturday in Tilburg, and $Y$ is the number
sold on the same Saturday in Groningen. The two shops have different
customers, so the founder assumes that $X$ and $Y$ are independent. The
marginal PMFs are:

| Value | 0 | 1 | 2 |
|---|---:|---:|---:|
| $p_X(x)$ | 0.4 | 0.5 | 0.1 |
| $p_Y(y)$ | 0.5 | 0.4 | 0.1 |

1. Build the joint PMF of $X$ and $Y$. Show the calculation for the cell
   $(x,y)=(1,1)$, and check that the nine entries sum to 1.
2. Calculate $E[X]$, $E[Y]$, and $E[XY]$. Calculate $E[XY]$ as a double sum
   over the cells of your table.
3. Calculate $\operatorname{Cov}(X,Y)$. Which statement of Lecture 5 gives
   this result without a calculation?
4. Calculate $E[X+Y]$ and $\operatorname{Var}(X+Y)$. Use
   $\operatorname{Var}(X)=0.41$ and $\operatorname{Var}(Y)=0.44$.
5. A friend suggests another joint table with the same two marginal PMFs. In
   the friend's table, $p_{X,Y}(2,2)=0.03$. Can $X$ and $Y$ be independent
   in the friend's table? Explain.

::: {.callout-note collapse="true"}
## Check your answer

1. For independent variables, every cell is the product of its two margins.
   For the cell $(1,1)$: $p_X(1)\,p_Y(1)=0.5\times0.4=0.20$.

   | $p_{X,Y}(x,y)$ | $Y=0$ | $Y=1$ | $Y=2$ | $p_X(x)$ |
   |---|---:|---:|---:|---:|
   | $X=0$ | 0.20 | 0.16 | 0.04 | 0.4 |
   | $X=1$ | 0.25 | 0.20 | 0.05 | 0.5 |
   | $X=2$ | 0.05 | 0.04 | 0.01 | 0.1 |
   | $p_Y(y)$ | 0.5 | 0.4 | 0.1 | 1 |

   The three rows sum to 0.4, 0.5, and 0.1, so the nine entries sum to 1.
2. $E[X]=0(0.4)+1(0.5)+2(0.1)=0.7$ and $E[Y]=0(0.5)+1(0.4)+2(0.1)=0.6$
   gift sets. In the double sum, every cell with $x=0$ or $y=0$ gives the
   product 0. The other four cells give
   $E[XY]=1\times1\times0.20+1\times2\times0.05+2\times1\times0.04+
   2\times2\times0.01=0.20+0.10+0.08+0.04=0.42$.
3. $\operatorname{Cov}(X,Y)=0.42-0.7\times0.6=0.42-0.42=0$. Lecture 5
   states that independence implies zero covariance, because
   $E[XY]=E[X]\,E[Y]$ for independent variables.
4. $E[X+Y]=0.7+0.6=1.3$ gift sets.
   $\operatorname{Var}(X+Y)=0.41+0.44+2\times0=0.85$. For independent
   variables the variances add.
5. No. Independence needs $p_{X,Y}(2,2)=p_X(2)\,p_Y(2)=0.1\times0.1=0.01$.
   The friend's entry of 0.03 is three times as large. One cell that does
   not match is enough, so $X$ and $Y$ are dependent in the friend's table.
   The two marginal PMFs alone do not determine the joint PMF.

Verification in R:

```{r}
#| label: bep02-soap-check
p_tilburg <- c(0.4, 0.5, 0.1)
p_groningen <- c(0.5, 0.4, 0.1)
soap_joint <- outer(p_tilburg, p_groningen)
dimnames(soap_joint) <- list(X = 0:2, Y = 0:2)
soap_joint
mean_tilburg <- sum(0:2 * p_tilburg)
mean_groningen <- sum(0:2 * p_groningen)
var_tilburg <- sum((0:2)^2 * p_tilburg) - mean_tilburg^2
var_groningen <- sum((0:2)^2 * p_groningen) - mean_groningen^2
round(c(total = sum(soap_joint), E_X = mean_tilburg, E_Y = mean_groningen,
        E_XY = sum(outer(0:2, 0:2) * soap_joint),
        Cov = sum(outer(0:2, 0:2) * soap_joint) - mean_tilburg * mean_groningen,
        Var_X = var_tilburg, Var_Y = var_groningen,
        E_sum = mean_tilburg + mean_groningen,
        Var_sum = var_tilburg + var_groningen), 4)
```
:::

### Task 10: The sum of two variables, with positive and negative covariance (extra practice for home)

An invented example. A founder sells ice cream from a cart in the park. Her
daily revenue from ice cream, $X$, has a mean of 320 euros and a standard
deviation of 50 euros. She wants to add a second product and compares two
options. With each option, the daily revenue from the second product, $Y$,
has a mean of 180 euros and a standard deviation of 30 euros.

- Lemonade sells well on the hot days on which ice cream also sells well:
  $\operatorname{Cov}(X,Y)=600$.
- Hot chocolate sells well on the cool days on which ice cream sells badly:
  $\operatorname{Cov}(X,Y)=-600$.

1. Calculate $\operatorname{Var}(X)$ and $\operatorname{Var}(Y)$.
2. Calculate the expected total revenue $E[X+Y]$ for each option.
3. Calculate the variance and the standard deviation of the total revenue
   with lemonade.
4. Calculate the variance and the standard deviation of the total revenue
   with hot chocolate.
5. Calculate the correlation of $X$ and $Y$ for each option.
6. Which option gives the more stable total revenue? Explain in one or two
   sentences why the two options have the same expected total and a
   different spread.

::: {.callout-note collapse="true"}
## Check your answer

1. $\operatorname{Var}(X)=50^2=2{,}500$ and $\operatorname{Var}(Y)=30^2=900$.
2. $E[X+Y]=320+180=500$ euros for both options. The rule for the expected
   value of a sum does not use the covariance.
3. Lemonade:
   $\operatorname{Var}(X+Y)=2{,}500+900+2\times600=4{,}600$, so
   $\operatorname{SD}(X+Y)=\sqrt{4{,}600}\approx67.82$ euros.
4. Hot chocolate:
   $\operatorname{Var}(X+Y)=2{,}500+900+2\times(-600)=2{,}200$, so
   $\operatorname{SD}(X+Y)=\sqrt{2{,}200}\approx46.90$ euros.
5. Lemonade: $\rho_{XY}=600/(50\times30)=0.4$. Hot chocolate:
   $\rho_{XY}=-600/(50\times30)=-0.4$. Both are moderate linear
   associations, one positive and one negative.
6. Hot chocolate gives the more stable total. Its standard deviation is about
   31% smaller (47 against 68 euros), and the expected total is 500 euros in
   both cases. With a negative covariance, a weak day for one product tends
   to be a good day for the other, so the two revenues partly cancel each
   other out. With a positive covariance, good days and bad days come together,
   and the total varies more.

Verification in R:

```{r}
#| label: bep02-icecream-check
var_icecream <- 50^2
var_second <- 30^2
var_lemonade <- var_icecream + var_second + 2 * 600
var_chocolate <- var_icecream + var_second + 2 * (-600)
c(var_icecream = var_icecream, var_second = var_second,
  mean_total = 320 + 180,
  var_lemonade = var_lemonade, sd_lemonade = sqrt(var_lemonade),
  var_chocolate = var_chocolate, sd_chocolate = sqrt(var_chocolate),
  corr_lemonade = 600 / (50 * 30), corr_chocolate = -600 / (50 * 30),
  sd_reduction = 1 - sqrt(var_chocolate) / sqrt(var_lemonade))
```
:::
:::

::: {.course-phase .phase-after}

## After class

### Individual practice

[Shark Tank data]{.shark-tag} R reports four values from the Shark Tank file. The first two are the shares
of pitches with an on-air deal in Seasons 1 to 8 and in Seasons 9 to 16. The
third is the correlation between `season` and the equity offered
(`equity_offered_pct`). The fourth is the correlation between the equity
offered and the deal indicator (`deal_on_show`).

```{r}
#| label: bep02-timed-output
#| echo: false
c(
  early_deal_rate = mean(D[!L]),
  late_deal_rate = mean(D[L]),
  cor_season_equity = cor(sharks$season, sharks$equity_offered_pct),
  cor_equity_deal = cor(sharks$equity_offered_pct, sharks$deal_on_show)
)
```

Write at most 120 words that interpret all four values. For the two rates,
name the reference class. For the two correlations, give the direction and
the rough strength. Say how the outcome is measured and what you cannot
conclude about cause.

::: {.callout-note collapse="true"}
## Check a model answer

Among the pitches in Seasons 1 to 8, 53.3% reached an on-air deal. Among the
pitches in Seasons 9 to 16, 68.8% did. The difference of 15.5 percentage points
describes these records. It does not show a causal effect of the season.
Season and equity initially offered have a moderate negative linear
association ($r\approx-0.42$). Equity offered and the deal indicator have a
weak negative linear association ($r\approx-0.11$). The entrepreneurs chose
their offers themselves, so pitches with low and high offers may differ in
other ways too. The outcome is a deal on air, not a completed
investment.
:::

### Check your understanding

Complete from memory:

1. The conditioning event gives ...
2. Independence of events means ...
3. Covariance averages ...
4. Correlation standardises ...

::: {.callout-note collapse="true"}
## Check your answers

1. The conditioning event gives the denominator or reference class.
2. Events $A$ and $B$ are independent when
   $P(A\cap B)=P(A)P(B)$.
3. Covariance averages products of deviations from the two means.
4. Correlation standardises covariance by dividing by both standard
   deviations.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Name two events
that both matter for it, for example "customer returns" (comes back for
a second order) and "customer bought on discount". Write two sentences. First, state
which conditional probability you would most want to know, and why the
direction of conditioning matters. Then say whether you would expect the two
events to be independent, and what a dependence would, and would not, tell
you.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus venture that sells boxes of prepared meals. "I most want $P(\text{returns} \mid \text{bought on discount})$, not the reverse, because I choose whether to run discounts, and this direction tells me about the customers a discount attracts. I expect dependence (discount buyers come back less often), but dependence alone would not show that the discount causes it: customers who look for low prices may be different from the start."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant.
::::

### More practice

For extra exercises on the lectures of this tutorial, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Tutorial 2. Show the solution only after I send my answer."

[Continue to Lecture 6: Continuous distributions and the normal model](b06-continuous-normal.qmd).

:::
