# Tutorial 2: Conditioning and relationships {#b-ep02}

```{r}
#| label: bep02-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 2, you can</span>

- calculate and interpret conditional and joint probabilities.
- apply independence of events, total probability, and Bayes' rule
  (Lecture 4).
- recover marginal PMFs and conditional means from a joint PMF, and find the
  mean and the variance of a sum (Lecture 5).
- interpret covariance and correlation without making causal claims.
- read a table of means, standard deviations, and correlations as printed in
  a research paper.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the solution is discussed together.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 4 and 5. Complete these statements
without reopening the chapters:

1. In $P(A\mid B)$, the reference class is ...
2. Variance averages ...
3. Events $A$ and $B$ are independent when ...
4. A joint PMF records ...
5. Covariance averages ...

Use the [formula sheet](formula-sheet.qmd) up to and including Lecture 5.

Review [Mathematics Refresher: fractions, products, and
sums](../resources/math-refresher.html) if needed.

::: {.callout-note collapse="true"}
## Check your preparation

1. The reference class is $B$, the event after the vertical bar.
2. Variance averages squared deviations from a variable's mean.
3. $P(A\cap B)=P(A)P(B)$.
4. A joint PMF records the probability of every possible value pair.
5. Covariance averages products of deviations from the two means.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try each task before you open its answer. In every answer, name the formula
and the reference class, check the value, interpret it in context, and say
what the result does not show.

Tasks 1 to 3 use the Shark Tank file and the two events from Lectures 4 and
5, built in the same way:

```{r}
#| label: bep02-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
D <- sharks$deal_on_show == 1   # on-air deal
L <- sharks$season >= 9         # late season: Seasons 9-16
```

### Task 1: Conditional probability and Bayes

R reports the counts of the 1,441 pitches by period and by deal:

```{r}
#| label: bep02-table
#| echo: false
tab <- table(
  Period = factor(L, levels = c(FALSE, TRUE),
                  labels = c("Seasons 1-8", "Seasons 9-16")),
  Deal = factor(D, levels = c(FALSE, TRUE),
                labels = c("No on-air deal", "On-air deal"))
)
tab
```

1. Calculate and interpret $P(D\mid L)$.
2. Calculate and interpret $P(L\mid D)$.
3. Verify $P(D\cap L)=P(D\mid L)P(L)$.
4. Use the law of total probability with the partition $L^c$ and $L$ to
   recover $P(D)$.
5. Show how Bayes' rule gives $P(L\mid D)$ from $P(D\mid L)$.

::: {.callout-note collapse="true"}
## Check your answer

$P(D\mid L)=505/734\approx 0.688$: among late-season records, 68.8% reached
an on-air deal. $P(L\mid D)=505/882\approx 0.573$: among deal records, 57.3%
came from late seasons. The joint probability is
$P(D\mid L)P(L)=(505/734)(734/1441)=505/1441\approx 0.350$.

The law of total probability gives

$$
P(D)=\frac{377}{707}\frac{707}{1441}
     +\frac{505}{734}\frac{734}{1441}
     =\frac{882}{1441}\approx 0.612.
$$

Bayes' rule gives

$$
P(L\mid D)=\frac{P(D\mid L)P(L)}{P(D)}
=\frac{(505/734)(734/1441)}{882/1441}
=\frac{505}{882}\approx 0.573.
$$
:::

### Task 2: Independence of events

Lecture 4 defined independence of events in two equivalent ways: through
conditioning and through the product form. R reports:

```{r}
#| label: bep02-independence
#| echo: false
c(
  P_D = mean(D),
  P_D_given_L = mean(D[L]),
  P_D_and_L = mean(D & L),
  P_D_times_P_L = mean(D) * mean(L)
)
```

1. Use the definition with conditioning to decide whether $D$ and $L$ are
   independent in these records.
2. Use the product form to reach the same conclusion.
3. Explain why "not independent" does not mean "late season causes a deal."
4. Explain why the early and late events are disjoint but not independent.

::: {.callout-note collapse="true"}
## Check your answer

The deal probability is about 0.612 overall and 0.688 within late seasons, so
conditioning on $L$ changes it. In the same way, the observed joint
probability is about 0.350, while the product $P(D)P(L)$ is about 0.312. So
$D$ and $L$ are not independent in these records. This is an association in
televised pitches that the producers selected. It does not show that the
season causes deals, because other things also changed over the seasons.
Early and late are disjoint because they cannot occur together. Observing
one makes the other impossible, so disjoint events with positive probability
are dependent.
:::

### Task 3: Joint PMF, covariance, and correlation

Let $X$ be the 0/1 deal indicator and $Y$ the 0/1 late-period indicator,
built from the same events:

```{r}
#| label: bep02-indicators
x <- as.integer(D)
y <- as.integer(L)
```

R reports:

```{r}
#| label: bep02-joint
#| echo: false
joint <- prop.table(table(X = x, Y = y))
joint

c(
  E_X = mean(x),
  E_Y = mean(y),
  Var_X = mean((x - mean(x))^2),
  Var_Y = mean((y - mean(y))^2),
  E_XY = mean(x * y),
  covariance = mean(x * y) - mean(x) * mean(y),
  correlation = cor(x, y)
)
```

1. Verify that the joint PMF sums to 1.
2. Recover the marginal PMF of $X$ by summing rows.
3. Verify $\operatorname{Var}(X)=E[X^2]-E[X]^2$.
4. Verify $\operatorname{Cov}(X,Y)=E[XY]-E[X]E[Y]$.
5. Interpret the sign of the covariance.
6. Explain what the correlation adds to the covariance, and what it does not
   prove.
7. Calculate the conditional means $E[X\mid Y=1]$ and $E[X\mid Y=0]$ from the
   joint PMF, and say what they mean.
8. The sum $X+Y$ counts how many of the two events, deal and late season,
   happened for a pitch. Calculate $E[X+Y]$ and
   $\operatorname{Var}(X+Y)$ from the values in the output.

::: {.callout-note collapse="true"}
## Check your answer

1. The four joint probabilities sum to 1:
   $0.229+0.159+0.262+0.350=1$.
2. Summing each row across the two $Y$ columns gives
   $P(X=0)=0.229+0.159=0.388$ and $P(X=1)=0.262+0.350=0.612$. As counts,
   these are $559/1441$ and $882/1441$.
3. Because $X$ takes only the values 0 and 1, $X^2=X$ and $E[X^2]=E[X]$. So
   $\operatorname{Var}(X)=E[X]-E[X]^2=0.612-0.612^2\approx 0.237$, as in the
   output.
4. $E[XY]=P(X=1,Y=1)\approx 0.3505$, and
   $E[X]E[Y]\approx 0.6121\times0.5094\approx 0.3118$. The difference is
   about 0.0387, as in the output.
5. The covariance is positive. The value 1 occurs for both variables
   together a little more often than it would under independence: deals and
   late seasons go together.
6. The correlation divides the covariance by both standard deviations. The
   result, about 0.16, has no units and lies between $-1$ and $1$: a weak
   positive linear association. It does not show a cause, and it does not
   say how important the association is in practice. For two 0/1 variables
   a correlation of zero would mean independence. For variables with more
   than two values it would not, as the U-shaped pattern in Lecture 5 shows.
7. Divide each column of the joint PMF by its sum. The column sums are
   $P(Y=1)\approx 0.5094$ and $P(Y=0)\approx 0.4906$. Because $X$ is 0 or 1,
   only the $X=1$ entry counts:
   $E[X\mid Y=1]\approx 0.3505/0.5094\approx 0.688$ and
   $E[X\mid Y=0]\approx 0.2616/0.4906\approx 0.533$. Among late-season
   pitches the average of the deal indicator is 0.688, so 68.8% reached a
   deal. Among early-season pitches it is 53.3%.
8. $E[X+Y]=E[X]+E[Y]=0.612+0.509=1.121$.
   $\operatorname{Var}(X+Y)=\operatorname{Var}(X)+\operatorname{Var}(Y)
   +2\operatorname{Cov}(X,Y)=0.237+0.250+2\times0.039\approx 0.565$.
   Without the covariance term the result would be 0.487. The sum varies
   more because deals and late seasons go together.
:::

### Task 4: Read a correlation table from a paper

Research papers in management usually start their results with one table of
means, standard deviations (SD), and correlations. The table below is an
invented example. It was constructed for this course in the style of such a
table and does not come from a real study. It describes a survey of 240
employees with four variables:

- creative self-efficacy: how much an employee believes in his or her own
  ability to be creative, rated from 1 to 7.
- job tenure: the number of years in the current job.
- supervisor support: how much support the employee gets from the
  supervisor, rated from 1 to 7.
- creative performance: how creative the employee's work is, rated from 1
  to 7.

The columns "Mean" and "SD" describe each variable alone. The columns 1, 2,
and 3 contain correlations. Each number is the correlation between the
variable in that row and the variable with that number. For example, the
0.31 in row 3 and column 1 is the correlation between supervisor support and
creative self-efficacy.

| Variable | Mean | SD | 1 | 2 | 3 |
|---|---:|---:|---:|---:|---:|
| 1. Creative self-efficacy (1--7) | 5.10 | 0.90 | | | |
| 2. Job tenure (years) | 6.40 | 4.20 | 0.08 | | |
| 3. Supervisor support (1--7) | 4.80 | 1.10 | 0.31 | 0.05 | |
| 4. Creative performance (1--7) | 4.20 | 1.00 | 0.36 | 0.12 | 0.28 |

1. Read off the correlation between creative self-efficacy and creative
   performance, and interpret its sign and strength in one sentence.
2. Which pair of variables has the weakest linear association?
3. A manager concludes: "Raising supervisor support will raise creative
   performance." Name one third variable, or one other reason, that could
   produce the correlation of 0.28 without support driving performance.
4. State one thing this table cannot tell you about an individual employee.

::: {.callout-note collapse="true"}
## Check your answer

1. The correlation is 0.36, in row 4 and column 1: a moderate positive
   linear association. Employees with higher creative self-efficacy tended
   to be rated higher on creative performance in this sample.
2. The weakest association is between job tenure and supervisor support,
   with a correlation of 0.05.
3. Several other explanations fit the same number. Employees in creative
   roles may receive more support and also perform more creatively. Or
   supervisors may give more support to employees who already perform well.
   A correlation table cannot separate these explanations.
4. The table describes how the variables move together across 240
   employees. It says nothing certain about any one of them. Many employees
   with high self-efficacy have average performance ratings.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice: 10 minutes

R reports four values from the Shark Tank file. The first two are the shares
of pitches with an on-air deal in Seasons 1 to 8 and in Seasons 9 to 16. The
third is the correlation between `season` and the equity offered
(`equity_offered_pct`). The fourth is the correlation between the equity
offered and the deal indicator (`deal_on_show`).

```{r}
#| label: bep02-timed-output
#| echo: false
c(
  early_deal_rate = mean(D[!L]),
  late_deal_rate = mean(D[L]),
  cor_season_equity = cor(sharks$season, sharks$equity_offered_pct),
  cor_equity_deal = cor(sharks$equity_offered_pct, sharks$deal_on_show)
)
```

Write at most 120 words that interpret all four values. For the two rates,
name the reference class. For the two correlations, give the direction and
the rough strength. Say how the outcome is measured and what cannot be
concluded about cause.

::: {.callout-note collapse="true"}
## Check a model answer

Among the pitches in Seasons 1--8, 53.3% reached an on-air deal. Among the
pitches in Seasons 9--16, 68.8% did. The difference of 15.5 percentage points
describes these records. It does not show a causal effect of the season.
Season and equity initially offered have a moderate negative linear
association ($r\approx-0.42$). Equity offered and the deal indicator have a
weak negative linear association ($r\approx-0.11$). The entrepreneurs chose
their offers themselves, so pitches with low and high offers may differ in
other ways too. The outcome is an agreement on air, not a completed
investment.
:::

### Check your understanding

Complete from memory:

1. The conditioning event supplies ...
2. Independence of events means ...
3. Covariance averages ...
4. Correlation standardises ...

::: {.callout-note collapse="true"}
## Check your answers

1. The conditioning event supplies the denominator or reference class.
2. Events $A$ and $B$ are independent when
   $P(A\cap B)=P(A)P(B)$.
3. Covariance averages products of deviations from the two means.
4. Correlation standardises covariance by dividing by both standard
   deviations.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Name two events
that both matter for it (for example "customer returns" and "customer bought
on discount"), and write two sentences: first state which conditional
probability you would most want to know and why the direction of conditioning
matters, then say whether you would expect the two events to be independent
and what a dependence would, and would not, tell you.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "I most want $P(\text{returns} \mid \text{bought on discount})$, not the reverse, because I choose whether to run discounts, and this direction tells me about the customers a discount attracts. I expect dependence (discount buyers return less), but dependence alone would not show the discount causes it: bargain-driven customers may differ to begin with."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant.
::::

Next is [Lecture 6](b06-continuous-normal.qmd). It moves from discrete to
continuous random variables and introduces the normal distribution.

:::
