# Tutorial 3: Normal models, sampling, and confidence intervals {#b-ep03}

```{r}
#| label: bep03-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 3, you can</span>

- calculate probabilities and values from uniform and normal models
- calculate standard errors and probabilities for sample means and proportions
- construct confidence intervals and qualify conclusions when the sampling design is limited
:::

Tutorials are **exam preparation sessions**. The tasks below are exam-style
questions. You first try each task in pairs. You may use AI for this, with
the four steps Attempt, Ask, Verify, Explain. Then the class discusses the
solution together.

This tutorial covers Lectures 6--9: densities and the normal model, samples
and estimators, the sampling distribution of a mean and of a proportion, and
confidence intervals. Hypothesis tests follow in Lecture 10 and in
Tutorial 4.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 6--9. If you find the algebra
difficult, review [Mathematics Refresher: sums, fractions, and
roots](../resources/math-refresher.html).

Use the formula sheet with all formulas up to and including Lecture 9.

Complete from memory:

1. Standardising an observation $x$ from $N(\mu,\sigma^2)$ means
   calculating ...
2. A parameter describes ...
3. An estimator is ...
4. The standard deviation of the sample mean $\bar X$ is ... The standard
   error of the sample mean is ...
5. The CLT describes ...
6. A written answer for a confidence interval has five parts, in this
   order: ...
7. The confidence interval for a mean with $s$ has the formula ... It may
   be used when ...
8. The confidence interval for a proportion has the formula ... It may be
   used when ...

::: {.callout-note collapse="true"}
## Check your preparation

1. $z=(x-\mu)/\sigma$: the number of population SDs between $x$ and the
   mean.
2. A parameter describes a population.
3. An estimator is a rule that estimates a parameter from a sample. Its
   value is random, because the sample is random.
4. The standard deviation of the sample mean is
   $\operatorname{SD}(\bar X)=\sigma/\sqrt n$. The standard error is
   $\operatorname{SE}(\bar X)=S/\sqrt n$, with the value $s/\sqrt n$. It is
   the estimated standard deviation of $\bar X$, used when $\sigma$ is
   unknown.
5. The distribution of the sample mean across random samples. It is
   approximately normal for a large random sample, also when the population
   is not normal.
6. Formula, with the reason why it may be used. Given values. Calculations:
   the standard error, then the margin of error. Interval, with the unit.
   Conclusion in the words of the question.
7. $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. The population is normal, or
   $n$ is large.
8. $\widehat P\pm z_{1-\alpha/2}\sqrt{\widehat P(1-\widehat P)/n}$. The
   condition is $n\hat p\ge10$ and $n(1-\hat p)\ge10$.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Attempt each task before you open its answer. A complete written answer
names the quantity and shows the calculation. It reports the value with its
unit and interprets this value in context. It also states one important
limit of the model or the data.

Tasks 1 to 9 belong to the tutorial. Tasks 10 to 12 are extra practice for
home.

In practice $\sigma$ is unknown, so an interval for a mean uses $s$ and a t
value. In this book, the values you need are given in the task or on the
formula sheet.

### Task 1: A normal model in both directions

An invented example. A founder runs an escape room. The time that a team
needs to solve the room, $Y$, is modelled as $Y\sim N(45,5^2)$, in minutes.
The standard deviation is 5 minutes.

Supplied values: $P(Z\le0.8)=0.7881$, $P(Z\le1.2)=0.8849$, and the 0.95
quantile $z_{0.95}=1.645$.

1. Calculate the probability that a team needs at most 49 minutes.
2. Calculate the probability that a team needs between 39 and 49 minutes.
3. The founder wants a time limit that 5% of the teams exceed. Calculate
   this limit.
4. Say in one sentence what the supplied quantile 1.645 means for a
   standard normal variable $Z$.

::: {.callout-note collapse="true"}
## Check your answer

1. Standardise: $z=(49-45)/5=0.8$. So
   $P(Y\le49)=P(Z\le0.8)=0.7881$. About 79% of the teams need at most 49
   minutes.
2. The two z values are $(39-45)/5=-1.2$ and $0.8$. The supplied values
   contain only positive z values, so use the symmetry rule and then the
   complement rule: $P(Z\le-1.2)=P(Z>1.2)=1-0.8849=0.1151$. Then

   $$
   P(39\le Y\le49)=P(Z\le0.8)-P(Z\le-1.2)=0.7881-0.1151=0.6730.
   $$

   About 67% of the teams need between 39 and 49 minutes.
3. If 5% of the teams exceed the limit, 95% are at or below it. The limit
   is the 0.95 quantile of $Y$:
   $y=\mu+z\sigma=45+1.645\times5=53.23$ minutes. A limit of about 53
   minutes fits.
4. 95% of the values of $Z$ are at or below 1.645, and 5% are above it.

Parts 1 and 2 go from a value to a probability. Part 3 goes from a
probability to a value. One limit: the normal curve is a model. The answers
are only as good as the fit of this model to the recorded times.

Verification in R:

```{r}
#| label: bep03-normal-check
c(p_at_most_49 = pnorm(49, mean = 45, sd = 5),
  p_39_to_49 = pnorm(49, mean = 45, sd = 5) - pnorm(39, mean = 45, sd = 5),
  by_hand = 0.7881 - (1 - 0.8849),
  limit = qnorm(0.95, mean = 45, sd = 5),
  limit_by_hand = 45 + 1.645 * 5)
```
:::

### Task 2: Population, frame, and claim

A student randomly selects 100 rows from the Shark Tank file and estimates
the mean equity offered (`equity_offered_pct`).

1. State the population about which this sample allows direct
   conclusions.
2. Explain why this sample does not represent all *Shark Tank* applicants.
3. Explain why this sample does not represent all US entrepreneurs.
4. State one change needed to target all US entrepreneurs.

::: {.callout-note collapse="true"}
## Check your answer

The direct finite population is the 1,441 recorded pitches in completed
Seasons 1--16. Applicants who did not appear on television are not in the
file, so the frame does not represent all applicants. The show also selects a specialised
group of entrepreneurs, not a probability sample of US entrepreneurs.
To target all US entrepreneurs, you would need a suitable national
sampling frame and a random selection from that frame.
:::

### Task 3: Reproduce the realised sample

[Shark Tank data]{.shark-tag} The code below draws the sample with seed 123, exactly as in Lecture 7.
So you get the same 100 rows when you run it:

```{r}
#| label: bep03-sample-output
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

c(
  n = nrow(pitch_sample),
  mean_equity = mean(pitch_sample$equity_offered_pct),
  sd_equity = sd(pitch_sample$equity_offered_pct),
  q1_equity = unname(quantile(pitch_sample$equity_offered_pct, 0.25)),
  median_equity = median(pitch_sample$equity_offered_pct),
  q3_equity = unname(quantile(pitch_sample$equity_offered_pct, 0.75)),
  deal_proportion = mean(pitch_sample$deal_on_show)
)
```

A note on the quartiles in this output. When a quartile falls between two
ordered values, R's `quantile()` reports a point between them. Here the
75th ordered value is 17 and the 76th is 20, and R reports 17.75. The rule
of Lecture 1 takes the 75th of the 100 ordered values, which is 17.

1. Identify the parameter, estimator, and estimate for the mean equity
   offered (`equity_offered_pct`).
2. Identify the parameter, estimator, and estimate for the deal proportion
   (`deal_on_show`).
3. Calculate the interquartile range (IQR) of the equity offered in the
   sample from the R output and interpret it. The IQR is defined in
   [Lecture 1](b01-introduction-data-analysis.qmd#b01-quartiles).
4. Explain why a different seed can produce different estimates and a
   different IQR.

::: {.callout-note collapse="true"}
## Check your answer

For the equity offered, the parameter is the population mean $\mu$, the
estimator is $\bar X$, and the estimate is $\bar x=12.76$%. For the deal
outcome, the parameter is the population proportion $p$, the estimator is
$\widehat P$, and the estimate is $\hat p=0.64$. The sample quartiles in the
R output are $Q_1=8$ and $Q_3=17.75$, so
$\operatorname{IQR}=17.75-8=9.75$ percentage points: the middle half of the
initial equity offers in the sample lies between 8% and 17.75%. With the rule of
Lecture 1, $Q_3=17$ and the IQR is $17-8=9$ percentage points. Both answers
are correct when you name the rule that you used. A different seed selects
a different random set of rows. So all sample statistics, including the
IQR, can change.

The ordered values behind the two rules:

```{r}
#| label: bep03-quartile-rules
sort(pitch_sample$equity_offered_pct)[c(25, 26, 75, 76)]
```
:::

### Task 4: Standard errors and a larger sample

[Shark Tank data]{.shark-tag} Use the sample output of Task 3.

1. Calculate the standard error of the sample mean of the equity offered,
   $\operatorname{SE}(\bar X)$.
2. Calculate the standard error of the sample proportion of deals,
   $\operatorname{SE}(\widehat P)$ (Lecture 8).
3. Check the success-failure condition for $\widehat P$.
4. Predict how each standard error changes if $n$ increases from 100 to 400
   and the spread in the sample stays similar.
5. In part 1, R calculates the sample SD with `sd()`. This function divides
   by $n-1$. State the value of $n-1$ here, and explain in one sentence why
   `sd()` divides by it.
6. Explain why the results of parts 1 and 2 are standard errors and not the
   standard deviations of $\bar X$ and $\widehat P$.

::: {.callout-note collapse="true"}
## Check your answer

$$
\operatorname{SE}(\bar X)=\frac{s}{\sqrt n}=\frac{7.7079}{\sqrt{100}}=0.7708.
$$

$$
\operatorname{SE}(\widehat P)
=\sqrt{\frac{\hat p(1-\hat p)}{n}}
=\sqrt{\frac{0.64(0.36)}{100}}=0.0480.
$$

The standard error of the mean is about 0.77 percentage points of equity.
The standard error of the proportion is about 4.8 percentage points of the
deal rate. The sample contains 64 successes (deals) and 36 failures (no
deal). Both counts are at least 10, so the success-failure condition holds.
If $n$ becomes four times as large, both standard errors are divided by
$\sqrt 4=2$. The divisor is $n-1=99$. The deviations are measured from the
sample mean, so they must sum to zero. As a result, only 99 of them can
vary freely. Dividing by 99 removes the downward bias that this causes
(Lecture 7).

The standard deviations of the two estimators are $\sigma/\sqrt n$ and
$\sqrt{p(1-p)/n}$. They contain the population values $\sigma$ and $p$,
which a sample does not give. The standard errors use $s$ and $\hat p$ from
the sample in their place. A standard error is an estimated standard
deviation of an estimator.

Verification in R:

```{r}
#| label: bep03-standard-errors
s_equity <- sd(pitch_sample$equity_offered_pct)
p_hat <- mean(pitch_sample$deal_on_show)
c(se_mean = s_equity / sqrt(100),
  se_proportion = sqrt(p_hat * (1 - p_hat) / 100),
  deals = sum(pitch_sample$deal_on_show),
  no_deals = 100 - sum(pitch_sample$deal_on_show),
  se_mean_n400 = s_equity / sqrt(400),
  se_proportion_n400 = sqrt(p_hat * (1 - p_hat) / 400))
```
:::

### Task 5: One pitch versus one mean

[Shark Tank data]{.shark-tag} Suppose the population mean of the equity offered is $\mu=13.21$%, the
population SD is $\sigma=8.44$ percentage points. You draw independent
random samples of $n=100$ pitches. Supplied values: $P(Z\le 2.12)=0.983$ and
$P(Z\le 0.21)=0.583$.

1. State the approximate sampling distribution of $\bar X$, with the
   standard deviation of the sample mean.
2. Calculate $P(\bar X>15)$.
3. A classmate models a single pitch as $X\sim N(13.21, 8.44^2)$ and
   calculates $P(X>15)$. Compute that value too, and explain why the two
   answers are so different.
4. The equity offered is strongly right-skewed (Lecture 1). So which of the
   two calculations uses a bad model? And why does the CLT still justify
   the calculation in part 2?

::: {.callout-note collapse="true"}
## Check your answer

The CLT gives approximately
$\bar X\mathrel{\dot\sim}N(13.21,\,8.44^2/100)$. The standard deviation of
the sample mean is $\operatorname{SD}(\bar X)=8.44/\sqrt{100}=0.844$
percentage points. Therefore

$$
P(\bar X>15)
=1-P\!\left(Z\le\frac{15-13.21}{0.844}\right)
=1-P(Z\le2.12)
\approx0.017.
$$

For a single pitch, $z=(15-13.21)/8.44\approx0.21$, so
$P(X>15)\approx1-0.583\approx0.42$. The mean of 100 pitches varies ten
times less than one pitch ($\sigma/\sqrt{100}$), so 15 is an ordinary value
for one pitch but an extreme value for a mean of 100. A normal curve is
symmetric and the equity offered is strongly right-skewed. So the normal
model for a single pitch in part 3 is a bad model. The CLT concerns the
sample mean. The mean of 100 independent draws has a nearly normal
sampling distribution, even when the population is skewed. This is why
the calculation in part 2 is justified.

Verification in R:

```{r}
#| label: bep03-pitch-mean-check
sd_mean <- 8.44 / sqrt(100)
c(sd_of_mean = sd_mean,
  z_mean = (15 - 13.21) / sd_mean,
  p_mean_above_15 = 1 - pnorm(15, mean = 13.21, sd = sd_mean),
  z_single = (15 - 13.21) / 8.44,
  p_single_above_15 = 1 - pnorm(15, mean = 13.21, sd = 8.44))
```
:::

### Task 6: The exact normal case and the CLT

An invented example. A founder makes scented candles. The burn time of one
candle, $X$, is modelled as $X\sim N(40,6^2)$, in hours. A tester burns a
random sample of $n=9$ candles and records the sample mean $\bar X$.
Supplied values for the whole task:

```text
P(Z <= 0.6) = 0.7257
P(Z <= 1.8) = 0.9641
P(Z <= 2.5) = 0.9938
```

1. State the distribution of $\bar X$ for $n=9$, with its mean and its
   standard deviation. Explain why you may use this distribution although
   the sample is small.
2. Calculate the probability that one candle burns less than 36.4 hours.
3. Calculate the probability that the mean burn time of the 9 candles is
   below 36.4 hours.
4. Explain in one sentence why the answers to parts 2 and 3 differ.

The founder also looks at the orders in her web shop. The value of one
order is strongly right-skewed: most orders are small and a few are very
large. The mean is $\mu=45$ euros and the standard deviation is
$\sigma=30$ euros.

5. For a random sample of $n=100$ orders, calculate the probability that
   the sample mean is above 52.50 euros. Name the result (the theorem)
   that allows this calculation.
6. Explain why you cannot do the same calculation for a random sample of 9
   orders.

::: {.callout-note collapse="true"}
## Check your answer

1. The population is normal. Then the sample mean is exactly normal for
   every sample size, also for a small one (Lecture 8):
   $\bar X\sim N(40,\,6^2/9)$. Its mean is 40 hours and its standard
   deviation is $\operatorname{SD}(\bar X)=6/\sqrt9=2$ hours.
2. One candle: $z=(36.4-40)/6=-0.6$. With the symmetry rule and the
   complement rule, $P(X<36.4)=P(Z\le-0.6)=1-0.7257=0.2743$.
3. The mean of 9 candles: $z=(36.4-40)/2=-1.8$. So
   $P(\bar X<36.4)=P(Z\le-1.8)=1-0.9641=0.0359$.
4. The mean of 9 candles varies three times less than one candle, so 36.4
   hours is an ordinary value for one candle (about 27%) and an unusual
   value for the mean of 9 candles (about 4%).
5. The population is not normal, but $n=100$ is large. The CLT gives
   approximately $\bar X\mathrel{\dot\sim}N(45,\,30^2/100)$, with
   $\operatorname{SD}(\bar X)=30/\sqrt{100}=3$ euros. Then
   $z=(52.50-45)/3=2.5$ and $P(\bar X>52.50)\approx1-0.9938=0.0062$.
6. The order values are not normal, so the exact normal case does not hold.
   A sample of 9 is too small for the CLT. No result of Lecture 8 gives the
   distribution of $\bar X$ in this case.

Verification in R:

```{r}
#| label: bep03-candles-check
c(sd_mean_9 = 6 / sqrt(9),
  p_one_below_36.4 = pnorm(36.4, mean = 40, sd = 6),
  p_mean_below_36.4 = pnorm(36.4, mean = 40, sd = 6 / sqrt(9)),
  sd_mean_100 = 30 / sqrt(100),
  p_mean_above_52.50 = 1 - pnorm(52.50, mean = 45, sd = 30 / sqrt(100)))
```
:::

### Task 7: A proportion through the CLT

[Shark Tank data]{.shark-tag} The deal proportion (`deal_on_show`) in the Shark Tank file is $p=0.612$. A
random sample of $n=100$ pitches is planned, with sample deal proportion
$\widehat P$. Supplied value: $P(Z\le -2.30)=0.011$.

1. Calculate the standard deviation of the sample proportion,
   $\operatorname{SD}(\widehat P)$, and check the success-failure
   condition.
2. State the approximate sampling distribution of $\widehat P$.
3. Approximate the probability that the sample shows a deal rate of at
   most 50%, that is, $P(\widehat P\le0.50)$.
4. Interpret the result in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

Here $p$ is given, so the result is a standard deviation and not a standard
error: $\operatorname{SD}(\widehat P)=\sqrt{0.612(0.388)/100}\approx0.0487$.
The expected counts are $np=61.2$ and $n(1-p)=38.8$, both at least 10.
So $\widehat P\mathrel{\dot\sim}N(0.612,\,0.0487^2)$. Standardise:

$$
z=\frac{0.50-0.612}{0.0487}\approx-2.30,
\qquad
P(\widehat P\le0.50)\approx P(Z\le-2.30)\approx0.011.
$$

If the deal proportion in the file is 61.2%, only about 1% of random
samples of 100 pitches would show a deal rate at or below 50%. Such a sample would be
surprising. Lecture 10 builds the hypothesis test on this idea.

Verification in R:

```{r}
#| label: bep03-proportion-clt-check
sd_phat <- sqrt(0.612 * 0.388 / 100)
c(sd_of_phat = sd_phat,
  expected_deals = 100 * 0.612,
  expected_no_deals = 100 * 0.388,
  z = (0.50 - 0.612) / sd_phat,
  p_at_most_050 = pnorm(0.50, mean = 0.612, sd = sd_phat))
```
:::

### Task 8: A confidence interval for a mean

An invented example. A founder runs a bouldering gym. Her business plan
assumes that a visit lasts 120 minutes on average. She draws a random sample
of 16 visits from the 5,000 visits of last year in the check-in system. The
sample has a mean of 136 minutes and a standard deviation of $s=28$
minutes. The visit times are roughly normally distributed. Supplied values:
$t_{0.95,\,15}=1.753$, $t_{0.975,\,15}=2.131$, and $t_{0.995,\,15}=2.947$.
The formula sheet does not list $df=15$, so the task gives these values
here.

1. Construct a 95% confidence interval for the mean visit time of last
   year. Write the five parts of the exam procedure for a mean (Lecture 9).
   In part 5, say what the interval tells you about the 120 minutes in the
   business plan.
2. Construct the 99% confidence interval. Only parts 2 to 5 change.
3. Construct the 90% confidence interval. Again only parts 2 to 5 change.
4. Compare the three conclusions about the 120 minutes, and explain how the
   width of the interval changes with the confidence level.

::: {.callout-note collapse="true"}
## Check your answer

The 95% interval in five parts:

1. **Formula.** $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used
   because the visit times are roughly normal. The sample is small, so a
   large $n$ is not the reason here.
2. **Given values.** $\bar x=136$ minutes, $s=28$ minutes, $n=16$,
   $df=15$. The confidence level is 95%, so $\alpha=0.05$ and the critical
   value is $t_{0.975,\,15}=2.131$.
3. **Calculations.** The standard error is $s/\sqrt n=28/\sqrt{16}=7$
   minutes. The margin of error is $2.131\times7=14.92$ minutes.
4. **Interval.** $(136-14.92,\ 136+14.92)=(121.08,\ 150.92)$ minutes.
5. **Conclusion.** The value 120 minutes lies outside the interval. At the
   5% level we can conclude that the mean visit time of last year differs
   from 120 minutes. The whole interval lies above 120, so visits last
   longer than the plan assumes. The lower end is 121.08 minutes, only
   about 1% above 120. So the difference may be small.

The 99% interval, parts 2 to 5:

2. **Given values.** As before, but the confidence level is 99%, so
   $\alpha=0.01$ and the critical value is $t_{0.995,\,15}=2.947$.
3. **Calculations.** The standard error stays 7 minutes. The margin of
   error is $2.947\times7=20.63$ minutes.
4. **Interval.** $(136-20.63,\ 136+20.63)=(115.37,\ 156.63)$ minutes.
5. **Conclusion.** The value 120 minutes lies inside the interval. At the
   1% level we cannot conclude that the mean visit time differs from 120
   minutes.

The 90% interval, parts 2 to 5:

2. **Given values.** As before, but the confidence level is 90%, so
   $\alpha=0.10$ and the critical value is $t_{0.95,\,15}=1.753$.
3. **Calculations.** The standard error stays 7 minutes. The margin of
   error is $1.753\times7=12.27$ minutes.
4. **Interval.** $(136-12.27,\ 136+12.27)=(123.73,\ 148.27)$ minutes.
5. **Conclusion.** The value 120 minutes lies outside the interval. At the
   10% level we can conclude that the mean visit time differs from 120
   minutes, and the whole interval lies above 120.

A higher confidence level needs a larger critical value, so the margin of
error grows from 12.27 minutes at 90% to 14.92 minutes at 95% and 20.63
minutes at 99%. The 99% interval is about 1.7 times as wide as the 90%
interval. With the same sample, we can conclude that the mean differs from
120 minutes at the 10% and the 5% level, but not at the stricter 1% level.
For this reason every conclusion names its level. One limit: the interval describes
the visits of last year, not the visits of next year.

Verification in R, with the supplied t values:

```{r}
#| label: bep03-interval-mean-check
se_visit <- 28 / sqrt(16)
c(se = se_visit,
  margin_95 = 2.131 * se_visit,
  lower_95 = 136 - 2.131 * se_visit, upper_95 = 136 + 2.131 * se_visit,
  margin_99 = 2.947 * se_visit,
  lower_99 = 136 - 2.947 * se_visit, upper_99 = 136 + 2.947 * se_visit,
  margin_90 = 1.753 * se_visit,
  lower_90 = 136 - 1.753 * se_visit, upper_90 = 136 + 1.753 * se_visit,
  width_99_over_90 = 2.947 / 1.753)
c(t_95 = qt(0.95, df = 15), t_975 = qt(0.975, df = 15),
  t_995 = qt(0.995, df = 15),
  low_end_above_120_in_percent = 100 * (121.08 - 120) / 120)
```
:::

### Task 9: A confidence interval for a proportion

An invented example. A founder sells sneakers online. Her parcel service
reports that customers return 15% of all sneaker orders. Last year her shop
had 6,000 orders. She draws a random sample of 400 of these orders.
Customers returned 40 of these 400 orders. Let
$p$ be the proportion of returned orders among all 6,000 orders.

1. Construct a 95% confidence interval for $p$. Write the five parts of the
   exam procedure for a proportion (Lecture 9), with the check of the
   condition. Check the 10% condition first. In part 5, say what the
   interval tells you about the 15% reported by the parcel service.
2. The founder says: "My return rate is 10%, so I am 5 percentage points
   below the 15%." Assess this sentence.
3. State the formula of the standard deviation of $\widehat P$ and explain
   why you cannot calculate it here.

::: {.callout-note collapse="true"}
## Check your answer

The sample contains 400 of the 6,000 orders, that is 6.7%. This is at most
10%, so we may treat the sample as iid.

The interval of part 1 in five parts:

1. **Formula.**
   $\widehat P\pm z_{1-\alpha/2}\sqrt{\widehat P(1-\widehat P)/n}$. The
   condition holds: $n\hat p=40$ and $n(1-\hat p)=360$ are both at least
   10.
2. **Given values.** 40 returned orders, $n=400$, $\hat p=40/400=0.10$. The
   confidence level is 95%, so $z_{0.975}=1.960$.
3. **Calculations.** The standard error is
   $\sqrt{0.10\times0.90/400}=\sqrt{0.000225}=0.015$. The margin of error
   is $1.960\times0.015=0.0294$.
4. **Interval.** $(0.10-0.0294,\ 0.10+0.0294)=(0.071,\ 0.129)$, that is,
   from 7.1% to 12.9% of the orders.
5. **Conclusion.** The value 0.15 lies outside the interval. At the 5%
   level we can conclude that her proportion of returned orders differs
   from 15%. The whole interval lies below 0.15.

Part 2. The sentence treats the estimate as the exact rate. The 10% is the
estimate from 400 orders. The plausible values run from 7.1% to 12.9%. So
her return rate is between 2.1 and 7.9 percentage points below the 15%.

Part 3. The standard deviation of the sample proportion is
$\operatorname{SD}(\widehat P)=\sqrt{p(1-p)/n}$. It contains the unknown
$p$, which is the number that the interval tries to estimate. The standard error
uses $\hat p$ in its place.

Verification in R:

```{r}
#| label: bep03-interval-proportion-check
p_hat_returns <- 40 / 400
se_returns <- sqrt(p_hat_returns * (1 - p_hat_returns) / 400)
c(p_hat = p_hat_returns,
  successes = 400 * p_hat_returns, failures = 400 * (1 - p_hat_returns),
  se = se_returns, margin = 1.960 * se_returns,
  lower = p_hat_returns - 1.960 * se_returns,
  upper = p_hat_returns + 1.960 * se_returns,
  sample_share_of_orders = 400 / 6000,
  below_15_at_most = 0.15 - (p_hat_returns - 1.960 * se_returns),
  below_15_at_least = 0.15 - (p_hat_returns + 1.960 * se_returns))
```
:::

### Task 10: A uniform model (extra practice for home)

An invented example. A founder runs a studio that prints T-shirts for
student associations. The time to print one order, $X$, is equally likely
to be anywhere between 20 and 60 minutes: $X\sim U(20,60)$.

1. A colleague sketches the density as a flat line at height 0.02 between
   20 and 60 minutes, and zero elsewhere. Show that this is not a valid
   density, and give the correct height.
2. Calculate $P(30\le X\le45)$.
3. Calculate $P(X>50)$.
4. Calculate $P(10\le X\le30)$. Note that a part of this interval lies
   outside the possible values of $X$.
5. State $P(X=40)$. Then explain why the height of the density is not a
   probability.

::: {.callout-note collapse="true"}
## Check your answer

1. A valid density is never negative and has a total area of 1. The sketch
   is never negative, but its area is $(60-20)\times0.02=0.8$, not 1. The
   correct height is $1/(60-20)=0.025$ per minute. Then the area is
   $40\times0.025=1$.
2. A probability is an area: width times height. So
   $P(30\le X\le45)=(45-30)\times0.025=0.375$.
3. $P(X>50)=(60-50)\times0.025=0.25$.
4. The density is zero below 20 minutes, so only the part from 20 to 30
   counts: $P(10\le X\le30)=(30-20)\times0.025=0.25$. A calculation with
   the full width of 20 minutes gives 0.5, which is wrong.
5. $P(X=40)=0$. A single value has width zero, so the area above it is
   zero. The height 0.025 is a probability per minute. It becomes a
   probability only when you multiply it by a width in minutes.

Verification in R:

```{r}
#| label: bep03-uniform-check
c(area_of_sketch = (60 - 20) * 0.02,
  correct_height = 1 / (60 - 20),
  p_30_to_45 = punif(45, min = 20, max = 60) - punif(30, min = 20, max = 60),
  p_above_50 = 1 - punif(50, min = 20, max = 60),
  p_10_to_30 = punif(30, min = 20, max = 60) - punif(10, min = 20, max = 60))
```
:::

### Task 11: A sample size for a target margin of error (extra practice for home)

The founder of the bouldering gym in Task 8 plans a new and larger sample.
She wants to estimate the mean visit time with a margin of error of at most
5 minutes. She uses the sample SD of Task 8 as a planning value:
$\sigma=28$ minutes.

1. Calculate the sample size that she needs at the 95% level
   ($z_{0.975}=1.960$).
2. Calculate the sample size for a margin of 2.5 minutes at the 95% level.
   Compare it with part 1.
3. Calculate the sample size for a margin of 5 minutes at the 99% level
   ($z_{0.995}=2.576$).

::: {.callout-note collapse="true"}
## Check your answer

The formula of Lecture 9, with $m$ the margin of error, is

$$
n\ge\left(\frac{z_{1-\alpha/2}\,\sigma}{m}\right)^2 .
$$

1. $n\ge(1.960\times28/5)^2=10.976^2\approx120.5$, so she needs 121 visits.
   Always round up: with 120 visits the margin would be slightly above 5
   minutes.
2. $n\ge(1.960\times28/2.5)^2=21.952^2\approx481.9$, so 482 visits. Half
   the margin needs about four times the sample.
3. $n\ge(2.576\times28/5)^2=14.4256^2\approx208.1$, so 209 visits. More
   confidence with the same margin needs more observations: 209 in place
   of 121.

The planning value is a guess. If the true $\sigma$ is larger than 28
minutes, the margin of the new sample will be larger than planned.

Verification in R:

```{r}
#| label: bep03-sample-size-check
n_needed <- c(margin_5_at_95 = (1.960 * 28 / 5)^2,
              margin_2.5_at_95 = (1.960 * 28 / 2.5)^2,
              margin_5_at_99 = (2.576 * 28 / 5)^2)
n_needed
ceiling(n_needed)
```
:::

### Task 12: A value for a sample mean (extra practice for home)

An invented example. A founder sells granola in bags of 250 grams. The
filling machine puts $X\sim N(250,4^2)$ grams into one bag. Every hour she
weighs a random sample of $n=16$ bags and calculates the sample mean
$\bar X$. She wants an alarm value $a$: she checks the machine when
$\bar X>a$. When the machine works as described, only 5% of the hourly
samples may give an alarm. Supplied value: the 0.95 quantile of $Z$ is
$z_{0.95}=1.645$.

1. State the distribution of $\bar X$, with its standard deviation.
2. Calculate the value $a$ with $P(\bar X>a)=0.05$.
3. Would the same rule make sense with single bags instead of the mean of
   16 bags? Calculate the value for a single bag and compare.

::: {.callout-note collapse="true"}
## Check your answer

1. The population is normal, so the sample mean is exactly normal:
   $\bar X\sim N(250,\,4^2/16)$, with
   $\operatorname{SD}(\bar X)=4/\sqrt{16}=1$ gram.
2. If 5% of the sample means lie above $a$, then 95% lie at or below it.
   So $a$ is the 0.95 quantile of $\bar X$:
   $a=250+1.645\times1=251.65$ grams.
3. For a single bag the value is $250+1.645\times4=256.58$ grams. Single
   bags vary four times more than the mean of 16 bags. So the mean of 16
   bags detects a change in the filling of the machine much earlier. Its
   alarm value is only 1.65 grams above 250, about 0.7% of the bag weight.

Verification in R:

```{r}
#| label: bep03-alarm-check
c(sd_mean = 4 / sqrt(16),
  alarm_mean = qnorm(0.95, mean = 250, sd = 4 / sqrt(16)),
  alarm_by_hand = 250 + 1.645 * 1,
  alarm_single = 250 + 1.645 * 4,
  shift_in_percent = 100 * 1.645 / 250)
```
:::

:::

::: {.course-phase .phase-after}

## After class

### Individual practice

Do this task alone: no notes, no AI, only the formula sheet. An invented example. A random sample of 64 customers
has a mean monthly spending of €82 and a sample SD of €24. Of the 64
customers, 40 renewed their subscription.

1. Calculate the standard error of the mean spending.
2. Calculate the standard error of the renewal proportion and check its
   condition.
3. State what would happen to both standard errors if the sample size were
   $n=256$.
4. Construct a 95% confidence interval for the mean monthly spending of all
   customers, in the five parts of the exam procedure. The degrees of
   freedom are 63. Use the closest value on the formula sheet:
   $t_{0.975,\,60}=2.000$. The business plan of the founder assumes a mean
   spending of €90. Say in part 5 what the interval tells you about this
   value.
5. Write one sentence that explains the difference between sampling
   uncertainty and selection bias.
6. Treat $\mu=82$ and $\sigma=24$ as the population values. Use the CLT to
   approximate the probability that a new sample of 64 customers shows a
   mean above €85. Supplied value: $P(Z\le 1)=0.841$.
7. Use $\sigma=24$ as a planning value. How many customers must a new
   survey include to estimate the mean spending with a margin of error of
   at most €3, at the 95% level
   ($z_{0.975}=1.96$)?

::: {.callout-note collapse="true"}
## Check a model answer

The standard error of the mean is $24/\sqrt{64}=3$ euros. The renewal
estimate is $\hat p=40/64=0.625$, with standard error
$\sqrt{0.625(0.375)/64}\approx0.0605$. The sample has 40 successes and 24
failures, so the success-failure condition holds. Increasing the sample
size from 64 to 256 would halve both standard errors if the spread
stayed similar.

The interval for the mean spending in five parts:

1. **Formula.** $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used
   because $n=64$ is large.
2. **Given values.** $\bar x=82$ euros, $s=24$ euros, $n=64$, $df=63$. The
   confidence level is 95%, and the closest value on the formula sheet is
   $t_{0.975,\,60}=2.000$.
3. **Calculations.** The standard error is $24/\sqrt{64}=3$ euros. The
   margin of error is $2.000\times3=6$ euros.
4. **Interval.** $(82-6,\ 82+6)=(76,\ 88)$ euros.
5. **Conclusion.** The value 90 euros lies outside the interval. At the 5%
   level we can conclude that the mean monthly spending differs from 90
   euros. The whole interval lies below 90, so customers spend less than
   the plan assumes: at least 2 euros less, about 2% of 90 euros.

The standard errors and the interval describe random variation from sample
to sample, not bias created by an unrepresentative sampling frame. For the
new sample, the standard deviation of the sample mean is
$\operatorname{SD}(\bar X)=24/\sqrt{64}=3$ euros and
$\bar X\mathrel{\dot\sim}N(82,3^2)$, so
$z=(85-82)/3=1$ and $P(\bar X>85)\approx1-P(Z\le1)\approx0.16$. For a margin
of €3, $n\ge(1.96\times24/3)^2=15.68^2\approx245.9$, so the survey needs 246
customers (Lecture 9).

Verification in R:

```{r}
#| label: bep03-timed-check
p_hat_renew <- 40 / 64
c(se_mean = 24 / sqrt(64),
  se_proportion = sqrt(p_hat_renew * (1 - p_hat_renew) / 64),
  t_exact_df63 = qt(0.975, df = 63),
  lower = 82 - 2.000 * 3, upper = 82 + 2.000 * 3,
  below_90_in_percent = 100 * (90 - 88) / 90,
  p_mean_above_85 = 1 - pnorm(85, mean = 82, sd = 3),
  n_for_margin_3 = (1.96 * 24 / 3)^2)
```
:::

### Check your understanding

Complete from memory:

1. A sampling distribution describes ...
2. The standard deviation of a sample mean equals ... The standard error of
   a sample mean equals ...
3. The success-failure condition requires ...
4. The CLT says ... becomes approximately normal.
5. A larger sample does not fix ...
6. A claimed value that lies outside a 95% confidence interval means ... A
   claimed value that lies inside means ...

::: {.callout-note collapse="true"}
## Check your answers

1. How a statistic would vary across repeated random samples.
2. $\sigma/\sqrt n$. The standard error is the estimated version,
   $s/\sqrt n$.
3. $np\ge10$ and $n(1-p)\ge10$. For a confidence interval the check uses
   $\hat p$.
4. The sample mean. For a large sample its sampling distribution is
   approximately normal. The single observations do not become normal.
5. Selection bias, which comes from the way units entered the frame or the
   sample.
6. Outside: at the 5% level we can conclude that the parameter differs from
   the claimed value. Inside: we cannot conclude that it differs, which is
   not proof that the claimed value is true.
:::

### Make it yours

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Pick one rate
or average that you could estimate from about 100 observations, for
example a conversion rate, the average value of a purchase, or a renewal
share. Then write two sentences. First state the standard error that you
would roughly expect, and the 95% interval that follows from it. Then name one claimed value that your interval would
rule out.
Lecture 10 turns this idea into a hypothesis test.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student venture that sells boxes of ready-made meals on campus. "If 60 of 100 customers renew, the standard error of the renewal share is about 5 percentage points, so my 95% interval runs from roughly 50% to 70%. A claim that only 45% renew would lie outside this interval."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant: ask it to check the standard error and
the interval for your own numbers.
::::

### More practice

For extra exercises on the lectures that this tutorial covers, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Tutorial 3. Show the solution only after I send my answer."

[Continue to Lecture 10: Testing decisions, errors, power, and a first look at regression](b10-testing-errors-power.qmd).

:::
