# Tutorial 3: Normal models, sampling, and standard errors {#b-ep03}

```{r}
#| label: bep03-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 3, you can</span>

- identify a valid population and sampling frame
- distinguish parameters, estimators, and estimates
- calculate standard errors for a mean and a proportion
- calculate and interpret quartiles and the IQR of an observed sample
- use normal-model and CLT probabilities without confusing observations
  and statistics
- qualify conclusions when the sampling design is limited.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the solution is discussed together.

This tutorial covers Lectures 6--8. The t distribution belongs to Lecture 9
and is practised in Tutorial 4. Everything here uses the normal model and
the CLT.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 6--8. If the algebra is slowing
you down, review [Mathematics Refresher: sums, fractions, and
roots](../resources/math-refresher.html).

Use the cumulative formula sheet through Lecture 8.

Complete from memory:

1. Standardising an observation $x$ from $N(\mu,\sigma^2)$ means
   calculating ...
2. A parameter describes ...
3. An estimator is ...
4. The standard error of $\bar X$ is ...
5. The CLT describes ...

::: {.callout-note collapse="true"}
## Check your preparation

1. $z=(x-\mu)/\sigma$: the number of population SDs between $x$ and the
   mean.
2. A parameter describes a population.
3. An estimator is a sample-based random rule used to estimate a parameter.
4. $\sigma/\sqrt n$, estimated by $S/\sqrt n$ when $\sigma$ is unknown.
5. The distribution of the sample mean across random samples. It is
   approximately normal for a large independent sample.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Attempt each task before opening its answer. A complete written answer
names the quantity, shows the calculation, reports the value and units,
interprets it in context, and states one important limit of the model or
the data.

### Task 1: Population, frame, and claim

A student randomly selects 100 rows from the Shark Tank file and estimates
the mean equity offered (`equity_offered_pct`).

1. State the population about which this sample directly supports
   conclusions.
2. Explain why it does not represent all *Shark Tank* applicants.
3. Explain why it does not represent all US entrepreneurs.
4. State one change needed to target all US entrepreneurs.

::: {.callout-note collapse="true"}
## Check your answer

The direct finite population is the 1,441 recorded pitches in completed
Seasons 1--16. Applicants who were not televised are absent, so the frame
does not represent all applicants. The show also selects a specialised
group of entrepreneurs, not a probability sample of US entrepreneurs.
Targeting all US entrepreneurs would require an appropriate national
sampling frame and random selection from that frame.
:::

### Task 2: Reproduce the realised sample

The sample is drawn with a visible seed, exactly as in Lecture 7, so every
reader gets the same 100 rows:

```{r}
#| label: bep03-sample-output
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

c(
  n = nrow(pitch_sample),
  mean_equity = mean(pitch_sample$equity_offered_pct),
  sd_equity = sd(pitch_sample$equity_offered_pct),
  q1_equity = unname(quantile(pitch_sample$equity_offered_pct, 0.25)),
  median_equity = median(pitch_sample$equity_offered_pct),
  q3_equity = unname(quantile(pitch_sample$equity_offered_pct, 0.75)),
  deal_proportion = mean(pitch_sample$deal_on_show)
)
```

1. Identify the parameter, estimator, and estimate for the mean equity
   offered (`equity_offered_pct`).
2. Identify the parameter, estimator, and estimate for the deal proportion
   (`deal_on_show`).
3. Calculate the interquartile range (IQR) of the equity offered in the
   sample and interpret it. The IQR is defined in
   [Lecture 1](b01-introduction-data-analysis.qmd#b01-quartiles).
4. Explain why a different seed can produce different estimates and a
   different IQR.

::: {.callout-note collapse="true"}
## Check your answer

For the equity offered, the parameter is the population mean $\mu$, the
estimator is $\bar X$, and the estimate is $\bar x=12.76$%. For the deal
outcome, the
parameter is the population proportion $p$, the estimator is $\widehat P$,
and the estimate is $\hat p=0.64$. The sample quartiles are $Q_1=8$ and
$Q_3=17.75$, so $\operatorname{IQR}=17.75-8=9.75$ percentage points: the
middle half of the sampled initial equity offers lies from 8% to 17.75%. A
different seed selects a different random set of rows, so all realised
sample statistics, including the IQR, can change.
:::

### Task 3: Standard errors and sample size

Use the sample output above.

1. Calculate the estimated standard error of the mean equity offered,
   $\widehat{\operatorname{SE}}(\bar X)$.
2. Calculate the estimated standard error of the deal proportion,
   $\widehat{\operatorname{SE}}(\widehat P)$ (Lecture 8).
3. Check the success-failure condition for $\widehat P$.
4. Predict how each SE changes if $n$ increases from 100 to 400 while the
   estimated spread remains similar.
5. The sample SD in part 1 was calculated with `sd()`, which divides by
   $n-1$. State that number here and explain it in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$$
\widehat{\operatorname{SE}}(\bar X)=7.7079/\sqrt{100}=0.7708.
$$

$$
\widehat{\operatorname{SE}}(\widehat P)
=\sqrt{0.64(0.36)/100}=0.0480.
$$

The first is about 0.77 percentage points of equity, the second about 4.8
percentage points of the deal rate. The sample contains 64 successes and 36
failures, so both are at least 10. Increasing $n$ fourfold divides both SEs by $\sqrt 4=2$. The divisor is
$n-1=99$: the deviations are measured from the estimated sample mean and
must sum to zero, so only 99 of them vary freely, and dividing by 99
removes the resulting downward bias (Lecture 7).
:::

### Task 4: One pitch versus one mean

Suppose the population mean of the equity offered is $\mu=13.21$%, the
population SD is $\sigma=8.44$ percentage points, and samples of $n=100$ are
drawn independently. Supplied values: $P(Z\le 2.12)=0.983$ and
$P(Z\le 0.21)=0.583$.

1. State the approximate sampling distribution of $\bar X$.
2. Calculate $P(\bar X>15)$.
3. A classmate instead models a *single* pitch as
   $X\sim N(13.21, 8.44^2)$ and calculates $P(X>15)$. Compute that value
   too, and explain why the two answers are so different.
4. Which of the two normal models did Lecture 6 reject for the equity
   offered, and why does the CLT still justify the calculation in part 2?

::: {.callout-note collapse="true"}
## Check your answer

The CLT gives approximately
$\bar X\mathrel{\dot\sim}N(13.21,\,8.44^2/100)$, with SE $0.844$.
Therefore

$$
P(\bar X>15)
=1-P\!\left(Z\le\frac{15-13.21}{0.844}\right)
=1-P(Z\le2.12)
\approx0.017.
$$

For a single pitch, $z=(15-13.21)/8.44\approx0.21$, so
$P(X>15)\approx0.42$. The mean of 100 pitches varies ten times less than
one pitch ($\sigma/\sqrt{100}$), so 15 is an ordinary value for one pitch
but an extreme value for a mean of 100. Lecture 6 rejected the normal
model for single pitches, because the equity offered is strongly
right-skewed. So the model in part 3 is a bad model. The CLT concerns the
sample mean. Averaging 100 independent draws gives a nearly normal sampling
distribution even from a skewed population, which is why the calculation in
part 2 is justified.
:::

### Task 5: A proportion through the CLT

The deal proportion (`deal_on_show`) in the Shark Tank file is $p=0.612$. A
random sample of $n=100$ pitches is planned, with sample deal proportion
$\widehat P$. Supplied value: $P(Z\le -2.30)=0.011$.

1. Calculate $\operatorname{SE}(\widehat P)$ and check the
   success-failure condition.
2. State the approximate sampling distribution of $\widehat P$.
3. Approximate the probability that the sample shows a deal rate of at
   most 50%, that is, $P(\widehat P\le0.50)$.
4. Interpret the result in one sentence.

::: {.callout-note collapse="true"}
## Check your answer

$\operatorname{SE}(\widehat P)=\sqrt{0.612(0.388)/100}\approx0.0487$.
The expected counts are $np=61.2$ and $n(1-p)=38.8$, both at least 10.
Hence $\widehat P\mathrel{\dot\sim}N(0.612,\,0.0487^2)$. Standardise:

$$
z=\frac{0.50-0.612}{0.0487}\approx-2.30,
\qquad
P(\widehat P\le0.50)\approx P(Z\le-2.30)\approx0.011.
$$

If the deal proportion in the file is 61.2%, only about 1% of random samples of
100 would show a deal rate at or below 50%. Such a sample would be
surprising. Lecture 10 builds the hypothesis test on this idea.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice: 15 minutes

Do this task alone and against the clock: no notes, no AI, formula sheet
only, as in the exam. An invented example: a random sample of 64 customers
has mean monthly spending €82 and sample SD €24. Among the 64 customers, 40
renewed their subscription.

1. Estimate the SE of mean spending.
2. Estimate the SE of the renewal proportion and check its condition.
3. State what would happen to both SEs at $n=256$.
4. Write one sentence distinguishing sampling uncertainty from selection
   bias.
5. Treating $\mu=82$ and $\sigma=24$ as the population values, use the CLT
   to approximate the probability that a *new* sample of 64 shows a mean
   above €85. Supplied value: $P(Z\le 1)=0.841$.
6. Use $\sigma=24$ as a planning value. How many customers must a new
   survey include to estimate mean spending to within €3, at the 95% level
   ($z_{0.975}=1.96$)?

::: {.callout-note collapse="true"}
## Check a model answer

The estimated SE of the mean is $24/\sqrt{64}=3$ euros. The renewal
estimate is $40/64=0.625$, with estimated SE
$\sqrt{0.625(0.375)/64}\approx0.0605$. The sample has 40 successes and 24
failures, so the success-failure condition holds. Increasing the sample
size from 64 to 256 would halve both SEs if the spread remained similar.
These SEs describe random sample-to-sample variation, not bias created by
an unrepresentative sampling frame. For the new sample,
$\bar X\mathrel{\dot\sim}N(82,3^2)$, so
$z=(85-82)/3=1$ and $P(\bar X>85)\approx1-P(Z\le1)\approx0.16$. For a margin
of €3, $n\ge(1.96\times24/3)^2=15.68^2\approx245.9$, so the survey needs 246
customers (Lecture 8).
:::

### Check your understanding

Complete from memory:

1. A sampling distribution describes ...
2. The standard error of a mean equals ...
3. The success-failure condition requires ...
4. The CLT says ... becomes approximately normal.
5. A larger sample does not fix ...

::: {.callout-note collapse="true"}
## Check your answers

1. How a statistic would vary across repeated random samples.
2. $\sigma/\sqrt n$ (estimated by $S/\sqrt n$).
3. $np\ge10$ and $n(1-p)\ge10$.
4. The sample mean. For a large sample its sampling distribution is
   approximately normal. The single observations do not become normal.
5. Selection bias: how units entered the frame or the sample.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Pick one rate
or average you could estimate from about 100 observations (conversion
rate, average basket, renewal share), and write two sentences. First state
the SE you would roughly expect and what range of sample outcomes would
therefore not surprise you. Then say what observed value would surprise you.
Lecture 10 turns this idea into a hypothesis test.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "From 100 orders, a renewal share near 60% has an SE of about 5 points, so observed shares between roughly 50% and 70% would not surprise me. A share of 45% would surprise me, because it is more than two SEs below what I believe."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant: ask it to check the SE you guessed for
your own numbers.
::::

[Continue to Lecture 9: The t distribution and confidence intervals](b09-t-confidence-intervals.qmd).

:::
