# Tutorial 4: Hypothesis tests, errors, and power {#b-ep04}

```{r}
#| label: bep04-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 4, you can</span>

- carry out a two-sided test for a mean and for a proportion in five
  steps, with a critical value and a rejection region
- write the conclusion in the words of the question, also when $H_0$ is not
  rejected
- connect a test decision with the confidence interval and with the
  p-value
- describe a Type I error and a Type II error in context, and say how the
  significance level and the sample size can change a decision
- read a row of regression coefficients without claiming more than the
  row shows.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You first try each task in pairs. You may use AI for this, with
the four steps Attempt, Ask, Verify, Explain. Then the class discusses the
solution together.

This tutorial covers Lecture 10: the five steps of a test, the link between
a test and a confidence interval, the p-value, the two errors, and power.
The tutorial also uses the confidence intervals of Lecture 9 again. All
tests in the exam are two-sided.

::: {.course-phase .phase-before}

## Before class

<span class="time-budget">Total preparation time: about 15 minutes.</span>

Bring your after-class work from Lectures 9 and 10. If you need it, review
[Mathematics Refresher: intervals, degrees of freedom, and
standardisation](../resources/math-refresher.html).

Use the formula sheet with all formulas up to and including Lecture 10.

Write from memory:

1. the five steps of a test, in order
2. the test statistic for a mean when $s$ is used, and its distribution if
   $H_0$ is true
3. the test statistic for a proportion, and its distribution if $H_0$ is
   true
4. the rejection rule of a two-sided test at the significance level
   $\alpha$
5. the definition of a two-sided p-value, and how it gives the same
   decision as the critical value
6. the definitions of a Type I error and a Type II error.

::: {.callout-note collapse="true"}
## Check your preparation

1. Hypotheses, with the significance level. Test statistic, with its
   distribution if $H_0$ is true and the condition. Critical value and
   rejection region. Observed value. Conclusion.
2. $T=(\bar X-\mu_0)/(S/\sqrt n)$. If $H_0$ is true, $T$ has the $t_{n-1}$
   distribution. The condition: the population is normal, or $n$ is large.
3. $Z=(\widehat P-p_0)/\sqrt{p_0(1-p_0)/n}$. If $H_0$ is true, $Z$ is
   approximately $N(0,1)$, provided $np_0\ge10$ and $n(1-p_0)\ge10$. The
   denominator is the standard deviation of $\widehat P$ under $H_0$.
4. Reject $H_0$ if $|t_{obs}|\ge t_{1-\alpha/2,\,n-1}$, or if
   $|z_{obs}|\ge z_{1-\alpha/2}$ for a proportion. In words: the observed
   value is at or below minus the critical value, or at or above the
   critical value.
5. The probability of a test statistic at least as far from zero as the
   observed value, in either direction. You calculate this probability as
   if $H_0$ were true: $2\,P(T\ge|t_{obs}|)$. The p-value is at most $\alpha$ exactly when the
   observed value lies in the rejection region.
6. A Type I error is rejecting $H_0$ when $H_0$ is true. Its probability is
   at most $\alpha$. A Type II error is not rejecting $H_0$ when $H_1$ is
   true.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try every task before you open its answer. Every conclusion must name the
parameter, the units, the population, and the direction of the difference.
It must also name the limits of the evidence. Write every test in the five steps and every confidence interval
in the five parts, numbered 1 to 5.

Every task shows a time in minutes for your attempt and the discussion.
Tasks 1 to 4, 6, and 7 take about 76 minutes together. Task 5 is optional
and is extra practice for home.

In the exam, a test for a mean always uses $s$ and a t value. If a t value
is not on the formula sheet, the task gives it as a supplied value.

Tasks 1, 2, and 5 use the random sample of 100 pitches (seed 123) from
Lecture 7. The code below draws it in the same way:

```{r}
#| label: bep04-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]

c(n = nrow(pitch_sample),
  mean_equity = mean(pitch_sample$equity_offered_pct),
  sd_equity = sd(pitch_sample$equity_offered_pct),
  deals = sum(pitch_sample$deal_on_show))
```

### Task 1: A test for a mean and its confidence interval (20 minutes)

[Shark Tank data]{.shark-tag} An advisor says that pitches offer 14.5% equity on average. Let $\mu$ be
the mean equity offered (`equity_offered_pct`) among all 1,441 recorded
pitches. The sample of 100 pitches has the mean $\bar x=12.76$% and the
standard deviation $s=7.71$ percentage points. The degrees of freedom are
99. Use the closest values on the formula sheet: $t_{0.975,\,100}=1.984$
and $t_{0.995,\,100}=2.626$.

1. Test $H_0:\mu=14.5$ against $H_1:\mu\ne14.5$ at the 5% level. Write the
   five steps.
2. Repeat steps 1, 3, and 5 of the test at the 1% level. State the decision
   and say whether it changes.
3. Construct the 95% confidence interval for $\mu$ in five parts.
4. Explain how the interval of part 3 shows the decision of part 1. Also
   explain what the interval tells you in addition to the decision.
5. R reports a two-sided p-value of 0.026 for this test. Show that the
   p-value gives the same decisions at the 5% level and at the 1% level.
6. The equity offered is strongly right-skewed. Explain why the test may
   still use a t critical value with 99 degrees of freedom.

::: {.callout-note collapse="true"}
## Check your answer

Part 1, the test in five steps:

1. **Hypotheses.** $H_0:\mu=14.5$ against $H_1:\mu\ne14.5$. Here $\mu$ is
   the mean equity offered among all 1,441 recorded pitches, in percent.
   The significance level is $\alpha=0.05$.
2. **Test statistic.** $T=(\bar X-14.5)/(S/\sqrt n)$. If $H_0$ is true,
   $T$ has the t distribution with $n-1=99$ degrees of freedom. The
   population is not normal, but $n=100$ is large.
3. **Critical value and rejection region.** The critical value is
   $t_{0.975,\,99}\approx1.984$. Reject $H_0$ if $t_{obs}\le-1.984$ or
   $t_{obs}\ge1.984$.
4. **Observed value.** The standard error is $s/\sqrt n=7.71/\sqrt{100}
   =0.771$. Then $t_{obs}=(12.76-14.5)/0.771=-1.74/0.771=-2.26$.
5. **Conclusion.** The observed value $-2.26$ is below $-1.984$, so it lies
   in the rejection region. Reject $H_0$. At the 5% level we can conclude
   that the mean equity offered among the recorded pitches differs from
   14.5%. The sign of $t_{obs}$ is negative, so the mean lies below 14.5%.

Part 2, the test at the 1% level:

Step 1. **Hypotheses.** As before, with $\alpha=0.01$.

Step 3. **Critical value and rejection region.** The critical value is
$t_{0.995,\,99}\approx2.626$. Reject $H_0$ if $t_{obs}\le-2.626$ or
$t_{obs}\ge2.626$.

Step 5. **Conclusion.** The observed value $-2.26$ lies between $-2.626$ and
2.626, so it does not lie in the rejection region. Do not reject $H_0$.
At the 1% level we cannot conclude that the mean equity offered differs
from 14.5%.

The decision changes with $\alpha$. The stricter 1% level needs stronger
evidence, and the same sample does not give it.

Part 3, the interval in five parts:

1. **Formula.** $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used
   because $n=100$ is large.
2. **Given values.** $\bar x=12.76$%, $s=7.71$ percentage points, $n=100$,
   $df=99$. The confidence level is 95%, so the critical value is
   $t_{0.975,\,99}\approx1.984$.
3. **Calculations.** The standard error is $7.71/\sqrt{100}=0.771$. The
   margin of error is $1.984\times0.771=1.53$ percentage points.
4. **Interval.** $(12.76-1.53,\ 12.76+1.53)=(11.23,\ 14.29)$ percent.
5. **Conclusion.** The value 14.5 lies outside the interval. At the 5%
   level we can conclude that the mean equity offered differs from 14.5%,
   and the whole interval lies below 14.5%.

Part 4. The test and the interval use the same critical value and the same
standard error. So the test rejects $H_0:\mu=14.5$ at the 5% level exactly
when 14.5 lies outside the 95% interval. The interval also shows the size of
the difference: the mean lies between 0.21 and 3.27 percentage points below
the 14.5% of the advisor. The lower end, 0.21 points, is about 1.5% of
14.5, so the difference may be small.

Part 5. The p-value 0.026 is below 0.05, so we reject $H_0$ at the 5%
level. The p-value is above 0.01, so we do not reject $H_0$ at the 1%
level. The critical values give the same decisions. The p-value is at most
$\alpha$ exactly when the observed value lies in the rejection region.

Part 6. The test uses a t value because $\sigma$ is unknown and $s$ is used
in its place. The sample is large, so by the CLT the test statistic is
approximately normal, although single offers are skewed. With 99 degrees of
freedom the t distribution is almost the same as the standard normal: the
critical value is 1.984 against 1.960. The rule is: with $s$, always use
the t distribution with $n-1$ degrees of freedom.

One limit: the conclusion is about the recorded pitches in the file, not
about all entrepreneurs.

Verification in R:

```{r}
#| label: bep04-mean-test
x <- pitch_sample$equity_offered_pct
n <- length(x)
x_bar <- mean(x)
s <- sd(x)
se <- s / sqrt(n)
t_obs <- (x_bar - 14.5) / se
ci <- x_bar + c(-1, 1) * qt(0.975, df = n - 1) * se

c(mean = x_bar, sd = s, se = se, df = n - 1, t_obs = t_obs,
  critical_5 = qt(0.975, df = n - 1), critical_1 = qt(0.995, df = n - 1),
  p_value = 2 * pt(-abs(t_obs), df = n - 1))
c(lower = ci[1], upper = ci[2],
  below_14.5_at_least = 14.5 - ci[2], below_14.5_at_most = 14.5 - ci[1],
  low_end_in_percent = 100 * (14.5 - ci[2]) / 14.5)
c(se_by_hand = 7.71 / sqrt(100), t_by_hand = (12.76 - 14.5) / 0.771,
  margin_by_hand = 1.984 * 0.771)
t.test(x, mu = 14.5)
```
:::

### Task 2: A test for a proportion and its confidence interval (15 minutes)

[Shark Tank data]{.shark-tag} A commentator claims that three quarters of all pitches end with a deal.
Let $p$ be the proportion of pitches with a deal (`deal_on_show`) among all
1,441 recorded pitches. In the sample of 100 pitches, 64 ended with a deal.

1. Test $H_0:p=0.75$ against $H_1:p\ne0.75$ at the 5% level. Write the five
   steps, with the check of the condition.
2. Construct the 95% confidence interval for $p$ in five parts, with the
   check of the condition.
3. The test divides by 0.0433. The interval uses 0.048. Name these two
   quantities and explain why their values differ.
4. Does the interval lead to the same conclusion about 0.75 as the test?
   Is this always so for a proportion?

::: {.callout-note collapse="true"}
## Check your answer

Part 1, the test in five steps:

1. **Hypotheses.** $H_0:p=0.75$ against $H_1:p\ne0.75$. Here $p$ is the
   proportion of pitches with a deal among all 1,441 recorded pitches. The
   significance level is $\alpha=0.05$.
2. **Test statistic.** $Z=(\widehat P-p_0)/\sqrt{p_0(1-p_0)/n}$. If $H_0$
   is true, $Z$ is approximately $N(0,1)$. The condition holds:
   $np_0=100\times0.75=75$ and $n(1-p_0)=100\times0.25=25$ are both at
   least 10.
3. **Critical value and rejection region.** The critical value is
   $z_{0.975}=1.960$. Reject $H_0$ if $z_{obs}\le-1.960$ or
   $z_{obs}\ge1.960$.
4. **Observed value.** The denominator is the standard deviation of
   $\widehat P$ under $H_0$: $\sqrt{0.75\times0.25/100}=0.0433$. The
   estimate is $\hat p=64/100=0.64$. Then
   $z_{obs}=(0.64-0.75)/0.0433=-2.54$.
5. **Conclusion.** The observed value $-2.54$ is below $-1.960$, so it lies
   in the rejection region. Reject $H_0$. At the 5% level we can conclude
   that the proportion of deals among the recorded pitches differs from
   75%. The sign of $z_{obs}$ is negative, so the proportion lies below
   75%.

Part 2, the interval in five parts:

1. **Formula.**
   $\widehat P\pm z_{1-\alpha/2}\sqrt{\widehat P(1-\widehat P)/n}$. The
   condition holds: $n\hat p=64$ and $n(1-\hat p)=36$ are both at least 10.
2. **Given values.** 64 deals, $n=100$, $\hat p=0.64$. The confidence level
   is 95%, so $z_{0.975}=1.960$.
3. **Calculations.** The standard error is
   $\sqrt{0.64\times0.36/100}=0.048$. The margin of error is
   $1.960\times0.048=0.094$.
4. **Interval.** $(0.64-0.094,\ 0.64+0.094)=(0.546,\ 0.734)$, that is, from
   54.6% to 73.4% of the pitches.
5. **Conclusion.** The value 0.75 lies outside the interval. At the 5%
   level we can conclude that the proportion of deals differs from 75%, and
   the whole interval lies below 75%.

Part 3. The value 0.0433 is the standard deviation of $\widehat P$ under
$H_0$, $\sqrt{p_0(1-p_0)/n}$. The test asks how surprising the sample would
be if $H_0$ were true. So the test takes $p=p_0=0.75$ as given and does
not estimate anything. The value 0.048 is the standard error,
$\sqrt{\hat p(1-\hat p)/n}$. The interval has no claimed value of $p$, so
it uses the estimate $\hat p=0.64$.

Part 4. Here the interval and the test agree: 0.75 lies outside the
interval, and the test rejects $H_0$. For a proportion this agreement is
not certain, because the two denominators differ. When the observed value
is very close to the critical value, the test and the interval can give
different answers. Use the procedure that the question names.

R reports a p-value of about 0.011 for this test. This p-value is below
0.05, so it gives the same decision.

Verification in R:

```{r}
#| label: bep04-proportion-test
n <- nrow(pitch_sample)
p_hat <- mean(pitch_sample$deal_on_show)
p0 <- 0.75

sd_null <- sqrt(p0 * (1 - p0) / n)
z_obs <- (p_hat - p0) / sd_null
se_hat <- sqrt(p_hat * (1 - p_hat) / n)
ci_p <- p_hat + c(-1, 1) * 1.960 * se_hat

c(successes_H0 = n * p0, failures_H0 = n * (1 - p0),
  p_hat = p_hat, sd_under_H0 = sd_null, z_obs = z_obs,
  p_value = 2 * pnorm(-abs(z_obs)))
c(standard_error = se_hat, margin = 1.960 * se_hat,
  lower = ci_p[1], upper = ci_p[2])
```
:::

### Task 3: Read a regression coefficient row (8 minutes)

Suppose that you regress the equity offered (in percentage points) on the
amount requested. The regression output shows this row for the amount
requested. The sample is large, so the critical
value at the 5% level is about 1.96.

| Predictor | Estimate | Std. Error | t value | p-value | 95% interval |
|---|---:|---:|---:|---:|---:|
| Amount requested, per \$100,000 | -0.290 | 0.058 | -5.00 | <0.001 | (-0.404, -0.176) |

1. State the tested hypotheses.
2. Explain every number in the row, with its units.
3. Give the statistical conclusion at the 5% level.
4. State two claims that the row cannot justify.

::: {.callout-note collapse="true"}
## Check your answer

Write $\beta_1$ for the population coefficient of the amount requested.
The test is $H_0:\beta_1=0$ against $H_1:\beta_1\ne0$. The estimate
$\hat\beta_1=-0.290$ says that each extra \$100,000 requested goes together
with 0.290 percentage points less equity offered. The standard error,
0.058, measures the sampling uncertainty of the estimate in the same units.
The observed value of the test statistic is $t_{obs}=-0.290/0.058=-5.00$:
the estimate lies five standard errors below zero. The interval says that
plausible values of the coefficient run from $-0.404$ to $-0.176$
percentage points per \$100,000.

The observed value $-5.00$ is far below $-1.96$, so it lies in the
rejection region. Reject $H_0$ at the 5% level: the coefficient differs
from zero, and the negative sign shows that it lies below zero. The p-value below
0.001 and the 95% interval, which does not contain zero, show the same
decision. The row alone does not show that asking for more money causes
lower equity offers, and it does not show that the conditions of the model
hold.

Verification in R:

```{r}
#| label: bep04-row-check
c(t_obs = -0.290 / 0.058,
  lower = -0.290 - 1.96 * 0.058, upper = -0.290 + 1.96 * 0.058,
  p_value = 2 * pnorm(-5))
```
:::

### Task 4: Errors, power, and what changes a decision (15 minutes)

Parts 1 to 5 are about the test of Task 3:

1. describe a Type I error
2. describe a Type II error
3. state how a larger sample affects the standard error and the power
4. explain why raising $\alpha$ has a cost as well as a benefit
5. state which of the two errors is possible after the decision in Task 3.

Parts 6 to 9 use an invented example. A founder runs a barbershop and plans
40 minutes for one haircut. She draws a random sample of 36 haircuts from
the booking system of last month. The sample has a mean of 44 minutes and a
standard deviation of $s=12$ minutes. Haircut times are roughly normally
distributed. She tests $H_0:\mu=40$ against $H_1:\mu\ne40$, where $\mu$ is
the mean time of all haircuts of last month. Supplied values:
$t_{0.95,\,35}=1.690$ and $t_{0.995,\,35}=2.724$.

6. Calculate the observed value $t_{obs}$.
7. Decide at the 10% level and at the 1% level. State the critical value
   and the rejection region each time.
8. Suppose that the same mean and the same standard deviation come from a
   sample of 144 haircuts. Calculate $t_{obs}$ again, and decide at the 1%
   level. The degrees of freedom are 143. Use the closest value on the
   formula sheet: $t_{0.995,\,120}=2.617$.
9. Explain how $\alpha$ and $n$ changed the decision. State which error is
   possible after each of the three decisions.

::: {.callout-note collapse="true"}
## Check your answer

Parts 1 to 5. A Type I error is concluding that the population coefficient
of the amount requested differs from zero when it equals zero. A Type II
error is not rejecting $H_0$ when the population coefficient is in fact not
zero. A larger independent sample usually reduces the standard error and
raises the power. Raising $\alpha$ also raises the power, but it raises the
probability of a Type I error at the same time. A wider rejection region
leads to more rejections, both when $H_1$ is true and when $H_0$ is true.
In Task 3 we rejected the null hypothesis, so only a Type I error is
possible. A Type II error is possible only after a decision not to reject.

Part 6. Step 4. The standard error is $s/\sqrt n=12/\sqrt{36}=2$ minutes.
Then $t_{obs}=(44-40)/2=2.00$.

Part 7. At the 10% level: Step 3. Each tail holds 5%, so the critical value
is $t_{0.95,\,35}=1.690$. Reject $H_0$ if $t_{obs}\le-1.690$ or
$t_{obs}\ge1.690$. Step 5. The observed value 2.00 lies in the rejection
region. Reject $H_0$: at the 10% level we can conclude that the mean
haircut time differs from 40 minutes. The positive sign shows that the
mean lies above 40 minutes. At the 1% level: Step 3. The critical value is
$t_{0.995,\,35}=2.724$. Reject $H_0$ if $t_{obs}\le-2.724$ or
$t_{obs}\ge2.724$. Step 5. The observed value 2.00 does not lie in this
rejection region. Do not reject $H_0$: at the 1% level we cannot conclude
that the mean haircut time differs from 40 minutes.

Part 8. Step 3. The critical value is $t_{0.995,\,120}=2.617$, the closest
listed value for $df=143$. Reject $H_0$ if $t_{obs}\le-2.617$ or
$t_{obs}\ge2.617$. Step 4. With 144 haircuts the standard error is
$12/\sqrt{144}=1$ minute, so $t_{obs}=(44-40)/1=4.00$. Step 5. The observed value 4.00
lies in the rejection region, so reject $H_0$. At the 1% level we can
conclude that the mean haircut time differs from 40 minutes. The positive
sign shows that the mean lies above 40 minutes.

Part 9. A smaller $\alpha$ moves the critical value away from zero, from
1.690 to 2.724. With the same data, we then no longer reject $H_0$. This
protects against a Type I error, but it makes a Type II error more
likely. A larger
sample makes the standard error smaller, from 2 minutes to 1 minute. The
same difference of 4 minutes, 10% of the planned 40 minutes, then gives a
larger observed value, 4.00 instead of 2.00. This means higher power. After
the two rejections only a Type I error is possible. After the decision not
to reject, with 36 haircuts at the 1% level, only a Type II error is
possible.

Verification in R:

```{r}
#| label: bep04-alpha-n-check
c(se_36 = 12 / sqrt(36), t_obs_36 = (44 - 40) / (12 / sqrt(36)),
  critical_10 = qt(0.95, df = 35), critical_1 = qt(0.995, df = 35),
  se_144 = 12 / sqrt(144), t_obs_144 = (44 - 40) / (12 / sqrt(144)),
  critical_1_df143 = qt(0.995, df = 143),
  difference_in_percent = 100 * (44 - 40) / 40)
```
:::

### Task 5 (optional): One-sided alternatives and the exam convention (8 minutes, extra practice for home)

[Shark Tank data]{.shark-tag} This task belongs to the optional box on one-sided alternatives in
Lecture 10. It is not part of the exam.

Task 1 gave $t_{obs}=-2.26$ with $df=99$ and a two-sided p-value of about
0.026.

1. Give the p-values for the one-sided alternatives $H_1:\mu<14.5$ and
   $H_1:\mu>14.5$. Assume that the researcher chose each alternative before
   looking at the data.
2. A student first sees that the sample mean is below 14.5. Then the student
   wants to report the left-sided p-value "because the direction is now
   known". Explain the error.
3. State the convention of this course for tests and one reason for it.

::: {.callout-note collapse="true"}
## Check your answer

Left-sided: $P(T\le-2.26)\approx0.013$, half of the two-sided p-value.
Right-sided: $P(T\ge-2.26)\approx0.987$, one minus that half. Suppose a
researcher chooses the tail after seeing the estimate. Then the 5% always
lies on the side where the estimate happens to fall. The true Type I error
rate is then about 10%, not 5%. A one-sided alternative must follow from
the research question. The researcher must choose it before looking at
the sample. The convention of this course: all
tests are two-sided. In business questions a difference in either direction
usually matters. Also, with this convention nobody can choose the tail
after seeing the data.

Verification in R, with the objects of Task 1:

```{r}
#| label: bep04-one-sided-check
c(left_sided = pt(t_obs, df = 99), right_sided = 1 - pt(t_obs, df = 99))
```
:::

### Task 6: Repair the conclusion (6 minutes)

The following conclusion about the row of Task 3 contains two errors. Find
them and write a better conclusion.

> The coefficient is significant, so asking for more money causes founders
> to offer less equity, and there is less than a 0.1% chance that the null
> hypothesis is true.

::: {.callout-note collapse="true"}
## Check your answer

The p-value is not the probability that the null hypothesis is true, and a
significant result does not show a cause. A better conclusion: "Each extra
\$100,000 requested goes together with about 0.29 percentage points less
equity offered (95% interval from 0.18 to 0.40 points less). If the
population coefficient were zero, an estimate this far from zero would occur
in fewer than 0.1% of samples. Nobody assigned the requested amounts to
the pitches at random, so this is an association, not a causal effect."
:::

### Task 7: A test that does not reject (12 minutes)

An invented example. A supplier of snack vending machines states in its
brochure that one machine sells 120 items per week on average. A founder
is thinking about signing a contract with this supplier. She draws a random sample of 25
of the 800 machines of the supplier. Last week these 25 machines sold 114
items on average, with a standard deviation of $s=20$ items. Weekly sales
per machine are roughly normally distributed. Supplied value:
$t_{0.975,\,24}=2.064$.

1. Test at the 5% level whether the mean weekly sales of all 800 machines
   differ from 120 items. Write the five steps.
2. The founder writes: "The test shows that the brochure is right. The
   mean is 120 items." Assess this conclusion and write a better one.
3. Calculate the 95% confidence interval for the mean weekly sales. Parts
   3 and 4 of the procedure are enough. Use the interval to support your
   answer to part 2.
4. State which error is possible after the decision of part 1, and describe
   it in the context of the founder.

::: {.callout-note collapse="true"}
## Check your answer

The test in five steps:

1. **Hypotheses.** $H_0:\mu=120$ against $H_1:\mu\ne120$. Here $\mu$ is the
   mean number of items that one machine sold last week, among all 800
   machines of the supplier. The significance level is $\alpha=0.05$.
2. **Test statistic.** $T=(\bar X-120)/(S/\sqrt n)$. If $H_0$ is true, $T$
   has the t distribution with $n-1=24$ degrees of freedom. The sample is
   small, so the condition is that the population is roughly normal.
3. **Critical value and rejection region.** The critical value is
   $t_{0.975,\,24}=2.064$. Reject $H_0$ if $t_{obs}\le-2.064$ or
   $t_{obs}\ge2.064$.
4. **Observed value.** The standard error is $s/\sqrt n=20/\sqrt{25}=4$
   items. Then $t_{obs}=(114-120)/4=-1.50$.
5. **Conclusion.** The observed value $-1.50$ lies between $-2.064$ and
   2.064, so it does not lie in the rejection region. Do not reject $H_0$.
   At the 5% level we cannot conclude that the mean weekly sales per
   machine differ from 120 items.

The conclusion of the founder claims too much. "Do not reject $H_0$" is not
the same as "accept $H_0$". The sample gave too little evidence against
120 items. The sample did not show that the mean is 120 items. A better conclusion: "The sample
of 25 machines gives no evidence at the 5% level that the mean weekly sales
differ from the 120 items of the brochure. But the sample is also in line
with clearly lower values."

The interval supports this. The margin of error is $2.064\times4=8.26$
items, so the 95% interval is $(114-8.26,\ 114+8.26)=(105.74,\ 122.26)$
items. It contains 120, and it also contains 106. A mean of 106 items would
be 14 items below the brochure, about 12% less. With this sample, you
cannot tell which of these two values is closer to the truth.

After a decision not to reject, only a Type II error is possible. Here it
would mean: the mean weekly sales of the 800 machines differ from 120
items, but the test did not detect this difference. With 25 machines the standard error
is 4 items. A larger sample has a smaller standard error and more power.

Verification in R:

```{r}
#| label: bep04-no-reject-check
se_items <- 20 / sqrt(25)
c(se = se_items, t_obs = (114 - 120) / se_items,
  critical = qt(0.975, df = 24),
  margin = 2.064 * se_items,
  lower = 114 - 2.064 * se_items, upper = 114 + 2.064 * se_items,
  below_120_in_percent = 100 * (120 - 106) / 120)
```
:::

:::

::: {.course-phase .phase-after}

## After class

<span class="time-budget">Total practice time: about 50 minutes.</span>

### Timed individual practice: 40 minutes

Solve this task alone and with a timer, under exam conditions: no notes, no
AI, only the formula sheet. An invented example. A random sample of 81
customers of a subscription service gives a mean spending of 42 euros and a
sample SD of 9 euros. Of the 81 customers, 55 renew their subscription.

1. Let $\mu$ be the population mean spending. Test $H_0:\mu=40$ against
   $H_1:\mu\ne40$ at the 5% level, in five steps. Supplied value:
   $t_{0.975,\,80}=1.990$.
2. Construct the 95% confidence interval for $\mu$ in five parts. Say in
   part 5 how the interval fits the decision of part 1.
3. Let $p$ be the population renewal proportion. Test $H_0:p=0.60$ against
   $H_1:p\ne0.60$ at the 5% level, in five steps. Then calculate the 95%
   confidence interval for $p$. For the interval, parts 3 and 4 of the
   procedure are enough.
4. Supplied values: the two-sided p-value is 0.049 for the test of part 1
   and 0.15 for the test of part 3. Show that the p-values give the same
   decisions as your critical values.
5. State the Type I error for each test. State which error is possible
   after each of your two decisions.
6. The sample mean lies 2 euros above 40 euros. Explain why the test
   result for the mean is still only just significant.
7. Write a report of six sentences about the mean spending. Write one
   sentence on each of these topics: the estimate, the uncertainty, the
   decision, the practical importance, the population, and causality.

::: {.callout-note collapse="true"}
## Check a model answer

Part 1, the test for the mean:

1. **Hypotheses.** $H_0:\mu=40$ against $H_1:\mu\ne40$, where $\mu$ is the
   mean spending of all customers, in euros. The significance level is
   $\alpha=0.05$.
2. **Test statistic.** $T=(\bar X-40)/(S/\sqrt n)$. If $H_0$ is true, $T$
   has the t distribution with $n-1=80$ degrees of freedom. The sample of
   81 customers is large.
3. **Critical value and rejection region.** The critical value is
   $t_{0.975,\,80}=1.990$. Reject $H_0$ if $t_{obs}\le-1.990$ or
   $t_{obs}\ge1.990$.
4. **Observed value.** The standard error is $9/\sqrt{81}=1$ euro. Then
   $t_{obs}=(42-40)/1=2.00$.
5. **Conclusion.** The observed value 2.00 is above 1.990, so it lies in
   the rejection region, but only just. Reject $H_0$. At the 5% level we
   can conclude that the mean spending differs from 40 euros, and the
   positive sign places it above 40 euros.

Part 2, the interval for the mean:

1. **Formula.** $\bar X\pm t_{1-\alpha/2,\,n-1}\,S/\sqrt n$. It may be used
   because $n=81$ is large.
2. **Given values.** $\bar x=42$ euros, $s=9$ euros, $n=81$, $df=80$. The
   confidence level is 95%, so the critical value is
   $t_{0.975,\,80}=1.990$.
3. **Calculations.** The standard error is 1 euro. The margin of error is
   $1.990\times1=1.99$ euros.
4. **Interval.** $(42-1.99,\ 42+1.99)=(40.01,\ 43.99)$ euros.
5. **Conclusion.** The value 40 euros lies just outside the interval. This
   fits the decision of part 1: at the 5% level the mean spending differs
   from 40 euros, and the interval lies above 40.

Part 3, the test for the proportion:

1. **Hypotheses.** $H_0:p=0.60$ against $H_1:p\ne0.60$, where $p$ is the
   proportion of all customers who renew. The significance level is
   $\alpha=0.05$.
2. **Test statistic.** $Z=(\widehat P-p_0)/\sqrt{p_0(1-p_0)/n}$. If $H_0$
   is true, $Z$ is approximately $N(0,1)$. The condition holds:
   $81\times0.60=48.6$ and $81\times0.40=32.4$ are both at least 10.
3. **Critical value and rejection region.** The critical value is
   $z_{0.975}=1.960$. Reject $H_0$ if $z_{obs}\le-1.960$ or
   $z_{obs}\ge1.960$.
4. **Observed value.** The standard deviation of $\widehat P$ under $H_0$
   is $\sqrt{0.60\times0.40/81}=0.0544$. The estimate is
   $\hat p=55/81=0.679$. Then $z_{obs}=(0.679-0.60)/0.0544=1.45$.
5. **Conclusion.** The observed value 1.45 lies between $-1.960$ and
   1.960, so it does not lie in the rejection region. Do not reject $H_0$.
   At the 5% level we cannot conclude that the renewal proportion differs
   from 60%.

The interval for the proportion uses the standard error:
$\sqrt{0.679\times0.321/81}=0.0519$. The margin of error is
$1.960\times0.0519=0.102$, so the 95% interval is $(0.577,\ 0.781)$. It
contains 0.60, in line with the decision of the test.

Part 4. The p-value 0.049 is below 0.05, so the test for the mean rejects
$H_0$. The p-value 0.15 is above 0.05, so the test for the proportion does
not reject $H_0$. These are the decisions of the critical values.

Part 5. A Type I error for the mean is to conclude that the mean spending
differs from 40 euros when $\mu=40$. A Type I error for the proportion is to
conclude that the renewal proportion differs from 60% when $p=0.60$. The
test for the mean rejected $H_0$, so only a Type I error is possible there.
The test for the proportion did not reject $H_0$, so only a Type II error
is possible there.

Part 6. The standard error is 1 euro, so the sample mean lies exactly two
standard errors above 40 euros. The critical value is 1.990. The decision
depends on the difference divided by its standard error, not on the size
of the difference alone.

Part 7. A model report: "The 81 sampled customers spend 42 euros on
average, 2 euros above the benchmark of 40 euros. The 95% interval for the
mean spending runs from 40.01 to 43.99 euros. The test rejects
$H_0:\mu=40$ at the 5% level, but only just: the observed value is 2.00
against the critical value 1.990. A difference of 2 euros is 5% of the
benchmark, so its importance depends on the margin per customer. The result
holds for the population from which we drew the customers at random. It
does not show what causes the higher spending."

Verification in R:

```{r}
#| label: bep04-timed-check
se_spending <- 9 / sqrt(81)
c(se = se_spending, t_obs = (42 - 40) / se_spending,
  critical = qt(0.975, df = 80), p_value = 2 * pt(-2, df = 80),
  lower = 42 - 1.990 * se_spending, upper = 42 + 1.990 * se_spending,
  difference_in_percent = 100 * (42 - 40) / 40)

p_hat_renew <- 55 / 81
sd_null_renew <- sqrt(0.60 * 0.40 / 81)
z_renew <- (p_hat_renew - 0.60) / sd_null_renew
se_renew <- sqrt(p_hat_renew * (1 - p_hat_renew) / 81)
c(successes_H0 = 81 * 0.60, failures_H0 = 81 * 0.40,
  p_hat = p_hat_renew, sd_under_H0 = sd_null_renew, z_obs = z_renew,
  p_value = 2 * pnorm(-abs(z_renew)))
c(standard_error = se_renew, margin = 1.960 * se_renew,
  lower = p_hat_renew - 1.960 * se_renew,
  upper = p_hat_renew + 1.960 * se_renew)
```
:::

### Check your understanding (5 minutes)

Complete from memory:

1. A 95% confidence interval is correctly described as ...
2. Reject $H_0:\mu=\mu_0$ at the level $\alpha$ exactly when the confidence
   interval with level $1-\alpha$ ...
3. A Type I error is ... Power is ...
4. "Do not reject $H_0$" is different from "accept $H_0$" because ...
5. The test for a proportion divides by ... The interval for a proportion
   uses ...
6. A regression coefficient row tests ...

::: {.callout-note collapse="true"}
## Check your answers

1. The result of a procedure that contains the fixed parameter in about
   95% of random samples. It is never "a 95% chance that the parameter is
   inside".
2. Does not contain $\mu_0$ (for a two-sided t test and a t interval from
   the same sample).
3. Rejecting $H_0$ when it is true. The probability of rejecting $H_0$
   when $H_1$ is true.
4. A sample may fail to reject $H_0$ only because it is too small or too
   variable. A real difference from the value of $H_0$ can still exist.
5. The standard deviation of $\widehat P$ under $H_0$,
   $\sqrt{p_0(1-p_0)/n}$. The standard error,
   $\sqrt{\hat p(1-\hat p)/n}$.
6. $H_0:\beta_j=0$ against $H_1:\beta_j\ne0$.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Two sentences about your own venture</span>

Think of your EiA venture, or any business you know well. Take the one
claim about it that you would most like to defend in front of an investor.
Then write two sentences. First write the claim as $H_0$ and $H_1$ about
one parameter, with a realistic sample. State which error would cost you
more: the false alarm (Type I) or the missed difference (Type II). Then say
which interval would make you change a real business decision, and not
just a slide in your pitch.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student venture that sells boxes of ready-made meals on campus. "I want to show that my average order value differs from the €9 break-even, so I test $H_0:\mu=9$ against $H_1:\mu\ne9$ on my next 150 orders. The missed difference would cost me more, because I would give up a business that can make a profit. An interval that lies entirely above €9 would make me buy a stock of packaging, not just update a slide."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant.
::::

### More practice

For extra exercises on the lectures that this tutorial covers, with worked solutions, ask **Tilly**,
the course AI assistant. For example: "Give me an extra exercise for
Tutorial 4. Show the solution only after I send my answer."

You have now completed all lectures and tutorials of QM Basic. Next, use
the formula sheet and attempt the [QM Basic Mock Exam](mock-exam.qmd). Continue to QM Intermediate when that course begins.

[Continue to QM Basic Formula Sheet](formula-sheet.qmd).

:::
