# Tutorial 4: Inference and tests against zero {#b-ep04}

```{r}
#| label: bep04-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 4, you can</span>

- construct and interpret intervals for a mean and a proportion
- complete two-sided tests of zero-valued contrasts
- connect a test decision with its confidence interval
- interpret the t distribution, degrees of freedom, errors, and power
- read a regression coefficient test without claiming more than it shows.
:::

Tutorials are **exam preparation sessions**. Every task below follows the
format of the final exam. You first try each task in pairs. You may use AI for this, following
Attempt, Ask, Verify, Explain. Then the solution is discussed together.

This tutorial covers Lectures 9--10: the t distribution and confidence
intervals from Lecture 9, and the p-value, test decisions, errors, and power
from Lecture 10. All tests are two-sided.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 9 and 10. Review
[Mathematics Refresher: intervals, degrees of freedom, and
standardisation](../resources/math-refresher.html) if needed.

Use the cumulative formula sheet through Lecture 10.

Write from memory:

1. the t confidence-interval form
2. the t statistic for testing a mean contrast against zero
3. why the critical value comes from the t distribution rather than the
   normal, and the meaning of $df=n-1$
4. the definition of a two-sided p-value
5. the 95%-interval/5%-test decision link.

::: {.callout-note collapse="true"}
## Check your preparation

1. $\widehat\delta\pm t_{0.975,\,n-1}\widehat{\operatorname{SE}}(\widehat\delta)$.
2. $T=\widehat\delta/\widehat{\operatorname{SE}}(\widehat\delta)$.
3. The population SD is replaced by the sample SD $S$, which itself varies
   across samples. The heavier tails of the t distribution account for that
   extra uncertainty. Once the sample mean is estimated, only $n-1$
   deviations vary freely when spread is estimated.
4. Under $H_0$, the probability of a statistic at least as far from zero
   as the observed statistic in either direction.
5. For a two-sided t test and the t interval with the same SE, reject
   $H_0$ at 5% exactly when the 95% interval excludes zero.
:::

:::

::: {.course-phase .phase-in-class}

## In the tutorial

Try every task before you open its answer. Every conclusion must name the
parameter, the units, the population, the direction, and the limits of the
evidence.

Tasks 1, 2, and 5 use the random sample of 100 pitches (seed 123) and the
equity gap, built as in Lectures 7 and 9:

```{r}
#| label: bep04-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
pitch_sample$equity_gap_pp <- pitch_sample$equity_offered_pct - 15
```

### Task 1: Mean contrast

For the sample of 100 pitches, define the gap between the equity offered
(`equity_offered_pct`) and the benchmark of 15 percentage points:

$$
G=\text{equity offered}-15.
$$

1. Calculate $\bar g$, $s_g$, the SE, and the degrees of freedom.
2. Construct a 95% interval for $\delta=E[G]$.
3. Test $H_0:\delta=0$ against $H_1:\delta\ne0$.
4. Explain why the interval and test agree.
5. Explain why the critical value comes from `qt(0.975, df = 99)` rather
   than `qnorm(0.975)`.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: bep04-mean-gap
g <- pitch_sample$equity_gap_pp
n <- length(g)
g_bar <- mean(g)
s_g <- sd(g)
se_g <- s_g / sqrt(n)
critical <- qt(0.975, df = n - 1)
ci_g <- g_bar + c(-1, 1) * critical * se_g
t_value <- g_bar / se_g
p_value <- 2 * pt(-abs(t_value), df = n - 1)

c(estimate = g_bar, sd = s_g, se = se_g, df = n - 1,
  lower = ci_g[1], upper = ci_g[2],
  t = t_value, p_value = p_value)
```

$\bar g=-2.24$, $s_g=7.7079$, $\widehat{\operatorname{SE}}=0.7708$, and
$df=99$, all in percentage points except $df$. The 95% interval is
approximately $(-3.77,\,-0.71)$. The statistic is $t=-2.906$ and the
two-sided p-value is about 0.0045, so reject the null hypothesis of a zero
gap. The mean equity offered in the recorded-pitch frame differs from 15
percentage points, and the negative estimate places it below. The interval
excludes zero because it uses the same t distribution and the same SE as the
test. The critical value is a t quantile because $\sigma$ was estimated by
$s_g$. With $df=99$ the value 1.984 is only slightly above the normal 1.96,
but the t value is the correct choice.
:::

### Task 2: Proportion contrast

Define $\delta_p=p-0.50$, where $p$ is the proportion of pitches that end
with a deal (`deal_on_show`) among all 1,441 recorded pitches. Use the same
sample of 100 pitches.

1. Estimate $\delta_p$.
2. Check the success-failure rule and test $H_0:\delta_p=0$ against
   $H_1:\delta_p\ne0$ with the large-sample z test.
3. Construct the large-sample 95% interval for $p$ and translate it into an
   interval for $\delta_p$.
4. State why the SE of the test and the SE of the interval differ.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: bep04-proportion-gap
n <- nrow(pitch_sample)
phat <- mean(pitch_sample$deal_on_show)
p0 <- 0.50
delta_hat <- phat - p0

null_se <- sqrt(p0 * (1 - p0) / n)
z_value <- delta_hat / null_se
p_value <- 2 * pnorm(-abs(z_value))

estimated_se <- sqrt(phat * (1 - phat) / n)
ci_p <- phat + c(-1, 1) * 1.96 * estimated_se
ci_delta <- ci_p - p0

c(phat = phat, estimated_gap = delta_hat,
  z = z_value, p_value = p_value)
c(proportion_lower = ci_p[1], proportion_upper = ci_p[2],
  gap_lower = ci_delta[1], gap_upper = ci_delta[2])
```

$\hat p=0.64$, so $\widehat\delta_p=0.14$. The success-failure rule holds
under the null: $np_0=n(1-p_0)=50\ge10$. Under the null,
$\operatorname{SE}_0=0.05$, $z=2.80$, and the two-sided p-value is about
0.005. Reject the null hypothesis of a zero gap. The deal proportion in the
recorded-pitch frame differs from 50%, and the positive estimate places it
above. The proportion interval is approximately $(0.546,\,0.734)$, which
corresponds to a gap interval of $(0.046,\,0.234)$. The test calculates the
SE with $p_0$, because it asks how surprising the sample would be if $H_0$
were true. The interval does not assume $H_0$, so it calculates the SE with
$\hat p$.
:::

### Task 3: Read a regression coefficient row

Suppose a regression of the equity offered (in percentage points) on the
amount requested reports this row:

| Predictor | Estimate | SE | t | p-value | 95% interval |
|---|---:|---:|---:|---:|---:|
| Amount requested, per $100,000 | -0.290 | 0.058 | -5.00 | <0.001 | (-0.404, -0.176) |

1. State the tested hypotheses.
2. Explain every number in units.
3. Give the statistical conclusion at 5%.
4. State two claims the row cannot justify.

::: {.callout-note collapse="true"}
## Check your answer

Write $\beta_{ask}$ for the population coefficient of the amount requested.
The test is $H_0:\beta_{ask}=0$ against $H_1:\beta_{ask}\ne0$. The estimate
$\hat\beta_{ask}=-0.290$ says that another \$100,000 requested goes together
with 0.290 percentage points less equity offered. The SE, 0.058, measures
the sampling uncertainty of the estimate in the same units. The statistic is
$-0.290/0.058=-5.00$. The interval says that plausible values of the
coefficient run from $-0.404$ to $-0.176$ percentage points per \$100,000.
The p-value is below 0.001 and the 95% interval excludes zero, so reject the
null hypothesis of a zero coefficient at 5%. The row alone does not show
that asking for more money causes lower equity offers, and it does not show
that the conditions of the model hold.
:::

### Task 4: Errors and power

For Task 3:

1. describe a Type I error
2. describe a Type II error
3. state how a larger sample affects SE and power
4. explain why raising $\alpha$ is a trade-off rather than a free gain.

::: {.callout-note collapse="true"}
## Check your answer

A Type I error is concluding that the population coefficient of the amount
requested differs from zero when it equals zero. A Type II error is not
rejecting $H_0$ when the population coefficient is in fact not zero. A
larger independent sample usually reduces the SE and raises power. Raising
$\alpha$ also raises power, but it raises the probability of a Type I error
at the same time. A wider rejection region leads to more rejections, both
when $H_1$ is true and when $H_0$ is true.
:::

### Task 5 (optional): One-sided alternatives and the exam convention

This task belongs to the optional box on one-sided alternatives in
Lecture 10. It is not examined.

Task 1 gave $t=-2.906$ with $df=99$ and a two-sided p-value of about
0.0045.

1. Give the p-values for the one-sided alternatives $H_1:\delta<0$ and
   $H_1:\delta>0$, each fixed before the data were inspected.
2. A student first sees the negative estimate and then proposes to report
   the left-sided p-value "since the direction is now known". Explain the
   error.
3. State the convention of this course for tests and one reason for it.

::: {.callout-note collapse="true"}
## Check your answer

Left-sided: $P(T_{99}\le-2.906)\approx0.0023$, half of the two-sided
p-value. Right-sided: $P(T_{99}\ge-2.906)\approx0.9977$, one minus that
half. If the tail is chosen after seeing the estimate, the 5% always lies on
the side where the estimate happens to fall. The true Type I error rate is
then about 10%, not 5%. A one-sided alternative must follow from the
research question and be fixed before the sample is inspected. The
convention of this course: all tests are two-sided. In business questions a
difference in either direction usually matters, and the convention leaves
no room for choosing the tail afterwards.
:::

### Task 6: Repair the conclusion

The following conclusion about the row of Task 3 contains two errors. Find
them and write a better conclusion.

> The coefficient is significant, so asking for more money causes founders
> to offer less equity, and there is less than a 0.1% chance that the null
> is true.

::: {.callout-note collapse="true"}
## Check your answer

The p-value is not the probability that the null hypothesis is true, and a
significant result does not show a cause. A better conclusion: "Another
\$100,000 requested goes together with about 0.29 percentage points less
equity offered (95% interval from 0.18 to 0.40 points less). If the
population coefficient were zero, an estimate this far from zero would occur
in fewer than 0.1% of samples. The pitches were not assigned to amounts at
random, so this is an association, not a causal effect."
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice: 25 minutes

Solve this task alone and with a timer, under exam conditions: no notes, no
AI, formula sheet only. The example is invented. A random sample of 81
customers of a subscription service gives a mean spending of 42 euros with
sample SD 9 euros. Of the 81 customers, 54 renew their subscription.

1. Let $\mu$ be the population mean spending and define
   $\delta_\mu=\mu-40$. Construct a 95% interval for $\delta_\mu$ (use
   $t_{0.975,80}\approx1.99$) and test $H_0:\delta_\mu=0$ against
   $H_1:\delta_\mu\ne0$ at 5%. Supplied value: the two-sided p-value for
   $t=2.00$ with $df=80$ is 0.049.
2. Let $p$ be the population renewal proportion and define
   $\delta_p=p-0.60$. Construct the large-sample 95% interval for $p$ and
   test $H_0:\delta_p=0$ against $H_1:\delta_p\ne0$ at 5%. Supplied value:
   the two-sided p-value for $z=1.22$ is 0.22.
3. State the Type I error for each test.
4. Explain why the mean result is only just significant although the
   estimated mean difference is 2 euros.
5. Write a report of six sentences. Give one sentence each to the estimate,
   the uncertainty, the decision, the practical importance, the population,
   and causality.

::: {.callout-note collapse="true"}
## Check a model answer

For the mean, $\widehat\delta_\mu=2$ euros,
$\widehat{\operatorname{SE}}=9/\sqrt{81}=1$ euro, and $df=80$. With
$t_{0.975,80}\approx1.99$, the gap interval is approximately
$(0.01,\,3.99)$ euros, which is the same as a mean interval of
$(40.01,\,43.99)$ euros. The statistic is $t=2.00$ and the two-sided
p-value is about 0.049, so reject the null hypothesis of a zero gap, but
only just.

For renewal, $\hat p=54/81=0.667$. The estimated gap is 0.067. The
large-sample interval for the proportion is approximately
$(0.564,\,0.769)$, so the gap interval is $(-0.036,\,0.169)$. The
success-failure rule holds under $H_0:p=0.60$, because $81(0.60)=48.6$ and
$81(0.40)=32.4$ are both at least 10. Then
$\operatorname{SE}_0\approx0.0544$ and $z\approx1.22$. The two-sided
p-value, about 0.22, is above 0.05. Do not reject $H_0$.

A Type I error would be to claim a non-zero gap in mean spending when
$\mu=40$, or a non-zero gap in the renewal proportion when $p=0.60$. The
mean result is only just significant because the estimate of 2 euros lies
almost exactly two SEs from zero. The p-value depends on the estimate
relative to its SE, not on the size of the estimate alone. A complete
report must also say whether a difference of 2 euros (5% of the benchmark of
40 euros) matters, and whether the sampling design supports a claim about
the intended population. Neither test shows a cause.

A model report for the mean: "The 81 sampled customers spend 42 euros on
average, 2 euros above the benchmark of 40. The 95% interval for the mean
runs from 40.01 to 43.99 euros. The test rejects a zero gap at 5%, but only
just (p-value 0.049). A gap of 2 euros is 5% of the benchmark, so its
importance depends on the margin per customer. The result holds for the
population from which the customers were drawn at random. It does not show
what causes the higher spending."
:::

### Check your understanding

Complete from memory:

1. A 95% confidence interval is correctly described as ...
2. Reject $H_0$ at 5% exactly when the matched 95% interval ...
3. A Type I error is .... Power is ...
4. "Do not reject $H_0$" is different from "accept $H_0$" because ...
5. A regression coefficient row tests ...

::: {.callout-note collapse="true"}
## Check your answers

1. The result of a procedure that captures the fixed parameter in about
   95% of random samples. It is never "a 95% chance that the parameter is
   inside".
2. Excludes zero (for a two-sided test and an interval that use the same
   SE).
3. Rejecting $H_0$ when it is true. The probability of rejecting $H_0$
   when $H_1$ is true.
4. A result that is not significant may only be too imprecise to tell a
   real non-zero parameter from zero.
5. $H_0:\beta_j=0$ against $H_1:\beta_j\ne0$.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture, or any business you know well. Take the one
claim about it that you would most like to defend in front of an investor,
and write two sentences: first phrase it as a zero-contrast test with a
realistic sample and state which error, the false alarm or the missed
effect, would cost you more, then say what interval outcome would make
you change a real decision rather than just a slide.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "The claim I must defend: average order value equals the €9 break-even, tested two-sided on my next 150 orders. The missed effect would cost me more, walking away from a viable business, so an interval sitting entirely above €9 would make me order packaging stock, not just update a slide."
:::

You can discuss this task, and any other task in this tutorial, with
**Tilly**, the course AI assistant.
::::

QM Basic is complete. Next, use the [QM Basic Formula Sheet](formula-sheet.qmd) and attempt the
[QM Basic Mock Exam](mock-exam.qmd). Continue to QM Intermediate when that course begins.

:::
