# Tutorial 4: Inference and tests against zero {#b-ep04}

```{r}
#| label: bep04-setup
#| include: false
# Chunk options only. Every object used in this chapter is constructed in
# visible code on this page, so that copying the code always works.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2)
```

::: {.outcomes}
<span class="label">By the end of Tutorial 4, you can</span>

- construct and interpret intervals for a mean and a proportion;
- complete two-sided tests of zero-valued contrasts;
- connect a test decision with its confidence interval;
- interpret the t distribution, degrees of freedom, errors, and power;
- explain when a one-sided alternative is legitimate and why the exam is
  two-sided;
- read a regression coefficient test without overclaiming.
:::

Tutorials are **exam preparation sessions**: every task below follows the
format of the final exam. The standing routine for each task is the same ---
the TA states the task, pairs work on it first for a few minutes with AI use
allowed, and the solution is then discussed in plenary, starting from the
pair attempts.

This tutorial covers Lectures 9--10: the t distribution and intervals from
Lecture 9, testing decisions and one-sided alternatives from Lecture 10.
Exam tests are two-sided.

::: {.course-phase .phase-before}

## Before class

Bring your after-class work from Lectures 9 and 10. Review
[Mathematics Refresher: intervals, degrees of freedom, and
standardisation](../resources/math-refresher.html) if needed.

Use the cumulative formula sheet through Lecture 10.

Write from memory:

1. the t confidence-interval form;
2. the t statistic for testing a mean contrast against zero;
3. why the critical value comes from the t distribution rather than the
   normal, and the meaning of $df=n-1$;
4. the definition of a two-sided p-value;
5. the 95%-interval/5%-test decision link.

::: {.callout-note collapse="true"}
## Check your preparation

1. $\widehat\delta\pm t_{0.975,\,n-1}\operatorname{SE}(\widehat\delta)$.
2. $T=\widehat\delta/\operatorname{SE}(\widehat\delta)$.
3. The population SD is replaced by the sample SD $S$, which itself varies
   across samples; the heavier-tailed t distribution prices in that extra
   uncertainty. Estimating the sample mean leaves $n-1$ freely varying
   deviations for estimating spread.
4. Under $H_0$, the probability of a statistic at least as far from zero
   as the observed statistic in either direction.
5. For matched two-sided t procedures, reject zero exactly when the 95%
   interval excludes zero.
:::

:::

::: {.course-phase .phase-in-class}

## In class

Attempt every task before opening its answer. Every conclusion must name
the parameter, units, population, direction, and evidence limit.

All tasks use the reproducible seed-123 sample of 100 and the equity gap,
built exactly as in Lectures 7 and 9:

```{r}
#| label: bep04-load
sharks <- read.csv("data/shark_tank_teaching.csv", stringsAsFactors = FALSE)
set.seed(123)
sample_rows <- sample(seq_len(nrow(sharks)), size = 100, replace = FALSE)
pitch_sample <- sharks[sample_rows, ]
pitch_sample$equity_gap_pp <- pitch_sample$equity_offered_pct - 15
```

### Task 1 --- Mean contrast

For the reproducible sample, define

$$
D=\text{initial equity offered}-15.
$$

1. Calculate $\bar d$, $s_d$, the SE, and the degrees of freedom.
2. Construct a 95% interval for $\delta=E[D]$.
3. Test $H_0:\delta=0$ against $H_1:\delta\ne0$.
4. Explain why the interval and test agree.
5. Explain why the critical value comes from `qt(0.975, df = 99)` rather
   than `qnorm(0.975)`.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: bep04-mean-gap
d <- pitch_sample$equity_gap_pp
n <- length(d)
d_bar <- mean(d)
s_d <- sd(d)
se_d <- s_d / sqrt(n)
critical <- qt(0.975, df = n - 1)
ci_d <- d_bar + c(-1, 1) * critical * se_d
t_value <- d_bar / se_d
p_value <- 2 * pt(-abs(t_value), df = n - 1)

c(estimate = d_bar, sd = s_d, se = se_d, df = n - 1,
  lower = ci_d[1], upper = ci_d[2],
  t = t_value, p_value = p_value)
```

$\bar d=-2.24$, $s_d=7.7079$, $SE=0.7708$, and $df=99$. The 95% interval
is approximately $(-3.77,\,-0.71)$. The statistic is $t=-2.906$ and the
two-sided p-value about 0.0045, so reject the zero-gap null. The interval
excludes zero because it uses the same t model and SE as the test. The
critical value is a t quantile because $\sigma$ was estimated by $s_d$;
with $df=99$ the value 1.984 is only slightly above the normal 1.96, but
the t choice is the correct procedure.
:::

### Task 2 --- Proportion contrast

Define $\delta_p=p-0.50$, where $p$ is the finite-frame deal proportion.

1. Estimate $\delta_p$.
2. Perform a two-sided large-sample z-test against zero.
3. Construct the course's large-sample 95% interval for $p$ and translate
   it into an interval for $\delta_p$.
4. State why the test and interval SEs differ.

::: {.callout-note collapse="true"}
## Check your answer

```{r}
#| label: bep04-proportion-gap
n <- nrow(pitch_sample)
phat <- mean(pitch_sample$deal_on_show)
p0 <- 0.50
delta_hat <- phat - p0

null_se <- sqrt(p0 * (1 - p0) / n)
z_value <- delta_hat / null_se
p_value <- 2 * pnorm(-abs(z_value))

estimated_se <- sqrt(phat * (1 - phat) / n)
ci_p <- phat + c(-1, 1) * 1.96 * estimated_se
ci_delta <- ci_p - p0

c(phat = phat, estimated_gap = delta_hat,
  z = z_value, p_value = p_value)
c(proportion_lower = ci_p[1], proportion_upper = ci_p[2],
  gap_lower = ci_delta[1], gap_upper = ci_delta[2])
```

$\hat p=0.64$, so $\widehat\delta_p=0.14$. Under the null, $SE_0=0.05$,
$z=2.80$, and the two-sided p-value is about 0.005. Reject the zero-gap
null. The proportion interval is approximately $(0.546,\,0.734)$,
corresponding to a gap interval $(0.046,\,0.234)$. The test calculates
variation under $p_0$ --- how surprising the sample is in the null's world
--- while the interval estimates it with $\hat p$, making no null
assumption.
:::

### Task 3 --- Read a regression coefficient row

Suppose a model of initial equity offered reports:

| Predictor | Estimate | SE | t | p-value | 95% interval |
|---|---:|---:|---:|---:|---:|
| Amount requested, per $100,000 | -0.290 | 0.058 | -5.00 | <0.001 | (-0.403, -0.177) |

1. State the tested hypotheses.
2. Explain every number in units.
3. Give the statistical conclusion at 5%.
4. State two claims the row cannot justify.

::: {.callout-note collapse="true"}
## Check your answer

The test is $H_0:\beta_{ask}=0$ against $H_1:\beta_{ask}\ne0$. The
estimate says that another \$100,000 requested is associated with 0.290
percentage points less fitted initial equity, holding the model's other
predictors fixed. The SE measures sampling uncertainty in the same
coefficient units. The statistic is $-0.290/0.058=-5.00$. The p-value is
below 0.001, and the 95% interval excludes zero, so reject the
zero-coefficient null. The row alone does not establish causality or show
that the model's assumptions hold.
:::

### Task 4 --- Errors and power

For Task 3:

1. describe a Type I error;
2. describe a Type II error;
3. state how a larger sample affects SE and power;
4. explain why raising $\alpha$ is a trade-off rather than a free gain.

::: {.callout-note collapse="true"}
## Check your answer

A Type I error is concluding that the population-model ask coefficient
differs from zero when it equals zero. A Type II error is failing to
reject zero when a specified non-zero coefficient is true. A larger
independent sample usually reduces the SE and raises power. Raising
$\alpha$ also raises power, but it increases the Type I error probability
by exactly the same mechanism: a wider rejection region catches more of
everything.
:::

### Task 5 --- One-sided alternatives and the exam convention

Task 1 gave $t=-2.906$ with $df=99$ and a two-sided p-value of about
0.0045.

1. Calculate the p-values for the pre-specified one-sided alternatives
   $H_1:\delta<0$ and $H_1:\delta>0$.
2. A student who saw the negative estimate first proposes to report the
   left-sided p-value "since the direction is now known." Explain the
   error.
3. State the course's exam convention and one reason behind it.

::: {.callout-note collapse="true"}
## Check your answer

Left-sided: $P(T_{99}\le-2.906)\approx0.0023$, the single left tail.
Right-sided: $P(T_{99}\ge-2.906)\approx0.9977$. Choosing the tail after
inspecting the estimate lets the 5% rejection mass follow the data to
whichever side the noise landed, which doubles the true Type I error rate
to about 10%; the alternative must be fixed by the research question
before the sample is examined. Exam convention: all tests are two-sided,
both because business questions usually make either direction relevant
and because the convention removes the temptation of tail-picking.
:::

### Task 6 --- Repair the conclusion

Critique:

> The coefficient is significant, so asking for more money causes founders
> to offer less equity, and there is less than a 0.1% chance that the null
> is true.

::: {.callout-note collapse="true"}
## Check your answer

The p-value is not the probability that the null is true, and significance
does not establish causality. A defensible conclusion reports a negative
adjusted association, its magnitude and interval, the evidence against a
zero coefficient under the model, and the observational design and
assumption limits.
:::

:::

::: {.course-phase .phase-after}

## After class

### Timed individual practice --- 25 minutes

Redo this unassisted and against the clock: no notes, no AI, formula sheet
only --- exam conditions. A random sample of 81 customers gives mean
spending 42 euros, sample SD 9 euros, and 54 renewals.

1. Define $\delta_\mu=\mu-40$, construct its 95% interval (use
   $t_{0.975,80}\approx1.99$), and test a two-sided zero null.
2. Define $\delta_p=p-0.60$, construct the course's 95% interval, and
   test a two-sided zero null.
3. State the Type I error for each test.
4. Explain why the mean result is only just significant while the
   estimated mean difference is still 2 euros.
5. Write a six-sentence report separating estimate, uncertainty, decision,
   practical importance, population, and causality.

::: {.callout-note collapse="true"}
## Check a model answer

For the mean, $\widehat\delta_\mu=2$, $SE=9/\sqrt{81}=1$, and $df=80$.
With $t_{0.975,80}\approx1.99$, the gap interval is approximately
$(0.01,\,3.99)$, equivalent to a mean interval $(40.01,\,43.99)$. The
statistic is $t=2.00$ and the two-sided p-value is about 0.049, so reject
the zero mean-gap null just narrowly.

For renewal, $\hat p=54/81=0.667$. The estimated gap is 0.067. The
large-sample proportion interval is approximately $(0.564,\,0.769)$, so
the gap interval is $(-0.036,\,0.169)$. Under $H_0:p=0.60$,
$SE_0\approx0.0544$, $z\approx1.22$, and the two-sided p-value is about
0.22; do not reject.

A Type I error would be claiming a non-zero mean spending gap when
$\mu=40$, or a non-zero renewal gap when $p=0.60$. The mean result is
borderline because the 2-euro estimate is almost exactly two SEs from
zero --- precision, not size, drives the p-value. A complete report must
still evaluate whether a 2-euro difference matters and whether the
sampling design supports the intended population claim. Neither test
identifies a causal explanation.
:::

### Check your understanding

Complete from memory:

1. A 95% confidence interval is correctly described as ...
2. Reject $H_0$ at 5% exactly when the matched 95% interval ...
3. A Type I error is ...; power is ...
4. A one-sided alternative is legitimate only when ...
5. A regression coefficient row tests ...

::: {.callout-note collapse="true"}
## Check your answers

1. The product of a procedure that captures the fixed parameter in about
   95% of repeated valid samples --- never "a 95% chance the parameter is
   inside."
2. Excludes zero (for matched two-sided procedures using the same SE).
3. Rejecting a true null; the probability of rejecting when a specified
   alternative is true.
4. It is fixed before the data are inspected and follows from the
   research question. (Exam tests are two-sided.)
5. $H_0:\beta_j=0$ against $H_1:\beta_j\ne0$, holding the model's other
   predictors fixed.
:::

### Make it yours (5 minutes)

:::: {.utility-prompt}
<span class="label">Utility: two sentences</span>

Think of your EiA venture --- or any business you know well. Take the one
claim about it that you would most like to defend in front of an investor,
and write two sentences: first phrase it as a zero-contrast test with a
realistic sample and state which error --- the false alarm or the missed
effect --- would cost you more, then say what interval outcome would make
you change a real decision rather than just a slide.

::: {.callout-note collapse="true"}
## See an example, then write your own

Take a student-run campus meal-prep box venture. "The claim I must defend: average order value equals the €9 break-even, tested two-sided on my next 150 orders. The missed effect would cost me more --- walking away from a viable business --- so an interval sitting entirely above €9 would make me order packaging stock, not just update a slide."
:::

You can discuss this task --- and any other task in this tutorial --- with
**Tilly**, the course AI assistant on Canvas.
::::

QM Basic is complete. Next, use the QM Basic Formula Sheet and attempt the
QM Basic Mock Exam. Continue to QM Intermediate when that course begins.

:::
